### 2020-01-03 doc/theorists_notebook/20191218gen_fcn_unreg.tex
Added notes on generating function for protein distributions

\documentclass[12pt]{article}%{amsart}
\usepackage[top = 1.0in, bottom = 1.0in, left = 1.0in, right = 1.0in]{geometry}
% See geometry.pdf to learn the layout options. There are lots.
%\geometry{letterpaper}               % ... or a4paper or a5paper or ...
%\geometry{landscape}             % Activate for for rotated page geometry
\usepackage[parfill]{parskip}    % Activate to begin paragraphs with an empty line rather than an indent
\usepackage{graphicx}
\usepackage{amssymb}
%\usepackage{epstopdf}
% \usepackage{caption}
\usepackage{amsmath}
% \usepackage{longtable}
% \usepackage{tabu}
% \usepackage{accents} %to get undertildes for vec & mat
\usepackage{color} % for colored text
% \usepackage[normalem]{ulem} % for editing: striking out text using 'sout' command

% \newcommand\mytablefigwidth{0.35\textwidth}

% allows to use .ai files directly w/o resaving as pdf
\DeclareGraphicsRule{.ai}{pdf}{.ai}{}

% Handy math macros!
\newcommand{\vect}[1]{\vec{#1}}
\newcommand{\matr}[1]{\mathbf{#1}}
\newcommand{\rate}[3]{{#1}_{#2}^{#3}}
\newcommand{\mmnote}[1]{\textcolor{cyan}{(MM:~#1)}}

% derivative macros. these are sneaky. the [{}] is an empty default 1st arg
% usage: provide 1 arg: \deriv{x} gives d/dx
% provide 2 args like so: \deriv[f]{x} gives df/dx
\newcommand{\deriv}[2][{}]{\frac{d #1}{d #2}}
\newcommand{\pderiv}[2][{}]{\frac{\partial #1}{\partial #2}}

%%%%% Referencing macros %%%%%
\newcommand{\fref}[1]{Figure~\ref{#1}}
\newcommand{\tref}[1]{Table~\ref{#1}}
\newcommand{\eref}[1]{Eq.~(\ref{#1})}
\newcommand{\erngref}[2]{Eq.~(\ref{#1}-\ref{#2})} % Equations
% \newcommand{\test}[1][]{%
% \ifthenelse{\isempty{#1}{}}{omitted}{#1}%
% }

%%%%%
\begin{document}

\title{Generating function solution for the bursty unregulated promoter}

\maketitle

\section{One-state promoter with bursts, mRNA only}

\subsection{From master equation to generating function}
Before tackling mRNA and protein together as in~\cite{Shahrezaei2008}, let's
just do mRNA (replicating Charlotte's notes; she says it's worked out
in~\cite{Paulsson2000}, but I don't see it, at least not in notation I can
comprehend). The master equation of interest is
\begin{align}
\deriv{t}p(m,t) = (m+1)\gamma p(m+1,t) - m\gamma p(m,t) - r p(m,t)
        + r \sum_{m^\prime=0}^m G_{m-m^\prime}(\theta) p(m^\prime,t),
\label{eq:1state_unreg_003}
\end{align}
where $G_{k}(\theta)$ is the geometric distribution defined as
\begin{align}
G_{k}(\theta) = \theta(1 - \theta)^k, \, k\in\{0,1,2,\dots\}.
\end{align}
With this convention, the mean burst size $\beta = (1-\theta)/\theta$.
$\gamma$ and $r$ are mRNA degradation rates and transcription burst
initiation rates, resp. The last term represents all ways the system could
end up with $m$ mRNAs, having started the burst with $m^\prime$. Define
$\lambda = r/\gamma$ and nondimensionalize time by $\gamma$, giving
\begin{align}
\deriv{t}p(m,t) = (m+1)p(m+1,t) - m p(m,t) - \lambda p(m,t)
        + \lambda \sum_{m^\prime=0}^m G_{m-m^\prime}(\theta) p(m^\prime,t),
\end{align}
The probability generating function is defined as
\begin{align}
F(z,t) = \sum_m z^m p(m,t).
\end{align}
Multiply both sides of the CME by $z^m$ and sum over all $m$ to get
\begin{align}
\pderiv[F]{t} = (1 - z) \pderiv[F]{z}
        + \left(\frac{\theta}{1-z(1-\theta)}-1\right)\lambda F.
\end{align}
(This is just like in Charlotte's notes; the tricky bit is to
recognize the need to reverse the double sum in the last term from
$\sum_{m=0}^\infty\sum_{m^\prime=0}^m$ to
$\sum_{m^\prime=0}^\infty\sum_{m=m^\prime}^\infty$.
The sum on $m$ is then an easy geometric series. The rest of the procedure
is just the standard reindexing tricks for sums over master equations.)
Changing variables to $\xi=1-\theta$ and simplifying gives
\begin{align}
\pderiv[F]{t} + (z - 1) \pderiv[F]{z} = \frac{(z-1)\xi}{1-z\xi}\lambda F.
\label{eq:1state_unreg_015}
\end{align}
\subsection{Stead-state}
At steady-state, the PDE reduces to the ODE
\begin{align}
\deriv[F]{z} = \frac{\xi}{1-z\xi}\lambda F,
\end{align}
which we can integrate as
\begin{align}
\int \frac{dF}{F} = \int \frac{\lambda\xi dz}{1-\xi z}.
\end{align}
The initial conditions for generating functions always confuse me,
especially since most authors play fast and loose and assume ``it's
trivial.'' The key fact: from the definition
$F(z,t) = \sum_m z^m p(m,t)$,
normalization requires that
$F(z=1^-,t) = \sum_m p(m,t) = 1$.\footnote{
Sometimes the generating function may be undefined \textit{at} $z=1$ but the
limit still holds. Also people tend to change variables from $z$ to other
things, so don't lose track of how this condition transforms.
}
Doing the integrals (and producing constant $c$) gives
\begin{align}
\ln F &= -\lambda \ln(1-\xi z) + c
\\
F &= \frac{c}{(1-\xi z)^\lambda}.
\end{align}
Only one choice for $c$ can satisfy initial conditions, producing
\begin{align}
F(z) = \left(\frac{1-\xi}{1-\xi z}\right)^\lambda
        = \left(\frac{\theta}{1 - z(1-\theta)}\right)^\lambda,
\end{align}
which is exactly the negative binomial's generating function, as expected.

\subsection{Time-dependent case}
Return to~\eref{eq:1state_unreg_015} and for convenience,
change variables to $v=z-1$, producing
\begin{align}
\frac{1}{v}\pderiv[F]{t} + \pderiv[F]{v} = \frac{\lambda\xi}{1-(1+v)\xi} F.
\end{align}
This can be solved with the method of characteristics. Parametrize the
characteristics by the new variable $s$. Initial conditions at $t=0$
correspond to $s_0$, $v_0$, and $F(v_0,t=0)$, where $s_0$ and $v_0$ are
constants to be found later. The characteristic equations are
\begin{align}
\deriv[t]{s} &= \frac{1}{v}
\\
\deriv[v]{s} &= 1
\\
\deriv[F]{s} &= \frac{\lambda\xi}{1-(1+v)\xi} F.
\end{align}
We could immediately write $v=s+c_v$ for some constant $c_v$ yet to be
found. Most authors drop the constant without justification. This works
because our set of ODEs only contains $ds$, not $s$ itself. So, in fact,
we need not even solve for $v$ in terms of $s$, since the second ODE
tells us $ds=dv$, we can immediately remove $s$ from the problem and we
are left with the ODEs
\begin{align}
\deriv[t]{v} &= \frac{1}{v}
\\
\deriv[F]{v} &= \frac{\lambda\xi}{1-(1+v)\xi} F.
\end{align}
The first is trivial. Using our initial conditions it gives
\begin{align}
v = v_0 e^t.
\end{align}
Set this aside and tackle the second ODE.
It still permits separation of variables which gives
\begin{align}
\int \frac{dF}{F} &= \int \frac{\lambda \xi dv}{1 - (1+v)\xi}
\\
\ln F &= -\lambda \ln(1-(1+v)\xi) + \text{constant}.
\end{align}
Initial conditions determine the constant, but this is subtle. It may be
tempting to transform back to $z$ and $t$, for which initial conditions
are more obvious, but $v$ is a more convenient variable since it is
parametrizes the characteristics. In other words, we can view $F$ as a
one variable function $F(v)$ or as a two variable function $F(z,t)$. It
is much easier to deal with before changing variables back, rather than
after. But, in terms of $v$, what are our initial conditions? We know
$t=0$ corresponds to some $v_0$ we have yet to determine, and this
corresponds to some $F(v_0)$ yet to be found. Any parametrization of the
constant that enforces this is acceptable. A clever idea is to break up
and reparametrize the constant as $c_1$ and $c_2$ and bury these in the
logs as
\begin{align}
\ln \frac{F(v)}{c_1} &= -\lambda \ln\frac{1-(1+v)\xi}{c_2},
\end{align}
where we are now explicitly showing the $v$ dependence of $F$ for clarity.
For this equality to remain true when $v=v_0$,
the only choice of $c_1$ and $c_2$ that works is
\begin{align}
\ln \frac{F(v)}{F(v_0)} &= -\lambda \ln\frac{1-(1+v)\xi}{1-(1+v_0)\xi}
\end{align}
which leads immediately to
\begin{align}
F(v) &= F(v_0) \left(\frac{1-(1+v_0)\xi}{1-(1+v)\xi}\right)^{\lambda}
\end{align}
Now we may transform variables back to $z$ and $t$.
Recall from earlier that $1+v=z$ and $v=v_0 e^t$, so $v_0 = (z-1)e^{-t}$.
These substitutions handle the term in parentheses. $F(v_0)$ is more subtle.
We must avoid the trap of thinking there exists some $z_0$ corresponding to $v_0$;
$v_0$ corresponds only to $t=0$. This is the magic of characteristics.
$F(v_0)$ corresponds to the initial condition on $F$, i.e., $F(z,t=0)$
\textit{not} $F(z_0,t=0)$ since $z_0$ does not exist.
But $F(z,t=0)$, and therefore $F(v_0)$, is by definition given by
\begin{align}
F(v_0) = F(z,t=0) = \sum_m z^m p(m, t=0),
\end{align}
where $p(m, t=0)$ is whatever initial probability distribution
we specify for the problem. Finally then
\begin{align}
F(z, t) &=  \left(\frac{1-(1+(z-1)e^{-t})\xi}{1-z\xi}\right)^{\lambda}
                \left(\sum_m z^m p(m, t=0)\right)
\\
F(z, t) &=  \left(\frac{1-(1+(z-1)e^{-t})(1-\theta)}{1-z(1-\theta)}\right)^{\lambda}
                \left(\sum_m z^m p(m, t=0)\right).
\end{align}
If the initial condition is $p(0,0)=1$, then the second term in
parentheses disappears and this clearly reduces to the usual negative
binomial generating function as $t\rightarrow\infty$. For any other
initial condition, this has to remain true but it's not obvious to me
how the math will work out, even for the next-simplest initial condition
like $p(m,0) = \delta_{mk}$.
\mmnote{Still need to work out some algebra details here to see what happens.}

\section{Adding translation}
If we include transcription and translation, the master equation becomes
very similar to Eq.~(1) in~\cite{Shahrezaei2008}, except with mRNA
production terms borrowed from our~\eref{eq:1state_unreg_003}.
As in~\cite{Shahrezaei2008}, it is more convenient to nondimensionalize
time by the protein degradation rate $\gamma_p$ rather than the mRNA
degradation rate $\gamma_m$.
Then $\gamma\equiv\gamma_m/\gamma_p$ is the dimensionless mRNA lifetime,
and $\lambda\equiv r_m/\gamma_p$ is the dimensionless rate at which
transcription bursts initiate.
The mean burst size $\beta = (1-\theta)/\theta$ as before.
The new parameter we need is the translation rate $r_p$,
which we immediately nondimensionalize as $r\equiv r_m/\gamma_p$.
The authors of~\cite{Shahrezaei2008} have three dimensionless parameters
$a$, $b$, and $\gamma$, which have a simple correspondence with ours:
$a=\lambda$, $b=r/\gamma$, and $\gamma=\gamma$.

With all this, the master equation takes the form
\begin{align}
\deriv{t}p(m,n,t) = 
        & \lambda \left[\sum_{m^\prime=0}^m G_{m-m^\prime}(\theta) p(m^\prime,n,t)
        - p(m,n,t) \right]
        \\
        & + r m [ p(m,n-1,t) - p(m,n,t) ]
        \\
        & + \gamma [ (m+1) p(m+1,n,t) - m p(m,n,t) ]
        \\
        & + (n+1) p(m,n+1,t) - n p(m,n,t).
\end{align}
The first line covers transcription, the second translation, the third
mRNA degradation, and the fourth protein degradation. Conceptually, this
is identical to Eq.~(1) in~\cite{Shahrezaei2008} except for the very
first term.\footnote{The equations also differ by nondimensionalization
and relabeling of parameters, but these are ``bookkeeping'' operations,
not physical/conceptual ones.}
The probability generating function is now defined as
\begin{align}
F(z,w,t) = \sum_{m,n=0}^\infty z^m w^n p(m,n,t),
\end{align}
and similarly to before, we transform the master equation by multiplying
both sides by $z^m w^n$ and summing over $m$ and $n$.
The first term is unaffected by summing over $n$ and proceeds exactly as
in the mRNA-only case. All other terms proceed exactly as
in~\cite{Shahrezaei2008}, so the PDE for the generating function is
\begin{align}
\pderiv[F]{t} = \left(\frac{\theta}{1-z(1-\theta)}-1\right)\lambda F
        + rz(w - 1) \pderiv[F]{z} + \gamma(1 - z) \pderiv[F]{z}
        + (1 - w) \pderiv[F]{w}.
\end{align}
For convenience, and analogously to the mRNA-only case, changing variables to
$u=z-1$, $v=w-1$, and $\xi=1-\theta$ and simplifying gives
\begin{align}
\pderiv[F]{t} = \frac{\lambda\xi u}{1-\xi(u+1)} F
        + \left[ rv(1+u) - \gamma u \right] \pderiv[F]{u}
        - v \pderiv[F]{v},
\end{align}
or, rearranging terms to emphasize the analogy with
Eq.~(2) in~\cite{Shahrezaei2008},
\begin{align}
\pderiv[F]{v} - \left[ r(1+u) - \gamma \frac{u}{v} \right] \pderiv[F]{u} 
        + \frac{1}{v}\pderiv[F]{t} = \frac{\lambda\xi}{1-\xi(u+1)} \frac{u}{v} F.
\end{align}
Again, the left-hand side is functionally identical to Eq.~(2)
in~\cite{Shahrezaei2008}, and only the right-hand side has new content.
Despite the more complicated right-hand side, this is still solvable
with the method of characteristics; parametrizing the characteristics by
$s$ leads to a set of four equations, namely
\begin{align}
\deriv[v]{s}=1
\\
\deriv[t]{s}=\frac{1}{v}
\\
\deriv[u]{s} = \gamma \frac{u}{v} - r(1+u)
\\
\deriv[F]{s} = \frac{\lambda\xi}{1-\xi(u+1)} \frac{u}{v} F.
\end{align}
The equations for $t$ and $v$ are identical to the mRNA-only case,
allowing us to immediately eliminate $s$ from the problem and replace it
with $v$. We also have
\begin{align}
v = v_0 e^t
\end{align}
as before, for some initial $v_0$ yet to be found corresponding to $t=0$.
The remaining two ODEs to be solved are
\begin{align}
\deriv[u]{v} = \gamma \frac{u}{v} - r(1+u)
\\
\deriv[F]{v} = \frac{\lambda\xi}{1-\xi(u+1)} \frac{u}{v} F.
\end{align}
Shahrezaei and Swain approximately solve the first for $\gamma \gg 1$.
\mmnote{Now I see why they defined $b$ as they did, so they could get
$\gamma$ alone on one side. That's a bit cleaner than the way I've
parametrized, but it shouldn't change the conclusion.} In fact, this
approximation should serve us even better than it did them: their yeast
genes of interest feature $\gamma\sim 2-6$, whereas for \textit{E. coli}
with protein ``degradation'' dominated by dilution we expect $\gamma\sim
5-15$. They find $u(v) \approx bv/(1-bv)$, which, translated to our
variables, reads
\begin{align}
u(v) \approx \frac{rv}{\gamma-rv}.
\end{align}
Plugging this into the ODE for $F$ gives
\begin{align}
\deriv[F]{v} \approx
        \frac{\lambda\xi}{1-\xi\frac{\gamma}{\gamma-rv}} \frac{r}{\gamma-rv} F
        = \frac{\lambda r \xi}{(1-\xi)\gamma - rv} F
        = \frac{\lambda r (1-\theta)}{\theta\gamma - rv} F,
\end{align}
where in the last step we changed variables back to $\theta=1-\xi$.
This is still solvable with separation of variables, as
\begin{align}
\int\frac{dF}{F} &\approx \int \frac{\lambda r (1-\theta) dv}{\theta\gamma - rv} F
\\
\ln \frac{F(v)}{F(v_0)} &\approx -\lambda (1-\theta)
                \ln\left(\frac{\theta\gamma - rv}{\theta\gamma - rv_0}\right),
\end{align}
where we handled initial conditions in close analogy to the mRNA-only case.
With some algebra, and recalling that $v=w-1$ and $v_0=(w-1)e^{-t}$, we have
\begin{align}
F(w,t) \approx \left(\frac{1-\frac{r}{\theta\gamma}(w-1)e^{-t}}
                {1-\frac{r}{\theta\gamma}(w-1)}\right)^{\lambda(1-\theta)}
                \left(\sum_n w^n p(n,t=0)\right).
\label{eq:1state_unreg_030}
\end{align}
Note that $z$, which corresponded to mRNA copy number, dropped out when
we substituted our approximate solution for $u$ as a function of $v$.
This gave us a generating function for the protein distribution alone,
with all stochasticity and burstiness of the mRNA (approximately)
incorporated. The second term is simply the initial condition set by our
initially specified probability distribution on protein.
The first term is exactly, up to relabeling parameters, Eq.~(7)
in~\cite{Shahrezaei2008}, which leads to a probability distribution that
resembles a negative binomial, but with time dependence, multiplied by a
hypergeometric funtion.

Is this result of use to us? As with Eq.~(8) in~\cite{Shahrezaei2008},
it is valid only for $\gamma \gg 1$ and $t\gg \gamma^{-1}$.
The former is easily satisfied for our system. The latter means the result
is only valid for times longer than the mRNA lifetime,
which is also satisfied in our problem.

But the time scale to approach steady state is set by the protein
lifetime itself, which, if dominated by dilution, means $F(w,t)$ will
never reach steady-state, not even to a crude approximation.
So~\eref{eq:1state_unreg_030} applies, but we \textit{cannot} take its
steady-state limit. For very special initial conditions, Shahrezaei and
Swain work out the full time-dependent probability distribution, Eq.~(8)
in~\cite{Shahrezaei2008}. But it is not clear if we can do this for the
binomially partitioned initial distribution that we would like to use.
\mmnote{Needs more thought.}

%%%%%%%%%%%%%%%%%%%%% APPENDICES %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\appendix

% \bibliographystyle{nature}
\bibliographystyle{abbrv}
\bibliography{../library}

\end{document}

### 2020-01-28 doc/theorists_notebook/20191218gen_fcn_unreg.tex
Extended constit FISH analysis with theory interpretation

@@
 \\
-F(z, t) &=  \left(\frac{1-(1+(z-1)e^{-t})(1-\theta)}{1-z(1-\theta)}\right)^{\lambda}
+F(z, t) &=  \left(\frac{1-(1-\theta)(1+(z-1)e^{-t})}
+                        {1-(1-\theta)z}\right)^{\lambda}
                 \left(\sum_m z^m p(m, t=0)\right).
@@
 \mmnote{Still need to work out some algebra details here to see what happens.}
+\mmnote{Our t-dependence is already different than theirs, even though
+the steady state is the same. For trivial $k=0$ IC, everything will work
+out similarly, I think. For anything else, I think the direct approach
+of differentiating will produce an impossible mess. But we might be able
+to use their idea of a ``propagator'' probability and do a convolution
+over that; in principle, getting to the same place, but that might be
+tidier??}
 

### 2020-04-14 tests/stan_tests.py
Stan testing of 2F1 I forgot to commit weeks ago

@@ -1,0 +1,38 @@
+from git import Repo #for directory convenience
+
+import numpy as np
+
+import cmdstanpy
+import arviz as az
+
+repo = Repo("./", search_parent_directories=True)
+# repo_rootdir holds the absolute path to the top-level of our repo                 
+repo_rootdir = repo.working_tree_dir
+
+sm_gaussF = cmdstanpy.CmdStanModel(
+    stan_file=f"{repo_rootdir}/code/stan/test_gaussF.stan",
+    compile=True,)
+
+# stan needs to know how many data points to generate,
+# so pick a representative promoter
+stan_data = dict(
+    a=22.35,
+    b=-17.3,
+    c=12.5,
+    z=-1.59,
+    )
+
+stan_output = sm_gaussF.sample(
+    data=stan_data,
+    fixed_param=True,
+    iter_sampling=1,
+)
+
+# Convert to ArviZ InferenceData object
+stan_output = az.from_cmdstanpy(
+    stan_output, 
+    posterior_predictive=["output"]
+)
+
+# uncomment for running live in ipython,
+stan_output.posterior_predictive.output 

### 2020-04-17 doc/gen_fcn_notes.tex
Add work on writeup of gen func solutions to mRNA dists

\documentclass[12pt]{article}%{amsart}
\usepackage[top = 1.0in, bottom = 1.0in, left = 1.0in, right = 1.0in]{geometry}
% See geometry.pdf to learn the layout options. There are lots.
%\geometry{letterpaper}               % ... or a4paper or a5paper or ...
%\geometry{landscape}             % Activate for for rotated page geometry
\usepackage[parfill]{parskip}    % Activate to begin paragraphs with an empty line rather than an indent
\usepackage{graphicx}
\usepackage{amssymb}
%\usepackage{epstopdf}
% \usepackage{caption}
\usepackage{amsmath}
% \usepackage{longtable}
% \usepackage{tabu}
% \usepackage{accents} %to get undertildes for vec & mat
\usepackage{color} % for colored text
% \usepackage[normalem]{ulem} % for editing: striking out text using 'sout' command

% \newcommand\mytablefigwidth{0.35\textwidth}

% allows to use .ai files directly w/o resaving as pdf
\DeclareGraphicsRule{.ai}{pdf}{.ai}{}

% Handy math macros!
\newcommand{\vect}[1]{\vec{#1}}
\newcommand{\matr}[1]{\mathbf{#1}}
\newcommand{\rate}[3]{{#1}_{#2}^{#3}}
\newcommand{\mmnote}[1]{\textcolor{cyan}{(MM:~#1)}}

% derivative macros. these are sneaky. the [{}] is an empty default 1st arg
% usage: provide 1 arg: \deriv{x} gives d/dx
% provide 2 args like so: \deriv[f]{x} gives df/dx
\newcommand{\deriv}[2][{}]{\frac{d #1}{d #2}}
\newcommand{\pderiv}[2][{}]{\frac{\partial #1}{\partial #2}}
\newcommand{\psecderiv}[2][{}]{\frac{\partial{^2} #1}{\partial #2{^2}}}

%%%%% Referencing macros %%%%%
\newcommand{\fref}[1]{Figure~\ref{#1}}
\newcommand{\tref}[1]{Table~\ref{#1}}
\newcommand{\eref}[1]{Eq.~(\ref{#1})}
\newcommand{\erngref}[2]{Eq.~(\ref{#1}-\ref{#2})} % Equations
% \newcommand{\test}[1][]{%
% \ifthenelse{\isempty{#1}{}}{omitted}{#1}%
% }

%%%%%
\begin{document}

\title{Generating function solution for bursty promoters}

\maketitle

\section{Constitutive promoter with bursts}

\subsection{From master equation to generating function}
Before tackling mRNA and repression together, let us
just consider mRNA \mmnote{replicating Charlotte's notes; she says it's worked out
in~\cite{Paulsson2000}, but I don't see it, at least not in notation I can
comprehend}. The master equation of interest is
\begin{align}
\deriv{t}p(m,t) = (m+1)\gamma p(m+1,t) - m\gamma p(m,t) - r p(m,t)
        + r \sum_{m^\prime=0}^m G_{m-m^\prime}(\theta) p(m^\prime,t),
\label{eq:1state_unreg_003}
\end{align}
where $G_{k}(\theta)$ is the geometric distribution defined as
\begin{align}
G_{k}(\theta) = \theta(1 - \theta)^k, \, k\in\{0,1,2,\dots\}.
\end{align}
With this convention, the mean burst size $b = (1-\theta)/\theta$.
$\gamma$ and $r$ are mRNA degradation rates and transcription burst
initiation rates, resp. The last term represents all ways the system could
end up with $m$ mRNAs, having started the burst with $m^\prime$. Define
$\lambda = r/\gamma$ and nondimensionalize time by $\gamma$, giving
\begin{align}
\deriv{t}p(m,t) = (m+1)p(m+1,t) - m p(m,t) - \lambda p(m,t)
        + \lambda \sum_{m^\prime=0}^m G_{m-m^\prime}(\theta) p(m^\prime,t),
\end{align}
The probability generating function is defined as
\begin{align}
F(z,t) = \sum_{m=0}^\infty z^m p(m,t).
\end{align}
Multiply both sides of the CME by $z^m$ and sum over all $m$ to get
\begin{align}
\pderiv[F]{t} = \sum_{m=0}^\infty z^m
\left(
(m+1)p(m+1,t) - m p(m,t) - \lambda p(m,t)
+ \lambda \sum_{m^\prime=0}^m G_{m-m^\prime}(\theta) p(m^\prime,t)
\right).
\end{align}
The first two terms can be rewritten as derivatives of the
generating function, i.e.,
\begin{align}
\sum_{m=0}^\infty z^m (m+1)p(m+1,t)
=
\pderiv{z} \sum_{m=0}^\infty z^{m+1} p(m+1,t)
=
\pderiv[F]{z}
\end{align}
and
\begin{align}
\sum_{m=0}^\infty z^m mp(m,t)
=
z\pderiv{z} \sum_{m=0}^\infty z^m p(m,t)
=
z\pderiv[F]{z}.
\end{align}
The final term is the most trouble. The trick is to reverse
the default order of the sums, as in
\begin{align}
&\lambda \sum_{m=0}^\infty \sum_{m^\prime=0}^m
        z^m \theta(1 - \theta)^{m-m^\prime} p(m^\prime,t)
\\=
&\lambda\theta \sum_{m^\prime=0}^\infty z^{m^\prime} p(m^\prime,t)
        \sum_{m=m^\prime}^\infty
        z^{m-m^\prime} (1 - \theta)^{m-m^\prime}
\\=
&\frac{\lambda\theta F}{1 - (1 - \theta)z}.
\end{align}
This works because after reordering the sums, the sum on $m$ is just
a geometric series. Its result is independent of $m^\prime$, so the sum on
$m^\prime$ reduces to the definition of the generating function.
Putting these results together, the PDE for the generation function is
\begin{align}
\pderiv[F]{t} = \pderiv[F]{z} - z\pderiv[F]{z} - \lambda F
        + \frac{\lambda\theta F}{1-z(1-\theta)}.
\end{align}
Changing variables to $\xi=1-\theta$ and simplifying gives
\begin{align}
\pderiv[F]{t} + (z - 1) \pderiv[F]{z} = \frac{(z-1)\xi}{1-z\xi}\lambda F.
\label{eq:1state_unreg_015}
\end{align}
\subsection{Stead-state}
At steady-state, the PDE reduces to the ODE
\begin{align}
\deriv[F]{z} = \frac{\xi}{1-z\xi}\lambda F,
\end{align}
which we can integrate as
\begin{align}
\int \frac{dF}{F} = \int \frac{\lambda\xi dz}{1-\xi z}.
\end{align}
The initial conditions for generating functions can be subtle and confusing.
The key fact follows from the definition
$F(z,t) = \sum_m z^m p(m,t)$,
normalization requires that
$F(z=1^-,t) = \sum_m p(m,t) = 1$.\footnote{
Sometimes the generating function may be undefined \textit{at} $z=1$ but the
limit still holds. Also many authors change variables from $z$ to other
things, so one must keep track of how the normalization condition transforms.
}
Doing the integrals (and producing constant $c$) gives
\begin{align}
\ln F &= -\lambda \ln(1-\xi z) + c
\\
F &= \frac{c}{(1-\xi z)^\lambda}.
\end{align}
Only one choice for $c$ can satisfy initial conditions, producing
\begin{align}
F(z) = \left(\frac{1-\xi}{1-\xi z}\right)^\lambda
        = \left(\frac{\theta}{1 - z(1-\theta)}\right)^\lambda,
\end{align}
which is exactly the negative binomial's generating function, as expected.

\section{Adding repression}
\subsection{Deriving the generating function for mRNA distribution}
Let us move from a one-state promoter to a two-state promoter, where one state
has repressor bound and the other produces transcriptional bursts as above.
Although now we have an equation for each promoter state, otherwise
the master equation reads similarly to the one-state case, except with
additional terms corresponding to transitions between promoter states, namely
\begin{align}
\deriv{t}p_R(m,t) =& k_R^+ p_A(m,t) - k_R^- p_R(m,t)
        + (m+1)\gamma p_R(m+1,t) - m\gamma p_R(m,t)
\\
\begin{split}
\deriv{t}p_A(m,t) =& - k_R^+ p_A(m,t) + k_R^- p_R(m,t)
        + (m+1)\gamma p_A(m+1,t) - m\gamma p_A(m,t) 
\\
&- k_i p_A(m,t) + k_i \sum_{m^\prime=0}^m \theta(1-\theta)^{m-m^\prime} p_A(m^\prime,t),
\end{split}
\end{align}
where $p_R(m,t)$ is the probability of the system having $m$ mRNA copies
and having repressor bound to the promoter at time $t$, and $p_A$ is an
analogous probability to find the promoter without repressor bound.
$k_R+$ and $k_R^-$ are, respectively, the rates at which repressors
bind and unbind to and from the promoter, and
$\gamma$ is the mRNA degradation rate.
$k_i$ is the rate at which bursts initiate, and as before,
the geometric distribution of burst sizes has mean $b=(1-\theta)/\theta$.\footnote{
$\theta$ can be viewed as a Bernoulli trial probability: after each transcript
in a burst is produced, $\theta$ is the probability that the burst
terminates and $(1-\theta)$ is the probability that the burst continues.
}

Interestingly, it turns out that this problem maps exactly onto the
three-stage promoter model considered by
Shahrezaei and Swain in~\cite{Shahrezaei2008}, with relabelings.
Their approximate solution for protein distributions amounts to
the same approximation we make here in regarding the duration
of mRNA synthesis bursts as instantaneous, so their solution
for protein distributions also solves our problem of mRNA distributions.
Let us examine the analogy more closely.
They consider a two-state promoter, as we do here, but they model
mRNA as being produced one at a time and degraded, with rates
$v_0$ and $d_0$.
Then they model translation as occuring with rate $v_1$,
and protein degradation with rate $d_1$.
Now consider the limit where $v_1, d_0\rightarrow\infty$ with
their ratio $v_1/d_0$ held constant.
$v_1/d_0$ resembles the average burst size of translation from a single
mRNA: these are the rates of two Poisson processes that compete over a
transcript, which matches the story of geometrically distributed burst sizes.
So after taking this limit, on timescales slow compared to $v_1$ and $d_0$,
it appears that transcription events fire at rate $v_0$ and produce a
geometrically distributed burst of translation of mean size $v_1/d_0$,
which intuitively matches the story we have told above for mRNA
with variables relabeled.

To verify this intuitively conjectured mapping between our problem and
the solution in~\cite{Shahrezaei2008}, we continue with a careful
solution for the mRNA distribution using probability generating functions,
following the ideas sketched in~\cite{Shahrezaei2008}.
It is natural to nondimensionalize rates in the problem by $\gamma$,
or equivalently, this amounts to measuring time in units of $\gamma^{-1}$.
We are also only interested in steady state, so we set the time
derivatives to zero, giving
\begin{align}
0 =& k_R^+ p_A(m) - k_R^- p_R(m) + (m+1) p_R(m+1) - m p_R(m)
\\
\begin{split}
0 =& - k_R^+ p_A(m) + k_R^- p_R(m) + (m+1) p_A(m+1) - m p_A(m) 
\\
&- k_i p_A(m) + k_i \sum_{m^\prime=0}^m \theta(1-\theta)^{m-m^\prime} p_A(m^\prime),
\end{split}
\end{align}

The probability generating function is defined as before in the
constitutive case, except now we must introduce a generating function
for each promoter state,
\begin{align}
f_A(z) = \sum_{m=0}^\infty z^m p_A(m),
\;
f_R(z) = \sum_{m=0}^\infty z^m p_R(m).
\end{align}
Our real objective is the generating function $f(z)$ that generates
the mRNA distribution $p(m)$, independent of what state the promoter is in.
But since $p(m) = p_A(m) + p_R(m)$, it follows too that $f(z) = f_A(z) + f_R(z)$.

So as before we multiply both equations by $z^m$ and sum over all $m$.
Each individual term transforms exactly as did an analogous term in
the constitutive case, so the coupled ODEs for the generating functions read
\begin{align}
0 =& k_R^+ f_A(z) - k_R^- f_R(z) + \pderiv{z} f_R(z) - z \pderiv{z} f_R(z)
\\
\begin{split}
0 =&  - k_R^+ f_A(z) + k_R^- f_R(z) + \pderiv{z} f_A(z) - z \pderiv{z} f_A(z)
\\
&- k_i f_A(z) + k_i \frac{\theta}{1-z(1-\theta)} f_A(z),
\end{split}
\end{align}
and after changing variables $\xi = 1 - \theta$ as before
and rearranging, we have
\begin{align}
0 &= k_R^+ f_A(z) - k_R^- f_R(z) + (1-z) \pderiv{z} f_R(z)
\\
0 &=  - k_R^+ f_A(z) + k_R^- f_R(z) + (1 - z) \pderiv{z} f_A(z)
+ k_i \frac{(z-1)\xi}{1-z\xi} f_A(z),
\end{align}
We can transform this problem from two coupled first-order ODEs to a
single second-order ODE by solving for $f_A$ in the first and plugging
into the second, giving
\begin{align}
\begin{split}
0 = (1&-z) \pderiv[f_R]{z}
+ \frac{1-z}{k_R^+}
        \left(k_R^- \pderiv[f_R]{z} + \pderiv[f_R]{z} +(z-1) \psecderiv[f_R]{z}\right)
\\
&+ \frac{k_i}{k_R^+} \frac{(z-1)\xi}{1-z\xi}
        \left(k_R^- f_R + (z-1) \pderiv[f_R]{z}\right),
\end{split}
\end{align}
where, to reduce notational clutter, we have dropped the explicit $z$
dependence of $f_A$ and $f_R$. Simplifying we have
\begin{align}
0 = \psecderiv[f_R]{z}
        - \left(\frac{k_i\xi}{1-z\xi}
                + \frac{1 + k_R^- + k_R^+}{1-z}
        \right)\pderiv[f_R]{z}
        + \frac{k_i k_R^- \xi}{(1-z\xi)(1-z)}f_R.
\end{align}
This can be recognized as the hypergeometric differential equation,
with singularities at $z=1$, $z=\xi^{-1}$, and $z=\infty$. The latter
can be verified by a change of variables from $z$ to $x=1/z$,
being careful with the chain rule, and noting that $z=\infty$ is a
singular point if and only if $x=1/z=0$ is a singular point.

The standard form of the hypergeometric differential equation has its
singularities at 0, 1, and $\infty$, so to take advantage of the
standard form solutions to this ODE, we first need to transform
variables to put it into a standard form.
However, this is subtle. While any such transformation should work in
principle, the solutions are expressed most simply in the neighborhood
of $z=0$, but the normalization condition that we need to enforce
corresponds to $z=1$. The easiest path, therefore, is to find a change
of variables that maps
1 to 0, $\infty$ to $\infty$, and $\xi^{-1}$ to 1.
This is most intuitively done in two steps.

First map the $z=1$ singularity to 0 by the change of variables $v=z-1$, giving
\begin{align}
0 = \psecderiv[f_R]{v}
        + \left(\frac{k_i\xi}{(1+v)\xi - 1}
                + \frac{1 + k_R^- + k_R^+}{v}
        \right)\pderiv[f_R]{v}
        + \frac{k_i k_R^- \xi}{((1+v)\xi - 1)v}f_R.
\end{align}
Now two singularities are at $v=0$ and $v=\infty$.
The third is determined by $(1+v)\xi -1 = 0$, or $v=\xi^{-1} - 1$.
We want another variable change that maps this third singularity to 1
(without moving 0 or infinity). Changing variables again to
$w=\frac{v}{\xi^{-1} - 1} = \frac{\xi}{1-\xi} v$ fits the bill.
In other words, the combined change of variables
\begin{align}
w = \frac{\xi}{1-\xi} (z-1)
\end{align}
maps $z = \{1, \xi^{-1}, \infty\}$ to $w =\{0, 1, \infty\}$ as desired.
Plugging in, being mindful of the chain rule and noting
$(1 + v)\xi - 1 = (1 - \xi)(w - 1)$ gives
\begin{align}
0 = \left(\frac{\xi}{1-\xi}\right)^2 \psecderiv[f_R]{w}
+ \left(
        \frac{\xi k_i}{(1-\xi)(w-1)} + \frac{\xi(1 + k_R^- + k_R^+)}{(1-\xi)w}
\right) \frac{\xi}{1-\xi} \pderiv[f_R]{w}
+ \frac{k_i k_R^- \xi^2}{(1-\xi)^2 w(w-1)}f_R.
\end{align}
This is close to the standard form of the hypergeometric differential
equation, and some cancellation and rearrangement gives
\begin{align}
0 = w(w-1)\psecderiv[f_R]{w}
+ \left(k_i w + (1 + k_R^- + k_R^+)(w-1)\right) \pderiv[f_R]{w}
+ k_i k_R^- f_R.
\end{align}
and a little more algebra produces
\begin{align}
0 = w(1-w)\psecderiv[f_R]{w}
+ \left(1 + k_R^- + k_R^+
        - (1 + k_i + k_R^- + k_R^+)w
\right) \pderiv[f_R]{w}
- k_i k_R^- f_R,
\end{align}
which is the standard form. From this we can read off the solution
in terms of hypergeometric functions ${_2F_1}$ from
any standard source, e.g.~\cite{Abramovitz}, and identify the
conventional parameters in terms of our model parameters.
We want the general solution in the neighborhood of $w=0$ ($z=1$),
which for a homogeneous linear second order ODE must be a sum of two
linearly independent solutions.
More precisely,
\begin{align}
f_R(w) = C_1 {_2F_1}(\alpha, \beta, \gamma; w)
+ C_2 w^{1-\gamma}{_2F_1}(1+\alpha-\gamma, 1+\beta-\gamma, 2-\gamma; w)
\end{align}
with parameters determined by
\begin{align}
\begin{split}
\alpha\beta &= k_i k_R^-
\\
1+\alpha+\beta &= 1+k_i+k_R^-+k_R^+
\\
\gamma &= 1 + k_R^- + k_R^+
\end{split}
\end{align}
and constants $C_1$ and $C_2$ to be set by boundary conditions.
Solving for $\alpha$ and $\beta$, we find\footnote{
Note that $\alpha$ and $\beta$ are interchangeable in the definition of
${_2F_1}$ and differ only in the sign preceeding the radical.
}
\begin{align}
\begin{split}
\alpha &= \frac{1}{2}
\left(k_i+k_R^-+k_R^+ + \sqrt{(k_i+k_R^-+k_R^+)^2 - 4k_i k_R^-}\right)
\\
\beta &= \frac{1}{2}
\left(k_i+k_R^-+k_R^+ - \sqrt{(k_i+k_R^-+k_R^+)^2 - 4k_i k_R^-}\right)
\\
\gamma &= 1 + k_R^- + k_R^+.
\end{split}
\end{align}
Since the normalization condition requires that $f_R$ be finite at $w=0$,
we can immediately set $C_2=0$ to discard the second solution.
This is because all the rate constants are strictly positive,
so $\gamma>1$ and therefore $w^{1-\gamma}$ blows up as $w\rightarrow0$.
Now that we have $f_R$, we would like to find the generating function
for the mRNA distribution, $f(z) = f_A(z) + f_R(z)$.
We can recover $f_A$ from our solution for $f_R$, namely
\begin{align}
f_A(z) = \frac{1}{k_R^+}\left(k_R^- f_R(z) + (z-1) \pderiv[f_R]{z}\right)
\end{align}
or
\begin{align}
f_A(w) = \frac{1}{k_R^+}\left(k_R^- f_R(w) + w \pderiv[f_R]{w}\right),
\end{align}
where in the second line we transformed our original relation between
$f_R$ and $f_A$ to our new, more convenient, variable $w$.
Plugging our solution for $f_R(w) = C_1{_2F_1}(\alpha, \beta, \gamma; w)$
into $f_A$, we will require the differentiation rule for ${_2F_1}$,
which tells us
\begin{align}
\pderiv[f_R]{w} = C_1\frac{\alpha\beta}{\gamma}
                {_2F_1}(\alpha+1, \beta+1, \gamma+1; w),
\end{align}
from which it follows that
\begin{align}
f_A(w) = \frac{C_1}{k_R^+}
\left(
k_R^- {_2F_1}(\alpha, \beta, \gamma; w)
+ w\frac{\alpha\beta}{\gamma} {_2F_1}(\alpha+1, \beta+1, \gamma+1; w)
\right)
\end{align}
and therefore
\begin{align}
f(w) = C_1\left(1 + \frac{k_R^-}{k_R^+}\right)
        {_2F_1}(\alpha, \beta, \gamma; w)
+ w \frac{C_1}{k_R^+} \frac{\alpha\beta}{\gamma}
        {_2F_1}(\alpha+1, \beta+1, \gamma+1; w).
\end{align}
To proceed, we need one of the (many) useful identities known for
hypergeometric functions, in particular
\begin{align}
w\frac{\alpha\beta}{\gamma} {_2F_1}(\alpha+1, \beta+1, \gamma+1; w)
=
(\gamma-1)\left(
{_2F_1}(\alpha, \beta, \gamma-1; w) - {_2F_1}(\alpha, \beta, \gamma; w)
\right).
\end{align}
Substituting this for the second term in $f(w)$, we find
\begin{align}
f(w) = \frac{C_1}{k_R^+}
\left[
        \left(k_R^+ + k_R^-\right)
        {_2F_1}(\alpha, \beta, \gamma; w)
+ (\gamma-1)\left(
        {_2F_1}(\alpha, \beta, \gamma-1; w) - {_2F_1}(\alpha, \beta, \gamma; w)
        \right)
\right],
\end{align}
and since $\gamma-1 = k_R^+ + k_R^-$, the first and third terms cancel,
leaving only
\begin{align}
f(w) = C_1\frac{k_R^+ + k_R^-}{k_R^+} {_2F_1}(\alpha, \beta, \gamma-1; w).
\end{align}
Now we enforce normalization, demanding $f(w=0) = f(z=1) = 1$.
${_2F_1}(\alpha, \beta, \gamma-1; 0) = 1$, so we must have
$C_1 = k_R^+ / (k_R^+ + k_R^-)$ and consequently
\begin{align}
f(w) =  {_2F_1}(\alpha, \beta, k_R^+ + k_R^-; w).
\end{align}
Recalling that the mean burst size $b = (1-\theta)/\theta = \xi/(1-\xi)$
and $w = \frac{\xi}{1-\xi} (z-1) = b (z-1)$,
we can transform back to the original variable $z$ to find the tidy result
\begin{align}
f(z) =  {_2F_1}(\alpha, \beta, k_R^+ + k_R^-; b(z-1)),
\end{align}
with $\alpha$ and $\beta$ given above by
\begin{align}
\begin{split}
\alpha &= \frac{1}{2}
\left(k_i+k_R^-+k_R^+ + \sqrt{(k_i+k_R^-+k_R^+)^2 - 4k_i k_R^-}\right)
\\
\beta &= \frac{1}{2}
\left(k_i+k_R^-+k_R^+ - \sqrt{(k_i+k_R^-+k_R^+)^2 - 4k_i k_R^-}\right).
\end{split}
\end{align}
Finally we are in sight of the original goal. We can generate the
steady-state probability distribution of interest by
differentiating the generating function,
\begin{align}
p(m) = m! \left.\frac{\partial^m}{\partial z^m} f(z) \right|_{z=0},
\end{align}
which follows easily from its definition. Some contemplation reveals
that repeated application of the derivative rule used above will produce
products of the form $\alpha(\alpha+1)(\alpha+2)\cdots(\alpha+m-1)$ in
the expression for $p(m)$ and similarly for $\beta$ and $\gamma$. These
resemble ratios of factorials, but since $\alpha$, $\beta$, and $\gamma$
are not necessarily integer, we should express the ratios using gamma
functions instead. More precisely, one finds
\begin{align}
p(m) = \frac{
        \Gamma(\alpha + m)\Gamma(\beta + m)\Gamma(k_R^+ + k_R^-)
        }
        {
        \Gamma(\alpha)\Gamma(\beta)\Gamma(k_R^+ + k_R^- + m)
        }
\frac{b^m}{m!}{_2F_1}(\alpha+m, \beta+m, k_R^++k_R^-+m; -b)
\label{eq:p_m_bursty+rep}
\end{align}
which is finally the probability distribution we sought to derive.

\subsection{Moments}
We can verify the correctness of~\eref{eq:p_m_bursty+rep}
by comparison with Gillespie simulation.
We can also easily compute moments of the distribution from
the generating function, and check these against raw moments computed
directly from the master equation.

The $j$-th raw moment of the distribution can be computed from the
generating function from the formula
\begin{align}
\langle m^j \rangle
= \left. \left(z \frac{\partial}{\partial z}\right)^j f(z)\right|_{z=1}.
\end{align}
Note that \textit{probabilities} are generated by differentiating
and evaluating at $z=0$, while \textit{moments} are generated by
differentiating and evaluating at $z=1$. This also aligns with the
normalization constraint, which can be thought of as the zeroth raw moment.

The mean mRNA is given by
\begin{align}
\langle m \rangle = \left.z \pderiv[f]{z} \right|_{z=1}
= \frac{\alpha\beta b}{k_R^+ + k_R^-}
= k_i b\frac{k_R^-}{k_R^+ + k_R^-},
\end{align}
which is intuitive and appealingly simple:
the mean mRNA is just the mean mRNA in the absense of repression
($k_i b$, the burst rate time mean burst size)
times the fraction of time the repressor is bound, $k_R^-/(k_R^+ + k_R^-)$.

To compute the variance, let us first compute
\begin{align}
\begin{split}
\langle m^2\rangle
&= \left .z\pderiv{z}\left(z\pderiv[f]{z}\right) \right|_{z=1}
= \left. z\left(\pderiv[f]{z} + z\psecderiv[f]{z}\right)\right|_{z=1}
\\
&= \frac{\alpha\beta b}{k_R^+ + k_R^-}
        + \frac{\alpha(\alpha+1)\beta(\beta+1) b^2}
                {(k_R^+ + k_R^-)(k_R^+ + k_R^- + 1)}
\\
&= \frac{k_i b k_R^-}{k_R^+ + k_R^-}
        + \frac{k_i k_R^- b^2 (1 + k_i + k_R^+ + k_R^- + k_i k_R^-)}
                {(k_R^+ + k_R^-)(k_R^+ + k_R^- + 1)},
\end{split}
\end{align}
from which the variance is
\begin{align}
\begin{split}
var(m) = \langle m^2\rangle - \langle m\rangle^2
&= \frac{k_i b k_R^-}{k_R^+ + k_R^-}
\left(
1 + \frac{b(1 + k_i + k_R^+ + k_R^- + k_i k_R^-)} {k_R^+ + k_R^- + 1}
- \frac{k_i b k_R^-}{k_R^+ + k_R^-}
\right)
\\
&= k_i b\frac{k_R^-}{k_R^+ + k_R^-}
\left(
1 + b + k_i b\frac{1 + k_R^-} {k_R^+ + k_R^- + 1}
- \frac{k_i b k_R^-}{k_R^+ + k_R^-}
\right)
\\
&= k_i b\frac{k_R^-}{k_R^+ + k_R^-}
\left(
1 + b + k_i b\frac{k_R^+}{(k_R^+ + k_R^-)(k_R^+ + k_R^- + 1)}
\right).
\end{split}
\end{align}
The Fano factor is the quantity in parentheses. This can be interpreted as
the variability arising from the negative binomial distribution of the
unregulated promoter $(1+b)$ plus additional noise arising from repression.

\mmnote{Still to writeup: moment derivation direct from CME to compare}

\subsection{Numerical considerations and recursion formulas}
We would like to carry out Bayesian parameter inference on FISH data
from~\cite{Jones2014}, using~\eref{eq:p_m_bursty+rep} as our
likelihood. This requires accurate (and preferably fast)
numerical evaluation of the hypergeometric function ${_2F_1}$,
which is a notoriously hard problem~\cite{Pearson2016, Gil2007},
and our particular needs here present an especial challenge as we show below.

The hypergeometric function is defined by its Taylor series as
\begin{align}
{_2F_1}(a,b,c;z) 
= \sum_{l=0}^\infty
\frac{\Gamma(a + l)\Gamma(b + l)\Gamma(c)}
        {\Gamma(a)\Gamma(b)\Gamma(c + l)}
\frac{z^l}{l!}
\end{align}
for $|z|<1$, and by analytic continuation elsewhere.
If $z\lesssim1/2$ and $\alpha$ and $\beta$ are not too large
(absolute value below 20 or 30),
then the series converges quickly and an accurate numerical representation is
easily computed by truncating the series after a reasonable number of terms.
Unfortunately, we need to evaluate ${_2F_1}$ over mRNA copy numbers fully out
to the tail of the distribution, which can easily reach 50, possibly 100.
From~\eref{eq:p_m_bursty+rep}, this means evaluating ${_2F_1}$
repeatedly for values of $a$, $b$, and $c$ spanning the full range
from $\mathcal{O}(1)$ to $\mathcal{O}(10^2)$,
even if $\alpha$, $\beta$, and $\gamma$
in~\eref{eq:p_m_bursty+rep} are small,
with the situation even worse if they are not small.
A naive numerical evaluation of the series definition will be
prone to overflow and, if any of $a,b,c<0$, then some successive terms in the
series have alternating signs which can lead to catastrophic cancellations.

The solution involves recursion relations originally explored by Gauss,
and studied extensively in~\cite{Pearson2016, Gil2007}.

%%%%%%%%%%%%%%%%%%%%% APPENDICES %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\appendix

% \bibliographystyle{nature}
\bibliographystyle{abbrv}
\bibliography{../library}

\end{document}

### 2020-04-17 code/exploratory/gillespie_moment_test.py

@@ -1,0 +1,96 @@
+# This script is to test my apparent error in derivation of PGF for
+# bursty transcription + repression. Conclusion: no, my PGF is fine,
+# I just screwed up somewhere in deriving moments from the PGF
+#%%
+import numpy as np
+import numba
+
+import matplotlib.pyplot as plt
+import bokeh
+import bokeh.io
+bokeh.io.output_notebook()
+
+import biocircuits
+import srep
+
+kRoff = 1.
+kRon  = 1.
+k_i   = 1.
+k_u   = 100.
+r     = 100.
+gamma = 1.
+
+@numba.njit
+def simple_propensity(propensities, population, t):
+    """Updates an array of propensities given a set of parameters
+    and an array of populations.
+    """
+    # Unpack population
+    rep, emp, act, m = population
+    
+    # Update propensities
+    propensities[0] = kRoff * rep # unbind repressor
+    propensities[1] = kRon * emp  # bind repressor
+    propensities[2] = k_i * emp   # initiate burst
+    propensities[3] = k_u * act   # terminate burst
+    propensities[4] = r * act     # make transcript
+    propensities[5] = gamma * m   # degrade transcript
+
+# Columns are rep bound, empty, active, and m count
+simple_update = np.array([[-1,1, 0, 0],  # unbind repressor
+                          [1,-1, 0, 0],  # bind repressor
+                          [0,-1, 1, 0],  # initiate burst
+                          [0, 1,-1, 0],  # terminate burst
+                          [0, 0, 0, 1],  # make transcript
+                          [0, 0, 0,-1]], # degrade transcript
+                         dtype=np.int)
+
+time_points = np.linspace(0,1000,100)
+samples = biocircuits.gillespie_ssa(
+    simple_propensity,
+    simple_update,
+    np.array([1,0,0,0]), # init pop
+    time_points,
+    # return_time_points=True,
+    size=500,
+    n_threads=2,
+    progress_bar=True,
+)
+
+# %%
+plots = bokeh.plotting.figure(plot_width=400,
+                              plot_height=400,
+                              x_axis_label='dimensionless time',
+                              y_axis_label='number of mRNAs')
+
+# Plot trajectories and mean
+for x in samples[::100,:,-1]:
+    plots.line(time_points, x, line_width=0.3, 
+                alpha=0.2, line_join='bevel')
[…]

### 2020-04-17 code/figures/fig3.py

@@ -150,10 +150,10 @@ ax[1,1].errorbar(koff_Oid[1], ham_kOid[0], xerr=err_Oid, yerr=ham_kOid[1],
                 fmt='og', capsize=2, label=r"Oid, $\gamma^{-1} = $5 min")
 
 ax[1,1].plot((0, 0.3),(0, 0.3), 'k--')
-ax[1,1].set_xlabel(r'$k_R^-$ min$^{-1}$ (this study)')
-ax[1,1].set_ylabel(r'$k_R^-$ min$^{-1}$ (Hammar et al 2014)')
+ax[1,1].set_xlabel(r'$k_R^-$ (min$^{-1})$ (this study)')
+ax[1,1].set_ylabel(r'$k_R^-$ (min$^{-1})$ (Hammar et al 2014)')
 ax[1,1].legend(fontsize='small')
 
-plt.savefig(f"{repo_rootdir}/figures/fig3/fig3_v3.pdf")
+plt.savefig(f"{repo_rootdir}/figures/fig3/fig3_base.pdf")
 
 # %%

### 2020-04-20 doc/gen_fcn_notes.tex
Added more to writeup on gen fcn + numerics for 2F1

@@
 \subsection{From master equation to generating function}
+
+\begin{figure}
+\includegraphics[width=0.9\textwidth]{../figures/fig2/fig2.pdf}
+\caption{
+\mmnote{This figure wouldn't appear in this section, instead
+just a reference to the cartoon so the reader knows
+what cartoon model we're doing all this math for.}
+}
+\end{figure}
+We would like to derive the full probability distribution of
+mRNA copy number per cell for a model combining simple repression
+with a promoter featuring bursty transcription initiation.
 Before tackling mRNA and repression together, let us
-just consider mRNA \mmnote{replicating Charlotte's notes; she says it's worked out
+just consider the bursty consititutive promoter,
+model 4 in Fig.~1A.
+\mmnote{replicating Charlotte's notes; she says it's worked out
 in~\cite{Paulsson2000}, but I don't see it, at least not in notation I can
-comprehend}. The master equation of interest is
+comprehend. Need to find a cite for this,
+or maybe just cite Charlotte, personal comm, since that's where I actually got it\dots}.
+The master equation of interest is
 \begin{align}
@@
 \subsection{Deriving the generating function for mRNA distribution}
+
+\begin{figure}
+\includegraphics[width=0.9\textwidth]{../figures/fig3/fig3.ai}
+\caption{
+\mmnote{This figure wouldn't appear in this section, instead
+just a reference to the cartoon so the reader knows
+what cartoon model we're doing all this math for.}
+}
+\end{figure}
 Let us move from a one-state promoter to a two-state promoter, where one state
 has repressor bound and the other produces transcriptional bursts as above.
+A schematic of this model is shown in Fig.\ 2A.
 Although now we have an equation for each promoter state, otherwise
@@
 in terms of hypergeometric functions ${_2F_1}$ from
-any standard source, e.g.~\cite{Abramovitz}, and identify the
+any standard source, e.g.~\cite{Abramowitz1964}, and identify the
 conventional parameters in terms of our model parameters.
@@
 \subsection{Numerical considerations and recursion formulas}
+\subsubsection{Generalities}
 We would like to carry out Bayesian parameter inference on FISH data
@@
 numerical evaluation of the hypergeometric function ${_2F_1}$,
-which is a notoriously hard problem~\cite{Pearson2016, Gil2007},
+which is a notoriously hard problem~\cite{Pearson2017, Gil2007},
 and our particular needs here present an especial challenge as we show below.
@@
 
+One solution is to evaluate ${_2F_1}$ using arbitrary precision arithmetic
+instead of floating point arithmetic,
+e.g., using the \texttt{mpmath} library in Python.
+This is accurate but incredibly slow computationally.
+To quantify how slow, we found that
+evaluating the likelihood defined by~\eref{eq:p_m_bursty+rep} $\sim50$ times
+(for a typical dataset of interest from~\cite{Jones2014},
+with $m$ values spanning 0 to $\sim50$)
+using arbitrary precision arithmetic is 100-1000 fold slower than
+evaluating a negative binomial likelihood for the corresponding
+constitutive promoter dataset.
+
+To claw back $\gtrsim30$ fold of that slowdown, we can exploit
+one of the many catalogued symmetries involving ${_2F_1}$.
 The solution involves recursion relations originally explored by Gauss,
-and studied extensively in~\cite{Pearson2016, Gil2007}.
+and studied extensively in~\cite{Pearson2017, Gil2007}.
+They are sometimes known as contiguous relations and relate the values
+of any set of 3 hypergeometric functions whose arguments differ by integers.
+To rephrase this symbolically, consider a set of hypergeometric functions
+indexed by an integer $n$,
+\begin{align}
+f_n = {_2F_1}(a+\epsilon_i n, b+\epsilon_j n, c+\epsilon_k n; z),
+\end{align}
+for a fixed choice of $\epsilon_i, \epsilon_j, \epsilon_k \in \{0,\pm 1\}$
+(at least one of $\epsilon_i, \epsilon_j, \epsilon_k$ must be nonzero,
+else the set of $f_n$ would contain only a single element).
+Then there exist known recurrence relations of the form
+\begin{align}
+A_n f_{n-1} + B_n f_{n} + C_n f_{n+1} = 0,
+\end{align}
+where $A_n, B_n$, and $C_n$ are some functions of $a,b,c$, and $z$.
+In other words, for fixed $\epsilon_i, \epsilon_j, \epsilon_k, a, b,$ and $c$,
+if we can merely evaluate ${_2F_1}$ twice, say for $n^\prime$ and $n^\prime-1$,
+then we can easily and rapidly generate values for arbitrary $n$.
+
+This provides a convenient solution for our problem: we need repeated
+evaluations of ${_2F_1}(a+m, b+m, c+m; z)$
+for fixed $a,b$, and $c$ and many integer values of $m$.
+They idea is that we can use arbitrary precision arithmetic to evaluate
+${_2F_1}$ for just two particular values of $m$ and then generate
+${_2F_1}$ for the other 50-100 values of $m$ using the recurrence
+relation.\footnote{
+There are even more sophisticated ways of utilizing the recurrence
+relations that might have netted another factor of 2 speed-up, and
+possibly as much as a factor of 10, but the method described here had
+already reduced the computation time to an acceptable
+$\mathcal{O}(\text{1 min})$, so these more sophisticated approaches did
+not seem worth the time to pursue.
+}
+
+However, there two further wrinkles.
+The first is that
+a naive application of the recurrence relation is numerically unstable.
+Roughly, this is because the three term recurrence relations,
+like second order ODEs, admit two linearly independent solutions.
+In a certain eigenbasis, one of these solutions dominates the other
+as $n\rightarrow\infty$, and as $n\rightarrow-\infty$,
+the dominance is reversed.
+If we fail to work in this eigenbasis, our solution of the recurrence relation
+will be a mixture of these solutions and rapidly accumulate numerical error.
+For our purposes, it suffices to know that the authors of~\cite{Gil2007}
+derived the numerically stable solutions (so-called \textit{minimal solutions})
+for several possible choices of $\epsilon_i, \epsilon_j, \epsilon_k$.
+Running the recurrence in the proper direction using a minimal solution
+is numerically robust and can be done entirely in floating point arithmetic, 
+so that we only need to evaluate ${_2F_1}$ with arbitrary precision arithmetic
+to generate the seed values for the recursion.
+
+The second wrinkle is a corollary to the first.
+The minimal solutions are only minimal for certain ranges of the argument $z$,
+and not all of the 26 possible recurrence relations
+have minimal solutions for all $z$.
+This can be solved by using one of the many transformation formulae for
+${_2F_1}$ to convert to a different recurrence relation that has
+a minimal solution over the required domain of $z$, although
+this can require some trial and error to find the right transformation,
+the right recurrence relation, and the right minimal solution.
+
+\subsubsection{Particulars}
+Let us now demonstrate these generalities for our problem of interest.
+In order to evaluate the probability distribution of our
+model,~\eref{eq:p_m_bursty+rep}, we need to evaluate hypergeometric functions
+of the form ${_2F_1}(\alpha+m, \beta+m, \gamma+m; -b)$
+for values of $m$ ranging from $0$ to $\mathcal{O}(100)$.
+The authors of~\cite{Gil2007} did not derive a recursion relation
+for precisely this case. We could follow their methods and do so ourselves,
+but it is much easier to convert to a case that they did consider.
+The strategy is to look through the minimal solutions tabulated
+in~\cite{Gil2007} and search for a transformation we could apply to
+${_2F_1}(\alpha+m, \beta+m, \gamma+m; -b)$ that would place the $m$'s
+(the variable being incremented by the recursion)
+in the same arguments of ${_2F_1}$ as the minimal solution.
+After some ``guess and check,'' we found that the transformation
+\begin{align}
+{_2F_1}(\alpha+m, \beta+m, \gamma+m; -b)
+=
+(1+b)^{-\alpha-m}
+        {_2F_1}\left(\alpha+m, \gamma-\beta, \gamma+m; \frac{b}{1+b}\right),
+\label{eq:rec_euler_pretransform}
+\end{align}
+produces a ${_2F_1}$ on the right hand side that closely resembles
+the minimal solutions $y_{3,m}$ and $y_{4,m}$ in Eq.~4.3 in~\cite{Gil2007}.
+Explicitly, these solutions are
+\begin{align}
+y_{3,m}
+&\propto
+{_2F_1}\left(-\alpha^\prime + \gamma^\prime - m,
+                -\beta^\prime + \gamma^\prime,
+                1-\alpha^\prime-\beta^\prime+\gamma^\prime-m;
+                1-z\right)
+\\
+y_{4,m}
+&\propto
+{_2F_1}\left(\alpha^\prime + m,
+                \beta^\prime,
+                1+\alpha^\prime+\beta^\prime-\gamma^\prime+m;
+                1-z\right),
+\label{eq:minimal_soln_sans_prefac}
+\end{align}
+where we have omitted prefactors which are unimportant for now.
+Which of these two we should use depends on what values $z$ takes on.
+Equating $1-z=b/(1+b)$ gives $z=1/(1+b)$, and since $b$ is strictly positive,
+$z$ is bounded between 0 and 1.
+From Eq.~4.5 in~\cite{Gil2007}, $y_{4,m}$ is the minimal solution
+for real $z$ satisfying $0<z<2$, so this is the only minimal solution we need.
+
+Now that we have our minimal solution,
+what recurrence relation does it satisfy?
+Confusingly, the recurrence relation of which $y_{4,m}$ is a solution
+increments different arguments of ${_2F_1}$ that does $y_{4,m}$:
+it increments the first only, rather than first and third.
+This recurrence relation can be looked up, e.g., Eq.~15.2.10
+in~\cite{Abramowitz1964}, which is
+\begin{align}
+(\gamma^\prime - (\alpha^\prime + m)) f_{m-1}
++
+(2(\alpha^\prime+m) - \gamma^\prime + (\beta^\prime - \alpha^\prime)z)f_m
++ \alpha^\prime(z-1) f_{m+1} = 0.
+\label{eq:chosen_rec_rel}
+\end{align}
+Now we must solve for the parameters appearing in the recurrence relation
+in terms of our parameters, namely by setting
+\begin{align}
+\begin{split}
+\alpha^\prime &= \alpha
+\\
+\beta^\prime &= \gamma - \beta
+\\
+1 + \alpha^\prime + \beta^\prime - \gamma^\prime &= \gamma
+\\
+1 - z &= \frac{b}{1+b}
+\end{split}
+\end{align}
+and solving to find
+\begin{align}
+\begin{split}
+\alpha^\prime &= \alpha
+\\
+\beta^\prime &= \gamma - \beta
+\\
+\gamma^\prime &= 1 + \alpha - \beta
+\\
+z &= \frac{1}{1+b}
+.
+\end{split}
+\end{align}
+Finally we have everything we need. The minimal solution
+\begin{align}
+y_{4,m}
+=
+\frac{\Gamma(1+\alpha^\prime-\gamma^\prime+m)}
+        {\Gamma(1+\alpha^\prime+\beta^\prime-\gamma^\prime+m)}
+\times
+{_2F_1}\left(\alpha^\prime + m,
+                \beta^\prime,
+                1+\alpha^\prime+\beta^\prime-\gamma^\prime+m;
+                1-z\right),
+\end{align}
+where we have now included the necessary prefactors,
+is a numerically stable solution of the recurrence
+relation~\eref{eq:chosen_rec_rel} if the recursion is run
+from large $m$ to small $m$.
+So the procedure is to compute the value of ${_2F_1}$ for the two
+largest $m$ values of interest using arbitrary precision arithmetic,
+compute the prefactors to construct
+$y_{4,\text{max}(m)}$ and $y_{4,\text{max}(m)-1}$,
+recursively compute $y_{4,m}$ for all $m$ less than $\text{max}(m)$ down
+to $m=0$, then cancel off the prefactors of the resulting values of
+$y_{4,m}$ for all $m$ to produce ${_2F_1}$ for all desired $m$ values.
+\mmnote{Maybe I should reformat this paragraph into
+a pseudocode version of the algorithm?
+Instead of, or in addition to?}
+
+With ${_2F_1}$ computed, the only remaining numerical danger in computing
+$p(m)$ in~\eref{eq:p_m_bursty+rep} is overflow of the gamma functions.
+This is easily solved by taking the log of the entire expression
+and using standard routines to compute the log of the gamma functions,
+then exponentiating the entire expression at the end if $p(m)$
+is needed rather than $\log p(m)$.
 
@@
 \bibliographystyle{abbrv}
-\bibliography{../library}
+\bibliography{./library}
 

### 2020-05-09 doc/paper.tex
Add discussion of choice of constit model

--- 2020-04-20 doc/gen_fcn_notes.tex
@@
 %%%%% Referencing macros %%%%%
-\newcommand{\fref}[1]{Figure~\ref{#1}}
+\newcommand{\fig}[1]{Figure~\ref{#1}}
 \newcommand{\tref}[1]{Table~\ref{#1}}
-\newcommand{\eref}[1]{Eq.~(\ref{#1})}
-\newcommand{\erngref}[2]{Eq.~(\ref{#1}-\ref{#2})} % Equations
-% \newcommand{\test}[1][]{%
-% \ifthenelse{\isempty{#1}{}}{omitted}{#1}%
-% }
+\newcommand{\eq}[1]{Eq.~(\ref{#1})}
+\newcommand{\eqrange}[2]{Eq.~(\ref{#1}-\ref{#2})}
 
@@
 
-\title{Generating function solution for bursty promoters}
+\title{Simple repression draft}
 
@@
 
-\section{Constitutive promoter with bursts}
-
-\subsection{From master equation to generating function}
-
-\begin{figure}
+\section{Introduction}
+
+\section{Means}
+\subsection{Fold-changes are indistinguishable across models}
+As shown in Figure~1.
+\begin{figure}%[h!]
+\centering
+\includegraphics[width=\textwidth]{../figures/fig1/fig1.pdf}
+\caption{\textbf{
+    An overview of the simple repression motif at the level of means.}
+    (A) depicts the qualitative biological picture of simple repression.
+    (B) shows the ``master curve'' to which all cartoons in (C) and (D) collapse.
+    (C) shows two possible equilibrium pictures of simple repression, while
+    (D) shows four possible coarse-graining in a nonequilibrium picture.
+    }
+\label{fig1:means_cartoons}
+\end{figure}
+
+\section{Beyond means}
+\subsection{A one-state promoter with bursting is the ``best'' model of constitutive promoters}
+Before we can tackle simple repression, we need an adequate
+phenomenological model of constitutive expression.
+The literature abounds with options from which we can choose,
+and we show several potential kinetic models for constitutive promoters
+in~\fig{fig:constit_cartoons}.
+Let us consider the suitability of each model for our purposes in turn.
+
+\begin{figure}%[h!]
+\centering
+\includegraphics[width=0.9\textwidth]{../figures/fig1.5/fig1point5.ai}
+\caption{\textbf{Constitutive promoter cartoon comparison.}
+The left column depicts various plausible cartoons for constitutive promoters,
+along with literature references to their prior usage.
+In model (1), transcripts are produced in a Poisson process~\cite{Jones2014}.
+Model (2) features explicit modeling of RNAP binding/unbinding
+kinetics~\cite{Phillips2015a}.
+Model (3) is a more detailed generalization of model (2),
+treating transcription initiation as a multi-step process
+proceeding through closed and open complexes~\cite{Mitarai2015}.
+(4) a more coarse-grained version of (2) where the precise nature
+of active and inactive states are left ambiguous~\mmnote{Cite Manuel}, and
+(5) transcripts are produced in bursts,
+and initiation of bursts is a Poisson process.
+\mmnote{Haven't found model 5 formulated quite like this in the literature,
+usually burstiness is handled like model 4, but I should probably search a bit more.}
+The right column shows the Fano factor for each model.
+Note especially the crucial diagnostic: (2) and (3) have $\nu$
+strictly below 1, while only for (4) and (5) can $\nu$ exceed 1.
+Models with Fano factors $\le 1$ cannot produce the FISH data
+observed in~\cite{Jones2014} without introducing additional
+assumptions and model complexity.
+}
+\label{fig:constit_cartoons}
+\end{figure}
+
+\subsubsection{Model 1 - Poisson promoter.}
+This is the picture from Jones et.\ al.~\cite{Jones2014}, which
+assumes that transcripts are produced as a Poisson process from a
+single promoter state. This model insists that the ``true'' mRNA
+distribution is Poisson, implying the Fano factor $\nu$ must be 1.
+In~\cite{Jones2014}, the authors carefully attribute deviations
+from Fano = 1 to intensity variability in fluorescent spots, gene
+copy number variation, and copy number fluctuations of the
+transcription machinery, e.g., RNAP itself.
+In this picture, the master equation makes no appearance,
+and all the corrections to Poisson behavior are derived
+as additive corrections to the Fano factor.
+For disproving the ``universal noise curve'' from So et
+al~\cite{So2011}, this picture was excellent. It is appealing in
+its simplicity, with only two parameters, the initiation rate $r$
+and degradation rate $\gamma$. Since $\gamma$ is independently
+known from other experiments, and the mean mRNA copy number is
+$r/\gamma$, $r$ is easily inferred from data.
+In other words, the model is not excessively complex for the data at hand.
+But for many interesting questions, for instance in the recent
+work~\mmnote{cite Manuel's preprint once posted},
+knowledge of means and variances alone is insufficient, and a
+description of the full distribution of molecular counts is necessary.
+For this we need a( slightly) more complex model than model 1
+that would allow us to incorporate the non-Poissonian features of
+constitutive promoters directly into a master equation formulation.
+
+\subsubsection{Model 2 - Two-state promoter, RNAP bound or unbound.}
+This model was considered in, e.g.,~\cite{Phillips2015a} and ~\cite{Phillips2019}.
+Here transcription initiation proceeds from the bound to the unbound state,
+reflecting the microscopic reality that an RNAP that has begun to elongate
+a transcript is no longer available at the start site to begin another.
+The problem with this picture is that the Fano factor is
+\begin{align}
+    \nu = 1 -
+        \frac{r\rate{k}{P}{+}}
+            {\left(\rate{k}{P}{+} + \rate{k}{P}{-} + r\right)
+             \left(\gamma + \rate{k}{P}{+} + \rate{k}{P}{-} + r\right)},
+\end{align}
+which is always $<1$.
+To make contact with the experimental reality of $\nu>1$,
+we will have to do, at a minimum,
+the same corrections to Poisson behavior as in model 1 above.
+So while this model adds an appealing element of microscopic
+reality, we are forced to reject it as the additional complexity
+is unable to capture the phenomenology of interest.
+Obviously the promoter state does in fact proceed through cycles
+of RNAP binding, initiating, and elongating, but it seems that
+the super-Poissonian noise in mRNA copy number we want to
+model must be governed by other features of the system.
+
+\subsubsection{Model 3 - Three-state promoter, multistep
+transcription initiation and escape.}
+How might we remedy the deficits of model 2?
+It is known~\cite{DeHaseth1998} that once RNAP initially binds
+the promoter region, a multitude of distinct steps occur
+sequentially before RNAP finally escapes into the elongation
+phase. Perhaps adding some of this mechanistic detail might
+rescue model 2? The next simplest refinement of model 2 could
+consider open complex formation and promoter escape as separate
+steps rather than as a single effective step. In other words, we
+construct model 3 by adding a single extra state to model 2, and
+we will label the two RNAP-bound states as the closed and open
+complexes, despite the true biochemical details certainly being
+more complex. The authors of~\cite{Mitarai2015}, for instance,
+considered this model, although they added an additional
+repressor bound state and did not explicitly consider the limit
+with no repressor that we analyze here. Again, our goal here is
+not a complete accounting of all the relevant biochemical detail;
+this is an exploratory search for the important features our
+model needs to include.
+
+Unfortunately, as the authors of~\cite{Mitarai2015} hint, this
+model too has Fano factor $\nu<1$. This can be seen by starting
+with their Eq.~A1 for the Fano factor of the analogous model with
+repressor and taking the limit as repressor concentration goes to
+zero. After substantial algebra, the result is
+\begin{align}
+\nu = 1 - \frac{r k_O k_P^+}{\mathcal{Z}}
+\frac{k_P^+ + k_P^- + k_O + \gamma}
+        {\mathcal{Z} + \gamma(k_P^+ + k_P^- + k_O + \gamma) + \gamma^2},
+\end{align}
+where we defined $\mathcal{Z} = r(k_O + k_P^-) + k_P^+(k_O + r)$
+for notational tidiness. This is necessarily less than 1 for
+arbitrary rate constants.
+
+In fact, we suspect \textit{any} model in which transcription
+proceeds through a multistep cycle must necessarily have $\nu<1$.
+The intuitive argument compares the waiting time distribution to
+traverse the cycle with the waiting time for a Poisson promoter
+(model 1) with the same mean time. The latter is simply an
+exponential distribution. The former is a convolution of multiple
+exponentials, and intuitively should be more peaked with a
+smaller fractional width than an exponential with the same mean.\footnote{
+This can be made more precise with a result
+from~\cite{Stewart2007}, who showed that the convolution of
+multiple gamma distributions (of which the exponential
+distribution is a special case) is to a very good approximation
+also gamma distributed. Using their Eq.~2 for the distribution of
+the convolution, with shape parameters set to 1 to
+give exponential distributions, the total waiting time
+distribution has variance to squared mean ratio
+$\sigma^2/\mu^2 = \sum_i k_i^2/\left(\sum_i k_i\right)^2$,
+where the $k_i$ are the rates of the individual steps. Clearly
+this is less than 1 and therefore the total waiting time
+distribution is narrower than the corresponding exponential.} A
+less disperse waiting time distribution means transcription
+initations are more uniformly distributed in time relative to a
+Poisson process. Hence the distribution of mRNA over a population
+of cells should be less variable compared to Poisson, giving
+$\nu<1$.\footnote{
+This last step, while intuitive, can be argued by analogy to
+photon statistics where antibunching gives rise to sub-Poissonian
+noise~\cite{Paul1982, Zou1990}. Although loopholes exist, we do
+not expect they apply for our problem. Nevertheless we refrain
+from elevating to a ``theorem'' this cycle/sub-Poissonian equivalence.
+\mmnote{Antibunching is the ``obvious'' analogy for me but it'd
+be totally out of left field for most readers, any suggestions on
+a more comprehensible reference??}
+}
+
+Regardless of the merits of this model in describing the noise
+properties of constitutive transcription initation, it does not
+capture the dominant source of noise we want to include in our
+phenomenological model, so we must discard these details and
+search elsewhere.
+
+\subsubsection{Model 4 - Two-state promoter,
+``active'' and ``inactive'' states.}
+Inspired by~\cite{Razo-Mejia2020}, we revisit an analog of model
+2, but the interpretation of the two states is changed. Rather
+than explicitly viewing them as RNAP bound and unbound, we view
+them as ``active'' and ``inactive,'' which are able and unable to
+initiate transcripts, respectively. We are noncommittal as to the
+microscopic details of these states.
+
+One interpretation~\cite{Chong2014, Sevier2016} for the active
+and inactive states is that it represents the promoter's
+supercoiling state: transitions to the inactive state are caused
+by accumulation of positive supercoiling, which inhibits
+transcription, and transitions back to ``active'' are caused by
+gyrase or other topoisomerases relieving the supercoiling. This
+is an interesting possibility because it would mean the timescale
+for promoter state transitions is driven by topoisomerase
+kinetics, not by RNAP kinetics. From in vitro measurments, the
+former are suggested to be of order minutes~\cite{Chong2014}.
+Contrast this with model 2, where the state transitions are
+assumed to be governed by RNAP, which, assuming a copy number per
+cell of order $10^3$, has a diffusion-limited association rate
+$k_{on} \sim 10^2~\text{s}^{-1}$ to a target promoter. Combined
+with known $K_D$'s of order $\mu$M, this gives an RNAP
+dissociation rate $k_{off}$ of order $10^2$. As we will show
+below, however, there are some lingering puzzles with
+interpreting this supercoiling interpretation, so we leave it as
+a speculation and refrain from assigning definite physical
+meaning to the two states in this model.
+
+Intuitively one might expect that, since transcripts are produced
+as a Poisson process only when the promoter is in one of the two
+states in this model, transcription initiations should now be
+``bunched,'' in constrast to the ``anti-bunching'' of models 2
+and 3 above. One might further guess that this bunching would
+lead to super-Poissonian noise in the mRNA distribution over a
+population of cells. An honest calculation of the Fano factor produces
+\begin{align}
+\nu &= 1 + \frac{r k^-}{(k^+ + k^- + \gamma)(k^+ + k^-)},
+\end{align}
+which is strictly greater than 1, verifying the above intuition.
+Note we have dropped the $P$ label on the promoter switching rates to
+emphasize that these very likely do not represent kinetics of RNAP itself.
+This calculation can also be sidestepped by noting that the model
+is mathematically equivalent to the simple repression repression model
+from~\cite{Jones2014}, with states and rates relabeled and reinterpreted.
+
+How does this model compare to model 1 above?
+In model 1, all non-Poisson features of the mRNA
+distribution were handled as extrinsic corrections.
+By contrast, here the 3 parameter model is used to fit the full
+mRNA distribution as measured in mRNA FISH experiments.
+In essense, all variability in the mRNA distribution is regarded
+as ``intrinsic,'' arising either from stochastic initiation or
+from switching between the two coarse-grained promoter states.
+The advantage of this approach is that it fits neatly into the
+master equation picture, and the parameters thus inferred can be used as
+input for more complicated models with regulation by transcription factors.
+
+While this seems promising, there is a major drawback for our purposes
+which was already uncovered by the authors of~\cite{Razo-Mejia2020}:
+the statistical inference problem is nonidentifiable
+\mmnote{find a good cite for this usage},
+meaning it is impossible to infer the
+parameters $r$ and $k^-$ from the FISH data of~\cite{Jones2014}
+(as shown in Fig.~S2 of~\cite{Razo-Mejia2020}).
+Rather, only the ratio $r/k^-$ could be inferred. In that work,
+the problem was worked around with an informative prior on the
+ratio $k^-/k^+$. That approach is unlikely to work here, as,
+recall, our entire goal in modeling constitutive expression is to
+use it as the basis for a yet more complicated model, when we add
+on repression. But adding more complexity to a model that is
+already poorly identified is a fool's errand,
+so we will explore one more potential model.
+
+\subsubsection{Model 5 - One-state promoter with explicit bursts.}
+This model is inspired by the failure mode of model 4.
+The key observation above was that, as found
+in~\cite{Razo-Mejia2020}, only two parameters, $k^+$ and the
+ratio $r/k^-$, could be directly inferred from the FISH data
+from~\cite{Jones2014}.
+So let us take this seriously and imagine a model where these are
+the only two model parameters. What would this model look like?
+
+To develop some intuition, consider model 4 in the limit
+$k^+ \ll k^- \lesssim r$, which is roughly satisfied by the
+parameters inferred in~\cite{Razo-Mejia2020}.
+In this limit, the system spends the majority of its time in the inactive state, occasionally becoming active and making a burst of transcripts.
+This should call to mind the well-known phenomenon of transcriptional bursting, as reported in, e.g.,~\cite{Golding2005,Chong2014,Sevier2016}\mmnote{should probably add some more cites here}.
+Let us make this correspondence more precise.
+The mean dwell time in the active state is $1/k^-$.
+While in this state, transcripts are produced at a rate $r$ per
+unit time. So on average, $r/k^-$ transcripts are produced before
+the system switches to the inactive state. Once in the inactive
+state, the system dwells there for an average time $1/k^+$ before
+returning to the active state and repeating the process. $r/k^-$
+resembles an average burst size, and $1/k^+$ resembles the time
+interval between burst events. More precisely, $1/k^+$ is the
+mean time between the end of one burst and the start of the next,
+whereas $1/k^+ + 1/k^-$ would be the mean interval between the
+start of two successive burst events, but in the limit $k^+ \ll
+k^-$, $1/k^+ + 1/k^- \approx 1/k^+$. Note that this limit ensures
+that the waiting time between bursts is approximately
+exponentially distributed, with mean set by the only timescale
+left in the problem,
+$1/k^+$.\footnote{
+If instead it were the case that $k^+ \sim k^-$, then the waiting time
+$1/k^+ + 1/k^-$ would have a peaked distribution, but this does
+not appear to be the case for any of datasets
+from~\cite{Jones2014}.}
+
+Let us now verify this intuition with a precise derivation
+to check that $r/k^-$ is in fact the mean burst size
+and to obtain the full burst size distribution.
+Consider first a constant, known dwell time $T$ in the active state.
+Transcripts are produced at a rate $r$ per unit time, so the
+number of transcripts $n$ produced during $T$ fits exactly
+the ``story'' for the Poisson distribution, i.e.,
+\begin{equation}
+    P(n\mid T) = \frac{(rT)^n}{n!} \exp(-rT).
+\end{equation}
+Since the dwell time $T$ is unobservable, we actually want $P(n)$,
+the dwell time distribution with no conditioning on $T$.
+Basic rules of probability theory tell us we can write $P(n)$
+in terms of $P(n\mid T)$ as
+\begin{equation}
+    P(n) =\int_0^\infty P(n\mid T) P(T) dT.
+\end{equation}
+But we know the dwell time distribution $P(T)$,
+which is exponentially distributed according to
+\begin{equation}
+    P(T) = k^- \exp(-T k^-),
+\end{equation}
+so $P(n)$ can be written as
+\begin{equation}
+    P(n) = k^- \frac{r^n}{n!}
+            \int_0^\infty T^n\exp[-(r + k^-)T]\,dT.
+\end{equation}
+A standard integral table shows
+$\int_0^\infty x^n e^{-ax}\,dx = n!/a^{n+1}$, so
+\begin{equation}
+    P(n) = \frac{k^- r^n}{(k^- + r)^{n+1}}
+        = \frac{k^-}{k^- + r}
+            \left(\frac{r}{k^- + r}\right)^n
+        = \frac{k^-}{k^- + r}
+            \left(1 - \frac{k^-}{k^- + r}\right)^n,
+\end{equation}
+which is exactly the geometric distribution with standard parameter
+$\theta\equiv k^-/(k^- + r)$ and domain
+$n \in \{0, 1, 2, \dots\}$
+(this is one of two common conventions for the geometric distribution).
+The mean of the geometric distribution, with this convention, is
+\begin{align}
+\langle n\rangle = \frac{1 - \theta}{\theta}
+        = \left(1 - \frac{k^-}{k^- + r}\right)
+                    \frac{k^- + r}{k^-}
+        = \frac{r}{k^-},
+\end{align}
+exactly as we guessed intuitively above.
+
+So in taking the limit $r,k^-\rightarrow\infty$, $r/k^-\equiv b$,
+we obtain a model which effectively has only a single promoter state,
+which initiates bursts at rate $k^+$
+(transitions to the active state, in the model 4 picture).
+The master equation for mRNA copy number $m$ takes the form
+\begin{align}
+\begin{split}
+\deriv{t}{p(m,t)} = & (m+1)\gamma p(m+1,t) - m\gamma p(m,t) \\
+        &+ \sum_{j=0}^{m-1} k^+ p(j,t) Geom(m-j;b)
+         - \sum_{j=m+1}^\infty k^+ p(m,t) Geom(j-m;b),
+\end{split}
+\end{align}
+where $Geom(n;b)$ is the geometric distribution with mean~$b$, i.e.,
+$Geom(n;b) = \frac{1}{1+b}\left(\frac{b}{1+b}\right)^n$
+(with domain over nonnegative integers as above).
+The first two terms are the usual mRNA degradation terms.
+The third term enumerates all ways the system can produce
+a burst of transcripts and arrive at copy number $m$,
+given that it had copy number $j$ before the burst.
+The fourth term allows the system to start with copy number $m$,
+then produce a burst and end with copy number $j$.
+In fact this last sum has trivial $j$ dependence and
+simply enforces normalization of the geometric distribution.
+Carrying it out we have
+\begin{equation}
+\begin{split}
+\deriv{t}{p(m,t)} = & (m+1)\gamma p(m+1,t) - m\gamma p(m,t) \\
+        &+ \sum_{j=0}^{m-1} k^+ p(j,t) Geom(m-j;b)
+            - k^+ p(m,t),
+\end{split}
+\end{equation}
+This improves on model 4 in that now the parameters are easily inferred,
+as we will see later, and have clean interpretations.
+The non-Poissonian features are attributed to the emprically
+well-established phenomenological picture of bursty transcription.
+
+The big approximation in going from model 4 to 5 is that a burst
+is produced instantaneously rather than over a finite time.
+If the true burst duration is not short compared to transcription factor
+kinetic timescales, this could be a problem in that mean burst size
+in the presense and absence of repressors could change,
+rendering parameter inferences from the constitutive case inappropriate.
+Let us make some simple estimates of this.
+
+Consider the time delay between the first and final RNAPs
+in a burst initiating transcription
+(\textit{not} the time to complete transcripts,
+which potentially could be much longer.)
+If this timescale is short compared to the typical
+search timescale of transcription factors, then all is well.
+The estimates from deHaseth et.\ al.~\cite{DeHaseth1998}
+put RNAP's diffusion-limited on rate around
+$\sim\text{few}\times10^{-2}~\text{nM}^{-1}~\text{s}^{-1}$
+and polymerase loading as high as $1~\text{s}^{-1}$.
+Then for reasonable burst sizes of $<10$, it is reasonable to guess that
+bursts might finish initiating on a timescale of tens of seconds or less
+(with another 30-60 sec to finish elongation, but that does not matter here).
+A transcription factor with typical copy number of order 10 (or less)
+would have a diffusion-limited association rate of order
+$(10~\text{sec})^{-1}$.
+Higher copy number TFs tend to have many binding sites over the genome,
+which should serve to pull them out of circulation and keep their
+effective association rates from rising too large.
+Therefore, there is \textit{perhaps} a timescale separation
+possible between transcription factor association rates and burst
+durations, but this assumption could very well break down, so we
+will have to keep it in mind when we infer repressor rates from
+the Jones et.\ al.\ FISH data later.
+
+In reflecting on these 5 models, the reader may feel that
+exploring a multitude of potential models just to return to a
+very minimal phenomenological model of bursty transcription may
+seem highly pedantic. But the purpose of the exercise was to
+examine a host of models from the literature and understand why
+they are insufficient, one way or another, for our purposes.
+Along the way we have learned that the detailed kinetics of RNAP
+binding and initiating transcription are probably irrelevant for
+setting the population distribution of mRNA.
+The timescales are simply too fast, and as we will see later
+in~\fig{fig:constit_post}, the noise seems to be governed by
+slower timescales. Perhaps in hindsight this is not surprising:
+intuitively, the degradation rate $\gamma$ sets the fundamental
+timescale for mRNA dynamics, and any other processes that
+substantially modulate the mRNA distribution should not differ
+from $\gamma$ by orders of magnitude.
+
+\subsection{Parameter inference for constitutive promoters}
+Having established that a one-state bursty model achieves a
+Goldilocks level of complexity, we test it more thoroughly on all
+the data from constitutive promoters from~\cite{Jones2014}.
+\fig{fig:constit_post} shows the results.
+\begin{figure}%[h!]
+\centering
 \includegraphics[width=0.9\textwidth]{../figures/fig2/fig2.pdf}
-\caption{
-\mmnote{This figure wouldn't appear in this section, instead
-just a reference to the cartoon so the reader knows
-what cartoon model we're doing all this math for.}
+\caption{\textbf{Constitutive promoter model comparison.}
+    (A) ???
+    (B) shows the ECDF of the observed population distribution
+    of mRNA transcripts under the control of a constitutive lacUV5 promoter.
+    Fits of model (1), Poisson, and model (5), negative binomial, are shown.
+    \mmnote{Would like to add posterior predictive intervals,
+    to give sense of robustness of model.}
+    (C) 95\% highest posterior density regions for burst rate $k_i$ and
+    mean burst size $b$ for 18 promoters from Brewster 2012. Legend is
+    same as (d).
+    \mmnote{shared y axis??}
+    (D) Burst rate $k_i$ plotted vs.\ binding energy from energy
+    matrices from Brewster 2012.
+    \mmnote{Need to add fit line still. Also visibility of color palette is very bad, adjust.
+    There are 18 data points, maybe 6 colors and 3 glyphs/linestyles
+    cycled through? or 9 and 2?}
     }
+\label{fig:constit_post}
 \end{figure}
+\mmnote{Discuss puzzling comparison w/ Chong2014: if supercoiling is the thing, why are my burst sizes all the same but burst rates vary?
+And why is the duty cycle of my promoter so low, ie., bursts so short?
+COmpare their fig 7E; their $\beta/\alpha$ is my $k^+/k^-$.
+They have one or two genes with very small $\beta/\alpha$, does the \textit{galK} locus just happen to be that, or is there a deeper disagreement?
+Hard to say w/o more data.}
+
+\subsection{Transcription factor kinetics can be inferred from FISH measurements}
+Figure~3 presents inferred transcription factor rates, compares
+them to single-molecule measurements, and compares their ratios
+to gold-standard binding energies.
+\begin{figure}%[h!]
+\centering
+\includegraphics[width=0.9\textwidth]{../figures/fig3/fig3.ai}
+\caption{\textbf{Simple repression model comparison.}
+(A) depicts the cartoon of simple repression for which we infer parameters.
+(B) shows 50\% and 95\% contours of several 2D slices of the 9D posterior
+distribution, from fitting one unbinding rate for each operator (O1, O2, O3)
+and one binding rate for each aTc concentration (corresponding to an
+unknown mean repressor copy number).
+(C) compares ratios of our inferred unbinding rates with operator
+binding energy differences measured in Garcia \& Phillips 2011 (triangles)
+and Razo-Mejia et.\ al.\ 2018 (squares).
+(D) compares our inferred unbinding rates with single-molecule measurements
+of the same from Hammar et.\ at.\ 2014 \mmnote{Johan Elf's group}.
+Values are of similar order of magnitude,
+but a precise quantitative comparison unfortunately depends sensitively
+on the assumed mRNA lifetime, which was not measured.
+    }
+\label{fig3:kR_inferences}
+\end{figure}
+
+\begin{figure}%[h!]
+\centering
+\includegraphics[width=0.9\textwidth]{../figures/figSIxx/ppc_many_pooled.pdf}
+\caption{\textbf{Posterior predictive checks.}
+(A) duplicates the posterior from~\fig{fig3:kR_inferences}B.
+(B), (C), and (D) show samples from the posterior predictive distribution
+for each of the datasets used in the inference.
+Samples are sorted by operator and plotted separately simply for visual clarity.
+The unrepressed promoter, UV5, is shown with each as a reference point.
+\mmnote{
+Obviously the PPC is not perfect,
+but considering the relative simplicity of the model, I'm pleasantly surprised
+how well it worked and that the model was even identifiable at all.
+Interestingly, and perhaps not surprisingly, only 3 of the 9 datasets
+were identifiable if fit in isolation: otherwise,
+it was only by making assumptions that certain rates are equal
+across datasets was identifiability achieved.
+}
+    }
+\label{fig4:ppc}
+\end{figure}
+
+\appendix
+
+\section{Derivations of constitutive promoter models}
+\subsection{Model 2}
+\mmnote{Obtain Fano factor from CME.}
+\subsection{Model 3}
+\mmnote{Obtain Fano factor by taking limit from Mitarai et al's expression.}
+\subsection{Model 4}
+\mmnote{Obtain Fano factor by relabeling cartoon from Jones2014.}
+
+\section{Generating function solution for bursty promoters}
+
+\subsection{Constitutive promoter with bursts}
+
+\subsubsection{From master equation to generating function}
+
 We would like to derive the full probability distribution of
@@
 just consider the bursty consititutive promoter,
-model 4 in Fig.~1A.
+model 4 in~\fig{fig:constit_post}
 \mmnote{replicating Charlotte's notes; she says it's worked out
@@
 \end{align}
-\subsection{Stead-state}
+\subsubsection{Stead-state}
 At steady-state, the PDE reduces to the ODE
@@
 
-\section{Adding repression}
-\subsection{Deriving the generating function for mRNA distribution}
-
-\begin{figure}
-\includegraphics[width=0.9\textwidth]{../figures/fig3/fig3.ai}
-\caption{
-\mmnote{This figure wouldn't appear in this section, instead
-just a reference to the cartoon so the reader knows
-what cartoon model we're doing all this math for.}
-}
-\end{figure}
+\subsection{Adding repression}
+\subsubsection{Deriving the generating function for mRNA distribution}
+
 Let us move from a one-state promoter to a two-state promoter, where one state
 has repressor bound and the other produces transcriptional bursts as above.
-A schematic of this model is shown in Fig.\ 2A.
+A schematic of this model is shown in~\fig{fig3:kR_inferences}.
 Although now we have an equation for each promoter state, otherwise
@@
 
-\subsection{Moments}
-We can verify the correctness of~\eref{eq:p_m_bursty+rep}
+\subsubsection{Moments}
+We can verify the correctness of~\eq{eq:p_m_bursty+rep}
 by comparison with Gillespie simulation.
@@
 We would like to carry out Bayesian parameter inference on FISH data
-from~\cite{Jones2014}, using~\eref{eq:p_m_bursty+rep} as our
+from~\cite{Jones2014}, using~\eq{eq:p_m_bursty+rep} as our
 likelihood. This requires accurate (and preferably fast)
@@
 to the tail of the distribution, which can easily reach 50, possibly 100.
-From~\eref{eq:p_m_bursty+rep}, this means evaluating ${_2F_1}$
+From~\eq{eq:p_m_bursty+rep}, this means evaluating ${_2F_1}$
 repeatedly for values of $a$, $b$, and $c$ spanning the full range
@@
 even if $\alpha$, $\beta$, and $\gamma$
-in~\eref{eq:p_m_bursty+rep} are small,
+in~\eq{eq:p_m_bursty+rep} are small,
 with the situation even worse if they are not small.
@@
 To quantify how slow, we found that
-evaluating the likelihood defined by~\eref{eq:p_m_bursty+rep} $\sim50$ times
+evaluating the likelihood defined by~\eq{eq:p_m_bursty+rep} $\sim50$ times
 (for a typical dataset of interest from~\cite{Jones2014},
@@
 In order to evaluate the probability distribution of our
-model,~\eref{eq:p_m_bursty+rep}, we need to evaluate hypergeometric functions
+model,~\eq{eq:p_m_bursty+rep}, we need to evaluate hypergeometric functions
 of the form ${_2F_1}(\alpha+m, \beta+m, \gamma+m; -b)$
@@
 is a numerically stable solution of the recurrence
-relation~\eref{eq:chosen_rec_rel} if the recursion is run
+relation~\eq{eq:chosen_rec_rel} if the recursion is run
 from large $m$ to small $m$.
@@
 With ${_2F_1}$ computed, the only remaining numerical danger in computing
-$p(m)$ in~\eref{eq:p_m_bursty+rep} is overflow of the gamma functions.
+$p(m)$ in~\eq{eq:p_m_bursty+rep} is overflow of the gamma functions.
 This is easily solved by taking the log of the entire expression

### 2020-05-10 doc/paper.tex
Small edits + outline sketch of section on means & fold-change

@@
 \section{Introduction}
-
+\mmnote{Key ideas, no particular order, that I haven't written down before:
+\begin{itemize}
+\item Our goal is to build phenomenological models of input-output
+functions of genetic circuits. We want concepts, not mere facts.
+In particular a key question is: at this level of
+coarse graining, what microscopic details do I need to explicitly
+model, and how do we figure that out? For example, do I need to
+worry about all or even any of the steps that individual RNAPs go
+through each time they make a transcript? Turning the question
+around, can we see any imprint of those processes in the
+available data? If the answer is no, then those processes are
+irrelevant for our purposes. Forward modeling and inverse
+(statistical inferential) modeling can complement each other
+beautifully here.
+\end{itemize}}
 \section{Means}
@@
 As shown in Figure~1.
+\mmnote{Key ideas:
+\begin{itemize}
+\item Exclusive repression and occupancy hypothesis. Not
+necessarily steric hindrance, e.g., maybe kinked DNA, but it is
+key that there is zero transcription when repressor is bound.
+This is the key takeaway from the qualitative cartoon in Fig 1A.
+\item Change order in Fig~1 and order of presentation of noneq
+models to match my treatment of Fano factors. This doesn't do
+them in order of increasing math complexity, but it does do them
+in logical order: Poisson, RNAP bound/unbound states, RNAP
+unbound/closed/open states, ambiguous active/inactive, bursty
+Poisson. For each model, sketch the intuition for its
+construction and interpretation of the states/weights/rates,
+possibly calling ahead/deferring some arguments to the
+constitutive Fano section. Then quote the expression for mean
+mRNA (that will be derived more carefully in an appendix), then
+show the algebra to compute fold-change from mean mRNA and show
+how it fits into form of the master curve.
+\item Emphasize that tuning any of the parameters in $\rho$ is
+probably off-limits experimentally: it wrecks entire cell
+physiology and confounds everything. You can write down models of
+what you THINK the important steps are and aren't, and use in
+vitro data for the important steps, but without any way to
+experimentally perturb in vivo, you have no way to TEST whether
+your modeling assumptions are correct! Also point out
+supercoiling~\cite{Chong2014,Sevier2016} to emphasize that the
+potential features that are necessary to include are WAY more
+than just RNAP details. One possible tuning knob is maybe RNAP
+binding sites, but we don't have a terrific predictive
+understanding of how mutations there would influence any of the
+parameters inside $\rho$, so still not obvious how to make use of
+that. Brewster/Daniel2012 energy matrix is only good in one model
+framework and doesn't tell you anything about how model params in
+other pictures would respond.
+\item Weak promoter corresponds to $\rho\approx 1$. In principle,
+if repressor binding E or kinetics are known to high enough
+precision, deviations from $\rho=1$ become a matter of expt. In
+practice, I don't think it's at all realistic to measure
+repressor params or expression levels precisely enough to detect
+the expected tiny deviations from $\rho=1$. We can estimate: how
+good would determination of $\Delta F_R$ in Fig~1B have to be to
+get a given uncertainty in $\rho$? I think it's approx 1-to-1:
+e.g., an uncertainty of 0.1 (kT) in $\Delta F_R$ converts to a
+0.1 uncert in $\log\rho$, and since $\rho\approx 1$, that means
+$\log(1+\Delta\rho)\approx\Delta\rho\approx 0.1$.
+That's a really hard determination of $\Delta F_R$ for a very
+coarse determination of $\rho$. Doubtful that inference of
+kinetic rates would be any easier and probably a lot harder.Also,
+in vitro RNAP data says $\rho\approx 1$ should be true, and even
+if that's the wrong quantity to be talking about, Manuel's
+inference of $k^+$ and $k^-$ suggests that still $\rho$ is close
+to 1, and that inference is agnostic to the microscopics.
+\item Master curve exists b/c repressor is on its own leaf of
+graph. (But maybe don't use that language.) Arb noneq stuff can
+happen on rest of promoter state space, doesn't matter. Is this
+``one curve to rule them all'' a tautology? How do we probe the
+theory? Can't do it at level of means, there's no discriminatory
+power, need to go beyond. Test if free energy inferred from
+fold-change measurements agrees with TF rates inferred from
+single molec, and if those rates agree with rates inferred from
+mRNA population distributions.
+\end{itemize}
+}
 \begin{figure}%[h!]
@@
 \caption{\textbf{Constitutive promoter cartoon comparison.}
-The left column depicts various plausible cartoons for constitutive promoters,
-along with literature references to their prior usage.
+The left column depicts various plausible cartoons for constitutive promoters.
+We also list literature references where models have been used previously.
 In model (1), transcripts are produced in a Poisson process~\cite{Jones2014}.
@@
 proceeding through closed and open complexes~\cite{Mitarai2015}.
-(4) a more coarse-grained version of (2) where the precise nature
-of active and inactive states are left ambiguous~\mmnote{Cite Manuel}, and
-(5) transcripts are produced in bursts,
+Model (4) is somewhat analogous to (2) except with the precise nature
+of active and inactive states left ambiguous~\cite{Razo-Mejia2020}.
+Finally, model (5) can be viewed as a certain limit of model (4)
+in which transcripts are produced in bursts,
 and initiation of bursts is a Poisson process.
@@
 give exponential distributions, the total waiting time
-distribution has variance to squared mean ratio
+distribution has a ratio of variance to squared mean 
 $\sigma^2/\mu^2 = \sum_i k_i^2/\left(\sum_i k_i\right)^2$,
@@
 not expect they apply for our problem. Nevertheless we refrain
-from elevating to a ``theorem'' this cycle/sub-Poissonian equivalence.
+from elevating this cycle/sub-Poissonian equivalence to a ``theorem.''
 \mmnote{Antibunching is the ``obvious'' analogy for me but it'd
@@
 which was already uncovered by the authors of~\cite{Razo-Mejia2020}:
-the statistical inference problem is nonidentifiable
-\mmnote{find a good cite for this usage},
+the statistical inference problem is nonidentifiable,
+in the sense described in Section 4.3 of~\cite{Gelman2013},
 meaning it is impossible to infer the

### 2020-05-12 doc/paper.tex
Added much detail to means section, but more to do

@@
 \newcommand{\eq}[1]{Eq.~(\ref{#1})}
-\newcommand{\eqrange}[2]{Eq.~(\ref{#1}-\ref{#2})}
+\newcommand{\eqrange}[2]{Eqs.~(\ref{#1}-\ref{#2})}
 
@@
 \item Our goal is to build phenomenological models of input-output
-functions of genetic circuits. We want concepts, not mere facts.
+functions of genetic circuits. More precisely, we want phenomenology
+in the sense of coarse-graining away atomistic detail, but still
+retaining biophysical meaning, e.g., we don't want to coarse-grain
+as far as Hill functions. Why not? We are motivated by studies
+like~\cite{Razo-Mejia2020}, for which Hill functions are
+insufficient. Such work requires a quantitative theory of how
+biophysical changes at the molecular level propagate to
+input-output functions at the genetic circuit level.
+We want concepts, not mere facts.
 In particular a key question is: at this level of
@@
 \end{itemize}}
+
+\fig{fig1:means_cartoons}A shows the qualitative picture of
+simple repression. An operator, the binding site on the DNA for a
+repressor protein, may be found occupied by a repressor, in which
+case transcription is blocked from occurring. Or it may be found
+unoccupied, in which case RNA polymerase (RNAP) may bind and
+transcription can proceed.
+
+The key assumption we make is that binding of repressor and RNAP in the promoter region of interest is exclusive, meaning that one or the other may bind, but never may both be simultaneously bound.
+It is often imagined that when the repressor is bound to its operator,
+RNAP is sterically blocked from binding to its promoter sequence.
+Current evidence suggests this is sometimes, but not
+always the case~\mmnote{Cite that fig, from ??Nathan or Bill or
+Rob??, showing locations of repressor binding sites far upstream
+of promoter}, and it remains an interesting open question precisely
+how a repressor bound far upstream is able to repress transription.
+Suggestions include ``action-at-a-distance'' mediated by kinks in
+the DNA, formed when the repressor is bound, that prevent RNAP binding.
+Nevertheless, our modeling in this work is sufficiently
+coarse-grained that we simply assume exclusive binding and need
+not worry about such precise details.
+
+\mmnote{Add a notation aside somewhere that all association rates are written as zeroth-order rates.
+In other words, the concentration of the molecule is hidden inside, so for instance, $k_R^+$ is proportional to repressor copy number.}
+
 \section{Means}
 \subsection{Fold-changes are indistinguishable across models}
-As shown in Figure~1.
+As a first stop on our search for the ``right'' model of simple
+repression, let us consider what we can learn from theory and
+measurements of fold-change in mean expression. The fold-change is defined as
+\begin{equation}
+\text{fold-change}
+= \frac{\langle \text{gene expression with repressor present} \rangle}
+        {\langle \text{gene expression with repressor absent} \rangle}
+= \frac{\langle m\rangle(R\ne 0)}{\langle m\rangle(R = 0)}
+\label{eq:fc_def}
+\end{equation}
+where angle brackets denote the averge over a population of cells
+and mean mRNA $\langle m\rangle$ is viewed as a function of
+repressor copy number $R$.
+Also note the implicit assumption in equating ratios of gene
+expression, i.e., protein levels, to mRNA levels. The second
+equality in~\ref{eq:fc_def} amounts to an assumption that the
+translation efficiency, i.e., the number of proteins translated
+per mRNA, is the same in both conditions.
+This is reasonable since the cells in the two conditions are
+identical except for the presence of the transcription factor,
+and it is difficult to imagine the transcription factor having
+any direct effect on translation.
+
+Fold-change has proven a very convenient observable in past
+work~\cite{Garcia2011a, Razo-Mejia2018, Chure2019}.
+Part of its utility in dissecting transcriptional regulation is
+its ratiometric nature, which is always easier than making an
+absolute measurement. Also, by measuring otherwise identical cells
+with an without a transcription factor present, any biological
+noise common to both conditions can be made to cancel away.
+
+Panels C and D of \fig{fig1:means_cartoons} depict a smorgasbord
+of mathematized cartoons for simple repression that have appeared
+in previous literature. For each cartoon, we calculate the
+fold-change in mean gene expression as predicted by that model,
+deferring some algebraic details to the appendix.
+What we will find is that all cartoons collapse to a single
+master curve, shown in \fig{fig1:means_cartoons}B, which contains
+just two parameters. We label the parameters $\Delta F_R$, an
+effective free energy parametrizing the repressor-DNA
+interaction, and $\rho$, which subsumes all details of
+transcription in the absence of repressors.
+We will offer some intuition for why this master curve exists and discuss
+why it is likely impossible to discriminate ``right'' from ``wrong''
+cartoons given only measurements of fold-changes in expression.
+
+\mmnote{Snappy names for all these cartoons would really help the reader.
+Also naming them by number of promoter states is confusing to the reader because analogous models have different number of states when we're talking about constitutive version vs version with repressor.
+Thoughts??}
+
+\subsubsection{Equilibrium, 2-state}
+In this simplest model, the promoter is idealized as existing in
+one of two states, either repressor bound or repressor unbound.
+The rate of transcription, and therefore the level of gene
+expression, is assumed to be proportional to the fraction of time
+spent in the repressor unbound state.
+Mathematically, this amounts to the statement that the mean mRNA level is
+\begin{equation}
+\langle m \rangle = \frac{r}{\gamma}
+        \left(1 + \frac{R}{N_{NS}} e^{-\beta\Delta\epsilon_R}\right)^{-1},
+\end{equation}
+where $r$ is the transcription rate from the repressor unbound
+state, $\gamma$ is the mRNA degradation rate, $R$ is repressor
+copy number, $N_{NS}$ is the number of nonspecific binding sites
+in the genome (where repressors spend most of their time when not
+bound to the operator, their specifc binding site),
+$\beta=1/k_BT$, and $\Delta\epsilon_R$ is the binding energy of a
+repressor to its operator site. The derivation of this is deferred
+to the appendix.
+
+The fold-change is easily found as the ratio of mean mRNA with
+and without repressor. This gives
+\begin{equation}
+\text{fold-change}
+= \left(1 + \frac{R}{N_{NS}} e^{-\beta\Delta\epsilon_R}\right)^{-1},
+\end{equation}
+which clearly matches the form of the master curve
+in~\eq{fig1:means_cartoons} with $\rho=1$.
+
+In fact it was noted in~\cite{Chure2019} that this two-state
+model can be viewed as the coarse-graining of any equilibrium
+promoter model in which no transcriptionally active states have
+transcription factor bound. Put differently, there is no overlap
+between TF bound states and transcriptionally active states. We
+will see this explicitly in the 3-state equilibrium model below,
+but perhaps surprising is that something analogous carries over
+even to the nonequilbrium models we consider later.
+
+\subsubsection{Equilibrium, 3-state}
+Compared to the previous model, here we fine-grain the repressor
+unbound state into two separate states: empty, and RNAP bound.
+This picture was used in~\cite{Garcia2011a} as we use it here,
+and in~\cite{Razo-Mejia2018} and~\cite{Chure2019} it was generalized
+to incorporate small-molecule inducers that bind the repressor.
+Interestingly, the effect of this is simply to renormalize $R$
+from the total number of repressors to a smaller effective number,
+roughly (though not exactly) the number which are unbound by inducers.
+
+The mean mRNA, as derived in the appendix from a similar
+enumeration of states and weights as the previous model, is
+\begin{equation}
+\langle m \rangle = \frac{r}{\gamma}
+\frac{\frac{P}{N_{NS}} e^{-\beta\Delta\epsilon_P}}
+        {
+        1 + \frac{R}{N_{NS}} e^{-\beta\Delta\epsilon_R}
+        + \frac{P}{N_{NS}} e^{-\beta\Delta\epsilon_P}
+        },
+\end{equation}
+where the new variables are $\Delta\epsilon_P$, the difference in
+RNAP binding energy to its specific site (the promoter) relative
+to an average nonspecific background site, and the RNAP copy
+number, $P$. The fold-change again follows immediately as
+\begin{align}
+\text{fold-change}
+&= \frac{\frac{P}{N_{NS}} e^{-\beta\Delta\epsilon_P}}
+        {
+        1 + \frac{R}{N_{NS}} e^{-\beta\Delta\epsilon_R}
+        + \frac{P}{N_{NS}} e^{-\beta\Delta\epsilon_P}
+        }
+\frac{1 + \frac{P}{N_{NS}} e^{-\beta\Delta\epsilon_P}}
+        {\frac{P}{N_{NS}} e^{-\beta\Delta\epsilon_P}}
+\\
+&= \left(
+1 + \frac{\frac{R}{N_{NS}} e^{-\beta\Delta\epsilon_R}}
+        {1 + \frac{P}{N_{NS}} e^{-\beta\Delta\epsilon_P}}
+\right)^{-1}
+\\
+&= (1 + \exp(-\Delta F_R - \log\rho))^{-1},
+\end{align}
+with $\Delta F_R = \beta\Delta\epsilon_R - \log(R/N_{NS})$
+and $\rho = 1 + \frac{P}{N_{NS}}\mathrm{e}^{-\beta\Delta\varepsilon_P}$
+as shown in~\fig{fig1:means_cartoons}.
+
+\subsubsection{Nonequilibrium model 1 - Poisson promoter}
+For our first kinetic model, we consider the simplest possible
+picture with only two states, repressor bound and unbound.
+This is exactly the model used for the main results of~\cite{Jones2014}.
+In this picture, repressor association and dissociation rates
+from its operator site, $k_R^+$ and $k_R^-$, respectively, govern
+transitions between the two states. When the system is in the
+unbound state, transcriptions initiate at rate $r$, which
+represents a coarse-graining of all the downstream processes into
+a single effective rate. mRNA is again degraded at rate $\gamma$.
+
+Let $p_{m,R}(t)$ denote the joint probability of finding the
+system in the repressor bound state with $m$ mRNA molecules
+present at time $t$, and similarly define $p_{m,U}(t)$ for the
+repressor unbound state. 
+This model is governed by a master equation giving the time
+evolution of $p_{m,R}(t)$ and $p_{m,U}(t)$, which is
+\begin{align}
+\begin{split}
+\deriv{t}p_{m,R}(t) =& - k_R^- p_{m,R}(t) + k_R^+ p_{m,U}(t)
+                + (m+1)\gamma p_{m+1,R}(t) - \gamma p_{m,R}(t)
+\\
+\deriv{t}p_{m,U}(t) =&\; k_R^- p_{m,R}(t) - k_R^+ p_{m,U}(t)
+                        + rp_{m-1,U}(t) - rp_{m,U}(t)
+                \\
+                &+ (m+1)\gamma p_{m+1,U}(t) - \gamma p_{m,U}(t).
+\end{split}
+\end{align}
+In each equation, the first two terms describe transitions
+between promoter states due to repressors unbinding and binding,
+respectively. The final two terms describe degradation of mRNA,
+decreasing the copy number by one, and the terms with coefficient
+$r$ describe transcription initiation increasing the mRNA copy
+number by one.
+
+We can greatly simplify the notation, which will be especially
+useful for the more complicated models yet to come, by
+rexpressing the master equation in vector form. The promoter
+states are collected into a vector and the rate constants are
+collected into matrices as
+\begin{equation}
+\vect{p}_m = \begin{pmatrix} p_{m,R} \\ p_{m,U} \end{pmatrix},\
+\matr{K} = \begin{pmatrix} -k_R^- & k_R^+ \\ k_R^- & -k_R^+ \end{pmatrix},\
+\matr{R} = \begin{pmatrix} 0 & 0 \\ 0 & r \end{pmatrix},\
+\label{eq:2state_cme_matrices}
+\end{equation}
+so that the master equation may be condensed as
+\begin{equation}
+\deriv{t}\vect{p}_m(t) =
+\left( \matr{K} - \matr{R} - \gamma m \matr{I} \right) \vect{p}_m(t)
+                + \matr{R} \vect{p}_{m-1}(t) +
+                \gamma (m+1) \matr{I} \vect{p}_{m+1}(t),
+\label{eq:2state_rep_cme}
+\end{equation}
+where $\matr{I}$ is the identity matrix.
+Taking steady state by setting time derivatives to zero,
+the mean mRNA can be found to be
+\begin{equation}
+\langle m \rangle = \frac{r}{\gamma}
+        \left(1 + \frac{k_R^+}{k_R^-}\right)^{-1},
+\end{equation}
+with the algebra details again deferred to the appendix.
+Recall $k_R^+$ is proportional to the repressor copy number, so
+in computing fold-change, absence of repressor corresponds to
+$k_R^+\rightarrow0$.
+Therefore fold-change in this model is simply
+\begin{equation}
+\text{fold-change} = \left(1 + \frac{k_R^+}{k_R^-}\right)^{-1},
+\end{equation}
+again matching the master curve of~\fig{fig1:means_cartoons} with $\rho=1$.
+
+\subsubsection{Nonequilibrium model 2 - RNAP bound/unbound states}
+Analogous to the three-state equilibrium model, this model
+fine-grains the repressor unbound state of nonequilbrium model 1,
+resolving it into an empty promoter state and an RNAP-bound state.
+Note in this picture, in contrast with model 4 below,
+transcription initiation is accompanied by a promoter state
+change, in keeping with the interpretation as RNAP-bound and
+empty states: if an RNAP successfully escapes the promoter and
+proceeds to elongation of a transcript, clearly it is no longer
+bound at the promoter! Therefore another RNAP must bind before
+another transcript can be initiated.
+
+The master equation governing this model is analogous
+to~\eqrange{eq:2state_cme_matrices}{eq:2state_rep_cme} for model 1 above.
+The main subtlety arises since transcription initiation
+accompanies a promoter state change, but this matrix can be
+understood by analogy to the off-diagonal element of $\matr{K}$.
+The master equation for this model is
+\begin{equation}
+\deriv{t}\vect{p}_m(t) =
+\left( \matr{K} - \matr{R_{out}} - \gamma m \matr{I} \right) \vect{p}_m(t)
+                + \matr{R_{in}} \vect{p}_{m-1}(t) +
+                \gamma (m+1) \matr{I} \vect{p}_{m+1}(t),
+\label{eq:3state_rep_cme}
+\end{equation}
+with the state vector and promoter transition matrix defined as
+\begin{equation}
+\vect{p}_m = \begin{pmatrix} p_{m,R} \\ p_{m,E} \\ p_{m,P} \end{pmatrix},\
+\matr{K} = \begin{pmatrix} -k_R^- & k_R^+ & 0 \\
+                        k_R^- & -k_R^+ -k_P^+ & k_P^- \\
+                        0 & k_P^+ & -k_P^- 
+                \end{pmatrix},
+\label{eq:3state_cme_matrices_pt1}
+\end{equation}
+and the initiation matrices given by
+\begin{equation}
+\matr{R_{in}} = \begin{pmatrix}
+                0 & 0 & 0 \\ 0 & 0 & r \\ 0 & 0 & 0
+                \end{pmatrix},\
+\matr{R_{out}} = \begin{pmatrix}
+                0 & 0 & 0 \\ 0 & 0 & 0 \\ 0 & 0 & r
+                \end{pmatrix}.
+\label{eq:3state_cme_matrices_pt2}
+\end{equation}
+The elements of $\vect{p}_m$ encode the probabilities of having
+$m$ mRNA present along with the promoter having repressor bound,
+being empty, or having RNAP bound, respectively.
+$\matr{R_{in}}$ describes probability flux into the state
+$\vect{p}_m$ from a state with one fewer mRNA, namely
+$\vect{p}_{m-1}$, and $\matr{R_{out}}$ describes probability flux
+out of the state $\vect{p}_m$ for a state with one more mRNA,
+namely $\vect{p}_{m+1}$.
+$\matr{K}$ is closely analogous to model 1.
+
+Mean mRNA at steady state is found analogously to model 1, with the result
+\begin{equation}
+\langle m\rangle = \frac{r}{\gamma}
+        \frac{k_R^- k_P^+}
+        {k_R^- k_P^+ + k_R^- (k_P^- + r) + k_R^+ (k_P^- + r)}.
+\end{equation}
+Fold-change is again found from the ratio prescribed by~\eq{eq:fc_def},
+from which we have
+\begin{align}
+\text{fold-change}
+&=      \frac{k_R^- k_P^+}
+        {k_R^- k_P^+ + k_R^- (k_P^- + r) + k_R^+ (k_P^- + r)}
+        \frac{k_P^+ + k_P^- + r}{k_P^+}
+\\
+&=      \left(1 + \frac{k_R^+}{k_R^-}
+                \frac{k_P^- + r}{k_P^+ + k_P^- + r}
+        \right)^{-1}
+\\
+&=      \left(1 + \frac{k_R^+}{k_R^-}
+        \left(1 + \frac{k_P^+}{k_P^- + r}\right)^{-1}
+        \right)^{-1},
+\end{align}
+which follows the master curve with
+$\rho = 1 + k_P^+/(k_P^- + r)$ as claimed.
+
+\subsubsection{Nonequilibrium model 3
+- Multistep transcription initiation and escape}
+One might reasonably complain that the first two
+``nonequilbrium'' models we have considered are straw men. Their
+steady states necessarily satisfy detailed balance which is
+equivalent to thermodynamic equilibrium. Why is this the case? At
+steady state there is by definition no net probability flux in or
+out of each promoter state, but since the promoter states form a
+linear chain, there is only one way in or out of the repressor
+bound and RNAP bound states, implying each edge must actually
+have a net zero probability flux, which is the definition of
+detailed balance (usually phrased as equality of forward and
+reverse transition fluxes).
+
+Now we consider model 3 in~\eq{fig1:means_cartoons} which allows
+the possibility of true nonequilibrium steady-state fluxes
+through the promoter states.
+We point out that this model was considered previously in~\cite{Mitarai2015}
+where a comparison was made with model 1 as used in~\cite{Jones2014}.
+The authors of~\cite{Mitarai2015} argued that the additional
+complexity is essential to properly account for the noise in the
+mRNA distribution. We will weigh in on both models later when we
+consider observables beyond fold-change.
+
+The master equation governing this model is identical to model 2 above,
+\begin{equation}
+\deriv{t}\vect{p}_m(t) =
+\left( \matr{K} - \matr{R_{out}} - \gamma m \matr{I} \right) \vect{p}_m(t)
+                + \matr{R_{in}} \vect{p}_{m-1}(t) +
+                \gamma (m+1) \matr{I} \vect{p}_{m+1}(t),
+\end{equation}
+but obviously with different matrices.
+The state vector and promoter transition matrix are now
+\begin{equation}
+\vect{p}_m = \begin{pmatrix} p_{m,R} \\ p_{m,E} \\
+                             p_{m,C} \\ p_{m,O}\end{pmatrix},\
+\matr{K} = \begin{pmatrix} -k_R^- & k_R^+ & 0 & 0\\
+                        k_R^- & -k_R^+ -k_P^+ & k_P^- & 0 \\
+                        0 & k_P^+ & -k_P^- - k_O & 0 \\
+                        0 & 0 & k_O & 0
+                \end{pmatrix},
+\end{equation}
+with the four promoter states, in order, being repressor bound,
+empty, RNAP closed complex, and RNAP open complex.
+Besides increasing dimension by one again, the only new feature in
+$\matr{K}$ is the rate $k_O$, representing the (irreversible)
+rate of open complex formation from the closed complex.
+The initiation matrices are given by
+\begin{equation}
+\matr{R_{in}} = \begin{pmatrix}
+        0 & 0 & 0 & 0 \\ 0 & 0 & 0 & r \\ 0 & 0 & 0 & 0\\ 0 & 0 & 0 & 0
+                \end{pmatrix},\
+\matr{R_{out}} = \begin{pmatrix}
+        0 & 0 & 0 & 0 \\ 0 & 0 & 0 & 0 \\ 0 & 0 & 0 & 0\\ 0 & 0 & 0 & r
+                \end{pmatrix},
+\end{equation}
+again closely analogous to model 2.
+
+The expression for mean mRNA is substantially more complicated
+now, as worked out in the appendix where we find
+\begin{align}
+\langle m\rangle &= \frac{r}{\gamma}
+        \frac{k_R^- k_P^+ k_O}
+        {k_R^- [(k_P^+ (k_O + r) + r(k_P^- + k_O)] + k_R^+ r(k_P^- + k_O)}
+\\
+&= \frac{r}{\gamma}
+\frac{\frac{k_P^+ k_O}{r(k_O + k_P^-)}}
+        {1 + \frac{k_P^+ (k_O + r)}{r(k_O + k_P^-)} + \frac{k_R^+}{k_R^-}}.
+\end{align}
+The strategy is to isolate the terms involving the repressor, so that now the fold-change is seen to be simply
+\begin{align}
+\text{fold-change}
+&= \frac{\frac{k_P^+ k_O}{r(k_O + k_P^-)}}
+        {1 + \frac{k_P^+ (k_O + r)}{r(k_O + k_P^-)} + \frac{k_R^+}{k_R^-}}
+        \frac{1 + \frac{k_P^+ (k_O + r)}{r(k_O + k_P^-)}}
+                {\frac{k_P^+ k_O}{r(k_O + k_P^-)}}
+\\
+&= \left(
+        1 + \frac{k_R^+}{k_R^-}
+        \left(1 + \frac{k_P^+ (k_O + r)}{r(k_O + k_P^-)}\right)^{-1}
+\right)^{-1},
+\end{align}
+surprisingly reducing to the master curve once again, with
+$\rho = 1 + \frac{k_P^+ (k_O + r)}{r(k_O + k_P^-)}$.
+
+This example hints that an arbitrarily fine-grained model of
+downstream transcription steps may still be collapsed to the form
+of the master curve in~\fig{fig1:means_cartoons}B so long as the
+repressor binding is exclusive with transcriptionally active
+states. We offer this as a conjecture, and we suspect that a
+careful argument using the King-Altman diagram
+method~\cite{King1956, Hill1966} might furnish a ``proof.''
+Our focus here is not on full generality but rather to survey an
+assortment of plausible models that have been proposed in the
+literature for simple repression.
+
+\subsubsection{Nonequilibrium model 4 - ``Active'' and ``inactive'' states}
+This model is at the core of the theory in~\cite{Razo-Mejia2020}.
+At a glance the cartoon for this model may appear very similar to model 2,
+and mathematically it is, but the interpretation is rather different.
+In model 2, we interpreted the third state literally as an
+RNAP-bound promoter and modeled initiation of a transcript as
+triggering a promoter state change, making the hopefully
+uncontroversial assumption that an RNAP can only make one
+transcript at a time. In contrast, in the present model the
+promoter state does \textit{not} change when a transcript is
+initiated. So we no longer interpret these states as literally
+RNAP bound and unbound but instead as coarse-grained ``active''
+and ``inactive'' states, the details of which we leave
+unspecified for now. We will comment more on this model below
+when we discuss Fano factors of models.
+
+Mathematically this model is very similar to models 1 and 2.
+Like model 1, the matrix $R$ describing transcription initiation
+is diagonal, namely
+\begin{equation}
+\matr{R} = \begin{pmatrix}
+                0 & 0 & 0 \\ 0 & 0 & 0 \\ 0 & 0 & r
+        \end{pmatrix}.
+\end{equation}
+so the master equation takes verbatim the same form as it did for model
+1,~\eq{eq:2state_rep_cme}. Meanwhile the promoter transition matrix
+$K$ is the same as~\eq{eq:3state_cme_matrices_pt1} from model 2,
+although we relabel the rate constants from $k_P^\pm$ to $k^\pm$ to
+reiterate that these are not simply RNAP binding and unbinding rates.
+
+Carrying out the algebra, the mean mRNA can be found to be
+\begin{equation}
+\langle m\rangle = \frac{r}{\gamma}
+\frac{k_R^- k^+}
+{k_R^- k^+ + k_R^- k^- + k_R^+ k^-},
+\end{equation}
+and the fold-change readily follows,
+\begin{align}
+\text{fold-change}
+&=      \frac{k_R^- k^+}{k_R^- k^+ + k_R^- k^- + k_R^+ k^-}
+        \frac{k_R^- k^+ + k_R^- k^-}{k_R^- k^+}
+\\
+&=      \left(1 + \frac{k_R^+}{k_R^-}
+                \left(1 + \frac{k^+}{k^-}\right)^{-1}
+        \right)^{-1},
+\end{align}
+from which we see $\rho = 1 + k^+/k^-$ as shown in~\fig{fig1:means_cartoons}.
+
+\subsubsection{Nonequilibrium model 5 - Bursty promoter}
+We will have much more to say about this model in subsequent sections.
+For now we sketch it as an intuitive analog to model 1,
+with just two states, repressor bound or unbound, and transition rates
+between them of $k_R^+$ and $k_R^-$.
+In model 1, when in the unbound state, single mRNA transcripts
+are produced as a Poisson process with some characteristic rate
+$r$. This model by contrast produces, at some Poisson rate $k_i$,
+\textit{bursts} of transcripts with a geometrically distributed
+burst size and a mean burst size $b$.
+
+From this intuitive picture and by analogy to model 1, then, it should be plausible that the mean mRNA level is
+\begin{equation}
+\langle m\rangle = \frac{k_i b}{\gamma}
+        \left(1 + \frac{k_R^+}{k_R^-}\right)^{-1},
+\end{equation}
+which will turn out to be correct from a careful calculation.
+For now, we simply note that just like model 1, the fold-change becomes
+\begin{equation}
+\text{fold-change} = \left(1 + \frac{k_R^+}{k_R^-}\right)^{-1}
+\end{equation}
+with $\rho=1$ also like model 1.
+
+We will also see later how this model emerges as a natural limit of model 4.
+
+\subsubsection{Summary and discussion of results across models for fold-changes in mean expression}
 \mmnote{Key ideas:
@@
 This is the key takeaway from the qualitative cartoon in Fig 1A.
-\item Change order in Fig~1 and order of presentation of noneq
-models to match my treatment of Fano factors. This doesn't do
-them in order of increasing math complexity, but it does do them
-in logical order: Poisson, RNAP bound/unbound states, RNAP
-unbound/closed/open states, ambiguous active/inactive, bursty
-Poisson. For each model, sketch the intuition for its
-construction and interpretation of the states/weights/rates,
-possibly calling ahead/deferring some arguments to the
-constitutive Fano section. Then quote the expression for mean
-mRNA (that will be derived more carefully in an appendix), then
-show the algebra to compute fold-change from mean mRNA and show
-how it fits into form of the master curve.
+\item Key assumption: expression is proportional to occupation
+probability of whatever the ``active'' state is for a given
+model, whatever we mean by active.
+\item Also emphasize that parameters/rates are \textit{not}
+apples-to-apples between cartoons, especially the initiation rate $r$.
+Indeed, this is the entire point! Except for repressor
+energies/rates, the other parameters are not independently
+measurable or inferrable b/c we're not even sure what
+steps/states we should be writing in the model!
+(simple RNAP kinetics as inferred in vitro, or supercoiling, or
+RNAP cluster kinetics a la Ibrahim, or maybe
+gene-copy/RNAP-copy/other-copy number is really the key, or...)
+So all you get are coarse-grained combos like $\rho$, and even
+$\rho$ would be hard to infer, much less trying to infer its innards.
 \item Emphasize that tuning any of the parameters in $\rho$ is
@@
 other pictures would respond.
+In theory you could refit their raw data to different models, but
+as we see later, even having data of full mRNA distributions is
+insufficient to identify parameters in all but the simplest of
+the models in Figs.\ 1 and 2, so it seems doubtful that only
+measurements of means would fare better.
+??Double check, what did they actually measure in 2012 paper??
 \item Weak promoter corresponds to $\rho\approx 1$. In principle,
@@
 mRNA population distributions.
+\item Here's a fun puzzle: 2-state equilibrium model gets details
+of transcription so wrong, how can it be useful? Because
+fold-change is a ratio! Since details all collapse into $\rho$,
+fitting this 2-state model to fold-change measurements can
+\textit{still} give you a not-bad estimate of ratio of repressor
+rates, i.e., binding energies.
+\item 
 \end{itemize}
@@
 \centering
-\includegraphics[width=\textwidth]{../figures/fig1/fig1.pdf}
+\includegraphics[width=\textwidth]{../figures/fig1/fig1.ai}
 \caption{\textbf{
@@
     (C) shows two possible equilibrium pictures of simple repression, while
-    (D) shows four possible coarse-graining in a nonequilibrium picture.
+    (D) shows five possible coarse-grainings in a nonequilibrium picture.
     }
@@
 
+\section{Discussion and future work}
+\mmnote{
+Do all my approximations work simply because galK happens to have
+a low duty cycle of supercoiling, $\beta/\alpha$ in Chong2014
+picture? A potentially interesting test: put in the dozen or so
+Hernan constructs at some other locus, preferably one of the ones
+Chong identified as having a large duty cycle (large $\beta/\alpha$),
+i.e., the opposite limit of galK. Then just measure fold-changes.
+You don't even need to do FISH, although that would be even
+better. If their model is right, $\rho$ should be approximately
+set by the duty cycle of supercoiling, NOT by RNAP kinetics.
+\textit{This is a NICE prediction for future work.} \textit{cspE}
+is claimed nonessential, prediction is you drop HG reporter
+constructs in there and the inferred binding energies of all lacI
+operators shift the same amount, by about $\ln(3)\sim 1kT$. Not
+huge, but detectable, although might be confounding effects from
+moving to a new locus. With FISH measurements, $k^+$ and $k^-$ in
+Fig 2, model 4 should now be separately inferrable if Chong2014
+measurements of $\beta/\alpha$ are correct. This would require an
+obnoxious amount of cloning, but once you did that, the
+measurements would be same old stuff that RP lab does.}
+
 \appendix

### 2020-05-12 code/figures/fig1master_curve.py

@@ -11,5 +11,5 @@ fig, ax = plt.subplots(1, 1, figsize=(5, 3), dpi=150)
 x = np.linspace(-6,6)
 ax.plot(x, 1/(1+np.exp(-x)))
 ax.set_ylabel('fold-change')
-ax.set_xlabel(r"$\Delta F_R - \log\rho~(k_BT)$")
+ax.set_xlabel(r"$\Delta F_R + \log\rho~(k_BT)$")
 plt.savefig('../../figures/fig1/fig1_master_curve.pdf', bbox_inches='tight')

### 2020-05-13 doc/paper.tex
Edits from RP + other small edits

@@
 repressor copy number $R$.
-Also note the implicit assumption in equating ratios of gene
-expression, i.e., protein levels, to mRNA levels. The second
-equality in~\ref{eq:fc_def} amounts to an assumption that the
+The second
+equality in~\ref{eq:fc_def} follows from assuming that the
 translation efficiency, i.e., the number of proteins translated
 per mRNA, is the same in both conditions.
+In other words, we assume that mean protein level is proportional
+to mean mRNA level, and that the proportionality constant is the
+same in both conditions and therefore cancels out in the ratio.
 This is reasonable since the cells in the two conditions are
@@
 absolute measurement. Also, by measuring otherwise identical cells
-with an without a transcription factor present, any biological
+with and without a transcription factor present, any biological
 noise common to both conditions can be made to cancel away.
 
-Panels C and D of \fig{fig1:means_cartoons} depict a smorgasbord
+\fig{fig1:means_cartoons} depicts a smorgasbord
 of mathematized cartoons for simple repression that have appeared
@@
 What we will find is that all cartoons collapse to a single
-master curve, shown in \fig{fig1:means_cartoons}B, which contains
+master curve, shown in \fig{fig1:means_cartoons}c, which contains
 just two parameters. We label the parameters $\Delta F_R$, an
@@
 Mathematically, this amounts to the statement that the mean mRNA level is
+given by
 \begin{equation}
@@
 $\beta=1/k_BT$, and $\Delta\epsilon_R$ is the binding energy of a
-repressor to its operator site. The derivation of this is deferred
+repressor to its operator site. The derivation of this result is deferred
 to the appendix.
@@
 as shown in~\fig{fig1:means_cartoons}.
+
+\begin{figure}%[h!]
+\centering
+\includegraphics[width=\textwidth]{../figures/fig1/fig1.ai}
+\caption{
+\textbf{
+An overview of the simple repression motif at the level of means.}
+(A) depicts the qualitative biological picture of simple repression.
+(B) shows a variety of possible mathematized cartoons of simple
+repression, along with the corresponding complexity $\rho$ of the
+architecture, an effective parameter which subsumes all details
+of the architecture excluding the repressor.
+(C) shows the ``master curve'' to which all cartoons in (B) collapse.
+}
+\label{fig1:means_cartoons}
+\end{figure}
 
@@
 useful for the more complicated models yet to come, by
-rexpressing the master equation in vector form. The promoter
+re-expressing the master equation in vector form. The promoter
 states are collected into a vector and the rate constants are
@@
 \left( \matr{K} - \matr{R} - \gamma m \matr{I} \right) \vect{p}_m(t)
-                + \matr{R} \vect{p}_{m-1}(t) +
-                \gamma (m+1) \matr{I} \vect{p}_{m+1}(t),
+                + \matr{R} \vect{p}_{m-1}(t)
+                + \gamma (m+1) \matr{I} \vect{p}_{m+1}(t)
+                - \gamma m \matr{I} \vect{p}_m(t),
 \label{eq:2state_rep_cme}
@@
 The main subtlety arises since transcription initiation
-accompanies a promoter state change, but this matrix can be
-understood by analogy to the off-diagonal element of $\matr{K}$.
+accompanies a promoter state change.
+This can be understood by analogy to $\matr{K}$. The
+off-diagonals and diagonals of $\matr{K}$ correspond to
+transitions arriving at or departing from, respectively, the
+promoter state of interest. If transcription initiations are
+accompanied by promoter state changes, we must have separate
+matrices for arriving and departing transcription events since
+the arriving and departing transitions have different initial
+copy number of mRNA, unlike for $\matr{K}$ where they are the same.
 The master equation for this model is
@@
 \deriv{t}\vect{p}_m(t) =
-\left( \matr{K} - \matr{R_{out}} - \gamma m \matr{I} \right) \vect{p}_m(t)
-                + \matr{R_{in}} \vect{p}_{m-1}(t) +
-                \gamma (m+1) \matr{I} \vect{p}_{m+1}(t),
+\left( \matr{K} - \matr{R_D} - \gamma m \matr{I} \right) \vect{p}_m(t)
+                + \matr{R_A} \vect{p}_{m-1}(t) +
+                \gamma (m+1) \matr{I} \vect{p}_{m+1}(t)
+                - \gamma m \matr{I} \vect{p}_m(t),
 \label{eq:3state_rep_cme}
@@
 \begin{equation}
-\matr{R_{in}} = \begin{pmatrix}
+\matr{R_A} = \begin{pmatrix}
                 0 & 0 & 0 \\ 0 & 0 & r \\ 0 & 0 & 0
                 \end{pmatrix},\
-\matr{R_{out}} = \begin{pmatrix}
+\matr{R_D} = \begin{pmatrix}
                 0 & 0 & 0 \\ 0 & 0 & 0 \\ 0 & 0 & r
-                \end{pmatrix}.
+                \end{pmatrix},
 \label{eq:3state_cme_matrices_pt2}
@@
 being empty, or having RNAP bound, respectively.
-$\matr{R_{in}}$ describes probability flux into the state
+$\matr{R_A}$ describes probability flux \textit{arriving} at the state
 $\vect{p}_m$ from a state with one fewer mRNA, namely
-$\vect{p}_{m-1}$, and $\matr{R_{out}}$ describes probability flux
-out of the state $\vect{p}_m$ for a state with one more mRNA,
-namely $\vect{p}_{m+1}$.
+$\vect{p}_{m-1}$, and $\matr{R_D}$ describes probability flux
+\textit{departing} from the state $\vect{p}_m$
+for a state with one more mRNA, namely $\vect{p}_{m+1}$.
 $\matr{K}$ is closely analogous to model 1.
@@
 
-The master equation governing this model is identical to model 2 above,
+The master equation governing this model is
+identical in form to model 2 above, namely
 \begin{equation}
 \deriv{t}\vect{p}_m(t) =
-\left( \matr{K} - \matr{R_{out}} - \gamma m \matr{I} \right) \vect{p}_m(t)
-                + \matr{R_{in}} \vect{p}_{m-1}(t) +
-                \gamma (m+1) \matr{I} \vect{p}_{m+1}(t),
-\end{equation}
-but obviously with different matrices.
+\left( \matr{K} - \matr{R_D} - \gamma m \matr{I} \right) \vect{p}_m(t)
+                + \matr{R_A} \vect{p}_{m-1}(t) +
+                \gamma (m+1) \matr{I} \vect{p}_{m+1}(t)
+                - \gamma m \matr{I} \vect{p}_m(t),
+\end{equation}
+but obviously with a higher-dimensional state space and different matrices.
 The state vector and promoter transition matrix are now
@@
 \begin{equation}
-\matr{R_{in}} = \begin{pmatrix}
+\matr{R_A} = \begin{pmatrix}
         0 & 0 & 0 & 0 \\ 0 & 0 & 0 & r \\ 0 & 0 & 0 & 0\\ 0 & 0 & 0 & 0
                 \end{pmatrix},\
-\matr{R_{out}} = \begin{pmatrix}
+\matr{R_D} = \begin{pmatrix}
         0 & 0 & 0 & 0 \\ 0 & 0 & 0 & 0 \\ 0 & 0 & 0 & 0\\ 0 & 0 & 0 & r
@@
 now, as worked out in the appendix where we find
-\begin{align}
-\langle m\rangle &= \frac{r}{\gamma}
+\begin{equation}
+\langle m\rangle = \frac{r}{\gamma}
         \frac{k_R^- k_P^+ k_O}
-        {k_R^- [(k_P^+ (k_O + r) + r(k_P^- + k_O)] + k_R^+ r(k_P^- + k_O)}
-\\
-&= \frac{r}{\gamma}
+        {k_R^- [(k_P^+ (k_O + r) + r(k_P^- + k_O)] + k_R^+ r(k_P^- + k_O)},
+\end{equation}
+which can be simplified to
+\begin{equation}
+\langle m\rangle
+= \frac{r}{\gamma}
 \frac{\frac{k_P^+ k_O}{r(k_O + k_P^-)}}
         {1 + \frac{k_P^+ (k_O + r)}{r(k_O + k_P^-)} + \frac{k_R^+}{k_R^-}}.
-\end{align}
+\end{equation}
 The strategy is to isolate the terms involving the repressor, so that now the fold-change is seen to be simply
@@
 }
-\begin{figure}%[h!]
-\centering
-\includegraphics[width=\textwidth]{../figures/fig1/fig1.ai}
-\caption{\textbf{
-    An overview of the simple repression motif at the level of means.}
-    (A) depicts the qualitative biological picture of simple repression.
-    (B) shows the ``master curve'' to which all cartoons in (C) and (D) collapse.
-    (C) shows two possible equilibrium pictures of simple repression, while
-    (D) shows five possible coarse-grainings in a nonequilibrium picture.
-    }
-\label{fig1:means_cartoons}
-\end{figure}
 

### 2020-05-16 doc/paper.tex
Finished discussion of means + other small edits

@@
 
-\mmnote{Snappy names for all these cartoons would really help the reader.
-Also naming them by number of promoter states is confusing to the reader because analogous models have different number of states when we're talking about constitutive version vs version with repressor.
-Thoughts??}
-
 \subsubsection{Equilibrium, 2-state}
@@
 but perhaps surprising is that something analogous carries over
-even to the nonequilbrium models we consider later.
+even to the nonequilibrium models we consider later.
 
@@
 (B) shows a variety of possible mathematized cartoons of simple
-repression, along with the corresponding complexity $\rho$ of the
-architecture, an effective parameter which subsumes all details
+repression, along with the effective parameter $\rho$
+which subsumes all regulatory details
 of the architecture excluding the repressor.
@@
 Analogous to the three-state equilibrium model, this model
-fine-grains the repressor unbound state of nonequilbrium model 1,
+fine-grains the repressor unbound state of nonequilibrium model 1,
 resolving it into an empty promoter state and an RNAP-bound state.
@@
 One might reasonably complain that the first two
-``nonequilbrium'' models we have considered are straw men. Their
+``nonequilibrium'' models we have considered are straw men. Their
 steady states necessarily satisfy detailed balance which is
@@
 \subsubsection{Summary and discussion of results across models for fold-changes in mean expression}
-\mmnote{Key ideas:
-\begin{itemize}
-\item Exclusive repression and occupancy hypothesis. Not
-necessarily steric hindrance, e.g., maybe kinked DNA, but it is
-key that there is zero transcription when repressor is bound.
-This is the key takeaway from the qualitative cartoon in Fig 1A.
-\item Key assumption: expression is proportional to occupation
-probability of whatever the ``active'' state is for a given
-model, whatever we mean by active.
-\item Also emphasize that parameters/rates are \textit{not}
-apples-to-apples between cartoons, especially the initiation rate $r$.
-Indeed, this is the entire point! Except for repressor
-energies/rates, the other parameters are not independently
-measurable or inferrable b/c we're not even sure what
-steps/states we should be writing in the model!
-(simple RNAP kinetics as inferred in vitro, or supercoiling, or
-RNAP cluster kinetics a la Ibrahim, or maybe
-gene-copy/RNAP-copy/other-copy number is really the key, or...)
-So all you get are coarse-grained combos like $\rho$, and even
-$\rho$ would be hard to infer, much less trying to infer its innards.
-\item Emphasize that tuning any of the parameters in $\rho$ is
-probably off-limits experimentally: it wrecks entire cell
-physiology and confounds everything. You can write down models of
-what you THINK the important steps are and aren't, and use in
-vitro data for the important steps, but without any way to
-experimentally perturb in vivo, you have no way to TEST whether
-your modeling assumptions are correct! Also point out
-supercoiling~\cite{Chong2014,Sevier2016} to emphasize that the
-potential features that are necessary to include are WAY more
-than just RNAP details. One possible tuning knob is maybe RNAP
-binding sites, but we don't have a terrific predictive
-understanding of how mutations there would influence any of the
-parameters inside $\rho$, so still not obvious how to make use of
-that. Brewster/Daniel2012 energy matrix is only good in one model
-framework and doesn't tell you anything about how model params in
-other pictures would respond.
-In theory you could refit their raw data to different models, but
-as we see later, even having data of full mRNA distributions is
-insufficient to identify parameters in all but the simplest of
-the models in Figs.\ 1 and 2, so it seems doubtful that only
-measurements of means would fare better.
-??Double check, what did they actually measure in 2012 paper??
-\item Weak promoter corresponds to $\rho\approx 1$. In principle,
-if repressor binding E or kinetics are known to high enough
-precision, deviations from $\rho=1$ become a matter of expt. In
-practice, I don't think it's at all realistic to measure
-repressor params or expression levels precisely enough to detect
-the expected tiny deviations from $\rho=1$. We can estimate: how
-good would determination of $\Delta F_R$ in Fig~1B have to be to
-get a given uncertainty in $\rho$? I think it's approx 1-to-1:
-e.g., an uncertainty of 0.1 (kT) in $\Delta F_R$ converts to a
-0.1 uncert in $\log\rho$, and since $\rho\approx 1$, that means
-$\log(1+\Delta\rho)\approx\Delta\rho\approx 0.1$.
-That's a really hard determination of $\Delta F_R$ for a very
-coarse determination of $\rho$. Doubtful that inference of
-kinetic rates would be any easier and probably a lot harder.Also,
-in vitro RNAP data says $\rho\approx 1$ should be true, and even
-if that's the wrong quantity to be talking about, Manuel's
-inference of $k^+$ and $k^-$ suggests that still $\rho$ is close
-to 1, and that inference is agnostic to the microscopics.
-\item Master curve exists b/c repressor is on its own leaf of
-graph. (But maybe don't use that language.) Arb noneq stuff can
-happen on rest of promoter state space, doesn't matter. Is this
-``one curve to rule them all'' a tautology? How do we probe the
-theory? Can't do it at level of means, there's no discriminatory
-power, need to go beyond. Test if free energy inferred from
-fold-change measurements agrees with TF rates inferred from
-single molec, and if those rates agree with rates inferred from
+What have we learned from our exploration of the models
+in~\fig{fig1:means_cartoons} and their predictions of
+fold-changes in mean expression levels? The takeaway is the
+existence of a master curve (\fig{fig1:means_cartoons}C) to which
+the fold-change predictions of all the models collapse. This
+master curve is parametrized by only two effective parameters:
+$\Delta F_R$, which characterizes the repressor, and $\rho$,
+which characterizes all other features of the promoter
+architecture. The key assumption underpinning this result is that
+zero transcription occurs when a repressor is bound to its
+operator, though note we are agnostic about the molecular
+mechanism which achieves this.\footnote{
+Steric effects are the obvious mechanism, but, for instance,
+``action-at-a-distance'' mediated by kinked DNA due to
+repressors bound tens or hundreds of nucleotides upstream of a
+promoter is plausible as well.}
+
+Why does the master curve of~\fig{fig1:means_cartoons}C exist at all?
+This brings to mind the deep questions posed in,
+e.g.,~\cite{Frank2013} and~\cite{Frank2014a}, asking us to search
+for general ``symmetries'' of the problem instead of specifying
+generative models of the known microscopics of a system.
+In our case, the key feature seems to be the exclusive nature of
+repressor and RNAP binding, which allows the parameter describing
+the repressor, $\Delta F_R$, to cleanly separate from all other
+details of the promoter architecture, which are encapsulated in $\rho$.
+In the graph-theoretic language which was applied to similar
+problems in~\cite{Gunawardena2012} and~\cite{Ahsendorf2014}, the
+repressor-bound state is a leaf of the graph describing
+transitions between promoter states. Arbitrary nonequilibrium
+behavior can occur on the rest of the promoter state space, but
+it may all be swept up in the effective parameter $\rho$, to
+which the repressor makes no contribution.
+
+As suggested in~\cite{Chure2019}, we believe this master curve
+should generalize to architectures with multiple repressor
+binding sites, as long as the exclusivity of transcription factor
+binding and transcription initation is maintained. The
+interpretation of $\Delta F_R$ is then of an effective free
+energy of all repressor bound states.
+In an equilibrium picture this is simply given by the log of the
+sum of Boltzmann weights of all repressor bound states, which
+looks like the log of a partition function of a subsystem.
+In a nonequilibrium picture, while we can still mathematically gather
+terms and give the resulting collection the label $\Delta F_R$,
+it is unclear if the physical interpretation as an effective free energy
+makes sense, and we leave this an open problem for future work.\footnote{
+As is well known, free energies cannot be assigned unambiguously
+to states out of equilibrium because the free energy change along
+a generic path traversing the state space is path dependent,
+unlike at equilibrium.
+A consequence of this is that, out of equilibrium, $\Delta F_R$
+is no longer a simple sum of Boltzmann weights. Instead it
+resembles a restricted sum of King-Altman
+diagrams~\cite{King1956, Hill1966}. Following the work of
+Hill~\cite{Hill1989}, it may yet be possible to interpret this expression
+as an effective free energy, but this remains unclear to us.
+}
+
+If we relax the requirement of exclusive repressor-RNAP binding,
+one could imagine models in which repressor and RNAP doubly-bound
+states are allowed, where the repressor's effect is to
+\text{reduce} the transcription rate rather than setting it to zero.
+Our results do not strictly apply to such a model, although we
+note that if the repressor's reduction of the transcription rate
+is substantial, such a model might still be well-approximated by
+one of the models in~\fig{fig1:means_cartoons}.
+
+One may worry that our ``one curve to rule them all'' is a
+mathematical tautology. In fact we \textit{agree} with this
+criticism if $\Delta F_R$ is ``just a fitting parameter'' and cannot
+be meaningfully interpreted as a real, physical free energy.
+We believe our theory deserves to be scrapped unless it can be
+demonstrated that $\Delta F_R$ \textit{is} a physically
+meaningful and interpretable free energy.
+
+How do we test this and probe the theory with fold-change
+measurements? There is a fundamental limitation in that the
+master curve is essentially a one-parameter function of $\Delta
+F_R + \log\rho$. Worse, there are many \textit{a priori}
+plausible microscopic mechanisms that could contribute to the
+value of $\rho$, such as RNAP binding and escape
+kinetics~\cite{DeHaseth1998, Mitarai2015}, and/or supercoiling
+accumulation and release~\cite{Chong2014, Sevier2016}, and/or,
+more speculatively, RNAP clusters analogous to those obvserved in
+eukaryotes~\cite{Cisse2013}. Even if $\Delta F_R$ is measured to
+high precision, inferring the potential microscopic contributions
+to $\rho$, buried inside a log no less, from fold-change
+measurements seems hopeless. As a statistical inference problem
+it is entirely nonidentifiable, in the language
+of~\cite{Gelman2013}, Section 4.3.
+
+If we cannot simply infer values of $\rho$ from measurements of
+fold-change, can we perturb some of the parameters that make up
+$\rho$ and measure the change? Unfortunately this is probably
+off-limits experimentally: most of the potential contributors to
+$\rho$ are global processes that affect many or all genes. For
+instance, changing RNAP association rates by changing RNAP copy
+numbers, or changing supercoiling kinetics by changing
+topoisomerase copy numbers, would massively perturb the entire
+cell's physiology and confound any determination of $\rho$.
+
+One might instead imagine a bottom-up modeling approach, where we
+mathematize a model of what we hypothesize the important steps
+are and are not, use in vitro data for the steps deemed
+important, and \textit{predict} what $\rho$ should be.
+But again, because of the one-parameter nature of the master
+curve, many different models will likely make indistinguishable
+predictions, and without any way to experimentally perturb in
+vivo, there is no clear way to test whether the modeling
+assumptions are correct.
+
+In light of this, we prefer the view that parameters and rates
+are not directly comparable between cartoons
+in~\fig{fig1:means_cartoons}. Rather, parameters in the
+simpler cartoons represent coarse-grained combinations of
+parameters in the finer-grained models. For instance, by equating
+$\rho$ between any two models, one can derive various possible
+correspondences between the two models' parameters. Note that
+these correspondences are clearly not unique, since many possible
+associations could be made. It then is a choice as to what microscopic
+interpretations the model-builder prefers for the parameters in a
+particular cartoon, and as to which coarse-grainings lend
+intuition and which seem nonsensical. Indeed, since it remains an open question what microscopic features dominate $\rho$ (as suggested above, perhaps RNAP binding and escape
+kinetics~\cite{DeHaseth1998, Mitarai2015}, or supercoiling
+accumulation and release~\cite{Chong2014, Sevier2016}, or,
+something more exotic like RNAP clusters~\cite{Cisse2013}), we are hesitant to put too much weight on any one microscopic interpretation of model parameters that make up $\rho$.
+
+One possible tuning knob to probe $\rho$ that would not globally
+perturb the cell's physiology is to manipulate RNAP binding sites.
+Work such as~\cite{Kinney2010} has shown that models of
+sequence-dependent RNAP can be inferred from data, and the
+authors of~\cite{Brewster2012} showed that the model
+of~\cite{Kinney2010} has predictive power by using the model to
+\textit{design} binding sites of a desired affinity.
+But for our purposes, this begs the question: the authors
+of~\cite{Kinney2010} \textit{assumed} a particular model
+(essentially our 3-state equilibrium model but without the
+repressor), so it is unclear how or if such methods can be turned
+around to \textit{compare} different models of promoter function.
+
+We have already pointed out that the master curve
+of~\fig{fig1:means_cartoons} is essentially a one-parameter model,
+the one parameter being $\Delta F_R + \log\rho$.
+By now the reader may be alarmed as to how can we even determine
+$\Delta F_R$ and $\rho$ independently of each other, never mind
+shedding a lens on the internal structure of $\rho$ itself.
+A hint is provided by the weak promoter approximation, invoked
+repeatedly in prior studies~\cite{Bintu2005c, Garcia2011a,
+Razo-Mejia2018} of simple repression using the 3-state
+equilibrium model in~\fig{fig1:means_cartoons}.
+In that picture, the weak promoter approximation means
+$\frac{P}{N_{NS}}\exp(-\beta\Delta\epsilon_P) \ll 1$,
+meaning therefore $\rho\approx1$.
+This is suggestive, but how can we be sure that $\rho$ is not,
+for instance, actually $10^2$ and that $\Delta F_R$ hides a
+compensatory factor?
+A resolution is offered by an independent inference of $\rho$ in
+the absence of repressors. This was done in~\cite{Razo-Mejia2020}
+by fitting nonequilibrium model 4  in~\fig{fig1:means_cartoons},
+with zero repressor (looking ahead, this is equivalent to model 4
+in~\fig{fig:constit_cartoons}), to FISH data
+from~\cite{Brewster2014}. This provided a determination of $k^+$
+and $k^-$, from which their ratio is estimated to be no more than
+a few $10^{-1}$ and possibly as small as $10^{-2}$.
+
+The realization that $\rho\approx1$ to an excellent
+approximation, \textit{independent} of which model
+in~\ref{fig1:means_cartoons} one prefers, goes a long way towards
+explaining the surprising success of equilibrium models of simple
+repression. Even though our 2- and 3-state models get so many
+details of transcription wrong, it does not matter because
+fold-change is a cleverly designed ratio. Since $\rho$ subsumes
+all details except the repressor, and $\log\rho\approx0$, fitting
+these simple models to fold-change measurements can still give a
+surprisingly good estimate of repressor binding energies. So the
+ratiometric construction of fold-change fulfills its intended
+purpose of canceling out all features of the promoter
+architecture except the repressor itself. Nevertheless it is
+perhaps surprising how effectively it does so: \textit{a priori},
+one might not have expected $\rho$ to be quite so close to 1.
+
+We would also like to highlight the relevance
+of~\cite{Landman2019} here. Landman et.\ al.\ reanalyzed and
+compared in vivo and in vitro data on the lacI repressor's
+binding affinity to its various operator sequences. (The in vivo
+data was from, essentially, fitting our master curve to
+expression measurements.) They find broad agreement between the
+in vitro and in vivo values. This reinforces the suspicion that
+the equilibrium $\Delta\epsilon_R$ repressor binding energies do
+in fact represent real physical free energies.
+Again, \textit{a priori} this did not have to be the case, even
+knowing that $\rho\approx1$.
+
+In principle, if $\Delta F_R$ can be measured to sufficient
+precision, then deviations from $\rho=1$ become a
+testable matter of experiment.
+In practice, it is probably unrealistic to measure repressor
+rates $k_R^+$ or $k_R^-$ or fold-changes in expression levels
+(and hence $\Delta\epsilon_R$) precisely enough to detect the
+expected tiny deviations from $\rho=1$.
+We can estimate the
+requisite precision in $\Delta F_R$ to resolve a given
+$\Delta\rho$ by noting, since $\rho\approx1$, that
+$\log(1+\Delta\rho)\approx \Delta\rho$, so
+$\Delta(\Delta F_R) \approx \Delta\rho$.
+Suppose we are lucky and $\Delta\rho$ happens to be $\sim0.1$, on
+the high end of our range estimated above.
+A determination of $\Delta\epsilon_R/k_BT$ with an uncertainty of
+barely 0.1 was achieved in the excellent measurements
+of~\cite{Razo-Mejia2018}, so this requires a very difficult
+determination of $\Delta F_R$ for a very crude determination of $\rho$,
+which suggests, to put it lightly, this is not a
+promising path to pursue experimentally.
+It is doubtful that inference of repressor kinetic rates
+would be any easier.\footnote{
+\mmnote{
+In fact the uncertainties quoted in Hammar et.\
+al.~\cite{Hammar2014} from Elf's group are small enough to do
+this calculation, but HJ \& I think their error bars are insanely
+optimistic. Not sure I want to blatantly call them out on it though??}
+}
+
+Moving forward, we have weak evidence supporting the
+interpretation of $\Delta F_R$ as a physically real free
+energy~\cite{Landman2019} and other work casting
+doubt~\cite{Hammar2014}.
+How might we resolve the confusion?
+If there is no discriminatory power to test the theory and
+distinguish the various models with measurements of fold-changes
+in means, how do we probe the theory? Clearly to discriminate
+between the nonequilibrium models in~\fig{fig1:means_cartoons},
+we need to go beyond means to ask questions about kinetics, noise and even
+full distributions of mRNA copy numbers over a population of cells.
+If the ``one-curve-to-rule-them-all'' is more than a
+mathematical tautology, then the free energy of repressor binding
+inferred from fold-change measurements should agree with
+repressor binding and unbinding rates.
+In other words, the equilbrium and nonequilibrium definitions of
+$\Delta F_R$ must agree, meaning
+\begin{equation}
+\Delta F_R = \beta\Delta\epsilon_R - \log(R/N_{NS})
+        = - \log(k_R^+/k_R^-),
+\end{equation}
+must hold, where $\beta\Delta\epsilon_R$ is inferred from the
+master curve fit to fold-change measurements, and $k_R^+$ and
+$k_R^-$ are inferred in some orthogonal manner.
+Single molecule measurements such as from~\cite{Hammar2014} have
+directly observed these rates, and in the remainder of this work
+we explore a complementary approach: inferring repressor binding
+and unbinding rates $k_R^+$ and $k_R^-$ from FISH measurements of
 mRNA population distributions.
-\item Here's a fun puzzle: 2-state equilibrium model gets details
-of transcription so wrong, how can it be useful? Because
-fold-change is a ratio! Since details all collapse into $\rho$,
-fitting this 2-state model to fold-change measurements can
-\textit{still} give you a not-bad estimate of ratio of repressor
-rates, i.e., binding energies.
-\item 
-\end{itemize}
-}
 

### 2020-05-16 code/figures/fig1master_curve.py

@@ -5,7 +5,7 @@ import pandas as pd
 import matplotlib.pyplot as plt
 import srep.viz
 colors = srep.viz.color_selector('pboc')
-srep.viz.plotting_style()
+srep.viz.plotting_style(grid=False)
 
 fig, ax = plt.subplots(1, 1, figsize=(5, 3), dpi=150)
 x = np.linspace(-6,6)

### 2020-05-16 doc/paper.tex
Edits from RP & HG

@@
 noise common to both conditions can be made to cancel away.
+In addition, the presence of different operator sites is known to
+change the basal expression level in the absence of
+repressor~\mmnote{look up cite}.
+Computing fold-changes neatly cancels this effect as well.
 
@@
 zero transcription occurs when a repressor is bound to its
-operator, though note we are agnostic about the molecular
-mechanism which achieves this.\footnote{
-Steric effects are the obvious mechanism, but, for instance,
+operator. Note, however, that we are agnostic about the molecular
+mechanism which achieves this;
+steric effects are the obvious mechanism, but, for instance,
 ``action-at-a-distance'' mediated by kinked DNA due to
 repressors bound tens or hundreds of nucleotides upstream of a
-promoter is plausible as well.}
+promoter is plausible as well.
 
@@
 it is unclear if the physical interpretation as an effective free energy
-makes sense, and we leave this an open problem for future work.\footnote{
+makes sense.
 As is well known, free energies cannot be assigned unambiguously
@@
 as an effective free energy, but this remains unclear to us.
-}
+We leave this an open problem for future work.
 
@@
 be meaningfully interpreted as a real, physical free energy.
-We believe our theory deserves to be scrapped unless it can be
-demonstrated that $\Delta F_R$ \textit{is} a physically
+An analogy to Hill functions is appropriate here. One of their
+great strengths and weaknesses, depending on the use they are put
+to, is that their parameters coarse-grain many details and are
+generally not interpretable in terms of microscopic models, for
+deep reasons discussed at length in~\cite{Frank2013}.
+By contrast, our master curve claims to have the best of both worlds:
+a coarse-graining of all details besides the repressor into
+a single effective parameter $\rho$,
+while simultaneously retaining an interpretation of
+$\Delta F_R$ as a physically
 meaningful and interpretable free energy.
+Our task, then, is to prove or disprove this claim.
 
@@
 accumulation and release~\cite{Chong2014, Sevier2016}, and/or,
-more speculatively, RNAP clusters analogous to those obvserved in
-eukaryotes~\cite{Cisse2013}. Even if $\Delta F_R$ is measured to
+RNAP clusters analogous to those obvserved in
+eukaryotes~\cite{Cisse2013, Cho2016}
+and recently also observed in bacteria~\cite{Ladouceur2020}.
+Even if $\Delta F_R$ is measured to
 high precision, inferring the potential microscopic contributions
@@
 fold-change, can we perturb some of the parameters that make up
-$\rho$ and measure the change? Unfortunately this is probably
+$\rho$ and measure the change? Unfortunately we suspect this is
 off-limits experimentally: most of the potential contributors to
@@
 particular cartoon, and as to which coarse-grainings lend
-intuition and which seem nonsensical. Indeed, since it remains an open question what microscopic features dominate $\rho$ (as suggested above, perhaps RNAP binding and escape
-kinetics~\cite{DeHaseth1998, Mitarai2015}, or supercoiling
-accumulation and release~\cite{Chong2014, Sevier2016}, or,
-something more exotic like RNAP clusters~\cite{Cisse2013}), we are hesitant to put too much weight on any one microscopic interpretation of model parameters that make up $\rho$.
+intuition and which seem nonsensical.
+Indeed, since it remains an open question what microscopic
+features dominate $\rho$ (as suggested above, perhaps RNAP
+binding and escape kinetics~\cite{DeHaseth1998, Mitarai2015}, or
+supercoiling accumulation and release~\cite{Chong2014, Sevier2016},
+or, something more exotic like
+RNAP clusters~\cite{Cisse2013, Cho2016, Ladouceur2020}),
+we are hesitant to put too much weight on any one microscopic
+interpretation of model parameters that make up $\rho$.
 
@@
 approximation, \textit{independent} of which model
-in~\ref{fig1:means_cartoons} one prefers, goes a long way towards
+in~\fig{fig1:means_cartoons} one prefers, goes a long way towards
 explaining the surprising success of equilibrium models of simple

### 2020-05-17 doc/paper.tex
Overhaul & split constit inference fig, also hacked p-tile PPC plots

@@
 \centering
-\includegraphics[width=0.9\textwidth]{../figures/fig2/fig2.pdf}
-\caption{\textbf{Constitutive promoter model comparison.}
-    (A) ???
+\includegraphics[width=0.9\textwidth]{../figures/fig2/fig2pt1.pdf}
+\caption{\textbf{Constitutive promoter posterior
+        inference and predictive checks.}
+        (A) shows the joint posterior density from MCMC sampling of model 5,
+        bursty promoter with negative binomially-distributed steady state.
         (B) shows the ECDF of the observed population distribution
         of mRNA transcripts under the control of a constitutive lacUV5 promoter.
-    Fits of model (1), Poisson, and model (5), negative binomial, are shown.
-    \mmnote{Would like to add posterior predictive intervals,
-    to give sense of robustness of model.}
-    (C) 95\% highest posterior density regions for burst rate $k_i$ and
-    mean burst size $b$ for 18 promoters from Brewster 2012. Legend is
-    same as (d).
-    \mmnote{shared y axis??}
-    (D) Burst rate $k_i$ plotted vs.\ binding energy from energy
+        Posterior predictive percentiles are shown as colored bands for
+        model (1), Poisson, and model (5), negative binomial.
+        }
+\label{fig:uv5_post}
+\end{figure}
+
+\begin{figure}%[h!]
+\centering
+\includegraphics[width=0.9\textwidth]{../figures/fig2/fig2pt2.pdf}
+\caption{\textbf{Constitutive promoter model comparison.}
+    (A) Joint posterior distributions for burst rate $k_i$ and
+    mean burst size $b$ for 18 promoters from Jones 2014.
+    Each contour indicates the 95\% highest posterior density
+    region for a particular promoter, i.e., on a plot analogous
+    to~\fig{fig:uv5_post}, the contour would enclose
+    approximately 95\% of the samples, and thus 95\% of the
+    probability mass.
+    Note that the vertical axis is shared with (B).
+    Panel (B) plots the burst rate $k_i$ vs.\ binding energy from energy
     matrices from Brewster 2012.
-    \mmnote{Need to add fit line still. Also visibility of color palette is very bad, adjust.
-    There are 18 data points, maybe 6 colors and 3 glyphs/linestyles
-    cycled through? or 9 and 2?}
+    The dotted line shows the predicted slope according to
+    $\ln k_i/\gamma \sim \text{constant} - \beta\Delta\epsilon_P$,
+    described in text.
     }
@@
 
+%%%%%%%%%%%%%%%%%%%%% APPENDICES %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 \appendix
@@
 just consider the bursty consititutive promoter,
-model 4 in~\fig{fig:constit_post}
+model 5 in~\fig{fig:constit_cartoons}
 \mmnote{replicating Charlotte's notes; she says it's worked out
@@
 
-%%%%%%%%%%%%%%%%%%%%% APPENDICES %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\appendix
-
 % \bibliographystyle{nature}

### 2020-05-17 code/analysis/constit_post_inf.py

@@ -10,7 +10,7 @@ import arviz as az
 
 from bebi103.stan import disable_logging as be_quiet_stan
 from bebi103.stan import check_all_diagnostics
-from srep.data_loader import load_FISH_by_promoter
+from srep.utils import load_FISH_by_promoter
 
 repo = Repo("./", search_parent_directories=True)
 # repo_rootdir holds the absolute path to the top-level of our repo
@@ -42,6 +42,6 @@ for gene in df_unreg['experiment'].unique():
     check_all_diagnostics(all_samples[gene])
 
 # pickle the samples. ~20 separate netcdfs, only for use together? No thanks
-outfile = open(f"{repo_rootdir}/data/stan_samples/constit_post_inf.pkl", 'wb')
+outfile = open(f"{repo_rootdir}/data/mcmc_samples/constit_post_inf.pkl", 'wb')
 pickle.dump(all_samples, outfile)
 outfile.close()

### 2020-05-17 code/figures/fig2.py

@@ -1,92 +1,0 @@
-import pickle
-from git import Repo #for directory convenience
-
-import numpy as np
-import scipy.stats as st
-import pandas as pd
-import arviz as az
-
-import matplotlib.pyplot as plt
-import bebi103.viz
-
-import srep.viz
-from srep.data_loader import load_FISH_by_promoter
-
-
-repo = Repo("./", search_parent_directories=True)
-# repo_rootdir holds the absolute path to the top-level of our repo
-repo_rootdir = repo.working_tree_dir
-
-colors = srep.viz.color_selector('constit')
-srep.viz.plotting_style()
-
-pklfile = open(f"{repo_rootdir}/data/stan_samples/constit_post_inf.pkl", 'rb')
-all_samples = pickle.load(pklfile)
-pklfile.close()
-
-df_unreg, df_energies, = load_FISH_by_promoter(("unreg", "energies"))
-
-fig, ax = plt.subplots(2, 2, figsize=(9,9))
-
-# ###########################################################################
-# POSTERIOR PREDICTIVE FOR UV5
-# ###########################################################################
-# JB's util only works w/ Bokeh, so plotting PPC in matplotlib will take
-# some extra effort. Here's a quick and dirty data + MAP plot as placeholder
-# first plot UV5 data
-uv5_counts = df_unreg[df_unreg['experiment'] == "UV5"]["mRNA_cell"]
-x_data, y_data = bebi103.viz._ecdf_vals(uv5_counts, staircase=True)
-ax[0,1].plot(x_data, y_data, label="UV5", color=colors['UV5'])
-# next add Poisson
-y_poiss = st.poisson.cdf(x_data, 18.72)
-ax[0,1].plot(x_data, y_poiss, label="Poisson", color='k')
-# then add negbinom
-# UV5 MAP: alpha = 5.35, b = 3.5, note scipy's weird parametrization
-y_nbinom = st.nbinom.cdf(x_data, 5.35, 1/4.5)
-ax[0,1].plot(x_data, y_nbinom, '--', label="Neg Binom", color='k')
-ax[0,1].legend(loc='lower right', fontsize='small')
-ax[0,1].set_xlabel('mRNA counts per cell')
-ax[0,1].set_ylabel('CDF')
-
-
-# ###########################################################################
-# 95% HPD for all 18 constitutive promoters
-# ###########################################################################
[…]

### 2020-05-17 code/figures/fig2pt1.py

@@ -1,0 +1,114 @@
+#%%
+import pickle
+from git import Repo #for directory convenience
+
+import numpy as np
+import scipy.stats as st
+import pandas as pd
+import arviz as az
+
+import matplotlib.pyplot as plt
+import bebi103.viz
+
+import srep.viz
+from srep.utils import load_FISH_by_promoter
+
+#%%
+repo = Repo("./", search_parent_directories=True)
+# repo_rootdir holds the absolute path to the top-level of our repo
+repo_rootdir = repo.working_tree_dir
+
+colors = srep.viz.color_selector('constit')
+srep.viz.plotting_style()
+
+pklfile = open(f"{repo_rootdir}/data/mcmc_samples/constit_post_inf.pkl", 'rb')
+all_samples = pickle.load(pklfile)
+pklfile.close()
+
+df_unreg, df_energies, = load_FISH_by_promoter(("unreg", "energies"))
+df_energies.sort_values('Energy (kT)', inplace=True)
+
+# ###########################################################################
+# POSTERIOR FOR UV5
+# ###########################################################################
+# what was formerly a 2x2 subplots is now 2 separate 1x2 subplots
+fig, ax = plt.subplots(1, 2, figsize=(8,4))
+alpha_samples = all_samples['UV5'].posterior.alpha.values.flatten()
+b_samples = all_samples['UV5'].posterior.b.values.flatten()
+# thin samples before plotting, excessive # of glyphs otherwise
+ax[0].plot(
+    b_samples[::3],
+    alpha_samples[::3],
+    'k.',
+    markersize=3,
+    alpha=0.2,
+    label='UV5 posterior samples')
+ax[0].set_ylabel(r'$k_i$ (bursts per mRNA lifetime)')
+ax[0].set_xlabel(r'$b$ (transcripts per burst)')
+ax[0].legend()
+
+# ###########################################################################
+# POSTERIOR PREDICTIVE FOR UV5
+# ###########################################################################
+df_UV5 = df_unreg[df_unreg["experiment"] == "UV5"]
+n_samples = (
+    all_samples["UV5"].posterior_predictive.dims["chain"]
[…]

### 2020-05-17 code/figures/fig2pt2.py

@@ -1,0 +1,100 @@
+#%%
+import pickle
+from git import Repo #for directory convenience
+
+import numpy as np
+import scipy.stats as st
+import pandas as pd
+import arviz as az
+
+import matplotlib.pyplot as plt
+import bebi103.viz
+
+import srep.viz
+from srep.utils import load_FISH_by_promoter
+
+#%%
+repo = Repo("./", search_parent_directories=True)
+# repo_rootdir holds the absolute path to the top-level of our repo
+repo_rootdir = repo.working_tree_dir
+
+colors = srep.viz.color_selector('constit')
+srep.viz.plotting_style()
+
+pklfile = open(f"{repo_rootdir}/data/mcmc_samples/constit_post_inf.pkl", 'rb')
+all_samples = pickle.load(pklfile)
+pklfile.close()
+
+df_unreg, df_energies, = load_FISH_by_promoter(("unreg", "energies"))
+df_energies.sort_values('Energy (kT)', inplace=True)
+
+#%%
+# ###########################################################################
+# 95% HPD for all 18 constitutive promoters
+# ###########################################################################
+fig, ax = plt.subplots(1, 2, figsize=(8,4.5), sharey=True)
+# setup colors, linestyles, and marker styles for all 18 promoters
+cwheel_set = (
+    'red', 'blue', 'purple',
+    'green', 'brown', 'black',
+    'yellow', 'magenta', 'white'
+    )
+cwheel = 2 * colors
+# cwheel = 2 * cwheel_set
+lwheel_nested = (9*('-',), 9*('--',))
+lwheel = [item for sublist in lwheel_nested for item in sublist]
+mwheel_nested = (9*('o',), 9*('^',))
+mwheel = [item for sublist in mwheel_nested for item in sublist]
+
+# # loop thru df, not all_samples keys, so we get deterministic order!
+for i, promoter in enumerate(df_energies.Name):
+    alpha_samples = all_samples[promoter].posterior.alpha.values.flatten()
+    b_samples = all_samples[promoter].posterior.b.values.flatten()
+    x_contour, y_contour = bebi103.viz.contour_lines_from_samples(
+        b_samples, alpha_samples, levels=0.95, smooth=0.025
+    )
+    ax[0].loglog(
+        x_contour[0],
+        y_contour[0],
+        lwheel[i], #linestyle
+        label=promoter,
[…]

### 2020-05-17 srep/viz.py

@@ -4,11 +4,13 @@ import bokeh.layouts
 import bokeh.palettes
 import seaborn as sns
 import numpy as np
+import pandas as pd
+import xarray
 import os
 import matplotlib.pyplot as plt
 import matplotlib
 
-from bebi103.viz import predictive_ecdf
+import bebi103
 
 def plotting_style(grid=True):
     """
@@ -89,69 +91,255 @@ def color_selector(style):
             'black':'#060606'}
 
     elif style.lower() == 'constit':
-        colors = {
-            'UV5': '#7AA974',
-            'WTDL20v2': '#BFD598',
-            'WT': '#DCECCB',
-            'WTDL10': '#EAC264',
-            'WTDL20': '#F3DAA9',
-            'WTDL30': '#FFEDCE',
-            'WTDR30': '#738FC1',
-            '5DL1': '#A9BFE3',
-            '5DL5': '#C9D7EE',
-            '5DL10': '#D56C55',
-            '5DL20': '#E8B19D',
-            '5DL30': '#F1D4C9',
-            '5DR1': '#AB85AC',
-            '5DR5': '#D4C2D9',
-            '5DR1v2':'#7E9D90',
-            '5DR10':'#905426',
-            '5DR20':'#a8a8a8',
-            '5DR30':'#0f0f0f'}
+        colors = [
+            '#7AA974',
+            '#ffef14',
+            '#A9BFE3',
+            '#D56C55',
+            '#65c9b2',
+            '#AB85AC',
+            '#905426',
+            '#a8a8a8',
+            '#0f0f0f']
     return colors
 
-def ppc_ecdfs(posterior_samples, df):
+
+def predictive_ecdf(
+    samples,
+    data=None,
+    diff=False,
+    percentiles=[95, 75, 50, 25],
+    color="blue",
+    data_color="orange",
+    data_staircase=True,
+    data_size=2,
+    median_lw=1,
+    x=None,
+    discrete=False,
+    ax=None,
+    pred_label=None,
+    data_label=None,
+    **kwargs,
+):
+    """Plot a predictive ECDF from samples.
+    Code drawn directly from Justin Bois' bebi103 module, only mod is
+    changing plotting backend from bokeh -> mpl.
+
+    TODO:: UPDATE DOCS, currently outdated!!
+    Parameters
+    ----------
+    samples : Numpy array or xarray, shape (n_samples, n) or xarray DataArray
+        A Numpy array containing predictive samples.
[…]

### 2020-05-17 doc/paper.tex
Additions to text on constit promoter inference

@@
                 &+ (m+1)\gamma p_{m+1,U}(t) - \gamma p_{m,U}(t).
+\label{eq:poisson_promoter_cme}
 \end{split}
@@
                 + \matr{R} \vect{p}_{m-1}(t)
-                + \gamma (m+1) \matr{I} \vect{p}_{m+1}(t)
-                - \gamma m \matr{I} \vect{p}_m(t),
+                + \gamma (m+1) \matr{I} \vect{p}_{m+1}(t),
 \label{eq:2state_rep_cme}
@@
                 + \matr{R_A} \vect{p}_{m-1}(t) +
-                \gamma (m+1) \matr{I} \vect{p}_{m+1}(t)
-                - \gamma m \matr{I} \vect{p}_m(t),
+                \gamma (m+1) \matr{I} \vect{p}_{m+1}(t),
 \label{eq:3state_rep_cme}
@@
                 + \matr{R_A} \vect{p}_{m-1}(t) +
-                \gamma (m+1) \matr{I} \vect{p}_{m+1}(t)
-                - \gamma m \matr{I} \vect{p}_m(t),
+                \gamma (m+1) \matr{I} \vect{p}_{m+1}(t),
 \end{equation}
@@
 
+\section{Finding the ``right'' model: Bayesian parameter inference and model selection}
 \subsection{Parameter inference for constitutive promoters}
-Having established that a one-state bursty model achieves a
-Goldilocks level of complexity, we test it more thoroughly on all
-the data from constitutive promoters from~\cite{Jones2014}.
-\fig{fig:constit_post} shows the results.
+
+From consideration of Fano factors in the previous section, we
+suspect that model 5 in~\fig{fig:constit_cartoons},
+a one-state bursty model of constitutive promoters,
+achieves a Goldilocks level of complexity.
+Does this stand up to closer scrutiny, namely, comparison to
+full mRNA distributions rather than simply their moments?
+We will test this thoroughly on all the data
+from constitutive promoters from Jones et.\ al.~\cite{Jones2014}.
+
+It will be instructive, however, to first consider the
+Poisson promoter, model 1 in~\fig{fig:constit_cartoons}.
+This is a convenient proxy for models 2 and 3, whose exact
+steady-state mRNA distributions would be tedious to calculate
+and we will not explicitly consider.
+But since they have Fano factors $\nu\le 1$, they can
+fare no better than the simple Poisson model.
+We will also not explicitly consider model 4 from~\fig{fig:constit_cartoons}
+since it was already thoroughly analyzed in~\cite{Razo-Mejia2020},
+and since model 5 can be viewed as a special case of it.
+
+In performing parameter inference on the FISH data
+of~\cite{Jones2014}, building a generative model is
+straightforward. The data are single-cell mRNA counts, and we
+assume different cells to be independent and identically
+distributed from the underlying generative model, which is just
+the steady-state mRNA distribution computed from the master
+equation of the underlying model.
+
+\subsubsection{Model 1: Poisson promoter}
+\mmnote{A lot of the details I've written here might be excessive for main text and serve better as an intro-by-example to Bayes in the appendix, then reduce main text to a bit more bare bones?}
+For this model the master equation of interest
+is~\eq{eq:poisson_promoter_cme} with repressor set to zero, i.e.,
+\begin{equation}
+\deriv{t}p_{m,U}(t) = rp_{m-1,U}(t) - rp_{m,U}(t)
+        + (m+1)\gamma p_{m+1,U}(t) - \gamma p_{m,U}(t),
+\end{equation}
+whose steady-state distribution is Poisson, given by
+\begin{equation}
+p_{m,U} = \frac{\lambda^m e^{-\lambda}}{m!}
+\end{equation}
+with $\lambda=r/\gamma$. This steady state distribution defines our
+likelihood $p(m\mid\lambda)$, i.e.,
+\begin{equation}
+p(m\mid\lambda) = \frac{\lambda^m e^{-\lambda}}{m!}
+\label{eq:poisson_inference010}
+\end{equation}
+is the probability of observing
+a single cell with $m$ mRNAs given a known value of $\lambda$.
+
+With our likelihood in hand, we can invoke Bayes theorem to write down
+the posterior distribution
+on our model parameter $\lambda$, given by
+\begin{equation}
+p(\lambda\mid D) \propto p(D\mid\lambda) p(\lambda),
+\end{equation}
+where our data $D=\{m_1, m_2,\dots, m_N\}$ is a list of mRNA counts
+over a population of $N$ identical cells.
+As is standard we have neglected the factor $p(D)$ on the right
+hand side since it is independent of $\lambda$ and serves only as
+a normalization factor.
+Since we assume each cell's mRNA count is independent of others,
+the likelihood is simply a product of the
+single cell likelihoods given by \eq{eq:poisson_inference010} above, so
+\begin{equation}
+p(D\mid\lambda) = \prod_{k=1}^N \frac{\lambda^{m_k}e^{-\lambda}}{m_k!}.
+\end{equation}
+This is often notated simply as
+\begin{equation}
+D\mid\lambda \sim \text{Poisson}(\lambda),
+\end{equation}
+which is read as the data $D$, conditioned on a value of
+$\lambda$, is Poisson distributed with mean $\lambda$.
+The $\sim$ operator literally denotes ``is distributed according to.''
+
+To proceed we need to specify a prior. In this case we are extremely data-rich,
+as the dataset from Jones et.\ al~\cite{Jones2014} has of order
+1000-3000 single-cell measurements for each promoter,
+so our choice of prior matters little here,
+as long as it does not exclude the true value of $\lambda$.
+A convenient choice for our Poisson likelihood is its conjugate prior,
+the gamma distribution. Although they only exist for
+certain special likelihoods, conjugate priors have the nice
+feature that the posterior takes the same functional form as the
+prior, with updated parameters. This makes calculations
+analytically tractable and also offers a nice interpretation of
+the inference procedure as updating our knowledge about the model
+parameters.
+
+Putting a gamma prior on $\lambda$ introduces two new parameters
+$\alpha$ and $\beta$ which parametrize the gamma distribution
+itself, which we use to encode the range of $\lambda$ values we
+view as reasonable. Recall $\lambda$ is the mean steady-state
+mRNA count per cell, which \textit{a priori} could plausibly be
+anywhere from 0 to a few hundred. $\alpha=1$ and $\beta=1/50$
+achieve this, since the gamma distribution is strictly positive
+with mean $\alpha/\beta$ and standard deviation
+$\sqrt{\alpha}/\beta$. To be explicit, then, our prior is
+\begin{equation}
+\lambda\mid\alpha,\beta \sim \text{Gamma}(\alpha, \beta)
+\end{equation}
+
+As an aside, note that if we did not know that our prior was a
+conjugate prior, we could still write down our posterior
+distribution from its definition as
+\begin{equation}
+p(\lambda\mid D,\alpha,\beta)
+\propto p(D\mid\lambda) p(\lambda \mid\alpha,\beta)
+\propto \left(\prod_{k=1}^N \frac{\lambda^{m_k}e^{-\lambda}}{m_k!}\right)
+        \frac{\beta}{\Gamma(\alpha)}(\beta\lambda)^{\alpha-1} e^{-\beta\lambda}
+.
+\end{equation}
+Without foreknowledge that this in fact reduces to a gamma distribution,
+this expression might appear rather inscrutable. When conjugate
+priors are unavailable for the likelihood of interest - which is
+almost always the case for models with $>1$ model parameter -
+this inscrutability is the norm, and making sense of posteriors
+analytically is almost always impossible. Fortunately, MCMC
+sampling provides us a powerful method of constructing posteriors
+numerically which we will make use of extensively.
+
+Since we did use a conjugate prior, we may simply look up our
+posterior in any standard table of conjugate priors-likelihood
+pairs, from which we find that
+\begin{equation}
+\lambda\mid D,\alpha,\beta
+\sim \text{Gamma}\left(\alpha + \bar{m}N, \beta + N\right),
+\end{equation}
+where we defined the sample mean $\bar{m} = \frac{1}{N}\sum_k m_k$
+for notational convenience.
+A glance at the FISH data from~\cite{Jones2014} reveals that $N$ is
+$\mathcal{O}(10^3)$ and $\langle m\rangle \gtrsim 0.1$ for all
+constitutive strains in~\cite{Jones2014}, so
+$\bar{m}N \gtrsim 10^2$. Therefore as we suspected,
+our prior parameters are completely overwhelmed by the data.
+In fact, $\bar{m}N$ and $N$ are so large that we can,
+to an excellent approximation, ignore the $\alpha$ and $\beta$ dependence and
+approximate the gamma distribution as a Gaussian with mean $\bar{m}$ and
+standard deviation $\sqrt{\bar{m}/N}$, giving
+\begin{equation}
+\lambda\mid D
+\sim \text{Gamma}\left(\alpha + \bar{m}N, \beta + N\right)
+\approx \text{Normal}\left(\bar{m}, \sqrt{\frac{\bar{m}}{N}}\right).
+\end{equation}
+As an example with real numbers, for the \textit{lacUV5} promoter,
+Jones et.\ al~\cite{Jones2014} measured 2648 cells with an average
+mRNA count per cell of $\bar{m} \approx 18.7$. In this case then,
+our posterior is
+\begin{equation}
+\lambda\mid D_{UV5}
+\sim \text{Normal}\left(18.7, 0.08\right),
+\end{equation}
+which suggests we have inferred our model's one parameter to a precision of order 1\%.
+
+This is not wrong, but it is not the full story.
+The model's posterior is tightly constrained, but is it a good generative model?
+In other words, does the model generate data that ``look like'' our actual data, and is it therefore plausible that the model captures the important features of the data generating process?
+This intuitive notion can be codified with \textit{posterior predictive checks}, or PPCs, and we will see that this simple Poisson model fails badly.
 \begin{figure}%[h!]
@@
 \end{figure}
+
+\mmnote{Outline:
+\begin{itemize}
+\item Poisson model. Write Bayes, briefly explain conjugacy and it's
+convenience. immediately write down posterior, make Gaussian
+approx. Explain PPC sampling (actually PP dist is analytically
+known but you still have to draw RNGs from it. Also this obscures
+the intuitive meaning of PPC sampling from posterior and
+generating synthetic data, so I think it's more useful to just
+generate PP samples from posterior draws, this gives a more
+intuitive apples-to-apples comparison w/ model 5).
+\item Bursty model. Write Bayes, sketch story of NB likelihood, sketch choice
+of priors, show~\fig{fig:uv5_post}, see bursty $\gg$ Poisson model.
+\item All 18 promoters. Hey the model still works on things other than
+UV5, cool. Interesting and somewhat counterintuitive scaling b/w
+burst rate and binding energy. Puzzle in comparing w/ Chong2014
+supercoiling model. Are we missing important things?? Unclear, we
+leave as open question. Good enough for our purposes: posterior is identifiable, PPC is great, and of the models we've thought of it is unique in satisfying both.
+\end{itemize}
+}
 
@@
 \subsection{Transcription factor kinetics can be inferred from FISH measurements}
+\mmnote{Outline:
+\begin{itemize}
+\item Full model.
+Write Bayes, handwave 2F1 dist story in limits of weak, strong,
+and intermediate repression.
+\item Hey look it mostly works.
+With full model, 9D model is well identified, though most of the
+individual experiments unsurprisingly are not: only the ratio of
+repressor rates is identifiable. PPC is not perfect, it kinda
+misses some rep/op pairs but overall it's surprisingly predictive
+considering how strong were the assumptions that went into it.
+\item Do we chalk this up to experimental imperfections or is there
+interesting biology to be found in the disagreements?
+I'm inclined towards the former. Even just catching cells in
+truly apples-to-apples growth phases, or matching up parameters
+of imaging sessions weeks or months apart, is hardly trivial.
+\item The alternative is to refine our model and/or add on
+complexity, and it's not at all clear what additions to make to
+the model, especially how to produce the long tail of repressed
+strains over UV5. I see no way to get that without getting into
+complicated time-correlation/history effects, where
+transcription...rebounds?? After being repressed?? Sounds odd.
+\item Comparison between my inferred rates, equilibrium binding E, and single-molecule measurements.
+Not quite as slam-dunk as I'd hoped because of $\gamma$ dependence, but rates are w/in factor of 2 or better, seems quite good really!
+With exception of one Oid data data point, free energies are w/in about 0.5 kT also.
+That's not massively larger than the repeatability threshold of fold-change measurements, at maybe 0.1 to 0.3 kT.
+\item Surprising that so many model parameters can be inferred from just a few distributions that, by eye, don't have strong features!
+Does require some assumptions, e.g., about equivalence of rates across experiments, but that seems quite reasonable.
+\end{itemize}
+}
+
 Figure~3 presents inferred transcription factor rates, compares
@@
 (A) depicts the cartoon of simple repression for which we infer parameters.
+\mmnote{
+Not sure panel (A) is necessary, we've already shown it in at least
+2 other figs. We could instead show some alternative slices of 9D posterior
+as a companion to (B), or something totally different\dots??}
 (B) shows 50\% and 95\% contours of several 2D slices of the 9D posterior
@@
 \appendix
+
+\section{Bayesian inference}
+\mmnote{
+\begin{itemize}
+\item Bayes theorem.
+Elaborate on posterior as plausibility over
+space of model parameters.
+\item Choosing likelihoods.
+Easy here, just the steady-state distributions of our master
+equations, assuming iid single-cell measurements.
+\item Choosing priors.
+Not critical here because we are data-rich and nonhierarchical,
+as long as we don't exclude ground truth. In other applications,
+prior can matter a lot. This is a feature, not a bug: it's
+revealing hidden structure that was implied by your modeling
+assumptions and you just didn't realize it. Predictive checks to
+verify you're not excluding any plausible values.
+\item Marginalization/expectations.
+Analytically hard, trivial with sampling.
+\item MCMC.
+Lot's of fun here, and deep analogies to stat mech and
+Hamiltonian mechanics worth exploring. In a perfect world, you'd
+wave a magic wand and generate independent draws from target
+posterior. But we don't know how. Next best: generate a Markov
+chain of $N$ \textit{correlated} samples, estimate an
+autocorrelation time $\tau$, then you've got roughly $N/\tau$
+effective samples. Obviously with some assumptions.
+\item Posterior predictive checks.
+So the posterior is fine, but is the model actually not wrong?
+PPC is essential to verify that your model is a plausible
+generative process for the data.
+\item Discuss identifiability, degeneracy, and failed PPCs in main text. Better thru example anyway.
+\end{itemize}
+}
 

### 2020-05-17 code/figures/fig2pt1.py

@@ -69,7 +69,7 @@ for i, draw in enumerate(poiss_post_draws):
 # now plot PPC from Poisson samples
 srep.viz.predictive_ecdf(
     poiss_ppc_draws,
-    color='purple',
+    color='green',
     percentiles=ptiles,
     discrete=True,
     ax=ax[1],
@@ -88,7 +88,7 @@ srep.viz.predictive_ecdf(
     ax=ax[1],
     pred_label='Model 5 (N. Binom) PPC',
     data_label='UV5 data, Jones et. al.',
-    data_color='red',
+    data_color='orange',
     data_size=1 #linewidth
     )
 # first plot UV5 data

### 2020-05-18 doc/paper.tex
Added more text on constitutive inference, plus redid bibfile

@@
 
-This is not wrong, but it is not the full story.
-The model's posterior is tightly constrained, but is it a good generative model?
-In other words, does the model generate data that ``look like'' our actual data, and is it therefore plausible that the model captures the important features of the data generating process?
-This intuitive notion can be codified with \textit{posterior predictive checks}, or PPCs, and we will see that this simple Poisson model fails badly.
+This is not wrong, but it is not the full story. The model's
+posterior is tightly constrained, but is it a good generative
+model? In other words, does the model generate data that look similar
+to our actual data, and is it therefore plausible that the
+model captures the important features of the data generating
+process? This intuitive notion can be codified with
+\textit{posterior predictive checks}, or PPCs, and we will see
+that this simple Poisson model fails badly.
+
+The intuitive idea of posterior predictive checks is simple:
+\mmnote{Not sure if all this stuff is clearer by example,
+but then does it confuse the reader in generalizing to other models?}
+\begin{enumerate}
+\item Make a random draw of the model parameter $\lambda$
+from the posterior distribution.
+\item Plug that draw into the likelihood and generate
+a synthetic dataset $\{m_k\}$ conditioned on $\lambda$.
+\item Repeat many times.
+\end{enumerate}
+More formally, the posterior predictive distribution can be thought of
+as the distribution of future yet-to-be-observed data,
+conditioned on the data we have already observed.
+Clearly if those data appear quite different,
+the model has a problem. Put another way, if we suppose
+the generative model is true, then the synthetic datasets
+we generate should resemble the actual observed data, and if not,
+it suggests the model is missing important features.
+All the data we consider in this work are 1D (distributions of mRNA counts
+over a population) so ECDFs are an excellent visual means of comparing
+synthetic and observed datasets. In general for higher dimensional datasets,
+much of the challenge is in merely designed good visualizations
+that can actually show if synthetic and observed data are similar or not.
+
+For our example Poisson promoter model then, we merely draw many
+random numbers, say 1000, from the Gaussian posterior. For each one
+of those draws, we generate a dataset from the likelihood,
+i.e., we draw 2648 (the number of observed cells in the actual dataset)
+Poisson-distributed numbers for each of the 1000 posterior draws,
+for a total of 2648000 samples from the posterior predictive distribution.
+
+To compare so many samples with the actual observed data,
+one excellent visualization for 1D data is ECDFs of the quantiles,
+as shown for our Poisson model in~\fig{fig:uv5_post}.
+In this example, the median for each possible mRNA count is shown
+as a dark green line, while shaded bands of progressively lighter green
+contain 25\%, 50\%, 75\%, and 95\% of the posterior predictive samples.
+Plotting quantiles in this way gives us a sense of the range of data
+we might consider plausible, under the assumption that the model is true.
+In this case it is quite obvious that the observed data, plotted in orange,
+could not plausibly come from this Poisson generative model.
+The other model with a PPC plotted in~\fig{fig:uv5_post},
+bursty model 5 with a negative binomial steady-state mRNA distribution,
+looks like a good candidate by eye. We cover it in more detail later.
+
 \begin{figure}%[h!]
@@
 
-\mmnote{Outline:
+\mmnote{Outline, still to cover:
 \begin{itemize}
-\item Poisson model. Write Bayes, briefly explain conjugacy and it's
-convenience. immediately write down posterior, make Gaussian
-approx. Explain PPC sampling (actually PP dist is analytically
-known but you still have to draw RNGs from it. Also this obscures
-the intuitive meaning of PPC sampling from posterior and
-generating synthetic data, so I think it's more useful to just
-generate PP samples from posterior draws, this gives a more
-intuitive apples-to-apples comparison w/ model 5).
 \item Bursty model. Write Bayes, sketch story of NB likelihood, sketch choice
@@
 \bibliographystyle{abbrv}
-\bibliography{./library}
+\bibliography{./simple_rep}
 

### 2020-05-18 code/figures/fig2pt2.py

@@ -87,7 +87,7 @@ for i, promoter in enumerate(df_energies.Name):
 # add a guideline for the eye for the predicted log(k_i) ~ - binding E
 guide_x = np.linspace(-5.5,-2)
 guide_y = np.exp(-guide_x)/50
-ax[1].plot(guide_x, guide_y, 'k--')
+ax[1].plot(guide_x, guide_y, 'k--', label='predicted \n scaling')
 
 ax[1].set_xlabel(r'Binding energy $(k_BT)$')
 ax[1].set_yscale("log")

### 2020-05-19 doc/abstract.tex
initial splitting of document into several files

% !TEX root = ./busty_transcription.tex
\begin{abstract}
Abstract here.
\end{abstract}

### 2020-05-19 doc/appendix_01_bayesian_inference.tex
initial splitting of document into several files

--- 2020-05-17 doc/paper.tex
@@
-\documentclass[12pt]{article}%{amsart}
-\usepackage[top = 1.0in, bottom = 1.0in, left = 1.0in, right = 1.0in]{geometry}
-% See geometry.pdf to learn the layout options. There are lots.
-%\geometry{letterpaper}               % ... or a4paper or a5paper or ...
-%\geometry{landscape}             % Activate for for rotated page geometry
-\usepackage[parfill]{parskip}    % Activate to begin paragraphs with an empty line rather than an indent
-\usepackage{graphicx}
-\usepackage{amssymb}
-%\usepackage{epstopdf}
-% \usepackage{caption}
-\usepackage{amsmath}
-% \usepackage{longtable}
-% \usepackage{tabu}
-% \usepackage{accents} %to get undertildes for vec & mat
-\usepackage{color} % for colored text
-% \usepackage[normalem]{ulem} % for editing: striking out text using 'sout' command
-
-% \newcommand\mytablefigwidth{0.35\textwidth}
-
-% allows to use .ai files directly w/o resaving as pdf
-\DeclareGraphicsRule{.ai}{pdf}{.ai}{}
-
-% Handy math macros!
-\newcommand{\vect}[1]{\vec{#1}}
-\newcommand{\matr}[1]{\mathbf{#1}}
-\newcommand{\rate}[3]{{#1}_{#2}^{#3}}
-\newcommand{\mmnote}[1]{\textcolor{cyan}{(MM:~#1)}}
-
-% derivative macros. these are sneaky. the [{}] is an empty default 1st arg
-% usage: provide 1 arg: \deriv{x} gives d/dx
-% provide 2 args like so: \deriv[f]{x} gives df/dx
-\newcommand{\deriv}[2][{}]{\frac{d #1}{d #2}}
-\newcommand{\pderiv}[2][{}]{\frac{\partial #1}{\partial #2}}
-\newcommand{\psecderiv}[2][{}]{\frac{\partial{^2} #1}{\partial #2{^2}}}
-
-%%%%% Referencing macros %%%%%
-\newcommand{\fig}[1]{Figure~\ref{#1}}
-\newcommand{\tref}[1]{Table~\ref{#1}}
-\newcommand{\eq}[1]{Eq.~(\ref{#1})}
-\newcommand{\eqrange}[2]{Eqs.~(\ref{#1}-\ref{#2})}
-
-%%%%%
-\begin{document}
-
-\title{Simple repression draft}
-
-\maketitle
-
-\section{Introduction}
-\mmnote{Key ideas, no particular order, that I haven't written down before:
+% !TEX root = ./busty_transcription.tex
+\section{Bayesian inference}
+\mmnote{
 \begin{itemize}
-\item Our goal is to build phenomenological models of input-output
-functions of genetic circuits. More precisely, we want phenomenology
-in the sense of coarse-graining away atomistic detail, but still
-retaining biophysical meaning, e.g., we don't want to coarse-grain
-as far as Hill functions. Why not? We are motivated by studies
-like~\cite{Razo-Mejia2020}, for which Hill functions are
-insufficient. Such work requires a quantitative theory of how
-biophysical changes at the molecular level propagate to
-input-output functions at the genetic circuit level.
-We want concepts, not mere facts.
-In particular a key question is: at this level of
-coarse graining, what microscopic details do I need to explicitly
-model, and how do we figure that out? For example, do I need to
-worry about all or even any of the steps that individual RNAPs go
-through each time they make a transcript? Turning the question
-around, can we see any imprint of those processes in the
-available data? If the answer is no, then those processes are
-irrelevant for our purposes. Forward modeling and inverse
-(statistical inferential) modeling can complement each other
-beautifully here.
-\end{itemize}}
-
-\fig{fig1:means_cartoons}A shows the qualitative picture of
-simple repression. An operator, the binding site on the DNA for a
-repressor protein, may be found occupied by a repressor, in which
-case transcription is blocked from occurring. Or it may be found
-unoccupied, in which case RNA polymerase (RNAP) may bind and
-transcription can proceed.
-
-The key assumption we make is that binding of repressor and RNAP in the promoter region of interest is exclusive, meaning that one or the other may bind, but never may both be simultaneously bound.
-It is often imagined that when the repressor is bound to its operator,
-RNAP is sterically blocked from binding to its promoter sequence.
-Current evidence suggests this is sometimes, but not
-always the case~\mmnote{Cite that fig, from ??Nathan or Bill or
-Rob??, showing locations of repressor binding sites far upstream
-of promoter}, and it remains an interesting open question precisely
-how a repressor bound far upstream is able to repress transription.
-Suggestions include ``action-at-a-distance'' mediated by kinks in
-the DNA, formed when the repressor is bound, that prevent RNAP binding.
-Nevertheless, our modeling in this work is sufficiently
-coarse-grained that we simply assume exclusive binding and need
-not worry about such precise details.
-
-\mmnote{Add a notation aside somewhere that all association rates are written as zeroth-order rates.
-In other words, the concentration of the molecule is hidden inside, so for instance, $k_R^+$ is proportional to repressor copy number.}
-
-\section{Means}
-\subsection{Fold-changes are indistinguishable across models}
-As a first stop on our search for the ``right'' model of simple
-repression, let us consider what we can learn from theory and
-measurements of fold-change in mean expression. The fold-change is defined as
-\begin{equation}
-\text{fold-change}
-= \frac{\langle \text{gene expression with repressor present} \rangle}
-        {\langle \text{gene expression with repressor absent} \rangle}
-= \frac{\langle m\rangle(R\ne 0)}{\langle m\rangle(R = 0)}
-\label{eq:fc_def}
-\end{equation}
-where angle brackets denote the averge over a population of cells
-and mean mRNA $\langle m\rangle$ is viewed as a function of
-repressor copy number $R$.
-The second
-equality in~\ref{eq:fc_def} follows from assuming that the
-translation efficiency, i.e., the number of proteins translated
-per mRNA, is the same in both conditions.
-In other words, we assume that mean protein level is proportional
-to mean mRNA level, and that the proportionality constant is the
-same in both conditions and therefore cancels out in the ratio.
-This is reasonable since the cells in the two conditions are
-identical except for the presence of the transcription factor,
-and it is difficult to imagine the transcription factor having
-any direct effect on translation.
-
-Fold-change has proven a very convenient observable in past
-work~\cite{Garcia2011a, Razo-Mejia2018, Chure2019}.
-Part of its utility in dissecting transcriptional regulation is
-its ratiometric nature, which is always easier than making an
-absolute measurement. Also, by measuring otherwise identical cells
-with and without a transcription factor present, any biological
-noise common to both conditions can be made to cancel away.
-In addition, the presence of different operator sites is known to
-change the basal expression level in the absence of
-repressor~\mmnote{look up cite}.
-Computing fold-changes neatly cancels this effect as well.
-
-\fig{fig1:means_cartoons} depicts a smorgasbord
-of mathematized cartoons for simple repression that have appeared
-in previous literature. For each cartoon, we calculate the
-fold-change in mean gene expression as predicted by that model,
-deferring some algebraic details to the appendix.
-What we will find is that all cartoons collapse to a single
-master curve, shown in \fig{fig1:means_cartoons}c, which contains
-just two parameters. We label the parameters $\Delta F_R$, an
-effective free energy parametrizing the repressor-DNA
-interaction, and $\rho$, which subsumes all details of
-transcription in the absence of repressors.
-We will offer some intuition for why this master curve exists and discuss
-why it is likely impossible to discriminate ``right'' from ``wrong''
-cartoons given only measurements of fold-changes in expression.
-
-\subsubsection{Equilibrium, 2-state}
-In this simplest model, the promoter is idealized as existing in
-one of two states, either repressor bound or repressor unbound.
-The rate of transcription, and therefore the level of gene
-expression, is assumed to be proportional to the fraction of time
-spent in the repressor unbound state.
-Mathematically, this amounts to the statement that the mean mRNA level is
-given by
-\begin{equation}
-\langle m \rangle = \frac{r}{\gamma}
-        \left(1 + \frac{R}{N_{NS}} e^{-\beta\Delta\epsilon_R}\right)^{-1},
-\end{equation}
-where $r$ is the transcription rate from the repressor unbound
-state, $\gamma$ is the mRNA degradation rate, $R$ is repressor
-copy number, $N_{NS}$ is the number of nonspecific binding sites
-in the genome (where repressors spend most of their time when not
-bound to the operator, their specifc binding site),
-$\beta=1/k_BT$, and $\Delta\epsilon_R$ is the binding energy of a
-repressor to its operator site. The derivation of this result is deferred
-to the appendix.
-
-The fold-change is easily found as the ratio of mean mRNA with
-and without repressor. This gives
-\begin{equation}
-\text{fold-change}
-= \left(1 + \frac{R}{N_{NS}} e^{-\beta\Delta\epsilon_R}\right)^{-1},
-\end{equation}
-which clearly matches the form of the master curve
-in~\eq{fig1:means_cartoons} with $\rho=1$.
-
-In fact it was noted in~\cite{Chure2019} that this two-state
-model can be viewed as the coarse-graining of any equilibrium
-promoter model in which no transcriptionally active states have
-transcription factor bound. Put differently, there is no overlap
-between TF bound states and transcriptionally active states. We
-will see this explicitly in the 3-state equilibrium model below,
-but perhaps surprising is that something analogous carries over
-even to the nonequilibrium models we consider later.
-
-\subsubsection{Equilibrium, 3-state}
-Compared to the previous model, here we fine-grain the repressor
-unbound state into two separate states: empty, and RNAP bound.
-This picture was used in~\cite{Garcia2011a} as we use it here,
-and in~\cite{Razo-Mejia2018} and~\cite{Chure2019} it was generalized
-to incorporate small-molecule inducers that bind the repressor.
-Interestingly, the effect of this is simply to renormalize $R$
-from the total number of repressors to a smaller effective number,
-roughly (though not exactly) the number which are unbound by inducers.
-
-The mean mRNA, as derived in the appendix from a similar
-enumeration of states and weights as the previous model, is
-\begin{equation}
-\langle m \rangle = \frac{r}{\gamma}
-\frac{\frac{P}{N_{NS}} e^{-\beta\Delta\epsilon_P}}
-        {
-        1 + \frac{R}{N_{NS}} e^{-\beta\Delta\epsilon_R}
-        + \frac{P}{N_{NS}} e^{-\beta\Delta\epsilon_P}
-        },
-\end{equation}
-where the new variables are $\Delta\epsilon_P$, the difference in
-RNAP binding energy to its specific site (the promoter) relative
-to an average nonspecific background site, and the RNAP copy
-number, $P$. The fold-change again follows immediately as
-\begin{align}
-\text{fold-change}
-&= \frac{\frac{P}{N_{NS}} e^{-\beta\Delta\epsilon_P}}
-        {
-        1 + \frac{R}{N_{NS}} e^{-\beta\Delta\epsilon_R}
-        + \frac{P}{N_{NS}} e^{-\beta\Delta\epsilon_P}
+\item Bayes theorem.
+Elaborate on posterior as plausibility over
+space of model parameters.
+\item Choosing likelihoods.
+Easy here, just the steady-state distributions of our master
+equations, assuming iid single-cell measurements.
+\item Choosing priors.
+Not critical here because we are data-rich and nonhierarchical,
+as long as we don't exclude ground truth. In other applications,
+prior can matter a lot. This is a feature, not a bug: it's
+revealing hidden structure that was implied by your modeling
+assumptions and you just didn't realize it. Predictive checks to
+verify you're not excluding any plausible values.
+\item Marginalization/expectations.
+Analytically hard, trivial with sampling.
+\item MCMC.
+Lot's of fun here, and deep analogies to stat mech and
+Hamiltonian mechanics worth exploring. In a perfect world, you'd
+wave a magic wand and generate independent draws from target
+posterior. But we don't know how. Next best: generate a Markov
+chain of $N$ \textit{correlated} samples, estimate an
+autocorrelation time $\tau$, then you've got roughly $N/\tau$
+effective samples. Obviously with some assumptions.
+\item Posterior predictive checks.
+So the posterior is fine, but is the model actually not wrong?
+PPC is essential to verify that your model is a plausible
+generative process for the data.
+\item Discuss identifiability, degeneracy, and failed PPCs in main text. Better thru example anyway.
+\end{itemize}
 }
-\frac{1 + \frac{P}{N_{NS}} e^{-\beta\Delta\epsilon_P}}
-        {\frac{P}{N_{NS}} e^{-\beta\Delta\epsilon_P}}
-\\
-&= \left(
-1 + \frac{\frac{R}{N_{NS}} e^{-\beta\Delta\epsilon_R}}
-        {1 + \frac{P}{N_{NS}} e^{-\beta\Delta\epsilon_P}}
-\right)^{-1}
-\\
-&= (1 + \exp(-\Delta F_R - \log\rho))^{-1},
-\end{align}
-with $\Delta F_R = \beta\Delta\epsilon_R - \log(R/N_{NS})$
-and $\rho = 1 + \frac{P}{N_{NS}}\mathrm{e}^{-\beta\Delta\varepsilon_P}$
-as shown in~\fig{fig1:means_cartoons}.
-
-\begin{figure}%[h!]
-\centering
-\includegraphics[width=\textwidth]{../figures/fig1/fig1.ai}
-\caption{
-\textbf{
-An overview of the simple repression motif at the level of means.}
-(A) depicts the qualitative biological picture of simple repression.
-(B) shows a variety of possible mathematized cartoons of simple
-repression, along with the effective parameter $\rho$
-which subsumes all regulatory details
-of the architecture excluding the repressor.
-(C) shows the ``master curve'' to which all cartoons in (B) collapse.
-}
-\label{fig1:means_cartoons}
-\end{figure}
-
-\subsubsection{Nonequilibrium model 1 - Poisson promoter}
-For our first kinetic model, we consider the simplest possible
-picture with only two states, repressor bound and unbound.
-This is exactly the model used for the main results of~\cite{Jones2014}.
-In this picture, repressor association and dissociation rates
-from its operator site, $k_R^+$ and $k_R^-$, respectively, govern
-transitions between the two states. When the system is in the
-unbound state, transcriptions initiate at rate $r$, which
-represents a coarse-graining of all the downstream processes into
-a single effective rate. mRNA is again degraded at rate $\gamma$.
-
-Let $p_{m,R}(t)$ denote the joint probability of finding the
-system in the repressor bound state with $m$ mRNA molecules
-present at time $t$, and similarly define $p_{m,U}(t)$ for the
-repressor unbound state. 
-This model is governed by a master equation giving the time
-evolution of $p_{m,R}(t)$ and $p_{m,U}(t)$, which is
-\begin{align}
-\begin{split}
-\deriv{t}p_{m,R}(t) =& - k_R^- p_{m,R}(t) + k_R^+ p_{m,U}(t)
-                + (m+1)\gamma p_{m+1,R}(t) - \gamma p_{m,R}(t)
-\\
-\deriv{t}p_{m,U}(t) =&\; k_R^- p_{m,R}(t) - k_R^+ p_{m,U}(t)
-                        + rp_{m-1,U}(t) - rp_{m,U}(t)
-                \\
-                &+ (m+1)\gamma p_{m+1,U}(t) - \gamma p_{m,U}(t).
-\end{split}
-\end{align}
-In each equation, the first two terms describe transitions
-between promoter states due to repressors unbinding and binding,
-respectively. The final two terms describe degradation of mRNA,
-decreasing the copy number by one, and the terms with coefficient
-$r$ describe transcription initiation increasing the mRNA copy
-number by one.
-
-We can greatly simplify the notation, which will be especially
-useful for the more complicated models yet to come, by
-re-expressing the master equation in vector form. The promoter
-states are collected into a vector and the rate constants are
-collected into matrices as
-\begin{equation}
-\vect{p}_m = \begin{pmatrix} p_{m,R} \\ p_{m,U} \end{pmatrix},\
-\matr{K} = \begin{pmatrix} -k_R^- & k_R^+ \\ k_R^- & -k_R^+ \end{pmatrix},\
-\matr{R} = \begin{pmatrix} 0 & 0 \\ 0 & r \end{pmatrix},\
-\label{eq:2state_cme_matrices}
-\end{equation}
-so that the master equation may be condensed as
-\begin{equation}
-\deriv{t}\vect{p}_m(t) =
-\left( \matr{K} - \matr{R} - \gamma m \matr{I} \right) \vect{p}_m(t)
-                + \matr{R} \vect{p}_{m-1}(t)
-                + \gamma (m+1) \matr{I} \vect{p}_{m+1}(t)
-                - \gamma m \matr{I} \vect{p}_m(t),
-\label{eq:2state_rep_cme}
-\end{equation}
-where $\matr{I}$ is the identity matrix.
-Taking steady state by setting time derivatives to zero,
-the mean mRNA can be found to be
-\begin{equation}
-\langle m \rangle = \frac{r}{\gamma}
-        \left(1 + \frac{k_R^+}{k_R^-}\right)^{-1},
-\end{equation}
-with the algebra details again deferred to the appendix.
-Recall $k_R^+$ is proportional to the repressor copy number, so
-in computing fold-change, absence of repressor corresponds to
-$k_R^+\rightarrow0$.
-Therefore fold-change in this model is simply
-\begin{equation}
-\text{fold-change} = \left(1 + \frac{k_R^+}{k_R^-}\right)^{-1},
-\end{equation}
-again matching the master curve of~\fig{fig1:means_cartoons} with $\rho=1$.
-
-\subsubsection{Nonequilibrium model 2 - RNAP bound/unbound states}
-Analogous to the three-state equilibrium model, this model
-fine-grains the repressor unbound state of nonequilibrium model 1,
-resolving it into an empty promoter state and an RNAP-bound state.
-Note in this picture, in contrast with model 4 below,
-transcription initiation is accompanied by a promoter state
-change, in keeping with the interpretation as RNAP-bound and
-empty states: if an RNAP successfully escapes the promoter and
-proceeds to elongation of a transcript, clearly it is no longer
-bound at the promoter! Therefore another RNAP must bind before
-another transcript can be initiated.
-
-The master equation governing this model is analogous
-to~\eqrange{eq:2state_cme_matrices}{eq:2state_rep_cme} for model 1 above.
-The main subtlety arises since transcription initiation
-accompanies a promoter state change.
-This can be understood by analogy to $\matr{K}$. The
-off-diagonals and diagonals of $\matr{K}$ correspond to
-transitions arriving at or departing from, respectively, the
-promoter state of interest. If transcription initiations are
-accompanied by promoter state changes, we must have separate
-matrices for arriving and departing transcription events since
-the arriving and departing transitions have different initial
-copy number of mRNA, unlike for $\matr{K}$ where they are the same.
-The master equation for this model is
-\begin{equation}
-\deriv{t}\vect{p}_m(t) =
-\left( \matr{K} - \matr{R_D} - \gamma m \matr{I} \right) \vect{p}_m(t)
-                + \matr{R_A} \vect{p}_{m-1}(t) +
-                \gamma (m+1) \matr{I} \vect{p}_{m+1}(t)
-                - \gamma m \matr{I} \vect{p}_m(t),
-\label{eq:3state_rep_cme}
-\end{equation}
-with the state vector and promoter transition matrix defined as
-\begin{equation}
-\vect{p}_m = \begin{pmatrix} p_{m,R} \\ p_{m,E} \\ p_{m,P} \end{pmatrix},\
-\matr{K} = \begin{pmatrix} -k_R^- & k_R^+ & 0 \\
-                        k_R^- & -k_R^+ -k_P^+ & k_P^- \\
-                        0 & k_P^+ & -k_P^- 
-                \end{pmatrix},
-\label{eq:3state_cme_matrices_pt1}
-\end{equation}
-and the initiation matrices given by
-\begin{equation}
-\matr{R_A} = \begin{pmatrix}
-                0 & 0 & 0 \\ 0 & 0 & r \\ 0 & 0 & 0
-                \end{pmatrix},\
-\matr{R_D} = \begin{pmatrix}
-                0 & 0 & 0 \\ 0 & 0 & 0 \\ 0 & 0 & r
-                \end{pmatrix},
-\label{eq:3state_cme_matrices_pt2}
-\end{equation}
-The elements of $\vect{p}_m$ encode the probabilities of having
-$m$ mRNA present along with the promoter having repressor bound,
-being empty, or having RNAP bound, respectively.
-$\matr{R_A}$ describes probability flux \textit{arriving} at the state
-$\vect{p}_m$ from a state with one fewer mRNA, namely
-$\vect{p}_{m-1}$, and $\matr{R_D}$ describes probability flux
-\textit{departing} from the state $\vect{p}_m$
-for a state with one more mRNA, namely $\vect{p}_{m+1}$.
-$\matr{K}$ is closely analogous to model 1.
-
-Mean mRNA at steady state is found analogously to model 1, with the result
-\begin{equation}
-\langle m\rangle = \frac{r}{\gamma}
-        \frac{k_R^- k_P^+}
-        {k_R^- k_P^+ + k_R^- (k_P^- + r) + k_R^+ (k_P^- + r)}.
-\end{equation}
-Fold-change is again found from the ratio prescribed by~\eq{eq:fc_def},
-from which we have
-\begin{align}
-\text{fold-change}
-&=      \frac{k_R^- k_P^+}
-        {k_R^- k_P^+ + k_R^- (k_P^- + r) + k_R^+ (k_P^- + r)}
-        \frac{k_P^+ + k_P^- + r}{k_P^+}
-\\
-&=      \left(1 + \frac{k_R^+}{k_R^-}
-                \frac{k_P^- + r}{k_P^+ + k_P^- + r}
-        \right)^{-1}
-\\
-&=      \left(1 + \frac{k_R^+}{k_R^-}
-        \left(1 + \frac{k_P^+}{k_P^- + r}\right)^{-1}
-        \right)^{-1},
-\end{align}
-which follows the master curve with
-$\rho = 1 + k_P^+/(k_P^- + r)$ as claimed.
-
-\subsubsection{Nonequilibrium model 3
-- Multistep transcription initiation and escape}
-One might reasonably complain that the first two
-``nonequilibrium'' models we have considered are straw men. Their
-steady states necessarily satisfy detailed balance which is
-equivalent to thermodynamic equilibrium. Why is this the case? At
-steady state there is by definition no net probability flux in or
-out of each promoter state, but since the promoter states form a
-linear chain, there is only one way in or out of the repressor
-bound and RNAP bound states, implying each edge must actually
-have a net zero probability flux, which is the definition of
-detailed balance (usually phrased as equality of forward and
-reverse transition fluxes).
-
-Now we consider model 3 in~\eq{fig1:means_cartoons} which allows
-the possibility of true nonequilibrium steady-state fluxes
-through the promoter states.
-We point out that this model was considered previously in~\cite{Mitarai2015}
-where a comparison was made with model 1 as used in~\cite{Jones2014}.
-The authors of~\cite{Mitarai2015} argued that the additional
-complexity is essential to properly account for the noise in the
-mRNA distribution. We will weigh in on both models later when we
-consider observables beyond fold-change.
-
-The master equation governing this model is
-identical in form to model 2 above, namely
-\begin{equation}
-\deriv{t}\vect{p}_m(t) =
-\left( \matr{K} - \matr{R_D} - \gamma m \matr{I} \right) \vect{p}_m(t)
-                + \matr{R_A} \vect{p}_{m-1}(t) +
-                \gamma (m+1) \matr{I} \vect{p}_{m+1}(t)
-                - \gamma m \matr{I} \vect{p}_m(t),
-\end{equation}
-but obviously with a higher-dimensional state space and different matrices.
-The state vector and promoter transition matrix are now
-\begin{equation}
-\vect{p}_m = \begin{pmatrix} p_{m,R} \\ p_{m,E} \\
-                             p_{m,C} \\ p_{m,O}\end{pmatrix},\
-\matr{K} = \begin{pmatrix} -k_R^- & k_R^+ & 0 & 0\\
-                        k_R^- & -k_R^+ -k_P^+ & k_P^- & 0 \\
-                        0 & k_P^+ & -k_P^- - k_O & 0 \\
-                        0 & 0 & k_O & 0
-                \end{pmatrix},
-\end{equation}
-with the four promoter states, in order, being repressor bound,
-empty, RNAP closed complex, and RNAP open complex.
-Besides increasing dimension by one again, the only new feature in
-$\matr{K}$ is the rate $k_O$, representing the (irreversible)
-rate of open complex formation from the closed complex.
-The initiation matrices are given by
-\begin{equation}
-\matr{R_A} = \begin{pmatrix}
-        0 & 0 & 0 & 0 \\ 0 & 0 & 0 & r \\ 0 & 0 & 0 & 0\\ 0 & 0 & 0 & 0
-                \end{pmatrix},\
-\matr{R_D} = \begin{pmatrix}
-        0 & 0 & 0 & 0 \\ 0 & 0 & 0 & 0 \\ 0 & 0 & 0 & 0\\ 0 & 0 & 0 & r
-                \end{pmatrix},
-\end{equation}
-again closely analogous to model 2.
-
-The expression for mean mRNA is substantially more complicated
-now, as worked out in the appendix where we find
-\begin{equation}
-\langle m\rangle = \frac{r}{\gamma}
-        \frac{k_R^- k_P^+ k_O}
-        {k_R^- [(k_P^+ (k_O + r) + r(k_P^- + k_O)] + k_R^+ r(k_P^- + k_O)},
-\end{equation}
-which can be simplified to
-\begin{equation}
-\langle m\rangle
-= \frac{r}{\gamma}
-\frac{\frac{k_P^+ k_O}{r(k_O + k_P^-)}}
-        {1 + \frac{k_P^+ (k_O + r)}{r(k_O + k_P^-)} + \frac{k_R^+}{k_R^-}}.
-\end{equation}
-The strategy is to isolate the terms involving the repressor, so that now the fold-change is seen to be simply
-\begin{align}
-\text{fold-change}
-&= \frac{\frac{k_P^+ k_O}{r(k_O + k_P^-)}}
-        {1 + \frac{k_P^+ (k_O + r)}{r(k_O + k_P^-)} + \frac{k_R^+}{k_R^-}}
-        \frac{1 + \frac{k_P^+ (k_O + r)}{r(k_O + k_P^-)}}
-                {\frac{k_P^+ k_O}{r(k_O + k_P^-)}}
-\\
-&= \left(
-        1 + \frac{k_R^+}{k_R^-}
-        \left(1 + \frac{k_P^+ (k_O + r)}{r(k_O + k_P^-)}\right)^{-1}
-\right)^{-1},
-\end{align}
-surprisingly reducing to the master curve once again, with
-$\rho = 1 + \frac{k_P^+ (k_O + r)}{r(k_O + k_P^-)}$.
-
-This example hints that an arbitrarily fine-grained model of
-downstream transcription steps may still be collapsed to the form
-of the master curve in~\fig{fig1:means_cartoons}B so long as the
-repressor binding is exclusive with transcriptionally active
-states. We offer this as a conjecture, and we suspect that a
-careful argument using the King-Altman diagram
-method~\cite{King1956, Hill1966} might furnish a ``proof.''
-Our focus here is not on full generality but rather to survey an
-assortment of plausible models that have been proposed in the
-literature for simple repression.
-
-\subsubsection{Nonequilibrium model 4 - ``Active'' and ``inactive'' states}
-This model is at the core of the theory in~\cite{Razo-Mejia2020}.
-At a glance the cartoon for this model may appear very similar to model 2,
-and mathematically it is, but the interpretation is rather different.
-In model 2, we interpreted the third state literally as an
-RNAP-bound promoter and modeled initiation of a transcript as
-triggering a promoter state change, making the hopefully
-uncontroversial assumption that an RNAP can only make one
-transcript at a time. In contrast, in the present model the
-promoter state does \textit{not} change when a transcript is
-initiated. So we no longer interpret these states as literally
-RNAP bound and unbound but instead as coarse-grained ``active''
-and ``inactive'' states, the details of which we leave
-unspecified for now. We will comment more on this model below
-when we discuss Fano factors of models.
-
-Mathematically this model is very similar to models 1 and 2.
-Like model 1, the matrix $R$ describing transcription initiation
-is diagonal, namely
-\begin{equation}
-\matr{R} = \begin{pmatrix}
-                0 & 0 & 0 \\ 0 & 0 & 0 \\ 0 & 0 & r
-        \end{pmatrix}.
-\end{equation}
-so the master equation takes verbatim the same form as it did for model
-1,~\eq{eq:2state_rep_cme}. Meanwhile the promoter transition matrix
-$K$ is the same as~\eq{eq:3state_cme_matrices_pt1} from model 2,
-although we relabel the rate constants from $k_P^\pm$ to $k^\pm$ to
-reiterate that these are not simply RNAP binding and unbinding rates.
-
-Carrying out the algebra, the mean mRNA can be found to be
-\begin{equation}
-\langle m\rangle = \frac{r}{\gamma}
-\frac{k_R^- k^+}
-{k_R^- k^+ + k_R^- k^- + k_R^+ k^-},
-\end{equation}
-and the fold-change readily follows,
-\begin{align}
-\text{fold-change}
-&=      \frac{k_R^- k^+}{k_R^- k^+ + k_R^- k^- + k_R^+ k^-}
-        \frac{k_R^- k^+ + k_R^- k^-}{k_R^- k^+}
-\\
-&=      \left(1 + \frac{k_R^+}{k_R^-}
-                \left(1 + \frac{k^+}{k^-}\right)^{-1}
-        \right)^{-1},
-\end{align}
-from which we see $\rho = 1 + k^+/k^-$ as shown in~\fig{fig1:means_cartoons}.
-
-\subsubsection{Nonequilibrium model 5 - Bursty promoter}
-We will have much more to say about this model in subsequent sections.
-For now we sketch it as an intuitive analog to model 1,
-with just two states, repressor bound or unbound, and transition rates
-between them of $k_R^+$ and $k_R^-$.
-In model 1, when in the unbound state, single mRNA transcripts
-are produced as a Poisson process with some characteristic rate
-$r$. This model by contrast produces, at some Poisson rate $k_i$,
-\textit{bursts} of transcripts with a geometrically distributed
-burst size and a mean burst size $b$.
-
-From this intuitive picture and by analogy to model 1, then, it should be plausible that the mean mRNA level is
-\begin{equation}
-\langle m\rangle = \frac{k_i b}{\gamma}
-        \left(1 + \frac{k_R^+}{k_R^-}\right)^{-1},
-\end{equation}
-which will turn out to be correct from a careful calculation.
-For now, we simply note that just like model 1, the fold-change becomes
-\begin{equation}
-\text{fold-change} = \left(1 + \frac{k_R^+}{k_R^-}\right)^{-1}
-\end{equation}
-with $\rho=1$ also like model 1.
-
-We will also see later how this model emerges as a natural limit of model 4.
-
-\subsubsection{Summary and discussion of results across models for fold-changes in mean expression}
-What have we learned from our exploration of the models
-in~\fig{fig1:means_cartoons} and their predictions of
-fold-changes in mean expression levels? The takeaway is the
-existence of a master curve (\fig{fig1:means_cartoons}C) to which
-the fold-change predictions of all the models collapse. This
-master curve is parametrized by only two effective parameters:
-$\Delta F_R$, which characterizes the repressor, and $\rho$,
-which characterizes all other features of the promoter
-architecture. The key assumption underpinning this result is that
-zero transcription occurs when a repressor is bound to its
-operator. Note, however, that we are agnostic about the molecular
-mechanism which achieves this;
-steric effects are the obvious mechanism, but, for instance,
-``action-at-a-distance'' mediated by kinked DNA due to
-repressors bound tens or hundreds of nucleotides upstream of a
-promoter is plausible as well.
-
-Why does the master curve of~\fig{fig1:means_cartoons}C exist at all?
-This brings to mind the deep questions posed in,
-e.g.,~\cite{Frank2013} and~\cite{Frank2014a}, asking us to search
-for general ``symmetries'' of the problem instead of specifying
-generative models of the known microscopics of a system.
-In our case, the key feature seems to be the exclusive nature of
-repressor and RNAP binding, which allows the parameter describing
-the repressor, $\Delta F_R$, to cleanly separate from all other
-details of the promoter architecture, which are encapsulated in $\rho$.
-In the graph-theoretic language which was applied to similar
-problems in~\cite{Gunawardena2012} and~\cite{Ahsendorf2014}, the
-repressor-bound state is a leaf of the graph describing
-transitions between promoter states. Arbitrary nonequilibrium
-behavior can occur on the rest of the promoter state space, but
-it may all be swept up in the effective parameter $\rho$, to
-which the repressor makes no contribution.
-
-As suggested in~\cite{Chure2019}, we believe this master curve
-should generalize to architectures with multiple repressor
-binding sites, as long as the exclusivity of transcription factor
-binding and transcription initation is maintained. The
-interpretation of $\Delta F_R$ is then of an effective free
-energy of all repressor bound states.
-In an equilibrium picture this is simply given by the log of the
-sum of Boltzmann weights of all repressor bound states, which
-looks like the log of a partition function of a subsystem.
-In a nonequilibrium picture, while we can still mathematically gather
-terms and give the resulting collection the label $\Delta F_R$,
-it is unclear if the physical interpretation as an effective free energy
-makes sense.
-As is well known, free energies cannot be assigned unambiguously
-to states out of equilibrium because the free energy change along
-a generic path traversing the state space is path dependent,
-unlike at equilibrium.
-A consequence of this is that, out of equilibrium, $\Delta F_R$
-is no longer a simple sum of Boltzmann weights. Instead it
-resembles a restricted sum of King-Altman
-diagrams~\cite{King1956, Hill1966}. Following the work of
-Hill~\cite{Hill1989}, it may yet be possible to interpret this expression
-as an effective free energy, but this remains unclear to us.
-We leave this an open problem for future work.
-
-If we relax the requirement of exclusive repressor-RNAP binding,
-one could imagine models in which repressor and RNAP doubly-bound
-states are allowed, where the repressor's effect is to
-\text{reduce} the transcription rate rather than setting it to zero.
-Our results do not strictly apply to such a model, although we
-note that if the repressor's reduction of the transcription rate
-is substantial, such a model might still be well-approximated by
-one of the models in~\fig{fig1:means_cartoons}.
-
-One may worry that our ``one curve to rule them all'' is a
-mathematical tautology. In fact we \textit{agree} with this
-criticism if $\Delta F_R$ is ``just a fitting parameter'' and cannot
-be meaningfully interpreted as a real, physical free energy.
-An analogy to Hill functions is appropriate here. One of their
-great strengths and weaknesses, depending on the use they are put
-to, is that their parameters coarse-grain many details and are
-generally not interpretable in terms of microscopic models, for
-deep reasons discussed at length in~\cite{Frank2013}.
-By contrast, our master curve claims to have the best of both worlds:
-a coarse-graining of all details besides the repressor into
-a single effective parameter $\rho$,
-while simultaneously retaining an interpretation of
-$\Delta F_R$ as a physically
-meaningful and interpretable free energy.
-Our task, then, is to prove or disprove this claim.
-
-How do we test this and probe the theory with fold-change
-measurements? There is a fundamental limitation in that the
-master curve is essentially a one-parameter function of $\Delta
-F_R + \log\rho$. Worse, there are many \textit{a priori}
-plausible microscopic mechanisms that could contribute to the
-value of $\rho$, such as RNAP binding and escape
-kinetics~\cite{DeHaseth1998, Mitarai2015}, and/or supercoiling
-accumulation and release~\cite{Chong2014, Sevier2016}, and/or,
-RNAP clusters analogous to those obvserved in
-eukaryotes~\cite{Cisse2013, Cho2016}
-and recently also observed in bacteria~\cite{Ladouceur2020}.
-Even if $\Delta F_R$ is measured to
-high precision, inferring the potential microscopic contributions
-to $\rho$, buried inside a log no less, from fold-change
-measurements seems hopeless. As a statistical inference problem
-it is entirely nonidentifiable, in the language
-of~\cite{Gelman2013}, Section 4.3.
-
-If we cannot simply infer values of $\rho$ from measurements of
-fold-change, can we perturb some of the parameters that make up
-$\rho$ and measure the change? Unfortunately we suspect this is
-off-limits experimentally: most of the potential contributors to
-$\rho$ are global processes that affect many or all genes. For
-instance, changing RNAP association rates by changing RNAP copy
-numbers, or changing supercoiling kinetics by changing
-topoisomerase copy numbers, would massively perturb the entire
-cell's physiology and confound any determination of $\rho$.
-
-One might instead imagine a bottom-up modeling approach, where we
-mathematize a model of what we hypothesize the important steps
-are and are not, use in vitro data for the steps deemed
-important, and \textit{predict} what $\rho$ should be.
-But again, because of the one-parameter nature of the master
-curve, many different models will likely make indistinguishable
-predictions, and without any way to experimentally perturb in
-vivo, there is no clear way to test whether the modeling
-assumptions are correct.
-
-In light of this, we prefer the view that parameters and rates
-are not directly comparable between cartoons
-in~\fig{fig1:means_cartoons}. Rather, parameters in the
-simpler cartoons represent coarse-grained combinations of
-parameters in the finer-grained models. For instance, by equating
-$\rho$ between any two models, one can derive various possible
-correspondences between the two models' parameters. Note that
-these correspondences are clearly not unique, since many possible
-associations could be made. It then is a choice as to what microscopic
-interpretations the model-builder prefers for the parameters in a
-particular cartoon, and as to which coarse-grainings lend
-intuition and which seem nonsensical.
-Indeed, since it remains an open question what microscopic
-features dominate $\rho$ (as suggested above, perhaps RNAP
-binding and escape kinetics~\cite{DeHaseth1998, Mitarai2015}, or
-supercoiling accumulation and release~\cite{Chong2014, Sevier2016},
-or, something more exotic like
-RNAP clusters~\cite{Cisse2013, Cho2016, Ladouceur2020}),
-we are hesitant to put too much weight on any one microscopic
-interpretation of model parameters that make up $\rho$.
-
-One possible tuning knob to probe $\rho$ that would not globally
-perturb the cell's physiology is to manipulate RNAP binding sites.
-Work such as~\cite{Kinney2010} has shown that models of
-sequence-dependent RNAP can be inferred from data, and the
-authors of~\cite{Brewster2012} showed that the model
-of~\cite{Kinney2010} has predictive power by using the model to
-\textit{design} binding sites of a desired affinity.
-But for our purposes, this begs the question: the authors
-of~\cite{Kinney2010} \textit{assumed} a particular model
-(essentially our 3-state equilibrium model but without the
-repressor), so it is unclear how or if such methods can be turned
-around to \textit{compare} different models of promoter function.
-
-We have already pointed out that the master curve
-of~\fig{fig1:means_cartoons} is essentially a one-parameter model,
-the one parameter being $\Delta F_R + \log\rho$.
-By now the reader may be alarmed as to how can we even determine
-$\Delta F_R$ and $\rho$ independently of each other, never mind
-shedding a lens on the internal structure of $\rho$ itself.
-A hint is provided by the weak promoter approximation, invoked
-repeatedly in prior studies~\cite{Bintu2005c, Garcia2011a,
-Razo-Mejia2018} of simple repression using the 3-state
-equilibrium model in~\fig{fig1:means_cartoons}.
-In that picture, the weak promoter approximation means
-$\frac{P}{N_{NS}}\exp(-\beta\Delta\epsilon_P) \ll 1$,
-meaning therefore $\rho\approx1$.
-This is suggestive, but how can we be sure that $\rho$ is not,
-for instance, actually $10^2$ and that $\Delta F_R$ hides a
-compensatory factor?
-A resolution is offered by an independent inference of $\rho$ in
-the absence of repressors. This was done in~\cite{Razo-Mejia2020}
-by fitting nonequilibrium model 4  in~\fig{fig1:means_cartoons},
-with zero repressor (looking ahead, this is equivalent to model 4
-in~\fig{fig:constit_cartoons}), to FISH data
-from~\cite{Brewster2014}. This provided a determination of $k^+$
-and $k^-$, from which their ratio is estimated to be no more than
-a few $10^{-1}$ and possibly as small as $10^{-2}$.
-
-The realization that $\rho\approx1$ to an excellent
-approximation, \textit{independent} of which model
-in~\fig{fig1:means_cartoons} one prefers, goes a long way towards
-explaining the surprising success of equilibrium models of simple
-repression. Even though our 2- and 3-state models get so many
-details of transcription wrong, it does not matter because
-fold-change is a cleverly designed ratio. Since $\rho$ subsumes
-all details except the repressor, and $\log\rho\approx0$, fitting
-these simple models to fold-change measurements can still give a
-surprisingly good estimate of repressor binding energies. So the
-ratiometric construction of fold-change fulfills its intended
-purpose of canceling out all features of the promoter
-architecture except the repressor itself. Nevertheless it is
-perhaps surprising how effectively it does so: \textit{a priori},
-one might not have expected $\rho$ to be quite so close to 1.
-
-We would also like to highlight the relevance
-of~\cite{Landman2019} here. Landman et.\ al.\ reanalyzed and
-compared in vivo and in vitro data on the lacI repressor's
-binding affinity to its various operator sequences. (The in vivo
-data was from, essentially, fitting our master curve to
-expression measurements.) They find broad agreement between the
-in vitro and in vivo values. This reinforces the suspicion that
-the equilibrium $\Delta\epsilon_R$ repressor binding energies do
-in fact represent real physical free energies.
-Again, \textit{a priori} this did not have to be the case, even
-knowing that $\rho\approx1$.
-
-In principle, if $\Delta F_R$ can be measured to sufficient
-precision, then deviations from $\rho=1$ become a
-testable matter of experiment.
-In practice, it is probably unrealistic to measure repressor
-rates $k_R^+$ or $k_R^-$ or fold-changes in expression levels
-(and hence $\Delta\epsilon_R$) precisely enough to detect the
-expected tiny deviations from $\rho=1$.
-We can estimate the
-requisite precision in $\Delta F_R$ to resolve a given
-$\Delta\rho$ by noting, since $\rho\approx1$, that
-$\log(1+\Delta\rho)\approx \Delta\rho$, so
-$\Delta(\Delta F_R) \approx \Delta\rho$.
-Suppose we are lucky and $\Delta\rho$ happens to be $\sim0.1$, on
-the high end of our range estimated above.
-A determination of $\Delta\epsilon_R/k_BT$ with an uncertainty of
-barely 0.1 was achieved in the excellent measurements
-of~\cite{Razo-Mejia2018}, so this requires a very difficult
-determination of $\Delta F_R$ for a very crude determination of $\rho$,
-which suggests, to put it lightly, this is not a
-promising path to pursue experimentally.
-It is doubtful that inference of repressor kinetic rates
-would be any easier.\footnote{
-\mmnote{
-In fact the uncertainties quoted in Hammar et.\
-al.~\cite{Hammar2014} from Elf's group are small enough to do
-this calculation, but HJ \& I think their error bars are insanely
-optimistic. Not sure I want to blatantly call them out on it though??}
-}
-
-Moving forward, we have weak evidence supporting the
-interpretation of $\Delta F_R$ as a physically real free
-energy~\cite{Landman2019} and other work casting
-doubt~\cite{Hammar2014}.
-How might we resolve the confusion?
-If there is no discriminatory power to test the theory and
-distinguish the various models with measurements of fold-changes
-in means, how do we probe the theory? Clearly to discriminate
-between the nonequilibrium models in~\fig{fig1:means_cartoons},
-we need to go beyond means to ask questions about kinetics, noise and even
-full distributions of mRNA copy numbers over a population of cells.
-If the ``one-curve-to-rule-them-all'' is more than a
-mathematical tautology, then the free energy of repressor binding
-inferred from fold-change measurements should agree with
-repressor binding and unbinding rates.
-In other words, the equilbrium and nonequilibrium definitions of
-$\Delta F_R$ must agree, meaning
-\begin{equation}
-\Delta F_R = \beta\Delta\epsilon_R - \log(R/N_{NS})
-        = - \log(k_R^+/k_R^-),
-\end{equation}
-must hold, where $\beta\Delta\epsilon_R$ is inferred from the
-master curve fit to fold-change measurements, and $k_R^+$ and
-$k_R^-$ are inferred in some orthogonal manner.
-Single molecule measurements such as from~\cite{Hammar2014} have
-directly observed these rates, and in the remainder of this work
-we explore a complementary approach: inferring repressor binding
-and unbinding rates $k_R^+$ and $k_R^-$ from FISH measurements of
-mRNA population distributions.
-
-\section{Beyond means}
-\subsection{A one-state promoter with bursting is the ``best'' model of constitutive promoters}
-Before we can tackle simple repression, we need an adequate
-phenomenological model of constitutive expression.
-The literature abounds with options from which we can choose,
-and we show several potential kinetic models for constitutive promoters
-in~\fig{fig:constit_cartoons}.
-Let us consider the suitability of each model for our purposes in turn.
-
-\begin{figure}%[h!]
-\centering
-\includegraphics[width=0.9\textwidth]{../figures/fig1.5/fig1point5.ai}
-\caption{\textbf{Constitutive promoter cartoon comparison.}
-The left column depicts various plausible cartoons for constitutive promoters.
-We also list literature references where models have been used previously.
-In model (1), transcripts are produced in a Poisson process~\cite{Jones2014}.
-Model (2) features explicit modeling of RNAP binding/unbinding
-kinetics~\cite{Phillips2015a}.
-Model (3) is a more detailed generalization of model (2),
-treating transcription initiation as a multi-step process
-proceeding through closed and open complexes~\cite{Mitarai2015}.
-Model (4) is somewhat analogous to (2) except with the precise nature
-of active and inactive states left ambiguous~\cite{Razo-Mejia2020}.
-Finally, model (5) can be viewed as a certain limit of model (4)
-in which transcripts are produced in bursts,
-and initiation of bursts is a Poisson process.
-\mmnote{Haven't found model 5 formulated quite like this in the literature,
-usually burstiness is handled like model 4, but I should probably search a bit more.}
-The right column shows the Fano factor for each model.
-Note especially the crucial diagnostic: (2) and (3) have $\nu$
-strictly below 1, while only for (4) and (5) can $\nu$ exceed 1.
-Models with Fano factors $\le 1$ cannot produce the FISH data
-observed in~\cite{Jones2014} without introducing additional
-assumptions and model complexity.
-}
-\label{fig:constit_cartoons}
-\end{figure}
-
-\subsubsection{Model 1 - Poisson promoter.}
-This is the picture from Jones et.\ al.~\cite{Jones2014}, which
-assumes that transcripts are produced as a Poisson process from a
-single promoter state. This model insists that the ``true'' mRNA
-distribution is Poisson, implying the Fano factor $\nu$ must be 1.
-In~\cite{Jones2014}, the authors carefully attribute deviations
-from Fano = 1 to intensity variability in fluorescent spots, gene
-copy number variation, and copy number fluctuations of the
-transcription machinery, e.g., RNAP itself.
-In this picture, the master equation makes no appearance,
-and all the corrections to Poisson behavior are derived
-as additive corrections to the Fano factor.
-For disproving the ``universal noise curve'' from So et
-al~\cite{So2011}, this picture was excellent. It is appealing in
-its simplicity, with only two parameters, the initiation rate $r$
-and degradation rate $\gamma$. Since $\gamma$ is independently
-known from other experiments, and the mean mRNA copy number is
-$r/\gamma$, $r$ is easily inferred from data.
-In other words, the model is not excessively complex for the data at hand.
-But for many interesting questions, for instance in the recent
-work~\mmnote{cite Manuel's preprint once posted},
-knowledge of means and variances alone is insufficient, and a
-description of the full distribution of molecular counts is necessary.
-For this we need a( slightly) more complex model than model 1
-that would allow us to incorporate the non-Poissonian features of
-constitutive promoters directly into a master equation formulation.
-
-\subsubsection{Model 2 - Two-state promoter, RNAP bound or unbound.}
-This model was considered in, e.g.,~\cite{Phillips2015a} and ~\cite{Phillips2019}.
-Here transcription initiation proceeds from the bound to the unbound state,
-reflecting the microscopic reality that an RNAP that has begun to elongate
-a transcript is no longer available at the start site to begin another.
-The problem with this picture is that the Fano factor is
-\begin{align}
-    \nu = 1 -
-        \frac{r\rate{k}{P}{+}}
-            {\left(\rate{k}{P}{+} + \rate{k}{P}{-} + r\right)
-             \left(\gamma + \rate{k}{P}{+} + \rate{k}{P}{-} + r\right)},
-\end{align}
-which is always $<1$.
-To make contact with the experimental reality of $\nu>1$,
-we will have to do, at a minimum,
-the same corrections to Poisson behavior as in model 1 above.
-So while this model adds an appealing element of microscopic
-reality, we are forced to reject it as the additional complexity
-is unable to capture the phenomenology of interest.
-Obviously the promoter state does in fact proceed through cycles
-of RNAP binding, initiating, and elongating, but it seems that
-the super-Poissonian noise in mRNA copy number we want to
-model must be governed by other features of the system.
-
-\subsubsection{Model 3 - Three-state promoter, multistep
-transcription initiation and escape.}
-How might we remedy the deficits of model 2?
-It is known~\cite{DeHaseth1998} that once RNAP initially binds
-the promoter region, a multitude of distinct steps occur
-sequentially before RNAP finally escapes into the elongation
-phase. Perhaps adding some of this mechanistic detail might
-rescue model 2? The next simplest refinement of model 2 could
-consider open complex formation and promoter escape as separate
-steps rather than as a single effective step. In other words, we
-construct model 3 by adding a single extra state to model 2, and
-we will label the two RNAP-bound states as the closed and open
-complexes, despite the true biochemical details certainly being
-more complex. The authors of~\cite{Mitarai2015}, for instance,
-considered this model, although they added an additional
-repressor bound state and did not explicitly consider the limit
-with no repressor that we analyze here. Again, our goal here is
-not a complete accounting of all the relevant biochemical detail;
-this is an exploratory search for the important features our
-model needs to include.
-
-Unfortunately, as the authors of~\cite{Mitarai2015} hint, this
-model too has Fano factor $\nu<1$. This can be seen by starting
-with their Eq.~A1 for the Fano factor of the analogous model with
-repressor and taking the limit as repressor concentration goes to
-zero. After substantial algebra, the result is
-\begin{align}
-\nu = 1 - \frac{r k_O k_P^+}{\mathcal{Z}}
-\frac{k_P^+ + k_P^- + k_O + \gamma}
-        {\mathcal{Z} + \gamma(k_P^+ + k_P^- + k_O + \gamma) + \gamma^2},
-\end{align}
-where we defined $\mathcal{Z} = r(k_O + k_P^-) + k_P^+(k_O + r)$
-for notational tidiness. This is necessarily less than 1 for
-arbitrary rate constants.
-
-In fact, we suspect \textit{any} model in which transcription
-proceeds through a multistep cycle must necessarily have $\nu<1$.
-The intuitive argument compares the waiting time distribution to
-traverse the cycle with the waiting time for a Poisson promoter
-(model 1) with the same mean time. The latter is simply an
-exponential distribution. The former is a convolution of multiple
-exponentials, and intuitively should be more peaked with a
-smaller fractional width than an exponential with the same mean.\footnote{
-This can be made more precise with a result
-from~\cite{Stewart2007}, who showed that the convolution of
-multiple gamma distributions (of which the exponential
-distribution is a special case) is to a very good approximation
-also gamma distributed. Using their Eq.~2 for the distribution of
-the convolution, with shape parameters set to 1 to
-give exponential distributions, the total waiting time
-distribution has a ratio of variance to squared mean 
-$\sigma^2/\mu^2 = \sum_i k_i^2/\left(\sum_i k_i\right)^2$,
-where the $k_i$ are the rates of the individual steps. Clearly
-this is less than 1 and therefore the total waiting time
-distribution is narrower than the corresponding exponential.} A
-less disperse waiting time distribution means transcription
-initations are more uniformly distributed in time relative to a
-Poisson process. Hence the distribution of mRNA over a population
-of cells should be less variable compared to Poisson, giving
-$\nu<1$.\footnote{
-This last step, while intuitive, can be argued by analogy to
-photon statistics where antibunching gives rise to sub-Poissonian
-noise~\cite{Paul1982, Zou1990}. Although loopholes exist, we do
-not expect they apply for our problem. Nevertheless we refrain
-from elevating this cycle/sub-Poissonian equivalence to a ``theorem.''
-\mmnote{Antibunching is the ``obvious'' analogy for me but it'd
-be totally out of left field for most readers, any suggestions on
-a more comprehensible reference??}
-}
-
-Regardless of the merits of this model in describing the noise
-properties of constitutive transcription initation, it does not
-capture the dominant source of noise we want to include in our
-phenomenological model, so we must discard these details and
-search elsewhere.
-
-\subsubsection{Model 4 - Two-state promoter,
-``active'' and ``inactive'' states.}
-Inspired by~\cite{Razo-Mejia2020}, we revisit an analog of model
-2, but the interpretation of the two states is changed. Rather
-than explicitly viewing them as RNAP bound and unbound, we view
-them as ``active'' and ``inactive,'' which are able and unable to
-initiate transcripts, respectively. We are noncommittal as to the
-microscopic details of these states.
-
-One interpretation~\cite{Chong2014, Sevier2016} for the active
-and inactive states is that it represents the promoter's
-supercoiling state: transitions to the inactive state are caused
-by accumulation of positive supercoiling, which inhibits
-transcription, and transitions back to ``active'' are caused by
-gyrase or other topoisomerases relieving the supercoiling. This
-is an interesting possibility because it would mean the timescale
-for promoter state transitions is driven by topoisomerase
-kinetics, not by RNAP kinetics. From in vitro measurments, the
-former are suggested to be of order minutes~\cite{Chong2014}.
-Contrast this with model 2, where the state transitions are
-assumed to be governed by RNAP, which, assuming a copy number per
-cell of order $10^3$, has a diffusion-limited association rate
-$k_{on} \sim 10^2~\text{s}^{-1}$ to a target promoter. Combined
-with known $K_D$'s of order $\mu$M, this gives an RNAP
-dissociation rate $k_{off}$ of order $10^2$. As we will show
-below, however, there are some lingering puzzles with
-interpreting this supercoiling interpretation, so we leave it as
-a speculation and refrain from assigning definite physical
-meaning to the two states in this model.
-
-Intuitively one might expect that, since transcripts are produced
-as a Poisson process only when the promoter is in one of the two
-states in this model, transcription initiations should now be
-``bunched,'' in constrast to the ``anti-bunching'' of models 2
-and 3 above. One might further guess that this bunching would
-lead to super-Poissonian noise in the mRNA distribution over a
-population of cells. An honest calculation of the Fano factor produces
-\begin{align}
-\nu &= 1 + \frac{r k^-}{(k^+ + k^- + \gamma)(k^+ + k^-)},
-\end{align}
-which is strictly greater than 1, verifying the above intuition.
-Note we have dropped the $P$ label on the promoter switching rates to
-emphasize that these very likely do not represent kinetics of RNAP itself.
-This calculation can also be sidestepped by noting that the model
-is mathematically equivalent to the simple repression repression model
-from~\cite{Jones2014}, with states and rates relabeled and reinterpreted.
-
-How does this model compare to model 1 above?
-In model 1, all non-Poisson features of the mRNA
-distribution were handled as extrinsic corrections.
-By contrast, here the 3 parameter model is used to fit the full
-mRNA distribution as measured in mRNA FISH experiments.
-In essense, all variability in the mRNA distribution is regarded
-as ``intrinsic,'' arising either from stochastic initiation or
-from switching between the two coarse-grained promoter states.
-The advantage of this approach is that it fits neatly into the
-master equation picture, and the parameters thus inferred can be used as
-input for more complicated models with regulation by transcription factors.
-
-While this seems promising, there is a major drawback for our purposes
-which was already uncovered by the authors of~\cite{Razo-Mejia2020}:
-the statistical inference problem is nonidentifiable,
-in the sense described in Section 4.3 of~\cite{Gelman2013},
-meaning it is impossible to infer the
-parameters $r$ and $k^-$ from the FISH data of~\cite{Jones2014}
-(as shown in Fig.~S2 of~\cite{Razo-Mejia2020}).
-Rather, only the ratio $r/k^-$ could be inferred. In that work,
-the problem was worked around with an informative prior on the
-ratio $k^-/k^+$. That approach is unlikely to work here, as,
-recall, our entire goal in modeling constitutive expression is to
-use it as the basis for a yet more complicated model, when we add
-on repression. But adding more complexity to a model that is
-already poorly identified is a fool's errand,
-so we will explore one more potential model.
-
-\subsubsection{Model 5 - One-state promoter with explicit bursts.}
-This model is inspired by the failure mode of model 4.
-The key observation above was that, as found
-in~\cite{Razo-Mejia2020}, only two parameters, $k^+$ and the
-ratio $r/k^-$, could be directly inferred from the FISH data
-from~\cite{Jones2014}.
-So let us take this seriously and imagine a model where these are
-the only two model parameters. What would this model look like?
-
-To develop some intuition, consider model 4 in the limit
-$k^+ \ll k^- \lesssim r$, which is roughly satisfied by the
-parameters inferred in~\cite{Razo-Mejia2020}.
-In this limit, the system spends the majority of its time in the inactive state, occasionally becoming active and making a burst of transcripts.
-This should call to mind the well-known phenomenon of transcriptional bursting, as reported in, e.g.,~\cite{Golding2005,Chong2014,Sevier2016}\mmnote{should probably add some more cites here}.
-Let us make this correspondence more precise.
-The mean dwell time in the active state is $1/k^-$.
-While in this state, transcripts are produced at a rate $r$ per
-unit time. So on average, $r/k^-$ transcripts are produced before
-the system switches to the inactive state. Once in the inactive
-state, the system dwells there for an average time $1/k^+$ before
-returning to the active state and repeating the process. $r/k^-$
-resembles an average burst size, and $1/k^+$ resembles the time
-interval between burst events. More precisely, $1/k^+$ is the
-mean time between the end of one burst and the start of the next,
-whereas $1/k^+ + 1/k^-$ would be the mean interval between the
-start of two successive burst events, but in the limit $k^+ \ll
-k^-$, $1/k^+ + 1/k^- \approx 1/k^+$. Note that this limit ensures
-that the waiting time between bursts is approximately
-exponentially distributed, with mean set by the only timescale
-left in the problem,
-$1/k^+$.\footnote{
-If instead it were the case that $k^+ \sim k^-$, then the waiting time
-$1/k^+ + 1/k^-$ would have a peaked distribution, but this does
-not appear to be the case for any of datasets
-from~\cite{Jones2014}.}
-
-Let us now verify this intuition with a precise derivation
-to check that $r/k^-$ is in fact the mean burst size
-and to obtain the full burst size distribution.
-Consider first a constant, known dwell time $T$ in the active state.
-Transcripts are produced at a rate $r$ per unit time, so the
-number of transcripts $n$ produced during $T$ fits exactly
-the ``story'' for the Poisson distribution, i.e.,
-\begin{equation}
-    P(n\mid T) = \frac{(rT)^n}{n!} \exp(-rT).
-\end{equation}
-Since the dwell time $T$ is unobservable, we actually want $P(n)$,
-the dwell time distribution with no conditioning on $T$.
-Basic rules of probability theory tell us we can write $P(n)$
-in terms of $P(n\mid T)$ as
-\begin{equation}
-    P(n) =\int_0^\infty P(n\mid T) P(T) dT.
-\end{equation}
-But we know the dwell time distribution $P(T)$,
-which is exponentially distributed according to
-\begin{equation}
-    P(T) = k^- \exp(-T k^-),
-\end{equation}
-so $P(n)$ can be written as
-\begin{equation}
-    P(n) = k^- \frac{r^n}{n!}
-            \int_0^\infty T^n\exp[-(r + k^-)T]\,dT.
-\end{equation}
-A standard integral table shows
-$\int_0^\infty x^n e^{-ax}\,dx = n!/a^{n+1}$, so
-\begin{equation}
-    P(n) = \frac{k^- r^n}{(k^- + r)^{n+1}}
-        = \frac{k^-}{k^- + r}
-            \left(\frac{r}{k^- + r}\right)^n
-        = \frac{k^-}{k^- + r}
-            \left(1 - \frac{k^-}{k^- + r}\right)^n,
-\end{equation}
-which is exactly the geometric distribution with standard parameter
-$\theta\equiv k^-/(k^- + r)$ and domain
-$n \in \{0, 1, 2, \dots\}$
-(this is one of two common conventions for the geometric distribution).
-The mean of the geometric distribution, with this convention, is
-\begin{align}
-\langle n\rangle = \frac{1 - \theta}{\theta}
-        = \left(1 - \frac{k^-}{k^- + r}\right)
-                    \frac{k^- + r}{k^-}
-        = \frac{r}{k^-},
-\end{align}
-exactly as we guessed intuitively above.
-
-So in taking the limit $r,k^-\rightarrow\infty$, $r/k^-\equiv b$,
-we obtain a model which effectively has only a single promoter state,
-which initiates bursts at rate $k^+$
-(transitions to the active state, in the model 4 picture).
-The master equation for mRNA copy number $m$ takes the form
-\begin{align}
-\begin{split}
-\deriv{t}{p(m,t)} = & (m+1)\gamma p(m+1,t) - m\gamma p(m,t) \\
-        &+ \sum_{j=0}^{m-1} k^+ p(j,t) Geom(m-j;b)
-         - \sum_{j=m+1}^\infty k^+ p(m,t) Geom(j-m;b),
-\end{split}
-\end{align}
-where $Geom(n;b)$ is the geometric distribution with mean~$b$, i.e.,
-$Geom(n;b) = \frac{1}{1+b}\left(\frac{b}{1+b}\right)^n$
-(with domain over nonnegative integers as above).
-The first two terms are the usual mRNA degradation terms.
-The third term enumerates all ways the system can produce
-a burst of transcripts and arrive at copy number $m$,
-given that it had copy number $j$ before the burst.
-The fourth term allows the system to start with copy number $m$,
-then produce a burst and end with copy number $j$.
-In fact this last sum has trivial $j$ dependence and
-simply enforces normalization of the geometric distribution.
-Carrying it out we have
-\begin{equation}
-\begin{split}
-\deriv{t}{p(m,t)} = & (m+1)\gamma p(m+1,t) - m\gamma p(m,t) \\
-        &+ \sum_{j=0}^{m-1} k^+ p(j,t) Geom(m-j;b)
-            - k^+ p(m,t),
-\end{split}
-\end{equation}
-This improves on model 4 in that now the parameters are easily inferred,
-as we will see later, and have clean interpretations.
-The non-Poissonian features are attributed to the emprically
-well-established phenomenological picture of bursty transcription.
-
-The big approximation in going from model 4 to 5 is that a burst
-is produced instantaneously rather than over a finite time.
-If the true burst duration is not short compared to transcription factor
-kinetic timescales, this could be a problem in that mean burst size
-in the presense and absence of repressors could change,
-rendering parameter inferences from the constitutive case inappropriate.
-Let us make some simple estimates of this.
-
-Consider the time delay between the first and final RNAPs
-in a burst initiating transcription
-(\textit{not} the time to complete transcripts,
-which potentially could be much longer.)
-If this timescale is short compared to the typical
-search timescale of transcription factors, then all is well.
-The estimates from deHaseth et.\ al.~\cite{DeHaseth1998}
-put RNAP's diffusion-limited on rate around
-$\sim\text{few}\times10^{-2}~\text{nM}^{-1}~\text{s}^{-1}$
-and polymerase loading as high as $1~\text{s}^{-1}$.
-Then for reasonable burst sizes of $<10$, it is reasonable to guess that
-bursts might finish initiating on a timescale of tens of seconds or less
-(with another 30-60 sec to finish elongation, but that does not matter here).
-A transcription factor with typical copy number of order 10 (or less)
-would have a diffusion-limited association rate of order
-$(10~\text{sec})^{-1}$.
-Higher copy number TFs tend to have many binding sites over the genome,
-which should serve to pull them out of circulation and keep their
-effective association rates from rising too large.
-Therefore, there is \textit{perhaps} a timescale separation
-possible between transcription factor association rates and burst
-durations, but this assumption could very well break down, so we
-will have to keep it in mind when we infer repressor rates from
-the Jones et.\ al.\ FISH data later.
-
-In reflecting on these 5 models, the reader may feel that
-exploring a multitude of potential models just to return to a
-very minimal phenomenological model of bursty transcription may
-seem highly pedantic. But the purpose of the exercise was to
-examine a host of models from the literature and understand why
-they are insufficient, one way or another, for our purposes.
-Along the way we have learned that the detailed kinetics of RNAP
-binding and initiating transcription are probably irrelevant for
-setting the population distribution of mRNA.
-The timescales are simply too fast, and as we will see later
-in~\fig{fig:constit_post}, the noise seems to be governed by
-slower timescales. Perhaps in hindsight this is not surprising:
-intuitively, the degradation rate $\gamma$ sets the fundamental
-timescale for mRNA dynamics, and any other processes that
-substantially modulate the mRNA distribution should not differ
-from $\gamma$ by orders of magnitude.
-
-\subsection{Parameter inference for constitutive promoters}
-Having established that a one-state bursty model achieves a
-Goldilocks level of complexity, we test it more thoroughly on all
-the data from constitutive promoters from~\cite{Jones2014}.
-\fig{fig:constit_post} shows the results.
-\begin{figure}%[h!]
-\centering
-\includegraphics[width=0.9\textwidth]{../figures/fig2/fig2pt1.pdf}
-\caption{\textbf{Constitutive promoter posterior
-        inference and predictive checks.}
-        (A) shows the joint posterior density from MCMC sampling of model 5,
-        bursty promoter with negative binomially-distributed steady state.
-        (B) shows the ECDF of the observed population distribution
-        of mRNA transcripts under the control of a constitutive lacUV5 promoter.
-        Posterior predictive percentiles are shown as colored bands for
-        model (1), Poisson, and model (5), negative binomial.
-        }
-\label{fig:uv5_post}
-\end{figure}
-
-\begin{figure}%[h!]
-\centering
-\includegraphics[width=0.9\textwidth]{../figures/fig2/fig2pt2.pdf}
-\caption{\textbf{Constitutive promoter model comparison.}
-    (A) Joint posterior distributions for burst rate $k_i$ and
-    mean burst size $b$ for 18 promoters from Jones 2014.
-    Each contour indicates the 95\% highest posterior density
-    region for a particular promoter, i.e., on a plot analogous
-    to~\fig{fig:uv5_post}, the contour would enclose
-    approximately 95\% of the samples, and thus 95\% of the
-    probability mass.
-    Note that the vertical axis is shared with (B).
-    Panel (B) plots the burst rate $k_i$ vs.\ binding energy from energy
-    matrices from Brewster 2012.
-    The dotted line shows the predicted slope according to
-    $\ln k_i/\gamma \sim \text{constant} - \beta\Delta\epsilon_P$,
-    described in text.
-    }
-\label{fig:constit_post}
-\end{figure}
-\mmnote{Discuss puzzling comparison w/ Chong2014: if supercoiling is the thing, why are my burst sizes all the same but burst rates vary?
-And why is the duty cycle of my promoter so low, ie., bursts so short?
-COmpare their fig 7E; their $\beta/\alpha$ is my $k^+/k^-$.
-They have one or two genes with very small $\beta/\alpha$, does the \textit{galK} locus just happen to be that, or is there a deeper disagreement?
-Hard to say w/o more data.}
-
-\subsection{Transcription factor kinetics can be inferred from FISH measurements}
-Figure~3 presents inferred transcription factor rates, compares
-them to single-molecule measurements, and compares their ratios
-to gold-standard binding energies.
-\begin{figure}%[h!]
-\centering
-\includegraphics[width=0.9\textwidth]{../figures/fig3/fig3.ai}
-\caption{\textbf{Simple repression model comparison.}
-(A) depicts the cartoon of simple repression for which we infer parameters.
-(B) shows 50\% and 95\% contours of several 2D slices of the 9D posterior
-distribution, from fitting one unbinding rate for each operator (O1, O2, O3)
-and one binding rate for each aTc concentration (corresponding to an
-unknown mean repressor copy number).
-(C) compares ratios of our inferred unbinding rates with operator
-binding energy differences measured in Garcia \& Phillips 2011 (triangles)
-and Razo-Mejia et.\ al.\ 2018 (squares).
-(D) compares our inferred unbinding rates with single-molecule measurements
-of the same from Hammar et.\ at.\ 2014 \mmnote{Johan Elf's group}.
-Values are of similar order of magnitude,
-but a precise quantitative comparison unfortunately depends sensitively
-on the assumed mRNA lifetime, which was not measured.
-    }
-\label{fig3:kR_inferences}
-\end{figure}
-
-\begin{figure}%[h!]
-\centering
-\includegraphics[width=0.9\textwidth]{../figures/figSIxx/ppc_many_pooled.pdf}
-\caption{\textbf{Posterior predictive checks.}
-(A) duplicates the posterior from~\fig{fig3:kR_inferences}B.
-(B), (C), and (D) show samples from the posterior predictive distribution
-for each of the datasets used in the inference.
-Samples are sorted by operator and plotted separately simply for visual clarity.
-The unrepressed promoter, UV5, is shown with each as a reference point.
-\mmnote{
-Obviously the PPC is not perfect,
-but considering the relative simplicity of the model, I'm pleasantly surprised
-how well it worked and that the model was even identifiable at all.
-Interestingly, and perhaps not surprisingly, only 3 of the 9 datasets
-were identifiable if fit in isolation: otherwise,
-it was only by making assumptions that certain rates are equal
-across datasets was identifiability achieved.
-}
-    }
-\label{fig4:ppc}
-\end{figure}
-
-\section{Discussion and future work}
-\mmnote{
-Do all my approximations work simply because galK happens to have
-a low duty cycle of supercoiling, $\beta/\alpha$ in Chong2014
-picture? A potentially interesting test: put in the dozen or so
-Hernan constructs at some other locus, preferably one of the ones
-Chong identified as having a large duty cycle (large $\beta/\alpha$),
-i.e., the opposite limit of galK. Then just measure fold-changes.
-You don't even need to do FISH, although that would be even
-better. If their model is right, $\rho$ should be approximately
-set by the duty cycle of supercoiling, NOT by RNAP kinetics.
-\textit{This is a NICE prediction for future work.} \textit{cspE}
-is claimed nonessential, prediction is you drop HG reporter
-constructs in there and the inferred binding energies of all lacI
-operators shift the same amount, by about $\ln(3)\sim 1kT$. Not
-huge, but detectable, although might be confounding effects from
-moving to a new locus. With FISH measurements, $k^+$ and $k^-$ in
-Fig 2, model 4 should now be separately inferrable if Chong2014
-measurements of $\beta/\alpha$ are correct. This would require an
-obnoxious amount of cloning, but once you did that, the
-measurements would be same old stuff that RP lab does.}
-
-%%%%%%%%%%%%%%%%%%%%% APPENDICES %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\appendix
-
-\section{Derivations of constitutive promoter models}
-\subsection{Model 2}
-\mmnote{Obtain Fano factor from CME.}
-\subsection{Model 3}
-\mmnote{Obtain Fano factor by taking limit from Mitarai et al's expression.}
-\subsection{Model 4}
-\mmnote{Obtain Fano factor by relabeling cartoon from Jones2014.}
-
-\section{Generating function solution for bursty promoters}
-
-\subsection{Constitutive promoter with bursts}
-
-\subsubsection{From master equation to generating function}
-
-We would like to derive the full probability distribution of
-mRNA copy number per cell for a model combining simple repression
-with a promoter featuring bursty transcription initiation.
-Before tackling mRNA and repression together, let us
-just consider the bursty consititutive promoter,
-model 5 in~\fig{fig:constit_cartoons}
-\mmnote{replicating Charlotte's notes; she says it's worked out
-in~\cite{Paulsson2000}, but I don't see it, at least not in notation I can
-comprehend. Need to find a cite for this,
-or maybe just cite Charlotte, personal comm, since that's where I actually got it\dots}.
-The master equation of interest is
-\begin{align}
-\deriv{t}p(m,t) = (m+1)\gamma p(m+1,t) - m\gamma p(m,t) - r p(m,t)
-        + r \sum_{m^\prime=0}^m G_{m-m^\prime}(\theta) p(m^\prime,t),
-\label{eq:1state_unreg_003}
-\end{align}
-where $G_{k}(\theta)$ is the geometric distribution defined as
-\begin{align}
-G_{k}(\theta) = \theta(1 - \theta)^k, \, k\in\{0,1,2,\dots\}.
-\end{align}
-With this convention, the mean burst size $b = (1-\theta)/\theta$.
-$\gamma$ and $r$ are mRNA degradation rates and transcription burst
-initiation rates, resp. The last term represents all ways the system could
-end up with $m$ mRNAs, having started the burst with $m^\prime$. Define
-$\lambda = r/\gamma$ and nondimensionalize time by $\gamma$, giving
-\begin{align}
-\deriv{t}p(m,t) = (m+1)p(m+1,t) - m p(m,t) - \lambda p(m,t)
-        + \lambda \sum_{m^\prime=0}^m G_{m-m^\prime}(\theta) p(m^\prime,t),
-\end{align}
-The probability generating function is defined as
-\begin{align}
-F(z,t) = \sum_{m=0}^\infty z^m p(m,t).
-\end{align}
-Multiply both sides of the CME by $z^m$ and sum over all $m$ to get
-\begin{align}
-\pderiv[F]{t} = \sum_{m=0}^\infty z^m
-\left(
-(m+1)p(m+1,t) - m p(m,t) - \lambda p(m,t)
-+ \lambda \sum_{m^\prime=0}^m G_{m-m^\prime}(\theta) p(m^\prime,t)
-\right).
-\end{align}
-The first two terms can be rewritten as derivatives of the
-generating function, i.e.,
-\begin{align}
-\sum_{m=0}^\infty z^m (m+1)p(m+1,t)
-=
-\pderiv{z} \sum_{m=0}^\infty z^{m+1} p(m+1,t)
-=
-\pderiv[F]{z}
-\end{align}
-and
-\begin{align}
-\sum_{m=0}^\infty z^m mp(m,t)
-=
-z\pderiv{z} \sum_{m=0}^\infty z^m p(m,t)
-=
-z\pderiv[F]{z}.
-\end{align}
-The final term is the most trouble. The trick is to reverse
-the default order of the sums, as in
-\begin{align}
-&\lambda \sum_{m=0}^\infty \sum_{m^\prime=0}^m
-        z^m \theta(1 - \theta)^{m-m^\prime} p(m^\prime,t)
-\\=
-&\lambda\theta \sum_{m^\prime=0}^\infty z^{m^\prime} p(m^\prime,t)
-        \sum_{m=m^\prime}^\infty
-        z^{m-m^\prime} (1 - \theta)^{m-m^\prime}
-\\=
-&\frac{\lambda\theta F}{1 - (1 - \theta)z}.
-\end{align}
-This works because after reordering the sums, the sum on $m$ is just
-a geometric series. Its result is independent of $m^\prime$, so the sum on
-$m^\prime$ reduces to the definition of the generating function.
-Putting these results together, the PDE for the generation function is
-\begin{align}
-\pderiv[F]{t} = \pderiv[F]{z} - z\pderiv[F]{z} - \lambda F
-        + \frac{\lambda\theta F}{1-z(1-\theta)}.
-\end{align}
-Changing variables to $\xi=1-\theta$ and simplifying gives
-\begin{align}
-\pderiv[F]{t} + (z - 1) \pderiv[F]{z} = \frac{(z-1)\xi}{1-z\xi}\lambda F.
-\label{eq:1state_unreg_015}
-\end{align}
-\subsubsection{Stead-state}
-At steady-state, the PDE reduces to the ODE
-\begin{align}
-\deriv[F]{z} = \frac{\xi}{1-z\xi}\lambda F,
-\end{align}
-which we can integrate as
-\begin{align}
-\int \frac{dF}{F} = \int \frac{\lambda\xi dz}{1-\xi z}.
-\end{align}
-The initial conditions for generating functions can be subtle and confusing.
-The key fact follows from the definition
-$F(z,t) = \sum_m z^m p(m,t)$,
-normalization requires that
-$F(z=1^-,t) = \sum_m p(m,t) = 1$.\footnote{
-Sometimes the generating function may be undefined \textit{at} $z=1$ but the
-limit still holds. Also many authors change variables from $z$ to other
-things, so one must keep track of how the normalization condition transforms.
-}
-Doing the integrals (and producing constant $c$) gives
-\begin{align}
-\ln F &= -\lambda \ln(1-\xi z) + c
-\\
-F &= \frac{c}{(1-\xi z)^\lambda}.
-\end{align}
-Only one choice for $c$ can satisfy initial conditions, producing
-\begin{align}
-F(z) = \left(\frac{1-\xi}{1-\xi z}\right)^\lambda
-        = \left(\frac{\theta}{1 - z(1-\theta)}\right)^\lambda,
-\end{align}
-which is exactly the negative binomial's generating function, as expected.
-
-\subsection{Adding repression}
-\subsubsection{Deriving the generating function for mRNA distribution}
-
-Let us move from a one-state promoter to a two-state promoter, where one state
-has repressor bound and the other produces transcriptional bursts as above.
-A schematic of this model is shown in~\fig{fig3:kR_inferences}.
-Although now we have an equation for each promoter state, otherwise
-the master equation reads similarly to the one-state case, except with
-additional terms corresponding to transitions between promoter states, namely
-\begin{align}
-\deriv{t}p_R(m,t) =& k_R^+ p_A(m,t) - k_R^- p_R(m,t)
-        + (m+1)\gamma p_R(m+1,t) - m\gamma p_R(m,t)
-\\
-\begin{split}
-\deriv{t}p_A(m,t) =& - k_R^+ p_A(m,t) + k_R^- p_R(m,t)
-        + (m+1)\gamma p_A(m+1,t) - m\gamma p_A(m,t) 
-\\
-&- k_i p_A(m,t) + k_i \sum_{m^\prime=0}^m \theta(1-\theta)^{m-m^\prime} p_A(m^\prime,t),
-\end{split}
-\end{align}
-where $p_R(m,t)$ is the probability of the system having $m$ mRNA copies
-and having repressor bound to the promoter at time $t$, and $p_A$ is an
-analogous probability to find the promoter without repressor bound.
-$k_R+$ and $k_R^-$ are, respectively, the rates at which repressors
-bind and unbind to and from the promoter, and
-$\gamma$ is the mRNA degradation rate.
-$k_i$ is the rate at which bursts initiate, and as before,
-the geometric distribution of burst sizes has mean $b=(1-\theta)/\theta$.\footnote{
-$\theta$ can be viewed as a Bernoulli trial probability: after each transcript
-in a burst is produced, $\theta$ is the probability that the burst
-terminates and $(1-\theta)$ is the probability that the burst continues.
-}
-
-Interestingly, it turns out that this problem maps exactly onto the
-three-stage promoter model considered by
-Shahrezaei and Swain in~\cite{Shahrezaei2008}, with relabelings.
-Their approximate solution for protein distributions amounts to
-the same approximation we make here in regarding the duration
-of mRNA synthesis bursts as instantaneous, so their solution
-for protein distributions also solves our problem of mRNA distributions.
-Let us examine the analogy more closely.
-They consider a two-state promoter, as we do here, but they model
-mRNA as being produced one at a time and degraded, with rates
-$v_0$ and $d_0$.
-Then they model translation as occuring with rate $v_1$,
-and protein degradation with rate $d_1$.
-Now consider the limit where $v_1, d_0\rightarrow\infty$ with
-their ratio $v_1/d_0$ held constant.
-$v_1/d_0$ resembles the average burst size of translation from a single
-mRNA: these are the rates of two Poisson processes that compete over a
-transcript, which matches the story of geometrically distributed burst sizes.
-So after taking this limit, on timescales slow compared to $v_1$ and $d_0$,
-it appears that transcription events fire at rate $v_0$ and produce a
-geometrically distributed burst of translation of mean size $v_1/d_0$,
-which intuitively matches the story we have told above for mRNA
-with variables relabeled.
-
-To verify this intuitively conjectured mapping between our problem and
-the solution in~\cite{Shahrezaei2008}, we continue with a careful
-solution for the mRNA distribution using probability generating functions,
-following the ideas sketched in~\cite{Shahrezaei2008}.
-It is natural to nondimensionalize rates in the problem by $\gamma$,
-or equivalently, this amounts to measuring time in units of $\gamma^{-1}$.
-We are also only interested in steady state, so we set the time
-derivatives to zero, giving
-\begin{align}
-0 =& k_R^+ p_A(m) - k_R^- p_R(m) + (m+1) p_R(m+1) - m p_R(m)
-\\
-\begin{split}
-0 =& - k_R^+ p_A(m) + k_R^- p_R(m) + (m+1) p_A(m+1) - m p_A(m) 
-\\
-&- k_i p_A(m) + k_i \sum_{m^\prime=0}^m \theta(1-\theta)^{m-m^\prime} p_A(m^\prime),
-\end{split}
-\end{align}
-
-The probability generating function is defined as before in the
-constitutive case, except now we must introduce a generating function
-for each promoter state,
-\begin{align}
-f_A(z) = \sum_{m=0}^\infty z^m p_A(m),
-\;
-f_R(z) = \sum_{m=0}^\infty z^m p_R(m).
-\end{align}
-Our real objective is the generating function $f(z)$ that generates
-the mRNA distribution $p(m)$, independent of what state the promoter is in.
-But since $p(m) = p_A(m) + p_R(m)$, it follows too that $f(z) = f_A(z) + f_R(z)$.
-
-So as before we multiply both equations by $z^m$ and sum over all $m$.
-Each individual term transforms exactly as did an analogous term in
-the constitutive case, so the coupled ODEs for the generating functions read
-\begin{align}
-0 =& k_R^+ f_A(z) - k_R^- f_R(z) + \pderiv{z} f_R(z) - z \pderiv{z} f_R(z)
-\\
-\begin{split}
-0 =&  - k_R^+ f_A(z) + k_R^- f_R(z) + \pderiv{z} f_A(z) - z \pderiv{z} f_A(z)
-\\
-&- k_i f_A(z) + k_i \frac{\theta}{1-z(1-\theta)} f_A(z),
-\end{split}
-\end{align}
-and after changing variables $\xi = 1 - \theta$ as before
-and rearranging, we have
-\begin{align}
-0 &= k_R^+ f_A(z) - k_R^- f_R(z) + (1-z) \pderiv{z} f_R(z)
-\\
-0 &=  - k_R^+ f_A(z) + k_R^- f_R(z) + (1 - z) \pderiv{z} f_A(z)
-+ k_i \frac{(z-1)\xi}{1-z\xi} f_A(z),
-\end{align}
-We can transform this problem from two coupled first-order ODEs to a
-single second-order ODE by solving for $f_A$ in the first and plugging
-into the second, giving
-\begin{align}
-\begin{split}
-0 = (1&-z) \pderiv[f_R]{z}
-+ \frac{1-z}{k_R^+}
-        \left(k_R^- \pderiv[f_R]{z} + \pderiv[f_R]{z} +(z-1) \psecderiv[f_R]{z}\right)
-\\
-&+ \frac{k_i}{k_R^+} \frac{(z-1)\xi}{1-z\xi}
-        \left(k_R^- f_R + (z-1) \pderiv[f_R]{z}\right),
-\end{split}
-\end{align}
-where, to reduce notational clutter, we have dropped the explicit $z$
-dependence of $f_A$ and $f_R$. Simplifying we have
-\begin{align}
-0 = \psecderiv[f_R]{z}
-        - \left(\frac{k_i\xi}{1-z\xi}
-                + \frac{1 + k_R^- + k_R^+}{1-z}
-        \right)\pderiv[f_R]{z}
-        + \frac{k_i k_R^- \xi}{(1-z\xi)(1-z)}f_R.
-\end{align}
-This can be recognized as the hypergeometric differential equation,
-with singularities at $z=1$, $z=\xi^{-1}$, and $z=\infty$. The latter
-can be verified by a change of variables from $z$ to $x=1/z$,
-being careful with the chain rule, and noting that $z=\infty$ is a
-singular point if and only if $x=1/z=0$ is a singular point.
-
-The standard form of the hypergeometric differential equation has its
-singularities at 0, 1, and $\infty$, so to take advantage of the
-standard form solutions to this ODE, we first need to transform
-variables to put it into a standard form.
-However, this is subtle. While any such transformation should work in
-principle, the solutions are expressed most simply in the neighborhood
-of $z=0$, but the normalization condition that we need to enforce
-corresponds to $z=1$. The easiest path, therefore, is to find a change
-of variables that maps
-1 to 0, $\infty$ to $\infty$, and $\xi^{-1}$ to 1.
-This is most intuitively done in two steps.
-
-First map the $z=1$ singularity to 0 by the change of variables $v=z-1$, giving
-\begin{align}
-0 = \psecderiv[f_R]{v}
-        + \left(\frac{k_i\xi}{(1+v)\xi - 1}
-                + \frac{1 + k_R^- + k_R^+}{v}
-        \right)\pderiv[f_R]{v}
-        + \frac{k_i k_R^- \xi}{((1+v)\xi - 1)v}f_R.
-\end{align}
-Now two singularities are at $v=0$ and $v=\infty$.
-The third is determined by $(1+v)\xi -1 = 0$, or $v=\xi^{-1} - 1$.
-We want another variable change that maps this third singularity to 1
-(without moving 0 or infinity). Changing variables again to
-$w=\frac{v}{\xi^{-1} - 1} = \frac{\xi}{1-\xi} v$ fits the bill.
-In other words, the combined change of variables
-\begin{align}
-w = \frac{\xi}{1-\xi} (z-1)
-\end{align}
-maps $z = \{1, \xi^{-1}, \infty\}$ to $w =\{0, 1, \infty\}$ as desired.
-Plugging in, being mindful of the chain rule and noting
-$(1 + v)\xi - 1 = (1 - \xi)(w - 1)$ gives
-\begin{align}
-0 = \left(\frac{\xi}{1-\xi}\right)^2 \psecderiv[f_R]{w}
-+ \left(
-        \frac{\xi k_i}{(1-\xi)(w-1)} + \frac{\xi(1 + k_R^- + k_R^+)}{(1-\xi)w}
-\right) \frac{\xi}{1-\xi} \pderiv[f_R]{w}
-+ \frac{k_i k_R^- \xi^2}{(1-\xi)^2 w(w-1)}f_R.
-\end{align}
-This is close to the standard form of the hypergeometric differential
-equation, and some cancellation and rearrangement gives
-\begin{align}
-0 = w(w-1)\psecderiv[f_R]{w}
-+ \left(k_i w + (1 + k_R^- + k_R^+)(w-1)\right) \pderiv[f_R]{w}
-+ k_i k_R^- f_R.
-\end{align}
-and a little more algebra produces
-\begin{align}
-0 = w(1-w)\psecderiv[f_R]{w}
-+ \left(1 + k_R^- + k_R^+
-        - (1 + k_i + k_R^- + k_R^+)w
-\right) \pderiv[f_R]{w}
-- k_i k_R^- f_R,
-\end{align}
-which is the standard form. From this we can read off the solution
-in terms of hypergeometric functions ${_2F_1}$ from
-any standard source, e.g.~\cite{Abramowitz1964}, and identify the
-conventional parameters in terms of our model parameters.
-We want the general solution in the neighborhood of $w=0$ ($z=1$),
-which for a homogeneous linear second order ODE must be a sum of two
-linearly independent solutions.
-More precisely,
-\begin{align}
-f_R(w) = C_1 {_2F_1}(\alpha, \beta, \gamma; w)
-+ C_2 w^{1-\gamma}{_2F_1}(1+\alpha-\gamma, 1+\beta-\gamma, 2-\gamma; w)
-\end{align}
-with parameters determined by
-\begin{align}
-\begin{split}
-\alpha\beta &= k_i k_R^-
-\\
-1+\alpha+\beta &= 1+k_i+k_R^-+k_R^+
-\\
-\gamma &= 1 + k_R^- + k_R^+
-\end{split}
-\end{align}
-and constants $C_1$ and $C_2$ to be set by boundary conditions.
-Solving for $\alpha$ and $\beta$, we find\footnote{
-Note that $\alpha$ and $\beta$ are interchangeable in the definition of
-${_2F_1}$ and differ only in the sign preceeding the radical.
-}
-\begin{align}
-\begin{split}
-\alpha &= \frac{1}{2}
-\left(k_i+k_R^-+k_R^+ + \sqrt{(k_i+k_R^-+k_R^+)^2 - 4k_i k_R^-}\right)
-\\
-\beta &= \frac{1}{2}
-\left(k_i+k_R^-+k_R^+ - \sqrt{(k_i+k_R^-+k_R^+)^2 - 4k_i k_R^-}\right)
-\\
-\gamma &= 1 + k_R^- + k_R^+.
-\end{split}
-\end{align}
-Since the normalization condition requires that $f_R$ be finite at $w=0$,
-we can immediately set $C_2=0$ to discard the second solution.
-This is because all the rate constants are strictly positive,
-so $\gamma>1$ and therefore $w^{1-\gamma}$ blows up as $w\rightarrow0$.
-Now that we have $f_R$, we would like to find the generating function
-for the mRNA distribution, $f(z) = f_A(z) + f_R(z)$.
-We can recover $f_A$ from our solution for $f_R$, namely
-\begin{align}
-f_A(z) = \frac{1}{k_R^+}\left(k_R^- f_R(z) + (z-1) \pderiv[f_R]{z}\right)
-\end{align}
-or
-\begin{align}
-f_A(w) = \frac{1}{k_R^+}\left(k_R^- f_R(w) + w \pderiv[f_R]{w}\right),
-\end{align}
-where in the second line we transformed our original relation between
-$f_R$ and $f_A$ to our new, more convenient, variable $w$.
-Plugging our solution for $f_R(w) = C_1{_2F_1}(\alpha, \beta, \gamma; w)$
-into $f_A$, we will require the differentiation rule for ${_2F_1}$,
-which tells us
-\begin{align}
-\pderiv[f_R]{w} = C_1\frac{\alpha\beta}{\gamma}
-                {_2F_1}(\alpha+1, \beta+1, \gamma+1; w),
-\end{align}
-from which it follows that
-\begin{align}
-f_A(w) = \frac{C_1}{k_R^+}
-\left(
-k_R^- {_2F_1}(\alpha, \beta, \gamma; w)
-+ w\frac{\alpha\beta}{\gamma} {_2F_1}(\alpha+1, \beta+1, \gamma+1; w)
-\right)
-\end{align}
-and therefore
-\begin{align}
-f(w) = C_1\left(1 + \frac{k_R^-}{k_R^+}\right)
-        {_2F_1}(\alpha, \beta, \gamma; w)
-+ w \frac{C_1}{k_R^+} \frac{\alpha\beta}{\gamma}
-        {_2F_1}(\alpha+1, \beta+1, \gamma+1; w).
-\end{align}
-To proceed, we need one of the (many) useful identities known for
-hypergeometric functions, in particular
-\begin{align}
-w\frac{\alpha\beta}{\gamma} {_2F_1}(\alpha+1, \beta+1, \gamma+1; w)
-=
-(\gamma-1)\left(
-{_2F_1}(\alpha, \beta, \gamma-1; w) - {_2F_1}(\alpha, \beta, \gamma; w)
-\right).
-\end{align}
-Substituting this for the second term in $f(w)$, we find
-\begin{align}
-f(w) = \frac{C_1}{k_R^+}
-\left[
-        \left(k_R^+ + k_R^-\right)
-        {_2F_1}(\alpha, \beta, \gamma; w)
-+ (\gamma-1)\left(
-        {_2F_1}(\alpha, \beta, \gamma-1; w) - {_2F_1}(\alpha, \beta, \gamma; w)
-        \right)
-\right],
-\end{align}
-and since $\gamma-1 = k_R^+ + k_R^-$, the first and third terms cancel,
-leaving only
-\begin{align}
-f(w) = C_1\frac{k_R^+ + k_R^-}{k_R^+} {_2F_1}(\alpha, \beta, \gamma-1; w).
-\end{align}
-Now we enforce normalization, demanding $f(w=0) = f(z=1) = 1$.
-${_2F_1}(\alpha, \beta, \gamma-1; 0) = 1$, so we must have
-$C_1 = k_R^+ / (k_R^+ + k_R^-)$ and consequently
-\begin{align}
-f(w) =  {_2F_1}(\alpha, \beta, k_R^+ + k_R^-; w).
-\end{align}
-Recalling that the mean burst size $b = (1-\theta)/\theta = \xi/(1-\xi)$
-and $w = \frac{\xi}{1-\xi} (z-1) = b (z-1)$,
-we can transform back to the original variable $z$ to find the tidy result
-\begin{align}
-f(z) =  {_2F_1}(\alpha, \beta, k_R^+ + k_R^-; b(z-1)),
-\end{align}
-with $\alpha$ and $\beta$ given above by
-\begin{align}
-\begin{split}
-\alpha &= \frac{1}{2}
-\left(k_i+k_R^-+k_R^+ + \sqrt{(k_i+k_R^-+k_R^+)^2 - 4k_i k_R^-}\right)
-\\
-\beta &= \frac{1}{2}
-\left(k_i+k_R^-+k_R^+ - \sqrt{(k_i+k_R^-+k_R^+)^2 - 4k_i k_R^-}\right).
-\end{split}
-\end{align}
-Finally we are in sight of the original goal. We can generate the
-steady-state probability distribution of interest by
-differentiating the generating function,
-\begin{align}
-p(m) = m! \left.\frac{\partial^m}{\partial z^m} f(z) \right|_{z=0},
-\end{align}
-which follows easily from its definition. Some contemplation reveals
-that repeated application of the derivative rule used above will produce
-products of the form $\alpha(\alpha+1)(\alpha+2)\cdots(\alpha+m-1)$ in
-the expression for $p(m)$ and similarly for $\beta$ and $\gamma$. These
-resemble ratios of factorials, but since $\alpha$, $\beta$, and $\gamma$
-are not necessarily integer, we should express the ratios using gamma
-functions instead. More precisely, one finds
-\begin{align}
-p(m) = \frac{
-        \Gamma(\alpha + m)\Gamma(\beta + m)\Gamma(k_R^+ + k_R^-)
-        }
-        {
-        \Gamma(\alpha)\Gamma(\beta)\Gamma(k_R^+ + k_R^- + m)
-        }
-\frac{b^m}{m!}{_2F_1}(\alpha+m, \beta+m, k_R^++k_R^-+m; -b)
-\label{eq:p_m_bursty+rep}
-\end{align}
-which is finally the probability distribution we sought to derive.
-
-\subsubsection{Moments}
-We can verify the correctness of~\eq{eq:p_m_bursty+rep}
-by comparison with Gillespie simulation.
-We can also easily compute moments of the distribution from
-the generating function, and check these against raw moments computed
-directly from the master equation.
-
-The $j$-th raw moment of the distribution can be computed from the
-generating function from the formula
-\begin{align}
-\langle m^j \rangle
-= \left. \left(z \frac{\partial}{\partial z}\right)^j f(z)\right|_{z=1}.
-\end{align}
-Note that \textit{probabilities} are generated by differentiating
-and evaluating at $z=0$, while \textit{moments} are generated by
-differentiating and evaluating at $z=1$. This also aligns with the
-normalization constraint, which can be thought of as the zeroth raw moment.
-
-The mean mRNA is given by
-\begin{align}
-\langle m \rangle = \left.z \pderiv[f]{z} \right|_{z=1}
-= \frac{\alpha\beta b}{k_R^+ + k_R^-}
-= k_i b\frac{k_R^-}{k_R^+ + k_R^-},
-\end{align}
-which is intuitive and appealingly simple:
-the mean mRNA is just the mean mRNA in the absense of repression
-($k_i b$, the burst rate time mean burst size)
-times the fraction of time the repressor is bound, $k_R^-/(k_R^+ + k_R^-)$.
-
-To compute the variance, let us first compute
-\begin{align}
-\begin{split}
-\langle m^2\rangle
-&= \left .z\pderiv{z}\left(z\pderiv[f]{z}\right) \right|_{z=1}
-= \left. z\left(\pderiv[f]{z} + z\psecderiv[f]{z}\right)\right|_{z=1}
-\\
-&= \frac{\alpha\beta b}{k_R^+ + k_R^-}
-        + \frac{\alpha(\alpha+1)\beta(\beta+1) b^2}
-                {(k_R^+ + k_R^-)(k_R^+ + k_R^- + 1)}
-\\
-&= \frac{k_i b k_R^-}{k_R^+ + k_R^-}
-        + \frac{k_i k_R^- b^2 (1 + k_i + k_R^+ + k_R^- + k_i k_R^-)}
-                {(k_R^+ + k_R^-)(k_R^+ + k_R^- + 1)},
-\end{split}
-\end{align}
-from which the variance is
-\begin{align}
-\begin{split}
-var(m) = \langle m^2\rangle - \langle m\rangle^2
-&= \frac{k_i b k_R^-}{k_R^+ + k_R^-}
-\left(
-1 + \frac{b(1 + k_i + k_R^+ + k_R^- + k_i k_R^-)} {k_R^+ + k_R^- + 1}
-- \frac{k_i b k_R^-}{k_R^+ + k_R^-}
-\right)
-\\
-&= k_i b\frac{k_R^-}{k_R^+ + k_R^-}
-\left(
-1 + b + k_i b\frac{1 + k_R^-} {k_R^+ + k_R^- + 1}
-- \frac{k_i b k_R^-}{k_R^+ + k_R^-}
-\right)
-\\
-&= k_i b\frac{k_R^-}{k_R^+ + k_R^-}
-\left(
-1 + b + k_i b\frac{k_R^+}{(k_R^+ + k_R^-)(k_R^+ + k_R^- + 1)}
-\right).
-\end{split}
-\end{align}
-The Fano factor is the quantity in parentheses. This can be interpreted as
-the variability arising from the negative binomial distribution of the
-unregulated promoter $(1+b)$ plus additional noise arising from repression.
-
-\mmnote{Still to writeup: moment derivation direct from CME to compare}
-
-\subsection{Numerical considerations and recursion formulas}
-\subsubsection{Generalities}
-We would like to carry out Bayesian parameter inference on FISH data
-from~\cite{Jones2014}, using~\eq{eq:p_m_bursty+rep} as our
-likelihood. This requires accurate (and preferably fast)
-numerical evaluation of the hypergeometric function ${_2F_1}$,
-which is a notoriously hard problem~\cite{Pearson2017, Gil2007},
-and our particular needs here present an especial challenge as we show below.
-
-The hypergeometric function is defined by its Taylor series as
-\begin{align}
-{_2F_1}(a,b,c;z) 
-= \sum_{l=0}^\infty
-\frac{\Gamma(a + l)\Gamma(b + l)\Gamma(c)}
-        {\Gamma(a)\Gamma(b)\Gamma(c + l)}
-\frac{z^l}{l!}
-\end{align}
-for $|z|<1$, and by analytic continuation elsewhere.
-If $z\lesssim1/2$ and $\alpha$ and $\beta$ are not too large
-(absolute value below 20 or 30),
-then the series converges quickly and an accurate numerical representation is
-easily computed by truncating the series after a reasonable number of terms.
-Unfortunately, we need to evaluate ${_2F_1}$ over mRNA copy numbers fully out
-to the tail of the distribution, which can easily reach 50, possibly 100.
-From~\eq{eq:p_m_bursty+rep}, this means evaluating ${_2F_1}$
-repeatedly for values of $a$, $b$, and $c$ spanning the full range
-from $\mathcal{O}(1)$ to $\mathcal{O}(10^2)$,
-even if $\alpha$, $\beta$, and $\gamma$
-in~\eq{eq:p_m_bursty+rep} are small,
-with the situation even worse if they are not small.
-A naive numerical evaluation of the series definition will be
-prone to overflow and, if any of $a,b,c<0$, then some successive terms in the
-series have alternating signs which can lead to catastrophic cancellations.
-
-One solution is to evaluate ${_2F_1}$ using arbitrary precision arithmetic
-instead of floating point arithmetic,
-e.g., using the \texttt{mpmath} library in Python.
-This is accurate but incredibly slow computationally.
-To quantify how slow, we found that
-evaluating the likelihood defined by~\eq{eq:p_m_bursty+rep} $\sim50$ times
-(for a typical dataset of interest from~\cite{Jones2014},
-with $m$ values spanning 0 to $\sim50$)
-using arbitrary precision arithmetic is 100-1000 fold slower than
-evaluating a negative binomial likelihood for the corresponding
-constitutive promoter dataset.
-
-To claw back $\gtrsim30$ fold of that slowdown, we can exploit
-one of the many catalogued symmetries involving ${_2F_1}$.
-The solution involves recursion relations originally explored by Gauss,
-and studied extensively in~\cite{Pearson2017, Gil2007}.
-They are sometimes known as contiguous relations and relate the values
-of any set of 3 hypergeometric functions whose arguments differ by integers.
-To rephrase this symbolically, consider a set of hypergeometric functions
-indexed by an integer $n$,
-\begin{align}
-f_n = {_2F_1}(a+\epsilon_i n, b+\epsilon_j n, c+\epsilon_k n; z),
-\end{align}
-for a fixed choice of $\epsilon_i, \epsilon_j, \epsilon_k \in \{0,\pm 1\}$
-(at least one of $\epsilon_i, \epsilon_j, \epsilon_k$ must be nonzero,
-else the set of $f_n$ would contain only a single element).
-Then there exist known recurrence relations of the form
-\begin{align}
-A_n f_{n-1} + B_n f_{n} + C_n f_{n+1} = 0,
-\end{align}
-where $A_n, B_n$, and $C_n$ are some functions of $a,b,c$, and $z$.
-In other words, for fixed $\epsilon_i, \epsilon_j, \epsilon_k, a, b,$ and $c$,
-if we can merely evaluate ${_2F_1}$ twice, say for $n^\prime$ and $n^\prime-1$,
-then we can easily and rapidly generate values for arbitrary $n$.
-
-This provides a convenient solution for our problem: we need repeated
-evaluations of ${_2F_1}(a+m, b+m, c+m; z)$
-for fixed $a,b$, and $c$ and many integer values of $m$.
-They idea is that we can use arbitrary precision arithmetic to evaluate
-${_2F_1}$ for just two particular values of $m$ and then generate
-${_2F_1}$ for the other 50-100 values of $m$ using the recurrence
-relation.\footnote{
-There are even more sophisticated ways of utilizing the recurrence
-relations that might have netted another factor of 2 speed-up, and
-possibly as much as a factor of 10, but the method described here had
-already reduced the computation time to an acceptable
-$\mathcal{O}(\text{1 min})$, so these more sophisticated approaches did
-not seem worth the time to pursue.
-}
-
-However, there two further wrinkles.
-The first is that
-a naive application of the recurrence relation is numerically unstable.
-Roughly, this is because the three term recurrence relations,
-like second order ODEs, admit two linearly independent solutions.
-In a certain eigenbasis, one of these solutions dominates the other
-as $n\rightarrow\infty$, and as $n\rightarrow-\infty$,
-the dominance is reversed.
-If we fail to work in this eigenbasis, our solution of the recurrence relation
-will be a mixture of these solutions and rapidly accumulate numerical error.
-For our purposes, it suffices to know that the authors of~\cite{Gil2007}
-derived the numerically stable solutions (so-called \textit{minimal solutions})
-for several possible choices of $\epsilon_i, \epsilon_j, \epsilon_k$.
-Running the recurrence in the proper direction using a minimal solution
-is numerically robust and can be done entirely in floating point arithmetic, 
-so that we only need to evaluate ${_2F_1}$ with arbitrary precision arithmetic
-to generate the seed values for the recursion.
-
-The second wrinkle is a corollary to the first.
-The minimal solutions are only minimal for certain ranges of the argument $z$,
-and not all of the 26 possible recurrence relations
-have minimal solutions for all $z$.
-This can be solved by using one of the many transformation formulae for
-${_2F_1}$ to convert to a different recurrence relation that has
-a minimal solution over the required domain of $z$, although
-this can require some trial and error to find the right transformation,
-the right recurrence relation, and the right minimal solution.
-
-\subsubsection{Particulars}
-Let us now demonstrate these generalities for our problem of interest.
-In order to evaluate the probability distribution of our
-model,~\eq{eq:p_m_bursty+rep}, we need to evaluate hypergeometric functions
-of the form ${_2F_1}(\alpha+m, \beta+m, \gamma+m; -b)$
-for values of $m$ ranging from $0$ to $\mathcal{O}(100)$.
-The authors of~\cite{Gil2007} did not derive a recursion relation
-for precisely this case. We could follow their methods and do so ourselves,
-but it is much easier to convert to a case that they did consider.
-The strategy is to look through the minimal solutions tabulated
-in~\cite{Gil2007} and search for a transformation we could apply to
-${_2F_1}(\alpha+m, \beta+m, \gamma+m; -b)$ that would place the $m$'s
-(the variable being incremented by the recursion)
-in the same arguments of ${_2F_1}$ as the minimal solution.
-After some ``guess and check,'' we found that the transformation
-\begin{align}
-{_2F_1}(\alpha+m, \beta+m, \gamma+m; -b)
-=
-(1+b)^{-\alpha-m}
-        {_2F_1}\left(\alpha+m, \gamma-\beta, \gamma+m; \frac{b}{1+b}\right),
-\label{eq:rec_euler_pretransform}
-\end{align}
-produces a ${_2F_1}$ on the right hand side that closely resembles
-the minimal solutions $y_{3,m}$ and $y_{4,m}$ in Eq.~4.3 in~\cite{Gil2007}.
-Explicitly, these solutions are
-\begin{align}
-y_{3,m}
-&\propto
-{_2F_1}\left(-\alpha^\prime + \gamma^\prime - m,
-                -\beta^\prime + \gamma^\prime,
-                1-\alpha^\prime-\beta^\prime+\gamma^\prime-m;
-                1-z\right)
-\\
-y_{4,m}
-&\propto
-{_2F_1}\left(\alpha^\prime + m,
-                \beta^\prime,
-                1+\alpha^\prime+\beta^\prime-\gamma^\prime+m;
-                1-z\right),
-\label{eq:minimal_soln_sans_prefac}
-\end{align}
-where we have omitted prefactors which are unimportant for now.
-Which of these two we should use depends on what values $z$ takes on.
-Equating $1-z=b/(1+b)$ gives $z=1/(1+b)$, and since $b$ is strictly positive,
-$z$ is bounded between 0 and 1.
-From Eq.~4.5 in~\cite{Gil2007}, $y_{4,m}$ is the minimal solution
-for real $z$ satisfying $0<z<2$, so this is the only minimal solution we need.
-
-Now that we have our minimal solution,
-what recurrence relation does it satisfy?
-Confusingly, the recurrence relation of which $y_{4,m}$ is a solution
-increments different arguments of ${_2F_1}$ that does $y_{4,m}$:
-it increments the first only, rather than first and third.
-This recurrence relation can be looked up, e.g., Eq.~15.2.10
-in~\cite{Abramowitz1964}, which is
-\begin{align}
-(\gamma^\prime - (\alpha^\prime + m)) f_{m-1}
-+
-(2(\alpha^\prime+m) - \gamma^\prime + (\beta^\prime - \alpha^\prime)z)f_m
-+ \alpha^\prime(z-1) f_{m+1} = 0.
-\label{eq:chosen_rec_rel}
-\end{align}
-Now we must solve for the parameters appearing in the recurrence relation
-in terms of our parameters, namely by setting
-\begin{align}
-\begin{split}
-\alpha^\prime &= \alpha
-\\
-\beta^\prime &= \gamma - \beta
-\\
-1 + \alpha^\prime + \beta^\prime - \gamma^\prime &= \gamma
-\\
-1 - z &= \frac{b}{1+b}
-\end{split}
-\end{align}
-and solving to find
-\begin{align}
-\begin{split}
-\alpha^\prime &= \alpha
-\\
-\beta^\prime &= \gamma - \beta
-\\
-\gamma^\prime &= 1 + \alpha - \beta
-\\
-z &= \frac{1}{1+b}
-.
-\end{split}
-\end{align}
-Finally we have everything we need. The minimal solution
-\begin{align}
-y_{4,m}
-=
-\frac{\Gamma(1+\alpha^\prime-\gamma^\prime+m)}
-        {\Gamma(1+\alpha^\prime+\beta^\prime-\gamma^\prime+m)}
-\times
-{_2F_1}\left(\alpha^\prime + m,
-                \beta^\prime,
-                1+\alpha^\prime+\beta^\prime-\gamma^\prime+m;
-                1-z\right),
-\end{align}
-where we have now included the necessary prefactors,
-is a numerically stable solution of the recurrence
-relation~\eq{eq:chosen_rec_rel} if the recursion is run
-from large $m$ to small $m$.
-So the procedure is to compute the value of ${_2F_1}$ for the two
-largest $m$ values of interest using arbitrary precision arithmetic,
-compute the prefactors to construct
-$y_{4,\text{max}(m)}$ and $y_{4,\text{max}(m)-1}$,
-recursively compute $y_{4,m}$ for all $m$ less than $\text{max}(m)$ down
-to $m=0$, then cancel off the prefactors of the resulting values of
-$y_{4,m}$ for all $m$ to produce ${_2F_1}$ for all desired $m$ values.
-\mmnote{Maybe I should reformat this paragraph into
-a pseudocode version of the algorithm?
-Instead of, or in addition to?}
-
-With ${_2F_1}$ computed, the only remaining numerical danger in computing
-$p(m)$ in~\eq{eq:p_m_bursty+rep} is overflow of the gamma functions.
-This is easily solved by taking the log of the entire expression
-and using standard routines to compute the log of the gamma functions,
-then exponentiating the entire expression at the end if $p(m)$
-is needed rather than $\log p(m)$.
-
-% \bibliographystyle{nature}
-\bibliographystyle{abbrv}
-\bibliography{./library}
-
-\end{document}

### 2020-05-19 doc/appendix_02_constitutive_promoters.tex
initial splitting of document into several files

--- 2020-05-17 doc/paper.tex
@@
-\documentclass[12pt]{article}%{amsart}
-\usepackage[top = 1.0in, bottom = 1.0in, left = 1.0in, right = 1.0in]{geometry}
-% See geometry.pdf to learn the layout options. There are lots.
-%\geometry{letterpaper}               % ... or a4paper or a5paper or ...
-%\geometry{landscape}             % Activate for for rotated page geometry
-\usepackage[parfill]{parskip}    % Activate to begin paragraphs with an empty line rather than an indent
-\usepackage{graphicx}
-\usepackage{amssymb}
-%\usepackage{epstopdf}
-% \usepackage{caption}
-\usepackage{amsmath}
-% \usepackage{longtable}
-% \usepackage{tabu}
-% \usepackage{accents} %to get undertildes for vec & mat
-\usepackage{color} % for colored text
-% \usepackage[normalem]{ulem} % for editing: striking out text using 'sout' command
-
-% \newcommand\mytablefigwidth{0.35\textwidth}
-
-% allows to use .ai files directly w/o resaving as pdf
-\DeclareGraphicsRule{.ai}{pdf}{.ai}{}
-
-% Handy math macros!
-\newcommand{\vect}[1]{\vec{#1}}
-\newcommand{\matr}[1]{\mathbf{#1}}
-\newcommand{\rate}[3]{{#1}_{#2}^{#3}}
-\newcommand{\mmnote}[1]{\textcolor{cyan}{(MM:~#1)}}
-
-% derivative macros. these are sneaky. the [{}] is an empty default 1st arg
-% usage: provide 1 arg: \deriv{x} gives d/dx
-% provide 2 args like so: \deriv[f]{x} gives df/dx
-\newcommand{\deriv}[2][{}]{\frac{d #1}{d #2}}
-\newcommand{\pderiv}[2][{}]{\frac{\partial #1}{\partial #2}}
-\newcommand{\psecderiv}[2][{}]{\frac{\partial{^2} #1}{\partial #2{^2}}}
-
-%%%%% Referencing macros %%%%%
-\newcommand{\fig}[1]{Figure~\ref{#1}}
-\newcommand{\tref}[1]{Table~\ref{#1}}
-\newcommand{\eq}[1]{Eq.~(\ref{#1})}
-\newcommand{\eqrange}[2]{Eqs.~(\ref{#1}-\ref{#2})}
-
-%%%%%
-\begin{document}
-
-\title{Simple repression draft}
-
-\maketitle
-
-\section{Introduction}
-\mmnote{Key ideas, no particular order, that I haven't written down before:
-\begin{itemize}
-\item Our goal is to build phenomenological models of input-output
-functions of genetic circuits. More precisely, we want phenomenology
-in the sense of coarse-graining away atomistic detail, but still
-retaining biophysical meaning, e.g., we don't want to coarse-grain
-as far as Hill functions. Why not? We are motivated by studies
-like~\cite{Razo-Mejia2020}, for which Hill functions are
-insufficient. Such work requires a quantitative theory of how
-biophysical changes at the molecular level propagate to
-input-output functions at the genetic circuit level.
-We want concepts, not mere facts.
-In particular a key question is: at this level of
-coarse graining, what microscopic details do I need to explicitly
-model, and how do we figure that out? For example, do I need to
-worry about all or even any of the steps that individual RNAPs go
-through each time they make a transcript? Turning the question
-around, can we see any imprint of those processes in the
-available data? If the answer is no, then those processes are
-irrelevant for our purposes. Forward modeling and inverse
-(statistical inferential) modeling can complement each other
-beautifully here.
-\end{itemize}}
-
-\fig{fig1:means_cartoons}A shows the qualitative picture of
-simple repression. An operator, the binding site on the DNA for a
-repressor protein, may be found occupied by a repressor, in which
-case transcription is blocked from occurring. Or it may be found
-unoccupied, in which case RNA polymerase (RNAP) may bind and
-transcription can proceed.
-
-The key assumption we make is that binding of repressor and RNAP in the promoter region of interest is exclusive, meaning that one or the other may bind, but never may both be simultaneously bound.
-It is often imagined that when the repressor is bound to its operator,
-RNAP is sterically blocked from binding to its promoter sequence.
-Current evidence suggests this is sometimes, but not
-always the case~\mmnote{Cite that fig, from ??Nathan or Bill or
-Rob??, showing locations of repressor binding sites far upstream
-of promoter}, and it remains an interesting open question precisely
-how a repressor bound far upstream is able to repress transription.
-Suggestions include ``action-at-a-distance'' mediated by kinks in
-the DNA, formed when the repressor is bound, that prevent RNAP binding.
-Nevertheless, our modeling in this work is sufficiently
-coarse-grained that we simply assume exclusive binding and need
-not worry about such precise details.
-
-\mmnote{Add a notation aside somewhere that all association rates are written as zeroth-order rates.
-In other words, the concentration of the molecule is hidden inside, so for instance, $k_R^+$ is proportional to repressor copy number.}
-
-\section{Means}
-\subsection{Fold-changes are indistinguishable across models}
-As a first stop on our search for the ``right'' model of simple
-repression, let us consider what we can learn from theory and
-measurements of fold-change in mean expression. The fold-change is defined as
-\begin{equation}
-\text{fold-change}
-= \frac{\langle \text{gene expression with repressor present} \rangle}
-        {\langle \text{gene expression with repressor absent} \rangle}
-= \frac{\langle m\rangle(R\ne 0)}{\langle m\rangle(R = 0)}
-\label{eq:fc_def}
-\end{equation}
-where angle brackets denote the averge over a population of cells
-and mean mRNA $\langle m\rangle$ is viewed as a function of
-repressor copy number $R$.
-The second
-equality in~\ref{eq:fc_def} follows from assuming that the
-translation efficiency, i.e., the number of proteins translated
-per mRNA, is the same in both conditions.
-In other words, we assume that mean protein level is proportional
-to mean mRNA level, and that the proportionality constant is the
-same in both conditions and therefore cancels out in the ratio.
-This is reasonable since the cells in the two conditions are
-identical except for the presence of the transcription factor,
-and it is difficult to imagine the transcription factor having
-any direct effect on translation.
-
-Fold-change has proven a very convenient observable in past
-work~\cite{Garcia2011a, Razo-Mejia2018, Chure2019}.
-Part of its utility in dissecting transcriptional regulation is
-its ratiometric nature, which is always easier than making an
-absolute measurement. Also, by measuring otherwise identical cells
-with and without a transcription factor present, any biological
-noise common to both conditions can be made to cancel away.
-In addition, the presence of different operator sites is known to
-change the basal expression level in the absence of
-repressor~\mmnote{look up cite}.
-Computing fold-changes neatly cancels this effect as well.
-
-\fig{fig1:means_cartoons} depicts a smorgasbord
-of mathematized cartoons for simple repression that have appeared
-in previous literature. For each cartoon, we calculate the
-fold-change in mean gene expression as predicted by that model,
-deferring some algebraic details to the appendix.
-What we will find is that all cartoons collapse to a single
-master curve, shown in \fig{fig1:means_cartoons}c, which contains
-just two parameters. We label the parameters $\Delta F_R$, an
-effective free energy parametrizing the repressor-DNA
-interaction, and $\rho$, which subsumes all details of
-transcription in the absence of repressors.
-We will offer some intuition for why this master curve exists and discuss
-why it is likely impossible to discriminate ``right'' from ``wrong''
-cartoons given only measurements of fold-changes in expression.
-
-\subsubsection{Equilibrium, 2-state}
-In this simplest model, the promoter is idealized as existing in
-one of two states, either repressor bound or repressor unbound.
-The rate of transcription, and therefore the level of gene
-expression, is assumed to be proportional to the fraction of time
-spent in the repressor unbound state.
-Mathematically, this amounts to the statement that the mean mRNA level is
-given by
-\begin{equation}
-\langle m \rangle = \frac{r}{\gamma}
-        \left(1 + \frac{R}{N_{NS}} e^{-\beta\Delta\epsilon_R}\right)^{-1},
-\end{equation}
-where $r$ is the transcription rate from the repressor unbound
-state, $\gamma$ is the mRNA degradation rate, $R$ is repressor
-copy number, $N_{NS}$ is the number of nonspecific binding sites
-in the genome (where repressors spend most of their time when not
-bound to the operator, their specifc binding site),
-$\beta=1/k_BT$, and $\Delta\epsilon_R$ is the binding energy of a
-repressor to its operator site. The derivation of this result is deferred
-to the appendix.
-
-The fold-change is easily found as the ratio of mean mRNA with
-and without repressor. This gives
-\begin{equation}
-\text{fold-change}
-= \left(1 + \frac{R}{N_{NS}} e^{-\beta\Delta\epsilon_R}\right)^{-1},
-\end{equation}
-which clearly matches the form of the master curve
-in~\eq{fig1:means_cartoons} with $\rho=1$.
-
-In fact it was noted in~\cite{Chure2019} that this two-state
-model can be viewed as the coarse-graining of any equilibrium
-promoter model in which no transcriptionally active states have
-transcription factor bound. Put differently, there is no overlap
-between TF bound states and transcriptionally active states. We
-will see this explicitly in the 3-state equilibrium model below,
-but perhaps surprising is that something analogous carries over
-even to the nonequilibrium models we consider later.
-
-\subsubsection{Equilibrium, 3-state}
-Compared to the previous model, here we fine-grain the repressor
-unbound state into two separate states: empty, and RNAP bound.
-This picture was used in~\cite{Garcia2011a} as we use it here,
-and in~\cite{Razo-Mejia2018} and~\cite{Chure2019} it was generalized
-to incorporate small-molecule inducers that bind the repressor.
-Interestingly, the effect of this is simply to renormalize $R$
-from the total number of repressors to a smaller effective number,
-roughly (though not exactly) the number which are unbound by inducers.
-
-The mean mRNA, as derived in the appendix from a similar
-enumeration of states and weights as the previous model, is
-\begin{equation}
-\langle m \rangle = \frac{r}{\gamma}
-\frac{\frac{P}{N_{NS}} e^{-\beta\Delta\epsilon_P}}
-        {
-        1 + \frac{R}{N_{NS}} e^{-\beta\Delta\epsilon_R}
-        + \frac{P}{N_{NS}} e^{-\beta\Delta\epsilon_P}
-        },
-\end{equation}
-where the new variables are $\Delta\epsilon_P$, the difference in
-RNAP binding energy to its specific site (the promoter) relative
-to an average nonspecific background site, and the RNAP copy
-number, $P$. The fold-change again follows immediately as
-\begin{align}
-\text{fold-change}
-&= \frac{\frac{P}{N_{NS}} e^{-\beta\Delta\epsilon_P}}
-        {
-        1 + \frac{R}{N_{NS}} e^{-\beta\Delta\epsilon_R}
-        + \frac{P}{N_{NS}} e^{-\beta\Delta\epsilon_P}
-        }
-\frac{1 + \frac{P}{N_{NS}} e^{-\beta\Delta\epsilon_P}}
-        {\frac{P}{N_{NS}} e^{-\beta\Delta\epsilon_P}}
-\\
-&= \left(
-1 + \frac{\frac{R}{N_{NS}} e^{-\beta\Delta\epsilon_R}}
-        {1 + \frac{P}{N_{NS}} e^{-\beta\Delta\epsilon_P}}
-\right)^{-1}
-\\
-&= (1 + \exp(-\Delta F_R - \log\rho))^{-1},
-\end{align}
-with $\Delta F_R = \beta\Delta\epsilon_R - \log(R/N_{NS})$
-and $\rho = 1 + \frac{P}{N_{NS}}\mathrm{e}^{-\beta\Delta\varepsilon_P}$
-as shown in~\fig{fig1:means_cartoons}.
-
-\begin{figure}%[h!]
-\centering
-\includegraphics[width=\textwidth]{../figures/fig1/fig1.ai}
-\caption{
-\textbf{
-An overview of the simple repression motif at the level of means.}
-(A) depicts the qualitative biological picture of simple repression.
-(B) shows a variety of possible mathematized cartoons of simple
-repression, along with the effective parameter $\rho$
-which subsumes all regulatory details
-of the architecture excluding the repressor.
-(C) shows the ``master curve'' to which all cartoons in (B) collapse.
-}
-\label{fig1:means_cartoons}
-\end{figure}
-
-\subsubsection{Nonequilibrium model 1 - Poisson promoter}
-For our first kinetic model, we consider the simplest possible
-picture with only two states, repressor bound and unbound.
-This is exactly the model used for the main results of~\cite{Jones2014}.
-In this picture, repressor association and dissociation rates
-from its operator site, $k_R^+$ and $k_R^-$, respectively, govern
-transitions between the two states. When the system is in the
-unbound state, transcriptions initiate at rate $r$, which
-represents a coarse-graining of all the downstream processes into
-a single effective rate. mRNA is again degraded at rate $\gamma$.
-
-Let $p_{m,R}(t)$ denote the joint probability of finding the
-system in the repressor bound state with $m$ mRNA molecules
-present at time $t$, and similarly define $p_{m,U}(t)$ for the
-repressor unbound state. 
-This model is governed by a master equation giving the time
-evolution of $p_{m,R}(t)$ and $p_{m,U}(t)$, which is
-\begin{align}
-\begin{split}
-\deriv{t}p_{m,R}(t) =& - k_R^- p_{m,R}(t) + k_R^+ p_{m,U}(t)
-                + (m+1)\gamma p_{m+1,R}(t) - \gamma p_{m,R}(t)
-\\
-\deriv{t}p_{m,U}(t) =&\; k_R^- p_{m,R}(t) - k_R^+ p_{m,U}(t)
-                        + rp_{m-1,U}(t) - rp_{m,U}(t)
-                \\
-                &+ (m+1)\gamma p_{m+1,U}(t) - \gamma p_{m,U}(t).
-\end{split}
-\end{align}
-In each equation, the first two terms describe transitions
-between promoter states due to repressors unbinding and binding,
-respectively. The final two terms describe degradation of mRNA,
-decreasing the copy number by one, and the terms with coefficient
-$r$ describe transcription initiation increasing the mRNA copy
-number by one.
-
-We can greatly simplify the notation, which will be especially
-useful for the more complicated models yet to come, by
-re-expressing the master equation in vector form. The promoter
-states are collected into a vector and the rate constants are
-collected into matrices as
-\begin{equation}
-\vect{p}_m = \begin{pmatrix} p_{m,R} \\ p_{m,U} \end{pmatrix},\
-\matr{K} = \begin{pmatrix} -k_R^- & k_R^+ \\ k_R^- & -k_R^+ \end{pmatrix},\
-\matr{R} = \begin{pmatrix} 0 & 0 \\ 0 & r \end{pmatrix},\
-\label{eq:2state_cme_matrices}
-\end{equation}
-so that the master equation may be condensed as
-\begin{equation}
-\deriv{t}\vect{p}_m(t) =
-\left( \matr{K} - \matr{R} - \gamma m \matr{I} \right) \vect{p}_m(t)
-                + \matr{R} \vect{p}_{m-1}(t)
-                + \gamma (m+1) \matr{I} \vect{p}_{m+1}(t)
-                - \gamma m \matr{I} \vect{p}_m(t),
-\label{eq:2state_rep_cme}
-\end{equation}
-where $\matr{I}$ is the identity matrix.
-Taking steady state by setting time derivatives to zero,
-the mean mRNA can be found to be
-\begin{equation}
-\langle m \rangle = \frac{r}{\gamma}
-        \left(1 + \frac{k_R^+}{k_R^-}\right)^{-1},
-\end{equation}
-with the algebra details again deferred to the appendix.
-Recall $k_R^+$ is proportional to the repressor copy number, so
-in computing fold-change, absence of repressor corresponds to
-$k_R^+\rightarrow0$.
-Therefore fold-change in this model is simply
-\begin{equation}
-\text{fold-change} = \left(1 + \frac{k_R^+}{k_R^-}\right)^{-1},
-\end{equation}
-again matching the master curve of~\fig{fig1:means_cartoons} with $\rho=1$.
-
-\subsubsection{Nonequilibrium model 2 - RNAP bound/unbound states}
-Analogous to the three-state equilibrium model, this model
-fine-grains the repressor unbound state of nonequilibrium model 1,
-resolving it into an empty promoter state and an RNAP-bound state.
-Note in this picture, in contrast with model 4 below,
-transcription initiation is accompanied by a promoter state
-change, in keeping with the interpretation as RNAP-bound and
-empty states: if an RNAP successfully escapes the promoter and
-proceeds to elongation of a transcript, clearly it is no longer
-bound at the promoter! Therefore another RNAP must bind before
-another transcript can be initiated.
-
-The master equation governing this model is analogous
-to~\eqrange{eq:2state_cme_matrices}{eq:2state_rep_cme} for model 1 above.
-The main subtlety arises since transcription initiation
-accompanies a promoter state change.
-This can be understood by analogy to $\matr{K}$. The
-off-diagonals and diagonals of $\matr{K}$ correspond to
-transitions arriving at or departing from, respectively, the
-promoter state of interest. If transcription initiations are
-accompanied by promoter state changes, we must have separate
-matrices for arriving and departing transcription events since
-the arriving and departing transitions have different initial
-copy number of mRNA, unlike for $\matr{K}$ where they are the same.
-The master equation for this model is
-\begin{equation}
-\deriv{t}\vect{p}_m(t) =
-\left( \matr{K} - \matr{R_D} - \gamma m \matr{I} \right) \vect{p}_m(t)
-                + \matr{R_A} \vect{p}_{m-1}(t) +
-                \gamma (m+1) \matr{I} \vect{p}_{m+1}(t)
-                - \gamma m \matr{I} \vect{p}_m(t),
-\label{eq:3state_rep_cme}
-\end{equation}
-with the state vector and promoter transition matrix defined as
-\begin{equation}
-\vect{p}_m = \begin{pmatrix} p_{m,R} \\ p_{m,E} \\ p_{m,P} \end{pmatrix},\
-\matr{K} = \begin{pmatrix} -k_R^- & k_R^+ & 0 \\
-                        k_R^- & -k_R^+ -k_P^+ & k_P^- \\
-                        0 & k_P^+ & -k_P^- 
-                \end{pmatrix},
-\label{eq:3state_cme_matrices_pt1}
-\end{equation}
-and the initiation matrices given by
-\begin{equation}
-\matr{R_A} = \begin{pmatrix}
-                0 & 0 & 0 \\ 0 & 0 & r \\ 0 & 0 & 0
-                \end{pmatrix},\
-\matr{R_D} = \begin{pmatrix}
-                0 & 0 & 0 \\ 0 & 0 & 0 \\ 0 & 0 & r
-                \end{pmatrix},
-\label{eq:3state_cme_matrices_pt2}
-\end{equation}
-The elements of $\vect{p}_m$ encode the probabilities of having
-$m$ mRNA present along with the promoter having repressor bound,
-being empty, or having RNAP bound, respectively.
-$\matr{R_A}$ describes probability flux \textit{arriving} at the state
-$\vect{p}_m$ from a state with one fewer mRNA, namely
-$\vect{p}_{m-1}$, and $\matr{R_D}$ describes probability flux
-\textit{departing} from the state $\vect{p}_m$
-for a state with one more mRNA, namely $\vect{p}_{m+1}$.
-$\matr{K}$ is closely analogous to model 1.
-
-Mean mRNA at steady state is found analogously to model 1, with the result
-\begin{equation}
-\langle m\rangle = \frac{r}{\gamma}
-        \frac{k_R^- k_P^+}
-        {k_R^- k_P^+ + k_R^- (k_P^- + r) + k_R^+ (k_P^- + r)}.
-\end{equation}
-Fold-change is again found from the ratio prescribed by~\eq{eq:fc_def},
-from which we have
-\begin{align}
-\text{fold-change}
-&=      \frac{k_R^- k_P^+}
-        {k_R^- k_P^+ + k_R^- (k_P^- + r) + k_R^+ (k_P^- + r)}
-        \frac{k_P^+ + k_P^- + r}{k_P^+}
-\\
-&=      \left(1 + \frac{k_R^+}{k_R^-}
-                \frac{k_P^- + r}{k_P^+ + k_P^- + r}
-        \right)^{-1}
-\\
-&=      \left(1 + \frac{k_R^+}{k_R^-}
-        \left(1 + \frac{k_P^+}{k_P^- + r}\right)^{-1}
-        \right)^{-1},
-\end{align}
-which follows the master curve with
-$\rho = 1 + k_P^+/(k_P^- + r)$ as claimed.
-
-\subsubsection{Nonequilibrium model 3
-- Multistep transcription initiation and escape}
-One might reasonably complain that the first two
-``nonequilibrium'' models we have considered are straw men. Their
-steady states necessarily satisfy detailed balance which is
-equivalent to thermodynamic equilibrium. Why is this the case? At
-steady state there is by definition no net probability flux in or
-out of each promoter state, but since the promoter states form a
-linear chain, there is only one way in or out of the repressor
-bound and RNAP bound states, implying each edge must actually
-have a net zero probability flux, which is the definition of
-detailed balance (usually phrased as equality of forward and
-reverse transition fluxes).
-
-Now we consider model 3 in~\eq{fig1:means_cartoons} which allows
-the possibility of true nonequilibrium steady-state fluxes
-through the promoter states.
-We point out that this model was considered previously in~\cite{Mitarai2015}
-where a comparison was made with model 1 as used in~\cite{Jones2014}.
-The authors of~\cite{Mitarai2015} argued that the additional
-complexity is essential to properly account for the noise in the
-mRNA distribution. We will weigh in on both models later when we
-consider observables beyond fold-change.
-
-The master equation governing this model is
-identical in form to model 2 above, namely
-\begin{equation}
-\deriv{t}\vect{p}_m(t) =
-\left( \matr{K} - \matr{R_D} - \gamma m \matr{I} \right) \vect{p}_m(t)
-                + \matr{R_A} \vect{p}_{m-1}(t) +
-                \gamma (m+1) \matr{I} \vect{p}_{m+1}(t)
-                - \gamma m \matr{I} \vect{p}_m(t),
-\end{equation}
-but obviously with a higher-dimensional state space and different matrices.
-The state vector and promoter transition matrix are now
-\begin{equation}
-\vect{p}_m = \begin{pmatrix} p_{m,R} \\ p_{m,E} \\
-                             p_{m,C} \\ p_{m,O}\end{pmatrix},\
-\matr{K} = \begin{pmatrix} -k_R^- & k_R^+ & 0 & 0\\
-                        k_R^- & -k_R^+ -k_P^+ & k_P^- & 0 \\
-                        0 & k_P^+ & -k_P^- - k_O & 0 \\
-                        0 & 0 & k_O & 0
-                \end{pmatrix},
-\end{equation}
-with the four promoter states, in order, being repressor bound,
-empty, RNAP closed complex, and RNAP open complex.
-Besides increasing dimension by one again, the only new feature in
-$\matr{K}$ is the rate $k_O$, representing the (irreversible)
-rate of open complex formation from the closed complex.
-The initiation matrices are given by
-\begin{equation}
-\matr{R_A} = \begin{pmatrix}
-        0 & 0 & 0 & 0 \\ 0 & 0 & 0 & r \\ 0 & 0 & 0 & 0\\ 0 & 0 & 0 & 0
-                \end{pmatrix},\
-\matr{R_D} = \begin{pmatrix}
-        0 & 0 & 0 & 0 \\ 0 & 0 & 0 & 0 \\ 0 & 0 & 0 & 0\\ 0 & 0 & 0 & r
-                \end{pmatrix},
-\end{equation}
-again closely analogous to model 2.
-
-The expression for mean mRNA is substantially more complicated
-now, as worked out in the appendix where we find
-\begin{equation}
-\langle m\rangle = \frac{r}{\gamma}
-        \frac{k_R^- k_P^+ k_O}
-        {k_R^- [(k_P^+ (k_O + r) + r(k_P^- + k_O)] + k_R^+ r(k_P^- + k_O)},
-\end{equation}
-which can be simplified to
-\begin{equation}
-\langle m\rangle
-= \frac{r}{\gamma}
-\frac{\frac{k_P^+ k_O}{r(k_O + k_P^-)}}
-        {1 + \frac{k_P^+ (k_O + r)}{r(k_O + k_P^-)} + \frac{k_R^+}{k_R^-}}.
-\end{equation}
-The strategy is to isolate the terms involving the repressor, so that now the fold-change is seen to be simply
-\begin{align}
-\text{fold-change}
-&= \frac{\frac{k_P^+ k_O}{r(k_O + k_P^-)}}
-        {1 + \frac{k_P^+ (k_O + r)}{r(k_O + k_P^-)} + \frac{k_R^+}{k_R^-}}
-        \frac{1 + \frac{k_P^+ (k_O + r)}{r(k_O + k_P^-)}}
-                {\frac{k_P^+ k_O}{r(k_O + k_P^-)}}
-\\
-&= \left(
-        1 + \frac{k_R^+}{k_R^-}
-        \left(1 + \frac{k_P^+ (k_O + r)}{r(k_O + k_P^-)}\right)^{-1}
-\right)^{-1},
-\end{align}
-surprisingly reducing to the master curve once again, with
-$\rho = 1 + \frac{k_P^+ (k_O + r)}{r(k_O + k_P^-)}$.
-
-This example hints that an arbitrarily fine-grained model of
-downstream transcription steps may still be collapsed to the form
-of the master curve in~\fig{fig1:means_cartoons}B so long as the
-repressor binding is exclusive with transcriptionally active
-states. We offer this as a conjecture, and we suspect that a
-careful argument using the King-Altman diagram
-method~\cite{King1956, Hill1966} might furnish a ``proof.''
-Our focus here is not on full generality but rather to survey an
-assortment of plausible models that have been proposed in the
-literature for simple repression.
-
-\subsubsection{Nonequilibrium model 4 - ``Active'' and ``inactive'' states}
-This model is at the core of the theory in~\cite{Razo-Mejia2020}.
-At a glance the cartoon for this model may appear very similar to model 2,
-and mathematically it is, but the interpretation is rather different.
-In model 2, we interpreted the third state literally as an
-RNAP-bound promoter and modeled initiation of a transcript as
-triggering a promoter state change, making the hopefully
-uncontroversial assumption that an RNAP can only make one
-transcript at a time. In contrast, in the present model the
-promoter state does \textit{not} change when a transcript is
-initiated. So we no longer interpret these states as literally
-RNAP bound and unbound but instead as coarse-grained ``active''
-and ``inactive'' states, the details of which we leave
-unspecified for now. We will comment more on this model below
-when we discuss Fano factors of models.
-
-Mathematically this model is very similar to models 1 and 2.
-Like model 1, the matrix $R$ describing transcription initiation
-is diagonal, namely
-\begin{equation}
-\matr{R} = \begin{pmatrix}
-                0 & 0 & 0 \\ 0 & 0 & 0 \\ 0 & 0 & r
-        \end{pmatrix}.
-\end{equation}
-so the master equation takes verbatim the same form as it did for model
-1,~\eq{eq:2state_rep_cme}. Meanwhile the promoter transition matrix
-$K$ is the same as~\eq{eq:3state_cme_matrices_pt1} from model 2,
-although we relabel the rate constants from $k_P^\pm$ to $k^\pm$ to
-reiterate that these are not simply RNAP binding and unbinding rates.
-
-Carrying out the algebra, the mean mRNA can be found to be
-\begin{equation}
-\langle m\rangle = \frac{r}{\gamma}
-\frac{k_R^- k^+}
-{k_R^- k^+ + k_R^- k^- + k_R^+ k^-},
-\end{equation}
-and the fold-change readily follows,
-\begin{align}
-\text{fold-change}
-&=      \frac{k_R^- k^+}{k_R^- k^+ + k_R^- k^- + k_R^+ k^-}
-        \frac{k_R^- k^+ + k_R^- k^-}{k_R^- k^+}
-\\
-&=      \left(1 + \frac{k_R^+}{k_R^-}
-                \left(1 + \frac{k^+}{k^-}\right)^{-1}
-        \right)^{-1},
-\end{align}
-from which we see $\rho = 1 + k^+/k^-$ as shown in~\fig{fig1:means_cartoons}.
-
-\subsubsection{Nonequilibrium model 5 - Bursty promoter}
-We will have much more to say about this model in subsequent sections.
-For now we sketch it as an intuitive analog to model 1,
-with just two states, repressor bound or unbound, and transition rates
-between them of $k_R^+$ and $k_R^-$.
-In model 1, when in the unbound state, single mRNA transcripts
-are produced as a Poisson process with some characteristic rate
-$r$. This model by contrast produces, at some Poisson rate $k_i$,
-\textit{bursts} of transcripts with a geometrically distributed
-burst size and a mean burst size $b$.
-
-From this intuitive picture and by analogy to model 1, then, it should be plausible that the mean mRNA level is
-\begin{equation}
-\langle m\rangle = \frac{k_i b}{\gamma}
-        \left(1 + \frac{k_R^+}{k_R^-}\right)^{-1},
-\end{equation}
-which will turn out to be correct from a careful calculation.
-For now, we simply note that just like model 1, the fold-change becomes
-\begin{equation}
-\text{fold-change} = \left(1 + \frac{k_R^+}{k_R^-}\right)^{-1}
-\end{equation}
-with $\rho=1$ also like model 1.
-
-We will also see later how this model emerges as a natural limit of model 4.
-
-\subsubsection{Summary and discussion of results across models for fold-changes in mean expression}
-What have we learned from our exploration of the models
-in~\fig{fig1:means_cartoons} and their predictions of
-fold-changes in mean expression levels? The takeaway is the
-existence of a master curve (\fig{fig1:means_cartoons}C) to which
-the fold-change predictions of all the models collapse. This
-master curve is parametrized by only two effective parameters:
-$\Delta F_R$, which characterizes the repressor, and $\rho$,
-which characterizes all other features of the promoter
-architecture. The key assumption underpinning this result is that
-zero transcription occurs when a repressor is bound to its
-operator. Note, however, that we are agnostic about the molecular
-mechanism which achieves this;
-steric effects are the obvious mechanism, but, for instance,
-``action-at-a-distance'' mediated by kinked DNA due to
-repressors bound tens or hundreds of nucleotides upstream of a
-promoter is plausible as well.
-
-Why does the master curve of~\fig{fig1:means_cartoons}C exist at all?
-This brings to mind the deep questions posed in,
-e.g.,~\cite{Frank2013} and~\cite{Frank2014a}, asking us to search
-for general ``symmetries'' of the problem instead of specifying
-generative models of the known microscopics of a system.
-In our case, the key feature seems to be the exclusive nature of
-repressor and RNAP binding, which allows the parameter describing
-the repressor, $\Delta F_R$, to cleanly separate from all other
-details of the promoter architecture, which are encapsulated in $\rho$.
-In the graph-theoretic language which was applied to similar
-problems in~\cite{Gunawardena2012} and~\cite{Ahsendorf2014}, the
-repressor-bound state is a leaf of the graph describing
-transitions between promoter states. Arbitrary nonequilibrium
-behavior can occur on the rest of the promoter state space, but
-it may all be swept up in the effective parameter $\rho$, to
-which the repressor makes no contribution.
-
-As suggested in~\cite{Chure2019}, we believe this master curve
-should generalize to architectures with multiple repressor
-binding sites, as long as the exclusivity of transcription factor
-binding and transcription initation is maintained. The
-interpretation of $\Delta F_R$ is then of an effective free
-energy of all repressor bound states.
-In an equilibrium picture this is simply given by the log of the
-sum of Boltzmann weights of all repressor bound states, which
-looks like the log of a partition function of a subsystem.
-In a nonequilibrium picture, while we can still mathematically gather
-terms and give the resulting collection the label $\Delta F_R$,
-it is unclear if the physical interpretation as an effective free energy
-makes sense.
-As is well known, free energies cannot be assigned unambiguously
-to states out of equilibrium because the free energy change along
-a generic path traversing the state space is path dependent,
-unlike at equilibrium.
-A consequence of this is that, out of equilibrium, $\Delta F_R$
-is no longer a simple sum of Boltzmann weights. Instead it
-resembles a restricted sum of King-Altman
-diagrams~\cite{King1956, Hill1966}. Following the work of
-Hill~\cite{Hill1989}, it may yet be possible to interpret this expression
-as an effective free energy, but this remains unclear to us.
-We leave this an open problem for future work.
-
-If we relax the requirement of exclusive repressor-RNAP binding,
-one could imagine models in which repressor and RNAP doubly-bound
-states are allowed, where the repressor's effect is to
-\text{reduce} the transcription rate rather than setting it to zero.
-Our results do not strictly apply to such a model, although we
-note that if the repressor's reduction of the transcription rate
-is substantial, such a model might still be well-approximated by
-one of the models in~\fig{fig1:means_cartoons}.
-
-One may worry that our ``one curve to rule them all'' is a
-mathematical tautology. In fact we \textit{agree} with this
-criticism if $\Delta F_R$ is ``just a fitting parameter'' and cannot
-be meaningfully interpreted as a real, physical free energy.
-An analogy to Hill functions is appropriate here. One of their
-great strengths and weaknesses, depending on the use they are put
-to, is that their parameters coarse-grain many details and are
-generally not interpretable in terms of microscopic models, for
-deep reasons discussed at length in~\cite{Frank2013}.
-By contrast, our master curve claims to have the best of both worlds:
-a coarse-graining of all details besides the repressor into
-a single effective parameter $\rho$,
-while simultaneously retaining an interpretation of
-$\Delta F_R$ as a physically
-meaningful and interpretable free energy.
-Our task, then, is to prove or disprove this claim.
-
-How do we test this and probe the theory with fold-change
-measurements? There is a fundamental limitation in that the
-master curve is essentially a one-parameter function of $\Delta
-F_R + \log\rho$. Worse, there are many \textit{a priori}
-plausible microscopic mechanisms that could contribute to the
-value of $\rho$, such as RNAP binding and escape
-kinetics~\cite{DeHaseth1998, Mitarai2015}, and/or supercoiling
-accumulation and release~\cite{Chong2014, Sevier2016}, and/or,
-RNAP clusters analogous to those obvserved in
-eukaryotes~\cite{Cisse2013, Cho2016}
-and recently also observed in bacteria~\cite{Ladouceur2020}.
-Even if $\Delta F_R$ is measured to
-high precision, inferring the potential microscopic contributions
-to $\rho$, buried inside a log no less, from fold-change
-measurements seems hopeless. As a statistical inference problem
-it is entirely nonidentifiable, in the language
-of~\cite{Gelman2013}, Section 4.3.
-
-If we cannot simply infer values of $\rho$ from measurements of
-fold-change, can we perturb some of the parameters that make up
-$\rho$ and measure the change? Unfortunately we suspect this is
-off-limits experimentally: most of the potential contributors to
-$\rho$ are global processes that affect many or all genes. For
-instance, changing RNAP association rates by changing RNAP copy
-numbers, or changing supercoiling kinetics by changing
-topoisomerase copy numbers, would massively perturb the entire
-cell's physiology and confound any determination of $\rho$.
-
-One might instead imagine a bottom-up modeling approach, where we
-mathematize a model of what we hypothesize the important steps
-are and are not, use in vitro data for the steps deemed
-important, and \textit{predict} what $\rho$ should be.
-But again, because of the one-parameter nature of the master
-curve, many different models will likely make indistinguishable
-predictions, and without any way to experimentally perturb in
-vivo, there is no clear way to test whether the modeling
-assumptions are correct.
-
-In light of this, we prefer the view that parameters and rates
-are not directly comparable between cartoons
-in~\fig{fig1:means_cartoons}. Rather, parameters in the
-simpler cartoons represent coarse-grained combinations of
-parameters in the finer-grained models. For instance, by equating
-$\rho$ between any two models, one can derive various possible
-correspondences between the two models' parameters. Note that
-these correspondences are clearly not unique, since many possible
-associations could be made. It then is a choice as to what microscopic
-interpretations the model-builder prefers for the parameters in a
-particular cartoon, and as to which coarse-grainings lend
-intuition and which seem nonsensical.
-Indeed, since it remains an open question what microscopic
-features dominate $\rho$ (as suggested above, perhaps RNAP
-binding and escape kinetics~\cite{DeHaseth1998, Mitarai2015}, or
-supercoiling accumulation and release~\cite{Chong2014, Sevier2016},
-or, something more exotic like
-RNAP clusters~\cite{Cisse2013, Cho2016, Ladouceur2020}),
-we are hesitant to put too much weight on any one microscopic
-interpretation of model parameters that make up $\rho$.
-
-One possible tuning knob to probe $\rho$ that would not globally
-perturb the cell's physiology is to manipulate RNAP binding sites.
-Work such as~\cite{Kinney2010} has shown that models of
-sequence-dependent RNAP can be inferred from data, and the
-authors of~\cite{Brewster2012} showed that the model
-of~\cite{Kinney2010} has predictive power by using the model to
-\textit{design} binding sites of a desired affinity.
-But for our purposes, this begs the question: the authors
-of~\cite{Kinney2010} \textit{assumed} a particular model
-(essentially our 3-state equilibrium model but without the
-repressor), so it is unclear how or if such methods can be turned
-around to \textit{compare} different models of promoter function.
-
-We have already pointed out that the master curve
-of~\fig{fig1:means_cartoons} is essentially a one-parameter model,
-the one parameter being $\Delta F_R + \log\rho$.
-By now the reader may be alarmed as to how can we even determine
-$\Delta F_R$ and $\rho$ independently of each other, never mind
-shedding a lens on the internal structure of $\rho$ itself.
-A hint is provided by the weak promoter approximation, invoked
-repeatedly in prior studies~\cite{Bintu2005c, Garcia2011a,
-Razo-Mejia2018} of simple repression using the 3-state
-equilibrium model in~\fig{fig1:means_cartoons}.
-In that picture, the weak promoter approximation means
-$\frac{P}{N_{NS}}\exp(-\beta\Delta\epsilon_P) \ll 1$,
-meaning therefore $\rho\approx1$.
-This is suggestive, but how can we be sure that $\rho$ is not,
-for instance, actually $10^2$ and that $\Delta F_R$ hides a
-compensatory factor?
-A resolution is offered by an independent inference of $\rho$ in
-the absence of repressors. This was done in~\cite{Razo-Mejia2020}
-by fitting nonequilibrium model 4  in~\fig{fig1:means_cartoons},
-with zero repressor (looking ahead, this is equivalent to model 4
-in~\fig{fig:constit_cartoons}), to FISH data
-from~\cite{Brewster2014}. This provided a determination of $k^+$
-and $k^-$, from which their ratio is estimated to be no more than
-a few $10^{-1}$ and possibly as small as $10^{-2}$.
-
-The realization that $\rho\approx1$ to an excellent
-approximation, \textit{independent} of which model
-in~\fig{fig1:means_cartoons} one prefers, goes a long way towards
-explaining the surprising success of equilibrium models of simple
-repression. Even though our 2- and 3-state models get so many
-details of transcription wrong, it does not matter because
-fold-change is a cleverly designed ratio. Since $\rho$ subsumes
-all details except the repressor, and $\log\rho\approx0$, fitting
-these simple models to fold-change measurements can still give a
-surprisingly good estimate of repressor binding energies. So the
-ratiometric construction of fold-change fulfills its intended
-purpose of canceling out all features of the promoter
-architecture except the repressor itself. Nevertheless it is
-perhaps surprising how effectively it does so: \textit{a priori},
-one might not have expected $\rho$ to be quite so close to 1.
-
-We would also like to highlight the relevance
-of~\cite{Landman2019} here. Landman et.\ al.\ reanalyzed and
-compared in vivo and in vitro data on the lacI repressor's
-binding affinity to its various operator sequences. (The in vivo
-data was from, essentially, fitting our master curve to
-expression measurements.) They find broad agreement between the
-in vitro and in vivo values. This reinforces the suspicion that
-the equilibrium $\Delta\epsilon_R$ repressor binding energies do
-in fact represent real physical free energies.
-Again, \textit{a priori} this did not have to be the case, even
-knowing that $\rho\approx1$.
-
-In principle, if $\Delta F_R$ can be measured to sufficient
-precision, then deviations from $\rho=1$ become a
-testable matter of experiment.
-In practice, it is probably unrealistic to measure repressor
-rates $k_R^+$ or $k_R^-$ or fold-changes in expression levels
-(and hence $\Delta\epsilon_R$) precisely enough to detect the
-expected tiny deviations from $\rho=1$.
-We can estimate the
-requisite precision in $\Delta F_R$ to resolve a given
-$\Delta\rho$ by noting, since $\rho\approx1$, that
-$\log(1+\Delta\rho)\approx \Delta\rho$, so
-$\Delta(\Delta F_R) \approx \Delta\rho$.
-Suppose we are lucky and $\Delta\rho$ happens to be $\sim0.1$, on
-the high end of our range estimated above.
-A determination of $\Delta\epsilon_R/k_BT$ with an uncertainty of
-barely 0.1 was achieved in the excellent measurements
-of~\cite{Razo-Mejia2018}, so this requires a very difficult
-determination of $\Delta F_R$ for a very crude determination of $\rho$,
-which suggests, to put it lightly, this is not a
-promising path to pursue experimentally.
-It is doubtful that inference of repressor kinetic rates
-would be any easier.\footnote{
-\mmnote{
-In fact the uncertainties quoted in Hammar et.\
-al.~\cite{Hammar2014} from Elf's group are small enough to do
-this calculation, but HJ \& I think their error bars are insanely
-optimistic. Not sure I want to blatantly call them out on it though??}
-}
-
-Moving forward, we have weak evidence supporting the
-interpretation of $\Delta F_R$ as a physically real free
-energy~\cite{Landman2019} and other work casting
-doubt~\cite{Hammar2014}.
-How might we resolve the confusion?
-If there is no discriminatory power to test the theory and
-distinguish the various models with measurements of fold-changes
-in means, how do we probe the theory? Clearly to discriminate
-between the nonequilibrium models in~\fig{fig1:means_cartoons},
-we need to go beyond means to ask questions about kinetics, noise and even
-full distributions of mRNA copy numbers over a population of cells.
-If the ``one-curve-to-rule-them-all'' is more than a
-mathematical tautology, then the free energy of repressor binding
-inferred from fold-change measurements should agree with
-repressor binding and unbinding rates.
-In other words, the equilbrium and nonequilibrium definitions of
-$\Delta F_R$ must agree, meaning
-\begin{equation}
-\Delta F_R = \beta\Delta\epsilon_R - \log(R/N_{NS})
-        = - \log(k_R^+/k_R^-),
-\end{equation}
-must hold, where $\beta\Delta\epsilon_R$ is inferred from the
-master curve fit to fold-change measurements, and $k_R^+$ and
-$k_R^-$ are inferred in some orthogonal manner.
-Single molecule measurements such as from~\cite{Hammar2014} have
-directly observed these rates, and in the remainder of this work
-we explore a complementary approach: inferring repressor binding
-and unbinding rates $k_R^+$ and $k_R^-$ from FISH measurements of
-mRNA population distributions.
-
-\section{Beyond means}
-\subsection{A one-state promoter with bursting is the ``best'' model of constitutive promoters}
-Before we can tackle simple repression, we need an adequate
-phenomenological model of constitutive expression.
-The literature abounds with options from which we can choose,
-and we show several potential kinetic models for constitutive promoters
-in~\fig{fig:constit_cartoons}.
-Let us consider the suitability of each model for our purposes in turn.
-
-\begin{figure}%[h!]
-\centering
-\includegraphics[width=0.9\textwidth]{../figures/fig1.5/fig1point5.ai}
-\caption{\textbf{Constitutive promoter cartoon comparison.}
-The left column depicts various plausible cartoons for constitutive promoters.
-We also list literature references where models have been used previously.
-In model (1), transcripts are produced in a Poisson process~\cite{Jones2014}.
-Model (2) features explicit modeling of RNAP binding/unbinding
-kinetics~\cite{Phillips2015a}.
-Model (3) is a more detailed generalization of model (2),
-treating transcription initiation as a multi-step process
-proceeding through closed and open complexes~\cite{Mitarai2015}.
-Model (4) is somewhat analogous to (2) except with the precise nature
-of active and inactive states left ambiguous~\cite{Razo-Mejia2020}.
-Finally, model (5) can be viewed as a certain limit of model (4)
-in which transcripts are produced in bursts,
-and initiation of bursts is a Poisson process.
-\mmnote{Haven't found model 5 formulated quite like this in the literature,
-usually burstiness is handled like model 4, but I should probably search a bit more.}
-The right column shows the Fano factor for each model.
-Note especially the crucial diagnostic: (2) and (3) have $\nu$
-strictly below 1, while only for (4) and (5) can $\nu$ exceed 1.
-Models with Fano factors $\le 1$ cannot produce the FISH data
-observed in~\cite{Jones2014} without introducing additional
-assumptions and model complexity.
-}
-\label{fig:constit_cartoons}
-\end{figure}
-
-\subsubsection{Model 1 - Poisson promoter.}
-This is the picture from Jones et.\ al.~\cite{Jones2014}, which
-assumes that transcripts are produced as a Poisson process from a
-single promoter state. This model insists that the ``true'' mRNA
-distribution is Poisson, implying the Fano factor $\nu$ must be 1.
-In~\cite{Jones2014}, the authors carefully attribute deviations
-from Fano = 1 to intensity variability in fluorescent spots, gene
-copy number variation, and copy number fluctuations of the
-transcription machinery, e.g., RNAP itself.
-In this picture, the master equation makes no appearance,
-and all the corrections to Poisson behavior are derived
-as additive corrections to the Fano factor.
-For disproving the ``universal noise curve'' from So et
-al~\cite{So2011}, this picture was excellent. It is appealing in
-its simplicity, with only two parameters, the initiation rate $r$
-and degradation rate $\gamma$. Since $\gamma$ is independently
-known from other experiments, and the mean mRNA copy number is
-$r/\gamma$, $r$ is easily inferred from data.
-In other words, the model is not excessively complex for the data at hand.
-But for many interesting questions, for instance in the recent
-work~\mmnote{cite Manuel's preprint once posted},
-knowledge of means and variances alone is insufficient, and a
-description of the full distribution of molecular counts is necessary.
-For this we need a( slightly) more complex model than model 1
-that would allow us to incorporate the non-Poissonian features of
-constitutive promoters directly into a master equation formulation.
-
-\subsubsection{Model 2 - Two-state promoter, RNAP bound or unbound.}
-This model was considered in, e.g.,~\cite{Phillips2015a} and ~\cite{Phillips2019}.
-Here transcription initiation proceeds from the bound to the unbound state,
-reflecting the microscopic reality that an RNAP that has begun to elongate
-a transcript is no longer available at the start site to begin another.
-The problem with this picture is that the Fano factor is
-\begin{align}
-    \nu = 1 -
-        \frac{r\rate{k}{P}{+}}
-            {\left(\rate{k}{P}{+} + \rate{k}{P}{-} + r\right)
-             \left(\gamma + \rate{k}{P}{+} + \rate{k}{P}{-} + r\right)},
-\end{align}
-which is always $<1$.
-To make contact with the experimental reality of $\nu>1$,
-we will have to do, at a minimum,
-the same corrections to Poisson behavior as in model 1 above.
-So while this model adds an appealing element of microscopic
-reality, we are forced to reject it as the additional complexity
-is unable to capture the phenomenology of interest.
-Obviously the promoter state does in fact proceed through cycles
-of RNAP binding, initiating, and elongating, but it seems that
-the super-Poissonian noise in mRNA copy number we want to
-model must be governed by other features of the system.
-
-\subsubsection{Model 3 - Three-state promoter, multistep
-transcription initiation and escape.}
-How might we remedy the deficits of model 2?
-It is known~\cite{DeHaseth1998} that once RNAP initially binds
-the promoter region, a multitude of distinct steps occur
-sequentially before RNAP finally escapes into the elongation
-phase. Perhaps adding some of this mechanistic detail might
-rescue model 2? The next simplest refinement of model 2 could
-consider open complex formation and promoter escape as separate
-steps rather than as a single effective step. In other words, we
-construct model 3 by adding a single extra state to model 2, and
-we will label the two RNAP-bound states as the closed and open
-complexes, despite the true biochemical details certainly being
-more complex. The authors of~\cite{Mitarai2015}, for instance,
-considered this model, although they added an additional
-repressor bound state and did not explicitly consider the limit
-with no repressor that we analyze here. Again, our goal here is
-not a complete accounting of all the relevant biochemical detail;
-this is an exploratory search for the important features our
-model needs to include.
-
-Unfortunately, as the authors of~\cite{Mitarai2015} hint, this
-model too has Fano factor $\nu<1$. This can be seen by starting
-with their Eq.~A1 for the Fano factor of the analogous model with
-repressor and taking the limit as repressor concentration goes to
-zero. After substantial algebra, the result is
-\begin{align}
-\nu = 1 - \frac{r k_O k_P^+}{\mathcal{Z}}
-\frac{k_P^+ + k_P^- + k_O + \gamma}
-        {\mathcal{Z} + \gamma(k_P^+ + k_P^- + k_O + \gamma) + \gamma^2},
-\end{align}
-where we defined $\mathcal{Z} = r(k_O + k_P^-) + k_P^+(k_O + r)$
-for notational tidiness. This is necessarily less than 1 for
-arbitrary rate constants.
-
-In fact, we suspect \textit{any} model in which transcription
-proceeds through a multistep cycle must necessarily have $\nu<1$.
-The intuitive argument compares the waiting time distribution to
-traverse the cycle with the waiting time for a Poisson promoter
-(model 1) with the same mean time. The latter is simply an
-exponential distribution. The former is a convolution of multiple
-exponentials, and intuitively should be more peaked with a
-smaller fractional width than an exponential with the same mean.\footnote{
-This can be made more precise with a result
-from~\cite{Stewart2007}, who showed that the convolution of
-multiple gamma distributions (of which the exponential
-distribution is a special case) is to a very good approximation
-also gamma distributed. Using their Eq.~2 for the distribution of
-the convolution, with shape parameters set to 1 to
-give exponential distributions, the total waiting time
-distribution has a ratio of variance to squared mean 
-$\sigma^2/\mu^2 = \sum_i k_i^2/\left(\sum_i k_i\right)^2$,
-where the $k_i$ are the rates of the individual steps. Clearly
-this is less than 1 and therefore the total waiting time
-distribution is narrower than the corresponding exponential.} A
-less disperse waiting time distribution means transcription
-initations are more uniformly distributed in time relative to a
-Poisson process. Hence the distribution of mRNA over a population
-of cells should be less variable compared to Poisson, giving
-$\nu<1$.\footnote{
-This last step, while intuitive, can be argued by analogy to
-photon statistics where antibunching gives rise to sub-Poissonian
-noise~\cite{Paul1982, Zou1990}. Although loopholes exist, we do
-not expect they apply for our problem. Nevertheless we refrain
-from elevating this cycle/sub-Poissonian equivalence to a ``theorem.''
-\mmnote{Antibunching is the ``obvious'' analogy for me but it'd
-be totally out of left field for most readers, any suggestions on
-a more comprehensible reference??}
-}
-
-Regardless of the merits of this model in describing the noise
-properties of constitutive transcription initation, it does not
-capture the dominant source of noise we want to include in our
-phenomenological model, so we must discard these details and
-search elsewhere.
-
-\subsubsection{Model 4 - Two-state promoter,
-``active'' and ``inactive'' states.}
-Inspired by~\cite{Razo-Mejia2020}, we revisit an analog of model
-2, but the interpretation of the two states is changed. Rather
-than explicitly viewing them as RNAP bound and unbound, we view
-them as ``active'' and ``inactive,'' which are able and unable to
-initiate transcripts, respectively. We are noncommittal as to the
-microscopic details of these states.
-
-One interpretation~\cite{Chong2014, Sevier2016} for the active
-and inactive states is that it represents the promoter's
-supercoiling state: transitions to the inactive state are caused
-by accumulation of positive supercoiling, which inhibits
-transcription, and transitions back to ``active'' are caused by
-gyrase or other topoisomerases relieving the supercoiling. This
-is an interesting possibility because it would mean the timescale
-for promoter state transitions is driven by topoisomerase
-kinetics, not by RNAP kinetics. From in vitro measurments, the
-former are suggested to be of order minutes~\cite{Chong2014}.
-Contrast this with model 2, where the state transitions are
-assumed to be governed by RNAP, which, assuming a copy number per
-cell of order $10^3$, has a diffusion-limited association rate
-$k_{on} \sim 10^2~\text{s}^{-1}$ to a target promoter. Combined
-with known $K_D$'s of order $\mu$M, this gives an RNAP
-dissociation rate $k_{off}$ of order $10^2$. As we will show
-below, however, there are some lingering puzzles with
-interpreting this supercoiling interpretation, so we leave it as
-a speculation and refrain from assigning definite physical
-meaning to the two states in this model.
-
-Intuitively one might expect that, since transcripts are produced
-as a Poisson process only when the promoter is in one of the two
-states in this model, transcription initiations should now be
-``bunched,'' in constrast to the ``anti-bunching'' of models 2
-and 3 above. One might further guess that this bunching would
-lead to super-Poissonian noise in the mRNA distribution over a
-population of cells. An honest calculation of the Fano factor produces
-\begin{align}
-\nu &= 1 + \frac{r k^-}{(k^+ + k^- + \gamma)(k^+ + k^-)},
-\end{align}
-which is strictly greater than 1, verifying the above intuition.
-Note we have dropped the $P$ label on the promoter switching rates to
-emphasize that these very likely do not represent kinetics of RNAP itself.
-This calculation can also be sidestepped by noting that the model
-is mathematically equivalent to the simple repression repression model
-from~\cite{Jones2014}, with states and rates relabeled and reinterpreted.
-
-How does this model compare to model 1 above?
-In model 1, all non-Poisson features of the mRNA
-distribution were handled as extrinsic corrections.
-By contrast, here the 3 parameter model is used to fit the full
-mRNA distribution as measured in mRNA FISH experiments.
-In essense, all variability in the mRNA distribution is regarded
-as ``intrinsic,'' arising either from stochastic initiation or
-from switching between the two coarse-grained promoter states.
-The advantage of this approach is that it fits neatly into the
-master equation picture, and the parameters thus inferred can be used as
-input for more complicated models with regulation by transcription factors.
-
-While this seems promising, there is a major drawback for our purposes
-which was already uncovered by the authors of~\cite{Razo-Mejia2020}:
-the statistical inference problem is nonidentifiable,
-in the sense described in Section 4.3 of~\cite{Gelman2013},
-meaning it is impossible to infer the
-parameters $r$ and $k^-$ from the FISH data of~\cite{Jones2014}
-(as shown in Fig.~S2 of~\cite{Razo-Mejia2020}).
-Rather, only the ratio $r/k^-$ could be inferred. In that work,
-the problem was worked around with an informative prior on the
-ratio $k^-/k^+$. That approach is unlikely to work here, as,
-recall, our entire goal in modeling constitutive expression is to
-use it as the basis for a yet more complicated model, when we add
-on repression. But adding more complexity to a model that is
-already poorly identified is a fool's errand,
-so we will explore one more potential model.
-
-\subsubsection{Model 5 - One-state promoter with explicit bursts.}
-This model is inspired by the failure mode of model 4.
-The key observation above was that, as found
-in~\cite{Razo-Mejia2020}, only two parameters, $k^+$ and the
-ratio $r/k^-$, could be directly inferred from the FISH data
-from~\cite{Jones2014}.
-So let us take this seriously and imagine a model where these are
-the only two model parameters. What would this model look like?
-
-To develop some intuition, consider model 4 in the limit
-$k^+ \ll k^- \lesssim r$, which is roughly satisfied by the
-parameters inferred in~\cite{Razo-Mejia2020}.
-In this limit, the system spends the majority of its time in the inactive state, occasionally becoming active and making a burst of transcripts.
-This should call to mind the well-known phenomenon of transcriptional bursting, as reported in, e.g.,~\cite{Golding2005,Chong2014,Sevier2016}\mmnote{should probably add some more cites here}.
-Let us make this correspondence more precise.
-The mean dwell time in the active state is $1/k^-$.
-While in this state, transcripts are produced at a rate $r$ per
-unit time. So on average, $r/k^-$ transcripts are produced before
-the system switches to the inactive state. Once in the inactive
-state, the system dwells there for an average time $1/k^+$ before
-returning to the active state and repeating the process. $r/k^-$
-resembles an average burst size, and $1/k^+$ resembles the time
-interval between burst events. More precisely, $1/k^+$ is the
-mean time between the end of one burst and the start of the next,
-whereas $1/k^+ + 1/k^-$ would be the mean interval between the
-start of two successive burst events, but in the limit $k^+ \ll
-k^-$, $1/k^+ + 1/k^- \approx 1/k^+$. Note that this limit ensures
-that the waiting time between bursts is approximately
-exponentially distributed, with mean set by the only timescale
-left in the problem,
-$1/k^+$.\footnote{
-If instead it were the case that $k^+ \sim k^-$, then the waiting time
-$1/k^+ + 1/k^-$ would have a peaked distribution, but this does
-not appear to be the case for any of datasets
-from~\cite{Jones2014}.}
-
-Let us now verify this intuition with a precise derivation
-to check that $r/k^-$ is in fact the mean burst size
-and to obtain the full burst size distribution.
-Consider first a constant, known dwell time $T$ in the active state.
-Transcripts are produced at a rate $r$ per unit time, so the
-number of transcripts $n$ produced during $T$ fits exactly
-the ``story'' for the Poisson distribution, i.e.,
-\begin{equation}
-    P(n\mid T) = \frac{(rT)^n}{n!} \exp(-rT).
-\end{equation}
-Since the dwell time $T$ is unobservable, we actually want $P(n)$,
-the dwell time distribution with no conditioning on $T$.
-Basic rules of probability theory tell us we can write $P(n)$
-in terms of $P(n\mid T)$ as
-\begin{equation}
-    P(n) =\int_0^\infty P(n\mid T) P(T) dT.
-\end{equation}
-But we know the dwell time distribution $P(T)$,
-which is exponentially distributed according to
-\begin{equation}
-    P(T) = k^- \exp(-T k^-),
-\end{equation}
-so $P(n)$ can be written as
-\begin{equation}
-    P(n) = k^- \frac{r^n}{n!}
-            \int_0^\infty T^n\exp[-(r + k^-)T]\,dT.
-\end{equation}
-A standard integral table shows
-$\int_0^\infty x^n e^{-ax}\,dx = n!/a^{n+1}$, so
-\begin{equation}
-    P(n) = \frac{k^- r^n}{(k^- + r)^{n+1}}
-        = \frac{k^-}{k^- + r}
-            \left(\frac{r}{k^- + r}\right)^n
-        = \frac{k^-}{k^- + r}
-            \left(1 - \frac{k^-}{k^- + r}\right)^n,
-\end{equation}
-which is exactly the geometric distribution with standard parameter
-$\theta\equiv k^-/(k^- + r)$ and domain
-$n \in \{0, 1, 2, \dots\}$
-(this is one of two common conventions for the geometric distribution).
-The mean of the geometric distribution, with this convention, is
-\begin{align}
-\langle n\rangle = \frac{1 - \theta}{\theta}
-        = \left(1 - \frac{k^-}{k^- + r}\right)
-                    \frac{k^- + r}{k^-}
-        = \frac{r}{k^-},
-\end{align}
-exactly as we guessed intuitively above.
-
-So in taking the limit $r,k^-\rightarrow\infty$, $r/k^-\equiv b$,
-we obtain a model which effectively has only a single promoter state,
-which initiates bursts at rate $k^+$
-(transitions to the active state, in the model 4 picture).
-The master equation for mRNA copy number $m$ takes the form
-\begin{align}
-\begin{split}
-\deriv{t}{p(m,t)} = & (m+1)\gamma p(m+1,t) - m\gamma p(m,t) \\
-        &+ \sum_{j=0}^{m-1} k^+ p(j,t) Geom(m-j;b)
-         - \sum_{j=m+1}^\infty k^+ p(m,t) Geom(j-m;b),
-\end{split}
-\end{align}
-where $Geom(n;b)$ is the geometric distribution with mean~$b$, i.e.,
-$Geom(n;b) = \frac{1}{1+b}\left(\frac{b}{1+b}\right)^n$
-(with domain over nonnegative integers as above).
-The first two terms are the usual mRNA degradation terms.
-The third term enumerates all ways the system can produce
-a burst of transcripts and arrive at copy number $m$,
-given that it had copy number $j$ before the burst.
-The fourth term allows the system to start with copy number $m$,
-then produce a burst and end with copy number $j$.
-In fact this last sum has trivial $j$ dependence and
-simply enforces normalization of the geometric distribution.
-Carrying it out we have
-\begin{equation}
-\begin{split}
-\deriv{t}{p(m,t)} = & (m+1)\gamma p(m+1,t) - m\gamma p(m,t) \\
-        &+ \sum_{j=0}^{m-1} k^+ p(j,t) Geom(m-j;b)
-            - k^+ p(m,t),
-\end{split}
-\end{equation}
-This improves on model 4 in that now the parameters are easily inferred,
-as we will see later, and have clean interpretations.
-The non-Poissonian features are attributed to the emprically
-well-established phenomenological picture of bursty transcription.
-
-The big approximation in going from model 4 to 5 is that a burst
-is produced instantaneously rather than over a finite time.
-If the true burst duration is not short compared to transcription factor
-kinetic timescales, this could be a problem in that mean burst size
-in the presense and absence of repressors could change,
-rendering parameter inferences from the constitutive case inappropriate.
-Let us make some simple estimates of this.
-
-Consider the time delay between the first and final RNAPs
-in a burst initiating transcription
-(\textit{not} the time to complete transcripts,
-which potentially could be much longer.)
-If this timescale is short compared to the typical
-search timescale of transcription factors, then all is well.
-The estimates from deHaseth et.\ al.~\cite{DeHaseth1998}
-put RNAP's diffusion-limited on rate around
-$\sim\text{few}\times10^{-2}~\text{nM}^{-1}~\text{s}^{-1}$
-and polymerase loading as high as $1~\text{s}^{-1}$.
-Then for reasonable burst sizes of $<10$, it is reasonable to guess that
-bursts might finish initiating on a timescale of tens of seconds or less
-(with another 30-60 sec to finish elongation, but that does not matter here).
-A transcription factor with typical copy number of order 10 (or less)
-would have a diffusion-limited association rate of order
-$(10~\text{sec})^{-1}$.
-Higher copy number TFs tend to have many binding sites over the genome,
-which should serve to pull them out of circulation and keep their
-effective association rates from rising too large.
-Therefore, there is \textit{perhaps} a timescale separation
-possible between transcription factor association rates and burst
-durations, but this assumption could very well break down, so we
-will have to keep it in mind when we infer repressor rates from
-the Jones et.\ al.\ FISH data later.
-
-In reflecting on these 5 models, the reader may feel that
-exploring a multitude of potential models just to return to a
-very minimal phenomenological model of bursty transcription may
-seem highly pedantic. But the purpose of the exercise was to
-examine a host of models from the literature and understand why
-they are insufficient, one way or another, for our purposes.
-Along the way we have learned that the detailed kinetics of RNAP
-binding and initiating transcription are probably irrelevant for
-setting the population distribution of mRNA.
-The timescales are simply too fast, and as we will see later
-in~\fig{fig:constit_post}, the noise seems to be governed by
-slower timescales. Perhaps in hindsight this is not surprising:
-intuitively, the degradation rate $\gamma$ sets the fundamental
-timescale for mRNA dynamics, and any other processes that
-substantially modulate the mRNA distribution should not differ
-from $\gamma$ by orders of magnitude.
-
-\subsection{Parameter inference for constitutive promoters}
-Having established that a one-state bursty model achieves a
-Goldilocks level of complexity, we test it more thoroughly on all
-the data from constitutive promoters from~\cite{Jones2014}.
-\fig{fig:constit_post} shows the results.
-\begin{figure}%[h!]
-\centering
-\includegraphics[width=0.9\textwidth]{../figures/fig2/fig2pt1.pdf}
-\caption{\textbf{Constitutive promoter posterior
-        inference and predictive checks.}
-        (A) shows the joint posterior density from MCMC sampling of model 5,
-        bursty promoter with negative binomially-distributed steady state.
-        (B) shows the ECDF of the observed population distribution
-        of mRNA transcripts under the control of a constitutive lacUV5 promoter.
-        Posterior predictive percentiles are shown as colored bands for
-        model (1), Poisson, and model (5), negative binomial.
-        }
-\label{fig:uv5_post}
-\end{figure}
-
-\begin{figure}%[h!]
-\centering
-\includegraphics[width=0.9\textwidth]{../figures/fig2/fig2pt2.pdf}
-\caption{\textbf{Constitutive promoter model comparison.}
-    (A) Joint posterior distributions for burst rate $k_i$ and
-    mean burst size $b$ for 18 promoters from Jones 2014.
-    Each contour indicates the 95\% highest posterior density
-    region for a particular promoter, i.e., on a plot analogous
-    to~\fig{fig:uv5_post}, the contour would enclose
-    approximately 95\% of the samples, and thus 95\% of the
-    probability mass.
-    Note that the vertical axis is shared with (B).
-    Panel (B) plots the burst rate $k_i$ vs.\ binding energy from energy
-    matrices from Brewster 2012.
-    The dotted line shows the predicted slope according to
-    $\ln k_i/\gamma \sim \text{constant} - \beta\Delta\epsilon_P$,
-    described in text.
-    }
-\label{fig:constit_post}
-\end{figure}
-\mmnote{Discuss puzzling comparison w/ Chong2014: if supercoiling is the thing, why are my burst sizes all the same but burst rates vary?
-And why is the duty cycle of my promoter so low, ie., bursts so short?
-COmpare their fig 7E; their $\beta/\alpha$ is my $k^+/k^-$.
-They have one or two genes with very small $\beta/\alpha$, does the \textit{galK} locus just happen to be that, or is there a deeper disagreement?
-Hard to say w/o more data.}
-
-\subsection{Transcription factor kinetics can be inferred from FISH measurements}
-Figure~3 presents inferred transcription factor rates, compares
-them to single-molecule measurements, and compares their ratios
-to gold-standard binding energies.
-\begin{figure}%[h!]
-\centering
-\includegraphics[width=0.9\textwidth]{../figures/fig3/fig3.ai}
-\caption{\textbf{Simple repression model comparison.}
-(A) depicts the cartoon of simple repression for which we infer parameters.
-(B) shows 50\% and 95\% contours of several 2D slices of the 9D posterior
-distribution, from fitting one unbinding rate for each operator (O1, O2, O3)
-and one binding rate for each aTc concentration (corresponding to an
-unknown mean repressor copy number).
-(C) compares ratios of our inferred unbinding rates with operator
-binding energy differences measured in Garcia \& Phillips 2011 (triangles)
-and Razo-Mejia et.\ al.\ 2018 (squares).
-(D) compares our inferred unbinding rates with single-molecule measurements
-of the same from Hammar et.\ at.\ 2014 \mmnote{Johan Elf's group}.
-Values are of similar order of magnitude,
-but a precise quantitative comparison unfortunately depends sensitively
-on the assumed mRNA lifetime, which was not measured.
-    }
-\label{fig3:kR_inferences}
-\end{figure}
-
-\begin{figure}%[h!]
-\centering
-\includegraphics[width=0.9\textwidth]{../figures/figSIxx/ppc_many_pooled.pdf}
-\caption{\textbf{Posterior predictive checks.}
-(A) duplicates the posterior from~\fig{fig3:kR_inferences}B.
-(B), (C), and (D) show samples from the posterior predictive distribution
-for each of the datasets used in the inference.
-Samples are sorted by operator and plotted separately simply for visual clarity.
-The unrepressed promoter, UV5, is shown with each as a reference point.
-\mmnote{
-Obviously the PPC is not perfect,
-but considering the relative simplicity of the model, I'm pleasantly surprised
-how well it worked and that the model was even identifiable at all.
-Interestingly, and perhaps not surprisingly, only 3 of the 9 datasets
-were identifiable if fit in isolation: otherwise,
-it was only by making assumptions that certain rates are equal
-across datasets was identifiability achieved.
-}
-    }
-\label{fig4:ppc}
-\end{figure}
-
-\section{Discussion and future work}
-\mmnote{
-Do all my approximations work simply because galK happens to have
-a low duty cycle of supercoiling, $\beta/\alpha$ in Chong2014
-picture? A potentially interesting test: put in the dozen or so
-Hernan constructs at some other locus, preferably one of the ones
-Chong identified as having a large duty cycle (large $\beta/\alpha$),
-i.e., the opposite limit of galK. Then just measure fold-changes.
-You don't even need to do FISH, although that would be even
-better. If their model is right, $\rho$ should be approximately
-set by the duty cycle of supercoiling, NOT by RNAP kinetics.
-\textit{This is a NICE prediction for future work.} \textit{cspE}
-is claimed nonessential, prediction is you drop HG reporter
-constructs in there and the inferred binding energies of all lacI
-operators shift the same amount, by about $\ln(3)\sim 1kT$. Not
-huge, but detectable, although might be confounding effects from
-moving to a new locus. With FISH measurements, $k^+$ and $k^-$ in
-Fig 2, model 4 should now be separately inferrable if Chong2014
-measurements of $\beta/\alpha$ are correct. This would require an
-obnoxious amount of cloning, but once you did that, the
-measurements would be same old stuff that RP lab does.}
-
-%%%%%%%%%%%%%%%%%%%%% APPENDICES %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\appendix
-
+% !TEX root = ./busty_transcription.tex
 \section{Derivations of constitutive promoter models}
@@
 \mmnote{Obtain Fano factor by relabeling cartoon from Jones2014.}
-
-\section{Generating function solution for bursty promoters}
-
-\subsection{Constitutive promoter with bursts}
-
-\subsubsection{From master equation to generating function}
-
-We would like to derive the full probability distribution of
-mRNA copy number per cell for a model combining simple repression
-with a promoter featuring bursty transcription initiation.
-Before tackling mRNA and repression together, let us
-just consider the bursty consititutive promoter,
-model 5 in~\fig{fig:constit_cartoons}
-\mmnote{replicating Charlotte's notes; she says it's worked out
-in~\cite{Paulsson2000}, but I don't see it, at least not in notation I can
-comprehend. Need to find a cite for this,
-or maybe just cite Charlotte, personal comm, since that's where I actually got it\dots}.
-The master equation of interest is
-\begin{align}
-\deriv{t}p(m,t) = (m+1)\gamma p(m+1,t) - m\gamma p(m,t) - r p(m,t)
-        + r \sum_{m^\prime=0}^m G_{m-m^\prime}(\theta) p(m^\prime,t),
-\label{eq:1state_unreg_003}
-\end{align}
-where $G_{k}(\theta)$ is the geometric distribution defined as
-\begin{align}
-G_{k}(\theta) = \theta(1 - \theta)^k, \, k\in\{0,1,2,\dots\}.
-\end{align}
-With this convention, the mean burst size $b = (1-\theta)/\theta$.
-$\gamma$ and $r$ are mRNA degradation rates and transcription burst
-initiation rates, resp. The last term represents all ways the system could
-end up with $m$ mRNAs, having started the burst with $m^\prime$. Define
-$\lambda = r/\gamma$ and nondimensionalize time by $\gamma$, giving
-\begin{align}
-\deriv{t}p(m,t) = (m+1)p(m+1,t) - m p(m,t) - \lambda p(m,t)
-        + \lambda \sum_{m^\prime=0}^m G_{m-m^\prime}(\theta) p(m^\prime,t),
-\end{align}
-The probability generating function is defined as
-\begin{align}
-F(z,t) = \sum_{m=0}^\infty z^m p(m,t).
-\end{align}
-Multiply both sides of the CME by $z^m$ and sum over all $m$ to get
-\begin{align}
-\pderiv[F]{t} = \sum_{m=0}^\infty z^m
-\left(
-(m+1)p(m+1,t) - m p(m,t) - \lambda p(m,t)
-+ \lambda \sum_{m^\prime=0}^m G_{m-m^\prime}(\theta) p(m^\prime,t)
-\right).
-\end{align}
-The first two terms can be rewritten as derivatives of the
-generating function, i.e.,
-\begin{align}
-\sum_{m=0}^\infty z^m (m+1)p(m+1,t)
-=
-\pderiv{z} \sum_{m=0}^\infty z^{m+1} p(m+1,t)
-=
-\pderiv[F]{z}
-\end{align}
-and
-\begin{align}
-\sum_{m=0}^\infty z^m mp(m,t)
-=
-z\pderiv{z} \sum_{m=0}^\infty z^m p(m,t)
-=
-z\pderiv[F]{z}.
-\end{align}
-The final term is the most trouble. The trick is to reverse
-the default order of the sums, as in
-\begin{align}
-&\lambda \sum_{m=0}^\infty \sum_{m^\prime=0}^m
-        z^m \theta(1 - \theta)^{m-m^\prime} p(m^\prime,t)
-\\=
-&\lambda\theta \sum_{m^\prime=0}^\infty z^{m^\prime} p(m^\prime,t)
-        \sum_{m=m^\prime}^\infty
-        z^{m-m^\prime} (1 - \theta)^{m-m^\prime}
-\\=
-&\frac{\lambda\theta F}{1 - (1 - \theta)z}.
-\end{align}
-This works because after reordering the sums, the sum on $m$ is just
-a geometric series. Its result is independent of $m^\prime$, so the sum on
-$m^\prime$ reduces to the definition of the generating function.
-Putting these results together, the PDE for the generation function is
-\begin{align}
-\pderiv[F]{t} = \pderiv[F]{z} - z\pderiv[F]{z} - \lambda F
-        + \frac{\lambda\theta F}{1-z(1-\theta)}.
-\end{align}
-Changing variables to $\xi=1-\theta$ and simplifying gives
-\begin{align}
-\pderiv[F]{t} + (z - 1) \pderiv[F]{z} = \frac{(z-1)\xi}{1-z\xi}\lambda F.
-\label{eq:1state_unreg_015}
-\end{align}
-\subsubsection{Stead-state}
-At steady-state, the PDE reduces to the ODE
-\begin{align}
-\deriv[F]{z} = \frac{\xi}{1-z\xi}\lambda F,
-\end{align}
-which we can integrate as
-\begin{align}
-\int \frac{dF}{F} = \int \frac{\lambda\xi dz}{1-\xi z}.
-\end{align}
-The initial conditions for generating functions can be subtle and confusing.
-The key fact follows from the definition
-$F(z,t) = \sum_m z^m p(m,t)$,
-normalization requires that
-$F(z=1^-,t) = \sum_m p(m,t) = 1$.\footnote{
-Sometimes the generating function may be undefined \textit{at} $z=1$ but the
-limit still holds. Also many authors change variables from $z$ to other
-things, so one must keep track of how the normalization condition transforms.
-}
-Doing the integrals (and producing constant $c$) gives
-\begin{align}
-\ln F &= -\lambda \ln(1-\xi z) + c
-\\
-F &= \frac{c}{(1-\xi z)^\lambda}.
-\end{align}
-Only one choice for $c$ can satisfy initial conditions, producing
-\begin{align}
-F(z) = \left(\frac{1-\xi}{1-\xi z}\right)^\lambda
-        = \left(\frac{\theta}{1 - z(1-\theta)}\right)^\lambda,
-\end{align}
-which is exactly the negative binomial's generating function, as expected.
-
-\subsection{Adding repression}
-\subsubsection{Deriving the generating function for mRNA distribution}
-
-Let us move from a one-state promoter to a two-state promoter, where one state
-has repressor bound and the other produces transcriptional bursts as above.
-A schematic of this model is shown in~\fig{fig3:kR_inferences}.
-Although now we have an equation for each promoter state, otherwise
-the master equation reads similarly to the one-state case, except with
-additional terms corresponding to transitions between promoter states, namely
-\begin{align}
-\deriv{t}p_R(m,t) =& k_R^+ p_A(m,t) - k_R^- p_R(m,t)
-        + (m+1)\gamma p_R(m+1,t) - m\gamma p_R(m,t)
-\\
-\begin{split}
-\deriv{t}p_A(m,t) =& - k_R^+ p_A(m,t) + k_R^- p_R(m,t)
-        + (m+1)\gamma p_A(m+1,t) - m\gamma p_A(m,t) 
-\\
-&- k_i p_A(m,t) + k_i \sum_{m^\prime=0}^m \theta(1-\theta)^{m-m^\prime} p_A(m^\prime,t),
-\end{split}
-\end{align}
-where $p_R(m,t)$ is the probability of the system having $m$ mRNA copies
-and having repressor bound to the promoter at time $t$, and $p_A$ is an
-analogous probability to find the promoter without repressor bound.
-$k_R+$ and $k_R^-$ are, respectively, the rates at which repressors
-bind and unbind to and from the promoter, and
-$\gamma$ is the mRNA degradation rate.
-$k_i$ is the rate at which bursts initiate, and as before,
-the geometric distribution of burst sizes has mean $b=(1-\theta)/\theta$.\footnote{
-$\theta$ can be viewed as a Bernoulli trial probability: after each transcript
-in a burst is produced, $\theta$ is the probability that the burst
-terminates and $(1-\theta)$ is the probability that the burst continues.
-}
-
-Interestingly, it turns out that this problem maps exactly onto the
-three-stage promoter model considered by
-Shahrezaei and Swain in~\cite{Shahrezaei2008}, with relabelings.
-Their approximate solution for protein distributions amounts to
-the same approximation we make here in regarding the duration
-of mRNA synthesis bursts as instantaneous, so their solution
-for protein distributions also solves our problem of mRNA distributions.
-Let us examine the analogy more closely.
-They consider a two-state promoter, as we do here, but they model
-mRNA as being produced one at a time and degraded, with rates
-$v_0$ and $d_0$.
-Then they model translation as occuring with rate $v_1$,
-and protein degradation with rate $d_1$.
-Now consider the limit where $v_1, d_0\rightarrow\infty$ with
-their ratio $v_1/d_0$ held constant.
-$v_1/d_0$ resembles the average burst size of translation from a single
-mRNA: these are the rates of two Poisson processes that compete over a
-transcript, which matches the story of geometrically distributed burst sizes.
-So after taking this limit, on timescales slow compared to $v_1$ and $d_0$,
-it appears that transcription events fire at rate $v_0$ and produce a
-geometrically distributed burst of translation of mean size $v_1/d_0$,
-which intuitively matches the story we have told above for mRNA
-with variables relabeled.
-
-To verify this intuitively conjectured mapping between our problem and
-the solution in~\cite{Shahrezaei2008}, we continue with a careful
-solution for the mRNA distribution using probability generating functions,
-following the ideas sketched in~\cite{Shahrezaei2008}.
-It is natural to nondimensionalize rates in the problem by $\gamma$,
-or equivalently, this amounts to measuring time in units of $\gamma^{-1}$.
-We are also only interested in steady state, so we set the time
-derivatives to zero, giving
-\begin{align}
-0 =& k_R^+ p_A(m) - k_R^- p_R(m) + (m+1) p_R(m+1) - m p_R(m)
-\\
-\begin{split}
-0 =& - k_R^+ p_A(m) + k_R^- p_R(m) + (m+1) p_A(m+1) - m p_A(m) 
-\\
-&- k_i p_A(m) + k_i \sum_{m^\prime=0}^m \theta(1-\theta)^{m-m^\prime} p_A(m^\prime),
-\end{split}
-\end{align}
-
-The probability generating function is defined as before in the
-constitutive case, except now we must introduce a generating function
-for each promoter state,
-\begin{align}
-f_A(z) = \sum_{m=0}^\infty z^m p_A(m),
-\;
-f_R(z) = \sum_{m=0}^\infty z^m p_R(m).
-\end{align}
-Our real objective is the generating function $f(z)$ that generates
-the mRNA distribution $p(m)$, independent of what state the promoter is in.
-But since $p(m) = p_A(m) + p_R(m)$, it follows too that $f(z) = f_A(z) + f_R(z)$.
-
-So as before we multiply both equations by $z^m$ and sum over all $m$.
-Each individual term transforms exactly as did an analogous term in
-the constitutive case, so the coupled ODEs for the generating functions read
-\begin{align}
-0 =& k_R^+ f_A(z) - k_R^- f_R(z) + \pderiv{z} f_R(z) - z \pderiv{z} f_R(z)
-\\
-\begin{split}
-0 =&  - k_R^+ f_A(z) + k_R^- f_R(z) + \pderiv{z} f_A(z) - z \pderiv{z} f_A(z)
-\\
-&- k_i f_A(z) + k_i \frac{\theta}{1-z(1-\theta)} f_A(z),
-\end{split}
-\end{align}
-and after changing variables $\xi = 1 - \theta$ as before
-and rearranging, we have
-\begin{align}
-0 &= k_R^+ f_A(z) - k_R^- f_R(z) + (1-z) \pderiv{z} f_R(z)
-\\
-0 &=  - k_R^+ f_A(z) + k_R^- f_R(z) + (1 - z) \pderiv{z} f_A(z)
-+ k_i \frac{(z-1)\xi}{1-z\xi} f_A(z),
-\end{align}
-We can transform this problem from two coupled first-order ODEs to a
-single second-order ODE by solving for $f_A$ in the first and plugging
-into the second, giving
-\begin{align}
-\begin{split}
-0 = (1&-z) \pderiv[f_R]{z}
-+ \frac{1-z}{k_R^+}
-        \left(k_R^- \pderiv[f_R]{z} + \pderiv[f_R]{z} +(z-1) \psecderiv[f_R]{z}\right)
-\\
-&+ \frac{k_i}{k_R^+} \frac{(z-1)\xi}{1-z\xi}
-        \left(k_R^- f_R + (z-1) \pderiv[f_R]{z}\right),
-\end{split}
-\end{align}
-where, to reduce notational clutter, we have dropped the explicit $z$
-dependence of $f_A$ and $f_R$. Simplifying we have
-\begin{align}
-0 = \psecderiv[f_R]{z}
-        - \left(\frac{k_i\xi}{1-z\xi}
-                + \frac{1 + k_R^- + k_R^+}{1-z}
-        \right)\pderiv[f_R]{z}
-        + \frac{k_i k_R^- \xi}{(1-z\xi)(1-z)}f_R.
-\end{align}
-This can be recognized as the hypergeometric differential equation,
-with singularities at $z=1$, $z=\xi^{-1}$, and $z=\infty$. The latter
-can be verified by a change of variables from $z$ to $x=1/z$,
-being careful with the chain rule, and noting that $z=\infty$ is a
-singular point if and only if $x=1/z=0$ is a singular point.
-
-The standard form of the hypergeometric differential equation has its
-singularities at 0, 1, and $\infty$, so to take advantage of the
-standard form solutions to this ODE, we first need to transform
-variables to put it into a standard form.
-However, this is subtle. While any such transformation should work in
-principle, the solutions are expressed most simply in the neighborhood
-of $z=0$, but the normalization condition that we need to enforce
-corresponds to $z=1$. The easiest path, therefore, is to find a change
-of variables that maps
-1 to 0, $\infty$ to $\infty$, and $\xi^{-1}$ to 1.
-This is most intuitively done in two steps.
-
-First map the $z=1$ singularity to 0 by the change of variables $v=z-1$, giving
-\begin{align}
-0 = \psecderiv[f_R]{v}
-        + \left(\frac{k_i\xi}{(1+v)\xi - 1}
-                + \frac{1 + k_R^- + k_R^+}{v}
-        \right)\pderiv[f_R]{v}
-        + \frac{k_i k_R^- \xi}{((1+v)\xi - 1)v}f_R.
-\end{align}
-Now two singularities are at $v=0$ and $v=\infty$.
-The third is determined by $(1+v)\xi -1 = 0$, or $v=\xi^{-1} - 1$.
-We want another variable change that maps this third singularity to 1
-(without moving 0 or infinity). Changing variables again to
-$w=\frac{v}{\xi^{-1} - 1} = \frac{\xi}{1-\xi} v$ fits the bill.
-In other words, the combined change of variables
-\begin{align}
-w = \frac{\xi}{1-\xi} (z-1)
-\end{align}
-maps $z = \{1, \xi^{-1}, \infty\}$ to $w =\{0, 1, \infty\}$ as desired.
-Plugging in, being mindful of the chain rule and noting
-$(1 + v)\xi - 1 = (1 - \xi)(w - 1)$ gives
-\begin{align}
-0 = \left(\frac{\xi}{1-\xi}\right)^2 \psecderiv[f_R]{w}
-+ \left(
-        \frac{\xi k_i}{(1-\xi)(w-1)} + \frac{\xi(1 + k_R^- + k_R^+)}{(1-\xi)w}
-\right) \frac{\xi}{1-\xi} \pderiv[f_R]{w}
-+ \frac{k_i k_R^- \xi^2}{(1-\xi)^2 w(w-1)}f_R.
-\end{align}
-This is close to the standard form of the hypergeometric differential
-equation, and some cancellation and rearrangement gives
-\begin{align}
-0 = w(w-1)\psecderiv[f_R]{w}
-+ \left(k_i w + (1 + k_R^- + k_R^+)(w-1)\right) \pderiv[f_R]{w}
-+ k_i k_R^- f_R.
-\end{align}
-and a little more algebra produces
-\begin{align}
-0 = w(1-w)\psecderiv[f_R]{w}
-+ \left(1 + k_R^- + k_R^+
-        - (1 + k_i + k_R^- + k_R^+)w
-\right) \pderiv[f_R]{w}
-- k_i k_R^- f_R,
-\end{align}
-which is the standard form. From this we can read off the solution
-in terms of hypergeometric functions ${_2F_1}$ from
-any standard source, e.g.~\cite{Abramowitz1964}, and identify the
-conventional parameters in terms of our model parameters.
-We want the general solution in the neighborhood of $w=0$ ($z=1$),
-which for a homogeneous linear second order ODE must be a sum of two
-linearly independent solutions.
-More precisely,
-\begin{align}
-f_R(w) = C_1 {_2F_1}(\alpha, \beta, \gamma; w)
-+ C_2 w^{1-\gamma}{_2F_1}(1+\alpha-\gamma, 1+\beta-\gamma, 2-\gamma; w)
-\end{align}
-with parameters determined by
-\begin{align}
-\begin{split}
-\alpha\beta &= k_i k_R^-
-\\
-1+\alpha+\beta &= 1+k_i+k_R^-+k_R^+
-\\
-\gamma &= 1 + k_R^- + k_R^+
-\end{split}
-\end{align}
-and constants $C_1$ and $C_2$ to be set by boundary conditions.
-Solving for $\alpha$ and $\beta$, we find\footnote{
-Note that $\alpha$ and $\beta$ are interchangeable in the definition of
-${_2F_1}$ and differ only in the sign preceeding the radical.
-}
-\begin{align}
-\begin{split}
-\alpha &= \frac{1}{2}
-\left(k_i+k_R^-+k_R^+ + \sqrt{(k_i+k_R^-+k_R^+)^2 - 4k_i k_R^-}\right)
-\\
-\beta &= \frac{1}{2}
-\left(k_i+k_R^-+k_R^+ - \sqrt{(k_i+k_R^-+k_R^+)^2 - 4k_i k_R^-}\right)
-\\
-\gamma &= 1 + k_R^- + k_R^+.
-\end{split}
-\end{align}
-Since the normalization condition requires that $f_R$ be finite at $w=0$,
-we can immediately set $C_2=0$ to discard the second solution.
-This is because all the rate constants are strictly positive,
-so $\gamma>1$ and therefore $w^{1-\gamma}$ blows up as $w\rightarrow0$.
-Now that we have $f_R$, we would like to find the generating function
-for the mRNA distribution, $f(z) = f_A(z) + f_R(z)$.
-We can recover $f_A$ from our solution for $f_R$, namely
-\begin{align}
-f_A(z) = \frac{1}{k_R^+}\left(k_R^- f_R(z) + (z-1) \pderiv[f_R]{z}\right)
-\end{align}
-or
-\begin{align}
-f_A(w) = \frac{1}{k_R^+}\left(k_R^- f_R(w) + w \pderiv[f_R]{w}\right),
-\end{align}
-where in the second line we transformed our original relation between
-$f_R$ and $f_A$ to our new, more convenient, variable $w$.
-Plugging our solution for $f_R(w) = C_1{_2F_1}(\alpha, \beta, \gamma; w)$
-into $f_A$, we will require the differentiation rule for ${_2F_1}$,
-which tells us
-\begin{align}
-\pderiv[f_R]{w} = C_1\frac{\alpha\beta}{\gamma}
-                {_2F_1}(\alpha+1, \beta+1, \gamma+1; w),
-\end{align}
-from which it follows that
-\begin{align}
-f_A(w) = \frac{C_1}{k_R^+}
-\left(
-k_R^- {_2F_1}(\alpha, \beta, \gamma; w)
-+ w\frac{\alpha\beta}{\gamma} {_2F_1}(\alpha+1, \beta+1, \gamma+1; w)
-\right)
-\end{align}
-and therefore
-\begin{align}
-f(w) = C_1\left(1 + \frac{k_R^-}{k_R^+}\right)
-        {_2F_1}(\alpha, \beta, \gamma; w)
-+ w \frac{C_1}{k_R^+} \frac{\alpha\beta}{\gamma}
-        {_2F_1}(\alpha+1, \beta+1, \gamma+1; w).
-\end{align}
-To proceed, we need one of the (many) useful identities known for
-hypergeometric functions, in particular
-\begin{align}
-w\frac{\alpha\beta}{\gamma} {_2F_1}(\alpha+1, \beta+1, \gamma+1; w)
-=
-(\gamma-1)\left(
-{_2F_1}(\alpha, \beta, \gamma-1; w) - {_2F_1}(\alpha, \beta, \gamma; w)
-\right).
-\end{align}
-Substituting this for the second term in $f(w)$, we find
-\begin{align}
-f(w) = \frac{C_1}{k_R^+}
-\left[
-        \left(k_R^+ + k_R^-\right)
-        {_2F_1}(\alpha, \beta, \gamma; w)
-+ (\gamma-1)\left(
-        {_2F_1}(\alpha, \beta, \gamma-1; w) - {_2F_1}(\alpha, \beta, \gamma; w)
-        \right)
-\right],
-\end{align}
-and since $\gamma-1 = k_R^+ + k_R^-$, the first and third terms cancel,
-leaving only
-\begin{align}
-f(w) = C_1\frac{k_R^+ + k_R^-}{k_R^+} {_2F_1}(\alpha, \beta, \gamma-1; w).
-\end{align}
-Now we enforce normalization, demanding $f(w=0) = f(z=1) = 1$.
-${_2F_1}(\alpha, \beta, \gamma-1; 0) = 1$, so we must have
-$C_1 = k_R^+ / (k_R^+ + k_R^-)$ and consequently
-\begin{align}
-f(w) =  {_2F_1}(\alpha, \beta, k_R^+ + k_R^-; w).
-\end{align}
-Recalling that the mean burst size $b = (1-\theta)/\theta = \xi/(1-\xi)$
-and $w = \frac{\xi}{1-\xi} (z-1) = b (z-1)$,
-we can transform back to the original variable $z$ to find the tidy result
-\begin{align}
-f(z) =  {_2F_1}(\alpha, \beta, k_R^+ + k_R^-; b(z-1)),
-\end{align}
-with $\alpha$ and $\beta$ given above by
-\begin{align}
-\begin{split}
-\alpha &= \frac{1}{2}
-\left(k_i+k_R^-+k_R^+ + \sqrt{(k_i+k_R^-+k_R^+)^2 - 4k_i k_R^-}\right)
-\\
-\beta &= \frac{1}{2}
-\left(k_i+k_R^-+k_R^+ - \sqrt{(k_i+k_R^-+k_R^+)^2 - 4k_i k_R^-}\right).
-\end{split}
-\end{align}
-Finally we are in sight of the original goal. We can generate the
-steady-state probability distribution of interest by
-differentiating the generating function,
-\begin{align}
-p(m) = m! \left.\frac{\partial^m}{\partial z^m} f(z) \right|_{z=0},
-\end{align}
-which follows easily from its definition. Some contemplation reveals
-that repeated application of the derivative rule used above will produce
-products of the form $\alpha(\alpha+1)(\alpha+2)\cdots(\alpha+m-1)$ in
-the expression for $p(m)$ and similarly for $\beta$ and $\gamma$. These
-resemble ratios of factorials, but since $\alpha$, $\beta$, and $\gamma$
-are not necessarily integer, we should express the ratios using gamma
-functions instead. More precisely, one finds
-\begin{align}
-p(m) = \frac{
-        \Gamma(\alpha + m)\Gamma(\beta + m)\Gamma(k_R^+ + k_R^-)
-        }
-        {
-        \Gamma(\alpha)\Gamma(\beta)\Gamma(k_R^+ + k_R^- + m)
-        }
-\frac{b^m}{m!}{_2F_1}(\alpha+m, \beta+m, k_R^++k_R^-+m; -b)
-\label{eq:p_m_bursty+rep}
-\end{align}
-which is finally the probability distribution we sought to derive.
-
-\subsubsection{Moments}
-We can verify the correctness of~\eq{eq:p_m_bursty+rep}
-by comparison with Gillespie simulation.
-We can also easily compute moments of the distribution from
-the generating function, and check these against raw moments computed
-directly from the master equation.
-
-The $j$-th raw moment of the distribution can be computed from the
-generating function from the formula
-\begin{align}
-\langle m^j \rangle
-= \left. \left(z \frac{\partial}{\partial z}\right)^j f(z)\right|_{z=1}.
-\end{align}
-Note that \textit{probabilities} are generated by differentiating
-and evaluating at $z=0$, while \textit{moments} are generated by
-differentiating and evaluating at $z=1$. This also aligns with the
-normalization constraint, which can be thought of as the zeroth raw moment.
-
-The mean mRNA is given by
-\begin{align}
-\langle m \rangle = \left.z \pderiv[f]{z} \right|_{z=1}
-= \frac{\alpha\beta b}{k_R^+ + k_R^-}
-= k_i b\frac{k_R^-}{k_R^+ + k_R^-},
-\end{align}
-which is intuitive and appealingly simple:
-the mean mRNA is just the mean mRNA in the absense of repression
-($k_i b$, the burst rate time mean burst size)
-times the fraction of time the repressor is bound, $k_R^-/(k_R^+ + k_R^-)$.
-
-To compute the variance, let us first compute
-\begin{align}
-\begin{split}
-\langle m^2\rangle
-&= \left .z\pderiv{z}\left(z\pderiv[f]{z}\right) \right|_{z=1}
-= \left. z\left(\pderiv[f]{z} + z\psecderiv[f]{z}\right)\right|_{z=1}
-\\
-&= \frac{\alpha\beta b}{k_R^+ + k_R^-}
-        + \frac{\alpha(\alpha+1)\beta(\beta+1) b^2}
-                {(k_R^+ + k_R^-)(k_R^+ + k_R^- + 1)}
-\\
-&= \frac{k_i b k_R^-}{k_R^+ + k_R^-}
-        + \frac{k_i k_R^- b^2 (1 + k_i + k_R^+ + k_R^- + k_i k_R^-)}
-                {(k_R^+ + k_R^-)(k_R^+ + k_R^- + 1)},
-\end{split}
-\end{align}
-from which the variance is
-\begin{align}
-\begin{split}
-var(m) = \langle m^2\rangle - \langle m\rangle^2
-&= \frac{k_i b k_R^-}{k_R^+ + k_R^-}
-\left(
-1 + \frac{b(1 + k_i + k_R^+ + k_R^- + k_i k_R^-)} {k_R^+ + k_R^- + 1}
-- \frac{k_i b k_R^-}{k_R^+ + k_R^-}
-\right)
-\\
-&= k_i b\frac{k_R^-}{k_R^+ + k_R^-}
-\left(
-1 + b + k_i b\frac{1 + k_R^-} {k_R^+ + k_R^- + 1}
-- \frac{k_i b k_R^-}{k_R^+ + k_R^-}
-\right)
-\\
-&= k_i b\frac{k_R^-}{k_R^+ + k_R^-}
-\left(
-1 + b + k_i b\frac{k_R^+}{(k_R^+ + k_R^-)(k_R^+ + k_R^- + 1)}
-\right).
-\end{split}
-\end{align}
-The Fano factor is the quantity in parentheses. This can be interpreted as
-the variability arising from the negative binomial distribution of the
-unregulated promoter $(1+b)$ plus additional noise arising from repression.
-
-\mmnote{Still to writeup: moment derivation direct from CME to compare}
-
-\subsection{Numerical considerations and recursion formulas}
-\subsubsection{Generalities}
-We would like to carry out Bayesian parameter inference on FISH data
-from~\cite{Jones2014}, using~\eq{eq:p_m_bursty+rep} as our
-likelihood. This requires accurate (and preferably fast)
-numerical evaluation of the hypergeometric function ${_2F_1}$,
-which is a notoriously hard problem~\cite{Pearson2017, Gil2007},
-and our particular needs here present an especial challenge as we show below.
-
-The hypergeometric function is defined by its Taylor series as
-\begin{align}
-{_2F_1}(a,b,c;z) 
-= \sum_{l=0}^\infty
-\frac{\Gamma(a + l)\Gamma(b + l)\Gamma(c)}
-        {\Gamma(a)\Gamma(b)\Gamma(c + l)}
-\frac{z^l}{l!}
-\end{align}
-for $|z|<1$, and by analytic continuation elsewhere.
-If $z\lesssim1/2$ and $\alpha$ and $\beta$ are not too large
-(absolute value below 20 or 30),
-then the series converges quickly and an accurate numerical representation is
-easily computed by truncating the series after a reasonable number of terms.
-Unfortunately, we need to evaluate ${_2F_1}$ over mRNA copy numbers fully out
-to the tail of the distribution, which can easily reach 50, possibly 100.
-From~\eq{eq:p_m_bursty+rep}, this means evaluating ${_2F_1}$
-repeatedly for values of $a$, $b$, and $c$ spanning the full range
-from $\mathcal{O}(1)$ to $\mathcal{O}(10^2)$,
-even if $\alpha$, $\beta$, and $\gamma$
-in~\eq{eq:p_m_bursty+rep} are small,
-with the situation even worse if they are not small.
-A naive numerical evaluation of the series definition will be
-prone to overflow and, if any of $a,b,c<0$, then some successive terms in the
-series have alternating signs which can lead to catastrophic cancellations.
-
-One solution is to evaluate ${_2F_1}$ using arbitrary precision arithmetic
-instead of floating point arithmetic,
-e.g., using the \texttt{mpmath} library in Python.
-This is accurate but incredibly slow computationally.
-To quantify how slow, we found that
-evaluating the likelihood defined by~\eq{eq:p_m_bursty+rep} $\sim50$ times
-(for a typical dataset of interest from~\cite{Jones2014},
-with $m$ values spanning 0 to $\sim50$)
-using arbitrary precision arithmetic is 100-1000 fold slower than
-evaluating a negative binomial likelihood for the corresponding
-constitutive promoter dataset.
-
-To claw back $\gtrsim30$ fold of that slowdown, we can exploit
-one of the many catalogued symmetries involving ${_2F_1}$.
-The solution involves recursion relations originally explored by Gauss,
-and studied extensively in~\cite{Pearson2017, Gil2007}.
-They are sometimes known as contiguous relations and relate the values
-of any set of 3 hypergeometric functions whose arguments differ by integers.
-To rephrase this symbolically, consider a set of hypergeometric functions
-indexed by an integer $n$,
-\begin{align}
-f_n = {_2F_1}(a+\epsilon_i n, b+\epsilon_j n, c+\epsilon_k n; z),
-\end{align}
-for a fixed choice of $\epsilon_i, \epsilon_j, \epsilon_k \in \{0,\pm 1\}$
-(at least one of $\epsilon_i, \epsilon_j, \epsilon_k$ must be nonzero,
-else the set of $f_n$ would contain only a single element).
-Then there exist known recurrence relations of the form
-\begin{align}
-A_n f_{n-1} + B_n f_{n} + C_n f_{n+1} = 0,
-\end{align}
-where $A_n, B_n$, and $C_n$ are some functions of $a,b,c$, and $z$.
-In other words, for fixed $\epsilon_i, \epsilon_j, \epsilon_k, a, b,$ and $c$,
-if we can merely evaluate ${_2F_1}$ twice, say for $n^\prime$ and $n^\prime-1$,
-then we can easily and rapidly generate values for arbitrary $n$.
-
-This provides a convenient solution for our problem: we need repeated
-evaluations of ${_2F_1}(a+m, b+m, c+m; z)$
-for fixed $a,b$, and $c$ and many integer values of $m$.
-They idea is that we can use arbitrary precision arithmetic to evaluate
-${_2F_1}$ for just two particular values of $m$ and then generate
-${_2F_1}$ for the other 50-100 values of $m$ using the recurrence
-relation.\footnote{
-There are even more sophisticated ways of utilizing the recurrence
-relations that might have netted another factor of 2 speed-up, and
-possibly as much as a factor of 10, but the method described here had
-already reduced the computation time to an acceptable
-$\mathcal{O}(\text{1 min})$, so these more sophisticated approaches did
-not seem worth the time to pursue.
-}
-
-However, there two further wrinkles.
-The first is that
-a naive application of the recurrence relation is numerically unstable.
-Roughly, this is because the three term recurrence relations,
-like second order ODEs, admit two linearly independent solutions.
-In a certain eigenbasis, one of these solutions dominates the other
-as $n\rightarrow\infty$, and as $n\rightarrow-\infty$,
-the dominance is reversed.
-If we fail to work in this eigenbasis, our solution of the recurrence relation
-will be a mixture of these solutions and rapidly accumulate numerical error.
-For our purposes, it suffices to know that the authors of~\cite{Gil2007}
-derived the numerically stable solutions (so-called \textit{minimal solutions})
-for several possible choices of $\epsilon_i, \epsilon_j, \epsilon_k$.
-Running the recurrence in the proper direction using a minimal solution
-is numerically robust and can be done entirely in floating point arithmetic, 
-so that we only need to evaluate ${_2F_1}$ with arbitrary precision arithmetic
-to generate the seed values for the recursion.
-
-The second wrinkle is a corollary to the first.
-The minimal solutions are only minimal for certain ranges of the argument $z$,
-and not all of the 26 possible recurrence relations
-have minimal solutions for all $z$.
-This can be solved by using one of the many transformation formulae for
-${_2F_1}$ to convert to a different recurrence relation that has
-a minimal solution over the required domain of $z$, although
-this can require some trial and error to find the right transformation,
-the right recurrence relation, and the right minimal solution.
-
-\subsubsection{Particulars}
-Let us now demonstrate these generalities for our problem of interest.
-In order to evaluate the probability distribution of our
-model,~\eq{eq:p_m_bursty+rep}, we need to evaluate hypergeometric functions
-of the form ${_2F_1}(\alpha+m, \beta+m, \gamma+m; -b)$
-for values of $m$ ranging from $0$ to $\mathcal{O}(100)$.
-The authors of~\cite{Gil2007} did not derive a recursion relation
-for precisely this case. We could follow their methods and do so ourselves,
-but it is much easier to convert to a case that they did consider.
-The strategy is to look through the minimal solutions tabulated
-in~\cite{Gil2007} and search for a transformation we could apply to
-${_2F_1}(\alpha+m, \beta+m, \gamma+m; -b)$ that would place the $m$'s
-(the variable being incremented by the recursion)
-in the same arguments of ${_2F_1}$ as the minimal solution.
-After some ``guess and check,'' we found that the transformation
-\begin{align}
-{_2F_1}(\alpha+m, \beta+m, \gamma+m; -b)
-=
-(1+b)^{-\alpha-m}
-        {_2F_1}\left(\alpha+m, \gamma-\beta, \gamma+m; \frac{b}{1+b}\right),
-\label{eq:rec_euler_pretransform}
-\end{align}
-produces a ${_2F_1}$ on the right hand side that closely resembles
-the minimal solutions $y_{3,m}$ and $y_{4,m}$ in Eq.~4.3 in~\cite{Gil2007}.
-Explicitly, these solutions are
-\begin{align}
-y_{3,m}
-&\propto
-{_2F_1}\left(-\alpha^\prime + \gamma^\prime - m,
-                -\beta^\prime + \gamma^\prime,
-                1-\alpha^\prime-\beta^\prime+\gamma^\prime-m;
-                1-z\right)
-\\
-y_{4,m}
-&\propto
-{_2F_1}\left(\alpha^\prime + m,
-                \beta^\prime,
-                1+\alpha^\prime+\beta^\prime-\gamma^\prime+m;
-                1-z\right),
-\label{eq:minimal_soln_sans_prefac}
-\end{align}
-where we have omitted prefactors which are unimportant for now.
-Which of these two we should use depends on what values $z$ takes on.
-Equating $1-z=b/(1+b)$ gives $z=1/(1+b)$, and since $b$ is strictly positive,
-$z$ is bounded between 0 and 1.
-From Eq.~4.5 in~\cite{Gil2007}, $y_{4,m}$ is the minimal solution
-for real $z$ satisfying $0<z<2$, so this is the only minimal solution we need.
-
-Now that we have our minimal solution,
-what recurrence relation does it satisfy?
-Confusingly, the recurrence relation of which $y_{4,m}$ is a solution
-increments different arguments of ${_2F_1}$ that does $y_{4,m}$:
-it increments the first only, rather than first and third.
-This recurrence relation can be looked up, e.g., Eq.~15.2.10
-in~\cite{Abramowitz1964}, which is
-\begin{align}
-(\gamma^\prime - (\alpha^\prime + m)) f_{m-1}
-+
-(2(\alpha^\prime+m) - \gamma^\prime + (\beta^\prime - \alpha^\prime)z)f_m
-+ \alpha^\prime(z-1) f_{m+1} = 0.
-\label{eq:chosen_rec_rel}
-\end{align}
-Now we must solve for the parameters appearing in the recurrence relation
-in terms of our parameters, namely by setting
-\begin{align}
-\begin{split}
-\alpha^\prime &= \alpha
-\\
-\beta^\prime &= \gamma - \beta
-\\
-1 + \alpha^\prime + \beta^\prime - \gamma^\prime &= \gamma
-\\
-1 - z &= \frac{b}{1+b}
-\end{split}
-\end{align}
-and solving to find
-\begin{align}
-\begin{split}
-\alpha^\prime &= \alpha
-\\
-\beta^\prime &= \gamma - \beta
-\\
-\gamma^\prime &= 1 + \alpha - \beta
-\\
-z &= \frac{1}{1+b}
-.
-\end{split}
-\end{align}
-Finally we have everything we need. The minimal solution
-\begin{align}
-y_{4,m}
-=
-\frac{\Gamma(1+\alpha^\prime-\gamma^\prime+m)}
-        {\Gamma(1+\alpha^\prime+\beta^\prime-\gamma^\prime+m)}
-\times
-{_2F_1}\left(\alpha^\prime + m,
-                \beta^\prime,
-                1+\alpha^\prime+\beta^\prime-\gamma^\prime+m;
-                1-z\right),
-\end{align}
-where we have now included the necessary prefactors,
-is a numerically stable solution of the recurrence
-relation~\eq{eq:chosen_rec_rel} if the recursion is run
-from large $m$ to small $m$.
-So the procedure is to compute the value of ${_2F_1}$ for the two
-largest $m$ values of interest using arbitrary precision arithmetic,
-compute the prefactors to construct
-$y_{4,\text{max}(m)}$ and $y_{4,\text{max}(m)-1}$,
-recursively compute $y_{4,m}$ for all $m$ less than $\text{max}(m)$ down
-to $m=0$, then cancel off the prefactors of the resulting values of
-$y_{4,m}$ for all $m$ to produce ${_2F_1}$ for all desired $m$ values.
-\mmnote{Maybe I should reformat this paragraph into
-a pseudocode version of the algorithm?
-Instead of, or in addition to?}
-
-With ${_2F_1}$ computed, the only remaining numerical danger in computing
-$p(m)$ in~\eq{eq:p_m_bursty+rep} is overflow of the gamma functions.
-This is easily solved by taking the log of the entire expression
-and using standard routines to compute the log of the gamma functions,
-then exponentiating the entire expression at the end if $p(m)$
-is needed rather than $\log p(m)$.
-
-% \bibliographystyle{nature}
-\bibliographystyle{abbrv}
-\bibliography{./library}
-
-\end{document}

### 2020-05-19 doc/appendix_03_regulated_promoters.tex
initial splitting of document into several files

--- 2020-05-17 doc/paper.tex
@@
-\documentclass[12pt]{article}%{amsart}
-\usepackage[top = 1.0in, bottom = 1.0in, left = 1.0in, right = 1.0in]{geometry}
-% See geometry.pdf to learn the layout options. There are lots.
-%\geometry{letterpaper}               % ... or a4paper or a5paper or ...
-%\geometry{landscape}             % Activate for for rotated page geometry
-\usepackage[parfill]{parskip}    % Activate to begin paragraphs with an empty line rather than an indent
-\usepackage{graphicx}
-\usepackage{amssymb}
-%\usepackage{epstopdf}
-% \usepackage{caption}
-\usepackage{amsmath}
-% \usepackage{longtable}
-% \usepackage{tabu}
-% \usepackage{accents} %to get undertildes for vec & mat
-\usepackage{color} % for colored text
-% \usepackage[normalem]{ulem} % for editing: striking out text using 'sout' command
-
-% \newcommand\mytablefigwidth{0.35\textwidth}
-
-% allows to use .ai files directly w/o resaving as pdf
-\DeclareGraphicsRule{.ai}{pdf}{.ai}{}
-
-% Handy math macros!
-\newcommand{\vect}[1]{\vec{#1}}
-\newcommand{\matr}[1]{\mathbf{#1}}
-\newcommand{\rate}[3]{{#1}_{#2}^{#3}}
-\newcommand{\mmnote}[1]{\textcolor{cyan}{(MM:~#1)}}
-
-% derivative macros. these are sneaky. the [{}] is an empty default 1st arg
-% usage: provide 1 arg: \deriv{x} gives d/dx
-% provide 2 args like so: \deriv[f]{x} gives df/dx
-\newcommand{\deriv}[2][{}]{\frac{d #1}{d #2}}
-\newcommand{\pderiv}[2][{}]{\frac{\partial #1}{\partial #2}}
-\newcommand{\psecderiv}[2][{}]{\frac{\partial{^2} #1}{\partial #2{^2}}}
-
-%%%%% Referencing macros %%%%%
-\newcommand{\fig}[1]{Figure~\ref{#1}}
-\newcommand{\tref}[1]{Table~\ref{#1}}
-\newcommand{\eq}[1]{Eq.~(\ref{#1})}
-\newcommand{\eqrange}[2]{Eqs.~(\ref{#1}-\ref{#2})}
-
-%%%%%
-\begin{document}
-
-\title{Simple repression draft}
-
-\maketitle
-
-\section{Introduction}
-\mmnote{Key ideas, no particular order, that I haven't written down before:
-\begin{itemize}
-\item Our goal is to build phenomenological models of input-output
-functions of genetic circuits. More precisely, we want phenomenology
-in the sense of coarse-graining away atomistic detail, but still
-retaining biophysical meaning, e.g., we don't want to coarse-grain
-as far as Hill functions. Why not? We are motivated by studies
-like~\cite{Razo-Mejia2020}, for which Hill functions are
-insufficient. Such work requires a quantitative theory of how
-biophysical changes at the molecular level propagate to
-input-output functions at the genetic circuit level.
-We want concepts, not mere facts.
-In particular a key question is: at this level of
-coarse graining, what microscopic details do I need to explicitly
-model, and how do we figure that out? For example, do I need to
-worry about all or even any of the steps that individual RNAPs go
-through each time they make a transcript? Turning the question
-around, can we see any imprint of those processes in the
-available data? If the answer is no, then those processes are
-irrelevant for our purposes. Forward modeling and inverse
-(statistical inferential) modeling can complement each other
-beautifully here.
-\end{itemize}}
-
-\fig{fig1:means_cartoons}A shows the qualitative picture of
-simple repression. An operator, the binding site on the DNA for a
-repressor protein, may be found occupied by a repressor, in which
-case transcription is blocked from occurring. Or it may be found
-unoccupied, in which case RNA polymerase (RNAP) may bind and
-transcription can proceed.
-
-The key assumption we make is that binding of repressor and RNAP in the promoter region of interest is exclusive, meaning that one or the other may bind, but never may both be simultaneously bound.
-It is often imagined that when the repressor is bound to its operator,
-RNAP is sterically blocked from binding to its promoter sequence.
-Current evidence suggests this is sometimes, but not
-always the case~\mmnote{Cite that fig, from ??Nathan or Bill or
-Rob??, showing locations of repressor binding sites far upstream
-of promoter}, and it remains an interesting open question precisely
-how a repressor bound far upstream is able to repress transription.
-Suggestions include ``action-at-a-distance'' mediated by kinks in
-the DNA, formed when the repressor is bound, that prevent RNAP binding.
-Nevertheless, our modeling in this work is sufficiently
-coarse-grained that we simply assume exclusive binding and need
-not worry about such precise details.
-
-\mmnote{Add a notation aside somewhere that all association rates are written as zeroth-order rates.
-In other words, the concentration of the molecule is hidden inside, so for instance, $k_R^+$ is proportional to repressor copy number.}
-
-\section{Means}
-\subsection{Fold-changes are indistinguishable across models}
-As a first stop on our search for the ``right'' model of simple
-repression, let us consider what we can learn from theory and
-measurements of fold-change in mean expression. The fold-change is defined as
-\begin{equation}
-\text{fold-change}
-= \frac{\langle \text{gene expression with repressor present} \rangle}
-        {\langle \text{gene expression with repressor absent} \rangle}
-= \frac{\langle m\rangle(R\ne 0)}{\langle m\rangle(R = 0)}
-\label{eq:fc_def}
-\end{equation}
-where angle brackets denote the averge over a population of cells
-and mean mRNA $\langle m\rangle$ is viewed as a function of
-repressor copy number $R$.
-The second
-equality in~\ref{eq:fc_def} follows from assuming that the
-translation efficiency, i.e., the number of proteins translated
-per mRNA, is the same in both conditions.
-In other words, we assume that mean protein level is proportional
-to mean mRNA level, and that the proportionality constant is the
-same in both conditions and therefore cancels out in the ratio.
-This is reasonable since the cells in the two conditions are
-identical except for the presence of the transcription factor,
-and it is difficult to imagine the transcription factor having
-any direct effect on translation.
-
-Fold-change has proven a very convenient observable in past
-work~\cite{Garcia2011a, Razo-Mejia2018, Chure2019}.
-Part of its utility in dissecting transcriptional regulation is
-its ratiometric nature, which is always easier than making an
-absolute measurement. Also, by measuring otherwise identical cells
-with and without a transcription factor present, any biological
-noise common to both conditions can be made to cancel away.
-In addition, the presence of different operator sites is known to
-change the basal expression level in the absence of
-repressor~\mmnote{look up cite}.
-Computing fold-changes neatly cancels this effect as well.
-
-\fig{fig1:means_cartoons} depicts a smorgasbord
-of mathematized cartoons for simple repression that have appeared
-in previous literature. For each cartoon, we calculate the
-fold-change in mean gene expression as predicted by that model,
-deferring some algebraic details to the appendix.
-What we will find is that all cartoons collapse to a single
-master curve, shown in \fig{fig1:means_cartoons}c, which contains
-just two parameters. We label the parameters $\Delta F_R$, an
-effective free energy parametrizing the repressor-DNA
-interaction, and $\rho$, which subsumes all details of
-transcription in the absence of repressors.
-We will offer some intuition for why this master curve exists and discuss
-why it is likely impossible to discriminate ``right'' from ``wrong''
-cartoons given only measurements of fold-changes in expression.
-
-\subsubsection{Equilibrium, 2-state}
-In this simplest model, the promoter is idealized as existing in
-one of two states, either repressor bound or repressor unbound.
-The rate of transcription, and therefore the level of gene
-expression, is assumed to be proportional to the fraction of time
-spent in the repressor unbound state.
-Mathematically, this amounts to the statement that the mean mRNA level is
-given by
-\begin{equation}
-\langle m \rangle = \frac{r}{\gamma}
-        \left(1 + \frac{R}{N_{NS}} e^{-\beta\Delta\epsilon_R}\right)^{-1},
-\end{equation}
-where $r$ is the transcription rate from the repressor unbound
-state, $\gamma$ is the mRNA degradation rate, $R$ is repressor
-copy number, $N_{NS}$ is the number of nonspecific binding sites
-in the genome (where repressors spend most of their time when not
-bound to the operator, their specifc binding site),
-$\beta=1/k_BT$, and $\Delta\epsilon_R$ is the binding energy of a
-repressor to its operator site. The derivation of this result is deferred
-to the appendix.
-
-The fold-change is easily found as the ratio of mean mRNA with
-and without repressor. This gives
-\begin{equation}
-\text{fold-change}
-= \left(1 + \frac{R}{N_{NS}} e^{-\beta\Delta\epsilon_R}\right)^{-1},
-\end{equation}
-which clearly matches the form of the master curve
-in~\eq{fig1:means_cartoons} with $\rho=1$.
-
-In fact it was noted in~\cite{Chure2019} that this two-state
-model can be viewed as the coarse-graining of any equilibrium
-promoter model in which no transcriptionally active states have
-transcription factor bound. Put differently, there is no overlap
-between TF bound states and transcriptionally active states. We
-will see this explicitly in the 3-state equilibrium model below,
-but perhaps surprising is that something analogous carries over
-even to the nonequilibrium models we consider later.
-
-\subsubsection{Equilibrium, 3-state}
-Compared to the previous model, here we fine-grain the repressor
-unbound state into two separate states: empty, and RNAP bound.
-This picture was used in~\cite{Garcia2011a} as we use it here,
-and in~\cite{Razo-Mejia2018} and~\cite{Chure2019} it was generalized
-to incorporate small-molecule inducers that bind the repressor.
-Interestingly, the effect of this is simply to renormalize $R$
-from the total number of repressors to a smaller effective number,
-roughly (though not exactly) the number which are unbound by inducers.
-
-The mean mRNA, as derived in the appendix from a similar
-enumeration of states and weights as the previous model, is
-\begin{equation}
-\langle m \rangle = \frac{r}{\gamma}
-\frac{\frac{P}{N_{NS}} e^{-\beta\Delta\epsilon_P}}
-        {
-        1 + \frac{R}{N_{NS}} e^{-\beta\Delta\epsilon_R}
-        + \frac{P}{N_{NS}} e^{-\beta\Delta\epsilon_P}
-        },
-\end{equation}
-where the new variables are $\Delta\epsilon_P$, the difference in
-RNAP binding energy to its specific site (the promoter) relative
-to an average nonspecific background site, and the RNAP copy
-number, $P$. The fold-change again follows immediately as
-\begin{align}
-\text{fold-change}
-&= \frac{\frac{P}{N_{NS}} e^{-\beta\Delta\epsilon_P}}
-        {
-        1 + \frac{R}{N_{NS}} e^{-\beta\Delta\epsilon_R}
-        + \frac{P}{N_{NS}} e^{-\beta\Delta\epsilon_P}
-        }
-\frac{1 + \frac{P}{N_{NS}} e^{-\beta\Delta\epsilon_P}}
-        {\frac{P}{N_{NS}} e^{-\beta\Delta\epsilon_P}}
-\\
-&= \left(
-1 + \frac{\frac{R}{N_{NS}} e^{-\beta\Delta\epsilon_R}}
-        {1 + \frac{P}{N_{NS}} e^{-\beta\Delta\epsilon_P}}
-\right)^{-1}
-\\
-&= (1 + \exp(-\Delta F_R - \log\rho))^{-1},
-\end{align}
-with $\Delta F_R = \beta\Delta\epsilon_R - \log(R/N_{NS})$
-and $\rho = 1 + \frac{P}{N_{NS}}\mathrm{e}^{-\beta\Delta\varepsilon_P}$
-as shown in~\fig{fig1:means_cartoons}.
-
-\begin{figure}%[h!]
-\centering
-\includegraphics[width=\textwidth]{../figures/fig1/fig1.ai}
-\caption{
-\textbf{
-An overview of the simple repression motif at the level of means.}
-(A) depicts the qualitative biological picture of simple repression.
-(B) shows a variety of possible mathematized cartoons of simple
-repression, along with the effective parameter $\rho$
-which subsumes all regulatory details
-of the architecture excluding the repressor.
-(C) shows the ``master curve'' to which all cartoons in (B) collapse.
-}
-\label{fig1:means_cartoons}
-\end{figure}
-
-\subsubsection{Nonequilibrium model 1 - Poisson promoter}
-For our first kinetic model, we consider the simplest possible
-picture with only two states, repressor bound and unbound.
-This is exactly the model used for the main results of~\cite{Jones2014}.
-In this picture, repressor association and dissociation rates
-from its operator site, $k_R^+$ and $k_R^-$, respectively, govern
-transitions between the two states. When the system is in the
-unbound state, transcriptions initiate at rate $r$, which
-represents a coarse-graining of all the downstream processes into
-a single effective rate. mRNA is again degraded at rate $\gamma$.
-
-Let $p_{m,R}(t)$ denote the joint probability of finding the
-system in the repressor bound state with $m$ mRNA molecules
-present at time $t$, and similarly define $p_{m,U}(t)$ for the
-repressor unbound state. 
-This model is governed by a master equation giving the time
-evolution of $p_{m,R}(t)$ and $p_{m,U}(t)$, which is
-\begin{align}
-\begin{split}
-\deriv{t}p_{m,R}(t) =& - k_R^- p_{m,R}(t) + k_R^+ p_{m,U}(t)
-                + (m+1)\gamma p_{m+1,R}(t) - \gamma p_{m,R}(t)
-\\
-\deriv{t}p_{m,U}(t) =&\; k_R^- p_{m,R}(t) - k_R^+ p_{m,U}(t)
-                        + rp_{m-1,U}(t) - rp_{m,U}(t)
-                \\
-                &+ (m+1)\gamma p_{m+1,U}(t) - \gamma p_{m,U}(t).
-\end{split}
-\end{align}
-In each equation, the first two terms describe transitions
-between promoter states due to repressors unbinding and binding,
-respectively. The final two terms describe degradation of mRNA,
-decreasing the copy number by one, and the terms with coefficient
-$r$ describe transcription initiation increasing the mRNA copy
-number by one.
-
-We can greatly simplify the notation, which will be especially
-useful for the more complicated models yet to come, by
-re-expressing the master equation in vector form. The promoter
-states are collected into a vector and the rate constants are
-collected into matrices as
-\begin{equation}
-\vect{p}_m = \begin{pmatrix} p_{m,R} \\ p_{m,U} \end{pmatrix},\
-\matr{K} = \begin{pmatrix} -k_R^- & k_R^+ \\ k_R^- & -k_R^+ \end{pmatrix},\
-\matr{R} = \begin{pmatrix} 0 & 0 \\ 0 & r \end{pmatrix},\
-\label{eq:2state_cme_matrices}
-\end{equation}
-so that the master equation may be condensed as
-\begin{equation}
-\deriv{t}\vect{p}_m(t) =
-\left( \matr{K} - \matr{R} - \gamma m \matr{I} \right) \vect{p}_m(t)
-                + \matr{R} \vect{p}_{m-1}(t)
-                + \gamma (m+1) \matr{I} \vect{p}_{m+1}(t)
-                - \gamma m \matr{I} \vect{p}_m(t),
-\label{eq:2state_rep_cme}
-\end{equation}
-where $\matr{I}$ is the identity matrix.
-Taking steady state by setting time derivatives to zero,
-the mean mRNA can be found to be
-\begin{equation}
-\langle m \rangle = \frac{r}{\gamma}
-        \left(1 + \frac{k_R^+}{k_R^-}\right)^{-1},
-\end{equation}
-with the algebra details again deferred to the appendix.
-Recall $k_R^+$ is proportional to the repressor copy number, so
-in computing fold-change, absence of repressor corresponds to
-$k_R^+\rightarrow0$.
-Therefore fold-change in this model is simply
-\begin{equation}
-\text{fold-change} = \left(1 + \frac{k_R^+}{k_R^-}\right)^{-1},
-\end{equation}
-again matching the master curve of~\fig{fig1:means_cartoons} with $\rho=1$.
-
-\subsubsection{Nonequilibrium model 2 - RNAP bound/unbound states}
-Analogous to the three-state equilibrium model, this model
-fine-grains the repressor unbound state of nonequilibrium model 1,
-resolving it into an empty promoter state and an RNAP-bound state.
-Note in this picture, in contrast with model 4 below,
-transcription initiation is accompanied by a promoter state
-change, in keeping with the interpretation as RNAP-bound and
-empty states: if an RNAP successfully escapes the promoter and
-proceeds to elongation of a transcript, clearly it is no longer
-bound at the promoter! Therefore another RNAP must bind before
-another transcript can be initiated.
-
-The master equation governing this model is analogous
-to~\eqrange{eq:2state_cme_matrices}{eq:2state_rep_cme} for model 1 above.
-The main subtlety arises since transcription initiation
-accompanies a promoter state change.
-This can be understood by analogy to $\matr{K}$. The
-off-diagonals and diagonals of $\matr{K}$ correspond to
-transitions arriving at or departing from, respectively, the
-promoter state of interest. If transcription initiations are
-accompanied by promoter state changes, we must have separate
-matrices for arriving and departing transcription events since
-the arriving and departing transitions have different initial
-copy number of mRNA, unlike for $\matr{K}$ where they are the same.
-The master equation for this model is
-\begin{equation}
-\deriv{t}\vect{p}_m(t) =
-\left( \matr{K} - \matr{R_D} - \gamma m \matr{I} \right) \vect{p}_m(t)
-                + \matr{R_A} \vect{p}_{m-1}(t) +
-                \gamma (m+1) \matr{I} \vect{p}_{m+1}(t)
-                - \gamma m \matr{I} \vect{p}_m(t),
-\label{eq:3state_rep_cme}
-\end{equation}
-with the state vector and promoter transition matrix defined as
-\begin{equation}
-\vect{p}_m = \begin{pmatrix} p_{m,R} \\ p_{m,E} \\ p_{m,P} \end{pmatrix},\
-\matr{K} = \begin{pmatrix} -k_R^- & k_R^+ & 0 \\
-                        k_R^- & -k_R^+ -k_P^+ & k_P^- \\
-                        0 & k_P^+ & -k_P^- 
-                \end{pmatrix},
-\label{eq:3state_cme_matrices_pt1}
-\end{equation}
-and the initiation matrices given by
-\begin{equation}
-\matr{R_A} = \begin{pmatrix}
-                0 & 0 & 0 \\ 0 & 0 & r \\ 0 & 0 & 0
-                \end{pmatrix},\
-\matr{R_D} = \begin{pmatrix}
-                0 & 0 & 0 \\ 0 & 0 & 0 \\ 0 & 0 & r
-                \end{pmatrix},
-\label{eq:3state_cme_matrices_pt2}
-\end{equation}
-The elements of $\vect{p}_m$ encode the probabilities of having
-$m$ mRNA present along with the promoter having repressor bound,
-being empty, or having RNAP bound, respectively.
-$\matr{R_A}$ describes probability flux \textit{arriving} at the state
-$\vect{p}_m$ from a state with one fewer mRNA, namely
-$\vect{p}_{m-1}$, and $\matr{R_D}$ describes probability flux
-\textit{departing} from the state $\vect{p}_m$
-for a state with one more mRNA, namely $\vect{p}_{m+1}$.
-$\matr{K}$ is closely analogous to model 1.
-
-Mean mRNA at steady state is found analogously to model 1, with the result
-\begin{equation}
-\langle m\rangle = \frac{r}{\gamma}
-        \frac{k_R^- k_P^+}
-        {k_R^- k_P^+ + k_R^- (k_P^- + r) + k_R^+ (k_P^- + r)}.
-\end{equation}
-Fold-change is again found from the ratio prescribed by~\eq{eq:fc_def},
-from which we have
-\begin{align}
-\text{fold-change}
-&=      \frac{k_R^- k_P^+}
-        {k_R^- k_P^+ + k_R^- (k_P^- + r) + k_R^+ (k_P^- + r)}
-        \frac{k_P^+ + k_P^- + r}{k_P^+}
-\\
-&=      \left(1 + \frac{k_R^+}{k_R^-}
-                \frac{k_P^- + r}{k_P^+ + k_P^- + r}
-        \right)^{-1}
-\\
-&=      \left(1 + \frac{k_R^+}{k_R^-}
-        \left(1 + \frac{k_P^+}{k_P^- + r}\right)^{-1}
-        \right)^{-1},
-\end{align}
-which follows the master curve with
-$\rho = 1 + k_P^+/(k_P^- + r)$ as claimed.
-
-\subsubsection{Nonequilibrium model 3
-- Multistep transcription initiation and escape}
-One might reasonably complain that the first two
-``nonequilibrium'' models we have considered are straw men. Their
-steady states necessarily satisfy detailed balance which is
-equivalent to thermodynamic equilibrium. Why is this the case? At
-steady state there is by definition no net probability flux in or
-out of each promoter state, but since the promoter states form a
-linear chain, there is only one way in or out of the repressor
-bound and RNAP bound states, implying each edge must actually
-have a net zero probability flux, which is the definition of
-detailed balance (usually phrased as equality of forward and
-reverse transition fluxes).
-
-Now we consider model 3 in~\eq{fig1:means_cartoons} which allows
-the possibility of true nonequilibrium steady-state fluxes
-through the promoter states.
-We point out that this model was considered previously in~\cite{Mitarai2015}
-where a comparison was made with model 1 as used in~\cite{Jones2014}.
-The authors of~\cite{Mitarai2015} argued that the additional
-complexity is essential to properly account for the noise in the
-mRNA distribution. We will weigh in on both models later when we
-consider observables beyond fold-change.
-
-The master equation governing this model is
-identical in form to model 2 above, namely
-\begin{equation}
-\deriv{t}\vect{p}_m(t) =
-\left( \matr{K} - \matr{R_D} - \gamma m \matr{I} \right) \vect{p}_m(t)
-                + \matr{R_A} \vect{p}_{m-1}(t) +
-                \gamma (m+1) \matr{I} \vect{p}_{m+1}(t)
-                - \gamma m \matr{I} \vect{p}_m(t),
-\end{equation}
-but obviously with a higher-dimensional state space and different matrices.
-The state vector and promoter transition matrix are now
-\begin{equation}
-\vect{p}_m = \begin{pmatrix} p_{m,R} \\ p_{m,E} \\
-                             p_{m,C} \\ p_{m,O}\end{pmatrix},\
-\matr{K} = \begin{pmatrix} -k_R^- & k_R^+ & 0 & 0\\
-                        k_R^- & -k_R^+ -k_P^+ & k_P^- & 0 \\
-                        0 & k_P^+ & -k_P^- - k_O & 0 \\
-                        0 & 0 & k_O & 0
-                \end{pmatrix},
-\end{equation}
-with the four promoter states, in order, being repressor bound,
-empty, RNAP closed complex, and RNAP open complex.
-Besides increasing dimension by one again, the only new feature in
-$\matr{K}$ is the rate $k_O$, representing the (irreversible)
-rate of open complex formation from the closed complex.
-The initiation matrices are given by
-\begin{equation}
-\matr{R_A} = \begin{pmatrix}
-        0 & 0 & 0 & 0 \\ 0 & 0 & 0 & r \\ 0 & 0 & 0 & 0\\ 0 & 0 & 0 & 0
-                \end{pmatrix},\
-\matr{R_D} = \begin{pmatrix}
-        0 & 0 & 0 & 0 \\ 0 & 0 & 0 & 0 \\ 0 & 0 & 0 & 0\\ 0 & 0 & 0 & r
-                \end{pmatrix},
-\end{equation}
-again closely analogous to model 2.
-
-The expression for mean mRNA is substantially more complicated
-now, as worked out in the appendix where we find
-\begin{equation}
-\langle m\rangle = \frac{r}{\gamma}
-        \frac{k_R^- k_P^+ k_O}
-        {k_R^- [(k_P^+ (k_O + r) + r(k_P^- + k_O)] + k_R^+ r(k_P^- + k_O)},
-\end{equation}
-which can be simplified to
-\begin{equation}
-\langle m\rangle
-= \frac{r}{\gamma}
-\frac{\frac{k_P^+ k_O}{r(k_O + k_P^-)}}
-        {1 + \frac{k_P^+ (k_O + r)}{r(k_O + k_P^-)} + \frac{k_R^+}{k_R^-}}.
-\end{equation}
-The strategy is to isolate the terms involving the repressor, so that now the fold-change is seen to be simply
-\begin{align}
-\text{fold-change}
-&= \frac{\frac{k_P^+ k_O}{r(k_O + k_P^-)}}
-        {1 + \frac{k_P^+ (k_O + r)}{r(k_O + k_P^-)} + \frac{k_R^+}{k_R^-}}
-        \frac{1 + \frac{k_P^+ (k_O + r)}{r(k_O + k_P^-)}}
-                {\frac{k_P^+ k_O}{r(k_O + k_P^-)}}
-\\
-&= \left(
-        1 + \frac{k_R^+}{k_R^-}
-        \left(1 + \frac{k_P^+ (k_O + r)}{r(k_O + k_P^-)}\right)^{-1}
-\right)^{-1},
-\end{align}
-surprisingly reducing to the master curve once again, with
-$\rho = 1 + \frac{k_P^+ (k_O + r)}{r(k_O + k_P^-)}$.
-
-This example hints that an arbitrarily fine-grained model of
-downstream transcription steps may still be collapsed to the form
-of the master curve in~\fig{fig1:means_cartoons}B so long as the
-repressor binding is exclusive with transcriptionally active
-states. We offer this as a conjecture, and we suspect that a
-careful argument using the King-Altman diagram
-method~\cite{King1956, Hill1966} might furnish a ``proof.''
-Our focus here is not on full generality but rather to survey an
-assortment of plausible models that have been proposed in the
-literature for simple repression.
-
-\subsubsection{Nonequilibrium model 4 - ``Active'' and ``inactive'' states}
-This model is at the core of the theory in~\cite{Razo-Mejia2020}.
-At a glance the cartoon for this model may appear very similar to model 2,
-and mathematically it is, but the interpretation is rather different.
-In model 2, we interpreted the third state literally as an
-RNAP-bound promoter and modeled initiation of a transcript as
-triggering a promoter state change, making the hopefully
-uncontroversial assumption that an RNAP can only make one
-transcript at a time. In contrast, in the present model the
-promoter state does \textit{not} change when a transcript is
-initiated. So we no longer interpret these states as literally
-RNAP bound and unbound but instead as coarse-grained ``active''
-and ``inactive'' states, the details of which we leave
-unspecified for now. We will comment more on this model below
-when we discuss Fano factors of models.
-
-Mathematically this model is very similar to models 1 and 2.
-Like model 1, the matrix $R$ describing transcription initiation
-is diagonal, namely
-\begin{equation}
-\matr{R} = \begin{pmatrix}
-                0 & 0 & 0 \\ 0 & 0 & 0 \\ 0 & 0 & r
-        \end{pmatrix}.
-\end{equation}
-so the master equation takes verbatim the same form as it did for model
-1,~\eq{eq:2state_rep_cme}. Meanwhile the promoter transition matrix
-$K$ is the same as~\eq{eq:3state_cme_matrices_pt1} from model 2,
-although we relabel the rate constants from $k_P^\pm$ to $k^\pm$ to
-reiterate that these are not simply RNAP binding and unbinding rates.
-
-Carrying out the algebra, the mean mRNA can be found to be
-\begin{equation}
-\langle m\rangle = \frac{r}{\gamma}
-\frac{k_R^- k^+}
-{k_R^- k^+ + k_R^- k^- + k_R^+ k^-},
-\end{equation}
-and the fold-change readily follows,
-\begin{align}
-\text{fold-change}
-&=      \frac{k_R^- k^+}{k_R^- k^+ + k_R^- k^- + k_R^+ k^-}
-        \frac{k_R^- k^+ + k_R^- k^-}{k_R^- k^+}
-\\
-&=      \left(1 + \frac{k_R^+}{k_R^-}
-                \left(1 + \frac{k^+}{k^-}\right)^{-1}
-        \right)^{-1},
-\end{align}
-from which we see $\rho = 1 + k^+/k^-$ as shown in~\fig{fig1:means_cartoons}.
-
-\subsubsection{Nonequilibrium model 5 - Bursty promoter}
-We will have much more to say about this model in subsequent sections.
-For now we sketch it as an intuitive analog to model 1,
-with just two states, repressor bound or unbound, and transition rates
-between them of $k_R^+$ and $k_R^-$.
-In model 1, when in the unbound state, single mRNA transcripts
-are produced as a Poisson process with some characteristic rate
-$r$. This model by contrast produces, at some Poisson rate $k_i$,
-\textit{bursts} of transcripts with a geometrically distributed
-burst size and a mean burst size $b$.
-
-From this intuitive picture and by analogy to model 1, then, it should be plausible that the mean mRNA level is
-\begin{equation}
-\langle m\rangle = \frac{k_i b}{\gamma}
-        \left(1 + \frac{k_R^+}{k_R^-}\right)^{-1},
-\end{equation}
-which will turn out to be correct from a careful calculation.
-For now, we simply note that just like model 1, the fold-change becomes
-\begin{equation}
-\text{fold-change} = \left(1 + \frac{k_R^+}{k_R^-}\right)^{-1}
-\end{equation}
-with $\rho=1$ also like model 1.
-
-We will also see later how this model emerges as a natural limit of model 4.
-
-\subsubsection{Summary and discussion of results across models for fold-changes in mean expression}
-What have we learned from our exploration of the models
-in~\fig{fig1:means_cartoons} and their predictions of
-fold-changes in mean expression levels? The takeaway is the
-existence of a master curve (\fig{fig1:means_cartoons}C) to which
-the fold-change predictions of all the models collapse. This
-master curve is parametrized by only two effective parameters:
-$\Delta F_R$, which characterizes the repressor, and $\rho$,
-which characterizes all other features of the promoter
-architecture. The key assumption underpinning this result is that
-zero transcription occurs when a repressor is bound to its
-operator. Note, however, that we are agnostic about the molecular
-mechanism which achieves this;
-steric effects are the obvious mechanism, but, for instance,
-``action-at-a-distance'' mediated by kinked DNA due to
-repressors bound tens or hundreds of nucleotides upstream of a
-promoter is plausible as well.
-
-Why does the master curve of~\fig{fig1:means_cartoons}C exist at all?
-This brings to mind the deep questions posed in,
-e.g.,~\cite{Frank2013} and~\cite{Frank2014a}, asking us to search
-for general ``symmetries'' of the problem instead of specifying
-generative models of the known microscopics of a system.
-In our case, the key feature seems to be the exclusive nature of
-repressor and RNAP binding, which allows the parameter describing
-the repressor, $\Delta F_R$, to cleanly separate from all other
-details of the promoter architecture, which are encapsulated in $\rho$.
-In the graph-theoretic language which was applied to similar
-problems in~\cite{Gunawardena2012} and~\cite{Ahsendorf2014}, the
-repressor-bound state is a leaf of the graph describing
-transitions between promoter states. Arbitrary nonequilibrium
-behavior can occur on the rest of the promoter state space, but
-it may all be swept up in the effective parameter $\rho$, to
-which the repressor makes no contribution.
-
-As suggested in~\cite{Chure2019}, we believe this master curve
-should generalize to architectures with multiple repressor
-binding sites, as long as the exclusivity of transcription factor
-binding and transcription initation is maintained. The
-interpretation of $\Delta F_R$ is then of an effective free
-energy of all repressor bound states.
-In an equilibrium picture this is simply given by the log of the
-sum of Boltzmann weights of all repressor bound states, which
-looks like the log of a partition function of a subsystem.
-In a nonequilibrium picture, while we can still mathematically gather
-terms and give the resulting collection the label $\Delta F_R$,
-it is unclear if the physical interpretation as an effective free energy
-makes sense.
-As is well known, free energies cannot be assigned unambiguously
-to states out of equilibrium because the free energy change along
-a generic path traversing the state space is path dependent,
-unlike at equilibrium.
-A consequence of this is that, out of equilibrium, $\Delta F_R$
-is no longer a simple sum of Boltzmann weights. Instead it
-resembles a restricted sum of King-Altman
-diagrams~\cite{King1956, Hill1966}. Following the work of
-Hill~\cite{Hill1989}, it may yet be possible to interpret this expression
-as an effective free energy, but this remains unclear to us.
-We leave this an open problem for future work.
-
-If we relax the requirement of exclusive repressor-RNAP binding,
-one could imagine models in which repressor and RNAP doubly-bound
-states are allowed, where the repressor's effect is to
-\text{reduce} the transcription rate rather than setting it to zero.
-Our results do not strictly apply to such a model, although we
-note that if the repressor's reduction of the transcription rate
-is substantial, such a model might still be well-approximated by
-one of the models in~\fig{fig1:means_cartoons}.
-
-One may worry that our ``one curve to rule them all'' is a
-mathematical tautology. In fact we \textit{agree} with this
-criticism if $\Delta F_R$ is ``just a fitting parameter'' and cannot
-be meaningfully interpreted as a real, physical free energy.
-An analogy to Hill functions is appropriate here. One of their
-great strengths and weaknesses, depending on the use they are put
-to, is that their parameters coarse-grain many details and are
-generally not interpretable in terms of microscopic models, for
-deep reasons discussed at length in~\cite{Frank2013}.
-By contrast, our master curve claims to have the best of both worlds:
-a coarse-graining of all details besides the repressor into
-a single effective parameter $\rho$,
-while simultaneously retaining an interpretation of
-$\Delta F_R$ as a physically
-meaningful and interpretable free energy.
-Our task, then, is to prove or disprove this claim.
-
-How do we test this and probe the theory with fold-change
-measurements? There is a fundamental limitation in that the
-master curve is essentially a one-parameter function of $\Delta
-F_R + \log\rho$. Worse, there are many \textit{a priori}
-plausible microscopic mechanisms that could contribute to the
-value of $\rho$, such as RNAP binding and escape
-kinetics~\cite{DeHaseth1998, Mitarai2015}, and/or supercoiling
-accumulation and release~\cite{Chong2014, Sevier2016}, and/or,
-RNAP clusters analogous to those obvserved in
-eukaryotes~\cite{Cisse2013, Cho2016}
-and recently also observed in bacteria~\cite{Ladouceur2020}.
-Even if $\Delta F_R$ is measured to
-high precision, inferring the potential microscopic contributions
-to $\rho$, buried inside a log no less, from fold-change
-measurements seems hopeless. As a statistical inference problem
-it is entirely nonidentifiable, in the language
-of~\cite{Gelman2013}, Section 4.3.
-
-If we cannot simply infer values of $\rho$ from measurements of
-fold-change, can we perturb some of the parameters that make up
-$\rho$ and measure the change? Unfortunately we suspect this is
-off-limits experimentally: most of the potential contributors to
-$\rho$ are global processes that affect many or all genes. For
-instance, changing RNAP association rates by changing RNAP copy
-numbers, or changing supercoiling kinetics by changing
-topoisomerase copy numbers, would massively perturb the entire
-cell's physiology and confound any determination of $\rho$.
-
-One might instead imagine a bottom-up modeling approach, where we
-mathematize a model of what we hypothesize the important steps
-are and are not, use in vitro data for the steps deemed
-important, and \textit{predict} what $\rho$ should be.
-But again, because of the one-parameter nature of the master
-curve, many different models will likely make indistinguishable
-predictions, and without any way to experimentally perturb in
-vivo, there is no clear way to test whether the modeling
-assumptions are correct.
-
-In light of this, we prefer the view that parameters and rates
-are not directly comparable between cartoons
-in~\fig{fig1:means_cartoons}. Rather, parameters in the
-simpler cartoons represent coarse-grained combinations of
-parameters in the finer-grained models. For instance, by equating
-$\rho$ between any two models, one can derive various possible
-correspondences between the two models' parameters. Note that
-these correspondences are clearly not unique, since many possible
-associations could be made. It then is a choice as to what microscopic
-interpretations the model-builder prefers for the parameters in a
-particular cartoon, and as to which coarse-grainings lend
-intuition and which seem nonsensical.
-Indeed, since it remains an open question what microscopic
-features dominate $\rho$ (as suggested above, perhaps RNAP
-binding and escape kinetics~\cite{DeHaseth1998, Mitarai2015}, or
-supercoiling accumulation and release~\cite{Chong2014, Sevier2016},
-or, something more exotic like
-RNAP clusters~\cite{Cisse2013, Cho2016, Ladouceur2020}),
-we are hesitant to put too much weight on any one microscopic
-interpretation of model parameters that make up $\rho$.
-
-One possible tuning knob to probe $\rho$ that would not globally
-perturb the cell's physiology is to manipulate RNAP binding sites.
-Work such as~\cite{Kinney2010} has shown that models of
-sequence-dependent RNAP can be inferred from data, and the
-authors of~\cite{Brewster2012} showed that the model
-of~\cite{Kinney2010} has predictive power by using the model to
-\textit{design} binding sites of a desired affinity.
-But for our purposes, this begs the question: the authors
-of~\cite{Kinney2010} \textit{assumed} a particular model
-(essentially our 3-state equilibrium model but without the
-repressor), so it is unclear how or if such methods can be turned
-around to \textit{compare} different models of promoter function.
-
-We have already pointed out that the master curve
-of~\fig{fig1:means_cartoons} is essentially a one-parameter model,
-the one parameter being $\Delta F_R + \log\rho$.
-By now the reader may be alarmed as to how can we even determine
-$\Delta F_R$ and $\rho$ independently of each other, never mind
-shedding a lens on the internal structure of $\rho$ itself.
-A hint is provided by the weak promoter approximation, invoked
-repeatedly in prior studies~\cite{Bintu2005c, Garcia2011a,
-Razo-Mejia2018} of simple repression using the 3-state
-equilibrium model in~\fig{fig1:means_cartoons}.
-In that picture, the weak promoter approximation means
-$\frac{P}{N_{NS}}\exp(-\beta\Delta\epsilon_P) \ll 1$,
-meaning therefore $\rho\approx1$.
-This is suggestive, but how can we be sure that $\rho$ is not,
-for instance, actually $10^2$ and that $\Delta F_R$ hides a
-compensatory factor?
-A resolution is offered by an independent inference of $\rho$ in
-the absence of repressors. This was done in~\cite{Razo-Mejia2020}
-by fitting nonequilibrium model 4  in~\fig{fig1:means_cartoons},
-with zero repressor (looking ahead, this is equivalent to model 4
-in~\fig{fig:constit_cartoons}), to FISH data
-from~\cite{Brewster2014}. This provided a determination of $k^+$
-and $k^-$, from which their ratio is estimated to be no more than
-a few $10^{-1}$ and possibly as small as $10^{-2}$.
-
-The realization that $\rho\approx1$ to an excellent
-approximation, \textit{independent} of which model
-in~\fig{fig1:means_cartoons} one prefers, goes a long way towards
-explaining the surprising success of equilibrium models of simple
-repression. Even though our 2- and 3-state models get so many
-details of transcription wrong, it does not matter because
-fold-change is a cleverly designed ratio. Since $\rho$ subsumes
-all details except the repressor, and $\log\rho\approx0$, fitting
-these simple models to fold-change measurements can still give a
-surprisingly good estimate of repressor binding energies. So the
-ratiometric construction of fold-change fulfills its intended
-purpose of canceling out all features of the promoter
-architecture except the repressor itself. Nevertheless it is
-perhaps surprising how effectively it does so: \textit{a priori},
-one might not have expected $\rho$ to be quite so close to 1.
-
-We would also like to highlight the relevance
-of~\cite{Landman2019} here. Landman et.\ al.\ reanalyzed and
-compared in vivo and in vitro data on the lacI repressor's
-binding affinity to its various operator sequences. (The in vivo
-data was from, essentially, fitting our master curve to
-expression measurements.) They find broad agreement between the
-in vitro and in vivo values. This reinforces the suspicion that
-the equilibrium $\Delta\epsilon_R$ repressor binding energies do
-in fact represent real physical free energies.
-Again, \textit{a priori} this did not have to be the case, even
-knowing that $\rho\approx1$.
-
-In principle, if $\Delta F_R$ can be measured to sufficient
-precision, then deviations from $\rho=1$ become a
-testable matter of experiment.
-In practice, it is probably unrealistic to measure repressor
-rates $k_R^+$ or $k_R^-$ or fold-changes in expression levels
-(and hence $\Delta\epsilon_R$) precisely enough to detect the
-expected tiny deviations from $\rho=1$.
-We can estimate the
-requisite precision in $\Delta F_R$ to resolve a given
-$\Delta\rho$ by noting, since $\rho\approx1$, that
-$\log(1+\Delta\rho)\approx \Delta\rho$, so
-$\Delta(\Delta F_R) \approx \Delta\rho$.
-Suppose we are lucky and $\Delta\rho$ happens to be $\sim0.1$, on
-the high end of our range estimated above.
-A determination of $\Delta\epsilon_R/k_BT$ with an uncertainty of
-barely 0.1 was achieved in the excellent measurements
-of~\cite{Razo-Mejia2018}, so this requires a very difficult
-determination of $\Delta F_R$ for a very crude determination of $\rho$,
-which suggests, to put it lightly, this is not a
-promising path to pursue experimentally.
-It is doubtful that inference of repressor kinetic rates
-would be any easier.\footnote{
-\mmnote{
-In fact the uncertainties quoted in Hammar et.\
-al.~\cite{Hammar2014} from Elf's group are small enough to do
-this calculation, but HJ \& I think their error bars are insanely
-optimistic. Not sure I want to blatantly call them out on it though??}
-}
-
-Moving forward, we have weak evidence supporting the
-interpretation of $\Delta F_R$ as a physically real free
-energy~\cite{Landman2019} and other work casting
-doubt~\cite{Hammar2014}.
-How might we resolve the confusion?
-If there is no discriminatory power to test the theory and
-distinguish the various models with measurements of fold-changes
-in means, how do we probe the theory? Clearly to discriminate
-between the nonequilibrium models in~\fig{fig1:means_cartoons},
-we need to go beyond means to ask questions about kinetics, noise and even
-full distributions of mRNA copy numbers over a population of cells.
-If the ``one-curve-to-rule-them-all'' is more than a
-mathematical tautology, then the free energy of repressor binding
-inferred from fold-change measurements should agree with
-repressor binding and unbinding rates.
-In other words, the equilbrium and nonequilibrium definitions of
-$\Delta F_R$ must agree, meaning
-\begin{equation}
-\Delta F_R = \beta\Delta\epsilon_R - \log(R/N_{NS})
-        = - \log(k_R^+/k_R^-),
-\end{equation}
-must hold, where $\beta\Delta\epsilon_R$ is inferred from the
-master curve fit to fold-change measurements, and $k_R^+$ and
-$k_R^-$ are inferred in some orthogonal manner.
-Single molecule measurements such as from~\cite{Hammar2014} have
-directly observed these rates, and in the remainder of this work
-we explore a complementary approach: inferring repressor binding
-and unbinding rates $k_R^+$ and $k_R^-$ from FISH measurements of
-mRNA population distributions.
-
-\section{Beyond means}
-\subsection{A one-state promoter with bursting is the ``best'' model of constitutive promoters}
-Before we can tackle simple repression, we need an adequate
-phenomenological model of constitutive expression.
-The literature abounds with options from which we can choose,
-and we show several potential kinetic models for constitutive promoters
-in~\fig{fig:constit_cartoons}.
-Let us consider the suitability of each model for our purposes in turn.
-
-\begin{figure}%[h!]
-\centering
-\includegraphics[width=0.9\textwidth]{../figures/fig1.5/fig1point5.ai}
-\caption{\textbf{Constitutive promoter cartoon comparison.}
-The left column depicts various plausible cartoons for constitutive promoters.
-We also list literature references where models have been used previously.
-In model (1), transcripts are produced in a Poisson process~\cite{Jones2014}.
-Model (2) features explicit modeling of RNAP binding/unbinding
-kinetics~\cite{Phillips2015a}.
-Model (3) is a more detailed generalization of model (2),
-treating transcription initiation as a multi-step process
-proceeding through closed and open complexes~\cite{Mitarai2015}.
-Model (4) is somewhat analogous to (2) except with the precise nature
-of active and inactive states left ambiguous~\cite{Razo-Mejia2020}.
-Finally, model (5) can be viewed as a certain limit of model (4)
-in which transcripts are produced in bursts,
-and initiation of bursts is a Poisson process.
-\mmnote{Haven't found model 5 formulated quite like this in the literature,
-usually burstiness is handled like model 4, but I should probably search a bit more.}
-The right column shows the Fano factor for each model.
-Note especially the crucial diagnostic: (2) and (3) have $\nu$
-strictly below 1, while only for (4) and (5) can $\nu$ exceed 1.
-Models with Fano factors $\le 1$ cannot produce the FISH data
-observed in~\cite{Jones2014} without introducing additional
-assumptions and model complexity.
-}
-\label{fig:constit_cartoons}
-\end{figure}
-
-\subsubsection{Model 1 - Poisson promoter.}
-This is the picture from Jones et.\ al.~\cite{Jones2014}, which
-assumes that transcripts are produced as a Poisson process from a
-single promoter state. This model insists that the ``true'' mRNA
-distribution is Poisson, implying the Fano factor $\nu$ must be 1.
-In~\cite{Jones2014}, the authors carefully attribute deviations
-from Fano = 1 to intensity variability in fluorescent spots, gene
-copy number variation, and copy number fluctuations of the
-transcription machinery, e.g., RNAP itself.
-In this picture, the master equation makes no appearance,
-and all the corrections to Poisson behavior are derived
-as additive corrections to the Fano factor.
-For disproving the ``universal noise curve'' from So et
-al~\cite{So2011}, this picture was excellent. It is appealing in
-its simplicity, with only two parameters, the initiation rate $r$
-and degradation rate $\gamma$. Since $\gamma$ is independently
-known from other experiments, and the mean mRNA copy number is
-$r/\gamma$, $r$ is easily inferred from data.
-In other words, the model is not excessively complex for the data at hand.
-But for many interesting questions, for instance in the recent
-work~\mmnote{cite Manuel's preprint once posted},
-knowledge of means and variances alone is insufficient, and a
-description of the full distribution of molecular counts is necessary.
-For this we need a( slightly) more complex model than model 1
-that would allow us to incorporate the non-Poissonian features of
-constitutive promoters directly into a master equation formulation.
-
-\subsubsection{Model 2 - Two-state promoter, RNAP bound or unbound.}
-This model was considered in, e.g.,~\cite{Phillips2015a} and ~\cite{Phillips2019}.
-Here transcription initiation proceeds from the bound to the unbound state,
-reflecting the microscopic reality that an RNAP that has begun to elongate
-a transcript is no longer available at the start site to begin another.
-The problem with this picture is that the Fano factor is
-\begin{align}
-    \nu = 1 -
-        \frac{r\rate{k}{P}{+}}
-            {\left(\rate{k}{P}{+} + \rate{k}{P}{-} + r\right)
-             \left(\gamma + \rate{k}{P}{+} + \rate{k}{P}{-} + r\right)},
-\end{align}
-which is always $<1$.
-To make contact with the experimental reality of $\nu>1$,
-we will have to do, at a minimum,
-the same corrections to Poisson behavior as in model 1 above.
-So while this model adds an appealing element of microscopic
-reality, we are forced to reject it as the additional complexity
-is unable to capture the phenomenology of interest.
-Obviously the promoter state does in fact proceed through cycles
-of RNAP binding, initiating, and elongating, but it seems that
-the super-Poissonian noise in mRNA copy number we want to
-model must be governed by other features of the system.
-
-\subsubsection{Model 3 - Three-state promoter, multistep
-transcription initiation and escape.}
-How might we remedy the deficits of model 2?
-It is known~\cite{DeHaseth1998} that once RNAP initially binds
-the promoter region, a multitude of distinct steps occur
-sequentially before RNAP finally escapes into the elongation
-phase. Perhaps adding some of this mechanistic detail might
-rescue model 2? The next simplest refinement of model 2 could
-consider open complex formation and promoter escape as separate
-steps rather than as a single effective step. In other words, we
-construct model 3 by adding a single extra state to model 2, and
-we will label the two RNAP-bound states as the closed and open
-complexes, despite the true biochemical details certainly being
-more complex. The authors of~\cite{Mitarai2015}, for instance,
-considered this model, although they added an additional
-repressor bound state and did not explicitly consider the limit
-with no repressor that we analyze here. Again, our goal here is
-not a complete accounting of all the relevant biochemical detail;
-this is an exploratory search for the important features our
-model needs to include.
-
-Unfortunately, as the authors of~\cite{Mitarai2015} hint, this
-model too has Fano factor $\nu<1$. This can be seen by starting
-with their Eq.~A1 for the Fano factor of the analogous model with
-repressor and taking the limit as repressor concentration goes to
-zero. After substantial algebra, the result is
-\begin{align}
-\nu = 1 - \frac{r k_O k_P^+}{\mathcal{Z}}
-\frac{k_P^+ + k_P^- + k_O + \gamma}
-        {\mathcal{Z} + \gamma(k_P^+ + k_P^- + k_O + \gamma) + \gamma^2},
-\end{align}
-where we defined $\mathcal{Z} = r(k_O + k_P^-) + k_P^+(k_O + r)$
-for notational tidiness. This is necessarily less than 1 for
-arbitrary rate constants.
-
-In fact, we suspect \textit{any} model in which transcription
-proceeds through a multistep cycle must necessarily have $\nu<1$.
-The intuitive argument compares the waiting time distribution to
-traverse the cycle with the waiting time for a Poisson promoter
-(model 1) with the same mean time. The latter is simply an
-exponential distribution. The former is a convolution of multiple
-exponentials, and intuitively should be more peaked with a
-smaller fractional width than an exponential with the same mean.\footnote{
-This can be made more precise with a result
-from~\cite{Stewart2007}, who showed that the convolution of
-multiple gamma distributions (of which the exponential
-distribution is a special case) is to a very good approximation
-also gamma distributed. Using their Eq.~2 for the distribution of
-the convolution, with shape parameters set to 1 to
-give exponential distributions, the total waiting time
-distribution has a ratio of variance to squared mean 
-$\sigma^2/\mu^2 = \sum_i k_i^2/\left(\sum_i k_i\right)^2$,
-where the $k_i$ are the rates of the individual steps. Clearly
-this is less than 1 and therefore the total waiting time
-distribution is narrower than the corresponding exponential.} A
-less disperse waiting time distribution means transcription
-initations are more uniformly distributed in time relative to a
-Poisson process. Hence the distribution of mRNA over a population
-of cells should be less variable compared to Poisson, giving
-$\nu<1$.\footnote{
-This last step, while intuitive, can be argued by analogy to
-photon statistics where antibunching gives rise to sub-Poissonian
-noise~\cite{Paul1982, Zou1990}. Although loopholes exist, we do
-not expect they apply for our problem. Nevertheless we refrain
-from elevating this cycle/sub-Poissonian equivalence to a ``theorem.''
-\mmnote{Antibunching is the ``obvious'' analogy for me but it'd
-be totally out of left field for most readers, any suggestions on
-a more comprehensible reference??}
-}
-
-Regardless of the merits of this model in describing the noise
-properties of constitutive transcription initation, it does not
-capture the dominant source of noise we want to include in our
-phenomenological model, so we must discard these details and
-search elsewhere.
-
-\subsubsection{Model 4 - Two-state promoter,
-``active'' and ``inactive'' states.}
-Inspired by~\cite{Razo-Mejia2020}, we revisit an analog of model
-2, but the interpretation of the two states is changed. Rather
-than explicitly viewing them as RNAP bound and unbound, we view
-them as ``active'' and ``inactive,'' which are able and unable to
-initiate transcripts, respectively. We are noncommittal as to the
-microscopic details of these states.
-
-One interpretation~\cite{Chong2014, Sevier2016} for the active
-and inactive states is that it represents the promoter's
-supercoiling state: transitions to the inactive state are caused
-by accumulation of positive supercoiling, which inhibits
-transcription, and transitions back to ``active'' are caused by
-gyrase or other topoisomerases relieving the supercoiling. This
-is an interesting possibility because it would mean the timescale
-for promoter state transitions is driven by topoisomerase
-kinetics, not by RNAP kinetics. From in vitro measurments, the
-former are suggested to be of order minutes~\cite{Chong2014}.
-Contrast this with model 2, where the state transitions are
-assumed to be governed by RNAP, which, assuming a copy number per
-cell of order $10^3$, has a diffusion-limited association rate
-$k_{on} \sim 10^2~\text{s}^{-1}$ to a target promoter. Combined
-with known $K_D$'s of order $\mu$M, this gives an RNAP
-dissociation rate $k_{off}$ of order $10^2$. As we will show
-below, however, there are some lingering puzzles with
-interpreting this supercoiling interpretation, so we leave it as
-a speculation and refrain from assigning definite physical
-meaning to the two states in this model.
-
-Intuitively one might expect that, since transcripts are produced
-as a Poisson process only when the promoter is in one of the two
-states in this model, transcription initiations should now be
-``bunched,'' in constrast to the ``anti-bunching'' of models 2
-and 3 above. One might further guess that this bunching would
-lead to super-Poissonian noise in the mRNA distribution over a
-population of cells. An honest calculation of the Fano factor produces
-\begin{align}
-\nu &= 1 + \frac{r k^-}{(k^+ + k^- + \gamma)(k^+ + k^-)},
-\end{align}
-which is strictly greater than 1, verifying the above intuition.
-Note we have dropped the $P$ label on the promoter switching rates to
-emphasize that these very likely do not represent kinetics of RNAP itself.
-This calculation can also be sidestepped by noting that the model
-is mathematically equivalent to the simple repression repression model
-from~\cite{Jones2014}, with states and rates relabeled and reinterpreted.
-
-How does this model compare to model 1 above?
-In model 1, all non-Poisson features of the mRNA
-distribution were handled as extrinsic corrections.
-By contrast, here the 3 parameter model is used to fit the full
-mRNA distribution as measured in mRNA FISH experiments.
-In essense, all variability in the mRNA distribution is regarded
-as ``intrinsic,'' arising either from stochastic initiation or
-from switching between the two coarse-grained promoter states.
-The advantage of this approach is that it fits neatly into the
-master equation picture, and the parameters thus inferred can be used as
-input for more complicated models with regulation by transcription factors.
-
-While this seems promising, there is a major drawback for our purposes
-which was already uncovered by the authors of~\cite{Razo-Mejia2020}:
-the statistical inference problem is nonidentifiable,
-in the sense described in Section 4.3 of~\cite{Gelman2013},
-meaning it is impossible to infer the
-parameters $r$ and $k^-$ from the FISH data of~\cite{Jones2014}
-(as shown in Fig.~S2 of~\cite{Razo-Mejia2020}).
-Rather, only the ratio $r/k^-$ could be inferred. In that work,
-the problem was worked around with an informative prior on the
-ratio $k^-/k^+$. That approach is unlikely to work here, as,
-recall, our entire goal in modeling constitutive expression is to
-use it as the basis for a yet more complicated model, when we add
-on repression. But adding more complexity to a model that is
-already poorly identified is a fool's errand,
-so we will explore one more potential model.
-
-\subsubsection{Model 5 - One-state promoter with explicit bursts.}
-This model is inspired by the failure mode of model 4.
-The key observation above was that, as found
-in~\cite{Razo-Mejia2020}, only two parameters, $k^+$ and the
-ratio $r/k^-$, could be directly inferred from the FISH data
-from~\cite{Jones2014}.
-So let us take this seriously and imagine a model where these are
-the only two model parameters. What would this model look like?
-
-To develop some intuition, consider model 4 in the limit
-$k^+ \ll k^- \lesssim r$, which is roughly satisfied by the
-parameters inferred in~\cite{Razo-Mejia2020}.
-In this limit, the system spends the majority of its time in the inactive state, occasionally becoming active and making a burst of transcripts.
-This should call to mind the well-known phenomenon of transcriptional bursting, as reported in, e.g.,~\cite{Golding2005,Chong2014,Sevier2016}\mmnote{should probably add some more cites here}.
-Let us make this correspondence more precise.
-The mean dwell time in the active state is $1/k^-$.
-While in this state, transcripts are produced at a rate $r$ per
-unit time. So on average, $r/k^-$ transcripts are produced before
-the system switches to the inactive state. Once in the inactive
-state, the system dwells there for an average time $1/k^+$ before
-returning to the active state and repeating the process. $r/k^-$
-resembles an average burst size, and $1/k^+$ resembles the time
-interval between burst events. More precisely, $1/k^+$ is the
-mean time between the end of one burst and the start of the next,
-whereas $1/k^+ + 1/k^-$ would be the mean interval between the
-start of two successive burst events, but in the limit $k^+ \ll
-k^-$, $1/k^+ + 1/k^- \approx 1/k^+$. Note that this limit ensures
-that the waiting time between bursts is approximately
-exponentially distributed, with mean set by the only timescale
-left in the problem,
-$1/k^+$.\footnote{
-If instead it were the case that $k^+ \sim k^-$, then the waiting time
-$1/k^+ + 1/k^-$ would have a peaked distribution, but this does
-not appear to be the case for any of datasets
-from~\cite{Jones2014}.}
-
-Let us now verify this intuition with a precise derivation
-to check that $r/k^-$ is in fact the mean burst size
-and to obtain the full burst size distribution.
-Consider first a constant, known dwell time $T$ in the active state.
-Transcripts are produced at a rate $r$ per unit time, so the
-number of transcripts $n$ produced during $T$ fits exactly
-the ``story'' for the Poisson distribution, i.e.,
-\begin{equation}
-    P(n\mid T) = \frac{(rT)^n}{n!} \exp(-rT).
-\end{equation}
-Since the dwell time $T$ is unobservable, we actually want $P(n)$,
-the dwell time distribution with no conditioning on $T$.
-Basic rules of probability theory tell us we can write $P(n)$
-in terms of $P(n\mid T)$ as
-\begin{equation}
-    P(n) =\int_0^\infty P(n\mid T) P(T) dT.
-\end{equation}
-But we know the dwell time distribution $P(T)$,
-which is exponentially distributed according to
-\begin{equation}
-    P(T) = k^- \exp(-T k^-),
-\end{equation}
-so $P(n)$ can be written as
-\begin{equation}
-    P(n) = k^- \frac{r^n}{n!}
-            \int_0^\infty T^n\exp[-(r + k^-)T]\,dT.
-\end{equation}
-A standard integral table shows
-$\int_0^\infty x^n e^{-ax}\,dx = n!/a^{n+1}$, so
-\begin{equation}
-    P(n) = \frac{k^- r^n}{(k^- + r)^{n+1}}
-        = \frac{k^-}{k^- + r}
-            \left(\frac{r}{k^- + r}\right)^n
-        = \frac{k^-}{k^- + r}
-            \left(1 - \frac{k^-}{k^- + r}\right)^n,
-\end{equation}
-which is exactly the geometric distribution with standard parameter
-$\theta\equiv k^-/(k^- + r)$ and domain
-$n \in \{0, 1, 2, \dots\}$
-(this is one of two common conventions for the geometric distribution).
-The mean of the geometric distribution, with this convention, is
-\begin{align}
-\langle n\rangle = \frac{1 - \theta}{\theta}
-        = \left(1 - \frac{k^-}{k^- + r}\right)
-                    \frac{k^- + r}{k^-}
-        = \frac{r}{k^-},
-\end{align}
-exactly as we guessed intuitively above.
-
-So in taking the limit $r,k^-\rightarrow\infty$, $r/k^-\equiv b$,
-we obtain a model which effectively has only a single promoter state,
-which initiates bursts at rate $k^+$
-(transitions to the active state, in the model 4 picture).
-The master equation for mRNA copy number $m$ takes the form
-\begin{align}
-\begin{split}
-\deriv{t}{p(m,t)} = & (m+1)\gamma p(m+1,t) - m\gamma p(m,t) \\
-        &+ \sum_{j=0}^{m-1} k^+ p(j,t) Geom(m-j;b)
-         - \sum_{j=m+1}^\infty k^+ p(m,t) Geom(j-m;b),
-\end{split}
-\end{align}
-where $Geom(n;b)$ is the geometric distribution with mean~$b$, i.e.,
-$Geom(n;b) = \frac{1}{1+b}\left(\frac{b}{1+b}\right)^n$
-(with domain over nonnegative integers as above).
-The first two terms are the usual mRNA degradation terms.
-The third term enumerates all ways the system can produce
-a burst of transcripts and arrive at copy number $m$,
-given that it had copy number $j$ before the burst.
-The fourth term allows the system to start with copy number $m$,
-then produce a burst and end with copy number $j$.
-In fact this last sum has trivial $j$ dependence and
-simply enforces normalization of the geometric distribution.
-Carrying it out we have
-\begin{equation}
-\begin{split}
-\deriv{t}{p(m,t)} = & (m+1)\gamma p(m+1,t) - m\gamma p(m,t) \\
-        &+ \sum_{j=0}^{m-1} k^+ p(j,t) Geom(m-j;b)
-            - k^+ p(m,t),
-\end{split}
-\end{equation}
-This improves on model 4 in that now the parameters are easily inferred,
-as we will see later, and have clean interpretations.
-The non-Poissonian features are attributed to the emprically
-well-established phenomenological picture of bursty transcription.
-
-The big approximation in going from model 4 to 5 is that a burst
-is produced instantaneously rather than over a finite time.
-If the true burst duration is not short compared to transcription factor
-kinetic timescales, this could be a problem in that mean burst size
-in the presense and absence of repressors could change,
-rendering parameter inferences from the constitutive case inappropriate.
-Let us make some simple estimates of this.
-
-Consider the time delay between the first and final RNAPs
-in a burst initiating transcription
-(\textit{not} the time to complete transcripts,
-which potentially could be much longer.)
-If this timescale is short compared to the typical
-search timescale of transcription factors, then all is well.
-The estimates from deHaseth et.\ al.~\cite{DeHaseth1998}
-put RNAP's diffusion-limited on rate around
-$\sim\text{few}\times10^{-2}~\text{nM}^{-1}~\text{s}^{-1}$
-and polymerase loading as high as $1~\text{s}^{-1}$.
-Then for reasonable burst sizes of $<10$, it is reasonable to guess that
-bursts might finish initiating on a timescale of tens of seconds or less
-(with another 30-60 sec to finish elongation, but that does not matter here).
-A transcription factor with typical copy number of order 10 (or less)
-would have a diffusion-limited association rate of order
-$(10~\text{sec})^{-1}$.
-Higher copy number TFs tend to have many binding sites over the genome,
-which should serve to pull them out of circulation and keep their
-effective association rates from rising too large.
-Therefore, there is \textit{perhaps} a timescale separation
-possible between transcription factor association rates and burst
-durations, but this assumption could very well break down, so we
-will have to keep it in mind when we infer repressor rates from
-the Jones et.\ al.\ FISH data later.
-
-In reflecting on these 5 models, the reader may feel that
-exploring a multitude of potential models just to return to a
-very minimal phenomenological model of bursty transcription may
-seem highly pedantic. But the purpose of the exercise was to
-examine a host of models from the literature and understand why
-they are insufficient, one way or another, for our purposes.
-Along the way we have learned that the detailed kinetics of RNAP
-binding and initiating transcription are probably irrelevant for
-setting the population distribution of mRNA.
-The timescales are simply too fast, and as we will see later
-in~\fig{fig:constit_post}, the noise seems to be governed by
-slower timescales. Perhaps in hindsight this is not surprising:
-intuitively, the degradation rate $\gamma$ sets the fundamental
-timescale for mRNA dynamics, and any other processes that
-substantially modulate the mRNA distribution should not differ
-from $\gamma$ by orders of magnitude.
-
-\subsection{Parameter inference for constitutive promoters}
-Having established that a one-state bursty model achieves a
-Goldilocks level of complexity, we test it more thoroughly on all
-the data from constitutive promoters from~\cite{Jones2014}.
-\fig{fig:constit_post} shows the results.
-\begin{figure}%[h!]
-\centering
-\includegraphics[width=0.9\textwidth]{../figures/fig2/fig2pt1.pdf}
-\caption{\textbf{Constitutive promoter posterior
-        inference and predictive checks.}
-        (A) shows the joint posterior density from MCMC sampling of model 5,
-        bursty promoter with negative binomially-distributed steady state.
-        (B) shows the ECDF of the observed population distribution
-        of mRNA transcripts under the control of a constitutive lacUV5 promoter.
-        Posterior predictive percentiles are shown as colored bands for
-        model (1), Poisson, and model (5), negative binomial.
-        }
-\label{fig:uv5_post}
-\end{figure}
-
-\begin{figure}%[h!]
-\centering
-\includegraphics[width=0.9\textwidth]{../figures/fig2/fig2pt2.pdf}
-\caption{\textbf{Constitutive promoter model comparison.}
-    (A) Joint posterior distributions for burst rate $k_i$ and
-    mean burst size $b$ for 18 promoters from Jones 2014.
-    Each contour indicates the 95\% highest posterior density
-    region for a particular promoter, i.e., on a plot analogous
-    to~\fig{fig:uv5_post}, the contour would enclose
-    approximately 95\% of the samples, and thus 95\% of the
-    probability mass.
-    Note that the vertical axis is shared with (B).
-    Panel (B) plots the burst rate $k_i$ vs.\ binding energy from energy
-    matrices from Brewster 2012.
-    The dotted line shows the predicted slope according to
-    $\ln k_i/\gamma \sim \text{constant} - \beta\Delta\epsilon_P$,
-    described in text.
-    }
-\label{fig:constit_post}
-\end{figure}
-\mmnote{Discuss puzzling comparison w/ Chong2014: if supercoiling is the thing, why are my burst sizes all the same but burst rates vary?
-And why is the duty cycle of my promoter so low, ie., bursts so short?
-COmpare their fig 7E; their $\beta/\alpha$ is my $k^+/k^-$.
-They have one or two genes with very small $\beta/\alpha$, does the \textit{galK} locus just happen to be that, or is there a deeper disagreement?
-Hard to say w/o more data.}
-
-\subsection{Transcription factor kinetics can be inferred from FISH measurements}
-Figure~3 presents inferred transcription factor rates, compares
-them to single-molecule measurements, and compares their ratios
-to gold-standard binding energies.
-\begin{figure}%[h!]
-\centering
-\includegraphics[width=0.9\textwidth]{../figures/fig3/fig3.ai}
-\caption{\textbf{Simple repression model comparison.}
-(A) depicts the cartoon of simple repression for which we infer parameters.
-(B) shows 50\% and 95\% contours of several 2D slices of the 9D posterior
-distribution, from fitting one unbinding rate for each operator (O1, O2, O3)
-and one binding rate for each aTc concentration (corresponding to an
-unknown mean repressor copy number).
-(C) compares ratios of our inferred unbinding rates with operator
-binding energy differences measured in Garcia \& Phillips 2011 (triangles)
-and Razo-Mejia et.\ al.\ 2018 (squares).
-(D) compares our inferred unbinding rates with single-molecule measurements
-of the same from Hammar et.\ at.\ 2014 \mmnote{Johan Elf's group}.
-Values are of similar order of magnitude,
-but a precise quantitative comparison unfortunately depends sensitively
-on the assumed mRNA lifetime, which was not measured.
-    }
-\label{fig3:kR_inferences}
-\end{figure}
-
-\begin{figure}%[h!]
-\centering
-\includegraphics[width=0.9\textwidth]{../figures/figSIxx/ppc_many_pooled.pdf}
-\caption{\textbf{Posterior predictive checks.}
-(A) duplicates the posterior from~\fig{fig3:kR_inferences}B.
-(B), (C), and (D) show samples from the posterior predictive distribution
-for each of the datasets used in the inference.
-Samples are sorted by operator and plotted separately simply for visual clarity.
-The unrepressed promoter, UV5, is shown with each as a reference point.
-\mmnote{
-Obviously the PPC is not perfect,
-but considering the relative simplicity of the model, I'm pleasantly surprised
-how well it worked and that the model was even identifiable at all.
-Interestingly, and perhaps not surprisingly, only 3 of the 9 datasets
-were identifiable if fit in isolation: otherwise,
-it was only by making assumptions that certain rates are equal
-across datasets was identifiability achieved.
-}
-    }
-\label{fig4:ppc}
-\end{figure}
-
-\section{Discussion and future work}
-\mmnote{
-Do all my approximations work simply because galK happens to have
-a low duty cycle of supercoiling, $\beta/\alpha$ in Chong2014
-picture? A potentially interesting test: put in the dozen or so
-Hernan constructs at some other locus, preferably one of the ones
-Chong identified as having a large duty cycle (large $\beta/\alpha$),
-i.e., the opposite limit of galK. Then just measure fold-changes.
-You don't even need to do FISH, although that would be even
-better. If their model is right, $\rho$ should be approximately
-set by the duty cycle of supercoiling, NOT by RNAP kinetics.
-\textit{This is a NICE prediction for future work.} \textit{cspE}
-is claimed nonessential, prediction is you drop HG reporter
-constructs in there and the inferred binding energies of all lacI
-operators shift the same amount, by about $\ln(3)\sim 1kT$. Not
-huge, but detectable, although might be confounding effects from
-moving to a new locus. With FISH measurements, $k^+$ and $k^-$ in
-Fig 2, model 4 should now be separately inferrable if Chong2014
-measurements of $\beta/\alpha$ are correct. This would require an
-obnoxious amount of cloning, but once you did that, the
-measurements would be same old stuff that RP lab does.}
-
-%%%%%%%%%%%%%%%%%%%%% APPENDICES %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\appendix
-
-\section{Derivations of constitutive promoter models}
-\subsection{Model 2}
-\mmnote{Obtain Fano factor from CME.}
-\subsection{Model 3}
-\mmnote{Obtain Fano factor by taking limit from Mitarai et al's expression.}
-\subsection{Model 4}
-\mmnote{Obtain Fano factor by relabeling cartoon from Jones2014.}
-
+% !TEX root = ./busty_transcription.tex
 \section{Generating function solution for bursty promoters}
@@
 is needed rather than $\log p(m)$.
-
-% \bibliographystyle{nature}
-\bibliographystyle{abbrv}
-\bibliography{./library}
-
-\end{document}

### 2020-05-19 doc/section_01_introduction.tex
initial splitting of document into several files

--- 2020-05-17 doc/paper.tex
@@
-\documentclass[12pt]{article}%{amsart}
-\usepackage[top = 1.0in, bottom = 1.0in, left = 1.0in, right = 1.0in]{geometry}
-% See geometry.pdf to learn the layout options. There are lots.
-%\geometry{letterpaper}               % ... or a4paper or a5paper or ...
-%\geometry{landscape}             % Activate for for rotated page geometry
-\usepackage[parfill]{parskip}    % Activate to begin paragraphs with an empty line rather than an indent
-\usepackage{graphicx}
-\usepackage{amssymb}
-%\usepackage{epstopdf}
-% \usepackage{caption}
-\usepackage{amsmath}
-% \usepackage{longtable}
-% \usepackage{tabu}
-% \usepackage{accents} %to get undertildes for vec & mat
-\usepackage{color} % for colored text
-% \usepackage[normalem]{ulem} % for editing: striking out text using 'sout' command
-
-% \newcommand\mytablefigwidth{0.35\textwidth}
-
-% allows to use .ai files directly w/o resaving as pdf
-\DeclareGraphicsRule{.ai}{pdf}{.ai}{}
-
-% Handy math macros!
-\newcommand{\vect}[1]{\vec{#1}}
-\newcommand{\matr}[1]{\mathbf{#1}}
-\newcommand{\rate}[3]{{#1}_{#2}^{#3}}
-\newcommand{\mmnote}[1]{\textcolor{cyan}{(MM:~#1)}}
-
-% derivative macros. these are sneaky. the [{}] is an empty default 1st arg
-% usage: provide 1 arg: \deriv{x} gives d/dx
-% provide 2 args like so: \deriv[f]{x} gives df/dx
-\newcommand{\deriv}[2][{}]{\frac{d #1}{d #2}}
-\newcommand{\pderiv}[2][{}]{\frac{\partial #1}{\partial #2}}
-\newcommand{\psecderiv}[2][{}]{\frac{\partial{^2} #1}{\partial #2{^2}}}
-
-%%%%% Referencing macros %%%%%
-\newcommand{\fig}[1]{Figure~\ref{#1}}
-\newcommand{\tref}[1]{Table~\ref{#1}}
-\newcommand{\eq}[1]{Eq.~(\ref{#1})}
-\newcommand{\eqrange}[2]{Eqs.~(\ref{#1}-\ref{#2})}
-
-%%%%%
-\begin{document}
-
-\title{Simple repression draft}
-
-\maketitle
-
+% !TEX root = ./busty_transcription.tex
 \section{Introduction}
@@
 \begin{itemize}
-\item Our goal is to build phenomenological models of input-output
-functions of genetic circuits. More precisely, we want phenomenology
-in the sense of coarse-graining away atomistic detail, but still
-retaining biophysical meaning, e.g., we don't want to coarse-grain
-as far as Hill functions. Why not? We are motivated by studies
-like~\cite{Razo-Mejia2020}, for which Hill functions are
-insufficient. Such work requires a quantitative theory of how
-biophysical changes at the molecular level propagate to
-input-output functions at the genetic circuit level.
-We want concepts, not mere facts.
-In particular a key question is: at this level of
-coarse graining, what microscopic details do I need to explicitly
-model, and how do we figure that out? For example, do I need to
-worry about all or even any of the steps that individual RNAPs go
-through each time they make a transcript? Turning the question
-around, can we see any imprint of those processes in the
-available data? If the answer is no, then those processes are
-irrelevant for our purposes. Forward modeling and inverse
-(statistical inferential) modeling can complement each other
-beautifully here.
+\item Our goal is to build phenomenological models of input-output functions of
+genetic circuits. More precisely, we want phenomenology in the sense of
+coarse-graining away atomistic detail, but still retaining biophysical meaning,
+e.g., we don't want to coarse-grain as far as Hill functions. Why not? We are
+motivated by studies like~\cite{Razo-Mejia2020}, for which Hill functions are
+insufficient. Such work requires a quantitative theory of how biophysical
+changes at the molecular level propagate to input-output functions at the
+genetic circuit level. We want concepts, not mere facts. In particular a key
+question is: at this level of coarse graining, what microscopic details do I
+need to explicitly model, and how do we figure that out? For example, do I need
+to worry about all or even any of the steps that individual RNAPs go through
+each time they make a transcript? Turning the question around, can we see any
+imprint of those processes in the available data? If the answer is no, then
+those processes are irrelevant for our purposes. Forward modeling and inverse
+(statistical inferential) modeling can complement each other beautifully here.
 \end{itemize}}
 
-\fig{fig1:means_cartoons}A shows the qualitative picture of
-simple repression. An operator, the binding site on the DNA for a
-repressor protein, may be found occupied by a repressor, in which
-case transcription is blocked from occurring. Or it may be found
-unoccupied, in which case RNA polymerase (RNAP) may bind and
+\fig{fig1:means_cartoons}A shows the qualitative picture of simple repression.
+An operator, the binding site on the DNA for a repressor protein, may be found
+occupied by a repressor, in which case transcription is blocked from occurring.
+Or it may be found unoccupied, in which case RNA polymerase (RNAP) may bind and
 transcription can proceed.
 
-The key assumption we make is that binding of repressor and RNAP in the promoter region of interest is exclusive, meaning that one or the other may bind, but never may both be simultaneously bound.
-It is often imagined that when the repressor is bound to its operator,
-RNAP is sterically blocked from binding to its promoter sequence.
-Current evidence suggests this is sometimes, but not
-always the case~\mmnote{Cite that fig, from ??Nathan or Bill or
-Rob??, showing locations of repressor binding sites far upstream
-of promoter}, and it remains an interesting open question precisely
-how a repressor bound far upstream is able to repress transription.
-Suggestions include ``action-at-a-distance'' mediated by kinks in
-the DNA, formed when the repressor is bound, that prevent RNAP binding.
-Nevertheless, our modeling in this work is sufficiently
-coarse-grained that we simply assume exclusive binding and need
-not worry about such precise details.
+The key assumption we make is that binding of repressor and RNAP in the promoter
+region of interest is exclusive, meaning that one or the other may bind, but
+never may both be simultaneously bound. It is often imagined that when the
+repressor is bound to its operator, RNAP is sterically blocked from binding to
+its promoter sequence. Current evidence suggests this is sometimes, but not
+always the case~\mmnote{Cite that fig, from ??Nathan or Bill or Rob??, showing
+locations of repressor binding sites far upstream of promoter}, and it remains
+an interesting open question precisely how a repressor bound far upstream is
+able to repress transription. Suggestions include ``action-at-a-distance''
+mediated by kinks in the DNA, formed when the repressor is bound, that prevent
+RNAP binding. Nevertheless, our modeling in this work is sufficiently
+coarse-grained that we simply assume exclusive binding and need not worry about
+such precise details.
 
@@
 In other words, the concentration of the molecule is hidden inside, so for instance, $k_R^+$ is proportional to repressor copy number.}
-
-\section{Means}
-\subsection{Fold-changes are indistinguishable across models}
-As a first stop on our search for the ``right'' model of simple
-repression, let us consider what we can learn from theory and
-measurements of fold-change in mean expression. The fold-change is defined as
-\begin{equation}
-\text{fold-change}
-= \frac{\langle \text{gene expression with repressor present} \rangle}
-        {\langle \text{gene expression with repressor absent} \rangle}
-= \frac{\langle m\rangle(R\ne 0)}{\langle m\rangle(R = 0)}
-\label{eq:fc_def}
-\end{equation}
-where angle brackets denote the averge over a population of cells
-and mean mRNA $\langle m\rangle$ is viewed as a function of
-repressor copy number $R$.
-The second
-equality in~\ref{eq:fc_def} follows from assuming that the
-translation efficiency, i.e., the number of proteins translated
-per mRNA, is the same in both conditions.
-In other words, we assume that mean protein level is proportional
-to mean mRNA level, and that the proportionality constant is the
-same in both conditions and therefore cancels out in the ratio.
-This is reasonable since the cells in the two conditions are
-identical except for the presence of the transcription factor,
-and it is difficult to imagine the transcription factor having
-any direct effect on translation.
-
-Fold-change has proven a very convenient observable in past
-work~\cite{Garcia2011a, Razo-Mejia2018, Chure2019}.
-Part of its utility in dissecting transcriptional regulation is
-its ratiometric nature, which is always easier than making an
-absolute measurement. Also, by measuring otherwise identical cells
-with and without a transcription factor present, any biological
-noise common to both conditions can be made to cancel away.
-In addition, the presence of different operator sites is known to
-change the basal expression level in the absence of
-repressor~\mmnote{look up cite}.
-Computing fold-changes neatly cancels this effect as well.
-
-\fig{fig1:means_cartoons} depicts a smorgasbord
-of mathematized cartoons for simple repression that have appeared
-in previous literature. For each cartoon, we calculate the
-fold-change in mean gene expression as predicted by that model,
-deferring some algebraic details to the appendix.
-What we will find is that all cartoons collapse to a single
-master curve, shown in \fig{fig1:means_cartoons}c, which contains
-just two parameters. We label the parameters $\Delta F_R$, an
-effective free energy parametrizing the repressor-DNA
-interaction, and $\rho$, which subsumes all details of
-transcription in the absence of repressors.
-We will offer some intuition for why this master curve exists and discuss
-why it is likely impossible to discriminate ``right'' from ``wrong''
-cartoons given only measurements of fold-changes in expression.
-
-\subsubsection{Equilibrium, 2-state}
-In this simplest model, the promoter is idealized as existing in
-one of two states, either repressor bound or repressor unbound.
-The rate of transcription, and therefore the level of gene
-expression, is assumed to be proportional to the fraction of time
-spent in the repressor unbound state.
-Mathematically, this amounts to the statement that the mean mRNA level is
-given by
-\begin{equation}
-\langle m \rangle = \frac{r}{\gamma}
-        \left(1 + \frac{R}{N_{NS}} e^{-\beta\Delta\epsilon_R}\right)^{-1},
-\end{equation}
-where $r$ is the transcription rate from the repressor unbound
-state, $\gamma$ is the mRNA degradation rate, $R$ is repressor
-copy number, $N_{NS}$ is the number of nonspecific binding sites
-in the genome (where repressors spend most of their time when not
-bound to the operator, their specifc binding site),
-$\beta=1/k_BT$, and $\Delta\epsilon_R$ is the binding energy of a
-repressor to its operator site. The derivation of this result is deferred
-to the appendix.
-
-The fold-change is easily found as the ratio of mean mRNA with
-and without repressor. This gives
-\begin{equation}
-\text{fold-change}
-= \left(1 + \frac{R}{N_{NS}} e^{-\beta\Delta\epsilon_R}\right)^{-1},
-\end{equation}
-which clearly matches the form of the master curve
-in~\eq{fig1:means_cartoons} with $\rho=1$.
-
-In fact it was noted in~\cite{Chure2019} that this two-state
-model can be viewed as the coarse-graining of any equilibrium
-promoter model in which no transcriptionally active states have
-transcription factor bound. Put differently, there is no overlap
-between TF bound states and transcriptionally active states. We
-will see this explicitly in the 3-state equilibrium model below,
-but perhaps surprising is that something analogous carries over
-even to the nonequilibrium models we consider later.
-
-\subsubsection{Equilibrium, 3-state}
-Compared to the previous model, here we fine-grain the repressor
-unbound state into two separate states: empty, and RNAP bound.
-This picture was used in~\cite{Garcia2011a} as we use it here,
-and in~\cite{Razo-Mejia2018} and~\cite{Chure2019} it was generalized
-to incorporate small-molecule inducers that bind the repressor.
-Interestingly, the effect of this is simply to renormalize $R$
-from the total number of repressors to a smaller effective number,
-roughly (though not exactly) the number which are unbound by inducers.
-
-The mean mRNA, as derived in the appendix from a similar
-enumeration of states and weights as the previous model, is
-\begin{equation}
-\langle m \rangle = \frac{r}{\gamma}
-\frac{\frac{P}{N_{NS}} e^{-\beta\Delta\epsilon_P}}
-        {
-        1 + \frac{R}{N_{NS}} e^{-\beta\Delta\epsilon_R}
-        + \frac{P}{N_{NS}} e^{-\beta\Delta\epsilon_P}
-        },
-\end{equation}
-where the new variables are $\Delta\epsilon_P$, the difference in
-RNAP binding energy to its specific site (the promoter) relative
-to an average nonspecific background site, and the RNAP copy
-number, $P$. The fold-change again follows immediately as
-\begin{align}
-\text{fold-change}
-&= \frac{\frac{P}{N_{NS}} e^{-\beta\Delta\epsilon_P}}
-        {
-        1 + \frac{R}{N_{NS}} e^{-\beta\Delta\epsilon_R}
-        + \frac{P}{N_{NS}} e^{-\beta\Delta\epsilon_P}
-        }
-\frac{1 + \frac{P}{N_{NS}} e^{-\beta\Delta\epsilon_P}}
-        {\frac{P}{N_{NS}} e^{-\beta\Delta\epsilon_P}}
-\\
-&= \left(
-1 + \frac{\frac{R}{N_{NS}} e^{-\beta\Delta\epsilon_R}}
-        {1 + \frac{P}{N_{NS}} e^{-\beta\Delta\epsilon_P}}
-\right)^{-1}
-\\
-&= (1 + \exp(-\Delta F_R - \log\rho))^{-1},
-\end{align}
-with $\Delta F_R = \beta\Delta\epsilon_R - \log(R/N_{NS})$
-and $\rho = 1 + \frac{P}{N_{NS}}\mathrm{e}^{-\beta\Delta\varepsilon_P}$
-as shown in~\fig{fig1:means_cartoons}.
-
-\begin{figure}%[h!]
-\centering
-\includegraphics[width=\textwidth]{../figures/fig1/fig1.ai}
-\caption{
-\textbf{
-An overview of the simple repression motif at the level of means.}
-(A) depicts the qualitative biological picture of simple repression.
-(B) shows a variety of possible mathematized cartoons of simple
-repression, along with the effective parameter $\rho$
-which subsumes all regulatory details
-of the architecture excluding the repressor.
-(C) shows the ``master curve'' to which all cartoons in (B) collapse.
-}
-\label{fig1:means_cartoons}
-\end{figure}
-
-\subsubsection{Nonequilibrium model 1 - Poisson promoter}
-For our first kinetic model, we consider the simplest possible
-picture with only two states, repressor bound and unbound.
-This is exactly the model used for the main results of~\cite{Jones2014}.
-In this picture, repressor association and dissociation rates
-from its operator site, $k_R^+$ and $k_R^-$, respectively, govern
-transitions between the two states. When the system is in the
-unbound state, transcriptions initiate at rate $r$, which
-represents a coarse-graining of all the downstream processes into
-a single effective rate. mRNA is again degraded at rate $\gamma$.
-
-Let $p_{m,R}(t)$ denote the joint probability of finding the
-system in the repressor bound state with $m$ mRNA molecules
-present at time $t$, and similarly define $p_{m,U}(t)$ for the
-repressor unbound state. 
-This model is governed by a master equation giving the time
-evolution of $p_{m,R}(t)$ and $p_{m,U}(t)$, which is
-\begin{align}
-\begin{split}
-\deriv{t}p_{m,R}(t) =& - k_R^- p_{m,R}(t) + k_R^+ p_{m,U}(t)
-                + (m+1)\gamma p_{m+1,R}(t) - \gamma p_{m,R}(t)
-\\
-\deriv{t}p_{m,U}(t) =&\; k_R^- p_{m,R}(t) - k_R^+ p_{m,U}(t)
-                        + rp_{m-1,U}(t) - rp_{m,U}(t)
-                \\
-                &+ (m+1)\gamma p_{m+1,U}(t) - \gamma p_{m,U}(t).
-\end{split}
-\end{align}
-In each equation, the first two terms describe transitions
-between promoter states due to repressors unbinding and binding,
-respectively. The final two terms describe degradation of mRNA,
-decreasing the copy number by one, and the terms with coefficient
-$r$ describe transcription initiation increasing the mRNA copy
-number by one.
-
-We can greatly simplify the notation, which will be especially
-useful for the more complicated models yet to come, by
-re-expressing the master equation in vector form. The promoter
-states are collected into a vector and the rate constants are
-collected into matrices as
-\begin{equation}
-\vect{p}_m = \begin{pmatrix} p_{m,R} \\ p_{m,U} \end{pmatrix},\
-\matr{K} = \begin{pmatrix} -k_R^- & k_R^+ \\ k_R^- & -k_R^+ \end{pmatrix},\
-\matr{R} = \begin{pmatrix} 0 & 0 \\ 0 & r \end{pmatrix},\
-\label{eq:2state_cme_matrices}
-\end{equation}
-so that the master equation may be condensed as
-\begin{equation}
-\deriv{t}\vect{p}_m(t) =
-\left( \matr{K} - \matr{R} - \gamma m \matr{I} \right) \vect{p}_m(t)
-                + \matr{R} \vect{p}_{m-1}(t)
-                + \gamma (m+1) \matr{I} \vect{p}_{m+1}(t)
-                - \gamma m \matr{I} \vect{p}_m(t),
-\label{eq:2state_rep_cme}
-\end{equation}
-where $\matr{I}$ is the identity matrix.
-Taking steady state by setting time derivatives to zero,
-the mean mRNA can be found to be
-\begin{equation}
-\langle m \rangle = \frac{r}{\gamma}
-        \left(1 + \frac{k_R^+}{k_R^-}\right)^{-1},
-\end{equation}
-with the algebra details again deferred to the appendix.
-Recall $k_R^+$ is proportional to the repressor copy number, so
-in computing fold-change, absence of repressor corresponds to
-$k_R^+\rightarrow0$.
-Therefore fold-change in this model is simply
-\begin{equation}
-\text{fold-change} = \left(1 + \frac{k_R^+}{k_R^-}\right)^{-1},
-\end{equation}
-again matching the master curve of~\fig{fig1:means_cartoons} with $\rho=1$.
-
-\subsubsection{Nonequilibrium model 2 - RNAP bound/unbound states}
-Analogous to the three-state equilibrium model, this model
-fine-grains the repressor unbound state of nonequilibrium model 1,
-resolving it into an empty promoter state and an RNAP-bound state.
-Note in this picture, in contrast with model 4 below,
-transcription initiation is accompanied by a promoter state
-change, in keeping with the interpretation as RNAP-bound and
-empty states: if an RNAP successfully escapes the promoter and
-proceeds to elongation of a transcript, clearly it is no longer
-bound at the promoter! Therefore another RNAP must bind before
-another transcript can be initiated.
-
-The master equation governing this model is analogous
-to~\eqrange{eq:2state_cme_matrices}{eq:2state_rep_cme} for model 1 above.
-The main subtlety arises since transcription initiation
-accompanies a promoter state change.
-This can be understood by analogy to $\matr{K}$. The
-off-diagonals and diagonals of $\matr{K}$ correspond to
-transitions arriving at or departing from, respectively, the
-promoter state of interest. If transcription initiations are
-accompanied by promoter state changes, we must have separate
-matrices for arriving and departing transcription events since
-the arriving and departing transitions have different initial
-copy number of mRNA, unlike for $\matr{K}$ where they are the same.
-The master equation for this model is
-\begin{equation}
-\deriv{t}\vect{p}_m(t) =
-\left( \matr{K} - \matr{R_D} - \gamma m \matr{I} \right) \vect{p}_m(t)
-                + \matr{R_A} \vect{p}_{m-1}(t) +
-                \gamma (m+1) \matr{I} \vect{p}_{m+1}(t)
-                - \gamma m \matr{I} \vect{p}_m(t),
-\label{eq:3state_rep_cme}
-\end{equation}
-with the state vector and promoter transition matrix defined as
-\begin{equation}
-\vect{p}_m = \begin{pmatrix} p_{m,R} \\ p_{m,E} \\ p_{m,P} \end{pmatrix},\
-\matr{K} = \begin{pmatrix} -k_R^- & k_R^+ & 0 \\
-                        k_R^- & -k_R^+ -k_P^+ & k_P^- \\
-                        0 & k_P^+ & -k_P^- 
-                \end{pmatrix},
-\label{eq:3state_cme_matrices_pt1}
-\end{equation}
-and the initiation matrices given by
-\begin{equation}
-\matr{R_A} = \begin{pmatrix}
-                0 & 0 & 0 \\ 0 & 0 & r \\ 0 & 0 & 0
-                \end{pmatrix},\
-\matr{R_D} = \begin{pmatrix}
-                0 & 0 & 0 \\ 0 & 0 & 0 \\ 0 & 0 & r
-                \end{pmatrix},
-\label{eq:3state_cme_matrices_pt2}
-\end{equation}
-The elements of $\vect{p}_m$ encode the probabilities of having
-$m$ mRNA present along with the promoter having repressor bound,
-being empty, or having RNAP bound, respectively.
-$\matr{R_A}$ describes probability flux \textit{arriving} at the state
-$\vect{p}_m$ from a state with one fewer mRNA, namely
-$\vect{p}_{m-1}$, and $\matr{R_D}$ describes probability flux
-\textit{departing} from the state $\vect{p}_m$
-for a state with one more mRNA, namely $\vect{p}_{m+1}$.
-$\matr{K}$ is closely analogous to model 1.
-
-Mean mRNA at steady state is found analogously to model 1, with the result
-\begin{equation}
-\langle m\rangle = \frac{r}{\gamma}
-        \frac{k_R^- k_P^+}
-        {k_R^- k_P^+ + k_R^- (k_P^- + r) + k_R^+ (k_P^- + r)}.
-\end{equation}
-Fold-change is again found from the ratio prescribed by~\eq{eq:fc_def},
-from which we have
-\begin{align}
-\text{fold-change}
-&=      \frac{k_R^- k_P^+}
-        {k_R^- k_P^+ + k_R^- (k_P^- + r) + k_R^+ (k_P^- + r)}
-        \frac{k_P^+ + k_P^- + r}{k_P^+}
-\\
-&=      \left(1 + \frac{k_R^+}{k_R^-}
-                \frac{k_P^- + r}{k_P^+ + k_P^- + r}
-        \right)^{-1}
-\\
-&=      \left(1 + \frac{k_R^+}{k_R^-}
-        \left(1 + \frac{k_P^+}{k_P^- + r}\right)^{-1}
-        \right)^{-1},
-\end{align}
-which follows the master curve with
-$\rho = 1 + k_P^+/(k_P^- + r)$ as claimed.
-
-\subsubsection{Nonequilibrium model 3
-- Multistep transcription initiation and escape}
-One might reasonably complain that the first two
-``nonequilibrium'' models we have considered are straw men. Their
-steady states necessarily satisfy detailed balance which is
-equivalent to thermodynamic equilibrium. Why is this the case? At
-steady state there is by definition no net probability flux in or
-out of each promoter state, but since the promoter states form a
-linear chain, there is only one way in or out of the repressor
-bound and RNAP bound states, implying each edge must actually
-have a net zero probability flux, which is the definition of
-detailed balance (usually phrased as equality of forward and
-reverse transition fluxes).
-
-Now we consider model 3 in~\eq{fig1:means_cartoons} which allows
-the possibility of true nonequilibrium steady-state fluxes
-through the promoter states.
-We point out that this model was considered previously in~\cite{Mitarai2015}
-where a comparison was made with model 1 as used in~\cite{Jones2014}.
-The authors of~\cite{Mitarai2015} argued that the additional
-complexity is essential to properly account for the noise in the
-mRNA distribution. We will weigh in on both models later when we
-consider observables beyond fold-change.
-
-The master equation governing this model is
-identical in form to model 2 above, namely
-\begin{equation}
-\deriv{t}\vect{p}_m(t) =
-\left( \matr{K} - \matr{R_D} - \gamma m \matr{I} \right) \vect{p}_m(t)
-                + \matr{R_A} \vect{p}_{m-1}(t) +
-                \gamma (m+1) \matr{I} \vect{p}_{m+1}(t)
-                - \gamma m \matr{I} \vect{p}_m(t),
-\end{equation}
-but obviously with a higher-dimensional state space and different matrices.
-The state vector and promoter transition matrix are now
-\begin{equation}
-\vect{p}_m = \begin{pmatrix} p_{m,R} \\ p_{m,E} \\
-                             p_{m,C} \\ p_{m,O}\end{pmatrix},\
-\matr{K} = \begin{pmatrix} -k_R^- & k_R^+ & 0 & 0\\
-                        k_R^- & -k_R^+ -k_P^+ & k_P^- & 0 \\
-                        0 & k_P^+ & -k_P^- - k_O & 0 \\
-                        0 & 0 & k_O & 0
-                \end{pmatrix},
-\end{equation}
-with the four promoter states, in order, being repressor bound,
-empty, RNAP closed complex, and RNAP open complex.
-Besides increasing dimension by one again, the only new feature in
-$\matr{K}$ is the rate $k_O$, representing the (irreversible)
-rate of open complex formation from the closed complex.
-The initiation matrices are given by
-\begin{equation}
-\matr{R_A} = \begin{pmatrix}
-        0 & 0 & 0 & 0 \\ 0 & 0 & 0 & r \\ 0 & 0 & 0 & 0\\ 0 & 0 & 0 & 0
-                \end{pmatrix},\
-\matr{R_D} = \begin{pmatrix}
-        0 & 0 & 0 & 0 \\ 0 & 0 & 0 & 0 \\ 0 & 0 & 0 & 0\\ 0 & 0 & 0 & r
-                \end{pmatrix},
-\end{equation}
-again closely analogous to model 2.
-
-The expression for mean mRNA is substantially more complicated
-now, as worked out in the appendix where we find
-\begin{equation}
-\langle m\rangle = \frac{r}{\gamma}
-        \frac{k_R^- k_P^+ k_O}
-        {k_R^- [(k_P^+ (k_O + r) + r(k_P^- + k_O)] + k_R^+ r(k_P^- + k_O)},
-\end{equation}
-which can be simplified to
-\begin{equation}
-\langle m\rangle
-= \frac{r}{\gamma}
-\frac{\frac{k_P^+ k_O}{r(k_O + k_P^-)}}
-        {1 + \frac{k_P^+ (k_O + r)}{r(k_O + k_P^-)} + \frac{k_R^+}{k_R^-}}.
-\end{equation}
-The strategy is to isolate the terms involving the repressor, so that now the fold-change is seen to be simply
-\begin{align}
-\text{fold-change}
-&= \frac{\frac{k_P^+ k_O}{r(k_O + k_P^-)}}
-        {1 + \frac{k_P^+ (k_O + r)}{r(k_O + k_P^-)} + \frac{k_R^+}{k_R^-}}
-        \frac{1 + \frac{k_P^+ (k_O + r)}{r(k_O + k_P^-)}}
-                {\frac{k_P^+ k_O}{r(k_O + k_P^-)}}
-\\
-&= \left(
-        1 + \frac{k_R^+}{k_R^-}
-        \left(1 + \frac{k_P^+ (k_O + r)}{r(k_O + k_P^-)}\right)^{-1}
-\right)^{-1},
-\end{align}
-surprisingly reducing to the master curve once again, with
-$\rho = 1 + \frac{k_P^+ (k_O + r)}{r(k_O + k_P^-)}$.
-
-This example hints that an arbitrarily fine-grained model of
-downstream transcription steps may still be collapsed to the form
-of the master curve in~\fig{fig1:means_cartoons}B so long as the
-repressor binding is exclusive with transcriptionally active
-states. We offer this as a conjecture, and we suspect that a
-careful argument using the King-Altman diagram
-method~\cite{King1956, Hill1966} might furnish a ``proof.''
-Our focus here is not on full generality but rather to survey an
-assortment of plausible models that have been proposed in the
-literature for simple repression.
-
-\subsubsection{Nonequilibrium model 4 - ``Active'' and ``inactive'' states}
-This model is at the core of the theory in~\cite{Razo-Mejia2020}.
-At a glance the cartoon for this model may appear very similar to model 2,
-and mathematically it is, but the interpretation is rather different.
-In model 2, we interpreted the third state literally as an
-RNAP-bound promoter and modeled initiation of a transcript as
-triggering a promoter state change, making the hopefully
-uncontroversial assumption that an RNAP can only make one
-transcript at a time. In contrast, in the present model the
-promoter state does \textit{not} change when a transcript is
-initiated. So we no longer interpret these states as literally
-RNAP bound and unbound but instead as coarse-grained ``active''
-and ``inactive'' states, the details of which we leave
-unspecified for now. We will comment more on this model below
-when we discuss Fano factors of models.
-
-Mathematically this model is very similar to models 1 and 2.
-Like model 1, the matrix $R$ describing transcription initiation
-is diagonal, namely
-\begin{equation}
-\matr{R} = \begin{pmatrix}
-                0 & 0 & 0 \\ 0 & 0 & 0 \\ 0 & 0 & r
-        \end{pmatrix}.
-\end{equation}
-so the master equation takes verbatim the same form as it did for model
-1,~\eq{eq:2state_rep_cme}. Meanwhile the promoter transition matrix
-$K$ is the same as~\eq{eq:3state_cme_matrices_pt1} from model 2,
-although we relabel the rate constants from $k_P^\pm$ to $k^\pm$ to
-reiterate that these are not simply RNAP binding and unbinding rates.
-
-Carrying out the algebra, the mean mRNA can be found to be
-\begin{equation}
-\langle m\rangle = \frac{r}{\gamma}
-\frac{k_R^- k^+}
-{k_R^- k^+ + k_R^- k^- + k_R^+ k^-},
-\end{equation}
-and the fold-change readily follows,
-\begin{align}
-\text{fold-change}
-&=      \frac{k_R^- k^+}{k_R^- k^+ + k_R^- k^- + k_R^+ k^-}
-        \frac{k_R^- k^+ + k_R^- k^-}{k_R^- k^+}
-\\
-&=      \left(1 + \frac{k_R^+}{k_R^-}
-                \left(1 + \frac{k^+}{k^-}\right)^{-1}
-        \right)^{-1},
-\end{align}
-from which we see $\rho = 1 + k^+/k^-$ as shown in~\fig{fig1:means_cartoons}.
-
-\subsubsection{Nonequilibrium model 5 - Bursty promoter}
-We will have much more to say about this model in subsequent sections.
-For now we sketch it as an intuitive analog to model 1,
-with just two states, repressor bound or unbound, and transition rates
-between them of $k_R^+$ and $k_R^-$.
-In model 1, when in the unbound state, single mRNA transcripts
-are produced as a Poisson process with some characteristic rate
-$r$. This model by contrast produces, at some Poisson rate $k_i$,
-\textit{bursts} of transcripts with a geometrically distributed
-burst size and a mean burst size $b$.
-
-From this intuitive picture and by analogy to model 1, then, it should be plausible that the mean mRNA level is
-\begin{equation}
-\langle m\rangle = \frac{k_i b}{\gamma}
-        \left(1 + \frac{k_R^+}{k_R^-}\right)^{-1},
-\end{equation}
-which will turn out to be correct from a careful calculation.
-For now, we simply note that just like model 1, the fold-change becomes
-\begin{equation}
-\text{fold-change} = \left(1 + \frac{k_R^+}{k_R^-}\right)^{-1}
-\end{equation}
-with $\rho=1$ also like model 1.
-
-We will also see later how this model emerges as a natural limit of model 4.
-
-\subsubsection{Summary and discussion of results across models for fold-changes in mean expression}
-What have we learned from our exploration of the models
-in~\fig{fig1:means_cartoons} and their predictions of
-fold-changes in mean expression levels? The takeaway is the
-existence of a master curve (\fig{fig1:means_cartoons}C) to which
-the fold-change predictions of all the models collapse. This
-master curve is parametrized by only two effective parameters:
-$\Delta F_R$, which characterizes the repressor, and $\rho$,
-which characterizes all other features of the promoter
-architecture. The key assumption underpinning this result is that
-zero transcription occurs when a repressor is bound to its
-operator. Note, however, that we are agnostic about the molecular
-mechanism which achieves this;
-steric effects are the obvious mechanism, but, for instance,
-``action-at-a-distance'' mediated by kinked DNA due to
-repressors bound tens or hundreds of nucleotides upstream of a
-promoter is plausible as well.
-
-Why does the master curve of~\fig{fig1:means_cartoons}C exist at all?
-This brings to mind the deep questions posed in,
-e.g.,~\cite{Frank2013} and~\cite{Frank2014a}, asking us to search
-for general ``symmetries'' of the problem instead of specifying
-generative models of the known microscopics of a system.
-In our case, the key feature seems to be the exclusive nature of
-repressor and RNAP binding, which allows the parameter describing
-the repressor, $\Delta F_R$, to cleanly separate from all other
-details of the promoter architecture, which are encapsulated in $\rho$.
-In the graph-theoretic language which was applied to similar
-problems in~\cite{Gunawardena2012} and~\cite{Ahsendorf2014}, the
-repressor-bound state is a leaf of the graph describing
-transitions between promoter states. Arbitrary nonequilibrium
-behavior can occur on the rest of the promoter state space, but
-it may all be swept up in the effective parameter $\rho$, to
-which the repressor makes no contribution.
-
-As suggested in~\cite{Chure2019}, we believe this master curve
-should generalize to architectures with multiple repressor
-binding sites, as long as the exclusivity of transcription factor
-binding and transcription initation is maintained. The
-interpretation of $\Delta F_R$ is then of an effective free
-energy of all repressor bound states.
-In an equilibrium picture this is simply given by the log of the
-sum of Boltzmann weights of all repressor bound states, which
-looks like the log of a partition function of a subsystem.
-In a nonequilibrium picture, while we can still mathematically gather
-terms and give the resulting collection the label $\Delta F_R$,
-it is unclear if the physical interpretation as an effective free energy
-makes sense.
-As is well known, free energies cannot be assigned unambiguously
-to states out of equilibrium because the free energy change along
-a generic path traversing the state space is path dependent,
-unlike at equilibrium.
-A consequence of this is that, out of equilibrium, $\Delta F_R$
-is no longer a simple sum of Boltzmann weights. Instead it
-resembles a restricted sum of King-Altman
-diagrams~\cite{King1956, Hill1966}. Following the work of
-Hill~\cite{Hill1989}, it may yet be possible to interpret this expression
-as an effective free energy, but this remains unclear to us.
-We leave this an open problem for future work.
-
-If we relax the requirement of exclusive repressor-RNAP binding,
-one could imagine models in which repressor and RNAP doubly-bound
-states are allowed, where the repressor's effect is to
-\text{reduce} the transcription rate rather than setting it to zero.
-Our results do not strictly apply to such a model, although we
-note that if the repressor's reduction of the transcription rate
-is substantial, such a model might still be well-approximated by
-one of the models in~\fig{fig1:means_cartoons}.
-
-One may worry that our ``one curve to rule them all'' is a
-mathematical tautology. In fact we \textit{agree} with this
-criticism if $\Delta F_R$ is ``just a fitting parameter'' and cannot
-be meaningfully interpreted as a real, physical free energy.
-An analogy to Hill functions is appropriate here. One of their
-great strengths and weaknesses, depending on the use they are put
-to, is that their parameters coarse-grain many details and are
-generally not interpretable in terms of microscopic models, for
-deep reasons discussed at length in~\cite{Frank2013}.
-By contrast, our master curve claims to have the best of both worlds:
-a coarse-graining of all details besides the repressor into
-a single effective parameter $\rho$,
-while simultaneously retaining an interpretation of
-$\Delta F_R$ as a physically
-meaningful and interpretable free energy.
-Our task, then, is to prove or disprove this claim.
-
-How do we test this and probe the theory with fold-change
-measurements? There is a fundamental limitation in that the
-master curve is essentially a one-parameter function of $\Delta
-F_R + \log\rho$. Worse, there are many \textit{a priori}
-plausible microscopic mechanisms that could contribute to the
-value of $\rho$, such as RNAP binding and escape
-kinetics~\cite{DeHaseth1998, Mitarai2015}, and/or supercoiling
-accumulation and release~\cite{Chong2014, Sevier2016}, and/or,
-RNAP clusters analogous to those obvserved in
-eukaryotes~\cite{Cisse2013, Cho2016}
-and recently also observed in bacteria~\cite{Ladouceur2020}.
-Even if $\Delta F_R$ is measured to
-high precision, inferring the potential microscopic contributions
-to $\rho$, buried inside a log no less, from fold-change
-measurements seems hopeless. As a statistical inference problem
-it is entirely nonidentifiable, in the language
-of~\cite{Gelman2013}, Section 4.3.
-
-If we cannot simply infer values of $\rho$ from measurements of
-fold-change, can we perturb some of the parameters that make up
-$\rho$ and measure the change? Unfortunately we suspect this is
-off-limits experimentally: most of the potential contributors to
-$\rho$ are global processes that affect many or all genes. For
-instance, changing RNAP association rates by changing RNAP copy
-numbers, or changing supercoiling kinetics by changing
-topoisomerase copy numbers, would massively perturb the entire
-cell's physiology and confound any determination of $\rho$.
-
-One might instead imagine a bottom-up modeling approach, where we
-mathematize a model of what we hypothesize the important steps
-are and are not, use in vitro data for the steps deemed
-important, and \textit{predict} what $\rho$ should be.
-But again, because of the one-parameter nature of the master
-curve, many different models will likely make indistinguishable
-predictions, and without any way to experimentally perturb in
-vivo, there is no clear way to test whether the modeling
-assumptions are correct.
-
-In light of this, we prefer the view that parameters and rates
-are not directly comparable between cartoons
-in~\fig{fig1:means_cartoons}. Rather, parameters in the
-simpler cartoons represent coarse-grained combinations of
-parameters in the finer-grained models. For instance, by equating
-$\rho$ between any two models, one can derive various possible
-correspondences between the two models' parameters. Note that
-these correspondences are clearly not unique, since many possible
-associations could be made. It then is a choice as to what microscopic
-interpretations the model-builder prefers for the parameters in a
-particular cartoon, and as to which coarse-grainings lend
-intuition and which seem nonsensical.
-Indeed, since it remains an open question what microscopic
-features dominate $\rho$ (as suggested above, perhaps RNAP
-binding and escape kinetics~\cite{DeHaseth1998, Mitarai2015}, or
-supercoiling accumulation and release~\cite{Chong2014, Sevier2016},
-or, something more exotic like
-RNAP clusters~\cite{Cisse2013, Cho2016, Ladouceur2020}),
-we are hesitant to put too much weight on any one microscopic
-interpretation of model parameters that make up $\rho$.
-
-One possible tuning knob to probe $\rho$ that would not globally
-perturb the cell's physiology is to manipulate RNAP binding sites.
-Work such as~\cite{Kinney2010} has shown that models of
-sequence-dependent RNAP can be inferred from data, and the
-authors of~\cite{Brewster2012} showed that the model
-of~\cite{Kinney2010} has predictive power by using the model to
-\textit{design} binding sites of a desired affinity.
-But for our purposes, this begs the question: the authors
-of~\cite{Kinney2010} \textit{assumed} a particular model
-(essentially our 3-state equilibrium model but without the
-repressor), so it is unclear how or if such methods can be turned
-around to \textit{compare} different models of promoter function.
-
-We have already pointed out that the master curve
-of~\fig{fig1:means_cartoons} is essentially a one-parameter model,
-the one parameter being $\Delta F_R + \log\rho$.
-By now the reader may be alarmed as to how can we even determine
-$\Delta F_R$ and $\rho$ independently of each other, never mind
-shedding a lens on the internal structure of $\rho$ itself.
-A hint is provided by the weak promoter approximation, invoked
-repeatedly in prior studies~\cite{Bintu2005c, Garcia2011a,
-Razo-Mejia2018} of simple repression using the 3-state
-equilibrium model in~\fig{fig1:means_cartoons}.
-In that picture, the weak promoter approximation means
-$\frac{P}{N_{NS}}\exp(-\beta\Delta\epsilon_P) \ll 1$,
-meaning therefore $\rho\approx1$.
-This is suggestive, but how can we be sure that $\rho$ is not,
-for instance, actually $10^2$ and that $\Delta F_R$ hides a
-compensatory factor?
-A resolution is offered by an independent inference of $\rho$ in
-the absence of repressors. This was done in~\cite{Razo-Mejia2020}
-by fitting nonequilibrium model 4  in~\fig{fig1:means_cartoons},
-with zero repressor (looking ahead, this is equivalent to model 4
-in~\fig{fig:constit_cartoons}), to FISH data
-from~\cite{Brewster2014}. This provided a determination of $k^+$
-and $k^-$, from which their ratio is estimated to be no more than
-a few $10^{-1}$ and possibly as small as $10^{-2}$.
-
-The realization that $\rho\approx1$ to an excellent
-approximation, \textit{independent} of which model
-in~\fig{fig1:means_cartoons} one prefers, goes a long way towards
-explaining the surprising success of equilibrium models of simple
-repression. Even though our 2- and 3-state models get so many
-details of transcription wrong, it does not matter because
-fold-change is a cleverly designed ratio. Since $\rho$ subsumes
-all details except the repressor, and $\log\rho\approx0$, fitting
-these simple models to fold-change measurements can still give a
-surprisingly good estimate of repressor binding energies. So the
-ratiometric construction of fold-change fulfills its intended
-purpose of canceling out all features of the promoter
-architecture except the repressor itself. Nevertheless it is
-perhaps surprising how effectively it does so: \textit{a priori},
-one might not have expected $\rho$ to be quite so close to 1.
-
-We would also like to highlight the relevance
-of~\cite{Landman2019} here. Landman et.\ al.\ reanalyzed and
-compared in vivo and in vitro data on the lacI repressor's
-binding affinity to its various operator sequences. (The in vivo
-data was from, essentially, fitting our master curve to
-expression measurements.) They find broad agreement between the
-in vitro and in vivo values. This reinforces the suspicion that
-the equilibrium $\Delta\epsilon_R$ repressor binding energies do
-in fact represent real physical free energies.
-Again, \textit{a priori} this did not have to be the case, even
-knowing that $\rho\approx1$.
-
-In principle, if $\Delta F_R$ can be measured to sufficient
-precision, then deviations from $\rho=1$ become a
-testable matter of experiment.
-In practice, it is probably unrealistic to measure repressor
-rates $k_R^+$ or $k_R^-$ or fold-changes in expression levels
-(and hence $\Delta\epsilon_R$) precisely enough to detect the
-expected tiny deviations from $\rho=1$.
-We can estimate the
-requisite precision in $\Delta F_R$ to resolve a given
-$\Delta\rho$ by noting, since $\rho\approx1$, that
-$\log(1+\Delta\rho)\approx \Delta\rho$, so
-$\Delta(\Delta F_R) \approx \Delta\rho$.
-Suppose we are lucky and $\Delta\rho$ happens to be $\sim0.1$, on
-the high end of our range estimated above.
-A determination of $\Delta\epsilon_R/k_BT$ with an uncertainty of
-barely 0.1 was achieved in the excellent measurements
-of~\cite{Razo-Mejia2018}, so this requires a very difficult
-determination of $\Delta F_R$ for a very crude determination of $\rho$,
-which suggests, to put it lightly, this is not a
-promising path to pursue experimentally.
-It is doubtful that inference of repressor kinetic rates
-would be any easier.\footnote{
-\mmnote{
-In fact the uncertainties quoted in Hammar et.\
-al.~\cite{Hammar2014} from Elf's group are small enough to do
-this calculation, but HJ \& I think their error bars are insanely
-optimistic. Not sure I want to blatantly call them out on it though??}
-}
-
-Moving forward, we have weak evidence supporting the
-interpretation of $\Delta F_R$ as a physically real free
-energy~\cite{Landman2019} and other work casting
-doubt~\cite{Hammar2014}.
-How might we resolve the confusion?
-If there is no discriminatory power to test the theory and
-distinguish the various models with measurements of fold-changes
-in means, how do we probe the theory? Clearly to discriminate
-between the nonequilibrium models in~\fig{fig1:means_cartoons},
-we need to go beyond means to ask questions about kinetics, noise and even
-full distributions of mRNA copy numbers over a population of cells.
-If the ``one-curve-to-rule-them-all'' is more than a
-mathematical tautology, then the free energy of repressor binding
-inferred from fold-change measurements should agree with
-repressor binding and unbinding rates.
-In other words, the equilbrium and nonequilibrium definitions of
-$\Delta F_R$ must agree, meaning
-\begin{equation}
-\Delta F_R = \beta\Delta\epsilon_R - \log(R/N_{NS})
-        = - \log(k_R^+/k_R^-),
-\end{equation}
-must hold, where $\beta\Delta\epsilon_R$ is inferred from the
-master curve fit to fold-change measurements, and $k_R^+$ and
-$k_R^-$ are inferred in some orthogonal manner.
-Single molecule measurements such as from~\cite{Hammar2014} have
-directly observed these rates, and in the remainder of this work
-we explore a complementary approach: inferring repressor binding
-and unbinding rates $k_R^+$ and $k_R^-$ from FISH measurements of
-mRNA population distributions.
-
-\section{Beyond means}
-\subsection{A one-state promoter with bursting is the ``best'' model of constitutive promoters}
-Before we can tackle simple repression, we need an adequate
-phenomenological model of constitutive expression.
-The literature abounds with options from which we can choose,
-and we show several potential kinetic models for constitutive promoters
-in~\fig{fig:constit_cartoons}.
-Let us consider the suitability of each model for our purposes in turn.
-
-\begin{figure}%[h!]
-\centering
-\includegraphics[width=0.9\textwidth]{../figures/fig1.5/fig1point5.ai}
-\caption{\textbf{Constitutive promoter cartoon comparison.}
-The left column depicts various plausible cartoons for constitutive promoters.
-We also list literature references where models have been used previously.
-In model (1), transcripts are produced in a Poisson process~\cite{Jones2014}.
-Model (2) features explicit modeling of RNAP binding/unbinding
-kinetics~\cite{Phillips2015a}.
-Model (3) is a more detailed generalization of model (2),
-treating transcription initiation as a multi-step process
-proceeding through closed and open complexes~\cite{Mitarai2015}.
-Model (4) is somewhat analogous to (2) except with the precise nature
-of active and inactive states left ambiguous~\cite{Razo-Mejia2020}.
-Finally, model (5) can be viewed as a certain limit of model (4)
-in which transcripts are produced in bursts,
-and initiation of bursts is a Poisson process.
-\mmnote{Haven't found model 5 formulated quite like this in the literature,
-usually burstiness is handled like model 4, but I should probably search a bit more.}
-The right column shows the Fano factor for each model.
-Note especially the crucial diagnostic: (2) and (3) have $\nu$
-strictly below 1, while only for (4) and (5) can $\nu$ exceed 1.
-Models with Fano factors $\le 1$ cannot produce the FISH data
-observed in~\cite{Jones2014} without introducing additional
-assumptions and model complexity.
-}
-\label{fig:constit_cartoons}
-\end{figure}
-
-\subsubsection{Model 1 - Poisson promoter.}
-This is the picture from Jones et.\ al.~\cite{Jones2014}, which
-assumes that transcripts are produced as a Poisson process from a
-single promoter state. This model insists that the ``true'' mRNA
-distribution is Poisson, implying the Fano factor $\nu$ must be 1.
-In~\cite{Jones2014}, the authors carefully attribute deviations
-from Fano = 1 to intensity variability in fluorescent spots, gene
-copy number variation, and copy number fluctuations of the
-transcription machinery, e.g., RNAP itself.
-In this picture, the master equation makes no appearance,
-and all the corrections to Poisson behavior are derived
-as additive corrections to the Fano factor.
-For disproving the ``universal noise curve'' from So et
-al~\cite{So2011}, this picture was excellent. It is appealing in
-its simplicity, with only two parameters, the initiation rate $r$
-and degradation rate $\gamma$. Since $\gamma$ is independently
-known from other experiments, and the mean mRNA copy number is
-$r/\gamma$, $r$ is easily inferred from data.
-In other words, the model is not excessively complex for the data at hand.
-But for many interesting questions, for instance in the recent
-work~\mmnote{cite Manuel's preprint once posted},
-knowledge of means and variances alone is insufficient, and a
-description of the full distribution of molecular counts is necessary.
-For this we need a( slightly) more complex model than model 1
-that would allow us to incorporate the non-Poissonian features of
-constitutive promoters directly into a master equation formulation.
-
-\subsubsection{Model 2 - Two-state promoter, RNAP bound or unbound.}
-This model was considered in, e.g.,~\cite{Phillips2015a} and ~\cite{Phillips2019}.
-Here transcription initiation proceeds from the bound to the unbound state,
-reflecting the microscopic reality that an RNAP that has begun to elongate
-a transcript is no longer available at the start site to begin another.
-The problem with this picture is that the Fano factor is
-\begin{align}
-    \nu = 1 -
-        \frac{r\rate{k}{P}{+}}
-            {\left(\rate{k}{P}{+} + \rate{k}{P}{-} + r\right)
-             \left(\gamma + \rate{k}{P}{+} + \rate{k}{P}{-} + r\right)},
-\end{align}
-which is always $<1$.
-To make contact with the experimental reality of $\nu>1$,
-we will have to do, at a minimum,
-the same corrections to Poisson behavior as in model 1 above.
-So while this model adds an appealing element of microscopic
-reality, we are forced to reject it as the additional complexity
-is unable to capture the phenomenology of interest.
-Obviously the promoter state does in fact proceed through cycles
-of RNAP binding, initiating, and elongating, but it seems that
-the super-Poissonian noise in mRNA copy number we want to
-model must be governed by other features of the system.
-
-\subsubsection{Model 3 - Three-state promoter, multistep
-transcription initiation and escape.}
-How might we remedy the deficits of model 2?
-It is known~\cite{DeHaseth1998} that once RNAP initially binds
-the promoter region, a multitude of distinct steps occur
-sequentially before RNAP finally escapes into the elongation
-phase. Perhaps adding some of this mechanistic detail might
-rescue model 2? The next simplest refinement of model 2 could
-consider open complex formation and promoter escape as separate
-steps rather than as a single effective step. In other words, we
-construct model 3 by adding a single extra state to model 2, and
-we will label the two RNAP-bound states as the closed and open
-complexes, despite the true biochemical details certainly being
-more complex. The authors of~\cite{Mitarai2015}, for instance,
-considered this model, although they added an additional
-repressor bound state and did not explicitly consider the limit
-with no repressor that we analyze here. Again, our goal here is
-not a complete accounting of all the relevant biochemical detail;
-this is an exploratory search for the important features our
-model needs to include.
-
-Unfortunately, as the authors of~\cite{Mitarai2015} hint, this
-model too has Fano factor $\nu<1$. This can be seen by starting
-with their Eq.~A1 for the Fano factor of the analogous model with
-repressor and taking the limit as repressor concentration goes to
-zero. After substantial algebra, the result is
-\begin{align}
-\nu = 1 - \frac{r k_O k_P^+}{\mathcal{Z}}
-\frac{k_P^+ + k_P^- + k_O + \gamma}
-        {\mathcal{Z} + \gamma(k_P^+ + k_P^- + k_O + \gamma) + \gamma^2},
-\end{align}
-where we defined $\mathcal{Z} = r(k_O + k_P^-) + k_P^+(k_O + r)$
-for notational tidiness. This is necessarily less than 1 for
-arbitrary rate constants.
-
-In fact, we suspect \textit{any} model in which transcription
-proceeds through a multistep cycle must necessarily have $\nu<1$.
-The intuitive argument compares the waiting time distribution to
-traverse the cycle with the waiting time for a Poisson promoter
-(model 1) with the same mean time. The latter is simply an
-exponential distribution. The former is a convolution of multiple
-exponentials, and intuitively should be more peaked with a
-smaller fractional width than an exponential with the same mean.\footnote{
-This can be made more precise with a result
-from~\cite{Stewart2007}, who showed that the convolution of
-multiple gamma distributions (of which the exponential
-distribution is a special case) is to a very good approximation
-also gamma distributed. Using their Eq.~2 for the distribution of
-the convolution, with shape parameters set to 1 to
-give exponential distributions, the total waiting time
-distribution has a ratio of variance to squared mean 
-$\sigma^2/\mu^2 = \sum_i k_i^2/\left(\sum_i k_i\right)^2$,
-where the $k_i$ are the rates of the individual steps. Clearly
-this is less than 1 and therefore the total waiting time
-distribution is narrower than the corresponding exponential.} A
-less disperse waiting time distribution means transcription
-initations are more uniformly distributed in time relative to a
-Poisson process. Hence the distribution of mRNA over a population
-of cells should be less variable compared to Poisson, giving
-$\nu<1$.\footnote{
-This last step, while intuitive, can be argued by analogy to
-photon statistics where antibunching gives rise to sub-Poissonian
-noise~\cite{Paul1982, Zou1990}. Although loopholes exist, we do
-not expect they apply for our problem. Nevertheless we refrain
-from elevating this cycle/sub-Poissonian equivalence to a ``theorem.''
-\mmnote{Antibunching is the ``obvious'' analogy for me but it'd
-be totally out of left field for most readers, any suggestions on
-a more comprehensible reference??}
-}
-
-Regardless of the merits of this model in describing the noise
-properties of constitutive transcription initation, it does not
-capture the dominant source of noise we want to include in our
-phenomenological model, so we must discard these details and
-search elsewhere.
-
-\subsubsection{Model 4 - Two-state promoter,
-``active'' and ``inactive'' states.}
-Inspired by~\cite{Razo-Mejia2020}, we revisit an analog of model
-2, but the interpretation of the two states is changed. Rather
-than explicitly viewing them as RNAP bound and unbound, we view
-them as ``active'' and ``inactive,'' which are able and unable to
-initiate transcripts, respectively. We are noncommittal as to the
-microscopic details of these states.
-
-One interpretation~\cite{Chong2014, Sevier2016} for the active
-and inactive states is that it represents the promoter's
-supercoiling state: transitions to the inactive state are caused
-by accumulation of positive supercoiling, which inhibits
-transcription, and transitions back to ``active'' are caused by
-gyrase or other topoisomerases relieving the supercoiling. This
-is an interesting possibility because it would mean the timescale
-for promoter state transitions is driven by topoisomerase
-kinetics, not by RNAP kinetics. From in vitro measurments, the
-former are suggested to be of order minutes~\cite{Chong2014}.
-Contrast this with model 2, where the state transitions are
-assumed to be governed by RNAP, which, assuming a copy number per
-cell of order $10^3$, has a diffusion-limited association rate
-$k_{on} \sim 10^2~\text{s}^{-1}$ to a target promoter. Combined
-with known $K_D$'s of order $\mu$M, this gives an RNAP
-dissociation rate $k_{off}$ of order $10^2$. As we will show
-below, however, there are some lingering puzzles with
-interpreting this supercoiling interpretation, so we leave it as
-a speculation and refrain from assigning definite physical
-meaning to the two states in this model.
-
-Intuitively one might expect that, since transcripts are produced
-as a Poisson process only when the promoter is in one of the two
-states in this model, transcription initiations should now be
-``bunched,'' in constrast to the ``anti-bunching'' of models 2
-and 3 above. One might further guess that this bunching would
-lead to super-Poissonian noise in the mRNA distribution over a
-population of cells. An honest calculation of the Fano factor produces
-\begin{align}
-\nu &= 1 + \frac{r k^-}{(k^+ + k^- + \gamma)(k^+ + k^-)},
-\end{align}
-which is strictly greater than 1, verifying the above intuition.
-Note we have dropped the $P$ label on the promoter switching rates to
-emphasize that these very likely do not represent kinetics of RNAP itself.
-This calculation can also be sidestepped by noting that the model
-is mathematically equivalent to the simple repression repression model
-from~\cite{Jones2014}, with states and rates relabeled and reinterpreted.
-
-How does this model compare to model 1 above?
-In model 1, all non-Poisson features of the mRNA
-distribution were handled as extrinsic corrections.
-By contrast, here the 3 parameter model is used to fit the full
-mRNA distribution as measured in mRNA FISH experiments.
-In essense, all variability in the mRNA distribution is regarded
-as ``intrinsic,'' arising either from stochastic initiation or
-from switching between the two coarse-grained promoter states.
-The advantage of this approach is that it fits neatly into the
-master equation picture, and the parameters thus inferred can be used as
-input for more complicated models with regulation by transcription factors.
-
-While this seems promising, there is a major drawback for our purposes
-which was already uncovered by the authors of~\cite{Razo-Mejia2020}:
-the statistical inference problem is nonidentifiable,
-in the sense described in Section 4.3 of~\cite{Gelman2013},
-meaning it is impossible to infer the
-parameters $r$ and $k^-$ from the FISH data of~\cite{Jones2014}
-(as shown in Fig.~S2 of~\cite{Razo-Mejia2020}).
-Rather, only the ratio $r/k^-$ could be inferred. In that work,
-the problem was worked around with an informative prior on the
-ratio $k^-/k^+$. That approach is unlikely to work here, as,
-recall, our entire goal in modeling constitutive expression is to
-use it as the basis for a yet more complicated model, when we add
-on repression. But adding more complexity to a model that is
-already poorly identified is a fool's errand,
-so we will explore one more potential model.
-
-\subsubsection{Model 5 - One-state promoter with explicit bursts.}
-This model is inspired by the failure mode of model 4.
-The key observation above was that, as found
-in~\cite{Razo-Mejia2020}, only two parameters, $k^+$ and the
-ratio $r/k^-$, could be directly inferred from the FISH data
-from~\cite{Jones2014}.
-So let us take this seriously and imagine a model where these are
-the only two model parameters. What would this model look like?
-
-To develop some intuition, consider model 4 in the limit
-$k^+ \ll k^- \lesssim r$, which is roughly satisfied by the
-parameters inferred in~\cite{Razo-Mejia2020}.
-In this limit, the system spends the majority of its time in the inactive state, occasionally becoming active and making a burst of transcripts.
-This should call to mind the well-known phenomenon of transcriptional bursting, as reported in, e.g.,~\cite{Golding2005,Chong2014,Sevier2016}\mmnote{should probably add some more cites here}.
-Let us make this correspondence more precise.
-The mean dwell time in the active state is $1/k^-$.
-While in this state, transcripts are produced at a rate $r$ per
-unit time. So on average, $r/k^-$ transcripts are produced before
-the system switches to the inactive state. Once in the inactive
-state, the system dwells there for an average time $1/k^+$ before
-returning to the active state and repeating the process. $r/k^-$
-resembles an average burst size, and $1/k^+$ resembles the time
-interval between burst events. More precisely, $1/k^+$ is the
-mean time between the end of one burst and the start of the next,
-whereas $1/k^+ + 1/k^-$ would be the mean interval between the
-start of two successive burst events, but in the limit $k^+ \ll
-k^-$, $1/k^+ + 1/k^- \approx 1/k^+$. Note that this limit ensures
-that the waiting time between bursts is approximately
-exponentially distributed, with mean set by the only timescale
-left in the problem,
-$1/k^+$.\footnote{
-If instead it were the case that $k^+ \sim k^-$, then the waiting time
-$1/k^+ + 1/k^-$ would have a peaked distribution, but this does
-not appear to be the case for any of datasets
-from~\cite{Jones2014}.}
-
-Let us now verify this intuition with a precise derivation
-to check that $r/k^-$ is in fact the mean burst size
-and to obtain the full burst size distribution.
-Consider first a constant, known dwell time $T$ in the active state.
-Transcripts are produced at a rate $r$ per unit time, so the
-number of transcripts $n$ produced during $T$ fits exactly
-the ``story'' for the Poisson distribution, i.e.,
-\begin{equation}
-    P(n\mid T) = \frac{(rT)^n}{n!} \exp(-rT).
-\end{equation}
-Since the dwell time $T$ is unobservable, we actually want $P(n)$,
-the dwell time distribution with no conditioning on $T$.
-Basic rules of probability theory tell us we can write $P(n)$
-in terms of $P(n\mid T)$ as
-\begin{equation}
-    P(n) =\int_0^\infty P(n\mid T) P(T) dT.
-\end{equation}
-But we know the dwell time distribution $P(T)$,
-which is exponentially distributed according to
-\begin{equation}
-    P(T) = k^- \exp(-T k^-),
-\end{equation}
-so $P(n)$ can be written as
-\begin{equation}
-    P(n) = k^- \frac{r^n}{n!}
-            \int_0^\infty T^n\exp[-(r + k^-)T]\,dT.
-\end{equation}
-A standard integral table shows
-$\int_0^\infty x^n e^{-ax}\,dx = n!/a^{n+1}$, so
-\begin{equation}
-    P(n) = \frac{k^- r^n}{(k^- + r)^{n+1}}
-        = \frac{k^-}{k^- + r}
-            \left(\frac{r}{k^- + r}\right)^n
-        = \frac{k^-}{k^- + r}
-            \left(1 - \frac{k^-}{k^- + r}\right)^n,
-\end{equation}
-which is exactly the geometric distribution with standard parameter
-$\theta\equiv k^-/(k^- + r)$ and domain
-$n \in \{0, 1, 2, \dots\}$
-(this is one of two common conventions for the geometric distribution).
-The mean of the geometric distribution, with this convention, is
-\begin{align}
-\langle n\rangle = \frac{1 - \theta}{\theta}
-        = \left(1 - \frac{k^-}{k^- + r}\right)
-                    \frac{k^- + r}{k^-}
-        = \frac{r}{k^-},
-\end{align}
-exactly as we guessed intuitively above.
-
-So in taking the limit $r,k^-\rightarrow\infty$, $r/k^-\equiv b$,
-we obtain a model which effectively has only a single promoter state,
-which initiates bursts at rate $k^+$
-(transitions to the active state, in the model 4 picture).
-The master equation for mRNA copy number $m$ takes the form
-\begin{align}
-\begin{split}
-\deriv{t}{p(m,t)} = & (m+1)\gamma p(m+1,t) - m\gamma p(m,t) \\
-        &+ \sum_{j=0}^{m-1} k^+ p(j,t) Geom(m-j;b)
-         - \sum_{j=m+1}^\infty k^+ p(m,t) Geom(j-m;b),
-\end{split}
-\end{align}
-where $Geom(n;b)$ is the geometric distribution with mean~$b$, i.e.,
-$Geom(n;b) = \frac{1}{1+b}\left(\frac{b}{1+b}\right)^n$
-(with domain over nonnegative integers as above).
-The first two terms are the usual mRNA degradation terms.
-The third term enumerates all ways the system can produce
-a burst of transcripts and arrive at copy number $m$,
-given that it had copy number $j$ before the burst.
-The fourth term allows the system to start with copy number $m$,
-then produce a burst and end with copy number $j$.
-In fact this last sum has trivial $j$ dependence and
-simply enforces normalization of the geometric distribution.
-Carrying it out we have
-\begin{equation}
-\begin{split}
-\deriv{t}{p(m,t)} = & (m+1)\gamma p(m+1,t) - m\gamma p(m,t) \\
-        &+ \sum_{j=0}^{m-1} k^+ p(j,t) Geom(m-j;b)
-            - k^+ p(m,t),
-\end{split}
-\end{equation}
-This improves on model 4 in that now the parameters are easily inferred,
-as we will see later, and have clean interpretations.
-The non-Poissonian features are attributed to the emprically
-well-established phenomenological picture of bursty transcription.
-
-The big approximation in going from model 4 to 5 is that a burst
-is produced instantaneously rather than over a finite time.
-If the true burst duration is not short compared to transcription factor
-kinetic timescales, this could be a problem in that mean burst size
-in the presense and absence of repressors could change,
-rendering parameter inferences from the constitutive case inappropriate.
-Let us make some simple estimates of this.
-
-Consider the time delay between the first and final RNAPs
-in a burst initiating transcription
-(\textit{not} the time to complete transcripts,
-which potentially could be much longer.)
-If this timescale is short compared to the typical
-search timescale of transcription factors, then all is well.
-The estimates from deHaseth et.\ al.~\cite{DeHaseth1998}
-put RNAP's diffusion-limited on rate around
-$\sim\text{few}\times10^{-2}~\text{nM}^{-1}~\text{s}^{-1}$
-and polymerase loading as high as $1~\text{s}^{-1}$.
-Then for reasonable burst sizes of $<10$, it is reasonable to guess that
-bursts might finish initiating on a timescale of tens of seconds or less
-(with another 30-60 sec to finish elongation, but that does not matter here).
-A transcription factor with typical copy number of order 10 (or less)
-would have a diffusion-limited association rate of order
-$(10~\text{sec})^{-1}$.
-Higher copy number TFs tend to have many binding sites over the genome,
-which should serve to pull them out of circulation and keep their
-effective association rates from rising too large.
-Therefore, there is \textit{perhaps} a timescale separation
-possible between transcription factor association rates and burst
-durations, but this assumption could very well break down, so we
-will have to keep it in mind when we infer repressor rates from
-the Jones et.\ al.\ FISH data later.
-
-In reflecting on these 5 models, the reader may feel that
-exploring a multitude of potential models just to return to a
-very minimal phenomenological model of bursty transcription may
-seem highly pedantic. But the purpose of the exercise was to
-examine a host of models from the literature and understand why
-they are insufficient, one way or another, for our purposes.
-Along the way we have learned that the detailed kinetics of RNAP
-binding and initiating transcription are probably irrelevant for
-setting the population distribution of mRNA.
-The timescales are simply too fast, and as we will see later
-in~\fig{fig:constit_post}, the noise seems to be governed by
-slower timescales. Perhaps in hindsight this is not surprising:
-intuitively, the degradation rate $\gamma$ sets the fundamental
-timescale for mRNA dynamics, and any other processes that
-substantially modulate the mRNA distribution should not differ
-from $\gamma$ by orders of magnitude.
-
-\subsection{Parameter inference for constitutive promoters}
-Having established that a one-state bursty model achieves a
-Goldilocks level of complexity, we test it more thoroughly on all
-the data from constitutive promoters from~\cite{Jones2014}.
-\fig{fig:constit_post} shows the results.
-\begin{figure}%[h!]
-\centering
-\includegraphics[width=0.9\textwidth]{../figures/fig2/fig2pt1.pdf}
-\caption{\textbf{Constitutive promoter posterior
-        inference and predictive checks.}
-        (A) shows the joint posterior density from MCMC sampling of model 5,
-        bursty promoter with negative binomially-distributed steady state.
-        (B) shows the ECDF of the observed population distribution
-        of mRNA transcripts under the control of a constitutive lacUV5 promoter.
-        Posterior predictive percentiles are shown as colored bands for
-        model (1), Poisson, and model (5), negative binomial.
-        }
-\label{fig:uv5_post}
-\end{figure}
-
-\begin{figure}%[h!]
-\centering
-\includegraphics[width=0.9\textwidth]{../figures/fig2/fig2pt2.pdf}
-\caption{\textbf{Constitutive promoter model comparison.}
-    (A) Joint posterior distributions for burst rate $k_i$ and
-    mean burst size $b$ for 18 promoters from Jones 2014.
-    Each contour indicates the 95\% highest posterior density
-    region for a particular promoter, i.e., on a plot analogous
-    to~\fig{fig:uv5_post}, the contour would enclose
-    approximately 95\% of the samples, and thus 95\% of the
-    probability mass.
-    Note that the vertical axis is shared with (B).
-    Panel (B) plots the burst rate $k_i$ vs.\ binding energy from energy
-    matrices from Brewster 2012.
-    The dotted line shows the predicted slope according to
-    $\ln k_i/\gamma \sim \text{constant} - \beta\Delta\epsilon_P$,
-    described in text.
-    }
-\label{fig:constit_post}
-\end{figure}
-\mmnote{Discuss puzzling comparison w/ Chong2014: if supercoiling is the thing, why are my burst sizes all the same but burst rates vary?
-And why is the duty cycle of my promoter so low, ie., bursts so short?
-COmpare their fig 7E; their $\beta/\alpha$ is my $k^+/k^-$.
-They have one or two genes with very small $\beta/\alpha$, does the \textit{galK} locus just happen to be that, or is there a deeper disagreement?
-Hard to say w/o more data.}
-
-\subsection{Transcription factor kinetics can be inferred from FISH measurements}
-Figure~3 presents inferred transcription factor rates, compares
-them to single-molecule measurements, and compares their ratios
-to gold-standard binding energies.
-\begin{figure}%[h!]
-\centering
-\includegraphics[width=0.9\textwidth]{../figures/fig3/fig3.ai}
-\caption{\textbf{Simple repression model comparison.}
-(A) depicts the cartoon of simple repression for which we infer parameters.
-(B) shows 50\% and 95\% contours of several 2D slices of the 9D posterior
-distribution, from fitting one unbinding rate for each operator (O1, O2, O3)
-and one binding rate for each aTc concentration (corresponding to an
-unknown mean repressor copy number).
-(C) compares ratios of our inferred unbinding rates with operator
-binding energy differences measured in Garcia \& Phillips 2011 (triangles)
-and Razo-Mejia et.\ al.\ 2018 (squares).
-(D) compares our inferred unbinding rates with single-molecule measurements
-of the same from Hammar et.\ at.\ 2014 \mmnote{Johan Elf's group}.
-Values are of similar order of magnitude,
-but a precise quantitative comparison unfortunately depends sensitively
-on the assumed mRNA lifetime, which was not measured.
-    }
-\label{fig3:kR_inferences}
-\end{figure}
-
-\begin{figure}%[h!]
-\centering
-\includegraphics[width=0.9\textwidth]{../figures/figSIxx/ppc_many_pooled.pdf}
-\caption{\textbf{Posterior predictive checks.}
-(A) duplicates the posterior from~\fig{fig3:kR_inferences}B.
-(B), (C), and (D) show samples from the posterior predictive distribution
-for each of the datasets used in the inference.
-Samples are sorted by operator and plotted separately simply for visual clarity.
-The unrepressed promoter, UV5, is shown with each as a reference point.
-\mmnote{
-Obviously the PPC is not perfect,
-but considering the relative simplicity of the model, I'm pleasantly surprised
-how well it worked and that the model was even identifiable at all.
-Interestingly, and perhaps not surprisingly, only 3 of the 9 datasets
-were identifiable if fit in isolation: otherwise,
-it was only by making assumptions that certain rates are equal
-across datasets was identifiability achieved.
-}
-    }
-\label{fig4:ppc}
-\end{figure}
-
-\section{Discussion and future work}
-\mmnote{
-Do all my approximations work simply because galK happens to have
-a low duty cycle of supercoiling, $\beta/\alpha$ in Chong2014
-picture? A potentially interesting test: put in the dozen or so
-Hernan constructs at some other locus, preferably one of the ones
-Chong identified as having a large duty cycle (large $\beta/\alpha$),
-i.e., the opposite limit of galK. Then just measure fold-changes.
-You don't even need to do FISH, although that would be even
-better. If their model is right, $\rho$ should be approximately
-set by the duty cycle of supercoiling, NOT by RNAP kinetics.
-\textit{This is a NICE prediction for future work.} \textit{cspE}
-is claimed nonessential, prediction is you drop HG reporter
-constructs in there and the inferred binding energies of all lacI
-operators shift the same amount, by about $\ln(3)\sim 1kT$. Not
-huge, but detectable, although might be confounding effects from
-moving to a new locus. With FISH measurements, $k^+$ and $k^-$ in
-Fig 2, model 4 should now be separately inferrable if Chong2014
-measurements of $\beta/\alpha$ are correct. This would require an
-obnoxious amount of cloning, but once you did that, the
-measurements would be same old stuff that RP lab does.}
-
-%%%%%%%%%%%%%%%%%%%%% APPENDICES %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\appendix
-
-\section{Derivations of constitutive promoter models}
-\subsection{Model 2}
-\mmnote{Obtain Fano factor from CME.}
-\subsection{Model 3}
-\mmnote{Obtain Fano factor by taking limit from Mitarai et al's expression.}
-\subsection{Model 4}
-\mmnote{Obtain Fano factor by relabeling cartoon from Jones2014.}
-
-\section{Generating function solution for bursty promoters}
-
-\subsection{Constitutive promoter with bursts}
-
-\subsubsection{From master equation to generating function}
-
-We would like to derive the full probability distribution of
-mRNA copy number per cell for a model combining simple repression
-with a promoter featuring bursty transcription initiation.
-Before tackling mRNA and repression together, let us
-just consider the bursty consititutive promoter,
-model 5 in~\fig{fig:constit_cartoons}
-\mmnote{replicating Charlotte's notes; she says it's worked out
-in~\cite{Paulsson2000}, but I don't see it, at least not in notation I can
-comprehend. Need to find a cite for this,
-or maybe just cite Charlotte, personal comm, since that's where I actually got it\dots}.
-The master equation of interest is
-\begin{align}
-\deriv{t}p(m,t) = (m+1)\gamma p(m+1,t) - m\gamma p(m,t) - r p(m,t)
-        + r \sum_{m^\prime=0}^m G_{m-m^\prime}(\theta) p(m^\prime,t),
-\label{eq:1state_unreg_003}
-\end{align}
-where $G_{k}(\theta)$ is the geometric distribution defined as
-\begin{align}
-G_{k}(\theta) = \theta(1 - \theta)^k, \, k\in\{0,1,2,\dots\}.
-\end{align}
-With this convention, the mean burst size $b = (1-\theta)/\theta$.
-$\gamma$ and $r$ are mRNA degradation rates and transcription burst
-initiation rates, resp. The last term represents all ways the system could
-end up with $m$ mRNAs, having started the burst with $m^\prime$. Define
-$\lambda = r/\gamma$ and nondimensionalize time by $\gamma$, giving
-\begin{align}
-\deriv{t}p(m,t) = (m+1)p(m+1,t) - m p(m,t) - \lambda p(m,t)
-        + \lambda \sum_{m^\prime=0}^m G_{m-m^\prime}(\theta) p(m^\prime,t),
-\end{align}
-The probability generating function is defined as
-\begin{align}
-F(z,t) = \sum_{m=0}^\infty z^m p(m,t).
-\end{align}
-Multiply both sides of the CME by $z^m$ and sum over all $m$ to get
-\begin{align}
-\pderiv[F]{t} = \sum_{m=0}^\infty z^m
-\left(
-(m+1)p(m+1,t) - m p(m,t) - \lambda p(m,t)
-+ \lambda \sum_{m^\prime=0}^m G_{m-m^\prime}(\theta) p(m^\prime,t)
-\right).
-\end{align}
-The first two terms can be rewritten as derivatives of the
-generating function, i.e.,
-\begin{align}
-\sum_{m=0}^\infty z^m (m+1)p(m+1,t)
-=
-\pderiv{z} \sum_{m=0}^\infty z^{m+1} p(m+1,t)
-=
-\pderiv[F]{z}
-\end{align}
-and
-\begin{align}
-\sum_{m=0}^\infty z^m mp(m,t)
-=
-z\pderiv{z} \sum_{m=0}^\infty z^m p(m,t)
-=
-z\pderiv[F]{z}.
-\end{align}
-The final term is the most trouble. The trick is to reverse
-the default order of the sums, as in
-\begin{align}
-&\lambda \sum_{m=0}^\infty \sum_{m^\prime=0}^m
-        z^m \theta(1 - \theta)^{m-m^\prime} p(m^\prime,t)
-\\=
-&\lambda\theta \sum_{m^\prime=0}^\infty z^{m^\prime} p(m^\prime,t)
-        \sum_{m=m^\prime}^\infty
-        z^{m-m^\prime} (1 - \theta)^{m-m^\prime}
-\\=
-&\frac{\lambda\theta F}{1 - (1 - \theta)z}.
-\end{align}
-This works because after reordering the sums, the sum on $m$ is just
-a geometric series. Its result is independent of $m^\prime$, so the sum on
-$m^\prime$ reduces to the definition of the generating function.
-Putting these results together, the PDE for the generation function is
-\begin{align}
-\pderiv[F]{t} = \pderiv[F]{z} - z\pderiv[F]{z} - \lambda F
-        + \frac{\lambda\theta F}{1-z(1-\theta)}.
-\end{align}
-Changing variables to $\xi=1-\theta$ and simplifying gives
-\begin{align}
-\pderiv[F]{t} + (z - 1) \pderiv[F]{z} = \frac{(z-1)\xi}{1-z\xi}\lambda F.
-\label{eq:1state_unreg_015}
-\end{align}
-\subsubsection{Stead-state}
-At steady-state, the PDE reduces to the ODE
-\begin{align}
-\deriv[F]{z} = \frac{\xi}{1-z\xi}\lambda F,
-\end{align}
-which we can integrate as
-\begin{align}
-\int \frac{dF}{F} = \int \frac{\lambda\xi dz}{1-\xi z}.
-\end{align}
-The initial conditions for generating functions can be subtle and confusing.
-The key fact follows from the definition
-$F(z,t) = \sum_m z^m p(m,t)$,
-normalization requires that
-$F(z=1^-,t) = \sum_m p(m,t) = 1$.\footnote{
-Sometimes the generating function may be undefined \textit{at} $z=1$ but the
-limit still holds. Also many authors change variables from $z$ to other
-things, so one must keep track of how the normalization condition transforms.
-}
-Doing the integrals (and producing constant $c$) gives
-\begin{align}
-\ln F &= -\lambda \ln(1-\xi z) + c
-\\
-F &= \frac{c}{(1-\xi z)^\lambda}.
-\end{align}
-Only one choice for $c$ can satisfy initial conditions, producing
-\begin{align}
-F(z) = \left(\frac{1-\xi}{1-\xi z}\right)^\lambda
-        = \left(\frac{\theta}{1 - z(1-\theta)}\right)^\lambda,
-\end{align}
-which is exactly the negative binomial's generating function, as expected.
-
-\subsection{Adding repression}
-\subsubsection{Deriving the generating function for mRNA distribution}
-
-Let us move from a one-state promoter to a two-state promoter, where one state
-has repressor bound and the other produces transcriptional bursts as above.
-A schematic of this model is shown in~\fig{fig3:kR_inferences}.
-Although now we have an equation for each promoter state, otherwise
-the master equation reads similarly to the one-state case, except with
-additional terms corresponding to transitions between promoter states, namely
-\begin{align}
-\deriv{t}p_R(m,t) =& k_R^+ p_A(m,t) - k_R^- p_R(m,t)
-        + (m+1)\gamma p_R(m+1,t) - m\gamma p_R(m,t)
-\\
-\begin{split}
-\deriv{t}p_A(m,t) =& - k_R^+ p_A(m,t) + k_R^- p_R(m,t)
-        + (m+1)\gamma p_A(m+1,t) - m\gamma p_A(m,t) 
-\\
-&- k_i p_A(m,t) + k_i \sum_{m^\prime=0}^m \theta(1-\theta)^{m-m^\prime} p_A(m^\prime,t),
-\end{split}
-\end{align}
-where $p_R(m,t)$ is the probability of the system having $m$ mRNA copies
-and having repressor bound to the promoter at time $t$, and $p_A$ is an
-analogous probability to find the promoter without repressor bound.
-$k_R+$ and $k_R^-$ are, respectively, the rates at which repressors
-bind and unbind to and from the promoter, and
-$\gamma$ is the mRNA degradation rate.
-$k_i$ is the rate at which bursts initiate, and as before,
-the geometric distribution of burst sizes has mean $b=(1-\theta)/\theta$.\footnote{
-$\theta$ can be viewed as a Bernoulli trial probability: after each transcript
-in a burst is produced, $\theta$ is the probability that the burst
-terminates and $(1-\theta)$ is the probability that the burst continues.
-}
-
-Interestingly, it turns out that this problem maps exactly onto the
-three-stage promoter model considered by
-Shahrezaei and Swain in~\cite{Shahrezaei2008}, with relabelings.
-Their approximate solution for protein distributions amounts to
-the same approximation we make here in regarding the duration
-of mRNA synthesis bursts as instantaneous, so their solution
-for protein distributions also solves our problem of mRNA distributions.
-Let us examine the analogy more closely.
-They consider a two-state promoter, as we do here, but they model
-mRNA as being produced one at a time and degraded, with rates
-$v_0$ and $d_0$.
-Then they model translation as occuring with rate $v_1$,
-and protein degradation with rate $d_1$.
-Now consider the limit where $v_1, d_0\rightarrow\infty$ with
-their ratio $v_1/d_0$ held constant.
-$v_1/d_0$ resembles the average burst size of translation from a single
-mRNA: these are the rates of two Poisson processes that compete over a
-transcript, which matches the story of geometrically distributed burst sizes.
-So after taking this limit, on timescales slow compared to $v_1$ and $d_0$,
-it appears that transcription events fire at rate $v_0$ and produce a
-geometrically distributed burst of translation of mean size $v_1/d_0$,
-which intuitively matches the story we have told above for mRNA
-with variables relabeled.
-
-To verify this intuitively conjectured mapping between our problem and
-the solution in~\cite{Shahrezaei2008}, we continue with a careful
-solution for the mRNA distribution using probability generating functions,
-following the ideas sketched in~\cite{Shahrezaei2008}.
-It is natural to nondimensionalize rates in the problem by $\gamma$,
-or equivalently, this amounts to measuring time in units of $\gamma^{-1}$.
-We are also only interested in steady state, so we set the time
-derivatives to zero, giving
-\begin{align}
-0 =& k_R^+ p_A(m) - k_R^- p_R(m) + (m+1) p_R(m+1) - m p_R(m)
-\\
-\begin{split}
-0 =& - k_R^+ p_A(m) + k_R^- p_R(m) + (m+1) p_A(m+1) - m p_A(m) 
-\\
-&- k_i p_A(m) + k_i \sum_{m^\prime=0}^m \theta(1-\theta)^{m-m^\prime} p_A(m^\prime),
-\end{split}
-\end{align}
-
-The probability generating function is defined as before in the
-constitutive case, except now we must introduce a generating function
-for each promoter state,
-\begin{align}
-f_A(z) = \sum_{m=0}^\infty z^m p_A(m),
-\;
-f_R(z) = \sum_{m=0}^\infty z^m p_R(m).
-\end{align}
-Our real objective is the generating function $f(z)$ that generates
-the mRNA distribution $p(m)$, independent of what state the promoter is in.
-But since $p(m) = p_A(m) + p_R(m)$, it follows too that $f(z) = f_A(z) + f_R(z)$.
-
-So as before we multiply both equations by $z^m$ and sum over all $m$.
-Each individual term transforms exactly as did an analogous term in
-the constitutive case, so the coupled ODEs for the generating functions read
-\begin{align}
-0 =& k_R^+ f_A(z) - k_R^- f_R(z) + \pderiv{z} f_R(z) - z \pderiv{z} f_R(z)
-\\
-\begin{split}
-0 =&  - k_R^+ f_A(z) + k_R^- f_R(z) + \pderiv{z} f_A(z) - z \pderiv{z} f_A(z)
-\\
-&- k_i f_A(z) + k_i \frac{\theta}{1-z(1-\theta)} f_A(z),
-\end{split}
-\end{align}
-and after changing variables $\xi = 1 - \theta$ as before
-and rearranging, we have
-\begin{align}
-0 &= k_R^+ f_A(z) - k_R^- f_R(z) + (1-z) \pderiv{z} f_R(z)
-\\
-0 &=  - k_R^+ f_A(z) + k_R^- f_R(z) + (1 - z) \pderiv{z} f_A(z)
-+ k_i \frac{(z-1)\xi}{1-z\xi} f_A(z),
-\end{align}
-We can transform this problem from two coupled first-order ODEs to a
-single second-order ODE by solving for $f_A$ in the first and plugging
-into the second, giving
-\begin{align}
-\begin{split}
-0 = (1&-z) \pderiv[f_R]{z}
-+ \frac{1-z}{k_R^+}
-        \left(k_R^- \pderiv[f_R]{z} + \pderiv[f_R]{z} +(z-1) \psecderiv[f_R]{z}\right)
-\\
-&+ \frac{k_i}{k_R^+} \frac{(z-1)\xi}{1-z\xi}
-        \left(k_R^- f_R + (z-1) \pderiv[f_R]{z}\right),
-\end{split}
-\end{align}
-where, to reduce notational clutter, we have dropped the explicit $z$
-dependence of $f_A$ and $f_R$. Simplifying we have
-\begin{align}
-0 = \psecderiv[f_R]{z}
-        - \left(\frac{k_i\xi}{1-z\xi}
-                + \frac{1 + k_R^- + k_R^+}{1-z}
-        \right)\pderiv[f_R]{z}
-        + \frac{k_i k_R^- \xi}{(1-z\xi)(1-z)}f_R.
-\end{align}
-This can be recognized as the hypergeometric differential equation,
-with singularities at $z=1$, $z=\xi^{-1}$, and $z=\infty$. The latter
-can be verified by a change of variables from $z$ to $x=1/z$,
-being careful with the chain rule, and noting that $z=\infty$ is a
-singular point if and only if $x=1/z=0$ is a singular point.
-
-The standard form of the hypergeometric differential equation has its
-singularities at 0, 1, and $\infty$, so to take advantage of the
-standard form solutions to this ODE, we first need to transform
-variables to put it into a standard form.
-However, this is subtle. While any such transformation should work in
-principle, the solutions are expressed most simply in the neighborhood
-of $z=0$, but the normalization condition that we need to enforce
-corresponds to $z=1$. The easiest path, therefore, is to find a change
-of variables that maps
-1 to 0, $\infty$ to $\infty$, and $\xi^{-1}$ to 1.
-This is most intuitively done in two steps.
-
-First map the $z=1$ singularity to 0 by the change of variables $v=z-1$, giving
-\begin{align}
-0 = \psecderiv[f_R]{v}
-        + \left(\frac{k_i\xi}{(1+v)\xi - 1}
-                + \frac{1 + k_R^- + k_R^+}{v}
-        \right)\pderiv[f_R]{v}
-        + \frac{k_i k_R^- \xi}{((1+v)\xi - 1)v}f_R.
-\end{align}
-Now two singularities are at $v=0$ and $v=\infty$.
-The third is determined by $(1+v)\xi -1 = 0$, or $v=\xi^{-1} - 1$.
-We want another variable change that maps this third singularity to 1
-(without moving 0 or infinity). Changing variables again to
-$w=\frac{v}{\xi^{-1} - 1} = \frac{\xi}{1-\xi} v$ fits the bill.
-In other words, the combined change of variables
-\begin{align}
-w = \frac{\xi}{1-\xi} (z-1)
-\end{align}
-maps $z = \{1, \xi^{-1}, \infty\}$ to $w =\{0, 1, \infty\}$ as desired.
-Plugging in, being mindful of the chain rule and noting
-$(1 + v)\xi - 1 = (1 - \xi)(w - 1)$ gives
-\begin{align}
-0 = \left(\frac{\xi}{1-\xi}\right)^2 \psecderiv[f_R]{w}
-+ \left(
-        \frac{\xi k_i}{(1-\xi)(w-1)} + \frac{\xi(1 + k_R^- + k_R^+)}{(1-\xi)w}
-\right) \frac{\xi}{1-\xi} \pderiv[f_R]{w}
-+ \frac{k_i k_R^- \xi^2}{(1-\xi)^2 w(w-1)}f_R.
-\end{align}
-This is close to the standard form of the hypergeometric differential
-equation, and some cancellation and rearrangement gives
-\begin{align}
-0 = w(w-1)\psecderiv[f_R]{w}
-+ \left(k_i w + (1 + k_R^- + k_R^+)(w-1)\right) \pderiv[f_R]{w}
-+ k_i k_R^- f_R.
-\end{align}
-and a little more algebra produces
-\begin{align}
-0 = w(1-w)\psecderiv[f_R]{w}
-+ \left(1 + k_R^- + k_R^+
-        - (1 + k_i + k_R^- + k_R^+)w
-\right) \pderiv[f_R]{w}
-- k_i k_R^- f_R,
-\end{align}
-which is the standard form. From this we can read off the solution
-in terms of hypergeometric functions ${_2F_1}$ from
-any standard source, e.g.~\cite{Abramowitz1964}, and identify the
-conventional parameters in terms of our model parameters.
-We want the general solution in the neighborhood of $w=0$ ($z=1$),
-which for a homogeneous linear second order ODE must be a sum of two
-linearly independent solutions.
-More precisely,
-\begin{align}
-f_R(w) = C_1 {_2F_1}(\alpha, \beta, \gamma; w)
-+ C_2 w^{1-\gamma}{_2F_1}(1+\alpha-\gamma, 1+\beta-\gamma, 2-\gamma; w)
-\end{align}
-with parameters determined by
-\begin{align}
-\begin{split}
-\alpha\beta &= k_i k_R^-
-\\
-1+\alpha+\beta &= 1+k_i+k_R^-+k_R^+
-\\
-\gamma &= 1 + k_R^- + k_R^+
-\end{split}
-\end{align}
-and constants $C_1$ and $C_2$ to be set by boundary conditions.
-Solving for $\alpha$ and $\beta$, we find\footnote{
-Note that $\alpha$ and $\beta$ are interchangeable in the definition of
-${_2F_1}$ and differ only in the sign preceeding the radical.
-}
-\begin{align}
-\begin{split}
-\alpha &= \frac{1}{2}
-\left(k_i+k_R^-+k_R^+ + \sqrt{(k_i+k_R^-+k_R^+)^2 - 4k_i k_R^-}\right)
-\\
-\beta &= \frac{1}{2}
-\left(k_i+k_R^-+k_R^+ - \sqrt{(k_i+k_R^-+k_R^+)^2 - 4k_i k_R^-}\right)
-\\
-\gamma &= 1 + k_R^- + k_R^+.
-\end{split}
-\end{align}
-Since the normalization condition requires that $f_R$ be finite at $w=0$,
-we can immediately set $C_2=0$ to discard the second solution.
-This is because all the rate constants are strictly positive,
-so $\gamma>1$ and therefore $w^{1-\gamma}$ blows up as $w\rightarrow0$.
-Now that we have $f_R$, we would like to find the generating function
-for the mRNA distribution, $f(z) = f_A(z) + f_R(z)$.
-We can recover $f_A$ from our solution for $f_R$, namely
-\begin{align}
-f_A(z) = \frac{1}{k_R^+}\left(k_R^- f_R(z) + (z-1) \pderiv[f_R]{z}\right)
-\end{align}
-or
-\begin{align}
-f_A(w) = \frac{1}{k_R^+}\left(k_R^- f_R(w) + w \pderiv[f_R]{w}\right),
-\end{align}
-where in the second line we transformed our original relation between
-$f_R$ and $f_A$ to our new, more convenient, variable $w$.
-Plugging our solution for $f_R(w) = C_1{_2F_1}(\alpha, \beta, \gamma; w)$
-into $f_A$, we will require the differentiation rule for ${_2F_1}$,
-which tells us
-\begin{align}
-\pderiv[f_R]{w} = C_1\frac{\alpha\beta}{\gamma}
-                {_2F_1}(\alpha+1, \beta+1, \gamma+1; w),
-\end{align}
-from which it follows that
-\begin{align}
-f_A(w) = \frac{C_1}{k_R^+}
-\left(
-k_R^- {_2F_1}(\alpha, \beta, \gamma; w)
-+ w\frac{\alpha\beta}{\gamma} {_2F_1}(\alpha+1, \beta+1, \gamma+1; w)
-\right)
-\end{align}
-and therefore
-\begin{align}
-f(w) = C_1\left(1 + \frac{k_R^-}{k_R^+}\right)
-        {_2F_1}(\alpha, \beta, \gamma; w)
-+ w \frac{C_1}{k_R^+} \frac{\alpha\beta}{\gamma}
-        {_2F_1}(\alpha+1, \beta+1, \gamma+1; w).
-\end{align}
-To proceed, we need one of the (many) useful identities known for
-hypergeometric functions, in particular
-\begin{align}
-w\frac{\alpha\beta}{\gamma} {_2F_1}(\alpha+1, \beta+1, \gamma+1; w)
-=
-(\gamma-1)\left(
-{_2F_1}(\alpha, \beta, \gamma-1; w) - {_2F_1}(\alpha, \beta, \gamma; w)
-\right).
-\end{align}
-Substituting this for the second term in $f(w)$, we find
-\begin{align}
-f(w) = \frac{C_1}{k_R^+}
-\left[
-        \left(k_R^+ + k_R^-\right)
-        {_2F_1}(\alpha, \beta, \gamma; w)
-+ (\gamma-1)\left(
-        {_2F_1}(\alpha, \beta, \gamma-1; w) - {_2F_1}(\alpha, \beta, \gamma; w)
-        \right)
-\right],
-\end{align}
-and since $\gamma-1 = k_R^+ + k_R^-$, the first and third terms cancel,
-leaving only
-\begin{align}
-f(w) = C_1\frac{k_R^+ + k_R^-}{k_R^+} {_2F_1}(\alpha, \beta, \gamma-1; w).
-\end{align}
-Now we enforce normalization, demanding $f(w=0) = f(z=1) = 1$.
-${_2F_1}(\alpha, \beta, \gamma-1; 0) = 1$, so we must have
-$C_1 = k_R^+ / (k_R^+ + k_R^-)$ and consequently
-\begin{align}
-f(w) =  {_2F_1}(\alpha, \beta, k_R^+ + k_R^-; w).
-\end{align}
-Recalling that the mean burst size $b = (1-\theta)/\theta = \xi/(1-\xi)$
-and $w = \frac{\xi}{1-\xi} (z-1) = b (z-1)$,
-we can transform back to the original variable $z$ to find the tidy result
-\begin{align}
-f(z) =  {_2F_1}(\alpha, \beta, k_R^+ + k_R^-; b(z-1)),
-\end{align}
-with $\alpha$ and $\beta$ given above by
-\begin{align}
-\begin{split}
-\alpha &= \frac{1}{2}
-\left(k_i+k_R^-+k_R^+ + \sqrt{(k_i+k_R^-+k_R^+)^2 - 4k_i k_R^-}\right)
-\\
-\beta &= \frac{1}{2}
-\left(k_i+k_R^-+k_R^+ - \sqrt{(k_i+k_R^-+k_R^+)^2 - 4k_i k_R^-}\right).
-\end{split}
-\end{align}
-Finally we are in sight of the original goal. We can generate the
-steady-state probability distribution of interest by
-differentiating the generating function,
-\begin{align}
-p(m) = m! \left.\frac{\partial^m}{\partial z^m} f(z) \right|_{z=0},
-\end{align}
-which follows easily from its definition. Some contemplation reveals
-that repeated application of the derivative rule used above will produce
-products of the form $\alpha(\alpha+1)(\alpha+2)\cdots(\alpha+m-1)$ in
-the expression for $p(m)$ and similarly for $\beta$ and $\gamma$. These
-resemble ratios of factorials, but since $\alpha$, $\beta$, and $\gamma$
-are not necessarily integer, we should express the ratios using gamma
-functions instead. More precisely, one finds
-\begin{align}
-p(m) = \frac{
-        \Gamma(\alpha + m)\Gamma(\beta + m)\Gamma(k_R^+ + k_R^-)
-        }
-        {
-        \Gamma(\alpha)\Gamma(\beta)\Gamma(k_R^+ + k_R^- + m)
-        }
-\frac{b^m}{m!}{_2F_1}(\alpha+m, \beta+m, k_R^++k_R^-+m; -b)
-\label{eq:p_m_bursty+rep}
-\end{align}
-which is finally the probability distribution we sought to derive.
-
-\subsubsection{Moments}
-We can verify the correctness of~\eq{eq:p_m_bursty+rep}
-by comparison with Gillespie simulation.
-We can also easily compute moments of the distribution from
-the generating function, and check these against raw moments computed
-directly from the master equation.
-
-The $j$-th raw moment of the distribution can be computed from the
-generating function from the formula
-\begin{align}
-\langle m^j \rangle
-= \left. \left(z \frac{\partial}{\partial z}\right)^j f(z)\right|_{z=1}.
-\end{align}
-Note that \textit{probabilities} are generated by differentiating
-and evaluating at $z=0$, while \textit{moments} are generated by
-differentiating and evaluating at $z=1$. This also aligns with the
-normalization constraint, which can be thought of as the zeroth raw moment.
-
-The mean mRNA is given by
-\begin{align}
-\langle m \rangle = \left.z \pderiv[f]{z} \right|_{z=1}
-= \frac{\alpha\beta b}{k_R^+ + k_R^-}
-= k_i b\frac{k_R^-}{k_R^+ + k_R^-},
-\end{align}
-which is intuitive and appealingly simple:
-the mean mRNA is just the mean mRNA in the absense of repression
-($k_i b$, the burst rate time mean burst size)
-times the fraction of time the repressor is bound, $k_R^-/(k_R^+ + k_R^-)$.
-
-To compute the variance, let us first compute
-\begin{align}
-\begin{split}
-\langle m^2\rangle
-&= \left .z\pderiv{z}\left(z\pderiv[f]{z}\right) \right|_{z=1}
-= \left. z\left(\pderiv[f]{z} + z\psecderiv[f]{z}\right)\right|_{z=1}
-\\
-&= \frac{\alpha\beta b}{k_R^+ + k_R^-}
-        + \frac{\alpha(\alpha+1)\beta(\beta+1) b^2}
-                {(k_R^+ + k_R^-)(k_R^+ + k_R^- + 1)}
-\\
-&= \frac{k_i b k_R^-}{k_R^+ + k_R^-}
-        + \frac{k_i k_R^- b^2 (1 + k_i + k_R^+ + k_R^- + k_i k_R^-)}
-                {(k_R^+ + k_R^-)(k_R^+ + k_R^- + 1)},
-\end{split}
-\end{align}
-from which the variance is
-\begin{align}
-\begin{split}
-var(m) = \langle m^2\rangle - \langle m\rangle^2
-&= \frac{k_i b k_R^-}{k_R^+ + k_R^-}
-\left(
-1 + \frac{b(1 + k_i + k_R^+ + k_R^- + k_i k_R^-)} {k_R^+ + k_R^- + 1}
-- \frac{k_i b k_R^-}{k_R^+ + k_R^-}
-\right)
-\\
-&= k_i b\frac{k_R^-}{k_R^+ + k_R^-}
-\left(
-1 + b + k_i b\frac{1 + k_R^-} {k_R^+ + k_R^- + 1}
-- \frac{k_i b k_R^-}{k_R^+ + k_R^-}
-\right)
-\\
-&= k_i b\frac{k_R^-}{k_R^+ + k_R^-}
-\left(
-1 + b + k_i b\frac{k_R^+}{(k_R^+ + k_R^-)(k_R^+ + k_R^- + 1)}
-\right).
-\end{split}
-\end{align}
-The Fano factor is the quantity in parentheses. This can be interpreted as
-the variability arising from the negative binomial distribution of the
-unregulated promoter $(1+b)$ plus additional noise arising from repression.
-
-\mmnote{Still to writeup: moment derivation direct from CME to compare}
-
-\subsection{Numerical considerations and recursion formulas}
-\subsubsection{Generalities}
-We would like to carry out Bayesian parameter inference on FISH data
-from~\cite{Jones2014}, using~\eq{eq:p_m_bursty+rep} as our
-likelihood. This requires accurate (and preferably fast)
-numerical evaluation of the hypergeometric function ${_2F_1}$,
-which is a notoriously hard problem~\cite{Pearson2017, Gil2007},
-and our particular needs here present an especial challenge as we show below.
-
-The hypergeometric function is defined by its Taylor series as
-\begin{align}
-{_2F_1}(a,b,c;z) 
-= \sum_{l=0}^\infty
-\frac{\Gamma(a + l)\Gamma(b + l)\Gamma(c)}
-        {\Gamma(a)\Gamma(b)\Gamma(c + l)}
-\frac{z^l}{l!}
-\end{align}
-for $|z|<1$, and by analytic continuation elsewhere.
-If $z\lesssim1/2$ and $\alpha$ and $\beta$ are not too large
-(absolute value below 20 or 30),
-then the series converges quickly and an accurate numerical representation is
-easily computed by truncating the series after a reasonable number of terms.
-Unfortunately, we need to evaluate ${_2F_1}$ over mRNA copy numbers fully out
-to the tail of the distribution, which can easily reach 50, possibly 100.
-From~\eq{eq:p_m_bursty+rep}, this means evaluating ${_2F_1}$
-repeatedly for values of $a$, $b$, and $c$ spanning the full range
-from $\mathcal{O}(1)$ to $\mathcal{O}(10^2)$,
-even if $\alpha$, $\beta$, and $\gamma$
-in~\eq{eq:p_m_bursty+rep} are small,
-with the situation even worse if they are not small.
-A naive numerical evaluation of the series definition will be
-prone to overflow and, if any of $a,b,c<0$, then some successive terms in the
-series have alternating signs which can lead to catastrophic cancellations.
-
-One solution is to evaluate ${_2F_1}$ using arbitrary precision arithmetic
-instead of floating point arithmetic,
-e.g., using the \texttt{mpmath} library in Python.
-This is accurate but incredibly slow computationally.
-To quantify how slow, we found that
-evaluating the likelihood defined by~\eq{eq:p_m_bursty+rep} $\sim50$ times
-(for a typical dataset of interest from~\cite{Jones2014},
-with $m$ values spanning 0 to $\sim50$)
-using arbitrary precision arithmetic is 100-1000 fold slower than
-evaluating a negative binomial likelihood for the corresponding
-constitutive promoter dataset.
-
-To claw back $\gtrsim30$ fold of that slowdown, we can exploit
-one of the many catalogued symmetries involving ${_2F_1}$.
-The solution involves recursion relations originally explored by Gauss,
-and studied extensively in~\cite{Pearson2017, Gil2007}.
-They are sometimes known as contiguous relations and relate the values
-of any set of 3 hypergeometric functions whose arguments differ by integers.
-To rephrase this symbolically, consider a set of hypergeometric functions
-indexed by an integer $n$,
-\begin{align}
-f_n = {_2F_1}(a+\epsilon_i n, b+\epsilon_j n, c+\epsilon_k n; z),
-\end{align}
-for a fixed choice of $\epsilon_i, \epsilon_j, \epsilon_k \in \{0,\pm 1\}$
-(at least one of $\epsilon_i, \epsilon_j, \epsilon_k$ must be nonzero,
-else the set of $f_n$ would contain only a single element).
-Then there exist known recurrence relations of the form
-\begin{align}
-A_n f_{n-1} + B_n f_{n} + C_n f_{n+1} = 0,
-\end{align}
-where $A_n, B_n$, and $C_n$ are some functions of $a,b,c$, and $z$.
-In other words, for fixed $\epsilon_i, \epsilon_j, \epsilon_k, a, b,$ and $c$,
-if we can merely evaluate ${_2F_1}$ twice, say for $n^\prime$ and $n^\prime-1$,
-then we can easily and rapidly generate values for arbitrary $n$.
-
-This provides a convenient solution for our problem: we need repeated
-evaluations of ${_2F_1}(a+m, b+m, c+m; z)$
-for fixed $a,b$, and $c$ and many integer values of $m$.
-They idea is that we can use arbitrary precision arithmetic to evaluate
-${_2F_1}$ for just two particular values of $m$ and then generate
-${_2F_1}$ for the other 50-100 values of $m$ using the recurrence
-relation.\footnote{
-There are even more sophisticated ways of utilizing the recurrence
-relations that might have netted another factor of 2 speed-up, and
-possibly as much as a factor of 10, but the method described here had
-already reduced the computation time to an acceptable
-$\mathcal{O}(\text{1 min})$, so these more sophisticated approaches did
-not seem worth the time to pursue.
-}
-
-However, there two further wrinkles.
-The first is that
-a naive application of the recurrence relation is numerically unstable.
-Roughly, this is because the three term recurrence relations,
-like second order ODEs, admit two linearly independent solutions.
-In a certain eigenbasis, one of these solutions dominates the other
-as $n\rightarrow\infty$, and as $n\rightarrow-\infty$,
-the dominance is reversed.
-If we fail to work in this eigenbasis, our solution of the recurrence relation
-will be a mixture of these solutions and rapidly accumulate numerical error.
-For our purposes, it suffices to know that the authors of~\cite{Gil2007}
-derived the numerically stable solutions (so-called \textit{minimal solutions})
-for several possible choices of $\epsilon_i, \epsilon_j, \epsilon_k$.
-Running the recurrence in the proper direction using a minimal solution
-is numerically robust and can be done entirely in floating point arithmetic, 
-so that we only need to evaluate ${_2F_1}$ with arbitrary precision arithmetic
-to generate the seed values for the recursion.
-
-The second wrinkle is a corollary to the first.
-The minimal solutions are only minimal for certain ranges of the argument $z$,
-and not all of the 26 possible recurrence relations
-have minimal solutions for all $z$.
-This can be solved by using one of the many transformation formulae for
-${_2F_1}$ to convert to a different recurrence relation that has
-a minimal solution over the required domain of $z$, although
-this can require some trial and error to find the right transformation,
-the right recurrence relation, and the right minimal solution.
-
-\subsubsection{Particulars}
-Let us now demonstrate these generalities for our problem of interest.
-In order to evaluate the probability distribution of our
-model,~\eq{eq:p_m_bursty+rep}, we need to evaluate hypergeometric functions
-of the form ${_2F_1}(\alpha+m, \beta+m, \gamma+m; -b)$
-for values of $m$ ranging from $0$ to $\mathcal{O}(100)$.
-The authors of~\cite{Gil2007} did not derive a recursion relation
-for precisely this case. We could follow their methods and do so ourselves,
-but it is much easier to convert to a case that they did consider.
-The strategy is to look through the minimal solutions tabulated
-in~\cite{Gil2007} and search for a transformation we could apply to
-${_2F_1}(\alpha+m, \beta+m, \gamma+m; -b)$ that would place the $m$'s
-(the variable being incremented by the recursion)
-in the same arguments of ${_2F_1}$ as the minimal solution.
-After some ``guess and check,'' we found that the transformation
-\begin{align}
-{_2F_1}(\alpha+m, \beta+m, \gamma+m; -b)
-=
-(1+b)^{-\alpha-m}
-        {_2F_1}\left(\alpha+m, \gamma-\beta, \gamma+m; \frac{b}{1+b}\right),
-\label{eq:rec_euler_pretransform}
-\end{align}
-produces a ${_2F_1}$ on the right hand side that closely resembles
-the minimal solutions $y_{3,m}$ and $y_{4,m}$ in Eq.~4.3 in~\cite{Gil2007}.
-Explicitly, these solutions are
-\begin{align}
-y_{3,m}
-&\propto
-{_2F_1}\left(-\alpha^\prime + \gamma^\prime - m,
-                -\beta^\prime + \gamma^\prime,
-                1-\alpha^\prime-\beta^\prime+\gamma^\prime-m;
-                1-z\right)
-\\
-y_{4,m}
-&\propto
-{_2F_1}\left(\alpha^\prime + m,
-                \beta^\prime,
-                1+\alpha^\prime+\beta^\prime-\gamma^\prime+m;
-                1-z\right),
-\label{eq:minimal_soln_sans_prefac}
-\end{align}
-where we have omitted prefactors which are unimportant for now.
-Which of these two we should use depends on what values $z$ takes on.
-Equating $1-z=b/(1+b)$ gives $z=1/(1+b)$, and since $b$ is strictly positive,
-$z$ is bounded between 0 and 1.
-From Eq.~4.5 in~\cite{Gil2007}, $y_{4,m}$ is the minimal solution
-for real $z$ satisfying $0<z<2$, so this is the only minimal solution we need.
-
-Now that we have our minimal solution,
-what recurrence relation does it satisfy?
-Confusingly, the recurrence relation of which $y_{4,m}$ is a solution
-increments different arguments of ${_2F_1}$ that does $y_{4,m}$:
-it increments the first only, rather than first and third.
-This recurrence relation can be looked up, e.g., Eq.~15.2.10
-in~\cite{Abramowitz1964}, which is
-\begin{align}
-(\gamma^\prime - (\alpha^\prime + m)) f_{m-1}
-+
-(2(\alpha^\prime+m) - \gamma^\prime + (\beta^\prime - \alpha^\prime)z)f_m
-+ \alpha^\prime(z-1) f_{m+1} = 0.
-\label{eq:chosen_rec_rel}
-\end{align}
-Now we must solve for the parameters appearing in the recurrence relation
-in terms of our parameters, namely by setting
-\begin{align}
-\begin{split}
-\alpha^\prime &= \alpha
-\\
-\beta^\prime &= \gamma - \beta
-\\
-1 + \alpha^\prime + \beta^\prime - \gamma^\prime &= \gamma
-\\
-1 - z &= \frac{b}{1+b}
-\end{split}
-\end{align}
-and solving to find
-\begin{align}
-\begin{split}
-\alpha^\prime &= \alpha
-\\
-\beta^\prime &= \gamma - \beta
-\\
-\gamma^\prime &= 1 + \alpha - \beta
-\\
-z &= \frac{1}{1+b}
-.
-\end{split}
-\end{align}
-Finally we have everything we need. The minimal solution
-\begin{align}
-y_{4,m}
-=
-\frac{\Gamma(1+\alpha^\prime-\gamma^\prime+m)}
-        {\Gamma(1+\alpha^\prime+\beta^\prime-\gamma^\prime+m)}
-\times
-{_2F_1}\left(\alpha^\prime + m,
-                \beta^\prime,
-                1+\alpha^\prime+\beta^\prime-\gamma^\prime+m;
-                1-z\right),
-\end{align}
-where we have now included the necessary prefactors,
-is a numerically stable solution of the recurrence
-relation~\eq{eq:chosen_rec_rel} if the recursion is run
-from large $m$ to small $m$.
-So the procedure is to compute the value of ${_2F_1}$ for the two
-largest $m$ values of interest using arbitrary precision arithmetic,
-compute the prefactors to construct
-$y_{4,\text{max}(m)}$ and $y_{4,\text{max}(m)-1}$,
-recursively compute $y_{4,m}$ for all $m$ less than $\text{max}(m)$ down
-to $m=0$, then cancel off the prefactors of the resulting values of
-$y_{4,m}$ for all $m$ to produce ${_2F_1}$ for all desired $m$ values.
-\mmnote{Maybe I should reformat this paragraph into
-a pseudocode version of the algorithm?
-Instead of, or in addition to?}
-
-With ${_2F_1}$ computed, the only remaining numerical danger in computing
-$p(m)$ in~\eq{eq:p_m_bursty+rep} is overflow of the gamma functions.
-This is easily solved by taking the log of the entire expression
-and using standard routines to compute the log of the gamma functions,
-then exponentiating the entire expression at the end if $p(m)$
-is needed rather than $\log p(m)$.
-
-% \bibliographystyle{nature}
-\bibliographystyle{abbrv}
-\bibliography{./library}
-
-\end{document}

### 2020-05-19 doc/section_02_means.tex
initial splitting of document into several files

--- 2020-05-17 doc/paper.tex
@@
-\documentclass[12pt]{article}%{amsart}
-\usepackage[top = 1.0in, bottom = 1.0in, left = 1.0in, right = 1.0in]{geometry}
-% See geometry.pdf to learn the layout options. There are lots.
-%\geometry{letterpaper}               % ... or a4paper or a5paper or ...
-%\geometry{landscape}             % Activate for for rotated page geometry
-\usepackage[parfill]{parskip}    % Activate to begin paragraphs with an empty line rather than an indent
-\usepackage{graphicx}
-\usepackage{amssymb}
-%\usepackage{epstopdf}
-% \usepackage{caption}
-\usepackage{amsmath}
-% \usepackage{longtable}
-% \usepackage{tabu}
-% \usepackage{accents} %to get undertildes for vec & mat
-\usepackage{color} % for colored text
-% \usepackage[normalem]{ulem} % for editing: striking out text using 'sout' command
-
-% \newcommand\mytablefigwidth{0.35\textwidth}
-
-% allows to use .ai files directly w/o resaving as pdf
-\DeclareGraphicsRule{.ai}{pdf}{.ai}{}
-
-% Handy math macros!
-\newcommand{\vect}[1]{\vec{#1}}
-\newcommand{\matr}[1]{\mathbf{#1}}
-\newcommand{\rate}[3]{{#1}_{#2}^{#3}}
-\newcommand{\mmnote}[1]{\textcolor{cyan}{(MM:~#1)}}
-
-% derivative macros. these are sneaky. the [{}] is an empty default 1st arg
-% usage: provide 1 arg: \deriv{x} gives d/dx
-% provide 2 args like so: \deriv[f]{x} gives df/dx
-\newcommand{\deriv}[2][{}]{\frac{d #1}{d #2}}
-\newcommand{\pderiv}[2][{}]{\frac{\partial #1}{\partial #2}}
-\newcommand{\psecderiv}[2][{}]{\frac{\partial{^2} #1}{\partial #2{^2}}}
-
-%%%%% Referencing macros %%%%%
-\newcommand{\fig}[1]{Figure~\ref{#1}}
-\newcommand{\tref}[1]{Table~\ref{#1}}
-\newcommand{\eq}[1]{Eq.~(\ref{#1})}
-\newcommand{\eqrange}[2]{Eqs.~(\ref{#1}-\ref{#2})}
-
-%%%%%
-\begin{document}
-
-\title{Simple repression draft}
-
-\maketitle
-
-\section{Introduction}
-\mmnote{Key ideas, no particular order, that I haven't written down before:
-\begin{itemize}
-\item Our goal is to build phenomenological models of input-output
-functions of genetic circuits. More precisely, we want phenomenology
-in the sense of coarse-graining away atomistic detail, but still
-retaining biophysical meaning, e.g., we don't want to coarse-grain
-as far as Hill functions. Why not? We are motivated by studies
-like~\cite{Razo-Mejia2020}, for which Hill functions are
-insufficient. Such work requires a quantitative theory of how
-biophysical changes at the molecular level propagate to
-input-output functions at the genetic circuit level.
-We want concepts, not mere facts.
-In particular a key question is: at this level of
-coarse graining, what microscopic details do I need to explicitly
-model, and how do we figure that out? For example, do I need to
-worry about all or even any of the steps that individual RNAPs go
-through each time they make a transcript? Turning the question
-around, can we see any imprint of those processes in the
-available data? If the answer is no, then those processes are
-irrelevant for our purposes. Forward modeling and inverse
-(statistical inferential) modeling can complement each other
-beautifully here.
-\end{itemize}}
-
-\fig{fig1:means_cartoons}A shows the qualitative picture of
-simple repression. An operator, the binding site on the DNA for a
-repressor protein, may be found occupied by a repressor, in which
-case transcription is blocked from occurring. Or it may be found
-unoccupied, in which case RNA polymerase (RNAP) may bind and
-transcription can proceed.
-
-The key assumption we make is that binding of repressor and RNAP in the promoter region of interest is exclusive, meaning that one or the other may bind, but never may both be simultaneously bound.
-It is often imagined that when the repressor is bound to its operator,
-RNAP is sterically blocked from binding to its promoter sequence.
-Current evidence suggests this is sometimes, but not
-always the case~\mmnote{Cite that fig, from ??Nathan or Bill or
-Rob??, showing locations of repressor binding sites far upstream
-of promoter}, and it remains an interesting open question precisely
-how a repressor bound far upstream is able to repress transription.
-Suggestions include ``action-at-a-distance'' mediated by kinks in
-the DNA, formed when the repressor is bound, that prevent RNAP binding.
-Nevertheless, our modeling in this work is sufficiently
-coarse-grained that we simply assume exclusive binding and need
-not worry about such precise details.
-
-\mmnote{Add a notation aside somewhere that all association rates are written as zeroth-order rates.
-In other words, the concentration of the molecule is hidden inside, so for instance, $k_R^+$ is proportional to repressor copy number.}
-
+% !TEX root = ./busty_transcription.tex
 \section{Means}
 \subsection{Fold-changes are indistinguishable across models}
-As a first stop on our search for the ``right'' model of simple
-repression, let us consider what we can learn from theory and
-measurements of fold-change in mean expression. The fold-change is defined as
+As a first stop on our search for the ``right'' model of simple repression, let
+us consider what we can learn from theory and measurements of fold-change in
+mean expression. The fold-change is defined as
 \begin{equation}
@@
 \end{equation}
-where angle brackets denote the averge over a population of cells
-and mean mRNA $\langle m\rangle$ is viewed as a function of
-repressor copy number $R$.
-The second
-equality in~\ref{eq:fc_def} follows from assuming that the
-translation efficiency, i.e., the number of proteins translated
-per mRNA, is the same in both conditions.
-In other words, we assume that mean protein level is proportional
-to mean mRNA level, and that the proportionality constant is the
-same in both conditions and therefore cancels out in the ratio.
-This is reasonable since the cells in the two conditions are
-identical except for the presence of the transcription factor,
-and it is difficult to imagine the transcription factor having
-any direct effect on translation.
+where angle brackets denote the averge over a population of cells and mean mRNA
+$\langle m\rangle$ is viewed as a function of repressor copy number $R$. The
+second equality in~\ref{eq:fc_def} follows from assuming that the translation
+efficiency, i.e., the number of proteins translated per mRNA, is the same in
+both conditions. In other words, we assume that mean protein level is
+proportional to mean mRNA level, and that the proportionality constant is the
+same in both conditions and therefore cancels out in the ratio. This is
+reasonable since the cells in the two conditions are identical except for the
+presence of the transcription factor, and it is difficult to imagine the
+transcription factor having any direct effect on translation.
 
 Fold-change has proven a very convenient observable in past
-work~\cite{Garcia2011a, Razo-Mejia2018, Chure2019}.
-Part of its utility in dissecting transcriptional regulation is
-its ratiometric nature, which is always easier than making an
-absolute measurement. Also, by measuring otherwise identical cells
-with and without a transcription factor present, any biological
-noise common to both conditions can be made to cancel away.
-In addition, the presence of different operator sites is known to
-change the basal expression level in the absence of
-repressor~\mmnote{look up cite}.
-Computing fold-changes neatly cancels this effect as well.
-
-\fig{fig1:means_cartoons} depicts a smorgasbord
-of mathematized cartoons for simple repression that have appeared
-in previous literature. For each cartoon, we calculate the
-fold-change in mean gene expression as predicted by that model,
-deferring some algebraic details to the appendix.
-What we will find is that all cartoons collapse to a single
-master curve, shown in \fig{fig1:means_cartoons}c, which contains
-just two parameters. We label the parameters $\Delta F_R$, an
-effective free energy parametrizing the repressor-DNA
-interaction, and $\rho$, which subsumes all details of
-transcription in the absence of repressors.
-We will offer some intuition for why this master curve exists and discuss
-why it is likely impossible to discriminate ``right'' from ``wrong''
-cartoons given only measurements of fold-changes in expression.
+work~\cite{Garcia2011a, Razo-Mejia2018, Chure2019}. Part of its utility in
+dissecting transcriptional regulation is its ratiometric nature, which is always
+easier than making an absolute measurement. Also, by measuring otherwise
+identical cells with and without a transcription factor present, any biological
+noise common to both conditions can be made to cancel away. In addition, the
+presence of different operator sites is known to change the basal expression
+level in the absence of repressor~\mmnote{look up cite}. Computing fold-changes
+neatly cancels this effect as well.
+
+\fig{fig1:means_cartoons} depicts a smorgasbord of mathematized cartoons for
+simple repression that have appeared in previous literature. For each cartoon,
+we calculate the fold-change in mean gene expression as predicted by that model,
+deferring some algebraic details to the appendix. What we will find is that all
+cartoons collapse to a single master curve, shown in \fig{fig1:means_cartoons}c,
+which contains just two parameters. We label the parameters $\Delta F_R$, an
+effective free energy parametrizing the repressor-DNA interaction, and $\rho$,
+which subsumes all details of transcription in the absence of repressors. We
+will offer some intuition for why this master curve exists and discuss why it is
+likely impossible to discriminate ``right'' from ``wrong'' cartoons given only
+measurements of fold-changes in expression.
 
 \subsubsection{Equilibrium, 2-state}
-In this simplest model, the promoter is idealized as existing in
-one of two states, either repressor bound or repressor unbound.
-The rate of transcription, and therefore the level of gene
-expression, is assumed to be proportional to the fraction of time
-spent in the repressor unbound state.
-Mathematically, this amounts to the statement that the mean mRNA level is
-given by
+In this simplest model, the promoter is idealized as existing in one of two
+states, either repressor bound or repressor unbound. The rate of transcription,
+and therefore the level of gene expression, is assumed to be proportional to the
+fraction of time spent in the repressor unbound state. Mathematically, this
+amounts to the statement that the mean mRNA level is given by
 \begin{equation}
@@
 \end{equation}
-where $r$ is the transcription rate from the repressor unbound
-state, $\gamma$ is the mRNA degradation rate, $R$ is repressor
-copy number, $N_{NS}$ is the number of nonspecific binding sites
-in the genome (where repressors spend most of their time when not
-bound to the operator, their specifc binding site),
-$\beta=1/k_BT$, and $\Delta\epsilon_R$ is the binding energy of a
-repressor to its operator site. The derivation of this result is deferred
-to the appendix.
-
-The fold-change is easily found as the ratio of mean mRNA with
-and without repressor. This gives
+where $r$ is the transcription rate from the repressor unbound state, $\gamma$
+is the mRNA degradation rate, $R$ is repressor copy number, $N_{NS}$ is the
+number of nonspecific binding sites in the genome (where repressors spend most
+of their time when not bound to the operator, their specifc binding site),
+$\beta=1/k_BT$, and $\Delta\epsilon_R$ is the binding energy of a repressor to
+its operator site. The derivation of this result is deferred to the appendix.
+
+The fold-change is easily found as the ratio of mean mRNA with and without
+repressor. This gives
 \begin{equation}
@@
 \end{equation}
-which clearly matches the form of the master curve
-in~\eq{fig1:means_cartoons} with $\rho=1$.
-
-In fact it was noted in~\cite{Chure2019} that this two-state
-model can be viewed as the coarse-graining of any equilibrium
-promoter model in which no transcriptionally active states have
-transcription factor bound. Put differently, there is no overlap
-between TF bound states and transcriptionally active states. We
-will see this explicitly in the 3-state equilibrium model below,
-but perhaps surprising is that something analogous carries over
-even to the nonequilibrium models we consider later.
+which clearly matches the form of the master curve in~\eq{fig1:means_cartoons}
+with $\rho=1$.
+
+In fact it was noted in~\cite{Chure2019} that this two-state model can be viewed
+as the coarse-graining of any equilibrium promoter model in which no
+transcriptionally active states have transcription factor bound. Put
+differently, there is no overlap between TF bound states and transcriptionally
+active states. We will see this explicitly in the 3-state equilibrium model
+below, but perhaps surprising is that something analogous carries over even to
+the nonequilibrium models we consider later.
 
 \subsubsection{Equilibrium, 3-state}
-Compared to the previous model, here we fine-grain the repressor
-unbound state into two separate states: empty, and RNAP bound.
-This picture was used in~\cite{Garcia2011a} as we use it here,
-and in~\cite{Razo-Mejia2018} and~\cite{Chure2019} it was generalized
-to incorporate small-molecule inducers that bind the repressor.
-Interestingly, the effect of this is simply to renormalize $R$
-from the total number of repressors to a smaller effective number,
-roughly (though not exactly) the number which are unbound by inducers.
-
-The mean mRNA, as derived in the appendix from a similar
-enumeration of states and weights as the previous model, is
+Compared to the previous model, here we fine-grain the repressor unbound state
+into two separate states: empty, and RNAP bound. This picture was used
+in~\cite{Garcia2011a} as we use it here, and in~\cite{Razo-Mejia2018}
+and~\cite{Chure2019} it was generalized to incorporate small-molecule inducers
+that bind the repressor. Interestingly, the effect of this is simply to
+renormalize $R$ from the total number of repressors to a smaller effective
+number, roughly (though not exactly) the number which are unbound by inducers.
+
+The mean mRNA, as derived in the appendix from a similar enumeration of states
+and weights as the previous model, is
 \begin{equation}
@@
 \end{equation}
-where the new variables are $\Delta\epsilon_P$, the difference in
-RNAP binding energy to its specific site (the promoter) relative
-to an average nonspecific background site, and the RNAP copy
-number, $P$. The fold-change again follows immediately as
+where the new variables are $\Delta\epsilon_P$, the difference in RNAP binding
+energy to its specific site (the promoter) relative to an average nonspecific
+background site, and the RNAP copy number, $P$. The fold-change again follows
+immediately as
 \begin{align}
@@
 \end{align}
-with $\Delta F_R = \beta\Delta\epsilon_R - \log(R/N_{NS})$
-and $\rho = 1 + \frac{P}{N_{NS}}\mathrm{e}^{-\beta\Delta\varepsilon_P}$
-as shown in~\fig{fig1:means_cartoons}.
+with $\Delta F_R = \beta\Delta\epsilon_R - \log(R/N_{NS})$ and $\rho = 1 +
+\frac{P}{N_{NS}}\mathrm{e}^{-\beta\Delta\varepsilon_P}$ as shown
+in~\fig{fig1:means_cartoons}.
 
@@
 \includegraphics[width=\textwidth]{../figures/fig1/fig1.ai}
-\caption{
-\textbf{
-An overview of the simple repression motif at the level of means.}
-(A) depicts the qualitative biological picture of simple repression.
-(B) shows a variety of possible mathematized cartoons of simple
-repression, along with the effective parameter $\rho$
-which subsumes all regulatory details
-of the architecture excluding the repressor.
-(C) shows the ``master curve'' to which all cartoons in (B) collapse.
-}
+\caption{\textbf{An overview of the simple repression motif at the level of
+means.} (A) depicts the qualitative biological picture of simple repression. (B)
+shows a variety of possible mathematized cartoons of simple repression, along
+with the effective parameter $\rho$ which subsumes all regulatory details of the
+architecture excluding the repressor. (C) shows the ``master curve'' to which
+all cartoons in (B) collapse.}
 \label{fig1:means_cartoons}
@@
 \subsubsection{Nonequilibrium model 1 - Poisson promoter}
-For our first kinetic model, we consider the simplest possible
-picture with only two states, repressor bound and unbound.
-This is exactly the model used for the main results of~\cite{Jones2014}.
-In this picture, repressor association and dissociation rates
-from its operator site, $k_R^+$ and $k_R^-$, respectively, govern
-transitions between the two states. When the system is in the
-unbound state, transcriptions initiate at rate $r$, which
-represents a coarse-graining of all the downstream processes into
-a single effective rate. mRNA is again degraded at rate $\gamma$.
-
-Let $p_{m,R}(t)$ denote the joint probability of finding the
-system in the repressor bound state with $m$ mRNA molecules
-present at time $t$, and similarly define $p_{m,U}(t)$ for the
-repressor unbound state. 
-This model is governed by a master equation giving the time
-evolution of $p_{m,R}(t)$ and $p_{m,U}(t)$, which is
+For our first kinetic model, we consider the simplest possible picture with only
+two states, repressor bound and unbound. This is exactly the model used for the
+main results of~\cite{Jones2014}. In this picture, repressor association and
+dissociation rates from its operator site, $k_R^+$ and $k_R^-$, respectively,
+govern transitions between the two states. When the system is in the unbound
+state, transcriptions initiate at rate $r$, which represents a coarse-graining
+of all the downstream processes into a single effective rate. mRNA is again
+degraded at rate $\gamma$.
+
+Let $p_{m,R}(t)$ denote the joint probability of finding the system in the
+repressor bound state with $m$ mRNA molecules present at time $t$, and similarly
+define $p_{m,U}(t)$ for the repressor unbound state. This model is governed by a
+master equation giving the time evolution of $p_{m,R}(t)$ and $p_{m,U}(t)$,
+which is
 \begin{align}
@@
                 &+ (m+1)\gamma p_{m+1,U}(t) - \gamma p_{m,U}(t).
+\label{eq:poisson_promoter_cme}
 \end{split}
 \end{align}
-In each equation, the first two terms describe transitions
-between promoter states due to repressors unbinding and binding,
-respectively. The final two terms describe degradation of mRNA,
-decreasing the copy number by one, and the terms with coefficient
-$r$ describe transcription initiation increasing the mRNA copy
-number by one.
-
-We can greatly simplify the notation, which will be especially
-useful for the more complicated models yet to come, by
-re-expressing the master equation in vector form. The promoter
-states are collected into a vector and the rate constants are
-collected into matrices as
+In each equation, the first two terms describe transitions between promoter
+states due to repressors unbinding and binding, respectively. The final two
+terms describe degradation of mRNA, decreasing the copy number by one, and the
+terms with coefficient $r$ describe transcription initiation increasing the mRNA
+copy number by one.
+
+We can greatly simplify the notation, which will be especially useful for the
+more complicated models yet to come, by re-expressing the master equation in
+vector form. The promoter states are collected into a vector and the rate
+constants are collected into matrices as
 \begin{equation}
@@
                 + \matr{R} \vect{p}_{m-1}(t)
-                + \gamma (m+1) \matr{I} \vect{p}_{m+1}(t)
-                - \gamma m \matr{I} \vect{p}_m(t),
+                + \gamma (m+1) \matr{I} \vect{p}_{m+1}(t),
 \label{eq:2state_rep_cme}
 \end{equation}
-where $\matr{I}$ is the identity matrix.
-Taking steady state by setting time derivatives to zero,
-the mean mRNA can be found to be
+where $\matr{I}$ is the identity matrix. Taking steady state by setting time
+derivatives to zero, the mean mRNA can be found to be
 \begin{equation}
@@
 \end{equation}
-with the algebra details again deferred to the appendix.
-Recall $k_R^+$ is proportional to the repressor copy number, so
-in computing fold-change, absence of repressor corresponds to
-$k_R^+\rightarrow0$.
-Therefore fold-change in this model is simply
+with the algebra details again deferred to the appendix. Recall $k_R^+$ is
+proportional to the repressor copy number, so in computing fold-change, absence
+of repressor corresponds to $k_R^+\rightarrow0$. Therefore fold-change in this
+model is simply
 \begin{equation}
@@
 \subsubsection{Nonequilibrium model 2 - RNAP bound/unbound states}
-Analogous to the three-state equilibrium model, this model
-fine-grains the repressor unbound state of nonequilibrium model 1,
-resolving it into an empty promoter state and an RNAP-bound state.
-Note in this picture, in contrast with model 4 below,
-transcription initiation is accompanied by a promoter state
-change, in keeping with the interpretation as RNAP-bound and
-empty states: if an RNAP successfully escapes the promoter and
-proceeds to elongation of a transcript, clearly it is no longer
-bound at the promoter! Therefore another RNAP must bind before
-another transcript can be initiated.
+Analogous to the three-state equilibrium model, this model fine-grains the
+repressor unbound state of nonequilibrium model 1, resolving it into an empty
+promoter state and an RNAP-bound state. Note in this picture, in contrast with
+model 4 below, transcription initiation is accompanied by a promoter state
+change, in keeping with the interpretation as RNAP-bound and empty states: if an
+RNAP successfully escapes the promoter and proceeds to elongation of a
+transcript, clearly it is no longer bound at the promoter! Therefore another
+RNAP must bind before another transcript can be initiated.
 
 The master equation governing this model is analogous
-to~\eqrange{eq:2state_cme_matrices}{eq:2state_rep_cme} for model 1 above.
-The main subtlety arises since transcription initiation
-accompanies a promoter state change.
-This can be understood by analogy to $\matr{K}$. The
-off-diagonals and diagonals of $\matr{K}$ correspond to
-transitions arriving at or departing from, respectively, the
-promoter state of interest. If transcription initiations are
-accompanied by promoter state changes, we must have separate
-matrices for arriving and departing transcription events since
-the arriving and departing transitions have different initial
-copy number of mRNA, unlike for $\matr{K}$ where they are the same.
-The master equation for this model is
+to~\eqrange{eq:2state_cme_matrices}{eq:2state_rep_cme} for model 1 above. The
+main subtlety arises since transcription initiation accompanies a promoter state
+change. This can be understood by analogy to $\matr{K}$. The off-diagonals and
+diagonals of $\matr{K}$ correspond to transitions arriving at or departing from,
+respectively, the promoter state of interest. If transcription initiations are
+accompanied by promoter state changes, we must have separate matrices for
+arriving and departing transcription events since the arriving and departing
+transitions have different initial copy number of mRNA, unlike for $\matr{K}$
+where they are the same. The master equation for this model is
 \begin{equation}
@@
 \end{equation}
-The elements of $\vect{p}_m$ encode the probabilities of having
-$m$ mRNA present along with the promoter having repressor bound,
-being empty, or having RNAP bound, respectively.
-$\matr{R_A}$ describes probability flux \textit{arriving} at the state
-$\vect{p}_m$ from a state with one fewer mRNA, namely
-$\vect{p}_{m-1}$, and $\matr{R_D}$ describes probability flux
-\textit{departing} from the state $\vect{p}_m$
-for a state with one more mRNA, namely $\vect{p}_{m+1}$.
-$\matr{K}$ is closely analogous to model 1.
+The elements of $\vect{p}_m$ encode the probabilities of having $m$ mRNA present
+along with the promoter having repressor bound, being empty, or having RNAP
+bound, respectively. $\matr{R_A}$ describes probability flux \textit{arriving}
+at the state $\vect{p}_m$ from a state with one fewer mRNA, namely
+$\vect{p}_{m-1}$, and $\matr{R_D}$ describes probability flux \textit{departing}
+from the state $\vect{p}_m$ for a state with one more mRNA, namely
+$\vect{p}_{m+1}$. $\matr{K}$ is closely analogous to model 1.
 
@@
 \end{equation}
-Fold-change is again found from the ratio prescribed by~\eq{eq:fc_def},
-from which we have
+Fold-change is again found from the ratio prescribed by~\eq{eq:fc_def}, from
+which we have
 \begin{align}
@@
 \end{align}
-which follows the master curve with
-$\rho = 1 + k_P^+/(k_P^- + r)$ as claimed.
-
-\subsubsection{Nonequilibrium model 3
-- Multistep transcription initiation and escape}
-One might reasonably complain that the first two
-``nonequilibrium'' models we have considered are straw men. Their
-steady states necessarily satisfy detailed balance which is
-equivalent to thermodynamic equilibrium. Why is this the case? At
-steady state there is by definition no net probability flux in or
-out of each promoter state, but since the promoter states form a
-linear chain, there is only one way in or out of the repressor
-bound and RNAP bound states, implying each edge must actually
-have a net zero probability flux, which is the definition of
-detailed balance (usually phrased as equality of forward and
+which follows the master curve with $\rho = 1 + k_P^+/(k_P^- + r)$ as claimed.
+
+\subsubsection{Nonequilibrium model 3 - Multistep transcription initiation and
+escape} One might reasonably complain that the first two ``nonequilibrium''
+models we have considered are straw men. Their steady states necessarily satisfy
+detailed balance which is equivalent to thermodynamic equilibrium. Why is this
+the case? At steady state there is by definition no net probability flux in or
+out of each promoter state, but since the promoter states form a linear chain,
+there is only one way in or out of the repressor bound and RNAP bound states,
+implying each edge must actually have a net zero probability flux, which is the
+definition of detailed balance (usually phrased as equality of forward and
 reverse transition fluxes).
 
-Now we consider model 3 in~\eq{fig1:means_cartoons} which allows
-the possibility of true nonequilibrium steady-state fluxes
-through the promoter states.
-We point out that this model was considered previously in~\cite{Mitarai2015}
-where a comparison was made with model 1 as used in~\cite{Jones2014}.
-The authors of~\cite{Mitarai2015} argued that the additional
-complexity is essential to properly account for the noise in the
-mRNA distribution. We will weigh in on both models later when we
-consider observables beyond fold-change.
-
-The master equation governing this model is
-identical in form to model 2 above, namely
+Now we consider model 3 in~\eq{fig1:means_cartoons} which allows the possibility
+of true nonequilibrium steady-state fluxes through the promoter states. We point
+out that this model was considered previously in~\cite{Mitarai2015} where a
+comparison was made with model 1 as used in~\cite{Jones2014}. The authors
+of~\cite{Mitarai2015} argued that the additional complexity is essential to
+properly account for the noise in the mRNA distribution. We will weigh in on
+both models later when we consider observables beyond fold-change.
+
+The master equation governing this model is identical in form to model 2 above,
+namely
 \begin{equation}
@@
                 + \matr{R_A} \vect{p}_{m-1}(t) +
-                \gamma (m+1) \matr{I} \vect{p}_{m+1}(t)
-                - \gamma m \matr{I} \vect{p}_m(t),
-\end{equation}
-but obviously with a higher-dimensional state space and different matrices.
-The state vector and promoter transition matrix are now
+                \gamma (m+1) \matr{I} \vect{p}_{m+1}(t),
+\end{equation}
+but obviously with a higher-dimensional state space and different matrices. The
+state vector and promoter transition matrix are now
 \begin{equation}
@@
 \end{equation}
-with the four promoter states, in order, being repressor bound,
-empty, RNAP closed complex, and RNAP open complex.
-Besides increasing dimension by one again, the only new feature in
-$\matr{K}$ is the rate $k_O$, representing the (irreversible)
-rate of open complex formation from the closed complex.
-The initiation matrices are given by
+with the four promoter states, in order, being repressor bound, empty, RNAP
+closed complex, and RNAP open complex. Besides increasing dimension by one
+again, the only new feature in $\matr{K}$ is the rate $k_O$, representing the
+(irreversible) rate of open complex formation from the closed complex. The
+initiation matrices are given by
 \begin{equation}
@@
 
-The expression for mean mRNA is substantially more complicated
-now, as worked out in the appendix where we find
+The expression for mean mRNA is substantially more complicated now, as worked
+out in the appendix where we find
 \begin{equation}
@@
 \end{equation}
-The strategy is to isolate the terms involving the repressor, so that now the fold-change is seen to be simply
+The strategy is to isolate the terms involving the repressor, so that now the
+fold-change is seen to be simply
 \begin{align}
@@
 \end{align}
-surprisingly reducing to the master curve once again, with
-$\rho = 1 + \frac{k_P^+ (k_O + r)}{r(k_O + k_P^-)}$.
-
-This example hints that an arbitrarily fine-grained model of
-downstream transcription steps may still be collapsed to the form
-of the master curve in~\fig{fig1:means_cartoons}B so long as the
-repressor binding is exclusive with transcriptionally active
-states. We offer this as a conjecture, and we suspect that a
-careful argument using the King-Altman diagram
-method~\cite{King1956, Hill1966} might furnish a ``proof.''
-Our focus here is not on full generality but rather to survey an
-assortment of plausible models that have been proposed in the
-literature for simple repression.
+surprisingly reducing to the master curve once again, with $\rho = 1 +
+\frac{k_P^+ (k_O + r)}{r(k_O + k_P^-)}$.
+
+This example hints that an arbitrarily fine-grained model of downstream
+transcription steps may still be collapsed to the form of the master curve
+in~\fig{fig1:means_cartoons}B so long as the repressor binding is exclusive with
+transcriptionally active states. We offer this as a conjecture, and we suspect
+that a careful argument using the King-Altman diagram method~\cite{King1956,
+Hill1966} might furnish a ``proof.'' Our focus here is not on full generality
+but rather to survey an assortment of plausible models that have been proposed
+in the literature for simple repression.
 
 \subsubsection{Nonequilibrium model 4 - ``Active'' and ``inactive'' states}
-This model is at the core of the theory in~\cite{Razo-Mejia2020}.
-At a glance the cartoon for this model may appear very similar to model 2,
-and mathematically it is, but the interpretation is rather different.
-In model 2, we interpreted the third state literally as an
-RNAP-bound promoter and modeled initiation of a transcript as
-triggering a promoter state change, making the hopefully
-uncontroversial assumption that an RNAP can only make one
-transcript at a time. In contrast, in the present model the
-promoter state does \textit{not} change when a transcript is
-initiated. So we no longer interpret these states as literally
-RNAP bound and unbound but instead as coarse-grained ``active''
-and ``inactive'' states, the details of which we leave
-unspecified for now. We will comment more on this model below
-when we discuss Fano factors of models.
-
-Mathematically this model is very similar to models 1 and 2.
-Like model 1, the matrix $R$ describing transcription initiation
-is diagonal, namely
+This model is at the core of the theory in~\cite{Razo-Mejia2020}. At a glance
+the cartoon for this model may appear very similar to model 2, and
+mathematically it is, but the interpretation is rather different. In model 2, we
+interpreted the third state literally as an RNAP-bound promoter and modeled
+initiation of a transcript as triggering a promoter state change, making the
+hopefully uncontroversial assumption that an RNAP can only make one transcript
+at a time. In contrast, in the present model the promoter state does
+\textit{not} change when a transcript is initiated. So we no longer interpret
+these states as literally RNAP bound and unbound but instead as coarse-grained
+``active'' and ``inactive'' states, the details of which we leave unspecified
+for now. We will comment more on this model below when we discuss Fano factors
+of models.
+
+Mathematically this model is very similar to models 1 and 2. Like model 1, the
+matrix $R$ describing transcription initiation is diagonal, namely
 \begin{equation}
@@
 so the master equation takes verbatim the same form as it did for model
-1,~\eq{eq:2state_rep_cme}. Meanwhile the promoter transition matrix
-$K$ is the same as~\eq{eq:3state_cme_matrices_pt1} from model 2,
-although we relabel the rate constants from $k_P^\pm$ to $k^\pm$ to
-reiterate that these are not simply RNAP binding and unbinding rates.
+1,~\eq{eq:2state_rep_cme}. Meanwhile the promoter transition matrix $K$ is the
+same as~\eq{eq:3state_cme_matrices_pt1} from model 2, although we relabel the
+rate constants from $k_P^\pm$ to $k^\pm$ to reiterate that these are not simply
+RNAP binding and unbinding rates.
 
@@
 \subsubsection{Nonequilibrium model 5 - Bursty promoter}
-We will have much more to say about this model in subsequent sections.
-For now we sketch it as an intuitive analog to model 1,
-with just two states, repressor bound or unbound, and transition rates
-between them of $k_R^+$ and $k_R^-$.
-In model 1, when in the unbound state, single mRNA transcripts
-are produced as a Poisson process with some characteristic rate
-$r$. This model by contrast produces, at some Poisson rate $k_i$,
-\textit{bursts} of transcripts with a geometrically distributed
-burst size and a mean burst size $b$.
-
-From this intuitive picture and by analogy to model 1, then, it should be plausible that the mean mRNA level is
+We will have much more to say about this model in subsequent sections. For now
+we sketch it as an intuitive analog to model 1, with just two states, repressor
+bound or unbound, and transition rates between them of $k_R^+$ and $k_R^-$. In
+model 1, when in the unbound state, single mRNA transcripts are produced as a
+Poisson process with some characteristic rate $r$. This model by contrast
+produces, at some Poisson rate $k_i$, \textit{bursts} of transcripts with a
+geometrically distributed burst size and a mean burst size $b$.
+
+From this intuitive picture and by analogy to model 1, then, it should be
+plausible that the mean mRNA level is
 \begin{equation}
@@
 \end{equation}
-which will turn out to be correct from a careful calculation.
-For now, we simply note that just like model 1, the fold-change becomes
+which will turn out to be correct from a careful calculation. For now, we simply
+note that just like model 1, the fold-change becomes
 \begin{equation}
@@
 What have we learned from our exploration of the models
-in~\fig{fig1:means_cartoons} and their predictions of
-fold-changes in mean expression levels? The takeaway is the
-existence of a master curve (\fig{fig1:means_cartoons}C) to which
-the fold-change predictions of all the models collapse. This
-master curve is parametrized by only two effective parameters:
-$\Delta F_R$, which characterizes the repressor, and $\rho$,
-which characterizes all other features of the promoter
-architecture. The key assumption underpinning this result is that
-zero transcription occurs when a repressor is bound to its
-operator. Note, however, that we are agnostic about the molecular
-mechanism which achieves this;
-steric effects are the obvious mechanism, but, for instance,
-``action-at-a-distance'' mediated by kinked DNA due to
-repressors bound tens or hundreds of nucleotides upstream of a
-promoter is plausible as well.
-
-Why does the master curve of~\fig{fig1:means_cartoons}C exist at all?
-This brings to mind the deep questions posed in,
-e.g.,~\cite{Frank2013} and~\cite{Frank2014a}, asking us to search
-for general ``symmetries'' of the problem instead of specifying
-generative models of the known microscopics of a system.
-In our case, the key feature seems to be the exclusive nature of
-repressor and RNAP binding, which allows the parameter describing
-the repressor, $\Delta F_R$, to cleanly separate from all other
-details of the promoter architecture, which are encapsulated in $\rho$.
-In the graph-theoretic language which was applied to similar
-problems in~\cite{Gunawardena2012} and~\cite{Ahsendorf2014}, the
-repressor-bound state is a leaf of the graph describing
-transitions between promoter states. Arbitrary nonequilibrium
-behavior can occur on the rest of the promoter state space, but
-it may all be swept up in the effective parameter $\rho$, to
-which the repressor makes no contribution.
-
-As suggested in~\cite{Chure2019}, we believe this master curve
-should generalize to architectures with multiple repressor
-binding sites, as long as the exclusivity of transcription factor
-binding and transcription initation is maintained. The
-interpretation of $\Delta F_R$ is then of an effective free
-energy of all repressor bound states.
-In an equilibrium picture this is simply given by the log of the
-sum of Boltzmann weights of all repressor bound states, which
-looks like the log of a partition function of a subsystem.
-In a nonequilibrium picture, while we can still mathematically gather
-terms and give the resulting collection the label $\Delta F_R$,
-it is unclear if the physical interpretation as an effective free energy
-makes sense.
-As is well known, free energies cannot be assigned unambiguously
-to states out of equilibrium because the free energy change along
-a generic path traversing the state space is path dependent,
-unlike at equilibrium.
-A consequence of this is that, out of equilibrium, $\Delta F_R$
-is no longer a simple sum of Boltzmann weights. Instead it
-resembles a restricted sum of King-Altman
-diagrams~\cite{King1956, Hill1966}. Following the work of
-Hill~\cite{Hill1989}, it may yet be possible to interpret this expression
-as an effective free energy, but this remains unclear to us.
-We leave this an open problem for future work.
-
-If we relax the requirement of exclusive repressor-RNAP binding,
-one could imagine models in which repressor and RNAP doubly-bound
-states are allowed, where the repressor's effect is to
-\text{reduce} the transcription rate rather than setting it to zero.
-Our results do not strictly apply to such a model, although we
-note that if the repressor's reduction of the transcription rate
-is substantial, such a model might still be well-approximated by
-one of the models in~\fig{fig1:means_cartoons}.
-
-One may worry that our ``one curve to rule them all'' is a
-mathematical tautology. In fact we \textit{agree} with this
-criticism if $\Delta F_R$ is ``just a fitting parameter'' and cannot
-be meaningfully interpreted as a real, physical free energy.
-An analogy to Hill functions is appropriate here. One of their
-great strengths and weaknesses, depending on the use they are put
-to, is that their parameters coarse-grain many details and are
-generally not interpretable in terms of microscopic models, for
-deep reasons discussed at length in~\cite{Frank2013}.
-By contrast, our master curve claims to have the best of both worlds:
-a coarse-graining of all details besides the repressor into
-a single effective parameter $\rho$,
-while simultaneously retaining an interpretation of
-$\Delta F_R$ as a physically
-meaningful and interpretable free energy.
-Our task, then, is to prove or disprove this claim.
-
-How do we test this and probe the theory with fold-change
-measurements? There is a fundamental limitation in that the
-master curve is essentially a one-parameter function of $\Delta
-F_R + \log\rho$. Worse, there are many \textit{a priori}
-plausible microscopic mechanisms that could contribute to the
-value of $\rho$, such as RNAP binding and escape
-kinetics~\cite{DeHaseth1998, Mitarai2015}, and/or supercoiling
-accumulation and release~\cite{Chong2014, Sevier2016}, and/or,
-RNAP clusters analogous to those obvserved in
-eukaryotes~\cite{Cisse2013, Cho2016}
-and recently also observed in bacteria~\cite{Ladouceur2020}.
-Even if $\Delta F_R$ is measured to
-high precision, inferring the potential microscopic contributions
-to $\rho$, buried inside a log no less, from fold-change
-measurements seems hopeless. As a statistical inference problem
-it is entirely nonidentifiable, in the language
+in~\fig{fig1:means_cartoons} and their predictions of fold-changes in mean
+expression levels? The takeaway is the existence of a master curve
+(\fig{fig1:means_cartoons}C) to which the fold-change predictions of all the
+models collapse. This master curve is parametrized by only two effective
+parameters: $\Delta F_R$, which characterizes the repressor, and $\rho$, which
+characterizes all other features of the promoter architecture. The key
+assumption underpinning this result is that zero transcription occurs when a
+repressor is bound to its operator. Note, however, that we are agnostic about
+the molecular mechanism which achieves this; steric effects are the obvious
+mechanism, but, for instance, ``action-at-a-distance'' mediated by kinked DNA
+due to repressors bound tens or hundreds of nucleotides upstream of a promoter
+is plausible as well.
+
+Why does the master curve of~\fig{fig1:means_cartoons}C exist at all? This
+brings to mind the deep questions posed in, e.g.,~\cite{Frank2013}
+and~\cite{Frank2014a}, asking us to search for general ``symmetries'' of the
+problem instead of specifying generative models of the known microscopics of a
+system. In our case, the key feature seems to be the exclusive nature of
+repressor and RNAP binding, which allows the parameter describing the repressor,
+$\Delta F_R$, to cleanly separate from all other details of the promoter
+architecture, which are encapsulated in $\rho$. In the graph-theoretic language
+which was applied to similar problems in~\cite{Gunawardena2012}
+and~\cite{Ahsendorf2014}, the repressor-bound state is a leaf of the graph
+describing transitions between promoter states. Arbitrary nonequilibrium
+behavior can occur on the rest of the promoter state space, but it may all be
+swept up in the effective parameter $\rho$, to which the repressor makes no
+contribution.
+
+As suggested in~\cite{Chure2019}, we believe this master curve should generalize
+to architectures with multiple repressor binding sites, as long as the
+exclusivity of transcription factor binding and transcription initation is
+maintained. The interpretation of $\Delta F_R$ is then of an effective free
+energy of all repressor bound states. In an equilibrium picture this is simply
+given by the log of the sum of Boltzmann weights of all repressor bound states,
+which looks like the log of a partition function of a subsystem. In a
+nonequilibrium picture, while we can still mathematically gather terms and give
+the resulting collection the label $\Delta F_R$, it is unclear if the physical
+interpretation as an effective free energy makes sense. As is well known, free
+energies cannot be assigned unambiguously to states out of equilibrium because
+the free energy change along a generic path traversing the state space is path
+dependent, unlike at equilibrium. A consequence of this is that, out of
+equilibrium, $\Delta F_R$ is no longer a simple sum of Boltzmann weights.
+Instead it resembles a restricted sum of King-Altman diagrams~\cite{King1956,
+Hill1966}. Following the work of Hill~\cite{Hill1989}, it may yet be possible to
+interpret this expression as an effective free energy, but this remains unclear
+to us. We leave this an open problem for future work.
+
+If we relax the requirement of exclusive repressor-RNAP binding, one could
+imagine models in which repressor and RNAP doubly-bound states are allowed,
+where the repressor's effect is to \text{reduce} the transcription rate rather
+than setting it to zero. Our results do not strictly apply to such a model,
+although we note that if the repressor's reduction of the transcription rate is
+substantial, such a model might still be well-approximated by one of the models
+in~\fig{fig1:means_cartoons}.
+
+One may worry that our ``one curve to rule them all'' is a mathematical
+tautology. In fact we \textit{agree} with this criticism if $\Delta F_R$ is
+``just a fitting parameter'' and cannot be meaningfully interpreted as a real,
+physical free energy. An analogy to Hill functions is appropriate here. One of
+their great strengths and weaknesses, depending on the use they are put to, is
+that their parameters coarse-grain many details and are generally not
+interpretable in terms of microscopic models, for deep reasons discussed at
+length in~\cite{Frank2013}. By contrast, our master curve claims to have the
+best of both worlds: a coarse-graining of all details besides the repressor into
+a single effective parameter $\rho$, while simultaneously retaining an
+interpretation of $\Delta F_R$ as a physically meaningful and interpretable free
+energy. Our task, then, is to prove or disprove this claim.
+
+How do we test this and probe the theory with fold-change measurements? There is
+a fundamental limitation in that the master curve is essentially a one-parameter
+function of $\Delta F_R + \log\rho$. Worse, there are many \textit{a priori}
+plausible microscopic mechanisms that could contribute to the value of $\rho$,
+such as RNAP binding and escape kinetics~\cite{DeHaseth1998, Mitarai2015},
+and/or supercoiling accumulation and release~\cite{Chong2014, Sevier2016},
+and/or, RNAP clusters analogous to those obvserved in
+eukaryotes~\cite{Cisse2013, Cho2016} and recently also observed in
+bacteria~\cite{Ladouceur2020}. Even if $\Delta F_R$ is measured to high
+precision, inferring the potential microscopic contributions to $\rho$, buried
+inside a log no less, from fold-change measurements seems hopeless. As a
+statistical inference problem it is entirely nonidentifiable, in the language
 of~\cite{Gelman2013}, Section 4.3.
 
-If we cannot simply infer values of $\rho$ from measurements of
-fold-change, can we perturb some of the parameters that make up
-$\rho$ and measure the change? Unfortunately we suspect this is
-off-limits experimentally: most of the potential contributors to
-$\rho$ are global processes that affect many or all genes. For
-instance, changing RNAP association rates by changing RNAP copy
-numbers, or changing supercoiling kinetics by changing
-topoisomerase copy numbers, would massively perturb the entire
-cell's physiology and confound any determination of $\rho$.
-
-One might instead imagine a bottom-up modeling approach, where we
-mathematize a model of what we hypothesize the important steps
-are and are not, use in vitro data for the steps deemed
-important, and \textit{predict} what $\rho$ should be.
-But again, because of the one-parameter nature of the master
-curve, many different models will likely make indistinguishable
-predictions, and without any way to experimentally perturb in
-vivo, there is no clear way to test whether the modeling
-assumptions are correct.
-
-In light of this, we prefer the view that parameters and rates
-are not directly comparable between cartoons
-in~\fig{fig1:means_cartoons}. Rather, parameters in the
-simpler cartoons represent coarse-grained combinations of
-parameters in the finer-grained models. For instance, by equating
-$\rho$ between any two models, one can derive various possible
-correspondences between the two models' parameters. Note that
-these correspondences are clearly not unique, since many possible
-associations could be made. It then is a choice as to what microscopic
-interpretations the model-builder prefers for the parameters in a
-particular cartoon, and as to which coarse-grainings lend
-intuition and which seem nonsensical.
-Indeed, since it remains an open question what microscopic
-features dominate $\rho$ (as suggested above, perhaps RNAP
-binding and escape kinetics~\cite{DeHaseth1998, Mitarai2015}, or
-supercoiling accumulation and release~\cite{Chong2014, Sevier2016},
-or, something more exotic like
-RNAP clusters~\cite{Cisse2013, Cho2016, Ladouceur2020}),
-we are hesitant to put too much weight on any one microscopic
-interpretation of model parameters that make up $\rho$.
-
-One possible tuning knob to probe $\rho$ that would not globally
-perturb the cell's physiology is to manipulate RNAP binding sites.
-Work such as~\cite{Kinney2010} has shown that models of
-sequence-dependent RNAP can be inferred from data, and the
-authors of~\cite{Brewster2012} showed that the model
-of~\cite{Kinney2010} has predictive power by using the model to
-\textit{design} binding sites of a desired affinity.
-But for our purposes, this begs the question: the authors
-of~\cite{Kinney2010} \textit{assumed} a particular model
-(essentially our 3-state equilibrium model but without the
-repressor), so it is unclear how or if such methods can be turned
-around to \textit{compare} different models of promoter function.
-
-We have already pointed out that the master curve
-of~\fig{fig1:means_cartoons} is essentially a one-parameter model,
-the one parameter being $\Delta F_R + \log\rho$.
-By now the reader may be alarmed as to how can we even determine
-$\Delta F_R$ and $\rho$ independently of each other, never mind
-shedding a lens on the internal structure of $\rho$ itself.
-A hint is provided by the weak promoter approximation, invoked
-repeatedly in prior studies~\cite{Bintu2005c, Garcia2011a,
-Razo-Mejia2018} of simple repression using the 3-state
-equilibrium model in~\fig{fig1:means_cartoons}.
-In that picture, the weak promoter approximation means
-$\frac{P}{N_{NS}}\exp(-\beta\Delta\epsilon_P) \ll 1$,
-meaning therefore $\rho\approx1$.
-This is suggestive, but how can we be sure that $\rho$ is not,
-for instance, actually $10^2$ and that $\Delta F_R$ hides a
-compensatory factor?
-A resolution is offered by an independent inference of $\rho$ in
-the absence of repressors. This was done in~\cite{Razo-Mejia2020}
-by fitting nonequilibrium model 4  in~\fig{fig1:means_cartoons},
-with zero repressor (looking ahead, this is equivalent to model 4
-in~\fig{fig:constit_cartoons}), to FISH data
-from~\cite{Brewster2014}. This provided a determination of $k^+$
-and $k^-$, from which their ratio is estimated to be no more than
-a few $10^{-1}$ and possibly as small as $10^{-2}$.
-
-The realization that $\rho\approx1$ to an excellent
-approximation, \textit{independent} of which model
-in~\fig{fig1:means_cartoons} one prefers, goes a long way towards
-explaining the surprising success of equilibrium models of simple
-repression. Even though our 2- and 3-state models get so many
-details of transcription wrong, it does not matter because
-fold-change is a cleverly designed ratio. Since $\rho$ subsumes
-all details except the repressor, and $\log\rho\approx0$, fitting
-these simple models to fold-change measurements can still give a
-surprisingly good estimate of repressor binding energies. So the
-ratiometric construction of fold-change fulfills its intended
-purpose of canceling out all features of the promoter
-architecture except the repressor itself. Nevertheless it is
-perhaps surprising how effectively it does so: \textit{a priori},
-one might not have expected $\rho$ to be quite so close to 1.
-
-We would also like to highlight the relevance
-of~\cite{Landman2019} here. Landman et.\ al.\ reanalyzed and
-compared in vivo and in vitro data on the lacI repressor's
-binding affinity to its various operator sequences. (The in vivo
-data was from, essentially, fitting our master curve to
-expression measurements.) They find broad agreement between the
-in vitro and in vivo values. This reinforces the suspicion that
-the equilibrium $\Delta\epsilon_R$ repressor binding energies do
-in fact represent real physical free energies.
-Again, \textit{a priori} this did not have to be the case, even
-knowing that $\rho\approx1$.
-
-In principle, if $\Delta F_R$ can be measured to sufficient
-precision, then deviations from $\rho=1$ become a
-testable matter of experiment.
-In practice, it is probably unrealistic to measure repressor
-rates $k_R^+$ or $k_R^-$ or fold-changes in expression levels
-(and hence $\Delta\epsilon_R$) precisely enough to detect the
-expected tiny deviations from $\rho=1$.
-We can estimate the
-requisite precision in $\Delta F_R$ to resolve a given
-$\Delta\rho$ by noting, since $\rho\approx1$, that
-$\log(1+\Delta\rho)\approx \Delta\rho$, so
-$\Delta(\Delta F_R) \approx \Delta\rho$.
-Suppose we are lucky and $\Delta\rho$ happens to be $\sim0.1$, on
-the high end of our range estimated above.
-A determination of $\Delta\epsilon_R/k_BT$ with an uncertainty of
-barely 0.1 was achieved in the excellent measurements
-of~\cite{Razo-Mejia2018}, so this requires a very difficult
-determination of $\Delta F_R$ for a very crude determination of $\rho$,
-which suggests, to put it lightly, this is not a
-promising path to pursue experimentally.
-It is doubtful that inference of repressor kinetic rates
-would be any easier.\footnote{
-\mmnote{
-In fact the uncertainties quoted in Hammar et.\
-al.~\cite{Hammar2014} from Elf's group are small enough to do
-this calculation, but HJ \& I think their error bars are insanely
-optimistic. Not sure I want to blatantly call them out on it though??}
-}
-
-Moving forward, we have weak evidence supporting the
-interpretation of $\Delta F_R$ as a physically real free
-energy~\cite{Landman2019} and other work casting
-doubt~\cite{Hammar2014}.
-How might we resolve the confusion?
-If there is no discriminatory power to test the theory and
-distinguish the various models with measurements of fold-changes
-in means, how do we probe the theory? Clearly to discriminate
-between the nonequilibrium models in~\fig{fig1:means_cartoons},
-we need to go beyond means to ask questions about kinetics, noise and even
-full distributions of mRNA copy numbers over a population of cells.
-If the ``one-curve-to-rule-them-all'' is more than a
-mathematical tautology, then the free energy of repressor binding
-inferred from fold-change measurements should agree with
-repressor binding and unbinding rates.
-In other words, the equilbrium and nonequilibrium definitions of
-$\Delta F_R$ must agree, meaning
+If we cannot simply infer values of $\rho$ from measurements of fold-change, can
+we perturb some of the parameters that make up $\rho$ and measure the change?
+Unfortunately we suspect this is off-limits experimentally: most of the
+potential contributors to $\rho$ are global processes that affect many or all
+genes. For instance, changing RNAP association rates by changing RNAP copy
+numbers, or changing supercoiling kinetics by changing topoisomerase copy
+numbers, would massively perturb the entire cell's physiology and confound any
+determination of $\rho$.
+
+One might instead imagine a bottom-up modeling approach, where we mathematize a
+model of what we hypothesize the important steps are and are not, use in vitro
+data for the steps deemed important, and \textit{predict} what $\rho$ should be.
+But again, because of the one-parameter nature of the master curve, many
+different models will likely make indistinguishable predictions, and without any
+way to experimentally perturb in vivo, there is no clear way to test whether the
+modeling assumptions are correct.
+
+In light of this, we prefer the view that parameters and rates are not directly
+comparable between cartoons in~\fig{fig1:means_cartoons}. Rather, parameters in
+the simpler cartoons represent coarse-grained combinations of parameters in the
+finer-grained models. For instance, by equating $\rho$ between any two models,
+one can derive various possible correspondences between the two models'
+parameters. Note that these correspondences are clearly not unique, since many
+possible associations could be made. It then is a choice as to what microscopic
+interpretations the model-builder prefers for the parameters in a particular
+cartoon, and as to which coarse-grainings lend intuition and which seem
+nonsensical. Indeed, since it remains an open question what microscopic features
+dominate $\rho$ (as suggested above, perhaps RNAP binding and escape
+kinetics~\cite{DeHaseth1998, Mitarai2015}, or supercoiling accumulation and
+release~\cite{Chong2014, Sevier2016}, or, something more exotic like RNAP
+clusters~\cite{Cisse2013, Cho2016, Ladouceur2020}), we are hesitant to put too
+much weight on any one microscopic interpretation of model parameters that make
+up $\rho$.
+
+One possible tuning knob to probe $\rho$ that would not globally perturb the
+cell's physiology is to manipulate RNAP binding sites. Work such
+as~\cite{Kinney2010} has shown that models of sequence-dependent RNAP can be
+inferred from data, and the authors of~\cite{Brewster2012} showed that the model
+of~\cite{Kinney2010} has predictive power by using the model to \textit{design}
+binding sites of a desired affinity. But for our purposes, this begs the
+question: the authors of~\cite{Kinney2010} \textit{assumed} a particular model
+(essentially our 3-state equilibrium model but without the repressor), so it is
+unclear how or if such methods can be turned around to \textit{compare}
+different models of promoter function.
+
+We have already pointed out that the master curve of~\fig{fig1:means_cartoons}
+is essentially a one-parameter model, the one parameter being $\Delta F_R +
+\log\rho$. By now the reader may be alarmed as to how can we even determine
+$\Delta F_R$ and $\rho$ independently of each other, never mind shedding a lens
+on the internal structure of $\rho$ itself. A hint is provided by the weak
+promoter approximation, invoked repeatedly in prior studies~\cite{Bintu2005c,
+Garcia2011a, Razo-Mejia2018} of simple repression using the 3-state equilibrium
+model in~\fig{fig1:means_cartoons}. In that picture, the weak promoter
+approximation means $\frac{P}{N_{NS}}\exp(-\beta\Delta\epsilon_P) \ll 1$,
+meaning therefore $\rho\approx1$. This is suggestive, but how can we be sure
+that $\rho$ is not, for instance, actually $10^2$ and that $\Delta F_R$ hides a
+compensatory factor? A resolution is offered by an independent inference of
+$\rho$ in the absence of repressors. This was done in~\cite{Razo-Mejia2020} by
+fitting nonequilibrium model 4  in~\fig{fig1:means_cartoons}, with zero
+repressor (looking ahead, this is equivalent to model 4
+in~\fig{fig:constit_cartoons}), to FISH data from~\cite{Brewster2014}. This
+provided a determination of $k^+$ and $k^-$, from which their ratio is estimated
+to be no more than a few $10^{-1}$ and possibly as small as $10^{-2}$.
+
+The realization that $\rho\approx1$ to an excellent approximation,
+\textit{independent} of which model in~\fig{fig1:means_cartoons} one prefers,
+goes a long way towards explaining the surprising success of equilibrium models
+of simple repression. Even though our 2- and 3-state models get so many details
+of transcription wrong, it does not matter because fold-change is a cleverly
+designed ratio. Since $\rho$ subsumes all details except the repressor, and
+$\log\rho\approx0$, fitting these simple models to fold-change measurements can
+still give a surprisingly good estimate of repressor binding energies. So the
+ratiometric construction of fold-change fulfills its intended purpose of
+canceling out all features of the promoter architecture except the repressor
+itself. Nevertheless it is perhaps surprising how effectively it does so:
+\textit{a priori}, one might not have expected $\rho$ to be quite so close to 1.
+
+We would also like to highlight the relevance of~\cite{Landman2019} here.
+Landman et.\ al.\ reanalyzed and compared in vivo and in vitro data on the lacI
+repressor's binding affinity to its various operator sequences. (The in vivo
+data was from, essentially, fitting our master curve to expression
+measurements.) They find broad agreement between the in vitro and in vivo
+values. This reinforces the suspicion that the equilibrium $\Delta\epsilon_R$
+repressor binding energies do in fact represent real physical free energies.
+Again, \textit{a priori} this did not have to be the case, even knowing that
+$\rho\approx1$.
+
+In principle, if $\Delta F_R$ can be measured to sufficient precision, then
+deviations from $\rho=1$ become a testable matter of experiment. In practice, it
+is probably unrealistic to measure repressor rates $k_R^+$ or $k_R^-$ or
+fold-changes in expression levels (and hence $\Delta\epsilon_R$) precisely
+enough to detect the expected tiny deviations from $\rho=1$. We can estimate the
+requisite precision in $\Delta F_R$ to resolve a given $\Delta\rho$ by noting,
+since $\rho\approx1$, that $\log(1+\Delta\rho)\approx \Delta\rho$, so
+$\Delta(\Delta F_R) \approx \Delta\rho$. Suppose we are lucky and $\Delta\rho$
+happens to be $\sim0.1$, on the high end of our range estimated above. A
+determination of $\Delta\epsilon_R/k_BT$ with an uncertainty of barely 0.1 was
+achieved in the excellent measurements of~\cite{Razo-Mejia2018}, so this
+requires a very difficult determination of $\Delta F_R$ for a very crude
+determination of $\rho$, which suggests, to put it lightly, this is not a
+promising path to pursue experimentally. It is doubtful that inference of
+repressor kinetic rates would be any easier.\footnote{\mmnote{In fact the
+uncertainties quoted in Hammar et.\ al.~\cite{Hammar2014} from Elf's group are
+small enough to do this calculation, but HJ \& I think their error bars are
+insanely optimistic. Not sure I want to blatantly call them out on it though??}}
+
+Moving forward, we have weak evidence supporting the interpretation of $\Delta
+F_R$ as a physically real free energy~\cite{Landman2019} and other work casting
+doubt~\cite{Hammar2014}. How might we resolve the confusion? If there is no
+discriminatory power to test the theory and distinguish the various models with
+measurements of fold-changes in means, how do we probe the theory? Clearly to
+discriminate between the nonequilibrium models in~\fig{fig1:means_cartoons}, we
+need to go beyond means to ask questions about kinetics, noise and even full
+distributions of mRNA copy numbers over a population of cells. If the
+``one-curve-to-rule-them-all'' is more than a mathematical tautology, then the
+free energy of repressor binding inferred from fold-change measurements should
+agree with repressor binding and unbinding rates. In other words, the equilbrium
+and nonequilibrium definitions of $\Delta F_R$ must agree, meaning
 \begin{equation}
@@
 \end{equation}
-must hold, where $\beta\Delta\epsilon_R$ is inferred from the
-master curve fit to fold-change measurements, and $k_R^+$ and
-$k_R^-$ are inferred in some orthogonal manner.
-Single molecule measurements such as from~\cite{Hammar2014} have
-directly observed these rates, and in the remainder of this work
-we explore a complementary approach: inferring repressor binding
-and unbinding rates $k_R^+$ and $k_R^-$ from FISH measurements of
-mRNA population distributions.
-
-\section{Beyond means}
-\subsection{A one-state promoter with bursting is the ``best'' model of constitutive promoters}
-Before we can tackle simple repression, we need an adequate
-phenomenological model of constitutive expression.
-The literature abounds with options from which we can choose,
-and we show several potential kinetic models for constitutive promoters
-in~\fig{fig:constit_cartoons}.
-Let us consider the suitability of each model for our purposes in turn.
-
-\begin{figure}%[h!]
-\centering
-\includegraphics[width=0.9\textwidth]{../figures/fig1.5/fig1point5.ai}
-\caption{\textbf{Constitutive promoter cartoon comparison.}
-The left column depicts various plausible cartoons for constitutive promoters.
-We also list literature references where models have been used previously.
-In model (1), transcripts are produced in a Poisson process~\cite{Jones2014}.
-Model (2) features explicit modeling of RNAP binding/unbinding
-kinetics~\cite{Phillips2015a}.
-Model (3) is a more detailed generalization of model (2),
-treating transcription initiation as a multi-step process
-proceeding through closed and open complexes~\cite{Mitarai2015}.
-Model (4) is somewhat analogous to (2) except with the precise nature
-of active and inactive states left ambiguous~\cite{Razo-Mejia2020}.
-Finally, model (5) can be viewed as a certain limit of model (4)
-in which transcripts are produced in bursts,
-and initiation of bursts is a Poisson process.
-\mmnote{Haven't found model 5 formulated quite like this in the literature,
-usually burstiness is handled like model 4, but I should probably search a bit more.}
-The right column shows the Fano factor for each model.
-Note especially the crucial diagnostic: (2) and (3) have $\nu$
-strictly below 1, while only for (4) and (5) can $\nu$ exceed 1.
-Models with Fano factors $\le 1$ cannot produce the FISH data
-observed in~\cite{Jones2014} without introducing additional
-assumptions and model complexity.
-}
-\label{fig:constit_cartoons}
-\end{figure}
-
-\subsubsection{Model 1 - Poisson promoter.}
-This is the picture from Jones et.\ al.~\cite{Jones2014}, which
-assumes that transcripts are produced as a Poisson process from a
-single promoter state. This model insists that the ``true'' mRNA
-distribution is Poisson, implying the Fano factor $\nu$ must be 1.
-In~\cite{Jones2014}, the authors carefully attribute deviations
-from Fano = 1 to intensity variability in fluorescent spots, gene
-copy number variation, and copy number fluctuations of the
-transcription machinery, e.g., RNAP itself.
-In this picture, the master equation makes no appearance,
-and all the corrections to Poisson behavior are derived
-as additive corrections to the Fano factor.
-For disproving the ``universal noise curve'' from So et
-al~\cite{So2011}, this picture was excellent. It is appealing in
-its simplicity, with only two parameters, the initiation rate $r$
-and degradation rate $\gamma$. Since $\gamma$ is independently
-known from other experiments, and the mean mRNA copy number is
-$r/\gamma$, $r$ is easily inferred from data.
-In other words, the model is not excessively complex for the data at hand.
-But for many interesting questions, for instance in the recent
-work~\mmnote{cite Manuel's preprint once posted},
-knowledge of means and variances alone is insufficient, and a
-description of the full distribution of molecular counts is necessary.
-For this we need a( slightly) more complex model than model 1
-that would allow us to incorporate the non-Poissonian features of
-constitutive promoters directly into a master equation formulation.
-
-\subsubsection{Model 2 - Two-state promoter, RNAP bound or unbound.}
-This model was considered in, e.g.,~\cite{Phillips2015a} and ~\cite{Phillips2019}.
-Here transcription initiation proceeds from the bound to the unbound state,
-reflecting the microscopic reality that an RNAP that has begun to elongate
-a transcript is no longer available at the start site to begin another.
-The problem with this picture is that the Fano factor is
-\begin{align}
-    \nu = 1 -
-        \frac{r\rate{k}{P}{+}}
-            {\left(\rate{k}{P}{+} + \rate{k}{P}{-} + r\right)
-             \left(\gamma + \rate{k}{P}{+} + \rate{k}{P}{-} + r\right)},
-\end{align}
-which is always $<1$.
-To make contact with the experimental reality of $\nu>1$,
-we will have to do, at a minimum,
-the same corrections to Poisson behavior as in model 1 above.
-So while this model adds an appealing element of microscopic
-reality, we are forced to reject it as the additional complexity
-is unable to capture the phenomenology of interest.
-Obviously the promoter state does in fact proceed through cycles
-of RNAP binding, initiating, and elongating, but it seems that
-the super-Poissonian noise in mRNA copy number we want to
-model must be governed by other features of the system.
-
-\subsubsection{Model 3 - Three-state promoter, multistep
-transcription initiation and escape.}
-How might we remedy the deficits of model 2?
-It is known~\cite{DeHaseth1998} that once RNAP initially binds
-the promoter region, a multitude of distinct steps occur
-sequentially before RNAP finally escapes into the elongation
-phase. Perhaps adding some of this mechanistic detail might
-rescue model 2? The next simplest refinement of model 2 could
-consider open complex formation and promoter escape as separate
-steps rather than as a single effective step. In other words, we
-construct model 3 by adding a single extra state to model 2, and
-we will label the two RNAP-bound states as the closed and open
-complexes, despite the true biochemical details certainly being
-more complex. The authors of~\cite{Mitarai2015}, for instance,
-considered this model, although they added an additional
-repressor bound state and did not explicitly consider the limit
-with no repressor that we analyze here. Again, our goal here is
-not a complete accounting of all the relevant biochemical detail;
-this is an exploratory search for the important features our
-model needs to include.
-
-Unfortunately, as the authors of~\cite{Mitarai2015} hint, this
-model too has Fano factor $\nu<1$. This can be seen by starting
-with their Eq.~A1 for the Fano factor of the analogous model with
-repressor and taking the limit as repressor concentration goes to
-zero. After substantial algebra, the result is
-\begin{align}
-\nu = 1 - \frac{r k_O k_P^+}{\mathcal{Z}}
-\frac{k_P^+ + k_P^- + k_O + \gamma}
-        {\mathcal{Z} + \gamma(k_P^+ + k_P^- + k_O + \gamma) + \gamma^2},
-\end{align}
-where we defined $\mathcal{Z} = r(k_O + k_P^-) + k_P^+(k_O + r)$
-for notational tidiness. This is necessarily less than 1 for
-arbitrary rate constants.
-
-In fact, we suspect \textit{any} model in which transcription
-proceeds through a multistep cycle must necessarily have $\nu<1$.
-The intuitive argument compares the waiting time distribution to
-traverse the cycle with the waiting time for a Poisson promoter
-(model 1) with the same mean time. The latter is simply an
-exponential distribution. The former is a convolution of multiple
-exponentials, and intuitively should be more peaked with a
-smaller fractional width than an exponential with the same mean.\footnote{
-This can be made more precise with a result
-from~\cite{Stewart2007}, who showed that the convolution of
-multiple gamma distributions (of which the exponential
-distribution is a special case) is to a very good approximation
-also gamma distributed. Using their Eq.~2 for the distribution of
-the convolution, with shape parameters set to 1 to
-give exponential distributions, the total waiting time
-distribution has a ratio of variance to squared mean 
-$\sigma^2/\mu^2 = \sum_i k_i^2/\left(\sum_i k_i\right)^2$,
-where the $k_i$ are the rates of the individual steps. Clearly
-this is less than 1 and therefore the total waiting time
-distribution is narrower than the corresponding exponential.} A
-less disperse waiting time distribution means transcription
-initations are more uniformly distributed in time relative to a
-Poisson process. Hence the distribution of mRNA over a population
-of cells should be less variable compared to Poisson, giving
-$\nu<1$.\footnote{
-This last step, while intuitive, can be argued by analogy to
-photon statistics where antibunching gives rise to sub-Poissonian
-noise~\cite{Paul1982, Zou1990}. Although loopholes exist, we do
-not expect they apply for our problem. Nevertheless we refrain
-from elevating this cycle/sub-Poissonian equivalence to a ``theorem.''
-\mmnote{Antibunching is the ``obvious'' analogy for me but it'd
-be totally out of left field for most readers, any suggestions on
-a more comprehensible reference??}
-}
-
-Regardless of the merits of this model in describing the noise
-properties of constitutive transcription initation, it does not
-capture the dominant source of noise we want to include in our
-phenomenological model, so we must discard these details and
-search elsewhere.
-
-\subsubsection{Model 4 - Two-state promoter,
-``active'' and ``inactive'' states.}
-Inspired by~\cite{Razo-Mejia2020}, we revisit an analog of model
-2, but the interpretation of the two states is changed. Rather
-than explicitly viewing them as RNAP bound and unbound, we view
-them as ``active'' and ``inactive,'' which are able and unable to
-initiate transcripts, respectively. We are noncommittal as to the
-microscopic details of these states.
-
-One interpretation~\cite{Chong2014, Sevier2016} for the active
-and inactive states is that it represents the promoter's
-supercoiling state: transitions to the inactive state are caused
-by accumulation of positive supercoiling, which inhibits
-transcription, and transitions back to ``active'' are caused by
-gyrase or other topoisomerases relieving the supercoiling. This
-is an interesting possibility because it would mean the timescale
-for promoter state transitions is driven by topoisomerase
-kinetics, not by RNAP kinetics. From in vitro measurments, the
-former are suggested to be of order minutes~\cite{Chong2014}.
-Contrast this with model 2, where the state transitions are
-assumed to be governed by RNAP, which, assuming a copy number per
-cell of order $10^3$, has a diffusion-limited association rate
-$k_{on} \sim 10^2~\text{s}^{-1}$ to a target promoter. Combined
-with known $K_D$'s of order $\mu$M, this gives an RNAP
-dissociation rate $k_{off}$ of order $10^2$. As we will show
-below, however, there are some lingering puzzles with
-interpreting this supercoiling interpretation, so we leave it as
-a speculation and refrain from assigning definite physical
-meaning to the two states in this model.
-
-Intuitively one might expect that, since transcripts are produced
-as a Poisson process only when the promoter is in one of the two
-states in this model, transcription initiations should now be
-``bunched,'' in constrast to the ``anti-bunching'' of models 2
-and 3 above. One might further guess that this bunching would
-lead to super-Poissonian noise in the mRNA distribution over a
-population of cells. An honest calculation of the Fano factor produces
-\begin{align}
-\nu &= 1 + \frac{r k^-}{(k^+ + k^- + \gamma)(k^+ + k^-)},
-\end{align}
-which is strictly greater than 1, verifying the above intuition.
-Note we have dropped the $P$ label on the promoter switching rates to
-emphasize that these very likely do not represent kinetics of RNAP itself.
-This calculation can also be sidestepped by noting that the model
-is mathematically equivalent to the simple repression repression model
-from~\cite{Jones2014}, with states and rates relabeled and reinterpreted.
-
-How does this model compare to model 1 above?
-In model 1, all non-Poisson features of the mRNA
-distribution were handled as extrinsic corrections.
-By contrast, here the 3 parameter model is used to fit the full
-mRNA distribution as measured in mRNA FISH experiments.
-In essense, all variability in the mRNA distribution is regarded
-as ``intrinsic,'' arising either from stochastic initiation or
-from switching between the two coarse-grained promoter states.
-The advantage of this approach is that it fits neatly into the
-master equation picture, and the parameters thus inferred can be used as
-input for more complicated models with regulation by transcription factors.
-
-While this seems promising, there is a major drawback for our purposes
-which was already uncovered by the authors of~\cite{Razo-Mejia2020}:
-the statistical inference problem is nonidentifiable,
-in the sense described in Section 4.3 of~\cite{Gelman2013},
-meaning it is impossible to infer the
-parameters $r$ and $k^-$ from the FISH data of~\cite{Jones2014}
-(as shown in Fig.~S2 of~\cite{Razo-Mejia2020}).
-Rather, only the ratio $r/k^-$ could be inferred. In that work,
-the problem was worked around with an informative prior on the
-ratio $k^-/k^+$. That approach is unlikely to work here, as,
-recall, our entire goal in modeling constitutive expression is to
-use it as the basis for a yet more complicated model, when we add
-on repression. But adding more complexity to a model that is
-already poorly identified is a fool's errand,
-so we will explore one more potential model.
-
-\subsubsection{Model 5 - One-state promoter with explicit bursts.}
-This model is inspired by the failure mode of model 4.
-The key observation above was that, as found
-in~\cite{Razo-Mejia2020}, only two parameters, $k^+$ and the
-ratio $r/k^-$, could be directly inferred from the FISH data
-from~\cite{Jones2014}.
-So let us take this seriously and imagine a model where these are
-the only two model parameters. What would this model look like?
-
-To develop some intuition, consider model 4 in the limit
-$k^+ \ll k^- \lesssim r$, which is roughly satisfied by the
-parameters inferred in~\cite{Razo-Mejia2020}.
-In this limit, the system spends the majority of its time in the inactive state, occasionally becoming active and making a burst of transcripts.
-This should call to mind the well-known phenomenon of transcriptional bursting, as reported in, e.g.,~\cite{Golding2005,Chong2014,Sevier2016}\mmnote{should probably add some more cites here}.
-Let us make this correspondence more precise.
-The mean dwell time in the active state is $1/k^-$.
-While in this state, transcripts are produced at a rate $r$ per
-unit time. So on average, $r/k^-$ transcripts are produced before
-the system switches to the inactive state. Once in the inactive
-state, the system dwells there for an average time $1/k^+$ before
-returning to the active state and repeating the process. $r/k^-$
-resembles an average burst size, and $1/k^+$ resembles the time
-interval between burst events. More precisely, $1/k^+$ is the
-mean time between the end of one burst and the start of the next,
-whereas $1/k^+ + 1/k^-$ would be the mean interval between the
-start of two successive burst events, but in the limit $k^+ \ll
-k^-$, $1/k^+ + 1/k^- \approx 1/k^+$. Note that this limit ensures
-that the waiting time between bursts is approximately
-exponentially distributed, with mean set by the only timescale
-left in the problem,
-$1/k^+$.\footnote{
-If instead it were the case that $k^+ \sim k^-$, then the waiting time
-$1/k^+ + 1/k^-$ would have a peaked distribution, but this does
-not appear to be the case for any of datasets
-from~\cite{Jones2014}.}
-
-Let us now verify this intuition with a precise derivation
-to check that $r/k^-$ is in fact the mean burst size
-and to obtain the full burst size distribution.
-Consider first a constant, known dwell time $T$ in the active state.
-Transcripts are produced at a rate $r$ per unit time, so the
-number of transcripts $n$ produced during $T$ fits exactly
-the ``story'' for the Poisson distribution, i.e.,
-\begin{equation}
-    P(n\mid T) = \frac{(rT)^n}{n!} \exp(-rT).
-\end{equation}
-Since the dwell time $T$ is unobservable, we actually want $P(n)$,
-the dwell time distribution with no conditioning on $T$.
-Basic rules of probability theory tell us we can write $P(n)$
-in terms of $P(n\mid T)$ as
-\begin{equation}
-    P(n) =\int_0^\infty P(n\mid T) P(T) dT.
-\end{equation}
-But we know the dwell time distribution $P(T)$,
-which is exponentially distributed according to
-\begin{equation}
-    P(T) = k^- \exp(-T k^-),
-\end{equation}
-so $P(n)$ can be written as
-\begin{equation}
-    P(n) = k^- \frac{r^n}{n!}
-            \int_0^\infty T^n\exp[-(r + k^-)T]\,dT.
-\end{equation}
-A standard integral table shows
-$\int_0^\infty x^n e^{-ax}\,dx = n!/a^{n+1}$, so
-\begin{equation}
-    P(n) = \frac{k^- r^n}{(k^- + r)^{n+1}}
-        = \frac{k^-}{k^- + r}
-            \left(\frac{r}{k^- + r}\right)^n
-        = \frac{k^-}{k^- + r}
-            \left(1 - \frac{k^-}{k^- + r}\right)^n,
-\end{equation}
-which is exactly the geometric distribution with standard parameter
-$\theta\equiv k^-/(k^- + r)$ and domain
-$n \in \{0, 1, 2, \dots\}$
-(this is one of two common conventions for the geometric distribution).
-The mean of the geometric distribution, with this convention, is
-\begin{align}
-\langle n\rangle = \frac{1 - \theta}{\theta}
-        = \left(1 - \frac{k^-}{k^- + r}\right)
-                    \frac{k^- + r}{k^-}
-        = \frac{r}{k^-},
-\end{align}
-exactly as we guessed intuitively above.
-
-So in taking the limit $r,k^-\rightarrow\infty$, $r/k^-\equiv b$,
-we obtain a model which effectively has only a single promoter state,
-which initiates bursts at rate $k^+$
-(transitions to the active state, in the model 4 picture).
-The master equation for mRNA copy number $m$ takes the form
-\begin{align}
-\begin{split}
-\deriv{t}{p(m,t)} = & (m+1)\gamma p(m+1,t) - m\gamma p(m,t) \\
-        &+ \sum_{j=0}^{m-1} k^+ p(j,t) Geom(m-j;b)
-         - \sum_{j=m+1}^\infty k^+ p(m,t) Geom(j-m;b),
-\end{split}
-\end{align}
-where $Geom(n;b)$ is the geometric distribution with mean~$b$, i.e.,
-$Geom(n;b) = \frac{1}{1+b}\left(\frac{b}{1+b}\right)^n$
-(with domain over nonnegative integers as above).
-The first two terms are the usual mRNA degradation terms.
-The third term enumerates all ways the system can produce
-a burst of transcripts and arrive at copy number $m$,
-given that it had copy number $j$ before the burst.
-The fourth term allows the system to start with copy number $m$,
-then produce a burst and end with copy number $j$.
-In fact this last sum has trivial $j$ dependence and
-simply enforces normalization of the geometric distribution.
-Carrying it out we have
-\begin{equation}
-\begin{split}
-\deriv{t}{p(m,t)} = & (m+1)\gamma p(m+1,t) - m\gamma p(m,t) \\
-        &+ \sum_{j=0}^{m-1} k^+ p(j,t) Geom(m-j;b)
-            - k^+ p(m,t),
-\end{split}
-\end{equation}
-This improves on model 4 in that now the parameters are easily inferred,
-as we will see later, and have clean interpretations.
-The non-Poissonian features are attributed to the emprically
-well-established phenomenological picture of bursty transcription.
-
-The big approximation in going from model 4 to 5 is that a burst
-is produced instantaneously rather than over a finite time.
-If the true burst duration is not short compared to transcription factor
-kinetic timescales, this could be a problem in that mean burst size
-in the presense and absence of repressors could change,
-rendering parameter inferences from the constitutive case inappropriate.
-Let us make some simple estimates of this.
-
-Consider the time delay between the first and final RNAPs
-in a burst initiating transcription
-(\textit{not} the time to complete transcripts,
-which potentially could be much longer.)
-If this timescale is short compared to the typical
-search timescale of transcription factors, then all is well.
-The estimates from deHaseth et.\ al.~\cite{DeHaseth1998}
-put RNAP's diffusion-limited on rate around
-$\sim\text{few}\times10^{-2}~\text{nM}^{-1}~\text{s}^{-1}$
-and polymerase loading as high as $1~\text{s}^{-1}$.
-Then for reasonable burst sizes of $<10$, it is reasonable to guess that
-bursts might finish initiating on a timescale of tens of seconds or less
-(with another 30-60 sec to finish elongation, but that does not matter here).
-A transcription factor with typical copy number of order 10 (or less)
-would have a diffusion-limited association rate of order
-$(10~\text{sec})^{-1}$.
-Higher copy number TFs tend to have many binding sites over the genome,
-which should serve to pull them out of circulation and keep their
-effective association rates from rising too large.
-Therefore, there is \textit{perhaps} a timescale separation
-possible between transcription factor association rates and burst
-durations, but this assumption could very well break down, so we
-will have to keep it in mind when we infer repressor rates from
-the Jones et.\ al.\ FISH data later.
-
-In reflecting on these 5 models, the reader may feel that
-exploring a multitude of potential models just to return to a
-very minimal phenomenological model of bursty transcription may
-seem highly pedantic. But the purpose of the exercise was to
-examine a host of models from the literature and understand why
-they are insufficient, one way or another, for our purposes.
-Along the way we have learned that the detailed kinetics of RNAP
-binding and initiating transcription are probably irrelevant for
-setting the population distribution of mRNA.
-The timescales are simply too fast, and as we will see later
-in~\fig{fig:constit_post}, the noise seems to be governed by
-slower timescales. Perhaps in hindsight this is not surprising:
-intuitively, the degradation rate $\gamma$ sets the fundamental
-timescale for mRNA dynamics, and any other processes that
-substantially modulate the mRNA distribution should not differ
-from $\gamma$ by orders of magnitude.
-
-\subsection{Parameter inference for constitutive promoters}
-Having established that a one-state bursty model achieves a
-Goldilocks level of complexity, we test it more thoroughly on all
-the data from constitutive promoters from~\cite{Jones2014}.
-\fig{fig:constit_post} shows the results.
-\begin{figure}%[h!]
-\centering
-\includegraphics[width=0.9\textwidth]{../figures/fig2/fig2pt1.pdf}
-\caption{\textbf{Constitutive promoter posterior
-        inference and predictive checks.}
-        (A) shows the joint posterior density from MCMC sampling of model 5,
-        bursty promoter with negative binomially-distributed steady state.
-        (B) shows the ECDF of the observed population distribution
-        of mRNA transcripts under the control of a constitutive lacUV5 promoter.
-        Posterior predictive percentiles are shown as colored bands for
-        model (1), Poisson, and model (5), negative binomial.
-        }
-\label{fig:uv5_post}
-\end{figure}
-
-\begin{figure}%[h!]
-\centering
-\includegraphics[width=0.9\textwidth]{../figures/fig2/fig2pt2.pdf}
-\caption{\textbf{Constitutive promoter model comparison.}
-    (A) Joint posterior distributions for burst rate $k_i$ and
-    mean burst size $b$ for 18 promoters from Jones 2014.
-    Each contour indicates the 95\% highest posterior density
-    region for a particular promoter, i.e., on a plot analogous
-    to~\fig{fig:uv5_post}, the contour would enclose
-    approximately 95\% of the samples, and thus 95\% of the
-    probability mass.
-    Note that the vertical axis is shared with (B).
-    Panel (B) plots the burst rate $k_i$ vs.\ binding energy from energy
-    matrices from Brewster 2012.
-    The dotted line shows the predicted slope according to
-    $\ln k_i/\gamma \sim \text{constant} - \beta\Delta\epsilon_P$,
-    described in text.
-    }
-\label{fig:constit_post}
-\end{figure}
-\mmnote{Discuss puzzling comparison w/ Chong2014: if supercoiling is the thing, why are my burst sizes all the same but burst rates vary?
-And why is the duty cycle of my promoter so low, ie., bursts so short?
-COmpare their fig 7E; their $\beta/\alpha$ is my $k^+/k^-$.
-They have one or two genes with very small $\beta/\alpha$, does the \textit{galK} locus just happen to be that, or is there a deeper disagreement?
-Hard to say w/o more data.}
-
-\subsection{Transcription factor kinetics can be inferred from FISH measurements}
-Figure~3 presents inferred transcription factor rates, compares
-them to single-molecule measurements, and compares their ratios
-to gold-standard binding energies.
-\begin{figure}%[h!]
-\centering
-\includegraphics[width=0.9\textwidth]{../figures/fig3/fig3.ai}
-\caption{\textbf{Simple repression model comparison.}
-(A) depicts the cartoon of simple repression for which we infer parameters.
-(B) shows 50\% and 95\% contours of several 2D slices of the 9D posterior
-distribution, from fitting one unbinding rate for each operator (O1, O2, O3)
-and one binding rate for each aTc concentration (corresponding to an
-unknown mean repressor copy number).
-(C) compares ratios of our inferred unbinding rates with operator
-binding energy differences measured in Garcia \& Phillips 2011 (triangles)
-and Razo-Mejia et.\ al.\ 2018 (squares).
-(D) compares our inferred unbinding rates with single-molecule measurements
-of the same from Hammar et.\ at.\ 2014 \mmnote{Johan Elf's group}.
-Values are of similar order of magnitude,
-but a precise quantitative comparison unfortunately depends sensitively
-on the assumed mRNA lifetime, which was not measured.
-    }
-\label{fig3:kR_inferences}
-\end{figure}
-
-\begin{figure}%[h!]
-\centering
-\includegraphics[width=0.9\textwidth]{../figures/figSIxx/ppc_many_pooled.pdf}
-\caption{\textbf{Posterior predictive checks.}
-(A) duplicates the posterior from~\fig{fig3:kR_inferences}B.
-(B), (C), and (D) show samples from the posterior predictive distribution
-for each of the datasets used in the inference.
-Samples are sorted by operator and plotted separately simply for visual clarity.
-The unrepressed promoter, UV5, is shown with each as a reference point.
-\mmnote{
-Obviously the PPC is not perfect,
-but considering the relative simplicity of the model, I'm pleasantly surprised
-how well it worked and that the model was even identifiable at all.
-Interestingly, and perhaps not surprisingly, only 3 of the 9 datasets
-were identifiable if fit in isolation: otherwise,
-it was only by making assumptions that certain rates are equal
-across datasets was identifiability achieved.
-}
-    }
-\label{fig4:ppc}
-\end{figure}
-
-\section{Discussion and future work}
-\mmnote{
-Do all my approximations work simply because galK happens to have
-a low duty cycle of supercoiling, $\beta/\alpha$ in Chong2014
-picture? A potentially interesting test: put in the dozen or so
-Hernan constructs at some other locus, preferably one of the ones
-Chong identified as having a large duty cycle (large $\beta/\alpha$),
-i.e., the opposite limit of galK. Then just measure fold-changes.
-You don't even need to do FISH, although that would be even
-better. If their model is right, $\rho$ should be approximately
-set by the duty cycle of supercoiling, NOT by RNAP kinetics.
-\textit{This is a NICE prediction for future work.} \textit{cspE}
-is claimed nonessential, prediction is you drop HG reporter
-constructs in there and the inferred binding energies of all lacI
-operators shift the same amount, by about $\ln(3)\sim 1kT$. Not
-huge, but detectable, although might be confounding effects from
-moving to a new locus. With FISH measurements, $k^+$ and $k^-$ in
-Fig 2, model 4 should now be separately inferrable if Chong2014
-measurements of $\beta/\alpha$ are correct. This would require an
-obnoxious amount of cloning, but once you did that, the
-measurements would be same old stuff that RP lab does.}
-
-%%%%%%%%%%%%%%%%%%%%% APPENDICES %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\appendix
-
-\section{Derivations of constitutive promoter models}
-\subsection{Model 2}
-\mmnote{Obtain Fano factor from CME.}
-\subsection{Model 3}
-\mmnote{Obtain Fano factor by taking limit from Mitarai et al's expression.}
-\subsection{Model 4}
-\mmnote{Obtain Fano factor by relabeling cartoon from Jones2014.}
-
-\section{Generating function solution for bursty promoters}
-
-\subsection{Constitutive promoter with bursts}
-
-\subsubsection{From master equation to generating function}
-
-We would like to derive the full probability distribution of
-mRNA copy number per cell for a model combining simple repression
-with a promoter featuring bursty transcription initiation.
-Before tackling mRNA and repression together, let us
-just consider the bursty consititutive promoter,
-model 5 in~\fig{fig:constit_cartoons}
-\mmnote{replicating Charlotte's notes; she says it's worked out
-in~\cite{Paulsson2000}, but I don't see it, at least not in notation I can
-comprehend. Need to find a cite for this,
-or maybe just cite Charlotte, personal comm, since that's where I actually got it\dots}.
-The master equation of interest is
-\begin{align}
-\deriv{t}p(m,t) = (m+1)\gamma p(m+1,t) - m\gamma p(m,t) - r p(m,t)
-        + r \sum_{m^\prime=0}^m G_{m-m^\prime}(\theta) p(m^\prime,t),
-\label{eq:1state_unreg_003}
-\end{align}
-where $G_{k}(\theta)$ is the geometric distribution defined as
-\begin{align}
-G_{k}(\theta) = \theta(1 - \theta)^k, \, k\in\{0,1,2,\dots\}.
-\end{align}
-With this convention, the mean burst size $b = (1-\theta)/\theta$.
-$\gamma$ and $r$ are mRNA degradation rates and transcription burst
-initiation rates, resp. The last term represents all ways the system could
-end up with $m$ mRNAs, having started the burst with $m^\prime$. Define
-$\lambda = r/\gamma$ and nondimensionalize time by $\gamma$, giving
-\begin{align}
-\deriv{t}p(m,t) = (m+1)p(m+1,t) - m p(m,t) - \lambda p(m,t)
-        + \lambda \sum_{m^\prime=0}^m G_{m-m^\prime}(\theta) p(m^\prime,t),
-\end{align}
-The probability generating function is defined as
-\begin{align}
-F(z,t) = \sum_{m=0}^\infty z^m p(m,t).
-\end{align}
-Multiply both sides of the CME by $z^m$ and sum over all $m$ to get
-\begin{align}
-\pderiv[F]{t} = \sum_{m=0}^\infty z^m
-\left(
-(m+1)p(m+1,t) - m p(m,t) - \lambda p(m,t)
-+ \lambda \sum_{m^\prime=0}^m G_{m-m^\prime}(\theta) p(m^\prime,t)
-\right).
-\end{align}
-The first two terms can be rewritten as derivatives of the
-generating function, i.e.,
-\begin{align}
-\sum_{m=0}^\infty z^m (m+1)p(m+1,t)
-=
-\pderiv{z} \sum_{m=0}^\infty z^{m+1} p(m+1,t)
-=
-\pderiv[F]{z}
-\end{align}
-and
-\begin{align}
-\sum_{m=0}^\infty z^m mp(m,t)
-=
-z\pderiv{z} \sum_{m=0}^\infty z^m p(m,t)
-=
-z\pderiv[F]{z}.
-\end{align}
-The final term is the most trouble. The trick is to reverse
-the default order of the sums, as in
-\begin{align}
-&\lambda \sum_{m=0}^\infty \sum_{m^\prime=0}^m
-        z^m \theta(1 - \theta)^{m-m^\prime} p(m^\prime,t)
-\\=
-&\lambda\theta \sum_{m^\prime=0}^\infty z^{m^\prime} p(m^\prime,t)
-        \sum_{m=m^\prime}^\infty
-        z^{m-m^\prime} (1 - \theta)^{m-m^\prime}
-\\=
-&\frac{\lambda\theta F}{1 - (1 - \theta)z}.
-\end{align}
-This works because after reordering the sums, the sum on $m$ is just
-a geometric series. Its result is independent of $m^\prime$, so the sum on
-$m^\prime$ reduces to the definition of the generating function.
-Putting these results together, the PDE for the generation function is
-\begin{align}
-\pderiv[F]{t} = \pderiv[F]{z} - z\pderiv[F]{z} - \lambda F
-        + \frac{\lambda\theta F}{1-z(1-\theta)}.
-\end{align}
-Changing variables to $\xi=1-\theta$ and simplifying gives
-\begin{align}
-\pderiv[F]{t} + (z - 1) \pderiv[F]{z} = \frac{(z-1)\xi}{1-z\xi}\lambda F.
-\label{eq:1state_unreg_015}
-\end{align}
-\subsubsection{Stead-state}
-At steady-state, the PDE reduces to the ODE
-\begin{align}
-\deriv[F]{z} = \frac{\xi}{1-z\xi}\lambda F,
-\end{align}
-which we can integrate as
-\begin{align}
-\int \frac{dF}{F} = \int \frac{\lambda\xi dz}{1-\xi z}.
-\end{align}
-The initial conditions for generating functions can be subtle and confusing.
-The key fact follows from the definition
-$F(z,t) = \sum_m z^m p(m,t)$,
-normalization requires that
-$F(z=1^-,t) = \sum_m p(m,t) = 1$.\footnote{
-Sometimes the generating function may be undefined \textit{at} $z=1$ but the
-limit still holds. Also many authors change variables from $z$ to other
-things, so one must keep track of how the normalization condition transforms.
-}
-Doing the integrals (and producing constant $c$) gives
-\begin{align}
-\ln F &= -\lambda \ln(1-\xi z) + c
-\\
-F &= \frac{c}{(1-\xi z)^\lambda}.
-\end{align}
-Only one choice for $c$ can satisfy initial conditions, producing
-\begin{align}
-F(z) = \left(\frac{1-\xi}{1-\xi z}\right)^\lambda
-        = \left(\frac{\theta}{1 - z(1-\theta)}\right)^\lambda,
-\end{align}
-which is exactly the negative binomial's generating function, as expected.
-
-\subsection{Adding repression}
-\subsubsection{Deriving the generating function for mRNA distribution}
-
-Let us move from a one-state promoter to a two-state promoter, where one state
-has repressor bound and the other produces transcriptional bursts as above.
-A schematic of this model is shown in~\fig{fig3:kR_inferences}.
-Although now we have an equation for each promoter state, otherwise
-the master equation reads similarly to the one-state case, except with
-additional terms corresponding to transitions between promoter states, namely
-\begin{align}
-\deriv{t}p_R(m,t) =& k_R^+ p_A(m,t) - k_R^- p_R(m,t)
-        + (m+1)\gamma p_R(m+1,t) - m\gamma p_R(m,t)
-\\
-\begin{split}
-\deriv{t}p_A(m,t) =& - k_R^+ p_A(m,t) + k_R^- p_R(m,t)
-        + (m+1)\gamma p_A(m+1,t) - m\gamma p_A(m,t) 
-\\
-&- k_i p_A(m,t) + k_i \sum_{m^\prime=0}^m \theta(1-\theta)^{m-m^\prime} p_A(m^\prime,t),
-\end{split}
-\end{align}
-where $p_R(m,t)$ is the probability of the system having $m$ mRNA copies
-and having repressor bound to the promoter at time $t$, and $p_A$ is an
-analogous probability to find the promoter without repressor bound.
-$k_R+$ and $k_R^-$ are, respectively, the rates at which repressors
-bind and unbind to and from the promoter, and
-$\gamma$ is the mRNA degradation rate.
-$k_i$ is the rate at which bursts initiate, and as before,
-the geometric distribution of burst sizes has mean $b=(1-\theta)/\theta$.\footnote{
-$\theta$ can be viewed as a Bernoulli trial probability: after each transcript
-in a burst is produced, $\theta$ is the probability that the burst
-terminates and $(1-\theta)$ is the probability that the burst continues.
-}
-
-Interestingly, it turns out that this problem maps exactly onto the
-three-stage promoter model considered by
-Shahrezaei and Swain in~\cite{Shahrezaei2008}, with relabelings.
-Their approximate solution for protein distributions amounts to
-the same approximation we make here in regarding the duration
-of mRNA synthesis bursts as instantaneous, so their solution
-for protein distributions also solves our problem of mRNA distributions.
-Let us examine the analogy more closely.
-They consider a two-state promoter, as we do here, but they model
-mRNA as being produced one at a time and degraded, with rates
-$v_0$ and $d_0$.
-Then they model translation as occuring with rate $v_1$,
-and protein degradation with rate $d_1$.
-Now consider the limit where $v_1, d_0\rightarrow\infty$ with
-their ratio $v_1/d_0$ held constant.
-$v_1/d_0$ resembles the average burst size of translation from a single
-mRNA: these are the rates of two Poisson processes that compete over a
-transcript, which matches the story of geometrically distributed burst sizes.
-So after taking this limit, on timescales slow compared to $v_1$ and $d_0$,
-it appears that transcription events fire at rate $v_0$ and produce a
-geometrically distributed burst of translation of mean size $v_1/d_0$,
-which intuitively matches the story we have told above for mRNA
-with variables relabeled.
-
-To verify this intuitively conjectured mapping between our problem and
-the solution in~\cite{Shahrezaei2008}, we continue with a careful
-solution for the mRNA distribution using probability generating functions,
-following the ideas sketched in~\cite{Shahrezaei2008}.
-It is natural to nondimensionalize rates in the problem by $\gamma$,
-or equivalently, this amounts to measuring time in units of $\gamma^{-1}$.
-We are also only interested in steady state, so we set the time
-derivatives to zero, giving
-\begin{align}
-0 =& k_R^+ p_A(m) - k_R^- p_R(m) + (m+1) p_R(m+1) - m p_R(m)
-\\
-\begin{split}
-0 =& - k_R^+ p_A(m) + k_R^- p_R(m) + (m+1) p_A(m+1) - m p_A(m) 
-\\
-&- k_i p_A(m) + k_i \sum_{m^\prime=0}^m \theta(1-\theta)^{m-m^\prime} p_A(m^\prime),
-\end{split}
-\end{align}
-
-The probability generating function is defined as before in the
-constitutive case, except now we must introduce a generating function
-for each promoter state,
-\begin{align}
-f_A(z) = \sum_{m=0}^\infty z^m p_A(m),
-\;
-f_R(z) = \sum_{m=0}^\infty z^m p_R(m).
-\end{align}
-Our real objective is the generating function $f(z)$ that generates
-the mRNA distribution $p(m)$, independent of what state the promoter is in.
-But since $p(m) = p_A(m) + p_R(m)$, it follows too that $f(z) = f_A(z) + f_R(z)$.
-
-So as before we multiply both equations by $z^m$ and sum over all $m$.
-Each individual term transforms exactly as did an analogous term in
-the constitutive case, so the coupled ODEs for the generating functions read
-\begin{align}
-0 =& k_R^+ f_A(z) - k_R^- f_R(z) + \pderiv{z} f_R(z) - z \pderiv{z} f_R(z)
-\\
-\begin{split}
-0 =&  - k_R^+ f_A(z) + k_R^- f_R(z) + \pderiv{z} f_A(z) - z \pderiv{z} f_A(z)
-\\
-&- k_i f_A(z) + k_i \frac{\theta}{1-z(1-\theta)} f_A(z),
-\end{split}
-\end{align}
-and after changing variables $\xi = 1 - \theta$ as before
-and rearranging, we have
-\begin{align}
-0 &= k_R^+ f_A(z) - k_R^- f_R(z) + (1-z) \pderiv{z} f_R(z)
-\\
-0 &=  - k_R^+ f_A(z) + k_R^- f_R(z) + (1 - z) \pderiv{z} f_A(z)
-+ k_i \frac{(z-1)\xi}{1-z\xi} f_A(z),
-\end{align}
-We can transform this problem from two coupled first-order ODEs to a
-single second-order ODE by solving for $f_A$ in the first and plugging
-into the second, giving
-\begin{align}
-\begin{split}
-0 = (1&-z) \pderiv[f_R]{z}
-+ \frac{1-z}{k_R^+}
-        \left(k_R^- \pderiv[f_R]{z} + \pderiv[f_R]{z} +(z-1) \psecderiv[f_R]{z}\right)
-\\
-&+ \frac{k_i}{k_R^+} \frac{(z-1)\xi}{1-z\xi}
-        \left(k_R^- f_R + (z-1) \pderiv[f_R]{z}\right),
-\end{split}
-\end{align}
-where, to reduce notational clutter, we have dropped the explicit $z$
-dependence of $f_A$ and $f_R$. Simplifying we have
-\begin{align}
-0 = \psecderiv[f_R]{z}
-        - \left(\frac{k_i\xi}{1-z\xi}
-                + \frac{1 + k_R^- + k_R^+}{1-z}
-        \right)\pderiv[f_R]{z}
-        + \frac{k_i k_R^- \xi}{(1-z\xi)(1-z)}f_R.
-\end{align}
-This can be recognized as the hypergeometric differential equation,
-with singularities at $z=1$, $z=\xi^{-1}$, and $z=\infty$. The latter
-can be verified by a change of variables from $z$ to $x=1/z$,
-being careful with the chain rule, and noting that $z=\infty$ is a
-singular point if and only if $x=1/z=0$ is a singular point.
-
-The standard form of the hypergeometric differential equation has its
-singularities at 0, 1, and $\infty$, so to take advantage of the
-standard form solutions to this ODE, we first need to transform
-variables to put it into a standard form.
-However, this is subtle. While any such transformation should work in
-principle, the solutions are expressed most simply in the neighborhood
-of $z=0$, but the normalization condition that we need to enforce
-corresponds to $z=1$. The easiest path, therefore, is to find a change
-of variables that maps
-1 to 0, $\infty$ to $\infty$, and $\xi^{-1}$ to 1.
-This is most intuitively done in two steps.
-
-First map the $z=1$ singularity to 0 by the change of variables $v=z-1$, giving
-\begin{align}
-0 = \psecderiv[f_R]{v}
-        + \left(\frac{k_i\xi}{(1+v)\xi - 1}
-                + \frac{1 + k_R^- + k_R^+}{v}
-        \right)\pderiv[f_R]{v}
-        + \frac{k_i k_R^- \xi}{((1+v)\xi - 1)v}f_R.
-\end{align}
-Now two singularities are at $v=0$ and $v=\infty$.
-The third is determined by $(1+v)\xi -1 = 0$, or $v=\xi^{-1} - 1$.
-We want another variable change that maps this third singularity to 1
-(without moving 0 or infinity). Changing variables again to
-$w=\frac{v}{\xi^{-1} - 1} = \frac{\xi}{1-\xi} v$ fits the bill.
-In other words, the combined change of variables
-\begin{align}
-w = \frac{\xi}{1-\xi} (z-1)
-\end{align}
-maps $z = \{1, \xi^{-1}, \infty\}$ to $w =\{0, 1, \infty\}$ as desired.
-Plugging in, being mindful of the chain rule and noting
-$(1 + v)\xi - 1 = (1 - \xi)(w - 1)$ gives
-\begin{align}
-0 = \left(\frac{\xi}{1-\xi}\right)^2 \psecderiv[f_R]{w}
-+ \left(
-        \frac{\xi k_i}{(1-\xi)(w-1)} + \frac{\xi(1 + k_R^- + k_R^+)}{(1-\xi)w}
-\right) \frac{\xi}{1-\xi} \pderiv[f_R]{w}
-+ \frac{k_i k_R^- \xi^2}{(1-\xi)^2 w(w-1)}f_R.
-\end{align}
-This is close to the standard form of the hypergeometric differential
-equation, and some cancellation and rearrangement gives
-\begin{align}
-0 = w(w-1)\psecderiv[f_R]{w}
-+ \left(k_i w + (1 + k_R^- + k_R^+)(w-1)\right) \pderiv[f_R]{w}
-+ k_i k_R^- f_R.
-\end{align}
-and a little more algebra produces
-\begin{align}
-0 = w(1-w)\psecderiv[f_R]{w}
-+ \left(1 + k_R^- + k_R^+
-        - (1 + k_i + k_R^- + k_R^+)w
-\right) \pderiv[f_R]{w}
-- k_i k_R^- f_R,
-\end{align}
-which is the standard form. From this we can read off the solution
-in terms of hypergeometric functions ${_2F_1}$ from
-any standard source, e.g.~\cite{Abramowitz1964}, and identify the
-conventional parameters in terms of our model parameters.
-We want the general solution in the neighborhood of $w=0$ ($z=1$),
-which for a homogeneous linear second order ODE must be a sum of two
-linearly independent solutions.
-More precisely,
-\begin{align}
-f_R(w) = C_1 {_2F_1}(\alpha, \beta, \gamma; w)
-+ C_2 w^{1-\gamma}{_2F_1}(1+\alpha-\gamma, 1+\beta-\gamma, 2-\gamma; w)
-\end{align}
-with parameters determined by
-\begin{align}
-\begin{split}
-\alpha\beta &= k_i k_R^-
-\\
-1+\alpha+\beta &= 1+k_i+k_R^-+k_R^+
-\\
-\gamma &= 1 + k_R^- + k_R^+
-\end{split}
-\end{align}
-and constants $C_1$ and $C_2$ to be set by boundary conditions.
-Solving for $\alpha$ and $\beta$, we find\footnote{
-Note that $\alpha$ and $\beta$ are interchangeable in the definition of
-${_2F_1}$ and differ only in the sign preceeding the radical.
-}
-\begin{align}
-\begin{split}
-\alpha &= \frac{1}{2}
-\left(k_i+k_R^-+k_R^+ + \sqrt{(k_i+k_R^-+k_R^+)^2 - 4k_i k_R^-}\right)
-\\
-\beta &= \frac{1}{2}
-\left(k_i+k_R^-+k_R^+ - \sqrt{(k_i+k_R^-+k_R^+)^2 - 4k_i k_R^-}\right)
-\\
-\gamma &= 1 + k_R^- + k_R^+.
-\end{split}
-\end{align}
-Since the normalization condition requires that $f_R$ be finite at $w=0$,
-we can immediately set $C_2=0$ to discard the second solution.
-This is because all the rate constants are strictly positive,
-so $\gamma>1$ and therefore $w^{1-\gamma}$ blows up as $w\rightarrow0$.
-Now that we have $f_R$, we would like to find the generating function
-for the mRNA distribution, $f(z) = f_A(z) + f_R(z)$.
-We can recover $f_A$ from our solution for $f_R$, namely
-\begin{align}
-f_A(z) = \frac{1}{k_R^+}\left(k_R^- f_R(z) + (z-1) \pderiv[f_R]{z}\right)
-\end{align}
-or
-\begin{align}
-f_A(w) = \frac{1}{k_R^+}\left(k_R^- f_R(w) + w \pderiv[f_R]{w}\right),
-\end{align}
-where in the second line we transformed our original relation between
-$f_R$ and $f_A$ to our new, more convenient, variable $w$.
-Plugging our solution for $f_R(w) = C_1{_2F_1}(\alpha, \beta, \gamma; w)$
-into $f_A$, we will require the differentiation rule for ${_2F_1}$,
-which tells us
-\begin{align}
-\pderiv[f_R]{w} = C_1\frac{\alpha\beta}{\gamma}
-                {_2F_1}(\alpha+1, \beta+1, \gamma+1; w),
-\end{align}
-from which it follows that
-\begin{align}
-f_A(w) = \frac{C_1}{k_R^+}
-\left(
-k_R^- {_2F_1}(\alpha, \beta, \gamma; w)
-+ w\frac{\alpha\beta}{\gamma} {_2F_1}(\alpha+1, \beta+1, \gamma+1; w)
-\right)
-\end{align}
-and therefore
-\begin{align}
-f(w) = C_1\left(1 + \frac{k_R^-}{k_R^+}\right)
-        {_2F_1}(\alpha, \beta, \gamma; w)
-+ w \frac{C_1}{k_R^+} \frac{\alpha\beta}{\gamma}
-        {_2F_1}(\alpha+1, \beta+1, \gamma+1; w).
-\end{align}
-To proceed, we need one of the (many) useful identities known for
-hypergeometric functions, in particular
-\begin{align}
-w\frac{\alpha\beta}{\gamma} {_2F_1}(\alpha+1, \beta+1, \gamma+1; w)
-=
-(\gamma-1)\left(
-{_2F_1}(\alpha, \beta, \gamma-1; w) - {_2F_1}(\alpha, \beta, \gamma; w)
-\right).
-\end{align}
-Substituting this for the second term in $f(w)$, we find
-\begin{align}
-f(w) = \frac{C_1}{k_R^+}
-\left[
-        \left(k_R^+ + k_R^-\right)
-        {_2F_1}(\alpha, \beta, \gamma; w)
-+ (\gamma-1)\left(
-        {_2F_1}(\alpha, \beta, \gamma-1; w) - {_2F_1}(\alpha, \beta, \gamma; w)
-        \right)
-\right],
-\end{align}
-and since $\gamma-1 = k_R^+ + k_R^-$, the first and third terms cancel,
-leaving only
-\begin{align}
-f(w) = C_1\frac{k_R^+ + k_R^-}{k_R^+} {_2F_1}(\alpha, \beta, \gamma-1; w).
-\end{align}
-Now we enforce normalization, demanding $f(w=0) = f(z=1) = 1$.
-${_2F_1}(\alpha, \beta, \gamma-1; 0) = 1$, so we must have
-$C_1 = k_R^+ / (k_R^+ + k_R^-)$ and consequently
-\begin{align}
-f(w) =  {_2F_1}(\alpha, \beta, k_R^+ + k_R^-; w).
-\end{align}
-Recalling that the mean burst size $b = (1-\theta)/\theta = \xi/(1-\xi)$
-and $w = \frac{\xi}{1-\xi} (z-1) = b (z-1)$,
-we can transform back to the original variable $z$ to find the tidy result
-\begin{align}
-f(z) =  {_2F_1}(\alpha, \beta, k_R^+ + k_R^-; b(z-1)),
-\end{align}
-with $\alpha$ and $\beta$ given above by
-\begin{align}
-\begin{split}
-\alpha &= \frac{1}{2}
-\left(k_i+k_R^-+k_R^+ + \sqrt{(k_i+k_R^-+k_R^+)^2 - 4k_i k_R^-}\right)
-\\
-\beta &= \frac{1}{2}
-\left(k_i+k_R^-+k_R^+ - \sqrt{(k_i+k_R^-+k_R^+)^2 - 4k_i k_R^-}\right).
-\end{split}
-\end{align}
-Finally we are in sight of the original goal. We can generate the
-steady-state probability distribution of interest by
-differentiating the generating function,
-\begin{align}
-p(m) = m! \left.\frac{\partial^m}{\partial z^m} f(z) \right|_{z=0},
-\end{align}
-which follows easily from its definition. Some contemplation reveals
-that repeated application of the derivative rule used above will produce
-products of the form $\alpha(\alpha+1)(\alpha+2)\cdots(\alpha+m-1)$ in
-the expression for $p(m)$ and similarly for $\beta$ and $\gamma$. These
-resemble ratios of factorials, but since $\alpha$, $\beta$, and $\gamma$
-are not necessarily integer, we should express the ratios using gamma
-functions instead. More precisely, one finds
-\begin{align}
-p(m) = \frac{
-        \Gamma(\alpha + m)\Gamma(\beta + m)\Gamma(k_R^+ + k_R^-)
-        }
-        {
-        \Gamma(\alpha)\Gamma(\beta)\Gamma(k_R^+ + k_R^- + m)
-        }
-\frac{b^m}{m!}{_2F_1}(\alpha+m, \beta+m, k_R^++k_R^-+m; -b)
-\label{eq:p_m_bursty+rep}
-\end{align}
-which is finally the probability distribution we sought to derive.
-
-\subsubsection{Moments}
-We can verify the correctness of~\eq{eq:p_m_bursty+rep}
-by comparison with Gillespie simulation.
-We can also easily compute moments of the distribution from
-the generating function, and check these against raw moments computed
-directly from the master equation.
-
-The $j$-th raw moment of the distribution can be computed from the
-generating function from the formula
-\begin{align}
-\langle m^j \rangle
-= \left. \left(z \frac{\partial}{\partial z}\right)^j f(z)\right|_{z=1}.
-\end{align}
-Note that \textit{probabilities} are generated by differentiating
-and evaluating at $z=0$, while \textit{moments} are generated by
-differentiating and evaluating at $z=1$. This also aligns with the
-normalization constraint, which can be thought of as the zeroth raw moment.
-
-The mean mRNA is given by
-\begin{align}
-\langle m \rangle = \left.z \pderiv[f]{z} \right|_{z=1}
-= \frac{\alpha\beta b}{k_R^+ + k_R^-}
-= k_i b\frac{k_R^-}{k_R^+ + k_R^-},
-\end{align}
-which is intuitive and appealingly simple:
-the mean mRNA is just the mean mRNA in the absense of repression
-($k_i b$, the burst rate time mean burst size)
-times the fraction of time the repressor is bound, $k_R^-/(k_R^+ + k_R^-)$.
-
-To compute the variance, let us first compute
-\begin{align}
-\begin{split}
-\langle m^2\rangle
-&= \left .z\pderiv{z}\left(z\pderiv[f]{z}\right) \right|_{z=1}
-= \left. z\left(\pderiv[f]{z} + z\psecderiv[f]{z}\right)\right|_{z=1}
-\\
-&= \frac{\alpha\beta b}{k_R^+ + k_R^-}
-        + \frac{\alpha(\alpha+1)\beta(\beta+1) b^2}
-                {(k_R^+ + k_R^-)(k_R^+ + k_R^- + 1)}
-\\
-&= \frac{k_i b k_R^-}{k_R^+ + k_R^-}
-        + \frac{k_i k_R^- b^2 (1 + k_i + k_R^+ + k_R^- + k_i k_R^-)}
-                {(k_R^+ + k_R^-)(k_R^+ + k_R^- + 1)},
-\end{split}
-\end{align}
-from which the variance is
-\begin{align}
-\begin{split}
-var(m) = \langle m^2\rangle - \langle m\rangle^2
-&= \frac{k_i b k_R^-}{k_R^+ + k_R^-}
-\left(
-1 + \frac{b(1 + k_i + k_R^+ + k_R^- + k_i k_R^-)} {k_R^+ + k_R^- + 1}
-- \frac{k_i b k_R^-}{k_R^+ + k_R^-}
-\right)
-\\
-&= k_i b\frac{k_R^-}{k_R^+ + k_R^-}
-\left(
-1 + b + k_i b\frac{1 + k_R^-} {k_R^+ + k_R^- + 1}
-- \frac{k_i b k_R^-}{k_R^+ + k_R^-}
-\right)
-\\
-&= k_i b\frac{k_R^-}{k_R^+ + k_R^-}
-\left(
-1 + b + k_i b\frac{k_R^+}{(k_R^+ + k_R^-)(k_R^+ + k_R^- + 1)}
-\right).
-\end{split}
-\end{align}
-The Fano factor is the quantity in parentheses. This can be interpreted as
-the variability arising from the negative binomial distribution of the
-unregulated promoter $(1+b)$ plus additional noise arising from repression.
-
-\mmnote{Still to writeup: moment derivation direct from CME to compare}
-
-\subsection{Numerical considerations and recursion formulas}
-\subsubsection{Generalities}
-We would like to carry out Bayesian parameter inference on FISH data
-from~\cite{Jones2014}, using~\eq{eq:p_m_bursty+rep} as our
-likelihood. This requires accurate (and preferably fast)
-numerical evaluation of the hypergeometric function ${_2F_1}$,
-which is a notoriously hard problem~\cite{Pearson2017, Gil2007},
-and our particular needs here present an especial challenge as we show below.
-
-The hypergeometric function is defined by its Taylor series as
-\begin{align}
-{_2F_1}(a,b,c;z) 
-= \sum_{l=0}^\infty
-\frac{\Gamma(a + l)\Gamma(b + l)\Gamma(c)}
-        {\Gamma(a)\Gamma(b)\Gamma(c + l)}
-\frac{z^l}{l!}
-\end{align}
-for $|z|<1$, and by analytic continuation elsewhere.
-If $z\lesssim1/2$ and $\alpha$ and $\beta$ are not too large
-(absolute value below 20 or 30),
-then the series converges quickly and an accurate numerical representation is
-easily computed by truncating the series after a reasonable number of terms.
-Unfortunately, we need to evaluate ${_2F_1}$ over mRNA copy numbers fully out
-to the tail of the distribution, which can easily reach 50, possibly 100.
-From~\eq{eq:p_m_bursty+rep}, this means evaluating ${_2F_1}$
-repeatedly for values of $a$, $b$, and $c$ spanning the full range
-from $\mathcal{O}(1)$ to $\mathcal{O}(10^2)$,
-even if $\alpha$, $\beta$, and $\gamma$
-in~\eq{eq:p_m_bursty+rep} are small,
-with the situation even worse if they are not small.
-A naive numerical evaluation of the series definition will be
-prone to overflow and, if any of $a,b,c<0$, then some successive terms in the
-series have alternating signs which can lead to catastrophic cancellations.
-
-One solution is to evaluate ${_2F_1}$ using arbitrary precision arithmetic
-instead of floating point arithmetic,
-e.g., using the \texttt{mpmath} library in Python.
-This is accurate but incredibly slow computationally.
-To quantify how slow, we found that
-evaluating the likelihood defined by~\eq{eq:p_m_bursty+rep} $\sim50$ times
-(for a typical dataset of interest from~\cite{Jones2014},
-with $m$ values spanning 0 to $\sim50$)
-using arbitrary precision arithmetic is 100-1000 fold slower than
-evaluating a negative binomial likelihood for the corresponding
-constitutive promoter dataset.
-
-To claw back $\gtrsim30$ fold of that slowdown, we can exploit
-one of the many catalogued symmetries involving ${_2F_1}$.
-The solution involves recursion relations originally explored by Gauss,
-and studied extensively in~\cite{Pearson2017, Gil2007}.
-They are sometimes known as contiguous relations and relate the values
-of any set of 3 hypergeometric functions whose arguments differ by integers.
-To rephrase this symbolically, consider a set of hypergeometric functions
-indexed by an integer $n$,
-\begin{align}
-f_n = {_2F_1}(a+\epsilon_i n, b+\epsilon_j n, c+\epsilon_k n; z),
-\end{align}
-for a fixed choice of $\epsilon_i, \epsilon_j, \epsilon_k \in \{0,\pm 1\}$
-(at least one of $\epsilon_i, \epsilon_j, \epsilon_k$ must be nonzero,
-else the set of $f_n$ would contain only a single element).
-Then there exist known recurrence relations of the form
-\begin{align}
-A_n f_{n-1} + B_n f_{n} + C_n f_{n+1} = 0,
-\end{align}
-where $A_n, B_n$, and $C_n$ are some functions of $a,b,c$, and $z$.
-In other words, for fixed $\epsilon_i, \epsilon_j, \epsilon_k, a, b,$ and $c$,
-if we can merely evaluate ${_2F_1}$ twice, say for $n^\prime$ and $n^\prime-1$,
-then we can easily and rapidly generate values for arbitrary $n$.
-
-This provides a convenient solution for our problem: we need repeated
-evaluations of ${_2F_1}(a+m, b+m, c+m; z)$
-for fixed $a,b$, and $c$ and many integer values of $m$.
-They idea is that we can use arbitrary precision arithmetic to evaluate
-${_2F_1}$ for just two particular values of $m$ and then generate
-${_2F_1}$ for the other 50-100 values of $m$ using the recurrence
-relation.\footnote{
-There are even more sophisticated ways of utilizing the recurrence
-relations that might have netted another factor of 2 speed-up, and
-possibly as much as a factor of 10, but the method described here had
-already reduced the computation time to an acceptable
-$\mathcal{O}(\text{1 min})$, so these more sophisticated approaches did
-not seem worth the time to pursue.
-}
-
-However, there two further wrinkles.
-The first is that
-a naive application of the recurrence relation is numerically unstable.
-Roughly, this is because the three term recurrence relations,
-like second order ODEs, admit two linearly independent solutions.
-In a certain eigenbasis, one of these solutions dominates the other
-as $n\rightarrow\infty$, and as $n\rightarrow-\infty$,
-the dominance is reversed.
-If we fail to work in this eigenbasis, our solution of the recurrence relation
-will be a mixture of these solutions and rapidly accumulate numerical error.
-For our purposes, it suffices to know that the authors of~\cite{Gil2007}
-derived the numerically stable solutions (so-called \textit{minimal solutions})
-for several possible choices of $\epsilon_i, \epsilon_j, \epsilon_k$.
-Running the recurrence in the proper direction using a minimal solution
-is numerically robust and can be done entirely in floating point arithmetic, 
-so that we only need to evaluate ${_2F_1}$ with arbitrary precision arithmetic
-to generate the seed values for the recursion.
-
-The second wrinkle is a corollary to the first.
-The minimal solutions are only minimal for certain ranges of the argument $z$,
-and not all of the 26 possible recurrence relations
-have minimal solutions for all $z$.
-This can be solved by using one of the many transformation formulae for
-${_2F_1}$ to convert to a different recurrence relation that has
-a minimal solution over the required domain of $z$, although
-this can require some trial and error to find the right transformation,
-the right recurrence relation, and the right minimal solution.
-
-\subsubsection{Particulars}
-Let us now demonstrate these generalities for our problem of interest.
-In order to evaluate the probability distribution of our
-model,~\eq{eq:p_m_bursty+rep}, we need to evaluate hypergeometric functions
-of the form ${_2F_1}(\alpha+m, \beta+m, \gamma+m; -b)$
-for values of $m$ ranging from $0$ to $\mathcal{O}(100)$.
-The authors of~\cite{Gil2007} did not derive a recursion relation
-for precisely this case. We could follow their methods and do so ourselves,
-but it is much easier to convert to a case that they did consider.
-The strategy is to look through the minimal solutions tabulated
-in~\cite{Gil2007} and search for a transformation we could apply to
-${_2F_1}(\alpha+m, \beta+m, \gamma+m; -b)$ that would place the $m$'s
-(the variable being incremented by the recursion)
-in the same arguments of ${_2F_1}$ as the minimal solution.
-After some ``guess and check,'' we found that the transformation
-\begin{align}
-{_2F_1}(\alpha+m, \beta+m, \gamma+m; -b)
-=
-(1+b)^{-\alpha-m}
-        {_2F_1}\left(\alpha+m, \gamma-\beta, \gamma+m; \frac{b}{1+b}\right),
-\label{eq:rec_euler_pretransform}
-\end{align}
-produces a ${_2F_1}$ on the right hand side that closely resembles
-the minimal solutions $y_{3,m}$ and $y_{4,m}$ in Eq.~4.3 in~\cite{Gil2007}.
-Explicitly, these solutions are
-\begin{align}
-y_{3,m}
-&\propto
-{_2F_1}\left(-\alpha^\prime + \gamma^\prime - m,
-                -\beta^\prime + \gamma^\prime,
-                1-\alpha^\prime-\beta^\prime+\gamma^\prime-m;
-                1-z\right)
-\\
-y_{4,m}
-&\propto
-{_2F_1}\left(\alpha^\prime + m,
-                \beta^\prime,
-                1+\alpha^\prime+\beta^\prime-\gamma^\prime+m;
-                1-z\right),
-\label{eq:minimal_soln_sans_prefac}
-\end{align}
-where we have omitted prefactors which are unimportant for now.
-Which of these two we should use depends on what values $z$ takes on.
-Equating $1-z=b/(1+b)$ gives $z=1/(1+b)$, and since $b$ is strictly positive,
-$z$ is bounded between 0 and 1.
-From Eq.~4.5 in~\cite{Gil2007}, $y_{4,m}$ is the minimal solution
-for real $z$ satisfying $0<z<2$, so this is the only minimal solution we need.
-
-Now that we have our minimal solution,
-what recurrence relation does it satisfy?
-Confusingly, the recurrence relation of which $y_{4,m}$ is a solution
-increments different arguments of ${_2F_1}$ that does $y_{4,m}$:
-it increments the first only, rather than first and third.
-This recurrence relation can be looked up, e.g., Eq.~15.2.10
-in~\cite{Abramowitz1964}, which is
-\begin{align}
-(\gamma^\prime - (\alpha^\prime + m)) f_{m-1}
-+
-(2(\alpha^\prime+m) - \gamma^\prime + (\beta^\prime - \alpha^\prime)z)f_m
-+ \alpha^\prime(z-1) f_{m+1} = 0.
-\label{eq:chosen_rec_rel}
-\end{align}
-Now we must solve for the parameters appearing in the recurrence relation
-in terms of our parameters, namely by setting
-\begin{align}
-\begin{split}
-\alpha^\prime &= \alpha
-\\
-\beta^\prime &= \gamma - \beta
-\\
-1 + \alpha^\prime + \beta^\prime - \gamma^\prime &= \gamma
-\\
-1 - z &= \frac{b}{1+b}
-\end{split}
-\end{align}
-and solving to find
-\begin{align}
-\begin{split}
-\alpha^\prime &= \alpha
-\\
-\beta^\prime &= \gamma - \beta
-\\
-\gamma^\prime &= 1 + \alpha - \beta
-\\
-z &= \frac{1}{1+b}
-.
-\end{split}
-\end{align}
-Finally we have everything we need. The minimal solution
-\begin{align}
-y_{4,m}
-=
-\frac{\Gamma(1+\alpha^\prime-\gamma^\prime+m)}
-        {\Gamma(1+\alpha^\prime+\beta^\prime-\gamma^\prime+m)}
-\times
-{_2F_1}\left(\alpha^\prime + m,
-                \beta^\prime,
-                1+\alpha^\prime+\beta^\prime-\gamma^\prime+m;
-                1-z\right),
-\end{align}
-where we have now included the necessary prefactors,
-is a numerically stable solution of the recurrence
-relation~\eq{eq:chosen_rec_rel} if the recursion is run
-from large $m$ to small $m$.
-So the procedure is to compute the value of ${_2F_1}$ for the two
-largest $m$ values of interest using arbitrary precision arithmetic,
-compute the prefactors to construct
-$y_{4,\text{max}(m)}$ and $y_{4,\text{max}(m)-1}$,
-recursively compute $y_{4,m}$ for all $m$ less than $\text{max}(m)$ down
-to $m=0$, then cancel off the prefactors of the resulting values of
-$y_{4,m}$ for all $m$ to produce ${_2F_1}$ for all desired $m$ values.
-\mmnote{Maybe I should reformat this paragraph into
-a pseudocode version of the algorithm?
-Instead of, or in addition to?}
-
-With ${_2F_1}$ computed, the only remaining numerical danger in computing
-$p(m)$ in~\eq{eq:p_m_bursty+rep} is overflow of the gamma functions.
-This is easily solved by taking the log of the entire expression
-and using standard routines to compute the log of the gamma functions,
-then exponentiating the entire expression at the end if $p(m)$
-is needed rather than $\log p(m)$.
-
-% \bibliographystyle{nature}
-\bibliographystyle{abbrv}
-\bibliography{./library}
-
-\end{document}
+must hold, where $\beta\Delta\epsilon_R$ is inferred from the master curve fit
+to fold-change measurements, and $k_R^+$ and $k_R^-$ are inferred in some
+orthogonal manner. Single molecule measurements such as from~\cite{Hammar2014}
+have directly observed these rates, and in the remainder of this work we explore
+a complementary approach: inferring repressor binding and unbinding rates
+$k_R^+$ and $k_R^-$ from FISH measurements of mRNA population distributions.

### 2020-05-19 doc/section_03_beyond_means.tex
initial splitting of document into several files

--- 2020-05-17 doc/paper.tex
@@
-\documentclass[12pt]{article}%{amsart}
-\usepackage[top = 1.0in, bottom = 1.0in, left = 1.0in, right = 1.0in]{geometry}
-% See geometry.pdf to learn the layout options. There are lots.
-%\geometry{letterpaper}               % ... or a4paper or a5paper or ...
-%\geometry{landscape}             % Activate for for rotated page geometry
-\usepackage[parfill]{parskip}    % Activate to begin paragraphs with an empty line rather than an indent
-\usepackage{graphicx}
-\usepackage{amssymb}
-%\usepackage{epstopdf}
-% \usepackage{caption}
-\usepackage{amsmath}
-% \usepackage{longtable}
-% \usepackage{tabu}
-% \usepackage{accents} %to get undertildes for vec & mat
-\usepackage{color} % for colored text
-% \usepackage[normalem]{ulem} % for editing: striking out text using 'sout' command
-
-% \newcommand\mytablefigwidth{0.35\textwidth}
-
-% allows to use .ai files directly w/o resaving as pdf
-\DeclareGraphicsRule{.ai}{pdf}{.ai}{}
-
-% Handy math macros!
-\newcommand{\vect}[1]{\vec{#1}}
-\newcommand{\matr}[1]{\mathbf{#1}}
-\newcommand{\rate}[3]{{#1}_{#2}^{#3}}
-\newcommand{\mmnote}[1]{\textcolor{cyan}{(MM:~#1)}}
-
-% derivative macros. these are sneaky. the [{}] is an empty default 1st arg
-% usage: provide 1 arg: \deriv{x} gives d/dx
-% provide 2 args like so: \deriv[f]{x} gives df/dx
-\newcommand{\deriv}[2][{}]{\frac{d #1}{d #2}}
-\newcommand{\pderiv}[2][{}]{\frac{\partial #1}{\partial #2}}
-\newcommand{\psecderiv}[2][{}]{\frac{\partial{^2} #1}{\partial #2{^2}}}
-
-%%%%% Referencing macros %%%%%
-\newcommand{\fig}[1]{Figure~\ref{#1}}
-\newcommand{\tref}[1]{Table~\ref{#1}}
-\newcommand{\eq}[1]{Eq.~(\ref{#1})}
-\newcommand{\eqrange}[2]{Eqs.~(\ref{#1}-\ref{#2})}
-
-%%%%%
-\begin{document}
-
-\title{Simple repression draft}
-
-\maketitle
-
-\section{Introduction}
-\mmnote{Key ideas, no particular order, that I haven't written down before:
-\begin{itemize}
-\item Our goal is to build phenomenological models of input-output
-functions of genetic circuits. More precisely, we want phenomenology
-in the sense of coarse-graining away atomistic detail, but still
-retaining biophysical meaning, e.g., we don't want to coarse-grain
-as far as Hill functions. Why not? We are motivated by studies
-like~\cite{Razo-Mejia2020}, for which Hill functions are
-insufficient. Such work requires a quantitative theory of how
-biophysical changes at the molecular level propagate to
-input-output functions at the genetic circuit level.
-We want concepts, not mere facts.
-In particular a key question is: at this level of
-coarse graining, what microscopic details do I need to explicitly
-model, and how do we figure that out? For example, do I need to
-worry about all or even any of the steps that individual RNAPs go
-through each time they make a transcript? Turning the question
-around, can we see any imprint of those processes in the
-available data? If the answer is no, then those processes are
-irrelevant for our purposes. Forward modeling and inverse
-(statistical inferential) modeling can complement each other
-beautifully here.
-\end{itemize}}
-
-\fig{fig1:means_cartoons}A shows the qualitative picture of
-simple repression. An operator, the binding site on the DNA for a
-repressor protein, may be found occupied by a repressor, in which
-case transcription is blocked from occurring. Or it may be found
-unoccupied, in which case RNA polymerase (RNAP) may bind and
-transcription can proceed.
-
-The key assumption we make is that binding of repressor and RNAP in the promoter region of interest is exclusive, meaning that one or the other may bind, but never may both be simultaneously bound.
-It is often imagined that when the repressor is bound to its operator,
-RNAP is sterically blocked from binding to its promoter sequence.
-Current evidence suggests this is sometimes, but not
-always the case~\mmnote{Cite that fig, from ??Nathan or Bill or
-Rob??, showing locations of repressor binding sites far upstream
-of promoter}, and it remains an interesting open question precisely
-how a repressor bound far upstream is able to repress transription.
-Suggestions include ``action-at-a-distance'' mediated by kinks in
-the DNA, formed when the repressor is bound, that prevent RNAP binding.
-Nevertheless, our modeling in this work is sufficiently
-coarse-grained that we simply assume exclusive binding and need
-not worry about such precise details.
-
-\mmnote{Add a notation aside somewhere that all association rates are written as zeroth-order rates.
-In other words, the concentration of the molecule is hidden inside, so for instance, $k_R^+$ is proportional to repressor copy number.}
-
-\section{Means}
-\subsection{Fold-changes are indistinguishable across models}
-As a first stop on our search for the ``right'' model of simple
-repression, let us consider what we can learn from theory and
-measurements of fold-change in mean expression. The fold-change is defined as
-\begin{equation}
-\text{fold-change}
-= \frac{\langle \text{gene expression with repressor present} \rangle}
-        {\langle \text{gene expression with repressor absent} \rangle}
-= \frac{\langle m\rangle(R\ne 0)}{\langle m\rangle(R = 0)}
-\label{eq:fc_def}
-\end{equation}
-where angle brackets denote the averge over a population of cells
-and mean mRNA $\langle m\rangle$ is viewed as a function of
-repressor copy number $R$.
-The second
-equality in~\ref{eq:fc_def} follows from assuming that the
-translation efficiency, i.e., the number of proteins translated
-per mRNA, is the same in both conditions.
-In other words, we assume that mean protein level is proportional
-to mean mRNA level, and that the proportionality constant is the
-same in both conditions and therefore cancels out in the ratio.
-This is reasonable since the cells in the two conditions are
-identical except for the presence of the transcription factor,
-and it is difficult to imagine the transcription factor having
-any direct effect on translation.
-
-Fold-change has proven a very convenient observable in past
-work~\cite{Garcia2011a, Razo-Mejia2018, Chure2019}.
-Part of its utility in dissecting transcriptional regulation is
-its ratiometric nature, which is always easier than making an
-absolute measurement. Also, by measuring otherwise identical cells
-with and without a transcription factor present, any biological
-noise common to both conditions can be made to cancel away.
-In addition, the presence of different operator sites is known to
-change the basal expression level in the absence of
-repressor~\mmnote{look up cite}.
-Computing fold-changes neatly cancels this effect as well.
-
-\fig{fig1:means_cartoons} depicts a smorgasbord
-of mathematized cartoons for simple repression that have appeared
-in previous literature. For each cartoon, we calculate the
-fold-change in mean gene expression as predicted by that model,
-deferring some algebraic details to the appendix.
-What we will find is that all cartoons collapse to a single
-master curve, shown in \fig{fig1:means_cartoons}c, which contains
-just two parameters. We label the parameters $\Delta F_R$, an
-effective free energy parametrizing the repressor-DNA
-interaction, and $\rho$, which subsumes all details of
-transcription in the absence of repressors.
-We will offer some intuition for why this master curve exists and discuss
-why it is likely impossible to discriminate ``right'' from ``wrong''
-cartoons given only measurements of fold-changes in expression.
-
-\subsubsection{Equilibrium, 2-state}
-In this simplest model, the promoter is idealized as existing in
-one of two states, either repressor bound or repressor unbound.
-The rate of transcription, and therefore the level of gene
-expression, is assumed to be proportional to the fraction of time
-spent in the repressor unbound state.
-Mathematically, this amounts to the statement that the mean mRNA level is
-given by
-\begin{equation}
-\langle m \rangle = \frac{r}{\gamma}
-        \left(1 + \frac{R}{N_{NS}} e^{-\beta\Delta\epsilon_R}\right)^{-1},
-\end{equation}
-where $r$ is the transcription rate from the repressor unbound
-state, $\gamma$ is the mRNA degradation rate, $R$ is repressor
-copy number, $N_{NS}$ is the number of nonspecific binding sites
-in the genome (where repressors spend most of their time when not
-bound to the operator, their specifc binding site),
-$\beta=1/k_BT$, and $\Delta\epsilon_R$ is the binding energy of a
-repressor to its operator site. The derivation of this result is deferred
-to the appendix.
-
-The fold-change is easily found as the ratio of mean mRNA with
-and without repressor. This gives
-\begin{equation}
-\text{fold-change}
-= \left(1 + \frac{R}{N_{NS}} e^{-\beta\Delta\epsilon_R}\right)^{-1},
-\end{equation}
-which clearly matches the form of the master curve
-in~\eq{fig1:means_cartoons} with $\rho=1$.
-
-In fact it was noted in~\cite{Chure2019} that this two-state
-model can be viewed as the coarse-graining of any equilibrium
-promoter model in which no transcriptionally active states have
-transcription factor bound. Put differently, there is no overlap
-between TF bound states and transcriptionally active states. We
-will see this explicitly in the 3-state equilibrium model below,
-but perhaps surprising is that something analogous carries over
-even to the nonequilibrium models we consider later.
-
-\subsubsection{Equilibrium, 3-state}
-Compared to the previous model, here we fine-grain the repressor
-unbound state into two separate states: empty, and RNAP bound.
-This picture was used in~\cite{Garcia2011a} as we use it here,
-and in~\cite{Razo-Mejia2018} and~\cite{Chure2019} it was generalized
-to incorporate small-molecule inducers that bind the repressor.
-Interestingly, the effect of this is simply to renormalize $R$
-from the total number of repressors to a smaller effective number,
-roughly (though not exactly) the number which are unbound by inducers.
-
-The mean mRNA, as derived in the appendix from a similar
-enumeration of states and weights as the previous model, is
-\begin{equation}
-\langle m \rangle = \frac{r}{\gamma}
-\frac{\frac{P}{N_{NS}} e^{-\beta\Delta\epsilon_P}}
-        {
-        1 + \frac{R}{N_{NS}} e^{-\beta\Delta\epsilon_R}
-        + \frac{P}{N_{NS}} e^{-\beta\Delta\epsilon_P}
-        },
-\end{equation}
-where the new variables are $\Delta\epsilon_P$, the difference in
-RNAP binding energy to its specific site (the promoter) relative
-to an average nonspecific background site, and the RNAP copy
-number, $P$. The fold-change again follows immediately as
-\begin{align}
-\text{fold-change}
-&= \frac{\frac{P}{N_{NS}} e^{-\beta\Delta\epsilon_P}}
-        {
-        1 + \frac{R}{N_{NS}} e^{-\beta\Delta\epsilon_R}
-        + \frac{P}{N_{NS}} e^{-\beta\Delta\epsilon_P}
-        }
-\frac{1 + \frac{P}{N_{NS}} e^{-\beta\Delta\epsilon_P}}
-        {\frac{P}{N_{NS}} e^{-\beta\Delta\epsilon_P}}
-\\
-&= \left(
-1 + \frac{\frac{R}{N_{NS}} e^{-\beta\Delta\epsilon_R}}
-        {1 + \frac{P}{N_{NS}} e^{-\beta\Delta\epsilon_P}}
-\right)^{-1}
-\\
-&= (1 + \exp(-\Delta F_R - \log\rho))^{-1},
-\end{align}
-with $\Delta F_R = \beta\Delta\epsilon_R - \log(R/N_{NS})$
-and $\rho = 1 + \frac{P}{N_{NS}}\mathrm{e}^{-\beta\Delta\varepsilon_P}$
-as shown in~\fig{fig1:means_cartoons}.
-
-\begin{figure}%[h!]
-\centering
-\includegraphics[width=\textwidth]{../figures/fig1/fig1.ai}
-\caption{
-\textbf{
-An overview of the simple repression motif at the level of means.}
-(A) depicts the qualitative biological picture of simple repression.
-(B) shows a variety of possible mathematized cartoons of simple
-repression, along with the effective parameter $\rho$
-which subsumes all regulatory details
-of the architecture excluding the repressor.
-(C) shows the ``master curve'' to which all cartoons in (B) collapse.
-}
-\label{fig1:means_cartoons}
-\end{figure}
-
-\subsubsection{Nonequilibrium model 1 - Poisson promoter}
-For our first kinetic model, we consider the simplest possible
-picture with only two states, repressor bound and unbound.
-This is exactly the model used for the main results of~\cite{Jones2014}.
-In this picture, repressor association and dissociation rates
-from its operator site, $k_R^+$ and $k_R^-$, respectively, govern
-transitions between the two states. When the system is in the
-unbound state, transcriptions initiate at rate $r$, which
-represents a coarse-graining of all the downstream processes into
-a single effective rate. mRNA is again degraded at rate $\gamma$.
-
-Let $p_{m,R}(t)$ denote the joint probability of finding the
-system in the repressor bound state with $m$ mRNA molecules
-present at time $t$, and similarly define $p_{m,U}(t)$ for the
-repressor unbound state. 
-This model is governed by a master equation giving the time
-evolution of $p_{m,R}(t)$ and $p_{m,U}(t)$, which is
-\begin{align}
-\begin{split}
-\deriv{t}p_{m,R}(t) =& - k_R^- p_{m,R}(t) + k_R^+ p_{m,U}(t)
-                + (m+1)\gamma p_{m+1,R}(t) - \gamma p_{m,R}(t)
-\\
-\deriv{t}p_{m,U}(t) =&\; k_R^- p_{m,R}(t) - k_R^+ p_{m,U}(t)
-                        + rp_{m-1,U}(t) - rp_{m,U}(t)
-                \\
-                &+ (m+1)\gamma p_{m+1,U}(t) - \gamma p_{m,U}(t).
-\end{split}
-\end{align}
-In each equation, the first two terms describe transitions
-between promoter states due to repressors unbinding and binding,
-respectively. The final two terms describe degradation of mRNA,
-decreasing the copy number by one, and the terms with coefficient
-$r$ describe transcription initiation increasing the mRNA copy
-number by one.
-
-We can greatly simplify the notation, which will be especially
-useful for the more complicated models yet to come, by
-re-expressing the master equation in vector form. The promoter
-states are collected into a vector and the rate constants are
-collected into matrices as
-\begin{equation}
-\vect{p}_m = \begin{pmatrix} p_{m,R} \\ p_{m,U} \end{pmatrix},\
-\matr{K} = \begin{pmatrix} -k_R^- & k_R^+ \\ k_R^- & -k_R^+ \end{pmatrix},\
-\matr{R} = \begin{pmatrix} 0 & 0 \\ 0 & r \end{pmatrix},\
-\label{eq:2state_cme_matrices}
-\end{equation}
-so that the master equation may be condensed as
-\begin{equation}
-\deriv{t}\vect{p}_m(t) =
-\left( \matr{K} - \matr{R} - \gamma m \matr{I} \right) \vect{p}_m(t)
-                + \matr{R} \vect{p}_{m-1}(t)
-                + \gamma (m+1) \matr{I} \vect{p}_{m+1}(t)
-                - \gamma m \matr{I} \vect{p}_m(t),
-\label{eq:2state_rep_cme}
-\end{equation}
-where $\matr{I}$ is the identity matrix.
-Taking steady state by setting time derivatives to zero,
-the mean mRNA can be found to be
-\begin{equation}
-\langle m \rangle = \frac{r}{\gamma}
-        \left(1 + \frac{k_R^+}{k_R^-}\right)^{-1},
-\end{equation}
-with the algebra details again deferred to the appendix.
-Recall $k_R^+$ is proportional to the repressor copy number, so
-in computing fold-change, absence of repressor corresponds to
-$k_R^+\rightarrow0$.
-Therefore fold-change in this model is simply
-\begin{equation}
-\text{fold-change} = \left(1 + \frac{k_R^+}{k_R^-}\right)^{-1},
-\end{equation}
-again matching the master curve of~\fig{fig1:means_cartoons} with $\rho=1$.
-
-\subsubsection{Nonequilibrium model 2 - RNAP bound/unbound states}
-Analogous to the three-state equilibrium model, this model
-fine-grains the repressor unbound state of nonequilibrium model 1,
-resolving it into an empty promoter state and an RNAP-bound state.
-Note in this picture, in contrast with model 4 below,
-transcription initiation is accompanied by a promoter state
-change, in keeping with the interpretation as RNAP-bound and
-empty states: if an RNAP successfully escapes the promoter and
-proceeds to elongation of a transcript, clearly it is no longer
-bound at the promoter! Therefore another RNAP must bind before
-another transcript can be initiated.
-
-The master equation governing this model is analogous
-to~\eqrange{eq:2state_cme_matrices}{eq:2state_rep_cme} for model 1 above.
-The main subtlety arises since transcription initiation
-accompanies a promoter state change.
-This can be understood by analogy to $\matr{K}$. The
-off-diagonals and diagonals of $\matr{K}$ correspond to
-transitions arriving at or departing from, respectively, the
-promoter state of interest. If transcription initiations are
-accompanied by promoter state changes, we must have separate
-matrices for arriving and departing transcription events since
-the arriving and departing transitions have different initial
-copy number of mRNA, unlike for $\matr{K}$ where they are the same.
-The master equation for this model is
-\begin{equation}
-\deriv{t}\vect{p}_m(t) =
-\left( \matr{K} - \matr{R_D} - \gamma m \matr{I} \right) \vect{p}_m(t)
-                + \matr{R_A} \vect{p}_{m-1}(t) +
-                \gamma (m+1) \matr{I} \vect{p}_{m+1}(t)
-                - \gamma m \matr{I} \vect{p}_m(t),
-\label{eq:3state_rep_cme}
-\end{equation}
-with the state vector and promoter transition matrix defined as
-\begin{equation}
-\vect{p}_m = \begin{pmatrix} p_{m,R} \\ p_{m,E} \\ p_{m,P} \end{pmatrix},\
-\matr{K} = \begin{pmatrix} -k_R^- & k_R^+ & 0 \\
-                        k_R^- & -k_R^+ -k_P^+ & k_P^- \\
-                        0 & k_P^+ & -k_P^- 
-                \end{pmatrix},
-\label{eq:3state_cme_matrices_pt1}
-\end{equation}
-and the initiation matrices given by
-\begin{equation}
-\matr{R_A} = \begin{pmatrix}
-                0 & 0 & 0 \\ 0 & 0 & r \\ 0 & 0 & 0
-                \end{pmatrix},\
-\matr{R_D} = \begin{pmatrix}
-                0 & 0 & 0 \\ 0 & 0 & 0 \\ 0 & 0 & r
-                \end{pmatrix},
-\label{eq:3state_cme_matrices_pt2}
-\end{equation}
-The elements of $\vect{p}_m$ encode the probabilities of having
-$m$ mRNA present along with the promoter having repressor bound,
-being empty, or having RNAP bound, respectively.
-$\matr{R_A}$ describes probability flux \textit{arriving} at the state
-$\vect{p}_m$ from a state with one fewer mRNA, namely
-$\vect{p}_{m-1}$, and $\matr{R_D}$ describes probability flux
-\textit{departing} from the state $\vect{p}_m$
-for a state with one more mRNA, namely $\vect{p}_{m+1}$.
-$\matr{K}$ is closely analogous to model 1.
-
-Mean mRNA at steady state is found analogously to model 1, with the result
-\begin{equation}
-\langle m\rangle = \frac{r}{\gamma}
-        \frac{k_R^- k_P^+}
-        {k_R^- k_P^+ + k_R^- (k_P^- + r) + k_R^+ (k_P^- + r)}.
-\end{equation}
-Fold-change is again found from the ratio prescribed by~\eq{eq:fc_def},
-from which we have
-\begin{align}
-\text{fold-change}
-&=      \frac{k_R^- k_P^+}
-        {k_R^- k_P^+ + k_R^- (k_P^- + r) + k_R^+ (k_P^- + r)}
-        \frac{k_P^+ + k_P^- + r}{k_P^+}
-\\
-&=      \left(1 + \frac{k_R^+}{k_R^-}
-                \frac{k_P^- + r}{k_P^+ + k_P^- + r}
-        \right)^{-1}
-\\
-&=      \left(1 + \frac{k_R^+}{k_R^-}
-        \left(1 + \frac{k_P^+}{k_P^- + r}\right)^{-1}
-        \right)^{-1},
-\end{align}
-which follows the master curve with
-$\rho = 1 + k_P^+/(k_P^- + r)$ as claimed.
-
-\subsubsection{Nonequilibrium model 3
-- Multistep transcription initiation and escape}
-One might reasonably complain that the first two
-``nonequilibrium'' models we have considered are straw men. Their
-steady states necessarily satisfy detailed balance which is
-equivalent to thermodynamic equilibrium. Why is this the case? At
-steady state there is by definition no net probability flux in or
-out of each promoter state, but since the promoter states form a
-linear chain, there is only one way in or out of the repressor
-bound and RNAP bound states, implying each edge must actually
-have a net zero probability flux, which is the definition of
-detailed balance (usually phrased as equality of forward and
-reverse transition fluxes).
-
-Now we consider model 3 in~\eq{fig1:means_cartoons} which allows
-the possibility of true nonequilibrium steady-state fluxes
-through the promoter states.
-We point out that this model was considered previously in~\cite{Mitarai2015}
-where a comparison was made with model 1 as used in~\cite{Jones2014}.
-The authors of~\cite{Mitarai2015} argued that the additional
-complexity is essential to properly account for the noise in the
-mRNA distribution. We will weigh in on both models later when we
-consider observables beyond fold-change.
-
-The master equation governing this model is
-identical in form to model 2 above, namely
-\begin{equation}
-\deriv{t}\vect{p}_m(t) =
-\left( \matr{K} - \matr{R_D} - \gamma m \matr{I} \right) \vect{p}_m(t)
-                + \matr{R_A} \vect{p}_{m-1}(t) +
-                \gamma (m+1) \matr{I} \vect{p}_{m+1}(t)
-                - \gamma m \matr{I} \vect{p}_m(t),
-\end{equation}
-but obviously with a higher-dimensional state space and different matrices.
-The state vector and promoter transition matrix are now
-\begin{equation}
-\vect{p}_m = \begin{pmatrix} p_{m,R} \\ p_{m,E} \\
-                             p_{m,C} \\ p_{m,O}\end{pmatrix},\
-\matr{K} = \begin{pmatrix} -k_R^- & k_R^+ & 0 & 0\\
-                        k_R^- & -k_R^+ -k_P^+ & k_P^- & 0 \\
-                        0 & k_P^+ & -k_P^- - k_O & 0 \\
-                        0 & 0 & k_O & 0
-                \end{pmatrix},
-\end{equation}
-with the four promoter states, in order, being repressor bound,
-empty, RNAP closed complex, and RNAP open complex.
-Besides increasing dimension by one again, the only new feature in
-$\matr{K}$ is the rate $k_O$, representing the (irreversible)
-rate of open complex formation from the closed complex.
-The initiation matrices are given by
-\begin{equation}
-\matr{R_A} = \begin{pmatrix}
-        0 & 0 & 0 & 0 \\ 0 & 0 & 0 & r \\ 0 & 0 & 0 & 0\\ 0 & 0 & 0 & 0
-                \end{pmatrix},\
-\matr{R_D} = \begin{pmatrix}
-        0 & 0 & 0 & 0 \\ 0 & 0 & 0 & 0 \\ 0 & 0 & 0 & 0\\ 0 & 0 & 0 & r
-                \end{pmatrix},
-\end{equation}
-again closely analogous to model 2.
-
-The expression for mean mRNA is substantially more complicated
-now, as worked out in the appendix where we find
-\begin{equation}
-\langle m\rangle = \frac{r}{\gamma}
-        \frac{k_R^- k_P^+ k_O}
-        {k_R^- [(k_P^+ (k_O + r) + r(k_P^- + k_O)] + k_R^+ r(k_P^- + k_O)},
-\end{equation}
-which can be simplified to
-\begin{equation}
-\langle m\rangle
-= \frac{r}{\gamma}
-\frac{\frac{k_P^+ k_O}{r(k_O + k_P^-)}}
-        {1 + \frac{k_P^+ (k_O + r)}{r(k_O + k_P^-)} + \frac{k_R^+}{k_R^-}}.
-\end{equation}
-The strategy is to isolate the terms involving the repressor, so that now the fold-change is seen to be simply
-\begin{align}
-\text{fold-change}
-&= \frac{\frac{k_P^+ k_O}{r(k_O + k_P^-)}}
-        {1 + \frac{k_P^+ (k_O + r)}{r(k_O + k_P^-)} + \frac{k_R^+}{k_R^-}}
-        \frac{1 + \frac{k_P^+ (k_O + r)}{r(k_O + k_P^-)}}
-                {\frac{k_P^+ k_O}{r(k_O + k_P^-)}}
-\\
-&= \left(
-        1 + \frac{k_R^+}{k_R^-}
-        \left(1 + \frac{k_P^+ (k_O + r)}{r(k_O + k_P^-)}\right)^{-1}
-\right)^{-1},
-\end{align}
-surprisingly reducing to the master curve once again, with
-$\rho = 1 + \frac{k_P^+ (k_O + r)}{r(k_O + k_P^-)}$.
-
-This example hints that an arbitrarily fine-grained model of
-downstream transcription steps may still be collapsed to the form
-of the master curve in~\fig{fig1:means_cartoons}B so long as the
-repressor binding is exclusive with transcriptionally active
-states. We offer this as a conjecture, and we suspect that a
-careful argument using the King-Altman diagram
-method~\cite{King1956, Hill1966} might furnish a ``proof.''
-Our focus here is not on full generality but rather to survey an
-assortment of plausible models that have been proposed in the
-literature for simple repression.
-
-\subsubsection{Nonequilibrium model 4 - ``Active'' and ``inactive'' states}
-This model is at the core of the theory in~\cite{Razo-Mejia2020}.
-At a glance the cartoon for this model may appear very similar to model 2,
-and mathematically it is, but the interpretation is rather different.
-In model 2, we interpreted the third state literally as an
-RNAP-bound promoter and modeled initiation of a transcript as
-triggering a promoter state change, making the hopefully
-uncontroversial assumption that an RNAP can only make one
-transcript at a time. In contrast, in the present model the
-promoter state does \textit{not} change when a transcript is
-initiated. So we no longer interpret these states as literally
-RNAP bound and unbound but instead as coarse-grained ``active''
-and ``inactive'' states, the details of which we leave
-unspecified for now. We will comment more on this model below
-when we discuss Fano factors of models.
-
-Mathematically this model is very similar to models 1 and 2.
-Like model 1, the matrix $R$ describing transcription initiation
-is diagonal, namely
-\begin{equation}
-\matr{R} = \begin{pmatrix}
-                0 & 0 & 0 \\ 0 & 0 & 0 \\ 0 & 0 & r
-        \end{pmatrix}.
-\end{equation}
-so the master equation takes verbatim the same form as it did for model
-1,~\eq{eq:2state_rep_cme}. Meanwhile the promoter transition matrix
-$K$ is the same as~\eq{eq:3state_cme_matrices_pt1} from model 2,
-although we relabel the rate constants from $k_P^\pm$ to $k^\pm$ to
-reiterate that these are not simply RNAP binding and unbinding rates.
-
-Carrying out the algebra, the mean mRNA can be found to be
-\begin{equation}
-\langle m\rangle = \frac{r}{\gamma}
-\frac{k_R^- k^+}
-{k_R^- k^+ + k_R^- k^- + k_R^+ k^-},
-\end{equation}
-and the fold-change readily follows,
-\begin{align}
-\text{fold-change}
-&=      \frac{k_R^- k^+}{k_R^- k^+ + k_R^- k^- + k_R^+ k^-}
-        \frac{k_R^- k^+ + k_R^- k^-}{k_R^- k^+}
-\\
-&=      \left(1 + \frac{k_R^+}{k_R^-}
-                \left(1 + \frac{k^+}{k^-}\right)^{-1}
-        \right)^{-1},
-\end{align}
-from which we see $\rho = 1 + k^+/k^-$ as shown in~\fig{fig1:means_cartoons}.
-
-\subsubsection{Nonequilibrium model 5 - Bursty promoter}
-We will have much more to say about this model in subsequent sections.
-For now we sketch it as an intuitive analog to model 1,
-with just two states, repressor bound or unbound, and transition rates
-between them of $k_R^+$ and $k_R^-$.
-In model 1, when in the unbound state, single mRNA transcripts
-are produced as a Poisson process with some characteristic rate
-$r$. This model by contrast produces, at some Poisson rate $k_i$,
-\textit{bursts} of transcripts with a geometrically distributed
-burst size and a mean burst size $b$.
-
-From this intuitive picture and by analogy to model 1, then, it should be plausible that the mean mRNA level is
-\begin{equation}
-\langle m\rangle = \frac{k_i b}{\gamma}
-        \left(1 + \frac{k_R^+}{k_R^-}\right)^{-1},
-\end{equation}
-which will turn out to be correct from a careful calculation.
-For now, we simply note that just like model 1, the fold-change becomes
-\begin{equation}
-\text{fold-change} = \left(1 + \frac{k_R^+}{k_R^-}\right)^{-1}
-\end{equation}
-with $\rho=1$ also like model 1.
-
-We will also see later how this model emerges as a natural limit of model 4.
-
-\subsubsection{Summary and discussion of results across models for fold-changes in mean expression}
-What have we learned from our exploration of the models
-in~\fig{fig1:means_cartoons} and their predictions of
-fold-changes in mean expression levels? The takeaway is the
-existence of a master curve (\fig{fig1:means_cartoons}C) to which
-the fold-change predictions of all the models collapse. This
-master curve is parametrized by only two effective parameters:
-$\Delta F_R$, which characterizes the repressor, and $\rho$,
-which characterizes all other features of the promoter
-architecture. The key assumption underpinning this result is that
-zero transcription occurs when a repressor is bound to its
-operator. Note, however, that we are agnostic about the molecular
-mechanism which achieves this;
-steric effects are the obvious mechanism, but, for instance,
-``action-at-a-distance'' mediated by kinked DNA due to
-repressors bound tens or hundreds of nucleotides upstream of a
-promoter is plausible as well.
-
-Why does the master curve of~\fig{fig1:means_cartoons}C exist at all?
-This brings to mind the deep questions posed in,
-e.g.,~\cite{Frank2013} and~\cite{Frank2014a}, asking us to search
-for general ``symmetries'' of the problem instead of specifying
-generative models of the known microscopics of a system.
-In our case, the key feature seems to be the exclusive nature of
-repressor and RNAP binding, which allows the parameter describing
-the repressor, $\Delta F_R$, to cleanly separate from all other
-details of the promoter architecture, which are encapsulated in $\rho$.
-In the graph-theoretic language which was applied to similar
-problems in~\cite{Gunawardena2012} and~\cite{Ahsendorf2014}, the
-repressor-bound state is a leaf of the graph describing
-transitions between promoter states. Arbitrary nonequilibrium
-behavior can occur on the rest of the promoter state space, but
-it may all be swept up in the effective parameter $\rho$, to
-which the repressor makes no contribution.
-
-As suggested in~\cite{Chure2019}, we believe this master curve
-should generalize to architectures with multiple repressor
-binding sites, as long as the exclusivity of transcription factor
-binding and transcription initation is maintained. The
-interpretation of $\Delta F_R$ is then of an effective free
-energy of all repressor bound states.
-In an equilibrium picture this is simply given by the log of the
-sum of Boltzmann weights of all repressor bound states, which
-looks like the log of a partition function of a subsystem.
-In a nonequilibrium picture, while we can still mathematically gather
-terms and give the resulting collection the label $\Delta F_R$,
-it is unclear if the physical interpretation as an effective free energy
-makes sense.
-As is well known, free energies cannot be assigned unambiguously
-to states out of equilibrium because the free energy change along
-a generic path traversing the state space is path dependent,
-unlike at equilibrium.
-A consequence of this is that, out of equilibrium, $\Delta F_R$
-is no longer a simple sum of Boltzmann weights. Instead it
-resembles a restricted sum of King-Altman
-diagrams~\cite{King1956, Hill1966}. Following the work of
-Hill~\cite{Hill1989}, it may yet be possible to interpret this expression
-as an effective free energy, but this remains unclear to us.
-We leave this an open problem for future work.
-
-If we relax the requirement of exclusive repressor-RNAP binding,
-one could imagine models in which repressor and RNAP doubly-bound
-states are allowed, where the repressor's effect is to
-\text{reduce} the transcription rate rather than setting it to zero.
-Our results do not strictly apply to such a model, although we
-note that if the repressor's reduction of the transcription rate
-is substantial, such a model might still be well-approximated by
-one of the models in~\fig{fig1:means_cartoons}.
-
-One may worry that our ``one curve to rule them all'' is a
-mathematical tautology. In fact we \textit{agree} with this
-criticism if $\Delta F_R$ is ``just a fitting parameter'' and cannot
-be meaningfully interpreted as a real, physical free energy.
-An analogy to Hill functions is appropriate here. One of their
-great strengths and weaknesses, depending on the use they are put
-to, is that their parameters coarse-grain many details and are
-generally not interpretable in terms of microscopic models, for
-deep reasons discussed at length in~\cite{Frank2013}.
-By contrast, our master curve claims to have the best of both worlds:
-a coarse-graining of all details besides the repressor into
-a single effective parameter $\rho$,
-while simultaneously retaining an interpretation of
-$\Delta F_R$ as a physically
-meaningful and interpretable free energy.
-Our task, then, is to prove or disprove this claim.
-
-How do we test this and probe the theory with fold-change
-measurements? There is a fundamental limitation in that the
-master curve is essentially a one-parameter function of $\Delta
-F_R + \log\rho$. Worse, there are many \textit{a priori}
-plausible microscopic mechanisms that could contribute to the
-value of $\rho$, such as RNAP binding and escape
-kinetics~\cite{DeHaseth1998, Mitarai2015}, and/or supercoiling
-accumulation and release~\cite{Chong2014, Sevier2016}, and/or,
-RNAP clusters analogous to those obvserved in
-eukaryotes~\cite{Cisse2013, Cho2016}
-and recently also observed in bacteria~\cite{Ladouceur2020}.
-Even if $\Delta F_R$ is measured to
-high precision, inferring the potential microscopic contributions
-to $\rho$, buried inside a log no less, from fold-change
-measurements seems hopeless. As a statistical inference problem
-it is entirely nonidentifiable, in the language
-of~\cite{Gelman2013}, Section 4.3.
-
-If we cannot simply infer values of $\rho$ from measurements of
-fold-change, can we perturb some of the parameters that make up
-$\rho$ and measure the change? Unfortunately we suspect this is
-off-limits experimentally: most of the potential contributors to
-$\rho$ are global processes that affect many or all genes. For
-instance, changing RNAP association rates by changing RNAP copy
-numbers, or changing supercoiling kinetics by changing
-topoisomerase copy numbers, would massively perturb the entire
-cell's physiology and confound any determination of $\rho$.
-
-One might instead imagine a bottom-up modeling approach, where we
-mathematize a model of what we hypothesize the important steps
-are and are not, use in vitro data for the steps deemed
-important, and \textit{predict} what $\rho$ should be.
-But again, because of the one-parameter nature of the master
-curve, many different models will likely make indistinguishable
-predictions, and without any way to experimentally perturb in
-vivo, there is no clear way to test whether the modeling
-assumptions are correct.
-
-In light of this, we prefer the view that parameters and rates
-are not directly comparable between cartoons
-in~\fig{fig1:means_cartoons}. Rather, parameters in the
-simpler cartoons represent coarse-grained combinations of
-parameters in the finer-grained models. For instance, by equating
-$\rho$ between any two models, one can derive various possible
-correspondences between the two models' parameters. Note that
-these correspondences are clearly not unique, since many possible
-associations could be made. It then is a choice as to what microscopic
-interpretations the model-builder prefers for the parameters in a
-particular cartoon, and as to which coarse-grainings lend
-intuition and which seem nonsensical.
-Indeed, since it remains an open question what microscopic
-features dominate $\rho$ (as suggested above, perhaps RNAP
-binding and escape kinetics~\cite{DeHaseth1998, Mitarai2015}, or
-supercoiling accumulation and release~\cite{Chong2014, Sevier2016},
-or, something more exotic like
-RNAP clusters~\cite{Cisse2013, Cho2016, Ladouceur2020}),
-we are hesitant to put too much weight on any one microscopic
-interpretation of model parameters that make up $\rho$.
-
-One possible tuning knob to probe $\rho$ that would not globally
-perturb the cell's physiology is to manipulate RNAP binding sites.
-Work such as~\cite{Kinney2010} has shown that models of
-sequence-dependent RNAP can be inferred from data, and the
-authors of~\cite{Brewster2012} showed that the model
-of~\cite{Kinney2010} has predictive power by using the model to
-\textit{design} binding sites of a desired affinity.
-But for our purposes, this begs the question: the authors
-of~\cite{Kinney2010} \textit{assumed} a particular model
-(essentially our 3-state equilibrium model but without the
-repressor), so it is unclear how or if such methods can be turned
-around to \textit{compare} different models of promoter function.
-
-We have already pointed out that the master curve
-of~\fig{fig1:means_cartoons} is essentially a one-parameter model,
-the one parameter being $\Delta F_R + \log\rho$.
-By now the reader may be alarmed as to how can we even determine
-$\Delta F_R$ and $\rho$ independently of each other, never mind
-shedding a lens on the internal structure of $\rho$ itself.
-A hint is provided by the weak promoter approximation, invoked
-repeatedly in prior studies~\cite{Bintu2005c, Garcia2011a,
-Razo-Mejia2018} of simple repression using the 3-state
-equilibrium model in~\fig{fig1:means_cartoons}.
-In that picture, the weak promoter approximation means
-$\frac{P}{N_{NS}}\exp(-\beta\Delta\epsilon_P) \ll 1$,
-meaning therefore $\rho\approx1$.
-This is suggestive, but how can we be sure that $\rho$ is not,
-for instance, actually $10^2$ and that $\Delta F_R$ hides a
-compensatory factor?
-A resolution is offered by an independent inference of $\rho$ in
-the absence of repressors. This was done in~\cite{Razo-Mejia2020}
-by fitting nonequilibrium model 4  in~\fig{fig1:means_cartoons},
-with zero repressor (looking ahead, this is equivalent to model 4
-in~\fig{fig:constit_cartoons}), to FISH data
-from~\cite{Brewster2014}. This provided a determination of $k^+$
-and $k^-$, from which their ratio is estimated to be no more than
-a few $10^{-1}$ and possibly as small as $10^{-2}$.
-
-The realization that $\rho\approx1$ to an excellent
-approximation, \textit{independent} of which model
-in~\fig{fig1:means_cartoons} one prefers, goes a long way towards
-explaining the surprising success of equilibrium models of simple
-repression. Even though our 2- and 3-state models get so many
-details of transcription wrong, it does not matter because
-fold-change is a cleverly designed ratio. Since $\rho$ subsumes
-all details except the repressor, and $\log\rho\approx0$, fitting
-these simple models to fold-change measurements can still give a
-surprisingly good estimate of repressor binding energies. So the
-ratiometric construction of fold-change fulfills its intended
-purpose of canceling out all features of the promoter
-architecture except the repressor itself. Nevertheless it is
-perhaps surprising how effectively it does so: \textit{a priori},
-one might not have expected $\rho$ to be quite so close to 1.
-
-We would also like to highlight the relevance
-of~\cite{Landman2019} here. Landman et.\ al.\ reanalyzed and
-compared in vivo and in vitro data on the lacI repressor's
-binding affinity to its various operator sequences. (The in vivo
-data was from, essentially, fitting our master curve to
-expression measurements.) They find broad agreement between the
-in vitro and in vivo values. This reinforces the suspicion that
-the equilibrium $\Delta\epsilon_R$ repressor binding energies do
-in fact represent real physical free energies.
-Again, \textit{a priori} this did not have to be the case, even
-knowing that $\rho\approx1$.
-
-In principle, if $\Delta F_R$ can be measured to sufficient
-precision, then deviations from $\rho=1$ become a
-testable matter of experiment.
-In practice, it is probably unrealistic to measure repressor
-rates $k_R^+$ or $k_R^-$ or fold-changes in expression levels
-(and hence $\Delta\epsilon_R$) precisely enough to detect the
-expected tiny deviations from $\rho=1$.
-We can estimate the
-requisite precision in $\Delta F_R$ to resolve a given
-$\Delta\rho$ by noting, since $\rho\approx1$, that
-$\log(1+\Delta\rho)\approx \Delta\rho$, so
-$\Delta(\Delta F_R) \approx \Delta\rho$.
-Suppose we are lucky and $\Delta\rho$ happens to be $\sim0.1$, on
-the high end of our range estimated above.
-A determination of $\Delta\epsilon_R/k_BT$ with an uncertainty of
-barely 0.1 was achieved in the excellent measurements
-of~\cite{Razo-Mejia2018}, so this requires a very difficult
-determination of $\Delta F_R$ for a very crude determination of $\rho$,
-which suggests, to put it lightly, this is not a
-promising path to pursue experimentally.
-It is doubtful that inference of repressor kinetic rates
-would be any easier.\footnote{
-\mmnote{
-In fact the uncertainties quoted in Hammar et.\
-al.~\cite{Hammar2014} from Elf's group are small enough to do
-this calculation, but HJ \& I think their error bars are insanely
-optimistic. Not sure I want to blatantly call them out on it though??}
-}
-
-Moving forward, we have weak evidence supporting the
-interpretation of $\Delta F_R$ as a physically real free
-energy~\cite{Landman2019} and other work casting
-doubt~\cite{Hammar2014}.
-How might we resolve the confusion?
-If there is no discriminatory power to test the theory and
-distinguish the various models with measurements of fold-changes
-in means, how do we probe the theory? Clearly to discriminate
-between the nonequilibrium models in~\fig{fig1:means_cartoons},
-we need to go beyond means to ask questions about kinetics, noise and even
-full distributions of mRNA copy numbers over a population of cells.
-If the ``one-curve-to-rule-them-all'' is more than a
-mathematical tautology, then the free energy of repressor binding
-inferred from fold-change measurements should agree with
-repressor binding and unbinding rates.
-In other words, the equilbrium and nonequilibrium definitions of
-$\Delta F_R$ must agree, meaning
-\begin{equation}
-\Delta F_R = \beta\Delta\epsilon_R - \log(R/N_{NS})
-        = - \log(k_R^+/k_R^-),
-\end{equation}
-must hold, where $\beta\Delta\epsilon_R$ is inferred from the
-master curve fit to fold-change measurements, and $k_R^+$ and
-$k_R^-$ are inferred in some orthogonal manner.
-Single molecule measurements such as from~\cite{Hammar2014} have
-directly observed these rates, and in the remainder of this work
-we explore a complementary approach: inferring repressor binding
-and unbinding rates $k_R^+$ and $k_R^-$ from FISH measurements of
-mRNA population distributions.
-
+% !TEX root = ./busty_transcription.tex
 \section{Beyond means}
 \subsection{A one-state promoter with bursting is the ``best'' model of constitutive promoters}
-Before we can tackle simple repression, we need an adequate
-phenomenological model of constitutive expression.
-The literature abounds with options from which we can choose,
-and we show several potential kinetic models for constitutive promoters
-in~\fig{fig:constit_cartoons}.
-Let us consider the suitability of each model for our purposes in turn.
+Before we can tackle simple repression, we need an adequate phenomenological
+model of constitutive expression. The literature abounds with options from which
+we can choose, and we show several potential kinetic models for constitutive
+promoters in~\fig{fig:constit_cartoons}. Let us consider the suitability of each
+model for our purposes in turn.
 
@@
 The left column depicts various plausible cartoons for constitutive promoters.
-We also list literature references where models have been used previously.
-In model (1), transcripts are produced in a Poisson process~\cite{Jones2014}.
-Model (2) features explicit modeling of RNAP binding/unbinding
-kinetics~\cite{Phillips2015a}.
-Model (3) is a more detailed generalization of model (2),
-treating transcription initiation as a multi-step process
-proceeding through closed and open complexes~\cite{Mitarai2015}.
-Model (4) is somewhat analogous to (2) except with the precise nature
-of active and inactive states left ambiguous~\cite{Razo-Mejia2020}.
-Finally, model (5) can be viewed as a certain limit of model (4)
-in which transcripts are produced in bursts,
-and initiation of bursts is a Poisson process.
-\mmnote{Haven't found model 5 formulated quite like this in the literature,
-usually burstiness is handled like model 4, but I should probably search a bit more.}
-The right column shows the Fano factor for each model.
-Note especially the crucial diagnostic: (2) and (3) have $\nu$
-strictly below 1, while only for (4) and (5) can $\nu$ exceed 1.
-Models with Fano factors $\le 1$ cannot produce the FISH data
-observed in~\cite{Jones2014} without introducing additional
-assumptions and model complexity.
-}
+We also list literature references where models have been used previously. In
+model (1), transcripts are produced in a Poisson process~\cite{Jones2014}. Model
+(2) features explicit modeling of RNAP binding/unbinding
+kinetics~\cite{Phillips2015a}. Model (3) is a more detailed generalization of
+model (2), treating transcription initiation as a multi-step process proceeding
+through closed and open complexes~\cite{Mitarai2015}. Model (4) is somewhat
+analogous to (2) except with the precise nature of active and inactive states
+left ambiguous~\cite{Razo-Mejia2020}. Finally, model (5) can be viewed as a
+certain limit of model (4) in which transcripts are produced in bursts, and
+initiation of bursts is a Poisson process. \mmnote{Haven't found model 5
+formulated quite like this in the literature, usually burstiness is handled like
+model 4, but I should probably search a bit more.} The right column shows the
+Fano factor for each model. Note especially the crucial diagnostic: (2) and (3)
+have $\nu$ strictly below 1, while only for (4) and (5) can $\nu$ exceed 1.
+Models with Fano factors $\le 1$ cannot produce the FISH data observed
+in~\cite{Jones2014} without introducing additional assumptions and model
+complexity.}
 \label{fig:constit_cartoons}
@@
 \subsubsection{Model 1 - Poisson promoter.}
-This is the picture from Jones et.\ al.~\cite{Jones2014}, which
-assumes that transcripts are produced as a Poisson process from a
-single promoter state. This model insists that the ``true'' mRNA
-distribution is Poisson, implying the Fano factor $\nu$ must be 1.
-In~\cite{Jones2014}, the authors carefully attribute deviations
-from Fano = 1 to intensity variability in fluorescent spots, gene
-copy number variation, and copy number fluctuations of the
-transcription machinery, e.g., RNAP itself.
-In this picture, the master equation makes no appearance,
-and all the corrections to Poisson behavior are derived
-as additive corrections to the Fano factor.
-For disproving the ``universal noise curve'' from So et
-al~\cite{So2011}, this picture was excellent. It is appealing in
-its simplicity, with only two parameters, the initiation rate $r$
-and degradation rate $\gamma$. Since $\gamma$ is independently
-known from other experiments, and the mean mRNA copy number is
-$r/\gamma$, $r$ is easily inferred from data.
-In other words, the model is not excessively complex for the data at hand.
-But for many interesting questions, for instance in the recent
-work~\mmnote{cite Manuel's preprint once posted},
-knowledge of means and variances alone is insufficient, and a
-description of the full distribution of molecular counts is necessary.
-For this we need a( slightly) more complex model than model 1
-that would allow us to incorporate the non-Poissonian features of
-constitutive promoters directly into a master equation formulation.
+This is the picture from Jones et.\ al.~\cite{Jones2014}, which assumes that
+transcripts are produced as a Poisson process from a single promoter state. This
+model insists that the ``true'' mRNA distribution is Poisson, implying the Fano
+factor $\nu$ must be 1. In~\cite{Jones2014}, the authors carefully attribute
+deviations from Fano = 1 to intensity variability in fluorescent spots, gene
+copy number variation, and copy number fluctuations of the transcription
+machinery, e.g., RNAP itself. In this picture, the master equation makes no
+appearance, and all the corrections to Poisson behavior are derived as additive
+corrections to the Fano factor. For disproving the ``universal noise curve''
+from So et al~\cite{So2011}, this picture was excellent. It is appealing in its
+simplicity, with only two parameters, the initiation rate $r$ and degradation
+rate $\gamma$. Since $\gamma$ is independently known from other experiments, and
+the mean mRNA copy number is $r/\gamma$, $r$ is easily inferred from data. In
+other words, the model is not excessively complex for the data at hand. But for
+many interesting questions, for instance in the recent work~\mmnote{cite
+Manuel's preprint once posted}, knowledge of means and variances alone is
+insufficient, and a description of the full distribution of molecular counts is
+necessary. For this we need a( slightly) more complex model than model 1 that
+would allow us to incorporate the non-Poissonian features of constitutive
+promoters directly into a master equation formulation.
 
 \subsubsection{Model 2 - Two-state promoter, RNAP bound or unbound.}
-This model was considered in, e.g.,~\cite{Phillips2015a} and ~\cite{Phillips2019}.
-Here transcription initiation proceeds from the bound to the unbound state,
-reflecting the microscopic reality that an RNAP that has begun to elongate
-a transcript is no longer available at the start site to begin another.
-The problem with this picture is that the Fano factor is
+This model was considered in, e.g.,~\cite{Phillips2015a} and
+~\cite{Phillips2019}. Here transcription initiation proceeds from the bound to
+the unbound state, reflecting the microscopic reality that an RNAP that has
+begun to elongate a transcript is no longer available at the start site to begin
+another. The problem with this picture is that the Fano factor is
 \begin{align}
@@
 \end{align}
-which is always $<1$.
-To make contact with the experimental reality of $\nu>1$,
-we will have to do, at a minimum,
-the same corrections to Poisson behavior as in model 1 above.
-So while this model adds an appealing element of microscopic
-reality, we are forced to reject it as the additional complexity
-is unable to capture the phenomenology of interest.
-Obviously the promoter state does in fact proceed through cycles
-of RNAP binding, initiating, and elongating, but it seems that
-the super-Poissonian noise in mRNA copy number we want to
-model must be governed by other features of the system.
+which is always $<1$. To make contact with the experimental reality of $\nu>1$,
+we will have to do, at a minimum, the same corrections to Poisson behavior as in
+model 1 above. So while this model adds an appealing element of microscopic
+reality, we are forced to reject it as the additional complexity is unable to
+capture the phenomenology of interest. Obviously the promoter state does in fact
+proceed through cycles of RNAP binding, initiating, and elongating, but it seems
+that the super-Poissonian noise in mRNA copy number we want to model must be
+governed by other features of the system.
 
-\subsubsection{Model 3 - Three-state promoter, multistep
-transcription initiation and escape.}
-How might we remedy the deficits of model 2?
-It is known~\cite{DeHaseth1998} that once RNAP initially binds
-the promoter region, a multitude of distinct steps occur
-sequentially before RNAP finally escapes into the elongation
-phase. Perhaps adding some of this mechanistic detail might
-rescue model 2? The next simplest refinement of model 2 could
-consider open complex formation and promoter escape as separate
-steps rather than as a single effective step. In other words, we
-construct model 3 by adding a single extra state to model 2, and
-we will label the two RNAP-bound states as the closed and open
-complexes, despite the true biochemical details certainly being
-more complex. The authors of~\cite{Mitarai2015}, for instance,
-considered this model, although they added an additional
-repressor bound state and did not explicitly consider the limit
-with no repressor that we analyze here. Again, our goal here is
-not a complete accounting of all the relevant biochemical detail;
-this is an exploratory search for the important features our
-model needs to include.
+\subsubsection{Model 3 - Three-state promoter, multistep transcription
+initiation and escape.} How might we remedy the deficits of model 2? It is
+known~\cite{DeHaseth1998} that once RNAP initially binds the promoter region, a
+multitude of distinct steps occur sequentially before RNAP finally escapes into
+the elongation phase. Perhaps adding some of this mechanistic detail might
+rescue model 2? The next simplest refinement of model 2 could consider open
+complex formation and promoter escape as separate steps rather than as a single
+effective step. In other words, we construct model 3 by adding a single extra
+state to model 2, and we will label the two RNAP-bound states as the closed and
+open complexes, despite the true biochemical details certainly being more
+complex. The authors of~\cite{Mitarai2015}, for instance, considered this model,
+although they added an additional repressor bound state and did not explicitly
+consider the limit with no repressor that we analyze here. Again, our goal here
+is not a complete accounting of all the relevant biochemical detail; this is an
+exploratory search for the important features our model needs to include.
 
-Unfortunately, as the authors of~\cite{Mitarai2015} hint, this
-model too has Fano factor $\nu<1$. This can be seen by starting
-with their Eq.~A1 for the Fano factor of the analogous model with
-repressor and taking the limit as repressor concentration goes to
-zero. After substantial algebra, the result is
+Unfortunately, as the authors of~\cite{Mitarai2015} hint, this model too has
+Fano factor $\nu<1$. This can be seen by starting with their Eq.~A1 for the Fano
+factor of the analogous model with repressor and taking the limit as repressor
+concentration goes to zero. After substantial algebra, the result is
 \begin{align}
@@
 \end{align}
-where we defined $\mathcal{Z} = r(k_O + k_P^-) + k_P^+(k_O + r)$
-for notational tidiness. This is necessarily less than 1 for
-arbitrary rate constants.
+where we defined $\mathcal{Z} = r(k_O + k_P^-) + k_P^+(k_O + r)$ for notational
+tidiness. This is necessarily less than 1 for arbitrary rate constants.
 
-In fact, we suspect \textit{any} model in which transcription
-proceeds through a multistep cycle must necessarily have $\nu<1$.
-The intuitive argument compares the waiting time distribution to
-traverse the cycle with the waiting time for a Poisson promoter
-(model 1) with the same mean time. The latter is simply an
-exponential distribution. The former is a convolution of multiple
-exponentials, and intuitively should be more peaked with a
-smaller fractional width than an exponential with the same mean.\footnote{
-This can be made more precise with a result
-from~\cite{Stewart2007}, who showed that the convolution of
-multiple gamma distributions (of which the exponential
-distribution is a special case) is to a very good approximation
-also gamma distributed. Using their Eq.~2 for the distribution of
-the convolution, with shape parameters set to 1 to
-give exponential distributions, the total waiting time
-distribution has a ratio of variance to squared mean 
-$\sigma^2/\mu^2 = \sum_i k_i^2/\left(\sum_i k_i\right)^2$,
-where the $k_i$ are the rates of the individual steps. Clearly
-this is less than 1 and therefore the total waiting time
-distribution is narrower than the corresponding exponential.} A
-less disperse waiting time distribution means transcription
-initations are more uniformly distributed in time relative to a
-Poisson process. Hence the distribution of mRNA over a population
-of cells should be less variable compared to Poisson, giving
-$\nu<1$.\footnote{
-This last step, while intuitive, can be argued by analogy to
+In fact, we suspect \textit{any} model in which transcription proceeds through a
+multistep cycle must necessarily have $\nu<1$. The intuitive argument compares
+the waiting time distribution to traverse the cycle with the waiting time for a
+Poisson promoter (model 1) with the same mean time. The latter is simply an
+exponential distribution. The former is a convolution of multiple exponentials,
+and intuitively should be more peaked with a smaller fractional width than an
+exponential with the same mean.\footnote{This can be made more precise with a
+result from~\cite{Stewart2007}, who showed that the convolution of multiple
+gamma distributions (of which the exponential distribution is a special case) is
+to a very good approximation also gamma distributed. Using their Eq.~2 for the
+distribution of the convolution, with shape parameters set to 1 to give
+exponential distributions, the total waiting time distribution has a ratio of
+variance to squared mean $\sigma^2/\mu^2 = \sum_i k_i^2/\left(\sum_i
+k_i\right)^2$, where the $k_i$ are the rates of the individual steps. Clearly
+this is less than 1 and therefore the total waiting time distribution is
+narrower than the corresponding exponential.} A less disperse waiting time
+distribution means transcription initations are more uniformly distributed in
+time relative to a Poisson process. Hence the distribution of mRNA over a
+population of cells should be less variable compared to Poisson, giving
+$\nu<1$.\footnote{This last step, while intuitive, can be argued by analogy to
 photon statistics where antibunching gives rise to sub-Poissonian
-noise~\cite{Paul1982, Zou1990}. Although loopholes exist, we do
-not expect they apply for our problem. Nevertheless we refrain
-from elevating this cycle/sub-Poissonian equivalence to a ``theorem.''
-\mmnote{Antibunching is the ``obvious'' analogy for me but it'd
-be totally out of left field for most readers, any suggestions on
-a more comprehensible reference??}
-}
+noise~\cite{Paul1982, Zou1990}. Although loopholes exist, we do not expect they
+apply for our problem. Nevertheless we refrain from elevating this
+cycle/sub-Poissonian equivalence to a ``theorem.'' \mmnote{Antibunching is the
+``obvious'' analogy for me but it'd be totally out of left field for most
+readers, any suggestions on a more comprehensible reference??}}
 
-Regardless of the merits of this model in describing the noise
-properties of constitutive transcription initation, it does not
-capture the dominant source of noise we want to include in our
-phenomenological model, so we must discard these details and
-search elsewhere.
+Regardless of the merits of this model in describing the noise properties of
+constitutive transcription initation, it does not capture the dominant source of
+noise we want to include in our phenomenological model, so we must discard these
+details and search elsewhere.
 
-\subsubsection{Model 4 - Two-state promoter,
-``active'' and ``inactive'' states.}
-Inspired by~\cite{Razo-Mejia2020}, we revisit an analog of model
-2, but the interpretation of the two states is changed. Rather
-than explicitly viewing them as RNAP bound and unbound, we view
-them as ``active'' and ``inactive,'' which are able and unable to
-initiate transcripts, respectively. We are noncommittal as to the
-microscopic details of these states.
+\subsubsection{Model 4 - Two-state promoter, ``active'' and ``inactive''
+states.} Inspired by~\cite{Razo-Mejia2020}, we revisit an analog of model 2, but
+the interpretation of the two states is changed. Rather than explicitly viewing
+them as RNAP bound and unbound, we view them as ``active'' and ``inactive,''
+which are able and unable to initiate transcripts, respectively. We are
+noncommittal as to the microscopic details of these states.
 
-One interpretation~\cite{Chong2014, Sevier2016} for the active
-and inactive states is that it represents the promoter's
-supercoiling state: transitions to the inactive state are caused
-by accumulation of positive supercoiling, which inhibits
-transcription, and transitions back to ``active'' are caused by
-gyrase or other topoisomerases relieving the supercoiling. This
-is an interesting possibility because it would mean the timescale
-for promoter state transitions is driven by topoisomerase
-kinetics, not by RNAP kinetics. From in vitro measurments, the
-former are suggested to be of order minutes~\cite{Chong2014}.
-Contrast this with model 2, where the state transitions are
-assumed to be governed by RNAP, which, assuming a copy number per
-cell of order $10^3$, has a diffusion-limited association rate
-$k_{on} \sim 10^2~\text{s}^{-1}$ to a target promoter. Combined
-with known $K_D$'s of order $\mu$M, this gives an RNAP
-dissociation rate $k_{off}$ of order $10^2$. As we will show
-below, however, there are some lingering puzzles with
-interpreting this supercoiling interpretation, so we leave it as
-a speculation and refrain from assigning definite physical
-meaning to the two states in this model.
+One interpretation~\cite{Chong2014, Sevier2016} for the active and inactive
+states is that it represents the promoter's supercoiling state: transitions to
+the inactive state are caused by accumulation of positive supercoiling, which
+inhibits transcription, and transitions back to ``active'' are caused by gyrase
+or other topoisomerases relieving the supercoiling. This is an interesting
+possibility because it would mean the timescale for promoter state transitions
+is driven by topoisomerase kinetics, not by RNAP kinetics. From in vitro
+measurments, the former are suggested to be of order minutes~\cite{Chong2014}.
+Contrast this with model 2, where the state transitions are assumed to be
+governed by RNAP, which, assuming a copy number per cell of order $10^3$, has a
+diffusion-limited association rate $k_{on} \sim 10^2~\text{s}^{-1}$ to a target
+promoter. Combined with known $K_D$'s of order $\mu$M, this gives an RNAP
+dissociation rate $k_{off}$ of order $10^2$. As we will show below, however,
+there are some lingering puzzles with interpreting this supercoiling
+interpretation, so we leave it as a speculation and refrain from assigning
+definite physical meaning to the two states in this model.
 
-Intuitively one might expect that, since transcripts are produced
-as a Poisson process only when the promoter is in one of the two
-states in this model, transcription initiations should now be
-``bunched,'' in constrast to the ``anti-bunching'' of models 2
-and 3 above. One might further guess that this bunching would
-lead to super-Poissonian noise in the mRNA distribution over a
+Intuitively one might expect that, since transcripts are produced as a Poisson
+process only when the promoter is in one of the two states in this model,
+transcription initiations should now be ``bunched,'' in constrast to the
+``anti-bunching'' of models 2 and 3 above. One might further guess that this
+bunching would lead to super-Poissonian noise in the mRNA distribution over a
 population of cells. An honest calculation of the Fano factor produces
@@
 \end{align}
-which is strictly greater than 1, verifying the above intuition.
-Note we have dropped the $P$ label on the promoter switching rates to
-emphasize that these very likely do not represent kinetics of RNAP itself.
-This calculation can also be sidestepped by noting that the model
-is mathematically equivalent to the simple repression repression model
-from~\cite{Jones2014}, with states and rates relabeled and reinterpreted.
+which is strictly greater than 1, verifying the above intuition. Note we have
+dropped the $P$ label on the promoter switching rates to emphasize that these
+very likely do not represent kinetics of RNAP itself. This calculation can also
+be sidestepped by noting that the model is mathematically equivalent to the
+simple repression repression model from~\cite{Jones2014}, with states and rates
+relabeled and reinterpreted.
 
-How does this model compare to model 1 above?
-In model 1, all non-Poisson features of the mRNA
-distribution were handled as extrinsic corrections.
-By contrast, here the 3 parameter model is used to fit the full
-mRNA distribution as measured in mRNA FISH experiments.
-In essense, all variability in the mRNA distribution is regarded
-as ``intrinsic,'' arising either from stochastic initiation or
-from switching between the two coarse-grained promoter states.
-The advantage of this approach is that it fits neatly into the
-master equation picture, and the parameters thus inferred can be used as
-input for more complicated models with regulation by transcription factors.
+How does this model compare to model 1 above? In model 1, all non-Poisson
+features of the mRNA distribution were handled as extrinsic corrections. By
+contrast, here the 3 parameter model is used to fit the full mRNA distribution
+as measured in mRNA FISH experiments. In essense, all variability in the mRNA
+distribution is regarded as ``intrinsic,'' arising either from stochastic
+initiation or from switching between the two coarse-grained promoter states. The
+advantage of this approach is that it fits neatly into the master equation
+picture, and the parameters thus inferred can be used as input for more
+complicated models with regulation by transcription factors.
 
-While this seems promising, there is a major drawback for our purposes
-which was already uncovered by the authors of~\cite{Razo-Mejia2020}:
-the statistical inference problem is nonidentifiable,
-in the sense described in Section 4.3 of~\cite{Gelman2013},
-meaning it is impossible to infer the
-parameters $r$ and $k^-$ from the FISH data of~\cite{Jones2014}
-(as shown in Fig.~S2 of~\cite{Razo-Mejia2020}).
-Rather, only the ratio $r/k^-$ could be inferred. In that work,
-the problem was worked around with an informative prior on the
-ratio $k^-/k^+$. That approach is unlikely to work here, as,
-recall, our entire goal in modeling constitutive expression is to
-use it as the basis for a yet more complicated model, when we add
-on repression. But adding more complexity to a model that is
-already poorly identified is a fool's errand,
-so we will explore one more potential model.
+While this seems promising, there is a major drawback for our purposes which was
+already uncovered by the authors of~\cite{Razo-Mejia2020}: the statistical
+inference problem is nonidentifiable, in the sense described in Section 4.3
+of~\cite{Gelman2013}, meaning it is impossible to infer the parameters $r$ and
+$k^-$ from the FISH data of~\cite{Jones2014} (as shown in Fig.~S2
+of~\cite{Razo-Mejia2020}). Rather, only the ratio $r/k^-$ could be inferred. In
+that work, the problem was worked around with an informative prior on the ratio
+$k^-/k^+$. That approach is unlikely to work here, as, recall, our entire goal
+in modeling constitutive expression is to use it as the basis for a yet more
+complicated model, when we add on repression. But adding more complexity to a
+model that is already poorly identified is a fool's errand, so we will explore
+one more potential model.
 
 \subsubsection{Model 5 - One-state promoter with explicit bursts.}
-This model is inspired by the failure mode of model 4.
-The key observation above was that, as found
-in~\cite{Razo-Mejia2020}, only two parameters, $k^+$ and the
+This model is inspired by the failure mode of model 4. The key observation above
+was that, as found in~\cite{Razo-Mejia2020}, only two parameters, $k^+$ and the
 ratio $r/k^-$, could be directly inferred from the FISH data
-from~\cite{Jones2014}.
-So let us take this seriously and imagine a model where these are
-the only two model parameters. What would this model look like?
+from~\cite{Jones2014}. So let us take this seriously and imagine a model where
+these are the only two model parameters. What would this model look like?
 
-To develop some intuition, consider model 4 in the limit
-$k^+ \ll k^- \lesssim r$, which is roughly satisfied by the
-parameters inferred in~\cite{Razo-Mejia2020}.
-In this limit, the system spends the majority of its time in the inactive state, occasionally becoming active and making a burst of transcripts.
-This should call to mind the well-known phenomenon of transcriptional bursting, as reported in, e.g.,~\cite{Golding2005,Chong2014,Sevier2016}\mmnote{should probably add some more cites here}.
-Let us make this correspondence more precise.
-The mean dwell time in the active state is $1/k^-$.
-While in this state, transcripts are produced at a rate $r$ per
-unit time. So on average, $r/k^-$ transcripts are produced before
-the system switches to the inactive state. Once in the inactive
-state, the system dwells there for an average time $1/k^+$ before
-returning to the active state and repeating the process. $r/k^-$
-resembles an average burst size, and $1/k^+$ resembles the time
-interval between burst events. More precisely, $1/k^+$ is the
-mean time between the end of one burst and the start of the next,
-whereas $1/k^+ + 1/k^-$ would be the mean interval between the
-start of two successive burst events, but in the limit $k^+ \ll
-k^-$, $1/k^+ + 1/k^- \approx 1/k^+$. Note that this limit ensures
-that the waiting time between bursts is approximately
-exponentially distributed, with mean set by the only timescale
-left in the problem,
-$1/k^+$.\footnote{
-If instead it were the case that $k^+ \sim k^-$, then the waiting time
-$1/k^+ + 1/k^-$ would have a peaked distribution, but this does
-not appear to be the case for any of datasets
+To develop some intuition, consider model 4 in the limit $k^+ \ll k^- \lesssim
+r$, which is roughly satisfied by the parameters inferred
+in~\cite{Razo-Mejia2020}. In this limit, the system spends the majority of its
+time in the inactive state, occasionally becoming active and making a burst of
+transcripts. This should call to mind the well-known phenomenon of
+transcriptional bursting, as reported in,
+e.g.,~\cite{Golding2005,Chong2014,Sevier2016}\mmnote{should probably add some
+more cites here}. Let us make this correspondence more precise. The mean dwell
+time in the active state is $1/k^-$. While in this state, transcripts are
+produced at a rate $r$ per unit time. So on average, $r/k^-$ transcripts are
+produced before the system switches to the inactive state. Once in the inactive
+state, the system dwells there for an average time $1/k^+$ before returning to
+the active state and repeating the process. $r/k^-$ resembles an average burst
+size, and $1/k^+$ resembles the time interval between burst events. More
+precisely, $1/k^+$ is the mean time between the end of one burst and the start
+of the next, whereas $1/k^+ + 1/k^-$ would be the mean interval between the
+start of two successive burst events, but in the limit $k^+ \ll k^-$, $1/k^+ +
+1/k^- \approx 1/k^+$. Note that this limit ensures that the waiting time between
+bursts is approximately exponentially distributed, with mean set by the only
+timescale left in the problem, $1/k^+$.\footnote{If instead it were the case
+that $k^+ \sim k^-$, then the waiting time $1/k^+ + 1/k^-$ would have a peaked
+distribution, but this does not appear to be the case for any of datasets
 from~\cite{Jones2014}.}
 
-Let us now verify this intuition with a precise derivation
-to check that $r/k^-$ is in fact the mean burst size
-and to obtain the full burst size distribution.
-Consider first a constant, known dwell time $T$ in the active state.
-Transcripts are produced at a rate $r$ per unit time, so the
-number of transcripts $n$ produced during $T$ fits exactly
-the ``story'' for the Poisson distribution, i.e.,
+Let us now verify this intuition with a precise derivation to check that $r/k^-$
+is in fact the mean burst size and to obtain the full burst size distribution.
+Consider first a constant, known dwell time $T$ in the active state. Transcripts
+are produced at a rate $r$ per unit time, so the number of transcripts $n$
+produced during $T$ fits exactly the ``story'' for the Poisson distribution,
+i.e.,
 \begin{equation}
@@
 \end{equation}
-Since the dwell time $T$ is unobservable, we actually want $P(n)$,
-the dwell time distribution with no conditioning on $T$.
-Basic rules of probability theory tell us we can write $P(n)$
-in terms of $P(n\mid T)$ as
+Since the dwell time $T$ is unobservable, we actually want $P(n)$, the dwell
+time distribution with no conditioning on $T$. Basic rules of probability theory
+tell us we can write $P(n)$ in terms of $P(n\mid T)$ as
 \begin{equation}
@@
 \end{equation}
-But we know the dwell time distribution $P(T)$,
-which is exponentially distributed according to
+But we know the dwell time distribution $P(T)$, which is exponentially
+distributed according to
 \begin{equation}
@@
 \end{equation}
-A standard integral table shows
-$\int_0^\infty x^n e^{-ax}\,dx = n!/a^{n+1}$, so
+A standard integral table shows $\int_0^\infty x^n e^{-ax}\,dx = n!/a^{n+1}$, so
 \begin{equation}
@@
 which is exactly the geometric distribution with standard parameter
-$\theta\equiv k^-/(k^- + r)$ and domain
-$n \in \{0, 1, 2, \dots\}$
-(this is one of two common conventions for the geometric distribution).
-The mean of the geometric distribution, with this convention, is
+$\theta\equiv k^-/(k^- + r)$ and domain $n \in \{0, 1, 2, \dots\}$ (this is one
+of two common conventions for the geometric distribution). The mean of the
+geometric distribution, with this convention, is
 \begin{align}
@@
 
-So in taking the limit $r,k^-\rightarrow\infty$, $r/k^-\equiv b$,
-we obtain a model which effectively has only a single promoter state,
-which initiates bursts at rate $k^+$
-(transitions to the active state, in the model 4 picture).
-The master equation for mRNA copy number $m$ takes the form
+So in taking the limit $r,k^-\rightarrow\infty$, $r/k^-\equiv b$, we obtain a
+model which effectively has only a single promoter state, which initiates bursts
+at rate $k^+$ (transitions to the active state, in the model 4 picture). The
+master equation for mRNA copy number $m$ takes the form
 \begin{align}
@@
 \end{align}
-where $Geom(n;b)$ is the geometric distribution with mean~$b$, i.e.,
-$Geom(n;b) = \frac{1}{1+b}\left(\frac{b}{1+b}\right)^n$
-(with domain over nonnegative integers as above).
-The first two terms are the usual mRNA degradation terms.
-The third term enumerates all ways the system can produce
-a burst of transcripts and arrive at copy number $m$,
-given that it had copy number $j$ before the burst.
-The fourth term allows the system to start with copy number $m$,
-then produce a burst and end with copy number $j$.
-In fact this last sum has trivial $j$ dependence and
-simply enforces normalization of the geometric distribution.
+where $Geom(n;b)$ is the geometric distribution with mean~$b$, i.e., $Geom(n;b)
+= \frac{1}{1+b}\left(\frac{b}{1+b}\right)^n$ (with domain over nonnegative
+integers as above). The first two terms are the usual mRNA degradation terms.
+The third term enumerates all ways the system can produce a burst of transcripts
+and arrive at copy number $m$, given that it had copy number $j$ before the
+burst. The fourth term allows the system to start with copy number $m$, then
+produce a burst and end with copy number $j$. In fact this last sum has trivial
+$j$ dependence and simply enforces normalization of the geometric distribution.
 Carrying it out we have
@@
 \end{equation}
-This improves on model 4 in that now the parameters are easily inferred,
-as we will see later, and have clean interpretations.
-The non-Poissonian features are attributed to the emprically
-well-established phenomenological picture of bursty transcription.
+This improves on model 4 in that now the parameters are easily inferred, as we
+will see later, and have clean interpretations. The non-Poissonian features are
+attributed to the emprically well-established phenomenological picture of bursty
+transcription.
 
-The big approximation in going from model 4 to 5 is that a burst
-is produced instantaneously rather than over a finite time.
-If the true burst duration is not short compared to transcription factor
-kinetic timescales, this could be a problem in that mean burst size
-in the presense and absence of repressors could change,
-rendering parameter inferences from the constitutive case inappropriate.
+The big approximation in going from model 4 to 5 is that a burst is produced
+instantaneously rather than over a finite time. If the true burst duration is
+not short compared to transcription factor kinetic timescales, this could be a
+problem in that mean burst size in the presense and absence of repressors could
+change, rendering parameter inferences from the constitutive case inappropriate.
 Let us make some simple estimates of this.
 
-Consider the time delay between the first and final RNAPs
-in a burst initiating transcription
-(\textit{not} the time to complete transcripts,
-which potentially could be much longer.)
-If this timescale is short compared to the typical
-search timescale of transcription factors, then all is well.
-The estimates from deHaseth et.\ al.~\cite{DeHaseth1998}
-put RNAP's diffusion-limited on rate around
-$\sim\text{few}\times10^{-2}~\text{nM}^{-1}~\text{s}^{-1}$
-and polymerase loading as high as $1~\text{s}^{-1}$.
-Then for reasonable burst sizes of $<10$, it is reasonable to guess that
-bursts might finish initiating on a timescale of tens of seconds or less
-(with another 30-60 sec to finish elongation, but that does not matter here).
-A transcription factor with typical copy number of order 10 (or less)
-would have a diffusion-limited association rate of order
-$(10~\text{sec})^{-1}$.
-Higher copy number TFs tend to have many binding sites over the genome,
-which should serve to pull them out of circulation and keep their
-effective association rates from rising too large.
-Therefore, there is \textit{perhaps} a timescale separation
-possible between transcription factor association rates and burst
-durations, but this assumption could very well break down, so we
-will have to keep it in mind when we infer repressor rates from
-the Jones et.\ al.\ FISH data later.
+Consider the time delay between the first and final RNAPs in a burst initiating
+transcription (\textit{not} the time to complete transcripts, which potentially
+could be much longer.) If this timescale is short compared to the typical search
+timescale of transcription factors, then all is well. The estimates from
+deHaseth et.\ al.~\cite{DeHaseth1998} put RNAP's diffusion-limited on rate
+around $\sim\text{few}\times10^{-2}~\text{nM}^{-1}~\text{s}^{-1}$ and polymerase
+loading as high as $1~\text{s}^{-1}$. Then for reasonable burst sizes of $<10$,
+it is reasonable to guess that bursts might finish initiating on a timescale of
+tens of seconds or less (with another 30-60 sec to finish elongation, but that
+does not matter here). A transcription factor with typical copy number of order
+10 (or less) would have a diffusion-limited association rate of order
+$(10~\text{sec})^{-1}$. Higher copy number TFs tend to have many binding sites
+over the genome, which should serve to pull them out of circulation and keep
+their effective association rates from rising too large. Therefore, there is
+\textit{perhaps} a timescale separation possible between transcription factor
+association rates and burst durations, but this assumption could very well break
+down, so we will have to keep it in mind when we infer repressor rates from the
+Jones et.\ al.\ FISH data later.
 
-In reflecting on these 5 models, the reader may feel that
-exploring a multitude of potential models just to return to a
-very minimal phenomenological model of bursty transcription may
-seem highly pedantic. But the purpose of the exercise was to
-examine a host of models from the literature and understand why
-they are insufficient, one way or another, for our purposes.
-Along the way we have learned that the detailed kinetics of RNAP
-binding and initiating transcription are probably irrelevant for
-setting the population distribution of mRNA.
-The timescales are simply too fast, and as we will see later
-in~\fig{fig:constit_post}, the noise seems to be governed by
-slower timescales. Perhaps in hindsight this is not surprising:
-intuitively, the degradation rate $\gamma$ sets the fundamental
-timescale for mRNA dynamics, and any other processes that
-substantially modulate the mRNA distribution should not differ
+In reflecting on these 5 models, the reader may feel that exploring a multitude
+of potential models just to return to a very minimal phenomenological model of
+bursty transcription may seem highly pedantic. But the purpose of the exercise
+was to examine a host of models from the literature and understand why they are
+insufficient, one way or another, for our purposes. Along the way we have
+learned that the detailed kinetics of RNAP binding and initiating transcription
+are probably irrelevant for setting the population distribution of mRNA. The
+timescales are simply too fast, and as we will see later
+in~\fig{fig:constit_post}, the noise seems to be governed by slower timescales.
+Perhaps in hindsight this is not surprising: intuitively, the degradation rate
+$\gamma$ sets the fundamental timescale for mRNA dynamics, and any other
+processes that substantially modulate the mRNA distribution should not differ
 from $\gamma$ by orders of magnitude.
-
-\subsection{Parameter inference for constitutive promoters}
-Having established that a one-state bursty model achieves a
-Goldilocks level of complexity, we test it more thoroughly on all
-the data from constitutive promoters from~\cite{Jones2014}.
-\fig{fig:constit_post} shows the results.
-\begin{figure}%[h!]
-\centering
-\includegraphics[width=0.9\textwidth]{../figures/fig2/fig2pt1.pdf}
-\caption{\textbf{Constitutive promoter posterior
-        inference and predictive checks.}
-        (A) shows the joint posterior density from MCMC sampling of model 5,
-        bursty promoter with negative binomially-distributed steady state.
-        (B) shows the ECDF of the observed population distribution
-        of mRNA transcripts under the control of a constitutive lacUV5 promoter.
-        Posterior predictive percentiles are shown as colored bands for
-        model (1), Poisson, and model (5), negative binomial.
-        }
-\label{fig:uv5_post}
-\end{figure}
-
-\begin{figure}%[h!]
-\centering
-\includegraphics[width=0.9\textwidth]{../figures/fig2/fig2pt2.pdf}
-\caption{\textbf{Constitutive promoter model comparison.}
-    (A) Joint posterior distributions for burst rate $k_i$ and
-    mean burst size $b$ for 18 promoters from Jones 2014.
-    Each contour indicates the 95\% highest posterior density
-    region for a particular promoter, i.e., on a plot analogous
-    to~\fig{fig:uv5_post}, the contour would enclose
-    approximately 95\% of the samples, and thus 95\% of the
-    probability mass.
-    Note that the vertical axis is shared with (B).
-    Panel (B) plots the burst rate $k_i$ vs.\ binding energy from energy
-    matrices from Brewster 2012.
-    The dotted line shows the predicted slope according to
-    $\ln k_i/\gamma \sim \text{constant} - \beta\Delta\epsilon_P$,
-    described in text.
-    }
-\label{fig:constit_post}
-\end{figure}
-\mmnote{Discuss puzzling comparison w/ Chong2014: if supercoiling is the thing, why are my burst sizes all the same but burst rates vary?
-And why is the duty cycle of my promoter so low, ie., bursts so short?
-COmpare their fig 7E; their $\beta/\alpha$ is my $k^+/k^-$.
-They have one or two genes with very small $\beta/\alpha$, does the \textit{galK} locus just happen to be that, or is there a deeper disagreement?
-Hard to say w/o more data.}
-
-\subsection{Transcription factor kinetics can be inferred from FISH measurements}
-Figure~3 presents inferred transcription factor rates, compares
-them to single-molecule measurements, and compares their ratios
-to gold-standard binding energies.
-\begin{figure}%[h!]
-\centering
-\includegraphics[width=0.9\textwidth]{../figures/fig3/fig3.ai}
-\caption{\textbf{Simple repression model comparison.}
-(A) depicts the cartoon of simple repression for which we infer parameters.
-(B) shows 50\% and 95\% contours of several 2D slices of the 9D posterior
-distribution, from fitting one unbinding rate for each operator (O1, O2, O3)
-and one binding rate for each aTc concentration (corresponding to an
-unknown mean repressor copy number).
-(C) compares ratios of our inferred unbinding rates with operator
-binding energy differences measured in Garcia \& Phillips 2011 (triangles)
-and Razo-Mejia et.\ al.\ 2018 (squares).
-(D) compares our inferred unbinding rates with single-molecule measurements
-of the same from Hammar et.\ at.\ 2014 \mmnote{Johan Elf's group}.
-Values are of similar order of magnitude,
-but a precise quantitative comparison unfortunately depends sensitively
-on the assumed mRNA lifetime, which was not measured.
-    }
-\label{fig3:kR_inferences}
-\end{figure}
-
-\begin{figure}%[h!]
-\centering
-\includegraphics[width=0.9\textwidth]{../figures/figSIxx/ppc_many_pooled.pdf}
-\caption{\textbf{Posterior predictive checks.}
-(A) duplicates the posterior from~\fig{fig3:kR_inferences}B.
-(B), (C), and (D) show samples from the posterior predictive distribution
-for each of the datasets used in the inference.
-Samples are sorted by operator and plotted separately simply for visual clarity.
-The unrepressed promoter, UV5, is shown with each as a reference point.
-\mmnote{
-Obviously the PPC is not perfect,
-but considering the relative simplicity of the model, I'm pleasantly surprised
-how well it worked and that the model was even identifiable at all.
-Interestingly, and perhaps not surprisingly, only 3 of the 9 datasets
-were identifiable if fit in isolation: otherwise,
-it was only by making assumptions that certain rates are equal
-across datasets was identifiability achieved.
-}
-    }
-\label{fig4:ppc}
-\end{figure}
-
-\section{Discussion and future work}
-\mmnote{
-Do all my approximations work simply because galK happens to have
-a low duty cycle of supercoiling, $\beta/\alpha$ in Chong2014
-picture? A potentially interesting test: put in the dozen or so
-Hernan constructs at some other locus, preferably one of the ones
-Chong identified as having a large duty cycle (large $\beta/\alpha$),
-i.e., the opposite limit of galK. Then just measure fold-changes.
-You don't even need to do FISH, although that would be even
-better. If their model is right, $\rho$ should be approximately
-set by the duty cycle of supercoiling, NOT by RNAP kinetics.
-\textit{This is a NICE prediction for future work.} \textit{cspE}
-is claimed nonessential, prediction is you drop HG reporter
-constructs in there and the inferred binding energies of all lacI
-operators shift the same amount, by about $\ln(3)\sim 1kT$. Not
-huge, but detectable, although might be confounding effects from
-moving to a new locus. With FISH measurements, $k^+$ and $k^-$ in
-Fig 2, model 4 should now be separately inferrable if Chong2014
-measurements of $\beta/\alpha$ are correct. This would require an
-obnoxious amount of cloning, but once you did that, the
-measurements would be same old stuff that RP lab does.}
-
-%%%%%%%%%%%%%%%%%%%%% APPENDICES %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\appendix
-
-\section{Derivations of constitutive promoter models}
-\subsection{Model 2}
-\mmnote{Obtain Fano factor from CME.}
-\subsection{Model 3}
-\mmnote{Obtain Fano factor by taking limit from Mitarai et al's expression.}
-\subsection{Model 4}
-\mmnote{Obtain Fano factor by relabeling cartoon from Jones2014.}
-
-\section{Generating function solution for bursty promoters}
-
-\subsection{Constitutive promoter with bursts}
-
-\subsubsection{From master equation to generating function}
-
-We would like to derive the full probability distribution of
-mRNA copy number per cell for a model combining simple repression
-with a promoter featuring bursty transcription initiation.
-Before tackling mRNA and repression together, let us
-just consider the bursty consititutive promoter,
-model 5 in~\fig{fig:constit_cartoons}
-\mmnote{replicating Charlotte's notes; she says it's worked out
-in~\cite{Paulsson2000}, but I don't see it, at least not in notation I can
-comprehend. Need to find a cite for this,
-or maybe just cite Charlotte, personal comm, since that's where I actually got it\dots}.
-The master equation of interest is
-\begin{align}
-\deriv{t}p(m,t) = (m+1)\gamma p(m+1,t) - m\gamma p(m,t) - r p(m,t)
-        + r \sum_{m^\prime=0}^m G_{m-m^\prime}(\theta) p(m^\prime,t),
-\label{eq:1state_unreg_003}
-\end{align}
-where $G_{k}(\theta)$ is the geometric distribution defined as
-\begin{align}
-G_{k}(\theta) = \theta(1 - \theta)^k, \, k\in\{0,1,2,\dots\}.
-\end{align}
-With this convention, the mean burst size $b = (1-\theta)/\theta$.
-$\gamma$ and $r$ are mRNA degradation rates and transcription burst
-initiation rates, resp. The last term represents all ways the system could
-end up with $m$ mRNAs, having started the burst with $m^\prime$. Define
-$\lambda = r/\gamma$ and nondimensionalize time by $\gamma$, giving
-\begin{align}
-\deriv{t}p(m,t) = (m+1)p(m+1,t) - m p(m,t) - \lambda p(m,t)
-        + \lambda \sum_{m^\prime=0}^m G_{m-m^\prime}(\theta) p(m^\prime,t),
-\end{align}
-The probability generating function is defined as
-\begin{align}
-F(z,t) = \sum_{m=0}^\infty z^m p(m,t).
-\end{align}
-Multiply both sides of the CME by $z^m$ and sum over all $m$ to get
-\begin{align}
-\pderiv[F]{t} = \sum_{m=0}^\infty z^m
-\left(
-(m+1)p(m+1,t) - m p(m,t) - \lambda p(m,t)
-+ \lambda \sum_{m^\prime=0}^m G_{m-m^\prime}(\theta) p(m^\prime,t)
-\right).
-\end{align}
-The first two terms can be rewritten as derivatives of the
-generating function, i.e.,
-\begin{align}
-\sum_{m=0}^\infty z^m (m+1)p(m+1,t)
-=
-\pderiv{z} \sum_{m=0}^\infty z^{m+1} p(m+1,t)
-=
-\pderiv[F]{z}
-\end{align}
-and
-\begin{align}
-\sum_{m=0}^\infty z^m mp(m,t)
-=
-z\pderiv{z} \sum_{m=0}^\infty z^m p(m,t)
-=
-z\pderiv[F]{z}.
-\end{align}
-The final term is the most trouble. The trick is to reverse
-the default order of the sums, as in
-\begin{align}
-&\lambda \sum_{m=0}^\infty \sum_{m^\prime=0}^m
-        z^m \theta(1 - \theta)^{m-m^\prime} p(m^\prime,t)
-\\=
-&\lambda\theta \sum_{m^\prime=0}^\infty z^{m^\prime} p(m^\prime,t)
-        \sum_{m=m^\prime}^\infty
-        z^{m-m^\prime} (1 - \theta)^{m-m^\prime}
-\\=
-&\frac{\lambda\theta F}{1 - (1 - \theta)z}.
-\end{align}
-This works because after reordering the sums, the sum on $m$ is just
-a geometric series. Its result is independent of $m^\prime$, so the sum on
-$m^\prime$ reduces to the definition of the generating function.
-Putting these results together, the PDE for the generation function is
-\begin{align}
-\pderiv[F]{t} = \pderiv[F]{z} - z\pderiv[F]{z} - \lambda F
-        + \frac{\lambda\theta F}{1-z(1-\theta)}.
-\end{align}
-Changing variables to $\xi=1-\theta$ and simplifying gives
-\begin{align}
-\pderiv[F]{t} + (z - 1) \pderiv[F]{z} = \frac{(z-1)\xi}{1-z\xi}\lambda F.
-\label{eq:1state_unreg_015}
-\end{align}
-\subsubsection{Stead-state}
-At steady-state, the PDE reduces to the ODE
-\begin{align}
-\deriv[F]{z} = \frac{\xi}{1-z\xi}\lambda F,
-\end{align}
-which we can integrate as
-\begin{align}
-\int \frac{dF}{F} = \int \frac{\lambda\xi dz}{1-\xi z}.
-\end{align}
-The initial conditions for generating functions can be subtle and confusing.
-The key fact follows from the definition
-$F(z,t) = \sum_m z^m p(m,t)$,
-normalization requires that
-$F(z=1^-,t) = \sum_m p(m,t) = 1$.\footnote{
-Sometimes the generating function may be undefined \textit{at} $z=1$ but the
-limit still holds. Also many authors change variables from $z$ to other
-things, so one must keep track of how the normalization condition transforms.
-}
-Doing the integrals (and producing constant $c$) gives
-\begin{align}
-\ln F &= -\lambda \ln(1-\xi z) + c
-\\
-F &= \frac{c}{(1-\xi z)^\lambda}.
-\end{align}
-Only one choice for $c$ can satisfy initial conditions, producing
-\begin{align}
-F(z) = \left(\frac{1-\xi}{1-\xi z}\right)^\lambda
-        = \left(\frac{\theta}{1 - z(1-\theta)}\right)^\lambda,
-\end{align}
-which is exactly the negative binomial's generating function, as expected.
-
-\subsection{Adding repression}
-\subsubsection{Deriving the generating function for mRNA distribution}
-
-Let us move from a one-state promoter to a two-state promoter, where one state
-has repressor bound and the other produces transcriptional bursts as above.
-A schematic of this model is shown in~\fig{fig3:kR_inferences}.
-Although now we have an equation for each promoter state, otherwise
-the master equation reads similarly to the one-state case, except with
-additional terms corresponding to transitions between promoter states, namely
-\begin{align}
-\deriv{t}p_R(m,t) =& k_R^+ p_A(m,t) - k_R^- p_R(m,t)
-        + (m+1)\gamma p_R(m+1,t) - m\gamma p_R(m,t)
-\\
-\begin{split}
-\deriv{t}p_A(m,t) =& - k_R^+ p_A(m,t) + k_R^- p_R(m,t)
-        + (m+1)\gamma p_A(m+1,t) - m\gamma p_A(m,t) 
-\\
-&- k_i p_A(m,t) + k_i \sum_{m^\prime=0}^m \theta(1-\theta)^{m-m^\prime} p_A(m^\prime,t),
-\end{split}
-\end{align}
-where $p_R(m,t)$ is the probability of the system having $m$ mRNA copies
-and having repressor bound to the promoter at time $t$, and $p_A$ is an
-analogous probability to find the promoter without repressor bound.
-$k_R+$ and $k_R^-$ are, respectively, the rates at which repressors
-bind and unbind to and from the promoter, and
-$\gamma$ is the mRNA degradation rate.
-$k_i$ is the rate at which bursts initiate, and as before,
-the geometric distribution of burst sizes has mean $b=(1-\theta)/\theta$.\footnote{
-$\theta$ can be viewed as a Bernoulli trial probability: after each transcript
-in a burst is produced, $\theta$ is the probability that the burst
-terminates and $(1-\theta)$ is the probability that the burst continues.
-}
-
-Interestingly, it turns out that this problem maps exactly onto the
-three-stage promoter model considered by
-Shahrezaei and Swain in~\cite{Shahrezaei2008}, with relabelings.
-Their approximate solution for protein distributions amounts to
-the same approximation we make here in regarding the duration
-of mRNA synthesis bursts as instantaneous, so their solution
-for protein distributions also solves our problem of mRNA distributions.
-Let us examine the analogy more closely.
-They consider a two-state promoter, as we do here, but they model
-mRNA as being produced one at a time and degraded, with rates
-$v_0$ and $d_0$.
-Then they model translation as occuring with rate $v_1$,
-and protein degradation with rate $d_1$.
-Now consider the limit where $v_1, d_0\rightarrow\infty$ with
-their ratio $v_1/d_0$ held constant.
-$v_1/d_0$ resembles the average burst size of translation from a single
-mRNA: these are the rates of two Poisson processes that compete over a
-transcript, which matches the story of geometrically distributed burst sizes.
-So after taking this limit, on timescales slow compared to $v_1$ and $d_0$,
-it appears that transcription events fire at rate $v_0$ and produce a
-geometrically distributed burst of translation of mean size $v_1/d_0$,
-which intuitively matches the story we have told above for mRNA
-with variables relabeled.
-
-To verify this intuitively conjectured mapping between our problem and
-the solution in~\cite{Shahrezaei2008}, we continue with a careful
-solution for the mRNA distribution using probability generating functions,
-following the ideas sketched in~\cite{Shahrezaei2008}.
-It is natural to nondimensionalize rates in the problem by $\gamma$,
-or equivalently, this amounts to measuring time in units of $\gamma^{-1}$.
-We are also only interested in steady state, so we set the time
-derivatives to zero, giving
-\begin{align}
-0 =& k_R^+ p_A(m) - k_R^- p_R(m) + (m+1) p_R(m+1) - m p_R(m)
-\\
-\begin{split}
-0 =& - k_R^+ p_A(m) + k_R^- p_R(m) + (m+1) p_A(m+1) - m p_A(m) 
-\\
-&- k_i p_A(m) + k_i \sum_{m^\prime=0}^m \theta(1-\theta)^{m-m^\prime} p_A(m^\prime),
-\end{split}
-\end{align}
-
-The probability generating function is defined as before in the
-constitutive case, except now we must introduce a generating function
-for each promoter state,
-\begin{align}
-f_A(z) = \sum_{m=0}^\infty z^m p_A(m),
-\;
-f_R(z) = \sum_{m=0}^\infty z^m p_R(m).
-\end{align}
-Our real objective is the generating function $f(z)$ that generates
-the mRNA distribution $p(m)$, independent of what state the promoter is in.
-But since $p(m) = p_A(m) + p_R(m)$, it follows too that $f(z) = f_A(z) + f_R(z)$.
-
-So as before we multiply both equations by $z^m$ and sum over all $m$.
-Each individual term transforms exactly as did an analogous term in
-the constitutive case, so the coupled ODEs for the generating functions read
-\begin{align}
-0 =& k_R^+ f_A(z) - k_R^- f_R(z) + \pderiv{z} f_R(z) - z \pderiv{z} f_R(z)
-\\
-\begin{split}
-0 =&  - k_R^+ f_A(z) + k_R^- f_R(z) + \pderiv{z} f_A(z) - z \pderiv{z} f_A(z)
-\\
-&- k_i f_A(z) + k_i \frac{\theta}{1-z(1-\theta)} f_A(z),
-\end{split}
-\end{align}
-and after changing variables $\xi = 1 - \theta$ as before
-and rearranging, we have
-\begin{align}
-0 &= k_R^+ f_A(z) - k_R^- f_R(z) + (1-z) \pderiv{z} f_R(z)
-\\
-0 &=  - k_R^+ f_A(z) + k_R^- f_R(z) + (1 - z) \pderiv{z} f_A(z)
-+ k_i \frac{(z-1)\xi}{1-z\xi} f_A(z),
-\end{align}
-We can transform this problem from two coupled first-order ODEs to a
-single second-order ODE by solving for $f_A$ in the first and plugging
-into the second, giving
-\begin{align}
-\begin{split}
-0 = (1&-z) \pderiv[f_R]{z}
-+ \frac{1-z}{k_R^+}
-        \left(k_R^- \pderiv[f_R]{z} + \pderiv[f_R]{z} +(z-1) \psecderiv[f_R]{z}\right)
-\\
-&+ \frac{k_i}{k_R^+} \frac{(z-1)\xi}{1-z\xi}
-        \left(k_R^- f_R + (z-1) \pderiv[f_R]{z}\right),
-\end{split}
-\end{align}
-where, to reduce notational clutter, we have dropped the explicit $z$
-dependence of $f_A$ and $f_R$. Simplifying we have
-\begin{align}
-0 = \psecderiv[f_R]{z}
-        - \left(\frac{k_i\xi}{1-z\xi}
-                + \frac{1 + k_R^- + k_R^+}{1-z}
-        \right)\pderiv[f_R]{z}
-        + \frac{k_i k_R^- \xi}{(1-z\xi)(1-z)}f_R.
-\end{align}
-This can be recognized as the hypergeometric differential equation,
-with singularities at $z=1$, $z=\xi^{-1}$, and $z=\infty$. The latter
-can be verified by a change of variables from $z$ to $x=1/z$,
-being careful with the chain rule, and noting that $z=\infty$ is a
-singular point if and only if $x=1/z=0$ is a singular point.
-
-The standard form of the hypergeometric differential equation has its
-singularities at 0, 1, and $\infty$, so to take advantage of the
-standard form solutions to this ODE, we first need to transform
-variables to put it into a standard form.
-However, this is subtle. While any such transformation should work in
-principle, the solutions are expressed most simply in the neighborhood
-of $z=0$, but the normalization condition that we need to enforce
-corresponds to $z=1$. The easiest path, therefore, is to find a change
-of variables that maps
-1 to 0, $\infty$ to $\infty$, and $\xi^{-1}$ to 1.
-This is most intuitively done in two steps.
-
-First map the $z=1$ singularity to 0 by the change of variables $v=z-1$, giving
-\begin{align}
-0 = \psecderiv[f_R]{v}
-        + \left(\frac{k_i\xi}{(1+v)\xi - 1}
-                + \frac{1 + k_R^- + k_R^+}{v}
-        \right)\pderiv[f_R]{v}
-        + \frac{k_i k_R^- \xi}{((1+v)\xi - 1)v}f_R.
-\end{align}
-Now two singularities are at $v=0$ and $v=\infty$.
-The third is determined by $(1+v)\xi -1 = 0$, or $v=\xi^{-1} - 1$.
-We want another variable change that maps this third singularity to 1
-(without moving 0 or infinity). Changing variables again to
-$w=\frac{v}{\xi^{-1} - 1} = \frac{\xi}{1-\xi} v$ fits the bill.
-In other words, the combined change of variables
-\begin{align}
-w = \frac{\xi}{1-\xi} (z-1)
-\end{align}
-maps $z = \{1, \xi^{-1}, \infty\}$ to $w =\{0, 1, \infty\}$ as desired.
-Plugging in, being mindful of the chain rule and noting
-$(1 + v)\xi - 1 = (1 - \xi)(w - 1)$ gives
-\begin{align}
-0 = \left(\frac{\xi}{1-\xi}\right)^2 \psecderiv[f_R]{w}
-+ \left(
-        \frac{\xi k_i}{(1-\xi)(w-1)} + \frac{\xi(1 + k_R^- + k_R^+)}{(1-\xi)w}
-\right) \frac{\xi}{1-\xi} \pderiv[f_R]{w}
-+ \frac{k_i k_R^- \xi^2}{(1-\xi)^2 w(w-1)}f_R.
-\end{align}
-This is close to the standard form of the hypergeometric differential
-equation, and some cancellation and rearrangement gives
-\begin{align}
-0 = w(w-1)\psecderiv[f_R]{w}
-+ \left(k_i w + (1 + k_R^- + k_R^+)(w-1)\right) \pderiv[f_R]{w}
-+ k_i k_R^- f_R.
-\end{align}
-and a little more algebra produces
-\begin{align}
-0 = w(1-w)\psecderiv[f_R]{w}
-+ \left(1 + k_R^- + k_R^+
-        - (1 + k_i + k_R^- + k_R^+)w
-\right) \pderiv[f_R]{w}
-- k_i k_R^- f_R,
-\end{align}
-which is the standard form. From this we can read off the solution
-in terms of hypergeometric functions ${_2F_1}$ from
-any standard source, e.g.~\cite{Abramowitz1964}, and identify the
-conventional parameters in terms of our model parameters.
-We want the general solution in the neighborhood of $w=0$ ($z=1$),
-which for a homogeneous linear second order ODE must be a sum of two
-linearly independent solutions.
-More precisely,
-\begin{align}
-f_R(w) = C_1 {_2F_1}(\alpha, \beta, \gamma; w)
-+ C_2 w^{1-\gamma}{_2F_1}(1+\alpha-\gamma, 1+\beta-\gamma, 2-\gamma; w)
-\end{align}
-with parameters determined by
-\begin{align}
-\begin{split}
-\alpha\beta &= k_i k_R^-
-\\
-1+\alpha+\beta &= 1+k_i+k_R^-+k_R^+
-\\
-\gamma &= 1 + k_R^- + k_R^+
-\end{split}
-\end{align}
-and constants $C_1$ and $C_2$ to be set by boundary conditions.
-Solving for $\alpha$ and $\beta$, we find\footnote{
-Note that $\alpha$ and $\beta$ are interchangeable in the definition of
-${_2F_1}$ and differ only in the sign preceeding the radical.
-}
-\begin{align}
-\begin{split}
-\alpha &= \frac{1}{2}
-\left(k_i+k_R^-+k_R^+ + \sqrt{(k_i+k_R^-+k_R^+)^2 - 4k_i k_R^-}\right)
-\\
-\beta &= \frac{1}{2}
-\left(k_i+k_R^-+k_R^+ - \sqrt{(k_i+k_R^-+k_R^+)^2 - 4k_i k_R^-}\right)
-\\
-\gamma &= 1 + k_R^- + k_R^+.
-\end{split}
-\end{align}
-Since the normalization condition requires that $f_R$ be finite at $w=0$,
-we can immediately set $C_2=0$ to discard the second solution.
-This is because all the rate constants are strictly positive,
-so $\gamma>1$ and therefore $w^{1-\gamma}$ blows up as $w\rightarrow0$.
-Now that we have $f_R$, we would like to find the generating function
-for the mRNA distribution, $f(z) = f_A(z) + f_R(z)$.
-We can recover $f_A$ from our solution for $f_R$, namely
-\begin{align}
-f_A(z) = \frac{1}{k_R^+}\left(k_R^- f_R(z) + (z-1) \pderiv[f_R]{z}\right)
-\end{align}
-or
-\begin{align}
-f_A(w) = \frac{1}{k_R^+}\left(k_R^- f_R(w) + w \pderiv[f_R]{w}\right),
-\end{align}
-where in the second line we transformed our original relation between
-$f_R$ and $f_A$ to our new, more convenient, variable $w$.
-Plugging our solution for $f_R(w) = C_1{_2F_1}(\alpha, \beta, \gamma; w)$
-into $f_A$, we will require the differentiation rule for ${_2F_1}$,
-which tells us
-\begin{align}
-\pderiv[f_R]{w} = C_1\frac{\alpha\beta}{\gamma}
-                {_2F_1}(\alpha+1, \beta+1, \gamma+1; w),
-\end{align}
-from which it follows that
-\begin{align}
-f_A(w) = \frac{C_1}{k_R^+}
-\left(
-k_R^- {_2F_1}(\alpha, \beta, \gamma; w)
-+ w\frac{\alpha\beta}{\gamma} {_2F_1}(\alpha+1, \beta+1, \gamma+1; w)
-\right)
-\end{align}
-and therefore
-\begin{align}
-f(w) = C_1\left(1 + \frac{k_R^-}{k_R^+}\right)
-        {_2F_1}(\alpha, \beta, \gamma; w)
-+ w \frac{C_1}{k_R^+} \frac{\alpha\beta}{\gamma}
-        {_2F_1}(\alpha+1, \beta+1, \gamma+1; w).
-\end{align}
-To proceed, we need one of the (many) useful identities known for
-hypergeometric functions, in particular
-\begin{align}
-w\frac{\alpha\beta}{\gamma} {_2F_1}(\alpha+1, \beta+1, \gamma+1; w)
-=
-(\gamma-1)\left(
-{_2F_1}(\alpha, \beta, \gamma-1; w) - {_2F_1}(\alpha, \beta, \gamma; w)
-\right).
-\end{align}
-Substituting this for the second term in $f(w)$, we find
-\begin{align}
-f(w) = \frac{C_1}{k_R^+}
-\left[
-        \left(k_R^+ + k_R^-\right)
-        {_2F_1}(\alpha, \beta, \gamma; w)
-+ (\gamma-1)\left(
-        {_2F_1}(\alpha, \beta, \gamma-1; w) - {_2F_1}(\alpha, \beta, \gamma; w)
-        \right)
-\right],
-\end{align}
-and since $\gamma-1 = k_R^+ + k_R^-$, the first and third terms cancel,
-leaving only
-\begin{align}
-f(w) = C_1\frac{k_R^+ + k_R^-}{k_R^+} {_2F_1}(\alpha, \beta, \gamma-1; w).
-\end{align}
-Now we enforce normalization, demanding $f(w=0) = f(z=1) = 1$.
-${_2F_1}(\alpha, \beta, \gamma-1; 0) = 1$, so we must have
-$C_1 = k_R^+ / (k_R^+ + k_R^-)$ and consequently
-\begin{align}
-f(w) =  {_2F_1}(\alpha, \beta, k_R^+ + k_R^-; w).
-\end{align}
-Recalling that the mean burst size $b = (1-\theta)/\theta = \xi/(1-\xi)$
-and $w = \frac{\xi}{1-\xi} (z-1) = b (z-1)$,
-we can transform back to the original variable $z$ to find the tidy result
-\begin{align}
-f(z) =  {_2F_1}(\alpha, \beta, k_R^+ + k_R^-; b(z-1)),
-\end{align}
-with $\alpha$ and $\beta$ given above by
-\begin{align}
-\begin{split}
-\alpha &= \frac{1}{2}
-\left(k_i+k_R^-+k_R^+ + \sqrt{(k_i+k_R^-+k_R^+)^2 - 4k_i k_R^-}\right)
-\\
-\beta &= \frac{1}{2}
-\left(k_i+k_R^-+k_R^+ - \sqrt{(k_i+k_R^-+k_R^+)^2 - 4k_i k_R^-}\right).
-\end{split}
-\end{align}
-Finally we are in sight of the original goal. We can generate the
-steady-state probability distribution of interest by
-differentiating the generating function,
-\begin{align}
-p(m) = m! \left.\frac{\partial^m}{\partial z^m} f(z) \right|_{z=0},
-\end{align}
-which follows easily from its definition. Some contemplation reveals
-that repeated application of the derivative rule used above will produce
-products of the form $\alpha(\alpha+1)(\alpha+2)\cdots(\alpha+m-1)$ in
-the expression for $p(m)$ and similarly for $\beta$ and $\gamma$. These
-resemble ratios of factorials, but since $\alpha$, $\beta$, and $\gamma$
-are not necessarily integer, we should express the ratios using gamma
-functions instead. More precisely, one finds
-\begin{align}
-p(m) = \frac{
-        \Gamma(\alpha + m)\Gamma(\beta + m)\Gamma(k_R^+ + k_R^-)
-        }
-        {
-        \Gamma(\alpha)\Gamma(\beta)\Gamma(k_R^+ + k_R^- + m)
-        }
-\frac{b^m}{m!}{_2F_1}(\alpha+m, \beta+m, k_R^++k_R^-+m; -b)
-\label{eq:p_m_bursty+rep}
-\end{align}
-which is finally the probability distribution we sought to derive.
-
-\subsubsection{Moments}
-We can verify the correctness of~\eq{eq:p_m_bursty+rep}
-by comparison with Gillespie simulation.
-We can also easily compute moments of the distribution from
-the generating function, and check these against raw moments computed
-directly from the master equation.
-
-The $j$-th raw moment of the distribution can be computed from the
-generating function from the formula
-\begin{align}
-\langle m^j \rangle
-= \left. \left(z \frac{\partial}{\partial z}\right)^j f(z)\right|_{z=1}.
-\end{align}
-Note that \textit{probabilities} are generated by differentiating
-and evaluating at $z=0$, while \textit{moments} are generated by
-differentiating and evaluating at $z=1$. This also aligns with the
-normalization constraint, which can be thought of as the zeroth raw moment.
-
-The mean mRNA is given by
-\begin{align}
-\langle m \rangle = \left.z \pderiv[f]{z} \right|_{z=1}
-= \frac{\alpha\beta b}{k_R^+ + k_R^-}
-= k_i b\frac{k_R^-}{k_R^+ + k_R^-},
-\end{align}
-which is intuitive and appealingly simple:
-the mean mRNA is just the mean mRNA in the absense of repression
-($k_i b$, the burst rate time mean burst size)
-times the fraction of time the repressor is bound, $k_R^-/(k_R^+ + k_R^-)$.
-
-To compute the variance, let us first compute
-\begin{align}
-\begin{split}
-\langle m^2\rangle
-&= \left .z\pderiv{z}\left(z\pderiv[f]{z}\right) \right|_{z=1}
-= \left. z\left(\pderiv[f]{z} + z\psecderiv[f]{z}\right)\right|_{z=1}
-\\
-&= \frac{\alpha\beta b}{k_R^+ + k_R^-}
-        + \frac{\alpha(\alpha+1)\beta(\beta+1) b^2}
-                {(k_R^+ + k_R^-)(k_R^+ + k_R^- + 1)}
-\\
-&= \frac{k_i b k_R^-}{k_R^+ + k_R^-}
-        + \frac{k_i k_R^- b^2 (1 + k_i + k_R^+ + k_R^- + k_i k_R^-)}
-                {(k_R^+ + k_R^-)(k_R^+ + k_R^- + 1)},
-\end{split}
-\end{align}
-from which the variance is
-\begin{align}
-\begin{split}
-var(m) = \langle m^2\rangle - \langle m\rangle^2
-&= \frac{k_i b k_R^-}{k_R^+ + k_R^-}
-\left(
-1 + \frac{b(1 + k_i + k_R^+ + k_R^- + k_i k_R^-)} {k_R^+ + k_R^- + 1}
-- \frac{k_i b k_R^-}{k_R^+ + k_R^-}
-\right)
-\\
-&= k_i b\frac{k_R^-}{k_R^+ + k_R^-}
-\left(
-1 + b + k_i b\frac{1 + k_R^-} {k_R^+ + k_R^- + 1}
-- \frac{k_i b k_R^-}{k_R^+ + k_R^-}
-\right)
-\\
-&= k_i b\frac{k_R^-}{k_R^+ + k_R^-}
-\left(
-1 + b + k_i b\frac{k_R^+}{(k_R^+ + k_R^-)(k_R^+ + k_R^- + 1)}
-\right).
-\end{split}
-\end{align}
-The Fano factor is the quantity in parentheses. This can be interpreted as
-the variability arising from the negative binomial distribution of the
-unregulated promoter $(1+b)$ plus additional noise arising from repression.
-
-\mmnote{Still to writeup: moment derivation direct from CME to compare}
-
-\subsection{Numerical considerations and recursion formulas}
-\subsubsection{Generalities}
-We would like to carry out Bayesian parameter inference on FISH data
-from~\cite{Jones2014}, using~\eq{eq:p_m_bursty+rep} as our
-likelihood. This requires accurate (and preferably fast)
-numerical evaluation of the hypergeometric function ${_2F_1}$,
-which is a notoriously hard problem~\cite{Pearson2017, Gil2007},
-and our particular needs here present an especial challenge as we show below.
-
-The hypergeometric function is defined by its Taylor series as
-\begin{align}
-{_2F_1}(a,b,c;z) 
-= \sum_{l=0}^\infty
-\frac{\Gamma(a + l)\Gamma(b + l)\Gamma(c)}
-        {\Gamma(a)\Gamma(b)\Gamma(c + l)}
-\frac{z^l}{l!}
-\end{align}
-for $|z|<1$, and by analytic continuation elsewhere.
-If $z\lesssim1/2$ and $\alpha$ and $\beta$ are not too large
-(absolute value below 20 or 30),
-then the series converges quickly and an accurate numerical representation is
-easily computed by truncating the series after a reasonable number of terms.
-Unfortunately, we need to evaluate ${_2F_1}$ over mRNA copy numbers fully out
-to the tail of the distribution, which can easily reach 50, possibly 100.
-From~\eq{eq:p_m_bursty+rep}, this means evaluating ${_2F_1}$
-repeatedly for values of $a$, $b$, and $c$ spanning the full range
-from $\mathcal{O}(1)$ to $\mathcal{O}(10^2)$,
-even if $\alpha$, $\beta$, and $\gamma$
-in~\eq{eq:p_m_bursty+rep} are small,
-with the situation even worse if they are not small.
-A naive numerical evaluation of the series definition will be
-prone to overflow and, if any of $a,b,c<0$, then some successive terms in the
-series have alternating signs which can lead to catastrophic cancellations.
-
-One solution is to evaluate ${_2F_1}$ using arbitrary precision arithmetic
-instead of floating point arithmetic,
-e.g., using the \texttt{mpmath} library in Python.
-This is accurate but incredibly slow computationally.
-To quantify how slow, we found that
-evaluating the likelihood defined by~\eq{eq:p_m_bursty+rep} $\sim50$ times
-(for a typical dataset of interest from~\cite{Jones2014},
-with $m$ values spanning 0 to $\sim50$)
-using arbitrary precision arithmetic is 100-1000 fold slower than
-evaluating a negative binomial likelihood for the corresponding
-constitutive promoter dataset.
-
-To claw back $\gtrsim30$ fold of that slowdown, we can exploit
-one of the many catalogued symmetries involving ${_2F_1}$.
-The solution involves recursion relations originally explored by Gauss,
-and studied extensively in~\cite{Pearson2017, Gil2007}.
-They are sometimes known as contiguous relations and relate the values
-of any set of 3 hypergeometric functions whose arguments differ by integers.
-To rephrase this symbolically, consider a set of hypergeometric functions
-indexed by an integer $n$,
-\begin{align}
-f_n = {_2F_1}(a+\epsilon_i n, b+\epsilon_j n, c+\epsilon_k n; z),
-\end{align}
-for a fixed choice of $\epsilon_i, \epsilon_j, \epsilon_k \in \{0,\pm 1\}$
-(at least one of $\epsilon_i, \epsilon_j, \epsilon_k$ must be nonzero,
-else the set of $f_n$ would contain only a single element).
-Then there exist known recurrence relations of the form
-\begin{align}
-A_n f_{n-1} + B_n f_{n} + C_n f_{n+1} = 0,
-\end{align}
-where $A_n, B_n$, and $C_n$ are some functions of $a,b,c$, and $z$.
-In other words, for fixed $\epsilon_i, \epsilon_j, \epsilon_k, a, b,$ and $c$,
-if we can merely evaluate ${_2F_1}$ twice, say for $n^\prime$ and $n^\prime-1$,
-then we can easily and rapidly generate values for arbitrary $n$.
-
-This provides a convenient solution for our problem: we need repeated
-evaluations of ${_2F_1}(a+m, b+m, c+m; z)$
-for fixed $a,b$, and $c$ and many integer values of $m$.
-They idea is that we can use arbitrary precision arithmetic to evaluate
-${_2F_1}$ for just two particular values of $m$ and then generate
-${_2F_1}$ for the other 50-100 values of $m$ using the recurrence
-relation.\footnote{
-There are even more sophisticated ways of utilizing the recurrence
-relations that might have netted another factor of 2 speed-up, and
-possibly as much as a factor of 10, but the method described here had
-already reduced the computation time to an acceptable
-$\mathcal{O}(\text{1 min})$, so these more sophisticated approaches did
-not seem worth the time to pursue.
-}
-
-However, there two further wrinkles.
-The first is that
-a naive application of the recurrence relation is numerically unstable.
-Roughly, this is because the three term recurrence relations,
-like second order ODEs, admit two linearly independent solutions.
-In a certain eigenbasis, one of these solutions dominates the other
-as $n\rightarrow\infty$, and as $n\rightarrow-\infty$,
-the dominance is reversed.
-If we fail to work in this eigenbasis, our solution of the recurrence relation
-will be a mixture of these solutions and rapidly accumulate numerical error.
-For our purposes, it suffices to know that the authors of~\cite{Gil2007}
-derived the numerically stable solutions (so-called \textit{minimal solutions})
-for several possible choices of $\epsilon_i, \epsilon_j, \epsilon_k$.
-Running the recurrence in the proper direction using a minimal solution
-is numerically robust and can be done entirely in floating point arithmetic, 
-so that we only need to evaluate ${_2F_1}$ with arbitrary precision arithmetic
-to generate the seed values for the recursion.
-
-The second wrinkle is a corollary to the first.
-The minimal solutions are only minimal for certain ranges of the argument $z$,
-and not all of the 26 possible recurrence relations
-have minimal solutions for all $z$.
-This can be solved by using one of the many transformation formulae for
-${_2F_1}$ to convert to a different recurrence relation that has
-a minimal solution over the required domain of $z$, although
-this can require some trial and error to find the right transformation,
-the right recurrence relation, and the right minimal solution.
-
-\subsubsection{Particulars}
-Let us now demonstrate these generalities for our problem of interest.
-In order to evaluate the probability distribution of our
-model,~\eq{eq:p_m_bursty+rep}, we need to evaluate hypergeometric functions
-of the form ${_2F_1}(\alpha+m, \beta+m, \gamma+m; -b)$
-for values of $m$ ranging from $0$ to $\mathcal{O}(100)$.
-The authors of~\cite{Gil2007} did not derive a recursion relation
-for precisely this case. We could follow their methods and do so ourselves,
-but it is much easier to convert to a case that they did consider.
-The strategy is to look through the minimal solutions tabulated
-in~\cite{Gil2007} and search for a transformation we could apply to
-${_2F_1}(\alpha+m, \beta+m, \gamma+m; -b)$ that would place the $m$'s
-(the variable being incremented by the recursion)
-in the same arguments of ${_2F_1}$ as the minimal solution.
-After some ``guess and check,'' we found that the transformation
-\begin{align}
-{_2F_1}(\alpha+m, \beta+m, \gamma+m; -b)
-=
-(1+b)^{-\alpha-m}
-        {_2F_1}\left(\alpha+m, \gamma-\beta, \gamma+m; \frac{b}{1+b}\right),
-\label{eq:rec_euler_pretransform}
-\end{align}
-produces a ${_2F_1}$ on the right hand side that closely resembles
-the minimal solutions $y_{3,m}$ and $y_{4,m}$ in Eq.~4.3 in~\cite{Gil2007}.
-Explicitly, these solutions are
-\begin{align}
-y_{3,m}
-&\propto
-{_2F_1}\left(-\alpha^\prime + \gamma^\prime - m,
-                -\beta^\prime + \gamma^\prime,
-                1-\alpha^\prime-\beta^\prime+\gamma^\prime-m;
-                1-z\right)
-\\
-y_{4,m}
-&\propto
-{_2F_1}\left(\alpha^\prime + m,
-                \beta^\prime,
-                1+\alpha^\prime+\beta^\prime-\gamma^\prime+m;
-                1-z\right),
-\label{eq:minimal_soln_sans_prefac}
-\end{align}
-where we have omitted prefactors which are unimportant for now.
-Which of these two we should use depends on what values $z$ takes on.
-Equating $1-z=b/(1+b)$ gives $z=1/(1+b)$, and since $b$ is strictly positive,
-$z$ is bounded between 0 and 1.
-From Eq.~4.5 in~\cite{Gil2007}, $y_{4,m}$ is the minimal solution
-for real $z$ satisfying $0<z<2$, so this is the only minimal solution we need.
-
-Now that we have our minimal solution,
-what recurrence relation does it satisfy?
-Confusingly, the recurrence relation of which $y_{4,m}$ is a solution
-increments different arguments of ${_2F_1}$ that does $y_{4,m}$:
-it increments the first only, rather than first and third.
-This recurrence relation can be looked up, e.g., Eq.~15.2.10
-in~\cite{Abramowitz1964}, which is
-\begin{align}
-(\gamma^\prime - (\alpha^\prime + m)) f_{m-1}
-+
-(2(\alpha^\prime+m) - \gamma^\prime + (\beta^\prime - \alpha^\prime)z)f_m
-+ \alpha^\prime(z-1) f_{m+1} = 0.
-\label{eq:chosen_rec_rel}
-\end{align}
-Now we must solve for the parameters appearing in the recurrence relation
-in terms of our parameters, namely by setting
-\begin{align}
-\begin{split}
-\alpha^\prime &= \alpha
-\\
-\beta^\prime &= \gamma - \beta
-\\
-1 + \alpha^\prime + \beta^\prime - \gamma^\prime &= \gamma
-\\
-1 - z &= \frac{b}{1+b}
-\end{split}
-\end{align}
-and solving to find
-\begin{align}
-\begin{split}
-\alpha^\prime &= \alpha
-\\
-\beta^\prime &= \gamma - \beta
-\\
-\gamma^\prime &= 1 + \alpha - \beta
-\\
-z &= \frac{1}{1+b}
-.
-\end{split}
-\end{align}
-Finally we have everything we need. The minimal solution
-\begin{align}
-y_{4,m}
-=
-\frac{\Gamma(1+\alpha^\prime-\gamma^\prime+m)}
-        {\Gamma(1+\alpha^\prime+\beta^\prime-\gamma^\prime+m)}
-\times
-{_2F_1}\left(\alpha^\prime + m,
-                \beta^\prime,
-                1+\alpha^\prime+\beta^\prime-\gamma^\prime+m;
-                1-z\right),
-\end{align}
-where we have now included the necessary prefactors,
-is a numerically stable solution of the recurrence
-relation~\eq{eq:chosen_rec_rel} if the recursion is run
-from large $m$ to small $m$.
-So the procedure is to compute the value of ${_2F_1}$ for the two
-largest $m$ values of interest using arbitrary precision arithmetic,
-compute the prefactors to construct
-$y_{4,\text{max}(m)}$ and $y_{4,\text{max}(m)-1}$,
-recursively compute $y_{4,m}$ for all $m$ less than $\text{max}(m)$ down
-to $m=0$, then cancel off the prefactors of the resulting values of
-$y_{4,m}$ for all $m$ to produce ${_2F_1}$ for all desired $m$ values.
-\mmnote{Maybe I should reformat this paragraph into
-a pseudocode version of the algorithm?
-Instead of, or in addition to?}
-
-With ${_2F_1}$ computed, the only remaining numerical danger in computing
-$p(m)$ in~\eq{eq:p_m_bursty+rep} is overflow of the gamma functions.
-This is easily solved by taking the log of the entire expression
-and using standard routines to compute the log of the gamma functions,
-then exponentiating the entire expression at the end if $p(m)$
-is needed rather than $\log p(m)$.
-
-% \bibliographystyle{nature}
-\bibliographystyle{abbrv}
-\bibliography{./library}
-
-\end{document}

### 2020-05-19 doc/section_04_bayesian_inference.tex
initial splitting of document into several files

--- 2020-05-18 doc/paper.tex
@@
-\documentclass[12pt]{article}%{amsart}
-\usepackage[top = 1.0in, bottom = 1.0in, left = 1.0in, right = 1.0in]{geometry}
-% See geometry.pdf to learn the layout options. There are lots.
-%\geometry{letterpaper}               % ... or a4paper or a5paper or ...
-%\geometry{landscape}             % Activate for for rotated page geometry
-\usepackage[parfill]{parskip}    % Activate to begin paragraphs with an empty line rather than an indent
-\usepackage{graphicx}
-\usepackage{amssymb}
-%\usepackage{epstopdf}
-% \usepackage{caption}
-\usepackage{amsmath}
-% \usepackage{longtable}
-% \usepackage{tabu}
-% \usepackage{accents} %to get undertildes for vec & mat
-\usepackage{color} % for colored text
-% \usepackage[normalem]{ulem} % for editing: striking out text using 'sout' command
-
-% \newcommand\mytablefigwidth{0.35\textwidth}
-
-% allows to use .ai files directly w/o resaving as pdf
-\DeclareGraphicsRule{.ai}{pdf}{.ai}{}
-
-% Handy math macros!
-\newcommand{\vect}[1]{\vec{#1}}
-\newcommand{\matr}[1]{\mathbf{#1}}
-\newcommand{\rate}[3]{{#1}_{#2}^{#3}}
-\newcommand{\mmnote}[1]{\textcolor{cyan}{(MM:~#1)}}
-
-% derivative macros. these are sneaky. the [{}] is an empty default 1st arg
-% usage: provide 1 arg: \deriv{x} gives d/dx
-% provide 2 args like so: \deriv[f]{x} gives df/dx
-\newcommand{\deriv}[2][{}]{\frac{d #1}{d #2}}
-\newcommand{\pderiv}[2][{}]{\frac{\partial #1}{\partial #2}}
-\newcommand{\psecderiv}[2][{}]{\frac{\partial{^2} #1}{\partial #2{^2}}}
-
-%%%%% Referencing macros %%%%%
-\newcommand{\fig}[1]{Figure~\ref{#1}}
-\newcommand{\tref}[1]{Table~\ref{#1}}
-\newcommand{\eq}[1]{Eq.~(\ref{#1})}
-\newcommand{\eqrange}[2]{Eqs.~(\ref{#1}-\ref{#2})}
-
-%%%%%
-\begin{document}
-
-\title{Simple repression draft}
-
-\maketitle
-
-\section{Introduction}
-\mmnote{Key ideas, no particular order, that I haven't written down before:
-\begin{itemize}
-\item Our goal is to build phenomenological models of input-output
-functions of genetic circuits. More precisely, we want phenomenology
-in the sense of coarse-graining away atomistic detail, but still
-retaining biophysical meaning, e.g., we don't want to coarse-grain
-as far as Hill functions. Why not? We are motivated by studies
-like~\cite{Razo-Mejia2020}, for which Hill functions are
-insufficient. Such work requires a quantitative theory of how
-biophysical changes at the molecular level propagate to
-input-output functions at the genetic circuit level.
-We want concepts, not mere facts.
-In particular a key question is: at this level of
-coarse graining, what microscopic details do I need to explicitly
-model, and how do we figure that out? For example, do I need to
-worry about all or even any of the steps that individual RNAPs go
-through each time they make a transcript? Turning the question
-around, can we see any imprint of those processes in the
-available data? If the answer is no, then those processes are
-irrelevant for our purposes. Forward modeling and inverse
-(statistical inferential) modeling can complement each other
-beautifully here.
-\end{itemize}}
-
-\fig{fig1:means_cartoons}A shows the qualitative picture of
-simple repression. An operator, the binding site on the DNA for a
-repressor protein, may be found occupied by a repressor, in which
-case transcription is blocked from occurring. Or it may be found
-unoccupied, in which case RNA polymerase (RNAP) may bind and
-transcription can proceed.
-
-The key assumption we make is that binding of repressor and RNAP in the promoter region of interest is exclusive, meaning that one or the other may bind, but never may both be simultaneously bound.
-It is often imagined that when the repressor is bound to its operator,
-RNAP is sterically blocked from binding to its promoter sequence.
-Current evidence suggests this is sometimes, but not
-always the case~\mmnote{Cite that fig, from ??Nathan or Bill or
-Rob??, showing locations of repressor binding sites far upstream
-of promoter}, and it remains an interesting open question precisely
-how a repressor bound far upstream is able to repress transription.
-Suggestions include ``action-at-a-distance'' mediated by kinks in
-the DNA, formed when the repressor is bound, that prevent RNAP binding.
-Nevertheless, our modeling in this work is sufficiently
-coarse-grained that we simply assume exclusive binding and need
-not worry about such precise details.
-
-\mmnote{Add a notation aside somewhere that all association rates are written as zeroth-order rates.
-In other words, the concentration of the molecule is hidden inside, so for instance, $k_R^+$ is proportional to repressor copy number.}
-
-\section{Means}
-\subsection{Fold-changes are indistinguishable across models}
-As a first stop on our search for the ``right'' model of simple
-repression, let us consider what we can learn from theory and
-measurements of fold-change in mean expression. The fold-change is defined as
-\begin{equation}
-\text{fold-change}
-= \frac{\langle \text{gene expression with repressor present} \rangle}
-        {\langle \text{gene expression with repressor absent} \rangle}
-= \frac{\langle m\rangle(R\ne 0)}{\langle m\rangle(R = 0)}
-\label{eq:fc_def}
-\end{equation}
-where angle brackets denote the averge over a population of cells
-and mean mRNA $\langle m\rangle$ is viewed as a function of
-repressor copy number $R$.
-The second
-equality in~\ref{eq:fc_def} follows from assuming that the
-translation efficiency, i.e., the number of proteins translated
-per mRNA, is the same in both conditions.
-In other words, we assume that mean protein level is proportional
-to mean mRNA level, and that the proportionality constant is the
-same in both conditions and therefore cancels out in the ratio.
-This is reasonable since the cells in the two conditions are
-identical except for the presence of the transcription factor,
-and it is difficult to imagine the transcription factor having
-any direct effect on translation.
-
-Fold-change has proven a very convenient observable in past
-work~\cite{Garcia2011a, Razo-Mejia2018, Chure2019}.
-Part of its utility in dissecting transcriptional regulation is
-its ratiometric nature, which is always easier than making an
-absolute measurement. Also, by measuring otherwise identical cells
-with and without a transcription factor present, any biological
-noise common to both conditions can be made to cancel away.
-In addition, the presence of different operator sites is known to
-change the basal expression level in the absence of
-repressor~\mmnote{look up cite}.
-Computing fold-changes neatly cancels this effect as well.
-
-\fig{fig1:means_cartoons} depicts a smorgasbord
-of mathematized cartoons for simple repression that have appeared
-in previous literature. For each cartoon, we calculate the
-fold-change in mean gene expression as predicted by that model,
-deferring some algebraic details to the appendix.
-What we will find is that all cartoons collapse to a single
-master curve, shown in \fig{fig1:means_cartoons}c, which contains
-just two parameters. We label the parameters $\Delta F_R$, an
-effective free energy parametrizing the repressor-DNA
-interaction, and $\rho$, which subsumes all details of
-transcription in the absence of repressors.
-We will offer some intuition for why this master curve exists and discuss
-why it is likely impossible to discriminate ``right'' from ``wrong''
-cartoons given only measurements of fold-changes in expression.
-
-\subsubsection{Equilibrium, 2-state}
-In this simplest model, the promoter is idealized as existing in
-one of two states, either repressor bound or repressor unbound.
-The rate of transcription, and therefore the level of gene
-expression, is assumed to be proportional to the fraction of time
-spent in the repressor unbound state.
-Mathematically, this amounts to the statement that the mean mRNA level is
-given by
-\begin{equation}
-\langle m \rangle = \frac{r}{\gamma}
-        \left(1 + \frac{R}{N_{NS}} e^{-\beta\Delta\epsilon_R}\right)^{-1},
-\end{equation}
-where $r$ is the transcription rate from the repressor unbound
-state, $\gamma$ is the mRNA degradation rate, $R$ is repressor
-copy number, $N_{NS}$ is the number of nonspecific binding sites
-in the genome (where repressors spend most of their time when not
-bound to the operator, their specifc binding site),
-$\beta=1/k_BT$, and $\Delta\epsilon_R$ is the binding energy of a
-repressor to its operator site. The derivation of this result is deferred
-to the appendix.
-
-The fold-change is easily found as the ratio of mean mRNA with
-and without repressor. This gives
-\begin{equation}
-\text{fold-change}
-= \left(1 + \frac{R}{N_{NS}} e^{-\beta\Delta\epsilon_R}\right)^{-1},
-\end{equation}
-which clearly matches the form of the master curve
-in~\eq{fig1:means_cartoons} with $\rho=1$.
-
-In fact it was noted in~\cite{Chure2019} that this two-state
-model can be viewed as the coarse-graining of any equilibrium
-promoter model in which no transcriptionally active states have
-transcription factor bound. Put differently, there is no overlap
-between TF bound states and transcriptionally active states. We
-will see this explicitly in the 3-state equilibrium model below,
-but perhaps surprising is that something analogous carries over
-even to the nonequilibrium models we consider later.
-
-\subsubsection{Equilibrium, 3-state}
-Compared to the previous model, here we fine-grain the repressor
-unbound state into two separate states: empty, and RNAP bound.
-This picture was used in~\cite{Garcia2011a} as we use it here,
-and in~\cite{Razo-Mejia2018} and~\cite{Chure2019} it was generalized
-to incorporate small-molecule inducers that bind the repressor.
-Interestingly, the effect of this is simply to renormalize $R$
-from the total number of repressors to a smaller effective number,
-roughly (though not exactly) the number which are unbound by inducers.
-
-The mean mRNA, as derived in the appendix from a similar
-enumeration of states and weights as the previous model, is
-\begin{equation}
-\langle m \rangle = \frac{r}{\gamma}
-\frac{\frac{P}{N_{NS}} e^{-\beta\Delta\epsilon_P}}
-        {
-        1 + \frac{R}{N_{NS}} e^{-\beta\Delta\epsilon_R}
-        + \frac{P}{N_{NS}} e^{-\beta\Delta\epsilon_P}
-        },
-\end{equation}
-where the new variables are $\Delta\epsilon_P$, the difference in
-RNAP binding energy to its specific site (the promoter) relative
-to an average nonspecific background site, and the RNAP copy
-number, $P$. The fold-change again follows immediately as
-\begin{align}
-\text{fold-change}
-&= \frac{\frac{P}{N_{NS}} e^{-\beta\Delta\epsilon_P}}
-        {
-        1 + \frac{R}{N_{NS}} e^{-\beta\Delta\epsilon_R}
-        + \frac{P}{N_{NS}} e^{-\beta\Delta\epsilon_P}
-        }
-\frac{1 + \frac{P}{N_{NS}} e^{-\beta\Delta\epsilon_P}}
-        {\frac{P}{N_{NS}} e^{-\beta\Delta\epsilon_P}}
-\\
-&= \left(
-1 + \frac{\frac{R}{N_{NS}} e^{-\beta\Delta\epsilon_R}}
-        {1 + \frac{P}{N_{NS}} e^{-\beta\Delta\epsilon_P}}
-\right)^{-1}
-\\
-&= (1 + \exp(-\Delta F_R - \log\rho))^{-1},
-\end{align}
-with $\Delta F_R = \beta\Delta\epsilon_R - \log(R/N_{NS})$
-and $\rho = 1 + \frac{P}{N_{NS}}\mathrm{e}^{-\beta\Delta\varepsilon_P}$
-as shown in~\fig{fig1:means_cartoons}.
-
-\begin{figure}%[h!]
-\centering
-\includegraphics[width=\textwidth]{../figures/fig1/fig1.ai}
-\caption{
-\textbf{
-An overview of the simple repression motif at the level of means.}
-(A) depicts the qualitative biological picture of simple repression.
-(B) shows a variety of possible mathematized cartoons of simple
-repression, along with the effective parameter $\rho$
-which subsumes all regulatory details
-of the architecture excluding the repressor.
-(C) shows the ``master curve'' to which all cartoons in (B) collapse.
-}
-\label{fig1:means_cartoons}
-\end{figure}
-
-\subsubsection{Nonequilibrium model 1 - Poisson promoter}
-For our first kinetic model, we consider the simplest possible
-picture with only two states, repressor bound and unbound.
-This is exactly the model used for the main results of~\cite{Jones2014}.
-In this picture, repressor association and dissociation rates
-from its operator site, $k_R^+$ and $k_R^-$, respectively, govern
-transitions between the two states. When the system is in the
-unbound state, transcriptions initiate at rate $r$, which
-represents a coarse-graining of all the downstream processes into
-a single effective rate. mRNA is again degraded at rate $\gamma$.
-
-Let $p_{m,R}(t)$ denote the joint probability of finding the
-system in the repressor bound state with $m$ mRNA molecules
-present at time $t$, and similarly define $p_{m,U}(t)$ for the
-repressor unbound state. 
-This model is governed by a master equation giving the time
-evolution of $p_{m,R}(t)$ and $p_{m,U}(t)$, which is
-\begin{align}
-\begin{split}
-\deriv{t}p_{m,R}(t) =& - k_R^- p_{m,R}(t) + k_R^+ p_{m,U}(t)
-                + (m+1)\gamma p_{m+1,R}(t) - \gamma p_{m,R}(t)
-\\
-\deriv{t}p_{m,U}(t) =&\; k_R^- p_{m,R}(t) - k_R^+ p_{m,U}(t)
-                        + rp_{m-1,U}(t) - rp_{m,U}(t)
-                \\
-                &+ (m+1)\gamma p_{m+1,U}(t) - \gamma p_{m,U}(t).
-\label{eq:poisson_promoter_cme}
-\end{split}
-\end{align}
-In each equation, the first two terms describe transitions
-between promoter states due to repressors unbinding and binding,
-respectively. The final two terms describe degradation of mRNA,
-decreasing the copy number by one, and the terms with coefficient
-$r$ describe transcription initiation increasing the mRNA copy
-number by one.
-
-We can greatly simplify the notation, which will be especially
-useful for the more complicated models yet to come, by
-re-expressing the master equation in vector form. The promoter
-states are collected into a vector and the rate constants are
-collected into matrices as
-\begin{equation}
-\vect{p}_m = \begin{pmatrix} p_{m,R} \\ p_{m,U} \end{pmatrix},\
-\matr{K} = \begin{pmatrix} -k_R^- & k_R^+ \\ k_R^- & -k_R^+ \end{pmatrix},\
-\matr{R} = \begin{pmatrix} 0 & 0 \\ 0 & r \end{pmatrix},\
-\label{eq:2state_cme_matrices}
-\end{equation}
-so that the master equation may be condensed as
-\begin{equation}
-\deriv{t}\vect{p}_m(t) =
-\left( \matr{K} - \matr{R} - \gamma m \matr{I} \right) \vect{p}_m(t)
-                + \matr{R} \vect{p}_{m-1}(t)
-                + \gamma (m+1) \matr{I} \vect{p}_{m+1}(t),
-\label{eq:2state_rep_cme}
-\end{equation}
-where $\matr{I}$ is the identity matrix.
-Taking steady state by setting time derivatives to zero,
-the mean mRNA can be found to be
-\begin{equation}
-\langle m \rangle = \frac{r}{\gamma}
-        \left(1 + \frac{k_R^+}{k_R^-}\right)^{-1},
-\end{equation}
-with the algebra details again deferred to the appendix.
-Recall $k_R^+$ is proportional to the repressor copy number, so
-in computing fold-change, absence of repressor corresponds to
-$k_R^+\rightarrow0$.
-Therefore fold-change in this model is simply
-\begin{equation}
-\text{fold-change} = \left(1 + \frac{k_R^+}{k_R^-}\right)^{-1},
-\end{equation}
-again matching the master curve of~\fig{fig1:means_cartoons} with $\rho=1$.
-
-\subsubsection{Nonequilibrium model 2 - RNAP bound/unbound states}
-Analogous to the three-state equilibrium model, this model
-fine-grains the repressor unbound state of nonequilibrium model 1,
-resolving it into an empty promoter state and an RNAP-bound state.
-Note in this picture, in contrast with model 4 below,
-transcription initiation is accompanied by a promoter state
-change, in keeping with the interpretation as RNAP-bound and
-empty states: if an RNAP successfully escapes the promoter and
-proceeds to elongation of a transcript, clearly it is no longer
-bound at the promoter! Therefore another RNAP must bind before
-another transcript can be initiated.
-
-The master equation governing this model is analogous
-to~\eqrange{eq:2state_cme_matrices}{eq:2state_rep_cme} for model 1 above.
-The main subtlety arises since transcription initiation
-accompanies a promoter state change.
-This can be understood by analogy to $\matr{K}$. The
-off-diagonals and diagonals of $\matr{K}$ correspond to
-transitions arriving at or departing from, respectively, the
-promoter state of interest. If transcription initiations are
-accompanied by promoter state changes, we must have separate
-matrices for arriving and departing transcription events since
-the arriving and departing transitions have different initial
-copy number of mRNA, unlike for $\matr{K}$ where they are the same.
-The master equation for this model is
-\begin{equation}
-\deriv{t}\vect{p}_m(t) =
-\left( \matr{K} - \matr{R_D} - \gamma m \matr{I} \right) \vect{p}_m(t)
-                + \matr{R_A} \vect{p}_{m-1}(t) +
-                \gamma (m+1) \matr{I} \vect{p}_{m+1}(t),
-\label{eq:3state_rep_cme}
-\end{equation}
-with the state vector and promoter transition matrix defined as
-\begin{equation}
-\vect{p}_m = \begin{pmatrix} p_{m,R} \\ p_{m,E} \\ p_{m,P} \end{pmatrix},\
-\matr{K} = \begin{pmatrix} -k_R^- & k_R^+ & 0 \\
-                        k_R^- & -k_R^+ -k_P^+ & k_P^- \\
-                        0 & k_P^+ & -k_P^- 
-                \end{pmatrix},
-\label{eq:3state_cme_matrices_pt1}
-\end{equation}
-and the initiation matrices given by
-\begin{equation}
-\matr{R_A} = \begin{pmatrix}
-                0 & 0 & 0 \\ 0 & 0 & r \\ 0 & 0 & 0
-                \end{pmatrix},\
-\matr{R_D} = \begin{pmatrix}
-                0 & 0 & 0 \\ 0 & 0 & 0 \\ 0 & 0 & r
-                \end{pmatrix},
-\label{eq:3state_cme_matrices_pt2}
-\end{equation}
-The elements of $\vect{p}_m$ encode the probabilities of having
-$m$ mRNA present along with the promoter having repressor bound,
-being empty, or having RNAP bound, respectively.
-$\matr{R_A}$ describes probability flux \textit{arriving} at the state
-$\vect{p}_m$ from a state with one fewer mRNA, namely
-$\vect{p}_{m-1}$, and $\matr{R_D}$ describes probability flux
-\textit{departing} from the state $\vect{p}_m$
-for a state with one more mRNA, namely $\vect{p}_{m+1}$.
-$\matr{K}$ is closely analogous to model 1.
-
-Mean mRNA at steady state is found analogously to model 1, with the result
-\begin{equation}
-\langle m\rangle = \frac{r}{\gamma}
-        \frac{k_R^- k_P^+}
-        {k_R^- k_P^+ + k_R^- (k_P^- + r) + k_R^+ (k_P^- + r)}.
-\end{equation}
-Fold-change is again found from the ratio prescribed by~\eq{eq:fc_def},
-from which we have
-\begin{align}
-\text{fold-change}
-&=      \frac{k_R^- k_P^+}
-        {k_R^- k_P^+ + k_R^- (k_P^- + r) + k_R^+ (k_P^- + r)}
-        \frac{k_P^+ + k_P^- + r}{k_P^+}
-\\
-&=      \left(1 + \frac{k_R^+}{k_R^-}
-                \frac{k_P^- + r}{k_P^+ + k_P^- + r}
-        \right)^{-1}
-\\
-&=      \left(1 + \frac{k_R^+}{k_R^-}
-        \left(1 + \frac{k_P^+}{k_P^- + r}\right)^{-1}
-        \right)^{-1},
-\end{align}
-which follows the master curve with
-$\rho = 1 + k_P^+/(k_P^- + r)$ as claimed.
-
-\subsubsection{Nonequilibrium model 3
-- Multistep transcription initiation and escape}
-One might reasonably complain that the first two
-``nonequilibrium'' models we have considered are straw men. Their
-steady states necessarily satisfy detailed balance which is
-equivalent to thermodynamic equilibrium. Why is this the case? At
-steady state there is by definition no net probability flux in or
-out of each promoter state, but since the promoter states form a
-linear chain, there is only one way in or out of the repressor
-bound and RNAP bound states, implying each edge must actually
-have a net zero probability flux, which is the definition of
-detailed balance (usually phrased as equality of forward and
-reverse transition fluxes).
-
-Now we consider model 3 in~\eq{fig1:means_cartoons} which allows
-the possibility of true nonequilibrium steady-state fluxes
-through the promoter states.
-We point out that this model was considered previously in~\cite{Mitarai2015}
-where a comparison was made with model 1 as used in~\cite{Jones2014}.
-The authors of~\cite{Mitarai2015} argued that the additional
-complexity is essential to properly account for the noise in the
-mRNA distribution. We will weigh in on both models later when we
-consider observables beyond fold-change.
-
-The master equation governing this model is
-identical in form to model 2 above, namely
-\begin{equation}
-\deriv{t}\vect{p}_m(t) =
-\left( \matr{K} - \matr{R_D} - \gamma m \matr{I} \right) \vect{p}_m(t)
-                + \matr{R_A} \vect{p}_{m-1}(t) +
-                \gamma (m+1) \matr{I} \vect{p}_{m+1}(t),
-\end{equation}
-but obviously with a higher-dimensional state space and different matrices.
-The state vector and promoter transition matrix are now
-\begin{equation}
-\vect{p}_m = \begin{pmatrix} p_{m,R} \\ p_{m,E} \\
-                             p_{m,C} \\ p_{m,O}\end{pmatrix},\
-\matr{K} = \begin{pmatrix} -k_R^- & k_R^+ & 0 & 0\\
-                        k_R^- & -k_R^+ -k_P^+ & k_P^- & 0 \\
-                        0 & k_P^+ & -k_P^- - k_O & 0 \\
-                        0 & 0 & k_O & 0
-                \end{pmatrix},
-\end{equation}
-with the four promoter states, in order, being repressor bound,
-empty, RNAP closed complex, and RNAP open complex.
-Besides increasing dimension by one again, the only new feature in
-$\matr{K}$ is the rate $k_O$, representing the (irreversible)
-rate of open complex formation from the closed complex.
-The initiation matrices are given by
-\begin{equation}
-\matr{R_A} = \begin{pmatrix}
-        0 & 0 & 0 & 0 \\ 0 & 0 & 0 & r \\ 0 & 0 & 0 & 0\\ 0 & 0 & 0 & 0
-                \end{pmatrix},\
-\matr{R_D} = \begin{pmatrix}
-        0 & 0 & 0 & 0 \\ 0 & 0 & 0 & 0 \\ 0 & 0 & 0 & 0\\ 0 & 0 & 0 & r
-                \end{pmatrix},
-\end{equation}
-again closely analogous to model 2.
-
-The expression for mean mRNA is substantially more complicated
-now, as worked out in the appendix where we find
-\begin{equation}
-\langle m\rangle = \frac{r}{\gamma}
-        \frac{k_R^- k_P^+ k_O}
-        {k_R^- [(k_P^+ (k_O + r) + r(k_P^- + k_O)] + k_R^+ r(k_P^- + k_O)},
-\end{equation}
-which can be simplified to
-\begin{equation}
-\langle m\rangle
-= \frac{r}{\gamma}
-\frac{\frac{k_P^+ k_O}{r(k_O + k_P^-)}}
-        {1 + \frac{k_P^+ (k_O + r)}{r(k_O + k_P^-)} + \frac{k_R^+}{k_R^-}}.
-\end{equation}
-The strategy is to isolate the terms involving the repressor, so that now the fold-change is seen to be simply
-\begin{align}
-\text{fold-change}
-&= \frac{\frac{k_P^+ k_O}{r(k_O + k_P^-)}}
-        {1 + \frac{k_P^+ (k_O + r)}{r(k_O + k_P^-)} + \frac{k_R^+}{k_R^-}}
-        \frac{1 + \frac{k_P^+ (k_O + r)}{r(k_O + k_P^-)}}
-                {\frac{k_P^+ k_O}{r(k_O + k_P^-)}}
-\\
-&= \left(
-        1 + \frac{k_R^+}{k_R^-}
-        \left(1 + \frac{k_P^+ (k_O + r)}{r(k_O + k_P^-)}\right)^{-1}
-\right)^{-1},
-\end{align}
-surprisingly reducing to the master curve once again, with
-$\rho = 1 + \frac{k_P^+ (k_O + r)}{r(k_O + k_P^-)}$.
-
-This example hints that an arbitrarily fine-grained model of
-downstream transcription steps may still be collapsed to the form
-of the master curve in~\fig{fig1:means_cartoons}B so long as the
-repressor binding is exclusive with transcriptionally active
-states. We offer this as a conjecture, and we suspect that a
-careful argument using the King-Altman diagram
-method~\cite{King1956, Hill1966} might furnish a ``proof.''
-Our focus here is not on full generality but rather to survey an
-assortment of plausible models that have been proposed in the
-literature for simple repression.
-
-\subsubsection{Nonequilibrium model 4 - ``Active'' and ``inactive'' states}
-This model is at the core of the theory in~\cite{Razo-Mejia2020}.
-At a glance the cartoon for this model may appear very similar to model 2,
-and mathematically it is, but the interpretation is rather different.
-In model 2, we interpreted the third state literally as an
-RNAP-bound promoter and modeled initiation of a transcript as
-triggering a promoter state change, making the hopefully
-uncontroversial assumption that an RNAP can only make one
-transcript at a time. In contrast, in the present model the
-promoter state does \textit{not} change when a transcript is
-initiated. So we no longer interpret these states as literally
-RNAP bound and unbound but instead as coarse-grained ``active''
-and ``inactive'' states, the details of which we leave
-unspecified for now. We will comment more on this model below
-when we discuss Fano factors of models.
-
-Mathematically this model is very similar to models 1 and 2.
-Like model 1, the matrix $R$ describing transcription initiation
-is diagonal, namely
-\begin{equation}
-\matr{R} = \begin{pmatrix}
-                0 & 0 & 0 \\ 0 & 0 & 0 \\ 0 & 0 & r
-        \end{pmatrix}.
-\end{equation}
-so the master equation takes verbatim the same form as it did for model
-1,~\eq{eq:2state_rep_cme}. Meanwhile the promoter transition matrix
-$K$ is the same as~\eq{eq:3state_cme_matrices_pt1} from model 2,
-although we relabel the rate constants from $k_P^\pm$ to $k^\pm$ to
-reiterate that these are not simply RNAP binding and unbinding rates.
-
-Carrying out the algebra, the mean mRNA can be found to be
-\begin{equation}
-\langle m\rangle = \frac{r}{\gamma}
-\frac{k_R^- k^+}
-{k_R^- k^+ + k_R^- k^- + k_R^+ k^-},
-\end{equation}
-and the fold-change readily follows,
-\begin{align}
-\text{fold-change}
-&=      \frac{k_R^- k^+}{k_R^- k^+ + k_R^- k^- + k_R^+ k^-}
-        \frac{k_R^- k^+ + k_R^- k^-}{k_R^- k^+}
-\\
-&=      \left(1 + \frac{k_R^+}{k_R^-}
-                \left(1 + \frac{k^+}{k^-}\right)^{-1}
-        \right)^{-1},
-\end{align}
-from which we see $\rho = 1 + k^+/k^-$ as shown in~\fig{fig1:means_cartoons}.
-
-\subsubsection{Nonequilibrium model 5 - Bursty promoter}
-We will have much more to say about this model in subsequent sections.
-For now we sketch it as an intuitive analog to model 1,
-with just two states, repressor bound or unbound, and transition rates
-between them of $k_R^+$ and $k_R^-$.
-In model 1, when in the unbound state, single mRNA transcripts
-are produced as a Poisson process with some characteristic rate
-$r$. This model by contrast produces, at some Poisson rate $k_i$,
-\textit{bursts} of transcripts with a geometrically distributed
-burst size and a mean burst size $b$.
-
-From this intuitive picture and by analogy to model 1, then, it should be plausible that the mean mRNA level is
-\begin{equation}
-\langle m\rangle = \frac{k_i b}{\gamma}
-        \left(1 + \frac{k_R^+}{k_R^-}\right)^{-1},
-\end{equation}
-which will turn out to be correct from a careful calculation.
-For now, we simply note that just like model 1, the fold-change becomes
-\begin{equation}
-\text{fold-change} = \left(1 + \frac{k_R^+}{k_R^-}\right)^{-1}
-\end{equation}
-with $\rho=1$ also like model 1.
-
-We will also see later how this model emerges as a natural limit of model 4.
-
-\subsubsection{Summary and discussion of results across models for fold-changes in mean expression}
-What have we learned from our exploration of the models
-in~\fig{fig1:means_cartoons} and their predictions of
-fold-changes in mean expression levels? The takeaway is the
-existence of a master curve (\fig{fig1:means_cartoons}C) to which
-the fold-change predictions of all the models collapse. This
-master curve is parametrized by only two effective parameters:
-$\Delta F_R$, which characterizes the repressor, and $\rho$,
-which characterizes all other features of the promoter
-architecture. The key assumption underpinning this result is that
-zero transcription occurs when a repressor is bound to its
-operator. Note, however, that we are agnostic about the molecular
-mechanism which achieves this;
-steric effects are the obvious mechanism, but, for instance,
-``action-at-a-distance'' mediated by kinked DNA due to
-repressors bound tens or hundreds of nucleotides upstream of a
-promoter is plausible as well.
-
-Why does the master curve of~\fig{fig1:means_cartoons}C exist at all?
-This brings to mind the deep questions posed in,
-e.g.,~\cite{Frank2013} and~\cite{Frank2014a}, asking us to search
-for general ``symmetries'' of the problem instead of specifying
-generative models of the known microscopics of a system.
-In our case, the key feature seems to be the exclusive nature of
-repressor and RNAP binding, which allows the parameter describing
-the repressor, $\Delta F_R$, to cleanly separate from all other
-details of the promoter architecture, which are encapsulated in $\rho$.
-In the graph-theoretic language which was applied to similar
-problems in~\cite{Gunawardena2012} and~\cite{Ahsendorf2014}, the
-repressor-bound state is a leaf of the graph describing
-transitions between promoter states. Arbitrary nonequilibrium
-behavior can occur on the rest of the promoter state space, but
-it may all be swept up in the effective parameter $\rho$, to
-which the repressor makes no contribution.
-
-As suggested in~\cite{Chure2019}, we believe this master curve
-should generalize to architectures with multiple repressor
-binding sites, as long as the exclusivity of transcription factor
-binding and transcription initation is maintained. The
-interpretation of $\Delta F_R$ is then of an effective free
-energy of all repressor bound states.
-In an equilibrium picture this is simply given by the log of the
-sum of Boltzmann weights of all repressor bound states, which
-looks like the log of a partition function of a subsystem.
-In a nonequilibrium picture, while we can still mathematically gather
-terms and give the resulting collection the label $\Delta F_R$,
-it is unclear if the physical interpretation as an effective free energy
-makes sense.
-As is well known, free energies cannot be assigned unambiguously
-to states out of equilibrium because the free energy change along
-a generic path traversing the state space is path dependent,
-unlike at equilibrium.
-A consequence of this is that, out of equilibrium, $\Delta F_R$
-is no longer a simple sum of Boltzmann weights. Instead it
-resembles a restricted sum of King-Altman
-diagrams~\cite{King1956, Hill1966}. Following the work of
-Hill~\cite{Hill1989}, it may yet be possible to interpret this expression
-as an effective free energy, but this remains unclear to us.
-We leave this an open problem for future work.
-
-If we relax the requirement of exclusive repressor-RNAP binding,
-one could imagine models in which repressor and RNAP doubly-bound
-states are allowed, where the repressor's effect is to
-\text{reduce} the transcription rate rather than setting it to zero.
-Our results do not strictly apply to such a model, although we
-note that if the repressor's reduction of the transcription rate
-is substantial, such a model might still be well-approximated by
-one of the models in~\fig{fig1:means_cartoons}.
-
-One may worry that our ``one curve to rule them all'' is a
-mathematical tautology. In fact we \textit{agree} with this
-criticism if $\Delta F_R$ is ``just a fitting parameter'' and cannot
-be meaningfully interpreted as a real, physical free energy.
-An analogy to Hill functions is appropriate here. One of their
-great strengths and weaknesses, depending on the use they are put
-to, is that their parameters coarse-grain many details and are
-generally not interpretable in terms of microscopic models, for
-deep reasons discussed at length in~\cite{Frank2013}.
-By contrast, our master curve claims to have the best of both worlds:
-a coarse-graining of all details besides the repressor into
-a single effective parameter $\rho$,
-while simultaneously retaining an interpretation of
-$\Delta F_R$ as a physically
-meaningful and interpretable free energy.
-Our task, then, is to prove or disprove this claim.
-
-How do we test this and probe the theory with fold-change
-measurements? There is a fundamental limitation in that the
-master curve is essentially a one-parameter function of $\Delta
-F_R + \log\rho$. Worse, there are many \textit{a priori}
-plausible microscopic mechanisms that could contribute to the
-value of $\rho$, such as RNAP binding and escape
-kinetics~\cite{DeHaseth1998, Mitarai2015}, and/or supercoiling
-accumulation and release~\cite{Chong2014, Sevier2016}, and/or,
-RNAP clusters analogous to those obvserved in
-eukaryotes~\cite{Cisse2013, Cho2016}
-and recently also observed in bacteria~\cite{Ladouceur2020}.
-Even if $\Delta F_R$ is measured to
-high precision, inferring the potential microscopic contributions
-to $\rho$, buried inside a log no less, from fold-change
-measurements seems hopeless. As a statistical inference problem
-it is entirely nonidentifiable, in the language
-of~\cite{Gelman2013}, Section 4.3.
-
-If we cannot simply infer values of $\rho$ from measurements of
-fold-change, can we perturb some of the parameters that make up
-$\rho$ and measure the change? Unfortunately we suspect this is
-off-limits experimentally: most of the potential contributors to
-$\rho$ are global processes that affect many or all genes. For
-instance, changing RNAP association rates by changing RNAP copy
-numbers, or changing supercoiling kinetics by changing
-topoisomerase copy numbers, would massively perturb the entire
-cell's physiology and confound any determination of $\rho$.
-
-One might instead imagine a bottom-up modeling approach, where we
-mathematize a model of what we hypothesize the important steps
-are and are not, use in vitro data for the steps deemed
-important, and \textit{predict} what $\rho$ should be.
-But again, because of the one-parameter nature of the master
-curve, many different models will likely make indistinguishable
-predictions, and without any way to experimentally perturb in
-vivo, there is no clear way to test whether the modeling
-assumptions are correct.
-
-In light of this, we prefer the view that parameters and rates
-are not directly comparable between cartoons
-in~\fig{fig1:means_cartoons}. Rather, parameters in the
-simpler cartoons represent coarse-grained combinations of
-parameters in the finer-grained models. For instance, by equating
-$\rho$ between any two models, one can derive various possible
-correspondences between the two models' parameters. Note that
-these correspondences are clearly not unique, since many possible
-associations could be made. It then is a choice as to what microscopic
-interpretations the model-builder prefers for the parameters in a
-particular cartoon, and as to which coarse-grainings lend
-intuition and which seem nonsensical.
-Indeed, since it remains an open question what microscopic
-features dominate $\rho$ (as suggested above, perhaps RNAP
-binding and escape kinetics~\cite{DeHaseth1998, Mitarai2015}, or
-supercoiling accumulation and release~\cite{Chong2014, Sevier2016},
-or, something more exotic like
-RNAP clusters~\cite{Cisse2013, Cho2016, Ladouceur2020}),
-we are hesitant to put too much weight on any one microscopic
-interpretation of model parameters that make up $\rho$.
-
-One possible tuning knob to probe $\rho$ that would not globally
-perturb the cell's physiology is to manipulate RNAP binding sites.
-Work such as~\cite{Kinney2010} has shown that models of
-sequence-dependent RNAP can be inferred from data, and the
-authors of~\cite{Brewster2012} showed that the model
-of~\cite{Kinney2010} has predictive power by using the model to
-\textit{design} binding sites of a desired affinity.
-But for our purposes, this begs the question: the authors
-of~\cite{Kinney2010} \textit{assumed} a particular model
-(essentially our 3-state equilibrium model but without the
-repressor), so it is unclear how or if such methods can be turned
-around to \textit{compare} different models of promoter function.
-
-We have already pointed out that the master curve
-of~\fig{fig1:means_cartoons} is essentially a one-parameter model,
-the one parameter being $\Delta F_R + \log\rho$.
-By now the reader may be alarmed as to how can we even determine
-$\Delta F_R$ and $\rho$ independently of each other, never mind
-shedding a lens on the internal structure of $\rho$ itself.
-A hint is provided by the weak promoter approximation, invoked
-repeatedly in prior studies~\cite{Bintu2005c, Garcia2011a,
-Razo-Mejia2018} of simple repression using the 3-state
-equilibrium model in~\fig{fig1:means_cartoons}.
-In that picture, the weak promoter approximation means
-$\frac{P}{N_{NS}}\exp(-\beta\Delta\epsilon_P) \ll 1$,
-meaning therefore $\rho\approx1$.
-This is suggestive, but how can we be sure that $\rho$ is not,
-for instance, actually $10^2$ and that $\Delta F_R$ hides a
-compensatory factor?
-A resolution is offered by an independent inference of $\rho$ in
-the absence of repressors. This was done in~\cite{Razo-Mejia2020}
-by fitting nonequilibrium model 4  in~\fig{fig1:means_cartoons},
-with zero repressor (looking ahead, this is equivalent to model 4
-in~\fig{fig:constit_cartoons}), to FISH data
-from~\cite{Brewster2014}. This provided a determination of $k^+$
-and $k^-$, from which their ratio is estimated to be no more than
-a few $10^{-1}$ and possibly as small as $10^{-2}$.
-
-The realization that $\rho\approx1$ to an excellent
-approximation, \textit{independent} of which model
-in~\fig{fig1:means_cartoons} one prefers, goes a long way towards
-explaining the surprising success of equilibrium models of simple
-repression. Even though our 2- and 3-state models get so many
-details of transcription wrong, it does not matter because
-fold-change is a cleverly designed ratio. Since $\rho$ subsumes
-all details except the repressor, and $\log\rho\approx0$, fitting
-these simple models to fold-change measurements can still give a
-surprisingly good estimate of repressor binding energies. So the
-ratiometric construction of fold-change fulfills its intended
-purpose of canceling out all features of the promoter
-architecture except the repressor itself. Nevertheless it is
-perhaps surprising how effectively it does so: \textit{a priori},
-one might not have expected $\rho$ to be quite so close to 1.
-
-We would also like to highlight the relevance
-of~\cite{Landman2019} here. Landman et.\ al.\ reanalyzed and
-compared in vivo and in vitro data on the lacI repressor's
-binding affinity to its various operator sequences. (The in vivo
-data was from, essentially, fitting our master curve to
-expression measurements.) They find broad agreement between the
-in vitro and in vivo values. This reinforces the suspicion that
-the equilibrium $\Delta\epsilon_R$ repressor binding energies do
-in fact represent real physical free energies.
-Again, \textit{a priori} this did not have to be the case, even
-knowing that $\rho\approx1$.
-
-In principle, if $\Delta F_R$ can be measured to sufficient
-precision, then deviations from $\rho=1$ become a
-testable matter of experiment.
-In practice, it is probably unrealistic to measure repressor
-rates $k_R^+$ or $k_R^-$ or fold-changes in expression levels
-(and hence $\Delta\epsilon_R$) precisely enough to detect the
-expected tiny deviations from $\rho=1$.
-We can estimate the
-requisite precision in $\Delta F_R$ to resolve a given
-$\Delta\rho$ by noting, since $\rho\approx1$, that
-$\log(1+\Delta\rho)\approx \Delta\rho$, so
-$\Delta(\Delta F_R) \approx \Delta\rho$.
-Suppose we are lucky and $\Delta\rho$ happens to be $\sim0.1$, on
-the high end of our range estimated above.
-A determination of $\Delta\epsilon_R/k_BT$ with an uncertainty of
-barely 0.1 was achieved in the excellent measurements
-of~\cite{Razo-Mejia2018}, so this requires a very difficult
-determination of $\Delta F_R$ for a very crude determination of $\rho$,
-which suggests, to put it lightly, this is not a
-promising path to pursue experimentally.
-It is doubtful that inference of repressor kinetic rates
-would be any easier.\footnote{
-\mmnote{
-In fact the uncertainties quoted in Hammar et.\
-al.~\cite{Hammar2014} from Elf's group are small enough to do
-this calculation, but HJ \& I think their error bars are insanely
-optimistic. Not sure I want to blatantly call them out on it though??}
-}
-
-Moving forward, we have weak evidence supporting the
-interpretation of $\Delta F_R$ as a physically real free
-energy~\cite{Landman2019} and other work casting
-doubt~\cite{Hammar2014}.
-How might we resolve the confusion?
-If there is no discriminatory power to test the theory and
-distinguish the various models with measurements of fold-changes
-in means, how do we probe the theory? Clearly to discriminate
-between the nonequilibrium models in~\fig{fig1:means_cartoons},
-we need to go beyond means to ask questions about kinetics, noise and even
-full distributions of mRNA copy numbers over a population of cells.
-If the ``one-curve-to-rule-them-all'' is more than a
-mathematical tautology, then the free energy of repressor binding
-inferred from fold-change measurements should agree with
-repressor binding and unbinding rates.
-In other words, the equilbrium and nonequilibrium definitions of
-$\Delta F_R$ must agree, meaning
-\begin{equation}
-\Delta F_R = \beta\Delta\epsilon_R - \log(R/N_{NS})
-        = - \log(k_R^+/k_R^-),
-\end{equation}
-must hold, where $\beta\Delta\epsilon_R$ is inferred from the
-master curve fit to fold-change measurements, and $k_R^+$ and
-$k_R^-$ are inferred in some orthogonal manner.
-Single molecule measurements such as from~\cite{Hammar2014} have
-directly observed these rates, and in the remainder of this work
-we explore a complementary approach: inferring repressor binding
-and unbinding rates $k_R^+$ and $k_R^-$ from FISH measurements of
-mRNA population distributions.
-
-\section{Beyond means}
-\subsection{A one-state promoter with bursting is the ``best'' model of constitutive promoters}
-Before we can tackle simple repression, we need an adequate
-phenomenological model of constitutive expression.
-The literature abounds with options from which we can choose,
-and we show several potential kinetic models for constitutive promoters
-in~\fig{fig:constit_cartoons}.
-Let us consider the suitability of each model for our purposes in turn.
-
-\begin{figure}%[h!]
-\centering
-\includegraphics[width=0.9\textwidth]{../figures/fig1.5/fig1point5.ai}
-\caption{\textbf{Constitutive promoter cartoon comparison.}
-The left column depicts various plausible cartoons for constitutive promoters.
-We also list literature references where models have been used previously.
-In model (1), transcripts are produced in a Poisson process~\cite{Jones2014}.
-Model (2) features explicit modeling of RNAP binding/unbinding
-kinetics~\cite{Phillips2015a}.
-Model (3) is a more detailed generalization of model (2),
-treating transcription initiation as a multi-step process
-proceeding through closed and open complexes~\cite{Mitarai2015}.
-Model (4) is somewhat analogous to (2) except with the precise nature
-of active and inactive states left ambiguous~\cite{Razo-Mejia2020}.
-Finally, model (5) can be viewed as a certain limit of model (4)
-in which transcripts are produced in bursts,
-and initiation of bursts is a Poisson process.
-\mmnote{Haven't found model 5 formulated quite like this in the literature,
-usually burstiness is handled like model 4, but I should probably search a bit more.}
-The right column shows the Fano factor for each model.
-Note especially the crucial diagnostic: (2) and (3) have $\nu$
-strictly below 1, while only for (4) and (5) can $\nu$ exceed 1.
-Models with Fano factors $\le 1$ cannot produce the FISH data
-observed in~\cite{Jones2014} without introducing additional
-assumptions and model complexity.
-}
-\label{fig:constit_cartoons}
-\end{figure}
-
-\subsubsection{Model 1 - Poisson promoter.}
-This is the picture from Jones et.\ al.~\cite{Jones2014}, which
-assumes that transcripts are produced as a Poisson process from a
-single promoter state. This model insists that the ``true'' mRNA
-distribution is Poisson, implying the Fano factor $\nu$ must be 1.
-In~\cite{Jones2014}, the authors carefully attribute deviations
-from Fano = 1 to intensity variability in fluorescent spots, gene
-copy number variation, and copy number fluctuations of the
-transcription machinery, e.g., RNAP itself.
-In this picture, the master equation makes no appearance,
-and all the corrections to Poisson behavior are derived
-as additive corrections to the Fano factor.
-For disproving the ``universal noise curve'' from So et
-al~\cite{So2011}, this picture was excellent. It is appealing in
-its simplicity, with only two parameters, the initiation rate $r$
-and degradation rate $\gamma$. Since $\gamma$ is independently
-known from other experiments, and the mean mRNA copy number is
-$r/\gamma$, $r$ is easily inferred from data.
-In other words, the model is not excessively complex for the data at hand.
-But for many interesting questions, for instance in the recent
-work~\mmnote{cite Manuel's preprint once posted},
-knowledge of means and variances alone is insufficient, and a
-description of the full distribution of molecular counts is necessary.
-For this we need a( slightly) more complex model than model 1
-that would allow us to incorporate the non-Poissonian features of
-constitutive promoters directly into a master equation formulation.
-
-\subsubsection{Model 2 - Two-state promoter, RNAP bound or unbound.}
-This model was considered in, e.g.,~\cite{Phillips2015a} and ~\cite{Phillips2019}.
-Here transcription initiation proceeds from the bound to the unbound state,
-reflecting the microscopic reality that an RNAP that has begun to elongate
-a transcript is no longer available at the start site to begin another.
-The problem with this picture is that the Fano factor is
-\begin{align}
-    \nu = 1 -
-        \frac{r\rate{k}{P}{+}}
-            {\left(\rate{k}{P}{+} + \rate{k}{P}{-} + r\right)
-             \left(\gamma + \rate{k}{P}{+} + \rate{k}{P}{-} + r\right)},
-\end{align}
-which is always $<1$.
-To make contact with the experimental reality of $\nu>1$,
-we will have to do, at a minimum,
-the same corrections to Poisson behavior as in model 1 above.
-So while this model adds an appealing element of microscopic
-reality, we are forced to reject it as the additional complexity
-is unable to capture the phenomenology of interest.
-Obviously the promoter state does in fact proceed through cycles
-of RNAP binding, initiating, and elongating, but it seems that
-the super-Poissonian noise in mRNA copy number we want to
-model must be governed by other features of the system.
-
-\subsubsection{Model 3 - Three-state promoter, multistep
-transcription initiation and escape.}
-How might we remedy the deficits of model 2?
-It is known~\cite{DeHaseth1998} that once RNAP initially binds
-the promoter region, a multitude of distinct steps occur
-sequentially before RNAP finally escapes into the elongation
-phase. Perhaps adding some of this mechanistic detail might
-rescue model 2? The next simplest refinement of model 2 could
-consider open complex formation and promoter escape as separate
-steps rather than as a single effective step. In other words, we
-construct model 3 by adding a single extra state to model 2, and
-we will label the two RNAP-bound states as the closed and open
-complexes, despite the true biochemical details certainly being
-more complex. The authors of~\cite{Mitarai2015}, for instance,
-considered this model, although they added an additional
-repressor bound state and did not explicitly consider the limit
-with no repressor that we analyze here. Again, our goal here is
-not a complete accounting of all the relevant biochemical detail;
-this is an exploratory search for the important features our
-model needs to include.
-
-Unfortunately, as the authors of~\cite{Mitarai2015} hint, this
-model too has Fano factor $\nu<1$. This can be seen by starting
-with their Eq.~A1 for the Fano factor of the analogous model with
-repressor and taking the limit as repressor concentration goes to
-zero. After substantial algebra, the result is
-\begin{align}
-\nu = 1 - \frac{r k_O k_P^+}{\mathcal{Z}}
-\frac{k_P^+ + k_P^- + k_O + \gamma}
-        {\mathcal{Z} + \gamma(k_P^+ + k_P^- + k_O + \gamma) + \gamma^2},
-\end{align}
-where we defined $\mathcal{Z} = r(k_O + k_P^-) + k_P^+(k_O + r)$
-for notational tidiness. This is necessarily less than 1 for
-arbitrary rate constants.
-
-In fact, we suspect \textit{any} model in which transcription
-proceeds through a multistep cycle must necessarily have $\nu<1$.
-The intuitive argument compares the waiting time distribution to
-traverse the cycle with the waiting time for a Poisson promoter
-(model 1) with the same mean time. The latter is simply an
-exponential distribution. The former is a convolution of multiple
-exponentials, and intuitively should be more peaked with a
-smaller fractional width than an exponential with the same mean.\footnote{
-This can be made more precise with a result
-from~\cite{Stewart2007}, who showed that the convolution of
-multiple gamma distributions (of which the exponential
-distribution is a special case) is to a very good approximation
-also gamma distributed. Using their Eq.~2 for the distribution of
-the convolution, with shape parameters set to 1 to
-give exponential distributions, the total waiting time
-distribution has a ratio of variance to squared mean 
-$\sigma^2/\mu^2 = \sum_i k_i^2/\left(\sum_i k_i\right)^2$,
-where the $k_i$ are the rates of the individual steps. Clearly
-this is less than 1 and therefore the total waiting time
-distribution is narrower than the corresponding exponential.} A
-less disperse waiting time distribution means transcription
-initations are more uniformly distributed in time relative to a
-Poisson process. Hence the distribution of mRNA over a population
-of cells should be less variable compared to Poisson, giving
-$\nu<1$.\footnote{
-This last step, while intuitive, can be argued by analogy to
-photon statistics where antibunching gives rise to sub-Poissonian
-noise~\cite{Paul1982, Zou1990}. Although loopholes exist, we do
-not expect they apply for our problem. Nevertheless we refrain
-from elevating this cycle/sub-Poissonian equivalence to a ``theorem.''
-\mmnote{Antibunching is the ``obvious'' analogy for me but it'd
-be totally out of left field for most readers, any suggestions on
-a more comprehensible reference??}
-}
-
-Regardless of the merits of this model in describing the noise
-properties of constitutive transcription initation, it does not
-capture the dominant source of noise we want to include in our
-phenomenological model, so we must discard these details and
-search elsewhere.
-
-\subsubsection{Model 4 - Two-state promoter,
-``active'' and ``inactive'' states.}
-Inspired by~\cite{Razo-Mejia2020}, we revisit an analog of model
-2, but the interpretation of the two states is changed. Rather
-than explicitly viewing them as RNAP bound and unbound, we view
-them as ``active'' and ``inactive,'' which are able and unable to
-initiate transcripts, respectively. We are noncommittal as to the
-microscopic details of these states.
-
-One interpretation~\cite{Chong2014, Sevier2016} for the active
-and inactive states is that it represents the promoter's
-supercoiling state: transitions to the inactive state are caused
-by accumulation of positive supercoiling, which inhibits
-transcription, and transitions back to ``active'' are caused by
-gyrase or other topoisomerases relieving the supercoiling. This
-is an interesting possibility because it would mean the timescale
-for promoter state transitions is driven by topoisomerase
-kinetics, not by RNAP kinetics. From in vitro measurments, the
-former are suggested to be of order minutes~\cite{Chong2014}.
-Contrast this with model 2, where the state transitions are
-assumed to be governed by RNAP, which, assuming a copy number per
-cell of order $10^3$, has a diffusion-limited association rate
-$k_{on} \sim 10^2~\text{s}^{-1}$ to a target promoter. Combined
-with known $K_D$'s of order $\mu$M, this gives an RNAP
-dissociation rate $k_{off}$ of order $10^2$. As we will show
-below, however, there are some lingering puzzles with
-interpreting this supercoiling interpretation, so we leave it as
-a speculation and refrain from assigning definite physical
-meaning to the two states in this model.
-
-Intuitively one might expect that, since transcripts are produced
-as a Poisson process only when the promoter is in one of the two
-states in this model, transcription initiations should now be
-``bunched,'' in constrast to the ``anti-bunching'' of models 2
-and 3 above. One might further guess that this bunching would
-lead to super-Poissonian noise in the mRNA distribution over a
-population of cells. An honest calculation of the Fano factor produces
-\begin{align}
-\nu &= 1 + \frac{r k^-}{(k^+ + k^- + \gamma)(k^+ + k^-)},
-\end{align}
-which is strictly greater than 1, verifying the above intuition.
-Note we have dropped the $P$ label on the promoter switching rates to
-emphasize that these very likely do not represent kinetics of RNAP itself.
-This calculation can also be sidestepped by noting that the model
-is mathematically equivalent to the simple repression repression model
-from~\cite{Jones2014}, with states and rates relabeled and reinterpreted.
-
-How does this model compare to model 1 above?
-In model 1, all non-Poisson features of the mRNA
-distribution were handled as extrinsic corrections.
-By contrast, here the 3 parameter model is used to fit the full
-mRNA distribution as measured in mRNA FISH experiments.
-In essense, all variability in the mRNA distribution is regarded
-as ``intrinsic,'' arising either from stochastic initiation or
-from switching between the two coarse-grained promoter states.
-The advantage of this approach is that it fits neatly into the
-master equation picture, and the parameters thus inferred can be used as
-input for more complicated models with regulation by transcription factors.
-
-While this seems promising, there is a major drawback for our purposes
-which was already uncovered by the authors of~\cite{Razo-Mejia2020}:
-the statistical inference problem is nonidentifiable,
-in the sense described in Section 4.3 of~\cite{Gelman2013},
-meaning it is impossible to infer the
-parameters $r$ and $k^-$ from the FISH data of~\cite{Jones2014}
-(as shown in Fig.~S2 of~\cite{Razo-Mejia2020}).
-Rather, only the ratio $r/k^-$ could be inferred. In that work,
-the problem was worked around with an informative prior on the
-ratio $k^-/k^+$. That approach is unlikely to work here, as,
-recall, our entire goal in modeling constitutive expression is to
-use it as the basis for a yet more complicated model, when we add
-on repression. But adding more complexity to a model that is
-already poorly identified is a fool's errand,
-so we will explore one more potential model.
-
-\subsubsection{Model 5 - One-state promoter with explicit bursts.}
-This model is inspired by the failure mode of model 4.
-The key observation above was that, as found
-in~\cite{Razo-Mejia2020}, only two parameters, $k^+$ and the
-ratio $r/k^-$, could be directly inferred from the FISH data
-from~\cite{Jones2014}.
-So let us take this seriously and imagine a model where these are
-the only two model parameters. What would this model look like?
-
-To develop some intuition, consider model 4 in the limit
-$k^+ \ll k^- \lesssim r$, which is roughly satisfied by the
-parameters inferred in~\cite{Razo-Mejia2020}.
-In this limit, the system spends the majority of its time in the inactive state, occasionally becoming active and making a burst of transcripts.
-This should call to mind the well-known phenomenon of transcriptional bursting, as reported in, e.g.,~\cite{Golding2005,Chong2014,Sevier2016}\mmnote{should probably add some more cites here}.
-Let us make this correspondence more precise.
-The mean dwell time in the active state is $1/k^-$.
-While in this state, transcripts are produced at a rate $r$ per
-unit time. So on average, $r/k^-$ transcripts are produced before
-the system switches to the inactive state. Once in the inactive
-state, the system dwells there for an average time $1/k^+$ before
-returning to the active state and repeating the process. $r/k^-$
-resembles an average burst size, and $1/k^+$ resembles the time
-interval between burst events. More precisely, $1/k^+$ is the
-mean time between the end of one burst and the start of the next,
-whereas $1/k^+ + 1/k^-$ would be the mean interval between the
-start of two successive burst events, but in the limit $k^+ \ll
-k^-$, $1/k^+ + 1/k^- \approx 1/k^+$. Note that this limit ensures
-that the waiting time between bursts is approximately
-exponentially distributed, with mean set by the only timescale
-left in the problem,
-$1/k^+$.\footnote{
-If instead it were the case that $k^+ \sim k^-$, then the waiting time
-$1/k^+ + 1/k^-$ would have a peaked distribution, but this does
-not appear to be the case for any of datasets
-from~\cite{Jones2014}.}
-
-Let us now verify this intuition with a precise derivation
-to check that $r/k^-$ is in fact the mean burst size
-and to obtain the full burst size distribution.
-Consider first a constant, known dwell time $T$ in the active state.
-Transcripts are produced at a rate $r$ per unit time, so the
-number of transcripts $n$ produced during $T$ fits exactly
-the ``story'' for the Poisson distribution, i.e.,
-\begin{equation}
-    P(n\mid T) = \frac{(rT)^n}{n!} \exp(-rT).
-\end{equation}
-Since the dwell time $T$ is unobservable, we actually want $P(n)$,
-the dwell time distribution with no conditioning on $T$.
-Basic rules of probability theory tell us we can write $P(n)$
-in terms of $P(n\mid T)$ as
-\begin{equation}
-    P(n) =\int_0^\infty P(n\mid T) P(T) dT.
-\end{equation}
-But we know the dwell time distribution $P(T)$,
-which is exponentially distributed according to
-\begin{equation}
-    P(T) = k^- \exp(-T k^-),
-\end{equation}
-so $P(n)$ can be written as
-\begin{equation}
-    P(n) = k^- \frac{r^n}{n!}
-            \int_0^\infty T^n\exp[-(r + k^-)T]\,dT.
-\end{equation}
-A standard integral table shows
-$\int_0^\infty x^n e^{-ax}\,dx = n!/a^{n+1}$, so
-\begin{equation}
-    P(n) = \frac{k^- r^n}{(k^- + r)^{n+1}}
-        = \frac{k^-}{k^- + r}
-            \left(\frac{r}{k^- + r}\right)^n
-        = \frac{k^-}{k^- + r}
-            \left(1 - \frac{k^-}{k^- + r}\right)^n,
-\end{equation}
-which is exactly the geometric distribution with standard parameter
-$\theta\equiv k^-/(k^- + r)$ and domain
-$n \in \{0, 1, 2, \dots\}$
-(this is one of two common conventions for the geometric distribution).
-The mean of the geometric distribution, with this convention, is
-\begin{align}
-\langle n\rangle = \frac{1 - \theta}{\theta}
-        = \left(1 - \frac{k^-}{k^- + r}\right)
-                    \frac{k^- + r}{k^-}
-        = \frac{r}{k^-},
-\end{align}
-exactly as we guessed intuitively above.
-
-So in taking the limit $r,k^-\rightarrow\infty$, $r/k^-\equiv b$,
-we obtain a model which effectively has only a single promoter state,
-which initiates bursts at rate $k^+$
-(transitions to the active state, in the model 4 picture).
-The master equation for mRNA copy number $m$ takes the form
-\begin{align}
-\begin{split}
-\deriv{t}{p(m,t)} = & (m+1)\gamma p(m+1,t) - m\gamma p(m,t) \\
-        &+ \sum_{j=0}^{m-1} k^+ p(j,t) Geom(m-j;b)
-         - \sum_{j=m+1}^\infty k^+ p(m,t) Geom(j-m;b),
-\end{split}
-\end{align}
-where $Geom(n;b)$ is the geometric distribution with mean~$b$, i.e.,
-$Geom(n;b) = \frac{1}{1+b}\left(\frac{b}{1+b}\right)^n$
-(with domain over nonnegative integers as above).
-The first two terms are the usual mRNA degradation terms.
-The third term enumerates all ways the system can produce
-a burst of transcripts and arrive at copy number $m$,
-given that it had copy number $j$ before the burst.
-The fourth term allows the system to start with copy number $m$,
-then produce a burst and end with copy number $j$.
-In fact this last sum has trivial $j$ dependence and
-simply enforces normalization of the geometric distribution.
-Carrying it out we have
-\begin{equation}
-\begin{split}
-\deriv{t}{p(m,t)} = & (m+1)\gamma p(m+1,t) - m\gamma p(m,t) \\
-        &+ \sum_{j=0}^{m-1} k^+ p(j,t) Geom(m-j;b)
-            - k^+ p(m,t),
-\end{split}
-\end{equation}
-This improves on model 4 in that now the parameters are easily inferred,
-as we will see later, and have clean interpretations.
-The non-Poissonian features are attributed to the emprically
-well-established phenomenological picture of bursty transcription.
-
-The big approximation in going from model 4 to 5 is that a burst
-is produced instantaneously rather than over a finite time.
-If the true burst duration is not short compared to transcription factor
-kinetic timescales, this could be a problem in that mean burst size
-in the presense and absence of repressors could change,
-rendering parameter inferences from the constitutive case inappropriate.
-Let us make some simple estimates of this.
-
-Consider the time delay between the first and final RNAPs
-in a burst initiating transcription
-(\textit{not} the time to complete transcripts,
-which potentially could be much longer.)
-If this timescale is short compared to the typical
-search timescale of transcription factors, then all is well.
-The estimates from deHaseth et.\ al.~\cite{DeHaseth1998}
-put RNAP's diffusion-limited on rate around
-$\sim\text{few}\times10^{-2}~\text{nM}^{-1}~\text{s}^{-1}$
-and polymerase loading as high as $1~\text{s}^{-1}$.
-Then for reasonable burst sizes of $<10$, it is reasonable to guess that
-bursts might finish initiating on a timescale of tens of seconds or less
-(with another 30-60 sec to finish elongation, but that does not matter here).
-A transcription factor with typical copy number of order 10 (or less)
-would have a diffusion-limited association rate of order
-$(10~\text{sec})^{-1}$.
-Higher copy number TFs tend to have many binding sites over the genome,
-which should serve to pull them out of circulation and keep their
-effective association rates from rising too large.
-Therefore, there is \textit{perhaps} a timescale separation
-possible between transcription factor association rates and burst
-durations, but this assumption could very well break down, so we
-will have to keep it in mind when we infer repressor rates from
-the Jones et.\ al.\ FISH data later.
-
-In reflecting on these 5 models, the reader may feel that
-exploring a multitude of potential models just to return to a
-very minimal phenomenological model of bursty transcription may
-seem highly pedantic. But the purpose of the exercise was to
-examine a host of models from the literature and understand why
-they are insufficient, one way or another, for our purposes.
-Along the way we have learned that the detailed kinetics of RNAP
-binding and initiating transcription are probably irrelevant for
-setting the population distribution of mRNA.
-The timescales are simply too fast, and as we will see later
-in~\fig{fig:constit_post}, the noise seems to be governed by
-slower timescales. Perhaps in hindsight this is not surprising:
-intuitively, the degradation rate $\gamma$ sets the fundamental
-timescale for mRNA dynamics, and any other processes that
-substantially modulate the mRNA distribution should not differ
-from $\gamma$ by orders of magnitude.
-
+% !TEX root = ./busty_transcription.tex
 \section{Finding the ``right'' model: Bayesian parameter inference and model selection}
@@
 
-From consideration of Fano factors in the previous section, we
-suspect that model 5 in~\fig{fig:constit_cartoons},
-a one-state bursty model of constitutive promoters,
-achieves a Goldilocks level of complexity.
-Does this stand up to closer scrutiny, namely, comparison to
-full mRNA distributions rather than simply their moments?
-We will test this thoroughly on all the data
-from constitutive promoters from Jones et.\ al.~\cite{Jones2014}.
+From consideration of Fano factors in the previous section, we suspect that
+model 5 in~\fig{fig:constit_cartoons}, a one-state bursty model of constitutive
+promoters, achieves a Goldilocks level of complexity. Does this stand up to
+closer scrutiny, namely, comparison to full mRNA distributions rather than
+simply their moments? We will test this thoroughly on all the data from
+constitutive promoters from Jones et.\ al.~\cite{Jones2014}.
 
-It will be instructive, however, to first consider the
-Poisson promoter, model 1 in~\fig{fig:constit_cartoons}.
-This is a convenient proxy for models 2 and 3, whose exact
-steady-state mRNA distributions would be tedious to calculate
-and we will not explicitly consider.
-But since they have Fano factors $\nu\le 1$, they can
-fare no better than the simple Poisson model.
-We will also not explicitly consider model 4 from~\fig{fig:constit_cartoons}
-since it was already thoroughly analyzed in~\cite{Razo-Mejia2020},
-and since model 5 can be viewed as a special case of it.
+It will be instructive, however, to first consider the Poisson promoter, model 1
+in~\fig{fig:constit_cartoons}. This is a convenient proxy for models 2 and 3,
+whose exact steady-state mRNA distributions would be tedious to calculate and we
+will not explicitly consider. But since they have Fano factors $\nu\le 1$, they
+can fare no better than the simple Poisson model. We will also not explicitly
+consider model 4 from~\fig{fig:constit_cartoons} since it was already thoroughly
+analyzed in~\cite{Razo-Mejia2020}, and since model 5 can be viewed as a special
+case of it.
 
-In performing parameter inference on the FISH data
-of~\cite{Jones2014}, building a generative model is
-straightforward. The data are single-cell mRNA counts, and we
-assume different cells to be independent and identically
-distributed from the underlying generative model, which is just
-the steady-state mRNA distribution computed from the master
-equation of the underlying model.
+In performing parameter inference on the FISH data of~\cite{Jones2014}, building
+a generative model is straightforward. The data are single-cell mRNA counts, and
+we assume different cells to be independent and identically distributed from the
+underlying generative model, which is just the steady-state mRNA distribution
+computed from the master equation of the underlying model.
 
@@
 \mmnote{A lot of the details I've written here might be excessive for main text and serve better as an intro-by-example to Bayes in the appendix, then reduce main text to a bit more bare bones?}
-For this model the master equation of interest
-is~\eq{eq:poisson_promoter_cme} with repressor set to zero, i.e.,
+For this model the master equation of interest is~\eq{eq:poisson_promoter_cme}
+with repressor set to zero, i.e.,
 \begin{equation}
@@
 \end{equation}
-with $\lambda=r/\gamma$. This steady state distribution defines our
-likelihood $p(m\mid\lambda)$, i.e.,
+with $\lambda=r/\gamma$. This steady state distribution defines our likelihood
+$p(m\mid\lambda)$, i.e.,
 \begin{equation}
@@
 \end{equation}
-is the probability of observing
-a single cell with $m$ mRNAs given a known value of $\lambda$.
+is the probability of observing a single cell with $m$ mRNAs given a known value
+of $\lambda$.
 
-With our likelihood in hand, we can invoke Bayes theorem to write down
-the posterior distribution
-on our model parameter $\lambda$, given by
+With our likelihood in hand, we can invoke Bayes theorem to write down the
+posterior distribution on our model parameter $\lambda$, given by
 \begin{equation}
@@
 \end{equation}
-where our data $D=\{m_1, m_2,\dots, m_N\}$ is a list of mRNA counts
-over a population of $N$ identical cells.
-As is standard we have neglected the factor $p(D)$ on the right
-hand side since it is independent of $\lambda$ and serves only as
-a normalization factor.
-Since we assume each cell's mRNA count is independent of others,
-the likelihood is simply a product of the
-single cell likelihoods given by \eq{eq:poisson_inference010} above, so
+where our data $D=\{m_1, m_2,\dots, m_N\}$ is a list of mRNA counts over a
+population of $N$ identical cells. As is standard we have neglected the factor
+$p(D)$ on the right hand side since it is independent of $\lambda$ and serves
+only as a normalization factor. Since we assume each cell's mRNA count is
+independent of others, the likelihood is simply a product of the single cell
+likelihoods given by \eq{eq:poisson_inference010} above, so
 \begin{equation}
@@
 \end{equation}
-which is read as the data $D$, conditioned on a value of
-$\lambda$, is Poisson distributed with mean $\lambda$.
-The $\sim$ operator literally denotes ``is distributed according to.''
+which is read as the data $D$, conditioned on a value of $\lambda$, is Poisson
+distributed with mean $\lambda$. The $\sim$ operator literally denotes ``is
+distributed according to.''
 
 To proceed we need to specify a prior. In this case we are extremely data-rich,
-as the dataset from Jones et.\ al~\cite{Jones2014} has of order
-1000-3000 single-cell measurements for each promoter,
-so our choice of prior matters little here,
-as long as it does not exclude the true value of $\lambda$.
-A convenient choice for our Poisson likelihood is its conjugate prior,
-the gamma distribution. Although they only exist for
-certain special likelihoods, conjugate priors have the nice
-feature that the posterior takes the same functional form as the
-prior, with updated parameters. This makes calculations
-analytically tractable and also offers a nice interpretation of
-the inference procedure as updating our knowledge about the model
-parameters.
+as the dataset from Jones et.\ al~\cite{Jones2014} has of order 1000-3000
+single-cell measurements for each promoter, so our choice of prior matters
+little here, as long as it does not exclude the true value of $\lambda$. A
+convenient choice for our Poisson likelihood is its conjugate prior, the gamma
+distribution. Although they only exist for certain special likelihoods,
+conjugate priors have the nice feature that the posterior takes the same
+functional form as the prior, with updated parameters. This makes calculations
+analytically tractable and also offers a nice interpretation of the inference
+procedure as updating our knowledge about the model parameters.
 
-Putting a gamma prior on $\lambda$ introduces two new parameters
-$\alpha$ and $\beta$ which parametrize the gamma distribution
-itself, which we use to encode the range of $\lambda$ values we
-view as reasonable. Recall $\lambda$ is the mean steady-state
-mRNA count per cell, which \textit{a priori} could plausibly be
-anywhere from 0 to a few hundred. $\alpha=1$ and $\beta=1/50$
-achieve this, since the gamma distribution is strictly positive
-with mean $\alpha/\beta$ and standard deviation
-$\sqrt{\alpha}/\beta$. To be explicit, then, our prior is
+Putting a gamma prior on $\lambda$ introduces two new parameters $\alpha$ and
+$\beta$ which parametrize the gamma distribution itself, which we use to encode
+the range of $\lambda$ values we view as reasonable. Recall $\lambda$ is the
+mean steady-state mRNA count per cell, which \textit{a priori} could plausibly
+be anywhere from 0 to a few hundred. $\alpha=1$ and $\beta=1/50$ achieve this,
+since the gamma distribution is strictly positive with mean $\alpha/\beta$ and
+standard deviation $\sqrt{\alpha}/\beta$. To be explicit, then, our prior is
 \begin{equation}
@@
 
-As an aside, note that if we did not know that our prior was a
-conjugate prior, we could still write down our posterior
-distribution from its definition as
+As an aside, note that if we did not know that our prior was a conjugate prior,
+we could still write down our posterior distribution from its definition as
 \begin{equation}
@@
 \end{equation}
-Without foreknowledge that this in fact reduces to a gamma distribution,
-this expression might appear rather inscrutable. When conjugate
-priors are unavailable for the likelihood of interest - which is
-almost always the case for models with $>1$ model parameter -
-this inscrutability is the norm, and making sense of posteriors
-analytically is almost always impossible. Fortunately, MCMC
-sampling provides us a powerful method of constructing posteriors
-numerically which we will make use of extensively.
+Without foreknowledge that this in fact reduces to a gamma distribution, this
+expression might appear rather inscrutable. When conjugate priors are
+unavailable for the likelihood of interest - which is almost always the case for
+models with $>1$ model parameter - this inscrutability is the norm, and making
+sense of posteriors analytically is almost always impossible. Fortunately, MCMC
+sampling provides us a powerful method of constructing posteriors numerically
+which we will make use of extensively.
 
-Since we did use a conjugate prior, we may simply look up our
-posterior in any standard table of conjugate priors-likelihood
-pairs, from which we find that
+Since we did use a conjugate prior, we may simply look up our posterior in any
+standard table of conjugate priors-likelihood pairs, from which we find that
 \begin{equation}
@@
 \end{equation}
-where we defined the sample mean $\bar{m} = \frac{1}{N}\sum_k m_k$
-for notational convenience.
-A glance at the FISH data from~\cite{Jones2014} reveals that $N$ is
-$\mathcal{O}(10^3)$ and $\langle m\rangle \gtrsim 0.1$ for all
-constitutive strains in~\cite{Jones2014}, so
-$\bar{m}N \gtrsim 10^2$. Therefore as we suspected,
-our prior parameters are completely overwhelmed by the data.
-In fact, $\bar{m}N$ and $N$ are so large that we can,
-to an excellent approximation, ignore the $\alpha$ and $\beta$ dependence and
-approximate the gamma distribution as a Gaussian with mean $\bar{m}$ and
-standard deviation $\sqrt{\bar{m}/N}$, giving
+where we defined the sample mean $\bar{m} = \frac{1}{N}\sum_k m_k$ for
+notational convenience. A glance at the FISH data from~\cite{Jones2014} reveals
+that $N$ is $\mathcal{O}(10^3)$ and $\langle m\rangle \gtrsim 0.1$ for all
+constitutive strains in~\cite{Jones2014}, so $\bar{m}N \gtrsim 10^2$. Therefore
+as we suspected, our prior parameters are completely overwhelmed by the data. In
+fact, $\bar{m}N$ and $N$ are so large that we can, to an excellent
+approximation, ignore the $\alpha$ and $\beta$ dependence and approximate the
+gamma distribution as a Gaussian with mean $\bar{m}$ and standard deviation
+$\sqrt{\bar{m}/N}$, giving
 \begin{equation}
@@
 \end{equation}
-As an example with real numbers, for the \textit{lacUV5} promoter,
-Jones et.\ al~\cite{Jones2014} measured 2648 cells with an average
-mRNA count per cell of $\bar{m} \approx 18.7$. In this case then,
-our posterior is
+As an example with real numbers, for the \textit{lacUV5} promoter, Jones et.\
+al~\cite{Jones2014} measured 2648 cells with an average mRNA count per cell of
+$\bar{m} \approx 18.7$. In this case then, our posterior is
 \begin{equation}
@@
 \end{equation}
-which suggests we have inferred our model's one parameter to a precision of order 1\%.
+which suggests we have inferred our model's one parameter to a precision of
+order 1\%.
 
-This is not wrong, but it is not the full story. The model's
-posterior is tightly constrained, but is it a good generative
-model? In other words, does the model generate data that look similar
-to our actual data, and is it therefore plausible that the
-model captures the important features of the data generating
-process? This intuitive notion can be codified with
-\textit{posterior predictive checks}, or PPCs, and we will see
-that this simple Poisson model fails badly.
+This is not wrong, but it is not the full story. The model's posterior is
+tightly constrained, but is it a good generative model? In other words, does the
+model generate data that look similar to our actual data, and is it therefore
+plausible that the model captures the important features of the data generating
+process? This intuitive notion can be codified with \textit{posterior predictive
+checks}, or PPCs, and we will see that this simple Poisson model fails badly.
 
-The intuitive idea of posterior predictive checks is simple:
-\mmnote{Not sure if all this stuff is clearer by example,
-but then does it confuse the reader in generalizing to other models?}
+The intuitive idea of posterior predictive checks is simple: \mmnote{Not sure if
+all this stuff is clearer by example, but then does it confuse the reader in
+generalizing to other models?}
 \begin{enumerate}
-\item Make a random draw of the model parameter $\lambda$
-from the posterior distribution.
-\item Plug that draw into the likelihood and generate
-a synthetic dataset $\{m_k\}$ conditioned on $\lambda$.
+\item Make a random draw of the model parameter $\lambda$ from the posterior
+distribution.
+\item Plug that draw into the likelihood and generate a synthetic dataset
+$\{m_k\}$ conditioned on $\lambda$.
 \item Repeat many times.
 \end{enumerate}
-More formally, the posterior predictive distribution can be thought of
-as the distribution of future yet-to-be-observed data,
-conditioned on the data we have already observed.
-Clearly if those data appear quite different,
-the model has a problem. Put another way, if we suppose
-the generative model is true, then the synthetic datasets
-we generate should resemble the actual observed data, and if not,
-it suggests the model is missing important features.
-All the data we consider in this work are 1D (distributions of mRNA counts
-over a population) so ECDFs are an excellent visual means of comparing
-synthetic and observed datasets. In general for higher dimensional datasets,
-much of the challenge is in merely designed good visualizations
-that can actually show if synthetic and observed data are similar or not.
+More formally, the posterior predictive distribution can be thought of as the
+distribution of future yet-to-be-observed data, conditioned on the data we have
+already observed. Clearly if those data appear quite different, the model has a
+problem. Put another way, if we suppose the generative model is true, then the
+synthetic datasets we generate should resemble the actual observed data, and if
+not, it suggests the model is missing important features. All the data we
+consider in this work are 1D (distributions of mRNA counts over a population) so
+ECDFs are an excellent visual means of comparing synthetic and observed
+datasets. In general for higher dimensional datasets, much of the challenge is
+in merely designed good visualizations that can actually show if synthetic and
+observed data are similar or not.
 
-For our example Poisson promoter model then, we merely draw many
-random numbers, say 1000, from the Gaussian posterior. For each one
-of those draws, we generate a dataset from the likelihood,
-i.e., we draw 2648 (the number of observed cells in the actual dataset)
-Poisson-distributed numbers for each of the 1000 posterior draws,
-for a total of 2648000 samples from the posterior predictive distribution.
+For our example Poisson promoter model then, we merely draw many random numbers,
+say 1000, from the Gaussian posterior. For each one of those draws, we generate
+a dataset from the likelihood, i.e., we draw 2648 (the number of observed cells
+in the actual dataset) Poisson-distributed numbers for each of the 1000
+posterior draws, for a total of 2648000 samples from the posterior predictive
+distribution.
 
-To compare so many samples with the actual observed data,
-one excellent visualization for 1D data is ECDFs of the quantiles,
-as shown for our Poisson model in~\fig{fig:uv5_post}.
-In this example, the median for each possible mRNA count is shown
-as a dark green line, while shaded bands of progressively lighter green
-contain 25\%, 50\%, 75\%, and 95\% of the posterior predictive samples.
-Plotting quantiles in this way gives us a sense of the range of data
-we might consider plausible, under the assumption that the model is true.
-In this case it is quite obvious that the observed data, plotted in orange,
-could not plausibly come from this Poisson generative model.
-The other model with a PPC plotted in~\fig{fig:uv5_post},
-bursty model 5 with a negative binomial steady-state mRNA distribution,
-looks like a good candidate by eye. We cover it in more detail later.
+To compare so many samples with the actual observed data, one excellent
+visualization for 1D data is ECDFs of the quantiles, as shown for our Poisson
+model in~\fig{fig:uv5_post}. In this example, the median for each possible mRNA
+count is shown as a dark green line, while shaded bands of progressively lighter
+green contain 25\%, 50\%, 75\%, and 95\% of the posterior predictive samples.
+Plotting quantiles in this way gives us a sense of the range of data we might
+consider plausible, under the assumption that the model is true. In this case it
+is quite obvious that the observed data, plotted in orange, could not plausibly
+come from this Poisson generative model. The other model with a PPC plotted
+in~\fig{fig:uv5_post}, bursty model 5 with a negative binomial steady-state mRNA
+distribution, looks like a good candidate by eye. We cover it in more detail
+later.
 
@@
 \includegraphics[width=0.9\textwidth]{../figures/fig2/fig2pt1.pdf}
-\caption{\textbf{Constitutive promoter posterior
-        inference and predictive checks.}
-        (A) shows the joint posterior density from MCMC sampling of model 5,
-        bursty promoter with negative binomially-distributed steady state.
-        (B) shows the ECDF of the observed population distribution
-        of mRNA transcripts under the control of a constitutive lacUV5 promoter.
-        Posterior predictive percentiles are shown as colored bands for
-        model (1), Poisson, and model (5), negative binomial.
-        }
+\caption{\textbf{Constitutive promoter posterior inference and predictive
+        checks.} (A) shows the joint posterior density from MCMC sampling of
+        model 5, bursty promoter with negative binomially-distributed steady
+        state. (B) shows the ECDF of the observed population distribution of
+        mRNA transcripts under the control of a constitutive lacUV5 promoter.
+        Posterior predictive percentiles are shown as colored bands for model
+        (1), Poisson, and model (5), negative binomial.}
 \label{fig:uv5_post}
@@
 \begin{itemize}
-\item Bursty model. Write Bayes, sketch story of NB likelihood, sketch choice
-of priors, show~\fig{fig:uv5_post}, see bursty $\gg$ Poisson model.
-\item All 18 promoters. Hey the model still works on things other than
-UV5, cool. Interesting and somewhat counterintuitive scaling b/w
-burst rate and binding energy. Puzzle in comparing w/ Chong2014
-supercoiling model. Are we missing important things?? Unclear, we
-leave as open question. Good enough for our purposes: posterior is identifiable, PPC is great, and of the models we've thought of it is unique in satisfying both.
+\item Bursty model. Write Bayes, sketch story of NB likelihood, sketch choice of
+priors, show~\fig{fig:uv5_post}, see bursty $\gg$ Poisson model.
+\item All 18 promoters. Hey the model still works on things other than UV5,
+cool. Interesting and somewhat counterintuitive scaling b/w burst rate and
+binding energy. Puzzle in comparing w/ Chong2014 supercoiling model. Are we
+missing important things?? Unclear, we leave as open question. Good enough for
+our purposes: posterior is identifiable, PPC is great, and of the models we've
+thought of it is unique in satisfying both.
 \end{itemize}
@@
 \caption{\textbf{Constitutive promoter model comparison.}
-    (A) Joint posterior distributions for burst rate $k_i$ and
-    mean burst size $b$ for 18 promoters from Jones 2014.
-    Each contour indicates the 95\% highest posterior density
-    region for a particular promoter, i.e., on a plot analogous
-    to~\fig{fig:uv5_post}, the contour would enclose
-    approximately 95\% of the samples, and thus 95\% of the
-    probability mass.
-    Note that the vertical axis is shared with (B).
-    Panel (B) plots the burst rate $k_i$ vs.\ binding energy from energy
-    matrices from Brewster 2012.
-    The dotted line shows the predicted slope according to
-    $\ln k_i/\gamma \sim \text{constant} - \beta\Delta\epsilon_P$,
-    described in text.
-    }
+    (A) Joint posterior distributions for burst rate $k_i$ and mean burst size
+    $b$ for 18 promoters from Jones 2014. Each contour indicates the 95\%
+    highest posterior density region for a particular promoter, i.e., on a plot
+    analogous to~\fig{fig:uv5_post}, the contour would enclose approximately
+    95\% of the samples, and thus 95\% of the probability mass. Note that the
+    vertical axis is shared with (B). Panel (B) plots the burst rate $k_i$ vs.\
+    binding energy from energy matrices from Brewster 2012. The dotted line
+    shows the predicted slope according to $\ln k_i/\gamma \sim \text{constant}
+    - \beta\Delta\epsilon_P$, described in text.}
 \label{fig:constit_post}
@@
 \begin{itemize}
-\item Full model.
-Write Bayes, handwave 2F1 dist story in limits of weak, strong,
-and intermediate repression.
-\item Hey look it mostly works.
-With full model, 9D model is well identified, though most of the
-individual experiments unsurprisingly are not: only the ratio of
-repressor rates is identifiable. PPC is not perfect, it kinda
-misses some rep/op pairs but overall it's surprisingly predictive
-considering how strong were the assumptions that went into it.
-\item Do we chalk this up to experimental imperfections or is there
-interesting biology to be found in the disagreements?
-I'm inclined towards the former. Even just catching cells in
-truly apples-to-apples growth phases, or matching up parameters
-of imaging sessions weeks or months apart, is hardly trivial.
-\item The alternative is to refine our model and/or add on
-complexity, and it's not at all clear what additions to make to
-the model, especially how to produce the long tail of repressed
-strains over UV5. I see no way to get that without getting into
-complicated time-correlation/history effects, where
+\item Full model. Write Bayes, handwave 2F1 dist story in limits of weak,
+strong, and intermediate repression.
+\item Hey look it mostly works. With full model, 9D model is well identified,
+though most of the individual experiments unsurprisingly are not: only the ratio
+of repressor rates is identifiable. PPC is not perfect, it kinda misses some
+rep/op pairs but overall it's surprisingly predictive considering how strong
+were the assumptions that went into it.
+\item Do we chalk this up to experimental imperfections or is there interesting
+biology to be found in the disagreements? I'm inclined towards the former. Even
+just catching cells in truly apples-to-apples growth phases, or matching up
+parameters of imaging sessions weeks or months apart, is hardly trivial.
+\item The alternative is to refine our model and/or add on complexity, and it's
+not at all clear what additions to make to the model, especially how to produce
+the long tail of repressed strains over UV5. I see no way to get that without
+getting into complicated time-correlation/history effects, where
 transcription...rebounds?? After being repressed?? Sounds odd.
-\item Comparison between my inferred rates, equilibrium binding E, and single-molecule measurements.
-Not quite as slam-dunk as I'd hoped because of $\gamma$ dependence, but rates are w/in factor of 2 or better, seems quite good really!
-With exception of one Oid data data point, free energies are w/in about 0.5 kT also.
-That's not massively larger than the repeatability threshold of fold-change measurements, at maybe 0.1 to 0.3 kT.
-\item Surprising that so many model parameters can be inferred from just a few distributions that, by eye, don't have strong features!
-Does require some assumptions, e.g., about equivalence of rates across experiments, but that seems quite reasonable.
+\item Comparison between my inferred rates, equilibrium binding E, and
+single-molecule measurements. Not quite as slam-dunk as I'd hoped because of
+$\gamma$ dependence, but rates are w/in factor of 2 or better, seems quite good
+really! With exception of one Oid data data point, free energies are w/in about
+0.5 kT also. That's not massively larger than the repeatability threshold of
+fold-change measurements, at maybe 0.1 to 0.3 kT.
+\item Surprising that so many model parameters can be inferred from just a few
+distributions that, by eye, don't have strong features! Does require some
+assumptions, e.g., about equivalence of rates across experiments, but that seems
+quite reasonable.
 \end{itemize}
@@
 
-Figure~3 presents inferred transcription factor rates, compares
-them to single-molecule measurements, and compares their ratios
-to gold-standard binding energies.
+Figure~3 presents inferred transcription factor rates, compares them to
+single-molecule measurements, and compares their ratios to gold-standard binding
+energies.
 \begin{figure}%[h!]
@@
 (A) depicts the cartoon of simple repression for which we infer parameters.
-\mmnote{
-Not sure panel (A) is necessary, we've already shown it in at least
-2 other figs. We could instead show some alternative slices of 9D posterior
-as a companion to (B), or something totally different\dots??}
-(B) shows 50\% and 95\% contours of several 2D slices of the 9D posterior
-distribution, from fitting one unbinding rate for each operator (O1, O2, O3)
-and one binding rate for each aTc concentration (corresponding to an
-unknown mean repressor copy number).
-(C) compares ratios of our inferred unbinding rates with operator
-binding energy differences measured in Garcia \& Phillips 2011 (triangles)
-and Razo-Mejia et.\ al.\ 2018 (squares).
-(D) compares our inferred unbinding rates with single-molecule measurements
-of the same from Hammar et.\ at.\ 2014 \mmnote{Johan Elf's group}.
-Values are of similar order of magnitude,
-but a precise quantitative comparison unfortunately depends sensitively
-on the assumed mRNA lifetime, which was not measured.
-    }
+\mmnote{Not sure panel (A) is necessary, we've already shown it in at least 2
+other figs. We could instead show some alternative slices of 9D posterior as a
+companion to (B), or something totally different\dots??} (B) shows 50\% and 95\%
+contours of several 2D slices of the 9D posterior distribution, from fitting one
+unbinding rate for each operator (O1, O2, O3) and one binding rate for each aTc
+concentration (corresponding to an unknown mean repressor copy number). (C)
+compares ratios of our inferred unbinding rates with operator binding energy
+differences measured in Garcia \& Phillips 2011 (triangles) and Razo-Mejia et.\
+al.\ 2018 (squares). (D) compares our inferred unbinding rates with
+single-molecule measurements of the same from Hammar et.\ at.\ 2014
+\mmnote{Johan Elf's group}. Values are of similar order of magnitude, but a
+precise quantitative comparison unfortunately depends sensitively on the assumed
+mRNA lifetime, which was not measured.}
 \label{fig3:kR_inferences}
@@
 \caption{\textbf{Posterior predictive checks.}
-(A) duplicates the posterior from~\fig{fig3:kR_inferences}B.
-(B), (C), and (D) show samples from the posterior predictive distribution
-for each of the datasets used in the inference.
-Samples are sorted by operator and plotted separately simply for visual clarity.
-The unrepressed promoter, UV5, is shown with each as a reference point.
-\mmnote{
-Obviously the PPC is not perfect,
-but considering the relative simplicity of the model, I'm pleasantly surprised
-how well it worked and that the model was even identifiable at all.
-Interestingly, and perhaps not surprisingly, only 3 of the 9 datasets
-were identifiable if fit in isolation: otherwise,
-it was only by making assumptions that certain rates are equal
-across datasets was identifiability achieved.
-}
-    }
+(A) duplicates the posterior from~\fig{fig3:kR_inferences}B. (B), (C), and (D)
+show samples from the posterior predictive distribution for each of the datasets
+used in the inference. Samples are sorted by operator and plotted separately
+simply for visual clarity. The unrepressed promoter, UV5, is shown with each as
+a reference point. \mmnote{Obviously the PPC is not perfect, but considering the
+relative simplicity of the model, I'm pleasantly surprised how well it worked
+and that the model was even identifiable at all. Interestingly, and perhaps not
+surprisingly, only 3 of the 9 datasets were identifiable if fit in isolation:
+otherwise, it was only by making assumptions that certain rates are equal across
+datasets was identifiability achieved.}}
 \label{fig4:ppc}
 \end{figure}
-
-\section{Discussion and future work}
-\mmnote{
-Do all my approximations work simply because galK happens to have
-a low duty cycle of supercoiling, $\beta/\alpha$ in Chong2014
-picture? A potentially interesting test: put in the dozen or so
-Hernan constructs at some other locus, preferably one of the ones
-Chong identified as having a large duty cycle (large $\beta/\alpha$),
-i.e., the opposite limit of galK. Then just measure fold-changes.
-You don't even need to do FISH, although that would be even
-better. If their model is right, $\rho$ should be approximately
-set by the duty cycle of supercoiling, NOT by RNAP kinetics.
-\textit{This is a NICE prediction for future work.} \textit{cspE}
-is claimed nonessential, prediction is you drop HG reporter
-constructs in there and the inferred binding energies of all lacI
-operators shift the same amount, by about $\ln(3)\sim 1kT$. Not
-huge, but detectable, although might be confounding effects from
-moving to a new locus. With FISH measurements, $k^+$ and $k^-$ in
-Fig 2, model 4 should now be separately inferrable if Chong2014
-measurements of $\beta/\alpha$ are correct. This would require an
-obnoxious amount of cloning, but once you did that, the
-measurements would be same old stuff that RP lab does.}
-
-%%%%%%%%%%%%%%%%%%%%% APPENDICES %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\appendix
-
-\section{Bayesian inference}
-\mmnote{
-\begin{itemize}
-\item Bayes theorem.
-Elaborate on posterior as plausibility over
-space of model parameters.
-\item Choosing likelihoods.
-Easy here, just the steady-state distributions of our master
-equations, assuming iid single-cell measurements.
-\item Choosing priors.
-Not critical here because we are data-rich and nonhierarchical,
-as long as we don't exclude ground truth. In other applications,
-prior can matter a lot. This is a feature, not a bug: it's
-revealing hidden structure that was implied by your modeling
-assumptions and you just didn't realize it. Predictive checks to
-verify you're not excluding any plausible values.
-\item Marginalization/expectations.
-Analytically hard, trivial with sampling.
-\item MCMC.
-Lot's of fun here, and deep analogies to stat mech and
-Hamiltonian mechanics worth exploring. In a perfect world, you'd
-wave a magic wand and generate independent draws from target
-posterior. But we don't know how. Next best: generate a Markov
-chain of $N$ \textit{correlated} samples, estimate an
-autocorrelation time $\tau$, then you've got roughly $N/\tau$
-effective samples. Obviously with some assumptions.
-\item Posterior predictive checks.
-So the posterior is fine, but is the model actually not wrong?
-PPC is essential to verify that your model is a plausible
-generative process for the data.
-\item Discuss identifiability, degeneracy, and failed PPCs in main text. Better thru example anyway.
-\end{itemize}
-}
-
-\section{Derivations of constitutive promoter models}
-\subsection{Model 2}
-\mmnote{Obtain Fano factor from CME.}
-\subsection{Model 3}
-\mmnote{Obtain Fano factor by taking limit from Mitarai et al's expression.}
-\subsection{Model 4}
-\mmnote{Obtain Fano factor by relabeling cartoon from Jones2014.}
-
-\section{Generating function solution for bursty promoters}
-
-\subsection{Constitutive promoter with bursts}
-
-\subsubsection{From master equation to generating function}
-
-We would like to derive the full probability distribution of
-mRNA copy number per cell for a model combining simple repression
-with a promoter featuring bursty transcription initiation.
-Before tackling mRNA and repression together, let us
-just consider the bursty consititutive promoter,
-model 5 in~\fig{fig:constit_cartoons}
-\mmnote{replicating Charlotte's notes; she says it's worked out
-in~\cite{Paulsson2000}, but I don't see it, at least not in notation I can
-comprehend. Need to find a cite for this,
-or maybe just cite Charlotte, personal comm, since that's where I actually got it\dots}.
-The master equation of interest is
-\begin{align}
-\deriv{t}p(m,t) = (m+1)\gamma p(m+1,t) - m\gamma p(m,t) - r p(m,t)
-        + r \sum_{m^\prime=0}^m G_{m-m^\prime}(\theta) p(m^\prime,t),
-\label{eq:1state_unreg_003}
-\end{align}
-where $G_{k}(\theta)$ is the geometric distribution defined as
-\begin{align}
-G_{k}(\theta) = \theta(1 - \theta)^k, \, k\in\{0,1,2,\dots\}.
-\end{align}
-With this convention, the mean burst size $b = (1-\theta)/\theta$.
-$\gamma$ and $r$ are mRNA degradation rates and transcription burst
-initiation rates, resp. The last term represents all ways the system could
-end up with $m$ mRNAs, having started the burst with $m^\prime$. Define
-$\lambda = r/\gamma$ and nondimensionalize time by $\gamma$, giving
-\begin{align}
-\deriv{t}p(m,t) = (m+1)p(m+1,t) - m p(m,t) - \lambda p(m,t)
-        + \lambda \sum_{m^\prime=0}^m G_{m-m^\prime}(\theta) p(m^\prime,t),
-\end{align}
-The probability generating function is defined as
-\begin{align}
-F(z,t) = \sum_{m=0}^\infty z^m p(m,t).
-\end{align}
-Multiply both sides of the CME by $z^m$ and sum over all $m$ to get
-\begin{align}
-\pderiv[F]{t} = \sum_{m=0}^\infty z^m
-\left(
-(m+1)p(m+1,t) - m p(m,t) - \lambda p(m,t)
-+ \lambda \sum_{m^\prime=0}^m G_{m-m^\prime}(\theta) p(m^\prime,t)
-\right).
-\end{align}
-The first two terms can be rewritten as derivatives of the
-generating function, i.e.,
-\begin{align}
-\sum_{m=0}^\infty z^m (m+1)p(m+1,t)
-=
-\pderiv{z} \sum_{m=0}^\infty z^{m+1} p(m+1,t)
-=
-\pderiv[F]{z}
-\end{align}
-and
-\begin{align}
-\sum_{m=0}^\infty z^m mp(m,t)
-=
-z\pderiv{z} \sum_{m=0}^\infty z^m p(m,t)
-=
-z\pderiv[F]{z}.
-\end{align}
-The final term is the most trouble. The trick is to reverse
-the default order of the sums, as in
-\begin{align}
-&\lambda \sum_{m=0}^\infty \sum_{m^\prime=0}^m
-        z^m \theta(1 - \theta)^{m-m^\prime} p(m^\prime,t)
-\\=
-&\lambda\theta \sum_{m^\prime=0}^\infty z^{m^\prime} p(m^\prime,t)
-        \sum_{m=m^\prime}^\infty
-        z^{m-m^\prime} (1 - \theta)^{m-m^\prime}
-\\=
-&\frac{\lambda\theta F}{1 - (1 - \theta)z}.
-\end{align}
-This works because after reordering the sums, the sum on $m$ is just
-a geometric series. Its result is independent of $m^\prime$, so the sum on
-$m^\prime$ reduces to the definition of the generating function.
-Putting these results together, the PDE for the generation function is
-\begin{align}
-\pderiv[F]{t} = \pderiv[F]{z} - z\pderiv[F]{z} - \lambda F
-        + \frac{\lambda\theta F}{1-z(1-\theta)}.
-\end{align}
-Changing variables to $\xi=1-\theta$ and simplifying gives
-\begin{align}
-\pderiv[F]{t} + (z - 1) \pderiv[F]{z} = \frac{(z-1)\xi}{1-z\xi}\lambda F.
-\label{eq:1state_unreg_015}
-\end{align}
-\subsubsection{Stead-state}
-At steady-state, the PDE reduces to the ODE
-\begin{align}
-\deriv[F]{z} = \frac{\xi}{1-z\xi}\lambda F,
-\end{align}
-which we can integrate as
-\begin{align}
-\int \frac{dF}{F} = \int \frac{\lambda\xi dz}{1-\xi z}.
-\end{align}
-The initial conditions for generating functions can be subtle and confusing.
-The key fact follows from the definition
-$F(z,t) = \sum_m z^m p(m,t)$,
-normalization requires that
-$F(z=1^-,t) = \sum_m p(m,t) = 1$.\footnote{
-Sometimes the generating function may be undefined \textit{at} $z=1$ but the
-limit still holds. Also many authors change variables from $z$ to other
-things, so one must keep track of how the normalization condition transforms.
-}
-Doing the integrals (and producing constant $c$) gives
-\begin{align}
-\ln F &= -\lambda \ln(1-\xi z) + c
-\\
-F &= \frac{c}{(1-\xi z)^\lambda}.
-\end{align}
-Only one choice for $c$ can satisfy initial conditions, producing
-\begin{align}
-F(z) = \left(\frac{1-\xi}{1-\xi z}\right)^\lambda
-        = \left(\frac{\theta}{1 - z(1-\theta)}\right)^\lambda,
-\end{align}
-which is exactly the negative binomial's generating function, as expected.
-
-\subsection{Adding repression}
-\subsubsection{Deriving the generating function for mRNA distribution}
-
-Let us move from a one-state promoter to a two-state promoter, where one state
-has repressor bound and the other produces transcriptional bursts as above.
-A schematic of this model is shown in~\fig{fig3:kR_inferences}.
-Although now we have an equation for each promoter state, otherwise
-the master equation reads similarly to the one-state case, except with
-additional terms corresponding to transitions between promoter states, namely
-\begin{align}
-\deriv{t}p_R(m,t) =& k_R^+ p_A(m,t) - k_R^- p_R(m,t)
-        + (m+1)\gamma p_R(m+1,t) - m\gamma p_R(m,t)
-\\
-\begin{split}
-\deriv{t}p_A(m,t) =& - k_R^+ p_A(m,t) + k_R^- p_R(m,t)
-        + (m+1)\gamma p_A(m+1,t) - m\gamma p_A(m,t) 
-\\
-&- k_i p_A(m,t) + k_i \sum_{m^\prime=0}^m \theta(1-\theta)^{m-m^\prime} p_A(m^\prime,t),
-\end{split}
-\end{align}
-where $p_R(m,t)$ is the probability of the system having $m$ mRNA copies
-and having repressor bound to the promoter at time $t$, and $p_A$ is an
-analogous probability to find the promoter without repressor bound.
-$k_R+$ and $k_R^-$ are, respectively, the rates at which repressors
-bind and unbind to and from the promoter, and
-$\gamma$ is the mRNA degradation rate.
-$k_i$ is the rate at which bursts initiate, and as before,
-the geometric distribution of burst sizes has mean $b=(1-\theta)/\theta$.\footnote{
-$\theta$ can be viewed as a Bernoulli trial probability: after each transcript
-in a burst is produced, $\theta$ is the probability that the burst
-terminates and $(1-\theta)$ is the probability that the burst continues.
-}
-
-Interestingly, it turns out that this problem maps exactly onto the
-three-stage promoter model considered by
-Shahrezaei and Swain in~\cite{Shahrezaei2008}, with relabelings.
-Their approximate solution for protein distributions amounts to
-the same approximation we make here in regarding the duration
-of mRNA synthesis bursts as instantaneous, so their solution
-for protein distributions also solves our problem of mRNA distributions.
-Let us examine the analogy more closely.
-They consider a two-state promoter, as we do here, but they model
-mRNA as being produced one at a time and degraded, with rates
-$v_0$ and $d_0$.
-Then they model translation as occuring with rate $v_1$,
-and protein degradation with rate $d_1$.
-Now consider the limit where $v_1, d_0\rightarrow\infty$ with
-their ratio $v_1/d_0$ held constant.
-$v_1/d_0$ resembles the average burst size of translation from a single
-mRNA: these are the rates of two Poisson processes that compete over a
-transcript, which matches the story of geometrically distributed burst sizes.
-So after taking this limit, on timescales slow compared to $v_1$ and $d_0$,
-it appears that transcription events fire at rate $v_0$ and produce a
-geometrically distributed burst of translation of mean size $v_1/d_0$,
-which intuitively matches the story we have told above for mRNA
-with variables relabeled.
-
-To verify this intuitively conjectured mapping between our problem and
-the solution in~\cite{Shahrezaei2008}, we continue with a careful
-solution for the mRNA distribution using probability generating functions,
-following the ideas sketched in~\cite{Shahrezaei2008}.
-It is natural to nondimensionalize rates in the problem by $\gamma$,
-or equivalently, this amounts to measuring time in units of $\gamma^{-1}$.
-We are also only interested in steady state, so we set the time
-derivatives to zero, giving
-\begin{align}
-0 =& k_R^+ p_A(m) - k_R^- p_R(m) + (m+1) p_R(m+1) - m p_R(m)
-\\
-\begin{split}
-0 =& - k_R^+ p_A(m) + k_R^- p_R(m) + (m+1) p_A(m+1) - m p_A(m) 
-\\
-&- k_i p_A(m) + k_i \sum_{m^\prime=0}^m \theta(1-\theta)^{m-m^\prime} p_A(m^\prime),
-\end{split}
-\end{align}
-
-The probability generating function is defined as before in the
-constitutive case, except now we must introduce a generating function
-for each promoter state,
-\begin{align}
-f_A(z) = \sum_{m=0}^\infty z^m p_A(m),
-\;
-f_R(z) = \sum_{m=0}^\infty z^m p_R(m).
-\end{align}
-Our real objective is the generating function $f(z)$ that generates
-the mRNA distribution $p(m)$, independent of what state the promoter is in.
-But since $p(m) = p_A(m) + p_R(m)$, it follows too that $f(z) = f_A(z) + f_R(z)$.
-
-So as before we multiply both equations by $z^m$ and sum over all $m$.
-Each individual term transforms exactly as did an analogous term in
-the constitutive case, so the coupled ODEs for the generating functions read
-\begin{align}
-0 =& k_R^+ f_A(z) - k_R^- f_R(z) + \pderiv{z} f_R(z) - z \pderiv{z} f_R(z)
-\\
-\begin{split}
-0 =&  - k_R^+ f_A(z) + k_R^- f_R(z) + \pderiv{z} f_A(z) - z \pderiv{z} f_A(z)
-\\
-&- k_i f_A(z) + k_i \frac{\theta}{1-z(1-\theta)} f_A(z),
-\end{split}
-\end{align}
-and after changing variables $\xi = 1 - \theta$ as before
-and rearranging, we have
-\begin{align}
-0 &= k_R^+ f_A(z) - k_R^- f_R(z) + (1-z) \pderiv{z} f_R(z)
-\\
-0 &=  - k_R^+ f_A(z) + k_R^- f_R(z) + (1 - z) \pderiv{z} f_A(z)
-+ k_i \frac{(z-1)\xi}{1-z\xi} f_A(z),
-\end{align}
-We can transform this problem from two coupled first-order ODEs to a
-single second-order ODE by solving for $f_A$ in the first and plugging
-into the second, giving
-\begin{align}
-\begin{split}
-0 = (1&-z) \pderiv[f_R]{z}
-+ \frac{1-z}{k_R^+}
-        \left(k_R^- \pderiv[f_R]{z} + \pderiv[f_R]{z} +(z-1) \psecderiv[f_R]{z}\right)
-\\
-&+ \frac{k_i}{k_R^+} \frac{(z-1)\xi}{1-z\xi}
-        \left(k_R^- f_R + (z-1) \pderiv[f_R]{z}\right),
-\end{split}
-\end{align}
-where, to reduce notational clutter, we have dropped the explicit $z$
-dependence of $f_A$ and $f_R$. Simplifying we have
-\begin{align}
-0 = \psecderiv[f_R]{z}
-        - \left(\frac{k_i\xi}{1-z\xi}
-                + \frac{1 + k_R^- + k_R^+}{1-z}
-        \right)\pderiv[f_R]{z}
-        + \frac{k_i k_R^- \xi}{(1-z\xi)(1-z)}f_R.
-\end{align}
-This can be recognized as the hypergeometric differential equation,
-with singularities at $z=1$, $z=\xi^{-1}$, and $z=\infty$. The latter
-can be verified by a change of variables from $z$ to $x=1/z$,
-being careful with the chain rule, and noting that $z=\infty$ is a
-singular point if and only if $x=1/z=0$ is a singular point.
-
-The standard form of the hypergeometric differential equation has its
-singularities at 0, 1, and $\infty$, so to take advantage of the
-standard form solutions to this ODE, we first need to transform
-variables to put it into a standard form.
-However, this is subtle. While any such transformation should work in
-principle, the solutions are expressed most simply in the neighborhood
-of $z=0$, but the normalization condition that we need to enforce
-corresponds to $z=1$. The easiest path, therefore, is to find a change
-of variables that maps
-1 to 0, $\infty$ to $\infty$, and $\xi^{-1}$ to 1.
-This is most intuitively done in two steps.
-
-First map the $z=1$ singularity to 0 by the change of variables $v=z-1$, giving
-\begin{align}
-0 = \psecderiv[f_R]{v}
-        + \left(\frac{k_i\xi}{(1+v)\xi - 1}
-                + \frac{1 + k_R^- + k_R^+}{v}
-        \right)\pderiv[f_R]{v}
-        + \frac{k_i k_R^- \xi}{((1+v)\xi - 1)v}f_R.
-\end{align}
-Now two singularities are at $v=0$ and $v=\infty$.
-The third is determined by $(1+v)\xi -1 = 0$, or $v=\xi^{-1} - 1$.
-We want another variable change that maps this third singularity to 1
-(without moving 0 or infinity). Changing variables again to
-$w=\frac{v}{\xi^{-1} - 1} = \frac{\xi}{1-\xi} v$ fits the bill.
-In other words, the combined change of variables
-\begin{align}
-w = \frac{\xi}{1-\xi} (z-1)
-\end{align}
-maps $z = \{1, \xi^{-1}, \infty\}$ to $w =\{0, 1, \infty\}$ as desired.
-Plugging in, being mindful of the chain rule and noting
-$(1 + v)\xi - 1 = (1 - \xi)(w - 1)$ gives
-\begin{align}
-0 = \left(\frac{\xi}{1-\xi}\right)^2 \psecderiv[f_R]{w}
-+ \left(
-        \frac{\xi k_i}{(1-\xi)(w-1)} + \frac{\xi(1 + k_R^- + k_R^+)}{(1-\xi)w}
-\right) \frac{\xi}{1-\xi} \pderiv[f_R]{w}
-+ \frac{k_i k_R^- \xi^2}{(1-\xi)^2 w(w-1)}f_R.
-\end{align}
-This is close to the standard form of the hypergeometric differential
-equation, and some cancellation and rearrangement gives
-\begin{align}
-0 = w(w-1)\psecderiv[f_R]{w}
-+ \left(k_i w + (1 + k_R^- + k_R^+)(w-1)\right) \pderiv[f_R]{w}
-+ k_i k_R^- f_R.
-\end{align}
-and a little more algebra produces
-\begin{align}
-0 = w(1-w)\psecderiv[f_R]{w}
-+ \left(1 + k_R^- + k_R^+
-        - (1 + k_i + k_R^- + k_R^+)w
-\right) \pderiv[f_R]{w}
-- k_i k_R^- f_R,
-\end{align}
-which is the standard form. From this we can read off the solution
-in terms of hypergeometric functions ${_2F_1}$ from
-any standard source, e.g.~\cite{Abramowitz1964}, and identify the
-conventional parameters in terms of our model parameters.
-We want the general solution in the neighborhood of $w=0$ ($z=1$),
-which for a homogeneous linear second order ODE must be a sum of two
-linearly independent solutions.
-More precisely,
-\begin{align}
-f_R(w) = C_1 {_2F_1}(\alpha, \beta, \gamma; w)
-+ C_2 w^{1-\gamma}{_2F_1}(1+\alpha-\gamma, 1+\beta-\gamma, 2-\gamma; w)
-\end{align}
-with parameters determined by
-\begin{align}
-\begin{split}
-\alpha\beta &= k_i k_R^-
-\\
-1+\alpha+\beta &= 1+k_i+k_R^-+k_R^+
-\\
-\gamma &= 1 + k_R^- + k_R^+
-\end{split}
-\end{align}
-and constants $C_1$ and $C_2$ to be set by boundary conditions.
-Solving for $\alpha$ and $\beta$, we find\footnote{
-Note that $\alpha$ and $\beta$ are interchangeable in the definition of
-${_2F_1}$ and differ only in the sign preceeding the radical.
-}
-\begin{align}
-\begin{split}
-\alpha &= \frac{1}{2}
-\left(k_i+k_R^-+k_R^+ + \sqrt{(k_i+k_R^-+k_R^+)^2 - 4k_i k_R^-}\right)
-\\
-\beta &= \frac{1}{2}
-\left(k_i+k_R^-+k_R^+ - \sqrt{(k_i+k_R^-+k_R^+)^2 - 4k_i k_R^-}\right)
-\\
-\gamma &= 1 + k_R^- + k_R^+.
-\end{split}
-\end{align}
-Since the normalization condition requires that $f_R$ be finite at $w=0$,
-we can immediately set $C_2=0$ to discard the second solution.
-This is because all the rate constants are strictly positive,
-so $\gamma>1$ and therefore $w^{1-\gamma}$ blows up as $w\rightarrow0$.
-Now that we have $f_R$, we would like to find the generating function
-for the mRNA distribution, $f(z) = f_A(z) + f_R(z)$.
-We can recover $f_A$ from our solution for $f_R$, namely
-\begin{align}
-f_A(z) = \frac{1}{k_R^+}\left(k_R^- f_R(z) + (z-1) \pderiv[f_R]{z}\right)
-\end{align}
-or
-\begin{align}
-f_A(w) = \frac{1}{k_R^+}\left(k_R^- f_R(w) + w \pderiv[f_R]{w}\right),
-\end{align}
-where in the second line we transformed our original relation between
-$f_R$ and $f_A$ to our new, more convenient, variable $w$.
-Plugging our solution for $f_R(w) = C_1{_2F_1}(\alpha, \beta, \gamma; w)$
-into $f_A$, we will require the differentiation rule for ${_2F_1}$,
-which tells us
-\begin{align}
-\pderiv[f_R]{w} = C_1\frac{\alpha\beta}{\gamma}
-                {_2F_1}(\alpha+1, \beta+1, \gamma+1; w),
-\end{align}
-from which it follows that
-\begin{align}
-f_A(w) = \frac{C_1}{k_R^+}
-\left(
-k_R^- {_2F_1}(\alpha, \beta, \gamma; w)
-+ w\frac{\alpha\beta}{\gamma} {_2F_1}(\alpha+1, \beta+1, \gamma+1; w)
-\right)
-\end{align}
-and therefore
-\begin{align}
-f(w) = C_1\left(1 + \frac{k_R^-}{k_R^+}\right)
-        {_2F_1}(\alpha, \beta, \gamma; w)
-+ w \frac{C_1}{k_R^+} \frac{\alpha\beta}{\gamma}
-        {_2F_1}(\alpha+1, \beta+1, \gamma+1; w).
-\end{align}
-To proceed, we need one of the (many) useful identities known for
-hypergeometric functions, in particular
-\begin{align}
-w\frac{\alpha\beta}{\gamma} {_2F_1}(\alpha+1, \beta+1, \gamma+1; w)
-=
-(\gamma-1)\left(
-{_2F_1}(\alpha, \beta, \gamma-1; w) - {_2F_1}(\alpha, \beta, \gamma; w)
-\right).
-\end{align}
-Substituting this for the second term in $f(w)$, we find
-\begin{align}
-f(w) = \frac{C_1}{k_R^+}
-\left[
-        \left(k_R^+ + k_R^-\right)
-        {_2F_1}(\alpha, \beta, \gamma; w)
-+ (\gamma-1)\left(
-        {_2F_1}(\alpha, \beta, \gamma-1; w) - {_2F_1}(\alpha, \beta, \gamma; w)
-        \right)
-\right],
-\end{align}
-and since $\gamma-1 = k_R^+ + k_R^-$, the first and third terms cancel,
-leaving only
-\begin{align}
-f(w) = C_1\frac{k_R^+ + k_R^-}{k_R^+} {_2F_1}(\alpha, \beta, \gamma-1; w).
-\end{align}
-Now we enforce normalization, demanding $f(w=0) = f(z=1) = 1$.
-${_2F_1}(\alpha, \beta, \gamma-1; 0) = 1$, so we must have
-$C_1 = k_R^+ / (k_R^+ + k_R^-)$ and consequently
-\begin{align}
-f(w) =  {_2F_1}(\alpha, \beta, k_R^+ + k_R^-; w).
-\end{align}
-Recalling that the mean burst size $b = (1-\theta)/\theta = \xi/(1-\xi)$
-and $w = \frac{\xi}{1-\xi} (z-1) = b (z-1)$,
-we can transform back to the original variable $z$ to find the tidy result
-\begin{align}
-f(z) =  {_2F_1}(\alpha, \beta, k_R^+ + k_R^-; b(z-1)),
-\end{align}
-with $\alpha$ and $\beta$ given above by
-\begin{align}
-\begin{split}
-\alpha &= \frac{1}{2}
-\left(k_i+k_R^-+k_R^+ + \sqrt{(k_i+k_R^-+k_R^+)^2 - 4k_i k_R^-}\right)
-\\
-\beta &= \frac{1}{2}
-\left(k_i+k_R^-+k_R^+ - \sqrt{(k_i+k_R^-+k_R^+)^2 - 4k_i k_R^-}\right).
-\end{split}
-\end{align}
-Finally we are in sight of the original goal. We can generate the
-steady-state probability distribution of interest by
-differentiating the generating function,
-\begin{align}
-p(m) = m! \left.\frac{\partial^m}{\partial z^m} f(z) \right|_{z=0},
-\end{align}
-which follows easily from its definition. Some contemplation reveals
-that repeated application of the derivative rule used above will produce
-products of the form $\alpha(\alpha+1)(\alpha+2)\cdots(\alpha+m-1)$ in
-the expression for $p(m)$ and similarly for $\beta$ and $\gamma$. These
-resemble ratios of factorials, but since $\alpha$, $\beta$, and $\gamma$
-are not necessarily integer, we should express the ratios using gamma
-functions instead. More precisely, one finds
-\begin{align}
-p(m) = \frac{
-        \Gamma(\alpha + m)\Gamma(\beta + m)\Gamma(k_R^+ + k_R^-)
-        }
-        {
-        \Gamma(\alpha)\Gamma(\beta)\Gamma(k_R^+ + k_R^- + m)
-        }
-\frac{b^m}{m!}{_2F_1}(\alpha+m, \beta+m, k_R^++k_R^-+m; -b)
-\label{eq:p_m_bursty+rep}
-\end{align}
-which is finally the probability distribution we sought to derive.
-
-\subsubsection{Moments}
-We can verify the correctness of~\eq{eq:p_m_bursty+rep}
-by comparison with Gillespie simulation.
-We can also easily compute moments of the distribution from
-the generating function, and check these against raw moments computed
-directly from the master equation.
-
-The $j$-th raw moment of the distribution can be computed from the
-generating function from the formula
-\begin{align}
-\langle m^j \rangle
-= \left. \left(z \frac{\partial}{\partial z}\right)^j f(z)\right|_{z=1}.
-\end{align}
-Note that \textit{probabilities} are generated by differentiating
-and evaluating at $z=0$, while \textit{moments} are generated by
-differentiating and evaluating at $z=1$. This also aligns with the
-normalization constraint, which can be thought of as the zeroth raw moment.
-
-The mean mRNA is given by
-\begin{align}
-\langle m \rangle = \left.z \pderiv[f]{z} \right|_{z=1}
-= \frac{\alpha\beta b}{k_R^+ + k_R^-}
-= k_i b\frac{k_R^-}{k_R^+ + k_R^-},
-\end{align}
-which is intuitive and appealingly simple:
-the mean mRNA is just the mean mRNA in the absense of repression
-($k_i b$, the burst rate time mean burst size)
-times the fraction of time the repressor is bound, $k_R^-/(k_R^+ + k_R^-)$.
-
-To compute the variance, let us first compute
-\begin{align}
-\begin{split}
-\langle m^2\rangle
-&= \left .z\pderiv{z}\left(z\pderiv[f]{z}\right) \right|_{z=1}
-= \left. z\left(\pderiv[f]{z} + z\psecderiv[f]{z}\right)\right|_{z=1}
-\\
-&= \frac{\alpha\beta b}{k_R^+ + k_R^-}
-        + \frac{\alpha(\alpha+1)\beta(\beta+1) b^2}
-                {(k_R^+ + k_R^-)(k_R^+ + k_R^- + 1)}
-\\
-&= \frac{k_i b k_R^-}{k_R^+ + k_R^-}
-        + \frac{k_i k_R^- b^2 (1 + k_i + k_R^+ + k_R^- + k_i k_R^-)}
-                {(k_R^+ + k_R^-)(k_R^+ + k_R^- + 1)},
-\end{split}
-\end{align}
-from which the variance is
-\begin{align}
-\begin{split}
-var(m) = \langle m^2\rangle - \langle m\rangle^2
-&= \frac{k_i b k_R^-}{k_R^+ + k_R^-}
-\left(
-1 + \frac{b(1 + k_i + k_R^+ + k_R^- + k_i k_R^-)} {k_R^+ + k_R^- + 1}
-- \frac{k_i b k_R^-}{k_R^+ + k_R^-}
-\right)
-\\
-&= k_i b\frac{k_R^-}{k_R^+ + k_R^-}
-\left(
-1 + b + k_i b\frac{1 + k_R^-} {k_R^+ + k_R^- + 1}
-- \frac{k_i b k_R^-}{k_R^+ + k_R^-}
-\right)
-\\
-&= k_i b\frac{k_R^-}{k_R^+ + k_R^-}
-\left(
-1 + b + k_i b\frac{k_R^+}{(k_R^+ + k_R^-)(k_R^+ + k_R^- + 1)}
-\right).
-\end{split}
-\end{align}
-The Fano factor is the quantity in parentheses. This can be interpreted as
-the variability arising from the negative binomial distribution of the
-unregulated promoter $(1+b)$ plus additional noise arising from repression.
-
-\mmnote{Still to writeup: moment derivation direct from CME to compare}
-
-\subsection{Numerical considerations and recursion formulas}
-\subsubsection{Generalities}
-We would like to carry out Bayesian parameter inference on FISH data
-from~\cite{Jones2014}, using~\eq{eq:p_m_bursty+rep} as our
-likelihood. This requires accurate (and preferably fast)
-numerical evaluation of the hypergeometric function ${_2F_1}$,
-which is a notoriously hard problem~\cite{Pearson2017, Gil2007},
-and our particular needs here present an especial challenge as we show below.
-
-The hypergeometric function is defined by its Taylor series as
-\begin{align}
-{_2F_1}(a,b,c;z) 
-= \sum_{l=0}^\infty
-\frac{\Gamma(a + l)\Gamma(b + l)\Gamma(c)}
-        {\Gamma(a)\Gamma(b)\Gamma(c + l)}
-\frac{z^l}{l!}
-\end{align}
-for $|z|<1$, and by analytic continuation elsewhere.
-If $z\lesssim1/2$ and $\alpha$ and $\beta$ are not too large
-(absolute value below 20 or 30),
-then the series converges quickly and an accurate numerical representation is
-easily computed by truncating the series after a reasonable number of terms.
-Unfortunately, we need to evaluate ${_2F_1}$ over mRNA copy numbers fully out
-to the tail of the distribution, which can easily reach 50, possibly 100.
-From~\eq{eq:p_m_bursty+rep}, this means evaluating ${_2F_1}$
-repeatedly for values of $a$, $b$, and $c$ spanning the full range
-from $\mathcal{O}(1)$ to $\mathcal{O}(10^2)$,
-even if $\alpha$, $\beta$, and $\gamma$
-in~\eq{eq:p_m_bursty+rep} are small,
-with the situation even worse if they are not small.
-A naive numerical evaluation of the series definition will be
-prone to overflow and, if any of $a,b,c<0$, then some successive terms in the
-series have alternating signs which can lead to catastrophic cancellations.
-
-One solution is to evaluate ${_2F_1}$ using arbitrary precision arithmetic
-instead of floating point arithmetic,
-e.g., using the \texttt{mpmath} library in Python.
-This is accurate but incredibly slow computationally.
-To quantify how slow, we found that
-evaluating the likelihood defined by~\eq{eq:p_m_bursty+rep} $\sim50$ times
-(for a typical dataset of interest from~\cite{Jones2014},
-with $m$ values spanning 0 to $\sim50$)
-using arbitrary precision arithmetic is 100-1000 fold slower than
-evaluating a negative binomial likelihood for the corresponding
-constitutive promoter dataset.
-
-To claw back $\gtrsim30$ fold of that slowdown, we can exploit
-one of the many catalogued symmetries involving ${_2F_1}$.
-The solution involves recursion relations originally explored by Gauss,
-and studied extensively in~\cite{Pearson2017, Gil2007}.
-They are sometimes known as contiguous relations and relate the values
-of any set of 3 hypergeometric functions whose arguments differ by integers.
-To rephrase this symbolically, consider a set of hypergeometric functions
-indexed by an integer $n$,
-\begin{align}
-f_n = {_2F_1}(a+\epsilon_i n, b+\epsilon_j n, c+\epsilon_k n; z),
-\end{align}
-for a fixed choice of $\epsilon_i, \epsilon_j, \epsilon_k \in \{0,\pm 1\}$
-(at least one of $\epsilon_i, \epsilon_j, \epsilon_k$ must be nonzero,
-else the set of $f_n$ would contain only a single element).
-Then there exist known recurrence relations of the form
-\begin{align}
-A_n f_{n-1} + B_n f_{n} + C_n f_{n+1} = 0,
-\end{align}
-where $A_n, B_n$, and $C_n$ are some functions of $a,b,c$, and $z$.
-In other words, for fixed $\epsilon_i, \epsilon_j, \epsilon_k, a, b,$ and $c$,
-if we can merely evaluate ${_2F_1}$ twice, say for $n^\prime$ and $n^\prime-1$,
-then we can easily and rapidly generate values for arbitrary $n$.
-
-This provides a convenient solution for our problem: we need repeated
-evaluations of ${_2F_1}(a+m, b+m, c+m; z)$
-for fixed $a,b$, and $c$ and many integer values of $m$.
-They idea is that we can use arbitrary precision arithmetic to evaluate
-${_2F_1}$ for just two particular values of $m$ and then generate
-${_2F_1}$ for the other 50-100 values of $m$ using the recurrence
-relation.\footnote{
-There are even more sophisticated ways of utilizing the recurrence
-relations that might have netted another factor of 2 speed-up, and
-possibly as much as a factor of 10, but the method described here had
-already reduced the computation time to an acceptable
-$\mathcal{O}(\text{1 min})$, so these more sophisticated approaches did
-not seem worth the time to pursue.
-}
-
-However, there two further wrinkles.
-The first is that
-a naive application of the recurrence relation is numerically unstable.
-Roughly, this is because the three term recurrence relations,
-like second order ODEs, admit two linearly independent solutions.
-In a certain eigenbasis, one of these solutions dominates the other
-as $n\rightarrow\infty$, and as $n\rightarrow-\infty$,
-the dominance is reversed.
-If we fail to work in this eigenbasis, our solution of the recurrence relation
-will be a mixture of these solutions and rapidly accumulate numerical error.
-For our purposes, it suffices to know that the authors of~\cite{Gil2007}
-derived the numerically stable solutions (so-called \textit{minimal solutions})
-for several possible choices of $\epsilon_i, \epsilon_j, \epsilon_k$.
-Running the recurrence in the proper direction using a minimal solution
-is numerically robust and can be done entirely in floating point arithmetic, 
-so that we only need to evaluate ${_2F_1}$ with arbitrary precision arithmetic
-to generate the seed values for the recursion.
-
-The second wrinkle is a corollary to the first.
-The minimal solutions are only minimal for certain ranges of the argument $z$,
-and not all of the 26 possible recurrence relations
-have minimal solutions for all $z$.
-This can be solved by using one of the many transformation formulae for
-${_2F_1}$ to convert to a different recurrence relation that has
-a minimal solution over the required domain of $z$, although
-this can require some trial and error to find the right transformation,
-the right recurrence relation, and the right minimal solution.
-
-\subsubsection{Particulars}
-Let us now demonstrate these generalities for our problem of interest.
-In order to evaluate the probability distribution of our
-model,~\eq{eq:p_m_bursty+rep}, we need to evaluate hypergeometric functions
-of the form ${_2F_1}(\alpha+m, \beta+m, \gamma+m; -b)$
-for values of $m$ ranging from $0$ to $\mathcal{O}(100)$.
-The authors of~\cite{Gil2007} did not derive a recursion relation
-for precisely this case. We could follow their methods and do so ourselves,
-but it is much easier to convert to a case that they did consider.
-The strategy is to look through the minimal solutions tabulated
-in~\cite{Gil2007} and search for a transformation we could apply to
-${_2F_1}(\alpha+m, \beta+m, \gamma+m; -b)$ that would place the $m$'s
-(the variable being incremented by the recursion)
-in the same arguments of ${_2F_1}$ as the minimal solution.
-After some ``guess and check,'' we found that the transformation
-\begin{align}
-{_2F_1}(\alpha+m, \beta+m, \gamma+m; -b)
-=
-(1+b)^{-\alpha-m}
-        {_2F_1}\left(\alpha+m, \gamma-\beta, \gamma+m; \frac{b}{1+b}\right),
-\label{eq:rec_euler_pretransform}
-\end{align}
-produces a ${_2F_1}$ on the right hand side that closely resembles
-the minimal solutions $y_{3,m}$ and $y_{4,m}$ in Eq.~4.3 in~\cite{Gil2007}.
-Explicitly, these solutions are
-\begin{align}
-y_{3,m}
-&\propto
-{_2F_1}\left(-\alpha^\prime + \gamma^\prime - m,
-                -\beta^\prime + \gamma^\prime,
-                1-\alpha^\prime-\beta^\prime+\gamma^\prime-m;
-                1-z\right)
-\\
-y_{4,m}
-&\propto
-{_2F_1}\left(\alpha^\prime + m,
-                \beta^\prime,
-                1+\alpha^\prime+\beta^\prime-\gamma^\prime+m;
-                1-z\right),
-\label{eq:minimal_soln_sans_prefac}
-\end{align}
-where we have omitted prefactors which are unimportant for now.
-Which of these two we should use depends on what values $z$ takes on.
-Equating $1-z=b/(1+b)$ gives $z=1/(1+b)$, and since $b$ is strictly positive,
-$z$ is bounded between 0 and 1.
-From Eq.~4.5 in~\cite{Gil2007}, $y_{4,m}$ is the minimal solution
-for real $z$ satisfying $0<z<2$, so this is the only minimal solution we need.
-
-Now that we have our minimal solution,
-what recurrence relation does it satisfy?
-Confusingly, the recurrence relation of which $y_{4,m}$ is a solution
-increments different arguments of ${_2F_1}$ that does $y_{4,m}$:
-it increments the first only, rather than first and third.
-This recurrence relation can be looked up, e.g., Eq.~15.2.10
-in~\cite{Abramowitz1964}, which is
-\begin{align}
-(\gamma^\prime - (\alpha^\prime + m)) f_{m-1}
-+
-(2(\alpha^\prime+m) - \gamma^\prime + (\beta^\prime - \alpha^\prime)z)f_m
-+ \alpha^\prime(z-1) f_{m+1} = 0.
-\label{eq:chosen_rec_rel}
-\end{align}
-Now we must solve for the parameters appearing in the recurrence relation
-in terms of our parameters, namely by setting
-\begin{align}
-\begin{split}
-\alpha^\prime &= \alpha
-\\
-\beta^\prime &= \gamma - \beta
-\\
-1 + \alpha^\prime + \beta^\prime - \gamma^\prime &= \gamma
-\\
-1 - z &= \frac{b}{1+b}
-\end{split}
-\end{align}
-and solving to find
-\begin{align}
-\begin{split}
-\alpha^\prime &= \alpha
-\\
-\beta^\prime &= \gamma - \beta
-\\
-\gamma^\prime &= 1 + \alpha - \beta
-\\
-z &= \frac{1}{1+b}
-.
-\end{split}
-\end{align}
-Finally we have everything we need. The minimal solution
-\begin{align}
-y_{4,m}
-=
-\frac{\Gamma(1+\alpha^\prime-\gamma^\prime+m)}
-        {\Gamma(1+\alpha^\prime+\beta^\prime-\gamma^\prime+m)}
-\times
-{_2F_1}\left(\alpha^\prime + m,
-                \beta^\prime,
-                1+\alpha^\prime+\beta^\prime-\gamma^\prime+m;
-                1-z\right),
-\end{align}
-where we have now included the necessary prefactors,
-is a numerically stable solution of the recurrence
-relation~\eq{eq:chosen_rec_rel} if the recursion is run
-from large $m$ to small $m$.
-So the procedure is to compute the value of ${_2F_1}$ for the two
-largest $m$ values of interest using arbitrary precision arithmetic,
-compute the prefactors to construct
-$y_{4,\text{max}(m)}$ and $y_{4,\text{max}(m)-1}$,
-recursively compute $y_{4,m}$ for all $m$ less than $\text{max}(m)$ down
-to $m=0$, then cancel off the prefactors of the resulting values of
-$y_{4,m}$ for all $m$ to produce ${_2F_1}$ for all desired $m$ values.
-\mmnote{Maybe I should reformat this paragraph into
-a pseudocode version of the algorithm?
-Instead of, or in addition to?}
-
-With ${_2F_1}$ computed, the only remaining numerical danger in computing
-$p(m)$ in~\eq{eq:p_m_bursty+rep} is overflow of the gamma functions.
-This is easily solved by taking the log of the entire expression
-and using standard routines to compute the log of the gamma functions,
-then exponentiating the entire expression at the end if $p(m)$
-is needed rather than $\log p(m)$.
-
-% \bibliographystyle{nature}
-\bibliographystyle{abbrv}
-\bibliography{./simple_rep}
-
-\end{document}

### 2020-05-19 doc/section_05_discussion.tex
initial splitting of document into several files

--- 2020-05-17 doc/paper.tex
@@
-\documentclass[12pt]{article}%{amsart}
-\usepackage[top = 1.0in, bottom = 1.0in, left = 1.0in, right = 1.0in]{geometry}
-% See geometry.pdf to learn the layout options. There are lots.
-%\geometry{letterpaper}               % ... or a4paper or a5paper or ...
-%\geometry{landscape}             % Activate for for rotated page geometry
-\usepackage[parfill]{parskip}    % Activate to begin paragraphs with an empty line rather than an indent
-\usepackage{graphicx}
-\usepackage{amssymb}
-%\usepackage{epstopdf}
-% \usepackage{caption}
-\usepackage{amsmath}
-% \usepackage{longtable}
-% \usepackage{tabu}
-% \usepackage{accents} %to get undertildes for vec & mat
-\usepackage{color} % for colored text
-% \usepackage[normalem]{ulem} % for editing: striking out text using 'sout' command
-
-% \newcommand\mytablefigwidth{0.35\textwidth}
-
-% allows to use .ai files directly w/o resaving as pdf
-\DeclareGraphicsRule{.ai}{pdf}{.ai}{}
-
-% Handy math macros!
-\newcommand{\vect}[1]{\vec{#1}}
-\newcommand{\matr}[1]{\mathbf{#1}}
-\newcommand{\rate}[3]{{#1}_{#2}^{#3}}
-\newcommand{\mmnote}[1]{\textcolor{cyan}{(MM:~#1)}}
-
-% derivative macros. these are sneaky. the [{}] is an empty default 1st arg
-% usage: provide 1 arg: \deriv{x} gives d/dx
-% provide 2 args like so: \deriv[f]{x} gives df/dx
-\newcommand{\deriv}[2][{}]{\frac{d #1}{d #2}}
-\newcommand{\pderiv}[2][{}]{\frac{\partial #1}{\partial #2}}
-\newcommand{\psecderiv}[2][{}]{\frac{\partial{^2} #1}{\partial #2{^2}}}
-
-%%%%% Referencing macros %%%%%
-\newcommand{\fig}[1]{Figure~\ref{#1}}
-\newcommand{\tref}[1]{Table~\ref{#1}}
-\newcommand{\eq}[1]{Eq.~(\ref{#1})}
-\newcommand{\eqrange}[2]{Eqs.~(\ref{#1}-\ref{#2})}
-
-%%%%%
-\begin{document}
-
-\title{Simple repression draft}
-
-\maketitle
-
-\section{Introduction}
-\mmnote{Key ideas, no particular order, that I haven't written down before:
-\begin{itemize}
-\item Our goal is to build phenomenological models of input-output
-functions of genetic circuits. More precisely, we want phenomenology
-in the sense of coarse-graining away atomistic detail, but still
-retaining biophysical meaning, e.g., we don't want to coarse-grain
-as far as Hill functions. Why not? We are motivated by studies
-like~\cite{Razo-Mejia2020}, for which Hill functions are
-insufficient. Such work requires a quantitative theory of how
-biophysical changes at the molecular level propagate to
-input-output functions at the genetic circuit level.
-We want concepts, not mere facts.
-In particular a key question is: at this level of
-coarse graining, what microscopic details do I need to explicitly
-model, and how do we figure that out? For example, do I need to
-worry about all or even any of the steps that individual RNAPs go
-through each time they make a transcript? Turning the question
-around, can we see any imprint of those processes in the
-available data? If the answer is no, then those processes are
-irrelevant for our purposes. Forward modeling and inverse
-(statistical inferential) modeling can complement each other
-beautifully here.
-\end{itemize}}
-
-\fig{fig1:means_cartoons}A shows the qualitative picture of
-simple repression. An operator, the binding site on the DNA for a
-repressor protein, may be found occupied by a repressor, in which
-case transcription is blocked from occurring. Or it may be found
-unoccupied, in which case RNA polymerase (RNAP) may bind and
-transcription can proceed.
-
-The key assumption we make is that binding of repressor and RNAP in the promoter region of interest is exclusive, meaning that one or the other may bind, but never may both be simultaneously bound.
-It is often imagined that when the repressor is bound to its operator,
-RNAP is sterically blocked from binding to its promoter sequence.
-Current evidence suggests this is sometimes, but not
-always the case~\mmnote{Cite that fig, from ??Nathan or Bill or
-Rob??, showing locations of repressor binding sites far upstream
-of promoter}, and it remains an interesting open question precisely
-how a repressor bound far upstream is able to repress transription.
-Suggestions include ``action-at-a-distance'' mediated by kinks in
-the DNA, formed when the repressor is bound, that prevent RNAP binding.
-Nevertheless, our modeling in this work is sufficiently
-coarse-grained that we simply assume exclusive binding and need
-not worry about such precise details.
-
-\mmnote{Add a notation aside somewhere that all association rates are written as zeroth-order rates.
-In other words, the concentration of the molecule is hidden inside, so for instance, $k_R^+$ is proportional to repressor copy number.}
-
-\section{Means}
-\subsection{Fold-changes are indistinguishable across models}
-As a first stop on our search for the ``right'' model of simple
-repression, let us consider what we can learn from theory and
-measurements of fold-change in mean expression. The fold-change is defined as
-\begin{equation}
-\text{fold-change}
-= \frac{\langle \text{gene expression with repressor present} \rangle}
-        {\langle \text{gene expression with repressor absent} \rangle}
-= \frac{\langle m\rangle(R\ne 0)}{\langle m\rangle(R = 0)}
-\label{eq:fc_def}
-\end{equation}
-where angle brackets denote the averge over a population of cells
-and mean mRNA $\langle m\rangle$ is viewed as a function of
-repressor copy number $R$.
-The second
-equality in~\ref{eq:fc_def} follows from assuming that the
-translation efficiency, i.e., the number of proteins translated
-per mRNA, is the same in both conditions.
-In other words, we assume that mean protein level is proportional
-to mean mRNA level, and that the proportionality constant is the
-same in both conditions and therefore cancels out in the ratio.
-This is reasonable since the cells in the two conditions are
-identical except for the presence of the transcription factor,
-and it is difficult to imagine the transcription factor having
-any direct effect on translation.
-
-Fold-change has proven a very convenient observable in past
-work~\cite{Garcia2011a, Razo-Mejia2018, Chure2019}.
-Part of its utility in dissecting transcriptional regulation is
-its ratiometric nature, which is always easier than making an
-absolute measurement. Also, by measuring otherwise identical cells
-with and without a transcription factor present, any biological
-noise common to both conditions can be made to cancel away.
-In addition, the presence of different operator sites is known to
-change the basal expression level in the absence of
-repressor~\mmnote{look up cite}.
-Computing fold-changes neatly cancels this effect as well.
-
-\fig{fig1:means_cartoons} depicts a smorgasbord
-of mathematized cartoons for simple repression that have appeared
-in previous literature. For each cartoon, we calculate the
-fold-change in mean gene expression as predicted by that model,
-deferring some algebraic details to the appendix.
-What we will find is that all cartoons collapse to a single
-master curve, shown in \fig{fig1:means_cartoons}c, which contains
-just two parameters. We label the parameters $\Delta F_R$, an
-effective free energy parametrizing the repressor-DNA
-interaction, and $\rho$, which subsumes all details of
-transcription in the absence of repressors.
-We will offer some intuition for why this master curve exists and discuss
-why it is likely impossible to discriminate ``right'' from ``wrong''
-cartoons given only measurements of fold-changes in expression.
-
-\subsubsection{Equilibrium, 2-state}
-In this simplest model, the promoter is idealized as existing in
-one of two states, either repressor bound or repressor unbound.
-The rate of transcription, and therefore the level of gene
-expression, is assumed to be proportional to the fraction of time
-spent in the repressor unbound state.
-Mathematically, this amounts to the statement that the mean mRNA level is
-given by
-\begin{equation}
-\langle m \rangle = \frac{r}{\gamma}
-        \left(1 + \frac{R}{N_{NS}} e^{-\beta\Delta\epsilon_R}\right)^{-1},
-\end{equation}
-where $r$ is the transcription rate from the repressor unbound
-state, $\gamma$ is the mRNA degradation rate, $R$ is repressor
-copy number, $N_{NS}$ is the number of nonspecific binding sites
-in the genome (where repressors spend most of their time when not
-bound to the operator, their specifc binding site),
-$\beta=1/k_BT$, and $\Delta\epsilon_R$ is the binding energy of a
-repressor to its operator site. The derivation of this result is deferred
-to the appendix.
-
-The fold-change is easily found as the ratio of mean mRNA with
-and without repressor. This gives
-\begin{equation}
-\text{fold-change}
-= \left(1 + \frac{R}{N_{NS}} e^{-\beta\Delta\epsilon_R}\right)^{-1},
-\end{equation}
-which clearly matches the form of the master curve
-in~\eq{fig1:means_cartoons} with $\rho=1$.
-
-In fact it was noted in~\cite{Chure2019} that this two-state
-model can be viewed as the coarse-graining of any equilibrium
-promoter model in which no transcriptionally active states have
-transcription factor bound. Put differently, there is no overlap
-between TF bound states and transcriptionally active states. We
-will see this explicitly in the 3-state equilibrium model below,
-but perhaps surprising is that something analogous carries over
-even to the nonequilibrium models we consider later.
-
-\subsubsection{Equilibrium, 3-state}
-Compared to the previous model, here we fine-grain the repressor
-unbound state into two separate states: empty, and RNAP bound.
-This picture was used in~\cite{Garcia2011a} as we use it here,
-and in~\cite{Razo-Mejia2018} and~\cite{Chure2019} it was generalized
-to incorporate small-molecule inducers that bind the repressor.
-Interestingly, the effect of this is simply to renormalize $R$
-from the total number of repressors to a smaller effective number,
-roughly (though not exactly) the number which are unbound by inducers.
-
-The mean mRNA, as derived in the appendix from a similar
-enumeration of states and weights as the previous model, is
-\begin{equation}
-\langle m \rangle = \frac{r}{\gamma}
-\frac{\frac{P}{N_{NS}} e^{-\beta\Delta\epsilon_P}}
-        {
-        1 + \frac{R}{N_{NS}} e^{-\beta\Delta\epsilon_R}
-        + \frac{P}{N_{NS}} e^{-\beta\Delta\epsilon_P}
-        },
-\end{equation}
-where the new variables are $\Delta\epsilon_P$, the difference in
-RNAP binding energy to its specific site (the promoter) relative
-to an average nonspecific background site, and the RNAP copy
-number, $P$. The fold-change again follows immediately as
-\begin{align}
-\text{fold-change}
-&= \frac{\frac{P}{N_{NS}} e^{-\beta\Delta\epsilon_P}}
-        {
-        1 + \frac{R}{N_{NS}} e^{-\beta\Delta\epsilon_R}
-        + \frac{P}{N_{NS}} e^{-\beta\Delta\epsilon_P}
-        }
-\frac{1 + \frac{P}{N_{NS}} e^{-\beta\Delta\epsilon_P}}
-        {\frac{P}{N_{NS}} e^{-\beta\Delta\epsilon_P}}
-\\
-&= \left(
-1 + \frac{\frac{R}{N_{NS}} e^{-\beta\Delta\epsilon_R}}
-        {1 + \frac{P}{N_{NS}} e^{-\beta\Delta\epsilon_P}}
-\right)^{-1}
-\\
-&= (1 + \exp(-\Delta F_R - \log\rho))^{-1},
-\end{align}
-with $\Delta F_R = \beta\Delta\epsilon_R - \log(R/N_{NS})$
-and $\rho = 1 + \frac{P}{N_{NS}}\mathrm{e}^{-\beta\Delta\varepsilon_P}$
-as shown in~\fig{fig1:means_cartoons}.
-
-\begin{figure}%[h!]
-\centering
-\includegraphics[width=\textwidth]{../figures/fig1/fig1.ai}
-\caption{
-\textbf{
-An overview of the simple repression motif at the level of means.}
-(A) depicts the qualitative biological picture of simple repression.
-(B) shows a variety of possible mathematized cartoons of simple
-repression, along with the effective parameter $\rho$
-which subsumes all regulatory details
-of the architecture excluding the repressor.
-(C) shows the ``master curve'' to which all cartoons in (B) collapse.
-}
-\label{fig1:means_cartoons}
-\end{figure}
-
-\subsubsection{Nonequilibrium model 1 - Poisson promoter}
-For our first kinetic model, we consider the simplest possible
-picture with only two states, repressor bound and unbound.
-This is exactly the model used for the main results of~\cite{Jones2014}.
-In this picture, repressor association and dissociation rates
-from its operator site, $k_R^+$ and $k_R^-$, respectively, govern
-transitions between the two states. When the system is in the
-unbound state, transcriptions initiate at rate $r$, which
-represents a coarse-graining of all the downstream processes into
-a single effective rate. mRNA is again degraded at rate $\gamma$.
-
-Let $p_{m,R}(t)$ denote the joint probability of finding the
-system in the repressor bound state with $m$ mRNA molecules
-present at time $t$, and similarly define $p_{m,U}(t)$ for the
-repressor unbound state. 
-This model is governed by a master equation giving the time
-evolution of $p_{m,R}(t)$ and $p_{m,U}(t)$, which is
-\begin{align}
-\begin{split}
-\deriv{t}p_{m,R}(t) =& - k_R^- p_{m,R}(t) + k_R^+ p_{m,U}(t)
-                + (m+1)\gamma p_{m+1,R}(t) - \gamma p_{m,R}(t)
-\\
-\deriv{t}p_{m,U}(t) =&\; k_R^- p_{m,R}(t) - k_R^+ p_{m,U}(t)
-                        + rp_{m-1,U}(t) - rp_{m,U}(t)
-                \\
-                &+ (m+1)\gamma p_{m+1,U}(t) - \gamma p_{m,U}(t).
-\end{split}
-\end{align}
-In each equation, the first two terms describe transitions
-between promoter states due to repressors unbinding and binding,
-respectively. The final two terms describe degradation of mRNA,
-decreasing the copy number by one, and the terms with coefficient
-$r$ describe transcription initiation increasing the mRNA copy
-number by one.
-
-We can greatly simplify the notation, which will be especially
-useful for the more complicated models yet to come, by
-re-expressing the master equation in vector form. The promoter
-states are collected into a vector and the rate constants are
-collected into matrices as
-\begin{equation}
-\vect{p}_m = \begin{pmatrix} p_{m,R} \\ p_{m,U} \end{pmatrix},\
-\matr{K} = \begin{pmatrix} -k_R^- & k_R^+ \\ k_R^- & -k_R^+ \end{pmatrix},\
-\matr{R} = \begin{pmatrix} 0 & 0 \\ 0 & r \end{pmatrix},\
-\label{eq:2state_cme_matrices}
-\end{equation}
-so that the master equation may be condensed as
-\begin{equation}
-\deriv{t}\vect{p}_m(t) =
-\left( \matr{K} - \matr{R} - \gamma m \matr{I} \right) \vect{p}_m(t)
-                + \matr{R} \vect{p}_{m-1}(t)
-                + \gamma (m+1) \matr{I} \vect{p}_{m+1}(t)
-                - \gamma m \matr{I} \vect{p}_m(t),
-\label{eq:2state_rep_cme}
-\end{equation}
-where $\matr{I}$ is the identity matrix.
-Taking steady state by setting time derivatives to zero,
-the mean mRNA can be found to be
-\begin{equation}
-\langle m \rangle = \frac{r}{\gamma}
-        \left(1 + \frac{k_R^+}{k_R^-}\right)^{-1},
-\end{equation}
-with the algebra details again deferred to the appendix.
-Recall $k_R^+$ is proportional to the repressor copy number, so
-in computing fold-change, absence of repressor corresponds to
-$k_R^+\rightarrow0$.
-Therefore fold-change in this model is simply
-\begin{equation}
-\text{fold-change} = \left(1 + \frac{k_R^+}{k_R^-}\right)^{-1},
-\end{equation}
-again matching the master curve of~\fig{fig1:means_cartoons} with $\rho=1$.
-
-\subsubsection{Nonequilibrium model 2 - RNAP bound/unbound states}
-Analogous to the three-state equilibrium model, this model
-fine-grains the repressor unbound state of nonequilibrium model 1,
-resolving it into an empty promoter state and an RNAP-bound state.
-Note in this picture, in contrast with model 4 below,
-transcription initiation is accompanied by a promoter state
-change, in keeping with the interpretation as RNAP-bound and
-empty states: if an RNAP successfully escapes the promoter and
-proceeds to elongation of a transcript, clearly it is no longer
-bound at the promoter! Therefore another RNAP must bind before
-another transcript can be initiated.
-
-The master equation governing this model is analogous
-to~\eqrange{eq:2state_cme_matrices}{eq:2state_rep_cme} for model 1 above.
-The main subtlety arises since transcription initiation
-accompanies a promoter state change.
-This can be understood by analogy to $\matr{K}$. The
-off-diagonals and diagonals of $\matr{K}$ correspond to
-transitions arriving at or departing from, respectively, the
-promoter state of interest. If transcription initiations are
-accompanied by promoter state changes, we must have separate
-matrices for arriving and departing transcription events since
-the arriving and departing transitions have different initial
-copy number of mRNA, unlike for $\matr{K}$ where they are the same.
-The master equation for this model is
-\begin{equation}
-\deriv{t}\vect{p}_m(t) =
-\left( \matr{K} - \matr{R_D} - \gamma m \matr{I} \right) \vect{p}_m(t)
-                + \matr{R_A} \vect{p}_{m-1}(t) +
-                \gamma (m+1) \matr{I} \vect{p}_{m+1}(t)
-                - \gamma m \matr{I} \vect{p}_m(t),
-\label{eq:3state_rep_cme}
-\end{equation}
-with the state vector and promoter transition matrix defined as
-\begin{equation}
-\vect{p}_m = \begin{pmatrix} p_{m,R} \\ p_{m,E} \\ p_{m,P} \end{pmatrix},\
-\matr{K} = \begin{pmatrix} -k_R^- & k_R^+ & 0 \\
-                        k_R^- & -k_R^+ -k_P^+ & k_P^- \\
-                        0 & k_P^+ & -k_P^- 
-                \end{pmatrix},
-\label{eq:3state_cme_matrices_pt1}
-\end{equation}
-and the initiation matrices given by
-\begin{equation}
-\matr{R_A} = \begin{pmatrix}
-                0 & 0 & 0 \\ 0 & 0 & r \\ 0 & 0 & 0
-                \end{pmatrix},\
-\matr{R_D} = \begin{pmatrix}
-                0 & 0 & 0 \\ 0 & 0 & 0 \\ 0 & 0 & r
-                \end{pmatrix},
-\label{eq:3state_cme_matrices_pt2}
-\end{equation}
-The elements of $\vect{p}_m$ encode the probabilities of having
-$m$ mRNA present along with the promoter having repressor bound,
-being empty, or having RNAP bound, respectively.
-$\matr{R_A}$ describes probability flux \textit{arriving} at the state
-$\vect{p}_m$ from a state with one fewer mRNA, namely
-$\vect{p}_{m-1}$, and $\matr{R_D}$ describes probability flux
-\textit{departing} from the state $\vect{p}_m$
-for a state with one more mRNA, namely $\vect{p}_{m+1}$.
-$\matr{K}$ is closely analogous to model 1.
-
-Mean mRNA at steady state is found analogously to model 1, with the result
-\begin{equation}
-\langle m\rangle = \frac{r}{\gamma}
-        \frac{k_R^- k_P^+}
-        {k_R^- k_P^+ + k_R^- (k_P^- + r) + k_R^+ (k_P^- + r)}.
-\end{equation}
-Fold-change is again found from the ratio prescribed by~\eq{eq:fc_def},
-from which we have
-\begin{align}
-\text{fold-change}
-&=      \frac{k_R^- k_P^+}
-        {k_R^- k_P^+ + k_R^- (k_P^- + r) + k_R^+ (k_P^- + r)}
-        \frac{k_P^+ + k_P^- + r}{k_P^+}
-\\
-&=      \left(1 + \frac{k_R^+}{k_R^-}
-                \frac{k_P^- + r}{k_P^+ + k_P^- + r}
-        \right)^{-1}
-\\
-&=      \left(1 + \frac{k_R^+}{k_R^-}
-        \left(1 + \frac{k_P^+}{k_P^- + r}\right)^{-1}
-        \right)^{-1},
-\end{align}
-which follows the master curve with
-$\rho = 1 + k_P^+/(k_P^- + r)$ as claimed.
-
-\subsubsection{Nonequilibrium model 3
-- Multistep transcription initiation and escape}
-One might reasonably complain that the first two
-``nonequilibrium'' models we have considered are straw men. Their
-steady states necessarily satisfy detailed balance which is
-equivalent to thermodynamic equilibrium. Why is this the case? At
-steady state there is by definition no net probability flux in or
-out of each promoter state, but since the promoter states form a
-linear chain, there is only one way in or out of the repressor
-bound and RNAP bound states, implying each edge must actually
-have a net zero probability flux, which is the definition of
-detailed balance (usually phrased as equality of forward and
-reverse transition fluxes).
-
-Now we consider model 3 in~\eq{fig1:means_cartoons} which allows
-the possibility of true nonequilibrium steady-state fluxes
-through the promoter states.
-We point out that this model was considered previously in~\cite{Mitarai2015}
-where a comparison was made with model 1 as used in~\cite{Jones2014}.
-The authors of~\cite{Mitarai2015} argued that the additional
-complexity is essential to properly account for the noise in the
-mRNA distribution. We will weigh in on both models later when we
-consider observables beyond fold-change.
-
-The master equation governing this model is
-identical in form to model 2 above, namely
-\begin{equation}
-\deriv{t}\vect{p}_m(t) =
-\left( \matr{K} - \matr{R_D} - \gamma m \matr{I} \right) \vect{p}_m(t)
-                + \matr{R_A} \vect{p}_{m-1}(t) +
-                \gamma (m+1) \matr{I} \vect{p}_{m+1}(t)
-                - \gamma m \matr{I} \vect{p}_m(t),
-\end{equation}
-but obviously with a higher-dimensional state space and different matrices.
-The state vector and promoter transition matrix are now
-\begin{equation}
-\vect{p}_m = \begin{pmatrix} p_{m,R} \\ p_{m,E} \\
-                             p_{m,C} \\ p_{m,O}\end{pmatrix},\
-\matr{K} = \begin{pmatrix} -k_R^- & k_R^+ & 0 & 0\\
-                        k_R^- & -k_R^+ -k_P^+ & k_P^- & 0 \\
-                        0 & k_P^+ & -k_P^- - k_O & 0 \\
-                        0 & 0 & k_O & 0
-                \end{pmatrix},
-\end{equation}
-with the four promoter states, in order, being repressor bound,
-empty, RNAP closed complex, and RNAP open complex.
-Besides increasing dimension by one again, the only new feature in
-$\matr{K}$ is the rate $k_O$, representing the (irreversible)
-rate of open complex formation from the closed complex.
-The initiation matrices are given by
-\begin{equation}
-\matr{R_A} = \begin{pmatrix}
-        0 & 0 & 0 & 0 \\ 0 & 0 & 0 & r \\ 0 & 0 & 0 & 0\\ 0 & 0 & 0 & 0
-                \end{pmatrix},\
-\matr{R_D} = \begin{pmatrix}
-        0 & 0 & 0 & 0 \\ 0 & 0 & 0 & 0 \\ 0 & 0 & 0 & 0\\ 0 & 0 & 0 & r
-                \end{pmatrix},
-\end{equation}
-again closely analogous to model 2.
-
-The expression for mean mRNA is substantially more complicated
-now, as worked out in the appendix where we find
-\begin{equation}
-\langle m\rangle = \frac{r}{\gamma}
-        \frac{k_R^- k_P^+ k_O}
-        {k_R^- [(k_P^+ (k_O + r) + r(k_P^- + k_O)] + k_R^+ r(k_P^- + k_O)},
-\end{equation}
-which can be simplified to
-\begin{equation}
-\langle m\rangle
-= \frac{r}{\gamma}
-\frac{\frac{k_P^+ k_O}{r(k_O + k_P^-)}}
-        {1 + \frac{k_P^+ (k_O + r)}{r(k_O + k_P^-)} + \frac{k_R^+}{k_R^-}}.
-\end{equation}
-The strategy is to isolate the terms involving the repressor, so that now the fold-change is seen to be simply
-\begin{align}
-\text{fold-change}
-&= \frac{\frac{k_P^+ k_O}{r(k_O + k_P^-)}}
-        {1 + \frac{k_P^+ (k_O + r)}{r(k_O + k_P^-)} + \frac{k_R^+}{k_R^-}}
-        \frac{1 + \frac{k_P^+ (k_O + r)}{r(k_O + k_P^-)}}
-                {\frac{k_P^+ k_O}{r(k_O + k_P^-)}}
-\\
-&= \left(
-        1 + \frac{k_R^+}{k_R^-}
-        \left(1 + \frac{k_P^+ (k_O + r)}{r(k_O + k_P^-)}\right)^{-1}
-\right)^{-1},
-\end{align}
-surprisingly reducing to the master curve once again, with
-$\rho = 1 + \frac{k_P^+ (k_O + r)}{r(k_O + k_P^-)}$.
-
-This example hints that an arbitrarily fine-grained model of
-downstream transcription steps may still be collapsed to the form
-of the master curve in~\fig{fig1:means_cartoons}B so long as the
-repressor binding is exclusive with transcriptionally active
-states. We offer this as a conjecture, and we suspect that a
-careful argument using the King-Altman diagram
-method~\cite{King1956, Hill1966} might furnish a ``proof.''
-Our focus here is not on full generality but rather to survey an
-assortment of plausible models that have been proposed in the
-literature for simple repression.
-
-\subsubsection{Nonequilibrium model 4 - ``Active'' and ``inactive'' states}
-This model is at the core of the theory in~\cite{Razo-Mejia2020}.
-At a glance the cartoon for this model may appear very similar to model 2,
-and mathematically it is, but the interpretation is rather different.
-In model 2, we interpreted the third state literally as an
-RNAP-bound promoter and modeled initiation of a transcript as
-triggering a promoter state change, making the hopefully
-uncontroversial assumption that an RNAP can only make one
-transcript at a time. In contrast, in the present model the
-promoter state does \textit{not} change when a transcript is
-initiated. So we no longer interpret these states as literally
-RNAP bound and unbound but instead as coarse-grained ``active''
-and ``inactive'' states, the details of which we leave
-unspecified for now. We will comment more on this model below
-when we discuss Fano factors of models.
-
-Mathematically this model is very similar to models 1 and 2.
-Like model 1, the matrix $R$ describing transcription initiation
-is diagonal, namely
-\begin{equation}
-\matr{R} = \begin{pmatrix}
-                0 & 0 & 0 \\ 0 & 0 & 0 \\ 0 & 0 & r
-        \end{pmatrix}.
-\end{equation}
-so the master equation takes verbatim the same form as it did for model
-1,~\eq{eq:2state_rep_cme}. Meanwhile the promoter transition matrix
-$K$ is the same as~\eq{eq:3state_cme_matrices_pt1} from model 2,
-although we relabel the rate constants from $k_P^\pm$ to $k^\pm$ to
-reiterate that these are not simply RNAP binding and unbinding rates.
-
-Carrying out the algebra, the mean mRNA can be found to be
-\begin{equation}
-\langle m\rangle = \frac{r}{\gamma}
-\frac{k_R^- k^+}
-{k_R^- k^+ + k_R^- k^- + k_R^+ k^-},
-\end{equation}
-and the fold-change readily follows,
-\begin{align}
-\text{fold-change}
-&=      \frac{k_R^- k^+}{k_R^- k^+ + k_R^- k^- + k_R^+ k^-}
-        \frac{k_R^- k^+ + k_R^- k^-}{k_R^- k^+}
-\\
-&=      \left(1 + \frac{k_R^+}{k_R^-}
-                \left(1 + \frac{k^+}{k^-}\right)^{-1}
-        \right)^{-1},
-\end{align}
-from which we see $\rho = 1 + k^+/k^-$ as shown in~\fig{fig1:means_cartoons}.
-
-\subsubsection{Nonequilibrium model 5 - Bursty promoter}
-We will have much more to say about this model in subsequent sections.
-For now we sketch it as an intuitive analog to model 1,
-with just two states, repressor bound or unbound, and transition rates
-between them of $k_R^+$ and $k_R^-$.
-In model 1, when in the unbound state, single mRNA transcripts
-are produced as a Poisson process with some characteristic rate
-$r$. This model by contrast produces, at some Poisson rate $k_i$,
-\textit{bursts} of transcripts with a geometrically distributed
-burst size and a mean burst size $b$.
-
-From this intuitive picture and by analogy to model 1, then, it should be plausible that the mean mRNA level is
-\begin{equation}
-\langle m\rangle = \frac{k_i b}{\gamma}
-        \left(1 + \frac{k_R^+}{k_R^-}\right)^{-1},
-\end{equation}
-which will turn out to be correct from a careful calculation.
-For now, we simply note that just like model 1, the fold-change becomes
-\begin{equation}
-\text{fold-change} = \left(1 + \frac{k_R^+}{k_R^-}\right)^{-1}
-\end{equation}
-with $\rho=1$ also like model 1.
-
-We will also see later how this model emerges as a natural limit of model 4.
-
-\subsubsection{Summary and discussion of results across models for fold-changes in mean expression}
-What have we learned from our exploration of the models
-in~\fig{fig1:means_cartoons} and their predictions of
-fold-changes in mean expression levels? The takeaway is the
-existence of a master curve (\fig{fig1:means_cartoons}C) to which
-the fold-change predictions of all the models collapse. This
-master curve is parametrized by only two effective parameters:
-$\Delta F_R$, which characterizes the repressor, and $\rho$,
-which characterizes all other features of the promoter
-architecture. The key assumption underpinning this result is that
-zero transcription occurs when a repressor is bound to its
-operator. Note, however, that we are agnostic about the molecular
-mechanism which achieves this;
-steric effects are the obvious mechanism, but, for instance,
-``action-at-a-distance'' mediated by kinked DNA due to
-repressors bound tens or hundreds of nucleotides upstream of a
-promoter is plausible as well.
-
-Why does the master curve of~\fig{fig1:means_cartoons}C exist at all?
-This brings to mind the deep questions posed in,
-e.g.,~\cite{Frank2013} and~\cite{Frank2014a}, asking us to search
-for general ``symmetries'' of the problem instead of specifying
-generative models of the known microscopics of a system.
-In our case, the key feature seems to be the exclusive nature of
-repressor and RNAP binding, which allows the parameter describing
-the repressor, $\Delta F_R$, to cleanly separate from all other
-details of the promoter architecture, which are encapsulated in $\rho$.
-In the graph-theoretic language which was applied to similar
-problems in~\cite{Gunawardena2012} and~\cite{Ahsendorf2014}, the
-repressor-bound state is a leaf of the graph describing
-transitions between promoter states. Arbitrary nonequilibrium
-behavior can occur on the rest of the promoter state space, but
-it may all be swept up in the effective parameter $\rho$, to
-which the repressor makes no contribution.
-
-As suggested in~\cite{Chure2019}, we believe this master curve
-should generalize to architectures with multiple repressor
-binding sites, as long as the exclusivity of transcription factor
-binding and transcription initation is maintained. The
-interpretation of $\Delta F_R$ is then of an effective free
-energy of all repressor bound states.
-In an equilibrium picture this is simply given by the log of the
-sum of Boltzmann weights of all repressor bound states, which
-looks like the log of a partition function of a subsystem.
-In a nonequilibrium picture, while we can still mathematically gather
-terms and give the resulting collection the label $\Delta F_R$,
-it is unclear if the physical interpretation as an effective free energy
-makes sense.
-As is well known, free energies cannot be assigned unambiguously
-to states out of equilibrium because the free energy change along
-a generic path traversing the state space is path dependent,
-unlike at equilibrium.
-A consequence of this is that, out of equilibrium, $\Delta F_R$
-is no longer a simple sum of Boltzmann weights. Instead it
-resembles a restricted sum of King-Altman
-diagrams~\cite{King1956, Hill1966}. Following the work of
-Hill~\cite{Hill1989}, it may yet be possible to interpret this expression
-as an effective free energy, but this remains unclear to us.
-We leave this an open problem for future work.
-
-If we relax the requirement of exclusive repressor-RNAP binding,
-one could imagine models in which repressor and RNAP doubly-bound
-states are allowed, where the repressor's effect is to
-\text{reduce} the transcription rate rather than setting it to zero.
-Our results do not strictly apply to such a model, although we
-note that if the repressor's reduction of the transcription rate
-is substantial, such a model might still be well-approximated by
-one of the models in~\fig{fig1:means_cartoons}.
-
-One may worry that our ``one curve to rule them all'' is a
-mathematical tautology. In fact we \textit{agree} with this
-criticism if $\Delta F_R$ is ``just a fitting parameter'' and cannot
-be meaningfully interpreted as a real, physical free energy.
-An analogy to Hill functions is appropriate here. One of their
-great strengths and weaknesses, depending on the use they are put
-to, is that their parameters coarse-grain many details and are
-generally not interpretable in terms of microscopic models, for
-deep reasons discussed at length in~\cite{Frank2013}.
-By contrast, our master curve claims to have the best of both worlds:
-a coarse-graining of all details besides the repressor into
-a single effective parameter $\rho$,
-while simultaneously retaining an interpretation of
-$\Delta F_R$ as a physically
-meaningful and interpretable free energy.
-Our task, then, is to prove or disprove this claim.
-
-How do we test this and probe the theory with fold-change
-measurements? There is a fundamental limitation in that the
-master curve is essentially a one-parameter function of $\Delta
-F_R + \log\rho$. Worse, there are many \textit{a priori}
-plausible microscopic mechanisms that could contribute to the
-value of $\rho$, such as RNAP binding and escape
-kinetics~\cite{DeHaseth1998, Mitarai2015}, and/or supercoiling
-accumulation and release~\cite{Chong2014, Sevier2016}, and/or,
-RNAP clusters analogous to those obvserved in
-eukaryotes~\cite{Cisse2013, Cho2016}
-and recently also observed in bacteria~\cite{Ladouceur2020}.
-Even if $\Delta F_R$ is measured to
-high precision, inferring the potential microscopic contributions
-to $\rho$, buried inside a log no less, from fold-change
-measurements seems hopeless. As a statistical inference problem
-it is entirely nonidentifiable, in the language
-of~\cite{Gelman2013}, Section 4.3.
-
-If we cannot simply infer values of $\rho$ from measurements of
-fold-change, can we perturb some of the parameters that make up
-$\rho$ and measure the change? Unfortunately we suspect this is
-off-limits experimentally: most of the potential contributors to
-$\rho$ are global processes that affect many or all genes. For
-instance, changing RNAP association rates by changing RNAP copy
-numbers, or changing supercoiling kinetics by changing
-topoisomerase copy numbers, would massively perturb the entire
-cell's physiology and confound any determination of $\rho$.
-
-One might instead imagine a bottom-up modeling approach, where we
-mathematize a model of what we hypothesize the important steps
-are and are not, use in vitro data for the steps deemed
-important, and \textit{predict} what $\rho$ should be.
-But again, because of the one-parameter nature of the master
-curve, many different models will likely make indistinguishable
-predictions, and without any way to experimentally perturb in
-vivo, there is no clear way to test whether the modeling
-assumptions are correct.
-
-In light of this, we prefer the view that parameters and rates
-are not directly comparable between cartoons
-in~\fig{fig1:means_cartoons}. Rather, parameters in the
-simpler cartoons represent coarse-grained combinations of
-parameters in the finer-grained models. For instance, by equating
-$\rho$ between any two models, one can derive various possible
-correspondences between the two models' parameters. Note that
-these correspondences are clearly not unique, since many possible
-associations could be made. It then is a choice as to what microscopic
-interpretations the model-builder prefers for the parameters in a
-particular cartoon, and as to which coarse-grainings lend
-intuition and which seem nonsensical.
-Indeed, since it remains an open question what microscopic
-features dominate $\rho$ (as suggested above, perhaps RNAP
-binding and escape kinetics~\cite{DeHaseth1998, Mitarai2015}, or
-supercoiling accumulation and release~\cite{Chong2014, Sevier2016},
-or, something more exotic like
-RNAP clusters~\cite{Cisse2013, Cho2016, Ladouceur2020}),
-we are hesitant to put too much weight on any one microscopic
-interpretation of model parameters that make up $\rho$.
-
-One possible tuning knob to probe $\rho$ that would not globally
-perturb the cell's physiology is to manipulate RNAP binding sites.
-Work such as~\cite{Kinney2010} has shown that models of
-sequence-dependent RNAP can be inferred from data, and the
-authors of~\cite{Brewster2012} showed that the model
-of~\cite{Kinney2010} has predictive power by using the model to
-\textit{design} binding sites of a desired affinity.
-But for our purposes, this begs the question: the authors
-of~\cite{Kinney2010} \textit{assumed} a particular model
-(essentially our 3-state equilibrium model but without the
-repressor), so it is unclear how or if such methods can be turned
-around to \textit{compare} different models of promoter function.
-
-We have already pointed out that the master curve
-of~\fig{fig1:means_cartoons} is essentially a one-parameter model,
-the one parameter being $\Delta F_R + \log\rho$.
-By now the reader may be alarmed as to how can we even determine
-$\Delta F_R$ and $\rho$ independently of each other, never mind
-shedding a lens on the internal structure of $\rho$ itself.
-A hint is provided by the weak promoter approximation, invoked
-repeatedly in prior studies~\cite{Bintu2005c, Garcia2011a,
-Razo-Mejia2018} of simple repression using the 3-state
-equilibrium model in~\fig{fig1:means_cartoons}.
-In that picture, the weak promoter approximation means
-$\frac{P}{N_{NS}}\exp(-\beta\Delta\epsilon_P) \ll 1$,
-meaning therefore $\rho\approx1$.
-This is suggestive, but how can we be sure that $\rho$ is not,
-for instance, actually $10^2$ and that $\Delta F_R$ hides a
-compensatory factor?
-A resolution is offered by an independent inference of $\rho$ in
-the absence of repressors. This was done in~\cite{Razo-Mejia2020}
-by fitting nonequilibrium model 4  in~\fig{fig1:means_cartoons},
-with zero repressor (looking ahead, this is equivalent to model 4
-in~\fig{fig:constit_cartoons}), to FISH data
-from~\cite{Brewster2014}. This provided a determination of $k^+$
-and $k^-$, from which their ratio is estimated to be no more than
-a few $10^{-1}$ and possibly as small as $10^{-2}$.
-
-The realization that $\rho\approx1$ to an excellent
-approximation, \textit{independent} of which model
-in~\fig{fig1:means_cartoons} one prefers, goes a long way towards
-explaining the surprising success of equilibrium models of simple
-repression. Even though our 2- and 3-state models get so many
-details of transcription wrong, it does not matter because
-fold-change is a cleverly designed ratio. Since $\rho$ subsumes
-all details except the repressor, and $\log\rho\approx0$, fitting
-these simple models to fold-change measurements can still give a
-surprisingly good estimate of repressor binding energies. So the
-ratiometric construction of fold-change fulfills its intended
-purpose of canceling out all features of the promoter
-architecture except the repressor itself. Nevertheless it is
-perhaps surprising how effectively it does so: \textit{a priori},
-one might not have expected $\rho$ to be quite so close to 1.
-
-We would also like to highlight the relevance
-of~\cite{Landman2019} here. Landman et.\ al.\ reanalyzed and
-compared in vivo and in vitro data on the lacI repressor's
-binding affinity to its various operator sequences. (The in vivo
-data was from, essentially, fitting our master curve to
-expression measurements.) They find broad agreement between the
-in vitro and in vivo values. This reinforces the suspicion that
-the equilibrium $\Delta\epsilon_R$ repressor binding energies do
-in fact represent real physical free energies.
-Again, \textit{a priori} this did not have to be the case, even
-knowing that $\rho\approx1$.
-
-In principle, if $\Delta F_R$ can be measured to sufficient
-precision, then deviations from $\rho=1$ become a
-testable matter of experiment.
-In practice, it is probably unrealistic to measure repressor
-rates $k_R^+$ or $k_R^-$ or fold-changes in expression levels
-(and hence $\Delta\epsilon_R$) precisely enough to detect the
-expected tiny deviations from $\rho=1$.
-We can estimate the
-requisite precision in $\Delta F_R$ to resolve a given
-$\Delta\rho$ by noting, since $\rho\approx1$, that
-$\log(1+\Delta\rho)\approx \Delta\rho$, so
-$\Delta(\Delta F_R) \approx \Delta\rho$.
-Suppose we are lucky and $\Delta\rho$ happens to be $\sim0.1$, on
-the high end of our range estimated above.
-A determination of $\Delta\epsilon_R/k_BT$ with an uncertainty of
-barely 0.1 was achieved in the excellent measurements
-of~\cite{Razo-Mejia2018}, so this requires a very difficult
-determination of $\Delta F_R$ for a very crude determination of $\rho$,
-which suggests, to put it lightly, this is not a
-promising path to pursue experimentally.
-It is doubtful that inference of repressor kinetic rates
-would be any easier.\footnote{
-\mmnote{
-In fact the uncertainties quoted in Hammar et.\
-al.~\cite{Hammar2014} from Elf's group are small enough to do
-this calculation, but HJ \& I think their error bars are insanely
-optimistic. Not sure I want to blatantly call them out on it though??}
-}
-
-Moving forward, we have weak evidence supporting the
-interpretation of $\Delta F_R$ as a physically real free
-energy~\cite{Landman2019} and other work casting
-doubt~\cite{Hammar2014}.
-How might we resolve the confusion?
-If there is no discriminatory power to test the theory and
-distinguish the various models with measurements of fold-changes
-in means, how do we probe the theory? Clearly to discriminate
-between the nonequilibrium models in~\fig{fig1:means_cartoons},
-we need to go beyond means to ask questions about kinetics, noise and even
-full distributions of mRNA copy numbers over a population of cells.
-If the ``one-curve-to-rule-them-all'' is more than a
-mathematical tautology, then the free energy of repressor binding
-inferred from fold-change measurements should agree with
-repressor binding and unbinding rates.
-In other words, the equilbrium and nonequilibrium definitions of
-$\Delta F_R$ must agree, meaning
-\begin{equation}
-\Delta F_R = \beta\Delta\epsilon_R - \log(R/N_{NS})
-        = - \log(k_R^+/k_R^-),
-\end{equation}
-must hold, where $\beta\Delta\epsilon_R$ is inferred from the
-master curve fit to fold-change measurements, and $k_R^+$ and
-$k_R^-$ are inferred in some orthogonal manner.
-Single molecule measurements such as from~\cite{Hammar2014} have
-directly observed these rates, and in the remainder of this work
-we explore a complementary approach: inferring repressor binding
-and unbinding rates $k_R^+$ and $k_R^-$ from FISH measurements of
-mRNA population distributions.
-
-\section{Beyond means}
-\subsection{A one-state promoter with bursting is the ``best'' model of constitutive promoters}
-Before we can tackle simple repression, we need an adequate
-phenomenological model of constitutive expression.
-The literature abounds with options from which we can choose,
-and we show several potential kinetic models for constitutive promoters
-in~\fig{fig:constit_cartoons}.
-Let us consider the suitability of each model for our purposes in turn.
-
-\begin{figure}%[h!]
-\centering
-\includegraphics[width=0.9\textwidth]{../figures/fig1.5/fig1point5.ai}
-\caption{\textbf{Constitutive promoter cartoon comparison.}
-The left column depicts various plausible cartoons for constitutive promoters.
-We also list literature references where models have been used previously.
-In model (1), transcripts are produced in a Poisson process~\cite{Jones2014}.
-Model (2) features explicit modeling of RNAP binding/unbinding
-kinetics~\cite{Phillips2015a}.
-Model (3) is a more detailed generalization of model (2),
-treating transcription initiation as a multi-step process
-proceeding through closed and open complexes~\cite{Mitarai2015}.
-Model (4) is somewhat analogous to (2) except with the precise nature
-of active and inactive states left ambiguous~\cite{Razo-Mejia2020}.
-Finally, model (5) can be viewed as a certain limit of model (4)
-in which transcripts are produced in bursts,
-and initiation of bursts is a Poisson process.
-\mmnote{Haven't found model 5 formulated quite like this in the literature,
-usually burstiness is handled like model 4, but I should probably search a bit more.}
-The right column shows the Fano factor for each model.
-Note especially the crucial diagnostic: (2) and (3) have $\nu$
-strictly below 1, while only for (4) and (5) can $\nu$ exceed 1.
-Models with Fano factors $\le 1$ cannot produce the FISH data
-observed in~\cite{Jones2014} without introducing additional
-assumptions and model complexity.
-}
-\label{fig:constit_cartoons}
-\end{figure}
-
-\subsubsection{Model 1 - Poisson promoter.}
-This is the picture from Jones et.\ al.~\cite{Jones2014}, which
-assumes that transcripts are produced as a Poisson process from a
-single promoter state. This model insists that the ``true'' mRNA
-distribution is Poisson, implying the Fano factor $\nu$ must be 1.
-In~\cite{Jones2014}, the authors carefully attribute deviations
-from Fano = 1 to intensity variability in fluorescent spots, gene
-copy number variation, and copy number fluctuations of the
-transcription machinery, e.g., RNAP itself.
-In this picture, the master equation makes no appearance,
-and all the corrections to Poisson behavior are derived
-as additive corrections to the Fano factor.
-For disproving the ``universal noise curve'' from So et
-al~\cite{So2011}, this picture was excellent. It is appealing in
-its simplicity, with only two parameters, the initiation rate $r$
-and degradation rate $\gamma$. Since $\gamma$ is independently
-known from other experiments, and the mean mRNA copy number is
-$r/\gamma$, $r$ is easily inferred from data.
-In other words, the model is not excessively complex for the data at hand.
-But for many interesting questions, for instance in the recent
-work~\mmnote{cite Manuel's preprint once posted},
-knowledge of means and variances alone is insufficient, and a
-description of the full distribution of molecular counts is necessary.
-For this we need a( slightly) more complex model than model 1
-that would allow us to incorporate the non-Poissonian features of
-constitutive promoters directly into a master equation formulation.
-
-\subsubsection{Model 2 - Two-state promoter, RNAP bound or unbound.}
-This model was considered in, e.g.,~\cite{Phillips2015a} and ~\cite{Phillips2019}.
-Here transcription initiation proceeds from the bound to the unbound state,
-reflecting the microscopic reality that an RNAP that has begun to elongate
-a transcript is no longer available at the start site to begin another.
-The problem with this picture is that the Fano factor is
-\begin{align}
-    \nu = 1 -
-        \frac{r\rate{k}{P}{+}}
-            {\left(\rate{k}{P}{+} + \rate{k}{P}{-} + r\right)
-             \left(\gamma + \rate{k}{P}{+} + \rate{k}{P}{-} + r\right)},
-\end{align}
-which is always $<1$.
-To make contact with the experimental reality of $\nu>1$,
-we will have to do, at a minimum,
-the same corrections to Poisson behavior as in model 1 above.
-So while this model adds an appealing element of microscopic
-reality, we are forced to reject it as the additional complexity
-is unable to capture the phenomenology of interest.
-Obviously the promoter state does in fact proceed through cycles
-of RNAP binding, initiating, and elongating, but it seems that
-the super-Poissonian noise in mRNA copy number we want to
-model must be governed by other features of the system.
-
-\subsubsection{Model 3 - Three-state promoter, multistep
-transcription initiation and escape.}
-How might we remedy the deficits of model 2?
-It is known~\cite{DeHaseth1998} that once RNAP initially binds
-the promoter region, a multitude of distinct steps occur
-sequentially before RNAP finally escapes into the elongation
-phase. Perhaps adding some of this mechanistic detail might
-rescue model 2? The next simplest refinement of model 2 could
-consider open complex formation and promoter escape as separate
-steps rather than as a single effective step. In other words, we
-construct model 3 by adding a single extra state to model 2, and
-we will label the two RNAP-bound states as the closed and open
-complexes, despite the true biochemical details certainly being
-more complex. The authors of~\cite{Mitarai2015}, for instance,
-considered this model, although they added an additional
-repressor bound state and did not explicitly consider the limit
-with no repressor that we analyze here. Again, our goal here is
-not a complete accounting of all the relevant biochemical detail;
-this is an exploratory search for the important features our
-model needs to include.
-
-Unfortunately, as the authors of~\cite{Mitarai2015} hint, this
-model too has Fano factor $\nu<1$. This can be seen by starting
-with their Eq.~A1 for the Fano factor of the analogous model with
-repressor and taking the limit as repressor concentration goes to
-zero. After substantial algebra, the result is
-\begin{align}
-\nu = 1 - \frac{r k_O k_P^+}{\mathcal{Z}}
-\frac{k_P^+ + k_P^- + k_O + \gamma}
-        {\mathcal{Z} + \gamma(k_P^+ + k_P^- + k_O + \gamma) + \gamma^2},
-\end{align}
-where we defined $\mathcal{Z} = r(k_O + k_P^-) + k_P^+(k_O + r)$
-for notational tidiness. This is necessarily less than 1 for
-arbitrary rate constants.
-
-In fact, we suspect \textit{any} model in which transcription
-proceeds through a multistep cycle must necessarily have $\nu<1$.
-The intuitive argument compares the waiting time distribution to
-traverse the cycle with the waiting time for a Poisson promoter
-(model 1) with the same mean time. The latter is simply an
-exponential distribution. The former is a convolution of multiple
-exponentials, and intuitively should be more peaked with a
-smaller fractional width than an exponential with the same mean.\footnote{
-This can be made more precise with a result
-from~\cite{Stewart2007}, who showed that the convolution of
-multiple gamma distributions (of which the exponential
-distribution is a special case) is to a very good approximation
-also gamma distributed. Using their Eq.~2 for the distribution of
-the convolution, with shape parameters set to 1 to
-give exponential distributions, the total waiting time
-distribution has a ratio of variance to squared mean 
-$\sigma^2/\mu^2 = \sum_i k_i^2/\left(\sum_i k_i\right)^2$,
-where the $k_i$ are the rates of the individual steps. Clearly
-this is less than 1 and therefore the total waiting time
-distribution is narrower than the corresponding exponential.} A
-less disperse waiting time distribution means transcription
-initations are more uniformly distributed in time relative to a
-Poisson process. Hence the distribution of mRNA over a population
-of cells should be less variable compared to Poisson, giving
-$\nu<1$.\footnote{
-This last step, while intuitive, can be argued by analogy to
-photon statistics where antibunching gives rise to sub-Poissonian
-noise~\cite{Paul1982, Zou1990}. Although loopholes exist, we do
-not expect they apply for our problem. Nevertheless we refrain
-from elevating this cycle/sub-Poissonian equivalence to a ``theorem.''
-\mmnote{Antibunching is the ``obvious'' analogy for me but it'd
-be totally out of left field for most readers, any suggestions on
-a more comprehensible reference??}
-}
-
-Regardless of the merits of this model in describing the noise
-properties of constitutive transcription initation, it does not
-capture the dominant source of noise we want to include in our
-phenomenological model, so we must discard these details and
-search elsewhere.
-
-\subsubsection{Model 4 - Two-state promoter,
-``active'' and ``inactive'' states.}
-Inspired by~\cite{Razo-Mejia2020}, we revisit an analog of model
-2, but the interpretation of the two states is changed. Rather
-than explicitly viewing them as RNAP bound and unbound, we view
-them as ``active'' and ``inactive,'' which are able and unable to
-initiate transcripts, respectively. We are noncommittal as to the
-microscopic details of these states.
-
-One interpretation~\cite{Chong2014, Sevier2016} for the active
-and inactive states is that it represents the promoter's
-supercoiling state: transitions to the inactive state are caused
-by accumulation of positive supercoiling, which inhibits
-transcription, and transitions back to ``active'' are caused by
-gyrase or other topoisomerases relieving the supercoiling. This
-is an interesting possibility because it would mean the timescale
-for promoter state transitions is driven by topoisomerase
-kinetics, not by RNAP kinetics. From in vitro measurments, the
-former are suggested to be of order minutes~\cite{Chong2014}.
-Contrast this with model 2, where the state transitions are
-assumed to be governed by RNAP, which, assuming a copy number per
-cell of order $10^3$, has a diffusion-limited association rate
-$k_{on} \sim 10^2~\text{s}^{-1}$ to a target promoter. Combined
-with known $K_D$'s of order $\mu$M, this gives an RNAP
-dissociation rate $k_{off}$ of order $10^2$. As we will show
-below, however, there are some lingering puzzles with
-interpreting this supercoiling interpretation, so we leave it as
-a speculation and refrain from assigning definite physical
-meaning to the two states in this model.
-
-Intuitively one might expect that, since transcripts are produced
-as a Poisson process only when the promoter is in one of the two
-states in this model, transcription initiations should now be
-``bunched,'' in constrast to the ``anti-bunching'' of models 2
-and 3 above. One might further guess that this bunching would
-lead to super-Poissonian noise in the mRNA distribution over a
-population of cells. An honest calculation of the Fano factor produces
-\begin{align}
-\nu &= 1 + \frac{r k^-}{(k^+ + k^- + \gamma)(k^+ + k^-)},
-\end{align}
-which is strictly greater than 1, verifying the above intuition.
-Note we have dropped the $P$ label on the promoter switching rates to
-emphasize that these very likely do not represent kinetics of RNAP itself.
-This calculation can also be sidestepped by noting that the model
-is mathematically equivalent to the simple repression repression model
-from~\cite{Jones2014}, with states and rates relabeled and reinterpreted.
-
-How does this model compare to model 1 above?
-In model 1, all non-Poisson features of the mRNA
-distribution were handled as extrinsic corrections.
-By contrast, here the 3 parameter model is used to fit the full
-mRNA distribution as measured in mRNA FISH experiments.
-In essense, all variability in the mRNA distribution is regarded
-as ``intrinsic,'' arising either from stochastic initiation or
-from switching between the two coarse-grained promoter states.
-The advantage of this approach is that it fits neatly into the
-master equation picture, and the parameters thus inferred can be used as
-input for more complicated models with regulation by transcription factors.
-
-While this seems promising, there is a major drawback for our purposes
-which was already uncovered by the authors of~\cite{Razo-Mejia2020}:
-the statistical inference problem is nonidentifiable,
-in the sense described in Section 4.3 of~\cite{Gelman2013},
-meaning it is impossible to infer the
-parameters $r$ and $k^-$ from the FISH data of~\cite{Jones2014}
-(as shown in Fig.~S2 of~\cite{Razo-Mejia2020}).
-Rather, only the ratio $r/k^-$ could be inferred. In that work,
-the problem was worked around with an informative prior on the
-ratio $k^-/k^+$. That approach is unlikely to work here, as,
-recall, our entire goal in modeling constitutive expression is to
-use it as the basis for a yet more complicated model, when we add
-on repression. But adding more complexity to a model that is
-already poorly identified is a fool's errand,
-so we will explore one more potential model.
-
-\subsubsection{Model 5 - One-state promoter with explicit bursts.}
-This model is inspired by the failure mode of model 4.
-The key observation above was that, as found
-in~\cite{Razo-Mejia2020}, only two parameters, $k^+$ and the
-ratio $r/k^-$, could be directly inferred from the FISH data
-from~\cite{Jones2014}.
-So let us take this seriously and imagine a model where these are
-the only two model parameters. What would this model look like?
-
-To develop some intuition, consider model 4 in the limit
-$k^+ \ll k^- \lesssim r$, which is roughly satisfied by the
-parameters inferred in~\cite{Razo-Mejia2020}.
-In this limit, the system spends the majority of its time in the inactive state, occasionally becoming active and making a burst of transcripts.
-This should call to mind the well-known phenomenon of transcriptional bursting, as reported in, e.g.,~\cite{Golding2005,Chong2014,Sevier2016}\mmnote{should probably add some more cites here}.
-Let us make this correspondence more precise.
-The mean dwell time in the active state is $1/k^-$.
-While in this state, transcripts are produced at a rate $r$ per
-unit time. So on average, $r/k^-$ transcripts are produced before
-the system switches to the inactive state. Once in the inactive
-state, the system dwells there for an average time $1/k^+$ before
-returning to the active state and repeating the process. $r/k^-$
-resembles an average burst size, and $1/k^+$ resembles the time
-interval between burst events. More precisely, $1/k^+$ is the
-mean time between the end of one burst and the start of the next,
-whereas $1/k^+ + 1/k^-$ would be the mean interval between the
-start of two successive burst events, but in the limit $k^+ \ll
-k^-$, $1/k^+ + 1/k^- \approx 1/k^+$. Note that this limit ensures
-that the waiting time between bursts is approximately
-exponentially distributed, with mean set by the only timescale
-left in the problem,
-$1/k^+$.\footnote{
-If instead it were the case that $k^+ \sim k^-$, then the waiting time
-$1/k^+ + 1/k^-$ would have a peaked distribution, but this does
-not appear to be the case for any of datasets
-from~\cite{Jones2014}.}
-
-Let us now verify this intuition with a precise derivation
-to check that $r/k^-$ is in fact the mean burst size
-and to obtain the full burst size distribution.
-Consider first a constant, known dwell time $T$ in the active state.
-Transcripts are produced at a rate $r$ per unit time, so the
-number of transcripts $n$ produced during $T$ fits exactly
-the ``story'' for the Poisson distribution, i.e.,
-\begin{equation}
-    P(n\mid T) = \frac{(rT)^n}{n!} \exp(-rT).
-\end{equation}
-Since the dwell time $T$ is unobservable, we actually want $P(n)$,
-the dwell time distribution with no conditioning on $T$.
-Basic rules of probability theory tell us we can write $P(n)$
-in terms of $P(n\mid T)$ as
-\begin{equation}
-    P(n) =\int_0^\infty P(n\mid T) P(T) dT.
-\end{equation}
-But we know the dwell time distribution $P(T)$,
-which is exponentially distributed according to
-\begin{equation}
-    P(T) = k^- \exp(-T k^-),
-\end{equation}
-so $P(n)$ can be written as
-\begin{equation}
-    P(n) = k^- \frac{r^n}{n!}
-            \int_0^\infty T^n\exp[-(r + k^-)T]\,dT.
-\end{equation}
-A standard integral table shows
-$\int_0^\infty x^n e^{-ax}\,dx = n!/a^{n+1}$, so
-\begin{equation}
-    P(n) = \frac{k^- r^n}{(k^- + r)^{n+1}}
-        = \frac{k^-}{k^- + r}
-            \left(\frac{r}{k^- + r}\right)^n
-        = \frac{k^-}{k^- + r}
-            \left(1 - \frac{k^-}{k^- + r}\right)^n,
-\end{equation}
-which is exactly the geometric distribution with standard parameter
-$\theta\equiv k^-/(k^- + r)$ and domain
-$n \in \{0, 1, 2, \dots\}$
-(this is one of two common conventions for the geometric distribution).
-The mean of the geometric distribution, with this convention, is
-\begin{align}
-\langle n\rangle = \frac{1 - \theta}{\theta}
-        = \left(1 - \frac{k^-}{k^- + r}\right)
-                    \frac{k^- + r}{k^-}
-        = \frac{r}{k^-},
-\end{align}
-exactly as we guessed intuitively above.
-
-So in taking the limit $r,k^-\rightarrow\infty$, $r/k^-\equiv b$,
-we obtain a model which effectively has only a single promoter state,
-which initiates bursts at rate $k^+$
-(transitions to the active state, in the model 4 picture).
-The master equation for mRNA copy number $m$ takes the form
-\begin{align}
-\begin{split}
-\deriv{t}{p(m,t)} = & (m+1)\gamma p(m+1,t) - m\gamma p(m,t) \\
-        &+ \sum_{j=0}^{m-1} k^+ p(j,t) Geom(m-j;b)
-         - \sum_{j=m+1}^\infty k^+ p(m,t) Geom(j-m;b),
-\end{split}
-\end{align}
-where $Geom(n;b)$ is the geometric distribution with mean~$b$, i.e.,
-$Geom(n;b) = \frac{1}{1+b}\left(\frac{b}{1+b}\right)^n$
-(with domain over nonnegative integers as above).
-The first two terms are the usual mRNA degradation terms.
-The third term enumerates all ways the system can produce
-a burst of transcripts and arrive at copy number $m$,
-given that it had copy number $j$ before the burst.
-The fourth term allows the system to start with copy number $m$,
-then produce a burst and end with copy number $j$.
-In fact this last sum has trivial $j$ dependence and
-simply enforces normalization of the geometric distribution.
-Carrying it out we have
-\begin{equation}
-\begin{split}
-\deriv{t}{p(m,t)} = & (m+1)\gamma p(m+1,t) - m\gamma p(m,t) \\
-        &+ \sum_{j=0}^{m-1} k^+ p(j,t) Geom(m-j;b)
-            - k^+ p(m,t),
-\end{split}
-\end{equation}
-This improves on model 4 in that now the parameters are easily inferred,
-as we will see later, and have clean interpretations.
-The non-Poissonian features are attributed to the emprically
-well-established phenomenological picture of bursty transcription.
-
-The big approximation in going from model 4 to 5 is that a burst
-is produced instantaneously rather than over a finite time.
-If the true burst duration is not short compared to transcription factor
-kinetic timescales, this could be a problem in that mean burst size
-in the presense and absence of repressors could change,
-rendering parameter inferences from the constitutive case inappropriate.
-Let us make some simple estimates of this.
-
-Consider the time delay between the first and final RNAPs
-in a burst initiating transcription
-(\textit{not} the time to complete transcripts,
-which potentially could be much longer.)
-If this timescale is short compared to the typical
-search timescale of transcription factors, then all is well.
-The estimates from deHaseth et.\ al.~\cite{DeHaseth1998}
-put RNAP's diffusion-limited on rate around
-$\sim\text{few}\times10^{-2}~\text{nM}^{-1}~\text{s}^{-1}$
-and polymerase loading as high as $1~\text{s}^{-1}$.
-Then for reasonable burst sizes of $<10$, it is reasonable to guess that
-bursts might finish initiating on a timescale of tens of seconds or less
-(with another 30-60 sec to finish elongation, but that does not matter here).
-A transcription factor with typical copy number of order 10 (or less)
-would have a diffusion-limited association rate of order
-$(10~\text{sec})^{-1}$.
-Higher copy number TFs tend to have many binding sites over the genome,
-which should serve to pull them out of circulation and keep their
-effective association rates from rising too large.
-Therefore, there is \textit{perhaps} a timescale separation
-possible between transcription factor association rates and burst
-durations, but this assumption could very well break down, so we
-will have to keep it in mind when we infer repressor rates from
-the Jones et.\ al.\ FISH data later.
-
-In reflecting on these 5 models, the reader may feel that
-exploring a multitude of potential models just to return to a
-very minimal phenomenological model of bursty transcription may
-seem highly pedantic. But the purpose of the exercise was to
-examine a host of models from the literature and understand why
-they are insufficient, one way or another, for our purposes.
-Along the way we have learned that the detailed kinetics of RNAP
-binding and initiating transcription are probably irrelevant for
-setting the population distribution of mRNA.
-The timescales are simply too fast, and as we will see later
-in~\fig{fig:constit_post}, the noise seems to be governed by
-slower timescales. Perhaps in hindsight this is not surprising:
-intuitively, the degradation rate $\gamma$ sets the fundamental
-timescale for mRNA dynamics, and any other processes that
-substantially modulate the mRNA distribution should not differ
-from $\gamma$ by orders of magnitude.
-
-\subsection{Parameter inference for constitutive promoters}
-Having established that a one-state bursty model achieves a
-Goldilocks level of complexity, we test it more thoroughly on all
-the data from constitutive promoters from~\cite{Jones2014}.
-\fig{fig:constit_post} shows the results.
-\begin{figure}%[h!]
-\centering
-\includegraphics[width=0.9\textwidth]{../figures/fig2/fig2pt1.pdf}
-\caption{\textbf{Constitutive promoter posterior
-        inference and predictive checks.}
-        (A) shows the joint posterior density from MCMC sampling of model 5,
-        bursty promoter with negative binomially-distributed steady state.
-        (B) shows the ECDF of the observed population distribution
-        of mRNA transcripts under the control of a constitutive lacUV5 promoter.
-        Posterior predictive percentiles are shown as colored bands for
-        model (1), Poisson, and model (5), negative binomial.
-        }
-\label{fig:uv5_post}
-\end{figure}
-
-\begin{figure}%[h!]
-\centering
-\includegraphics[width=0.9\textwidth]{../figures/fig2/fig2pt2.pdf}
-\caption{\textbf{Constitutive promoter model comparison.}
-    (A) Joint posterior distributions for burst rate $k_i$ and
-    mean burst size $b$ for 18 promoters from Jones 2014.
-    Each contour indicates the 95\% highest posterior density
-    region for a particular promoter, i.e., on a plot analogous
-    to~\fig{fig:uv5_post}, the contour would enclose
-    approximately 95\% of the samples, and thus 95\% of the
-    probability mass.
-    Note that the vertical axis is shared with (B).
-    Panel (B) plots the burst rate $k_i$ vs.\ binding energy from energy
-    matrices from Brewster 2012.
-    The dotted line shows the predicted slope according to
-    $\ln k_i/\gamma \sim \text{constant} - \beta\Delta\epsilon_P$,
-    described in text.
-    }
-\label{fig:constit_post}
-\end{figure}
-\mmnote{Discuss puzzling comparison w/ Chong2014: if supercoiling is the thing, why are my burst sizes all the same but burst rates vary?
-And why is the duty cycle of my promoter so low, ie., bursts so short?
-COmpare their fig 7E; their $\beta/\alpha$ is my $k^+/k^-$.
-They have one or two genes with very small $\beta/\alpha$, does the \textit{galK} locus just happen to be that, or is there a deeper disagreement?
-Hard to say w/o more data.}
-
-\subsection{Transcription factor kinetics can be inferred from FISH measurements}
-Figure~3 presents inferred transcription factor rates, compares
-them to single-molecule measurements, and compares their ratios
-to gold-standard binding energies.
-\begin{figure}%[h!]
-\centering
-\includegraphics[width=0.9\textwidth]{../figures/fig3/fig3.ai}
-\caption{\textbf{Simple repression model comparison.}
-(A) depicts the cartoon of simple repression for which we infer parameters.
-(B) shows 50\% and 95\% contours of several 2D slices of the 9D posterior
-distribution, from fitting one unbinding rate for each operator (O1, O2, O3)
-and one binding rate for each aTc concentration (corresponding to an
-unknown mean repressor copy number).
-(C) compares ratios of our inferred unbinding rates with operator
-binding energy differences measured in Garcia \& Phillips 2011 (triangles)
-and Razo-Mejia et.\ al.\ 2018 (squares).
-(D) compares our inferred unbinding rates with single-molecule measurements
-of the same from Hammar et.\ at.\ 2014 \mmnote{Johan Elf's group}.
-Values are of similar order of magnitude,
-but a precise quantitative comparison unfortunately depends sensitively
-on the assumed mRNA lifetime, which was not measured.
-    }
-\label{fig3:kR_inferences}
-\end{figure}
-
-\begin{figure}%[h!]
-\centering
-\includegraphics[width=0.9\textwidth]{../figures/figSIxx/ppc_many_pooled.pdf}
-\caption{\textbf{Posterior predictive checks.}
-(A) duplicates the posterior from~\fig{fig3:kR_inferences}B.
-(B), (C), and (D) show samples from the posterior predictive distribution
-for each of the datasets used in the inference.
-Samples are sorted by operator and plotted separately simply for visual clarity.
-The unrepressed promoter, UV5, is shown with each as a reference point.
-\mmnote{
-Obviously the PPC is not perfect,
-but considering the relative simplicity of the model, I'm pleasantly surprised
-how well it worked and that the model was even identifiable at all.
-Interestingly, and perhaps not surprisingly, only 3 of the 9 datasets
-were identifiable if fit in isolation: otherwise,
-it was only by making assumptions that certain rates are equal
-across datasets was identifiability achieved.
-}
-    }
-\label{fig4:ppc}
-\end{figure}
-
+% !TEX root = ./busty_transcription.tex
 \section{Discussion and future work}
-\mmnote{
-Do all my approximations work simply because galK happens to have
-a low duty cycle of supercoiling, $\beta/\alpha$ in Chong2014
-picture? A potentially interesting test: put in the dozen or so
-Hernan constructs at some other locus, preferably one of the ones
-Chong identified as having a large duty cycle (large $\beta/\alpha$),
-i.e., the opposite limit of galK. Then just measure fold-changes.
-You don't even need to do FISH, although that would be even
-better. If their model is right, $\rho$ should be approximately
-set by the duty cycle of supercoiling, NOT by RNAP kinetics.
+\mmnote{Do all my approximations work simply because galK happens to have a low
+duty cycle of supercoiling, $\beta/\alpha$ in Chong2014 picture? A potentially
+interesting test: put in the dozen or so Hernan constructs at some other locus,
+preferably one of the ones Chong identified as having a large duty cycle (large
+$\beta/\alpha$), i.e., the opposite limit of galK. Then just measure
+fold-changes. You don't even need to do FISH, although that would be even
+better. If their model is right, $\rho$ should be approximately set by the duty
+cycle of supercoiling, NOT by RNAP kinetics.
 \textit{This is a NICE prediction for future work.} \textit{cspE}
-is claimed nonessential, prediction is you drop HG reporter
-constructs in there and the inferred binding energies of all lacI
-operators shift the same amount, by about $\ln(3)\sim 1kT$. Not
-huge, but detectable, although might be confounding effects from
-moving to a new locus. With FISH measurements, $k^+$ and $k^-$ in
-Fig 2, model 4 should now be separately inferrable if Chong2014
-measurements of $\beta/\alpha$ are correct. This would require an
-obnoxious amount of cloning, but once you did that, the
-measurements would be same old stuff that RP lab does.}
-
-%%%%%%%%%%%%%%%%%%%%% APPENDICES %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\appendix
-
-\section{Derivations of constitutive promoter models}
-\subsection{Model 2}
-\mmnote{Obtain Fano factor from CME.}
-\subsection{Model 3}
-\mmnote{Obtain Fano factor by taking limit from Mitarai et al's expression.}
-\subsection{Model 4}
-\mmnote{Obtain Fano factor by relabeling cartoon from Jones2014.}
-
-\section{Generating function solution for bursty promoters}
-
-\subsection{Constitutive promoter with bursts}
-
-\subsubsection{From master equation to generating function}
-
-We would like to derive the full probability distribution of
-mRNA copy number per cell for a model combining simple repression
-with a promoter featuring bursty transcription initiation.
-Before tackling mRNA and repression together, let us
-just consider the bursty consititutive promoter,
-model 5 in~\fig{fig:constit_cartoons}
-\mmnote{replicating Charlotte's notes; she says it's worked out
-in~\cite{Paulsson2000}, but I don't see it, at least not in notation I can
-comprehend. Need to find a cite for this,
-or maybe just cite Charlotte, personal comm, since that's where I actually got it\dots}.
-The master equation of interest is
-\begin{align}
-\deriv{t}p(m,t) = (m+1)\gamma p(m+1,t) - m\gamma p(m,t) - r p(m,t)
-        + r \sum_{m^\prime=0}^m G_{m-m^\prime}(\theta) p(m^\prime,t),
-\label{eq:1state_unreg_003}
-\end{align}
-where $G_{k}(\theta)$ is the geometric distribution defined as
-\begin{align}
-G_{k}(\theta) = \theta(1 - \theta)^k, \, k\in\{0,1,2,\dots\}.
-\end{align}
-With this convention, the mean burst size $b = (1-\theta)/\theta$.
-$\gamma$ and $r$ are mRNA degradation rates and transcription burst
-initiation rates, resp. The last term represents all ways the system could
-end up with $m$ mRNAs, having started the burst with $m^\prime$. Define
-$\lambda = r/\gamma$ and nondimensionalize time by $\gamma$, giving
-\begin{align}
-\deriv{t}p(m,t) = (m+1)p(m+1,t) - m p(m,t) - \lambda p(m,t)
-        + \lambda \sum_{m^\prime=0}^m G_{m-m^\prime}(\theta) p(m^\prime,t),
-\end{align}
-The probability generating function is defined as
-\begin{align}
-F(z,t) = \sum_{m=0}^\infty z^m p(m,t).
-\end{align}
-Multiply both sides of the CME by $z^m$ and sum over all $m$ to get
-\begin{align}
-\pderiv[F]{t} = \sum_{m=0}^\infty z^m
-\left(
-(m+1)p(m+1,t) - m p(m,t) - \lambda p(m,t)
-+ \lambda \sum_{m^\prime=0}^m G_{m-m^\prime}(\theta) p(m^\prime,t)
-\right).
-\end{align}
-The first two terms can be rewritten as derivatives of the
-generating function, i.e.,
-\begin{align}
-\sum_{m=0}^\infty z^m (m+1)p(m+1,t)
-=
-\pderiv{z} \sum_{m=0}^\infty z^{m+1} p(m+1,t)
-=
-\pderiv[F]{z}
-\end{align}
-and
-\begin{align}
-\sum_{m=0}^\infty z^m mp(m,t)
-=
-z\pderiv{z} \sum_{m=0}^\infty z^m p(m,t)
-=
-z\pderiv[F]{z}.
-\end{align}
-The final term is the most trouble. The trick is to reverse
-the default order of the sums, as in
-\begin{align}
-&\lambda \sum_{m=0}^\infty \sum_{m^\prime=0}^m
-        z^m \theta(1 - \theta)^{m-m^\prime} p(m^\prime,t)
-\\=
-&\lambda\theta \sum_{m^\prime=0}^\infty z^{m^\prime} p(m^\prime,t)
-        \sum_{m=m^\prime}^\infty
-        z^{m-m^\prime} (1 - \theta)^{m-m^\prime}
-\\=
-&\frac{\lambda\theta F}{1 - (1 - \theta)z}.
-\end{align}
-This works because after reordering the sums, the sum on $m$ is just
-a geometric series. Its result is independent of $m^\prime$, so the sum on
-$m^\prime$ reduces to the definition of the generating function.
-Putting these results together, the PDE for the generation function is
-\begin{align}
-\pderiv[F]{t} = \pderiv[F]{z} - z\pderiv[F]{z} - \lambda F
-        + \frac{\lambda\theta F}{1-z(1-\theta)}.
-\end{align}
-Changing variables to $\xi=1-\theta$ and simplifying gives
-\begin{align}
-\pderiv[F]{t} + (z - 1) \pderiv[F]{z} = \frac{(z-1)\xi}{1-z\xi}\lambda F.
-\label{eq:1state_unreg_015}
-\end{align}
-\subsubsection{Stead-state}
-At steady-state, the PDE reduces to the ODE
-\begin{align}
-\deriv[F]{z} = \frac{\xi}{1-z\xi}\lambda F,
-\end{align}
-which we can integrate as
-\begin{align}
-\int \frac{dF}{F} = \int \frac{\lambda\xi dz}{1-\xi z}.
-\end{align}
-The initial conditions for generating functions can be subtle and confusing.
-The key fact follows from the definition
-$F(z,t) = \sum_m z^m p(m,t)$,
-normalization requires that
-$F(z=1^-,t) = \sum_m p(m,t) = 1$.\footnote{
-Sometimes the generating function may be undefined \textit{at} $z=1$ but the
-limit still holds. Also many authors change variables from $z$ to other
-things, so one must keep track of how the normalization condition transforms.
-}
-Doing the integrals (and producing constant $c$) gives
-\begin{align}
-\ln F &= -\lambda \ln(1-\xi z) + c
-\\
-F &= \frac{c}{(1-\xi z)^\lambda}.
-\end{align}
-Only one choice for $c$ can satisfy initial conditions, producing
-\begin{align}
-F(z) = \left(\frac{1-\xi}{1-\xi z}\right)^\lambda
-        = \left(\frac{\theta}{1 - z(1-\theta)}\right)^\lambda,
-\end{align}
-which is exactly the negative binomial's generating function, as expected.
-
-\subsection{Adding repression}
-\subsubsection{Deriving the generating function for mRNA distribution}
-
-Let us move from a one-state promoter to a two-state promoter, where one state
-has repressor bound and the other produces transcriptional bursts as above.
-A schematic of this model is shown in~\fig{fig3:kR_inferences}.
-Although now we have an equation for each promoter state, otherwise
-the master equation reads similarly to the one-state case, except with
-additional terms corresponding to transitions between promoter states, namely
-\begin{align}
-\deriv{t}p_R(m,t) =& k_R^+ p_A(m,t) - k_R^- p_R(m,t)
-        + (m+1)\gamma p_R(m+1,t) - m\gamma p_R(m,t)
-\\
-\begin{split}
-\deriv{t}p_A(m,t) =& - k_R^+ p_A(m,t) + k_R^- p_R(m,t)
-        + (m+1)\gamma p_A(m+1,t) - m\gamma p_A(m,t) 
-\\
-&- k_i p_A(m,t) + k_i \sum_{m^\prime=0}^m \theta(1-\theta)^{m-m^\prime} p_A(m^\prime,t),
-\end{split}
-\end{align}
-where $p_R(m,t)$ is the probability of the system having $m$ mRNA copies
-and having repressor bound to the promoter at time $t$, and $p_A$ is an
-analogous probability to find the promoter without repressor bound.
-$k_R+$ and $k_R^-$ are, respectively, the rates at which repressors
-bind and unbind to and from the promoter, and
-$\gamma$ is the mRNA degradation rate.
-$k_i$ is the rate at which bursts initiate, and as before,
-the geometric distribution of burst sizes has mean $b=(1-\theta)/\theta$.\footnote{
-$\theta$ can be viewed as a Bernoulli trial probability: after each transcript
-in a burst is produced, $\theta$ is the probability that the burst
-terminates and $(1-\theta)$ is the probability that the burst continues.
-}
-
-Interestingly, it turns out that this problem maps exactly onto the
-three-stage promoter model considered by
-Shahrezaei and Swain in~\cite{Shahrezaei2008}, with relabelings.
-Their approximate solution for protein distributions amounts to
-the same approximation we make here in regarding the duration
-of mRNA synthesis bursts as instantaneous, so their solution
-for protein distributions also solves our problem of mRNA distributions.
-Let us examine the analogy more closely.
-They consider a two-state promoter, as we do here, but they model
-mRNA as being produced one at a time and degraded, with rates
-$v_0$ and $d_0$.
-Then they model translation as occuring with rate $v_1$,
-and protein degradation with rate $d_1$.
-Now consider the limit where $v_1, d_0\rightarrow\infty$ with
-their ratio $v_1/d_0$ held constant.
-$v_1/d_0$ resembles the average burst size of translation from a single
-mRNA: these are the rates of two Poisson processes that compete over a
-transcript, which matches the story of geometrically distributed burst sizes.
-So after taking this limit, on timescales slow compared to $v_1$ and $d_0$,
-it appears that transcription events fire at rate $v_0$ and produce a
-geometrically distributed burst of translation of mean size $v_1/d_0$,
-which intuitively matches the story we have told above for mRNA
-with variables relabeled.
-
-To verify this intuitively conjectured mapping between our problem and
-the solution in~\cite{Shahrezaei2008}, we continue with a careful
-solution for the mRNA distribution using probability generating functions,
-following the ideas sketched in~\cite{Shahrezaei2008}.
-It is natural to nondimensionalize rates in the problem by $\gamma$,
-or equivalently, this amounts to measuring time in units of $\gamma^{-1}$.
-We are also only interested in steady state, so we set the time
-derivatives to zero, giving
-\begin{align}
-0 =& k_R^+ p_A(m) - k_R^- p_R(m) + (m+1) p_R(m+1) - m p_R(m)
-\\
-\begin{split}
-0 =& - k_R^+ p_A(m) + k_R^- p_R(m) + (m+1) p_A(m+1) - m p_A(m) 
-\\
-&- k_i p_A(m) + k_i \sum_{m^\prime=0}^m \theta(1-\theta)^{m-m^\prime} p_A(m^\prime),
-\end{split}
-\end{align}
-
-The probability generating function is defined as before in the
-constitutive case, except now we must introduce a generating function
-for each promoter state,
-\begin{align}
-f_A(z) = \sum_{m=0}^\infty z^m p_A(m),
-\;
-f_R(z) = \sum_{m=0}^\infty z^m p_R(m).
-\end{align}
-Our real objective is the generating function $f(z)$ that generates
-the mRNA distribution $p(m)$, independent of what state the promoter is in.
-But since $p(m) = p_A(m) + p_R(m)$, it follows too that $f(z) = f_A(z) + f_R(z)$.
-
-So as before we multiply both equations by $z^m$ and sum over all $m$.
-Each individual term transforms exactly as did an analogous term in
-the constitutive case, so the coupled ODEs for the generating functions read
-\begin{align}
-0 =& k_R^+ f_A(z) - k_R^- f_R(z) + \pderiv{z} f_R(z) - z \pderiv{z} f_R(z)
-\\
-\begin{split}
-0 =&  - k_R^+ f_A(z) + k_R^- f_R(z) + \pderiv{z} f_A(z) - z \pderiv{z} f_A(z)
-\\
-&- k_i f_A(z) + k_i \frac{\theta}{1-z(1-\theta)} f_A(z),
-\end{split}
-\end{align}
-and after changing variables $\xi = 1 - \theta$ as before
-and rearranging, we have
-\begin{align}
-0 &= k_R^+ f_A(z) - k_R^- f_R(z) + (1-z) \pderiv{z} f_R(z)
-\\
-0 &=  - k_R^+ f_A(z) + k_R^- f_R(z) + (1 - z) \pderiv{z} f_A(z)
-+ k_i \frac{(z-1)\xi}{1-z\xi} f_A(z),
-\end{align}
-We can transform this problem from two coupled first-order ODEs to a
-single second-order ODE by solving for $f_A$ in the first and plugging
-into the second, giving
-\begin{align}
-\begin{split}
-0 = (1&-z) \pderiv[f_R]{z}
-+ \frac{1-z}{k_R^+}
-        \left(k_R^- \pderiv[f_R]{z} + \pderiv[f_R]{z} +(z-1) \psecderiv[f_R]{z}\right)
-\\
-&+ \frac{k_i}{k_R^+} \frac{(z-1)\xi}{1-z\xi}
-        \left(k_R^- f_R + (z-1) \pderiv[f_R]{z}\right),
-\end{split}
-\end{align}
-where, to reduce notational clutter, we have dropped the explicit $z$
-dependence of $f_A$ and $f_R$. Simplifying we have
-\begin{align}
-0 = \psecderiv[f_R]{z}
-        - \left(\frac{k_i\xi}{1-z\xi}
-                + \frac{1 + k_R^- + k_R^+}{1-z}
-        \right)\pderiv[f_R]{z}
-        + \frac{k_i k_R^- \xi}{(1-z\xi)(1-z)}f_R.
-\end{align}
-This can be recognized as the hypergeometric differential equation,
-with singularities at $z=1$, $z=\xi^{-1}$, and $z=\infty$. The latter
-can be verified by a change of variables from $z$ to $x=1/z$,
-being careful with the chain rule, and noting that $z=\infty$ is a
-singular point if and only if $x=1/z=0$ is a singular point.
-
-The standard form of the hypergeometric differential equation has its
-singularities at 0, 1, and $\infty$, so to take advantage of the
-standard form solutions to this ODE, we first need to transform
-variables to put it into a standard form.
-However, this is subtle. While any such transformation should work in
-principle, the solutions are expressed most simply in the neighborhood
-of $z=0$, but the normalization condition that we need to enforce
-corresponds to $z=1$. The easiest path, therefore, is to find a change
-of variables that maps
-1 to 0, $\infty$ to $\infty$, and $\xi^{-1}$ to 1.
-This is most intuitively done in two steps.
-
-First map the $z=1$ singularity to 0 by the change of variables $v=z-1$, giving
-\begin{align}
-0 = \psecderiv[f_R]{v}
-        + \left(\frac{k_i\xi}{(1+v)\xi - 1}
-                + \frac{1 + k_R^- + k_R^+}{v}
-        \right)\pderiv[f_R]{v}
-        + \frac{k_i k_R^- \xi}{((1+v)\xi - 1)v}f_R.
-\end{align}
-Now two singularities are at $v=0$ and $v=\infty$.
-The third is determined by $(1+v)\xi -1 = 0$, or $v=\xi^{-1} - 1$.
-We want another variable change that maps this third singularity to 1
-(without moving 0 or infinity). Changing variables again to
-$w=\frac{v}{\xi^{-1} - 1} = \frac{\xi}{1-\xi} v$ fits the bill.
-In other words, the combined change of variables
-\begin{align}
-w = \frac{\xi}{1-\xi} (z-1)
-\end{align}
-maps $z = \{1, \xi^{-1}, \infty\}$ to $w =\{0, 1, \infty\}$ as desired.
-Plugging in, being mindful of the chain rule and noting
-$(1 + v)\xi - 1 = (1 - \xi)(w - 1)$ gives
-\begin{align}
-0 = \left(\frac{\xi}{1-\xi}\right)^2 \psecderiv[f_R]{w}
-+ \left(
-        \frac{\xi k_i}{(1-\xi)(w-1)} + \frac{\xi(1 + k_R^- + k_R^+)}{(1-\xi)w}
-\right) \frac{\xi}{1-\xi} \pderiv[f_R]{w}
-+ \frac{k_i k_R^- \xi^2}{(1-\xi)^2 w(w-1)}f_R.
-\end{align}
-This is close to the standard form of the hypergeometric differential
-equation, and some cancellation and rearrangement gives
-\begin{align}
-0 = w(w-1)\psecderiv[f_R]{w}
-+ \left(k_i w + (1 + k_R^- + k_R^+)(w-1)\right) \pderiv[f_R]{w}
-+ k_i k_R^- f_R.
-\end{align}
-and a little more algebra produces
-\begin{align}
-0 = w(1-w)\psecderiv[f_R]{w}
-+ \left(1 + k_R^- + k_R^+
-        - (1 + k_i + k_R^- + k_R^+)w
-\right) \pderiv[f_R]{w}
-- k_i k_R^- f_R,
-\end{align}
-which is the standard form. From this we can read off the solution
-in terms of hypergeometric functions ${_2F_1}$ from
-any standard source, e.g.~\cite{Abramowitz1964}, and identify the
-conventional parameters in terms of our model parameters.
-We want the general solution in the neighborhood of $w=0$ ($z=1$),
-which for a homogeneous linear second order ODE must be a sum of two
-linearly independent solutions.
-More precisely,
-\begin{align}
-f_R(w) = C_1 {_2F_1}(\alpha, \beta, \gamma; w)
-+ C_2 w^{1-\gamma}{_2F_1}(1+\alpha-\gamma, 1+\beta-\gamma, 2-\gamma; w)
-\end{align}
-with parameters determined by
-\begin{align}
-\begin{split}
-\alpha\beta &= k_i k_R^-
-\\
-1+\alpha+\beta &= 1+k_i+k_R^-+k_R^+
-\\
-\gamma &= 1 + k_R^- + k_R^+
-\end{split}
-\end{align}
-and constants $C_1$ and $C_2$ to be set by boundary conditions.
-Solving for $\alpha$ and $\beta$, we find\footnote{
-Note that $\alpha$ and $\beta$ are interchangeable in the definition of
-${_2F_1}$ and differ only in the sign preceeding the radical.
-}
-\begin{align}
-\begin{split}
-\alpha &= \frac{1}{2}
-\left(k_i+k_R^-+k_R^+ + \sqrt{(k_i+k_R^-+k_R^+)^2 - 4k_i k_R^-}\right)
-\\
-\beta &= \frac{1}{2}
-\left(k_i+k_R^-+k_R^+ - \sqrt{(k_i+k_R^-+k_R^+)^2 - 4k_i k_R^-}\right)
-\\
-\gamma &= 1 + k_R^- + k_R^+.
-\end{split}
-\end{align}
-Since the normalization condition requires that $f_R$ be finite at $w=0$,
-we can immediately set $C_2=0$ to discard the second solution.
-This is because all the rate constants are strictly positive,
-so $\gamma>1$ and therefore $w^{1-\gamma}$ blows up as $w\rightarrow0$.
-Now that we have $f_R$, we would like to find the generating function
-for the mRNA distribution, $f(z) = f_A(z) + f_R(z)$.
-We can recover $f_A$ from our solution for $f_R$, namely
-\begin{align}
-f_A(z) = \frac{1}{k_R^+}\left(k_R^- f_R(z) + (z-1) \pderiv[f_R]{z}\right)
-\end{align}
-or
-\begin{align}
-f_A(w) = \frac{1}{k_R^+}\left(k_R^- f_R(w) + w \pderiv[f_R]{w}\right),
-\end{align}
-where in the second line we transformed our original relation between
-$f_R$ and $f_A$ to our new, more convenient, variable $w$.
-Plugging our solution for $f_R(w) = C_1{_2F_1}(\alpha, \beta, \gamma; w)$
-into $f_A$, we will require the differentiation rule for ${_2F_1}$,
-which tells us
-\begin{align}
-\pderiv[f_R]{w} = C_1\frac{\alpha\beta}{\gamma}
-                {_2F_1}(\alpha+1, \beta+1, \gamma+1; w),
-\end{align}
-from which it follows that
-\begin{align}
-f_A(w) = \frac{C_1}{k_R^+}
-\left(
-k_R^- {_2F_1}(\alpha, \beta, \gamma; w)
-+ w\frac{\alpha\beta}{\gamma} {_2F_1}(\alpha+1, \beta+1, \gamma+1; w)
-\right)
-\end{align}
-and therefore
-\begin{align}
-f(w) = C_1\left(1 + \frac{k_R^-}{k_R^+}\right)
-        {_2F_1}(\alpha, \beta, \gamma; w)
-+ w \frac{C_1}{k_R^+} \frac{\alpha\beta}{\gamma}
-        {_2F_1}(\alpha+1, \beta+1, \gamma+1; w).
-\end{align}
-To proceed, we need one of the (many) useful identities known for
-hypergeometric functions, in particular
-\begin{align}
-w\frac{\alpha\beta}{\gamma} {_2F_1}(\alpha+1, \beta+1, \gamma+1; w)
-=
-(\gamma-1)\left(
-{_2F_1}(\alpha, \beta, \gamma-1; w) - {_2F_1}(\alpha, \beta, \gamma; w)
-\right).
-\end{align}
-Substituting this for the second term in $f(w)$, we find
-\begin{align}
-f(w) = \frac{C_1}{k_R^+}
-\left[
-        \left(k_R^+ + k_R^-\right)
-        {_2F_1}(\alpha, \beta, \gamma; w)
-+ (\gamma-1)\left(
-        {_2F_1}(\alpha, \beta, \gamma-1; w) - {_2F_1}(\alpha, \beta, \gamma; w)
-        \right)
-\right],
-\end{align}
-and since $\gamma-1 = k_R^+ + k_R^-$, the first and third terms cancel,
-leaving only
-\begin{align}
-f(w) = C_1\frac{k_R^+ + k_R^-}{k_R^+} {_2F_1}(\alpha, \beta, \gamma-1; w).
-\end{align}
-Now we enforce normalization, demanding $f(w=0) = f(z=1) = 1$.
-${_2F_1}(\alpha, \beta, \gamma-1; 0) = 1$, so we must have
-$C_1 = k_R^+ / (k_R^+ + k_R^-)$ and consequently
-\begin{align}
-f(w) =  {_2F_1}(\alpha, \beta, k_R^+ + k_R^-; w).
-\end{align}
-Recalling that the mean burst size $b = (1-\theta)/\theta = \xi/(1-\xi)$
-and $w = \frac{\xi}{1-\xi} (z-1) = b (z-1)$,
-we can transform back to the original variable $z$ to find the tidy result
-\begin{align}
-f(z) =  {_2F_1}(\alpha, \beta, k_R^+ + k_R^-; b(z-1)),
-\end{align}
-with $\alpha$ and $\beta$ given above by
-\begin{align}
-\begin{split}
-\alpha &= \frac{1}{2}
-\left(k_i+k_R^-+k_R^+ + \sqrt{(k_i+k_R^-+k_R^+)^2 - 4k_i k_R^-}\right)
-\\
-\beta &= \frac{1}{2}
-\left(k_i+k_R^-+k_R^+ - \sqrt{(k_i+k_R^-+k_R^+)^2 - 4k_i k_R^-}\right).
-\end{split}
-\end{align}
-Finally we are in sight of the original goal. We can generate the
-steady-state probability distribution of interest by
-differentiating the generating function,
-\begin{align}
-p(m) = m! \left.\frac{\partial^m}{\partial z^m} f(z) \right|_{z=0},
-\end{align}
-which follows easily from its definition. Some contemplation reveals
-that repeated application of the derivative rule used above will produce
-products of the form $\alpha(\alpha+1)(\alpha+2)\cdots(\alpha+m-1)$ in
-the expression for $p(m)$ and similarly for $\beta$ and $\gamma$. These
-resemble ratios of factorials, but since $\alpha$, $\beta$, and $\gamma$
-are not necessarily integer, we should express the ratios using gamma
-functions instead. More precisely, one finds
-\begin{align}
-p(m) = \frac{
-        \Gamma(\alpha + m)\Gamma(\beta + m)\Gamma(k_R^+ + k_R^-)
-        }
-        {
-        \Gamma(\alpha)\Gamma(\beta)\Gamma(k_R^+ + k_R^- + m)
-        }
-\frac{b^m}{m!}{_2F_1}(\alpha+m, \beta+m, k_R^++k_R^-+m; -b)
-\label{eq:p_m_bursty+rep}
-\end{align}
-which is finally the probability distribution we sought to derive.
-
-\subsubsection{Moments}
-We can verify the correctness of~\eq{eq:p_m_bursty+rep}
-by comparison with Gillespie simulation.
-We can also easily compute moments of the distribution from
-the generating function, and check these against raw moments computed
-directly from the master equation.
-
-The $j$-th raw moment of the distribution can be computed from the
-generating function from the formula
-\begin{align}
-\langle m^j \rangle
-= \left. \left(z \frac{\partial}{\partial z}\right)^j f(z)\right|_{z=1}.
-\end{align}
-Note that \textit{probabilities} are generated by differentiating
-and evaluating at $z=0$, while \textit{moments} are generated by
-differentiating and evaluating at $z=1$. This also aligns with the
-normalization constraint, which can be thought of as the zeroth raw moment.
-
-The mean mRNA is given by
-\begin{align}
-\langle m \rangle = \left.z \pderiv[f]{z} \right|_{z=1}
-= \frac{\alpha\beta b}{k_R^+ + k_R^-}
-= k_i b\frac{k_R^-}{k_R^+ + k_R^-},
-\end{align}
-which is intuitive and appealingly simple:
-the mean mRNA is just the mean mRNA in the absense of repression
-($k_i b$, the burst rate time mean burst size)
-times the fraction of time the repressor is bound, $k_R^-/(k_R^+ + k_R^-)$.
-
-To compute the variance, let us first compute
-\begin{align}
-\begin{split}
-\langle m^2\rangle
-&= \left .z\pderiv{z}\left(z\pderiv[f]{z}\right) \right|_{z=1}
-= \left. z\left(\pderiv[f]{z} + z\psecderiv[f]{z}\right)\right|_{z=1}
-\\
-&= \frac{\alpha\beta b}{k_R^+ + k_R^-}
-        + \frac{\alpha(\alpha+1)\beta(\beta+1) b^2}
-                {(k_R^+ + k_R^-)(k_R^+ + k_R^- + 1)}
-\\
-&= \frac{k_i b k_R^-}{k_R^+ + k_R^-}
-        + \frac{k_i k_R^- b^2 (1 + k_i + k_R^+ + k_R^- + k_i k_R^-)}
-                {(k_R^+ + k_R^-)(k_R^+ + k_R^- + 1)},
-\end{split}
-\end{align}
-from which the variance is
-\begin{align}
-\begin{split}
-var(m) = \langle m^2\rangle - \langle m\rangle^2
-&= \frac{k_i b k_R^-}{k_R^+ + k_R^-}
-\left(
-1 + \frac{b(1 + k_i + k_R^+ + k_R^- + k_i k_R^-)} {k_R^+ + k_R^- + 1}
-- \frac{k_i b k_R^-}{k_R^+ + k_R^-}
-\right)
-\\
-&= k_i b\frac{k_R^-}{k_R^+ + k_R^-}
-\left(
-1 + b + k_i b\frac{1 + k_R^-} {k_R^+ + k_R^- + 1}
-- \frac{k_i b k_R^-}{k_R^+ + k_R^-}
-\right)
-\\
-&= k_i b\frac{k_R^-}{k_R^+ + k_R^-}
-\left(
-1 + b + k_i b\frac{k_R^+}{(k_R^+ + k_R^-)(k_R^+ + k_R^- + 1)}
-\right).
-\end{split}
-\end{align}
-The Fano factor is the quantity in parentheses. This can be interpreted as
-the variability arising from the negative binomial distribution of the
-unregulated promoter $(1+b)$ plus additional noise arising from repression.
-
-\mmnote{Still to writeup: moment derivation direct from CME to compare}
-
-\subsection{Numerical considerations and recursion formulas}
-\subsubsection{Generalities}
-We would like to carry out Bayesian parameter inference on FISH data
-from~\cite{Jones2014}, using~\eq{eq:p_m_bursty+rep} as our
-likelihood. This requires accurate (and preferably fast)
-numerical evaluation of the hypergeometric function ${_2F_1}$,
-which is a notoriously hard problem~\cite{Pearson2017, Gil2007},
-and our particular needs here present an especial challenge as we show below.
-
-The hypergeometric function is defined by its Taylor series as
-\begin{align}
-{_2F_1}(a,b,c;z) 
-= \sum_{l=0}^\infty
-\frac{\Gamma(a + l)\Gamma(b + l)\Gamma(c)}
-        {\Gamma(a)\Gamma(b)\Gamma(c + l)}
-\frac{z^l}{l!}
-\end{align}
-for $|z|<1$, and by analytic continuation elsewhere.
-If $z\lesssim1/2$ and $\alpha$ and $\beta$ are not too large
-(absolute value below 20 or 30),
-then the series converges quickly and an accurate numerical representation is
-easily computed by truncating the series after a reasonable number of terms.
-Unfortunately, we need to evaluate ${_2F_1}$ over mRNA copy numbers fully out
-to the tail of the distribution, which can easily reach 50, possibly 100.
-From~\eq{eq:p_m_bursty+rep}, this means evaluating ${_2F_1}$
-repeatedly for values of $a$, $b$, and $c$ spanning the full range
-from $\mathcal{O}(1)$ to $\mathcal{O}(10^2)$,
-even if $\alpha$, $\beta$, and $\gamma$
-in~\eq{eq:p_m_bursty+rep} are small,
-with the situation even worse if they are not small.
-A naive numerical evaluation of the series definition will be
-prone to overflow and, if any of $a,b,c<0$, then some successive terms in the
-series have alternating signs which can lead to catastrophic cancellations.
-
-One solution is to evaluate ${_2F_1}$ using arbitrary precision arithmetic
-instead of floating point arithmetic,
-e.g., using the \texttt{mpmath} library in Python.
-This is accurate but incredibly slow computationally.
-To quantify how slow, we found that
-evaluating the likelihood defined by~\eq{eq:p_m_bursty+rep} $\sim50$ times
-(for a typical dataset of interest from~\cite{Jones2014},
-with $m$ values spanning 0 to $\sim50$)
-using arbitrary precision arithmetic is 100-1000 fold slower than
-evaluating a negative binomial likelihood for the corresponding
-constitutive promoter dataset.
-
-To claw back $\gtrsim30$ fold of that slowdown, we can exploit
-one of the many catalogued symmetries involving ${_2F_1}$.
-The solution involves recursion relations originally explored by Gauss,
-and studied extensively in~\cite{Pearson2017, Gil2007}.
-They are sometimes known as contiguous relations and relate the values
-of any set of 3 hypergeometric functions whose arguments differ by integers.
-To rephrase this symbolically, consider a set of hypergeometric functions
-indexed by an integer $n$,
-\begin{align}
-f_n = {_2F_1}(a+\epsilon_i n, b+\epsilon_j n, c+\epsilon_k n; z),
-\end{align}
-for a fixed choice of $\epsilon_i, \epsilon_j, \epsilon_k \in \{0,\pm 1\}$
-(at least one of $\epsilon_i, \epsilon_j, \epsilon_k$ must be nonzero,
-else the set of $f_n$ would contain only a single element).
-Then there exist known recurrence relations of the form
-\begin{align}
-A_n f_{n-1} + B_n f_{n} + C_n f_{n+1} = 0,
-\end{align}
-where $A_n, B_n$, and $C_n$ are some functions of $a,b,c$, and $z$.
-In other words, for fixed $\epsilon_i, \epsilon_j, \epsilon_k, a, b,$ and $c$,
-if we can merely evaluate ${_2F_1}$ twice, say for $n^\prime$ and $n^\prime-1$,
-then we can easily and rapidly generate values for arbitrary $n$.
-
-This provides a convenient solution for our problem: we need repeated
-evaluations of ${_2F_1}(a+m, b+m, c+m; z)$
-for fixed $a,b$, and $c$ and many integer values of $m$.
-They idea is that we can use arbitrary precision arithmetic to evaluate
-${_2F_1}$ for just two particular values of $m$ and then generate
-${_2F_1}$ for the other 50-100 values of $m$ using the recurrence
-relation.\footnote{
-There are even more sophisticated ways of utilizing the recurrence
-relations that might have netted another factor of 2 speed-up, and
-possibly as much as a factor of 10, but the method described here had
-already reduced the computation time to an acceptable
-$\mathcal{O}(\text{1 min})$, so these more sophisticated approaches did
-not seem worth the time to pursue.
-}
-
-However, there two further wrinkles.
-The first is that
-a naive application of the recurrence relation is numerically unstable.
-Roughly, this is because the three term recurrence relations,
-like second order ODEs, admit two linearly independent solutions.
-In a certain eigenbasis, one of these solutions dominates the other
-as $n\rightarrow\infty$, and as $n\rightarrow-\infty$,
-the dominance is reversed.
-If we fail to work in this eigenbasis, our solution of the recurrence relation
-will be a mixture of these solutions and rapidly accumulate numerical error.
-For our purposes, it suffices to know that the authors of~\cite{Gil2007}
-derived the numerically stable solutions (so-called \textit{minimal solutions})
-for several possible choices of $\epsilon_i, \epsilon_j, \epsilon_k$.
-Running the recurrence in the proper direction using a minimal solution
-is numerically robust and can be done entirely in floating point arithmetic, 
-so that we only need to evaluate ${_2F_1}$ with arbitrary precision arithmetic
-to generate the seed values for the recursion.
-
-The second wrinkle is a corollary to the first.
-The minimal solutions are only minimal for certain ranges of the argument $z$,
-and not all of the 26 possible recurrence relations
-have minimal solutions for all $z$.
-This can be solved by using one of the many transformation formulae for
-${_2F_1}$ to convert to a different recurrence relation that has
-a minimal solution over the required domain of $z$, although
-this can require some trial and error to find the right transformation,
-the right recurrence relation, and the right minimal solution.
-
-\subsubsection{Particulars}
-Let us now demonstrate these generalities for our problem of interest.
-In order to evaluate the probability distribution of our
-model,~\eq{eq:p_m_bursty+rep}, we need to evaluate hypergeometric functions
-of the form ${_2F_1}(\alpha+m, \beta+m, \gamma+m; -b)$
-for values of $m$ ranging from $0$ to $\mathcal{O}(100)$.
-The authors of~\cite{Gil2007} did not derive a recursion relation
-for precisely this case. We could follow their methods and do so ourselves,
-but it is much easier to convert to a case that they did consider.
-The strategy is to look through the minimal solutions tabulated
-in~\cite{Gil2007} and search for a transformation we could apply to
-${_2F_1}(\alpha+m, \beta+m, \gamma+m; -b)$ that would place the $m$'s
-(the variable being incremented by the recursion)
-in the same arguments of ${_2F_1}$ as the minimal solution.
-After some ``guess and check,'' we found that the transformation
-\begin{align}
-{_2F_1}(\alpha+m, \beta+m, \gamma+m; -b)
-=
-(1+b)^{-\alpha-m}
-        {_2F_1}\left(\alpha+m, \gamma-\beta, \gamma+m; \frac{b}{1+b}\right),
-\label{eq:rec_euler_pretransform}
-\end{align}
-produces a ${_2F_1}$ on the right hand side that closely resembles
-the minimal solutions $y_{3,m}$ and $y_{4,m}$ in Eq.~4.3 in~\cite{Gil2007}.
-Explicitly, these solutions are
-\begin{align}
-y_{3,m}
-&\propto
-{_2F_1}\left(-\alpha^\prime + \gamma^\prime - m,
-                -\beta^\prime + \gamma^\prime,
-                1-\alpha^\prime-\beta^\prime+\gamma^\prime-m;
-                1-z\right)
-\\
-y_{4,m}
-&\propto
-{_2F_1}\left(\alpha^\prime + m,
-                \beta^\prime,
-                1+\alpha^\prime+\beta^\prime-\gamma^\prime+m;
-                1-z\right),
-\label{eq:minimal_soln_sans_prefac}
-\end{align}
-where we have omitted prefactors which are unimportant for now.
-Which of these two we should use depends on what values $z$ takes on.
-Equating $1-z=b/(1+b)$ gives $z=1/(1+b)$, and since $b$ is strictly positive,
-$z$ is bounded between 0 and 1.
-From Eq.~4.5 in~\cite{Gil2007}, $y_{4,m}$ is the minimal solution
-for real $z$ satisfying $0<z<2$, so this is the only minimal solution we need.
-
-Now that we have our minimal solution,
-what recurrence relation does it satisfy?
-Confusingly, the recurrence relation of which $y_{4,m}$ is a solution
-increments different arguments of ${_2F_1}$ that does $y_{4,m}$:
-it increments the first only, rather than first and third.
-This recurrence relation can be looked up, e.g., Eq.~15.2.10
-in~\cite{Abramowitz1964}, which is
-\begin{align}
-(\gamma^\prime - (\alpha^\prime + m)) f_{m-1}
-+
-(2(\alpha^\prime+m) - \gamma^\prime + (\beta^\prime - \alpha^\prime)z)f_m
-+ \alpha^\prime(z-1) f_{m+1} = 0.
-\label{eq:chosen_rec_rel}
-\end{align}
-Now we must solve for the parameters appearing in the recurrence relation
-in terms of our parameters, namely by setting
-\begin{align}
-\begin{split}
-\alpha^\prime &= \alpha
-\\
-\beta^\prime &= \gamma - \beta
-\\
-1 + \alpha^\prime + \beta^\prime - \gamma^\prime &= \gamma
-\\
-1 - z &= \frac{b}{1+b}
-\end{split}
-\end{align}
-and solving to find
-\begin{align}
-\begin{split}
-\alpha^\prime &= \alpha
-\\
-\beta^\prime &= \gamma - \beta
-\\
-\gamma^\prime &= 1 + \alpha - \beta
-\\
-z &= \frac{1}{1+b}
-.
-\end{split}
-\end{align}
-Finally we have everything we need. The minimal solution
-\begin{align}
-y_{4,m}
-=
-\frac{\Gamma(1+\alpha^\prime-\gamma^\prime+m)}
-        {\Gamma(1+\alpha^\prime+\beta^\prime-\gamma^\prime+m)}
-\times
-{_2F_1}\left(\alpha^\prime + m,
-                \beta^\prime,
-                1+\alpha^\prime+\beta^\prime-\gamma^\prime+m;
-                1-z\right),
-\end{align}
-where we have now included the necessary prefactors,
-is a numerically stable solution of the recurrence
-relation~\eq{eq:chosen_rec_rel} if the recursion is run
-from large $m$ to small $m$.
-So the procedure is to compute the value of ${_2F_1}$ for the two
-largest $m$ values of interest using arbitrary precision arithmetic,
-compute the prefactors to construct
-$y_{4,\text{max}(m)}$ and $y_{4,\text{max}(m)-1}$,
-recursively compute $y_{4,m}$ for all $m$ less than $\text{max}(m)$ down
-to $m=0$, then cancel off the prefactors of the resulting values of
-$y_{4,m}$ for all $m$ to produce ${_2F_1}$ for all desired $m$ values.
-\mmnote{Maybe I should reformat this paragraph into
-a pseudocode version of the algorithm?
-Instead of, or in addition to?}
-
-With ${_2F_1}$ computed, the only remaining numerical danger in computing
-$p(m)$ in~\eq{eq:p_m_bursty+rep} is overflow of the gamma functions.
-This is easily solved by taking the log of the entire expression
-and using standard routines to compute the log of the gamma functions,
-then exponentiating the entire expression at the end if $p(m)$
-is needed rather than $\log p(m)$.
-
-% \bibliographystyle{nature}
-\bibliographystyle{abbrv}
-\bibliography{./library}
-
-\end{document}
+is claimed nonessential, prediction is you drop HG reporter constructs in there
+and the inferred binding energies of all lacI operators shift the same amount,
+by about $\ln(3)\sim 1kT$. Not huge, but detectable, although might be
+confounding effects from moving to a new locus. With FISH measurements, $k^+$
+and $k^-$ in Fig 2, model 4 should now be separately inferrable if Chong2014
+measurements of $\beta/\alpha$ are correct. This would require an obnoxious
+amount of cloning, but once you did that, the measurements would be same old
+stuff that RP lab does.}

### 2020-05-20 doc/abstract.tex
RP report on May 19 morning
wrote abstract
went carefully through section 1 on introduction
went carefully through section 2 on means
addded many items to bib file

@@
 \begin{abstract}
-Abstract here.
+The study of transcription remains one of the centerpieces of modern biology with implications in settings from development to metabolism to evolution to disease.   Precision measurements
+using a host of different techniques including fluorescence and sequencing readouts have
+raised the bar for what it means to quantitatively understand transcriptional
+regulation.  The simplest example of a regulatory architecture is the repressor-operator
+model originally posited by Jacob and Monod in the 1960s, in which a single repressor binds to a promoter site and inhibits transcription.  Our understanding
+of this model architecture is sufficiently refined both experimentally
+and theoretically that it has become possible to carefully discriminate between different
+conceptual pictures of how simple repression works.  In this paper, we show how seven
+distinct models of the simple-repression motif, based both on equilibrium and kinetic
+thinking, can be used to derive the predicted level of mean expression and its variance.
+These different models are invoked to confront a variety of different data on both
+mean and variance in gene expression, illustrating the extent to which such
+models can be distinguished.
 \end{abstract}

### 2020-05-20 doc/section_01_introduction.tex
RP report on May 19 morning
wrote abstract
went carefully through section 1 on introduction
went carefully through section 2 on means
addded many items to bib file

@@
 \section{Introduction}
-\mmnote{Key ideas, no particular order, that I haven't written down before:
-\begin{itemize}
-\item Our goal is to build phenomenological models of input-output functions of
-genetic circuits. More precisely, we want phenomenology in the sense of
-coarse-graining away atomistic detail, but still retaining biophysical meaning,
-e.g., we don't want to coarse-grain as far as Hill functions. Why not? We are
-motivated by studies like~\cite{Razo-Mejia2020}, for which Hill functions are
-insufficient. Such work requires a quantitative theory of how biophysical
+
+Gene expression presides over much of the most important dynamism 
+of living organisms.   The level of expression of batteries of
+different genes is altered as a result of spatiotemporal cues that integrate
+chemical, mechanical and other types of signals.  The original repressor-operator model
+conceived by Jacob and Monod in the context of bacterial
+metabolism has now been transformed into the much
+broader subject of gene regulatory networks in living organisms
+of all kinds~\cite{Jacob1961, Britten1969, Ben-TabouDe-Leon2007}.  One of the remaining outstanding challenges
+to have emerged in the genomic era is our continued
+inability to predict the regulatory consequences of
+different regulatory architectures, stemming first and foremost
+from our ignorance about what those architectures even are with
+more than 60\% of the genes even in an ostensibly well understood organism
+such as  {\it E. coli} with no regulatory
+insights at all~\cite{Rydenfelt2014-2,Belliveau2018,Ghatak2019, Santos_Zavaleta2019}.
+But even once we have established the identity of key transcription
+factors and their binding sites of a given promoter architecture,  there remains the predictive challenge
+of understanding its input-output properties, an objective that can be met
+by a myriad of approaches using the tools of statistical physics~\cite{Ackers1982, Shea1985, Buchler2003,Vilar2003a,Vilar2003b, Bintu2005a,Bintu2005c, Gertz2009,Sherman2012, Saiz2013,
+Ko1991,Peccoud1995,Record1996,
+Kepler2001, Sanchez2008,Shahrezaei2008,Sanchez2011, Michel2010}.  One root
+to such predictive understanding is to focus on the simplest
+regulatory architecture and to push the theory-experiment
+dialogue as far and as hard as it can be pushed~\cite{Phillips2019}.  If we
+demonstrate that we can pass that test by successfully predicting
+both the means and variance in gene expression, then that provides
+a more solid foundation upon which to launch into 
+more complex problems.
+
+To that end, in this paper we examine a wide variety of distinct models
+for the simple repression regulatory architecture.  All of these
+models coarse-grain away some of the important microscopic
+features of these architectures that have been elucidated
+by generations of genetics, molecular biology and biochemistry.  
+Our aim is to understand the underlying principles of such
+coarse-graining such that future models of regulatory response
+can serve the powerful predictive role needed to take synthetic
+biology from what is often a brilliant exercise in enlightened 
+empiricism and recast it as the same kind of rational design process
+we are used to in other branches of engineering.  
+More precisely, we want phenomenology in the sense of
+coarse-graining away atomistic detail, but still retaining biophysical meaning.
+For example, we are not satisfied with the strictly phenomenological
+approach offered by 
+ Hill functions since we are motivated by studies like~\cite{Razo-Mejia2020}, for which Hill functions are
+clearly insufficient since each new situation
+requires a completely new set of parameters. Such work requires a quantitative theory of how biophysical
 changes at the molecular level propagate to input-output functions at the
 genetic circuit level. We want concepts, not mere facts. In particular a key
-question is: at this level of coarse graining, what microscopic details do I
-need to explicitly model, and how do we figure that out? For example, do I need
+question is: at this level of coarse graining, what microscopic details do we
+need to explicitly model, and how do we figure that out? For example, do we need
 to worry about all or even any of the steps that individual RNAPs go through
@@
 (statistical inferential) modeling can complement each other beautifully here.
-\end{itemize}}
 
-\fig{fig1:means_cartoons}A shows the qualitative picture of simple repression.
+Figure~\ref{fig1:means_cartoons}(A) shows the qualitative picture of simple repression that
+is implicit in the repressor-operator model.
 An operator, the binding site on the DNA for a repressor protein, may be found
 occupied by a repressor, in which case transcription is blocked from occurring.
-Or it may be found unoccupied, in which case RNA polymerase (RNAP) may bind and
+Alternatively, that binding site may be found unoccupied, in which case RNA polymerase (RNAP) may bind and
 transcription can proceed.
-
-The key assumption we make is that binding of repressor and RNAP in the promoter
+The key assumption we make in this simplest incarnation of
+the repressor-operator model is that binding of repressor and RNAP in the promoter
 region of interest is exclusive, meaning that one or the other may bind, but
@@
 its promoter sequence. Current evidence suggests this is sometimes, but not
-always the case~\mmnote{Cite that fig, from ??Nathan or Bill or Rob??, showing
-locations of repressor binding sites far upstream of promoter}, and it remains
+always the case, and it remains
 an interesting open question precisely how a repressor bound far upstream is
-able to repress transription. Suggestions include ``action-at-a-distance''
+able to repress transcription. 
+\marginpar{\it Cite that fig, from ??Nathan or Bill or Rob??, showing
+locations of repressor binding sites far upstream of promoter}
+Suggestions include ``action-at-a-distance''
 mediated by kinks in the DNA, formed when the repressor is bound, that prevent
 RNAP binding. Nevertheless, our modeling in this work is sufficiently
-coarse-grained that we simply assume exclusive binding and need not worry about
-such precise details.
+coarse-grained that we simply assume exclusive binding and leave explicit accounting
+of these details out of the problem.
 
-\mmnote{Add a notation aside somewhere that all association rates are written as zeroth-order rates.
-In other words, the concentration of the molecule is hidden inside, so for instance, $k_R^+$ is proportional to repressor copy number.}
+\begin{figure}%[h!]
+\centering
+\includegraphics[width=\textwidth]{../figures/fig1/fig1.ai}
+\caption{\textbf{An overview of the simple repression motif at the level of
+means.} (A) Schematic of the qualitative biological picture of simple repression. (B)
+A variety of possible mathematicized cartoons of simple repression, along
+with the effective parameter $\rho$ which subsumes all regulatory details of the
+architecture that do not directly involve the repressor. (C) The ``master curve'' to which
+all cartoons in (B) collapse.}
+\label{fig1:means_cartoons}
+\end{figure}
+
+The logic of the remainder of the paper is as follows.
+In section~\ref{section_02_means.tex}, we show how
+both thermodynamic models and kinetic models based upon
+the chemical master equation all culminate in the same underlying
+functional form for the fold-change in the average level of gene
+expression as shown in Figure~\ref{fig1:means_cartoons}(C).  Section~\ref{section_03_beyond_means.tex} goes
+beyond an analysis of the means by asking how the same models
+presented in Figure~\ref{fig1:means_cartoons} can be used
+to explore noise in gene expression. Of course, to make contact
+with experiment, all of these models must make a commitment
+to some numerical values for the key parameters found in each
+such model and in Section~\ref{section_04_bayesian_inference.tex}
+we explore the use of Bayesian inference to establish
+these parameters and to rigorously answer the question of how
+to discriminate between the different models.
+
+%
+%\mmnote{Key ideas, no particular order, that I haven't written down before:
+%\begin{itemize}
+%\item Our goal is to build phenomenological models of input-output functions of
+%genetic circuits. More precisely, we want phenomenology in the sense of
+%coarse-graining away atomistic detail, but still retaining biophysical meaning,
+%e.g., we don't want to coarse-grain as far as Hill functions. Why not? We are
+%motivated by studies like~\cite{Razo-Mejia2020}, for which Hill functions are
+%insufficient. Such work requires a quantitative theory of how biophysical
+%changes at the molecular level propagate to input-output functions at the
+%genetic circuit level. We want concepts, not mere facts. In particular a key
+%question is: at this level of coarse graining, what microscopic details do I
+%need to explicitly model, and how do we figure that out? For example, do I need
+%to worry about all or even any of the steps that individual RNAPs go through
+%each time they make a transcript? Turning the question around, can we see any
+%imprint of those processes in the available data? If the answer is no, then
+%those processes are irrelevant for our purposes. Forward modeling and inverse
+%(statistical inferential) modeling can complement each other beautifully here.
+%\end{itemize}}
+
+%Biochemistry tells us of a progression of steps, but theory tells us
+%that coarse graining can be exact

### 2020-05-20 doc/section_02_means.tex
RP report on May 19 morning
wrote abstract
went carefully through section 1 on introduction
went carefully through section 2 on means
addded many items to bib file

@@
 \section{Means}
+
+As noted in the previous section, there are two broad classes of models
+in play for computing the input-output functions of regulatory architectures 
+as shown in Figure~\ref{fig1:means_cartoons}.
+In both classes of model, the promoter is imagined to exist in a discrete
+set of states of occupancy, with each such state of occupancy
+accorded its own rate of transcription (and for many of these
+states, there is no transcription).  The models are probabilistic with
+each state assigned some probability and the overall rate of transcription
+given by 
+\begin{equation}
+\mbox{average rate of transcription} = \sum_i r_i p_i,
+\end{equation}
+where $i$ labels the distinct states, $p_i$ is the probability of
+the i$^{th}$ state and $r_i$ is the rate of transcription of
+that state.  Ultimately, the different models differ along several key
+axes: what states to consider and how to compute the probabilities of those
+states.
+
+The first class of models that are the focus of
+the present section on predicting the mean
+level of gene expression, sometimes known as thermodynamic models,
+invoke the tools of equilibrium statistical mechanics to compute the 
+probabilities~\cite{Ackers1982, Shea1985, Buchler2003,Vilar2003a,Vilar2003b, Bintu2005a,Bintu2005c, Gertz2009,Sherman2012, Saiz2013}.
+As seen in Figure~\ref{fig1:means_cartoons}(B), even within
+the class of thermodynamic models, we can make different commitments
+about the underlying microscopic states of the promoter.  Indeed,
+the list of options considered here does not at all exhaust the suite
+of different microscopic states we can assign to the promoter.
+
+The second class of models that allow us to access the mean
+expression use chemical master equations to compute the
+probabilities of the different microscopic states
+~\cite{Ko1991,Peccoud1995,Record1996,
+Kepler2001, Sanchez2008,Shahrezaei2008,Sanchez2011, Michel2010}.
+As seen in Figure~\ref{fig1:means_cartoons}(B), we consider a host of
+different nonequilibrium models, each of which will have its own 
+result for both the mean and noise in gene expression.
+
+%The second class of models based on chemical master equations compute the time-evolution
+%of the probability distribution for mRNA.
+
 \subsection{Fold-changes are indistinguishable across models}
 As a first stop on our search for the ``right'' model of simple repression, let
-us consider what we can learn from theory and measurements of fold-change in
+us consider what we can learn from theory and measurements on the
+average level of  gene expression in a population of cells.
+One experimental strategy that has been particularly useful (if incomplete since
+it misses out on gene expression dynamics) is to measure the  fold-change in
 mean expression. The fold-change is defined as
@@
 \end{equation}
-where angle brackets denote the averge over a population of cells and mean mRNA
+where angle brackets denote the average over a population of cells and mean mRNA
 $\langle m\rangle$ is viewed as a function of repressor copy number $R$. The
@@
 reasonable since the cells in the two conditions are identical except for the
-presence of the transcription factor, and it is difficult to imagine the
-transcription factor having any direct effect on translation.
+presence of the transcription factor, and the model assumes that the 
+transcription factor has no direct effect on translation.
 
@@
 work~\cite{Garcia2011a, Razo-Mejia2018, Chure2019}. Part of its utility in
-dissecting transcriptional regulation is its ratiometric nature, which is always
-easier than making an absolute measurement. Also, by measuring otherwise
+dissecting transcriptional regulation is its ratiometric nature, which removes many
+secondary effects that are present when making an absolute gene
+expression measurement.   Also, by measuring otherwise
 identical cells with and without a transcription factor present, any biological
@@
 
-\fig{fig1:means_cartoons} depicts a smorgasbord of mathematized cartoons for
-simple repression that have appeared in previous literature. For each cartoon,
+\fig{fig1:means_cartoons} depicts a smorgasbord of mathematicized cartoons for
+simple repression using both thermodynamic and kinetic models  that have appeared in previous literature. For each cartoon,
 we calculate the fold-change in mean gene expression as predicted by that model,
 deferring some algebraic details to the appendix. What we will find is that all
-cartoons collapse to a single master curve, shown in \fig{fig1:means_cartoons}c,
+cartoons collapse to a single master curve, shown in \fig{fig1:means_cartoons}(C),
 which contains just two parameters. We label the parameters $\Delta F_R$, an
@@
 which subsumes all details of transcription in the absence of repressors. We
-will offer some intuition for why this master curve exists and discuss why it is
-likely impossible to discriminate ``right'' from ``wrong'' cartoons given only
+will offer some intuition for why this master curve exists and discuss why at the level of
+the mean expression, we are unable   to discriminate ``right'' from ``wrong'' cartoons given only
 measurements of fold-changes in expression.
 
-\subsubsection{Equilibrium, 2-state}
+\subsubsection{The Two-State Equilibrium Model}
 In this simplest model, the promoter is idealized as existing in one of two
@@
 
-The fold-change is easily found as the ratio of mean mRNA with and without
-repressor. This gives
+The fold-change is  found as the ratio of mean mRNA with and without
+repressor as introduced in eqn.~\ref{eq:fc_def}. Invoking that definition results in
 \begin{equation}
@@
 \end{equation}
-which clearly matches the form of the master curve in~\eq{fig1:means_cartoons}
+which clearly matches the form of the master curve in Figure~\ref{fig1:means_cartoons}(C)
 with $\rho=1$.
@@
 active states. We will see this explicitly in the 3-state equilibrium model
-below, but perhaps surprising is that something analogous carries over even to
+below, but perhaps surprising is that an analogous result carries over even to
 the nonequilibrium models we consider later.
 
-\subsubsection{Equilibrium, 3-state}
+\subsubsection{The Three-State Equilibrium Model}
 Compared to the previous model, here we fine-grain the repressor unbound state
@@
 \frac{P}{N_{NS}}\mathrm{e}^{-\beta\Delta\varepsilon_P}$ as shown
-in~\fig{fig1:means_cartoons}.
+in~\fig{fig1:means_cartoons}.  Thus far,
+we see that the two thermodynamic models, despite making
+different coarse-graining commitments, result in the same functional
+form for the fold-change in mean gene expression.  We now
+explore how kinetic models fare when faced with computing
+the same observable.
 
-\begin{figure}%[h!]
-\centering
-\includegraphics[width=\textwidth]{../figures/fig1/fig1.ai}
-\caption{\textbf{An overview of the simple repression motif at the level of
-means.} (A) depicts the qualitative biological picture of simple repression. (B)
-shows a variety of possible mathematized cartoons of simple repression, along
-with the effective parameter $\rho$ which subsumes all regulatory details of the
-architecture excluding the repressor. (C) shows the ``master curve'' to which
-all cartoons in (B) collapse.}
-\label{fig1:means_cartoons}
-\end{figure}
-
-\subsubsection{Nonequilibrium model 1 - Poisson promoter}
-For our first kinetic model, we consider the simplest possible picture with only
+\subsubsection{The Poisson Promoter Nonequilibrium Model}
+For our first kinetic model, we  imitate the states considered in
+the Two-State Equilibrium Model and consider the simplest possible picture with only
 two states, repressor bound and unbound. This is exactly the model used for the
@@
 govern transitions between the two states. When the system is in the unbound
-state, transcriptions initiate at rate $r$, which represents a coarse-graining
-of all the downstream processes into a single effective rate. mRNA is again
-degraded at rate $\gamma$.
+state, transcription initiates at rate $r$, which represents a coarse-graining
+of all the downstream processes into a single effective rate. mRNA is 
+degraded at rate $\gamma$ as already exploited in the previous models.
 
@@
 repressor bound state with $m$ mRNA molecules present at time $t$, and similarly
-define $p_{m,U}(t)$ for the repressor unbound state. This model is governed by a
-master equation giving the time evolution of $p_{m,R}(t)$ and $p_{m,U}(t)$,
-which is
+define $p_{m,U}(t)$ for the repressor unbound state. This model is governed by coupled
+master equations giving the time evolution of $p_{m,R}(t)$ and 
+$p_{m,U}(t)$~\cite{Sanchez2008,Sanchez2011, Phillips2019}
+which we can write as 
 \begin{align}
@@
                 \\
-                &+ (m+1)\gamma p_{m+1,U}(t) - \gamma p_{m,U}(t).
+                &+ (m+1)\gamma p_{m+1,U}(t) - \gamma p_{m,U}(t),
 \label{eq:poisson_promoter_cme}
@@
 \end{align}
+where each term on the right corresponds to a transition between two
+states of the promoter.
 In each equation, the first two terms describe transitions between promoter
@@
 more complicated models yet to come, by re-expressing the master equation in
-vector form. The promoter states are collected into a vector and the rate
+vector form~\cite{Phillips2012}. The promoter states are collected into a vector and the rate
 constants are collected into matrices as
@@
 \end{equation}
-again matching the master curve of~\fig{fig1:means_cartoons} with $\rho=1$.
+again matching the master curve of~\fig{fig1:means_cartoons}(C) with $\rho=1$.
 
-\subsubsection{Nonequilibrium model 2 - RNAP bound/unbound states}
-Analogous to the three-state equilibrium model, this model fine-grains the
+\subsubsection{Nonequilibrium Model Two - RNAP Bound and Unbound States}
+Our second kinetic model mirrors the second equilibrium model of
+Figure~\ref{fig1:means_cartoons}(B) by fine-graining  the
 repressor unbound state of nonequilibrium model 1, resolving it into an empty
@@
 main subtlety arises since transcription initiation accompanies a promoter state
-change. This can be understood by analogy to $\matr{K}$. The off-diagonals and
-diagonals of $\matr{K}$ correspond to transitions arriving at or departing from,
-respectively, the promoter state of interest. If transcription initiations are
+change. This can be understood by analogy to $\matr{K}$. The off-diagonal and
+diagonal elements of $\matr{K}$ correspond to transitions arriving at or departing from,
+respectively, the promoter state of interest. If transcription initiation is 
 accompanied by promoter state changes, we must have separate matrices for
 arriving and departing transcription events since the arriving and departing
-transitions have different initial copy number of mRNA, unlike for $\matr{K}$
+transitions have different initial copy numbers of mRNA, unlike for $\matr{K}$
 where they are the same. The master equation for this model is
@@
 \end{align}
-which follows the master curve with $\rho = 1 + k_P^+/(k_P^- + r)$ as claimed.
+which follows the master curve of~\fig{fig1:means_cartoons}(C) with $\rho = 1 + k_P^+/(k_P^- + r)$ as claimed.
 
-\subsubsection{Nonequilibrium model 3 - Multistep transcription initiation and
-escape} One might reasonably complain that the first two ``nonequilibrium''
+\subsubsection{Nonequilibrium Model Three - Multistep Transcription Initiation and
+Escape}
+ One might reasonably complain that the first two ``nonequilibrium''
 models we have considered are straw men. Their steady states necessarily satisfy
@@
 
-Now we consider model 3 in~\eq{fig1:means_cartoons} which allows the possibility
+Now we consider model 3 in Figure~\ref{fig1:means_cartoons}(B) which allows the possibility
 of true nonequilibrium steady-state fluxes through the promoter states. We point
@@
 \end{equation}
-again closely analogous to model 2.
+again closely analogous to nonequilibrium model 2.
 
@@
 \end{align}
-surprisingly reducing to the master curve once again, with $\rho = 1 +
+surprisingly reducing to the master curve of~\fig{fig1:means_cartoons}(C) once again, with $\rho = 1 +
 \frac{k_P^+ (k_O + r)}{r(k_O + k_P^-)}$.
@@
 This example hints that an arbitrarily fine-grained model of downstream
-transcription steps may still be collapsed to the form of the master curve
-in~\fig{fig1:means_cartoons}B so long as the repressor binding is exclusive with
+transcription steps may still be collapsed to the form of the master curve for
+the means giving
+in~\fig{fig1:means_cartoons}(C) so long as the repressor binding is exclusive with
 transcriptionally active states. We offer this as a conjecture, and we suspect
@@
 Hill1966} might furnish a ``proof.'' Our focus here is not on full generality
-but rather to survey an assortment of plausible models that have been proposed
-in the literature for simple repression.
+but rather to survey an assortment of plausible models for simple
+repression (i.e. the repressor-operator model) that have been proposed
+in the literature.
 
-\subsubsection{Nonequilibrium model 4 - ``Active'' and ``inactive'' states}
-This model is at the core of the theory in~\cite{Razo-Mejia2020}. At a glance
+\subsubsection{Nonequilibrium Model Four - ``Active'' and ``Inactive'' States}
+This model, shown in Figure~\ref{fig1:means_cartoons}(B)   is at the core of the theory in~\cite{Razo-Mejia2020}. At a glance
 the cartoon for this model may appear very similar to model 2, and
@@
 
-\subsubsection{Nonequilibrium model 5 - Bursty promoter}
-We will have much more to say about this model in subsequent sections. For now
-we sketch it as an intuitive analog to model 1, with just two states, repressor
+\subsubsection{Nonequilibrium Model Five - Bursty Promoter}
+The final model we consider shown in Figure~\ref{fig1:means_cartoons}(B) 
+is an intuitive analog to model 1, with just two states, repressor
 bound or unbound, and transition rates between them of $k_R^+$ and $k_R^-$. In
 model 1, when in the unbound state, single mRNA transcripts are produced as a
-Poisson process with some characteristic rate $r$. This model by contrast
+Poisson process with some characteristic rate $r$. The current model by contrast
 produces, at some Poisson rate $k_i$, \textit{bursts} of transcripts with a
@@
 with $\rho=1$ also like model 1.
-
 We will also see later how this model emerges as a natural limit of model 4.
@@
 $k_R^+$ and $k_R^-$ from FISH measurements of mRNA population distributions.
+
+\mmnote{Add a notation aside somewhere that all association rates are written as zeroth-order rates.
+In other words, the concentration of the molecule is hidden inside, so for instance, $k_R^+$ is proportional to repressor copy number.}

### 2020-05-20 doc/section_02_means.tex
RP 9am Wed May 19
Finished going through section 2 on means.

Closing everything on my desktop until further notice.

@@
 
-\subsubsection{Summary and discussion of results across models for fold-changes in mean expression}
+\subsubsection{Summary of Results Ccross models for Fold-Changes in Mean Expression}
 What have we learned from our exploration of the models
 in~\fig{fig1:means_cartoons} and their predictions of fold-changes in mean
-expression levels? The takeaway is the existence of a master curve
-(\fig{fig1:means_cartoons}C) to which the fold-change predictions of all the
+expression levels? The key outcome of our analysis is the existence of a master curve
+ shown in \fig{fig1:means_cartoons}(C) to which the fold-change predictions of all the
 models collapse. This master curve is parametrized by only two effective
-parameters: $\Delta F_R$, which characterizes the repressor, and $\rho$, which
+parameters: $\Delta F_R$, which characterizes the number of repressors and
+their binding strength to the DNA, and $\rho$, which
 characterizes all other features of the promoter architecture. The key
-assumption underpinning this result is that zero transcription occurs when a
+assumption underpinning this result is that no transcription occurs when a
 repressor is bound to its operator. Note, however, that we are agnostic about
-the molecular mechanism which achieves this; steric effects are the obvious
+the molecular mechanism which achieves this; steric effects are one plausible
 mechanism, but, for instance, ``action-at-a-distance'' mediated by kinked DNA
@@
 
-Why does the master curve of~\fig{fig1:means_cartoons}C exist at all? This
+Why does the master curve of~\fig{fig1:means_cartoons}(C) exist at all? This
 brings to mind the deep questions posed in, e.g.,~\cite{Frank2013}
@@
 precision, inferring the potential microscopic contributions to $\rho$, buried
-inside a log no less, from fold-change measurements seems hopeless. As a
+inside a log no less, from fold-change measurements seems beyond reach. As a
 statistical inference problem it is entirely nonidentifiable, in the language
@@
 
-One might instead imagine a bottom-up modeling approach, where we mathematize a
+One might instead imagine a bottom-up modeling approach, where we mathematicize a
 model of what we hypothesize the important steps are and are not, use in vitro
@@
 approximation means $\frac{P}{N_{NS}}\exp(-\beta\Delta\epsilon_P) \ll 1$,
-meaning therefore $\rho\approx1$. This is suggestive, but how can we be sure
+meaning therefore $\rho\approx1$.  This approximation can be well
+justified on the basis of the number of RNAP per cell and the strength of
+binding of RNAP to DNA at weak promoters.  This is suggestive, but how can we be sure
 that $\rho$ is not, for instance, actually $10^2$ and that $\Delta F_R$ hides a
@@
 $k_R^+$ and $k_R^-$ from FISH measurements of mRNA population distributions.
-
-\mmnote{Add a notation aside somewhere that all association rates are written as zeroth-order rates.
+\marginpar{\it MM: Add a notation aside somewhere that all association rates are written as zeroth-order rates.
 In other words, the concentration of the molecule is hidden inside, so for instance, $k_R^+$ is proportional to repressor copy number.}

### 2020-05-20 code/figures/main/fig03.py
fused fig03 and 04 into single figure

@@ -1,0 +1,267 @@
+#%%
+import pickle
+from git import Repo #for directory convenience
+
+import numpy as np
+import scipy.stats as st
+import pandas as pd
+import arviz as az
+
+import matplotlib.pyplot as plt
+import bebi103.viz
+
+import srep.viz
+from srep.utils import load_FISH_by_promoter
+
+#%%
+repo = Repo("./", search_parent_directories=True)
+# repo_rootdir holds the absolute path to the top-level of our repo
+repo_rootdir = repo.working_tree_dir
+
+colors = srep.viz.color_selector('constit')
+srep.viz.plotting_style()
+#%%
+
+# Import necessary data
+
+repo = Repo("./", search_parent_directories=True)
+# repo_rootdir holds the absolute path to the top-level of our repo
+repo_rootdir = repo.working_tree_dir
+
+colors = srep.viz.color_selector('constit')
+srep.viz.plotting_style()
+
+pklfile = open(f"{repo_rootdir}/data/mcmc_samples/constit_post_inf.pkl", 'rb')
+all_samples = pickle.load(pklfile)
+pklfile.close()
+
+df_unreg, df_energies, = load_FISH_by_promoter(("unreg", "energies"))
+df_energies.sort_values('Energy (kT)', inplace=True)
+
+
+
+# %%
+
+# Generate composition figure to put all plot sections in a 
+# single mpl file
+
+# Initialize figure
+fig = plt.figure(constrained_layout=False, figsize=(6, 6))
+
+# Set gridspec for UV5 samples corner plot
+gs_a = fig.add_gridspec(
+    nrows=5, 
+    ncols=5, 
+    left=0.05, 
+    right=0.45,
+    top=1, 
+    bottom=0.55,
+    wspace=0,
+    hspace=0
+)
+
+ax_a1 = fig.add_subplot(gs_a[0, :-1])
+ax_a2 = fig.add_subplot(gs_a[1:, :-1])
+ax_a3 = fig.add_subplot(gs_a[1:, -1])
+# Set shared axes
+ax_a1.get_shared_x_axes().join(ax_a2, ax_a1)
+ax_a3.get_shared_y_axes().join(ax_a2, ax_a3)
+# turn axis off for marginal plots
+ax_a1.get_xaxis().set_visible(False)
+ax_a1.get_yaxis().set_visible(False)
[…]

### 2020-05-20 code/figures/main/fig03.py
add text for k_i scaling

@@ -258,6 +258,14 @@ for i, promoter in enumerate(df_energies.Name):
 guide_x = np.linspace(-5.5,-2)
 guide_y = np.exp(-guide_x)/50
 ax_d.plot(guide_x, guide_y, 'k--', label='predicted \n scaling')
+# Add text for scaling
+ax_d.text(
+    0.6, 
+    0.6, 
+    r"$\log(k_i) \sim \Delta\epsilon_r$",
+    transform=ax_d.transAxes,
+    rotation=-45,
+)
 
 ax_d.set_xlabel(r'Binding energy $(k_BT)$')
 ax_d.set_yscale("log")

### 2020-05-20 code/figures/main/fig03_v2.py
updating fig03 and adding alternative version

@@ -1,0 +1,306 @@
+#%%
+import pickle
+from git import Repo #for directory convenience
+
+import numpy as np
+import scipy.stats as st
+import pandas as pd
+import arviz as az
+
+import matplotlib.pyplot as plt
+import seaborn as sns
+import bebi103.viz
+
+import srep.viz
+from srep.utils import load_FISH_by_promoter
+
+#%%
+repo = Repo("./", search_parent_directories=True)
+# repo_rootdir holds the absolute path to the top-level of our repo
+repo_rootdir = repo.working_tree_dir
+
+colors = srep.viz.color_selector('constit')
+srep.viz.plotting_style()
+#%%
+
+# Import necessary data
+
+repo = Repo("./", search_parent_directories=True)
+# repo_rootdir holds the absolute path to the top-level of our repo
+repo_rootdir = repo.working_tree_dir
+
+colors = srep.viz.color_selector('constit')
+srep.viz.plotting_style()
+
+pklfile = open(f"{repo_rootdir}/data/mcmc_samples/constit_post_inf.pkl", 'rb')
+all_samples = pickle.load(pklfile)
+pklfile.close()
+
+df_unreg, df_energies, = load_FISH_by_promoter(("unreg", "energies"))
+df_energies.sort_values('Energy (kT)', inplace=True)
+
+
+
+# %%
+
+# Generate composition figure to put all plot sections in a 
+# single mpl file
+
+# Initialize figure
+fig = plt.figure(constrained_layout=False, figsize=(6, 6))
+
+# Set gridspec for UV5 samples corner plot
+gs_a = fig.add_gridspec(
+    nrows=5, 
+    ncols=5, 
+    left=0.05, 
+    right=0.45,
+    top=1, 
+    bottom=0.55,
+    wspace=0,
+    hspace=0
+)
+
+ax_a1 = fig.add_subplot(gs_a[0, :-1])
+ax_a2 = fig.add_subplot(gs_a[1:, :-1])
+ax_a3 = fig.add_subplot(gs_a[1:, -1])
+# Set shared axes
+ax_a1.get_shared_x_axes().join(ax_a2, ax_a1)
+ax_a3.get_shared_y_axes().join(ax_a2, ax_a3)
+# turn axis off for marginal plots
+ax_a1.get_xaxis().set_visible(False)
+ax_a1.get_yaxis().set_visible(False)
[…]

### 2020-05-20 code/figures/main/fig03.py
final version of fig03 ready

@@ -8,6 +8,8 @@ import pandas as pd
 import arviz as az
 
 import matplotlib.pyplot as plt
+import matplotlib
+import seaborn as sns
 import bebi103.viz
 
 import srep.viz
@@ -149,8 +151,6 @@ ax_a3.hist(alpha_samples,
           color="black",
           orientation="horizontal",
 )
-# set_xticks([3.25, 3.5, 3.75], minor=True)
-# ax_a1.yaxis.grid(True, which='minor')
 
 # (B)
 df_UV5 = df_unreg[df_unreg["experiment"] == "UV5"]
@@ -158,7 +158,7 @@ n_samples = (
     all_samples["UV5"].posterior_predictive.dims["chain"]
     * all_samples["UV5"].posterior_predictive.dims["draw"]
     )
-ptiles = (95, 75, 50, 25)
+ptiles = (95,)#, 75, 50, 25)
 
 # first compute params in analytical gamma-Poisson posterior.
 # ignore prior b/c data completely(!) overwhelms it.
@@ -201,17 +201,14 @@ ax_b.set_ylabel('ECDF')
 ax_b.set_xlim(right=60)
 
 # (C)
-cwheel_set = (
-    'red', 'blue', 'purple',
-    'green', 'brown', 'black',
-    'yellow', 'magenta', 'white'
-    )
-cwheel = 2 * colors
-# cwheel = 2 * cwheel_set
-lwheel_nested = (9*('-',), 9*('--',))
-lwheel = [item for sublist in lwheel_nested for item in sublist]
-mwheel_nested = (9*('o',), 9*('^',))
-mwheel = [item for sublist in mwheel_nested for item in sublist]
+# Set function to normalize colors to energy range
+
+col_norm = matplotlib.colors.Normalize(
+    vmin=df_energies["Energy (kT)"].min() - 2,
+    vmax=df_energies["Energy (kT)"].max() + 2,
+)
+# Chose colormap for energies
+cmap = matplotlib.cm.get_cmap('magma')
 
 # # loop thru df, not all_samples keys, so we get deterministic order!
 for i, promoter in enumerate(df_energies.Name):
@@ -223,22 +220,28 @@ for i, promoter in enumerate(df_energies.Name):
     ax_c.loglog(
         x_contour[0],
         y_contour[0],
-        lwheel[i], #linestyle
         label=promoter,
[…]

### 2020-05-20 code/figures/main/fig03_v2.py

@@ -8,7 +8,6 @@ import pandas as pd
 import arviz as az
 
 import matplotlib.pyplot as plt
-import seaborn as sns
 import bebi103.viz
 
 import srep.viz
@@ -214,8 +213,6 @@ lwheel = [item for sublist in lwheel_nested for item in sublist]
 mwheel_nested = (9*('o',), 9*('^',))
 mwheel = [item for sublist in mwheel_nested for item in sublist]
 
-colors = sns.color_palette("viridis", n_colors=len(df_energies))
-col_dict = dict(zip(df_energies["Name"], colors))
 # # loop thru df, not all_samples keys, so we get deterministic order!
 for i, promoter in enumerate(df_energies.Name):
     alpha_samples = all_samples[promoter].posterior.alpha.values.flatten()
@@ -229,14 +226,8 @@ for i, promoter in enumerate(df_energies.Name):
         lwheel[i], #linestyle
         label=promoter,
         linewidth=1.0,
-        color=col_dict[promoter],
-    )
-    ax_c.annotate(
-        f"{i + 1}",
-        (np.mean(x_contour[0]),
-        np.mean(y_contour),),
-        fontsize=7,
-    )
+        color=cwheel[i],
+        )
 ax_c.set_xlim(right=1.2e1)
 ax_c.set_ylim(top=1e1)
 ax_c.set_ylabel(r'$k_i$ (bursts per mRNA lifetime)')
@@ -261,28 +252,8 @@ for i, promoter in enumerate(df_energies.Name):
         marker=mwheel[i],
         markersize=4.0,
         label=promoter,
-        color=col_dict[promoter],
+        color=cwheel[i]
         )
-    ax_d.annotate(
-        f"{i + 1}",
-        (df_energies[df_energies.Name == promoter]["Energy (kT)"],
-        ptile_med,),
-        fontsize=7,
-    )
-# Add colorbar
-im = ax_d.scatter(
-    df_energies["Energy (kT)"].values,
-    range(0, 18),
-    c=df_energies["Energy (kT)"].values, 
-    vmin=df_energies["Energy (kT)"].min(),
-    vmax=df_energies["Energy (kT)"].max(),
-    s=35,
-    cmap=plt.cm.get_cmap('viridis')
-)
-im.set_visible(False)
-cbar = fig.colorbar(im, ax=ax_d, pad=0.01)
[…]

### 2020-05-21 doc/section_01_introduction.tex
MRM 20200520 pass through abstract and intro

@@
 
-Gene expression presides over much of the most important dynamism 
-of living organisms.   The level of expression of batteries of
-different genes is altered as a result of spatiotemporal cues that integrate
-chemical, mechanical and other types of signals.  The original repressor-operator model
-conceived by Jacob and Monod in the context of bacterial
-metabolism has now been transformed into the much
-broader subject of gene regulatory networks in living organisms
-of all kinds~\cite{Jacob1961, Britten1969, Ben-TabouDe-Leon2007}.  One of the remaining outstanding challenges
-to have emerged in the genomic era is our continued
-inability to predict the regulatory consequences of
-different regulatory architectures, stemming first and foremost
-from our ignorance about what those architectures even are with
-more than 60\% of the genes even in an ostensibly well understood organism
-such as  {\it E. coli} with no regulatory
-insights at all~\cite{Rydenfelt2014-2,Belliveau2018,Ghatak2019, Santos_Zavaleta2019}.
-But even once we have established the identity of key transcription
-factors and their binding sites of a given promoter architecture,  there remains the predictive challenge
-of understanding its input-output properties, an objective that can be met
-by a myriad of approaches using the tools of statistical physics~\cite{Ackers1982, Shea1985, Buchler2003,Vilar2003a,Vilar2003b, Bintu2005a,Bintu2005c, Gertz2009,Sherman2012, Saiz2013,
-Ko1991,Peccoud1995,Record1996,
-Kepler2001, Sanchez2008,Shahrezaei2008,Sanchez2011, Michel2010}.  One root
-to such predictive understanding is to focus on the simplest
-regulatory architecture and to push the theory-experiment
-dialogue as far and as hard as it can be pushed~\cite{Phillips2019}.  If we
-demonstrate that we can pass that test by successfully predicting
-both the means and variance in gene expression, then that provides
-a more solid foundation upon which to launch into 
-more complex problems.
+Gene expression presides over much of the most important dynamism of living
+organisms.   The level of expression of batteries of different genes is altered
+as a result of spatiotemporal cues that integrate chemical, mechanical and other
+types of signals.  The original repressor-operator model conceived by Jacob and
+Monod in the context of bacterial metabolism has now been transformed into the
+much broader subject of gene regulatory networks in living organisms of all
+kinds~\cite{Jacob1961, Britten1969, Ben-TabouDe-Leon2007}.  One of the remaining
+outstanding challenges to have emerged in the genomic era is our continued
+inability to predict the regulatory consequences of different regulatory
+architectures. This challenge stems first and foremost from our ignorance about
+what those architectures even are with more than 60\% of the genes even in an
+ostensibly well understood organism such as  {\it E. coli} with no regulatory
+insights at all~\cite{Rydenfelt2014-2,Belliveau2018,Ghatak2019,
+Santos_Zavaleta2019}. But even once we have established the identity of key
+transcription factors and their binding sites of a given promoter architecture,
+there remains the predictive challenge of understanding its input-output
+properties, an objective that can be met by a myriad of approaches using the
+tools of statistical physics~\cite{Ackers1982, Shea1985,
+Buchler2003,Vilar2003a,Vilar2003b, Bintu2005a,Bintu2005c, Gertz2009,Sherman2012,
+Saiz2013, Ko1991,Peccoud1995,Record1996, Kepler2001,
+Sanchez2008,Shahrezaei2008,Sanchez2011, Michel2010}.  One root to such
+predictive understanding is to focus on the simplest regulatory architecture and
+to push the theory-experiment dialogue as far and as hard as it can be
+pushed~\cite{Phillips2019}.  If we demonstrate that we can pass that test by
+successfully predicting both the means and variance in gene expression, then
+that provides a more solid foundation upon which to launch into more complex
+problems.
 
-To that end, in this paper we examine a wide variety of distinct models
-for the simple repression regulatory architecture.  All of these
-models coarse-grain away some of the important microscopic
-features of these architectures that have been elucidated
-by generations of genetics, molecular biology and biochemistry.  
-Our aim is to understand the underlying principles of such
-coarse-graining such that future models of regulatory response
-can serve the powerful predictive role needed to take synthetic
-biology from what is often a brilliant exercise in enlightened 
-empiricism and recast it as the same kind of rational design process
-we are used to in other branches of engineering.  
-More precisely, we want phenomenology in the sense of
-coarse-graining away atomistic detail, but still retaining biophysical meaning.
-For example, we are not satisfied with the strictly phenomenological
-approach offered by 
- Hill functions since we are motivated by studies like~\cite{Razo-Mejia2020}, for which Hill functions are
-clearly insufficient since each new situation
-requires a completely new set of parameters. Such work requires a quantitative theory of how biophysical
-changes at the molecular level propagate to input-output functions at the
-genetic circuit level. We want concepts, not mere facts. In particular a key
-question is: at this level of coarse graining, what microscopic details do we
-need to explicitly model, and how do we figure that out? For example, do we need
-to worry about all or even any of the steps that individual RNAPs go through
-each time they make a transcript? Turning the question around, can we see any
-imprint of those processes in the available data? If the answer is no, then
-those processes are irrelevant for our purposes. Forward modeling and inverse
-(statistical inferential) modeling can complement each other beautifully here.
+To that end, in this paper we examine a wide variety of distinct models for the
+simple repression regulatory architecture. This genetic architecture consists of
+a DNA promoter regulated by a transcriptional repressor that binds to a single
+binding site \cite{Garcia2011a}. All of the proposed models models coarse-grain
+away some of the important microscopic features of this architecture that have
+been elucidated by generations of geneticists, molecular biologists and
+biochemists.
+\mrm{Our aim is to understand the underlying principles of such coarse-graining
+in order to increase the predictive power of these models. In doing so we hope
+future models of regulatory response will be able to serve the powerful
+predictive role needed to take synthetic biology from a brilliant exercise in
+enlightened empiricism to a rational design framework as any other branch of
+engineering.}
+% Our aim is to understand the underlying principles of such
+% coarse-graining such that future models of regulatory response can serve the
+% powerful predictive role needed to take synthetic biology from what is often a
+% brilliant exercise in enlightened empiricism and recast it as the same kind of
+% rational design process we are used to in other branches of engineering. 
+More precisely, we want phenomenology in the sense of coarse-graining away
+atomistic detail, but still retaining biophysical meaning. For example, we are
+not satisfied with the strictly phenomenological approach offered by the
+commonly used Hill functions. Studies like~\cite{Razo-Mejia2018} have
+demonstrated that Hill functions are clearly insufficient since each new
+situation requires a completely new set of parameters. Such work requires a
+quantitative theory of how biophysical changes at the molecular level propagate
+to input-output functions at the genetic circuit level. We want concepts, not
+mere facts. In particular a key question is: at this level of coarse graining,
+what microscopic details do we need to explicitly model, and how do we figure
+that out? For example, do we need to worry about all or even any of the steps
+that individual RNAPs go through each time they make a transcript? Turning the
+question around, can we see any imprint of those processes in the available
+data? If the answer is no, then those processes are irrelevant for our purposes.
+Forward modeling and inverse (statistical inferential) modeling are necessary to 
+tackles such questions.
 
-Figure~\ref{fig1:means_cartoons}(A) shows the qualitative picture of simple repression that
-is implicit in the repressor-operator model.
-An operator, the binding site on the DNA for a repressor protein, may be found
-occupied by a repressor, in which case transcription is blocked from occurring.
-Alternatively, that binding site may be found unoccupied, in which case RNA polymerase (RNAP) may bind and
-transcription can proceed.
-The key assumption we make in this simplest incarnation of
-the repressor-operator model is that binding of repressor and RNAP in the promoter
-region of interest is exclusive, meaning that one or the other may bind, but
-never may both be simultaneously bound. It is often imagined that when the
-repressor is bound to its operator, RNAP is sterically blocked from binding to
-its promoter sequence. Current evidence suggests this is sometimes, but not
-always the case, and it remains
-an interesting open question precisely how a repressor bound far upstream is
-able to repress transcription. 
-\marginpar{\it Cite that fig, from ??Nathan or Bill or Rob??, showing
-locations of repressor binding sites far upstream of promoter}
-Suggestions include ``action-at-a-distance''
-mediated by kinks in the DNA, formed when the repressor is bound, that prevent
-RNAP binding. Nevertheless, our modeling in this work is sufficiently
-coarse-grained that we simply assume exclusive binding and leave explicit accounting
-of these details out of the problem.
+Figure~\ref{fig1:means_cartoons}(A) shows the qualitative picture of simple
+repression that is implicit in the repressor-operator model. An operator, the
+binding site on the DNA for a repressor protein, may be found occupied by a
+repressor, in which case transcription is blocked from occurring. Alternatively,
+that binding site may be found unoccupied, in which case RNA polymerase (RNAP)
+may bind and transcription can proceed. The key assumption we make in this
+simplest incarnation of the repressor-operator model is that binding of
+repressor and RNAP in the promoter region of interest is exclusive, meaning that
+one or the other may bind, but never may both be simultaneously bound. It is
+often imagined that when the repressor is bound to its operator, RNAP is
+sterically blocked from binding to its promoter sequence. Current evidence
+suggests this is sometimes, but not always the case, and it remains an
+interesting open question precisely how a repressor bound far upstream is able
+to repress transcription. \marginpar{\it Cite that fig, from ??Nathan or Bill or
+Rob??, showing locations of repressor binding sites far upstream of promoter}
+Suggestions include ``action-at-a-distance'' mediated by kinks in the DNA,
+formed when the repressor is bound, that prevent RNAP binding. Nevertheless, our
+modeling in this work is sufficiently coarse-grained that we simply assume
+exclusive binding and leave explicit accounting of these details out of the
+problem.
 
-\begin{figure}%[h!]
+\afterpage{\clearpage}
+\begin{figure}[p]
 \centering
-\includegraphics[width=\textwidth]{../figures/fig1/fig1.ai}
+\includegraphics[width=\textwidth]{../figures/main/fig01.pdf}
 \caption{\textbf{An overview of the simple repression motif at the level of
-means.} (A) Schematic of the qualitative biological picture of simple repression. (B)
-A variety of possible mathematicized cartoons of simple repression, along
-with the effective parameter $\rho$ which subsumes all regulatory details of the
-architecture that do not directly involve the repressor. (C) The ``master curve'' to which
-all cartoons in (B) collapse.}
+means.} (A) Schematic of the qualitative biological picture of the simple
+repression genetic architecture. (B) and (C) A variety of possible
+mathematicized cartoons of simple repression, along with the effective parameter
+$\rho$ which subsumes all regulatory details of the architecture that do not
+directly involve the repressor. (B) Simple repression models from an equilibrium
+perspective. (C) Equivalent models cast in chemical kinetics language. (D) The
+``master curve'' to which all cartoons in (B) and (C) collapse.}
 \label{fig1:means_cartoons}
@@
 
-The logic of the remainder of the paper is as follows.
-In section~\ref{section_02_means.tex}, we show how
-both thermodynamic models and kinetic models based upon
-the chemical master equation all culminate in the same underlying
-functional form for the fold-change in the average level of gene
-expression as shown in Figure~\ref{fig1:means_cartoons}(C).  Section~\ref{section_03_beyond_means.tex} goes
-beyond an analysis of the means by asking how the same models
-presented in Figure~\ref{fig1:means_cartoons} can be used
-to explore noise in gene expression. Of course, to make contact
-with experiment, all of these models must make a commitment
-to some numerical values for the key parameters found in each
-such model and in Section~\ref{section_04_bayesian_inference.tex}
-we explore the use of Bayesian inference to establish
-these parameters and to rigorously answer the question of how
-to discriminate between the different models.
+The logic of the remainder of the paper is as follows. In
+section~\ref{section_02_means}, we show how both thermodynamic models and
+kinetic models based upon the chemical master equation all culminate in the same
+underlying functional form for the fold-change in the average level of gene
+expression as shown in Figure~\ref{fig1:means_cartoons}(D).
+Section~\ref{section_03_beyond_means.tex} goes beyond an analysis of the mean
+gene expression by asking how the same models presented in
+Figure~\ref{fig1:means_cartoons}(C) can be used to explore noise in gene
+expression. To make contact with experiment, all of these models must make a
+commitment to some numerical values for the key parameters found in each such
+model. Therefore in Section~\ref{section_04_bayesian_inference.tex} we explore
+the use of Bayesian inference to establish these parameters and to rigorously
+answer the question of how to discriminate between the different models.
 
-%
 %\mmnote{Key ideas, no particular order, that I haven't written down before:

### 2020-05-21 doc/appendix_03_regulated_promoters.tex
Added to constit prom Bayesian inference + minor file reorg

@@
 % !TEX root = ./busty_transcription.tex
-\section{Generating function solution for bursty promoters}
+\section{Bursty promoter models - generating function solutions and numerics}
+\label{sec:gen_fcn_appdx}
 
@@
 which is exactly the negative binomial's generating function, as expected.
+Now we may recover the mRNA probability distribution of interest
+by differentiating the generating function, or by consulting a
+standard reference\mmnote{maybe van Kampen, check},
+from which we find that
+\begin{equation}
+p(m) = \frac{\Gamma(m+\lambda)}{\Gamma(m+1)\Gamma(\lambda)}
+        \theta^\lambda (1-\theta)^m.
+\end{equation}
+Note that the ratio of gamma functions is often expressed as a
+binomial coefficient, but since $\lambda$ may be non-integer,
+this would be ill-defined.
+Re-expressing this exclusively in our variables of interest,
+burst rate $\lambda$ and mean burst size $b$, we have
+\begin{equation}
+p(m) = \frac{\Gamma(m+\lambda)}{\Gamma(m+1)\Gamma(\lambda)}
+        \left(\frac{1}{1+b}\right)^\lambda
+        \left(\frac{b}{1+b}\right)^m.
+\label{eq:nbinom_deriv_final}
+\end{equation}
 

### 2020-05-21 doc/section_04_bayesian_inference.tex
Added to constit prom Bayesian inference + minor file reorg

@@
 It will be instructive, however, to first consider the Poisson promoter, model 1
-in~\fig{fig:constit_cartoons}. This is a convenient proxy for models 2 and 3,
-whose exact steady-state mRNA distributions would be tedious to calculate and we
-will not explicitly consider. But since they have Fano factors $\nu\le 1$, they
+in~\fig{fig:constit_cartoons}. As we alluded to earlier, since the Poisson
+distribution has its Fano factor $\nu$ strictly equal to 1,
+and all of the observed data has Fano factor $\nu>1$, we might
+already suspect that this model is incapable of fitting the data.
+We will verify that this in fact the case. As a consequence
+of ruling out model 1, models 2 and 3 will also be ruled out.
+Why? Again all the data have $\nu>1$, meaning the mRNA distributions
+are overdisperse compared with a Poisson distribution,
+while models 2 and 3 have Fano factors $\nu\le 1$ meaning they
+are underdisperse relative to Poisson. So models 2 and 3
 can fare no better than the simple Poisson model. We will also not explicitly
@@
 \begin{equation}
-p_{m,U} = \frac{\lambda^m e^{-\lambda}}{m!}
+p_U(m) = \frac{\lambda^m e^{-\lambda}}{m!}
 \end{equation}
 with $\lambda=r/\gamma$. This steady state distribution defines our likelihood
-$p(m\mid\lambda)$, i.e.,
+$p(m\mid\lambda)$, i.e.
 \begin{equation}
@@
 is the probability of observing a single cell with $m$ mRNAs given a known value
-of $\lambda$.
+of $\lambda$, written to make the conditioning explicit.
 
@@
 \begin{equation}
-D\mid\lambda \sim \text{Poisson}(\lambda),
+D \sim \text{Poisson}(\lambda),
 \end{equation}
-which is read as the data $D$, conditioned on a value of $\lambda$, is Poisson
+which is read as the data $D$ is Poisson
 distributed with mean $\lambda$. The $\sim$ operator literally denotes ``is
 distributed according to.''
+The conditioning on $\lambda$ is no longer explictily notated,
+only implied by its appearance on the right hand side.
 
@@
 little here, as long as it does not exclude the true value of $\lambda$. A
-convenient choice for our Poisson likelihood is its conjugate prior, the gamma
-distribution. Although they only exist for certain special likelihoods,
-conjugate priors have the nice feature that the posterior takes the same
-functional form as the prior, with updated parameters. This makes calculations
+convenient choice for our problem is to use a \textit{conjugate} prior.
+A conjugate prior is a special prior that causes the posterior to
+have the same functional form as the prior, simply with updated
+model parameters. This makes calculations
 analytically tractable and also offers a nice interpretation of the inference
 procedure as updating our knowledge about the model parameters.
+This makes conjugate priors very useful when they exist.
+The caveat is that conjugate priors only exist for a very limited
+number of likelihoods, mostly with only one or two model
+parameters, so in almost all other Bayesian inference problems,
+we must tackle the posterior numerically.
 
+But, for the problem at hand, a conjugate prior does in fact exist.
+For a Poisson likelihood of identical and identically distributed
+data, the conjugate prior is a gamma distribution, as
+can be looked up in, e.g.,~\cite{Gelman2013}, Section 2.6.
 Putting a gamma prior on $\lambda$ introduces two new parameters $\alpha$ and
@@
 \begin{equation}
-\lambda\mid\alpha,\beta \sim \text{Gamma}(\alpha, \beta)
+\lambda \sim \text{Gamma}(\alpha, \beta)
 \end{equation}
@@
 Since we did use a conjugate prior, we may simply look up our posterior in any
-standard table of conjugate priors-likelihood pairs, from which we find that
+standard reference such as~\cite{Gelman2013}, Section 2.6,
+from which we find that
 \begin{equation}
-\lambda\mid D,\alpha,\beta
+\lambda
 \sim \text{Gamma}\left(\alpha + \bar{m}N, \beta + N\right),
@@
 constitutive strains in~\cite{Jones2014}, so $\bar{m}N \gtrsim 10^2$. Therefore
-as we suspected, our prior parameters are completely overwhelmed by the data. In
+as we suspected, our prior parameters are completely overwhelmed by the data.
+The prior behaves, in a sense, like $\beta$ extra ``data points''
+with a mean value of $(\alpha-1)/\beta$~\cite{Gelman2013}, which
+gives us some intuition for how much data is needed to overwhelm
+the prior in this case: enough data $N$ such that $\beta\ll N$
+and $\alpha/\beta \ll \bar{m}$. In
 fact, $\bar{m}N$ and $N$ are so large that we can, to an excellent
@@
 \begin{equation}
-\lambda\mid D
+\lambda
 \sim \text{Gamma}\left(\alpha + \bar{m}N, \beta + N\right)
@@
 \begin{equation}
-\lambda\mid D_{UV5}
+\lambda
 \sim \text{Normal}\left(18.7, 0.08\right),
@@
 datasets. In general for higher dimensional datasets, much of the challenge is
-in merely designed good visualizations that can actually show if synthetic and
+in merely designing good visualizations that can actually show if synthetic and
 observed data are similar or not.
@@
 visualization for 1D data is ECDFs of the quantiles, as shown for our Poisson
-model in~\fig{fig:uv5_post}. In this example, the median for each possible mRNA
-count is shown as a dark green line, while shaded bands of progressively lighter
-green contain 25\%, 50\%, 75\%, and 95\% of the posterior predictive samples.
+model in~\fig{fig:constit_post_full}.
+In this example, the median for each possible mRNA
+count is shown as a dark green line, while shaded bands of lighter
+green contain 95\% of the posterior predictive samples.
 Plotting quantiles in this way gives us a sense of the range of data we might
@@
 come from this Poisson generative model. The other model with a PPC plotted
-in~\fig{fig:uv5_post}, bursty model 5 with a negative binomial steady-state mRNA
+in~\fig{fig:constit_post_full}, bursty model 5 with
+a negative binomial steady-state mRNA
 distribution, looks like a good candidate by eye. We cover it in more detail
@@
 
+\subsubsection{Nonequilibrium Model Five - Bursty promoter}
+Let us now consider the problem of parameter inference from FISH data
+for model five from~\fig{fig1:means_cartoons}(C). As derived
+in Appendix~\ref{sec:gen_fcn_appdx}, the steady-state mRNA distribution
+in this model is a negative binomial distribution,
+given by~\eq{eq:nbinom_deriv_final}, which is
+\begin{equation}
+p(m) = \frac{\Gamma(m+k_i)}{\Gamma(m+1)\Gamma(k_i)}
+        \left(\frac{1}{1+b}\right)^{k_i}
+        \left(\frac{b}{1+b}\right)^m,
+\end{equation}
+where $b$ is the mean burst size and $k_i$ is the burst rate
+nondimensionalized by the mRNA degradation rate $\gamma$.
+As sketched earlier, the story of this distribution is of
+geometrically-distributed bursts of mRNA, where the arrival of bursts is a
+Poisson process with rate $k_i$ and the mean size of a burst is $b$.
+
+As for the Poisson promoter model, this expression for the steady-state
+mRNA distribution is exactly the likelihood we want to use in Bayes theorem.
+Again denoting the single-cell mRNA count data as $D=\{m_1, m_2,\dots, m_N\}$,
+here Bayes theorem takes the form
+\begin{equation}
+p(k_i, b \mid D) \propto p(D\mid k_i,b)p(k_i, b).
+\end{equation}
+We already have our likelihood so we only need to choose priors
+on $k_i$ and $b$. For the datasets from~\cite{Jones2014} that we
+are analyzing, as for the Poisson promoter model above we are
+still data-rich so the prior's influence remains weak, but not
+nearly as weak because the dimensionality of our model has
+increased from one to two.
+
+We follow the guidance of~\cite{Gelman2013}, Section 2.9 in
+opting for weakly-informative priors on $k_i$ and $b$ (conjugate
+priors do not exist for this problem), and we find
+``street-fighting estimates''~\cite{Mahajan2010} to be an ideal
+way of constructing such priors. The idea of weakly informative
+priors is to allow all remotely plausible values of model
+parameters while excluding the completely absurd or unphysical.
+
+Consider $k_i$. The fastest known bacterial promoters control
+rRNA genes and initiate transcripts no faster than $\sim
+1/\text{sec}$. It would be exceedingly strange if any of the
+constitutive promoters from~\cite{Jones2014} were stronger than
+that, so we can take that as an upper bound. For a lower bound,
+if transcripts are produced too rarely, there would be nothing to
+see with FISH. The datasets for each strain contain of order
+$10^3$ cells, and if the $\langle m \rangle = k_i b/\gamma
+\lesssim 10^{-2}$, then the total number of expected mRNA
+detections would be single-digits or less and we would have
+essentially no data on which to carry out inference. So assuming
+$b$ is not too different from 1, justified next, and an mRNA
+lifetime of $\gamma^{-1}\sim 3-5~\text{min}$, this gives us soft
+bounds on $k_i/\gamma$ of perhaps $10^{-2}$ and $3\times 10^1$.
+
+Next consider mean burst size $b$. This parametrization of the
+geometric distribution allows bursts of size zero (which could
+representing aborted transcripts and initiations), but it would
+be quite strange for the mean burst size $b$ to be below
+$\sim10^{-1}$, for which nearly all bursts would be of size zero
+or one. For an upper bound, if transcripts are initiating at a
+rate somewhat slower than rRNA promoters, then it would probably
+take a time comparable to the lifetime of an mRNA to produce a
+burst larger than 10-20 transcripts, which would invalidate the
+approximation of the model that the duration of bursts are
+instantaneous compared to other timescales in the problem. So we
+will take soft bounds of $10^{-1}$ and $10^1$ for $b$.
+
+Note that the natural scale for these ``street-fighting estimates''
+was a log scale. This is commonly the case that our prior sense
+of reasonable and unreasonable parameters is set on a log scale.
+A natural way to enforce these soft bounds is therefore to use a
+lognormal prior distribution, with the soft bounds set $\pm2$
+standard deviations from the mean.
+
+With this, we are ready to write our full generative model as
+\begin{equation}
+\begin{split}
+\ln k_i \sim \text{Normal}(-0.5, 2)
+\\
+\ln b \sim \text{Normal}(0.5, 1)
+\\
+m \sim \text{NBinom}(k_i, b).
+\end{split}
+\end{equation}
+We carried out MCMC sampling on the posterior of this model, starting
+with the constitutive \textit{lacUV5} dataset from~\cite{Jones2014}.
+The resulting MCMC samples are shown in~\fig{fig:constit_post_full}(A).
+In constrast to the active/inactive constitutive model considered
+in~\cite{Razo-Mejia2020} (nonequilibrium model four
+in~\ref{fig:constit_cartoons}), this model is well-identified
+with both parameters determined to a fractional uncertainty of 5-10\%.
+The strong correlation reflects the fact that their product sets
+the mean of the mRNA distribution, which is tightly constrained
+by the data, but there is weak ``sloppiness''~\cite{Transtrum2015}
+along a set of values with a similar product.
+
+Flush with success having found the model's posterior to be
+well-identified, the next step was posterior predictive sampling.
+The only procedural difference compared to the Poisson promoter
+model is that, since we have no analytical posterior expression
+but only samples from the posterior, we generate a synthetic
+dataset (i.e., posterior predictive samples) for each posterior
+sample from the MCMC sampling. \fig{fig:constit_post_full}(B)
+shows the resulting predictive ECDF; similarly to the Poisson
+promoter model, the median posterior predictive ECDF is plotted
+as a dark blue line, while a lighter blue shaded region
+encompasses 95\% of the posterior predictive samples. Unlike the
+Poisson promoter model, the experimental ECDF closely tracks the
+posterior predictive ECDF, indicating this model is actually able
+to generate the observed data and increasing our confidence that
+this model is at least not wrong.
+
+\textit{lacUV5} is our primary target here, since it forms the
+core of all the simple repression constructs of~\cite{Jones2014}
+that we consider in Section~\ref{sec:rep_kinetics_inference}.
+Nevertheless, we thought it wise to apply our bursty promoter
+model to the other 17 constitutive promoters available in the
+FISH dataset from~\cite{Jones2014} as a test that the model is
+capturing the essential phenomenology. If the model fit well to
+all the different promoters, this would increase our confidence
+that it would serve well as a foundation for inferring repressor
+kinetics later in Section~\ref{sec:rep_kinetics_inference}.
+Conversely, were the model to fail on more than a couple of the
+other promoters, it would give us pause.
+
 \begin{figure}%[h!]
 \centering
-\includegraphics[width=0.9\textwidth]{../figures/fig2/fig2pt1.pdf}
-\caption{\textbf{Constitutive promoter posterior inference and predictive
-        checks.} (A) shows the joint posterior density from MCMC sampling of
-        model 5, bursty promoter with negative binomially-distributed steady
-        state. (B) shows the ECDF of the observed population distribution of
-        mRNA transcripts under the control of a constitutive lacUV5 promoter.
-        Posterior predictive percentiles are shown as colored bands for model
-        (1), Poisson, and model (5), negative binomial.}
-\label{fig:uv5_post}
+\includegraphics[width=\textwidth]{../figures/main/fig03.pdf}
+\caption{\textbf{Constitutive promoter posterior inference and model
+        comparison.}
+        (A) The joint posterior density of model 5, the
+        bursty promoter with negative binomially-distributed
+        steady state, is plotted with MCMC samples.
+        1D marginal densities are plotted as flanking histograms.
+        The model was fit on \textit{lacUV5} data from~\cite{Jones2014}.
+        (B) The ECDF of the observed population distribution of
+        mRNA transcripts under the control of a constitutive
+        lacUV5 promoter is shown in orange.
+        The median posterior predictive ECDFs for models (1),
+        Poisson, and (5), negative binomial, are plotted in
+        green and blue, respectively.
+        Lighter green and blue regions enclose 95\% of all
+        posterior predictive samples from their respective models.
+        Model (1) is in obvious contradiction with the data
+        while model (5) is not.
+        FISH data is again from~\cite{Jones2014}.
+    (C) Joint posterior distributions for burst rate $k_i$ and mean burst size
+    $b$ for 18 promoters from~\cite{Jones2014}. Each contour indicates the 95\%
+    highest posterior density region for a particular promoter. Note that the
+    vertical axis is shared with (D). Panel (D) plots the burst rate $k_i$ vs.\
+    the binding energy for each promoter from energy matrices
+    from~\cite{Brewster2012}.
+    The dotted line shows the predicted slope according to
+    $\ln k_i/\gamma \sim \text{constant} - \beta\Delta\epsilon_P$,
+    described in text.}
+\label{fig:constit_post_full}
 \end{figure}
+
+\fig{fig:constit_post_full}(C) shows the results, plotting the
+posterior distribution from individually MCMC sampling all 18
+constitutive promoter datasets from~\cite{Jones2014}. To aid
+visualization, rather than plotting samples for each promoter's
+posterior as we did in \fig{fig:constit_post_full}(A), for each
+posterior we find and plot the curve that surrounds the 95\%
+highest probability density region, i.e., on a plot analogous
+to~(A), each contour would enclose approximately 95\% of the
+samples, and thus 95\% of the probability mass,
+of its posterior distribution.
+Posterior predictive ECDFs (not shown) display a similar level of
+agreement between data and predictive samples as for the bursty
+model with \textit{lacUV5} in \fig{fig:constit_post_full}(B).
+\mmnote{Might be worth showing some/all of these PPCs in
+supplement for completeness? We could do postage stamp sized diff
+plots and probably fit all 18 promoters on 1-2 pages.}
+
+One interesting feature from \fig{fig:constit_post_full}(C) is
+that burst rate varies far more widely, over a range of
+$\sim10^2$, than burst size, confined to a range of
+$\lesssim10^1$ (and with the exception of promoter 6, just a span
+of 3-5x). This suggests that $k_i$, not $b$, is the key dynamic
+variable that promoter binding site tunes.
+
+It is interesting to connect this observation to the work
+of~\cite{Brewster2012}, where these same 18 promoters were
+considered through the lens of the three-state equilibrium model
+and binding energies $\Delta\epsilon_P$ were predicted from an
+energy matrix model derived from~\cite{Kinney2010}. The connection
+is made simply by equating mean mRNA levels in both theories,
+\mmnote{derivation is hasty, needs elaboration}
+\begin{equation}
+\langle m \rangle = \frac{k_i b}{\gamma}
+        = \frac{r}{\gamma}
+        \frac{\frac{P}{N_{NS}}\exp(-\beta\Delta\epsilon_P)}
+                {1+\frac{P}{N_{NS}}\exp(-\beta\Delta\epsilon_P)},
+\end{equation}
+and taking the weak-promoter limit in the equilibrium model,
+\begin{equation}
+\langle m \rangle = \frac{k_i b}{\gamma}
+        = \frac{r}{\gamma} \frac{P}{N_{NS}}\exp(-\beta\Delta\epsilon_P),
+\end{equation}
+valid for all the binding energies considered here.
+
+How are the two coarse-grainings related? The only association that makes dimensional sense and produces the correct order-of-magnitude for the known parameters is to take
+\begin{equation}
+\frac{k_i}{\gamma} = \exp(-\beta\Delta\epsilon_P)
+\label{eq:bursty_equil_corresp1}
+\end{equation}
+and
+\begin{equation}
+b = \frac{r}{\gamma}\frac{P}{N_{NS}}.
+\label{eq:bursty_equil_corresp2}
+\end{equation}
+\fig{fig:constit_post_full}(D) shows that this linear scaling
+between $\ln k_i$ and $-\beta\Delta\epsilon_P$ is approximately
+true for all 18 constitutive promoters considered. The intercept
+is an uninteresting function of the unknown model parameters, so
+we only consider the slope of the scaling in
+\fig{fig:constit_post_full}(D).
+\mmnote{Wait, is that true or can we do something with the intercept, related to the burst size in some useful way? Needs more careful thought.}
+
+While the associations represented by
+\eq{eq:bursty_equil_corresp1} and \eq{eq:bursty_equil_corresp2}
+appear to be born out by the scaling $\ln k_i \sim
+-\beta\Delta\epsilon_P$, we do not find the association of
+parameters encoded by
+\eqrange{eq:bursty_equil_corresp1}{eq:bursty_equil_corresp2}
+intuitive. If some deeper insight is waiting to be gleaned from it,
+we leave it as an open problem.
 
@@
 \begin{itemize}
-\item Bursty model. Write Bayes, sketch story of NB likelihood, sketch choice of
-priors, show~\fig{fig:uv5_post}, see bursty $\gg$ Poisson model.
 \item All 18 promoters. Hey the model still works on things other than UV5,
@@
 
-\begin{figure}%[h!]
-\centering
-\includegraphics[width=0.9\textwidth]{../figures/fig2/fig2pt2.pdf}
-\caption{\textbf{Constitutive promoter model comparison.}
-    (A) Joint posterior distributions for burst rate $k_i$ and mean burst size
-    $b$ for 18 promoters from Jones 2014. Each contour indicates the 95\%
-    highest posterior density region for a particular promoter, i.e., on a plot
-    analogous to~\fig{fig:uv5_post}, the contour would enclose approximately
-    95\% of the samples, and thus 95\% of the probability mass. Note that the
-    vertical axis is shared with (B). Panel (B) plots the burst rate $k_i$ vs.\
-    binding energy from energy matrices from Brewster 2012. The dotted line
-    shows the predicted slope according to $\ln k_i/\gamma \sim \text{constant}
-    - \beta\Delta\epsilon_P$, described in text.}
-\label{fig:constit_post}
-\end{figure}
 \mmnote{Discuss puzzling comparison w/ Chong2014: if supercoiling is the thing, why are my burst sizes all the same but burst rates vary?
@@
 \subsection{Transcription factor kinetics can be inferred from FISH measurements}
+\label{sec:rep_kinetics_inference}
 \mmnote{Outline:
@@
 
-Figure~3 presents inferred transcription factor rates, compares them to
+\fig{fig3:kR_inferences} presents inferred
+transcription factor rates, compares them to
 single-molecule measurements, and compares their ratios to gold-standard binding
@@
 \caption{\textbf{Simple repression model comparison.}
-(A) depicts the cartoon of simple repression for which we infer parameters.
+Panel (A) depicts the cartoon of simple repression for which we infer parameters.
 \mmnote{Not sure panel (A) is necessary, we've already shown it in at least 2
 other figs. We could instead show some alternative slices of 9D posterior as a
-companion to (B), or something totally different\dots??} (B) shows 50\% and 95\%
-contours of several 2D slices of the 9D posterior distribution, from fitting one
+companion to (B), or something totally different\dots??}
+(B) Contours which enclose 50\% and 95\% of the posterior
+probability mass are shown for of several 2D slices of
+the 9D posterior distribution, from fitting one
 unbinding rate for each operator (O1, O2, O3) and one binding rate for each aTc
 concentration (corresponding to an unknown mean repressor copy number). (C)
-compares ratios of our inferred unbinding rates with operator binding energy
+Ratios of our inferred unbinding rates are compared with operator binding energy
 differences measured in Garcia \& Phillips 2011 (triangles) and Razo-Mejia et.\
-al.\ 2018 (squares). (D) compares our inferred unbinding rates with
+al.\ 2018 (squares). (D) Unbinding rates inferred in this work are compared with
 single-molecule measurements of the same from Hammar et.\ at.\ 2014
@@
 \caption{\textbf{Posterior predictive checks.}
-(A) duplicates the posterior from~\fig{fig3:kR_inferences}B. (B), (C), and (D)
-show samples from the posterior predictive distribution for each of the datasets
+(A) The posterior distribution from~\fig{fig3:kR_inferences}B is
+shown again for context. (B), (C), and (D) Samples from the
+posterior predictive distribution are shown for each of the datasets
 used in the inference. Samples are sorted by operator and plotted separately

### 2020-05-21 doc/section_05_discussion.tex
Added to constit prom Bayesian inference + minor file reorg

@@
 stuff that RP lab does.}
+
+It is interesting to speculate what microscopic details are being
+coarse-grained by our burst rate and burst size in model 5.
+Chromosomal locus is one possible influence we have not addressed
+in this work, as all the FISH data from~\cite{Jones2014} that we
+considered was from a construct integrated at the \textit{galK} locus.
+The results of~\cite{Chong2014} indicate that
+transcription-induced supercoiling contributes substantially in
+driving transcriptional bursting, and furthermore, their Figure~7
+suggests that the parameters describing the rate, duration, and
+size of bursts vary substantially for transcription from
+different genomic loci.
+Although the authors of~\cite{Englaender2017} do not address
+noise, they note enormous differences in mean expression levels
+when an identical construct is integrated at different genomic loci.
+The authors of~\cite{Engl2020} attribute noise and burstiness in
+their FISH data to the influence of different sigma factors,
+which is a reasonable conclusion from their data.
+Could the difference also be due to the different chromosomal
+locations of the two operons?
+
+It seems clear therefore that chromosomal locus plays a major
+role in mean expression levels as well as noise. What features of
+different loci are and are not important? Sigma factors could
+certainly play a part, but it seems many other mechanisms are at
+play: supercoiling, and more generally 3D organization of the
+chromosome, read-through of neighboring genes, and surely some
+fascinating unknown unknowns.
+Could our preferred coarse-grained model capture the variability
+across different loci? If so, and we were to repeat the parameter
+inference as done in this work, is there a simple theoretical model
+we could build to understand the resulting parameters?
+
+A direct \mmnote{if cloning intensive} future experiment would be
+to design a payload with a promoter driving a FISH reporter,
+integrate that payload at a variety of loci scattered around the genome,
+and carry out FISH measurements of the resulting strains.
+\mmnote{
+Did Brewster already do this? I remember he integrated arrays of
+competitor binding sites all over the genome, but did he put
+lacUV5 + reporter scattered around?}
+Obviously integration loci would need to be carefully chosen so
+that one could plausibly hope the deletion and integration did
+not disturb the cell's physiology too badly.

### 2020-05-21 doc/section_03_beyond_means.tex
Added to constit prom Bayesian inference + minor file reorg
Update section_03_beyond_means.tex

@@
 % !TEX root = ./busty_transcription.tex
-\section{Beyond means}
+\section{Beyond Means in Gene Expression}
+
+In this section, our objective is to explore the same models considered in
+the previous section, but now with reference to the the question of
+how well they describe the distribution of gene expression levels, with
+special reference to the variance in these distributions.  To that end,
+we repeat the same pattern as in the previous section by examining
+the models one by one.
+
 \subsection{A one-state promoter with bursting is the ``best'' model of constitutive promoters}
@@
 
-\subsubsection{Model 1 - Poisson promoter.}
-This is the picture from Jones et.\ al.~\cite{Jones2014}, which assumes that
-transcripts are produced as a Poisson process from a single promoter state. This
+\subsubsection{Noise in the Poisson Promoter Model}
+The simplest model of constitutive expression that we can imagine
+is shown as model 1 in Figure~\ref{fig:constit_cartoons} and
+assumes that
+transcripts are produced as a Poisson process from a single promoter state. 
+This is the picture from Jones et.\ al.~\cite{Jones2014} that was used to
+interpret a systematic study of gene expression noise over a series of
+promoters designed to have different strengths.   This
 model insists that the ``true'' mRNA distribution is Poisson, implying the Fano
@@
 insufficient, and a description of the full distribution of molecular counts is
-necessary. For this we need a( slightly) more complex model than model 1 that
+necessary. For this we need a (slightly) more complex model than model 1 that
 would allow us to incorporate the non-Poissonian features of constitutive

### 2020-05-21 doc/abstract.tex

--- 2020-05-19 doc/abstract.tex
@@
 \begin{abstract}
-Abstract here.
+The study of transcription remains one of the centerpieces of modern biology
+with implications in settings from development to metabolism to evolution to
+disease. Precision measurements using a host of different techniques including
+fluorescence and sequencing readouts have raised the bar for what it means to
+quantitatively understand transcriptional regulation. In particular our
+understanding of the simplest genetic circuit is sufficiently refined both
+experimentally and theoretically that it has become possible to carefully
+discriminate between different conceptual pictures of how this regulatory system
+works. This regulatory motif, originally posited by Jacob and Monod in the
+1960s, consists of a single transcriptional repressor binding to a promoter site
+and inhibiting transcription. In this paper, we show how seven distinct models
+of this so-called simple-repression motif, based both on equilibrium and kinetic
+thinking, can be used to derive the predicted levels of gene expression. These
+different models are invoked to confront a variety of different data on mean,
+variance and full gene expression distribution, illustrating the extent to which
+such models can be distinguished.
 \end{abstract}

### 2020-05-21 doc/appendix_02_constitutive_promoters.tex

@@
 % !TEX root = ./busty_transcription.tex
-\section{Derivations of constitutive promoter models}
-\subsection{Model 2}
-\mmnote{Obtain Fano factor from CME.}
-\subsection{Model 3}
+\section{Derivations for non-bursty promoter models}
+\subsection{Nonequilibrium Model One - Poisson Promoter}
+\mmnote{Derive mean mRNA referred to in means section.}
+
+\subsection{Nonequilibrium Model Two - RNAP Bound and Unbound States}
+\mmnote{Obtain mean mRNA and Fano factor from CME.}
+
+\subsection{Nonequilibrium Model Three
+    - Multistep Transcription Initiation and Escape}
+\mmnote{Derive mean mRNA from CME.}
+\newline
 \mmnote{Obtain Fano factor by taking limit from Mitarai et al's expression.}
-\subsection{Model 4}
+
+\subsection{Nonequilibrium Model Four
+    - ``Active'' and ``Inactive'' States}
+\mmnote{Derive mean mRNA from CME.}
+\newline
 \mmnote{Obtain Fano factor by relabeling cartoon from Jones2014.}

### 2020-05-21 doc/section_02_means.tex

@@
 % !TEX root = ./busty_transcription.tex
-\section{Means}
+\section{Means}\label{section_02_means}
 
-As noted in the previous section, there are two broad classes of models
-in play for computing the input-output functions of regulatory architectures 
-as shown in Figure~\ref{fig1:means_cartoons}.
-In both classes of model, the promoter is imagined to exist in a discrete
-set of states of occupancy, with each such state of occupancy
-accorded its own rate of transcription (and for many of these
-states, there is no transcription).  The models are probabilistic with
-each state assigned some probability and the overall rate of transcription
-given by 
+As noted in the previous section, there are two broad classes of models in play
+for computing the input-output functions of regulatory architectures as shown in
+Figure~\ref{fig1:means_cartoons}. In both classes of model, the promoter is
+imagined to exist in a discrete set of states of occupancy, with each such state
+of occupancy accorded its own rate of transcription -- including no 
+transcription for many of these states. The models are probabilistic with each
+state assigned some probability and the overall rate of transcription given by 
 \begin{equation}
@@
 \end{equation}
-where $i$ labels the distinct states, $p_i$ is the probability of
-the i$^{th}$ state and $r_i$ is the rate of transcription of
-that state.  Ultimately, the different models differ along several key
-axes: what states to consider and how to compute the probabilities of those
-states.
+where $i$ labels the distinct states, $p_i$ is the probability of the
+$i^{\text{th}}$ state, and $r_i$ is the rate of transcription of that state.
+Ultimately, the different models differ along several key axes: what states to
+consider and how to compute the probabilities of those states.
 
-The first class of models that are the focus of
-the present section on predicting the mean
-level of gene expression, sometimes known as thermodynamic models,
-invoke the tools of equilibrium statistical mechanics to compute the 
-probabilities~\cite{Ackers1982, Shea1985, Buchler2003,Vilar2003a,Vilar2003b, Bintu2005a,Bintu2005c, Gertz2009,Sherman2012, Saiz2013}.
-As seen in Figure~\ref{fig1:means_cartoons}(B), even within
-the class of thermodynamic models, we can make different commitments
-about the underlying microscopic states of the promoter.  Indeed,
-the list of options considered here does not at all exhaust the suite
-of different microscopic states we can assign to the promoter.
+The first class of models that are the focus of the present section on
+predicting the mean level of gene expression, sometimes known as thermodynamic
+models, invoke the tools of equilibrium statistical mechanics to compute the
+probabilities~\cite{Ackers1982, Shea1985, Buchler2003, Vilar2003a, Vilar2003b,
+Bintu2005a, Bintu2005c, Gertz2009, Sherman2012, Saiz2013}. As seen in
+Figure~\ref{fig1:means_cartoons}(B), even within the class of thermodynamic
+models, we can make different commitments about the underlying microscopic
+states of the promoter.  Indeed, the list of options considered here does not at
+all exhaust the suite of different microscopic states we can assign to the
+promoter.
 
-The second class of models that allow us to access the mean
-expression use chemical master equations to compute the
-probabilities of the different microscopic states
-~\cite{Ko1991,Peccoud1995,Record1996,
-Kepler2001, Sanchez2008,Shahrezaei2008,Sanchez2011, Michel2010}.
-As seen in Figure~\ref{fig1:means_cartoons}(B), we consider a host of
-different nonequilibrium models, each of which will have its own 
-result for both the mean and noise in gene expression.
+The second class of models that allow us to access the mean gene expression use
+chemical master equations to compute the probabilities of the different
+microscopic states ~\cite{Ko1991, Peccoud1995, Record1996, Kepler2001,
+Sanchez2008, Shahrezaei2008, Sanchez2011, Michel2010}. As seen in
+Figure~\ref{fig1:means_cartoons}(C), we consider a host of different
+nonequilibrium models, each of which will have its own result for both the mean
+(this section) and noise (next section) in gene expression.
 
@@
 As a first stop on our search for the ``right'' model of simple repression, let
-us consider what we can learn from theory and measurements on the
-average level of  gene expression in a population of cells.
-One experimental strategy that has been particularly useful (if incomplete since
-it misses out on gene expression dynamics) is to measure the  fold-change in
-mean expression. The fold-change is defined as
+us consider what we can learn from theory and experimental measurements on the
+average level of gene expression in a population of cells. One experimental
+strategy that has been particularly useful (if incomplete since it misses out on
+gene expression dynamics) is to measure the fold-change in mean expression. The
+fold-change is defined as
 \begin{equation}
@@
         {\langle \text{gene expression with repressor absent} \rangle}
-= \frac{\langle m\rangle(R\ne 0)}{\langle m\rangle(R = 0)}
+= \frac{\langle m (R > 0) \rangle}{\langle m (R = 0) \rangle}
 \label{eq:fc_def}
 \end{equation}
-where angle brackets denote the average over a population of cells and mean mRNA
-$\langle m\rangle$ is viewed as a function of repressor copy number $R$. The
-second equality in~\ref{eq:fc_def} follows from assuming that the translation
-efficiency, i.e., the number of proteins translated per mRNA, is the same in
-both conditions. In other words, we assume that mean protein level is
-proportional to mean mRNA level, and that the proportionality constant is the
-same in both conditions and therefore cancels out in the ratio. This is
-reasonable since the cells in the two conditions are identical except for the
-presence of the transcription factor, and the model assumes that the 
+where angle brackets $\left\langle \cdot \right\rangle$ denote the average over
+a population of cells and mean mRNA $\langle m\rangle$ is viewed as a function
+of repressor copy number $R$. What this means is that the fold-change in gene
+expression is a relative measurement of the effect of the transcriptional 
+repressor ($R > 0$) on the gene expression level compared to an unregulated
+promoter ($R = 0$). The second equality in~\ref{eq:fc_def} follows from assuming
+that the translation efficiency, i.e., the number of proteins translated per
+mRNA, is the same in both conditions. In other words, we assume that mean
+protein level is proportional to mean mRNA level, and that the proportionality
+constant is the same in both conditions and therefore cancels out in the ratio.
+This is reasonable since the cells in the two conditions are identical except
+for the presence of the transcription factor, and the model assumes that the
 transcription factor has no direct effect on translation.
@@
 work~\cite{Garcia2011a, Razo-Mejia2018, Chure2019}. Part of its utility in
-dissecting transcriptional regulation is its ratiometric nature, which removes many
-secondary effects that are present when making an absolute gene
-expression measurement.   Also, by measuring otherwise
-identical cells with and without a transcription factor present, any biological
-noise common to both conditions can be made to cancel away. In addition, the
-presence of different operator sites is known to change the basal expression
-level in the absence of repressor~\mmnote{look up cite}. Computing fold-changes
-neatly cancels this effect as well.
+dissecting transcriptional regulation is its ratiometric nature, which removes
+many secondary effects that are present when making an absolute gene expression
+measurement. Also, by measuring otherwise identical cells with and without a
+transcription factor present, any biological noise common to both conditions can
+be made to cancel away. In addition, the presence of different operator sites,
+i.e. binding sites for the repressor, is known to change the basal expression
+level in the absence of repressor~\mmnote{look up cite}\mrm{does reg-seq mention
+anything like this in their SI? I definitely saw that on my own experiments, but
+didn't mention it in the paper}. Computing fold-changes neatly cancels this
+effect as well.
 
 \fig{fig1:means_cartoons} depicts a smorgasbord of mathematicized cartoons for
-simple repression using both thermodynamic and kinetic models  that have appeared in previous literature. For each cartoon,
-we calculate the fold-change in mean gene expression as predicted by that model,
-deferring some algebraic details to the appendix. What we will find is that all
-cartoons collapse to a single master curve, shown in \fig{fig1:means_cartoons}(C),
-which contains just two parameters. We label the parameters $\Delta F_R$, an
-effective free energy parametrizing the repressor-DNA interaction, and $\rho$,
-which subsumes all details of transcription in the absence of repressors. We
-will offer some intuition for why this master curve exists and discuss why at the level of
-the mean expression, we are unable   to discriminate ``right'' from ``wrong'' cartoons given only
+simple repression using both thermodynamic and kinetic models that have
+appeared in previous literature. For each cartoon, we calculate the fold-change
+in mean gene expression as predicted by that model, deferring some algebraic
+details to the appendix \mrm{cite specific appendix section}. What we will find
+is that all cartoons collapse to a single master curve, shown in
+\fig{fig1:means_cartoons}(D), which contains just two parameters. We label the
+parameters $\Delta F_R$, an effective free energy parametrizing the
+repressor-DNA interaction, and $\rho$, which subsumes all details of
+transcription in the absence of repressors. We will offer some intuition for why
+this master curve exists and discuss why at the level of the mean expression, we
+are unable to discriminate ``right'' from ``wrong'' cartoons given only
 measurements of fold-changes in expression.
@@
 \subsubsection{The Two-State Equilibrium Model}
-In this simplest model, the promoter is idealized as existing in one of two
-states, either repressor bound or repressor unbound. The rate of transcription,
-and therefore the level of gene expression, is assumed to be proportional to the
-fraction of time spent in the repressor unbound state. Mathematically, this
-amounts to the statement that the mean mRNA level is given by
+In this simplest model, depicted as (1) in Figure~\ref{fig1:means_cartoons}(B),
+the promoter is idealized as existing in one of two states, either repressor
+bound or repressor unbound. The rate of transcription, and therefore the level
+of gene expression, is assumed to be proportional to the fraction of time spent
+in the repressor unbound state. Mathematically, this amounts to the statement
+that the mean mRNA level is given by
 \begin{equation}
 \langle m \rangle = \frac{r}{\gamma}
-        \left(1 + \frac{R}{N_{NS}} e^{-\beta\Delta\epsilon_R}\right)^{-1},
+        \left(1 + \frac{R}{N_{NS}} e^{-\beta\Delta\varepsilon_R}\right)^{-1},
 \end{equation}
@@
 is the mRNA degradation rate, $R$ is repressor copy number, $N_{NS}$ is the
-number of nonspecific binding sites in the genome (where repressors spend most
-of their time when not bound to the operator, their specifc binding site),
-$\beta=1/k_BT$, and $\Delta\epsilon_R$ is the binding energy of a repressor to
-its operator site. The derivation of this result is deferred to the appendix.
+number of nonspecific binding sites in the genome where repressors spend most
+of their time when not bound to the operator, $\beta \equiv 1/k_BT$, and
+$\Delta\varepsilon_R$ is the binding energy of a repressor to its operator site.
+The derivation of this result is deferred to the appendix \mrm{cite specific 
+section}.
 
-The fold-change is  found as the ratio of mean mRNA with and without
-repressor as introduced in eqn.~\ref{eq:fc_def}. Invoking that definition results in
+The fold-change is  found as the ratio of mean mRNA with and without repressor
+as introduced in eqn.~\ref{eq:fc_def}. Invoking that definition results in
 \begin{equation}
 \text{fold-change}
-= \left(1 + \frac{R}{N_{NS}} e^{-\beta\Delta\epsilon_R}\right)^{-1},
+= \left(1 + \frac{R}{N_{NS}} e^{-\beta\Delta\varepsilon_R}\right)^{-1},
 \end{equation}
-which clearly matches the form of the master curve in Figure~\ref{fig1:means_cartoons}(C)
-with $\rho=1$.
+which clearly matches the form of the master curve in
+Figure~\ref{fig1:means_cartoons}(C) with $\rho=1$ and $\Delta F_R = 
+\beta\Delta\varepsilon_r - \log (R / N_{NS})$.
 
@@
 Compared to the previous model, here we fine-grain the repressor unbound state
-into two separate states: empty, and RNAP bound. This picture was used
-in~\cite{Garcia2011a} as we use it here, and in~\cite{Razo-Mejia2018}
-and~\cite{Chure2019} it was generalized to incorporate small-molecule inducers
-that bind the repressor. Interestingly, the effect of this is simply to
-renormalize $R$ from the total number of repressors to a smaller effective
-number, roughly (though not exactly) the number which are unbound by inducers.
+into two separate states: empty, and RNAP bound as shown in (2) in
+Figure~\ref{fig1:means_cartoons}(B). This picture was used in~\cite{Garcia2011a}
+as we use it here, and in~\cite{Razo-Mejia2018} and~\cite{Chure2019} it was
+generalized to incorporate small-molecule inducers that bind the repressor. 
+% Interestingly, the effect of this generalization is
+% simply to renormalize $R$ from the total number of repressors to a smaller
+% effective number, roughly (though not exactly) the number which are unbound by
+% inducers.
 
@@
 \langle m \rangle = \frac{r}{\gamma}
-\frac{\frac{P}{N_{NS}} e^{-\beta\Delta\epsilon_P}}
+\frac{\frac{P}{N_{NS}} e^{-\beta\Delta\varepsilon_P}}
         {
-        1 + \frac{R}{N_{NS}} e^{-\beta\Delta\epsilon_R}
-        + \frac{P}{N_{NS}} e^{-\beta\Delta\epsilon_P}
+        1 + \frac{R}{N_{NS}} e^{-\beta\Delta\varepsilon_R}
+        + \frac{P}{N_{NS}} e^{-\beta\Delta\varepsilon_P}
         },
 \end{equation}
-where the new variables are $\Delta\epsilon_P$, the difference in RNAP binding
-energy to its specific site (the promoter) relative to an average nonspecific
-background site, and the RNAP copy number, $P$. The fold-change again follows
-immediately as
+where the new variables are $\Delta\varepsilon_P$, the difference in RNAP
+binding energy to its specific site (the promoter) relative to an average
+nonspecific background site, and the RNAP copy number, $P$. The fold-change
+again follows immediately as
 \begin{align}
 \text{fold-change}
-&= \frac{\frac{P}{N_{NS}} e^{-\beta\Delta\epsilon_P}}
+&= \frac{\frac{P}{N_{NS}} e^{-\beta\Delta\varepsilon_P}}
         {
-        1 + \frac{R}{N_{NS}} e^{-\beta\Delta\epsilon_R}
-        + \frac{P}{N_{NS}} e^{-\beta\Delta\epsilon_P}
+        1 + \frac{R}{N_{NS}} e^{-\beta\Delta\varepsilon_R}
+        + \frac{P}{N_{NS}} e^{-\beta\Delta\varepsilon_P}
         }
-\frac{1 + \frac{P}{N_{NS}} e^{-\beta\Delta\epsilon_P}}
-        {\frac{P}{N_{NS}} e^{-\beta\Delta\epsilon_P}}
+\frac{1 + \frac{P}{N_{NS}} e^{-\beta\Delta\varepsilon_P}}
+        {\frac{P}{N_{NS}} e^{-\beta\Delta\varepsilon_P}}
 \\
 &= \left(
-1 + \frac{\frac{R}{N_{NS}} e^{-\beta\Delta\epsilon_R}}
-        {1 + \frac{P}{N_{NS}} e^{-\beta\Delta\epsilon_P}}
+1 + \frac{\frac{R}{N_{NS}} e^{-\beta\Delta\varepsilon_R}}
+        {1 + \frac{P}{N_{NS}} e^{-\beta\Delta\varepsilon_P}}
 \right)^{-1}
@@
 \end{align}
-with $\Delta F_R = \beta\Delta\epsilon_R - \log(R/N_{NS})$ and $\rho = 1 +
+with $\Delta F_R = \beta\Delta\varepsilon_R - \log(R/N_{NS})$ and $\rho = 1 +
 \frac{P}{N_{NS}}\mathrm{e}^{-\beta\Delta\varepsilon_P}$ as shown
-in~\fig{fig1:means_cartoons}.  Thus far,
-we see that the two thermodynamic models, despite making
-different coarse-graining commitments, result in the same functional
-form for the fold-change in mean gene expression.  We now
-explore how kinetic models fare when faced with computing
-the same observable.
+in~\fig{fig1:means_cartoons}(B). Thus far, we see that the two thermodynamic
+models, despite making different coarse-graining commitments, result in the same
+functional form for the fold-change in mean gene expression.  We now explore how
+kinetic models fare when faced with computing the same observable.
 
 \subsubsection{The Poisson Promoter Nonequilibrium Model}
-For our first kinetic model, we  imitate the states considered in
-the Two-State Equilibrium Model and consider the simplest possible picture with only
-two states, repressor bound and unbound. This is exactly the model used for the
-main results of~\cite{Jones2014}. In this picture, repressor association and
+For our first kinetic model, we  imitate the states considered in the Two-State
+Equilibrium Model and consider the simplest possible picture with only two
+states, repressor bound and unbound. This is exactly the model used for the main
+results of~\cite{Jones2014}. In this picture, repressor association and
 dissociation rates from its operator site, $k_R^+$ and $k_R^-$, respectively,
@@
 state, transcription initiates at rate $r$, which represents a coarse-graining
-of all the downstream processes into a single effective rate. mRNA is 
-degraded at rate $\gamma$ as already exploited in the previous models.
+of all the downstream processes into a single effective rate. mRNA is degraded
+at rate $\gamma$ as already exploited in the previous models.
 
 Let $p_{m,R}(t)$ denote the joint probability of finding the system in the
-repressor bound state with $m$ mRNA molecules present at time $t$, and similarly
-define $p_{m,U}(t)$ for the repressor unbound state. This model is governed by coupled
-master equations giving the time evolution of $p_{m,R}(t)$ and 
-$p_{m,U}(t)$~\cite{Sanchez2008,Sanchez2011, Phillips2019}
-which we can write as 
+repressor bound state $R$ with $m$ mRNA molecules present at time $t$. Similarly
+define $p_{m,U}(t)$ for the repressor unbound state $U$. This model is governed
+by coupled master equations giving the time evolution of $p_{m,R}(t)$ and
+$p_{m,U}(t)$~\cite{Sanchez2008, Sanchez2011, Phillips2019} which we can write as
 \begin{align}
 \begin{split}
-\deriv{t}p_{m,R}(t) =& - k_R^- p_{m,R}(t) + k_R^+ p_{m,U}(t)
-                + (m+1)\gamma p_{m+1,R}(t) - \gamma p_{m,R}(t)
+\deriv{t}p_{m,R}(t) =& 
+- \overbrace{k_R^- p_{m,R}(t)}^{R \rightarrow U}
++ \overbrace{k_R^+ p_{m,U}(t)}^{U \rightarrow R}
++ \overbrace{(m+1)\gamma p_{m+1,R}(t)}^{m + 1 \rightarrow m}
+- \overbrace{\gamma p_{m,R}(t)}^{m \rightarrow m - 1}
 \\
-\deriv{t}p_{m,U}(t) =&\; k_R^- p_{m,R}(t) - k_R^+ p_{m,U}(t)
-                        + rp_{m-1,U}(t) - rp_{m,U}(t)
+\deriv{t}p_{m,U}(t) =&\; 
+\overbrace{k_R^- p_{m,R}(t)}^{R \rightarrow U}
+- \overbrace{k_R^+ p_{m,U}(t)}^{U \rightarrow R}
++ \overbrace{rp_{m-1,U}(t)}^{m-1 \rightarrow m}
+- \overbrace{rp_{m,U}(t)}^{m \rightarrow m + 1}
 \\
-                &+ (m+1)\gamma p_{m+1,U}(t) - \gamma p_{m,U}(t),
+&+ \overbrace{(m+1)\gamma p_{m+1,U}(t)}^{m + 1 \rightarrow m}
+- \overbrace{\gamma p_{m,U}(t)}^{m \rightarrow m - 1},
 \label{eq:poisson_promoter_cme}
@@
 \end{align}
-where each term on the right corresponds to a transition between two
-states of the promoter.
-In each equation, the first two terms describe transitions between promoter
-states due to repressors unbinding and binding, respectively. The final two
-terms describe degradation of mRNA, decreasing the copy number by one, and the
-terms with coefficient $r$ describe transcription initiation increasing the mRNA
-copy number by one.
+where each term on the right corresponds to a transition between two states of
+the promoter as indicated by the overbrace label. In each equation, the first
+two terms describe transitions between promoter states due to repressors
+unbinding and binding, respectively. The final two terms describe degradation of
+mRNA, decreasing the copy number by one, and the terms with coefficient $r$
+describe transcription initiation increasing the mRNA copy number by one.
 
@@
 more complicated models yet to come, by re-expressing the master equation in
-vector form~\cite{Phillips2012}. The promoter states are collected into a vector and the rate
-constants are collected into matrices as
+vector form~\cite{Phillips2012}. The promoter states are collected into a vector
+and the rate constants are collected into matrices as
 \begin{equation}
@@
 \end{equation}
-with the algebra details again deferred to the appendix. Recall $k_R^+$ is
-proportional to the repressor copy number, so in computing fold-change, absence
-of repressor corresponds to $k_R^+\rightarrow0$. Therefore fold-change in this
-model is simply
+with the algebra details again deferred to the appendix \mrm{cite specific
+location}. Recall $k_R^+$ is proportional to the repressor copy number, so in
+computing fold-change, absence of repressor corresponds to $k_R^+\rightarrow0$.
+Therefore fold-change in this model is simply
 \begin{equation}
@@
 \end{equation}
-again matching the master curve of~\fig{fig1:means_cartoons}(C) with $\rho=1$.
+again matching the master curve of~\fig{fig1:means_cartoons}(D) with $\rho=1$.
 
 \subsubsection{Nonequilibrium Model Two - RNAP Bound and Unbound States}
-Our second kinetic model mirrors the second equilibrium model of
-Figure~\ref{fig1:means_cartoons}(B) by fine-graining  the
-repressor unbound state of nonequilibrium model 1, resolving it into an empty
-promoter state and an RNAP-bound state. Note in this picture, in contrast with
-model 4 below, transcription initiation is accompanied by a promoter state
-change, in keeping with the interpretation as RNAP-bound and empty states: if an
-RNAP successfully escapes the promoter and proceeds to elongation of a
-transcript, clearly it is no longer bound at the promoter! Therefore another
-RNAP must bind before another transcript can be initiated.
+Our second kinetic model depicted in Figure~\ref{fig1:means_cartoons}(C) mirrors
+the second equilibrium model of Figure~\ref{fig1:means_cartoons}(B) by
+fine-graining  the repressor unbound state of nonequilibrium model 1, resolving
+it into an empty promoter state and an RNAP-bound state. Note in this picture,
+in contrast with model 4 below, transcription initiation is accompanied by a
+promoter state change, in keeping with the interpretation as RNAP-bound and
+empty states: if an RNAP successfully escapes the promoter and proceeds to
+elongation of a transcript, clearly it is no longer bound at the promoter.
+Therefore another RNAP must bind before another transcript can be initiated.
 
@@
 change. This can be understood by analogy to $\matr{K}$. The off-diagonal and
-diagonal elements of $\matr{K}$ correspond to transitions arriving at or departing from,
-respectively, the promoter state of interest. If transcription initiation is 
-accompanied by promoter state changes, we must have separate matrices for
-arriving and departing transcription events since the arriving and departing
-transitions have different initial copy numbers of mRNA, unlike for $\matr{K}$
-where they are the same. The master equation for this model is
+diagonal elements of $\matr{K}$ correspond to transitions arriving at or
+departing from, respectively, the promoter state of interest. If transcription
+initiation is accompanied by promoter state changes, we must have separate
+matrices for arriving and departing transcription events since the arriving and
+departing transitions have different initial copy numbers of mRNA, unlike for
+$\matr{K}$ where they are the same. The master equation for this model is
 \begin{equation}
@@
 \matr{R_A} = \begin{pmatrix}
-                0 & 0 & 0 \\ 0 & 0 & r \\ 0 & 0 & 0
+                0 & 0 & 0 \\ 
+                0 & 0 & r \\ 
+                0 & 0 & 0
                 \end{pmatrix},\
 \matr{R_D} = \begin{pmatrix}
-                0 & 0 & 0 \\ 0 & 0 & 0 \\ 0 & 0 & r
-                \end{pmatrix},
+                0 & 0 & 0 \\ 
+                0 & 0 & 0 \\ 
+                0 & 0 & r
+                \end{pmatrix}.
 \label{eq:3state_cme_matrices_pt2}
@@
 The elements of $\vect{p}_m$ encode the probabilities of having $m$ mRNA present
-along with the promoter having repressor bound, being empty, or having RNAP
-bound, respectively. $\matr{R_A}$ describes probability flux \textit{arriving}
-at the state $\vect{p}_m$ from a state with one fewer mRNA, namely
-$\vect{p}_{m-1}$, and $\matr{R_D}$ describes probability flux \textit{departing}
-from the state $\vect{p}_m$ for a state with one more mRNA, namely
-$\vect{p}_{m+1}$. $\matr{K}$ is closely analogous to model 1.
+along with the promoter having repressor bound ($R$), being empty ($E$), or
+having RNAP bound ($P$), respectively. $\matr{R_A}$ describes probability flux
+\textit{arriving} at the state $\vect{p}_m$ from a state with one fewer mRNA,
+namely $\vect{p}_{m-1}$, and $\matr{R_D}$ describes probability flux
+\textit{departing} from the state $\vect{p}_m$ for a state with one more mRNA,
+namely $\vect{p}_{m+1}$. $\matr{K}$ is closely analogous to model 1.
 
@@
 \end{align}
-which follows the master curve of~\fig{fig1:means_cartoons}(C) with $\rho = 1 + k_P^+/(k_P^- + r)$ as claimed.
+which follows the master curve of~\fig{fig1:means_cartoons}(D) with $\rho = 1 +
+k_P^+/(k_P^- + r)$ as claimed.
 
-\subsubsection{Nonequilibrium Model Three - Multistep Transcription Initiation and
-Escape}
- One might reasonably complain that the first two ``nonequilibrium''
-models we have considered are straw men. Their steady states necessarily satisfy
-detailed balance which is equivalent to thermodynamic equilibrium. Why is this
-the case? At steady state there is by definition no net probability flux in or
-out of each promoter state, but since the promoter states form a linear chain,
-there is only one way in or out of the repressor bound and RNAP bound states,
-implying each edge must actually have a net zero probability flux, which is the
-definition of detailed balance (usually phrased as equality of forward and
-reverse transition fluxes).
+\subsubsection{Nonequilibrium Model Three - Multistep Transcription Initiation
+and Escape}
+One might reasonably complain that the first two ``nonequilibrium'' models we
+have considered are straw men. Their steady states necessarily satisfy detailed
+balance which is equivalent to thermodynamic equilibrium. Why is this the case?
+At steady state there is by definition no net probability flux in or out of each
+promoter state, but since the promoter states form a linear chain, there is only
+one way in or out of the repressor bound and RNAP bound states, implying each
+edge must actually have a net zero probability flux, which is the definition of
+detailed balance (usually phrased as equality of forward and reverse transition
+fluxes).
 
-Now we consider model 3 in Figure~\ref{fig1:means_cartoons}(B) which allows the possibility
-of true nonequilibrium steady-state fluxes through the promoter states. We point
-out that this model was considered previously in~\cite{Mitarai2015} where a
-comparison was made with model 1 as used in~\cite{Jones2014}. The authors
-of~\cite{Mitarai2015} argued that the additional complexity is essential to
-properly account for the noise in the mRNA distribution. We will weigh in on
-both models later when we consider observables beyond fold-change.
+Now we consider model 3 in Figure~\ref{fig1:means_cartoons}(C) which allows the
+possibility of true nonequilibrium steady-state fluxes through the promoter
+states. We point out that this model was considered previously
+in~\cite{Mitarai2015} where a comparison was made with model 1 as used
+in~\cite{Jones2014}. The authors of~\cite{Mitarai2015} argued that the
+additional complexity is essential to properly account for the noise in the mRNA
+distribution. We will weigh in on both models later when we consider observables
+beyond fold-change.
 
@@
 \end{equation}
-with the four promoter states, in order, being repressor bound, empty, RNAP
-closed complex, and RNAP open complex. Besides increasing dimension by one
-again, the only new feature in $\matr{K}$ is the rate $k_O$, representing the
-(irreversible) rate of open complex formation from the closed complex. The
-initiation matrices are given by
+with the four promoter states, in order, being repressor bound ($R$), empty
+($E$), RNAP closed complex ($C$), and RNAP open complex ($O$). Besides
+increasing dimension by one, the only new feature in $\matr{K}$ is the
+rate $k_O$, representing the (irreversible) rate of open complex formation from
+the closed complex. The initiation matrices are given by
 \begin{equation}
 \matr{R_A} = \begin{pmatrix}
-        0 & 0 & 0 & 0 \\ 0 & 0 & 0 & r \\ 0 & 0 & 0 & 0\\ 0 & 0 & 0 & 0
+        0 & 0 & 0 & 0 \\ 
+        0 & 0 & 0 & r \\ 
+        0 & 0 & 0 & 0 \\ 
+        0 & 0 & 0 & 0
                 \end{pmatrix},\
 \matr{R_D} = \begin{pmatrix}
-        0 & 0 & 0 & 0 \\ 0 & 0 & 0 & 0 \\ 0 & 0 & 0 & 0\\ 0 & 0 & 0 & r
+        0 & 0 & 0 & 0 \\ 
+        0 & 0 & 0 & 0 \\ 
+        0 & 0 & 0 & 0 \\ 
+        0 & 0 & 0 & r
                 \end{pmatrix},
@@
 The expression for mean mRNA is substantially more complicated now, as worked
-out in the appendix where we find
+out in the appendix \mrm{cite section} where we find
 \begin{equation}
@@
 \end{align}
-surprisingly reducing to the master curve of~\fig{fig1:means_cartoons}(C) once again, with $\rho = 1 +
-\frac{k_P^+ (k_O + r)}{r(k_O + k_P^-)}$.
+surprisingly reducing to the master curve of~\fig{fig1:means_cartoons}(D) once
+again, with $\rho = 1 + \frac{k_P^+ (k_O + r)}{r(k_O + k_P^-)}$.
 
@@
 transcription steps may still be collapsed to the form of the master curve for
-the means giving
-in~\fig{fig1:means_cartoons}(C) so long as the repressor binding is exclusive with
-transcriptionally active states. We offer this as a conjecture, and we suspect
-that a careful argument using the King-Altman diagram method~\cite{King1956,
-Hill1966} might furnish a ``proof.'' Our focus here is not on full generality
-but rather to survey an assortment of plausible models for simple
-repression (i.e. the repressor-operator model) that have been proposed
-in the literature.
+the means giving in~\fig{fig1:means_cartoons}(D) so long as the repressor
+binding is exclusive with transcriptionally active states. We offer this as a
+conjecture, and we suspect that a careful argument using the King-Altman diagram
+method~\cite{King1956, Hill1966} might furnish a ``proof.'' Our focus here is
+not on full generality but rather to survey an assortment of plausible models
+for simple repression  that have been proposed in the literature.
 
 \subsubsection{Nonequilibrium Model Four - ``Active'' and ``Inactive'' States}
-This model, shown in Figure~\ref{fig1:means_cartoons}(B)   is at the core of the theory in~\cite{Razo-Mejia2020}. At a glance
-the cartoon for this model may appear very similar to model 2, and
-mathematically it is, but the interpretation is rather different. In model 2, we
-interpreted the third state literally as an RNAP-bound promoter and modeled
-initiation of a transcript as triggering a promoter state change, making the
-hopefully uncontroversial assumption that an RNAP can only make one transcript
-at a time. In contrast, in the present model the promoter state does
-\textit{not} change when a transcript is initiated. So we no longer interpret
-these states as literally RNAP bound and unbound but instead as coarse-grained
-``active'' and ``inactive'' states, the details of which we leave unspecified
-for now. We will comment more on this model below when we discuss Fano factors
-of models.
+This model, shown as (4) in Figure~\ref{fig1:means_cartoons}(C) is at the core
+of the theory in~\cite{Razo-Mejia2020}. At a glance the cartoon for this model
+may appear very similar to model 2, and mathematically it is, but the
+interpretation is rather different. In model 2, we interpreted the third state
+literally as an RNAP-bound promoter and modeled initiation of a transcript as
+triggering a promoter state change, making the hopefully uncontroversial
+assumption that an RNAP can only make one transcript at a time. In contrast, in
+the present model the promoter state does \textit{not} change when a transcript
+is initiated. So we no longer interpret these states as literally RNAP bound and
+unbound but instead as coarse-grained ``active'' and ``inactive'' states, the
+details of which we leave unspecified for now. We will comment more on this
+model below when we discuss Fano factors of models.
 
@@
 \matr{R} = \begin{pmatrix}
-                0 & 0 & 0 \\ 0 & 0 & 0 \\ 0 & 0 & r
+                0 & 0 & 0 \\ 
+                0 & 0 & 0 \\ 
+                0 & 0 & r
         \end{pmatrix}.
 \end{equation}
-so the master equation takes verbatim the same form as it did for model
+The master equation takes verbatim the same form as it did for model
 1,~\eq{eq:2state_rep_cme}. Meanwhile the promoter transition matrix $K$ is the
@@
 \end{align}
-from which we see $\rho = 1 + k^+/k^-$ as shown in~\fig{fig1:means_cartoons}.
+from which we see $\rho = 1 + k^+/k^-$ as shown in~\fig{fig1:means_cartoons}(C).
 
 \subsubsection{Nonequilibrium Model Five - Bursty Promoter}
-The final model we consider shown in Figure~\ref{fig1:means_cartoons}(B) 
-is an intuitive analog to model 1, with just two states, repressor
-bound or unbound, and transition rates between them of $k_R^+$ and $k_R^-$. In
-model 1, when in the unbound state, single mRNA transcripts are produced as a
-Poisson process with some characteristic rate $r$. The current model by contrast
-produces, at some Poisson rate $k_i$, \textit{bursts} of transcripts with a
+The final model we consider shown in Figure~\ref{fig1:means_cartoons}(C) is an
+intuitive analog to model 1, with just two states, repressor bound or unbound,
+and transition rates between them of $k_R^+$ and $k_R^-$. In model 1, when in
+the unbound state, single mRNA transcripts are produced as a Poisson process
+with some characteristic rate $r$. The current model by contrast produces, at
+some Poisson rate $k_i$, \textit{bursts} of mRNA transcripts with a
 geometrically distributed burst size and a mean burst size $b$.
@@
 
-\subsubsection{Summary of Results Ccross models for Fold-Changes in Mean Expression}
+\subsubsection{Summary of Results Ccross models for Fold-Changes in Mean
+Expression}
 What have we learned from our exploration of the models
 in~\fig{fig1:means_cartoons} and their predictions of fold-changes in mean
-expression levels? The key outcome of our analysis is the existence of a master curve
- shown in \fig{fig1:means_cartoons}(C) to which the fold-change predictions of all the
-models collapse. This master curve is parametrized by only two effective
-parameters: $\Delta F_R$, which characterizes the number of repressors and
-their binding strength to the DNA, and $\rho$, which
-characterizes all other features of the promoter architecture. The key
-assumption underpinning this result is that no transcription occurs when a
-repressor is bound to its operator. Note, however, that we are agnostic about
-the molecular mechanism which achieves this; steric effects are one plausible
-mechanism, but, for instance, ``action-at-a-distance'' mediated by kinked DNA
-due to repressors bound tens or hundreds of nucleotides upstream of a promoter
-is plausible as well.
+expression levels? The key outcome of our analysis is the existence of a master
+curve shown in \fig{fig1:means_cartoons}(D) to which the fold-change predictions
+of all the models collapse. This master curve is parametrized by only two
+effective parameters: $\Delta F_R$, which characterizes the number of repressors
+and their binding strength to the DNA, and $\rho$, which characterizes all other
+features of the promoter architecture. The key assumption underpinning this
+result is that no transcription occurs when a repressor is bound to its
+operator. Note, however, that we are agnostic about the molecular mechanism
+which achieves this; steric effects are one plausible mechanism, but, for
+instance, ``action-at-a-distance'' mediated by kinked DNA due to repressors
+bound tens or hundreds of nucleotides upstream of a promoter is plausible as
+well.
 
-Why does the master curve of~\fig{fig1:means_cartoons}(C) exist at all? This
+Why does the master curve of~\fig{fig1:means_cartoons}(D) exist at all? This
 brings to mind the deep questions posed in, e.g.,~\cite{Frank2013}
@@
 to architectures with multiple repressor binding sites, as long as the
-exclusivity of transcription factor binding and transcription initation is
+exclusivity of transcription factor binding and transcription initiation is
 maintained. The interpretation of $\Delta F_R$ is then of an effective free
@@
 
-One might instead imagine a bottom-up modeling approach, where we mathematicize a
-model of what we hypothesize the important steps are and are not, use in vitro
+One might instead imagine a bottom-up modeling approach, where we mathematicize
+a model of what we hypothesize the important steps are and are not, use in vitro
 data for the steps deemed important, and \textit{predict} what $\rho$ should be.
@@
 Garcia2011a, Razo-Mejia2018} of simple repression using the 3-state equilibrium
-model in~\fig{fig1:means_cartoons}. In that picture, the weak promoter
-approximation means $\frac{P}{N_{NS}}\exp(-\beta\Delta\epsilon_P) \ll 1$,
-meaning therefore $\rho\approx1$.  This approximation can be well
-justified on the basis of the number of RNAP per cell and the strength of
-binding of RNAP to DNA at weak promoters.  This is suggestive, but how can we be sure
-that $\rho$ is not, for instance, actually $10^2$ and that $\Delta F_R$ hides a
-compensatory factor? A resolution is offered by an independent inference of
-$\rho$ in the absence of repressors. This was done in~\cite{Razo-Mejia2020} by
-fitting nonequilibrium model 4  in~\fig{fig1:means_cartoons}, with zero
-repressor (looking ahead, this is equivalent to model 4
-in~\fig{fig:constit_cartoons}), to FISH data from~\cite{Brewster2014}. This
-provided a determination of $k^+$ and $k^-$, from which their ratio is estimated
-to be no more than a few $10^{-1}$ and possibly as small as $10^{-2}$.
+model in~\fig{fig1:means_cartoons}(B). In that picture, the weak promoter
+approximation means $\frac{P}{N_{NS}}\exp(-\beta\Delta\varepsilon_P) \ll 1$,
+meaning therefore $\rho\approx1$.  This approximation can be well justified on
+the basis of the number of RNAP per cell and the strength of binding of RNAP to
+DNA at weak promoters. This is suggestive, but how can we be sure that $\rho$
+is not, for instance, actually $10^2$ and that $\Delta F_R$ hides a compensatory
+factor? A resolution is offered by an independent inference of $\rho$ in the
+absence of repressors. This was done in~\cite{Razo-Mejia2020} by fitting
+nonequilibrium model 4 in~\fig{fig1:means_cartoons}(C), with zero repressor
+(looking ahead, this is equivalent to model 4 in~\fig{fig:constit_cartoons}), to
+FISH data from~\cite{Brewster2014}. This provided a determination of $k^+$ and
+$k^-$, from which their ratio is estimated to be no more than a few $10^{-1}$
+and possibly as small as $10^{-2}$.
 
@@
 measurements.) They find broad agreement between the in vitro and in vivo
-values. This reinforces the suspicion that the equilibrium $\Delta\epsilon_R$
+values. This reinforces the suspicion that the equilibrium $\Delta\varepsilon_R$
 repressor binding energies do in fact represent real physical free energies.
@@
 is probably unrealistic to measure repressor rates $k_R^+$ or $k_R^-$ or
-fold-changes in expression levels (and hence $\Delta\epsilon_R$) precisely
+fold-changes in expression levels (and hence $\Delta\varepsilon_R$) precisely
 enough to detect the expected tiny deviations from $\rho=1$. We can estimate the
@@
 happens to be $\sim0.1$, on the high end of our range estimated above. A
-determination of $\Delta\epsilon_R/k_BT$ with an uncertainty of barely 0.1 was
-achieved in the excellent measurements of~\cite{Razo-Mejia2018}, so this
+determination of $\Delta\varepsilon_R/k_BT$ with an uncertainty of barely 0.1
+was achieved in the excellent measurements of~\cite{Razo-Mejia2018}, so this
 requires a very difficult determination of $\Delta F_R$ for a very crude
@@
 free energy of repressor binding inferred from fold-change measurements should
-agree with repressor binding and unbinding rates. In other words, the equilbrium
-and nonequilibrium definitions of $\Delta F_R$ must agree, meaning
+agree with repressor binding and unbinding rates. In other words, the
+equilibrium and nonequilibrium definitions of $\Delta F_R$ must agree, meaning
 \begin{equation}
-\Delta F_R = \beta\Delta\epsilon_R - \log(R/N_{NS})
+\Delta F_R = \beta\Delta\varepsilon_R - \log(R/N_{NS})
         = - \log(k_R^+/k_R^-),
 \end{equation}
-must hold, where $\beta\Delta\epsilon_R$ is inferred from the master curve fit
-to fold-change measurements, and $k_R^+$ and $k_R^-$ are inferred in some
+must hold, where $\beta\Delta\varepsilon_R$ is inferred from the master curve
+fit to fold-change measurements, and $k_R^+$ and $k_R^-$ are inferred in some
 orthogonal manner. Single molecule measurements such as from~\cite{Hammar2014}

### 2020-05-21 doc/appendix_01_bayesian_inference.tex
MRM pass through section 4 on 20200521

@@
 % !TEX root = ./busty_transcription.tex
-\section{Bayesian inference}
+\section{Bayesian inference}\label{sec:bayesian}
 \mmnote{
@@
 }
+
+\subsection{Parameter inference for constitutive promoters}
+
+To proceed we need to specify a prior. In this case we are extremely data-rich,
+as the dataset from Jones et.\ al~\cite{Jones2014} has of order 1000-3000
+single-cell measurements for each promoter, so our choice of prior matters
+little here, as long as it is sufficiently broad. A convenient choice for our
+problem is to use a \textit{conjugate} prior. A conjugate prior is a special
+prior that causes the posterior to have the same functional form as the prior,
+simply with updated model parameters. This makes calculations analytically
+tractable and also offers a nice interpretation of the inference procedure as
+updating our knowledge about the model parameters. This makes conjugate priors
+very useful when they exist. The caveat is that conjugate priors only exist for
+a very limited number of likelihoods, mostly with only one or two model
+parameters, so in almost all other Bayesian inference problems, we must tackle
+the posterior numerically.
+
+But, for the problem at hand, a conjugate prior does in fact exist. For a
+Poisson likelihood of identical and identically distributed data, the conjugate
+prior is a gamma distribution, as can be looked up in, e.g.,~\cite{Gelman2013},
+Section 2.6. Putting a gamma prior on $\lambda$ introduces two new parameters
+$\alpha$ and $\beta$ which parametrize the gamma distribution itself, which we
+use to encode the range of $\lambda$ values we view as reasonable. Recall
+$\lambda$ is the mean steady-state mRNA count per cell, which \textit{a priori}
+could plausibly be anywhere from 0 to a few hundred. $\alpha=1$ and $\beta=1/50$
+achieve this, since the gamma distribution is strictly positive with mean
+$\alpha/\beta$ and standard deviation $\sqrt{\alpha}/\beta$. To be explicit,
+then, our prior is
+\begin{equation}
+\lambda \sim \text{Gamma}(\alpha, \beta)
+\end{equation}
+
+As an aside, note that if we did not know that our prior was a conjugate prior,
+we could still write down our posterior distribution from its definition as
+\begin{equation}
+p(\lambda\mid D,\alpha,\beta)
+\propto p(D\mid\lambda) p(\lambda \mid\alpha,\beta)
+\propto \left(\prod_{k=1}^N \frac{\lambda^{m_k}e^{-\lambda}}{m_k!}\right)
+        \frac{\beta}{\Gamma(\alpha)}(\beta\lambda)^{\alpha-1} e^{-\beta\lambda}
+.
+\end{equation}
+Without foreknowledge that this in fact reduces to a gamma distribution, this
+expression might appear rather inscrutable. When conjugate priors are
+unavailable for the likelihood of interest - which is almost always the case for
+models with $>1$ model parameter - this inscrutability is the norm, and making
+sense of posteriors analytically is almost always impossible. Fortunately, MCMC
+sampling provides us a powerful method of constructing posteriors numerically
+which we will make use of extensively.
+
+Since we did use a conjugate prior, we may simply look up our posterior in any
+standard reference such as~\cite{Gelman2013}, Section 2.6,
+from which we find that
+\begin{equation}
+\lambda
+\sim \text{Gamma}\left(\alpha + \bar{m}N, \beta + N\right),
+\end{equation}
+where we defined the sample mean $\bar{m} = \frac{1}{N}\sum_k m_k$ for
+notational convenience. A glance at the FISH data from~\cite{Jones2014} reveals
+that $N$ is $\mathcal{O}(10^3)$ and $\langle m\rangle \gtrsim 0.1$ for all
+constitutive strains in~\cite{Jones2014}, so $\bar{m}N \gtrsim 10^2$. Therefore
+as we suspected, our prior parameters are completely overwhelmed by the data.
+The prior behaves, in a sense, like $\beta$ extra ``data points''
+with a mean value of $(\alpha-1)/\beta$~\cite{Gelman2013}, which
+gives us some intuition for how much data is needed to overwhelm
+the prior in this case: enough data $N$ such that $\beta\ll N$
+and $\alpha/\beta \ll \bar{m}$. In
+fact, $\bar{m}N$ and $N$ are so large that we can, to an excellent
+approximation, ignore the $\alpha$ and $\beta$ dependence and approximate the
+gamma distribution as a Gaussian with mean $\bar{m}$ and standard deviation
+$\sqrt{\bar{m}/N}$, giving
+\begin{equation}
+\lambda
+\sim \text{Gamma}\left(\alpha + \bar{m}N, \beta + N\right)
+\approx \text{Normal}\left(\bar{m}, \sqrt{\frac{\bar{m}}{N}}\right).
+\end{equation}
+As an example with real numbers, for the \textit{lacUV5} promoter, Jones et.\
+al~\cite{Jones2014} measured 2648 cells with an average mRNA count per cell of
+$\bar{m} \approx 18.7$. In this case then, our posterior is
+\begin{equation}
+\lambda
+\sim \text{Normal}\left(18.7, 0.08\right),
+\end{equation}
+which suggests we have inferred our model's one parameter to a precision of
+order 1\%.
+
+This is not wrong, but it is not the full story. The model's posterior is
+tightly constrained, but is it a good generative model? In other words, does the
+model generate data that look similar to our actual data, and is it therefore
+plausible that the model captures the important features of the data generating
+process? This intuitive notion can be codified with \textit{posterior predictive
+checks}, or PPCs, and we will see that this simple Poisson model fails badly.
+
+The intuitive idea of posterior predictive checks is simple: \mmnote{Not sure if
+all this stuff is clearer by example, but then does it confuse the reader in
+generalizing to other models?}
+\begin{enumerate}
+\item Make a random draw of the model parameter $\lambda$ from the posterior
+distribution.
+\item Plug that draw into the likelihood and generate a synthetic dataset
+$\{m_k\}$ conditioned on $\lambda$.
+\item Repeat many times.
+\end{enumerate}
+More formally, the posterior predictive distribution can be thought of as the
+distribution of future yet-to-be-observed data, conditioned on the data we have
+already observed. Clearly if those data appear quite different, the model has a
+problem. Put another way, if we suppose the generative model is true, then the
+synthetic datasets we generate should resemble the actual observed data, and if
+not, it suggests the model is missing important features. All the data we
+consider in this work are 1D (distributions of mRNA counts over a population) so
+ECDFs are an excellent visual means of comparing synthetic and observed
+datasets. In general for higher dimensional datasets, much of the challenge is
+in merely designing good visualizations that can actually show if synthetic and
+observed data are similar or not.
+
+For our example Poisson promoter model then, we merely draw many random numbers,
+say 1000, from the Gaussian posterior. For each one of those draws, we generate
+a dataset from the likelihood, i.e., we draw 2648 (the number of observed cells
+in the actual dataset) Poisson-distributed numbers for each of the 1000
+posterior draws, for a total of 2648000 samples from the posterior predictive
+distribution.
+
+To compare so many samples with the actual observed data, one excellent
+visualization for 1D data is ECDFs of the quantiles, as shown for our Poisson
+model in~\fig{fig:constit_post_full}.
+In this example, the median for each possible mRNA
+count is shown as a dark green line, while shaded bands of lighter
+green contain 95\% of the posterior predictive samples.
+Plotting quantiles in this way gives us a sense of the range of data we might
+consider plausible, under the assumption that the model is true. In this case it
+is quite obvious that the observed data, plotted in orange, could not plausibly
+come from this Poisson generative model. The other model with a PPC plotted
+in~\fig{fig:constit_post_full}, bursty model 5 with
+a negative binomial steady-state mRNA
+distribution, looks like a good candidate by eye. We cover it in more detail
+later.
+
+\subsubsection{Model 5 - Bursty promoter}
+Let us now consider the problem of parameter inference from FISH data for model
+five from~\fig{fig1:means_cartoons}(C). As derived in
+Appendix~\ref{sec:gen_fcn_appdx}, the steady-state mRNA distribution in this
+model is a negative binomial distribution, given by
+\begin{equation}
+p(m) = \frac{\Gamma(m+k_i)}{\Gamma(m+1)\Gamma(k_i)}
+        \left(\frac{1}{1+b}\right)^{k_i}
+        \left(\frac{b}{1+b}\right)^m,
+\label{eq:neg_bionom}
+\end{equation}
+where $b$ is the mean burst size and $k_i$ is the burst rate nondimensionalized
+by the mRNA degradation rate $\gamma$. As sketched earlier, the story of this
+distribution is of geometrically-distributed bursts of mRNA, where the arrival
+of bursts is a Poisson process with rate $k_i$ and the mean size of a burst is
+$b$.
+
+As for the Poisson promoter model, this expression for the steady-state mRNA
+distribution is exactly the likelihood we want to use in Bayes theorem. Again
+denoting the single-cell mRNA count data as $D=\{m_1, m_2,\dots, m_N\}$, here
+Bayes theorem takes the form
+\begin{equation}
+p(k_i, b \mid D) \propto p(D\mid k_i,b)p(k_i, b).
+\end{equation}
+We already have our likelihood --the product of $N$ negative binomials as
+Eq.~\ref{eq:neg_bionom}--  so we only need to choose priors on $k_i$ and $b$.
+For the datasets from~\cite{Jones2014} that we are analyzing, as for the Poisson
+promoter model above we are still data-rich so the prior's influence remains
+weak, but not nearly as weak because the dimensionality of our model has
+increased from one to two.
+
+We follow the guidance of~\cite{Gelman2013}, Section 2.9 in
+opting for weakly-informative priors on $k_i$ and $b$ (conjugate
+priors do not exist for this problem), and we find
+``street-fighting estimates''~\cite{Mahajan2010} to be an ideal
+way of constructing such priors. The idea of weakly informative
+priors is to allow all remotely plausible values of model
+parameters while excluding the completely absurd or unphysical.
+
+Consider $k_i$. The fastest known bacterial promoters control
+rRNA genes and initiate transcripts no faster than $\sim
+1/\text{sec}$. It would be exceedingly strange if any of the
+constitutive promoters from~\cite{Jones2014} were stronger than
+that, so we can take that as an upper bound. For a lower bound,
+if transcripts are produced too rarely, there would be nothing to
+see with FISH. The datasets for each strain contain of order
+$10^3$ cells, and if the $\langle m \rangle = k_i b/\gamma
+\lesssim 10^{-2}$, then the total number of expected mRNA
+detections would be single-digits or less and we would have
+essentially no data on which to carry out inference. So assuming
+$b$ is not too different from 1, justified next, and an mRNA
+lifetime of $\gamma^{-1}\sim 3-5~\text{min}$, this gives us soft
+bounds on $k_i/\gamma$ of perhaps $10^{-2}$ and $3\times 10^1$.
+
+Next consider mean burst size $b$. This parametrization of the
+geometric distribution allows bursts of size zero (which could
+representing aborted transcripts and initiations), but it would
+be quite strange for the mean burst size $b$ to be below
+$\sim10^{-1}$, for which nearly all bursts would be of size zero
+or one. For an upper bound, if transcripts are initiating at a
+rate somewhat slower than rRNA promoters, then it would probably
+take a time comparable to the lifetime of an mRNA to produce a
+burst larger than 10-20 transcripts, which would invalidate the
+approximation of the model that the duration of bursts are
+instantaneous compared to other timescales in the problem. So we
+will take soft bounds of $10^{-1}$ and $10^1$ for $b$.
+
+Note that the natural scale for these ``street-fighting estimates''
+was a log scale. This is commonly the case that our prior sense
+of reasonable and unreasonable parameters is set on a log scale.
+A natural way to enforce these soft bounds is therefore to use a
+lognormal prior distribution, with the soft bounds set $\pm2$
+standard deviations from the mean.
+
+With this, we are ready to write our full generative model as
+\begin{equation}
+\begin{split}
+\ln k_i \sim \text{Normal}(-0.5, 2)
+\\
+\ln b \sim \text{Normal}(0.5, 1)
+\\
+m \sim \text{NBinom}(k_i, b).
+\end{split}
+\end{equation}
+We carried out MCMC sampling on the posterior of this model, starting
+with the constitutive \textit{lacUV5} dataset from~\cite{Jones2014}.
+The resulting MCMC samples are shown in~\fig{fig:constit_post_full}(A).
+In constrast to the active/inactive constitutive model considered
+in~\cite{Razo-Mejia2020} (nonequilibrium model four
+in~\ref{fig:constit_cartoons}), this model is well-identified
+with both parameters determined to a fractional uncertainty of 5-10\%.
+The strong correlation reflects the fact that their product sets
+the mean of the mRNA distribution, which is tightly constrained
+by the data, but there is weak ``sloppiness''~\cite{Transtrum2015}
+along a set of values with a similar product.
+
+Flush with success having found the model's posterior to be
+well-identified, the next step was posterior predictive sampling.
+The only procedural difference compared to the Poisson promoter
+model is that, since we have no analytical posterior expression
+but only samples from the posterior, we generate a synthetic
+dataset (i.e., posterior predictive samples) for each posterior
+sample from the MCMC sampling. \fig{fig:constit_post_full}(B)
+shows the resulting predictive ECDF; similarly to the Poisson
+promoter model, the median posterior predictive ECDF is plotted
+as a dark blue line, while a lighter blue shaded region
+encompasses 95\% of the posterior predictive samples. Unlike the
+Poisson promoter model, the experimental ECDF closely tracks the
+posterior predictive ECDF, indicating this model is actually able
+to generate the observed data and increasing our confidence that
+this model is at least not wrong.
+
+\textit{lacUV5} is our primary target here, since it forms the
+core of all the simple repression constructs of~\cite{Jones2014}
+that we consider in Section~\ref{sec:rep_kinetics_inference}.
+Nevertheless, we thought it wise to apply our bursty promoter
+model to the other 17 constitutive promoters available in the
+FISH dataset from~\cite{Jones2014} as a test that the model is
+capturing the essential phenomenology. If the model fit well to
+all the different promoters, this would increase our confidence
+that it would serve well as a foundation for inferring repressor
+kinetics later in Section~\ref{sec:rep_kinetics_inference}.
+Conversely, were the model to fail on more than a couple of the
+other promoters, it would give us pause.
+
+\begin{figure}%[h!]
+\centering
+\includegraphics[width=\textwidth]{../figures/main/fig03.pdf}
+\caption{\textbf{Constitutive promoter posterior inference and model
+        comparison.}
+        (A) The joint posterior density of model 5, the
+        bursty promoter with negative binomially-distributed
+        steady state, is plotted with MCMC samples.
+        1D marginal densities are plotted as flanking histograms.
+        The model was fit on \textit{lacUV5} data from~\cite{Jones2014}.
+        (B) The ECDF of the observed population distribution of
+        mRNA transcripts under the control of a constitutive
+        lacUV5 promoter is shown in orange.
+        The median posterior predictive ECDFs for models (1),
+        Poisson, and (5), negative binomial, are plotted in
+        green and blue, respectively.
+        Lighter green and blue regions enclose 95\% of all
+        posterior predictive samples from their respective models.
+        Model (1) is in obvious contradiction with the data
+        while model (5) is not.
+        FISH data is again from~\cite{Jones2014}.
+    (C) Joint posterior distributions for burst rate $k_i$ and mean burst size
+    $b$ for 18 promoters from~\cite{Jones2014}. Each contour indicates the 95\%
+    highest posterior density region for a particular promoter. Note that the
+    vertical axis is shared with (D). Panel (D) plots the burst rate $k_i$ vs.\
+    the binding energy for each promoter from energy matrices
+    from~\cite{Brewster2012}.
+    The dotted line shows the predicted slope according to
+    $\ln k_i/\gamma \sim \text{constant} - \beta\Delta\epsilon_P$,
+    described in text.}
+\label{fig:constit_post_full}
+\end{figure}
+
+\fig{fig:constit_post_full}(C) shows the results, plotting the
+posterior distribution from individually MCMC sampling all 18
+constitutive promoter datasets from~\cite{Jones2014}. To aid
+visualization, rather than plotting samples for each promoter's
+posterior as we did in \fig{fig:constit_post_full}(A), for each
+posterior we find and plot the curve that surrounds the 95\%
+highest probability density region, i.e., on a plot analogous
+to~(A), each contour would enclose approximately 95\% of the
+samples, and thus 95\% of the probability mass,
+of its posterior distribution.
+Posterior predictive ECDFs (not shown) display a similar level of
+agreement between data and predictive samples as for the bursty
+model with \textit{lacUV5} in \fig{fig:constit_post_full}(B).
+\mmnote{Might be worth showing some/all of these PPCs in
+supplement for completeness? We could do postage stamp sized diff
+plots and probably fit all 18 promoters on 1-2 pages.}
+
+One interesting feature from \fig{fig:constit_post_full}(C) is
+that burst rate varies far more widely, over a range of
+$\sim10^2$, than burst size, confined to a range of
+$\lesssim10^1$ (and with the exception of promoter 6, just a span
+of 3-5x). This suggests that $k_i$, not $b$, is the key dynamic
+variable that promoter binding site tunes.
+
+It is interesting to connect this observation to the work
+of~\cite{Brewster2012}, where these same 18 promoters were
+considered through the lens of the three-state equilibrium model
+and binding energies $\Delta\epsilon_P$ were predicted from an
+energy matrix model derived from~\cite{Kinney2010}. The connection
+is made simply by equating mean mRNA levels in both theories,
+\mmnote{derivation is hasty, needs elaboration}
+\begin{equation}
+\langle m \rangle = \frac{k_i b}{\gamma}
+        = \frac{r}{\gamma}
+        \frac{\frac{P}{N_{NS}}\exp(-\beta\Delta\epsilon_P)}
+                {1+\frac{P}{N_{NS}}\exp(-\beta\Delta\epsilon_P)},
+\end{equation}
+and taking the weak-promoter limit in the equilibrium model,
+\begin{equation}
+\langle m \rangle = \frac{k_i b}{\gamma}
+        = \frac{r}{\gamma} \frac{P}{N_{NS}}\exp(-\beta\Delta\epsilon_P),
+\end{equation}
+valid for all the binding energies considered here.
+
+How are the two coarse-grainings related? The only association that makes dimensional sense and produces the correct order-of-magnitude for the known parameters is to take
+\begin{equation}
+\frac{k_i}{\gamma} = \exp(-\beta\Delta\epsilon_P)
+\label{eq:bursty_equil_corresp1}
+\end{equation}
+and
+\begin{equation}
+b = \frac{r}{\gamma}\frac{P}{N_{NS}}.
+\label{eq:bursty_equil_corresp2}
+\end{equation}
+\fig{fig:constit_post_full}(D) shows that this linear scaling
+between $\ln k_i$ and $-\beta\Delta\epsilon_P$ is approximately
+true for all 18 constitutive promoters considered. The intercept
+is an uninteresting function of the unknown model parameters, so
+we only consider the slope of the scaling in
+\fig{fig:constit_post_full}(D).
+\mmnote{Wait, is that true or can we do something with the intercept, related to the burst size in some useful way? Needs more careful thought.}
+
+While the associations represented by
+\eq{eq:bursty_equil_corresp1} and \eq{eq:bursty_equil_corresp2}
+appear to be born out by the scaling $\ln k_i \sim
+-\beta\Delta\epsilon_P$, we do not find the association of
+parameters encoded by
+\eqrange{eq:bursty_equil_corresp1}{eq:bursty_equil_corresp2}
+intuitive. If some deeper insight is waiting to be gleaned from it,
+we leave it as an open problem.
+
+\mmnote{Outline, still to cover:
+\begin{itemize}
+\item All 18 promoters. Hey the model still works on things other than UV5,
+cool. Interesting and somewhat counterintuitive scaling b/w burst rate and
+binding energy. Puzzle in comparing w/ Chong2014 supercoiling model. Are we
+missing important things?? Unclear, we leave as open question. Good enough for
+our purposes: posterior is identifiable, PPC is great, and of the models we've
+thought of it is unique in satisfying both.
+\end{itemize}
+}
+
+\mmnote{Discuss puzzling comparison w/ Chong2014: if supercoiling is the thing, why are my burst sizes all the same but burst rates vary?
+And why is the duty cycle of my promoter so low, ie., bursts so short?
+COmpare their fig 7E; their $\beta/\alpha$ is my $k^+/k^-$.
+They have one or two genes with very small $\beta/\alpha$, does the \textit{galK} locus just happen to be that, or is there a deeper disagreement?
+Hard to say w/o more data.}
+
+\subsection{Transcription factor kinetics can be inferred from FISH

### 2020-05-21 doc/section_04_bayesian_inference.tex
MRM pass through section 4 on 20200521

@@
-% !TEX root = ./busty_transcription.tex
-\section{Finding the ``right'' model: Bayesian parameter inference and model selection}
+% !TEX root = ./bursty_transcription.tex
+\section{Finding the ``right'' model: Bayesian parameter inference and model 
+selection}
 \subsection{Parameter inference for constitutive promoters}
@@
 From consideration of Fano factors in the previous section, we suspect that
-model 5 in~\fig{fig:constit_cartoons}, a one-state bursty model of constitutive
-promoters, achieves a Goldilocks level of complexity. Does this stand up to
-closer scrutiny, namely, comparison to full mRNA distributions rather than
-simply their moments? We will test this thoroughly on all the data from
-constitutive promoters from Jones et.\ al.~\cite{Jones2014}.
+model 5 in Figure~\ref{fig:constit_cartoons}, a one-state bursty model of
+constitutive promoters, achieves a Goldilocks level of complexity. Does this
+stand up to closer scrutiny, namely, comparison to full mRNA distributions
+rather than simply their moments? We will test this thoroughly on
+single-cell mRNA counts for different unregulated promoters from Jones et.\
+al.~\cite{Jones2014}.
 
 It will be instructive, however, to first consider the Poisson promoter, model 1
-in~\fig{fig:constit_cartoons}. As we alluded to earlier, since the Poisson
-distribution has its Fano factor $\nu$ strictly equal to 1,
-and all of the observed data has Fano factor $\nu>1$, we might
-already suspect that this model is incapable of fitting the data.
-We will verify that this in fact the case. As a consequence
-of ruling out model 1, models 2 and 3 will also be ruled out.
-Why? Again all the data have $\nu>1$, meaning the mRNA distributions
-are overdisperse compared with a Poisson distribution,
-while models 2 and 3 have Fano factors $\nu\le 1$ meaning they
-are underdisperse relative to Poisson. So models 2 and 3
-can fare no better than the simple Poisson model. We will also not explicitly
+in Figure~\ref{fig:constit_cartoons}. As we alluded to earlier, since the
+Poisson distribution has a Fano factor $\nu$ strictly equal to 1, and all of
+the observed data has Fano factor $\nu>1$, we might already suspect that this
+model is incapable of fitting the data. We will verify that this in fact the
+case. Using the same argument we can immediately rule out models 2 and 3 from 
+Figure~\ref{fig:constit_cartoons}. these models have Fano factors $\nu\le 1$
+meaning they are underdisperse relative to Poisson. We will also not explicitly
 consider model 4 from~\fig{fig:constit_cartoons} since it was already thoroughly
@@
 
-In performing parameter inference on the FISH data of~\cite{Jones2014}, building
-a generative model is straightforward. The data are single-cell mRNA counts, and
-we assume different cells to be independent and identically distributed from the
-underlying generative model, which is just the steady-state mRNA distribution
-computed from the master equation of the underlying model.
+In performing parameter inference on the single-molecule data
+of~\cite{Jones2014}, building a generative model is straightforward. The data
+are single-cell mRNA counts, and we assume different cells to be independent and
+identically distributed from the underlying generative model. The generative
+model for our case is the steady-state mRNA distribution computed from the
+master equation under consideration.
 
 \subsubsection{Model 1: Poisson promoter}
-\mmnote{A lot of the details I've written here might be excessive for main text and serve better as an intro-by-example to Bayes in the appendix, then reduce main text to a bit more bare bones?}
-For this model the master equation of interest is~\eq{eq:poisson_promoter_cme}
-with repressor set to zero, i.e.,
+% \mmnote{A lot of the details I've written here might be excessive for main text and serve better as an intro-by-example to Bayes in the appendix, then reduce main text to a bit more bare bones?}
+For this model the master equation of interest is
+Eq.~\ref{eq:poisson_promoter_cme} with repressor set to zero, i.e.,
 \begin{equation}
@@
 \end{equation}
-whose steady-state distribution is Poisson, given by
-\begin{equation}
-p_U(m) = \frac{\lambda^m e^{-\lambda}}{m!}
-\end{equation}
-with $\lambda=r/\gamma$. This steady state distribution defines our likelihood
-$p(m\mid\lambda)$, i.e.
+whose steady-state solution is given by a Poisson distribution with parameter
+$\lambda \equiv r / \gamma$. This steady state distribution defines the
+likelihood $p(m\mid\lambda)$ for our Bayesian inference, i.e. 
 \begin{equation}
@@
 \end{equation}
-is the probability of observing a single cell with $m$ mRNAs given a known value
-of $\lambda$, written to make the conditioning explicit.
+is the probability of observing a single cell with $m$ mRNAs given a value of
+$\lambda$, written to make the conditioning explicit.
 
@@
 only as a normalization factor. Since we assume each cell's mRNA count is
-independent of others, the likelihood is simply a product of the single cell
-likelihoods given by \eq{eq:poisson_inference010} above, so
+independent of others, the likelihood of the full inference problem
+$p(D\mid\lambda)$ is simply a product of the single cell likelihoods given by
+\eq{eq:poisson_inference010} above, so
 \begin{equation}
@@
 \end{equation}
-which is read as the data $D$ is Poisson
-distributed with mean $\lambda$. The $\sim$ operator literally denotes ``is
-distributed according to.''
-The conditioning on $\lambda$ is no longer explictily notated,
-only implied by its appearance on the right hand side.
+which is read as the data $D$ is Poisson distributed with mean $\lambda$. The
+$\sim$ operator literally denotes ``is distributed according to.'' The
+conditioning on $\lambda$ is no longer explicitly notated, only implied by its
+appearance on the right hand side.
 
@@
 single-cell measurements for each promoter, so our choice of prior matters
-little here, as long as it does not exclude the true value of $\lambda$. A
-convenient choice for our problem is to use a \textit{conjugate} prior.
-A conjugate prior is a special prior that causes the posterior to
-have the same functional form as the prior, simply with updated
-model parameters. This makes calculations
-analytically tractable and also offers a nice interpretation of the inference
-procedure as updating our knowledge about the model parameters.
-This makes conjugate priors very useful when they exist.
-The caveat is that conjugate priors only exist for a very limited
-number of likelihoods, mostly with only one or two model
-parameters, so in almost all other Bayesian inference problems,
-we must tackle the posterior numerically.
+little here, as long as it is sufficiently broad. For details on the prior 
+selection we refer the reader to Appendix~\ref{sec:bayesian}. For our
+purpose here it suffices to specify that we use a prior of the form $\lambda
+\sim \text{Gamma}(\alpha, \beta)$. Putting a gamma prior on $\lambda$ introduces
+two new parameters $\alpha$ and $\beta$ which parametrize the gamma distribution
+itself, which we use to encode the range of $\lambda$ values we view as
+reasonable. Recall $\lambda$ is the mean steady-state mRNA count per cell, which
+\textit{a priori} could plausibly be anywhere from 0 to a few hundred.
+$\alpha=1$ and $\beta=1/50$ achieve this, since the gamma distribution is
+strictly positive with mean $\alpha/\beta$ and standard deviation
+$\sqrt{\alpha}/\beta$.
 
-But, for the problem at hand, a conjugate prior does in fact exist.
-For a Poisson likelihood of identical and identically distributed
-data, the conjugate prior is a gamma distribution, as
-can be looked up in, e.g.,~\cite{Gelman2013}, Section 2.6.
-Putting a gamma prior on $\lambda$ introduces two new parameters $\alpha$ and
-$\beta$ which parametrize the gamma distribution itself, which we use to encode
-the range of $\lambda$ values we view as reasonable. Recall $\lambda$ is the
-mean steady-state mRNA count per cell, which \textit{a priori} could plausibly
-be anywhere from 0 to a few hundred. $\alpha=1$ and $\beta=1/50$ achieve this,
-since the gamma distribution is strictly positive with mean $\alpha/\beta$ and
-standard deviation $\sqrt{\alpha}/\beta$. To be explicit, then, our prior is
-\begin{equation}
-\lambda \sim \text{Gamma}(\alpha, \beta)
-\end{equation}
-
-As an aside, note that if we did not know that our prior was a conjugate prior,
-we could still write down our posterior distribution from its definition as
-\begin{equation}
-p(\lambda\mid D,\alpha,\beta)
-\propto p(D\mid\lambda) p(\lambda \mid\alpha,\beta)
-\propto \left(\prod_{k=1}^N \frac{\lambda^{m_k}e^{-\lambda}}{m_k!}\right)
-        \frac{\beta}{\Gamma(\alpha)}(\beta\lambda)^{\alpha-1} e^{-\beta\lambda}
-.
-\end{equation}
-Without foreknowledge that this in fact reduces to a gamma distribution, this
-expression might appear rather inscrutable. When conjugate priors are
-unavailable for the likelihood of interest - which is almost always the case for
-models with $>1$ model parameter - this inscrutability is the norm, and making
-sense of posteriors analytically is almost always impossible. Fortunately, MCMC
-sampling provides us a powerful method of constructing posteriors numerically
-which we will make use of extensively.
-
-Since we did use a conjugate prior, we may simply look up our posterior in any
-standard reference such as~\cite{Gelman2013}, Section 2.6,
-from which we find that
+As detailed in Appendix~\ref{sec:bayesian} this particular choice of prior
+known as the \textit{conjugate} prior for a Poisson likelihood allows us to 
+get a closed form for the posterior distribution $p(\lambda \mid D)$. In 
+particular it can be shown that~\cite{Gelman2013}
 \begin{equation}
@@
 where we defined the sample mean $\bar{m} = \frac{1}{N}\sum_k m_k$ for
-notational convenience. A glance at the FISH data from~\cite{Jones2014} reveals
-that $N$ is $\mathcal{O}(10^3)$ and $\langle m\rangle \gtrsim 0.1$ for all
-constitutive strains in~\cite{Jones2014}, so $\bar{m}N \gtrsim 10^2$. Therefore
-as we suspected, our prior parameters are completely overwhelmed by the data.
-The prior behaves, in a sense, like $\beta$ extra ``data points''
-with a mean value of $(\alpha-1)/\beta$~\cite{Gelman2013}, which
-gives us some intuition for how much data is needed to overwhelm
-the prior in this case: enough data $N$ such that $\beta\ll N$
-and $\alpha/\beta \ll \bar{m}$. In
-fact, $\bar{m}N$ and $N$ are so large that we can, to an excellent
-approximation, ignore the $\alpha$ and $\beta$ dependence and approximate the
-gamma distribution as a Gaussian with mean $\bar{m}$ and standard deviation
-$\sqrt{\bar{m}/N}$, giving
+notational convenience. Furthermore given that $N$ is $\mathcal{O}(10^3)$ and
+$\langle m\rangle \gtrsim 0.1$ for all promoters measured in~\cite{Jones2014}
+our data easily overwhelms the choice of prior, and allows us to make the 
+excellent approximation 
 \begin{equation}
@@
 \sim \text{Normal}\left(18.7, 0.08\right),
+\label{eq:posterior_poisson}
 \end{equation}
@@
 
-This is not wrong, but it is not the full story. The model's posterior is
-tightly constrained, but is it a good generative model? In other words, does the
-model generate data that look similar to our actual data, and is it therefore
-plausible that the model captures the important features of the data generating
-process? This intuitive notion can be codified with \textit{posterior predictive
-checks}, or PPCs, and we will see that this simple Poisson model fails badly.
+\mrm{
+We alert the reader not to confuse a tightly constrained posterior distribution
+with a good generative model. In other words, not because the single parameter
+for our Poisson likelihood can only take a small range of values constrained by
+the data, this implies that such Poisson distribution characterizes the
+underlying cell-to-cell variability we observe in the data. To systematically
+check the validity of the generative model we can make use of \textit{posterior
+predictive checks}. These consists on generating synthetic data using our
+generative model for many plausible values of our parameter. For the case of our
+\textit{lacUV5} promoter this means that we sample a value for $\lambda$ out of
+the posterior distribution (Eq.~\ref{eq:posterior_poisson}), and generate 2648
+samples out of a Poisson distribution with such $\lambda$ value. We then repeat
+this procedure 1000 times. Clearly if those data appear quite different, the
+model has a problem. Put another way, if we suppose the generative model is
+true, then the synthetic datasets we generate should resemble the actual
+observed data, and if not, it suggests the model is missing important features.}
 
-The intuitive idea of posterior predictive checks is simple: \mmnote{Not sure if
-all this stuff is clearer by example, but then does it confuse the reader in
-generalizing to other models?}
-\begin{enumerate}
-\item Make a random draw of the model parameter $\lambda$ from the posterior
-distribution.
-\item Plug that draw into the likelihood and generate a synthetic dataset
-$\{m_k\}$ conditioned on $\lambda$.
-\item Repeat many times.
-\end{enumerate}
-More formally, the posterior predictive distribution can be thought of as the
-distribution of future yet-to-be-observed data, conditioned on the data we have
-already observed. Clearly if those data appear quite different, the model has a
-problem. Put another way, if we suppose the generative model is true, then the
-synthetic datasets we generate should resemble the actual observed data, and if
-not, it suggests the model is missing important features. All the data we
-consider in this work are 1D (distributions of mRNA counts over a population) so
-ECDFs are an excellent visual means of comparing synthetic and observed
-datasets. In general for higher dimensional datasets, much of the challenge is
-in merely designing good visualizations that can actually show if synthetic and
-observed data are similar or not.
+\mrm{
+To compare so many samples with the actual observed data, we can use empirical
+cumulative distribution functions (ECDF) of the distribution quantiles. This 
+representation is shown in Figure~\ref{fig:constit_post_full}(B). In this 
+example, the median for each possible mRNA count for the Poisson distribution
+is shown as a dark green line, while the lighter green contains 95\% of the
+posterior predictive samples. This way of representing the posterior predictive
+checks gives us a sense of the range of data we might consider plausible, under
+the assumption that the model is true. In this case it is quite obvious that the
+observed data, plotted in orange, could not plausibly come from this Poisson
+generative model. We will cover the other model shown in
+Figure~\ref{fig:constit_post_full}(B) next.
+}
 
-For our example Poisson promoter model then, we merely draw many random numbers,
-say 1000, from the Gaussian posterior. For each one of those draws, we generate
-a dataset from the likelihood, i.e., we draw 2648 (the number of observed cells
-in the actual dataset) Poisson-distributed numbers for each of the 1000
-posterior draws, for a total of 2648000 samples from the posterior predictive
-distribution.
-
-To compare so many samples with the actual observed data, one excellent
-visualization for 1D data is ECDFs of the quantiles, as shown for our Poisson
-model in~\fig{fig:constit_post_full}.
-In this example, the median for each possible mRNA
-count is shown as a dark green line, while shaded bands of lighter
-green contain 95\% of the posterior predictive samples.
-Plotting quantiles in this way gives us a sense of the range of data we might
-consider plausible, under the assumption that the model is true. In this case it
-is quite obvious that the observed data, plotted in orange, could not plausibly
-come from this Poisson generative model. The other model with a PPC plotted
-in~\fig{fig:constit_post_full}, bursty model 5 with
-a negative binomial steady-state mRNA
-distribution, looks like a good candidate by eye. We cover it in more detail
-later.
-
-\subsubsection{Nonequilibrium Model Five - Bursty promoter}
-Let us now consider the problem of parameter inference from FISH data
-for model five from~\fig{fig1:means_cartoons}(C). As derived
-in Appendix~\ref{sec:gen_fcn_appdx}, the steady-state mRNA distribution
-in this model is a negative binomial distribution,
-given by~\eq{eq:nbinom_deriv_final}, which is
+\subsubsection{Model 5 - Bursty promoter}
+Let us now consider the problem of parameter inference from FISH data for model
+five from~\fig{fig1:means_cartoons}(C). As derived in
+Appendix~\ref{sec:gen_fcn_appdx}, the steady-state mRNA distribution in this
+model is a negative binomial distribution, given by
 \begin{equation}
@@
         \left(\frac{b}{1+b}\right)^m,
+\label{eq:neg_bionom}
 \end{equation}
-where $b$ is the mean burst size and $k_i$ is the burst rate
-nondimensionalized by the mRNA degradation rate $\gamma$.
-As sketched earlier, the story of this distribution is of
-geometrically-distributed bursts of mRNA, where the arrival of bursts is a
-Poisson process with rate $k_i$ and the mean size of a burst is $b$.
+where $b$ is the mean burst size and $k_i$ is the burst rate nondimensionalized
+by the mRNA degradation rate $\gamma$. As sketched earlier, the story of this
+distribution is of geometrically-distributed bursts of mRNA, where the arrival
+of bursts is a Poisson process with rate $k_i$ and the mean size of a burst is
+$b$.
 
-As for the Poisson promoter model, this expression for the steady-state
-mRNA distribution is exactly the likelihood we want to use in Bayes theorem.
-Again denoting the single-cell mRNA count data as $D=\{m_1, m_2,\dots, m_N\}$,
-here Bayes theorem takes the form
+As for the Poisson promoter model, this expression for the steady-state mRNA
+distribution is exactly the likelihood we want to use in Bayes theorem. Again
+denoting the single-cell mRNA count data as $D=\{m_1, m_2,\dots, m_N\}$, here
+Bayes theorem takes the form
 \begin{equation}
@@
 \end{equation}
-We already have our likelihood so we only need to choose priors
-on $k_i$ and $b$. For the datasets from~\cite{Jones2014} that we
-are analyzing, as for the Poisson promoter model above we are
-still data-rich so the prior's influence remains weak, but not
-nearly as weak because the dimensionality of our model has
+We already have our likelihood --the product of $N$ negative binomials as
+Eq.~\ref{eq:neg_bionom}--  so we only need to choose priors on $k_i$ and $b$.
+For the datasets from~\cite{Jones2014} that we are analyzing, as for the Poisson
+promoter model above we are still data-rich so the prior's influence remains
+weak, but not nearly as weak because the dimensionality of our model has
 increased from one to two.
-
-We follow the guidance of~\cite{Gelman2013}, Section 2.9 in
-opting for weakly-informative priors on $k_i$ and $b$ (conjugate
-priors do not exist for this problem), and we find
-``street-fighting estimates''~\cite{Mahajan2010} to be an ideal
-way of constructing such priors. The idea of weakly informative
-priors is to allow all remotely plausible values of model
-parameters while excluding the completely absurd or unphysical.
-
-Consider $k_i$. The fastest known bacterial promoters control
-rRNA genes and initiate transcripts no faster than $\sim
-1/\text{sec}$. It would be exceedingly strange if any of the
-constitutive promoters from~\cite{Jones2014} were stronger than
-that, so we can take that as an upper bound. For a lower bound,
-if transcripts are produced too rarely, there would be nothing to
-see with FISH. The datasets for each strain contain of order
-$10^3$ cells, and if the $\langle m \rangle = k_i b/\gamma
-\lesssim 10^{-2}$, then the total number of expected mRNA
-detections would be single-digits or less and we would have
-essentially no data on which to carry out inference. So assuming
-$b$ is not too different from 1, justified next, and an mRNA
-lifetime of $\gamma^{-1}\sim 3-5~\text{min}$, this gives us soft
-bounds on $k_i/\gamma$ of perhaps $10^{-2}$ and $3\times 10^1$.
-
-Next consider mean burst size $b$. This parametrization of the
-geometric distribution allows bursts of size zero (which could
-representing aborted transcripts and initiations), but it would
-be quite strange for the mean burst size $b$ to be below
-$\sim10^{-1}$, for which nearly all bursts would be of size zero
-or one. For an upper bound, if transcripts are initiating at a
-rate somewhat slower than rRNA promoters, then it would probably
-take a time comparable to the lifetime of an mRNA to produce a
-burst larger than 10-20 transcripts, which would invalidate the
-approximation of the model that the duration of bursts are
-instantaneous compared to other timescales in the problem. So we
-will take soft bounds of $10^{-1}$ and $10^1$ for $b$.
-
-Note that the natural scale for these ``street-fighting estimates''
-was a log scale. This is commonly the case that our prior sense
-of reasonable and unreasonable parameters is set on a log scale.
-A natural way to enforce these soft bounds is therefore to use a
-lognormal prior distribution, with the soft bounds set $\pm2$
-standard deviations from the mean.
-
-With this, we are ready to write our full generative model as
+\mrm{
+Details on the arguments behind our prior distribution selection are left for
+Appendix~\ref{sec:bayesian}. Our full generative model can be specified as
 \begin{equation}
 \begin{split}
-\ln k_i \sim \text{Normal}(-0.5, 2)
-\\
-\ln b \sim \text{Normal}(0.5, 1)
-\\
+\ln k_i \sim \text{Normal}(-0.5, 2) \\
+\ln b \sim \text{Normal}(0.5, 1) \\
 m \sim \text{NBinom}(k_i, b).
@@
 \end{equation}
-We carried out MCMC sampling on the posterior of this model, starting
-with the constitutive \textit{lacUV5} dataset from~\cite{Jones2014}.
-The resulting MCMC samples are shown in~\fig{fig:constit_post_full}(A).
-In constrast to the active/inactive constitutive model considered
-in~\cite{Razo-Mejia2020} (nonequilibrium model four
-in~\ref{fig:constit_cartoons}), this model is well-identified
-with both parameters determined to a fractional uncertainty of 5-10\%.
-The strong correlation reflects the fact that their product sets
-the mean of the mRNA distribution, which is tightly constrained
-by the data, but there is weak ``sloppiness''~\cite{Transtrum2015}
-along a set of values with a similar product.
+Note that we specify a distribution over the logarithm of our parameters $k_i$
+and $b$. This is commonly the case for parameters for which our previous 
+knowledge based on our domain expertise spans several orders of magnitude.
+}
 
-Flush with success having found the model's posterior to be
-well-identified, the next step was posterior predictive sampling.
-The only procedural difference compared to the Poisson promoter
-model is that, since we have no analytical posterior expression
-but only samples from the posterior, we generate a synthetic
-dataset (i.e., posterior predictive samples) for each posterior
-sample from the MCMC sampling. \fig{fig:constit_post_full}(B)
-shows the resulting predictive ECDF; similarly to the Poisson
-promoter model, the median posterior predictive ECDF is plotted
-as a dark blue line, while a lighter blue shaded region
-encompasses 95\% of the posterior predictive samples. Unlike the
-Poisson promoter model, the experimental ECDF closely tracks the
-posterior predictive ECDF, indicating this model is actually able
-to generate the observed data and increasing our confidence that
-this model is at least not wrong.
+We carried out MCMC sampling on the posterior of this model, starting with the
+constitutive \textit{lacUV5} dataset from~\cite{Jones2014}. The resulting MCMC
+samples are shown in~\fig{fig:constit_post_full}(A). In contrast to the
+active/inactive constitutive model considered in~\cite{Razo-Mejia2020}
+(nonequilibrium model 4 in~\ref{fig:constit_cartoons}), this model is
+well-identified with both parameters determined to a fractional uncertainty of
+5-10\%. The strong correlation reflects the fact that their product sets the
+mean of the mRNA distribution, which is tightly constrained by the data, but
+there is weak ``sloppiness''~\cite{Transtrum2015} along a set of values with a
+similar product.
 
-\textit{lacUV5} is our primary target here, since it forms the
-core of all the simple repression constructs of~\cite{Jones2014}
-that we consider in Section~\ref{sec:rep_kinetics_inference}.
-Nevertheless, we thought it wise to apply our bursty promoter
-model to the other 17 constitutive promoters available in the
-FISH dataset from~\cite{Jones2014} as a test that the model is
-capturing the essential phenomenology. If the model fit well to
-all the different promoters, this would increase our confidence
-that it would serve well as a foundation for inferring repressor
-kinetics later in Section~\ref{sec:rep_kinetics_inference}.
-Conversely, were the model to fail on more than a couple of the
-other promoters, it would give us pause.
+Flush with success having found the model's posterior to be well-identified, the
+next step is posterior predictive sampling. The only procedural difference
+compared to the Poisson promoter model is that, since we have no analytical
+posterior expression but only samples from the posterior, we generate a
+synthetic dataset (i.e., posterior predictive samples) for each posterior sample
+from the MCMC sampling. \fig{fig:constit_post_full}(B) shows the resulting
+predictive ECDF; similarly to the Poisson promoter model, the median posterior
+predictive ECDF is plotted as a dark blue line, while a lighter blue shaded
+region encompasses 95\% of the posterior predictive samples. Unlike the Poisson
+promoter model, the experimental ECDF closely tracks the posterior predictive
+ECDF, indicating this model is actually able to generate the observed data and
+increasing our confidence that this model is at least not wrong.
+
+\textit{lacUV5} is our primary target here, since it forms the core of all the
+simple repression constructs of~\cite{Jones2014} that we consider in
+Section~\ref{sec:rep_kinetics_inference}. Nevertheless, we thought it wise to
+apply our bursty promoter model to the other 17 unregulated promoters available
+in the single-cell mRNA count dataset from~\cite{Jones2014} as a test that the
+model is capturing the essential phenomenology. If the model fit well to all the
+different promoters, this would increase our confidence that it would serve well
+as a foundation for inferring repressor kinetics later in
+Section~\ref{sec:rep_kinetics_inference}. Conversely, were the model to fail on
+more than a couple of the other promoters, it would give us pause.
 
@@
 \caption{\textbf{Constitutive promoter posterior inference and model
-        comparison.}
-        (A) The joint posterior density of model 5, the
-        bursty promoter with negative binomially-distributed
-        steady state, is plotted with MCMC samples.
-        1D marginal densities are plotted as flanking histograms.
-        The model was fit on \textit{lacUV5} data from~\cite{Jones2014}.
-        (B) The ECDF of the observed population distribution of
-        mRNA transcripts under the control of a constitutive
-        lacUV5 promoter is shown in orange.
-        The median posterior predictive ECDFs for models (1),
-        Poisson, and (5), negative binomial, are plotted in
-        green and blue, respectively.
-        Lighter green and blue regions enclose 95\% of all
-        posterior predictive samples from their respective models.
-        Model (1) is in obvious contradiction with the data
-        while model (5) is not.
-        FISH data is again from~\cite{Jones2014}.
-    (C) Joint posterior distributions for burst rate $k_i$ and mean burst size
-    $b$ for 18 promoters from~\cite{Jones2014}. Each contour indicates the 95\%
-    highest posterior density region for a particular promoter. Note that the
-    vertical axis is shared with (D). Panel (D) plots the burst rate $k_i$ vs.\
-    the binding energy for each promoter from energy matrices
-    from~\cite{Brewster2012}.
-    The dotted line shows the predicted slope according to
-    $\ln k_i/\gamma \sim \text{constant} - \beta\Delta\epsilon_P$,
-    described in text.}
+comparison.} (A) The joint posterior density of model 5, the bursty promoter
+with negative binomially-distributed steady state, is plotted with MCMC samples.
+1D marginal probability densities are plotted as flanking histograms. The model
+was fit on \textit{lacUV5} data from~\cite{Jones2014}. (B) The empirical
+distribution function (ECDF) of the observed population distribution of mRNA
+transcripts under the control of a constitutive \textit{lacUV5} promoter is
+shown in orange. The median posterior predictive ECDFs for models (1), Poisson,
+and (5), negative binomial, are plotted in dark green and dark blue,
+respectively. Lighter green and blue regions enclose 95\% of all posterior
+predictive samples from their respective models. Model (1) is in obvious
+contradiction with the data while model (5) is not. single-cell mRNA count data
+is again from~\cite{Jones2014}. (C) Joint posterior distributions for burst rate
+$k_i$ and mean burst size $b$ for 18 unregulated promoters
+from~\cite{Jones2014}. Each contour indicates the 95\% highest posterior
+probability density region for a particular promoter. Note that the vertical
+axis is shared with (D). (D) Plots of the burst rate $k_i$ vs.\ the binding
+energy for each promoter as predicted in~\cite{Brewster2012}. The dotted line
+shows the predicted slope according to $\ln k_i/\gamma \sim \text{constant} -
+\beta\Delta\varepsilon_P$, described in text. Each individual promoter is
+labeled with a unique number in both (C) and (D) for cross comparison.}
 \label{fig:constit_post_full}
@@
 
-\fig{fig:constit_post_full}(C) shows the results, plotting the
-posterior distribution from individually MCMC sampling all 18
-constitutive promoter datasets from~\cite{Jones2014}. To aid
-visualization, rather than plotting samples for each promoter's
-posterior as we did in \fig{fig:constit_post_full}(A), for each
-posterior we find and plot the curve that surrounds the 95\%
-highest probability density region, i.e., on a plot analogous
-to~(A), each contour would enclose approximately 95\% of the
-samples, and thus 95\% of the probability mass,
-of its posterior distribution.
-Posterior predictive ECDFs (not shown) display a similar level of
-agreement between data and predictive samples as for the bursty
-model with \textit{lacUV5} in \fig{fig:constit_post_full}(B).
-\mmnote{Might be worth showing some/all of these PPCs in
-supplement for completeness? We could do postage stamp sized diff
-plots and probably fit all 18 promoters on 1-2 pages.}
+\fig{fig:constit_post_full}(C) shows the results, plotting the posterior
+distribution from individually MCMC sampling all 18 constitutive promoter
+datasets from~\cite{Jones2014}. To aid visualization, rather than plotting
+samples for each promoter's posterior as in \fig{fig:constit_post_full}(A), for
+each posterior we find and plot the curve that surrounds the 95\% highest
+probability density region. What this means is that each contour 
+encloses approximately 95\% of the samples, and thus 95\% of the probability
+mass, of its posterior distribution. Posterior predictive ECDFs (not shown)
+display a similar level of agreement between data and predictive samples as for
+the bursty model with \textit{lacUV5} in \fig{fig:constit_post_full}(B).
+\mmnote{Might be worth showing some/all of these PPCs in supplement for
+completeness? We could do postage stamp sized diff plots and probably fit all 18
+promoters on 1-2 pages.}\mrm{Definitely need to show them in SI.}
 
-One interesting feature from \fig{fig:constit_post_full}(C) is
-that burst rate varies far more widely, over a range of
-$\sim10^2$, than burst size, confined to a range of
-$\lesssim10^1$ (and with the exception of promoter 6, just a span
-of 3-5x). This suggests that $k_i$, not $b$, is the key dynamic
-variable that promoter binding site tunes.
+One interesting feature from \fig{fig:constit_post_full}(C) is that burst rate
+varies far more widely, over a range of $\sim10^2$, than burst size, confined to
+a range of $\lesssim10^1$ (and with the exception of promoter 6, just a span of
+3-5x). This suggests that $k_i$, not $b$, is the key dynamic variable that
+promoter sequence tunes.
 
 It is interesting to connect this observation to the work
-of~\cite{Brewster2012}, where these same 18 promoters were
-considered through the lens of the three-state equilibrium model
-and binding energies $\Delta\epsilon_P$ were predicted from an
-energy matrix model derived from~\cite{Kinney2010}. The connection
-is made simply by equating mean mRNA levels in both theories,
-\mmnote{derivation is hasty, needs elaboration}
+of~\cite{Brewster2012}, where these same 18 promoters were considered through
+the lens of the three-state equilibrium model (model 2 in
+Figure~\ref{fig1:means_cartoons}(B)) and binding energies $\Delta\varepsilon_P$
+were predicted from an energy matrix model derived from~\cite{Kinney2010}. As
+previously discussed the thermodynamic models of gene regulation can only make
+statements about the mean gene expression. This implies that we can draw the
+connection between both frameworks by equating the mean mRNA $\left\langle m \right\rangle$. This results in
 \begin{equation}
@@
         = \frac{r}{\gamma}
-        \frac{\frac{P}{N_{NS}}\exp(-\beta\Delta\epsilon_P)}
-                {1+\frac{P}{N_{NS}}\exp(-\beta\Delta\epsilon_P)},
+        \frac{\frac{P}{N_{NS}}\exp(-\beta\Delta\varepsilon_P)}
+                {1+\frac{P}{N_{NS}}\exp(-\beta\Delta\varepsilon_P)}.
 \end{equation}
-and taking the weak-promoter limit in the equilibrium model,
+By taking the weak promoter approximation for the equilibrium model ($P/N_{NS} 
+\exp(-\beta\Delta\varepsilon_r \ll 1)$) results in
 \begin{equation}
 \langle m \rangle = \frac{k_i b}{\gamma}
-        = \frac{r}{\gamma} \frac{P}{N_{NS}}\exp(-\beta\Delta\epsilon_P),
+        = \frac{r}{\gamma} \frac{P}{N_{NS}}\exp(-\beta\Delta\varepsilon_P),
 \end{equation}
@@
 
-How are the two coarse-grainings related? The only association that makes dimensional sense and produces the correct order-of-magnitude for the known parameters is to take
+Given this result, how are the two coarse-grainings related? The only
+association that makes dimensional sense and produces the correct
+order-of-magnitude for the known parameters is to take
 \begin{equation}
-\frac{k_i}{\gamma} = \exp(-\beta\Delta\epsilon_P)
+\frac{k_i}{\gamma} = \exp(-\beta\Delta\varepsilon_P)
 \label{eq:bursty_equil_corresp1}
@@
 \end{equation}
-\fig{fig:constit_post_full}(D) shows that this linear scaling
-between $\ln k_i$ and $-\beta\Delta\epsilon_P$ is approximately
-true for all 18 constitutive promoters considered. The intercept
-is an uninteresting function of the unknown model parameters, so
-we only consider the slope of the scaling in
+Figure~\ref{fig:constit_post_full}(D) shows that this linear scaling between
+$\ln k_i$ and $-\beta\Delta\varepsilon_P$ is approximately true for all 18
+constitutive promoters considered. The intercept is an uninteresting function of
+the unknown model parameters, so we only consider the slope of the scaling in
 \fig{fig:constit_post_full}(D).
@@
 While the associations represented by
-\eq{eq:bursty_equil_corresp1} and \eq{eq:bursty_equil_corresp2}
-appear to be born out by the scaling $\ln k_i \sim
--\beta\Delta\epsilon_P$, we do not find the association of
-parameters encoded by
+Eq.~\ref{eq:bursty_equil_corresp1} and Eq.~\ref{eq:bursty_equil_corresp2}
+appear to be born out by the scaling $\ln k_i \sim -\beta\Delta\varepsilon_P$,
+we do not find the association of parameters encoded by
 \eqrange{eq:bursty_equil_corresp1}{eq:bursty_equil_corresp2}
-intuitive. If some deeper insight is waiting to be gleaned from it,
-we leave it as an open problem.
+intuitive. If some deeper insight is waiting to be gleaned from it, we leave it
+as an open problem.
 

### 2020-05-21 doc/section_03_beyond_means.tex
RP Beyond Means Thursday
I have finished a first pass through beyond the Means.

@@
 \centering
-\includegraphics[width=0.9\textwidth]{../figures/fig1.5/fig1point5.ai}
-\caption{\textbf{Constitutive promoter cartoon comparison.}
-The left column depicts various plausible cartoons for constitutive promoters.
-We also list literature references where models have been used previously. In
+\includegraphics[width=\textwidth]{../figures/main/fig02.pdf}
+\caption{\textbf{Comparison of different models for noise in the constitutive promoter.}
+The left column depicts various plausible models for the dynamics of constitutive promoters.
+%We also list literature references where models have been used previously. 
+In
 model (1), transcripts are produced in a Poisson process~\cite{Jones2014}. Model
@@
 certain limit of model (4) in which transcripts are produced in bursts, and
-initiation of bursts is a Poisson process. \mmnote{Haven't found model 5
-formulated quite like this in the literature, usually burstiness is handled like
-model 4, but I should probably search a bit more.} The right column shows the
+initiation of bursts is a Poisson process. 
+%\mmnote{Haven't found model 5
+%formulated quite like this in the literature, usually burstiness is handled like
+%model 4, but I should probably search a bit more.} 
+The right column shows the
 Fano factor for each model. Note especially the crucial diagnostic: (2) and (3)
@@
 
-\subsubsection{Model 2 - Two-state promoter, RNAP bound or unbound.}
-This model was considered in, e.g.,~\cite{Phillips2015a} and
-~\cite{Phillips2019}. Here transcription initiation proceeds from the bound to
+\subsubsection{Noise in the Two-State Promoter, RNAP Bound or Unbound.}
+Our second model of constitutive transcription posits an architecture
+in which the promoter is either empty or bound by RNAP
+~\cite{Phillips2015a, Phillips2019}. Here,  as shown in 
+the second row of Figure~\ref{fig:constit_cartoons},  transcription initiation results in
+a state transition from the bound to
 the unbound state, reflecting the microscopic reality that an RNAP that has
 begun to elongate a transcript is no longer available at the start site to begin
-another. The problem with this picture is that the Fano factor is
+another.  As shown in the appendix, the Fano factor in this model is given by
 \begin{align}
@@
             {\left(\rate{k}{P}{+} + \rate{k}{P}{-} + r\right)
-             \left(\gamma + \rate{k}{P}{+} + \rate{k}{P}{-} + r\right)},
+             \left(\gamma + \rate{k}{P}{+} + \rate{k}{P}{-} + r\right)}.
 \end{align}
-which is always $<1$. To make contact with the experimental reality of $\nu>1$,
-we will have to do, at a minimum, the same corrections to Poisson behavior as in
-model 1 above. So while this model adds an appealing element of microscopic
+The problem with this picture is that the Fano factor is
+ always $<1$. 
+ To make contact with the experimental reality of $\nu>1$,
+clearly some corrections will be needed.
+While this model adds an appealing element of microscopic
 reality, we are forced to reject it as the additional complexity is unable to
@@
 
-\subsubsection{Model 3 - Three-state promoter, multistep transcription
-initiation and escape.} How might we remedy the deficits of model 2? It is
+\subsubsection{Noise in the Three-State Promoter, Multistep Transcription
+Initiation and Escape.} How might we remedy the deficits of model 2? It is
 known~\cite{DeHaseth1998} that once RNAP initially binds the promoter region, a
 multitude of distinct steps occur sequentially before RNAP finally escapes into
-the elongation phase. Perhaps adding some of this mechanistic detail might
-rescue model 2? The next simplest refinement of model 2 could consider open
+the elongation phase. Perhaps adding some of this mechanistic detail as shown
+in Figure~\ref{fig:constit_cartoons} might
+rescue the previous model   The next simplest refinement of that model could consider open
 complex formation and promoter escape as separate steps rather than as a single
@@
 open complexes, despite the true biochemical details certainly being more
-complex. The authors of~\cite{Mitarai2015}, for instance, considered this model,
-although they added an additional repressor bound state and did not explicitly
-consider the limit with no repressor that we analyze here. Again, our goal here
+complex.  For example, earlier work extended this model by adding an additional
+repressor bound state and did not explicitly consider the limit with no repressor
+that we analyze here~\cite{Mitarai2015}.  Again, our goal here
 is not a complete accounting of all the relevant biochemical detail; this is an
-exploratory search for the important features our model needs to include.
+exploratory search for the important features that a model needs to include
+to square with the known experimental reality of constitutive expression.
 
-Unfortunately, as the authors of~\cite{Mitarai2015} hint, this model too has
+Unfortunately, as hinted at in earlier work~\cite{Mitarai2015}, this model too has
 Fano factor $\nu<1$. This can be seen by starting with their Eq.~A1 for the Fano
@@
 
+\marginpar{\it RP: very fun next paragraph - I like the argument, need to think more}
 In fact, we suspect \textit{any} model in which transcription proceeds through a
@@
 Regardless of the merits of this model in describing the noise properties of
-constitutive transcription initation, it does not capture the dominant source of
-noise we want to include in our phenomenological model, so we must discard these
-details and search elsewhere.
+constitutive transcription initiation, it ultimately fails the simplest quantitative
+feature of the data, namely that the Fano factor $> 1$ and hence  we must discard this
+mechanistic picture
+ and search elsewhere.
 
-\subsubsection{Model 4 - Two-state promoter, ``active'' and ``inactive''
-states.} Inspired by~\cite{Razo-Mejia2020}, we revisit an analog of model 2, but
+\subsubsection{Noise in a Two-State Promoter with ``Active'' and ``Inactive''
+States}
+ Inspired by~\cite{Razo-Mejia2020}, we revisit an analog of model 2, but as with
+ the models considered in the section on mean expression,
 the interpretation of the two states is changed. Rather than explicitly viewing
@@
 One interpretation~\cite{Chong2014, Sevier2016} for the active and inactive
-states is that it represents the promoter's supercoiling state: transitions to
+states is that they represent the promoter's supercoiling state: transitions to
 the inactive state are caused by accumulation of positive supercoiling, which
@@
 is driven by topoisomerase kinetics, not by RNAP kinetics. From in vitro
-measurments, the former are suggested to be of order minutes~\cite{Chong2014}.
+measurements, the former are suggested to be of order minutes~\cite{Chong2014}.
 Contrast this with model 2, where the state transitions are assumed to be
@@
 diffusion-limited association rate $k_{on} \sim 10^2~\text{s}^{-1}$ to a target
-promoter. Combined with known $K_D$'s of order $\mu$M, this gives an RNAP
-dissociation rate $k_{off}$ of order $10^2$. As we will show below, however,
+promoter. Combined with known $K_d$'s of order $\mu$M, this gives an RNAP
+dissociation rate $k_{off}$ of order $10^2$~s$^{-1}$. As we will show below, however,
 there are some lingering puzzles with interpreting this supercoiling
-interpretation, so we leave it as a speculation and refrain from assigning
+hypothesis, so we leave it as a speculation and refrain from assigning
 definite physical meaning to the two states in this model.
@@
 bunching would lead to super-Poissonian noise in the mRNA distribution over a
-population of cells. An honest calculation of the Fano factor produces
+population of cells.  Indeed, as shown in the appendix, a calculation of the Fano factor produces
 \begin{align}
@@
 be sidestepped by noting that the model is mathematically equivalent to the
-simple repression repression model from~\cite{Jones2014}, with states and rates
+simple repression model from~\cite{Jones2014}, with states and rates
 relabeled and reinterpreted.
@@
 contrast, here the 3 parameter model is used to fit the full mRNA distribution
-as measured in mRNA FISH experiments. In essense, all variability in the mRNA
+as measured in mRNA FISH experiments. In essence, all variability in the mRNA
 distribution is regarded as ``intrinsic,'' arising either from stochastic
@@
 
-\subsubsection{Model 5 - One-state promoter with explicit bursts.}
-This model is inspired by the failure mode of model 4. The key observation above
+\subsubsection{Noise Model for One-State Promoter with Explicit Bursts}
+The final model we consider is 
+ inspired by the failure mode of model 4. The key observation above
 was that, as found in~\cite{Razo-Mejia2020}, only two parameters, $k^+$ and the
@@
 
+\marginpar{\it RP: best description I have ever seen of this.  Good job}
 To develop some intuition, consider model 4 in the limit $k^+ \ll k^- \lesssim

### 2020-05-22 code/analysis/post_inf_pooled.py
Refactored PPCs for pooled repressed inference

@@ -66,7 +66,7 @@ model = srep.models.pooledInferenceModel(
 n_dim = len(var_labels)
 n_walkers = 35
 n_burn = 300
-n_steps = 200
+n_steps = 30
 
 # init walkers like prior but w/ narrower spread
 p0 = multinormal.rvs(
@@ -76,7 +76,7 @@ p0 = multinormal.rvs(
 
 #%%
 # run the sampler
-with Pool(processes=37) as pool:
+with Pool(processes=6) as pool:
 # instantiate sampler
     sampler = emcee.EnsembleSampler(
         n_walkers,
@@ -87,11 +87,26 @@ with Pool(processes=37) as pool:
     )
     pos, prob, state = sampler.run_mcmc(p0, n_burn, store=False, progress=True)
     _ = sampler.run_mcmc(pos, n_steps, progress=True, thin_by=40);
+
+# with posterior in hand, generate all the posterior predictive samples
+ppc_uv5 = srep.models.post_pred_bursty_rep(
+    sampler, n_pred=sum(data_uv5[1]),
+    kon_idx='nbinom', koff_idx='nbinom'
+    )
+ppc_rep = []
+for expt in expts:
+    # first look up indices
+    i = model.expts.index(expt)
+    kon_idx, koff_idx = model.expt_idx_to_rate_idx(i)
+    ppc_rep.append(
+        srep.models.post_pred_bursty_rep(
+            sampler, n_pred=sum(data_rep[i][1]),
+            kon_idx=kon_idx, koff_idx=koff_idx))
 del sampler.pool; # otherwise unpickling fails, even though pickling is fine
 
 #%%
-outfile = open(f"{repo_rootdir}/data/mcmc_samples/many_pooled_test.pkl", 'wb')
-dill.dump((model, sampler), outfile)
+outfile = open(f"{repo_rootdir}/data/mcmc_samples/repression_pooled_expts.pkl", 'wb')
+dill.dump((model, sampler, ppc_uv5, ppc_rep), outfile)
 outfile.close()
 
 print(f"Autocorr time: {sampler.get_autocorr_time()}")

### 2020-05-22 code/figures/figSIxx_ppc_arb_pooled.py

@@ -12,16 +12,10 @@ import emcee
 import arviz as az
 
 import matplotlib.pyplot as plt
-import bokeh
-import bokeh.io
 import bebi103.viz
 
-# from srep.data_loader import load_FISH_by_promoter
-# from srep.viz import ecdf
-# from srep.viz import plotting_style
 import srep
 
-bokeh.io.output_notebook()
 srep.viz.plotting_style()
 pboc_colors = srep.viz.color_selector('pboc')
 
@@ -33,9 +27,9 @@ repo_rootdir = repo.working_tree_dir
 
 # load in the pickled samples
 pklfile = open(
-    f"{repo_rootdir}/data/mcmc_samples/many_pooled_test.pkl", 'rb'
+    f"{repo_rootdir}/data/mcmc_samples/repression_pooled_expts.pkl", 'rb'
     )
-model, sampler = dill.load(pklfile)
+model, sampler, ppc_uv5, ppc_rep = dill.load(pklfile)
 pklfile.close()
 inf_dat = az.convert_to_inference_data(
     sampler, var_names=model.var_labels
@@ -46,73 +40,8 @@ data_uv5, data_rep = srep.utils.condense_data(model.expts)
 n_dim = np.shape(model.var_labels)
 
 #%%
-emcee_output = az.convert_to_inference_data(
-    sampler, var_names=model.var_labels
-    )
-bokeh.io.show(
-    bebi103.viz.corner(
-        emcee_output, pars=model.var_labels[-5:], plot_ecdf=True
-        )
-    )
-
-#%%
 
 fig, axes = plt.subplots(2,2, figsize=(9,9))
-
-# ##########################################################################
-# PLOTTING ALL PAIRWISE POSTERIORS FROM POOLED DATA MODEL
-# ##########################################################################
-lstyle = {'O2':'-', 'O1':'-.', 'Oid':'--'}
-lcolor = {'0p5ngmL':'green', '1ngmL':'purple', '2ngmL':'blue', '10ngmL':'red'}
-for i, expt in enumerate(model.expts):
-    # parse op/aTc values
-    op, aTc = re.split('_', expt)
-    # look up rates
-    kRon_label, kRoff_label = model.expt_rates[expt]
-    kR_on_samples = inf_dat.posterior[kRon_label].values.flatten()
-    kR_off_samples = inf_dat.posterior[kRoff_label].values.flatten()
-    x_contour, y_contour = bebi103.viz.contour_lines_from_samples(
[…]

### 2020-05-22 code/analysis/post_inf_pooled.py
Hacked p-tile predictive ECDFs + fixed my bug in fig03

@@ -66,7 +66,7 @@ model = srep.models.pooledInferenceModel(
 n_dim = len(var_labels)
 n_walkers = 35
 n_burn = 300
-n_steps = 30
+n_steps = 200
 
 # init walkers like prior but w/ narrower spread
 p0 = multinormal.rvs(
@@ -76,7 +76,7 @@ p0 = multinormal.rvs(
 
 #%%
 # run the sampler
-with Pool(processes=6) as pool:
+with Pool(processes=36) as pool:
 # instantiate sampler
     sampler = emcee.EnsembleSampler(
         n_walkers,

### 2020-05-22 code/figures/main/fig03.py

@@ -165,10 +165,10 @@ ptiles = (95,)#, 75, 50, 25)
 alpha = df_UV5["mRNA_cell"].sum()
 beta = len(df_UV5)
 # approx gamma posterior as normal b/c alpha is so huge
-poiss_post_draws = st.norm.rvs(alpha/beta, np.sqrt(alpha)/beta, size=400)
-poiss_ppc_draws = np.empty((len(df_UV5), len(poiss_post_draws)))
+poiss_post_draws = st.norm.rvs(alpha/beta, np.sqrt(alpha)/beta, size=n_samples)
+poiss_ppc_draws = np.empty((len(poiss_post_draws), len(df_UV5)))
 for i, draw in enumerate(poiss_post_draws):
-    poiss_ppc_draws[:,i] = st.poisson.rvs(draw, size=len(df_UV5))
+    poiss_ppc_draws[i,:] = st.poisson.rvs(draw, size=len(df_UV5))
 # now plot PPC from Poisson samples
 srep.viz.predictive_ecdf(
     poiss_ppc_draws,

### 2020-05-22 code/figures/main/pooled_repression_ppc.py

@@ -48,23 +48,23 @@ fig, axes = plt.subplots(2,2, figsize=(9,9))
 # ##########################################################################
 
 # organize all the options upfront
-plotting_draws = 75
 all_expts = (
     ("_"), # a blank b/c 1st index is skipped
     ("Oid_2ngmL", "Oid_1ngmL"),
     ("O1_1ngmL", "O1_2ngmL", "O1_10ngmL"),
     ("O2_0p5ngmL", "O2_1ngmL", "O2_2ngmL", "O2_10ngmL")
     )
-ppc_colors = ('blue', 'purple', 'red', 'yellow')
+ppc_colors = ('blue', 'purple', 'betancourt', 'orange')
 data_colors = ('black', 'black', 'black', 'black')
 # ppc_labels = ('0.5 ng/mL aTc', '1 ng/mL aTc', '2 ng/mL aTc', '10 ng/mL aTc')
 uv5_colors = {'ppc':'green', 'data':'black'}
 # convert labels to hex colors
-ppc_colors = [pboc_colors[label] for label in ppc_colors]
-data_colors = [pboc_colors[label] for label in data_colors]
+# ppc_colors = [pboc_colors[label] for label in ppc_colors]
+# data_colors = [pboc_colors[label] for label in data_colors]
 ppc_alpha = 0.3
 ppc_lw = 0.2
 data_lw = 0.6
+ptiles = (95,)
 # then (nearly?) all the rest below will not need changing
 
 for k, ax in enumerate(fig.axes):
@@ -72,51 +72,33 @@ for k, ax in enumerate(fig.axes):
         continue # upper left panel is handled separately
     ppc_labels = all_expts[k]
 
-    # now we start plotting. UV5 1st
-    for i in range(plotting_draws):
-        ax.plot(
-            *srep.viz.ecdf(ppc_uv5[i]),
-            color=pboc_colors[uv5_colors['ppc']],
-            alpha=ppc_alpha,
-            lw=ppc_lw
-            )
     # plot UV5 data later below to fix legend order
 
     # now loop over repressed datasets & plot ppc + observed data
     for j, expt in enumerate(ppc_labels):
         expt_idx = model.expts.index(expt)
-        for i in range(plotting_draws):
-            ax.plot(
-                *srep.viz.ecdf(ppc_rep[expt_idx][i]),
-                color=ppc_colors[j],
-                alpha=ppc_alpha,
-                lw=ppc_lw
-                )
-    # separate loop to plot data so it won't get buried by any ppc's
[…]

### 2020-05-22 srep/utils.py

@@ -4,6 +4,7 @@ from git import Repo #for directory convenience
 
 import numpy as np
 import pandas as pd
+from numba import njit
 
 repo = Repo("./", search_parent_directories=True)
 # repo_rootdir holds the absolute path to the top-level of our repo
@@ -73,3 +74,29 @@ def condense_data(expts):
             np.unique(df['mRNA_cell'], return_counts=True)
             )
     return data_uv5, rep_data
+
+@njit
+def uncondense_valuescounts(vc_tuple):
+    # assume vc_tuple is (values, counts) which are arrays of
+    # mRNA values and occurences in a dataset, respectively
+    values, counts = vc_tuple
+    list_of_lists = [values[i] * np.ones(counts[i]) for i,_ in enumerate(counts)]
+    # return np.array(list(itertools.chain(*list_of_lists)))
+    return np.array([cell for sublist in list_of_lists for cell in sublist])
+
+    # return np.array([ [ x * y for x in range(n) ] for y in range(n) ])
+
+@njit
+def uncondense_ppc(ppc_samples):
+    """
+    This function exists to solve the following problem:
+    My bursty_rep_rng generates samples, then returns
+    np.unique(samples, return_counts=True).
+    That's convenient mostly, and keeps .pkl size down, but
+    srep.viz.predictive_ecdf, borrowed from JB, needs raw samples as input.
+    This funtion regenerates the raw samples from the condensed form.
+    """
+    output_samples = np.empty((len(ppc_samples), ppc_samples[0][1].sum()))
+    for i in range(len(ppc_samples)):
+        output_samples[i,:] = uncondense_valuescounts(ppc_samples[i])
+    return output_samples

### 2020-05-22 srep/viz.py

@@ -130,6 +130,9 @@ def predictive_ecdf(
     ----------
     samples : Numpy array or xarray, shape (n_samples, n) or xarray DataArray
         A Numpy array containing predictive samples.
+        n_samples is the number of posterior samples, and n is the number
+        of observed data points, i.e., n is the number of posterior
+        predictive samples for each of the posterior samples
     data : Numpy array, shape (n,) or xarray DataArray
         If not None, ECDF of measured data is overlaid with predictive
         ECDF.
@@ -265,15 +268,6 @@ def predictive_ecdf(
             x = df_ecdf["x"]
             y1 = df_ecdf[ptile]
             y2 = df_ecdf[ptiles_str[-i - 1]]
-        # fill_between(
-        #     x,
-        #     y1,
-        #     x,
-        #     y2,
-        #     p=p,
-        #     show_line=False,
-        #     patch_kwargs=dict(color=colors[color][i]),
-        # )
         ax.fill_between(x, y1, y2, color=colors[color][i],)
 
     # The median as a solid line

### 2020-05-22 doc/section_05_discussion.tex
RP update main paper

@@
 \section{Discussion and future work}
+
+The study of gene expression is one of the dominant themes of modern
+biology, made all the more urgent by the dizzying pace at which
+genomes are being sequenced.  But there is a troubling Achilles heel
+buried in all of that genomic data and that is our inability to find
+and interpret regulatory sequence.  In many cases, this is not
+possible even qualitatively, let alone the possibility of quantitative
+dissection of the regulatory parts of genomes in a predictive fashion.
+Other recent work has tackled the challenge of finding and
+annotating the regulatory part of genomes.  Once we have determined
+the regulatory part of the genome, we are then faced with the
+next class of questions which are sharpened by formulating
+them in mathematical terms, namely, what are the input-output
+properties of these regulatory circuits and what knobs control them?
+
+The present work has tackled that question in the context of
+the first regulatory architecture hypothesized in the molecular
+biology era, namely, the repressor-operator model of Jacob and
+Monod.  Regulation in that architecture is the result of a 
+competition between a repressor which inhibits transcription
+and RNAP polymerase which undertakes it.  Through
+the labors of generations of geneticists, molecular biologists
+and biochemists, an overwhelming amount of information
+and insight has been garnered into this simple regulatory
+motif, licensing it as what one might call the ``hydrogen atom''
+of regulatory biology.   It is from that perspective that the present
+paper explores the extent to which some of the different models
+that have been articulated to describe that motif allow
+us to understand both the average level of gene expression
+found in a population of cells as well as the noise in that
+expression found in that population as would be reported in
+a FISH experiment, for example.
+
+Our key insights can be summarized as follows.  First, as shown in Figure
+Figure~\ref{fig1:means_cartoons}, the mean expression in the
+simple repression architecture is captured by a master curve
+in which the action of repressor and the details of the polymerase
+interaction with the promoter appear separately and additively
+in an effective free energy.  Interestingly, as has been shown
+elsewhere in the context of the Monod-Wyman-Changeux model,
+these kinds of coarse-graining results are an exact mathematical
+result and do not constitute hopeful approximations or biological
+naivete.  To further dissect the relative merits of the different models,
+we must appeal to higher moments of the gene expression probability
+distribution.  To that end, our second set of insights focus on gene
+expression noise, where it is seen that a treatment of
+the constitutive promoter already reveals that some models have Fano
+factors that are less than one, at odds with any and all experimental
+data that we are aware of.   This refinement allows us to determine
+parameters for the constitutive promoter that can then be used
+to attack gene expression noise in the context of
+the simple repression architecture.
+
 \mmnote{Do all my approximations work simply because galK happens to have a low

### 2020-05-23 srep/viz.py
update functions necessary for Fig 04

@@ -9,6 +9,10 @@ import xarray
 import os
 import matplotlib.pyplot as plt
 import matplotlib
+from mpl_toolkits.axes_grid1 import make_axes_locatable
+from matplotlib.path import Path
+from matplotlib.patches import BoxStyle
+from matplotlib.offsetbox import AnchoredText
 
 import bebi103
 
@@ -39,8 +43,11 @@ def plotting_style(grid=True):
            'xtick.color': 'k',
            'ytick.color': 'k'}
     plt.rc('text.latex', preamble=r'\usepackage{sfmath}')
-    plt.rc('mathtext', fontset='stixsans', sf='sans')
+    plt.rc('mathtext', fontset='stixsans', sf='sansserif')
+    plt.rc('legend', title_fontsize='8', frameon=True,
+            facecolor='#E3DCD0', framealpha=1)
     sns.set_style('darkgrid', rc=rc)
+    sns.set_palette("colorblind", color_codes=True)
 
 def color_selector(style):
     """
@@ -372,3 +379,47 @@ def traceplot(sampler, labels=None):
         ax.yaxis.set_label_coords(-0.1, 0.5)
     axes[-1].set_xlabel("step number");
     return fig, axes
+
+def titlebox(
+    ax, text, color='black', bgcolor=None, size=8, boxsize=0.1, pad=0.05, loc=10, **kwargs
+):
+    """Sets a colored box about the title with the width of the plot.
+   To use, simply call `titlebox(ax, 'plot title', bgcolor='blue')
+   """
+    boxsize=str(boxsize * 100)  + '%'
+    divider = make_axes_locatable(ax)
+    cax = divider.append_axes("top", size=boxsize, pad=pad)
+    cax.get_xaxis().set_visible(False)
+    cax.get_yaxis().set_visible(False)
+    cax.spines["top"].set_visible(False)
+    cax.spines["right"].set_visible(False)
+    cax.spines["bottom"].set_visible(False)
+    cax.spines["left"].set_visible(False)
+    plt.setp(cax.spines.values(), color=color)
+    if bgcolor != None:
+        cax.set_facecolor(bgcolor)
+    else:
+        cax.set_facecolor("white")
+    at = AnchoredText(text, loc=loc, frameon=False, prop=dict(size=size, color=color))
+    cax.add_artist(at)
+
+def bebi103_colors():
+    """
[…]

### 2020-05-23 code/figures/main/fig04.py
temptative final version for Fig 04

@@ -1,0 +1,336 @@
+# %%
+import re
+import dill
+from git import Repo #for directory convenience
+
+import numpy as np
+import pandas as pd
+
+import emcee
+import arviz as az
+
+import matplotlib.pyplot as plt
+import bebi103.viz
+
+import srep
+
+srep.viz.plotting_style()
+pboc_colors = srep.viz.color_selector('pboc')
+
+# %%
+# Initialize figure
+fig = plt.figure(constrained_layout=False, figsize=(6, 5))
+# Set gridspec for top row
+gs_top = fig.add_gridspec(
+    nrows=7, 
+    ncols=2, 
+    left=0.05, 
+    right=0.98,
+    top=1, 
+    bottom=0.55,
+    wspace=.55,
+    hspace=0.1,
+)
+
+ax_0 = fig.add_subplot(gs_top[:, 0])
+ax_1 = fig.add_subplot(gs_top[:3, 1])
+ax_2 = fig.add_subplot(gs_top[4:, 1])
+
+# Set gridspec for bottom row
+gs_bottom = fig.add_gridspec(
+    nrows=1, 
+    ncols=3, 
+    left=0.03, 
+    right=0.98,
+    top=0.4, 
+    bottom=0,
+    wspace=0.05,
+    hspace=0.2
+)
+
+ax_3 = fig.add_subplot(gs_bottom[0])
+ax_4 = fig.add_subplot(gs_bottom[1])
+ax_5 = fig.add_subplot(gs_bottom[2])
+
+# Set shared axes
+ax_4.get_shared_y_axes().join(ax_3, ax_4)
+ax_5.get_shared_y_axes().join(ax_3, ax_5)
+
+# Turn off axis
+ax_4.yaxis.set_ticklabels([])
+ax_5.yaxis.set_ticklabels([])
+
+# Pack all ax together
+ax = [ax_0, ax_1, ax_2, ax_3, ax_4, ax_5]
+# Modify tick font size
+for a in ax:
+    a.tick_params(axis="both", which="major", labelsize=8)
+
+# Set global colors for aTc concentrations
+aTc_colors = ('blue', 'betancourt', 'green', 'orange')
+aTc_col_dict = dict(zip(["0p5ngmL","1ngmL", "2ngmL", "10ngmL"], aTc_colors))
+# TOP ROW
+repo = Repo("./", search_parent_directories=True)
+# repo_rootdir holds the absolute path to the top-level of our repo
[…]

### 2020-05-23 code/figures/main/fig04.py
update marker shapes

@@ -10,6 +10,7 @@ import emcee
 import arviz as az
 
 import matplotlib.pyplot as plt
+import seaborn as sns
 import bebi103.viz
 
 import srep
@@ -69,6 +70,8 @@ for a in ax:
 # Set global colors for aTc concentrations
 aTc_colors = ('blue', 'betancourt', 'green', 'orange')
 aTc_col_dict = dict(zip(["0p5ngmL","1ngmL", "2ngmL", "10ngmL"], aTc_colors))
+
+
 # TOP ROW
 repo = Repo("./", search_parent_directories=True)
 # repo_rootdir holds the absolute path to the top-level of our repo
@@ -156,6 +159,9 @@ ax[0].set_xlabel(r'$log_{10}(k_R^-/\gamma)$', fontsize=8)
 # ##########################################################################
 # COMPARING k_off RATIOS TO BINDING E DIFFERENCES
 # ##########################################################################
+# Define colors for markesr
+col_markers = sns.color_palette("colorblind", n_colors=4)
+
 hg_rp_ep = (-17., -15.3, -13.9) # Oid, O1, O2
 hg_rp_err = 0.2
 mwc_soc_ep = (-17.7, -15.2, -13.6)
@@ -176,20 +182,23 @@ o1_oid_err = np.mean(np.diff(o1_oid))
 o2_o1_err  = np.mean(np.diff(o2_o1))
 
 ax[1].errorbar(o2_o1[1], hg_diff_ep[1], xerr=o2_o1_err, yerr=hg_rp_err,
-                fmt='^b', capsize=0, label='O2-O1, HG & RP 2011')
+                fmt='^', capsize=0, label='O2-O1, HG & RP 2011', 
+                color=col_markers[0])
 ax[1].errorbar(o1_oid[1], hg_diff_ep[0], xerr=o1_oid_err, yerr=hg_rp_err,
-                fmt='^g', capsize=0, label='O1-Oid, HG & RP 2011')
+                fmt='^', capsize=0, label='O1-Oid, HG & RP 2011',
+                color=col_markers[1])
 ax[1].errorbar(o2_o1[1], soc_diff_ep[1], xerr=o2_o1_err, yerr=mwc_soc_err,
-                fmt='ob', capsize=0, label='O2-O1, MRM et al 2018')
[…]

### 2020-05-23 code/figures/main/fig04_v2.py
add script for version 2 of Fig04 with numbered symbols

@@ -1,0 +1,422 @@
+# %%
+import re
+import dill
+from git import Repo #for directory convenience
+
+import numpy as np
+import pandas as pd
+
+import emcee
+import arviz as az
+
+import matplotlib.pyplot as plt
+import seaborn as sns
+import bebi103.viz
+
+import srep
+
+srep.viz.plotting_style()
+pboc_colors = srep.viz.color_selector('pboc')
+
+# %%
+# Initialize figure
+fig = plt.figure(constrained_layout=False, figsize=(6, 5))
+# Set gridspec for top row
+gs_top = fig.add_gridspec(
+    nrows=7, 
+    ncols=2, 
+    left=0.05, 
+    right=0.98,
+    top=1, 
+    bottom=0.55,
+    wspace=.55,
+    hspace=0.1,
+)
+
+ax_0 = fig.add_subplot(gs_top[:, 0])
+ax_1 = fig.add_subplot(gs_top[:3, 1])
+ax_2 = fig.add_subplot(gs_top[4:, 1])
+
+# Set gridspec for bottom row
+gs_bottom = fig.add_gridspec(
+    nrows=1, 
+    ncols=3, 
+    left=0.03, 
+    right=0.98,
+    top=0.4, 
+    bottom=0,
+    wspace=0.05,
+    hspace=0.2
+)
+
+ax_3 = fig.add_subplot(gs_bottom[0])
+ax_4 = fig.add_subplot(gs_bottom[1])
+ax_5 = fig.add_subplot(gs_bottom[2])
+
+# Set shared axes
+ax_4.get_shared_y_axes().join(ax_3, ax_4)
+ax_5.get_shared_y_axes().join(ax_3, ax_5)
+
+# Turn off axis
+ax_4.yaxis.set_ticklabels([])
+ax_5.yaxis.set_ticklabels([])
+
+# Pack all ax together
+ax = [ax_0, ax_1, ax_2, ax_3, ax_4, ax_5]
+# Modify tick font size
+for a in ax:
+    a.tick_params(axis="both", which="major", labelsize=8)
+
+# Set global colors for aTc concentrations
+aTc_colors = ('blue', 'betancourt', 'green', 'orange')
+aTc_col_dict = dict(zip(["0p5ngmL","1ngmL", "2ngmL", "10ngmL"], aTc_colors))
+
+
+# TOP ROW
+repo = Repo("./", search_parent_directories=True)
[…]

### 2020-05-23 doc/appendix_01_bayesian_inference.tex
Ironed out merge conflict
Fixed my error in Fig 3D scaling, add MRM's new Fig4
Found an arithmetic error in log(k_i)~deps_P argument,
realized intercept has meaning after all. Updated plot
and text to reflect. Also added MRM's beautiful new Fig 4

@@
 % !TEX root = ./busty_transcription.tex
-\section{Bayesian inference}\label{sec:bayesian}
+\section{Bayesian inference}
+\label{sec:bayesian}
 \mmnote{
@@
 
-\begin{figure}%[h!]
-\centering
-\includegraphics[width=\textwidth]{../figures/main/fig03.pdf}
-\caption{\textbf{Constitutive promoter posterior inference and model
-        comparison.}
-        (A) The joint posterior density of model 5, the
-        bursty promoter with negative binomially-distributed
-        steady state, is plotted with MCMC samples.
-        1D marginal densities are plotted as flanking histograms.
-        The model was fit on \textit{lacUV5} data from~\cite{Jones2014}.
-        (B) The ECDF of the observed population distribution of
-        mRNA transcripts under the control of a constitutive
-        lacUV5 promoter is shown in orange.
-        The median posterior predictive ECDFs for models (1),
-        Poisson, and (5), negative binomial, are plotted in
-        green and blue, respectively.
-        Lighter green and blue regions enclose 95\% of all
-        posterior predictive samples from their respective models.
-        Model (1) is in obvious contradiction with the data
-        while model (5) is not.
-        FISH data is again from~\cite{Jones2014}.
-    (C) Joint posterior distributions for burst rate $k_i$ and mean burst size
-    $b$ for 18 promoters from~\cite{Jones2014}. Each contour indicates the 95\%
-    highest posterior density region for a particular promoter. Note that the
-    vertical axis is shared with (D). Panel (D) plots the burst rate $k_i$ vs.\
-    the binding energy for each promoter from energy matrices
-    from~\cite{Brewster2012}.
-    The dotted line shows the predicted slope according to
-    $\ln k_i/\gamma \sim \text{constant} - \beta\Delta\epsilon_P$,
-    described in text.}
-\label{fig:constit_post_full}
-\end{figure}
-
 \fig{fig:constit_post_full}(C) shows the results, plotting the
@@
 \frac{k_i}{\gamma} = \exp(-\beta\Delta\epsilon_P)
-\label{eq:bursty_equil_corresp1}
+\label{eq:bursty_equil_corresp1_appdx}
 \end{equation}
@@
 b = \frac{r}{\gamma}\frac{P}{N_{NS}}.
-\label{eq:bursty_equil_corresp2}
+\label{eq:bursty_equil_corresp2_appdx}
 \end{equation}
@@
 While the associations represented by
-\eq{eq:bursty_equil_corresp1} and \eq{eq:bursty_equil_corresp2}
+\eq{eq:bursty_equil_corresp1_appdx} and \eq{eq:bursty_equil_corresp2_appdx}
 appear to be born out by the scaling $\ln k_i \sim
@@
 parameters encoded by
-\eqrange{eq:bursty_equil_corresp1}{eq:bursty_equil_corresp2}
+\eqrange{eq:bursty_equil_corresp1_appdx}{eq:bursty_equil_corresp2_appdx}
 intuitive. If some deeper insight is waiting to be gleaned from it,
@@
 
-\subsection{Transcription factor kinetics can be inferred from FISH
+\subsection{Transcription factor kinetics can be inferred from FISH}

### 2020-05-23 doc/section_04_bayesian_inference.tex
Fixed my error in Fig 3D scaling, add MRM's new Fig4
Found an arithmetic error in log(k_i)~deps_P argument,
realized intercept has meaning after all. Updated plot
and text to reflect. Also added MRM's beautiful new Fig 4

@@
 predictive samples from their respective models. Model (1) is in obvious
-contradiction with the data while model (5) is not. single-cell mRNA count data
+contradiction with the data while model (5) is not. Single-cell mRNA count data
 is again from~\cite{Jones2014}. (C) Joint posterior distributions for burst rate
@@
 energy for each promoter as predicted in~\cite{Brewster2012}. The dotted line
-shows the predicted slope according to $\ln k_i/\gamma \sim \text{constant} -
-\beta\Delta\varepsilon_P$, described in text. Each individual promoter is
+shows the predicted slope according to Eq.~\ref{eq:bursty_equil_corresp1},
+described in text. Each individual promoter is
 labeled with a unique number in both (C) and (D) for cross comparison.}
@@
 
-Given this result, how are the two coarse-grainings related? The only
+Given this result, how are the two coarse-grainings related?
+A quick estimate can shed some light.
+Consider for instance the \textit{lacUV5} promoter, which we see from
+Figure~\ref{fig:constit_post_full}(A) has $k_i/\gamma \sim b \sim \text{few}$,
+from Figure~\ref{fig:constit_post_full}(B) has $\langle m \rangle \sim 20$,
+and from~\cite{Brewster2012} has $\beta\Delta\epsilon_P \sim - 6.5$.
+Further we generally assume $P/N_{NS} \sim 10^{-3}$ since
+$N_{NS}\approx4.6\times10^6$ and $P\sim10^3$.
+After some guess-and-check with these values, one finds the only
 association that makes dimensional sense and produces the correct
@@
 \begin{equation}
-\frac{k_i}{\gamma} = \exp(-\beta\Delta\varepsilon_P)
+\frac{k_i}{\gamma} = \frac{P}{N_{NS}} \exp(-\beta\Delta\varepsilon_P)
 \label{eq:bursty_equil_corresp1}
@@
 \begin{equation}
-b = \frac{r}{\gamma}\frac{P}{N_{NS}}.
+b = \frac{r}{\gamma}.
 \label{eq:bursty_equil_corresp2}
@@
 $\ln k_i$ and $-\beta\Delta\varepsilon_P$ is approximately true for all 18
-constitutive promoters considered. The intercept is an uninteresting function of
-the unknown model parameters, so we only consider the slope of the scaling in
-\fig{fig:constit_post_full}(D).
-\mmnote{Wait, is that true or can we do something with the intercept, related to the burst size in some useful way? Needs more careful thought.}
+constitutive promoters considered. The plotted line is simply
+Eq.~\ref{eq:bursty_equil_corresp1} and assumes $P\approx 5000$.
 
@@
 Eq.~\ref{eq:bursty_equil_corresp1} and Eq.~\ref{eq:bursty_equil_corresp2}
-appear to be born out by the scaling $\ln k_i \sim -\beta\Delta\varepsilon_P$,
-we do not find the association of parameters encoded by
-\eqrange{eq:bursty_equil_corresp1}{eq:bursty_equil_corresp2}
-intuitive. If some deeper insight is waiting to be gleaned from it, we leave it
-as an open problem.
+appear to be born out by the data in Figure~\ref{fig:constit_post_full},
+we do not find the association of parameters they imply to be intuitive.
+We are also cautious to ascribe too much physical reality to the parameters.
+Indeed, part of our point in comparing the various constitutive
+promoter models is to demonstrate that these models each provide
+an internally self-consistent framework that adequately describes
+the data, but attempting to translate between models reveals
+the dubious physical interpretation of their parameters.
 
@@
 \begin{itemize}
-\item All 18 promoters. Hey the model still works on things other than UV5,
-cool. Interesting and somewhat counterintuitive scaling b/w burst rate and
-binding energy. Puzzle in comparing w/ Chong2014 supercoiling model. Are we
+\item There is a puzzle in comparing w/ Chong2014 supercoiling
+model: if supercoiling is the thing, why are my burst sizes all
+the same but burst rates vary? And why is the duty cycle of my
+promoter so low, ie., bursts so short? COmpare their fig 7E;
+their $\beta/\alpha$ is my $k^+/k^-$. They have one or two genes
+with very small $\beta/\alpha$, does the \textit{galK} locus just
+happen to be that, or is there a deeper disagreement? Hard to say
+w/o more data. Are we
 missing important things?? Unclear, we leave as open question. Good enough for
@@
 
-\mmnote{Discuss puzzling comparison w/ Chong2014: if supercoiling is the thing, why are my burst sizes all the same but burst rates vary?
-And why is the duty cycle of my promoter so low, ie., bursts so short?
-COmpare their fig 7E; their $\beta/\alpha$ is my $k^+/k^-$.
-They have one or two genes with very small $\beta/\alpha$, does the \textit{galK} locus just happen to be that, or is there a deeper disagreement?
-Hard to say w/o more data.}
-
 \subsection{Transcription factor kinetics can be inferred from FISH measurements}
 \label{sec:rep_kinetics_inference}
+
+\begin{figure}%[h!]
+\centering
+\includegraphics[width=\textwidth]{../figures/main/fig04.pdf}
+\caption{\textbf{Simple repression parameter inference and comparison.}
+(A) Contours which enclose 50\% and 95\% of the posterior
+probability mass are shown for each of several 2D slices of
+the 9D posterior distribution. The model assumes one
+unbinding rate for each operator (O1, O2, O3) and one binding rate for each aTc
+concentration (corresponding to an unknown mean repressor copy number).
+(B, upper panel)
+Ratios of our inferred unbinding rates are compared with
+operator binding energy differences measured by
+Garcia and Phillips~\cite{Garcia2011a} (triangles) and
+Razo-Mejia et.\ al.~\cite{Razo-Mejia2018} (circles).
+Blue glyphs compare O2-O1, while orange compare O1-Oid.
+(B, lower panel)
+Unbinding rates for O1 (cyan) and Oid (red) inferred in this work are compared with
+single-molecule measurements of the same from Hammar et.\ at.\ 2014.
+The comparison requires an assumption about mRNA lifetime
+$\gamma^{-1}$, which was not measured, and we plot representative
+choices assuming $\gamma^{-1}=3$~min (triangles) or $\gamma^{-1}=5$~min (squares).
+(C) Posterior predictive checks are shown for each of the
+datasets used in the inference of the model in (A). Observed FISH
+data from~\cite{Jones2014} is plotted as black lines. The median
+posterior predictive ECDF for each condition is plotted as a dark
+colored line. Lighter colored bands enclose 95\% of all posterior
+predictive samples for a given operator/repressor copy number pair.
+Samples are sorted by operator and plotted separately for visual clarity.
+The unrepressed promoter, UV5, is shown with each as a reference point.
+}
+\label{fig4:repressed_post_full}
+\end{figure}
+
 \mmnote{Outline:
@@
 \item Hey look it mostly works. With full model, 9D model is well identified,
-though most of the individual experiments unsurprisingly are not: only the ratio
-of repressor rates is identifiable. PPC is not perfect, it kinda misses some
+though most of the individual experiments unsurprisingly are not:
+only 3 of the 9 datasets were identifiable if fit in isolation
+and for the rest only the ratio of repressor rates is identifiable.
+Only by making assumptions that certain rates
+are equal across datasets was identifiability achieved.
+PPC is not perfect, it kinda misses some
 rep/op pairs but overall it's surprisingly predictive considering how strong
@@
 }
-
-\fig{fig3:kR_inferences} presents inferred
-transcription factor rates, compares them to
-single-molecule measurements, and compares their ratios to gold-standard binding
-energies.
-\begin{figure}%[h!]
-\centering
-\includegraphics[width=0.9\textwidth]{../figures/fig3/fig3.ai}
-\caption{\textbf{Simple repression model comparison.}
-Panel (A) depicts the cartoon of simple repression for which we infer parameters.
-\mmnote{Not sure panel (A) is necessary, we've already shown it in at least 2
-other figs. We could instead show some alternative slices of 9D posterior as a
-companion to (B), or something totally different\dots??}
-(B) Contours which enclose 50\% and 95\% of the posterior
-probability mass are shown for of several 2D slices of
-the 9D posterior distribution, from fitting one
-unbinding rate for each operator (O1, O2, O3) and one binding rate for each aTc
-concentration (corresponding to an unknown mean repressor copy number). (C)
-Ratios of our inferred unbinding rates are compared with operator binding energy
-differences measured in Garcia \& Phillips 2011 (triangles) and Razo-Mejia et.\
-al.\ 2018 (squares). (D) Unbinding rates inferred in this work are compared with
-single-molecule measurements of the same from Hammar et.\ at.\ 2014
-\mmnote{Johan Elf's group}. Values are of similar order of magnitude, but a
-precise quantitative comparison unfortunately depends sensitively on the assumed
-mRNA lifetime, which was not measured.}
-\label{fig3:kR_inferences}
-\end{figure}
-
-\begin{figure}%[h!]
-\centering
-\includegraphics[width=0.9\textwidth]{../figures/figSIxx/ppc_many_pooled.pdf}
-\caption{\textbf{Posterior predictive checks.}
-(A) The posterior distribution from~\fig{fig3:kR_inferences}B is
-shown again for context. (B), (C), and (D) Samples from the
-posterior predictive distribution are shown for each of the datasets
-used in the inference. Samples are sorted by operator and plotted separately
-simply for visual clarity. The unrepressed promoter, UV5, is shown with each as
-a reference point. \mmnote{Obviously the PPC is not perfect, but considering the
-relative simplicity of the model, I'm pleasantly surprised how well it worked
-and that the model was even identifiable at all. Interestingly, and perhaps not
-surprisingly, only 3 of the 9 datasets were identifiable if fit in isolation:
-otherwise, it was only by making assumptions that certain rates are equal across
-datasets was identifiability achieved.}}
-\label{fig4:ppc}
-\end{figure}

### 2020-05-23 code/figures/main/fig03.py

@@ -253,7 +253,7 @@ for i, promoter in enumerate(df_energies.Name):
         yerr=np.array((err_lower, err_upper)).reshape((2,1)),
         fmt="o",
         markersize=4.0,
-        label=promoter,
+        # label=promoter,
         c=cmap(col_norm(df_energies[df_energies.Name == promoter]["Energy (kT)"].values[0])),
     )
     # Append y position
@@ -283,14 +283,14 @@ cbar = fig.colorbar(im, ax=ax_d, pad=0.01)
 cbar.set_label(r"$\Delta\epsilon_r \; (k_BT)$")
 
 # add a guideline for the eye for the predicted log(k_i) ~ - binding E
-guide_x = np.linspace(-5.5,-2)
-guide_y = np.exp(-guide_x)/50
+guide_x = np.linspace(-7,-3.5)
+guide_y = np.exp(-guide_x)/9e2
 ax_d.plot(guide_x, guide_y, 'k--', label='predicted \n scaling')
 # Add text for scaling
 ax_d.text(
-    0.6, 
-    0.6, 
-    r"$\log(k_i) \sim - \Delta\epsilon_r$",
+    0.29, 
+    0.04, 
+    r"$\log k_i \sim - \Delta\epsilon_P$",
     transform=ax_d.transAxes,
     rotation=-45,
 )

### 2020-05-23 doc/appendix_03_regulated_promoters.tex
Text on posterior inference on repressed data

@@
 \frac{b^m}{m!}{_2F_1}(\alpha+m, \beta+m, k_R^++k_R^-+m; -b)
-\label{eq:p_m_bursty+rep}
+\label{eq:p_m_bursty+rep_appdx}
 \end{align}
@@
 \subsubsection{Moments}
-We can verify the correctness of~\eq{eq:p_m_bursty+rep}
+We can verify the correctness of~\eq{eq:p_m_bursty+rep_appdx}
 by comparison with Gillespie simulation.
@@
 We would like to carry out Bayesian parameter inference on FISH data
-from~\cite{Jones2014}, using~\eq{eq:p_m_bursty+rep} as our
+from~\cite{Jones2014}, using~\eq{eq:p_m_bursty+rep_appdx} as our
 likelihood. This requires accurate (and preferably fast)
@@
 to the tail of the distribution, which can easily reach 50, possibly 100.
-From~\eq{eq:p_m_bursty+rep}, this means evaluating ${_2F_1}$
+From~\eq{eq:p_m_bursty+rep_appdx}, this means evaluating ${_2F_1}$
 repeatedly for values of $a$, $b$, and $c$ spanning the full range
@@
 even if $\alpha$, $\beta$, and $\gamma$
-in~\eq{eq:p_m_bursty+rep} are small,
+in~\eq{eq:p_m_bursty+rep_appdx} are small,
 with the situation even worse if they are not small.
@@
 To quantify how slow, we found that
-evaluating the likelihood defined by~\eq{eq:p_m_bursty+rep} $\sim50$ times
+evaluating the likelihood defined by~\eq{eq:p_m_bursty+rep_appdx} $\sim50$ times
 (for a typical dataset of interest from~\cite{Jones2014},
@@
 In order to evaluate the probability distribution of our
-model,~\eq{eq:p_m_bursty+rep}, we need to evaluate hypergeometric functions
+model,~\eq{eq:p_m_bursty+rep_appdx}, we need to evaluate hypergeometric functions
 of the form ${_2F_1}(\alpha+m, \beta+m, \gamma+m; -b)$
@@
 With ${_2F_1}$ computed, the only remaining numerical danger in computing
-$p(m)$ in~\eq{eq:p_m_bursty+rep} is overflow of the gamma functions.
+$p(m)$ in~\eq{eq:p_m_bursty+rep_appdx} is overflow of the gamma functions.
 This is easily solved by taking the log of the entire expression

### 2020-05-23 doc/section_04_bayesian_inference.tex
Text on posterior inference on repressed data

@@
 
+Now that we have a satisfactory model in hand for constitutive
+promoters, we would like to return to the main thread: can we
+infer repressor binding and unbinding rates from mRNA distributions
+over a population of cells as measured by FISH in~\cite{Jones2014}?
+If so, how do these inferred rates compare to direct
+single-molecule measurements such as from~\cite{Hammar2014} and
+to binding energies such as from~\cite{Garcia2011a}
+and~\cite{Razo-Mejia2018}, which were inferred under the assumptions
+of the equilibrium models in Figure~\ref{fig1:means_cartoons}(B)?
+And can this comparison shed light on the unreasonable effectiveness
+of the equilibrium models?
+
 \begin{figure}%[h!]
@@
 
+As we found in Section~\ref{sec:beyond_means}, for our purposes
+the ``right'' model of a constitutive promoter is the bursty
+picture, model five in Figure~\ref{fig:constit_cartoons}.
+Therefore our starting point here is the analogous model with
+repressor added, model 5 in Figure~\ref{fig1:means_cartoons}.
+For a given repressor binding site and copy number,
+this model has four rate parameters to be inferred:
+the repressor binding and unbinding rates $k_R^+$, and $k_R^-$,
+the initiation rate of bursts, $k_i$, and the mean burst size $b$
+(we nondimensionalize all of these by the mRNA degradation rate $\gamma$).
+
+Before showing the mathematical formulation of our statistical
+inference model, we would like to sketch the intuitive structure.
+The dataset from~\cite{Jones2014} we consider consists of
+single-cell mRNA FISH measurements of nine different conditions,
+spanning several combinations of three unique binding sites and
+four unique repressor copy numbers. We assume that the values of
+$k_i$ and $b$ are known, since we have already cleanly inferred
+them from constitutive promoter data, and further we assume that
+these values are the same across datasets with different
+repressor binding sites and copy numbers. We assume that there is
+one unbinding rate parameter for each repressor binding site,
+which is shared across conditions, and likewise one binding rate
+for each unique repressor copy number. This makes our model seven
+dimensional, or nine if one counts $k_i$ and $b$ as well.
+
+Formally now, denote the set of seven repressor rates to be inferred as
+\begin{equation}
+\vect{k} =\{k_{Oid}^-, k_{O1}^-, k_{O2}^-,
+k_{0.5}^+, k_{1}^+, k_{2}^+, k_{10}^+\}.
+\end{equation}
+Note the repressor copy numbers are not known directly, so we label their
+association rates by the concentration of a small molecule inducer aTc
+which indirectly controls the transcription of $\textit{lacI}$.
+Also note that the authors of~\cite{Jones2014} report estimates
+of LacI copy number per cell rather than direct measurements.
+However, these estimates were made \textit{assuming} the validity
+of the equilibrium models in Figure~\ref{fig1:means_cartoons},
+and since \textit{testing} these models is our goal, we will make
+no assumptions about the LacI copy number for given aTc concentrations.
+
+Bayes theorem reads simply
+\begin{equation}
+p(\vect{k}, k_i, b \mid D)
+\propto
+p(D \mid\vect{k}, k_i, b) p(\vect{k}, k_i, b),
+\end{equation}
+where $D$ is the set of all $N$ observed single-cell mRNA counts
+across the various conditions. We assume that individual single-cell
+measurements are independent so that the likelihood factorizes as
+\begin{equation}
+p(D \mid\vect{k}, k_i, b)
+= \prod_{j=1}^N p(m\mid \vect{k}, k_i, b)
+= \prod_{j=1}^N p(m\mid k_j^+, k_j^-, k_i, b)
+\end{equation}
+where $k_j^\pm$ represent the appropriate binding and unbinding
+rates for the $j$-th measured cell. The probability
+$p(m\mid k_j^+, k_j^-, k_i, b)$ appearing in the last expression
+is exactly Eq.~\ref{eq:p_m_bursty+rep_appdx}, the steady-state
+distribution for our bursty model with repression derived in
+Section~\ref{sec:gen_fcn_appdx}, which for completeness we reproduce here as
+\begin{equation}
+\begin{split}
+p(m \mid k_R^+, k_R^-, k_i, b) = & ~\frac{
+        \Gamma(\alpha + m)\Gamma(\beta + m)\Gamma(k_R^+ + k_R^-)
+        }
+        {
+        \Gamma(\alpha)\Gamma(\beta)\Gamma(k_R^+ + k_R^- + m)
+        }
+\frac{b^m}{m!}
+\\
+&\times {_2F_1}(\alpha+m, \beta+m, k_R^++k_R^-+m; -b).
+\end{split}
+\label{eq:p_m_bursty+rep}
+\end{equation}
+where $\alpha$ and $\beta$, defined for notational convenience, are
+\begin{align}
+\begin{split}
+\alpha &= \frac{1}{2}
+\left(k_i+k_R^-+k_R^+ + \sqrt{(k_i+k_R^-+k_R^+)^2 - 4k_i k_R^-}\right)
+\\
+\beta &= \frac{1}{2}
+\left(k_i+k_R^-+k_R^+ - \sqrt{(k_i+k_R^-+k_R^+)^2 - 4k_i k_R^-}\right).
+\end{split}
+\end{align}
+
+\mmnote{Probably SI most of this paragraph?}
+This likelihood is rather inscrutable. We did not find any of the
+known analytical approximations for ${_2F_1}$ terribly useful in
+gaining intuition, so we instead resorted to numerics. One insight we found
+\mmnote{a plot would probably help explain, maybe SI, need to explain better}
+was that for very strong or very weak repression, the distribution in
+Eq.~\ref{eq:p_m_bursty+rep} is well approximated by a negative binomial
+with burst size $b$ and burst rate $k_i$ equal to their
+constitutive \textit{lacUV5} values, except with $k_i$ multiplied
+by the fold-change $\left(1+k_R^+/k_R^-\right)^{-1}$.
+In other words, once again only the \textit{ratio} $k_R^+/k_R^-$
+was detectable. But for intermediate repression, the distribution
+was visibly broadened with Fano factor greater than $1+b$, the
+value for the corresponding constitutive case. This indicates
+that the repressor rates had left an imprint on the distribution,
+and perhaps intuitively, this intermediate regime occurs for values of
+$k_R^\pm$ comparable to the burst rate $k_i$. Put another way,
+if the repressor rates are much faster or much slower than $k_i$,
+then there is a timescale separation and effectively only one
+timescale remains, $k_i\left(1+k_R^+/k_R^-\right)^{-1}$.
+Only when all three rates in the problem are comparable does the
+mRNA distribution retain detectable information about them.
+
+Next we specify priors. As for the constitutive model, weakly
+informative lognormal priors are a natural choice for all our
+rates. We found that if the priors were too weak, our MCMC
+sampler would often become stuck in regions of parameter space
+with very low probability density, unable to move. We struck a
+balance in choosing our prior widths between helping the sampler
+run while simultaneously verifying that the marginal posteriors
+for each parameter were not artificially constrained or distorted
+by the presence of the prior. The only exception to this is the
+highly informative priors we placed on $k_i$ and $b$, since we
+have strong knowledge of them from our inference of constitutive
+promoters above.
+
+With priors and likelihood specified we may write down our complete generative model as
+\begin{equation}
+\begin{split}
+\log_{10}k_i &\sim \text{Normal}(0.725, 0.025)\\
+\log_{10}b   &\sim \text{Normal}(0.55, 0.025)\\
+\log_{10}k_{0.5}^+ &\sim \text{Normal}(-0.45, 0.3)\\
+\log_{10}k_{1}^+   &\sim \text{Normal}(0.6, 0.3)\\
+\log_{10}k_{2}^+   &\sim \text{Normal}(1.15, 0.3)\\
+\log_{10}k_{10}^+  &\sim \text{Normal}(1.5, 0.3)\\
+\log_{10}k_{Oid}^- &\sim \text{Normal}(-0.25, 0.3)\\
+\log_{10}k_{O1}^-  &\sim \text{Normal}(0.1, 0.3)\\
+\log_{10}k_{O2}^-  &\sim \text{Normal}(0.45, 0.3)\\
+m &\sim \text{Likelihood}(k_R^+, k_R^-, k_i, b),
+\end{split}
+\end{equation}
+where the likelihood is specified by Eq.~\ref{eq:p_m_bursty+rep}.
+\mmnote{Now that I've typed this out, this notation seems totally
+pointless here, especially when the likelihood is not a standard
+distribution. Maybe I should make a table or something to record
+the prior values, and maybe even that should go to SI?}
+We ran MCMC sampling on the full nine dimensional posterior
+specified by this generative model. To attempt to visualize this
+object, in Figure~\ref{fig4:repressed_post_full}(A) we plot
+several two-dimensional slices as contour plots, analogous to
+Figure~\ref{fig:constit_post_full}(C). Each of these nine slices
+corresponds to the $(k_R^+, k_R^-)$ pair of rates for one of the
+conditions from the dataset used to fit the model and gives a
+sense of the uncertainty and correlations in the posterior.
+
 \mmnote{Outline:
 \begin{itemize}
-\item Full model. Write Bayes, handwave 2F1 dist story in limits of weak,
-strong, and intermediate repression.
 \item Hey look it mostly works. With full model, 9D model is well identified,

### 2020-05-25 doc/appendix_01_bayesian_inference.tex
update minor things such as figure naming

@@
 in~\cite{Razo-Mejia2020} (nonequilibrium model four
-in~\ref{fig:constit_cartoons}), this model is well-identified
+in~\ref{fig2:constit_cartoons}), this model is well-identified
 with both parameters determined to a fractional uncertainty of 5-10\%.

### 2020-05-25 doc/section_02_means.tex
update minor things such as figure naming

@@
 nonequilibrium model 4 in~\fig{fig1:means_cartoons}(C), with zero repressor
-(looking ahead, this is equivalent to model 4 in~\fig{fig:constit_cartoons}), to
+(looking ahead, this is equivalent to model 4 in~\fig{fig2:constit_cartoons}), to
 FISH data from~\cite{Brewster2014}. This provided a determination of $k^+$ and

### 2020-05-25 doc/section_03_beyond_means.tex
Fixed my error in Fig 3D scaling, add MRM's new Fig4
Found an arithmetic error in log(k_i)~deps_P argument,
realized intercept has meaning after all. Updated plot
and text to reflect. Also added MRM's beautiful new Fig 4
update minor things such as figure naming

@@
 \section{Beyond Means in Gene Expression}
+\label{sec:beyond_means}
 
@@
 we can choose, and we show several potential kinetic models for constitutive
-promoters in~\fig{fig:constit_cartoons}. Let us consider the suitability of each
+promoters in~\fig{fig2:constit_cartoons}. Let us consider the suitability of each
 model for our purposes in turn.
@@
 complexity.}
-\label{fig:constit_cartoons}
+\label{fig2:constit_cartoons}
 \end{figure}
@@
 The simplest model of constitutive expression that we can imagine
-is shown as model 1 in Figure~\ref{fig:constit_cartoons} and
+is shown as model 1 in Figure~\ref{fig2:constit_cartoons} and
 assumes that
@@
 ~\cite{Phillips2015a, Phillips2019}. Here,  as shown in 
-the second row of Figure~\ref{fig:constit_cartoons},  transcription initiation results in
+the second row of Figure~\ref{fig2:constit_cartoons},  transcription initiation results in
 a state transition from the bound to
@@
 the elongation phase. Perhaps adding some of this mechanistic detail as shown
-in Figure~\ref{fig:constit_cartoons} might
+in Figure~\ref{fig2:constit_cartoons} might
 rescue the previous model   The next simplest refinement of that model could consider open

### 2020-05-25 doc/section_04_bayesian_inference.tex
update minor things such as figure naming

@@
 From consideration of Fano factors in the previous section, we suspect that
-model 5 in Figure~\ref{fig:constit_cartoons}, a one-state bursty model of
+model 5 in Figure~\ref{fig2:constit_cartoons}, a one-state bursty model of
 constitutive promoters, achieves a Goldilocks level of complexity. Does this
@@
 It will be instructive, however, to first consider the Poisson promoter, model 1
-in Figure~\ref{fig:constit_cartoons}. As we alluded to earlier, since the
+in Figure~\ref{fig2:constit_cartoons}. As we alluded to earlier, since the
 Poisson distribution has a Fano factor $\nu$ strictly equal to 1, and all of
@@
 case. Using the same argument we can immediately rule out models 2 and 3 from 
-Figure~\ref{fig:constit_cartoons}. these models have Fano factors $\nu\le 1$
+Figure~\ref{fig2:constit_cartoons}. these models have Fano factors $\nu\le 1$
 meaning they are underdisperse relative to Poisson. We will also not explicitly
-consider model 4 from~\fig{fig:constit_cartoons} since it was already thoroughly
+consider model 4 from~\fig{fig2:constit_cartoons} since it was already thoroughly
 analyzed in~\cite{Razo-Mejia2020}, and since model 5 can be viewed as a special
@@
 active/inactive constitutive model considered in~\cite{Razo-Mejia2020}
-(nonequilibrium model 4 in~\ref{fig:constit_cartoons}), this model is
+(nonequilibrium model 4 in~\ref{fig2:constit_cartoons}), this model is
 well-identified with both parameters determined to a fractional uncertainty of
@@
 the ``right'' model of a constitutive promoter is the bursty
-picture, model five in Figure~\ref{fig:constit_cartoons}.
+picture, model five in Figure~\ref{fig2:constit_cartoons}.
 Therefore our starting point here is the analogous model with

### 2020-05-25 doc/section_05_discussion.tex
RP new version of discussion
MRM pass through discussion

--- 2020-05-21 doc/section_05_discussion.tex
@@
 \section{Discussion and future work}
+
+The study of gene expression is one of the dominant themes of modern biology,
+made all the more urgent by the dizzying pace at which genomes are being
+sequenced. But there is a troubling Achilles heel buried in all of that genomic
+data and that is our inability to find and interpret regulatory sequence. In
+many cases, this is not possible even qualitatively, let alone the possibility
+of quantitative dissection of the regulatory parts of genomes in a predictive
+fashion. Other recent work has tackled the challenge of finding and annotating
+the regulatory part of genomes \cite{Belliveau2018, Ireland2020} \mrm{add other
+non-self referential}. Once we have determined the architecture of the
+regulatory part of the genome, we are then faced with the next class of
+questions which are sharpened by formulating them in mathematical terms, namely,
+what are the input-output properties of these regulatory circuits and what knobs
+control them?
+
+The present work has tackled that question in the context of the first
+regulatory architecture hypothesized in the molecular biology era, namely, the
+repressor-operator model of Jacob and Monod. Regulation in that architecture is
+the result of a competition between a repressor which inhibits transcription and
+RNAP polymerase which undertakes it. Through the labors of generations of
+geneticists, molecular biologists and biochemists, an overwhelming amount of
+information and insight has been garnered into this simple regulatory motif,
+licensing it as what one might call the ``hydrogen atom'' of regulatory biology.
+It is from that perspective that the present paper explores the extent to which
+some of the different models that have been articulated to describe that motif
+allow us to understand both the average level of gene expression found in a
+population of cells, the intrinsic cell-to-cell variability, and the full gene
+expression distribution found in such population as would be reported in a FISH
+experiment, for example.
+
+Our key insights can be summarized as follows. First, as shown in 
+Figure~\ref{fig1:means_cartoons}, the mean expression in the simple repression
+architecture is captured by a master curve in which the action of repressor and
+the details of the polymerase interaction with the promoter appear separately
+and additively in an effective free energy. Interestingly, as has been shown
+elsewhere in the context of the Monod-Wyman-Changeux model, these kinds of
+coarse-graining results are an exact mathematical result and do not constitute
+hopeful approximations or biological naivete \cite{Razo-Mejia2018, Chure2019}.
+To further dissect the relative merits of the different models, we must appeal
+to higher moments of the gene expression probability distribution. To that end,
+our second set of insights focus on gene expression noise, where it is seen that
+a treatment of the constitutive promoter already reveals that some models have
+Fano factors (standard deviation/mean) that are less than one, at odds with
+any and all experimental data that we are aware of.
+\mrm{
+This theoretical result allows us to directly discard a subset of the models
+(models 1-3 in Figure~\ref{fig2:fig2:constit_cartoons}) since they cannot be
+reconciled with experimental observations. The two models left (models 4 and 5
+in Figure~\ref{fig2:fig2:constit_cartoons}) seem to provide a rich enough
+phenomenology to be able to reproduce the data. A previous exploration of one 
+of these models demonstrated the sloppy nature of the model in which single-cell
+mRNA counts data alone cannot constrain the value of all parameters 
+simultaneously \cite{Razo-Mejia2020}. Here we demonstrate that the proposed
+one-state bursty promoter model (model 5 in 
+Figure~\ref{fig2:fig2:constit_cartoons}) is a limit of the commonly used
+two-state promoter model \cite{Peccoud1995}. We put at test the idea that this
+level of coarse-graining is descriptive enough to reproduce previous 
+experimental observations. In particular we perform Bayesian inference on the
+two parameters describing the full steady-state mRNA distribution, finding that
+the model is able to capture a plethora of promoter sequences with different
+mean levels of expression and noise.
+}
+
+\mrm{
+Here we should include a discussion of the regulated case.
+}
+
+It is interesting to speculate what microscopic details are being coarse-grained
+by our burst rate and burst size in Figure~\ref{fig2:constit_cartoons} model 5.
+Chromosomal locus is one possible influence we have not addressed in this work,
+as all the FISH data from~\cite{Jones2014} that we considered was from a
+construct integrated at the \textit{galK} locus. The results of~\cite{Chong2014}
+indicate that transcription-induced supercoiling contributes substantially in
+driving transcriptional bursting, and furthermore, their Figure~7 suggests that
+the parameters describing the rate, duration, and size of bursts vary
+substantially for transcription from different genomic loci. Although the
+authors of~\cite{Englaender2017} do not address noise, they note enormous
+differences in mean expression levels when an identical construct is integrated
+at different genomic loci. The authors of~\cite{Engl2020} attribute noise and
+burstiness in their FISH data to the influence of different sigma factors, which
+is a reasonable conclusion from their data. Could the difference also be due to
+the different chromosomal locations of the two operons?
+
+It seems clear therefore that chromosomal locus plays a major role in mean
+expression levels as well as noise. What features of different loci are and are
+not important? Sigma factors could certainly play a part, but it seems many
+other mechanisms are at play: supercoiling, and more generally 3D organization
+of the chromosome, read-through of neighboring genes, and surely some
+fascinating unknown unknowns. Could our preferred coarse-grained model capture
+the variability across different loci? If so, and we were to repeat the
+parameter inference as done in this work, is there a simple theoretical model we
+could build to understand the resulting parameters?
+
 \mmnote{Do all my approximations work simply because galK happens to have a low
@@
 
-It is interesting to speculate what microscopic details are being
-coarse-grained by our burst rate and burst size in model 5.
-Chromosomal locus is one possible influence we have not addressed
-in this work, as all the FISH data from~\cite{Jones2014} that we
-considered was from a construct integrated at the \textit{galK} locus.
-The results of~\cite{Chong2014} indicate that
-transcription-induced supercoiling contributes substantially in
-driving transcriptional bursting, and furthermore, their Figure~7
-suggests that the parameters describing the rate, duration, and
-size of bursts vary substantially for transcription from
-different genomic loci.
-Although the authors of~\cite{Englaender2017} do not address
-noise, they note enormous differences in mean expression levels
-when an identical construct is integrated at different genomic loci.
-The authors of~\cite{Engl2020} attribute noise and burstiness in
-their FISH data to the influence of different sigma factors,
-which is a reasonable conclusion from their data.
-Could the difference also be due to the different chromosomal
-locations of the two operons?
-
-It seems clear therefore that chromosomal locus plays a major
-role in mean expression levels as well as noise. What features of
-different loci are and are not important? Sigma factors could
-certainly play a part, but it seems many other mechanisms are at
-play: supercoiling, and more generally 3D organization of the
-chromosome, read-through of neighboring genes, and surely some
-fascinating unknown unknowns.
-Could our preferred coarse-grained model capture the variability
-across different loci? If so, and we were to repeat the parameter
-inference as done in this work, is there a simple theoretical model
-we could build to understand the resulting parameters?
-
 A direct \mmnote{if cloning intensive} future experiment would be

### 2020-05-25 code/figures/main/fig03.py
turning legends into legendary legends

@@ -142,7 +142,7 @@ ax_a1.hist(b_samples,
 )
 # Add gridlines
 [ax_a3.axhline(x, color="white", linewidth=0.5) 
-for x in [5, 5.5, 6]]
+for x in [4.75, 5, 5.25, 5.5, 5.75, 6]]
 
 ax_a3.hist(alpha_samples,
           50,
@@ -176,7 +176,7 @@ srep.viz.predictive_ecdf(
     percentiles=ptiles,
     discrete=True,
     ax=ax_b,
-    pred_label='Model 1 (Poisson) PPC',
+    pred_label='Model 1 (Poisson)',
     )
 
 # next neg binom model + data overlaid
@@ -187,18 +187,35 @@ srep.viz.predictive_ecdf(
     data=df_UV5["mRNA_cell"],
     percentiles=ptiles,
     discrete=True,
-    # diff=True,
     ax=ax_b,
-    pred_label='Model 5 (N. Binom) PPC',
+    pred_label='Model 5 (N. Binom)',
     data_label='UV5 data, Jones et. al.',
-    data_color='orange',
-    data_size=1 #linewidth
+    data_color='black',
+    data_size=1, #linewidth
     )
+# Add plots for composite legend
+# 1. Extract colors
+blues = srep.viz.bebi103_colors()["blue"]
+greens = srep.viz.bebi103_colors()["green"]
+# 2. Declarre plots
+poisson_1 = ax_b.plot([], [], color=greens[-1])
+poisson_2= ax_b.fill([], [], color=greens[0])
+neg_binom_1 = ax_b.plot([], [], color=blues[-1])
+neg_binom_2= ax_b.fill([], [], color=blues[0])
+uv5 = ax_b.plot([], [], color="black")
 
-ax_b.legend(loc='lower right', fontsize='small')
+ax_b.legend(
+    [(poisson_2[0], poisson_1[0]), (neg_binom_2[0], neg_binom_1[0]), (uv5[0],),], 
+    ["Model 1 (Poisson)", "Model 5 (N. Binom)", "UV5 data, Jones et. al."],
+    fontsize=8,
+    loc="lower right",
+)
+
+# ax_b.legend(loc='lower right', fontsize='small')
 ax_b.set_xlabel('mRNA counts per cell')
 ax_b.set_ylabel('ECDF')
 ax_b.set_xlim(right=60)
+
 
 # (C)
[…]

### 2020-05-25 code/figures/main/fig04.py

@@ -334,8 +334,42 @@ for k, ax in enumerate([ax_3, ax_4, ax_5]):
     # ax.legend(loc='lower right', fontsize='small')
 # Set yaxis label for first plot
 ax_3.set_ylabel('ECDF')
+
+# Add UV5 to dictionary
+aTc_col_dict["UV5"] = "purple"
+# Generate legend with shaded region symbols
+# 1. Initialize list to save plots
+plot_list = list()
+# 2. Initialize lsit to save labels
+label_list = list()
+# 3. Loop through concentrations
+for key, item in aTc_col_dict.items():
+    # 4. Extract colors
+    colors = srep.viz.bebi103_colors()[item]
+    # 5. Generate plots
+    p1 = ax_5.plot([], [], color=colors[-1])
+    p2= ax_5.fill([], [], color=colors[0])
+    # 6. Fix label name
+    label = expt.split("_")[1]
+    label = label.replace("ngmL", " ng/mL")
+    label = label.replace("0p5", "0.5")
+    # 7. Append to lists
+    plot_list.append((p2[0], p1[0],))
+    label_list.append(label)
+# Append data plot
+p = ax_5.plot([], [], color="black")
+plot_list.append((p[0],))
+label_list.append("data")
+
 # Set legend for last plot
-ax_5.legend(loc="lower right", bbox_to_anchor=(1, -0.4), ncol=6, fontsize=8)
+ax_5.legend(
+    plot_list,
+    label_list,
+    loc="lower right",
+    bbox_to_anchor=(1, -0.4),
+    ncol=6,
+    fontsize=8
+)
 
 # Add plot labels
 plt.gcf().text(-0.03, 1, "(A)", fontsize=14)

### 2020-05-25 code/figures/main/fig04_v2.py

@@ -407,8 +407,42 @@ for k, ax in enumerate([ax_3, ax_4, ax_5]):
     # ax.legend(loc='lower right', fontsize='small')
 # Set yaxis label for first plot
 ax_3.set_ylabel('ECDF')
+
+# Add UV5 to dictionary
+aTc_col_dict["UV5"] = "purple"
+# Generate legend with shaded region symbols
+# 1. Initialize list to save plots
+plot_list = list()
+# 2. Initialize lsit to save labels
+label_list = list()
+# 3. Loop through concentrations
+for key, item in aTc_col_dict.items():
+    # 4. Extract colors
+    colors = srep.viz.bebi103_colors()[item]
+    # 5. Generate plots
+    p1 = ax_5.plot([], [], color=colors[-1])
+    p2= ax_5.fill([], [], color=colors[0])
+    # 6. Fix label name
+    label = expt.split("_")[1]
+    label = label.replace("ngmL", " ng/mL")
+    label = label.replace("0p5", "0.5")
+    # 7. Append to lists
+    plot_list.append((p2[0], p1[0],))
+    label_list.append(label)
+# Append data plot
+p = ax_5.plot([], [], color="black")
+plot_list.append((p[0],))
+label_list.append("data")
+
 # Set legend for last plot
-ax_5.legend(loc="lower right", bbox_to_anchor=(1, -0.4), ncol=6, fontsize=8)
+ax_5.legend(
+    plot_list,
+    label_list,
+    loc="lower right",
+    bbox_to_anchor=(1, -0.4),
+    ncol=6,
+    fontsize=8
+)
 
 # Add plot labels
 plt.gcf().text(-0.03, 1, "(A)", fontsize=14)

### 2020-05-26 doc/appendix_03_regulated_promoters.tex
update minor things such as figure naming
making lots of progress on the bursty promoter derivation

@@
-% !TEX root = ./busty_transcription.tex
+% !TEX root = ./bursty_transcription.tex
 \section{Bursty promoter models - generating function solutions and numerics}
@@
 
-We would like to derive the full probability distribution of
-mRNA copy number per cell for a model combining simple repression
-with a promoter featuring bursty transcription initiation.
-Before tackling mRNA and repression together, let us
-just consider the bursty consititutive promoter,
-model 5 in~\fig{fig:constit_cartoons}
-\mmnote{replicating Charlotte's notes; she says it's worked out
-in~\cite{Paulsson2000}, but I don't see it, at least not in notation I can
-comprehend. Need to find a cite for this,
-or maybe just cite Charlotte, personal comm, since that's where I actually got it\dots}.
-The master equation of interest is
+The objective of this section is to write down the steady-state mRNA
+distribution for model 5 in Figure~\ref{fig2:constit_cartoons}. Our claim is
+that this model is rich enough that it can capture the expression pattern of
+bacteria unregulated promoters regulated by the $\sigma_{70}$ sigma factor.
+Figure~\ref{figS1:bursty_one_state} shows two different schematic
+representations of the model. Figure~\ref{figS1:bursty_one_state}(A) shows the
+promoter cartoon model with burst initiation rate $k_i$, mRNA degradation rate
+$\gamma$, and mean burst size $b$. For our derivation of the chemical master
+equation we will focus more on Figure~\ref{figS1:bursty_one_state}(B). This
+representation is intended to highlight that bursty gene expression allows
+transitions between mRNA count $m$ and $m'$ even with $m - m' > 1$.
+
+\begin{figure}[h!]
+\centering
+\includegraphics{../figures/si/figS0X_bursty_states.pdf}
+\caption{
+\textbf{Bursty transcription for unregulated promoter.}
+(A) Schematic of the one-state bursty transcription model. Rate $k_i$ is the
+bursty initiation rate, $\gamma$ is the mRNA degradation rate, and $b$ is the
+mean burst size. (B) Schematic depiction of the mRNA count state transitions.
+The model in (A) allows for transitions of $> 1$ mRNA counts with probability
+$G_{m-m'}$.}
+\label{figS1:bursty_one_state}
+\end{figure}
+
+To derive the master equation we begin by thinking the possible state
+transitions to ``enter'' state $m$. There are two possible paths to jump from an
+mRNA count $m' \neq m$ to a state $m$ in a small time window $\Delta t$:
+\begin{enumerate}
+        \item By degradation of a single mRNA, jumping from $m+1$ to $m$.
+        \item By producing $m-m'$ mRNA for $m' \in \{0, 1, \ldots, m-1\}$.
+\end{enumerate}
+For the ``exit'' states from $m$ into $m' \neq m$ during a small time window
+$\Delta t$ we also have two possibilities:
+\begin{enumerate}
+        \item By degradation of a single mRNA, jumping from $m$ to $m-1$.
+        \item By producing $m'-m$ mRNA for $m'-m \in \{1, 2, \ldots\}$.
+\end{enumerate}
+This implies that the probability of having $m$ mRNA at time $t + \Delta t$ can
+be written as
+\begin{equation}
+\begin{split}
+p(m, t + \Delta t) = &p(m, t)
++ \overbrace{\gamma \Delta t (m + 1) p(m + 1, t)}^{m + 1 \rightarrow m}
+- \overbrace{\gamma \Delta t m p(m, t)}^{m \rightarrow m - 1} \\
+&+ \overbrace{k_i \Delta t \sum_{m'=0}^{m-1} G_{m-m'} p(m', t)}^
+{m'\in \{0, 1, \ldots m-1\} \rightarrow m}
+- \overbrace{k_i \Delta t \sum_{m'=m + 1}^{\infty} G_{m'-m} p(m, t)}^
+{m \rightarrow m'\in \{m+1, m+2, \ldots\}},
+\end{split}
+\label{eq:si_master_deltat}
+\end{equation}
+where we indicate $G_{m'-m}$ as the probability of having a burst of size
+$m'-m$. We suggestively use the letter $G$ as we will assume that these bursts
+sizes are geometrically distributed with parameter $\theta$. This is written as
+\begin{equation}
+G_{k} = \theta (1 - \theta)^k\; \text{for } k \in \{0, 1, 2, \ldots \}.
+\end{equation}
+What this implies is that for a geometrically distributed burst size we have a
+mean burst size $b$ of the form
+\begin{equation}
+b \equiv \left\langle m' - m \right\rangle 
+= \sum_{k=0}^\infty k \theta (1 - \theta)^k = {1 - \theta \over \theta}.
+\end{equation}
+
+To clean up Equation~\ref{eq:si_master_deltat} we can send the first term on the
+right hand side to the left, and divide both sides by $\Delta t$. Upon taking
+the limit where $\Delta t \rightarrow 0$ we can write
+\begin{equation}
+{d \over dt}p(m, t) = (m + 1) \gamma p(m + 1, t)
+- m \gamma p(m, t)
++ k_i \sum_{m'=0}^{m-1} G_{m-m'} p(m', t) 
+- k_i \sum_{m'=m + 1}^{\infty} G_{m'-m} p(m, t).
+\end{equation}
+Furthermore, given that the timescale for this equation is set by the mRNA
+degradation rate $\gamma$ we can divide both sides by this rate, obtaining
+\begin{equation}
+{d \over d\tau}p(m, \tau) = (m + 1) p(m + 1, \tau)
+- m p(m, \tau)
++ \lambda \sum_{m'=0}^{m-1} G_{m-m'} p(m', \tau) 
+- \lambda \sum_{m'=m + 1}^{\infty} G_{m'-m} p(m, \tau),
+\label{eq:si_master_ode}
+\end{equation}
+where we defined $\tau \equiv t \times \gamma$, and $\lambda \equiv k_i/\gamma$.
+The last term in Eq.~\ref{eq:si_master_ode} sums all burst sizes except for a 
+burst of size zero. We can re-index the sum to include this term, obtaining
+\begin{equation}
+\lambda \sum_{m'=m + 1}^{\infty} G_{m'-m} p(m, \tau) = \lambda p(m, t) \left[
+\underbrace{\sum_{m'=\textcolor{red}{m}}^{\infty}G_{m'-m}}
+_{\text{re-index sum to include burst size zero}} -
+\underbrace{G_0}_{\text{subtract extra added term}}\right].
+\end{equation}
+Given the normalization of the geometric distribution, adding the probability of
+all possible burst sizes -- including size zero since we re-indexed the sum --
+allows us to write
+\begin{equation}
+\sum_{m'=m}^{\infty}G_{m'-m} - G_0 = 1 - G_0.
+\end{equation}
+Substituting this into Eq.~\ref{eq:si_master_ode} results in
+\begin{equation}
+{d \over d\tau}p(m, \tau) = (m + 1) p(m + 1, \tau)
+- m p(m, \tau)
++ \lambda \sum_{m'=0}^{m-1} G_{m-m'} p(m', \tau) 
+- \lambda p(m, \tau) \left[ 1 - G_0 \right].
+\label{eq:si_master_ode_2}
+\end{equation}
+To finally get at a more compact version of the equation notice that the third
+term in Eq.~\ref{eq:si_master_ode_2} includes burst from size $m'-m = 1$ to size
+$m' - m = m$. We can include the term $p(m, t) G_0$ in the sum which allows
+bursts of size $m' - m = 0$. This results in our final form for the chemical
+master equation
 \begin{align}
-\deriv{t}p(m,t) = (m+1)\gamma p(m+1,t) - m\gamma p(m,t) - r p(m,t)
-        + r \sum_{m^\prime=0}^m G_{m-m^\prime}(\theta) p(m^\prime,t),
-\label{eq:1state_unreg_003}
+{d \over d\tau}p(m, \tau) = 
+(m + 1) p(m+1, \tau)
+- m p(m, \tau) - 
+\lambda p(m, \tau)
++ \lambda \sum_{m^\prime=0}^m G_{m-m^\prime} p(m^\prime, \tau).
+\label{eq:si_master_unreg}
 \end{align}
-where $G_{k}(\theta)$ is the geometric distribution defined as
+
+In order to solve Eq.~\ref{eq:si_master_unreg} we will use the generating
+function method \mrm{reference?}. The probability generating function is defined
+as
 \begin{align}
-G_{k}(\theta) = \theta(1 - \theta)^k, \, k\in\{0,1,2,\dots\}.
+F(z,t) = \sum_{m=0}^\infty z^m p(m,t),
 \end{align}
-With this convention, the mean burst size $b = (1-\theta)/\theta$.
-$\gamma$ and $r$ are mRNA degradation rates and transcription burst
-initiation rates, resp. The last term represents all ways the system could
-end up with $m$ mRNAs, having started the burst with $m^\prime$. Define
-$\lambda = r/\gamma$ and nondimensionalize time by $\gamma$, giving
+where $z$ is just a dummy variable that will help us later on to obtain the
+moments of the distribution. Applying $F(z, t)$ on both sides of 
+Eq.~\ref{eq:si_master_unreg} results in
+\begin{equation}
+\sum_m z^m {d \over d\tau} p(m, \tau) = 
+\sum_m z^m \left[ 
+- m p(m, \tau) 
++ (m + 1) p(m + 1, \tau) 
++ \lambda \sum_{m' = 0}^m G_{m-m'} p(m', \tau) - \lambda p(m, \tau)
+\right],
+\end{equation}
+where we use $\sum_m \equiv \sum_{m=0}^\infty$. We can distribute the sum and
+use the definition of $F(z, t)$ to obtain
+\begin{equation}
+{d F(z, \tau) \over d\tau} =
+- \sum_m z^m m p(m, \tau) +
++ \sum_m z^m (m + 1) p(m + 1, \tau)
++ \lambda \sum_m z^m \sum_{m'=0}^m G_{m-m'} p(m', \tau)
+- \lambda F(z, \tau).
+\label{eq:si_generating_01}
+\end{equation}
+We can make use of properties of the generating function to write everything in
+terms of $F(z, \tau)$: the first term on the right hand side of
+Eq.~\ref{eq:si_generating_01} can be rewritten as
 \begin{align}
-\deriv{t}p(m,t) = (m+1)p(m+1,t) - m p(m,t) - \lambda p(m,t)
-        + \lambda \sum_{m^\prime=0}^m G_{m-m^\prime}(\theta) p(m^\prime,t),
+\sum_{m} z^{m} \cdot m \cdot p(m, \tau) &=
+\sum_{m} z \frac{\partial z^{m}}{\partial z} p(m, \tau), \\
+&=\sum_{m} z \frac{\partial}{\partial z}\left(z^{m} p(m, \tau)\right), \\
+&=z \frac{\partial}{\partial z}\left(\sum_{m} z^{m} p(m, \tau)\right), \\
+&= z {\partial F(z, \tau) \over \partial z}.
 \end{align}
-The probability generating function is defined as
+For the second term on the right hand side of Eq.~\ref{eq:si_generating_01} we
+define $k \equiv m + 1$. This allows us to write
 \begin{align}
-F(z,t) = \sum_{m=0}^\infty z^m p(m,t).
+\sum_{m=0}^{\infty} z^{m} \cdot(m+1) \cdot p(m+1, \tau) &=
+\sum_{k=1}^{\infty} z^{k-1} \cdot k \cdot p(k, \tau), \\
+&=z^{-1} \sum_{k=1}^{\infty} z^{k} \cdot k \cdot p(k, \tau), \\
+&=z^{-1} \sum_{k=0}^{\infty} z^{k} \cdot k \cdot p(k, \tau), \\
+&=z^{-1} \left(z \frac{\partial F(z)}{\partial z}\right), \\
+&=\frac{\partial F(z)}{\partial z}.
 \end{align}
-Multiply both sides of the CME by $z^m$ and sum over all $m$ to get
+The third term in Eq.~\ref{eq:si_generating_01} is the most trouble. The trick
+is to reverse the default order of the sums as
+\begin{equation}
+\sum_{m=0}^{\infty} \sum_{m'=0}^{m} = \sum_{m'=0}^{\infty} \sum_{m=m'}^{\infty}.
+\end{equation}
+This results in
 \begin{align}
-\pderiv[F]{t} = \sum_{m=0}^\infty z^m
-\left(
-(m+1)p(m+1,t) - m p(m,t) - \lambda p(m,t)
-+ \lambda \sum_{m^\prime=0}^m G_{m-m^\prime}(\theta) p(m^\prime,t)
-\right).
+\lambda \sum_m z^m \sum_{m'=0}^m G_{m-m'} p(m', \tau) =
+\lambda \sum_{m'=0}^{\infty} \sum_{m=m'}^{\infty} z^m 
+\theta (1 - \theta)^{m-m'} p(m', \tau),
 \end{align}
-The first two terms can be rewritten as derivatives of the
-generating function, i.e.,
+where we also substituted the definition of the geometric distribution $G_{k} =
+\theta (1 - \theta)^k$. Redistributing the sums we can write
 \begin{align}
-\sum_{m=0}^\infty z^m (m+1)p(m+1,t)
-=
-\pderiv{z} \sum_{m=0}^\infty z^{m+1} p(m+1,t)
-=
-\pderiv[F]{z}
+\lambda \sum_{m'=0}^{\infty} \sum_{m=m'}^{\infty} z^m 
+\theta (1 - \theta)^{m-m'} p(m', \tau) = 
+\lambda \theta \sum_{n=0}^{\infty}(1-\theta)^{m'} P(m', \tau) 
+\sum_{m=m'}^{\infty} \left[z (1-\theta)\right]^{m}.
+\label{eq:si_generating_02}
 \end{align}
-and
+We know that $1 - \theta \in (0, 1)$, and that $z$ is a dummy variable that we
+usually evaluate as $z=1$. With these two facts we can safely state that
+$| z (1 - \theta) | < 1$. Defining $n \equiv m - m'$ we rewrite the last sum in 
+Eq.~\ref{eq:si_generating_02} as
 \begin{align}
-\sum_{m=0}^\infty z^m mp(m,t)
-=
-z\pderiv{z} \sum_{m=0}^\infty z^m p(m,t)
-=
-z\pderiv[F]{z}.
+\sum_{m=m'}^{\infty} \left[z (1-\theta)\right]^{m} &= 
+\sum_{n=0}^{\infty} \left[z (1-\theta)\right]^{n + m'} \\
+&= \left[ z (1 - \theta) \right]^{m'} 
+\sum_{n=0}^{\infty} \left[ z (1 - \theta) \right]^{m'} \\
+&= \left[ z (1 - \theta) \right]^{m'} 
+\left( {1 \over 1 - z (1 - \theta)} \right),
 \end{align}
-The final term is the most trouble. The trick is to reverse
-the default order of the sums, as in
-\begin{align}
-&\lambda \sum_{m=0}^\infty \sum_{m^\prime=0}^m
-        z^m \theta(1 - \theta)^{m-m^\prime} p(m^\prime,t)
-\\=
-&\lambda\theta \sum_{m^\prime=0}^\infty z^{m^\prime} p(m^\prime,t)
-        \sum_{m=m^\prime}^\infty
-        z^{m-m^\prime} (1 - \theta)^{m-m^\prime}
-\\=
-&\frac{\lambda\theta F}{1 - (1 - \theta)z}.
-\end{align}
-This works because after reordering the sums, the sum on $m$ is just
-a geometric series. Its result is independent of $m^\prime$, so the sum on
-$m^\prime$ reduces to the definition of the generating function.
+where we use the geometric series since, as stated before,$| z (1 - \theta) | <
+1$
 Putting these results together, the PDE for the generation function is

### 2020-05-26 doc/appendix_03_regulated_promoters.tex
MRM pass through bursty promoter derivation on 20200525

@@
 where we use the geometric series since, as stated before,$| z (1 - \theta) | <
-1$
-Putting these results together, the PDE for the generation function is
-\begin{align}
-\pderiv[F]{t} = \pderiv[F]{z} - z\pderiv[F]{z} - \lambda F
+1$. Putting these results together, the PDE for the generation function is
+\begin{equation}
+{\partial F \over \partial \tau} = 
+{\partial F \over\partial z}
+- z {\partial F \over \partial z} - \lambda F
 + \frac{\lambda\theta F}{1-z(1-\theta)}.
-\end{align}
+\end{equation}
 Changing variables to $\xi=1-\theta$ and simplifying gives
 \begin{align}
-\pderiv[F]{t} + (z - 1) \pderiv[F]{z} = \frac{(z-1)\xi}{1-z\xi}\lambda F.
+{\partial F \over \partial \tau} + (z - 1) {\partial F \over \partial z} = 
+\frac{(z-1)\xi}{1-z\xi}\lambda F.
 \label{eq:1state_unreg_015}
 \end{align}
+
 \subsubsection{Stead-state}
-At steady-state, the PDE reduces to the ODE
+
+To get at the mRNA distribution at steady state we first must solve
+Eq.~\ref{eq:1state_unreg_015} setting the time derivative to zero. At
+steady-state, the PDE reduces to the ODE
 \begin{align}
@@
 \end{align}
-The initial conditions for generating functions can be subtle and confusing.
-The key fact follows from the definition
+The initial conditions for generating functions can be subtle and confusing. The
+key fact follows from the definition
 $F(z,t) = \sum_m z^m p(m,t)$,
 normalization requires that
-$F(z=1^-,t) = \sum_m p(m,t) = 1$.\footnote{
+$F(z=1, t) = \sum_m p(m,t) = 1$.
+\footnote{
 Sometimes the generating function may be undefined \textit{at} $z=1$ but the
@@
 \end{align}
-which is exactly the negative binomial's generating function, as expected.
-Now we may recover the mRNA probability distribution of interest
-by differentiating the generating function, or by consulting a
-standard reference\mmnote{maybe van Kampen, check},
-from which we find that
+which is exactly the negative binomial's generating function. Now we may recover
+the mRNA probability distribution of interest by differentiating the generating
+function, or by consulting a standard reference\mmnote{maybe van Kampen, check},
+from which we find that\mrm{I have a series of notes where I go through some 
+of the derivatives to find the general pattern, arriving to this exact result 
+by ``scratch and sniff'' without the black magic. Should I add it to this 
+section?}
 \begin{equation}
@@
 \end{equation}
-Note that the ratio of gamma functions is often expressed as a
-binomial coefficient, but since $\lambda$ may be non-integer,
-this would be ill-defined.
-Re-expressing this exclusively in our variables of interest,
-burst rate $\lambda$ and mean burst size $b$, we have
+Note that the ratio of gamma functions is often expressed as a binomial
+coefficient, but since $\lambda$ may be non-integer, this would be ill-defined.
+Re-expressing this exclusively in our variables of interest, burst rate
+$\lambda$ and mean burst size $b$, we have
 \begin{equation}

### 2020-05-26 doc/abstract.tex
RP work on May 26
I worked on the Abstract, the Intro, the Discussion and put in a label in Section 4 so when it was called the section got the right number.

@@
 variance and full gene expression distribution, illustrating the extent to which
-such models can be distinguished.
+such models can be distinguished, and suggesting a two-state model with a distribution
+of burst sizes as the most potent of the seven for describing the simple-repression motif.
 \end{abstract}

### 2020-05-26 doc/section_01_introduction.tex
RP work on May 26
I worked on the Abstract, the Intro, the Discussion and put in a label in Section 4 so when it was called the section got the right number.

@@
 biochemists.
-\mrm{Our aim is to understand the underlying principles of such coarse-graining
+Our aim is to understand the underlying principles of such coarse-graining
 in order to increase the predictive power of these models. In doing so we hope
@@
 predictive role needed to take synthetic biology from a brilliant exercise in
-enlightened empiricism to a rational design framework as any other branch of
-engineering.}
+enlightened empiricism to a rational design framework as  in any other branch of
+engineering.
 % Our aim is to understand the underlying principles of such
@@
 expression as shown in Figure~\ref{fig1:means_cartoons}(D).
-Section~\ref{section_03_beyond_means.tex} goes beyond an analysis of the mean
+Section~\ref{sec:beyond_means} goes beyond an analysis of the mean
 gene expression by asking how the same models presented in
@@
 commitment to some numerical values for the key parameters found in each such
-model. Therefore in Section~\ref{section_04_bayesian_inference.tex} we explore
+model. Therefore in Section~\ref{section_04_bayesian_inference} we explore
 the use of Bayesian inference to establish these parameters and to rigorously

### 2020-05-26 doc/section_04_bayesian_inference.tex
RP work on May 26
I worked on the Abstract, the Intro, the Discussion and put in a label in Section 4 so when it was called the section got the right number.

@@
 selection}
+\label{section_04_bayesian_inference}
 \subsection{Parameter inference for constitutive promoters}
@@
 sense of the uncertainty and correlations in the posterior.
-
-\mmnote{Outline:
-\begin{itemize}
-\item Hey look it mostly works. With full model, 9D model is well identified,
-though most of the individual experiments unsurprisingly are not:
-only 3 of the 9 datasets were identifiable if fit in isolation
-and for the rest only the ratio of repressor rates is identifiable.
-Only by making assumptions that certain rates
-are equal across datasets was identifiability achieved.
-PPC is not perfect, it kinda misses some
-rep/op pairs but overall it's surprisingly predictive considering how strong
-were the assumptions that went into it.
-\item Do we chalk this up to experimental imperfections or is there interesting
-biology to be found in the disagreements? I'm inclined towards the former. Even
-just catching cells in truly apples-to-apples growth phases, or matching up
-parameters of imaging sessions weeks or months apart, is hardly trivial.
-\item The alternative is to refine our model and/or add on complexity, and it's
-not at all clear what additions to make to the model, especially how to produce
-the long tail of repressed strains over UV5. I see no way to get that without
-getting into complicated time-correlation/history effects, where
-transcription...rebounds?? After being repressed?? Sounds odd.
-\item Comparison between my inferred rates, equilibrium binding E, and
-single-molecule measurements. Not quite as slam-dunk as I'd hoped because of
-$\gamma$ dependence, but rates are w/in factor of 2 or better, seems quite good
-really! With exception of one Oid data data point, free energies are w/in about
-0.5 kT also. That's not massively larger than the repeatability threshold of
-fold-change measurements, at maybe 0.1 to 0.3 kT.
-\item Surprising that so many model parameters can be inferred from just a few
-distributions that, by eye, don't have strong features! Does require some
-assumptions, e.g., about equivalence of rates across experiments, but that seems
-quite reasonable.
-\end{itemize}
-}

### 2020-05-26 doc/section_05_discussion.tex
RP work on May 26
I worked on the Abstract, the Intro, the Discussion and put in a label in Section 4 so when it was called the section got the right number.

@@
 any and all experimental data that we are aware of.
-\mrm{
 This theoretical result allows us to directly discard a subset of the models
-(models 1-3 in Figure~\ref{fig2:fig2:constit_cartoons}) since they cannot be
-reconciled with experimental observations. The two models left (models 4 and 5
-in Figure~\ref{fig2:fig2:constit_cartoons}) seem to provide a rich enough
-phenomenology to be able to reproduce the data. A previous exploration of one 
-of these models demonstrated the sloppy nature of the model in which single-cell
-mRNA counts data alone cannot constrain the value of all parameters 
+(models 1-3 in Figure~\ref{fig2:constit_cartoons}) since they cannot be
+reconciled with experimental observations. The two remaining models (models 4 and 5
+in Figure~\ref{fig2:constit_cartoons}) appear to contain enough microscopic
+realistm to be able to reproduce the data. A previous exploration of one 
+of these models demonstrated the sloppy nature of the model in which data on single-cell
+mRNA counts alone cannot constrain the value of all parameters 
 simultaneously \cite{Razo-Mejia2020}. Here we demonstrate that the proposed
 one-state bursty promoter model (model 5 in 
-Figure~\ref{fig2:fig2:constit_cartoons}) is a limit of the commonly used
-two-state promoter model \cite{Peccoud1995}. We put at test the idea that this
-level of coarse-graining is descriptive enough to reproduce previous 
-experimental observations. In particular we perform Bayesian inference on the
+Figure~\ref{fig2:constit_cartoons}) emerges as  a limit of the commonly used
+two-state promoter model \cite{Peccoud1995}. We put  the idea to the test that this
+level of coarse-graining is rich enough to reproduce previous 
+experimental observations. In particular we perform Bayesian inference to determine the
 two parameters describing the full steady-state mRNA distribution, finding that
-the model is able to capture a plethora of promoter sequences with different
+the model is able to provide a quantitative description of  a plethora of promoter sequences with different
 mean levels of expression and noise.
-}
 
 \mrm{
-Here we should include a discussion of the regulated case.
+Here we should include a discussion of the regulated case.  RP: I made a first
+brief stab at it.
 }
+With the results of the constitutive promoter in hand, we then fix the parameters
+associated with this class of promoters and use them as input for evaluating
+the noise in gene expression for the simple repression motif itself.  This analysis
+allows us to provide a single overarching analysis of both the constitutive
+and simple repression architectures using one simple model and
+corresponding set of self-consistent parameters, demonstrating not only
+a predictive framework, but also reconciling the equilibrium and non-equilibrium
+views of the same simple repression constructs.
 
@@
 not disturb the cell's physiology too badly.
+
+\mmnote{Outline:
+\begin{itemize}
+\item Hey look it mostly works. With full model, 9D model is well identified,
+though most of the individual experiments unsurprisingly are not:
+only 3 of the 9 datasets were identifiable if fit in isolation
+and for the rest only the ratio of repressor rates is identifiable.
+Only by making assumptions that certain rates
+are equal across datasets was identifiability achieved.
+PPC is not perfect, it kinda misses some
+rep/op pairs but overall it's surprisingly predictive considering how strong
+were the assumptions that went into it.
+\item Do we chalk this up to experimental imperfections or is there interesting
+biology to be found in the disagreements? I'm inclined towards the former. Even
+just catching cells in truly apples-to-apples growth phases, or matching up
+parameters of imaging sessions weeks or months apart, is hardly trivial.
+\item The alternative is to refine our model and/or add on complexity, and it's
+not at all clear what additions to make to the model, especially how to produce
+the long tail of repressed strains over UV5. I see no way to get that without
+getting into complicated time-correlation/history effects, where
+transcription...rebounds?? After being repressed?? Sounds odd.
+\item Comparison between my inferred rates, equilibrium binding E, and
+single-molecule measurements. Not quite as slam-dunk as I'd hoped because of
+$\gamma$ dependence, but rates are w/in factor of 2 or better, seems quite good
+really! With exception of one Oid data data point, free energies are w/in about
+0.5 kT also. That's not massively larger than the repeatability threshold of
+fold-change measurements, at maybe 0.1 to 0.3 kT.
+\item Surprising that so many model parameters can be inferred from just a few
+distributions that, by eye, don't have strong features! Does require some
+assumptions, e.g., about equivalence of rates across experiments, but that seems
+quite reasonable.
+\end{itemize}
+}

### 2020-05-26 doc/section_05_discussion.tex
fix definition of fano factor

@@
 a treatment of the constitutive promoter already reveals that some models have
-Fano factors (standard deviation/mean) that are less than one, at odds with
+Fano factors (variance/mean) that are less than one, at odds with
 any and all experimental data that we are aware of.

### 2020-05-26 doc/section_03_beyond_means.tex
MRM pass through beyond means on 20200526

@@
 
-In this section, our objective is to explore the same models considered in
-the previous section, but now with reference to the the question of
-how well they describe the distribution of gene expression levels, with
-special reference to the variance in these distributions.  To that end,
-we repeat the same pattern as in the previous section by examining
-the models one by one.
+In this section, our objective is to explore the same models considered in the
+previous section, but now with reference to the the question of how well they
+describe the distribution of gene expression levels, with special reference to
+the variance in these distributions. To that end, we repeat the same pattern as
+in the previous section by examining the models one by one. \mrm{In particular
+we will focus on the Fano factor, defined as the variance/mean. This metric
+serves as a discriminatory tool from the null model that the steady-state mRNA
+distribution must be Poisson, giving a Fano factor of one.}
 
-\subsection{A one-state promoter with bursting is the ``best'' model of constitutive promoters}
+\subsection{A one-state promoter with bursting is the ``best'' model of
+constitutive promoters}
+
 Before we can tackle simple repression, we need an adequate phenomenological
@@
 we can choose, and we show several potential kinetic models for constitutive
-promoters in~\fig{fig2:constit_cartoons}. Let us consider the suitability of each
-model for our purposes in turn.
+promoters in~\fig{fig2:constit_cartoons}. Let us consider the suitability of
+each model for our purposes in turn.
 
@@
 \caption{\textbf{Comparison of different models for noise in the constitutive promoter.}
-The left column depicts various plausible models for the dynamics of constitutive promoters.
-%We also list literature references where models have been used previously. 
-In
-model (1), transcripts are produced in a Poisson process~\cite{Jones2014}. Model
-(2) features explicit modeling of RNAP binding/unbinding
-kinetics~\cite{Phillips2015a}. Model (3) is a more detailed generalization of
-model (2), treating transcription initiation as a multi-step process proceeding
-through closed and open complexes~\cite{Mitarai2015}. Model (4) is somewhat
-analogous to (2) except with the precise nature of active and inactive states
-left ambiguous~\cite{Razo-Mejia2020}. Finally, model (5) can be viewed as a
-certain limit of model (4) in which transcripts are produced in bursts, and
-initiation of bursts is a Poisson process. 
-%\mmnote{Haven't found model 5
-%formulated quite like this in the literature, usually burstiness is handled like
-%model 4, but I should probably search a bit more.} 
-The right column shows the
-Fano factor for each model. Note especially the crucial diagnostic: (2) and (3)
-have $\nu$ strictly below 1, while only for (4) and (5) can $\nu$ exceed 1.
-Models with Fano factors $\le 1$ cannot produce the FISH data observed
-in~\cite{Jones2014} without introducing additional assumptions and model
-complexity.}
+The left column depicts various plausible models for the dynamics of
+constitutive promoters. In model (1), transcripts are produced in a Poisson
+process~\cite{Sanchez2013, Jones2014}. Model (2) features explicit modeling of
+RNAP binding/unbinding kinetics~\cite{Phillips2015a}. Model (3) is a more
+detailed generalization of model (2), treating transcription initiation as a
+multi-step process proceeding through closed and open
+complexes~\cite{Mitarai2015}. Model (4) is somewhat analogous to (2) except with
+the precise nature of active and inactive states left
+ambiguous~\cite{Peccoud1995, Shahrezaei2008, Razo-Mejia2020}. Finally, model (5)
+can be viewed as a certain limit of model (4) in which transcripts are produced
+in bursts, and initiation of bursts is a Poisson process. 
+% \mmnote{Haven't found model 5 formulated quite like this in the literature,
+% usually burstiness is handled like model 4, but I should probably search a bit
+% more.} 
+The right column shows the Fano factor $\nu$ (variance/mean) for each model.
+Note especially the crucial diagnostic: (2) and (3) have $\nu$ strictly below 1,
+while only for (4) and (5) can $\nu$ exceed 1. Models with Fano factors $\le 1$
+cannot produce the FISH data observed in~\cite{Jones2014} without introducing
+additional assumptions and model complexity.}
 \label{fig2:constit_cartoons}
@@
 \subsubsection{Noise in the Poisson Promoter Model}
-The simplest model of constitutive expression that we can imagine
-is shown as model 1 in Figure~\ref{fig2:constit_cartoons} and
-assumes that
-transcripts are produced as a Poisson process from a single promoter state. 
-This is the picture from Jones et.\ al.~\cite{Jones2014} that was used to
-interpret a systematic study of gene expression noise over a series of
-promoters designed to have different strengths.   This
-model insists that the ``true'' mRNA distribution is Poisson, implying the Fano
-factor $\nu$ must be 1. In~\cite{Jones2014}, the authors carefully attribute
-deviations from Fano = 1 to intensity variability in fluorescent spots, gene
-copy number variation, and copy number fluctuations of the transcription
-machinery, e.g., RNAP itself. In this picture, the master equation makes no
-appearance, and all the corrections to Poisson behavior are derived as additive
-corrections to the Fano factor. For disproving the ``universal noise curve''
-from So et al~\cite{So2011}, this picture was excellent. It is appealing in its
-simplicity, with only two parameters, the initiation rate $r$ and degradation
-rate $\gamma$. Since $\gamma$ is independently known from other experiments, and
-the mean mRNA copy number is $r/\gamma$, $r$ is easily inferred from data. In
-other words, the model is not excessively complex for the data at hand. But for
-many interesting questions, for instance in the recent work~\mmnote{cite
-Manuel's preprint once posted}, knowledge of means and variances alone is
-insufficient, and a description of the full distribution of molecular counts is
-necessary. For this we need a (slightly) more complex model than model 1 that
-would allow us to incorporate the non-Poissonian features of constitutive
-promoters directly into a master equation formulation.
+
+The simplest model of constitutive expression that we can imagine is shown as
+model 1 in Figure~\ref{fig2:constit_cartoons} and assumes that transcripts are
+produced as a Poisson process from a single promoter state. This is the picture
+from Jones et.\ al.~\cite{Jones2014} that was used to interpret a systematic
+study of gene expression noise over a series of promoters designed to have
+different strengths. This model insists that the ``true'' mRNA distribution is
+Poisson, implying the Fano factor $\nu$ must be 1. In~\cite{Jones2014}, the
+authors carefully attribute measured deviations from Fano = 1 to intensity
+variability in fluorescent spots, gene copy number variation, and copy number
+fluctuations of the transcription machinery, e.g., RNAP itself. In this picture,
+the master equation makes no appearance, and all the corrections to Poisson
+behavior are derived as additive corrections to the Fano factor. For disproving
+the ``universal noise curve'' from So et al~\cite{So2011}, this picture was
+excellent. It is appealing in its simplicity, with only two parameters, the
+initiation rate $r$ and degradation rate $\gamma$. Since $\gamma$ is
+independently known from other experiments, and the mean mRNA copy number is
+$r/\gamma$, $r$ is easily inferred from data. In other words, the model is not
+excessively complex for the data at hand. But for many interesting questions,
+for instance in the recent work~\cite{Razo-Mejia2020}, knowledge of means and
+variances alone is insufficient, and a description of the full distribution of
+molecular counts is necessary. For this we need a (slightly) more complex model
+than model 1 that would allow us to incorporate the non-Poissonian features of
+constitutive promoters directly into a master equation formulation.
 
 \subsubsection{Noise in the Two-State Promoter, RNAP Bound or Unbound.}
-Our second model of constitutive transcription posits an architecture
-in which the promoter is either empty or bound by RNAP
-~\cite{Phillips2015a, Phillips2019}. Here,  as shown in 
-the second row of Figure~\ref{fig2:constit_cartoons},  transcription initiation results in
-a state transition from the bound to
-the unbound state, reflecting the microscopic reality that an RNAP that has
-begun to elongate a transcript is no longer available at the start site to begin
-another.  As shown in the appendix, the Fano factor in this model is given by
+
+Our second model of constitutive transcription posits an architecture in which
+the promoter is either empty or bound by RNAP ~\cite{Phillips2015a,
+Phillips2019}. Here,  as shown in model 2 of Figure~\ref{fig2:constit_cartoons},
+transcription initiation results in a state transition from the bound to the
+unbound state, reflecting the microscopic reality that an RNAP that has begun to
+elongate a transcript is no longer available at the start site to begin another.
+As shown in Appendix~\ref{sec:unreg_promoters}, the Fano factor in this model is
+given by
 \begin{align}
@@
 \end{align}
-The problem with this picture is that the Fano factor is
- always $<1$. 
- To make contact with the experimental reality of $\nu>1$,
-clearly some corrections will be needed.
-While this model adds an appealing element of microscopic
-reality, we are forced to reject it as the additional complexity is unable to
-capture the phenomenology of interest. Obviously the promoter state does in fact
-proceed through cycles of RNAP binding, initiating, and elongating, but it seems
-that the super-Poissonian noise in mRNA copy number we want to model must be
-governed by other features of the system.
+The problem with this picture is that the Fano factor is always $<1$. To make
+ contact with the experimental reality of $\nu>1$, clearly some corrections will
+ be needed. While this model adds an appealing element of microscopic reality,
+ we are forced to reject it as the additional complexity is unable to capture
+ the phenomenology of interest. Obviously the promoter state does in fact
+ proceed through cycles of RNAP binding, initiating, and elongating, but it
+ seems that the super-Poissonian noise in mRNA copy number we want to model must
+ be governed by other features of the system.
 
 \subsubsection{Noise in the Three-State Promoter, Multistep Transcription
-Initiation and Escape.} How might we remedy the deficits of model 2? It is
-known~\cite{DeHaseth1998} that once RNAP initially binds the promoter region, a
-multitude of distinct steps occur sequentially before RNAP finally escapes into
-the elongation phase. Perhaps adding some of this mechanistic detail as shown
-in Figure~\ref{fig2:constit_cartoons} might
-rescue the previous model   The next simplest refinement of that model could consider open
-complex formation and promoter escape as separate steps rather than as a single
-effective step. In other words, we construct model 3 by adding a single extra
-state to model 2, and we will label the two RNAP-bound states as the closed and
-open complexes, despite the true biochemical details certainly being more
-complex.  For example, earlier work extended this model by adding an additional
-repressor bound state and did not explicitly consider the limit with no repressor
-that we analyze here~\cite{Mitarai2015}.  Again, our goal here
-is not a complete accounting of all the relevant biochemical detail; this is an
-exploratory search for the important features that a model needs to include
-to square with the known experimental reality of constitutive expression.
+Initiation and Escape.}
 
-Unfortunately, as hinted at in earlier work~\cite{Mitarai2015}, this model too has
-Fano factor $\nu<1$. This can be seen by starting with their Eq.~A1 for the Fano
-factor of the analogous model with repressor and taking the limit as repressor
-concentration goes to zero. After substantial algebra, the result is
+How might we remedy the deficits of model 2? It is known~\cite{DeHaseth1998}
+that once RNAP initially binds the promoter region, a multitude of distinct
+steps occur sequentially before RNAP finally escapes into the elongation phase.
+Perhaps adding some of this mechanistic detail as shown in model 3 of
+Figure~\ref{fig2:constit_cartoons} might rescue the previous model. The next
+simplest refinement of that model could consider open complex formation and
+promoter escape as separate steps rather than as a single effective step. In
+other words, we construct model 3 by adding a single extra state to model 2, and
+we will label the two RNAP-bound states as the closed and open complexes,
+despite the true biochemical details certainly being more complex. For example,
+earlier work extended this model by adding an additional repressor bound state
+and did not explicitly consider the limit with no repressor that we analyze
+here~\cite{Mitarai2015}. Again, our goal here is not a complete accounting of
+all the relevant biochemical detail; this is an exploratory search for the
+important features that a model needs to include to square with the known
+experimental reality of constitutive expression.
+
+Unfortunately, as hinted at in earlier work~\cite{Mitarai2015}, this model too
+has Fano factor $\nu<1$. This can be seen by starting with their Eq.~A1 for the
+Fano factor of the analogous model with repressor and taking the limit as
+repressor concentration goes to zero. After substantial algebra, the result is
 \begin{align}
@@
 and intuitively should be more peaked with a smaller fractional width than an
-exponential with the same mean.\footnote{This can be made more precise with a
-result from~\cite{Stewart2007}, who showed that the convolution of multiple
-gamma distributions (of which the exponential distribution is a special case) is
-to a very good approximation also gamma distributed. Using their Eq.~2 for the
-distribution of the convolution, with shape parameters set to 1 to give
-exponential distributions, the total waiting time distribution has a ratio of
-variance to squared mean $\sigma^2/\mu^2 = \sum_i k_i^2/\left(\sum_i
-k_i\right)^2$, where the $k_i$ are the rates of the individual steps. Clearly
-this is less than 1 and therefore the total waiting time distribution is
-narrower than the corresponding exponential.} A less disperse waiting time
-distribution means transcription initations are more uniformly distributed in
-time relative to a Poisson process. Hence the distribution of mRNA over a
-population of cells should be less variable compared to Poisson, giving
-$\nu<1$.\footnote{This last step, while intuitive, can be argued by analogy to
-photon statistics where antibunching gives rise to sub-Poissonian
-noise~\cite{Paul1982, Zou1990}. Although loopholes exist, we do not expect they
-apply for our problem. Nevertheless we refrain from elevating this
-cycle/sub-Poissonian equivalence to a ``theorem.'' \mmnote{Antibunching is the
-``obvious'' analogy for me but it'd be totally out of left field for most
-readers, any suggestions on a more comprehensible reference??}}
+exponential with the same mean.
+\footnote{This can be made more precise with a result from~\cite{Stewart2007},
+who showed that the convolution of multiple gamma distributions (of which the
+exponential distribution is a special case) is, to a very good approximation,
+also gamma distributed. Using their Eq.~2 for the distribution of the
+convolution, with shape parameters set to 1 to give exponential distributions,
+the total waiting time distribution has a ratio of variance to squared mean
+$\sigma^2/\mu^2 = \sum_i k_i^2/\left(\sum_i k_i\right)^2$, where the $k_i$ are
+the rates of the individual steps. Clearly this is less than 1 and therefore the
+total waiting time distribution is narrower than the corresponding exponential.}
+A less disperse waiting time distribution means transcription initations are
+more uniformly distributed in time relative to a Poisson process. Hence the
+distribution of mRNA over a population of cells should be less variable compared
+to Poisson, giving $\nu<1$.
+\footnote{This last step, while intuitive, can be argued by analogy to photon
+statistics where antibunching gives rise to sub-Poissonian noise~\cite{Paul1982,
+Zou1990}. Although loopholes exist, we do not expect they apply for our problem.
+Nevertheless we refrain from elevating this cycle/sub-Poissonian equivalence to
+a ``theorem.'' \mmnote{Antibunching is the ``obvious'' analogy for me but it'd
+be totally out of left field for most readers, any suggestions on a more
+comprehensible reference??}}
 
 Regardless of the merits of this model in describing the noise properties of
-constitutive transcription initiation, it ultimately fails the simplest quantitative
-feature of the data, namely that the Fano factor $> 1$ and hence  we must discard this
-mechanistic picture
- and search elsewhere.
+constitutive transcription initiation, it ultimately fails the simplest
+quantitative feature of the data, namely that the Fano factor $> 1$ and hence
+we must discard this mechanistic picture and search elsewhere.
 
@@
 States}
+
 Inspired by~\cite{Razo-Mejia2020}, we revisit an analog of model 2, but as with
- the models considered in the section on mean expression,
-the interpretation of the two states is changed. Rather than explicitly viewing
-them as RNAP bound and unbound, we view them as ``active'' and ``inactive,''
-which are able and unable to initiate transcripts, respectively. We are
-noncommittal as to the microscopic details of these states.
+the models considered in \ref{section_02_means}, the interpretation of the two
+states is changed. Rather than explicitly viewing them as RNAP bound and
+unbound, we view them as ``active'' and ``inactive,'' which are able and unable
+to initiate transcripts, respectively. We are noncommittal as to the microscopic
+details of these states.
 
@@
 promoter. Combined with known $K_d$'s of order $\mu$M, this gives an RNAP
-dissociation rate $k_{off}$ of order $10^2$~s$^{-1}$. As we will show below, however,
-there are some lingering puzzles with interpreting this supercoiling
-hypothesis, so we leave it as a speculation and refrain from assigning
-definite physical meaning to the two states in this model.
+dissociation rate $k_{off}$ of order $10^2$~s$^{-1}$. As we will show below,
+however, there are some lingering puzzles with interpreting this supercoiling
+hypothesis, so we leave it as a speculation and refrain from assigning definite
+physical meaning to the two states in this model.
 
@@
 process only when the promoter is in one of the two states in this model,
-transcription initiations should now be ``bunched,'' in constrast to the
+transcription initiations should now be ``bunched,'' in contrast to the
 ``anti-bunching'' of models 2 and 3 above. One might further guess that this
 bunching would lead to super-Poissonian noise in the mRNA distribution over a
-population of cells.  Indeed, as shown in the appendix, a calculation of the Fano factor produces
+population of cells.  Indeed, as shown in Appendix~\ref{sec:unreg_promoters}, a
+calculation of the Fano factor produces
 \begin{align}
@@
 be sidestepped by noting that the model is mathematically equivalent to the
-simple repression model from~\cite{Jones2014}, with states and rates
-relabeled and reinterpreted.
+simple repression model from~\cite{Jones2014}, with states and rates relabeled
+and reinterpreted.
 
@@
 inference problem is nonidentifiable, in the sense described in Section 4.3
-of~\cite{Gelman2013}, meaning it is impossible to infer the parameters $r$ and
-$k^-$ from the FISH data of~\cite{Jones2014} (as shown in Fig.~S2
-of~\cite{Razo-Mejia2020}). Rather, only the ratio $r/k^-$ could be inferred. In
-that work, the problem was worked around with an informative prior on the ratio
-$k^-/k^+$. That approach is unlikely to work here, as, recall, our entire goal
-in modeling constitutive expression is to use it as the basis for a yet more
-complicated model, when we add on repression. But adding more complexity to a
-model that is already poorly identified is a fool's errand, so we will explore
-one more potential model.
+of~\cite{Gelman2013}. What this means is that it is impossible to infer the
+parameters $r$ and $k^-$ from the FISH data of~\cite{Jones2014} (as shown in
+Fig.~S2 of~\cite{Razo-Mejia2020}). Rather, only the ratio $r/k^-$ could be
+inferred. In that work, the problem was worked around with an informative prior
+on the ratio $k^-/k^+$. That approach is unlikely to work here, as, recall, our
+entire goal in modeling constitutive expression is to use it as the basis for a
+yet more complicated model, when we add on repression. But adding more
+complexity to a model that is already poorly identified is a fool's errand, so
+we will explore one more potential model.
 
 \subsubsection{Noise Model for One-State Promoter with Explicit Bursts}
+
 The final model we consider is 
@@
 bursts is approximately exponentially distributed, with mean set by the only
-timescale left in the problem, $1/k^+$.\footnote{If instead it were the case
-that $k^+ \sim k^-$, then the waiting time $1/k^+ + 1/k^-$ would have a peaked
-distribution, but this does not appear to be the case for any of datasets
-from~\cite{Jones2014}.}
+timescale left in the problem, $1/k^+$.
+\footnote{If instead it were the case that $k^+ \sim k^-$, then the waiting time
+$1/k^+ + 1/k^-$ would have a peaked distribution, but this does not appear to be
+the case for any of datasets from~\cite{Jones2014}.}
 
@@
 at rate $k^+$ (transitions to the active state, in the model 4 picture). The
-master equation for mRNA copy number $m$ takes the form
+master equation for mRNA copy number $m$ as derived in
+Appendix~\ref{sec:unreg_promoters} takes the form
 \begin{align}
 \begin{split}
-\deriv{t}{p(m,t)} = & (m+1)\gamma p(m+1,t) - m\gamma p(m,t) \\
-        &+ \sum_{j=0}^{m-1} k^+ p(j,t) Geom(m-j;b)
-         - \sum_{j=m+1}^\infty k^+ p(m,t) Geom(j-m;b),
+{d \over dt}p(m,t) = & (m+1)\gamma p(m+1,t) - m\gamma p(m,t) \\
+        &+ \sum_{m'=0}^{m-1} k_i p(m',t) \text{Geom}(m-m';b)
+         - \sum_{m'=m+1}^\infty k_i p(m,t) \text{Geom}(m'-m;b),
 \end{split}
 \end{align}
-where $Geom(n;b)$ is the geometric distribution with mean~$b$, i.e., $Geom(n;b)
-= \frac{1}{1+b}\left(\frac{b}{1+b}\right)^n$ (with domain over nonnegative
+\mrm{notice I changed $k^+$ to $k_i$ to be consistent with the figure.}
+where we use $k_i$ to denote the burst initiation rate, $\text{Geom}(n;b)$ is the
+geometric distribution with mean~$b$, i.e., $\text{Geom}(n;b) =
+\frac{1}{1+b}\left(\frac{b}{1+b}\right)^n$ (with domain over nonnegative
 integers as above). The first two terms are the usual mRNA degradation terms.
 The third term enumerates all ways the system can produce a burst of transcripts
-and arrive at copy number $m$, given that it had copy number $j$ before the
+and arrive at copy number $m$, given that it had copy number $m'$ before the
 burst. The fourth term allows the system to start with copy number $m$, then
-produce a burst and end with copy number $j$. In fact this last sum has trivial
-$j$ dependence and simply enforces normalization of the geometric distribution.
+produce a burst and end with copy number $m'$. In fact this last sum has trivial
+$m'$ dependence and simply enforces normalization of the geometric distribution.
 Carrying it out we have
@@
 \begin{split}
-\deriv{t}{p(m,t)} = & (m+1)\gamma p(m+1,t) - m\gamma p(m,t) \\
-        &+ \sum_{j=0}^{m-1} k^+ p(j,t) Geom(m-j;b)
-            - k^+ p(m,t),
+{d \over dt} p(m,t) = & (m+1)\gamma p(m+1,t) - m\gamma p(m,t) \\
+        &+ \sum_{m'=0}^{m-1} k_i p(m',t) \text{Geom}(m-m';b)
+            - k_i p(m,t),
 \end{split}
 \end{equation}
+We direct readers again to Appendix~\ref{sec:beyond_means} for further details.
 This improves on model 4 in that now the parameters are easily inferred, as we
 will see later, and have clean interpretations. The non-Poissonian features are
-attributed to the emprically well-established phenomenological picture of bursty
-transcription.
+attributed to the empirically well-established phenomenological picture of
+bursty transcription.
 
@@
 not short compared to transcription factor kinetic timescales, this could be a
-problem in that mean burst size in the presense and absence of repressors could
+problem in that mean burst size in the presence and absence of repressors could
 change, rendering parameter inferences from the constitutive case inappropriate.
@@
 10 (or less) would have a diffusion-limited association rate of order
-$(10~\text{sec})^{-1}$. Higher copy number TFs tend to have many binding sites
-over the genome, which should serve to pull them out of circulation and keep
-their effective association rates from rising too large. Therefore, there is
-\textit{perhaps} a timescale separation possible between transcription factor
-association rates and burst durations, but this assumption could very well break
-down, so we will have to keep it in mind when we infer repressor rates from the
-Jones et.\ al.\ FISH data later.
+$(10~\text{sec})^{-1}$ \mrm{cite Elf?}. Higher copy number TFs tend to have many
+binding sites over the genome, which should serve to pull them out of
+circulation and keep their effective association rates from rising too large.
+Therefore, there is \textit{perhaps} a timescale separation possible between
+transcription factor association rates and burst durations, but this assumption
+could very well break down, so we will have to keep it in mind when we infer
+repressor rates from the Jones et.\ al.\ FISH data later~\cite{Jones2014}.
 

### 2020-05-27 code/figures/main/fig02B.py
fig02B script

@@ -1,0 +1,136 @@
+# %%
+import pickle
+from git import Repo #for directory convenience
+
+import numpy as np
+import scipy.stats as st
+import pandas as pd
+import arviz as az
+
+import matplotlib.pyplot as plt
+import matplotlib
+import seaborn as sns
+import bebi103.viz
+
+import srep.viz
+from srep.utils import load_FISH_by_promoter
+
+# Set PBoC plotting style
+srep.viz.plotting_style()
+# %%
+repo = Repo("./", search_parent_directories=True)
+# repo_rootdir holds the absolute path to the top-level of our repo
+repo_rootdir = repo.working_tree_dir
+
+# Import sm-FISH data along with promoter information
+df_unreg, df_energies, = load_FISH_by_promoter(("unreg", "energies"))
+df_energies.sort_values('Energy (kT)', inplace=True)
+
+# %%
+# Group FISH data by promoters
+df_group = df_unreg.groupby("experiment")
+
+# Initialize dataframe to save mean expression and Fano factor
+names = ["promoter", "mean", "var", "fano", "energy_kT"]
+df_summary = pd.DataFrame([], columns=names)
+
+# Loop through promoters computing mean expression and Fano
+for prom, data in df_group:
+    # Compute mean expression
+    mean_m = data["mRNA_cell"].mean()
+    # Compute variance
+    var_m = data["mRNA_cell"].var()
+    # Compute Fano factor
+    fano_m = var_m / mean_m
+    # Extract binding energy
+    energy = df_energies[df_energies["Name"] == prom]["Energy (kT)"].values[0]
+    # Append to dataframe
+    df_summary = df_summary.append(
+        pd.Series([prom, mean_m, var_m, fano_m, energy], index=names),
+        ignore_index=True
+    )
+# Sort dataframe by energy
+df_summary.sort_values("energy_kT", inplace=True)
+
+# %%
+# Set function to normalize colors to energy range
+col_norm = matplotlib.colors.Normalize(
+    vmin=df_energies["Energy (kT)"].min() - 2,
+    vmax=df_energies["Energy (kT)"].max() + 2,
+)
+
+# Initialize figure
+fig, ax = plt.subplots(1, 1, figsize=(6, 2))
+
[…]

### 2020-05-27 doc/section_03_beyond_means.tex
update caption

@@
 \caption{\textbf{Comparison of different models for noise in the constitutive promoter.}
-The left column depicts various plausible models for the dynamics of
+(A) The left column depicts various plausible models for the dynamics of
 constitutive promoters. In model (1), transcripts are produced in a Poisson
@@
 while only for (4) and (5) can $\nu$ exceed 1. Models with Fano factors $\le 1$
-cannot produce the FISH data observed in~\cite{Jones2014} without introducing
-additional assumptions and model complexity.}
+cannot produce the FISH data observed in part (B) without introducing
+additional assumptions and model complexity. (B) Data from \cite{Jones2014}.
+Mean mRNA count vs. Fano factor (variance/mean) for different promoters as
+determined with single-molecule mRNA FISH. The colorbar indicates the predicted
+binding affinity of RNAP to the promoter sequence as determined in
+\cite{Brewster2012}. Numbers serve for cross comparison with data presented in 
+Figure 3.}
 \label{fig2:constit_cartoons}

### 2020-05-27 doc/section_04_bayesian_inference.tex
RP on inference
Made it up to and through Figure 3

@@
 the observed data has Fano factor $\nu>1$, we might already suspect that this
-model is incapable of fitting the data. We will verify that this in fact the
+model is incapable of fitting the data. We will verify that this is in fact the
 case. Using the same argument we can immediately rule out models 2 and 3 from 
-Figure~\ref{fig2:constit_cartoons}. these models have Fano factors $\nu\le 1$
-meaning they are underdisperse relative to Poisson. We will also not explicitly
+Figure~\ref{fig2:constit_cartoons}.  These models have Fano factors $\nu\le 1$
+meaning they are underdispersed relative to the Poisson distribution. We will also not explicitly
 consider model 4 from~\fig{fig2:constit_cartoons} since it was already thoroughly
@@
 
-\mrm{
 We alert the reader not to confuse a tightly constrained posterior distribution
@@
 check the validity of the generative model we can make use of \textit{posterior
-predictive checks}. These consists on generating synthetic data using our
+predictive checks}. This consists in generating synthetic data using our
 generative model for many plausible values of our parameter. For the case of our
@@
 the posterior distribution (Eq.~\ref{eq:posterior_poisson}), and generate 2648
-samples out of a Poisson distribution with such $\lambda$ value. We then repeat
+samples drawn from a Poisson distribution with  $\lambda$ chosen appropriately. We then repeat
 this procedure 1000 times. Clearly if those data appear quite different, the
@@
 true, then the synthetic datasets we generate should resemble the actual
-observed data, and if not, it suggests the model is missing important features.}
+observed data, and if not, it suggests the model is missing important features.
 
-\mrm{
 To compare so many samples with the actual observed data, we can use empirical
@@
 Figure~\ref{fig:constit_post_full}(B) next.
-}
 
@@
 increased from one to two.
-\mrm{
 Details on the arguments behind our prior distribution selection are left for
@@
 knowledge based on our domain expertise spans several orders of magnitude.
-}
 

### 2020-05-27 code/figures/main/fig03.py
change label color to gray to increase visibility

@@ -281,6 +281,7 @@ for i, promoter in enumerate(df_energies.Name):
         (df_energies[df_energies.Name == promoter]["Energy (kT)"],
         ptile_med,),
         fontsize=7,
+        color="gray",
     )
 
 # Add colorbar
@@ -303,16 +304,7 @@ cbar.set_label(r"$\Delta\epsilon_r \; (k_BT)$")
 guide_x = np.linspace(-7,-3.5)
 guide_y = np.exp(-guide_x)/9e2
 ax_d.plot(guide_x, guide_y, 'k--', label='predicted \n scaling')
-# Add text for scaling
-# Muir's location
-# ax_d.text(
-#     0.29, 
-#     0.04, 
-#     r"$\log k_i \sim - \Delta\epsilon_P$",
-#     transform=ax_d.transAxes,
-#     rotation=-45,
-# )
-# Manuel's location
+
 ax_d.text(
     0.29, 
     0.25, 

### 2020-05-27 doc/section_04_bayesian_inference.tex
MRM rewrote Bayesian inference (unreg promoter) with more transparent language

@@
 From consideration of Fano factors in the previous section, we suspect that
-model 5 in Figure~\ref{fig2:constit_cartoons}, a one-state bursty model of
+model 5 in Figure~\ref{fig2:constit_cartoons}(A), a one-state bursty model of
 constitutive promoters, achieves a Goldilocks level of complexity. Does this
@@
 Poisson distribution has a Fano factor $\nu$ strictly equal to 1, and all of
-the observed data has Fano factor $\nu>1$, we might already suspect that this
-model is incapable of fitting the data. We will verify that this is in fact the
-case. Using the same argument we can immediately rule out models 2 and 3 from 
-Figure~\ref{fig2:constit_cartoons}.  These models have Fano factors $\nu\le 1$
-meaning they are underdispersed relative to the Poisson distribution. We will also not explicitly
-consider model 4 from~\fig{fig2:constit_cartoons} since it was already thoroughly
-analyzed in~\cite{Razo-Mejia2020}, and since model 5 can be viewed as a special
-case of it.
+the observed data in Figure~\ref{fig2:constit_cartoons}(B) has Fano factor
+$\nu>1$, we might already suspect that this model is incapable of fitting the
+data. We will verify that this is in fact the case. Using the same argument we
+can immediately rule out models 2 and 3 from
+Figure~\ref{fig2:constit_cartoons}(A). These models have Fano factors $\nu\le 1$
+meaning they are underdispersed relative to the Poisson distribution. We will
+also not explicitly consider model 4 from~\fig{fig2:constit_cartoons}(A) since
+it was already thoroughly analyzed in~\cite{Razo-Mejia2020}, and since model 5
+can be viewed as a special case of it.
 
-In performing parameter inference on the single-molecule data
-of~\cite{Jones2014}, building a generative model is straightforward. The data
-are single-cell mRNA counts, and we assume different cells to be independent and
-identically distributed from the underlying generative model. The generative
-model for our case is the steady-state mRNA distribution computed from the
-master equation under consideration.
+Our objective for this section will then be to assess whether or not model 5 is
+quantitatively able to reproduce experimental data. In other words, if our claim
+is that the level of coarse graining in this model is capable of capturing the
+relevant features of the data, then we should be able to find values for the
+model parameters that can match theoretical predictions with single-molecule
+mRNA count distributions. A natural language for this parameter inference
+problem is that of Bayesian probability. We will then build a Bayesian inference
+pipeline to fit the model parameters to data. To gain intuition on how this
+analysis is done we will begin with the ``wrong'' model 1 in
+Figure~\ref{fig2:constit_cartoons}(A). We will use the full dataset of
+single-cell mRNA counts from~\cite{Jones2014} used in
+Figure~\ref{fig2:constit_cartoons}(B).
 
 \subsubsection{Model 1: Poisson promoter}
-% \mmnote{A lot of the details I've written here might be excessive for main text and serve better as an intro-by-example to Bayes in the appendix, then reduce main text to a bit more bare bones?}
+
 For this model the master equation of interest is
@@
 \begin{equation}
-\deriv{t}p_{m,U}(t) = rp_{m-1,U}(t) - rp_{m,U}(t)
-        + (m+1)\gamma p_{m+1,U}(t) - \gamma p_{m,U}(t),
+{d\over dt}p_{m,U}(t) = 
+        rp_{m-1,U}(t) 
+        - rp_{m,U}(t)
+        + (m+1)\gamma p_{m+1,U}(t) 
+        - \gamma p_{m,U}(t),
 \end{equation}
 whose steady-state solution is given by a Poisson distribution with parameter
-$\lambda \equiv r / \gamma$. This steady state distribution defines the
-likelihood $p(m\mid\lambda)$ for our Bayesian inference, i.e. 
+$\lambda \equiv r / \gamma$~\cite{Sanchez2013}. The goal of our inference 
+problem is then to find the probability distribution for the parameter value
+$\lambda$ given the experimental data. By Bayes theorem this can be written as
 \begin{equation}
-p(m\mid\lambda) = \frac{\lambda^m e^{-\lambda}}{m!}
+p(\lambda \mid D) = {p(D \mid \lambda) p(\lambda) \over p(D)},
+\end{equation}
+where $D = \{m_1, m_2, \ldots, m_N \}$ are the single-cell mRNA experimental
+counts. As is standard we will neglected the denominator $p(D)$ on the right
+hand side since it is independent of $\lambda$ and serves only as a
+normalization factor.
+
+The steady-state solution for the master equation defines the likelihood
+term for a single cell $p(m \mid \lambda)$. What this means is that for a given
+choice of parameter $\lambda$, under model 1 of
+Figure~\ref{fig2:constit_cartoons}(A), we expect to observe $m$ mRNAs in a
+single cell with probability
+\begin{equation}
+p(m\mid\lambda) = \frac{\lambda^m e^{-\lambda}}{m!}.
 \label{eq:poisson_inference010}
 \end{equation}
-is the probability of observing a single cell with $m$ mRNAs given a value of
-$\lambda$, written to make the conditioning explicit.
-
-With our likelihood in hand, we can invoke Bayes theorem to write down the
-posterior distribution on our model parameter $\lambda$, given by
-\begin{equation}
-p(\lambda\mid D) \propto p(D\mid\lambda) p(\lambda),
-\end{equation}
-where our data $D=\{m_1, m_2,\dots, m_N\}$ is a list of mRNA counts over a
-population of $N$ identical cells. As is standard we have neglected the factor
-$p(D)$ on the right hand side since it is independent of $\lambda$ and serves
-only as a normalization factor. Since we assume each cell's mRNA count is
-independent of others, the likelihood of the full inference problem
-$p(D\mid\lambda)$ is simply a product of the single cell likelihoods given by
-\eq{eq:poisson_inference010} above, so
+Assuming each cell's mRNA count in our dataset is independent of others, the
+likelihood of the full inference problem $p(D\mid\lambda)$ is simply a product
+of the single cell likelihoods given by \eq{eq:poisson_inference010} above, so
 \begin{equation}
@@
 \end{equation}
-This is often notated simply as
-\begin{equation}
-D \sim \text{Poisson}(\lambda),
-\end{equation}
-which is read as the data $D$ is Poisson distributed with mean $\lambda$. The
-$\sim$ operator literally denotes ``is distributed according to.'' The
-conditioning on $\lambda$ is no longer explicitly notated, only implied by its
-appearance on the right hand side.
 
-To proceed we need to specify a prior. In this case we are extremely data-rich,
-as the dataset from Jones et.\ al~\cite{Jones2014} has of order 1000-3000
-single-cell measurements for each promoter, so our choice of prior matters
-little here, as long as it is sufficiently broad. For details on the prior 
-selection we refer the reader to Appendix~\ref{sec:bayesian}. For our
-purpose here it suffices to specify that we use a prior of the form $\lambda
-\sim \text{Gamma}(\alpha, \beta)$. Putting a gamma prior on $\lambda$ introduces
-two new parameters $\alpha$ and $\beta$ which parametrize the gamma distribution
-itself, which we use to encode the range of $\lambda$ values we view as
-reasonable. Recall $\lambda$ is the mean steady-state mRNA count per cell, which
-\textit{a priori} could plausibly be anywhere from 0 to a few hundred.
-$\alpha=1$ and $\beta=1/50$ achieve this, since the gamma distribution is
-strictly positive with mean $\alpha/\beta$ and standard deviation
-$\sqrt{\alpha}/\beta$.
+To proceed we need to specify a prior distribution $p(\lambda)$. In this case we
+are extremely data-rich, as the dataset from Jones et.\ al~\cite{Jones2014} has
+of order 1000-3000 single-cell measurements for each promoter, so our choice of
+prior matters little here, as long as it is sufficiently broad. For details on
+the prior selection we refer the reader to Appendix~\ref{sec:bayesian}. For our
+purpose here it suffices to specify that we use as prior a Gamma distribution.
+This particular choice of prior introduces two new parameters, $\alpha$ and
+$\beta$, which parametrize the gamma distribution itself, which we use to encode
+the range of $\lambda$ values we view as reasonable. Recall $\lambda$ is the
+mean steady-state mRNA count per cell, which \textit{a priori} could plausibly
+be anywhere from 0 to a few hundred. $\alpha=1$ and $\beta=1/50$ achieve this,
+since the gamma distribution is strictly positive with mean $\alpha/\beta$ and
+standard deviation $\sqrt{\alpha}/\beta$.
 
-As detailed in Appendix~\ref{sec:bayesian} this particular choice of prior
-known as the \textit{conjugate} prior for a Poisson likelihood allows us to 
+As detailed in Appendix~\ref{sec:bayesian} this particular choice of prior known
+as the \textit{conjugate} prior for a Poisson likelihood. This type of priors
+have nice properties to work with, in particular a conjugate prior allows us to
 get a closed form for the posterior distribution $p(\lambda \mid D)$. In
-particular it can be shown that~\cite{Gelman2013}
-\begin{equation}
-\lambda
-\sim \text{Gamma}\left(\alpha + \bar{m}N, \beta + N\right),
-\end{equation}
-where we defined the sample mean $\bar{m} = \frac{1}{N}\sum_k m_k$ for
-notational convenience. Furthermore given that $N$ is $\mathcal{O}(10^3)$ and
+particular it can be shown that for a Poisson distribution likelihood with its
+Gamma distribution conjugate prior, the posterior distribution is also a Gamma
+distribution ~\cite{Gelman2013}. In particular the two parameters $\alpha'$ and
+$\beta'$ for this posterior distribution take the form $\alpha' = \alpha +
+\bar{m} N$ and $\beta' = \beta + N$, where we defined the sample mean $\bar{m} =
+\frac{1}{N}\sum_k m_k$ for notational convenience, and $N$ is the number of
+cells n our dataset. Furthermore given that $N$ is $\mathcal{O}(10^3)$ and
 $\langle m\rangle \gtrsim 0.1$ for all promoters measured in~\cite{Jones2014}
-our data easily overwhelms the choice of prior, and allows us to make the 
-excellent approximation 
-\begin{equation}
-\lambda
-\sim \text{Gamma}\left(\alpha + \bar{m}N, \beta + N\right)
-\approx \text{Normal}\left(\bar{m}, \sqrt{\frac{\bar{m}}{N}}\right).
-\end{equation}
-As an example with real numbers, for the \textit{lacUV5} promoter, Jones et.\
-al~\cite{Jones2014} measured 2648 cells with an average mRNA count per cell of
-$\bar{m} \approx 18.7$. In this case then, our posterior is
-\begin{equation}
-\lambda
-\sim \text{Normal}\left(18.7, 0.08\right),
-\label{eq:posterior_poisson}
-\end{equation}
-which suggests we have inferred our model's one parameter to a precision of
-order 1\%.
+our data easily overwhelms the choice of prior, and allows us to approximate the
+Gamma distribution with a Gaussian distribution with mean $\bar{m}$ and variance
+$\bar{m} / N$ with marginal errors. As an example with real numbers, for the
+\textit{lacUV5} promoter, Jones et.\ al~\cite{Jones2014} measured 2648 cells
+with an average mRNA count per cell of $\bar{m} \approx 18.7$. For this case our
+posterior distribution $P(\lambda \mid D)$ would be a Gaussian distribution with
+mean $\mu = 18.7$, and a standard deviation $\sigma \approx 0.08$. This suggests
+we have inferred our model's one parameter to a precision of order 1\%.
 
-We alert the reader not to confuse a tightly constrained posterior distribution
-with a good generative model. In other words, not because the single parameter
-for our Poisson likelihood can only take a small range of values constrained by
-the data, this implies that such Poisson distribution characterizes the
-underlying cell-to-cell variability we observe in the data. To systematically
-check the validity of the generative model we can make use of \textit{posterior
-predictive checks}. This consists in generating synthetic data using our
-generative model for many plausible values of our parameter. For the case of our
-\textit{lacUV5} promoter this means that we sample a value for $\lambda$ out of
-the posterior distribution (Eq.~\ref{eq:posterior_poisson}), and generate 2648
-samples drawn from a Poisson distribution with  $\lambda$ chosen appropriately. We then repeat
-this procedure 1000 times. Clearly if those data appear quite different, the
-model has a problem. Put another way, if we suppose the generative model is
-true, then the synthetic datasets we generate should resemble the actual
-observed data, and if not, it suggests the model is missing important features.
-
-To compare so many samples with the actual observed data, we can use empirical
-cumulative distribution functions (ECDF) of the distribution quantiles. This 
-representation is shown in Figure~\ref{fig:constit_post_full}(B). In this 
-example, the median for each possible mRNA count for the Poisson distribution
-is shown as a dark green line, while the lighter green contains 95\% of the
-posterior predictive samples. This way of representing the posterior predictive
-checks gives us a sense of the range of data we might consider plausible, under
-the assumption that the model is true. In this case it is quite obvious that the
-observed data, plotted in orange, could not plausibly come from this Poisson
-generative model. We will cover the other model shown in
-Figure~\ref{fig:constit_post_full}(B) next.
+We remind the reader that we began this section claiming that the Poisson model
+was ``wrong'' since it could not reproduce features of the data such as a Fano
+factor $> 1$. The fact that we got such a narrow posterior distribution for our
+parameter $P(\lambda \mid D)$ does not equate to the model being adequate to
+describe the data. What this means is that given the data $D$, the parameter 
+$\lambda$ for our model is highly constrained to take a narrow range of values,
+but a narrow posterior distribution does not speak to the goodness of fit of the
+model. As we will see later in Figure~\ref{fig:constit_post_full} after
+exploring the bursty promoter model, indeed there is not good correspondence
+when contrasting the Poisson model with the experimental .
 
 \subsubsection{Model 5 - Bursty promoter}
-Let us now consider the problem of parameter inference from FISH data for model
-five from~\fig{fig1:means_cartoons}(C). As derived in
+
+Let us now consider the problem of parameter inference for model five
+from~\fig{fig1:means_cartoons}(C). As derived in
 Appendix~\ref{sec:gen_fcn_appdx}, the steady-state mRNA distribution in this
@@
 \end{equation}
-where $b$ is the mean burst size and $k_i$ is the burst rate nondimensionalized
-by the mRNA degradation rate $\gamma$. As sketched earlier, the story of this
-distribution is of geometrically-distributed bursts of mRNA, where the arrival
-of bursts is a Poisson process with rate $k_i$ and the mean size of a burst is
-$b$.
+where $b$ is the mean burst size and $k_i$ is the burst rate in units of the
+mRNA degradation rate $\gamma$. As sketched earlier, to think of the negative
+binomial distribution in terms of an intuitive story, we imagine the arrival of
+bursts as a Poisson process with rate $k_i$, where each burst has a 
+geometrically-distributed size with mean size $b$.
 
 As for the Poisson promoter model, this expression for the steady-state mRNA
-distribution is exactly the likelihood we want to use in Bayes theorem. Again
-denoting the single-cell mRNA count data as $D=\{m_1, m_2,\dots, m_N\}$, here
-Bayes theorem takes the form
+distribution is exactly the likelihood we want to use when stating Bayes
+theorem. Again denoting the single-cell mRNA count data as $D=\{m_1, m_2,\dots,
+m_N\}$, here Bayes theorem takes the form
 \begin{equation}
@@
 \end{equation}
-We already have our likelihood --the product of $N$ negative binomials as
-Eq.~\ref{eq:neg_bionom}--  so we only need to choose priors on $k_i$ and $b$.
+We already have our likelihood -- the product of $N$ negative binomials as
+Eq.~\ref{eq:neg_bionom} -- so we only need to choose priors on $k_i$ and $b$.
 For the datasets from~\cite{Jones2014} that we are analyzing, as for the Poisson
@@
 weak, but not nearly as weak because the dimensionality of our model has
-increased from one to two.
-Details on the arguments behind our prior distribution selection are left for
-Appendix~\ref{sec:bayesian}. Our full generative model can be specified as
-\begin{equation}
-\begin{split}
-\ln k_i \sim \text{Normal}(-0.5, 2) \\
-\ln b \sim \text{Normal}(0.5, 1) \\
-m \sim \text{NBinom}(k_i, b).
-\end{split}
-\end{equation}
-Note that we specify a distribution over the logarithm of our parameters $k_i$
-and $b$. This is commonly the case for parameters for which our previous 
-knowledge based on our domain expertise spans several orders of magnitude.
+increased from one parameter to two. Details on the arguments behind our prior
+distribution selection are left for Appendix~\ref{sec:bayesian}. We state here
+that the natural scale to explore these parameters is logarithmic. This is
+commonly the case for parameters for which our previous knowledge based on our
+domain expertise spans several orders of magnitude. For this we chose log-Normal
+distributions for both $k_i$ and $b$. Details on the mean and variance of these
+distributions can be found in Appendix~\ref{sec:bayesian}.
 
@@
 constitutive \textit{lacUV5} dataset from~\cite{Jones2014}. The resulting MCMC
-samples are shown in~\fig{fig:constit_post_full}(A). In contrast to the
+samples are shown in Figure~\ref{fig:constit_post_full}(A). In contrast to the
 active/inactive constitutive model considered in~\cite{Razo-Mejia2020}
@@
 
-Flush with success having found the model's posterior to be well-identified, the
-next step is posterior predictive sampling. The only procedural difference
-compared to the Poisson promoter model is that, since we have no analytical
-posterior expression but only samples from the posterior, we generate a
-synthetic dataset (i.e., posterior predictive samples) for each posterior sample
-from the MCMC sampling. \fig{fig:constit_post_full}(B) shows the resulting
-predictive ECDF; similarly to the Poisson promoter model, the median posterior
-predictive ECDF is plotted as a dark blue line, while a lighter blue shaded
-region encompasses 95\% of the posterior predictive samples. Unlike the Poisson
-promoter model, the experimental ECDF closely tracks the posterior predictive
-ECDF, indicating this model is actually able to generate the observed data and
-increasing our confidence that this model is at least not wrong.
+Having found the model's posterior to be well-identified as with the Poisson
+promoter, the next step is to compare both models with experimental data. To do
+this for the case of the bursty promoter, for each of the parameter samples
+shown in Figure~\ref{fig:constit_post_full}(A) we generated negative
+bionomial-distributed mRNA counts. As MCMC samples parameter space
+proportionally to the posterior distribution, this set of random samples span
+the range of possible values that we would expect given the correspondence
+between our theoretical model and the experimental data. A similar procedure can
+be applied to the Poisson promoter. To compare so many samples with the actual
+observed data, we can use empirical cumulative distribution functions (ECDF) of
+the distribution quantiles. This representation is shown in
+Figure~\ref{fig:constit_post_full}(B). In this example, the median for each
+possible mRNA count for the Poisson distribution is shown as a dark green line,
+while the lighter green contains 95\% of the randomly generated samples. This
+way of representing the fit of the model to the data gives us a sense of the
+range of data we might consider plausible, under the assumption that the model
+is true. For this case, as we expected given our premise of the Poisson promoter
+being wrong, it is quite obvious that the observed data, plotted in black. An 
+equivalent plot for the bursty promoter model is shown in blue. Again the darker
+tone shows the median, while the lighter color encompasses 95\% of the randomly
+generated samples. Unlike the Poisson promoter model, the experimental ECDF
+closely tracks the posterior predictive ECDF, indicating this model is actually
+able to generate the observed data and increasing our confidence that this model
+is at least not wrong.
 
@@
 shows the predicted slope according to Eq.~\ref{eq:bursty_equil_corresp1},
-described in text. Each individual promoter is
-labeled with a unique number in both (C) and (D) for cross comparison.}
+described in text. Each individual promoter is labeled with a unique number in
+both (C) and (D) for cross comparison and for comparison with 
+Figure~\ref{fig2:constit_cartoons}(B).}
 \label{fig:constit_post_full}
@@
 encloses approximately 95\% of the samples, and thus 95\% of the probability
-mass, of its posterior distribution. Posterior predictive ECDFs (not shown)
+mass, of its posterior distribution. Theory-experiment comparisons (not shown)
 display a similar level of agreement between data and predictive samples as for
@@
 statements about the mean gene expression. This implies that we can draw the
-connection between both frameworks by equating the mean mRNA $\left\langle m \right\rangle$. This results in
+connection between both frameworks by equating the mean mRNA $\left\langle m
+\right\rangle$. This results in
 \begin{equation}
@@
 By taking the weak promoter approximation for the equilibrium model ($P/N_{NS} 
-\exp(-\beta\Delta\varepsilon_r \ll 1)$) results in
+\exp(-\beta\Delta\varepsilon_r) \ll 1$) results in
 \begin{equation}
@@
 
-Given this result, how are the two coarse-grainings related?
-A quick estimate can shed some light.
-Consider for instance the \textit{lacUV5} promoter, which we see from
-Figure~\ref{fig:constit_post_full}(A) has $k_i/\gamma \sim b \sim \text{few}$,
-from Figure~\ref{fig:constit_post_full}(B) has $\langle m \rangle \sim 20$,
-and from~\cite{Brewster2012} has $\beta\Delta\epsilon_P \sim - 6.5$.
+Given this result, how are the two coarse-grainings related? A quick estimate
+can shed some light. Consider for instance the \textit{lacUV5} promoter, which
+we see from Figure~\ref{fig:constit_post_full}(A) has $k_i/\gamma \sim b \sim
+\text{few}$, from Figure~\ref{fig:constit_post_full}(B) has $\langle m \rangle
+\sim 20$, and from~\cite{Brewster2012} has $\beta\Delta\epsilon_P \sim - 6.5$.
 Further we generally assume $P/N_{NS} \sim 10^{-3}$ since
-$N_{NS}\approx4.6\times10^6$ and $P\sim10^3$.
-After some guess-and-check with these values, one finds the only
-association that makes dimensional sense and produces the correct
-order-of-magnitude for the known parameters is to take
+$N_{NS}\approx4.6\times10^6$ and $P\sim10^3$. After some guess-and-check with
+these values, one finds the only association that makes dimensional sense and
+produces the correct order-of-magnitude for the known parameters is to take
 \begin{equation}
@@
 
-While the associations represented by
-Eq.~\ref{eq:bursty_equil_corresp1} and Eq.~\ref{eq:bursty_equil_corresp2}
-appear to be born out by the data in Figure~\ref{fig:constit_post_full},
-we do not find the association of parameters they imply to be intuitive.
-We are also cautious to ascribe too much physical reality to the parameters.
-Indeed, part of our point in comparing the various constitutive
-promoter models is to demonstrate that these models each provide
-an internally self-consistent framework that adequately describes
-the data, but attempting to translate between models reveals
-the dubious physical interpretation of their parameters.
+While the associations represented by Eq.~\ref{eq:bursty_equil_corresp1} and
+Eq.~\ref{eq:bursty_equil_corresp2} appear to be born out by the data in
+Figure~\ref{fig:constit_post_full}, we do not find the association of parameters
+they imply to be intuitive. We are also cautious to ascribe too much physical
+reality to the parameters. Indeed, part of our point in comparing the various
+constitutive promoter models is to demonstrate that these models each provide an
+internally self-consistent framework that adequately describes the data, but
+attempting to translate between models reveals the dubious physical
+interpretation of their parameters.
 

### 2020-05-27 doc/section_04_bayesian_inference.tex
MRM pass through the entire Bayesian inference section. Still is missing last part.

@@
 
-\subsection{Transcription factor kinetics can be inferred from FISH measurements}
-\label{sec:rep_kinetics_inference}
+\subsection{Transcription factor kinetics can be inferred from FISH
+measurements}\label{sec:rep_kinetics_inference}
 
-Now that we have a satisfactory model in hand for constitutive
-promoters, we would like to return to the main thread: can we
-infer repressor binding and unbinding rates from mRNA distributions
-over a population of cells as measured by FISH in~\cite{Jones2014}?
-If so, how do these inferred rates compare to direct
-single-molecule measurements such as from~\cite{Hammar2014} and
-to binding energies such as from~\cite{Garcia2011a}
-and~\cite{Razo-Mejia2018}, which were inferred under the assumptions
-of the equilibrium models in Figure~\ref{fig1:means_cartoons}(B)?
-And can this comparison shed light on the unreasonable effectiveness
-of the equilibrium models?
+Now that we have a satisfactory model in hand for constitutive promoters, we
+would like to return to the main thread: can we infer repressor binding and
+unbinding rates from mRNA distributions over a population of cells as measured
+by single-molecule FISH in~\cite{Jones2014}? If so, how do these inferred rates
+compare to direct single-molecule measurements such as from~\cite{Hammar2014}
+and to binding energies such as from~\cite{Garcia2011a}
+and~\cite{Razo-Mejia2018}, which were inferred under the assumptions of the
+equilibrium models in Figure~\ref{fig1:means_cartoons}(B)? And can this
+comparison shed light on the unreasonable effectiveness of the equilibrium
+models?
 
@@
 \caption{\textbf{Simple repression parameter inference and comparison.}
-(A) Contours which enclose 50\% and 95\% of the posterior
-probability mass are shown for each of several 2D slices of
-the 9D posterior distribution. The model assumes one
-unbinding rate for each operator (O1, O2, O3) and one binding rate for each aTc
-concentration (corresponding to an unknown mean repressor copy number).
-(B, upper panel)
-Ratios of our inferred unbinding rates are compared with
-operator binding energy differences measured by
-Garcia and Phillips~\cite{Garcia2011a} (triangles) and
-Razo-Mejia et.\ al.~\cite{Razo-Mejia2018} (circles).
-Blue glyphs compare O2-O1, while orange compare O1-Oid.
-(B, lower panel)
-Unbinding rates for O1 (cyan) and Oid (red) inferred in this work are compared with
-single-molecule measurements of the same from Hammar et.\ at.\ 2014.
-The comparison requires an assumption about mRNA lifetime
-$\gamma^{-1}$, which was not measured, and we plot representative
-choices assuming $\gamma^{-1}=3$~min (triangles) or $\gamma^{-1}=5$~min (squares).
-(C) Posterior predictive checks are shown for each of the
-datasets used in the inference of the model in (A). Observed FISH
-data from~\cite{Jones2014} is plotted as black lines. The median
-posterior predictive ECDF for each condition is plotted as a dark
-colored line. Lighter colored bands enclose 95\% of all posterior
-predictive samples for a given operator/repressor copy number pair.
-Samples are sorted by operator and plotted separately for visual clarity.
-The unrepressed promoter, UV5, is shown with each as a reference point.
-}
+(A) Contours which enclose 50\% and 95\% of the posterior probability mass are
+shown for each of several 2D slices of the 9D posterior distribution. The model
+assumes one unbinding rate for each operator (Oid, O1, O2) and one binding rate
+for each aTc induction concentration (corresponding to an unknown mean repressor
+copy number). (B, upper panel) Ratios of our inferred unbinding rates are
+compared with operator binding energy differences measured by Garcia and
+Phillips~\cite{Garcia2011a} (triangles) and Razo-Mejia et.\
+al.~\cite{Razo-Mejia2018} (circles). Blue glyphs compare O2-O1, while orange
+compare O1-Oid. (B, lower panel) Unbinding rates for O1 (cyan) and Oid (red)
+inferred in this work are compared with single-molecule measurements of the same
+from Hammar et.\ at.\ 2014. The comparison requires an assumption about mRNA
+lifetime $\gamma^{-1}$, which was not measured, and we plot representative
+choices assuming $\gamma^{-1}=3$~min (triangles) or $\gamma^{-1}=5$~min
+(squares). (C) Theory-experiment comparison are shown for each of the datasets
+used in the inference of the model in (A). Observed single-molecule mRNA counts
+data from~\cite{Jones2014} is plotted as black lines. The median of the randomly
+generated samples for each condition is plotted as a dark colored line. Lighter
+colored bands enclose 95\% of all samples for a given operator/repressor copy
+number pair. Samples are sorted by operator and plotted separately for visual
+clarity. The unregulated promoter, UV5, is shown with each as a reference
+point.}
 \label{fig4:repressed_post_full}
@@
 
-As we found in Section~\ref{sec:beyond_means}, for our purposes
-the ``right'' model of a constitutive promoter is the bursty
-picture, model five in Figure~\ref{fig2:constit_cartoons}.
-Therefore our starting point here is the analogous model with
-repressor added, model 5 in Figure~\ref{fig1:means_cartoons}.
-For a given repressor binding site and copy number,
-this model has four rate parameters to be inferred:
-the repressor binding and unbinding rates $k_R^+$, and $k_R^-$,
-the initiation rate of bursts, $k_i$, and the mean burst size $b$
-(we nondimensionalize all of these by the mRNA degradation rate $\gamma$).
+As we found in Section~\ref{sec:beyond_means}, for our purposes the ``right''
+model of a constitutive promoter is the bursty picture, model five in
+Figure~\ref{fig2:constit_cartoons}. Therefore our starting point here is the
+analogous model with repressor added, model 5 in
+Figure~\ref{fig1:means_cartoons}(C). For a given repressor binding site and copy
+number, this model has four rate parameters to be inferred: the repressor
+binding and unbinding rates $k_R^+$, and $k_R^-$, the initiation rate of bursts,
+$k_i$, and the mean burst size $b$ (we nondimensionalize all of these by the
+mRNA degradation rate $\gamma$).
 
-Before showing the mathematical formulation of our statistical
-inference model, we would like to sketch the intuitive structure.
-The dataset from~\cite{Jones2014} we consider consists of
-single-cell mRNA FISH measurements of nine different conditions,
-spanning several combinations of three unique binding sites and
-four unique repressor copy numbers. We assume that the values of
-$k_i$ and $b$ are known, since we have already cleanly inferred
-them from constitutive promoter data, and further we assume that
-these values are the same across datasets with different
-repressor binding sites and copy numbers. We assume that there is
-one unbinding rate parameter for each repressor binding site,
-which is shared across conditions, and likewise one binding rate
-for each unique repressor copy number. This makes our model seven
-dimensional, or nine if one counts $k_i$ and $b$ as well.
+Before showing the mathematical formulation of our statistical inference model,
+we would like to sketch the intuitive structure. The dataset
+from~\cite{Jones2014} we consider consists of single-cell mRNA FISH measurements
+of nine different conditions, spanning several combinations of three unique
+repressor binding sites and four unique repressor copy numbers. We assume that
+the values of $k_i$ and $b$ are known, since we have already cleanly inferred
+them from constitutive promoter data, and further we assume that these values
+are the same across datasets with different repressor binding sites and copy
+numbers. In other words, we assume that the regulation of the transcription
+factor does not affect the mean burst size nor the burst initiation rate. The
+regulation happens as the promoter is taken away from the transcriptionally
+active state when the promoter is bound. We assume that there is one unbinding
+rate parameter for each repressor binding site, which is shared across aTc
+inducer conditions, and likewise one binding rate for each unique repressor copy
+number. This makes our model nine dimensional, or nine if one counts $k_i$ and
+$b$ as well.
 
@@
 \vect{k} =\{k_{Oid}^-, k_{O1}^-, k_{O2}^-,
-k_{0.5}^+, k_{1}^+, k_{2}^+, k_{10}^+\}.
+k_{0.5}^+, k_{1}^+, k_{2}^+, k_{10}^+\},
 \end{equation}
-Note the repressor copy numbers are not known directly, so we label their
-association rates by the concentration of a small molecule inducer aTc
-which indirectly controls the transcription of $\textit{lacI}$.
-Also note that the authors of~\cite{Jones2014} report estimates
-of LacI copy number per cell rather than direct measurements.
-However, these estimates were made \textit{assuming} the validity
-of the equilibrium models in Figure~\ref{fig1:means_cartoons},
-and since \textit{testing} these models is our goal, we will make
-no assumptions about the LacI copy number for given aTc concentrations.
+where subscripts for dissociation rates $k^-$  indicate the different repressor
+binding sites, and subscripts for association rates $k^+$ indicate the aTc
+concentration at which the repressor itself was induced. We emphasize that for
+this particular dataset the repressor copy numbers are not known directly, so we
+label their association rates by the concentration of a small molecule inducer
+aTc which indirectly controls the transcription of $\textit{lacI}$. Also note
+that the authors of~\cite{Jones2014} report estimates of LacI copy number per
+cell rather than direct measurements. However, these estimates were made
+\textit{assuming} the validity of the equilibrium models in
+Figure~\ref{fig1:means_cartoons}, and since \textit{testing} these models is our
+goal, we will make no assumptions about the LacI copy number for given aTc
+concentrations.
 
-Bayes theorem reads simply
+Having stated the problem, Bayes theorem reads simply
 \begin{equation}
@@
 \end{equation}
-where $D$ is the set of all $N$ observed single-cell mRNA counts
+where $D$ is again the set of all $N$ observed single-cell mRNA counts
 across the various conditions. We assume that individual single-cell
@@
 where $k_j^\pm$ represent the appropriate binding and unbinding
-rates for the $j$-th measured cell. The probability
+rates out of $\vect{k}$ for the $j$-th measured cell. The probability
 $p(m\mid k_j^+, k_j^-, k_i, b)$ appearing in the last expression
@@
 \mmnote{Probably SI most of this paragraph?}
-This likelihood is rather inscrutable. We did not find any of the
-known analytical approximations for ${_2F_1}$ terribly useful in
-gaining intuition, so we instead resorted to numerics. One insight we found
+This likelihood is rather inscrutable. We did not find any of the known
+analytical approximations for ${_2F_1}$ terribly useful in gaining intuition, so
+we instead resorted to numerics. One insight we found
 \mmnote{a plot would probably help explain, maybe SI, need to explain better}
 was that for very strong or very weak repression, the distribution in
-Eq.~\ref{eq:p_m_bursty+rep} is well approximated by a negative binomial
-with burst size $b$ and burst rate $k_i$ equal to their
-constitutive \textit{lacUV5} values, except with $k_i$ multiplied
-by the fold-change $\left(1+k_R^+/k_R^-\right)^{-1}$.
-In other words, once again only the \textit{ratio} $k_R^+/k_R^-$
-was detectable. But for intermediate repression, the distribution
-was visibly broadened with Fano factor greater than $1+b$, the
-value for the corresponding constitutive case. This indicates
-that the repressor rates had left an imprint on the distribution,
-and perhaps intuitively, this intermediate regime occurs for values of
-$k_R^\pm$ comparable to the burst rate $k_i$. Put another way,
-if the repressor rates are much faster or much slower than $k_i$,
-then there is a timescale separation and effectively only one
-timescale remains, $k_i\left(1+k_R^+/k_R^-\right)^{-1}$.
-Only when all three rates in the problem are comparable does the
-mRNA distribution retain detectable information about them.
+Eq.~\ref{eq:p_m_bursty+rep} is well approximated by a negative binomial with
+burst size $b$ and burst rate $k_i$ equal to their constitutive \textit{lacUV5}
+values, except with $k_i$ multiplied by the fold-change
+$\left(1+k_R^+/k_R^-\right)^{-1}$. In other words, once again only the
+\textit{ratio} $k_R^+/k_R^-$ was detectable. But for intermediate repression,
+the distribution was visibly broadened with Fano factor greater than $1+b$, the
+value for the corresponding constitutive case. This indicates that the repressor
+rates had left an imprint on the distribution, and perhaps intuitively, this
+intermediate regime occurs for values of $k_R^\pm$ comparable to the burst rate
+$k_i$. Put another way, if the repressor rates are much faster or much slower
+than $k_i$, then there is a timescale separation and effectively only one
+timescale remains, $k_i\left(1+k_R^+/k_R^-\right)^{-1}$. Only when all three
+rates in the problem are comparable does the mRNA distribution retain detectable
+information about them.
 
-Next we specify priors. As for the constitutive model, weakly
-informative lognormal priors are a natural choice for all our
-rates. We found that if the priors were too weak, our MCMC
-sampler would often become stuck in regions of parameter space
-with very low probability density, unable to move. We struck a
-balance in choosing our prior widths between helping the sampler
-run while simultaneously verifying that the marginal posteriors
-for each parameter were not artificially constrained or distorted
-by the presence of the prior. The only exception to this is the
-highly informative priors we placed on $k_i$ and $b$, since we
-have strong knowledge of them from our inference of constitutive
-promoters above.
+Next we specify priors. As for the constitutive model, weakly informative
+lognormal priors are a natural choice for all our rates. We found that if the
+priors were too weak, our MCMC sampler would often become stuck in regions of
+parameter space with very low probability density, unable to move. We struck a
+balance in choosing our prior widths between helping the sampler run while
+simultaneously verifying that the marginal posteriors for each parameter were
+not artificially constrained or distorted by the presence of the prior. The only
+exception to this is the highly informative priors we placed on $k_i$ and $b$,
+since we have strong knowledge of them from our inference of constitutive
+promoters above. All details for our prior distributions are listed in 
+Appendix~\ref{sec:bayes}.
 
-With priors and likelihood specified we may write down our complete generative model as
-\begin{equation}
-\begin{split}
-\log_{10}k_i &\sim \text{Normal}(0.725, 0.025)\\
-\log_{10}b   &\sim \text{Normal}(0.55, 0.025)\\
-\log_{10}k_{0.5}^+ &\sim \text{Normal}(-0.45, 0.3)\\
-\log_{10}k_{1}^+   &\sim \text{Normal}(0.6, 0.3)\\
-\log_{10}k_{2}^+   &\sim \text{Normal}(1.15, 0.3)\\
-\log_{10}k_{10}^+  &\sim \text{Normal}(1.5, 0.3)\\
-\log_{10}k_{Oid}^- &\sim \text{Normal}(-0.25, 0.3)\\
-\log_{10}k_{O1}^-  &\sim \text{Normal}(0.1, 0.3)\\
-\log_{10}k_{O2}^-  &\sim \text{Normal}(0.45, 0.3)\\
-m &\sim \text{Likelihood}(k_R^+, k_R^-, k_i, b),
-\end{split}
-\end{equation}
-where the likelihood is specified by Eq.~\ref{eq:p_m_bursty+rep}.
-\mmnote{Now that I've typed this out, this notation seems totally
-pointless here, especially when the likelihood is not a standard
-distribution. Maybe I should make a table or something to record
-the prior values, and maybe even that should go to SI?}
-We ran MCMC sampling on the full nine dimensional posterior
-specified by this generative model. To attempt to visualize this
-object, in Figure~\ref{fig4:repressed_post_full}(A) we plot
-several two-dimensional slices as contour plots, analogous to
-Figure~\ref{fig:constit_post_full}(C). Each of these nine slices
-corresponds to the $(k_R^+, k_R^-)$ pair of rates for one of the
-conditions from the dataset used to fit the model and gives a
+We ran MCMC sampling on the full nine dimensional posterior specified by this
+generative model. To attempt to visualize this object, in
+Figure~\ref{fig4:repressed_post_full}(A) we plot several two-dimensional slices
+as contour plots, analogous to Figure~\ref{fig:constit_post_full}(C). Each of
+these nine slices corresponds to the $(k_R^+, k_R^-)$ pair of rates for one of
+the conditions from the dataset used to fit the model and gives a
 sense of the uncertainty and correlations in the posterior.
+\mrm{before invoking part (C) of the figure we are missing th explanation for 
+part (B). I don't fully understand how these computations were made, so I'll 
+let Jupiter detail this.}
+
+% With priors and likelihood specified we may write down our complete generative model as
+% \begin{equation}
+% \begin{split}
+% \log_{10}k_i &\sim \text{Normal}(0.725, 0.025)\\
+% \log_{10}b   &\sim \text{Normal}(0.55, 0.025)\\
+% \log_{10}k_{0.5}^+ &\sim \text{Normal}(-0.45, 0.3)\\
+% \log_{10}k_{1}^+   &\sim \text{Normal}(0.6, 0.3)\\
+% \log_{10}k_{2}^+   &\sim \text{Normal}(1.15, 0.3)\\
+% \log_{10}k_{10}^+  &\sim \text{Normal}(1.5, 0.3)\\
+% \log_{10}k_{Oid}^- &\sim \text{Normal}(-0.25, 0.3)\\
+% \log_{10}k_{O1}^-  &\sim \text{Normal}(0.1, 0.3)\\
+% \log_{10}k_{O2}^-  &\sim \text{Normal}(0.45, 0.3)\\
+% m &\sim \text{Likelihood}(k_R^+, k_R^-, k_i, b),
+% \end{split}
+% \end{equation}
+% where the likelihood is specified by Eq.~\ref{eq:p_m_bursty+rep}.
+% \mmnote{Now that I've typed this out, this notation seems totally
+% pointless here, especially when the likelihood is not a standard
+% distribution. Maybe I should make a table or something to record
+% the prior values, and maybe even that should go to SI?}

### 2020-05-28 doc/appendix_01_bayesian_inference.tex
MRM starting the intro section to the Bayesian inference appendix

@@
 \label{sec:bayesian}
+
+\subsection{The problem of parameter inference}
+
+\mmnote{Bayes theorem. Elaborate on posterior as plausibility over space of
+model parameters.}
+
+One could argue that the whole goal of formulating theoretical models about
+nature is to sharpen our understanding from qualitative statements to precise
+quantitative assertions about the relevant features of the natural phenomena in
+question \cite{Gunawardena2014}. It is in these models that we intend to distill
+the essential parts of the object of study. Writing down such models leads to a
+propagation of mathematical variables that parametrize our models. By assigning
+numerical values to these parameters we can compute concrete predictions that
+can be contrasted with experimental data. But for these predictions to match the
+data the parameter values have to carefully be chosen from the whole parameter
+space. But how do we go about asserting the effectiveness of different regions
+of parameter space to speak to the ability of our model to reproduce the
+experimental observations? The language of probability, and more specifically of
+Bayesian statistics is -- we think -- the natural language to tackle this
+question.
+
+\subsubsection{Bayes theorem}
+
+Bayes theorem is a simple mathematical statement that can apply to \textit{any}
+logical conjecture. For two particular events $A$ and $B$ that potentially 
+depend on each other Bayes theorem gives us a recipe for how to update our 
+beliefs about one, let's say $B$, given that we get to observe something about
+$A$. In its most classic form Bayes theorem is written as
+\begin{equation}
+P(B \mid A) = {P(A \mid B) P(B) \over P(A)},
+\end{equation}
+where the vertical line $\mid$ is read as ``given that''. So $P(B \mid A)$ is
+read as probability of $B$ given that $A$ took place. $A$ and $B$ can be any
+logical assertion. In particular the problem of Bayesian inference focuses on
+the question of finding the probability distribution of a particular parameter
+value given the data.
+
+For a given model with a set of parameters $\vec{\theta} = (\theta_1, \theta_2,
+\ldots, \theta_n)$, the so-called \textit{posterior distribution} 
+$P(\vec{\theta} \mid D)$, where $D$ is the experimental data, quantifies the
+plausibility of a set of parameter values given that we get to observe some
+particular dataset. In other words, through the application of Bayes formula we
+update our beliefs on the possible values that parameters can take upon learning
+the outcome of a particular experiment. We specify the word ``update'' as we
+come to every inference problem with prior information about the plausibility of
+particular regions of parameter space even before performing any experiment.
+Even when we claim as researchers that we are totally ignorant about the values
+that the parameters in our models can take, we always come to a problem with
+domain expertise that can be exploited. If this was not the case, it is likely
+that the formulation of our model is not going to capture the phenomena we claim
+to want to understand. This prior information is caputerd in the \textit{prior
+probability} $P(\vec{\theta})$. The relationship between how parameter values
+can connect with the data is enconded in the \textit{likelihood function} $P(D
+\mid \vec{\theta})$. Our theoretical model, whether deterministic or
+probabilistic, is encoded in this term that can be intuitively understood as the
+probability of having observed the particular experimental data we have at hand
+given that our model is parametrized with the concrete values $\vec{\theta}$. 
+Implicitly here we are also conditioning on the fact that our theoretical model
+is ``true,'' i.e. the model itself if evaluated or simulated in the computer is
+capable of generating equivalent datasets to the one we got to observe in an 
+experiment. In this way Bayesian inference consists of applying Bayes formula 
+as 
+\begin{equation}
+P(\vec{\theta} \mid D) \propto P(D \mid \vec{\theta}) P(\vec{\theta}).
+\end{equation}
+Notice than rather than writing the full form of Bayes theorem, we limit 
+ourselves to the terms that depend on our quantity of interest -- that is the 
+parameter values themselves $\vec{\theta}$ -- as the denominator $P(D)$ only
+serves as a normalization constant.
+
+\subsubsection{The likelihood function}
+
+\mmnote{Choosing likelihoods. Easy here, just the steady-state distributions of
+our master equations, assuming iid single-cell measurements.}
+
+As we alluded in the previous section it is through the likelihood function 
+$P(D \mid \vec{\theta})$ that we encode the connection between our parameter 
+values and the experimental observables. Broadly speaking there are two classes
+of models that we might need to encode into our likelihood function:
+\begin{itemize}
+        \item Deterministic models: Models for which a concrete selection of
+        parameter values give a single output. Say differently, rigid models 
+        with a one-to-one mapping between inputs and outputs.
+        \item Probabilistic models: As the name says it, models that rather than
+        having a one-to-one mapping, describe the full probability distribution
+        of possible outputs.
+\end{itemize}
+In this paper we focus on inference done with probabilistic models. After all
+the chemical master equations we wrote down describe the time evolutions of the
+mRNA probability distribution. So all our terms $P(\vec{\theta} \mid D)$ will 
+be given by the steady-state solution of the corresponding chemical master 
+equation in question. This is rather convenient as we do not have to worry about
+adding a statistical model on top of our model to describe deviations from the
+predictions; our models themselves focus on predicting such variations from 
+cell count to cell count.
+
+\subsubsection{Prior selection}
+
+\mmnote{Choosing priors. Not critical here because we are data-rich and
+nonhierarchical, as long as we don't exclude ground truth. In other
+applications, prior can matter a lot. This is a feature, not a bug: it's
+revealing hidden structure that was implied by your modeling assumptions and you
+just didn't realize it. Predictive checks to verify you're not excluding any
+plausible values.}
+
+The different models explored in this work embraced different levels of 
+coarse-graining that resulted in a diverse set of number of parameters. For each
+of these model configurations Bayes theorem demands from us to represent our 
+preconceptions on the possible parameter values in the form of the prior 
+$P(\vec{\theta})$. Throughout this work for models with $> 1$ parameter we 
+assign independent priors to each of the parameters; this is
+\begin{equation}
+P(\vec{\theta}) = \prod_{i=1}^n P(\theta_i).
+\end{equation}
+Although it is common practice to use non-informative, or maximally
+uninformative priors, we are of the mindset that this is a disservice to the
+philosophical and practical implications of Bayes theorem. It sounds almost
+contradictory to claim that can we represent our thinking about a natural
+phenomena in the form of a mathematical model -- which itself is a bold claim of
+deeply understanding the features that matter in the object of study -- but we
+have absolutely no idea what the parameter values ought to be. We therefore make
+use of our own expertise, many times in the form of order-of-magnitude
+estimates, to write down prior distributions for our parameters.
+
 \mmnote{
 \begin{itemize}
-\item Bayes theorem.
-Elaborate on posterior as plausibility over
-space of model parameters.
-\item Choosing likelihoods.
-Easy here, just the steady-state distributions of our master
-equations, assuming iid single-cell measurements.
-\item Choosing priors.
-Not critical here because we are data-rich and nonhierarchical,
-as long as we don't exclude ground truth. In other applications,
-prior can matter a lot. This is a feature, not a bug: it's
-revealing hidden structure that was implied by your modeling
-assumptions and you just didn't realize it. Predictive checks to
-verify you're not excluding any plausible values.
+% \item Bayes theorem.
+% Elaborate on posterior as plausibility over
+% space of model parameters.
+% \item Choosing likelihoods.
+% Easy here, just the steady-state distributions of our master
+% equations, assuming iid single-cell measurements.
+% \item Choosing priors.
+% Not critical here because we are data-rich and nonhierarchical,
+% as long as we don't exclude ground truth. In other applications,
+% prior can matter a lot. This is a feature, not a bug: it's
+% revealing hidden structure that was implied by your modeling
+% assumptions and you just didn't realize it. Predictive checks to
+% verify you're not excluding any plausible values.
 \item Marginalization/expectations.
@@
 
-\subsection{Parameter inference for constitutive promoters}
+\subsection{Bayesian inference on constitutive promoters}
+\label{sec:si_bayes_unreg}
 

### 2020-05-28 doc/section_04_bayesian_inference.tex

@@
 \label{section_04_bayesian_inference}
+
+\textcolor{red}{In this section of the paper, we continue our program of providing
+one complete description of the entire broad sweep of studies
+that have been made in the context of the repressor-operator model,
+dating all the way back to the original work of Jacob and Monod
+and including the visionary quantitative work of M\"{u}ller-Hill and collaborators
+and up to more recent studies.   In addition, the aim is to reconcile the 
+equilibrium and non-equilibrium perspectives that have been brought
+to bear on this problem.  To that end, we require a self-consistent means
+of determining the parameters in these different theoretical analyses and
+the current section does that, first by examining the constitutive promoter
+and then turning to the full case of simple repression. }
+
 \subsection{Parameter inference for constitutive promoters}
@@
 
-As detailed in Appendix~\ref{sec:bayesian} this particular choice of prior known
+As detailed in Appendix~\ref{sec:bayesian} this particular choice of prior is known
 as the \textit{conjugate} prior for a Poisson likelihood. This type of priors
@@
 \frac{1}{N}\sum_k m_k$ for notational convenience, and $N$ is the number of
-cells n our dataset. Furthermore given that $N$ is $\mathcal{O}(10^3)$ and
+cells in our dataset. Furthermore, given that $N$ is $\mathcal{O}(10^3)$ and
 $\langle m\rangle \gtrsim 0.1$ for all promoters measured in~\cite{Jones2014}
@@
 exploring the bursty promoter model, indeed there is not good correspondence
-when contrasting the Poisson model with the experimental .
+when contrasting the Poisson model with the experimental data.
 
@@
 is true. For this case, as we expected given our premise of the Poisson promoter
-being wrong, it is quite obvious that the observed data, plotted in black. An 
+being wrong, it is quite obvious that the observed data, plotted in black is not consistent with
+the Poisson promoter model. An 
 equivalent plot for the bursty promoter model is shown in blue. Again the darker

### 2020-05-28 doc/appendix_02_bursty.tex
update organization file

--- 2020-05-26 doc/appendix_03_regulated_promoters.tex
@@
 &= \left[ z (1 - \theta) \right]^{m'} 
-\sum_{n=0}^{\infty} \left[ z (1 - \theta) \right]^{m'} \\
+\sum_{n=0}^{\infty} \left[ z (1 - \theta) \right]^{n} \\
 &= \left[ z (1 - \theta) \right]^{m'} 
@@
 \end{align}
-where we use the geometric series since, as stated before,$| z (1 - \theta) | <
+where we use the geometric series since, as stated before, $| z (1 - \theta) | <
 1$. Putting these results together, the PDE for the generation function is
@@
 has repressor bound and the other produces transcriptional bursts as above.
-A schematic of this model is shown in~\fig{fig3:kR_inferences}.
-Although now we have an equation for each promoter state, otherwise
-the master equation reads similarly to the one-state case, except with
-additional terms corresponding to transitions between promoter states, namely
+A schematic of this model is shown as model 5 in
+Figure~\ref{fig1:means_cartoons}(C). Although now we have an equation for each
+promoter state, otherwise the master equation reads similarly to the one-state
+case, except with additional terms corresponding to transitions between promoter
+states, namely
 \begin{align}
-\deriv{t}p_R(m,t) =& k_R^+ p_A(m,t) - k_R^- p_R(m,t)
+{d\over dt}p_R(m,t) =& 
+k_R^+ p_A(m,t) - k_R^- p_R(m,t)
         + (m+1)\gamma p_R(m+1,t) - m\gamma p_R(m,t)
@@
 \begin{split}
-\deriv{t}p_A(m,t) =& - k_R^+ p_A(m,t) + k_R^- p_R(m,t)
+{d\over dt}p_A(m,t) =& - k_R^+ p_A(m,t) + k_R^- p_R(m,t)
         + (m+1)\gamma p_A(m+1,t) - m\gamma p_A(m,t) 
@@
 \end{align}
-where $p_R(m,t)$ is the probability of the system having $m$ mRNA copies
-and having repressor bound to the promoter at time $t$, and $p_A$ is an
-analogous probability to find the promoter without repressor bound.
-$k_R+$ and $k_R^-$ are, respectively, the rates at which repressors
-bind and unbind to and from the promoter, and
-$\gamma$ is the mRNA degradation rate.
-$k_i$ is the rate at which bursts initiate, and as before,
-the geometric distribution of burst sizes has mean $b=(1-\theta)/\theta$.\footnote{
+where $p_R(m,t)$ is the probability of the system having $m$ mRNA copies and
+having repressor bound to the promoter at time $t$, and $p_A$ is an analogous
+probability to find the promoter without repressor bound. $k_R+$ and $k_R^-$
+are, respectively, the rates at which repressors bind and unbind to and from the
+promoter, and $\gamma$ is the mRNA degradation rate. $k_i$ is the rate at which
+bursts initiate, and as before, the geometric distribution of burst sizes has
+mean $b=(1-\theta)/\theta$.
+\footnote{
 $\theta$ can be viewed as a Bernoulli trial probability: after each transcript
-in a burst is produced, $\theta$ is the probability that the burst
-terminates and $(1-\theta)$ is the probability that the burst continues.
+in a burst is produced, $\theta$ is the probability that the burst terminates
+and $(1-\theta)$ is the probability that the burst continues.
 }
 
-Interestingly, it turns out that this problem maps exactly onto the
-three-stage promoter model considered by
-Shahrezaei and Swain in~\cite{Shahrezaei2008}, with relabelings.
-Their approximate solution for protein distributions amounts to
-the same approximation we make here in regarding the duration
-of mRNA synthesis bursts as instantaneous, so their solution
-for protein distributions also solves our problem of mRNA distributions.
-Let us examine the analogy more closely.
-They consider a two-state promoter, as we do here, but they model
-mRNA as being produced one at a time and degraded, with rates
-$v_0$ and $d_0$.
-Then they model translation as occuring with rate $v_1$,
-and protein degradation with rate $d_1$.
-Now consider the limit where $v_1, d_0\rightarrow\infty$ with
-their ratio $v_1/d_0$ held constant.
-$v_1/d_0$ resembles the average burst size of translation from a single
-mRNA: these are the rates of two Poisson processes that compete over a
-transcript, which matches the story of geometrically distributed burst sizes.
-So after taking this limit, on timescales slow compared to $v_1$ and $d_0$,
-it appears that transcription events fire at rate $v_0$ and produce a
-geometrically distributed burst of translation of mean size $v_1/d_0$,
-which intuitively matches the story we have told above for mRNA
-with variables relabeled.
+Interestingly, it turns out that this problem maps exactly onto the three-stage
+promoter model considered by Shahrezaei and Swain in~\cite{Shahrezaei2008}, with
+relabelings. Their approximate solution for protein distributions amounts to the
+same approximation we make here in regarding the duration of mRNA synthesis
+bursts as instantaneous, so their solution for protein distributions also solves
+our problem of mRNA distributions. Let us examine the analogy more closely. They
+consider a two-state promoter, as we do here, but they model mRNA as being
+produced one at a time and degraded, with rates $v_0$ and $d_0$. Then they model
+translation as occurring with rate $v_1$, and protein degradation with rate
+$d_1$ as shown in Figure~\ref{fig:shahrezaei}. Now consider the limit where
+$v_1, d_0\rightarrow\infty$ with their ratio $v_1/d_0$ held constant. $v_1/d_0$
+resembles the average burst size of translation from a single mRNA: these are
+the rates of two Poisson processes that compete over a transcript, which matches
+the story of geometrically distributed burst sizes. In other words, in our 
+bursty promoter model we can think of the parameter $\theta$ as determining one
+competing process to end the burst and $(1 - \theta)$ as a process wanting to
+continue the burst. So after taking this limit, on timescales slow compared to
+$v_1$ and $d_0$, it appears that transcription events fire at rate $v_0$ and
+produce a geometrically distributed burst of translation of mean size $v_1/d_0$,
+which intuitively matches the story we have told above for mRNA with variables
+relabeled.
 
-To verify this intuitively conjectured mapping between our problem and
-the solution in~\cite{Shahrezaei2008}, we continue with a careful
-solution for the mRNA distribution using probability generating functions,
-following the ideas sketched in~\cite{Shahrezaei2008}.
-It is natural to nondimensionalize rates in the problem by $\gamma$,
-or equivalently, this amounts to measuring time in units of $\gamma^{-1}$.
-We are also only interested in steady state, so we set the time
-derivatives to zero, giving
+\begin{figure}
+\centering
+\includegraphics{../figures/si/figS0X_Shahrezaei_promoter.pdf}
+\caption{\textbf{Schematic of three-stage promoter from~\cite{Shahrezaei2008}.}
+Adapted from Shahrezaei \& Swain~\cite{Shahrezaei2008}. In their paper they
+derive a closed form solution for the protein distribution. Our two-state bursty
+promoter at the mRNA level can be mapped into their solution with some
+relabeling.}
+\label{fig:shahrezaei}
+\end{figure}
+
+To verify this intuitively conjectured mapping between our problem and the
+solution in~\cite{Shahrezaei2008}, we continue with a careful solution for the
+mRNA distribution using probability generating functions, following the ideas
+sketched in~\cite{Shahrezaei2008}. It is natural to nondimensionalize rates in
+the problem by $\gamma$, or equivalently, this amounts to measuring time in
+units of $\gamma^{-1}$. We are also only interested in steady state, so we set
+the time derivatives to zero, giving
 \begin{align}
@@
 \end{align}
+where for convenience we kept the same notation for all rates, but these are
+now expressed in units of mean mRNA lifetime $\gamma^{-1}$.
 
-The probability generating function is defined as before in the
-constitutive case, except now we must introduce a generating function
-for each promoter state,
+The probability generating function is defined as before in the constitutive
+case, except now we must introduce a generating function for each promoter
+state,
 \begin{align}
@@
 \end{align}
-Our real objective is the generating function $f(z)$ that generates
-the mRNA distribution $p(m)$, independent of what state the promoter is in.
-But since $p(m) = p_A(m) + p_R(m)$, it follows too that $f(z) = f_A(z) + f_R(z)$.
+Our real objective is the generating function $f(z)$ that generates the mRNA
+distribution $p(m)$, independent of what state the promoter is in. But since
+$p(m) = p_A(m) + p_R(m)$, it follows too that $f(z) = f_A(z) + f_R(z)$.
 
-So as before we multiply both equations by $z^m$ and sum over all $m$.
-Each individual term transforms exactly as did an analogous term in
-the constitutive case, so the coupled ODEs for the generating functions read
+As before we multiply both equations by $z^m$ and sum over all $m$. Each
+individual term transforms exactly as did an analogous term in the constitutive
+case, so the coupled ODEs for the generating functions read
 \begin{align}
@@
 \end{align}
-and after changing variables $\xi = 1 - \theta$ as before
-and rearranging, we have
+and after changing variables $\xi = 1 - \theta$ as before and rearranging, we
+have
 \begin{align}
@@
 \end{align}
-We can transform this problem from two coupled first-order ODEs to a
-single second-order ODE by solving for $f_A$ in the first and plugging
-into the second, giving
+We can transform this problem from two coupled first-order ODEs to a single
+second-order ODE by solving for $f_A$ in the first and plugging into the second,
+giving
 \begin{align}
@@
 \end{align}
-where, to reduce notational clutter, we have dropped the explicit $z$
-dependence of $f_A$ and $f_R$. Simplifying we have
+where, to reduce notational clutter, we have dropped the explicit $z$ dependence
+of $f_A$ and $f_R$. Simplifying we have
 \begin{align}
@@
 \end{align}
-This can be recognized as the hypergeometric differential equation,
-with singularities at $z=1$, $z=\xi^{-1}$, and $z=\infty$. The latter
-can be verified by a change of variables from $z$ to $x=1/z$,
-being careful with the chain rule, and noting that $z=\infty$ is a
-singular point if and only if $x=1/z=0$ is a singular point.
+This can be recognized as the hypergeometric differential equation, with
+singularities at $z=1$, $z=\xi^{-1}$, and $z=\infty$. The latter can be verified
+by a change of variables from $z$ to $x=1/z$, being careful with the chain rule,
+and noting that $z=\infty$ is a singular point if and only if $x=1/z=0$ is a
+singular point.
 
 The standard form of the hypergeometric differential equation has its
-singularities at 0, 1, and $\infty$, so to take advantage of the
-standard form solutions to this ODE, we first need to transform
-variables to put it into a standard form.
-However, this is subtle. While any such transformation should work in
-principle, the solutions are expressed most simply in the neighborhood
-of $z=0$, but the normalization condition that we need to enforce
-corresponds to $z=1$. The easiest path, therefore, is to find a change
-of variables that maps
-1 to 0, $\infty$ to $\infty$, and $\xi^{-1}$ to 1.
-This is most intuitively done in two steps.
+singularities at 0, 1, and $\infty$, so to take advantage of the standard form
+solutions to this ODE, we first need to transform variables to put it into a
+standard form. However, this is subtle. While any such transformation should
+work in principle, the solutions are expressed most simply in the neighborhood
+of $z=0$, but the normalization condition that we need to enforce corresponds to
+$z=1$. The easiest path, therefore, is to find a change of variables that maps 1
+to 0, $\infty$ to $\infty$, and $\xi^{-1}$ to 1. This is most intuitively done
+in two steps.
 
@@
 \end{align}
-Now two singularities are at $v=0$ and $v=\infty$.
-The third is determined by $(1+v)\xi -1 = 0$, or $v=\xi^{-1} - 1$.
-We want another variable change that maps this third singularity to 1
-(without moving 0 or infinity). Changing variables again to
-$w=\frac{v}{\xi^{-1} - 1} = \frac{\xi}{1-\xi} v$ fits the bill.
-In other words, the combined change of variables
+Now two singularities are at $v=0$ and $v=\infty$. The third is determined by
+$(1+v)\xi -1 = 0$, or $v=\xi^{-1} - 1$. We want another variable change that
+maps this third singularity to 1 (without moving 0 or infinity). Changing
+variables again to $w=\frac{v}{\xi^{-1} - 1} = \frac{\xi}{1-\xi} v$ fits the
+bill. In other words, the combined change of variables
 \begin{align}
@@
 \end{align}
-maps $z = \{1, \xi^{-1}, \infty\}$ to $w =\{0, 1, \infty\}$ as desired.
-Plugging in, being mindful of the chain rule and noting
+maps $z = \{1, \xi^{-1}, \infty\}$ to $w =\{0, 1, \infty\}$ as desired. Plugging
+in, being mindful of the chain rule and noting
 $(1 + v)\xi - 1 = (1 - \xi)(w - 1)$ gives
@@
 \end{align}
-This is close to the standard form of the hypergeometric differential
-equation, and some cancellation and rearrangement gives
+This is close to the standard form of the hypergeometric differential equation,
+and some cancellation and rearrangement gives
 \begin{align}
@@
 \end{align}
-which is the standard form. From this we can read off the solution
-in terms of hypergeometric functions ${_2F_1}$ from
-any standard source, e.g.~\cite{Abramowitz1964}, and identify the
-conventional parameters in terms of our model parameters.
-We want the general solution in the neighborhood of $w=0$ ($z=1$),
-which for a homogeneous linear second order ODE must be a sum of two
-linearly independent solutions.
-More precisely,
+which is the standard form. From this we can read off the solution in terms of
+hypergeometric functions ${_2F_1}$ from any standard source,
+e.g.~\cite{Abramowitz1964}, and identify the conventional parameters in terms of
+our model parameters. We want the general solution in the neighborhood of $w=0$
+($z=1$), which for a homogeneous linear second order ODE must be a sum of two
+linearly independent solutions. More precisely, 
+\mrm{the use of variable $\gamma$ here is confusing. Is it okay to change it to
+$\delta$ or something else in order to avoid associating it with the mRNA
+degradation rate?}
 \begin{align}
@@
 \end{align}
-and constants $C_1$ and $C_2$ to be set by boundary conditions.
-Solving for $\alpha$ and $\beta$, we find\footnote{
+and constants $C_1$ and $C_2$ to be set by boundary conditions. Solving for
+$\alpha$ and $\beta$, we find
+\footnote{
 Note that $\alpha$ and $\beta$ are interchangeable in the definition of
@@
 \end{align}
-Finally we are in sight of the original goal. We can generate the
-steady-state probability distribution of interest by
-differentiating the generating function,
+Finally we are in sight of the original goal. We can generate the steady-state
+probability distribution of interest by differentiating the generating function,
 \begin{align}
@@
 \end{align}
-which follows easily from its definition. Some contemplation reveals
-that repeated application of the derivative rule used above will produce
-products of the form $\alpha(\alpha+1)(\alpha+2)\cdots(\alpha+m-1)$ in
-the expression for $p(m)$ and similarly for $\beta$ and $\gamma$. These
-resemble ratios of factorials, but since $\alpha$, $\beta$, and $\gamma$
-are not necessarily integer, we should express the ratios using gamma
-functions instead. More precisely, one finds
+which follows easily from its definition. Some contemplation reveals that
+repeated application of the derivative rule used above will produce products of
+the form $\alpha(\alpha+1)(\alpha+2)\cdots(\alpha+m-1)$ in the expression for
+$p(m)$ and similarly for $\beta$ and $\gamma$. These resemble ratios of
+factorials, but since $\alpha$, $\beta$, and $\gamma$ are not necessarily
+integer, we should express the ratios using gamma functions instead. More
+precisely, one finds
 \begin{align}
@@
 
-\subsubsection{Moments}
-We can verify the correctness of~\eq{eq:p_m_bursty+rep_appdx}
-by comparison with Gillespie simulation.
-We can also easily compute moments of the distribution from
-the generating function, and check these against raw moments computed
+\subsubsection{Moments of two-state bursty promoter mRNA distribution}
+
+We can verify the correctness of~\eq{eq:p_m_bursty+rep_appdx} by comparison with
+Gillespie simulation. We can also easily compute moments of the distribution
+from the generating function, and check these against raw moments computed
 directly from the master equation.
 
-The $j$-th raw moment of the distribution can be computed from the
-generating function from the formula
+The $j$-th raw moment of the distribution can be computed from the generating
+function from the formula
 \begin{align}
@@
 \end{align}
-Note that \textit{probabilities} are generated by differentiating
-and evaluating at $z=0$, while \textit{moments} are generated by
-differentiating and evaluating at $z=1$. This also aligns with the
-normalization constraint, which can be thought of as the zeroth raw moment.
+Note that \textit{probabilities} are generated by differentiating and evaluating
+at $z=0$, while \textit{moments} are generated by differentiating and evaluating
+at $z=1$. This also aligns with the normalization constraint, which can be
+thought of as the zeroth raw moment.
 
@@
 \end{align}
-which is intuitive and appealingly simple:
-the mean mRNA is just the mean mRNA in the absense of repression
-($k_i b$, the burst rate time mean burst size)
+which is intuitive and appealingly simple: the mean mRNA is just the mean mRNA
+in the absence of repression ($k_i b$, the burst rate time mean burst size)
 times the fraction of time the repressor is bound, $k_R^-/(k_R^+ + k_R^-)$.
@@
 \mmnote{Still to writeup: moment derivation direct from CME to compare}
+\mrm{stopped reading here}
 

### 2020-05-29 doc/appendix_02_bursty.tex
MRM going through RP comments on bursty dist. derivation

@@
 that this model is rich enough that it can capture the expression pattern of
-bacteria unregulated promoters regulated by the $\sigma_{70}$ sigma factor.
-Figure~\ref{figS1:bursty_one_state} shows two different schematic
-representations of the model. Figure~\ref{figS1:bursty_one_state}(A) shows the
-promoter cartoon model with burst initiation rate $k_i$, mRNA degradation rate
-$\gamma$, and mean burst size $b$. For our derivation of the chemical master
-equation we will focus more on Figure~\ref{figS1:bursty_one_state}(B). This
-representation is intended to highlight that bursty gene expression allows
-transitions between mRNA count $m$ and $m'$ even with $m - m' > 1$.
+bacteria constitutive promoters. Figure~\ref{figS1:bursty_one_state} shows two
+different schematic representations of the model.
+Figure~\ref{figS1:bursty_one_state}(A) shows the promoter cartoon model with
+burst initiation rate $k_i$, mRNA degradation rate $\gamma$, and mean burst size
+$b$. For our derivation of the chemical master equation we will focus more on
+Figure~\ref{figS1:bursty_one_state}(B). This representation is intended to
+highlight that bursty gene expression allows transitions between mRNA count $m$
+and $m'$ even with $m - m' > 1$.
 
@@
 The model in (A) allows for transitions of $> 1$ mRNA counts with probability
-$G_{m-m'}$.}
+$G_{m-m'}$, where the state jumps from having $m'$ mRNA to having $m$ mRNA in a
+single burst of gene expression.}
 \label{figS1:bursty_one_state}
@@
 
-To derive the master equation we begin by thinking the possible state
+To derive the master equation we begin by considering the possible state
 transitions to ``enter'' state $m$. There are two possible paths to jump from an
@@
 where we indicate $G_{m'-m}$ as the probability of having a burst of size
-$m'-m$. We suggestively use the letter $G$ as we will assume that these bursts
-sizes are geometrically distributed with parameter $\theta$. This is written as
+$m'-m$, i.e. when the number of mRNAs jump from $m$ to $m' > m$ due to a single
+mRNA transcription burst. We suggestively use the letter $G$ as we will assume
+that these bursts sizes are geometrically distributed with parameter $\theta$.
+This is written as
 \begin{equation}
@@
 \end{equation}
-What this implies is that for a geometrically distributed burst size we have a
-mean burst size $b$ of the form
+In Section~\ref{sec:beyond_means} of the main text we derive this functional
+form for the burst size distribution. An intuitive way to think about it is that
+for transcription initiation events that take place instantaneously there are
+two competing possibilities: Producing another mRNA with probability $(1 -
+\theta)$, or ending the burst with probability $\theta$. What this implies is
+that for a geometrically distributed burst size we have a mean burst size $b$ of
+the form
 \begin{equation}
@@
 \end{equation}
-Given the normalization of the geometric distribution, adding the probability of
-all possible burst sizes -- including size zero since we re-indexed the sum --
-allows us to write
+Given the normalization constraint of the geometric distribution, adding the
+probability of all possible burst sizes -- including size zero since we
+re-indexed the sum -- allows us to write
 \begin{equation}
@@
 where $z$ is just a dummy variable that will help us later on to obtain the
-moments of the distribution. Applying $F(z, t)$ on both sides of 
-Eq.~\ref{eq:si_master_unreg} results in
+moments of the distribution. Let us now multiply both sides of
+Eq.~\ref{eq:si_master_unreg} by $z^m$ and sum over all $m$
 \begin{equation}
@@
 \end{align}
+
+\begin{figure}[h!]
+\centering
+\includegraphics{../figures/si/figS0X_reindex_sum.pdf}
+\caption{\textbf{Reindexing double sum.} Schematic for how to proceed with the
+reindexing of the sum $\sum_{m=0}^\infty \sum_{m'=0}^m$. The trick consists on
+noticing that this double sum adds up all of the $(m, m')$ pairs in the lower
+triangle part of the $m, m'$ plane.}
+\label{figS2:sum_reindex}
+\end{figure}
+
 The third term in Eq.~\ref{eq:si_generating_01} is the most trouble. The trick
@@
 \end{equation}
-This results in
+To see the logic of the sum we point the reader to
+Figure~\ref{figS2:sum_reindex}. The key is to notice that the double sum
+$\sum_{m=0}^\infty \sum_{m'=0}^m$ is adding all possible pairs $(m, m')$ in the
+lower triangle, so we can add the terms vertically as the original sum
+indexing suggests, i.e.
+\begin{equation}
+\sum_{m=0}^{\infty} \sum_{m'=0}^{m} x_{(m, m')}= 
+x_{(0, 0)} + x_{(1, 0)} + x_{(1, 1)} + x_{(2, 0)} + x_{(2, 1)} + x_{(2, 2)} + 
+\ldots,
+\end{equation}
+where the variable $x$ is just a placeholder to indicate the order in which the
+sum is taking place. But we can also add the terms horizontally as
+\begin{equation}
+\sum_{m'=0}^{\infty} \sum_{m=m'}^{\infty} x_{(m, m')} =
+x_{(0, 0)} + x_{(1, 0)} + x_{(2, 0)} + \ldots + x_{(1,1)} + x_{(2, 1)} + \ldots,
+\end{equation}
+which still adds all of the lower triangle terms. Applying this reindexing
+results in
 \begin{align}
@@
 \end{align}
-We know that $1 - \theta \in (0, 1)$, and that $z$ is a dummy variable that we
-usually evaluate as $z=1$. With these two facts we can safely state that
-$| z (1 - \theta) | < 1$. Defining $n \equiv m - m'$ we rewrite the last sum in 
+
+The next step requires us to look slightly ahead into what we expect to obtain.
+We are working on deriving an equation for the generating function $F(z, \tau)$
+that when solved will allow us to compute what we care about, i.e. the
+probability function $p(m, \tau)$. Upon finding the function for $F(z, \tau)$,
+we will recover this probability distribution by evaluating $F(z=1, \tau)$ in
+other cases we can evaluate $F(z=0, \tau)$ to recover instead the moments of the
+distribution. The point here is that we know that when all the dust settles we
+will evaluate $z$ to be less than one \mrm{someone needs to fact check me}.
+Furthermore, we know that the parameter of the geometric distribution $\theta$
+must be between zero and one. With these two facts we can safely state that $| z
+(1 - \theta) | < 1$. Defining $n \equiv m - m'$ we rewrite the last sum in
 Eq.~\ref{eq:si_generating_02} as
@@
 
-\subsubsection{Stead-state}
+\subsubsection{Steady-state}
 
@@
         = \left(\frac{\theta}{1 - z(1-\theta)}\right)^\lambda,
+\label{eq:gen_fn}
 \end{align}
-which is exactly the negative binomial's generating function. Now we may recover
-the mRNA probability distribution of interest by differentiating the generating
-function, or by consulting a standard reference\mmnote{maybe van Kampen, check},
-from which we find that\mrm{I have a series of notes where I go through some 
-of the derivatives to find the general pattern, arriving to this exact result 
-by ``scratch and sniff'' without the black magic. Should I add it to this 
-section?}
+
+\subsubsection{Recovering the steady-state probability distribution}
+
+To obtain the steady state mRNA distribution $p(m)$ we are aiming for we need to
+extract it from the generating function
+\begin{equation}
+F(z) = \sum_m z^m p(m).
+\end{equation}
+Taking a derivative with respect to $z$ results in
+\begin{equation}
+{d F(z) \over dz} = \sum_m m z^{m - 1} p(m).
+\end{equation}
+Setting $z = 0$ leaves one term in the sum when $m = 1$ 
+\begin{equation}
+\left.\frac{d F(z)}{d z}\right|_{z=0} = 
+\left(0 \cdot 0^{-1} \cdot p(0) 
++ 1 \cdot 0^0 \cdot p(1) 
++ 2 \cdot 0^1 \cdot p(2)
++ \cdots \right) = p(1),
+\end{equation}
+since in the limit $\lim_{x \rightarrow 0^+} x^x = 1$. A second derivative of
+the generating function would result in
+\begin{equation}
+\frac{d^{2} F(z)}{d z^{2}} = \sum_{m=0}^{\infty} m(m-1) z^{m-2} p(m).
+\end{equation}
+Again evaluating at $z = 0$ gives 
+\begin{equation}
+\left.\frac{d^{2} F(z)}{d z}\right|_{z=0} = 2 p(z).
+\end{equation}
+In general any $p(m)$ is obtained from the generating function as
+\begin{equation}
+p(m) = {1 \over m!} \left. {d^m F(z) \over dz} \right\vert_{z=0}.
+\label{eq:prob_from_gen}
+\end{equation}
+
+Let's now look at the general form of the derivative for our generating function
+in Eq.~\ref{eq:gen_fn}. For $p(0)$ we simply evaluate $F(z=0)$ directly, 
+obtaining
+\begin{equation}
+p(0) = F(z=0) = \theta^{\lambda}.
+\end{equation}
+The first derivative results in
+\begin{equation}
+\begin{aligned}
+\frac{d F(z)}{d z} &=\theta^{\lambda} \frac{d}{d z}(1-z(1-\theta))^{-\lambda} \\
+&=\theta^{\lambda}\left[-\lambda(1-z(1-f))^{-\lambda-1} \cdot(\theta-1)\right]\\
+&=\theta^{\lambda}\left[\lambda(1-z(1-\theta))^{-\lambda-1}(1-\theta)\right].
+\end{aligned}
+\end{equation}
+Evaluating this at $z=0$ as required to get $p(1)$ gives
+\begin{equation}
+\left.\frac{d F(z)}{d z}\right|_{z=0}=\theta^{\lambda} \lambda(1-\theta)
+\end{equation}
+For the second derivative we find
+\begin{equation}
+\frac{d^{2} F(z)}{d z^{2}} = \theta^{\lambda}
+\left[\lambda(\lambda+1)(1-z(1-\theta))^{-\lambda-2}(1-\theta)^{2}\right].
+\end{equation}
+Again evaluating $z = 0$ gives
+\begin{equation}
+\left.\frac{d^{2} F(z)}{d z^{2}}\right|_{z=0} = 
+\theta^{\lambda} \lambda(\lambda+1)(1-\theta)^{2}.
+\end{equation}
+Let's go for one more derivative to see the pattern. The third derivative of the
+generating function gives
+\begin{equation}
+\frac{d^{3} F(z)}{d z^{3}} = 
+\theta^{\lambda} 
+\left[\lambda(\lambda+1)
+(\lambda+2)(1-z(1-\theta))^{-\lambda-3}(1-\theta)^{3}\right],
+\end{equation}
+which again we evaluate at $z=0$
+\begin{equation}
+\left.\frac{d^{3} F(z)}{d z^{3}}\right|_{z=1} =
+\theta^{\lambda}\left[\lambda(\lambda+1)(\lambda+2)(1-\theta)^{3}\right].
+\end{equation}
+If $\lambda$ was an integer we could write this as
+\begin{equation}
+\left.\frac{d^{3} F(z)}{d z^{3}}\right|_{z=0} = 
+\frac{(\lambda+2) !}{(\lambda-1) !} \theta^{\lambda}(1-\theta)^{3}.
+\end{equation}
+Since $\lambda$ might not be an integer we can write this using Gamma functions
+as
+\begin{equation}
+\left.\frac{d^{3} F(z)}{d z^{3}}\right|_{z=0} = 
+\frac{\Gamma(\lambda+3)}{\Gamma(\lambda)} \theta^{\lambda}(1-\theta)^{3}.
+\end{equation}
+Generalizing the pattern we then have that the $m$-th derivative takes the form
+\begin{equation}
+\left.\frac{d^{m} F(z)}{d z^{m}}\right|_{z=0} =
+\frac{\Gamma(\lambda+m)}{\Gamma(\lambda)} \theta^{\lambda}(1-\theta)^{m}.
+\end{equation}
+With this result we can use Eq.~\ref{eq:prob_from_gen} to obtain the desired
+steady-state probability distribution function
 \begin{equation}

### 2020-05-29 doc/appendix_03_bayesian_inference.tex
MRM pass through MM section on inference for unreg promoters for SI

--- 2020-05-28 doc/appendix_01_bayesian_inference.tex
@@
 the chemical master equations we wrote down describe the time evolutions of the
-mRNA probability distribution. So all our terms $P(\vec{\theta} \mid D)$ will 
-be given by the steady-state solution of the corresponding chemical master 
-equation in question. This is rather convenient as we do not have to worry about
-adding a statistical model on top of our model to describe deviations from the
-predictions; our models themselves focus on predicting such variations from 
-cell count to cell count.
+mRNA probability distribution. So all our terms $P(\vec{\theta} \mid D)$ will be
+given by the steady-state solution of the corresponding chemical master equation
+in question. This is rather convenient as we do not have to worry about adding a
+statistical model on top of our model to describe deviations from the
+predictions; our models themselves focus on predicting such variations from cell
+count to cell count.
 
@@
 The different models explored in this work embraced different levels of
-coarse-graining that resulted in a diverse set of number of parameters. For each
-of these model configurations Bayes theorem demands from us to represent our 
-preconceptions on the possible parameter values in the form of the prior 
-$P(\vec{\theta})$. Throughout this work for models with $> 1$ parameter we 
+coarse-graining that resulted in a diverse number of parameters for different
+models. For each of these model configurations Bayes theorem demands from us to
+represent our preconceptions on the possible parameter values in the form of the
+prior $P(\vec{\theta})$. Throughout this work for models with $> 1$ parameter we
 assign independent priors to each of the parameters; this is
@@
 use of our own expertise, many times in the form of order-of-magnitude
-estimates, to write down prior distributions for our parameters.
+estimates, to write down prior distributions for our parameters. In summary then
+we don’t want to be maximally uninformative because we aren’t completely
+uninformed.
+
+\mrm{Still need to write something about the intuition of being data-rich and 
+how then the prior matters little.}
 
@@
 
-To proceed we need to specify a prior. In this case we are extremely data-rich,
-as the dataset from Jones et.\ al~\cite{Jones2014} has of order 1000-3000
-single-cell measurements for each promoter, so our choice of prior matters
-little here, as long as it is sufficiently broad. A convenient choice for our
-problem is to use a \textit{conjugate} prior. A conjugate prior is a special
-prior that causes the posterior to have the same functional form as the prior,
-simply with updated model parameters. This makes calculations analytically
-tractable and also offers a nice interpretation of the inference procedure as
-updating our knowledge about the model parameters. This makes conjugate priors
-very useful when they exist. The caveat is that conjugate priors only exist for
-a very limited number of likelihoods, mostly with only one or two model
-parameters, so in almost all other Bayesian inference problems, we must tackle
-the posterior numerically.
+Having introduced the ideas behind Bayesian inference we are ready to apply the
+theoretical machinery to our non-equilibrium models. In particular in this
+section we will focus on model 1 and model 5 in
+Figure~\ref{fig2:constit_cartoons}(A). Model 1, the Poisson promoter, will help
+us build practical intuition into the implementation of the Bayesian inference
+pipeline as we noted in Section~\ref{sec:beyond_means} of the main text that
+this model cannot be reconciled with experimental data from observables such as
+the Fano factor. In other words, we acknowledge that this model is ``wrong,''
+but we still see value in going through the analysis since the simple nature of
+the model translates into a neat statistical analysis.
+
+\subsubsection{Model 1 - Poisson promoter}
+
+As specified in the main test, the mRNA steady-state distribution for model 1 in
+Figure~\ref{fig2:constit_cartoons}(A) is Poisson with parameter $\lambda$.
+Throughout this Appendix we will appeal to the convenient notation for
+probability distributions of the form
+\begin{equation}
+m \sim \text{Poisson}(\lambda),
+\end{equation}
+where the simbol ``$\sim$'' can be read as \textit{is distributed according to}.
+So the previous equation can be read as: the mRNA copy number $m$ is distributed
+according to a Poisson distribution with parameter $\lambda$. Our objective then
+is to compute the posterior probability distribution $P(\lambda \mid D)$, where,
+as in the main text, $D = \{ m_1, m_2, \ldots, m_N \}$ are the data consisting
+of single-cell mRNA counts. Since we can assume that each of the cells mRNA
+counts are independent of any other cells, our likelihood function $P(D \mid
+\lambda)$ consists of the product of $N$ Poisson distributions.
+
+To proceed with the inference problem we need to specify a prior. In this case
+we are extremely data-rich, as the dataset from Jones et.\ al~\cite{Jones2014}
+has of order 1000-3000 single-cell measurements for each promoter, so our choice
+of prior matters little here, as long as it is sufficiently broad. A convenient
+choice for our problem is to use a \textit{conjugate} prior. A conjugate prior
+is a special prior that causes the posterior to have the same functional form as
+the prior, simply with updated model parameters. This makes calculations
+analytically tractable and also offers a nice interpretation of the inference
+procedure as updating our knowledge about the model parameters. This makes
+conjugate priors very useful when they exist. The caveat is that conjugate
+priors only exist for a very limited number of likelihoods, mostly with only one
+or two model parameters, so in almost all other Bayesian inference problems, we
+must tackle the posterior numerically.
 
@@
 \sim \text{Normal}\left(18.7, 0.08\right),
+\label{eq:gauss_posterior}
 \end{equation}
@@
 
-This is not wrong, but it is not the full story. The model's posterior is
-tightly constrained, but is it a good generative model? In other words, does the
-model generate data that look similar to our actual data, and is it therefore
+This is not wrong, but it is not the full story. The model's posterior
+distribution is tightly constrained, but is it a good generative model? In other
+words, if we use the model to generate synthetic data in the computer does it
+generate data that look similar to our actual data, and is it therefore
 plausible that the model captures the important features of the data generating
@@
 
-The intuitive idea of posterior predictive checks is simple: \mmnote{Not sure if
-all this stuff is clearer by example, but then does it confuse the reader in
-generalizing to other models?}
+The intuitive idea of posterior predictive checks is simple: 
 \begin{enumerate}
@@
 already observed. Clearly if those data appear quite different, the model has a
-problem. Put another way, if we suppose the generative model is true, then the
-synthetic datasets we generate should resemble the actual observed data, and if
-not, it suggests the model is missing important features. All the data we
-consider in this work are 1D (distributions of mRNA counts over a population) so
-ECDFs are an excellent visual means of comparing synthetic and observed
-datasets. In general for higher dimensional datasets, much of the challenge is
-in merely designing good visualizations that can actually show if synthetic and
-observed data are similar or not.
+problem. Put another way, if we suppose the generative model is true, i.e. we
+claim that our model explains the process through which our observed
+experimental data was generated, then the synthetic datasets we generate should
+resemble the actual observed data. If this is not the case, it suggests the
+model is missing important features. All the data we consider in this work are
+1D (distributions of mRNA counts over a population) so empirical cumulative
+distribution functions ECDFs are an excellent visual means of comparing
+synthetic and observed datasets. In general for higher dimensional datasets,
+much of the challenge is in merely designing good visualizations that can
+actually show if synthetic and observed data are similar or not.
 
 For our example Poisson promoter model then, we merely draw many random numbers,
-say 1000, from the Gaussian posterior. For each one of those draws, we generate
-a dataset from the likelihood, i.e., we draw 2648 (the number of observed cells
-in the actual dataset) Poisson-distributed numbers for each of the 1000
-posterior draws, for a total of 2648000 samples from the posterior predictive
-distribution.
+say 1000, from the Gaussian posterior in Eq.~\ref{eq:gauss_posterior}. For each
+one of those draws, we generate a dataset from the likelihood, i.e., we draw
+2648 (the number of observed cells in the actual dataset) Poisson-distributed
+numbers for each of the 1000 posterior draws, for a total of 2648000 samples
+from the posterior predictive distribution.
 
@@
 visualization for 1D data is ECDFs of the quantiles, as shown for our Poisson
-model in~\fig{fig:constit_post_full}.
-In this example, the median for each possible mRNA
-count is shown as a dark green line, while shaded bands of lighter
-green contain 95\% of the posterior predictive samples.
-Plotting quantiles in this way gives us a sense of the range of data we might
-consider plausible, under the assumption that the model is true. In this case it
-is quite obvious that the observed data, plotted in orange, could not plausibly
-come from this Poisson generative model. The other model with a PPC plotted
-in~\fig{fig:constit_post_full}, bursty model 5 with
-a negative binomial steady-state mRNA
-distribution, looks like a good candidate by eye. We cover it in more detail
-later.
+model in~\fig{fig:constit_post_full}(B) in the main text.
 
 \subsubsection{Model 5 - Bursty promoter}
+
 Let us now consider the problem of parameter inference from FISH data for model
@@
         \left(\frac{b}{1+b}\right)^m,
-\label{eq:neg_bionom}
+\label{eq:si_neg_bionom}
 \end{equation}
 where $b$ is the mean burst size and $k_i$ is the burst rate nondimensionalized
-by the mRNA degradation rate $\gamma$. As sketched earlier, the story of this
-distribution is of geometrically-distributed bursts of mRNA, where the arrival
-of bursts is a Poisson process with rate $k_i$ and the mean size of a burst is
-$b$.
+by the mRNA degradation rate $\gamma$. As sketched earlier, we can intuitively
+think about this distribution through a simple story. The story of this
+distribution is that the promoter undergoes geometrically-distributed bursts of
+mRNA, where the arrival of bursts is a Poisson process with rate $k_i$ and the
+mean size of a burst is $b$.
 
@@
 \begin{equation}
-p(k_i, b \mid D) \propto p(D\mid k_i,b)p(k_i, b).
+p(k_i, b \mid D) \propto p(D\mid k_i,b)p(k_i, b),
 \end{equation}
-We already have our likelihood --the product of $N$ negative binomials as
-Eq.~\ref{eq:neg_bionom}--  so we only need to choose priors on $k_i$ and $b$.
-For the datasets from~\cite{Jones2014} that we are analyzing, as for the Poisson
-promoter model above we are still data-rich so the prior's influence remains
-weak, but not nearly as weak because the dimensionality of our model has
-increased from one to two.
+where the likelihood $p(D\mid k_i,b)$ is given by the product of $N$ negative
+binomials as in Eq.~\ref{eq:si_neg_bionom}. We only need to choose priors on
+$k_i$ and $b$. For the datasets from~\cite{Jones2014} that we are analyzing, as
+for the Poisson promoter model above we are still data-rich so the prior's
+influence remains weak, but not nearly as weak because the dimensionality of our
+model has increased from one to two.
 
-We follow the guidance of~\cite{Gelman2013}, Section 2.9 in
-opting for weakly-informative priors on $k_i$ and $b$ (conjugate
-priors do not exist for this problem), and we find
-``street-fighting estimates''~\cite{Mahajan2010} to be an ideal
-way of constructing such priors. The idea of weakly informative
-priors is to allow all remotely plausible values of model
-parameters while excluding the completely absurd or unphysical.
+We follow the guidance of~\cite{Gelman2013}, Section 2.9 in opting for
+weakly-informative priors on $k_i$ and $b$ (conjugate priors do not exist for
+this problem), and we find ``street-fighting estimates''~\cite{Mahajan2010} to
+be an ideal way of constructing such priors. The idea of weakly informative
+priors is to allow all remotely plausible values of model parameters while
+excluding the completely absurd or unphysical.
 
-Consider $k_i$. The fastest known bacterial promoters control
-rRNA genes and initiate transcripts no faster than $\sim
-1/\text{sec}$. It would be exceedingly strange if any of the
-constitutive promoters from~\cite{Jones2014} were stronger than
-that, so we can take that as an upper bound. For a lower bound,
-if transcripts are produced too rarely, there would be nothing to
-see with FISH. The datasets for each strain contain of order
-$10^3$ cells, and if the $\langle m \rangle = k_i b/\gamma
-\lesssim 10^{-2}$, then the total number of expected mRNA
-detections would be single-digits or less and we would have
-essentially no data on which to carry out inference. So assuming
-$b$ is not too different from 1, justified next, and an mRNA
-lifetime of $\gamma^{-1}\sim 3-5~\text{min}$, this gives us soft
-bounds on $k_i/\gamma$ of perhaps $10^{-2}$ and $3\times 10^1$.
+Consider $k_i$. Some of the strongest known bacterial promoters control rRNA
+genes and initiate transcripts no faster than $\sim 1/\text{sec}$. It would be
+exceedingly strange if any of the constitutive promoters from~\cite{Jones2014}
+were stronger than that, so we can take that as an upper bound. For a lower
+bound, if transcripts are produced too rarely, there would be nothing to see
+with FISH. The datasets for each strain contain of order $10^3$ cells, and if
+the $\langle m \rangle = k_i b/\gamma \lesssim 10^{-2}$, then the total number
+of expected mRNA detections would be single-digits or less and we would have
+essentially no data on which to carry out inference. So assuming $b$ is not too
+different from 1, justified next, and an mRNA lifetime of $\gamma^{-1}\sim
+3-5~\text{min}$, this gives us soft bounds on $k_i/\gamma$ of perhaps $10^{-2}$
+and $3\times 10^1$.
 
-Next consider mean burst size $b$. This parametrization of the
-geometric distribution allows bursts of size zero (which could
-representing aborted transcripts and initiations), but it would
-be quite strange for the mean burst size $b$ to be below
-$\sim10^{-1}$, for which nearly all bursts would be of size zero
-or one. For an upper bound, if transcripts are initiating at a
-rate somewhat slower than rRNA promoters, then it would probably
-take a time comparable to the lifetime of an mRNA to produce a
-burst larger than 10-20 transcripts, which would invalidate the
-approximation of the model that the duration of bursts are
-instantaneous compared to other timescales in the problem. So we
-will take soft bounds of $10^{-1}$ and $10^1$ for $b$.
+Next consider mean burst size $b$. This parametrization of the geometric
+distribution allows bursts of size zero (which could representing aborted
+transcripts and initiations), but it would be quite strange for the mean burst
+size $b$ to be below $\sim10^{-1}$, for which nearly all bursts would be of size
+zero or one. For an upper bound, if transcripts are initiating at a rate
+somewhat slower than rRNA promoters, then it would probably take a time
+comparable to the lifetime of an mRNA to produce a burst larger than 10-20
+transcripts, which would invalidate the approximation of the model that the
+duration of bursts are instantaneous compared to other timescales in the
+problem. So we will take soft bounds of $10^{-1}$ and $10^1$ for $b$.
 
-Note that the natural scale for these ``street-fighting estimates''
-was a log scale. This is commonly the case that our prior sense
-of reasonable and unreasonable parameters is set on a log scale.
-A natural way to enforce these soft bounds is therefore to use a
-lognormal prior distribution, with the soft bounds set $\pm2$
-standard deviations from the mean.
+Note that the natural scale for these ``street-fighting estimates'' was a log
+scale. This is commonly the case that our prior sense of reasonable and
+unreasonable parameters is set on a log scale. A natural way to enforce these
+soft bounds is therefore to use a lognormal prior distribution, with the soft
+bounds set $\pm2$ standard deviations from the mean.
 
@@
 \begin{split}
-\ln k_i \sim \text{Normal}(-0.5, 2)
+\ln k_i \sim \text{Normal}(-0.5, 2),
 \\
-\ln b \sim \text{Normal}(0.5, 1)
+\ln b \sim \text{Normal}(0.5, 1),
 \\
@@
 \end{equation}
-We carried out MCMC sampling on the posterior of this model, starting
-with the constitutive \textit{lacUV5} dataset from~\cite{Jones2014}.
-The resulting MCMC samples are shown in~\fig{fig:constit_post_full}(A).
-In constrast to the active/inactive constitutive model considered
-in~\cite{Razo-Mejia2020} (nonequilibrium model four
-in~\ref{fig2:constit_cartoons}), this model is well-identified
-with both parameters determined to a fractional uncertainty of 5-10\%.
-The strong correlation reflects the fact that their product sets
-the mean of the mRNA distribution, which is tightly constrained
-by the data, but there is weak ``sloppiness''~\cite{Transtrum2015}
-along a set of values with a similar product.
+Section ~\ref{section_04_bayesian_inference} in the main text details the
+results of applying this inference to the single-cell mRNA counts data.
 
-Flush with success having found the model's posterior to be
-well-identified, the next step was posterior predictive sampling.
-The only procedural difference compared to the Poisson promoter
-model is that, since we have no analytical posterior expression
-but only samples from the posterior, we generate a synthetic
-dataset (i.e., posterior predictive samples) for each posterior
-sample from the MCMC sampling. \fig{fig:constit_post_full}(B)
-shows the resulting predictive ECDF; similarly to the Poisson
-promoter model, the median posterior predictive ECDF is plotted
-as a dark blue line, while a lighter blue shaded region
-encompasses 95\% of the posterior predictive samples. Unlike the
-Poisson promoter model, the experimental ECDF closely tracks the
-posterior predictive ECDF, indicating this model is actually able
-to generate the observed data and increasing our confidence that
-this model is at least not wrong.
-
-\textit{lacUV5} is our primary target here, since it forms the
-core of all the simple repression constructs of~\cite{Jones2014}
-that we consider in Section~\ref{sec:rep_kinetics_inference}.
-Nevertheless, we thought it wise to apply our bursty promoter
-model to the other 17 constitutive promoters available in the
-FISH dataset from~\cite{Jones2014} as a test that the model is
-capturing the essential phenomenology. If the model fit well to
-all the different promoters, this would increase our confidence
-that it would serve well as a foundation for inferring repressor
-kinetics later in Section~\ref{sec:rep_kinetics_inference}.
-Conversely, were the model to fail on more than a couple of the
-other promoters, it would give us pause.
-
-\fig{fig:constit_post_full}(C) shows the results, plotting the
-posterior distribution from individually MCMC sampling all 18
-constitutive promoter datasets from~\cite{Jones2014}. To aid
-visualization, rather than plotting samples for each promoter's
-posterior as we did in \fig{fig:constit_post_full}(A), for each
-posterior we find and plot the curve that surrounds the 95\%
-highest probability density region, i.e., on a plot analogous
-to~(A), each contour would enclose approximately 95\% of the
-samples, and thus 95\% of the probability mass,
-of its posterior distribution.
-Posterior predictive ECDFs (not shown) display a similar level of
-agreement between data and predictive samples as for the bursty
-model with \textit{lacUV5} in \fig{fig:constit_post_full}(B).
-\mmnote{Might be worth showing some/all of these PPCs in
-supplement for completeness? We could do postage stamp sized diff
-plots and probably fit all 18 promoters on 1-2 pages.}
-
-One interesting feature from \fig{fig:constit_post_full}(C) is
-that burst rate varies far more widely, over a range of
-$\sim10^2$, than burst size, confined to a range of
-$\lesssim10^1$ (and with the exception of promoter 6, just a span
-of 3-5x). This suggests that $k_i$, not $b$, is the key dynamic
-variable that promoter binding site tunes.
-
-It is interesting to connect this observation to the work
-of~\cite{Brewster2012}, where these same 18 promoters were
-considered through the lens of the three-state equilibrium model
-and binding energies $\Delta\epsilon_P$ were predicted from an
-energy matrix model derived from~\cite{Kinney2010}. The connection
-is made simply by equating mean mRNA levels in both theories,
-\mmnote{derivation is hasty, needs elaboration}
-\begin{equation}
-\langle m \rangle = \frac{k_i b}{\gamma}
-        = \frac{r}{\gamma}
-        \frac{\frac{P}{N_{NS}}\exp(-\beta\Delta\epsilon_P)}
-                {1+\frac{P}{N_{NS}}\exp(-\beta\Delta\epsilon_P)},
-\end{equation}
-and taking the weak-promoter limit in the equilibrium model,
-\begin{equation}
-\langle m \rangle = \frac{k_i b}{\gamma}
-        = \frac{r}{\gamma} \frac{P}{N_{NS}}\exp(-\beta\Delta\epsilon_P),
-\end{equation}
-valid for all the binding energies considered here.
-
-How are the two coarse-grainings related? The only association that makes dimensional sense and produces the correct order-of-magnitude for the known parameters is to take
-\begin{equation}
-\frac{k_i}{\gamma} = \exp(-\beta\Delta\epsilon_P)
-\label{eq:bursty_equil_corresp1_appdx}
-\end{equation}
-and
-\begin{equation}
-b = \frac{r}{\gamma}\frac{P}{N_{NS}}.
-\label{eq:bursty_equil_corresp2_appdx}
-\end{equation}
-\fig{fig:constit_post_full}(D) shows that this linear scaling
-between $\ln k_i$ and $-\beta\Delta\epsilon_P$ is approximately
-true for all 18 constitutive promoters considered. The intercept
-is an uninteresting function of the unknown model parameters, so
-we only consider the slope of the scaling in
-\fig{fig:constit_post_full}(D).
-\mmnote{Wait, is that true or can we do something with the intercept, related to the burst size in some useful way? Needs more careful thought.}
-
-While the associations represented by
-\eq{eq:bursty_equil_corresp1_appdx} and \eq{eq:bursty_equil_corresp2_appdx}
-appear to be born out by the scaling $\ln k_i \sim
--\beta\Delta\epsilon_P$, we do not find the association of
-parameters encoded by
-\eqrange{eq:bursty_equil_corresp1_appdx}{eq:bursty_equil_corresp2_appdx}
-intuitive. If some deeper insight is waiting to be gleaned from it,
-we leave it as an open problem.
-
-\mmnote{Outline, still to cover:
-\begin{itemize}
-\item All 18 promoters. Hey the model still works on things other than UV5,
-cool. Interesting and somewhat counterintuitive scaling b/w burst rate and
-binding energy. Puzzle in comparing w/ Chong2014 supercoiling model. Are we
-missing important things?? Unclear, we leave as open question. Good enough for
-our purposes: posterior is identifiable, PPC is great, and of the models we've
-thought of it is unique in satisfying both.
-\end{itemize}
-}
-
-\mmnote{Discuss puzzling comparison w/ Chong2014: if supercoiling is the thing, why are my burst sizes all the same but burst rates vary?
-And why is the duty cycle of my promoter so low, ie., bursts so short?
-COmpare their fig 7E; their $\beta/\alpha$ is my $k^+/k^-$.
-They have one or two genes with very small $\beta/\alpha$, does the \textit{galK} locus just happen to be that, or is there a deeper disagreement?
-Hard to say w/o more data.}
-
-\subsection{Transcription factor kinetics can be inferred from FISH}
+\subsection{Bayesian inference on the simple-repression architecture}

### 2020-05-29 doc/appendix_03_bayesian_inference.tex
recover MM original text for inference on regulated promoters.  Still need to clean it up.

@@
 \subsection{Bayesian inference on the simple-repression architecture}
+
+Now that we have a satisfactory model in hand for constitutive
+promoters, we would like to return to the main thread: can we
+infer repressor binding and unbinding rates from mRNA distributions
+over a population of cells as measured by FISH in~\cite{Jones2014}?
+If so, how do these inferred rates compare to direct
+single-molecule measurements such as from~\cite{Hammar2014} and
+to binding energies such as from~\cite{Garcia2011a}
+and~\cite{Razo-Mejia2018}, which were inferred under the assumptions
+of the equilibrium models in Figure~\ref{fig1:means_cartoons}(B)?
+And can this comparison shed light on the unreasonable effectiveness
+of the equilibrium models?
+
+As we found in Section~\ref{sec:beyond_means}, for our purposes
+the ``right'' model of a constitutive promoter is the bursty
+picture, model five in Figure~\ref{fig:constit_cartoons}.
+Therefore our starting point here is the analogous model with
+repressor added, model 5 in Figure~\ref{fig1:means_cartoons}.
+For a given repressor binding site and copy number,
+this model has four rate parameters to be inferred:
+the repressor binding and unbinding rates $k_R^+$, and $k_R^-$,
+the initiation rate of bursts, $k_i$, and the mean burst size $b$
+(we nondimensionalize all of these by the mRNA degradation rate $\gamma$).
+
+Before showing the mathematical formulation of our statistical
+inference model, we would like to sketch the intuitive structure.
+The dataset from~\cite{Jones2014} we consider consists of
+single-cell mRNA FISH measurements of nine different conditions,
+spanning several combinations of three unique binding sites and
+four unique repressor copy numbers. We assume that the values of
+$k_i$ and $b$ are known, since we have already cleanly inferred
+them from constitutive promoter data, and further we assume that
+these values are the same across datasets with different
+repressor binding sites and copy numbers. We assume that there is
+one unbinding rate parameter for each repressor binding site,
+which is shared across conditions, and likewise one binding rate
+for each unique repressor copy number. This makes our model seven
+dimensional, or nine if one counts $k_i$ and $b$ as well.
+
+Formally now, denote the set of seven repressor rates to be inferred as
+\begin{equation}
+\vect{k} =\{k_{Oid}^-, k_{O1}^-, k_{O2}^-,
+k_{0.5}^+, k_{1}^+, k_{2}^+, k_{10}^+\}.
+\end{equation}
+Note the repressor copy numbers are not known directly, so we label their
+association rates by the concentration of a small molecule inducer aTc
+which indirectly controls the transcription of $\textit{lacI}$.
+Also note that the authors of~\cite{Jones2014} report estimates
+of LacI copy number per cell rather than direct measurements.
+However, these estimates were made \textit{assuming} the validity
+of the equilibrium models in Figure~\ref{fig1:means_cartoons},
+and since \textit{testing} these models is our goal, we will make
+no assumptions about the LacI copy number for given aTc concentrations.
+
+Bayes theorem reads simply
+\begin{equation}
+p(\vect{k}, k_i, b \mid D)
+\propto
+p(D \mid\vect{k}, k_i, b) p(\vect{k}, k_i, b),
+\end{equation}
+where $D$ is the set of all $N$ observed single-cell mRNA counts
+across the various conditions. We assume that individual single-cell
+measurements are independent so that the likelihood factorizes as
+\begin{equation}
+p(D \mid\vect{k}, k_i, b)
+= \prod_{j=1}^N p(m\mid \vect{k}, k_i, b)
+= \prod_{j=1}^N p(m\mid k_j^+, k_j^-, k_i, b)
+\end{equation}
+where $k_j^\pm$ represent the appropriate binding and unbinding
+rates for the $j$-th measured cell. The probability
+$p(m\mid k_j^+, k_j^-, k_i, b)$ appearing in the last expression
+is exactly Eq.~\ref{eq:p_m_bursty+rep_appdx}, the steady-state
+distribution for our bursty model with repression derived in
+Section~\ref{sec:gen_fcn_appdx}, which for completeness we reproduce here as
+\begin{equation}
+\begin{split}
+p(m \mid k_R^+, k_R^-, k_i, b) = & ~\frac{
+        \Gamma(\alpha + m)\Gamma(\beta + m)\Gamma(k_R^+ + k_R^-)
+        }
+        {
+        \Gamma(\alpha)\Gamma(\beta)\Gamma(k_R^+ + k_R^- + m)
+        }
+\frac{b^m}{m!}
+\\
+&\times {_2F_1}(\alpha+m, \beta+m, k_R^++k_R^-+m; -b).
+\end{split}
+\label{eq:p_m_bursty+rep}
+\end{equation}
+where $\alpha$ and $\beta$, defined for notational convenience, are
+\begin{align}
+\begin{split}
+\alpha &= \frac{1}{2}
+\left(k_i+k_R^-+k_R^+ + \sqrt{(k_i+k_R^-+k_R^+)^2 - 4k_i k_R^-}\right)
+\\
+\beta &= \frac{1}{2}
+\left(k_i+k_R^-+k_R^+ - \sqrt{(k_i+k_R^-+k_R^+)^2 - 4k_i k_R^-}\right).
+\end{split}
+\end{align}
+
+\mmnote{Probably SI most of this paragraph?}
+This likelihood is rather inscrutable. We did not find any of the
+known analytical approximations for ${_2F_1}$ terribly useful in
+gaining intuition, so we instead resorted to numerics. One insight we found
+\mmnote{a plot would probably help explain, maybe SI, need to explain better}
+was that for very strong or very weak repression, the distribution in
+Eq.~\ref{eq:p_m_bursty+rep} is well approximated by a negative binomial
+with burst size $b$ and burst rate $k_i$ equal to their
+constitutive \textit{lacUV5} values, except with $k_i$ multiplied
+by the fold-change $\left(1+k_R^+/k_R^-\right)^{-1}$.
+In other words, once again only the \textit{ratio} $k_R^+/k_R^-$
+was detectable. But for intermediate repression, the distribution
+was visibly broadened with Fano factor greater than $1+b$, the
+value for the corresponding constitutive case. This indicates
+that the repressor rates had left an imprint on the distribution,
+and perhaps intuitively, this intermediate regime occurs for values of
+$k_R^\pm$ comparable to the burst rate $k_i$. Put another way,
+if the repressor rates are much faster or much slower than $k_i$,
+then there is a timescale separation and effectively only one
+timescale remains, $k_i\left(1+k_R^+/k_R^-\right)^{-1}$.
+Only when all three rates in the problem are comparable does the
+mRNA distribution retain detectable information about them.
+
+Next we specify priors. As for the constitutive model, weakly
+informative lognormal priors are a natural choice for all our
+rates. We found that if the priors were too weak, our MCMC
+sampler would often become stuck in regions of parameter space
+with very low probability density, unable to move. We struck a
+balance in choosing our prior widths between helping the sampler
+run while simultaneously verifying that the marginal posteriors
+for each parameter were not artificially constrained or distorted
+by the presence of the prior. The only exception to this is the
+highly informative priors we placed on $k_i$ and $b$, since we
+have strong knowledge of them from our inference of constitutive
+promoters above.
+
+With priors and likelihood specified we may write down our complete generative model as
+\begin{equation}
+\begin{split}
+\log_{10}k_i &\sim \text{Normal}(0.725, 0.025)\\
+\log_{10}b   &\sim \text{Normal}(0.55, 0.025)\\
+\log_{10}k_{0.5}^+ &\sim \text{Normal}(-0.45, 0.3)\\
+\log_{10}k_{1}^+   &\sim \text{Normal}(0.6, 0.3)\\
+\log_{10}k_{2}^+   &\sim \text{Normal}(1.15, 0.3)\\
+\log_{10}k_{10}^+  &\sim \text{Normal}(1.5, 0.3)\\
+\log_{10}k_{Oid}^- &\sim \text{Normal}(-0.25, 0.3)\\
+\log_{10}k_{O1}^-  &\sim \text{Normal}(0.1, 0.3)\\
+\log_{10}k_{O2}^-  &\sim \text{Normal}(0.45, 0.3)\\
+m &\sim \text{Likelihood}(k_R^+, k_R^-, k_i, b),
+\end{split}
+\end{equation}
+where the likelihood is specified by Eq.~\ref{eq:p_m_bursty+rep}.
+\mmnote{Now that I've typed this out, this notation seems totally
+pointless here, especially when the likelihood is not a standard
+distribution. Maybe I should make a table or something to record
+the prior values, and maybe even that should go to SI?}
+We ran MCMC sampling on the full nine dimensional posterior
+specified by this generative model. To attempt to visualize this
+object, in Figure~\ref{fig4:repressed_post_full}(A) we plot
+several two-dimensional slices as contour plots, analogous to
+Figure~\ref{fig:constit_post_full}(C). Each of these nine slices
+corresponds to the $(k_R^+, k_R^-)$ pair of rates for one of the
+conditions from the dataset used to fit the model and gives a
+sense of the uncertainty and correlations in the posterior.
+
+\mmnote{Outline:
+\begin{itemize}
+\item Hey look it mostly works. With full model, 9D model is well identified,
+though most of the individual experiments unsurprisingly are not:
+only 3 of the 9 datasets were identifiable if fit in isolation
+and for the rest only the ratio of repressor rates is identifiable.
+Only by making assumptions that certain rates
+are equal across datasets was identifiability achieved.
+PPC is not perfect, it kinda misses some
+rep/op pairs but overall it's surprisingly predictive considering how strong
+were the assumptions that went into it.
+\item Do we chalk this up to experimental imperfections or is there interesting
+biology to be found in the disagreements? I'm inclined towards the former. Even
+just catching cells in truly apples-to-apples growth phases, or matching up
+parameters of imaging sessions weeks or months apart, is hardly trivial.
+\item The alternative is to refine our model and/or add on complexity, and it's
+not at all clear what additions to make to the model, especially how to produce
+the long tail of repressed strains over UV5. I see no way to get that without
+getting into complicated time-correlation/history effects, where
+transcription...rebounds?? After being repressed?? Sounds odd.
+\item Comparison between my inferred rates, equilibrium binding E, and
+single-molecule measurements. Not quite as slam-dunk as I'd hoped because of
+$\gamma$ dependence, but rates are w/in factor of 2 or better, seems quite good
+really! With exception of one Oid data data point, free energies are w/in about
+0.5 kT also. That's not massively larger than the repeatability threshold of
+fold-change measurements, at maybe 0.1 to 0.3 kT.
+\item Surprising that so many model parameters can be inferred from just a few
+distributions that, by eye, don't have strong features! Does require some
+assumptions, e.g., about equivalence of rates across experiments, but that seems
+quite reasonable.
+\end{itemize}
+}

### 2020-05-29 doc/section_02_means.tex
Many additions to main text Bayesian inference

@@
 
-\subsubsection{Summary of Results Ccross models for Fold-Changes in Mean
+\subsubsection{Summary of Results across models for Fold-Changes in Mean
 Expression}
@@
         = - \log(k_R^+/k_R^-),
+\label{eq:deltaFR_eq_noneq_equiv}
 \end{equation}

### 2020-05-29 doc/section_03_beyond_means.tex
Many additions to main text Bayesian inference

@@
 the variance in these distributions. To that end, we repeat the same pattern as
-in the previous section by examining the models one by one. \mrm{In particular
+in the previous section by examining the models one by one. In particular
 we will focus on the Fano factor, defined as the variance/mean. This metric
-serves as a discriminatory tool from the null model that the steady-state mRNA
-distribution must be Poisson, giving a Fano factor of one.}
+serves as a powerful discriminatory tool from the null model that the
+steady-state mRNA distribution must be Poisson, giving a Fano factor of one.
 

### 2020-05-29 doc/section_04_bayesian_inference.tex
Many additions to main text Bayesian inference

@@
 
-\textcolor{red}{In this section of the paper, we continue our program of providing
+In this section of the paper, we continue our program of providing
 one complete description of the entire broad sweep of studies
@@
 equilibrium and non-equilibrium perspectives that have been brought
-to bear on this problem.  To that end, we require a self-consistent means
-of determining the parameters in these different theoretical analyses and
-the current section does that, first by examining the constitutive promoter
-and then turning to the full case of simple repression. }
+to bear on this problem. From Section~\ref{section_02_means}, this reconciliation
+depends on a key quantitative question as codified by Eq.~\ref{eq:deltaFR_eq_noneq_equiv}: does the free energy of repressor binding, as described in the equilibrium models and indirectly inferred from gene expression measurements, agree with the corresponding values of repressor binding and unbinding rates in the non-equilibrium picture, measured or inferred more directly?
+In this section we tackle the statistical inference problem of inferring these repressor rates from mRNA FISH data.
+But before we can turn to the full case of simple repression, we must choose an appropriate model of the constitutive promoter and infer the parameter values in that model.
+This is the problem we address first.
 
@@
 model 5 in Figure~\ref{fig2:constit_cartoons}(A), a one-state bursty model of
-constitutive promoters, achieves a Goldilocks level of complexity. Does this
+constitutive promoters, achieves the right balance of complexity
+and simplicity, by allowing both Fano factor $\nu>1$, but also by
+remedying, by design, the problems of parameter degeneracy than
+model 4 in Figure~\ref{fig2:constit_cartoons} suffered. Does this
 stand up to closer scrutiny, namely, comparison to full mRNA distributions
@@
 
-It is interesting to connect this observation to the work
+\subsubsection{Connecting inferred parameters to prior work}
+It is interesting to connect these speculations on $k_i$ and $b$ to the work
 of~\cite{Brewster2012}, where these same 18 promoters were considered through
@@
 measurements}\label{sec:rep_kinetics_inference}
+\subsubsection{Building the model and performing parameter inference}
 
 Now that we have a satisfactory model in hand for constitutive promoters, we
-would like to return to the main thread: can we infer repressor binding and
+would like to return to the main thread: can we reconcile the equilibrium
+and nonequilibrium models by putting to the test
+Eq.~\ref{eq:deltaFR_eq_noneq_equiv}, the correspondence between indirectly
+inferred equilibrium binding energies and nonequilibrium kinetic rates?
+To make this comparison, is it possible to infer repressor binding and
 unbinding rates from mRNA distributions over a population of cells as measured
@@
 comparison shed light on the unreasonable effectiveness of the equilibrium
-models?
+models, for instance, in their application in~\cite{Chure2019, Chure2019a}?
 
@@
 model of a constitutive promoter is the bursty picture, model five in
-Figure~\ref{fig2:constit_cartoons}. Therefore our starting point here is the
+Figure~\ref{fig2:constit_cartoons}A. Therefore our starting point here is the
 analogous model with repressor added, model 5 in
@@
 regulation happens as the promoter is taken away from the transcriptionally
-active state when the promoter is bound. We assume that there is one unbinding
+active state when the promoter is bound by repressor. We assume that there is one unbinding
 rate parameter for each repressor binding site, which is shared across aTc
 inducer conditions, and likewise one binding rate for each unique repressor copy
-number. This makes our model nine dimensional, or nine if one counts $k_i$ and
-$b$ as well.
+number. This makes our model seven dimensional, or nine if one counts $k_i$ and
+$b$ as well. Note that we use only a subset of the datasets from
+Jones et.\ al.~\cite{Jones2014}, as discussed more in Appendix~\ref{sec:bayes}.
 
@@
 label their association rates by the concentration of a small molecule inducer
-aTc which indirectly controls the transcription of $\textit{lacI}$. Also note
+aTc which indirectly controls the transcription of $\textit{lacI}$, as
+described in detail in~\cite{Jones2014}. Also note
 that the authors of~\cite{Jones2014} report estimates of LacI copy number per
@@
 Figure~\ref{fig1:means_cartoons}, and since \textit{testing} these models is our
-goal, we will make no assumptions about the LacI copy number for given aTc
+present goal, it would be circular logic if we were to make the same assumption.
+Therefore we will make no assumptions about the LacI copy number for given aTc
 concentrations.
@@
 We ran MCMC sampling on the full nine dimensional posterior specified by this
-generative model. To attempt to visualize this object, in
+model. To attempt to visualize this object, in
 Figure~\ref{fig4:repressed_post_full}(A) we plot several two-dimensional slices
@@
 sense of the uncertainty and correlations in the posterior.
-\mrm{before invoking part (C) of the figure we are missing th explanation for 
-part (B). I don't fully understand how these computations were made, so I'll 
-let Jupiter detail this.}
+We note that the 95\% uncertainties of all the rates span about $\sim0.3$
+log units, or about a factor of two, with the exception of $k_{0.5}^+$, the
+association rate for the lowest repressor copy number which is somewhat larger.
+
+\mmnote{These next 2 paragraphs should be SI, I think. Leaving here for
+now to avoid a merge conflict w/ MRM.}
+
+We found that fitting a single operator/aTc concentration at a
+time with a single binding and unbinding rate did not yield a
+stable inference for most of the possible operator/aTc
+combinations. In other words, a single dataset could not
+independently resolve the binding and unbinding rates, only their
+ratio as set by the mean fold-change in
+Figure~\ref{fig1:means_cartoons}. Only by making the assumption
+of a single unique binding rate for each repressor copy number
+and a single unique unbinding rate for each binding site, as done
+in Figure~\ref{fig4:repressed_post_full}A, was it possible to
+independently resolve the rates and not merely their ratios.
+
+We also note that we found it necessary to exclude the very
+weakly and very strongly repressed datasets from Jones et.\
+al.~\cite{Jones2014}. In both cases there was, in a sense, not
+enough information in the distriubtions for our inference
+algorithm to extract, and their inclusion simply caused problems
+for the MCMC sampler without yielding any new insight. For the
+strongly repressed data (Oid, 10~ng/mL aTc), with >95\% of cells
+with zero mRNA, there was quite literally very little data from
+which to infer rates. And the weakly repressed data, all with the
+repressor binding site O3, had an unbinding rate so fast that the
+sampler essentially sampled from the prior; the likelihood had
+negligible influence, meaning the data was not informing the
+sampler in any meaninful way, so no inference was possible.
+
+\subsubsection{Comparison with prior measurements of repressor binding energies}
+Our primary goal in this work is to reconcile the kinetic
+and equilibrium pictures of simple repression.
+Towards this end we would like to compare the repressor kinetic rates we
+have inferred with the repressor binding energies inferred through
+multiple methods in~\cite{Garcia2011a} and~\cite{Razo-Mejia2018}.
+If the agreement is close, then it suggests that the equilibrium
+models are not wrong and the repressor binding energies they contain
+correspond to physically real free energies, not mere fit parameters.
+
+Figure~\ref{fig4:repressed_post_full}(B) shows both comparisons,
+with the top panel comparing to equilbrium binding energies
+and the bottom panel comparing to single-molecule measurements.
+
+First consider the top panel and its comparison between repressor
+kinetic rates and binding energies.
+As described in section~\ref{section_02_means}, if the
+equilibrium binding energies from~\cite{Garcia2011a}
+and~\cite{Razo-Mejia2018} indeed are the physically real binding
+energies we believe them to be, then they should be related to
+the repressor kinetic rates via
+Eq.~\ref{eq:deltaFR_eq_noneq_equiv}, which we reprint here,
+\begin{equation}
+\Delta F_R = \beta\Delta\varepsilon_R - \log(R/N_{NS})
+        = - \log(k_R^+/k_R^-).
+\label{eq:deltaFR_eq_noneq_equiv_repeat}
+\end{equation}
+Assuming mass action
+kinetics implies that $k_R^+$ is proportional to repressor copy
+number $R$, or more precisely, it can be thought of as repressor
+copy number times some intrinsic per molecule association rate.
+But since $R$ is not directly known for our data
+from~\cite{Jones2014}, we cannot use this equation directly.
+Instead we can consider two different repressor binding sites and
+compute the \textit{difference} in binding energy between them,
+since this difference depends only on the unbinding rates and not
+on the binding rates. This can be seen by evaluating
+Eq.~\ref{eq:deltaFR_eq_noneq_equiv_repeat} for two different
+repressor binding sites, labeled (1) and (2), but with the same
+repressor copy number $R$, and taking the difference to find
+\begin{equation}
+\Delta F_R^{(1)} - \Delta F_R^{(2)}
+= \beta\Delta\varepsilon_1 - \beta\Delta\varepsilon_2
+= - \log(k_R^+/k_1^-) + \log(k_R^+/k_2^-),
+\end{equation}
+or simply
+\begin{equation}
+\beta\Delta\varepsilon_1 - \beta\Delta\varepsilon_2
+= \log(k_2^-/k_1^-).
+\end{equation}
+The left and right hand sides of this equation are exactly the horizontal
+and vertical axes of the top panel of Figure~\ref{fig4:repressed_post_full}.
+Since we inferred rates for three repressor binding sites (O1, O2, and Oid),
+there are only two independent differences that can be constructed, and we
+arbitrarily chose to plot O2-O1 and O1-Oid in
+Figure~\ref{fig4:repressed_post_full}(B).
+Numerically, we compute values of $k_{O1}^- / k_{Oid}^-$ and
+$k_{O2}^- / k_{O1}^-$ directly from our full posterior samples,
+which conveniently provides uncertainties as well, as detailed in
+Appendix~\ref{sec:bayesian}.
+\mmnote{Explain in appendix how marginalization with sampling makes this
+``trivial'' to compute from samples of the joint posterior, b/c
+sampling makes marginalization trivial.}
+We then compare these ratios of rates to the binding energy differences
+$\Delta\epsilon_{O1} - \Delta\epsilon_{Oid}$ and from
+$\Delta\epsilon_{O2} - \Delta\epsilon_{O1}$ as computed from the values
+from both~\cite{Garcia2011a} and~\cite{Razo-Mejia2018}.
+Three of the four values are within $\sim0.5 k_BT$ of the diagonal
+representing perfect agreement, which is comparable to the
+$\sim\text{few}\times0.1 k_BT$ variability between the independent
+determinations of the same quantities between~\cite{Garcia2011a}
+and~\cite{Razo-Mejia2018}. The only outlier involves Oid measurements
+from~\cite{Razo-Mejia2018}, and as the authors of~\cite{Razo-Mejia2018} note,
+this is a difficult measurement of low fluorescence signal against
+high background since Oid represses so strongly. We are therefore
+inclined to regard the failure of this point to fall near the
+diagonal as a testament to the difficulty of the measurement and
+not as a failure of our theory.
+\mmnote{Does this seem a fair characterization?}
+
+On the whole then, we regard this as striking confirmation of the
+validity of the equilibrium models. Their lynchpin parameter is a
+phenomenological free energy of repressor binding that has
+previously only been inferred indirectly. Our result shows that
+the microscopic interpretation of this free energy, as the log of
+a ratio of transition rates, does indeed hold true.
+\mmnote{Does this deserve qualifiers, like ``to a good approximation,'' or ``as a first order model'' or something?
+I like the simple finality without, but I worry it's too brazen or subtlely misleading.}
+
+\subsubsection{Comparison with prior measurements of repressor kinetics}
+In the previous section we established the equivalence between
+the equilibrium models' binding energies and the repressor
+kinetics we infer from mRNA population distributions.
+But one might worry that the repressor rates we infer
+from mRNA distributions are \textit{themselves} merely
+fitting parameters and that they do not actually correspond to the
+binding and unbinding rates of the repressor in vivo.
+To verify that this is not the case, we next compare our kinetic rates with a
+different measurement of the same rates using a radically different method:
+single molecule measurements as performed in Hammar et.\ al.~\cite{Hammar2014}.
+This is plotted in the lower panel of Figure~\ref{fig4:repressed_post_full}B.
+
+Since we do not have access to repressor copy number for either
+the FISH data from~\cite{Jones2014} or the single-molecule data
+from~\cite{Hammar2014}, we cannot make an apples-to-apples
+comparison of association rates $k_R^+$. Further, while Hammar
+et.\ al.\ directly measure the dissociation rates $k_R^-$, our
+inference procedure returns $k_R^-/\gamma$, i.e., the repressor
+dissociation rate nondimensionalized by the mRNA degradation rate
+$\gamma$. So to make the comparison, we must make an assumption
+for the value of $\gamma$ since it was not directly measured. For
+most mRNAs in \textit{E.\ coli}, quoted values for the typical mRNA
+lifetime $\gamma^{-1}$ range between about 2.5~min~\cite{Chen2015}
+to 8~min. We chose $\gamma^{-1} = 3$~min and $\gamma^{-1} =
+5$~min as representative values and plot a comparison of
+$k_{O1}^-$ and $k_{Oid}^-$ from our inference with corresponding values
+reported in~\cite{Hammar2014} for both these choices of $\gamma$.
+
+The degree of quantitative agreement in the lower panel of
+Figure~\ref{fig4:repressed_post_full}B clearly depends on the
+precise choice of $\gamma$.
+Nevertheless we find this comparison very satisfying, when two wildly different approaches to a measurement of the same quantity yield broadly compatible results.
+We emphasize the agreement between our rates
+and the rates reported in~\cite{Hammar2014} for any reasonable
+$\gamma$: values differ by at most a factor of 2 and possibly
+agree to within our uncertainties of 10-20\%. From this we feel
+confident asserting that the parameters we have inferred from
+Jones et.\ al.'s FISH data do in fact correspond to repressor
+binding and unbinding rates, and therefore our conclusions on the
+agreement of these rates with binding energies
+from~\cite{Garcia2011a} and~\cite{Razo-Mejia2018} are valid.
+
+\subsubsection{Model checking}
+In Figure~\ref{fig:constit_post_full}B we saw that the simple
+Poisson model of a constitutive promoter, despite having a well
+behaved posterior, was clearly insufficient to describe the data.
+It behooves us to carry out a similar check for our model of
+simple repression, codified by Eq.~\ref{eq:p_m_bursty+rep} for
+the steady-state mRNA copy number distribution.
+As derived in Sections~\ref{section_02_means}
+and~\ref{sec:beyond_means}, we have compelling theoretical reasons
+to believe it is a good model, but if it nevertheless turned out
+to be badly contradicted by the data we should like to know.
+
+The details are deferred to Appendix~\ref{sec:bayesian}, and here
+we only attempt to summarize the intuitive ideas, as detailed at
+greater length by Jaynes~\cite{Jaynes2003} as well as Gelman and
+coauthors~\cite{Gelman2013,Gelman2013a}. From our samples of the
+posterior distribution, plotted in
+Figure~\ref{fig4:repressed_post_full}A, we generate many
+replicate data using a random number generator.
+In Figure~\ref{fig4:repressed_post_full}C, we plot ECDFs of the
+middle 95\% quantiles of these replicate data with the actual
+experimental data from Jones et.\ al.~\cite{Jones2014} overlaid,
+covering all ten experimental conditions spanning repressor
+binding sites and copy numbers (as well as the constitutive
+baseline UV5).
+
+The purpose of Figure~\ref{fig4:repressed_post_full}C is simply a
+graphical, qualitative assessment of the model: do the
+experimental data systematically disagree with the simulated
+data, which would suggest that our model is missing important
+features? A further question is not just whether there is a
+detectable difference between simulated and experimental data,
+but whether this difference is likely to materially affect the
+conclusions we draw from the posterior in
+Figure~\ref{fig4:repressed_post_full}A. More rigorous and
+quantitative statistical tests are possible~\cite{Gelman2013},
+but their quantitativeness does not necessarily make them more
+useful. As stated in~\cite{Gelman2013a}, we often find this
+graphical comparison more enlightening because it better engages
+our intuitition for the model, not merely telling \textit{if} the
+model is wrong but suggesting \textit{how} the model may be
+incomplete.
+
+Our broad brush takeaway from
+Figure~\ref{fig4:repressed_post_full}C is overall of good
+agreement. There some oddities, in particular the long tails in
+the data for Oid, 1~ng/mL, and O2, 0.5~ng/mL. The latter is
+especially odd since it extends beyond the tail of the
+unregulated UV5 distribution. This is a relatively small number
+of cells, however, so whether this is a peculiarity of the
+experimental data, a statistical fluke of small numbers, or a
+real biological effect is unclear. It is conceivable that there
+is some very slow timescale switching dynamics that could cause
+this bimodality, although it is unclear why it would only appear
+for specific repressor copy numbers. There is also a small offset
+between experiment and simulation for O2 at the higher repressor
+copy numbers, especially at 2~ng/mL. We suspect this may be
+caused by the very large number of zero mRNA counts for Oid,
+2~ng/mL skewing it's partner datasets through the shared
+association rate, and so we are not too concerned by this
+discrepancy either. None of these fairly minor quibbles cause us
+to seriously doubt the overall correctness of our model, which
+further validates its use to compare the equilbrium models'
+binding energies to the nonequilibrium models' repressor
+kinetics, as we originally set out to do.
 

### 2020-05-29 doc/section_05_discussion.tex
Many additions to main text Bayesian inference

@@
 not disturb the cell's physiology too badly.
-
-\mmnote{Outline:
-\begin{itemize}
-\item Hey look it mostly works. With full model, 9D model is well identified,
-though most of the individual experiments unsurprisingly are not:
-only 3 of the 9 datasets were identifiable if fit in isolation
-and for the rest only the ratio of repressor rates is identifiable.
-Only by making assumptions that certain rates
-are equal across datasets was identifiability achieved.
-PPC is not perfect, it kinda misses some
-rep/op pairs but overall it's surprisingly predictive considering how strong
-were the assumptions that went into it.
-\item Do we chalk this up to experimental imperfections or is there interesting
-biology to be found in the disagreements? I'm inclined towards the former. Even
-just catching cells in truly apples-to-apples growth phases, or matching up
-parameters of imaging sessions weeks or months apart, is hardly trivial.
-\item The alternative is to refine our model and/or add on complexity, and it's
-not at all clear what additions to make to the model, especially how to produce
-the long tail of repressed strains over UV5. I see no way to get that without
-getting into complicated time-correlation/history effects, where
-transcription...rebounds?? After being repressed?? Sounds odd.
-\item Comparison between my inferred rates, equilibrium binding E, and
-single-molecule measurements. Not quite as slam-dunk as I'd hoped because of
-$\gamma$ dependence, but rates are w/in factor of 2 or better, seems quite good
-really! With exception of one Oid data data point, free energies are w/in about
-0.5 kT also. That's not massively larger than the repeatability threshold of
-fold-change measurements, at maybe 0.1 to 0.3 kT.
-\item Surprising that so many model parameters can be inferred from just a few
-distributions that, by eye, don't have strong features! Does require some
-assumptions, e.g., about equivalence of rates across experiments, but that seems
-quite reasonable.
-\end{itemize}
-}

### 2020-05-29 doc/section_04_bayesian_inference.tex
RP carefully through inference
did repressor section

@@
 for the MCMC sampler without yielding any new insight. For the
-strongly repressed data (Oid, 10~ng/mL aTc), with >95\% of cells
+strongly repressed data (Oid, 10~ng/mL aTc), with $>$ 95\% of cells
 with zero mRNA, there was quite literally very little data from
@@
 and the bottom panel comparing to single-molecule measurements.
-
 First consider the top panel and its comparison between repressor
@@
 the repressor kinetic rates via
-Eq.~\ref{eq:deltaFR_eq_noneq_equiv}, which we reprint here,
+Eq.~\ref{eq:deltaFR_eq_noneq_equiv}, which we restate here,
 \begin{equation}
@@
 not as a failure of our theory.
-\mmnote{Does this seem a fair characterization?}
+\mmnote{Does this seem a fair characterization?   - RP: thanks, that was very helpful
+for me.}
 
@@
 the microscopic interpretation of this free energy, as the log of
-a ratio of transition rates, does indeed hold true.
-\mmnote{Does this deserve qualifiers, like ``to a good approximation,'' or ``as a first order model'' or something?
+a ratio of transition rates, does indeed hold true to within the inherent
+uncertainties that remain in the entire theory-experiment dialogue.
+%\mmnote{Does this deserve qualifiers, like ``to a good approximation,'' or ``as a first order model'' or something?
 I like the simple finality without, but I worry it's too brazen or subtlely misleading.}

### 2020-05-29 code/figures/main/fig02B.py
fix number visibility

@@ -9,6 +9,7 @@ import arviz as az
 
 import matplotlib.pyplot as plt
 import matplotlib
+import matplotlib.patheffects as path_effects
 import seaborn as sns
 import bebi103.viz
 
@@ -76,7 +77,7 @@ ax.text(
 # super-Poissonian
 ax.axhspan(1, 10, facecolor='#E8B19D', alpha=0.5, zorder=-100)
 ax.text(
-    5,
+    5
     2.5,
     "super-Poissonian",
     fontsize=10,
@@ -114,11 +115,15 @@ cbar.set_label(r"$\Delta\epsilon_r \; (k_BT)$")
 
 # Annotate points
 for i, (row, data) in enumerate(df_summary.iterrows()):
-    ax.annotate(
+    text = ax.annotate(
         f"{i + 1}",
         (data["mean"], data["fano"],),
         fontsize=7,
-        color="gray",
+        color="white",
+    )
+    text.set_path_effects(
+        [path_effects.Stroke(linewidth=1.5, foreground="black"),
+         path_effects.Normal()]
     )
 
 # Label axis

### 2020-05-29 code/figures/main/fig03.py

@@ -9,6 +9,7 @@ import arviz as az
 
 import matplotlib.pyplot as plt
 import matplotlib
+import matplotlib.patheffects as path_effects
 import seaborn as sns
 import bebi103.viz
 
@@ -241,12 +242,18 @@ for i, promoter in enumerate(df_energies.Name):
         linewidth=1.0,
         c=cmap(col_norm(df_energies[df_energies.Name == promoter]["Energy (kT)"].values[0])),
     )
-    ax_c.annotate(
+    text = ax_c.annotate(
         f"{i + 1}",
         (np.mean(x_contour[0]),
         np.mean(y_contour),),
         fontsize=7,
+        color="white"
     )
+    text.set_path_effects(
+        [path_effects.Stroke(linewidth=1.5, foreground="black"),
+         path_effects.Normal()]
+    )
+
 ax_c.set_xlim(right=1.2e1)
 ax_c.set_ylim(top=1e1)
 ax_c.set_ylabel(r'$k_i$ (bursts per mRNA lifetime)')
@@ -276,12 +283,16 @@ for i, promoter in enumerate(df_energies.Name):
     # Append y position
     y_pos.append(ptile_med)
     
-    ax_d.annotate(
+    text = ax_d.annotate(
         f"{i + 1}",
         (df_energies[df_energies.Name == promoter]["Energy (kT)"],
         ptile_med,),
         fontsize=7,
-        color="gray",
+        color="white"
+    )
+    text.set_path_effects(
+        [path_effects.Stroke(linewidth=1.5, foreground="black"),
+         path_effects.Normal()]
     )
 
 # Add colorbar

### 2020-05-30 doc/section_04_bayesian_inference.tex
MRM pass through section 04. No big changes

@@
 
-In this section of the paper, we continue our program of providing
-one complete description of the entire broad sweep of studies
-that have been made in the context of the repressor-operator model,
-dating all the way back to the original work of Jacob and Monod
-and including the visionary quantitative work of M\"{u}ller-Hill and collaborators
-and up to more recent studies.   In addition, the aim is to reconcile the 
-equilibrium and non-equilibrium perspectives that have been brought
-to bear on this problem. From Section~\ref{section_02_means}, this reconciliation
-depends on a key quantitative question as codified by Eq.~\ref{eq:deltaFR_eq_noneq_equiv}: does the free energy of repressor binding, as described in the equilibrium models and indirectly inferred from gene expression measurements, agree with the corresponding values of repressor binding and unbinding rates in the non-equilibrium picture, measured or inferred more directly?
-In this section we tackle the statistical inference problem of inferring these repressor rates from mRNA FISH data.
-But before we can turn to the full case of simple repression, we must choose an appropriate model of the constitutive promoter and infer the parameter values in that model.
-This is the problem we address first.
+In this section of the paper, we continue our program of providing one complete
+description of the entire broad sweep of studies that have been made in the
+context of the repressor-operator model, dating all the way back to the original
+work of Jacob and Monod and including the visionary quantitative work of
+M\"{u}ller-Hill and collaborators~\cite{Oehler1990} and up to more recent
+studies \cite{Garcia2011a}. In addition, the aim is to reconcile the equilibrium
+and non-equilibrium perspectives that have been brought to bear on this problem.
+From Section~\ref{section_02_means}, this reconciliation depends on a key
+quantitative question as codified by Eq.~\ref{eq:deltaFR_eq_noneq_equiv}: does
+the free energy of repressor binding, as described in the equilibrium models and
+indirectly inferred from gene expression measurements, agree with the
+corresponding values of repressor binding and unbinding rates in the
+non-equilibrium picture, measured or inferred more directly? In this section we
+tackle the statistical inference problem of inferring these repressor rates from
+mRNA FISH data. But before we can turn to the full case of simple repression, we
+must choose an appropriate model of the constitutive promoter and infer the
+parameter values in that model. This is the problem we address first.
 
@@
 model 5 in Figure~\ref{fig2:constit_cartoons}(A), a one-state bursty model of
-constitutive promoters, achieves the right balance of complexity
-and simplicity, by allowing both Fano factor $\nu>1$, but also by
-remedying, by design, the problems of parameter degeneracy than
-model 4 in Figure~\ref{fig2:constit_cartoons} suffered. Does this
-stand up to closer scrutiny, namely, comparison to full mRNA distributions
-rather than simply their moments? We will test this thoroughly on
-single-cell mRNA counts for different unregulated promoters from Jones et.\
-al.~\cite{Jones2014}.
+constitutive promoters, achieves the right balance of complexity and simplicity,
+by allowing both Fano factor $\nu>1$, but also by remedying, by design, the
+problems of parameter degeneracy than model 4 in
+Figure~\ref{fig2:constit_cartoons} suffered. Does this stand up to closer
+scrutiny, namely, comparison to full mRNA distributions rather than simply their
+moments? We will test this thoroughly on single-cell mRNA counts for different
+unregulated promoters from Jones et.\ al.~\cite{Jones2014}.
 
@@
 in Figure~\ref{fig2:constit_cartoons}. As we alluded to earlier, since the
-Poisson distribution has a Fano factor $\nu$ strictly equal to 1, and all of
-the observed data in Figure~\ref{fig2:constit_cartoons}(B) has Fano factor
-$\nu>1$, we might already suspect that this model is incapable of fitting the
-data. We will verify that this is in fact the case. Using the same argument we
-can immediately rule out models 2 and 3 from
-Figure~\ref{fig2:constit_cartoons}(A). These models have Fano factors $\nu\le 1$
-meaning they are underdispersed relative to the Poisson distribution. We will
-also not explicitly consider model 4 from~\fig{fig2:constit_cartoons}(A) since
-it was already thoroughly analyzed in~\cite{Razo-Mejia2020}, and since model 5
-can be viewed as a special case of it.
+Poisson distribution has a Fano factor $\nu$ strictly equal to 1, and all of the
+observed data in Figure~\ref{fig2:constit_cartoons}(B) has Fano factor $\nu>1$,
+we might already suspect that this model is incapable of fitting the data. We
+will verify that this is in fact the case. Using the same argument we can
+immediately rule out models 2 and 3 from Figure~\ref{fig2:constit_cartoons}(A).
+These models have Fano factors $\nu\le 1$ meaning they are underdispersed
+relative to the Poisson distribution. We will also not explicitly consider model
+4 from~\fig{fig2:constit_cartoons}(A) since it was already thoroughly analyzed
+in~\cite{Razo-Mejia2020}, and since model 5 can be viewed as a special case of
+it.
 
@@
 
-The steady-state solution for the master equation defines the likelihood
-term for a single cell $p(m \mid \lambda)$. What this means is that for a given
+The steady-state solution for the master equation defines the likelihood term
+for a single cell $p(m \mid \lambda)$. What this means is that for a given
 choice of parameter $\lambda$, under model 1 of
@@
 
-As detailed in Appendix~\ref{sec:bayesian} this particular choice of prior is known
-as the \textit{conjugate} prior for a Poisson likelihood. This type of priors
-have nice properties to work with, in particular a conjugate prior allows us to
-get a closed form for the posterior distribution $p(\lambda \mid D)$. In
-particular it can be shown that for a Poisson distribution likelihood with its
+As detailed in Appendix~\ref{sec:bayesian} this particular choice of prior is
+known as the \textit{conjugate} prior for a Poisson likelihood. This type of
+priors have nice properties to work with, in particular a conjugate prior allows
+us to get a closed form for the posterior distribution $p(\lambda \mid D)$. For
+our case it can be shown that for a Poisson distribution likelihood with its
 Gamma distribution conjugate prior, the posterior distribution is also a Gamma
-distribution ~\cite{Gelman2013}. In particular the two parameters $\alpha'$ and
+distribution ~\cite{Gelman2013}. Specifically the two parameters $\alpha'$ and
 $\beta'$ for this posterior distribution take the form $\alpha' = \alpha +
@@
 is true. For this case, as we expected given our premise of the Poisson promoter
-being wrong, it is quite obvious that the observed data, plotted in black is not consistent with
-the Poisson promoter model. An 
-equivalent plot for the bursty promoter model is shown in blue. Again the darker
-tone shows the median, while the lighter color encompasses 95\% of the randomly
-generated samples. Unlike the Poisson promoter model, the experimental ECDF
-closely tracks the posterior predictive ECDF, indicating this model is actually
-able to generate the observed data and increasing our confidence that this model
-is at least not wrong.
+being wrong, it is quite obvious that the observed data, plotted in black is not
+consistent with the Poisson promoter model. An equivalent plot for the bursty
+promoter model is shown in blue. Again the darker tone shows the median, while
+the lighter color encompasses 95\% of the randomly generated samples. Unlike the
+Poisson promoter model, the experimental ECDF closely tracks the posterior
+predictive ECDF, indicating this model is actually able to generate the observed
+data and increasing our confidence that this model is at least not wrong.
 
@@
 \text{few}$, from Figure~\ref{fig:constit_post_full}(B) has $\langle m \rangle
-\sim 20$, and from~\cite{Brewster2012} has $\beta\Delta\epsilon_P \sim - 6.5$.
-Further we generally assume $P/N_{NS} \sim 10^{-3}$ since
+\sim 20$, and from~\cite{Brewster2012} has $\beta\Delta\varepsilon_P \sim -
+6.5$. Further we generally assume $P/N_{NS} \sim 10^{-3}$ since
 $N_{NS}\approx4.6\times10^6$ and $P\sim10^3$. After some guess-and-check with
@@
 Now that we have a satisfactory model in hand for constitutive promoters, we
-would like to return to the main thread: can we reconcile the equilibrium
-and nonequilibrium models by putting to the test
+would like to return to the main thread: can we reconcile the equilibrium and
+nonequilibrium models by putting to the test
 Eq.~\ref{eq:deltaFR_eq_noneq_equiv}, the correspondence between indirectly
-inferred equilibrium binding energies and nonequilibrium kinetic rates?
-To make this comparison, is it possible to infer repressor binding and
-unbinding rates from mRNA distributions over a population of cells as measured
-by single-molecule FISH in~\cite{Jones2014}? If so, how do these inferred rates
+inferred equilibrium binding energies and nonequilibrium kinetic rates? To make
+this comparison, is it possible to infer repressor binding and unbinding rates
+from mRNA distributions over a population of cells as measured by
+single-molecule FISH in~\cite{Jones2014}? If so, how do these inferred rates
 compare to direct single-molecule measurements such as from~\cite{Hammar2014}
@@
 this particular dataset the repressor copy numbers are not known directly, so we
-label their association rates by the concentration of a small molecule inducer
+label their association rates by the concentration of a small inducer molecule 
 aTc which indirectly controls the transcription of $\textit{lacI}$, as
@@
 \end{equation}
-where $\alpha$ and $\beta$, defined for notational convenience, are
+where $_2F_1$ is the confluent hypergeometric function of the second kind,
+$\alpha$ and $\beta$, defined for notational convenience, are
 \begin{align}
@@
 
-\mmnote{These next 2 paragraphs should be SI, I think. Leaving here for
-now to avoid a merge conflict w/ MRM.}
-
-We found that fitting a single operator/aTc concentration at a
-time with a single binding and unbinding rate did not yield a
-stable inference for most of the possible operator/aTc
-combinations. In other words, a single dataset could not
-independently resolve the binding and unbinding rates, only their
-ratio as set by the mean fold-change in
-Figure~\ref{fig1:means_cartoons}. Only by making the assumption
-of a single unique binding rate for each repressor copy number
-and a single unique unbinding rate for each binding site, as done
-in Figure~\ref{fig4:repressed_post_full}A, was it possible to
-independently resolve the rates and not merely their ratios.
-
-We also note that we found it necessary to exclude the very
-weakly and very strongly repressed datasets from Jones et.\
-al.~\cite{Jones2014}. In both cases there was, in a sense, not
-enough information in the distriubtions for our inference
-algorithm to extract, and their inclusion simply caused problems
-for the MCMC sampler without yielding any new insight. For the
-strongly repressed data (Oid, 10~ng/mL aTc), with $>$ 95\% of cells
-with zero mRNA, there was quite literally very little data from
-which to infer rates. And the weakly repressed data, all with the
-repressor binding site O3, had an unbinding rate so fast that the
-sampler essentially sampled from the prior; the likelihood had
-negligible influence, meaning the data was not informing the
-sampler in any meaninful way, so no inference was possible.
-
 \subsubsection{Comparison with prior measurements of repressor binding energies}
-Our primary goal in this work is to reconcile the kinetic
-and equilibrium pictures of simple repression.
-Towards this end we would like to compare the repressor kinetic rates we
-have inferred with the repressor binding energies inferred through
-multiple methods in~\cite{Garcia2011a} and~\cite{Razo-Mejia2018}.
-If the agreement is close, then it suggests that the equilibrium
-models are not wrong and the repressor binding energies they contain
+Our primary goal in this work is to reconcile the kinetic and equilibrium
+pictures of simple repression. Towards this end we would like to compare the
+repressor kinetic rates we have inferred with the repressor binding energies
+inferred through multiple methods in~\cite{Garcia2011a}
+and~\cite{Razo-Mejia2018}. If the agreement is close, then it suggests that the
+equilibrium models are not wrong and the repressor binding energies they contain
 correspond to physically real free energies, not mere fit parameters.
 
-Figure~\ref{fig4:repressed_post_full}(B) shows both comparisons,
-with the top panel comparing to equilbrium binding energies
-and the bottom panel comparing to single-molecule measurements.
-First consider the top panel and its comparison between repressor
-kinetic rates and binding energies.
-As described in section~\ref{section_02_means}, if the
-equilibrium binding energies from~\cite{Garcia2011a}
-and~\cite{Razo-Mejia2018} indeed are the physically real binding
-energies we believe them to be, then they should be related to
-the repressor kinetic rates via
-Eq.~\ref{eq:deltaFR_eq_noneq_equiv}, which we restate here,
+Figure~\ref{fig4:repressed_post_full}(B) shows both comparisons, with the top
+panel comparing to equilibrium binding energies and the bottom panel comparing
+to single-molecule measurements. First consider the top panel and its comparison
+between repressor kinetic rates and binding energies. As described in
+section~\ref{section_02_means}, if the equilibrium binding energies
+from~\cite{Garcia2011a} and~\cite{Razo-Mejia2018} indeed are the physically real
+binding energies we believe them to be, then they should be related to the
+repressor kinetic rates via Eq.~\ref{eq:deltaFR_eq_noneq_equiv}, which we
+restate here,
 \begin{equation}
@@
 \end{equation}
-Assuming mass action
-kinetics implies that $k_R^+$ is proportional to repressor copy
-number $R$, or more precisely, it can be thought of as repressor
-copy number times some intrinsic per molecule association rate.
-But since $R$ is not directly known for our data
-from~\cite{Jones2014}, we cannot use this equation directly.
-Instead we can consider two different repressor binding sites and
-compute the \textit{difference} in binding energy between them,
-since this difference depends only on the unbinding rates and not
-on the binding rates. This can be seen by evaluating
-Eq.~\ref{eq:deltaFR_eq_noneq_equiv_repeat} for two different
-repressor binding sites, labeled (1) and (2), but with the same
+Assuming mass action kinetics implies that $k_R^+$ is proportional to repressor
+copy number $R$, or more precisely, it can be thought of as repressor copy
+number times some intrinsic per molecule association rate. But since $R$ is not
+directly known for our data from~\cite{Jones2014}, we cannot use this equation
+directly. Instead we can consider two different repressor binding sites and
+compute the \textit{difference} in binding energy between them, since this
+difference depends only on the unbinding rates and not on the binding rates.
+This can be seen by evaluating Eq.~\ref{eq:deltaFR_eq_noneq_equiv_repeat} for
+two different repressor binding sites, labeled (1) and (2), but with the same
 repressor copy number $R$, and taking the difference to find
@@
 \end{equation}
-The left and right hand sides of this equation are exactly the horizontal
-and vertical axes of the top panel of Figure~\ref{fig4:repressed_post_full}.
-Since we inferred rates for three repressor binding sites (O1, O2, and Oid),
-there are only two independent differences that can be constructed, and we
-arbitrarily chose to plot O2-O1 and O1-Oid in
-Figure~\ref{fig4:repressed_post_full}(B).
-Numerically, we compute values of $k_{O1}^- / k_{Oid}^-$ and
-$k_{O2}^- / k_{O1}^-$ directly from our full posterior samples,
-which conveniently provides uncertainties as well, as detailed in
-Appendix~\ref{sec:bayesian}.
+The left and right hand sides of this equation are exactly the horizontal and
+vertical axes of the top panel of Figure~\ref{fig4:repressed_post_full}. Since
+we inferred rates for three repressor binding sites (O1, O2, and Oid), there are
+only two independent differences that can be constructed, and we arbitrarily
+chose to plot O2-O1 and O1-Oid in Figure~\ref{fig4:repressed_post_full}(B).
+Numerically, we compute values of $k_{O1}^- / k_{Oid}^-$ and $k_{O2}^- /
+k_{O1}^-$ directly from our full posterior samples, which conveniently provides
+uncertainties as well, as detailed in Appendix~\ref{sec:bayesian}.
 \mmnote{Explain in appendix how marginalization with sampling makes this
-``trivial'' to compute from samples of the joint posterior, b/c
-sampling makes marginalization trivial.}
-We then compare these ratios of rates to the binding energy differences
-$\Delta\epsilon_{O1} - \Delta\epsilon_{Oid}$ and from
-$\Delta\epsilon_{O2} - \Delta\epsilon_{O1}$ as computed from the values
-from both~\cite{Garcia2011a} and~\cite{Razo-Mejia2018}.
-Three of the four values are within $\sim0.5 k_BT$ of the diagonal
-representing perfect agreement, which is comparable to the
-$\sim\text{few}\times0.1 k_BT$ variability between the independent
-determinations of the same quantities between~\cite{Garcia2011a}
+``trivial'' to compute from samples of the joint posterior, b/c sampling makes
+marginalization trivial.} We then compare these log ratios of rates to the
+binding energy differences $\Delta\varepsilon_{O1} - \Delta\varepsilon_{Oid}$
+and from $\Delta\varepsilon_{O2} - \Delta\varepsilon_{O1}$ as computed from the
+values from both~\cite{Garcia2011a} and~\cite{Razo-Mejia2018}. Three of the four
+values are within $\sim0.5 k_BT$ of the diagonal representing perfect agreement,
+which is comparable to the $\sim\text{few}\times0.1 k_BT$ variability between
+the independent determinations of the same quantities between~\cite{Garcia2011a}
 and~\cite{Razo-Mejia2018}. The only outlier involves Oid measurements
 from~\cite{Razo-Mejia2018}, and as the authors of~\cite{Razo-Mejia2018} note,
-this is a difficult measurement of low fluorescence signal against
-high background since Oid represses so strongly. We are therefore
-inclined to regard the failure of this point to fall near the
-diagonal as a testament to the difficulty of the measurement and
-not as a failure of our theory.
-\mmnote{Does this seem a fair characterization?   - RP: thanks, that was very helpful
-for me.}
+this is a difficult measurement of low fluorescence signal against high
+background since Oid represses so strongly. We are therefore inclined to regard
+the failure of this point to fall near the diagonal as a testament to the
+difficulty of the measurement and not as a failure of our theory. \mmnote{Does
+this seem a fair characterization?   - RP: thanks, that was very helpful for
+me.}
 
-On the whole then, we regard this as striking confirmation of the
-validity of the equilibrium models. Their lynchpin parameter is a
-phenomenological free energy of repressor binding that has
-previously only been inferred indirectly. Our result shows that
-the microscopic interpretation of this free energy, as the log of
-a ratio of transition rates, does indeed hold true to within the inherent
+On the whole then, we regard this as striking confirmation of the validity of
+the equilibrium models. Their lynchpin parameter is a phenomenological free
+energy of repressor binding that has previously only been inferred indirectly.
+Our result shows that the microscopic interpretation of this free energy, as the
+log of a ratio of transition rates, does indeed hold true to within the inherent
 uncertainties that remain in the entire theory-experiment dialogue.
@@
 \subsubsection{Comparison with prior measurements of repressor kinetics}
-In the previous section we established the equivalence between
-the equilibrium models' binding energies and the repressor
-kinetics we infer from mRNA population distributions.
-But one might worry that the repressor rates we infer
-from mRNA distributions are \textit{themselves} merely
-fitting parameters and that they do not actually correspond to the
-binding and unbinding rates of the repressor in vivo.
-To verify that this is not the case, we next compare our kinetic rates with a
-different measurement of the same rates using a radically different method:
-single molecule measurements as performed in Hammar et.\ al.~\cite{Hammar2014}.
-This is plotted in the lower panel of Figure~\ref{fig4:repressed_post_full}B.
+In the previous section we established the equivalence between the equilibrium
+models' binding energies and the repressor kinetics we infer from mRNA
+population distributions. But one might worry that the repressor rates we infer
+from mRNA distributions are \textit{themselves} merely fitting parameters and
+that they do not actually correspond to the binding and unbinding rates of the
+repressor in vivo. To verify that this is not the case, we next compare our
+kinetic rates with a different measurement of the same rates using a radically
+different method: single molecule measurements as performed in Hammar et.\
+al.~\cite{Hammar2014}. This is plotted in the lower panel of
+Figure~\ref{fig4:repressed_post_full}B.
 
-Since we do not have access to repressor copy number for either
-the FISH data from~\cite{Jones2014} or the single-molecule data
-from~\cite{Hammar2014}, we cannot make an apples-to-apples
-comparison of association rates $k_R^+$. Further, while Hammar
-et.\ al.\ directly measure the dissociation rates $k_R^-$, our
-inference procedure returns $k_R^-/\gamma$, i.e., the repressor
-dissociation rate nondimensionalized by the mRNA degradation rate
-$\gamma$. So to make the comparison, we must make an assumption
-for the value of $\gamma$ since it was not directly measured. For
-most mRNAs in \textit{E.\ coli}, quoted values for the typical mRNA
-lifetime $\gamma^{-1}$ range between about 2.5~min~\cite{Chen2015}
-to 8~min. We chose $\gamma^{-1} = 3$~min and $\gamma^{-1} =
-5$~min as representative values and plot a comparison of
-$k_{O1}^-$ and $k_{Oid}^-$ from our inference with corresponding values
+Since we do not have access to repressor copy number for either the
+single-cell mRNA data from~\cite{Jones2014} or the single-molecule data
+from~\cite{Hammar2014}, we cannot make an apples-to-apples comparison of
+association rates $k_R^+$. Further, while Hammar et.\ al.\ directly measure the
+dissociation rates $k_R^-$, our inference procedure returns $k_R^-/\gamma$,
+i.e., the repressor dissociation rate nondimensionalized by the mRNA degradation
+rate $\gamma$. So to make the comparison, we must make an assumption for the
+value of $\gamma$ since it was not directly measured. For most mRNAs in
+\textit{E.\ coli}, quoted values for the typical mRNA lifetime $\gamma^{-1}$
+range between about 2.5~min~\cite{Chen2015} to 8~min. We chose $\gamma^{-1} =
+3$~min and $\gamma^{-1} = 5$~min as representative values and plot a comparison
+of $k_{O1}^-$ and $k_{Oid}^-$ from our inference with corresponding values
 reported in~\cite{Hammar2014} for both these choices of $\gamma$.
@@
 The degree of quantitative agreement in the lower panel of
-Figure~\ref{fig4:repressed_post_full}B clearly depends on the
-precise choice of $\gamma$.
-Nevertheless we find this comparison very satisfying, when two wildly different approaches to a measurement of the same quantity yield broadly compatible results.
-We emphasize the agreement between our rates
-and the rates reported in~\cite{Hammar2014} for any reasonable
-$\gamma$: values differ by at most a factor of 2 and possibly
-agree to within our uncertainties of 10-20\%. From this we feel
-confident asserting that the parameters we have inferred from
-Jones et.\ al.'s FISH data do in fact correspond to repressor
-binding and unbinding rates, and therefore our conclusions on the
-agreement of these rates with binding energies
-from~\cite{Garcia2011a} and~\cite{Razo-Mejia2018} are valid.
+Figure~\ref{fig4:repressed_post_full}B clearly depends on the precise choice of
+$\gamma$. Nevertheless we find this comparison very satisfying, when two wildly
+different approaches to a measurement of the same quantity yield broadly
+compatible results. We emphasize the agreement between our rates and the rates
+reported in~\cite{Hammar2014} for any reasonable $\gamma$: values differ by at
+most a factor of 2 and possibly agree to within our uncertainties of 10-20\%.
+From this we feel confident asserting that the parameters we have inferred from
+Jones et.\ al.'s single-cell mRNA counts data do in fact correspond to repressor
+binding and unbinding rates, and therefore our conclusions on the agreement of
+these rates with binding energies from~\cite{Garcia2011a}
+and~\cite{Razo-Mejia2018} are valid.
 
 \subsubsection{Model checking}
-In Figure~\ref{fig:constit_post_full}B we saw that the simple
-Poisson model of a constitutive promoter, despite having a well
-behaved posterior, was clearly insufficient to describe the data.
-It behooves us to carry out a similar check for our model of
-simple repression, codified by Eq.~\ref{eq:p_m_bursty+rep} for
-the steady-state mRNA copy number distribution.
-As derived in Sections~\ref{section_02_means}
-and~\ref{sec:beyond_means}, we have compelling theoretical reasons
-to believe it is a good model, but if it nevertheless turned out
-to be badly contradicted by the data we should like to know.
+In Figure~\ref{fig:constit_post_full}B we saw that the simple Poisson model of a
+constitutive promoter, despite having a well behaved posterior, was clearly
+insufficient to describe the data. It behooves us to carry out a similar check
+for our model of simple repression, codified by Eq.~\ref{eq:p_m_bursty+rep} for
+the steady-state mRNA copy number distribution. As derived in
+Sections~\ref{section_02_means} and~\ref{sec:beyond_means}, we have compelling
+theoretical reasons to believe it is a good model, but if it nevertheless turned
+out to be badly contradicted by the data we should like to know.
 
-The details are deferred to Appendix~\ref{sec:bayesian}, and here
-we only attempt to summarize the intuitive ideas, as detailed at
-greater length by Jaynes~\cite{Jaynes2003} as well as Gelman and
-coauthors~\cite{Gelman2013,Gelman2013a}. From our samples of the
-posterior distribution, plotted in
-Figure~\ref{fig4:repressed_post_full}A, we generate many
-replicate data using a random number generator.
-In Figure~\ref{fig4:repressed_post_full}C, we plot ECDFs of the
-middle 95\% quantiles of these replicate data with the actual
-experimental data from Jones et.\ al.~\cite{Jones2014} overlaid,
-covering all ten experimental conditions spanning repressor
-binding sites and copy numbers (as well as the constitutive
-baseline UV5).
+The details are deferred to Appendix~\ref{sec:bayesian}, and here we only
+attempt to summarize the intuitive ideas, as detailed at greater length by
+Jaynes~\cite{Jaynes2003} as well as Gelman and
+coauthors~\cite{Gelman2013,Gelman2013a}. From our samples of the posterior
+distribution, plotted in Figure~\ref{fig4:repressed_post_full}A, we generate
+many replicate data using a random number generator. In
+Figure~\ref{fig4:repressed_post_full}C, we plot empirical cumulative
+distribution functions of the middle 95\% quantiles of these replicate data with
+the actual experimental data from Jones et.\ al.~\cite{Jones2014} overlaid,
+covering all ten experimental conditions spanning repressor binding sites and
+copy numbers (as well as the constitutive baseline UV5).
 
-The purpose of Figure~\ref{fig4:repressed_post_full}C is simply a
-graphical, qualitative assessment of the model: do the
-experimental data systematically disagree with the simulated
-data, which would suggest that our model is missing important
-features? A further question is not just whether there is a
-detectable difference between simulated and experimental data,
-but whether this difference is likely to materially affect the
-conclusions we draw from the posterior in
-Figure~\ref{fig4:repressed_post_full}A. More rigorous and
-quantitative statistical tests are possible~\cite{Gelman2013},
-but their quantitativeness does not necessarily make them more
-useful. As stated in~\cite{Gelman2013a}, we often find this
-graphical comparison more enlightening because it better engages
-our intuitition for the model, not merely telling \textit{if} the
-model is wrong but suggesting \textit{how} the model may be
-incomplete.
+The purpose of Figure~\ref{fig4:repressed_post_full}C is simply a graphical,
+qualitative assessment of the model: do the experimental data systematically
+disagree with the simulated data, which would suggest that our model is missing
+important features? A further question is not just whether there is a detectable
+difference between simulated and experimental data, but whether this difference
+is likely to materially affect the conclusions we draw from the posterior in
+Figure~\ref{fig4:repressed_post_full}A. More rigorous and quantitative
+statistical tests are possible~\cite{Gelman2013}, but their quantitativeness
+does not necessarily make them more useful. As stated in~\cite{Gelman2013a}, we
+often find this graphical comparison more enlightening because it better engages
+our intuition for the model, not merely telling \textit{if} the model is wrong
+but suggesting \textit{how} the model may be incomplete.
 
-Our broad brush takeaway from
-Figure~\ref{fig4:repressed_post_full}C is overall of good
-agreement. There some oddities, in particular the long tails in
-the data for Oid, 1~ng/mL, and O2, 0.5~ng/mL. The latter is
-especially odd since it extends beyond the tail of the
-unregulated UV5 distribution. This is a relatively small number
-of cells, however, so whether this is a peculiarity of the
-experimental data, a statistical fluke of small numbers, or a
-real biological effect is unclear. It is conceivable that there
-is some very slow timescale switching dynamics that could cause
-this bimodality, although it is unclear why it would only appear
-for specific repressor copy numbers. There is also a small offset
-between experiment and simulation for O2 at the higher repressor
-copy numbers, especially at 2~ng/mL. We suspect this may be
-caused by the very large number of zero mRNA counts for Oid,
-2~ng/mL skewing it's partner datasets through the shared
-association rate, and so we are not too concerned by this
-discrepancy either. None of these fairly minor quibbles cause us
-to seriously doubt the overall correctness of our model, which
-further validates its use to compare the equilbrium models'
-binding energies to the nonequilibrium models' repressor
-kinetics, as we originally set out to do.
+Our broad brush takeaway from Figure~\ref{fig4:repressed_post_full}C is overall
+of good agreement. There some oddities, in particular the long tails in the data
+for Oid, 1~ng/mL, and O2, 0.5~ng/mL. The latter is especially odd since it
+extends beyond the tail of the unregulated UV5 distribution. This is a
+relatively small number of cells, however, so whether this is a peculiarity of
+the experimental data, a statistical fluke of small numbers, or a real
+biological effect is unclear. It is conceivable that there is some very slow
+timescale switching dynamics that could cause this bimodality, although it is
+unclear why it would only appear for specific repressor copy numbers. There is
+also a small offset between experiment and simulation for O2 at the higher
+repressor copy numbers, especially at 2~ng/mL. We suspect this may be caused by
+the very large number of zero mRNA counts for Oid, 2~ng/mL skewing it's partner
+datasets through the shared association rate, and so we are not too concerned by
+this discrepancy either. None of these fairly minor quibbles cause us to
+seriously doubt the overall correctness of our model, which further validates
+its use to compare the equilibrium models' binding energies to the
+nonequilibrium models' repressor kinetics, as we originally set out to do.
 

### 2020-05-30 doc/appendix_03_bayesian_inference.tex
transferred paragraphs from main text to SI

@@
 }
+
+\mmnote{These next 2 paragraphs should be SI, I think. Leaving here for
+now to avoid a merge conflict w/ MRM.}
+
+We found that fitting a single operator/aTc concentration at a
+time with a single binding and unbinding rate did not yield a
+stable inference for most of the possible operator/aTc
+combinations. In other words, a single dataset could not
+independently resolve the binding and unbinding rates, only their
+ratio as set by the mean fold-change in
+Figure~\ref{fig1:means_cartoons}. Only by making the assumption
+of a single unique binding rate for each repressor copy number
+and a single unique unbinding rate for each binding site, as done
+in Figure~\ref{fig4:repressed_post_full}A, was it possible to
+independently resolve the rates and not merely their ratios.
+
+We also note that we found it necessary to exclude the very
+weakly and very strongly repressed datasets from Jones et.\
+al.~\cite{Jones2014}. In both cases there was, in a sense, not
+enough information in the distriubtions for our inference
+algorithm to extract, and their inclusion simply caused problems
+for the MCMC sampler without yielding any new insight. For the
+strongly repressed data (Oid, 10~ng/mL aTc), with $>$ 95\% of cells
+with zero mRNA, there was quite literally very little data from
+which to infer rates. And the weakly repressed data, all with the
+repressor binding site O3, had an unbinding rate so fast that the
+sampler essentially sampled from the prior; the likelihood had
+negligible influence, meaning the data was not informing the
+sampler in any meaninful way, so no inference was possible.

### 2020-05-30 doc/section_02_means.tex
small fixes. Starting to remove color notes

@@
 \fig{fig1:means_cartoons} depicts a smorgasbord of mathematicized cartoons for
-simple repression using both thermodynamic and kinetic models that have
-appeared in previous literature. For each cartoon, we calculate the fold-change
-in mean gene expression as predicted by that model, deferring some algebraic
-details to the appendix \mrm{cite specific appendix section}. What we will find
-is that all cartoons collapse to a single master curve, shown in
-\fig{fig1:means_cartoons}(D), which contains just two parameters. We label the
-parameters $\Delta F_R$, an effective free energy parametrizing the
-repressor-DNA interaction, and $\rho$, which subsumes all details of
-transcription in the absence of repressors. We will offer some intuition for why
-this master curve exists and discuss why at the level of the mean expression, we
-are unable to discriminate ``right'' from ``wrong'' cartoons given only
-measurements of fold-changes in expression.
+simple repression using both thermodynamic and kinetic models that have appeared
+in previous literature. For each cartoon, we calculate the fold-change in mean
+gene expression as predicted by that model, deferring some algebraic details to
+Appendix~\ref{sec:non_bursty}. What we will find is that all cartoons collapse
+to a single master curve, shown in \fig{fig1:means_cartoons}(D), which contains
+just two parameters. We label the parameters $\Delta F_R$, an effective free
+energy parametrizing the repressor-DNA interaction, and $\rho$, which subsumes
+all details of transcription in the absence of repressors. We will offer some
+intuition for why this master curve exists and discuss why at the level of the
+mean expression, we are unable to discriminate ``right'' from ``wrong'' cartoons
+given only measurements of fold-changes in expression.
 
@@
 $\Delta\varepsilon_R$ is the binding energy of a repressor to its operator site.
-The derivation of this result is deferred to the appendix \mrm{cite specific 
-section}.
+The derivation of this result is deferred to Appendix~\ref{sec:non_bursty}.
 
@@
 \end{equation}
-with the algebra details again deferred to the appendix \mrm{cite specific
-location}. Recall $k_R^+$ is proportional to the repressor copy number, so in
-computing fold-change, absence of repressor corresponds to $k_R^+\rightarrow0$.
-Therefore fold-change in this model is simply
+with the algebra details again deferred to Appendix~\ref{sec:non_bursty}. Recall
+$k_R^+$ is proportional to the repressor copy number, so in computing
+fold-change, absence of repressor corresponds to $k_R^+\rightarrow0$. Therefore
+fold-change in this model is simply
 \begin{equation}
@@
 The expression for mean mRNA is substantially more complicated now, as worked
-out in the appendix \mrm{cite section} where we find
+out in Appendix~\ref{sec:non_bursty} where we find
 \begin{equation}

### 2020-05-30 doc/section_03_beyond_means.tex
small fixes. Starting to remove color notes

@@
 elongate a transcript is no longer available at the start site to begin another.
-As shown in Appendix~\ref{sec:unreg_promoters}, the Fano factor in this model is
+As shown in Appendix~\ref{sec:non_bursty}, the Fano factor in this model is
 given by
@@
 bunching would lead to super-Poissonian noise in the mRNA distribution over a
-population of cells.  Indeed, as shown in Appendix~\ref{sec:unreg_promoters}, a
+population of cells.  Indeed, as shown in Appendix~\ref{sec:non_bursty}, a
 calculation of the Fano factor produces
@@
 master equation for mRNA copy number $m$ as derived in
-Appendix~\ref{sec:unreg_promoters} takes the form
+Appendix~\ref{sec:non_bursty} takes the form
 \begin{align}
@@
 \end{align}
-\mrm{notice I changed $k^+$ to $k_i$ to be consistent with the figure.}
 where we use $k_i$ to denote the burst initiation rate, $\text{Geom}(n;b)$ is the

### 2020-05-30 doc/appendix_03_bayesian_inference.tex
MRM progress towards cleaning the simple-repression inference section in the SI (with new figure)

@@
 
-\mrm{Still need to write something about the intuition of being data-rich and 
-how then the prior matters little.}
+For our particular case all of the datasets from~\cite{Jones2014} used in this
+paper have $\mathcal{O}(10^3)$ data points. What this implies is that our
+particular choice of priors will not significantly affect our inference as long
+as they are broad enough. A way to see why this is the case is to simply look at
+Bayes theorem. For $N ~ 1000-3000$ datum all of the independent of each other
+and $n \ll 10^3$ parameters Bayes theorem reads as
+\begin{equation}
+P(\vec{\theta} \mid D) \propto \prod_{k=1}^{N} P(d_k \mid \vec{\theta})
+\prod_{i=1}^n P(\theta_i),
+\end{equation}
+where $d_k$ represents the $k$-th datum. That means that if our priors span a
+wide range of parameter space, the posterior distribution would be dominated by
+the likelihood function.
 
@@
 \end{equation}
-Section ~\ref{section_04_bayesian_inference} in the main text details the
+Section~\ref{section_04_bayesian_inference} in the main text details the
 results of applying this inference to the single-cell mRNA counts data.
@@
 
-Now that we have a satisfactory model in hand for constitutive
-promoters, we would like to return to the main thread: can we
-infer repressor binding and unbinding rates from mRNA distributions
-over a population of cells as measured by FISH in~\cite{Jones2014}?
-If so, how do these inferred rates compare to direct
-single-molecule measurements such as from~\cite{Hammar2014} and
-to binding energies such as from~\cite{Garcia2011a}
-and~\cite{Razo-Mejia2018}, which were inferred under the assumptions
-of the equilibrium models in Figure~\ref{fig1:means_cartoons}(B)?
-And can this comparison shed light on the unreasonable effectiveness
-of the equilibrium models?
+As detailed in~\ref{section_04_bayesian_inference} in the main text the
+inference on the unregulated promoter served as a stepping stone towards our
+ultimate goal of inferring repressor rates from the steady-state mRNA
+distributions of simple-repression architectures. For this we expand the
+one-state bursty promoter model to a two-state promoter as schematized in
+Figure~\ref{fig1:means_cartoons}(C) as model 5. This model adds two new
+parameters: the repressor binding rate $k^+$, solely function of the repressor
+concentration, and the repressor dissociation rate $k^-$, solely a function of
+the repressor-DNA binding affinity.
+
+The structure of the data in~\cite{Jones2014} for regulated promoters tuned
+these two parameters independently. In their work the production of the LacI
+repressor was under the control of an inducible promoter regulated by the TetR
+repressor as schematized in Figre~\ref{figS:aTc_circuit}. When TetR binds to the
+small molecule anhydrotetracycline (aTc), it shifts to an inactive conformation
+unable to bind to the DNA. This translates into an increase in gene expression
+level. In other words, the higher the concentration of aTc added to the media,
+the less TetR repressors that can control the expression of the \textit{lacI}
+gene, so the higher the concentration of LacI repressors in the cell.
+
+\begin{figure}[h!]
+\centering
+\includegraphics{../figures/si/figS0X_aTc_circuit.pdf}
+\caption{\textbf{aTc controlled expression of LacI repressor.} Schematic of the
+circuit used in~\cite{Jones2014} to control the expression of the LacI
+repressor. The \textit{lacI} gene is under the control of the TetR repressor. As
+the TetR repressor is inactivated upon binding of anhydrotetracycline or aTc,
+the more aTc added to the media were cells are growing, the less TetR repressors
+available to control the expression of the \textit{lacI} gene, resulting in more
+LacI repressors per cell. LacI simultaneously controls the expression of the
+mRNA on which single-molecule mRNA FISH was performed for gene expression
+quantification.}
+\label{figS:aTc_circuit}
+\end{figure}
 

### 2020-05-30 doc/appendix_03_bayesian_inference.tex
MRM cleaning inference section in SI. Still needs input from MM

@@
 the less TetR repressors that can control the expression of the \textit{lacI}
-gene, so the higher the concentration of LacI repressors in the cell.
+gene, so the higher the concentration of LacI repressors in the cell. So by 
+tuning the amount of aTc in the media where the experimental strains were grown
+they effectively tune $k^+$ in our simple theoretical model. On the other hand
+to tune $k^-$ the authors swap three different binding sites for the LacI 
+repressor, each with different repressor-DNA binding affinities previously 
+characterized \cite{Garcia2011a}.
 
@@
 
-As we found in Section~\ref{sec:beyond_means}, for our purposes
-the ``right'' model of a constitutive promoter is the bursty
-picture, model five in Figure~\ref{fig:constit_cartoons}.
-Therefore our starting point here is the analogous model with
-repressor added, model 5 in Figure~\ref{fig1:means_cartoons}.
-For a given repressor binding site and copy number,
-this model has four rate parameters to be inferred:
-the repressor binding and unbinding rates $k_R^+$, and $k_R^-$,
-the initiation rate of bursts, $k_i$, and the mean burst size $b$
-(we nondimensionalize all of these by the mRNA degradation rate $\gamma$).
-
-Before showing the mathematical formulation of our statistical
-inference model, we would like to sketch the intuitive structure.
-The dataset from~\cite{Jones2014} we consider consists of
-single-cell mRNA FISH measurements of nine different conditions,
-spanning several combinations of three unique binding sites and
-four unique repressor copy numbers. We assume that the values of
-$k_i$ and $b$ are known, since we have already cleanly inferred
-them from constitutive promoter data, and further we assume that
-these values are the same across datasets with different
-repressor binding sites and copy numbers. We assume that there is
-one unbinding rate parameter for each repressor binding site,
-which is shared across conditions, and likewise one binding rate
-for each unique repressor copy number. This makes our model seven
-dimensional, or nine if one counts $k_i$ and $b$ as well.
+What this means is that we have access to data with different combinations of
+$k^-$ and $k^+$. We could naively try to fit the kinetic parameters individually
+for each of the datasets, but there is no reason to believe that the binding
+site identity for the LacI repressor somehow affects its expression level
+controlled from a completely different location in the genome, nor vice versa.
+In other words, what makes the most sense it to fit all datasets together to
+obtain a single value for each of the association and dissociation rates. What
+this means, as described in~\ref{section_04_bayesian_inference} of the main text
+is that we have a seven dimensional parameter space with four possible
+association rates $k^+$ given the four available aTc concentrations, and three
+possible dissociation rates $k^-$ given the three different binding sites
+available in the dataset.
 
@@
 \end{equation}
-Note the repressor copy numbers are not known directly, so we label their
-association rates by the concentration of a small molecule inducer aTc
-which indirectly controls the transcription of $\textit{lacI}$.
-Also note that the authors of~\cite{Jones2014} report estimates
-of LacI copy number per cell rather than direct measurements.
-However, these estimates were made \textit{assuming} the validity
-of the equilibrium models in Figure~\ref{fig1:means_cartoons},
-and since \textit{testing} these models is our goal, we will make
-no assumptions about the LacI copy number for given aTc concentrations.
-
-Bayes theorem reads simply
+Note that since the repressor copy numbers are not known directly as explained
+before, we label their association rates by the concentration of aTc. Bayes
+theorem reads simply
 \begin{equation}
@@
 \end{equation}
-where $D$ is the set of all $N$ observed single-cell mRNA counts
-across the various conditions. We assume that individual single-cell
-measurements are independent so that the likelihood factorizes as
+where $D$ is the set of all $N$ observed single-cell mRNA counts across the
+various conditions. We assume that individual single-cell measurements are
+independent so that the likelihood factorizes as
 \begin{equation}
@@
 \end{equation}
-where $k_j^\pm$ represent the appropriate binding and unbinding
-rates for the $j$-th measured cell. The probability
-$p(m\mid k_j^+, k_j^-, k_i, b)$ appearing in the last expression
-is exactly Eq.~\ref{eq:p_m_bursty+rep_appdx}, the steady-state
-distribution for our bursty model with repression derived in
-Section~\ref{sec:gen_fcn_appdx}, which for completeness we reproduce here as
+where $k_j^\pm$ represent the appropriate binding and unbinding rates for the
+$j$-th measured cell. Our likelihood function, previously derived in
+Appendix~\ref{sec:gen_fcn_appdx}, is given by the rather complicated result in
+Eq.~\ref{eq:p_m_bursty+rep_appdx}, which for completeness we reproduce here as
 \begin{equation}
@@
 
-\mmnote{Probably SI most of this paragraph?}
-This likelihood is rather inscrutable. We did not find any of the
-known analytical approximations for ${_2F_1}$ terribly useful in
-gaining intuition, so we instead resorted to numerics. One insight we found
+\mrm{This paragraph needs its own subsection in my opinion if it is to be 
+included in the SI.}
+This likelihood is rather inscrutable. We did not find any of the known
+analytical approximations for ${_2F_1}$ terribly useful in gaining intuition, so
+we instead resorted to numerics. One insight we found
 \mmnote{a plot would probably help explain, maybe SI, need to explain better}
 was that for very strong or very weak repression, the distribution in
-Eq.~\ref{eq:p_m_bursty+rep} is well approximated by a negative binomial
-with burst size $b$ and burst rate $k_i$ equal to their
-constitutive \textit{lacUV5} values, except with $k_i$ multiplied
-by the fold-change $\left(1+k_R^+/k_R^-\right)^{-1}$.
-In other words, once again only the \textit{ratio} $k_R^+/k_R^-$
-was detectable. But for intermediate repression, the distribution
-was visibly broadened with Fano factor greater than $1+b$, the
-value for the corresponding constitutive case. This indicates
-that the repressor rates had left an imprint on the distribution,
-and perhaps intuitively, this intermediate regime occurs for values of
-$k_R^\pm$ comparable to the burst rate $k_i$. Put another way,
-if the repressor rates are much faster or much slower than $k_i$,
-then there is a timescale separation and effectively only one
-timescale remains, $k_i\left(1+k_R^+/k_R^-\right)^{-1}$.
-Only when all three rates in the problem are comparable does the
-mRNA distribution retain detectable information about them.
+Eq.~\ref{eq:p_m_bursty+rep} is well approximated by a negative binomial with
+burst size $b$ and burst rate $k_i$ equal to their constitutive \textit{lacUV5}
+values, except with $k_i$ multiplied by the fold-change
+$\left(1+k_R^+/k_R^-\right)^{-1}$. In other words, once again only the
+\textit{ratio} $k_R^+/k_R^-$ was detectable. But for intermediate repression,
+the distribution was visibly broadened with Fano factor greater than $1+b$, the
+value for the corresponding constitutive case. This indicates that the repressor
+rates had left an imprint on the distribution, and perhaps intuitively, this
+intermediate regime occurs for values of $k_R^\pm$ comparable to the burst rate
+$k_i$. Put another way, if the repressor rates are much faster or much slower
+than $k_i$, then there is a timescale separation and effectively only one
+timescale remains, $k_i\left(1+k_R^+/k_R^-\right)^{-1}$. Only when all three
+rates in the problem are comparable does the mRNA distribution retain detectable
+information about them.
 
-Next we specify priors. As for the constitutive model, weakly
-informative lognormal priors are a natural choice for all our
-rates. We found that if the priors were too weak, our MCMC
-sampler would often become stuck in regions of parameter space
-with very low probability density, unable to move. We struck a
-balance in choosing our prior widths between helping the sampler
-run while simultaneously verifying that the marginal posteriors
-for each parameter were not artificially constrained or distorted
-by the presence of the prior. The only exception to this is the
-highly informative priors we placed on $k_i$ and $b$, since we
-have strong knowledge of them from our inference of constitutive
+Next we specify priors. As for the constitutive model, weakly informative
+lognormal priors are a natural choice for all our rates. We found that if the
+priors were too weak, our MCMC sampler would often become stuck in regions of
+parameter space with very low probability density, unable to move. We struck a
+balance in choosing our prior widths between helping the sampler run while
+simultaneously verifying that the marginal posteriors for each parameter were
+not artificially constrained or distorted by the presence of the prior. The only
+exception to this is the highly informative priors we placed on $k_i$ and $b$,
+since we have strong knowledge of them from our inference of constitutive
 promoters above.
@@
 \end{equation}
-where the likelihood is specified by Eq.~\ref{eq:p_m_bursty+rep}.
-\mmnote{Now that I've typed this out, this notation seems totally
-pointless here, especially when the likelihood is not a standard
-distribution. Maybe I should make a table or something to record
-the prior values, and maybe even that should go to SI?}
-We ran MCMC sampling on the full nine dimensional posterior
-specified by this generative model. To attempt to visualize this
-object, in Figure~\ref{fig4:repressed_post_full}(A) we plot
-several two-dimensional slices as contour plots, analogous to
-Figure~\ref{fig:constit_post_full}(C). Each of these nine slices
-corresponds to the $(k_R^+, k_R^-)$ pair of rates for one of the
-conditions from the dataset used to fit the model and gives a
-sense of the uncertainty and correlations in the posterior.
+where the likelihood is specified by Eq.~\ref{eq:p_m_bursty+rep}. We ran MCMC
+sampling on the full nine dimensional posterior specified by this generative
+model.
+
+We found that fitting a single operator/aTc concentration at a time with a
+single binding and unbinding rate did not yield a stable inference for most of
+the possible operator/aTc combinations. In other words, a single dataset could
+not independently resolve the binding and unbinding rates, only their ratio as
+set by the mean fold-change in Figure~\ref{fig1:means_cartoons} in the main
+text. Only by making the assumption of a single unique binding rate for each
+repressor copy number and a single unique unbinding rate for each binding site,
+as done in Figure~\ref{fig4:repressed_post_full}(A), was it possible to
+independently resolve the rates and not merely their ratios.
+
+We also note that we found it necessary to exclude the very weakly and very
+strongly repressed datasets from Jones et.\ al.~\cite{Jones2014}. In both cases
+there was, in a sense, not enough information in the distributions for our
+inference algorithm to extract, and their inclusion simply caused problems for
+the MCMC sampler without yielding any new insight. For the strongly repressed
+data (Oid, 10~ng/mL aTc), with $>$ 95\% of cells with zero mRNA, there was quite
+literally very little data from which to infer rates. And the weakly repressed
+data, all with the repressor binding site O3, had an unbinding rate so fast that
+the sampler essentially sampled from the prior; the likelihood had negligible
+influence, meaning the data was not informing the sampler in any meaningful way,
+so no inference was possible.
 
@@
 }
-
-\mmnote{These next 2 paragraphs should be SI, I think. Leaving here for
-now to avoid a merge conflict w/ MRM.}
-
-We found that fitting a single operator/aTc concentration at a
-time with a single binding and unbinding rate did not yield a
-stable inference for most of the possible operator/aTc
-combinations. In other words, a single dataset could not
-independently resolve the binding and unbinding rates, only their
-ratio as set by the mean fold-change in
-Figure~\ref{fig1:means_cartoons}. Only by making the assumption
-of a single unique binding rate for each repressor copy number
-and a single unique unbinding rate for each binding site, as done
-in Figure~\ref{fig4:repressed_post_full}A, was it possible to
-independently resolve the rates and not merely their ratios.
-
-We also note that we found it necessary to exclude the very
-weakly and very strongly repressed datasets from Jones et.\
-al.~\cite{Jones2014}. In both cases there was, in a sense, not
-enough information in the distriubtions for our inference
-algorithm to extract, and their inclusion simply caused problems
-for the MCMC sampler without yielding any new insight. For the
-strongly repressed data (Oid, 10~ng/mL aTc), with $>$ 95\% of cells
-with zero mRNA, there was quite literally very little data from
-which to infer rates. And the weakly repressed data, all with the
-repressor binding site O3, had an unbinding rate so fast that the
-sampler essentially sampled from the prior; the likelihood had
-negligible influence, meaning the data was not informing the
-sampler in any meaninful way, so no inference was possible.

### 2020-06-01 doc/appendix_01_nonbursty.tex
small fixes. Starting to remove color notes
Generalities section of Appendix 1 done

--- 2020-05-21 doc/appendix_02_constitutive_promoters.tex
@@
 \section{Derivations for non-bursty promoter models}
+\label{sec:non_bursty}
+
+In this section we detail the calculation of mean mRNA levels,
+fold-changes in expression, and Fano factors for nonequilibrium
+promoter models one through four in Figure~\ref{fig1:means_cartoons}.
+These are the results that were quoted but not derived in
+Sections~\ref{section_02_means} and \ref{sec:beyond_means} of the
+main text. In each of these four models, the natural
+mathematization of their cartoons is as a chemical master
+equation such as Eq.~\ref{eq:2state_rep_cme} for model one, which
+for completeness we reproduce here as
+\begin{align}
+\begin{split}
+\deriv{t}p_R(m,t) =& 
+- \overbrace{k_R^- p_R(m,t)}^{R \rightarrow U}
++ \overbrace{k_R^+ p_U(m,t)}^{U \rightarrow R}
++ \overbrace{(m+1)\gamma p_R(m+1,t)}^{m + 1 \rightarrow m}
+- \overbrace{\gamma mp_R(m,t)}^{m \rightarrow m - 1}
+\\
+\deriv{t}p_U(m,t) =&\; 
+\overbrace{k_R^- p_R(m,t)}^{R \rightarrow U}
+- \overbrace{k_R^+ p_U(m,t)}^{U \rightarrow R}
++ \overbrace{rp_U(m-1,t)}^{m-1 \rightarrow m}
+- \overbrace{rp_U(m,t)}^{m \rightarrow m + 1}
+\\
+&+ \overbrace{(m+1)\gamma p_U(m+1,t)}^{m + 1 \rightarrow m}
+- \overbrace{\gamma mp_U(m,t)}^{m \rightarrow m - 1}.
+\label{eq:poisson_promoter_cme_appdx}
+\end{split}
+\end{align}
+Here $p_R(m,t)$ and $p_U(m,t)$ are the probabilities of finding
+the system with $m$ mRNA molecules at time $t$ either in the
+repressor bound or unbound states, respectively. $r$ is the
+probability per unit time that a transcript will be initiated
+when the repressor is unbound, and $\gamma$ is the probability
+per unit time for a given mRNA to be degraded. $k_R^-$ is the
+probability per unit time that a bound repressor will unbind,
+while $k_R^+$ is the probability per unit time that an unbound
+operator will become bound by a repressor. Assuming mass action
+kinetics, $k_R^+$ is proportional to repressor copy number $R$.
+
+Next consider the cartoon for nonequilibrium model two in
+Figure~\ref{fig1:means_cartoons}. Now we must track probabilities
+$p_R$, $p_P$, and $p_E$ for the repressor bound, polymerase
+bound, and unbound states, respectively. By analogy
+to~\ref{eq:poisson_promoter_cme_appdx}, the master equation for
+model two can be written
+\begin{align}
+\begin{split}
+\deriv{t}p_R(m,t) =& 
+- \overbrace{k_R^- p_R(m,t)}^{R \rightarrow U}
++ \overbrace{k_R^+ p_E(m,t)}^{U \rightarrow R}
++ \overbrace{(m+1)\gamma p_R(m+1,t)}^{m + 1 \rightarrow m}
+- \overbrace{\gamma mp_R(m,t)}^{m \rightarrow m - 1}
+\\
+\deriv{t}p_E(m,t) =&\; 
+  \overbrace{k_R^- p_R(m,t)}^{R \rightarrow U}
+- \overbrace{k_R^+ p_E(m,t)}^{U \rightarrow R}
++ \overbrace{(m+1)\gamma p_E(m+1,t)}^{m + 1 \rightarrow m}
+- \overbrace{\gamma mp_E(m,t)}^{m \rightarrow m - 1}.
+\\
+&
++ \overbrace{k_P^- p_P(m,t)}^{A \rightarrow U}
+- \overbrace{k_P^+ p_E(m,t)}^{U \rightarrow A}
++ \overbrace{rp_P(m-1,t)}^{m-1 \rightarrow m,\;A \rightarrow U}
+\\
+\deriv{t}p_P(m,t) =&\; 
+- \overbrace{k_P^- p_P(m,t)}^{A \rightarrow U}
++ \overbrace{k_P^+ p_E(m,t)}^{U \rightarrow A}
++ \overbrace{(m+1)\gamma p_P(m+1,t)}^{m + 1 \rightarrow m}
+- \overbrace{\gamma mp_P(m,t)}^{m \rightarrow m - 1}.
+\\
+&- \overbrace{rp_P(m,t)}^{m \rightarrow m + 1,\;A \rightarrow U}.
+\label{eq:model2_cme_appdx}
+\end{split}
+\end{align}
+$k_P^+$ and $k_P^-$ are defined in close analogy to $k_R^+$ and
+$k_R^-$, except for RNAP binding and unbinding instead of
+repressor. Similarly $p_P(m,t)$ is defined for the active
+(RNAP-bound) state exactly as are $p_R(m,t)$ and $p_E(m,t)$ for
+the repressor bound and unbound states, respectively.
+The new subtlety Eq.~\ref{eq:model2_cme_appdx} introduces compared to
+Eq.~\ref{eq:poisson_promoter_cme_appdx} is that when mRNAs are
+produced, the promoter state also changes. This is encoded by the
+terms involving $r$, the last term in each of the equations for
+$p_E$ and $p_P$. The former accounts for arrivals in the
+unbound state and the latter accounts for departures from the
+RNAP-bound state.
+
+To condense and clarify the unwieldy notation of
+Eq.~\ref{eq:model2_cme_appdx}, it can be rewriten in matrix form. We define the column vector $\vec{p}(m,t)$ as
+\begin{equation}
+\vec{p}(m,t)
+= \begin{pmatrix} p_R(m,t) \\ p_E(m,t) \\ p_P(m,t) \end{pmatrix}
+\end{equation}
+to gather, for a given $m$, the probabilities of finding the
+system in the three possible promoter states. Then all the
+transition rates may be condensed into matrices which multiply
+this vector. The first matrix is
+\begin{equation}
+\mathbf{K} = \begin{pmatrix} -k_R^- & k_R^+ & 0 \\
+                        k_R^- & -k_R^+ -k_P^+ & k_P^- \\
+                        0 & k_P^+ & -k_P^- 
+                \end{pmatrix},
+% \label{eq:3state_cme_matrices_pt1}
+\end{equation}
+which tracks all promoter state changes in
+Eq.~\ref{eq:model2_cme_appdx} that are \textit{not} accompanied
+by a change in the mRNA copy number. The two terms accounting for
+transcription, the only transition that increases mRNA copy
+number, must be handled by two separate matrices given by
+\begin{equation}
+\mathbf{R_A} = \begin{pmatrix}
+                0 & 0 & 0 \\ 
+                0 & 0 & r \\ 
+                0 & 0 & 0
+                \end{pmatrix},\
+\mathbf{R_D} = \begin{pmatrix}
+                0 & 0 & 0 \\ 
+                0 & 0 & 0 \\ 
+                0 & 0 & r
+                \end{pmatrix}.
+% \label{eq:3state_cme_matrices_pt2}
+\end{equation}
+With these definitions, we can condense Eq.~\ref{eq:model2_cme_appdx} into
+the single equation
+\begin{equation}
+\frac{d}{dt} \vec{p}(m,t) =
+\left( \mathbf{K} - \mathbf{R_D} - \gamma m \mathbf{I} \right) \vec{p}(m,t)
+                + \mathbf{R_A} \vec{p}(m-1,t) +
+                \gamma (m+1) \mathbf{I} \vec{p}(m+1,t),
+\label{eq:generic_cme_appdx}
+\end{equation}
+which is just Eq.~\ref{eq:3state_rep_cme} in the main text.
+Straightforward if tedious algebra verifies that
+Eqs.~\ref{eq:3state_rep_cme} and~\ref{eq:generic_cme_appdx} are
+in fact equivalent.
+
+Although we derived Eq.~\ref{eq:generic_cme_appdx} for the particular case
+of nonequilibrium model two in Figure~\ref{fig1:means_cartoons},
+in fact the chemical master equations for all of the
+nonequilibrium models in Figure~\ref{fig1:means_cartoons} except
+for model five can be cast in this form. (We treat model five
+separately in Section~\ref{sec:gen_fcn_appdx}.)
+Model three introduces no new subtleties beyond model two and
+Eq.~\ref{eq:generic_cme_appdx} applies equally well to it, simply
+with different matrices of dimension four instead of three.
+Models one and four are both handled by Eq.~\ref{eq:2state_rep_cme}
+in the main text, which is just Eq.~\ref{eq:generic_cme_appdx}
+except in the special case of $\mathbf{R_D} = \mathbf{R_A} \equiv
+\mathbf{R}$, since in these two models transcription initiation
+events do not change promoter state.
+
+Recalling that our goal in this section is to derive expressions
+for mean mRNA and Fano factor for nonequilibrium models one
+through four in Figure~\ref{fig1:means_cartoons}, and since all
+four of these models are described by Eq.~\ref{eq:generic_cme_appdx},
+we can save substantial effort by deriving general formulas for
+mean mRNA and Fano factor from Eq.~\ref{eq:generic_cme_appdx}
+once and for all. Then for each model we can simply plug in the
+appropriate matrices for $\mathbf{K}$, $\mathbf{R_D}$, and
+$\mathbf{R_A}$ and carry out the remaining algebra.
+
+For our purposes it will suffice to derive the first and second
+moments of the mRNA distribution from this master equation,
+similar to the treatment in~\cite{Sanchez2011}, but we refer the
+interested reader to~\cite{Razo-Mejia2020} for an analogous
+treatment demonstrating an analytical solution for arbitrary moments.
+
+\subsection{General forms for mean mRNA and Fano factor}
+Our task now is to derive expressions for the first two moments of the
+steady-state mRNA distribution from Eq.~\ref{eq:generic_cme_appdx}.
+Our treatment of this is analogous to that given in~\cite{Sanchez2011}
+and~\cite{Razo-Mejia2020}. We first obtain the steady-state limit of
+Eq.~\ref{eq:generic_cme_appdx} in which the time derivative vanishes, giving
+\begin{equation}
+0 =
+\left( \mathbf{K} - \mathbf{R_D} - \gamma m \mathbf{I} \right) \vec{p}(m)
+                + \mathbf{R_A} \vec{p}(m-1) +
+                \gamma (m+1) \mathbf{I} \vec{p}(m+1),
+\label{eq:generic_cme_ss}
+\end{equation}
+From this, we want to compute
+\begin{equation}
+\langle{m}\rangle = \sum_S \sum_{m=0}^\infty m p_S(m)
+\end{equation}
+and
+\begin{equation}
+\langle{m^2}\rangle = \sum_S \sum_{m=0}^\infty m^2 p_S(m)
+\end{equation}
+which define the average values of $m$ and $m^2$ at steady state, where the
+averaging is over all possible mRNA copy numbers and promoter states $S$.
+For example, for model one in Figure~\ref{fig1:means_cartoons}, the sum on $S$
+would cover repressor bound and unbound states ($R$ and $U$ respectively),
+for model two, the sum would cover repressor bound, polymerase bound,
+and empty states ($R$, $P$, and $E$), and so on for the other models.
+
+Along the way it will be convenient to define the following
+\textit{conditional} moments as
+\begin{equation}
+\langle\vec{m}\rangle = \sum_{m=0}^\infty m \vec{p}(m),
+\end{equation}
+and
+\begin{equation}
+\langle\vec{m}^2\rangle = \sum_{m=0}^\infty m^2 \vec{p}(m).
+\end{equation}
+These objects are vectors of the same size as $\vect{p}(m)$, and each
+component can be thought of as the expected value of the mRNA copy number,
+or copy number squared, conditional on the promoter being in a certain state.
+For example, for model one in Figure~\ref{fig1:means_cartoons} which has
+repressor bound and unbound states labeled $R$ and $U$,
+$\langle\vec{m}^2\rangle$ would be
+\begin{equation}
+\langle\vec{m}^2\rangle
+= \begin{pmatrix} \sum_{m=0}^\infty m^2 p_R(m)
+                \\ \sum_{m=0}^\infty m^2 p_U(m) \end{pmatrix}.
+\end{equation}
+Analogously to $\langle\vec{m}\rangle$ and $\langle\vec{m}^2\rangle$,
+it is convenient to define the vector
+\begin{equation}
+\langle\vec{m}^0\rangle = \sum_{m=0}^\infty \vec{p}(m),
+\end{equation}
+whose elements are simply the probabilities of finding the system
+in each of the possible promoter states. To reduce notational clutter
+in what follows, sums over $m$ will be assumed to run from 0 to $\infty$
+unless otherwise specified, i.e., $\sum_{m=0}^\infty$ will be denoted
+simply as $\sum_m$.
+It will also be convenient to denote by $\vec{1}^\dagger$ a row
+vector of the same length as $\vec{p}$ whose elements are all
+1, such that, for instance,
+$\vec{1}^\dagger \cdot \langle\vec{m}^0\rangle = 1$,
+$\vec{1}^\dagger \cdot \langle\vec{m}\rangle = \langle{m}\rangle$, etc.
+
+\subsubsection{Promoter state probabilities $\langle\vec{m}^0\rangle$}
+To begin, we will find the promoter state probabilities
+$\langle\vec{m}^0\rangle$ from Eq.~\ref{eq:generic_cme_ss} by summing over all
+mRNA copy numbers $m$, producing
+\begin{equation}
+0 = \sum_m \left[
+    \left( \mathbf{K} - \mathbf{R_D} - \gamma m \mathbf{I} \right) \vec{p}(m)
+                + \mathbf{R_A} \vec{p}(m-1) +
+                \gamma (m+1) \mathbf{I} \vec{p}(m+1)
+\right]
+\end{equation}
+Using the definitions of $\langle\vec{m}^0\rangle$ and $\langle\vec{m}\rangle$,
+and noting the matrices $\mathbf{K}$, $\mathbf{R_D}$, and $\mathbf{R_A}$
+are all independent of $m$ and can be moved outside the sum, this simplifies to
+\begin{equation}
+0 = (\mathbf{K} - \mathbf{R_D}) \langle\vec{m}^0\rangle
+    - \gamma \langle\vec{m}\rangle + \mathbf{R_A} \sum_m \vec{p}(m-1)
+    + \gamma \sum_m (m+1)\vec{p}(m+1).
+\label{eq:generic_cme_deriv_020}
+\end{equation}
+The last two terms can be handled by reindexing the summations, transforming
+them to match the definitions of $\langle\vec{m}^0\rangle$ and
+$\langle\vec{m}\rangle$. For the first, noting $\vec{p}(-1)=0$ since
+negative numbers of mRNA are nonsensical, we have
+\begin{equation}
+\sum_{m=0}^\infty \vec{p}(m-1)
+= \sum_{m=-1}^\infty \vec{p}(m)
+= \sum_{m=0}^\infty \vec{p}(m) = \langle\vec{m}^0\rangle.
+\end{equation}
+Similarly for the second,
+\begin{equation}
+\sum_{m=0}^\infty (m+1)\vec{p}(m+1)
+= \sum_{m=1}^\infty m\vec{p}(m)
+= \sum_{m=0}^\infty m\vec{p}(m) = \langle\vec{m}\rangle,
+\end{equation}
+which holds since in extending the lower limit from $m=1$ to $m=0$,
+the extra term we added to the sum is zero. Substituting these back in
+Eq.~\ref{eq:generic_cme_deriv_020}, we have
+\begin{equation}
+0 = (\mathbf{K} - \mathbf{R_D}) \langle\vec{m}^0\rangle
+    - \gamma \langle\vec{m}\rangle + \mathbf{R_A} \langle\vec{m}^0\rangle
+    + \gamma \langle\vec{m}\rangle,
+\end{equation}
+or simply
+\begin{equation}
+0 = (\mathbf{K} - \mathbf{R_D} + \mathbf{R_A}) \langle\vec{m}^0\rangle.
+\label{eq:generic_cme_vecm0}
+\end{equation}
+So given matrices $\mathbf{K}$, $\mathbf{R_D}$, and $\mathbf{R_A}$ describing
+a promoter, finding $\langle\vec{m}^0\rangle$ simply amounts to solving this
+set of linear equations, subject to the normalization constraint
+$\vec{1}^\dagger \cdot \langle\vec{m}^0\rangle = 1$.
+
+\subsubsection{First moments $\langle\vec{m}\rangle$ and $\langle{m}\rangle$}
+By analogy to the above procedure for finding $\langle\vec{m}^0\rangle$,
+we may find $\langle\vec{m}\rangle$ by first multiplying
+Eq.~\ref{eq:generic_cme_ss} by $m$ and then summing over $m$.
+Carrying out this procedure we have
+\begin{equation}
+0 = \sum_m m \left[
+\left( \mathbf{K} - \mathbf{R_D} - \gamma m \mathbf{I} \right) \vec{p}(m)
+            + \mathbf{R_A} \vec{p}(m-1) +
+            \gamma (m+1) \mathbf{I} \vec{p}(m+1)
+\right],
+\end{equation}
+and now identifying $\langle\vec{m}\rangle$ and $\langle\vec{m}^2\rangle$ gives
+\begin{equation}
+0 = (\mathbf{K} - \mathbf{R_D}) \langle\vec{m}\rangle
+    - \gamma \langle\vec{m}^2\rangle + \mathbf{R_A} \sum_m m\vec{p}(m-1)
+    + \gamma \sum_m m(m+1)\vec{p}(m+1).
+\label{eq:generic_cme_deriv_030}
+\end{equation}
+The summations in the last two terms can be reindexed just as we did for
+$\langle\vec{m}^0\rangle$, freely adding or removing terms from
+the sum which are zero. For the first,
+\begin{equation}
+\sum_{m=0}^\infty m\vec{p}(m-1)
+= \sum_{m=-1}^\infty (m+1)\vec{p}(m)
+= \sum_{m=0}^\infty (m+1)\vec{p}(m)
+= \langle\vec{m}\rangle + \langle\vec{m}^0\rangle,
+\end{equation}
+and for the second,
+\begin{equation}
+\sum_{m=0}^\infty m(m+1)\vec{p}(m+1)
+= \sum_{m=1}^\infty (m-1)m\vec{p}(m)
+= \sum_{m=0}^\infty (m-1) m\vec{p}(m)
+= \langle\vec{m}^2\rangle - \langle\vec{m}\rangle.
+\end{equation}
+Substituting back in Eq.~\ref{eq:generic_cme_deriv_030} then produces
+\begin{equation}
+0 = (\mathbf{K} - \mathbf{R_D}) \langle\vec{m}\rangle
+- \gamma \langle\vec{m}^2\rangle
++ \mathbf{R_A} (\langle\vec{m}\rangle + \langle\vec{m}^0\rangle)
++ \gamma (\langle\vec{m}^2\rangle - \langle\vec{m}\rangle),
+\end{equation}
+or after simplifying
+\begin{equation}
+0 = (\mathbf{K} - \mathbf{R_D} + \mathbf{R_A} - \gamma) \langle\vec{m}\rangle
++ \mathbf{R_A} \langle\vec{m}^0\rangle.
+\label{eq:generic_cme_deriv_040}
+\end{equation}
+So like $\langle\vec{m}^0\rangle$, $\langle\vec{m}\rangle$ is also found
+by simply solving a set of linear equations after first solving for
+$\langle\vec{m}^0\rangle$ from Eq.~\ref{eq:generic_cme_vecm0}.
+
+Next we can find the mean mRNA copy number $\langle{m}\rangle$ from
+$\langle\vec{m}\rangle$ according to
+\begin{equation}
+\langle{m}\rangle = \vec{1}^\dagger\cdot\langle\vec{m}\rangle,
+\label{eq:m_from_vecm}
+\end{equation}
+where $\vec{1}^\dagger$ is a row vector whose elements are all 1.
+Eq.~\ref{eq:m_from_vecm} holds since the $i$-th element of the column vector
+$\langle\vec{m}\rangle$ is the mean mRNA value conditional on the system
+occupying the $i-$th promoter state, so the dot product with $\vec{1}^\dagger$
+amounts to simply summing the elements of $\langle\vec{m}\rangle$.
+Rather than solving Eq.~\ref{eq:generic_cme_deriv_040} for $\langle\vec{m}\rangle$
+and then computing $\vec{1}^\dagger\cdot\langle\vec{m}\rangle$, we may
+take a shortcut by muliplying Eq.~\ref{eq:generic_cme_deriv_040} by
+$\vec{1}^\dagger$ first. The key observation that makes this useful is that
+\begin{equation}
+\vec{1}^\dagger \cdot (\mathbf{K} - \mathbf{R_D} + \mathbf{R_A}) = 0.
+\end{equation}
+Intuitively, this equality holds because each column of this matrix
+represents transitions to and from a given promoter state. In any given column,
+the diagonal encodes all departing transitions and off-diagonals encode all
+arriving transitions. Conservation of probability means that each column
+sums to zero, and summing columns is exactly the operation that multiplying by
+$\vec{1}^\dagger$ carries out.
+
+Proceeding then in multiplying Eq.~\ref{eq:generic_cme_deriv_040} by
+$\vec{1}^\dagger$ produces
+\begin{equation}
+0 = -\gamma \vec{1}^\dagger\cdot\langle\vec{m}\rangle
++ \vec{1}^\dagger\cdot\mathbf{R_A}\langle\vec{m}^0\rangle,
+\end{equation}
+or simply
+\begin{equation}
+\langle{m}\rangle
+= \frac{1}{\gamma} \vec{1}^\dagger\cdot\mathbf{R_A}\langle\vec{m}^0\rangle.
+\label{eq:generic_mean_m_appdx}
+\end{equation}
+We note that the in equilibrium models of transcription such as in
+Figure~\ref{fig1:means_cartoons}, it is usually \textit{assumed} that the
+mean mRNA level is given by the ratio of initiation rate $r$ to degradation
+rate $\gamma$ multiplied by the probability of finding the system in the
+RNAP-bound state. Reassuringly, we have recovered exactly this result
+from the master equation picture: the product
+$\vec{1}^\dagger\cdot\mathbf{R_A}\langle\vec{m}^0\rangle$ picks out the
+probability of the active promoter state from $\langle\vec{m}^0\rangle$
+and multiplies it by the initiation rate $r$.
+
+\subsubsection{Second moment $\langle{m}^2\rangle$ and Fano factor $\nu$}
+Continuing the pattern of the zeroth and first moments, we now find
+$\langle\vec{m}^2\rangle$ by multiplying Eq.~\ref{eq:generic_cme_ss} by $m^2$
+and then summing over $m$, which explicitly is
+\begin{equation}
+0 = \sum_m m^2 \left[
+\left( \mathbf{K} - \mathbf{R_D} - \gamma m \mathbf{I} \right) \vec{p}(m)
+            + \mathbf{R_A} \vec{p}(m-1) +
+            \gamma (m+1) \mathbf{I} \vec{p}(m+1)
+\right].
+\end{equation}
+Identifying the moments $\langle\vec{m}^2\rangle$ and $\langle\vec{m}^3\rangle$
+in the first term simplifies this to
+\begin{equation}
+0 = (\mathbf{K} - \mathbf{R_D}) \langle\vec{m}^2\rangle
+    - \gamma \langle\vec{m}^3\rangle + \mathbf{R_A} \sum_m m^2\vec{p}(m-1)
+    + \gamma \sum_m m^2(m+1)\vec{p}(m+1).
+\label{eq:generic_cme_deriv_050}
+\end{equation}
+Reindexing the sums of the last two terms proceeds just as it did for the
+zeroth and first moments. Explicitly, we have
+\begin{equation}
+\sum_{m=0}^\infty m^2\vec{p}(m-1)
+= \sum_{m=-1}^\infty (m+1)^2\vec{p}(m)
+= \sum_{m=0}^\infty (m+1)^2\vec{p}(m)
+= \langle\vec{m}^2\rangle + 2\langle\vec{m}\rangle + \langle\vec{m}^0\rangle,
+\end{equation}
+for the first sum and
+\begin{equation}
+\sum_{m=0}^\infty m^2(m+1)\vec{p}(m+1)
+= \sum_{m=1}^\infty (m-1)^2m\vec{p}(m)
+= \sum_{m=0}^\infty (m-1)^2 m\vec{p}(m)
+= \langle\vec{m}^3\rangle - 2\langle\vec{m}^2\rangle + \langle\vec{m}\rangle
+\end{equation}
+for the second. Substituting the results of the sums back in
+Eq.~\ref{eq:generic_cme_deriv_050} gives
+\begin{equation}
+0 = (\mathbf{K} - \mathbf{R_D}) \langle\vec{m}^2\rangle
+- \gamma \langle\vec{m}^3\rangle
++ \mathbf{R_A}
+    (\langle\vec{m}^2\rangle + 2\langle\vec{m}\rangle + \langle\vec{m}^0\rangle)
++ \gamma
+    (\langle\vec{m}^3\rangle - 2\langle\vec{m}^2\rangle + \langle\vec{m}\rangle),
+\end{equation}
+and after grouping like powers of $m$ we have
+\begin{equation}
+0 = (\mathbf{K} - \mathbf{R_D} + \mathbf{R_A} - 2\gamma) \langle\vec{m}^2\rangle
++ (2\mathbf{R_A} + \gamma) \langle\vec{m}\rangle
++ \mathbf{R_A} \langle\vec{m}^0\rangle.
+\label{eq:generic_cme_deriv_060}
+\end{equation}
+As we found when computing $\langle{m}\rangle$ from $\langle\vec{m}\rangle$,
+we can spare ourselves some algebra by multiplying
+Eq.~\ref{eq:generic_cme_deriv_060} by $\vect{1}^\dagger$, which then reduces to
+\begin{equation}
+0 = - 2\gamma \langle{m}^2\rangle
++ \vec{1}^\dagger\cdot(2\mathbf{R_A} + \gamma) \langle\vec{m}\rangle
++ \vec{1}^\dagger\cdot\mathbf{R_A} \langle\vec{m}^0\rangle,
+\end{equation}
+and noting from Eq.~\ref{eq:generic_mean_m_appdx} that
+$\vec{1}^\dagger\cdot\mathbf{R_A} \langle\vec{m}^0\rangle
+= \gamma\langle{m}\rangle$, we have the tidy result
+\begin{equation}
+\langle{m}^2\rangle
+= \langle{m}\rangle + \frac{1}{\gamma}
+        \vec{1}^\dagger\cdot\mathbf{R_A} \langle\vec{m}\rangle.
+\end{equation}
+
+Finally we have all the preliminary results needed to write a general expression for the Fano factor $\nu$. The Fano factor is defined as the ratio of variance to mean, which can be written as
+\begin{equation}
+\nu = \frac{\langle{m}^2\rangle - \langle{m}\rangle^2}{\langle{m}\rangle}
+= \frac{
+    \langle{m}\rangle + \frac{1}{\gamma}
+        \vec{1}^\dagger\cdot\mathbf{R_A} \langle\vec{m}\rangle
+    - \langle{m}\rangle^2
+    }{\langle{m}\rangle}
+\end{equation}
+and simplified to
+\begin{equation}
+\nu = 1 - \langle{m}\rangle
++ \frac{\vec{1}^\dagger\cdot \mathbf{R_A}\langle\vec{m}\rangle}
+        {\gamma \langle{m}\rangle}.
+\label{eq:generic_fano_appdx}
+\end{equation}
+Note a subtle notational trap here: $\langle{m}\rangle
+= \frac{1}{\gamma} \vec{1}^\dagger\cdot\mathbf{R_A}\langle\vec{m}^0\rangle$
+rather than the by-eye similar but wrong expression $\langle{m}\rangle
+\ne \frac{1}{\gamma} \vec{1}^\dagger\cdot\mathbf{R_A}\langle\vec{m}\rangle$,
+so the last term in Eq.~\ref{eq:generic_fano_appdx} is in general quite nontrivial.
+For a generic promoter, Eq.~\ref{eq:generic_fano_appdx} may be greater than, less than, or equal to one, as asserted in Section~\ref{sec:beyond_means}.
+We have not found the general form Eq.~\ref{eq:generic_fano_appdx} terribly intuitive and instead defer discussion to specific examples.
+
+\subsubsection{Summary of general results}
+For ease of reference, we collect and reprint here the key results derived in this section that are used in the main text and subsequent subsections.
+Mean mRNA copy number and Fano factor are given by Eqs.~\ref{eq:generic_mean_m_appdx} and \ref{eq:generic_fano_appdx}, which are
+\begin{equation}
+\langle{m}\rangle
+= \frac{1}{\gamma} \vec{1}^\dagger\cdot\mathbf{R_A}\langle\vec{m}^0\rangle
+\end{equation}
+and
+\begin{equation}
+\nu = 1 - \langle{m}\rangle
++ \frac{\vec{1}^\dagger\cdot \mathbf{R_A}\langle\vec{m}\rangle}
+        {\gamma \langle{m}\rangle},
+\end{equation}
+respectively.
+To compute these two quantities, we need the expressions for
+$\langle\vec{m}^0\rangle$ and $\langle\vec{m}\rangle$ given by solving
+Eqs.~\ref{eq:generic_cme_vecm0} and \ref{eq:generic_cme_deriv_040},
+respectively, which are
+\begin{equation}
+(\mathbf{K} - \mathbf{R_D} + \mathbf{R_A}) \langle\vec{m}^0\rangle = 0
+\end{equation}
+and
+\begin{equation}
+(\mathbf{K} - \mathbf{R_D}
++ \mathbf{R_A} - \gamma\mathbf{I}) \langle\vec{m}\rangle
+= - \mathbf{R_A} \langle\vec{m}^0\rangle.
+\end{equation}
+
 \subsection{Nonequilibrium Model One - Poisson Promoter}

### 2020-06-01 doc/section_01_introduction.tex
MRM pass through entire main text. Removed side notes. Still lots left

@@
 inability to predict the regulatory consequences of different regulatory
-architectures. This challenge stems first and foremost from our ignorance about
-what those architectures even are with more than 60\% of the genes even in an
-ostensibly well understood organism such as  {\it E. coli} with no regulatory
-insights at all~\cite{Rydenfelt2014-2,Belliveau2018,Ghatak2019,
+architectures \mrm{, i.e. the arangement and effect of binding sites for
+transcription factors in the DNA}. This challenge stems first and foremost from
+our ignorance about what those architectures even are with more than 60\% of the
+genes even in an ostensibly well understood organism such as  {\it E. coli} with
+no regulatory insights at all~\cite{Rydenfelt2014-2,Belliveau2018,Ghatak2019,
 Santos_Zavaleta2019}. But even once we have established the identity of key

### 2020-06-01 doc/section_02_means.tex
RP
MRM pass through entire main text. Removed side notes. Still lots left

@@
 % !TEX root = ./busty_transcription.tex
-\section{Means}\label{section_02_means}
+\section{Mean Gene Expression}\label{section_02_means}
 
@@
 Fold-change has proven a very convenient observable in past
-work~\cite{Garcia2011a, Razo-Mejia2018, Chure2019}. Part of its utility in
-dissecting transcriptional regulation is its ratiometric nature, which removes
-many secondary effects that are present when making an absolute gene expression
-measurement. Also, by measuring otherwise identical cells with and without a
-transcription factor present, any biological noise common to both conditions can
-be made to cancel away. In addition, the presence of different operator sites,
-i.e. binding sites for the repressor, is known to change the basal expression
-level in the absence of repressor~\mmnote{look up cite}\mrm{does reg-seq mention
-anything like this in their SI? I definitely saw that on my own experiments, but
-didn't mention it in the paper}. Computing fold-changes neatly cancels this
-effect as well.
+work~\cite{Garcia2011a, Brewster2014, Razo-Mejia2018, Chure2019}. Part of its
+utility in dissecting transcriptional regulation is its ratiometric nature,
+which removes many secondary effects that are present when making an absolute
+gene expression measurement. Also, by measuring otherwise identical cells with
+and without a transcription factor present, any biological noise common to both
+conditions can be made to cancel away. In addition, the presence of different
+operator sites, i.e. binding sites for the repressor, is known to change the
+basal expression level in the absence of repressor~\mmnote{look up
+cite}\mrm{does reg-seq mention anything like this in their SI? I definitely saw
+that on my own experiments, but didn't mention it in the paper}. Computing
+fold-changes neatly cancels this effect as well.
 
@@
 departing transitions have different initial copy numbers of mRNA, unlike for
-$\matr{K}$ where they are the same. The master equation for this model is
+$\matr{K}$ where they are the same (see Appendix~\cite{sec:non_bursty}). The
+master equation for this model is
 \begin{equation}
@@
 \subsubsection{Nonequilibrium Model Four - ``Active'' and ``Inactive'' States}
-This model, shown as (4) in Figure~\ref{fig1:means_cartoons}(C) is at the core
-of the theory in~\cite{Razo-Mejia2020}. At a glance the cartoon for this model
-may appear very similar to model 2, and mathematically it is, but the
-interpretation is rather different. In model 2, we interpreted the third state
-literally as an RNAP-bound promoter and modeled initiation of a transcript as
-triggering a promoter state change, making the hopefully uncontroversial
-assumption that an RNAP can only make one transcript at a time. In contrast, in
-the present model the promoter state does \textit{not} change when a transcript
-is initiated. So we no longer interpret these states as literally RNAP bound and
-unbound but instead as coarse-grained ``active'' and ``inactive'' states, the
-details of which we leave unspecified for now. We will comment more on this
-model below when we discuss Fano factors of models.
+Model 4 in Figure~\ref{fig1:means_cartoons}(C) is at the core of the theory
+in~\cite{Razo-Mejia2020}. At a glance the cartoon for this model may appear very
+similar to model 2, and mathematically it is, but the interpretation is rather
+different. In model 2, we interpreted the third state literally as an RNAP-bound
+promoter and modeled initiation of a transcript as triggering a promoter state
+change, making the hopefully uncontroversial assumption that an RNAP can only
+make one transcript at a time. In contrast, in the present model the promoter
+state does \textit{not} change when a transcript is initiated. So we no longer
+interpret these states as literally RNAP bound and unbound but instead as
+coarse-grained ``active'' and ``inactive'' states, the details of which we leave
+unspecified for now. We will comment more on this model below when we discuss
+Fano factors of models.
 
@@
 
-Why does the master curve of~\fig{fig1:means_cartoons}(D) exist at all? This
-brings to mind the deep questions posed in, e.g.,~\cite{Frank2013}
-and~\cite{Frank2014a}, asking us to search for general ``symmetries'' of the
-problem instead of specifying generative models of the known microscopics of a
-system. In our case, the key feature seems to be the exclusive nature of
-repressor and RNAP binding, which allows the parameter describing the repressor,
-$\Delta F_R$, to cleanly separate from all other details of the promoter
-architecture, which are encapsulated in $\rho$. In the graph-theoretic language
-which was applied to similar problems in~\cite{Gunawardena2012}
-and~\cite{Ahsendorf2014}, the repressor-bound state is a leaf of the graph
-describing transitions between promoter states. Arbitrary nonequilibrium
-behavior can occur on the rest of the promoter state space, but it may all be
-swept up in the effective parameter $\rho$, to which the repressor makes no
-contribution.
+Why does the master curve of~\fig{fig1:means_cartoons}(D) exist at all?
+\textcolor{blue}{This brings to mind the deep questions posed in,
+e.g.,~\cite{Frank2013} and~\cite{Frank2014a}, asking us to search for general
+``symmetries'' of the problem instead of specifying generative models of the
+known microscopics of a system} \mrm{I don't understand this last sentence. Can
+we make it more transparent?}. In our case, the key feature seems to be the
+exclusive nature of repressor and RNAP binding, which allows the parameter
+describing the repressor, $\Delta F_R$, to cleanly separate from all other
+details of the promoter architecture, which are encapsulated in $\rho$. In the
+graph-theoretic language which was applied to similar problems
+in~\cite{Gunawardena2012} and~\cite{Ahsendorf2014}, the repressor-bound state is
+a leaf of the graph describing transitions between promoter states. Arbitrary
+nonequilibrium behavior can occur on the rest of the promoter state space, but
+it may all be swept up in the effective parameter $\rho$, to which the repressor
+makes no contribution.
 
@@
 the basis of the number of RNAP per cell and the strength of binding of RNAP to
-DNA at weak promoters. This is suggestive, but how can we be sure that $\rho$
-is not, for instance, actually $10^2$ and that $\Delta F_R$ hides a compensatory
+DNA at weak promoters. This is suggestive, but how can we be sure that $\rho$ is
+not, for instance, actually $10^2$ and that $\Delta F_R$ hides a compensatory
 factor? A resolution is offered by an independent inference of $\rho$ in the
@@
 nonequilibrium model 4 in~\fig{fig1:means_cartoons}(C), with zero repressor
-(looking ahead, this is equivalent to model 4 in~\fig{fig2:constit_cartoons}), to
-FISH data from~\cite{Brewster2014}. This provided a determination of $k^+$ and
-$k^-$, from which their ratio is estimated to be no more than a few $10^{-1}$
-and possibly as small as $10^{-2}$.
+(looking ahead, this is equivalent to model 4 in~\fig{fig2:constit_cartoons}),
+to single-cell mRNA counts data from~\cite{Brewster2014}. This provided a
+determination of $k^+$ and $k^-$, from which their ratio is estimated to be no
+more than a few $10^{-1}$ and possibly as small as $10^{-2}$.
 
@@
 a complementary approach: inferring repressor binding and unbinding rates
-$k_R^+$ and $k_R^-$ from FISH measurements of mRNA population distributions.
+$k_R^+$ and $k_R^-$ from single-molecule measurements of mRNA population
+distributions.
 \marginpar{\it MM: Add a notation aside somewhere that all association rates are written as zeroth-order rates.

### 2020-06-01 doc/section_03_beyond_means.tex
MRM pass through entire main text. Removed side notes. Still lots left

@@
 
-\subsection{A one-state promoter with bursting is the ``best'' model of
-constitutive promoters}
+\subsection{Kinetic models for unregulated promoter noise}
 
@@
 while only for (4) and (5) can $\nu$ exceed 1. Models with Fano factors $\le 1$
-cannot produce the FISH data observed in part (B) without introducing
+cannot produce the single-cell data observed in part (B) without introducing
 additional assumptions and model complexity. (B) Data from \cite{Jones2014}.
 Mean mRNA count vs. Fano factor (variance/mean) for different promoters as
-determined with single-molecule mRNA FISH. The colorbar indicates the predicted
-binding affinity of RNAP to the promoter sequence as determined in
-\cite{Brewster2012}. Numbers serve for cross comparison with data presented in 
-Figure 3.}
+determined with single-molecule mRNA Fluorescence \textit{in situ}
+Hybridization. The colorbar indicates the predicted binding affinity of RNAP to
+the promoter sequence as determined in \cite{Brewster2012}. Numbers serve for
+cross comparison with data presented in Figure 3.}
 \label{fig2:constit_cartoons}
@@
 study of gene expression noise over a series of promoters designed to have
-different strengths. This model insists that the ``true'' mRNA distribution is
-Poisson, implying the Fano factor $\nu$ must be 1. In~\cite{Jones2014}, the
-authors carefully attribute measured deviations from Fano = 1 to intensity
-variability in fluorescent spots, gene copy number variation, and copy number
-fluctuations of the transcription machinery, e.g., RNAP itself. In this picture,
-the master equation makes no appearance, and all the corrections to Poisson
-behavior are derived as additive corrections to the Fano factor. For disproving
-the ``universal noise curve'' from So et al~\cite{So2011}, this picture was
-excellent. It is appealing in its simplicity, with only two parameters, the
-initiation rate $r$ and degradation rate $\gamma$. Since $\gamma$ is
-independently known from other experiments, and the mean mRNA copy number is
-$r/\gamma$, $r$ is easily inferred from data. In other words, the model is not
-excessively complex for the data at hand. But for many interesting questions,
-for instance in the recent work~\cite{Razo-Mejia2020}, knowledge of means and
-variances alone is insufficient, and a description of the full distribution of
-molecular counts is necessary. For this we need a (slightly) more complex model
-than model 1 that would allow us to incorporate the non-Poissonian features of
-constitutive promoters directly into a master equation formulation.
+different strengths. This model insists that the ``true'' steady-state mRNA
+distribution is Poisson, implying the Fano factor $\nu$ must be 1.
+In~\cite{Jones2014}, the authors carefully attribute measured deviations from
+Fano = 1 to intensity variability in fluorescent spots, gene copy number
+variation, and copy number fluctuations of the transcription machinery, e.g.,
+RNAP itself. In this picture, the master equation makes no appearance, and all
+the corrections to Poisson behavior are derived as additive corrections to the
+Fano factor. For disproving the ``universal noise curve'' from So et
+al~\cite{So2011}, this picture was excellent. It is appealing in its simplicity,
+with only two parameters, the initiation rate $r$ and degradation rate $\gamma$.
+Since $\gamma$ is independently known from other experiments, and the mean mRNA
+copy number is $r/\gamma$, $r$ is easily inferred from data. In other words, the
+model is not excessively complex for the data at hand. But for many interesting
+questions, for instance in the recent work~\cite{Razo-Mejia2020}, knowledge of
+means and variances alone is insufficient, and a description of the full
+distribution of molecular counts is necessary. For this we need a (slightly)
+more complex model than model 1 that would allow us to incorporate the
+non-Poissonian features of constitutive promoters directly into a master
+equation formulation.
 
@@
 The problem with this picture is that the Fano factor is always $<1$. To make
- contact with the experimental reality of $\nu>1$, clearly some corrections will
- be needed. While this model adds an appealing element of microscopic reality,
- we are forced to reject it as the additional complexity is unable to capture
- the phenomenology of interest. Obviously the promoter state does in fact
- proceed through cycles of RNAP binding, initiating, and elongating, but it
- seems that the super-Poissonian noise in mRNA copy number we want to model must
- be governed by other features of the system.
+ contact with the experimental reality of $\nu>1$ as shown in
+ Figure~\ref{fig2:constit_cartoons}(B), clearly some corrections will be needed.
+ While this model adds an appealing element of microscopic reality, we are
+ forced to reject it as the additional complexity is unable to capture the
+ phenomenology of interest. Obviously the promoter state does in fact proceed
+ through cycles of RNAP binding, initiating, and elongating, but it seems that
+ the super-Poissonian noise in mRNA copy number we want to model must be
+ governed by other features of the system.
 
@@
 comprehensible reference??}}
-
 Regardless of the merits of this model in describing the noise properties of
@@
 of~\cite{Gelman2013}. What this means is that it is impossible to infer the
-parameters $r$ and $k^-$ from the FISH data of~\cite{Jones2014} (as shown in
-Fig.~S2 of~\cite{Razo-Mejia2020}). Rather, only the ratio $r/k^-$ could be
-inferred. In that work, the problem was worked around with an informative prior
-on the ratio $k^-/k^+$. That approach is unlikely to work here, as, recall, our
-entire goal in modeling constitutive expression is to use it as the basis for a
-yet more complicated model, when we add on repression. But adding more
-complexity to a model that is already poorly identified is a fool's errand, so
-we will explore one more potential model.
+parameters $r$ and $k^-$ from the single-cell mRNA counts data
+of~\cite{Jones2014} (as shown in Fig.~S2 of~\cite{Razo-Mejia2020}). Rather, only
+the ratio $r/k^-$ could be inferred. In that work, the problem was worked around
+with an informative prior on the ratio $k^-/k^+$. That approach is unlikely to
+work here, as, recall, our entire goal in modeling constitutive expression is to
+use it as the basis for a yet more complicated model, when we add on repression.
+But adding more complexity to a model that is already poorly identified is a
+fool's errand, so we will explore one more potential model.
 
@@
 was that, as found in~\cite{Razo-Mejia2020}, only two parameters, $k^+$ and the
-ratio $r/k^-$, could be directly inferred from the FISH data
+ratio $r/k^-$, could be directly inferred from the single-cell mRNA data
 from~\cite{Jones2014}. So let us take this seriously and imagine a model where
@@
 are produced at a rate $r$ per unit time, so the number of transcripts $n$
-produced during $T$ fits exactly the ``story'' for the Poisson distribution,
-i.e.,
+produced during $T$ is Poisson distributed with mean $rT$, i.e.,
 \begin{equation}
@@
 \end{equation}
-We direct readers again to Appendix~\ref{sec:beyond_means} for further details.
+We direct readers again to Appendix~\ref{sec:gen_fcn_appdx} for further details.
 This improves on model 4 in that now the parameters are easily inferred, as we
@@
 10 (or less) would have a diffusion-limited association rate of order
-$(10~\text{sec})^{-1}$ \mrm{cite Elf?}. Higher copy number TFs tend to have many
-binding sites over the genome, which should serve to pull them out of
+$(10~\text{sec})^{-1}$ \cite{Hammar2014}. Higher copy number TFs tend to have
+many binding sites over the genome, which should serve to pull them out of
 circulation and keep their effective association rates from rising too large.
@@
 could very well break down, so we will have to keep it in mind when we infer
-repressor rates from the Jones et.\ al.\ FISH data later~\cite{Jones2014}.
+repressor rates from the Jones et.\ al.\ single-cell mRNA counts data
+later~\cite{Jones2014}.
 

### 2020-06-01 doc/section_04_bayesian_inference.tex
MRM pass through entire main text. Removed side notes. Still lots left

@@
 % !TEX root = ./bursty_transcription.tex
-\section{Finding the ``right'' model: Bayesian parameter inference and model 
-selection}
+\section{Finding the ``right'' model: Bayesian parameter inference}
 \label{section_04_bayesian_inference}
@@
 tackle the statistical inference problem of inferring these repressor rates from
-mRNA FISH data. But before we can turn to the full case of simple repression, we
-must choose an appropriate model of the constitutive promoter and infer the
-parameter values in that model. This is the problem we address first.
+single-cell mRNA counts data. But before we can turn to the full case of simple
+repression, we must choose an appropriate model of the constitutive promoter and
+infer the parameter values in that model. This is the problem we address first.
 
@@
 problems of parameter degeneracy than model 4 in
-Figure~\ref{fig2:constit_cartoons} suffered. Does this stand up to closer
-scrutiny, namely, comparison to full mRNA distributions rather than simply their
-moments? We will test this thoroughly on single-cell mRNA counts for different
-unregulated promoters from Jones et.\ al.~\cite{Jones2014}.
+Figure~\ref{fig2:constit_cartoons} suffered~\cite{Razo-Mejia2020}. Does this
+stand up to closer scrutiny, namely, comparison to full mRNA distributions
+rather than simply their moments? We will test this thoroughly on single-cell
+mRNA counts for different unregulated promoters from Jones et.\
+al.~\cite{Jones2014}.
 
@@
 \bar{m} N$ and $\beta' = \beta + N$, where we defined the sample mean $\bar{m} =
-\frac{1}{N}\sum_k m_k$ for notational convenience, and $N$ is the number of
-cells in our dataset. Furthermore, given that $N$ is $\mathcal{O}(10^3)$ and
+\frac{1}{N}\sum_{k=1}^N m_k$ for notational convenience, and $N$ is the number
+of cells in our dataset. Furthermore, given that $N$ is $\mathcal{O}(10^3)$ and
 $\langle m\rangle \gtrsim 0.1$ for all promoters measured in~\cite{Jones2014}
@@
 
-We carried out MCMC sampling on the posterior of this model, starting with the
-constitutive \textit{lacUV5} dataset from~\cite{Jones2014}. The resulting MCMC
-samples are shown in Figure~\ref{fig:constit_post_full}(A). In contrast to the
-active/inactive constitutive model considered in~\cite{Razo-Mejia2020}
-(nonequilibrium model 4 in~\ref{fig2:constit_cartoons}), this model is
-well-identified with both parameters determined to a fractional uncertainty of
-5-10\%. The strong correlation reflects the fact that their product sets the
-mean of the mRNA distribution, which is tightly constrained by the data, but
-there is weak ``sloppiness''~\cite{Transtrum2015} along a set of values with a
-similar product.
+We carried out Markov-Chain Monte Carlo (MCMC) sampling on the posterior of this
+model, starting with the constitutive \textit{lacUV5} dataset
+from~\cite{Jones2014}. The resulting MCMC samples are shown in
+Figure~\ref{fig:constit_post_full}(A). In contrast to the active/inactive
+constitutive model considered in~\cite{Razo-Mejia2020} (nonequilibrium model 4
+in~\ref{fig2:constit_cartoons}), this model is well-identified with both
+parameters determined to a fractional uncertainty of 5-10\%. The strong
+correlation reflects the fact that their product sets the mean of the mRNA
+distribution, which is tightly constrained by the data, but there is weak
+``sloppiness''~\cite{Transtrum2015} along a set of values with a similar
+product.
 
@@
 transcripts under the control of a constitutive \textit{lacUV5} promoter is
-shown in orange. The median posterior predictive ECDFs for models (1), Poisson,
+shown in black. The median posterior predictive ECDFs for models (1), Poisson,
 and (5), negative binomial, are plotted in dark green and dark blue,
@@
 \subsubsection{Connecting inferred parameters to prior work}
-It is interesting to connect these speculations on $k_i$ and $b$ to the work
+It is interesting to connect these inferences on $k_i$ and $b$ to the work
 of~\cite{Brewster2012}, where these same 18 promoters were considered through
@@
 
-\subsection{Transcription factor kinetics can be inferred from FISH
-measurements}\label{sec:rep_kinetics_inference}
+\subsection{Transcription factor kinetics can be inferred from single-cell mRNA
+distribution measurements}\label{sec:rep_kinetics_inference}
 \subsubsection{Building the model and performing parameter inference}
@@
 from mRNA distributions over a population of cells as measured by
-single-molecule FISH in~\cite{Jones2014}? If so, how do these inferred rates
-compare to direct single-molecule measurements such as from~\cite{Hammar2014}
-and to binding energies such as from~\cite{Garcia2011a}
-and~\cite{Razo-Mejia2018}, which were inferred under the assumptions of the
-equilibrium models in Figure~\ref{fig1:means_cartoons}(B)? And can this
-comparison shed light on the unreasonable effectiveness of the equilibrium
-models, for instance, in their application in~\cite{Chure2019, Chure2019a}?
+single-molecule Fluorescence \textit{in situ} Hybridization in~\cite{Jones2014}?
+If so, how do these inferred rates compare to direct single-molecule
+measurements such as from~\cite{Hammar2014} and to binding energies such as
+from~\cite{Garcia2011a} and~\cite{Razo-Mejia2018}, which were inferred under the
+assumptions of the equilibrium models in Figure~\ref{fig1:means_cartoons}(B)?
+And can this comparison shed light on the unreasonable effectiveness of the
+equilibrium models, for instance, in their application in~\cite{Chure2019,
+Chure2019a}?
 
@@
 model of a constitutive promoter is the bursty picture, model five in
-Figure~\ref{fig2:constit_cartoons}A. Therefore our starting point here is the
+Figure~\ref{fig2:constit_cartoons}(A). Therefore our starting point here is the
 analogous model with repressor added, model 5 in
@@
 we would like to sketch the intuitive structure. The dataset
-from~\cite{Jones2014} we consider consists of single-cell mRNA FISH measurements
-of nine different conditions, spanning several combinations of three unique
+from~\cite{Jones2014} we consider consists of single-cell mRNA counts data of
+nine different conditions, spanning several combinations of three unique
 repressor binding sites and four unique repressor copy numbers. We assume that
@@
 regulation happens as the promoter is taken away from the transcriptionally
-active state when the promoter is bound by repressor. We assume that there is one unbinding
-rate parameter for each repressor binding site, which is shared across aTc
-inducer conditions, and likewise one binding rate for each unique repressor copy
-number. This makes our model seven dimensional, or nine if one counts $k_i$ and
-$b$ as well. Note that we use only a subset of the datasets from
-Jones et.\ al.~\cite{Jones2014}, as discussed more in Appendix~\ref{sec:bayes}.
+active state when the promoter is bound by repressor. We assume that there is
+one unbinding rate parameter for each repressor binding site, and likewise one
+binding rate for each unique repressor copy number. This makes our model seven
+dimensional, or nine if one counts $k_i$ and $b$ as well. Note that we use only
+a subset of the datasets from Jones et.\ al.~\cite{Jones2014}, as discussed more
+in Appendix~\ref{sec:bayes}.
 
@@
 where subscripts for dissociation rates $k^-$  indicate the different repressor
-binding sites, and subscripts for association rates $k^+$ indicate the aTc
-concentration at which the repressor itself was induced. We emphasize that for
-this particular dataset the repressor copy numbers are not known directly, so we
-label their association rates by the concentration of a small inducer molecule 
-aTc which indirectly controls the transcription of $\textit{lacI}$, as
-described in detail in~\cite{Jones2014}. Also note
-that the authors of~\cite{Jones2014} report estimates of LacI copy number per
-cell rather than direct measurements. However, these estimates were made
-\textit{assuming} the validity of the equilibrium models in
-Figure~\ref{fig1:means_cartoons}, and since \textit{testing} these models is our
-present goal, it would be circular logic if we were to make the same assumption.
-Therefore we will make no assumptions about the LacI copy number for given aTc
-concentrations.
+binding sites, and subscripts for association rates $k^+$ indicate the
+concentration of the small-molecule that controlled the expression of the LacI
+repressor (see Appendix~\ref{sec:bayes}). This is because for this particular
+dataset the repressor copy numbers were not measured directly, but it is safe to
+assume that a given concentration of the inducer resulted in a specific mean
+repressor copy number\cite{Chure2019a}. Also note that the authors
+of~\cite{Jones2014} report estimates of LacI copy number per cell rather than
+direct measurements. However, these estimates were made assuming the validity of
+the equilibrium models in Figure~\ref{fig1:means_cartoons}, and since testing
+these models is our present goal, it would be circular logic if we were to make
+the same assumption. Therefore we will make no assumptions about the LacI copy
+number for a given inducer concentrations.
 
@@
 
-\mmnote{Probably SI most of this paragraph?}
 This likelihood is rather inscrutable. We did not find any of the known
@@
 values, except with $k_i$ multiplied by the fold-change
-$\left(1+k_R^+/k_R^-\right)^{-1}$. In other words, once again only the
-\textit{ratio} $k_R^+/k_R^-$ was detectable. But for intermediate repression,
-the distribution was visibly broadened with Fano factor greater than $1+b$, the
-value for the corresponding constitutive case. This indicates that the repressor
-rates had left an imprint on the distribution, and perhaps intuitively, this
-intermediate regime occurs for values of $k_R^\pm$ comparable to the burst rate
-$k_i$. Put another way, if the repressor rates are much faster or much slower
-than $k_i$, then there is a timescale separation and effectively only one
-timescale remains, $k_i\left(1+k_R^+/k_R^-\right)^{-1}$. Only when all three
-rates in the problem are comparable does the mRNA distribution retain detectable
-information about them.
+$\left(1+k_R^+/k_R^-\right)^{-1}$. In other words, once again only the ratio
+$k_R^+/k_R^-$ was detectable. But for intermediate repression, the distribution
+was visibly broadened with Fano factor greater than $1+b$, the value for the
+corresponding constitutive case. This indicates that the repressor rates had
+left an imprint on the distribution, and perhaps intuitively, this intermediate
+regime occurs for values of $k_R^\pm$ comparable to the burst rate $k_i$. Put
+another way, if the repressor rates are much faster or much slower than $k_i$,
+then there is a timescale separation and effectively only one timescale remains,
+$k_i\left(1+k_R^+/k_R^-\right)^{-1}$. Only when all three rates in the problem
+are comparable does the mRNA distribution retain detectable information about
+them.
 
@@
 the failure of this point to fall near the diagonal as a testament to the
-difficulty of the measurement and not as a failure of our theory. \mmnote{Does
-this seem a fair characterization?   - RP: thanks, that was very helpful for
-me.}
+difficulty of the measurement and not as a failure of our theory.
 
@@
 al.~\cite{Hammar2014}. This is plotted in the lower panel of
-Figure~\ref{fig4:repressed_post_full}B.
+Figure~\ref{fig4:repressed_post_full}(B).
 
@@
 The degree of quantitative agreement in the lower panel of
-Figure~\ref{fig4:repressed_post_full}B clearly depends on the precise choice of
-$\gamma$. Nevertheless we find this comparison very satisfying, when two wildly
-different approaches to a measurement of the same quantity yield broadly
+Figure~\ref{fig4:repressed_post_full}(B) clearly depends on the precise choice
+of $\gamma$. Nevertheless we find this comparison very satisfying, when two
+wildly different approaches to a measurement of the same quantity yield broadly
 compatible results. We emphasize the agreement between our rates and the rates
@@
 \subsubsection{Model checking}
-In Figure~\ref{fig:constit_post_full}B we saw that the simple Poisson model of a
-constitutive promoter, despite having a well behaved posterior, was clearly
+In Figure~\ref{fig:constit_post_full}(B) we saw that the simple Poisson model of
+a constitutive promoter, despite having a well behaved posterior, was clearly
 insufficient to describe the data. It behooves us to carry out a similar check
@@
 coauthors~\cite{Gelman2013,Gelman2013a}. From our samples of the posterior
-distribution, plotted in Figure~\ref{fig4:repressed_post_full}A, we generate
+distribution, plotted in Figure~\ref{fig4:repressed_post_full}(A), we generate
 many replicate data using a random number generator. In
-Figure~\ref{fig4:repressed_post_full}C, we plot empirical cumulative
+Figure~\ref{fig4:repressed_post_full}(C), we plot empirical cumulative
 distribution functions of the middle 95\% quantiles of these replicate data with
@@
 
-The purpose of Figure~\ref{fig4:repressed_post_full}C is simply a graphical,
+The purpose of Figure~\ref{fig4:repressed_post_full}(C) is simply a graphical,
 qualitative assessment of the model: do the experimental data systematically
@@
 is likely to materially affect the conclusions we draw from the posterior in
-Figure~\ref{fig4:repressed_post_full}A. More rigorous and quantitative
+Figure~\ref{fig4:repressed_post_full}(A). More rigorous and quantitative
 statistical tests are possible~\cite{Gelman2013}, but their quantitativeness
@@
 
-Our broad brush takeaway from Figure~\ref{fig4:repressed_post_full}C is overall
-of good agreement. There some oddities, in particular the long tails in the data
-for Oid, 1~ng/mL, and O2, 0.5~ng/mL. The latter is especially odd since it
-extends beyond the tail of the unregulated UV5 distribution. This is a
+Our broad brush takeaway from Figure~\ref{fig4:repressed_post_full}(C) is
+overall of good agreement. There some oddities, in particular the long tails in
+the data for Oid, 1~ng/mL, and O2, 0.5~ng/mL. The latter is especially odd since
+it extends beyond the tail of the unregulated UV5 distribution. This is a
 relatively small number of cells, however, so whether this is a peculiarity of
@@
 nonequilibrium models' repressor kinetics, as we originally set out to do.
-
-% With priors and likelihood specified we may write down our complete generative model as
-% \begin{equation}
-% \begin{split}
-% \log_{10}k_i &\sim \text{Normal}(0.725, 0.025)\\
-% \log_{10}b   &\sim \text{Normal}(0.55, 0.025)\\
-% \log_{10}k_{0.5}^+ &\sim \text{Normal}(-0.45, 0.3)\\
-% \log_{10}k_{1}^+   &\sim \text{Normal}(0.6, 0.3)\\
-% \log_{10}k_{2}^+   &\sim \text{Normal}(1.15, 0.3)\\
-% \log_{10}k_{10}^+  &\sim \text{Normal}(1.5, 0.3)\\
-% \log_{10}k_{Oid}^- &\sim \text{Normal}(-0.25, 0.3)\\
-% \log_{10}k_{O1}^-  &\sim \text{Normal}(0.1, 0.3)\\
-% \log_{10}k_{O2}^-  &\sim \text{Normal}(0.45, 0.3)\\
-% m &\sim \text{Likelihood}(k_R^+, k_R^-, k_i, b),
-% \end{split}
-% \end{equation}
-% where the likelihood is specified by Eq.~\ref{eq:p_m_bursty+rep}.
-% \mmnote{Now that I've typed this out, this notation seems totally
-% pointless here, especially when the likelihood is not a standard
-% distribution. Maybe I should make a table or something to record
-% the prior values, and maybe even that should go to SI?}

### 2020-06-01 doc/section_05_discussion.tex
MRM pass through entire main text. Removed side notes. Still lots left

@@
 population of cells, the intrinsic cell-to-cell variability, and the full gene
-expression distribution found in such population as would be reported in a FISH
-experiment, for example.
+expression distribution found in such population as would be reported in a
+single molecule Fluorescence in situ Hybridization experiment, for example.
 
@@
 a treatment of the constitutive promoter already reveals that some models have
-Fano factors (variance/mean) that are less than one, at odds with
-any and all experimental data that we are aware of.
-This theoretical result allows us to directly discard a subset of the models
-(models 1-3 in Figure~\ref{fig2:constit_cartoons}) since they cannot be
-reconciled with experimental observations. The two remaining models (models 4 and 5
-in Figure~\ref{fig2:constit_cartoons}) appear to contain enough microscopic
-realistm to be able to reproduce the data. A previous exploration of one 
-of these models demonstrated the sloppy nature of the model in which data on single-cell
-mRNA counts alone cannot constrain the value of all parameters 
+Fano factors (variance/mean) that are less than one, at odds with any and all
+experimental data that we are aware of. This theoretical result allows us to
+directly discard a subset of the models (models 1-3 in
+Figure~\ref{fig2:constit_cartoons}(A)) since they cannot be reconciled with
+experimental observations. The two remaining models (models 4 and 5 in
+Figure~\ref{fig2:constit_cartoons}) appear to contain enough microscopic
+realistm to be able to reproduce the data. A previous exploration of one of
+these models demonstrated the sloppy nature of the model in which data on
+single-cell mRNA counts alone cannot constrain the value of all parameters
 simultaneously \cite{Razo-Mejia2020}. Here we demonstrate that the proposed
-one-state bursty promoter model (model 5 in 
-Figure~\ref{fig2:constit_cartoons}) emerges as  a limit of the commonly used
-two-state promoter model \cite{Peccoud1995}. We put  the idea to the test that this
-level of coarse-graining is rich enough to reproduce previous 
-experimental observations. In particular we perform Bayesian inference to determine the
-two parameters describing the full steady-state mRNA distribution, finding that
-the model is able to provide a quantitative description of  a plethora of promoter sequences with different
-mean levels of expression and noise.
+one-state bursty promoter model (model 5 in Figure~\ref{fig2:constit_cartoons})
+emerges as  a limit of the commonly used two-state promoter model
+\cite{Peccoud1995}. We put  the idea to the test that this level of
+coarse-graining is rich enough to reproduce previous experimental observations.
+In particular we perform Bayesian inference to determine the two parameters
+describing the full steady-state mRNA distribution, finding that the model is
+able to provide a quantitative description of  a plethora of promoter sequences
+with different mean levels of expression and noise.
 
-\mrm{
-Here we should include a discussion of the regulated case.  RP: I made a first
-brief stab at it.
-}
-With the results of the constitutive promoter in hand, we then fix the parameters
-associated with this class of promoters and use them as input for evaluating
-the noise in gene expression for the simple repression motif itself.  This analysis
-allows us to provide a single overarching analysis of both the constitutive
-and simple repression architectures using one simple model and
-corresponding set of self-consistent parameters, demonstrating not only
-a predictive framework, but also reconciling the equilibrium and non-equilibrium
+With the results of the constitutive promoter in hand, we then fix the
+parameters associated with this class of promoters and use them as input for
+evaluating the noise in gene expression for the simple repression motif itself.
+This analysis allows us to provide a single overarching analysis of both the
+constitutive and simple repression architectures using one simple model and
+corresponding set of self-consistent parameters, demonstrating not only a
+predictive framework, but also reconciling the equilibrium and non-equilibrium
 views of the same simple repression constructs.

### 2020-06-02 doc/section_01_introduction.tex
MRM pass through intro and discussion to add more citations

--- 2020-05-26 doc/section_01_introduction.tex
@@
 inability to predict the regulatory consequences of different regulatory
-architectures. This challenge stems first and foremost from our ignorance about
-what those architectures even are with more than 60\% of the genes even in an
-ostensibly well understood organism such as  {\it E. coli} with no regulatory
-insights at all~\cite{Rydenfelt2014-2,Belliveau2018,Ghatak2019,
-Santos_Zavaleta2019}. But even once we have established the identity of key
-transcription factors and their binding sites of a given promoter architecture,
-there remains the predictive challenge of understanding its input-output
-properties, an objective that can be met by a myriad of approaches using the
-tools of statistical physics~\cite{Ackers1982, Shea1985,
-Buchler2003,Vilar2003a,Vilar2003b, Bintu2005a,Bintu2005c, Gertz2009,Sherman2012,
-Saiz2013, Ko1991,Peccoud1995,Record1996, Kepler2001,
-Sanchez2008,Shahrezaei2008,Sanchez2011, Michel2010}.  One root to such
+architectures, i.e. the arrangement and affinity of binding sites for
+transcription factors and RNA polymerases on the DNA. This challenge stems
+first and foremost from our ignorance about what those architectures even are
+with more than 60\% of the genes even in an ostensibly well understood organism
+such as  {\it E. coli} with no regulatory insights at all~\cite{Rydenfelt2014-2,
+Belliveau2018, Ghatak2019, Santos_Zavaleta2019}. But even once we have
+established the identity of key transcription factors and their binding sites of
+a given promoter architecture, there remains the predictive challenge of
+understanding its input-output properties, an objective that can be met by a
+myriad of approaches using the tools of statistical physics~\cite{Ackers1982,
+Shea1985, Buchler2003, Vilar2003a, Vilar2003b, Bintu2005a, Bintu2005c,
+Gertz2009,Sherman2012, Saiz2013, Ko1991, Peccoud1995, Record1996, Kepler2001,
+Sanchez2008, Shahrezaei2008, Sanchez2011, Michel2010}.  One root to such
 predictive understanding is to focus on the simplest regulatory architecture and
@@
 been elucidated by generations of geneticists, molecular biologists and
-biochemists.
-Our aim is to understand the underlying principles of such coarse-graining
-in order to increase the predictive power of these models. In doing so we hope
-future models of regulatory response will be able to serve the powerful
-predictive role needed to take synthetic biology from a brilliant exercise in
-enlightened empiricism to a rational design framework as  in any other branch of
-engineering.
-% Our aim is to understand the underlying principles of such
-% coarse-graining such that future models of regulatory response can serve the
-% powerful predictive role needed to take synthetic biology from what is often a
-% brilliant exercise in enlightened empiricism and recast it as the same kind of
-% rational design process we are used to in other branches of engineering. 
-More precisely, we want phenomenology in the sense of coarse-graining away
-atomistic detail, but still retaining biophysical meaning. For example, we are
-not satisfied with the strictly phenomenological approach offered by the
-commonly used Hill functions. Studies like~\cite{Razo-Mejia2018} have
-demonstrated that Hill functions are clearly insufficient since each new
-situation requires a completely new set of parameters. Such work requires a
-quantitative theory of how biophysical changes at the molecular level propagate
-to input-output functions at the genetic circuit level. We want concepts, not
-mere facts. In particular a key question is: at this level of coarse graining,
-what microscopic details do we need to explicitly model, and how do we figure
-that out? For example, do we need to worry about all or even any of the steps
-that individual RNAPs go through each time they make a transcript? Turning the
-question around, can we see any imprint of those processes in the available
-data? If the answer is no, then those processes are irrelevant for our purposes.
-Forward modeling and inverse (statistical inferential) modeling are necessary to 
-tackles such questions.
+biochemists. Our aim is to understand the underlying principles of such
+coarse-graining in order to increase the predictive power of these models. In
+doing so we hope future models of regulatory response will be able to serve the
+powerful predictive role needed to take synthetic biology from a brilliant
+exercise in enlightened empiricism to a rational design framework as  in any
+other branch of engineering. More precisely, we want phenomenology in the sense
+of coarse-graining away atomistic detail, but still retaining biophysical
+meaning. For example, we are not satisfied with the strictly phenomenological
+approach offered by the commonly used Hill functions~\cite{Frank2013}.
+Studies like~\cite{Razo-Mejia2018} have demonstrated that Hill functions are
+clearly insufficient since each new situation requires a completely new set of
+parameters. Such work requires a quantitative theory of how biophysical changes
+at the molecular level propagate to input-output functions at the genetic
+circuit level. We want concepts, not mere facts. In particular a key question
+is: at this level of coarse graining, what microscopic details do we need to
+explicitly model, and how do we figure that out? For example, do we need to
+worry about all or even any of the steps that individual RNAPs go through each
+time they make a transcript? Turning the question around, can we see any imprint
+of those processes in the available data? If the answer is no, then those
+processes are irrelevant for our purposes. Forward modeling and inverse
+(statistical inferential) modeling are necessary to tackles such questions.
 
@@
 interesting open question precisely how a repressor bound far upstream is able
-to repress transcription. \marginpar{\it Cite that fig, from ??Nathan or Bill or
-Rob??, showing locations of repressor binding sites far upstream of promoter}
-Suggestions include ``action-at-a-distance'' mediated by kinks in the DNA,
-formed when the repressor is bound, that prevent RNAP binding. Nevertheless, our
-modeling in this work is sufficiently coarse-grained that we simply assume
-exclusive binding and leave explicit accounting of these details out of the
-problem.
+to repress transcription~\cite{Rydenfelt2014-2}. Suggestions include
+``action-at-a-distance'' mediated by kinks in the DNA, formed when the repressor
+is bound, that prevent RNAP binding. Nevertheless, our modeling in this work is
+sufficiently coarse-grained that we simply assume exclusive binding and leave
+explicit accounting of these details out of the problem.
 

### 2020-06-02 doc/section_05_discussion.tex
MRM pass through intro and discussion to add more citations

@@
 fashion. Other recent work has tackled the challenge of finding and annotating
-the regulatory part of genomes \cite{Belliveau2018, Ireland2020} \mrm{add other
-non-self referential}. Once we have determined the architecture of the
-regulatory part of the genome, we are then faced with the next class of
-questions which are sharpened by formulating them in mathematical terms, namely,
-what are the input-output properties of these regulatory circuits and what knobs
-control them?
+the regulatory part of genomes \cite{Belliveau2018, Ireland2020}. Once we have
+determined the architecture of the regulatory part of the genome, we are then
+faced with the next class of questions which are sharpened by formulating them
+in mathematical terms, namely, what are the input-output properties of these
+regulatory circuits and what knobs control them?
 
@@
 regulatory architecture hypothesized in the molecular biology era, namely, the
-repressor-operator model of Jacob and Monod. Regulation in that architecture is
-the result of a competition between a repressor which inhibits transcription and
-RNAP polymerase which undertakes it. Through the labors of generations of
-geneticists, molecular biologists and biochemists, an overwhelming amount of
-information and insight has been garnered into this simple regulatory motif,
-licensing it as what one might call the ``hydrogen atom'' of regulatory biology.
-It is from that perspective that the present paper explores the extent to which
-some of the different models that have been articulated to describe that motif
-allow us to understand both the average level of gene expression found in a
-population of cells, the intrinsic cell-to-cell variability, and the full gene
-expression distribution found in such population as would be reported in a
-single molecule Fluorescence in situ Hybridization experiment, for example.
+repressor-operator model of Jacob and Monod~\cite{Jacob1961}. Regulation in that
+architecture is the result of a competition between a repressor which inhibits
+transcription and RNAP polymerase which undertakes it. Through the labors of
+generations of geneticists, molecular biologists and biochemists, an
+overwhelming amount of information and insight has been garnered into this
+simple regulatory motif, licensing it as what one might call the ``hydrogen
+atom'' of regulatory biology. It is from that perspective that the present paper
+explores the extent to which some of the different models that have been
+articulated to describe that motif allow us to understand both the average level
+of gene expression found in a population of cells, the intrinsic cell-to-cell
+variability, and the full gene expression distribution found in such population
+as would be reported in a single molecule Fluorescence in situ Hybridization
+experiment, for example.
 
@@
 Fano factors (variance/mean) that are less than one, at odds with any and all
-experimental data that we are aware of. This theoretical result allows us to
-directly discard a subset of the models (models 1-3 in
+experimental data that we are aware of\cite{So2011, Jones2014}. This theoretical
+result allows us to directly discard a subset of the models (models 1-3 in
 Figure~\ref{fig2:constit_cartoons}(A)) since they cannot be reconciled with
 experimental observations. The two remaining models (models 4 and 5 in
-Figure~\ref{fig2:constit_cartoons}) appear to contain enough microscopic
-realistm to be able to reproduce the data. A previous exploration of one of
-these models demonstrated the sloppy nature of the model in which data on
-single-cell mRNA counts alone cannot constrain the value of all parameters
-simultaneously \cite{Razo-Mejia2020}. Here we demonstrate that the proposed
-one-state bursty promoter model (model 5 in Figure~\ref{fig2:constit_cartoons})
-emerges as  a limit of the commonly used two-state promoter model
-\cite{Peccoud1995}. We put  the idea to the test that this level of
-coarse-graining is rich enough to reproduce previous experimental observations.
-In particular we perform Bayesian inference to determine the two parameters
-describing the full steady-state mRNA distribution, finding that the model is
-able to provide a quantitative description of  a plethora of promoter sequences
-with different mean levels of expression and noise.
+Figure~\ref{fig2:constit_cartoons}) appear to contain enough microscopic realism
+to be able to reproduce the data. A previous exploration of model 4 demonstrated
+the sloppy nature of the model in which data on single-cell mRNA counts alone
+cannot constrain the value of all parameters simultaneously
+\cite{Razo-Mejia2020}. Here we demonstrate that the proposed one-state bursty
+promoter model (model 5 in Figure~\ref{fig2:constit_cartoons}) emerges as  a
+limit of the commonly used two-state promoter model \cite{Peccoud1995,
+Shahrezaei2008, So2011, Sanchez2013, Jones2014}. We put the idea to the test
+that this level of coarse-graining is rich enough to reproduce previous
+experimental observations. In particular we perform Bayesian inference to
+determine the two parameters describing the full steady-state mRNA distribution,
+finding that the model is able to provide a quantitative description of a
+plethora of promoter sequences with different mean levels of expression and
+noise.
 
@@
 evaluating the noise in gene expression for the simple repression motif itself.
-This analysis allows us to provide a single overarching analysis of both the
-constitutive and simple repression architectures using one simple model and
-corresponding set of self-consistent parameters, demonstrating not only a
-predictive framework, but also reconciling the equilibrium and non-equilibrium
-views of the same simple repression constructs.
+This allows us to provide a single overarching analysis of both the constitutive
+and simple repression architectures using one simple model and corresponding set
+of self-consistent parameters, demonstrating not only a predictive framework,
+but also reconciling the equilibrium and non-equilibrium views of the same
+simple repression constructs. More specifically, we obtained values for the
+transcription factor association and dissociation rates by performing Bayesian
+inference on the ful mRNA distribution for data obtained from simple-repression
+promoters with varying number of transcription factors per cell and affinity of
+such transcription factors for the binding site. The free energy value obtained
+from these kinetic rates -- computed as the log ratio of the rates -- agrees
+with previous inferences performed only from mean gene expression measurements,
+that assumed an equilibrium rather than a kinetic
+framework~\cite{Garcia2011a,Razo-Mejia2018}.
 
@@
 Chromosomal locus is one possible influence we have not addressed in this work,
-as all the FISH data from~\cite{Jones2014} that we considered was from a
-construct integrated at the \textit{galK} locus. The results of~\cite{Chong2014}
-indicate that transcription-induced supercoiling contributes substantially in
-driving transcriptional bursting, and furthermore, their Figure~7 suggests that
-the parameters describing the rate, duration, and size of bursts vary
-substantially for transcription from different genomic loci. Although the
-authors of~\cite{Englaender2017} do not address noise, they note enormous
-differences in mean expression levels when an identical construct is integrated
-at different genomic loci. The authors of~\cite{Engl2020} attribute noise and
-burstiness in their FISH data to the influence of different sigma factors, which
-is a reasonable conclusion from their data. Could the difference also be due to
-the different chromosomal locations of the two operons?
+as all the single-molecule mRNA data from~\cite{Jones2014} that we considered
+was from a construct integrated at the \textit{galK} locus. The results
+of~\cite{Chong2014} indicate that transcription-induced supercoiling contributes
+substantially in driving transcriptional bursting, and furthermore, their
+Figure~7 suggests that the parameters describing the rate, duration, and size of
+bursts vary substantially for transcription from different genomic loci.
+Although the authors of~\cite{Englaender2017} do not address noise, they note
+enormous differences in mean expression levels when an identical construct is
+integrated at different genomic loci. The authors of~\cite{Engl2020} attribute
+noise and burstiness in their single-molecule mRNA data to the influence of
+different sigma factors, which is a reasonable conclusion from their data. Could
+the difference also be due to the different chromosomal locations of the two
+operons?
 
@@
 
+\mrm{We need to decide ASAP if this is going to be included in the discussion or not. My vote is no, but Jupiter must decide.}
 \mmnote{Do all my approximations work simply because galK happens to have a low
@@
 not disturb the cell's physiology too badly.
+
+\mrm{Rather than proposing more experiments, I feel that we are missing a closing paragraph that puts all these findings in context.
+Where do we go from here as a community?
+Things such as the next building block, i.e.
+simple activation, and then from there scale up the complexity until we can get a full gene regulatory network kind of thing.
+Rather than going small into tiny details specific to this, let's step back and look at the big picture.}

### 2020-06-02 doc/section_05_discussion.tex
RP June 2
Intro has references to Muller-Hill
Discussion expanded
bib has many new entries

@@
 transcription factor association and dissociation rates by performing Bayesian
-inference on the ful mRNA distribution for data obtained from simple-repression
+inference on the full mRNA distribution for data obtained from simple-repression
 promoters with varying number of transcription factors per cell and affinity of
@@
 
-\mrm{We need to decide ASAP if this is going to be included in the discussion or not. My vote is no, but Jupiter must decide.}
-\mmnote{Do all my approximations work simply because galK happens to have a low
-duty cycle of supercoiling, $\beta/\alpha$ in Chong2014 picture? A potentially
-interesting test: put in the dozen or so Hernan constructs at some other locus,
-preferably one of the ones Chong identified as having a large duty cycle (large
-$\beta/\alpha$), i.e., the opposite limit of galK. Then just measure
-fold-changes. You don't even need to do FISH, although that would be even
-better. If their model is right, $\rho$ should be approximately set by the duty
-cycle of supercoiling, NOT by RNAP kinetics.
-\textit{This is a NICE prediction for future work.} \textit{cspE}
-is claimed nonessential, prediction is you drop HG reporter constructs in there
-and the inferred binding energies of all lacI operators shift the same amount,
-by about $\ln(3)\sim 1kT$. Not huge, but detectable, although might be
-confounding effects from moving to a new locus. With FISH measurements, $k^+$
-and $k^-$ in Fig 2, model 4 should now be separately inferrable if Chong2014
-measurements of $\beta/\alpha$ are correct. This would require an obnoxious
-amount of cloning, but once you did that, the measurements would be same old
-stuff that RP lab does.}
+\textcolor{red}{RP: I relegated all remaining comments to the background.  Muir
+can look at them to see if he really wants them.  I agreed with Manuel that the stuff
+below should not be included in the Discussion.  ONE THING FOR SURE - Muller-Hill
+and Oehler need to get a lot of love in references at beginning.}
+In summary, this work took up the challenge of exploring the extent
+to which a single specific mechanistic model of the simple-repression regulatory
+architecture suffices to explain the broad sweep of experimental
+data for this system.  Pioneering early experimental efforts from
+the M\"{u}ller-Hill lab established the simple-repression motif
+as an arena for the quantitative dissection of regulatory response in
+bacteria, with similar beautiful work emerging in examples such as the
+arabinose and gal operons as well~\cite{Dunn1984b, Oehler1990, Oehler1994, Schleif2000,Weickert1993, Semsey2002, SwintKruse2009}.  In light of a new generation of precision measurements
+on these systems, the definition of what it means to understand them can now
+be formulated as a rigorous quantitative question, and many aspects of
+the simple repression motif seem to merit the claim that they are sufficiently
+well understood to use that understanding to design new versions of that
+architecture based upon predictions about how repressor copy number and DNA binding site strength
+control expression.  In our view, the next step in the progression is
+to first perform similar rigorous analyses of the fundamental ``basis set'' of regulatory
+architectures such as the simple activation motif that is regulated by
+a single activator binding site, the repressor-activator architecture, mediated by
+the binding of both a single activator and a single repressor and beyond.
+With the individual input-output functions in hand, similar quantitative
+dissections including the rigorous analysis of their tuning parameters can be undertaken
+for the ``basis set'' of full gene-regulatory networks such as switches, feed-forward architectures and
+oscillators for example, building upon the recent impressive bonanza of efforts from
+systems biologists and synthetic biologists~\cite{Milo2002,Alon2007}.
 
-A direct \mmnote{if cloning intensive} future experiment would be
-to design a payload with a promoter driving a FISH reporter,
-integrate that payload at a variety of loci scattered around the genome,
-and carry out FISH measurements of the resulting strains.
-\mmnote{
-Did Brewster already do this? I remember he integrated arrays of
-competitor binding sites all over the genome, but did he put
-lacUV5 + reporter scattered around?}
-Obviously integration loci would need to be carefully chosen so
-that one could plausibly hope the deletion and integration did
-not disturb the cell's physiology too badly.
-
-\mrm{Rather than proposing more experiments, I feel that we are missing a closing paragraph that puts all these findings in context.
+%\mrm{We need to decide ASAP if this is going to be included in the discussion or not. My vote is no, but Jupiter must decide.}
+%\mmnote{Do all my approximations work simply because galK happens to have a low
+%duty cycle of supercoiling, $\beta/\alpha$ in Chong2014 picture? A potentially
+%interesting test: put in the dozen or so Hernan constructs at some other locus,
+%preferably one of the ones Chong identified as having a large duty cycle (large
+%$\beta/\alpha$), i.e., the opposite limit of galK. Then just measure
+%fold-changes. You don't even need to do FISH, although that would be even
+%better. If their model is right, $\rho$ should be approximately set by the duty
+%cycle of supercoiling, NOT by RNAP kinetics.
+%\textit{This is a NICE prediction for future work.} \textit{cspE}
+%is claimed nonessential, prediction is you drop HG reporter constructs in there
+%and the inferred binding energies of all lacI operators shift the same amount,
+%by about $\ln(3)\sim 1kT$. Not huge, but detectable, although might be
+%confounding effects from moving to a new locus. With FISH measurements, $k^+$
+%and $k^-$ in Fig 2, model 4 should now be separately inferrable if Chong2014
+%measurements of $\beta/\alpha$ are correct. This would require an obnoxious
+%amount of cloning, but once you did that, the measurements would be same old
+%stuff that RP lab does.}
+%
+%A direct \mmnote{if cloning intensive} future experiment would be
+%to design a payload with a promoter driving a FISH reporter,
+%integrate that payload at a variety of loci scattered around the genome,
+%and carry out FISH measurements of the resulting strains.
+%\mmnote{
+%Did Brewster already do this? I remember he integrated arrays of
+%competitor binding sites all over the genome, but did he put
+%lacUV5 + reporter scattered around?}
+%Obviously integration loci would need to be carefully chosen so
+%that one could plausibly hope the deletion and integration did
+%not disturb the cell's physiology too badly.
+%
+%\mrm{Rather than proposing more experiments, I feel that we are missing a closing paragraph that puts all these findings in context.
 Where do we go from here as a community?

### 2020-06-03 doc/appendix_01_nonbursty.tex
Finish appendix on mean and Fanos for nonbursty models

@@
 Figure~\ref{fig1:means_cartoons}. Now we must track probabilities
-$p_R$, $p_P$, and $p_E$ for the repressor bound, polymerase
-bound, and unbound states, respectively. By analogy
+$p_R$, $p_P$, and $p_E$ for the repressor bound, empty, and
+polymerase bound states, respectively. By analogy
 to~\ref{eq:poisson_promoter_cme_appdx}, the master equation for
@@
 \end{equation}
+$\mathbf{R_A}$ accounts for transitions \textit{arriving} in a given state
+while $\mathbf{R_D}$ tracks \textit{departing} transitions.
 With these definitions, we can condense Eq.~\ref{eq:model2_cme_appdx} into
@@
 Straightforward if tedious algebra verifies that
-Eqs.~\ref{eq:3state_rep_cme} and~\ref{eq:generic_cme_appdx} are
+Eqs.~\ref{eq:model2_cme_appdx} and~\ref{eq:generic_cme_appdx} are
 in fact equivalent.
@@
 \subsubsection{Promoter state probabilities $\langle\vec{m}^0\rangle$}
+\label{sec:m0_appdx}
 To begin, we will find the promoter state probabilities
@@
 $\vec{1}^\dagger \cdot \langle\vec{m}^0\rangle = 1$.
+It will turn out to be the case that, if the matrix
+$\mathbf{K} - \mathbf{R_D} + \mathbf{R_A}$ is $n$ dimensional, it will always
+have only $n-1$ linearly independent equations. Including the normalization
+condition provides the $n$-th linearly independent equation, ensuring a
+unique solution. So when using this equation to solve for
+$\langle\vec{m}^0\rangle$, we may always drop one row of the matrix equation
+at our convenience and supplement the system with the normalization
+condition instead.
 
@@
 \subsubsection{Summary of general results}
-For ease of reference, we collect and reprint here the key results derived in this section that are used in the main text and subsequent subsections.
-Mean mRNA copy number and Fano factor are given by Eqs.~\ref{eq:generic_mean_m_appdx} and \ref{eq:generic_fano_appdx}, which are
+For ease of reference, we collect and reprint here the key results derived
+in this section that are used in the main text and subsequent subsections.
+Mean mRNA copy number and Fano factor are given by
+Eqs.~\ref{eq:generic_mean_m_appdx} and \ref{eq:generic_fano_appdx}, which are
 \begin{equation}
@@
 = \frac{1}{\gamma} \vec{1}^\dagger\cdot\mathbf{R_A}\langle\vec{m}^0\rangle
+\label{eq:mean_m_appdx_ref}
 \end{equation}
@@
         {\gamma \langle{m}\rangle},
+\label{eq:fano_appdx_ref}
 \end{equation}
@@
 (\mathbf{K} - \mathbf{R_D} + \mathbf{R_A}) \langle\vec{m}^0\rangle = 0
+\label{eq:vecm0_appdx_ref}
 \end{equation}
@@
 = - \mathbf{R_A} \langle\vec{m}^0\rangle.
+\label{eq:vecm_appdx_ref}
 \end{equation}
+Some comments are in order before we consider particular models. First,
+note that to obtain $\langle\vec{m}\rangle$ and $\nu$, we need not bother
+solving for all components of the vectors $\langle\vec{m}^0\rangle$ and
+$\langle\vec{m}\rangle$, but only the components which are multiplied by
+nonzero elements of $\mathbf{R_A}$. The only component of
+$\langle\vec{m}^0\rangle$ that ever survives is the transciptionally active
+state, and for the models we consider here, there is only ever one such state.
+This will save us some amount of algebra below.
+
+Also note that we are computing Fano factors to verify the results of
+Section~\ref{sec:beyond_means}, concerning the constitutive promoter models
+in Figure~\ref{fig2:constit_cartoons} which are analogs of the simple
+repression models in Figure~\ref{fig1:means_cartoons}. We can translate
+the matrices from the simple repression models to the constitutive case
+by simply all occurences of repressor rates and removing the row and column
+corresponding to the repressor bound state. The results for
+$\langle{m}\rangle$ computed in the repressed case can be easily translated
+to the constitutive case, rather than recalculating from scratch, by taking
+the limit $k_R^+\rightarrow 0$, since this amounts to sending repressor
+copy number to zero.
+
+Finally, we point out that it would be possible to compute
+$\langle\vec{m}^0\rangle$ more simply using the diagram methods from King
+and Altman~\cite{King1956} (also independently discovered by
+Hill~\cite{Hill1966}). But to our knowledge this method cannot be
+applied to compute $\langle\vec{m}\rangle$ or $\nu$, so since we would need
+to resort to solving the matrix equations anyways for $\langle\vec{m}\rangle$,
+we choose not to introduce the extra conceptual burden of the diagram
+methods simply for computing $\langle\vec{m}^0\rangle$.
 
 \subsection{Nonequilibrium Model One - Poisson Promoter}
-\mmnote{Derive mean mRNA referred to in means section.}
+\subsubsection{Mean mRNA}
+For nonequilibrium model one in Figure~\ref{fig1:means_cartoons},
+we have already shown the full master equation in
+Eq.~\ref{eq:poisson_promoter_cme} and Eq.~\ref{eq:poisson_promoter_cme_appdx},
+but for completeness we reprint it again as
+\begin{align}
+\begin{split}
+\deriv{t}p_R(m,t) =& 
+- \overbrace{k_R^- p_R(m,t)}^{R \rightarrow U}
++ \overbrace{k_R^+ p_U(m,t)}^{U \rightarrow R}
++ \overbrace{(m+1)\gamma p_R(m+1,t)}^{m + 1 \rightarrow m}
+- \overbrace{\gamma mp_R(m,t)}^{m \rightarrow m - 1}
+\\
+\deriv{t}p_U(m,t) =&\; 
+\overbrace{k_R^- p_R(m,t)}^{R \rightarrow U}
+- \overbrace{k_R^+ p_U(m,t)}^{U \rightarrow R}
++ \overbrace{rp_U(m-1,t)}^{m-1 \rightarrow m}
+- \overbrace{rp_U(m,t)}^{m \rightarrow m + 1}
+\\
+&+ \overbrace{(m+1)\gamma p_U(m+1,t)}^{m + 1 \rightarrow m}
+- \overbrace{\gamma mp_U(m,t)}^{m \rightarrow m - 1}.
+\end{split}
+\end{align}
+This is a direct transcription of the states and rates in
+Figure~\ref{fig1:means_cartoons}. This may be converted to the matrix form
+of the master equation shown in Eq.~\ref{eq:generic_cme_appdx} with matrices
+\begin{equation}
+\vec{p}(m) = \begin{pmatrix} p_R(m) \\ p_U(m) \end{pmatrix},\
+\mathbf{K} = \begin{pmatrix} -k_R^- & k_R^+ \\ k_R^- & -k_R^+ \end{pmatrix},\
+\mathbf{R} = \begin{pmatrix} 0 & 0 \\ 0 & r \end{pmatrix},\
+\end{equation}
+where $\mathbf{R_A}$ and $\mathbf{R_D}$ are equal, so we drop the subscript
+and denote both simply by $\mathbf{R}$. Since our interest is only in
+steady-state we dropped the time dependence as well.
+
+First we need $\langle\vec{m}^0\rangle$. Label its components as $p_R$ and
+$p_U$, the probabilities of finding the system in either promoter state,
+and note that only $p_U$ survives multiplication by $\mathbf{R}$, since
+\begin{equation}
+\mathbf{R} \langle\vec{m}^0\rangle
+= \begin{pmatrix} 0 & 0 \\ 0 & r \end{pmatrix}
+    \begin{pmatrix} p_R \\ p_U \end{pmatrix}
+= \begin{pmatrix} 0 \\ r p_U \end{pmatrix},
+\label{eq:Rm0_model1_appdx}
+\end{equation}
+so we need not bother finding $p_R$. Then we have
+\begin{equation}
+(\mathbf{K} - \mathbf{R_D} + \mathbf{R_A}) \langle\vec{m}^0\rangle
+= \begin{pmatrix} -k_R^- & k_R^+ \\ k_R^- & -k_R^+ \end{pmatrix}
+    \begin{pmatrix} p_R \\ p_U \end{pmatrix} = 0.
+\end{equation}
+As mentioned earlier in Section~\ref{sec:m0_appdx}, the two rows are linearly
+dependent, so taking only the first row and using normalization to set
+$p_R = 1-p_U$ gives
+\begin{equation}
+-k_R^- (1-p_U) + k_R^+ p_U = 0,
+\end{equation}
+which is easily solved to find
+\begin{equation}
+p_U = \frac{k_R^-}{k_R^- + k_R^+}.
+\end{equation}
+Substituting this into Eq.~\ref{eq:Rm0_model1_appdx}, and the result of that
+into Eq.~\ref{eq:mean_m_appdx_ref}, we have
+\begin{equation}
+\langle{m}\rangle = \frac{r}{\gamma} \frac{k_R^-}{k_R^- + k_R^+}
+\end{equation}
+as asserted in Eq.~\ref{eq:mean_m_model1} of the main text.
+
+\subsubsection{Fano factor}
+To verify that the Fano factor for model one in
+Figure~\ref{fig2:constit_cartoons} is in fact 1 as claimed in the main text,
+note that in this limit $p_U = 1$ and $\langle{m}\rangle = r/\gamma$. All
+elements of $\mathbf{K}$ are zero, and $\mathbf{R_A}-\mathbf{R_D} = 0$,
+so Eq.~\ref{eq:vecm_appdx_ref} reduces to
+\begin{equation}
+- \gamma \langle\vec{m}\rangle = - r,
+\end{equation}
+or, in other words, since there is only one promoter state,
+$\langle\vec{m}\rangle = \langle{m}\rangle$. Then it follows that
+\begin{equation}
+\nu = 1 -\frac{r}{\gamma}
+    + \frac{r \langle{m}\rangle}{\gamma \langle{m}\rangle}
+= 1
+\end{equation}
+as claimed.
 
 \subsection{Nonequilibrium Model Two - RNAP Bound and Unbound States}
-\mmnote{Obtain mean mRNA and Fano factor from CME.}
+\subsubsection{Mean mRNA}
+As shown earlier, the full master equation for model two in
+Figure~\ref{fig1:means_cartoons} is
+\begin{align}
+\begin{split}
+\deriv{t}p_R(m,t) =& 
+- \overbrace{k_R^- p_R(m,t)}^{R \rightarrow U}
++ \overbrace{k_R^+ p_E(m,t)}^{U \rightarrow R}
++ \overbrace{(m+1)\gamma p_R(m+1,t)}^{m + 1 \rightarrow m}
+- \overbrace{\gamma mp_R(m,t)}^{m \rightarrow m - 1}
+\\
+\deriv{t}p_E(m,t) =&\; 
+    \overbrace{k_R^- p_R(m,t)}^{R \rightarrow U}
+- \overbrace{k_R^+ p_E(m,t)}^{U \rightarrow R}
++ \overbrace{(m+1)\gamma p_E(m+1,t)}^{m + 1 \rightarrow m}
+- \overbrace{\gamma mp_E(m,t)}^{m \rightarrow m - 1}.
+\\
+&
++ \overbrace{k_P^- p_P(m,t)}^{A \rightarrow U}
+- \overbrace{k_P^+ p_E(m,t)}^{U \rightarrow A}
++ \overbrace{rp_P(m-1,t)}^{m-1 \rightarrow m,\;A \rightarrow U}
+\\
+\deriv{t}p_P(m,t) =&\; 
+- \overbrace{k_P^- p_P(m,t)}^{A \rightarrow U}
++ \overbrace{k_P^+ p_E(m,t)}^{U \rightarrow A}
++ \overbrace{(m+1)\gamma p_P(m+1,t)}^{m + 1 \rightarrow m}
+- \overbrace{\gamma mp_P(m,t)}^{m \rightarrow m - 1}.
+\\
+&- \overbrace{rp_P(m,t)}^{m \rightarrow m + 1,\;A \rightarrow U},
+\end{split}
+\end{align}
+which can be condensed to the matrix form of Eq.~\ref{eq:generic_cme_appdx}
+with matrices given by
+\begin{equation}
+\mathbf{K} = \begin{pmatrix} -k_R^- & k_R^+ & 0 \\
+                        k_R^- & -k_R^+ -k_P^+ & k_P^- \\
+                        0 & k_P^+ & -k_P^- 
+                \end{pmatrix},\
+\mathbf{R_A} = \begin{pmatrix}
+                0 & 0 & 0 \\ 
+                0 & 0 & r \\ 
+                0 & 0 & 0
+                \end{pmatrix},\
+\mathbf{R_D} = \begin{pmatrix}
+                0 & 0 & 0 \\ 
+                0 & 0 & 0 \\ 
+                0 & 0 & r
+                \end{pmatrix}.
+\label{eq:model2_matrices_appdx}
+\end{equation}
+As for model one, we must first find $\mathbf{R_A} \langle\vect{m}^0\rangle$.
+Denote its components as $p_R$, $p_E$, $p_P$, the probabilities of being
+found in repressor bound, empty, or RNAP-bound states, respectively.
+Only $p_P$ is necessary to find since
+\begin{equation}
+\mathbf{R_A} \langle\vec{m}^0\rangle
+= \begin{pmatrix} 0 \\ r p_P \\ 0 \end{pmatrix}.
+\label{eq:model2_deriv_appdx_020}
+\end{equation}
+Then Eq.~\ref{eq:vecm0_appdx_ref} for $\langle\vect{m}\rangle$ reads
+\begin{equation}
+\begin{pmatrix} -k_R^- & k_R^+ & 0 \\
+        k_R^- & -k_R^+ -k_P^+ & k_P^- + r\\
+        0 & k_P^+ & -k_P^- - r
+\end{pmatrix}
+\begin{pmatrix}
+    p_R \\ p_E \\ p_P
+\end{pmatrix}
+= 0.
+\label{eq:model2_K-R+R_for_m0}
+\end{equation}
+Discarding the middle row as redundant and incorporating the normalization
+condition leads to a set of three linearly independent equations, namely
+\begin{align}
+-k_R^- p_R + k_R^+ p_E &= 0 \\
+k_P^+ p_E + (-k_P^- - r) p_P &= 0 \\
+p_R + p_E + p_P &= 1.
+\end{align}
+Using $p_R = 1 - p_E - p_P$ to eliminate $p_R$ in the first and solving the
+resulting equation for $p_E$ gives
+$p_E = (1 - p_P){k_R^-}/{(k_R^- + k_R^+)}$.
+Substituting this for $p_E$ in the second equation gives an equation in
+$p_P$ alone which is
+\begin{equation}
+k_P^+ k_R^- (1-p_P) - (k_P^- + r)(k_R^+ + k_R^-) p_P = 0
+\end{equation}
+and solving for $p_P$ gives
+\begin{equation}
+p_P = \frac{k_P^+ k_R^-}{k_P^+ k_R^- + (k_P^- + r)(k_R^+ + k_R^-)}.
+\end{equation}
+Substituting this in Eq.~\ref{eq:model2_deriv_appdx_020} and multiplying by $\mathbf{R_A}$ produces
+\begin{equation}
+\mathbf{R_A} \langle\vec{m}^0\rangle
+= r \frac{k_P^+ k_R^-} {k_P^+ k_R^- + (k_P^- + r)(k_R^+ + k_R^-)}
+\begin{pmatrix} 0 \\ 1 \\ 0 \end{pmatrix}
+\end{equation}
+from which $\langle{m}\rangle$ follows readily,
+\begin{equation}
+\langle{m}\rangle = \frac{r}{\gamma}
+        \frac{k_P^+ k_R^-} {k_P^+ k_R^- + (k_P^- + r)(k_R^+ + k_R^-)},
+\label{eq:model2_meanm_appdx}
+\end{equation}
+as claimed in Eq.~\ref{eq:model2_meanm} in the main text.
+
+\subsubsection{Fano factor}
+To compute the Fano factor, we first remove the repressor bound state from
+the matrices describing the model, which reduce to
+\begin{equation}
+\mathbf{K} = \begin{pmatrix}
+                -k_P^+ & k_P^- \\
+                 k_P^+ &-k_P^- 
+                \end{pmatrix},\
+\mathbf{R_A} = \begin{pmatrix}
+                0 & r \\ 
+                0 & 0
+                \end{pmatrix},\
+\mathbf{R_D} = \begin{pmatrix}
+                0 & 0 \\ 
+                0 & r
+                \end{pmatrix}.    
+\end{equation}
+Similarly we remove the repressor bound state from
+$\mathbf{R_A} \langle\vec{m}^0\rangle$ and take the $k_R^+\rightarrow 0$
+limit, which simplifies to 
+\begin{equation}
+\mathbf{R_A} \langle\vec{m}^0\rangle
+= r \frac{k_P^+ } {k_P^+ + k_P^- + r}
+\begin{pmatrix} 1 \\ 0 \end{pmatrix}.
+\end{equation}
+Then we must compute $\langle\vec{m}\rangle$ from Eq.~\ref{eq:vecm_appdx_ref},
+which with these matrices reads
+\begin{equation}
+(\mathbf{K} - \mathbf{R_D} + \mathbf{R_A} - \gamma\mathbf{I})
+    \langle\vec{m}\rangle
+= \begin{pmatrix}
+    -k_P^+ -\gamma & k_P^- + r\\
+    k_P^+          &-k_P^- - r - \gamma
+    \end{pmatrix}
+    \begin{pmatrix} m_E \\ m_P \end{pmatrix}
+= r \frac{k_P^+ } {k_P^+ + k_P^- + r}
+    \begin{pmatrix} 1 \\ 0 \end{pmatrix},
+\end{equation}
+where we labeled the components of $\langle\vec{m}\rangle$ as $m_E$ and $m_P$,
+since they are the mean mRNA counts conditional upon the system residing in
+the empty or polymerase bound states, respectively. Unlike for
+$\langle\vec{m}^0\rangle$, the rows of this matrix are linearly independent
+so we simply solve this matrix equation as is. We can immediately eliminate
+$m_E$ since $m_E = m_P (k_P^- + r + \gamma)/k_P^+$ from the second row, and
+substituting into the first row gives an equation for $m_P$ alone, which is
+\begin{equation}
+\left[-(k_P^+ + \gamma)(k_P^- + r + \gamma) + k_P^+(k_P^- + r)\right] m_P
+= - \frac{r (k_P^+)^2}{k_P^+ + k_P^- + r}.
+\end{equation}
+Expanding the products cancels several terms, and solving for $m_P$ gives
+\begin{equation}
+m_P = \frac{r (k_P^+)^2}
+            {\gamma(k_P^+ + k_P^- + r)(k_P^+ + k_P^- + r + \gamma)}.
+\end{equation}
+Note then that $\vec{1}^\dagger\cdot\mathbf{R_A}\langle\vec{m}\rangle = rm_P$.
+We also need the constitutive limit of $\langle{m}\rangle$ from
+Eq.~\ref{eq:model2_meanm_appdx}, again found by taking $k_R^+\rightarrow0$,
+which is
+\begin{equation}
+\langle{m}\rangle = \frac{r}{\gamma} \frac{k_P^+ } {k_P^+ + k_P^- + r}
+\end{equation}
+and substituting this along with
+$\vec{1}^\dagger\cdot\mathbf{R_A}\langle\vec{m}\rangle = rm_P$ into
+Eq.~\ref{eq:fano_appdx_ref} for the Fano factor $\nu$, we find
+\begin{equation}
+\nu = 1 - \frac{r}{\gamma} \frac{k_P^+ } {k_P^+ + k_P^- + r}
+    + \frac{r}{\gamma}\frac{r (k_P^+)^2}{\gamma(k_P^+ + k_P^- + r)
+                                        (k_P^+ + k_P^- + r + \gamma)}
+\left(\frac{r}{\gamma} \frac{k_P^+ } {k_P^+ + k_P^- + r}\right)^{-1}.
+\end{equation}
+This simplifies to
+\begin{equation}
+\nu = 1 - \frac{r}{\gamma}
+    \left(
+        \frac{k_P^+ } {k_P^+ + k_P^- + r}
+        - \frac{k_P^+ } {k_P^+ + k_P^- + r + \gamma}
+    \right),
+\end{equation}
+which further simplifies to
+\begin{equation}
+\nu = 1 - \frac{r k_P^+ } {(k_P^+ + k_P^- + r)(k_P^+ + k_P^- + r + \gamma)},
+\end{equation}
+exactly Eq.~\ref{eq:model2_fano} in the main text.
 
@@
     - Multistep Transcription Initiation and Escape}
-\mmnote{Derive mean mRNA from CME.}
-\newline
-\mmnote{Obtain Fano factor by taking limit from Mitarai et al's expression.}
+\subsubsection{Mean mRNA}
+In close analogy to model two above, nonequilibrium model three from
+Figure~\ref{fig1:means_cartoons} can be described by our generic master equation
+Eq.~\ref{eq:generic_cme_appdx} with promoter transition matrix given by
+\begin{equation}
+\mathbf{K} =
+\begin{pmatrix} -k_R^- & k_R^+ & 0 & 0\\
+        k_R^- & -k_R^+ -k_P^+ & k_P^- & 0 \\
+        0 & k_P^+ & -k_P^- - k_O & 0 \\
+        0 & 0 & k_O & 0
+\end{pmatrix}
+\end{equation}
+and transcription matrices given by
+\begin{equation}
+\mathbf{R_A} =
+\begin{pmatrix}
+        0 & 0 & 0 & 0 \\ 
+        0 & 0 & 0 & r \\ 
+        0 & 0 & 0 & 0 \\ 
+        0 & 0 & 0 & 0
+\end{pmatrix},\
+\mathbf{R_D} =
+\begin{pmatrix}
+        0 & 0 & 0 & 0 \\ 
+        0 & 0 & 0 & 0 \\ 
+        0 & 0 & 0 & 0 \\ 
+        0 & 0 & 0 & r
+\end{pmatrix}.
+\end{equation}
+$\langle\vec{m}^0\rangle$ is again given by Eq.~\ref{eq:vecm0_appdx_ref},
+which in this case takes the form
+\begin{equation}
+(\mathbf{K} - \mathbf{R_D} + \mathbf{R_A}) \langle\vec{m}^0\rangle =
+\begin{pmatrix} -k_R^- & k_R^+ & 0 & 0\\
+    k_R^- & -k_R^+ -k_P^+ & k_P^- & r \\
+    0 & k_P^+ & -k_P^- - k_O & 0 \\
+    0 & 0 & k_O & - r
+\end{pmatrix}
+\begin{pmatrix} p_R \\ p_E \\ p_C \\ p_O
+\end{pmatrix} = 0,
+\end{equation}
+where the four components of $\langle\vec{m}^0\rangle$ correspond to the
+four promoter states repressor bound, empty, RNAP-bound closed complex, and
+RNAP-bound open complex. As explained in Section~\ref{sec:m0_appdx}, we
+are free to discard one linearly dependent row from this matrix and replace
+it with the normalization condition $p_R + p_E + p_C + p_O = 1$.
+Using normalization to eliminate $p_R$ from the first row gives
+\begin{equation}
+p_E = (1 - p_C - p_O)\frac{k_R^-}{k_R^- + k_R^+}.
+\end{equation}
+If we substitute this in the third row, then the last two rows constitute
+two equations in $p_C$ and $p_O$ given by
+\begin{align}
+k_P^+k_R^-(1-p_C-p_O) - (k_P^- + k_O)(k_R^+ + k_R^-) p_C &= 0
+\\
+k_O p_C - r p_O &= 0.
+\end{align}
+Solving for $p_C = p_O r/k_O$ in the second and substituting into the first
+gives us our desired single equation in the single variable $p_O$, which is
+\begin{equation}
+k_P^+k_R^- - k_P^+k_R^-\left(1 + \frac{r}{k_O}\right)p_O
+            - (k_P^- + k_O)(k_R^+ + k_R^-) \frac{r}{k_O}p_O = 0,
+\end{equation}
+and solving for $p_O$ we find
+\begin{equation}
+p_O = \frac{k_P^+ k_R^- k_O}{k_P^+ k_R^- k_O + r k_P^+ k_R^- +
+                            r (k_P^- + k_O) (k_R^+ + k_R^-)}.
+\label{eq:model3_pO}
+\end{equation}
+Once again $p_O$, the transcriptionally active state, is the only component
+of $\langle\vec{m}^0\rangle$ that survives multiplication by $\mathbf{R_A}$,
+and $\mathbf{R_A}\langle\vec{m}^0\rangle = r p_O$. So
+\begin{equation}
+\langle{m}\rangle =
+    \frac{1}{\gamma}\vec{1}^\dagger\cdot\mathbf{R_A}\langle\vec{m}^0\rangle
+= \frac{r}{\gamma}
+    \frac{k_P^+ k_R^- k_O}{k_P^+ k_R^- k_O + r k_P^+ k_R^- +
+                            r (k_P^- + k_O) (k_R^+ + k_R^-)},
+\end{equation}
+which equals Eq.~\ref{eq:model3_mean_m} in the main text.
+
+\subsubsection{Fano factor}
+To compute the Fano factor of the analogous constitutive promoter, we first
+excise the repressor states and rates from the problem. More precisely, we
+construct the matrix
+$(\mathbf{K} - \mathbf{R_D} + \mathbf{R_A} - \gamma\mathbf{I})$ and
+substitute it into Eq.~\ref{eq:vecm_appdx_ref} which is now
+\begin{equation}
+(\mathbf{K} - \mathbf{R_D} + \mathbf{R_A} - \gamma\mathbf{I})
+    \langle\vec{m}\rangle
+= \begin{pmatrix}
+    -k_P^+ - \gamma & k_P^- & r \\
+     k_P^+ & -k_P^- - k_O - \gamma & 0 \\
+     0 & k_O & - r- \gamma
+\end{pmatrix}
+\begin{pmatrix} m_E \\ m_C \\ m_O \end{pmatrix}
+= -r p_O \begin{pmatrix}1 \\ 0 \\ 0 \end{pmatrix}
+\end{equation}
+where we labeled the unbound, closed complex, and open complex componenets of
+$\langle\vec{m}\rangle$ as $m_E$, $m_C$, and $m_O$, respectively. $p_O$
+is given by the limit of Eq.~\ref{eq:model3_pO} as $k_R^+\rightarrow 0$,
+which is
+\begin{equation}
+p_O = \frac{k_P^+  k_O}{k_P^+ (k_O + r) + r (k_P^- + k_O)}
+\equiv \frac{k_P^+  k_O}{\mathcal{Z}},
+\end{equation}
+where we define $\mathcal{Z}$ for upcoming convenience as this sum of terms
+will appear repeatedly. We can use the second equation to eliminate $m_E$,
+finding $m_E = m_C(k_P^- + k_O + \gamma)/k_P^+$, and the third to eliminate
+$m_C$, which is simply $m_C = m_O(r+\gamma)/k_O$. Substituting these both into
+the first equation gives a single equation for the variable of interest, $m_O$,
+\begin{equation}
+-(k_P^+ + \gamma) (k_P^- + k_O + \gamma) (r + \gamma) m_O
+    + k_P^- k_P^+ (r + \gamma) m_O + r k_P^+ k_O m_O = - r k_P^+ k_O p_O,
+\end{equation}
+which is solved for $m_O$ to give
+\begin{equation}
+m_O = p_O \frac{r k_P^+ k_O}
+    {(k_P^+ + \gamma) (k_P^- + k_O + \gamma) (r + \gamma)
+        -r k_P^+ k_O - k_P^- k_P^+ (r + \gamma)}.
+\end{equation}
+Expanding the denominator and canceling terms leads to
+\begin{equation}
+m_O = p_O \frac{r}{\gamma} \frac{k_P^+ k_O}
+    {\mathcal{Z} + \gamma(k_P^+ + k_P^- + k_O + r) + \gamma^2}.
+\end{equation}
+Now $\vec{1}^\dagger\cdot\mathbf{R_A}\langle\vec{m}\rangle = r m_O$, and
+$\langle{m}\rangle = rp_O/\gamma$, so if we substitute these two quantities
+into Eq.~\ref{eq:fano_appdx_ref}, we will readily obtain the Fano factor as
+\begin{equation}
+\nu = 1 - \langle{m}\rangle
+    + \frac{\vec{1}^\dagger\cdot\mathbf{R_A}\langle\vec{m}\rangle}
+            {\gamma \langle{m}\rangle}
+= 1 - \frac{r}{\gamma}p_O + \frac{m_O}{p_O}.
+\end{equation}
+Substituting, we see that
+\begin{equation}
+\nu = 1 - \frac{r}{\gamma} \frac{k_P^+ k_O}{\mathcal{Z}}
+    + \frac{r}{\gamma}
+    \frac{k_P^+ k_O}
+            {\mathcal{Z} + \gamma(k_P^+ + k_P^- + k_O + r) + \gamma^2},
+\end{equation}
+and after simplifying, we obtain
+\begin{equation}
+\nu = 1 - \frac{r k_P^+ k_O}{\mathcal{Z}}
+        \frac{k_P^+ + k_P^- + k_O + r + \gamma}
+            {\mathcal{Z} + \gamma(k_P^+ + k_P^- + k_O + r) + \gamma^2},
+\end{equation}
+as stated in Eq.~\ref{eq:model3_fano} in the main text.
 
@@
     - ``Active'' and ``Inactive'' States}
-\mmnote{Derive mean mRNA from CME.}
-\newline
-\mmnote{Obtain Fano factor by relabeling cartoon from Jones2014.}
+\subsubsection{Mean mRNA}
+The mathematical specification of this model is almost identical to model two.
+The matrix $\mathbf{K}$ is identical, as is $\mathbf{R_D}$. The only
+difference is that now $\mathbf{R_A}=\mathbf{R_D}$, i.e., both are diagonal,
+in contrast to model two where $\mathbf{R_A}$ has an off-diagonal element, as in
+Eq.~\ref{eq:model2_matrices_appdx}.
+Then the analog of Eq.~\ref{eq:model2_K-R+R_for_m0} for finding
+$\langle{m}^0\rangle$ is
+\begin{equation}
+\begin{pmatrix} -k_R^- & k_R^+ & 0 \\
+        k_R^- & -k_R^+ -k_P^+ & k_P^-\\
+        0 & k_P^+ & -k_P^-
+\end{pmatrix}
+\begin{pmatrix}
+    p_R \\ p_E \\ p_P
+\end{pmatrix}
+= 0.
+\end{equation}
+In fact we need not do this calculation explicitly and can instead recycle
+the calculation of mean mRNA $\langle{m}\rangle$ from model two. The
+matrices are identical except for the relabeling
+$k_P^- \longleftrightarrow (k_P^- + r)$, and a careful look through the
+derivation of $\langle{m}\rangle$ for model two shows that the parameters
+$k_P^-$ and $r$ only ever appear as the sum $k_P^- + r$. So taking
+$\langle{m}\rangle$ from model two, Eq.~\ref{eq:model2_meanm_appdx}, and
+relabeling $(k_P^- + r) \rightarrow k_P^-$ gives us our answer for model
+four, simply
+\begin{equation}
+\langle{m}\rangle = \frac{r}{\gamma}
+        \frac{k_P^+ k_R^-} {k_P^+ k_R^- + k_P^- (k_R^+ + k_R^-)}.
+\end{equation}
+
+\subsubsection{Fano factor}
+Likewise, for computing the Fano factor of this model we may take a shortcut.
+Consider the constitutive model four from Figure~\ref{fig2:constit_cartoons}
+for which we want to compute the Fano factor and compare it to nonequilibrium
+model one of simple repression in Figure~\ref{fig1:means_cartoons}.
+Mathematically these are exactly the same model, just with rates labeled
+differently and the meaning of the promoter states interpreted differently.
+Furthermore, nonequilibrium model one from Figure~\ref{fig1:means_cartoons}
+was the model considered by Jones et.\ al.~\cite{Jones2014}, where they
+derived the Fano factor for that model to be
+\begin{equation}
+\nu = 1 + \frac{r k_R^+}{(k_R^+ + k_R^-)(k_R^+ + k_R^- + \gamma)}.
+\end{equation}
+So recognizing that the relabelings $k_R^+ \rightarrow k^-$ and
+$k_R^- \rightarrow k^+$ will translate this result to our model four from
+Figure~\ref{fig2:constit_cartoons}, we can immediately write down our Fano
+factor as
+\begin{equation}
+\nu = 1 + \frac{r k^-}{(k^- + k^+)(k^- + k^+ + \gamma)},
+\end{equation}
+as quoted in Eq.~\ref{eq:model4_fano} and in Figure~\ref{fig2:constit_cartoons}.

### 2020-06-03 doc/section_01_introduction.tex

@@
 to push the theory-experiment dialogue as far and as hard as it can be
-pushed~\cite{Phillips2019}.  If we demonstrate that we can pass that test by
+pushed~\cite{Garcia2011,Phillips2019}.  If we demonstrate that we can pass that test by
 successfully predicting both the means and variance in gene expression, then
@@
 a DNA promoter regulated by a transcriptional repressor that binds to a single
-binding site \cite{Garcia2011a}. All of the proposed models models coarse-grain
+binding site as developed in pioneering early work on
+the quantitative dissection of transcription \cite{Oehler1994, Oehler1990}. All of the proposed models models coarse-grain
 away some of the important microscopic features of this architecture that have

### 2020-06-03 doc/section_01_introduction.tex
MRM pass through intro and discussion. removed most color and commented text.

@@
 Gene expression presides over much of the most important dynamism of living
-organisms.   The level of expression of batteries of different genes is altered
-as a result of spatiotemporal cues that integrate chemical, mechanical and other
+organisms. The level of expression of batteries of different genes is altered as
+a result of spatiotemporal cues that integrate chemical, mechanical and other
 types of signals. The original repressor-operator model conceived by Jacob and
@@
 architectures, i.e. the arrangement and affinity of binding sites for
-transcription factors and RNA polymerases on the DNA. This challenge stems
-first and foremost from our ignorance about what those architectures even are
-with more than 60\% of the genes even in an ostensibly well understood organism
-such as  {\it E. coli} with no regulatory insights at all~\cite{Rydenfelt2014-2,
+transcription factors and RNA polymerases on the DNA. This challenge stems first
+and foremost from our ignorance about what those architectures even are with
+more than 60\% of the genes even in an ostensibly well understood organism such
+as {\it E. coli} with no regulatory insights at all~\cite{Rydenfelt2014-2,
 Belliveau2018, Ghatak2019, Santos_Zavaleta2019}. But even once we have
@@
 Shea1985, Buchler2003, Vilar2003a, Vilar2003b, Bintu2005a, Bintu2005c,
-Gertz2009,Sherman2012, Saiz2013, Ko1991, Peccoud1995, Record1996, Kepler2001,
+Gertz2009, Sherman2012, Saiz2013, Ko1991, Peccoud1995, Record1996, Kepler2001,
 Sanchez2008, Shahrezaei2008, Sanchez2011, Michel2010}. One root to such
@@
 to push the theory-experiment dialogue as far and as hard as it can be
-pushed~\cite{Garcia2011,Phillips2019}.  If we demonstrate that we can pass that test by
-successfully predicting both the means and variance in gene expression, then
-that provides a more solid foundation upon which to launch into more complex
-problems.
+pushed~\cite{Garcia2011, Phillips2019}. If we demonstrate that we can pass that
+test by successfully predicting both the means and variance in gene expression,
+then that provides a more solid foundation upon which to launch into more
+complex problems.
 
@@
 a DNA promoter regulated by a transcriptional repressor that binds to a single
-binding site as developed in pioneering early work on
-the quantitative dissection of transcription \cite{Oehler1994, Oehler1990}. All of the proposed models models coarse-grain
-away some of the important microscopic features of this architecture that have
-been elucidated by generations of geneticists, molecular biologists and
-biochemists. Our aim is to understand the underlying principles of such
-coarse-graining in order to increase the predictive power of these models. In
-doing so we hope future models of regulatory response will be able to serve the
-powerful predictive role needed to take synthetic biology from a brilliant
-exercise in enlightened empiricism to a rational design framework as  in any
-other branch of engineering. More precisely, we want phenomenology in the sense
-of coarse-graining away atomistic detail, but still retaining biophysical
+binding site as developed in pioneering early work on the quantitative
+dissection of transcription \cite{Oehler1994, Oehler1990}. All of the proposed
+models models coarse-grain away some of the important microscopic features of
+this architecture that have been elucidated by generations of geneticists,
+molecular biologists and biochemists. Our aim is to understand the underlying
+principles of such coarse-graining in order to increase the predictive power of
+these models. In doing so we hope future models of regulatory response will be
+able to serve the powerful predictive role needed to take synthetic biology from
+a brilliant exercise in enlightened empiricism to a rational design framework as
+in any other branch of engineering. More precisely, we want phenomenology in the
+sense of coarse-graining away atomistic detail, but still retaining biophysical
 meaning. For example, we are not satisfied with the strictly phenomenological
-approach offered by the commonly used Hill functions~\cite{Frank2013}.
-Studies like~\cite{Razo-Mejia2018} have demonstrated that Hill functions are
-clearly insufficient since each new situation requires a completely new set of
+approach offered by the commonly used Hill functions~\cite{Frank2013}. Studies
+like~\cite{Razo-Mejia2018} have demonstrated that Hill functions are clearly
+insufficient since each new situation requires a completely new set of
 parameters. Such work requires a quantitative theory of how biophysical changes
@@
 circuit level. We want concepts, not mere facts. In particular a key question
-is: at this level of coarse graining, what microscopic details do we need to
+is: at this level of coarse-graining, what microscopic details do we need to
 explicitly model, and how do we figure that out? For example, do we need to
-worry about all or even any of the steps that individual RNAPs go through each
-time they make a transcript? Turning the question around, can we see any imprint
-of those processes in the available data? If the answer is no, then those
-processes are irrelevant for our purposes. Forward modeling and inverse
-(statistical inferential) modeling are necessary to tackles such questions.
+worry about all or even any of the steps that individual RNA polymerases go
+through each time they make a transcript? Turning the question around, can we
+see any imprint of those processes in the available data? If the answer is no,
+then those processes are irrelevant for our purposes. Forward modeling and
+inverse (statistical inferential) modeling are necessary to tackles such
+questions.
 
@@
 expression as shown in Figure~\ref{fig1:means_cartoons}(D).
-Section~\ref{sec:beyond_means} goes beyond an analysis of the mean
-gene expression by asking how the same models presented in
+Section~\ref{sec:beyond_means} goes beyond an analysis of the mean gene
+expression by asking how the same models presented in
 Figure~\ref{fig1:means_cartoons}(C) can be used to explore noise in gene
@@
 commitment to some numerical values for the key parameters found in each such
-model. Therefore in Section~\ref{section_04_bayesian_inference} we explore
-the use of Bayesian inference to establish these parameters and to rigorously
-answer the question of how to discriminate between the different models.
+model. Therefore in Section~\ref{section_04_bayesian_inference} we explore the
+use of Bayesian inference to establish these parameters and to rigorously answer
+the question of how to discriminate between the different models.
 

### 2020-06-03 doc/section_05_discussion.tex
MRM pass through intro and discussion. removed most color and commented text.

--- 2020-06-02 doc/section_05_discussion.tex
@@
 variability, and the full gene expression distribution found in such population
-as would be reported in a single molecule Fluorescence in situ Hybridization
-experiment, for example.
+as would be reported in a single molecule mRNA Fluorescence in situ
+Hybridization experiment, for example.
 
@@
 architecture is captured by a master curve in which the action of repressor and
-the details of the polymerase interaction with the promoter appear separately
-and additively in an effective free energy. Interestingly, as has been shown
+the details of the RNAP interaction with the promoter appear separately and
+additively in an effective free energy. Interestingly, as has been shown
 elsewhere in the context of the Monod-Wyman-Changeux model, these kinds of
@@
 that assumed an equilibrium rather than a kinetic
-framework~\cite{Garcia2011a,Razo-Mejia2018}.
+framework~\cite{Garcia2011a, Razo-Mejia2018}.
 
@@
 
-\mrm{We need to decide ASAP if this is going to be included in the discussion or not. My vote is no, but Jupiter must decide.}
-\mmnote{Do all my approximations work simply because galK happens to have a low
-duty cycle of supercoiling, $\beta/\alpha$ in Chong2014 picture? A potentially
-interesting test: put in the dozen or so Hernan constructs at some other locus,
-preferably one of the ones Chong identified as having a large duty cycle (large
-$\beta/\alpha$), i.e., the opposite limit of galK. Then just measure
-fold-changes. You don't even need to do FISH, although that would be even
-better. If their model is right, $\rho$ should be approximately set by the duty
-cycle of supercoiling, NOT by RNAP kinetics.
-\textit{This is a NICE prediction for future work.} \textit{cspE}
-is claimed nonessential, prediction is you drop HG reporter constructs in there
-and the inferred binding energies of all lacI operators shift the same amount,
-by about $\ln(3)\sim 1kT$. Not huge, but detectable, although might be
-confounding effects from moving to a new locus. With FISH measurements, $k^+$
-and $k^-$ in Fig 2, model 4 should now be separately inferrable if Chong2014
-measurements of $\beta/\alpha$ are correct. This would require an obnoxious
-amount of cloning, but once you did that, the measurements would be same old
-stuff that RP lab does.}
-
-A direct \mmnote{if cloning intensive} future experiment would be
-to design a payload with a promoter driving a FISH reporter,
-integrate that payload at a variety of loci scattered around the genome,
-and carry out FISH measurements of the resulting strains.
-\mmnote{
-Did Brewster already do this? I remember he integrated arrays of
-competitor binding sites all over the genome, but did he put
-lacUV5 + reporter scattered around?}
-Obviously integration loci would need to be carefully chosen so
-that one could plausibly hope the deletion and integration did
-not disturb the cell's physiology too badly.
-
-\mrm{Rather than proposing more experiments, I feel that we are missing a closing paragraph that puts all these findings in context.
-Where do we go from here as a community?
-Things such as the next building block, i.e.
-simple activation, and then from there scale up the complexity until we can get a full gene regulatory network kind of thing.
-Rather than going small into tiny details specific to this, let's step back and look at the big picture.}
+In summary, this work took up the challenge of exploring the extent to which a
+single specific mechanistic model of the simple-repression regulatory
+architecture suffices to explain the broad sweep of experimental data for this
+system. Pioneering early experimental efforts from the M\"{u}ller-Hill lab
+established the simple-repression motif as an arena for the quantitative
+dissection of regulatory response in bacteria, with similar beautiful work
+emerging in examples such as the arabinose and gal operons as
+well~\cite{Dunn1984b, Oehler1990, Weickert1993, Oehler1994, Schleif2000,
+Semsey2002, SwintKruse2009}. In light of a new generation of precision
+measurements on these systems, the definition of what it means to understand
+them can now be formulated as a rigorous quantitative question. In particular,
+many aspects of the simple repression motif seem to merit the claim that they
+are sufficiently well understood to use that understanding to design new
+versions of that architecture based upon predictions about how repressor copy
+number and DNA binding site strength control expression. In our view, the next
+step in the progression is to first perform similar rigorous analyses of the
+fundamental ``basis set'' of regulatory architectures. Natural followups to this
+work is the exploration of motifs such as simple activation that is
+regulated by a single activator binding site, and the repressor-activator
+architecture, mediated by the binding of both a single activator and a single
+repressor and beyond. With the individual input-output functions in hand,
+similar quantitative dissections including the rigorous analysis of their tuning
+parameters can be undertaken for the ``basis set'' of full gene-regulatory
+networks such as switches, feed-forward architectures and oscillators for
+example, building upon the recent impressive bonanza of efforts from systems
+biologists and synthetic biologists~\cite{Milo2002, Alon2007}.

### 2020-06-03 doc/appendix_01_nonbursty.tex
MRM quick pass through appendix 1. looks really good

@@
 
-In this section we detail the calculation of mean mRNA levels,
-fold-changes in expression, and Fano factors for nonequilibrium
-promoter models one through four in Figure~\ref{fig1:means_cartoons}.
-These are the results that were quoted but not derived in
-Sections~\ref{section_02_means} and \ref{sec:beyond_means} of the
-main text. In each of these four models, the natural
-mathematization of their cartoons is as a chemical master
-equation such as Eq.~\ref{eq:2state_rep_cme} for model one, which
-for completeness we reproduce here as
+\mrm{I personally think that the paper so far has done a great job at being very 
+pedagogical. That is why I am of the opinion that before jumping into the 
+derivation of moments and so on there needs to be a $p(m, t + \Delta t)$ kind of 
+derivation.}
+
+In this section we detail the calculation of mean mRNA levels, fold-changes in
+expression, and Fano factors for nonequilibrium promoter models one through four
+in Figure~\ref{fig1:means_cartoons}. These are the results that were quoted but
+not derived in Sections~\ref{section_02_means} and \ref{sec:beyond_means} of the
+main text. In each of these four models, the natural mathematization of their
+cartoons is as a chemical master equation such as Eq.~\ref{eq:2state_rep_cme}
+for model one, which for completeness we reproduce here as
 \begin{align}
@@
 \end{align}
-Here $p_R(m,t)$ and $p_U(m,t)$ are the probabilities of finding
-the system with $m$ mRNA molecules at time $t$ either in the
-repressor bound or unbound states, respectively. $r$ is the
-probability per unit time that a transcript will be initiated
-when the repressor is unbound, and $\gamma$ is the probability
-per unit time for a given mRNA to be degraded. $k_R^-$ is the
-probability per unit time that a bound repressor will unbind,
-while $k_R^+$ is the probability per unit time that an unbound
-operator will become bound by a repressor. Assuming mass action
-kinetics, $k_R^+$ is proportional to repressor copy number $R$.
+Here $p_R(m,t)$ and $p_U(m,t)$ are the probabilities of finding the system with
+$m$ mRNA molecules at time $t$ either in the repressor bound or unbound states,
+respectively. $r$ is the probability per unit time that a transcript will be
+initiated when the repressor is unbound, and $\gamma$ is the probability per
+unit time for a given mRNA to be degraded. $k_R^-$ is the probability per unit
+time that a bound repressor will unbind, while $k_R^+$ is the probability per
+unit time that an unbound operator will become bound by a repressor. Assuming
+mass action kinetics, $k_R^+$ is proportional to repressor copy number $R$.
 
 Next consider the cartoon for nonequilibrium model two in
-Figure~\ref{fig1:means_cartoons}. Now we must track probabilities
-$p_R$, $p_P$, and $p_E$ for the repressor bound, empty, and
-polymerase bound states, respectively. By analogy
-to~\ref{eq:poisson_promoter_cme_appdx}, the master equation for
-model two can be written
+Figure~\ref{fig1:means_cartoons}(C). Now we must track probabilities $p_R$,
+$p_P$, and $p_E$ for the repressor bound, empty, and polymerase bound states,
+respectively. By analogy to~\ref{eq:poisson_promoter_cme_appdx}, the master
+equation for model two can be written
 \begin{align}
@@
 \end{align}
-$k_P^+$ and $k_P^-$ are defined in close analogy to $k_R^+$ and
-$k_R^-$, except for RNAP binding and unbinding instead of
-repressor. Similarly $p_P(m,t)$ is defined for the active
-(RNAP-bound) state exactly as are $p_R(m,t)$ and $p_E(m,t)$ for
-the repressor bound and unbound states, respectively.
-The new subtlety Eq.~\ref{eq:model2_cme_appdx} introduces compared to
-Eq.~\ref{eq:poisson_promoter_cme_appdx} is that when mRNAs are
-produced, the promoter state also changes. This is encoded by the
-terms involving $r$, the last term in each of the equations for
-$p_E$ and $p_P$. The former accounts for arrivals in the
-unbound state and the latter accounts for departures from the
+$k_P^+$ and $k_P^-$ are defined in close analogy to $k_R^+$ and $k_R^-$, except
+for RNAP binding and unbinding instead of repressor. Similarly $p_P(m,t)$ is
+defined for the active (RNAP-bound) state exactly as are $p_R(m,t)$ and
+$p_E(m,t)$ for the repressor bound and unbound states, respectively. The new
+subtlety Eq.~\ref{eq:model2_cme_appdx} introduces compared to
+Eq.~\ref{eq:poisson_promoter_cme_appdx} is that when mRNAs are produced, the
+promoter state also changes. This is encoded by the terms involving $r$, the
+last term in each of the equations for $p_E$ and $p_P$. The former accounts for
+arrivals in the unbound state and the latter accounts for departures from the
 RNAP-bound state.
 
-To condense and clarify the unwieldy notation of
-Eq.~\ref{eq:model2_cme_appdx}, it can be rewriten in matrix form. We define the column vector $\vec{p}(m,t)$ as
+To condense and clarify the unwieldy notation of Eq.~\ref{eq:model2_cme_appdx},
+it can be rewritten in matrix form. We define the column vector $\vec{p}(m,t)$
+as
 \begin{equation}
@@
 \end{equation}
-to gather, for a given $m$, the probabilities of finding the
-system in the three possible promoter states. Then all the
-transition rates may be condensed into matrices which multiply
-this vector. The first matrix is
+to gather, for a given $m$, the probabilities of finding the system in the three
+possible promoter states. Then all the transition rates may be condensed into
+matrices which multiply this vector. The first matrix is
 \begin{equation}
@@
 \end{equation}
-which tracks all promoter state changes in
-Eq.~\ref{eq:model2_cme_appdx} that are \textit{not} accompanied
-by a change in the mRNA copy number. The two terms accounting for
-transcription, the only transition that increases mRNA copy
+which tracks all promoter state changes in Eq.~\ref{eq:model2_cme_appdx} that
+are \textit{not} accompanied by a change in the mRNA copy number. The two terms
+accounting for transcription, the only transition that increases mRNA copy
 number, must be handled by two separate matrices given by
@@
 \end{equation}
-$\mathbf{R_A}$ accounts for transitions \textit{arriving} in a given state
-while $\mathbf{R_D}$ tracks \textit{departing} transitions.
-With these definitions, we can condense Eq.~\ref{eq:model2_cme_appdx} into
-the single equation
+$\mathbf{R_A}$ accounts for transitions \textit{arriving} in a given state while
+$\mathbf{R_D}$ tracks \textit{departing} transitions. With these definitions, we
+can condense Eq.~\ref{eq:model2_cme_appdx} into the single equation
 \begin{equation}
@@
 \end{equation}
-which is just Eq.~\ref{eq:3state_rep_cme} in the main text.
-Straightforward if tedious algebra verifies that
-Eqs.~\ref{eq:model2_cme_appdx} and~\ref{eq:generic_cme_appdx} are
-in fact equivalent.
-
-Although we derived Eq.~\ref{eq:generic_cme_appdx} for the particular case
-of nonequilibrium model two in Figure~\ref{fig1:means_cartoons},
-in fact the chemical master equations for all of the
-nonequilibrium models in Figure~\ref{fig1:means_cartoons} except
-for model five can be cast in this form. (We treat model five
-separately in Section~\ref{sec:gen_fcn_appdx}.)
-Model three introduces no new subtleties beyond model two and
-Eq.~\ref{eq:generic_cme_appdx} applies equally well to it, simply
-with different matrices of dimension four instead of three.
-Models one and four are both handled by Eq.~\ref{eq:2state_rep_cme}
-in the main text, which is just Eq.~\ref{eq:generic_cme_appdx}
-except in the special case of $\mathbf{R_D} = \mathbf{R_A} \equiv
-\mathbf{R}$, since in these two models transcription initiation
-events do not change promoter state.
-
-Recalling that our goal in this section is to derive expressions
-for mean mRNA and Fano factor for nonequilibrium models one
-through four in Figure~\ref{fig1:means_cartoons}, and since all
-four of these models are described by Eq.~\ref{eq:generic_cme_appdx},
-we can save substantial effort by deriving general formulas for
-mean mRNA and Fano factor from Eq.~\ref{eq:generic_cme_appdx}
-once and for all. Then for each model we can simply plug in the
-appropriate matrices for $\mathbf{K}$, $\mathbf{R_D}$, and
+which is just Eq.~\ref{eq:3state_rep_cme} in the main text. Straightforward if
+tedious algebra verifies that Eqs.~\ref{eq:model2_cme_appdx}
+and~\ref{eq:generic_cme_appdx} are in fact equivalent.
+
+Although we derived Eq.~\ref{eq:generic_cme_appdx} for the particular case of
+nonequilibrium model two in Figure~\ref{fig1:means_cartoons}, in fact the
+chemical master equations for all of the nonequilibrium models in
+Figure~\ref{fig1:means_cartoons} except for model five can be cast in this form.
+(We treat model five separately in Appendix~\ref{sec:gen_fcn_appdx}.) Model
+three introduces no new subtleties beyond model two and
+Eq.~\ref{eq:generic_cme_appdx} applies equally well to it, simply with different
+matrices of dimension four instead of three. Models one and four are both
+handled by Eq.~\ref{eq:2state_rep_cme} in the main text, which is just
+Eq.~\ref{eq:generic_cme_appdx} except in the special case of $\mathbf{R_D} =
+\mathbf{R_A} \equiv \mathbf{R}$, since in these two models transcription
+initiation events do not change promoter state.
+
+Recalling that our goal in this section is to derive expressions for mean mRNA
+and Fano factor for nonequilibrium models one through four in
+Figure~\ref{fig1:means_cartoons}, and since all four of these models are
+described by Eq.~\ref{eq:generic_cme_appdx}, we can save substantial effort by
+deriving general formulas for mean mRNA and Fano factor from
+Eq.~\ref{eq:generic_cme_appdx} once and for all. Then for each model we can
+simply plug in the appropriate matrices for $\mathbf{K}$, $\mathbf{R_D}$, and
 $\mathbf{R_A}$ and carry out the remaining algebra.
 
-For our purposes it will suffice to derive the first and second
-moments of the mRNA distribution from this master equation,
-similar to the treatment in~\cite{Sanchez2011}, but we refer the
-interested reader to~\cite{Razo-Mejia2020} for an analogous
-treatment demonstrating an analytical solution for arbitrary moments.
+For our purposes it will suffice to derive the first and second moments of the
+mRNA distribution from this master equation, similar to the treatment
+in~\cite{Sanchez2011}, but we refer the interested reader
+to~\cite{Razo-Mejia2020} for an analogous treatment demonstrating an analytical
+solution for arbitrary moments.
 
@@
 Our task now is to derive expressions for the first two moments of the
-steady-state mRNA distribution from Eq.~\ref{eq:generic_cme_appdx}.
-Our treatment of this is analogous to that given in~\cite{Sanchez2011}
+steady-state mRNA distribution from Eq.~\ref{eq:generic_cme_appdx}. Our
+treatment of this is analogous to that given in~\cite{Sanchez2011}
 and~\cite{Razo-Mejia2020}. We first obtain the steady-state limit of
@@
 \begin{equation}
-\langle{m}\rangle = \sum_S \sum_{m=0}^\infty m p_S(m)
+\langle{m}\rangle = \sum_S \sum_{m=0}^\infty m \, p_S(m)
 \end{equation}
@@
 which define the average values of $m$ and $m^2$ at steady state, where the
-averaging is over all possible mRNA copy numbers and promoter states $S$.
-For example, for model one in Figure~\ref{fig1:means_cartoons}, the sum on $S$
-would cover repressor bound and unbound states ($R$ and $U$ respectively),
-for model two, the sum would cover repressor bound, polymerase bound,
-and empty states ($R$, $P$, and $E$), and so on for the other models.
+averaging is over all possible mRNA copy numbers and promoter states $S$. For
+example, for model one in Figure~\ref{fig1:means_cartoons}(C), the sum on $S$
+would cover repressor bound and unbound states ($R$ and $U$ respectively), for
+model two, the sum would cover repressor bound, polymerase bound, and empty
+states ($R$, $P$, and $E$), and so on for the other models.
 
@@
 \end{equation}
-These objects are vectors of the same size as $\vect{p}(m)$, and each
-component can be thought of as the expected value of the mRNA copy number,
-or copy number squared, conditional on the promoter being in a certain state.
-For example, for model one in Figure~\ref{fig1:means_cartoons} which has
-repressor bound and unbound states labeled $R$ and $U$,
-$\langle\vec{m}^2\rangle$ would be
+These objects are vectors of the same size as $\vect{p}(m)$, and each component
+can be thought of as the expected value of the mRNA copy number, or copy number
+squared, conditional on the promoter being in a certain state. For example, for
+model one in Figure~\ref{fig1:means_cartoons} which has repressor bound and
+unbound states labeled $R$ and $U$, $\langle\vec{m}^2\rangle$ would be
 \begin{equation}
@@
 \end{equation}
-whose elements are simply the probabilities of finding the system
-in each of the possible promoter states. To reduce notational clutter
-in what follows, sums over $m$ will be assumed to run from 0 to $\infty$
-unless otherwise specified, i.e., $\sum_{m=0}^\infty$ will be denoted
-simply as $\sum_m$.
-It will also be convenient to denote by $\vec{1}^\dagger$ a row
-vector of the same length as $\vec{p}$ whose elements are all
-1, such that, for instance,
-$\vec{1}^\dagger \cdot \langle\vec{m}^0\rangle = 1$,
-$\vec{1}^\dagger \cdot \langle\vec{m}\rangle = \langle{m}\rangle$, etc.
+whose elements are simply the probabilities of finding the system in each of the
+possible promoter states. To reduce notational clutter in what follows, sums
+over $m$ will be assumed to run from 0 to $\infty$ unless otherwise specified,
+i.e., $\sum_{m=0}^\infty$ will be denoted simply as $\sum_m$. It will also be
+convenient to denote by $\vec{1}^\dagger$ a row vector of the same length as
+$\vec{p}$ whose elements are all 1, such that, for instance, $\vec{1}^\dagger
+\cdot \langle\vec{m}^0\rangle = 1$, $\vec{1}^\dagger \cdot \langle\vec{m}\rangle
+= \langle{m}\rangle$, etc.
 
@@
 them to match the definitions of $\langle\vec{m}^0\rangle$ and
-$\langle\vec{m}\rangle$. For the first, noting $\vec{p}(-1)=0$ since
-negative numbers of mRNA are nonsensical, we have
+$\langle\vec{m}\rangle$. For the first, noting $\vec{p}(-1)=0$ since negative
+numbers of mRNA are nonsensical, we have
 \begin{equation}
@@
 \end{equation}
-which holds since in extending the lower limit from $m=1$ to $m=0$,
-the extra term we added to the sum is zero. Substituting these back in
+which holds since in extending the lower limit from $m=1$ to $m=0$, the extra
+term we added to the sum is zero. Substituting these back in
 Eq.~\ref{eq:generic_cme_deriv_020}, we have
@@
 \end{equation}
-So given matrices $\mathbf{K}$, $\mathbf{R_D}$, and $\mathbf{R_A}$ describing
-a promoter, finding $\langle\vec{m}^0\rangle$ simply amounts to solving this
-set of linear equations, subject to the normalization constraint
-$\vec{1}^\dagger \cdot \langle\vec{m}^0\rangle = 1$.
-It will turn out to be the case that, if the matrix
-$\mathbf{K} - \mathbf{R_D} + \mathbf{R_A}$ is $n$ dimensional, it will always
-have only $n-1$ linearly independent equations. Including the normalization
-condition provides the $n$-th linearly independent equation, ensuring a
-unique solution. So when using this equation to solve for
-$\langle\vec{m}^0\rangle$, we may always drop one row of the matrix equation
-at our convenience and supplement the system with the normalization
-condition instead.
+So given matrices $\mathbf{K}$, $\mathbf{R_D}$, and $\mathbf{R_A}$ describing a
+promoter, finding $\langle\vec{m}^0\rangle$ simply amounts to solving this set
+of linear equations, subject to the normalization constraint $\vec{1}^\dagger
+\cdot \langle\vec{m}^0\rangle = 1$. It will turn out to be the case that, if the
+matrix $\mathbf{K} - \mathbf{R_D} + \mathbf{R_A}$ is $n$ dimensional, it will
+always have only $n-1$ linearly independent equations. Including the
+normalization condition provides the $n$-th linearly independent equation,
+ensuring a unique solution. So when using this equation to solve for
+$\langle\vec{m}^0\rangle$, we may always drop one row of the matrix equation at
+our convenience and supplement the system with the normalization condition
+instead.
 
 \subsubsection{First moments $\langle\vec{m}\rangle$ and $\langle{m}\rangle$}
-By analogy to the above procedure for finding $\langle\vec{m}^0\rangle$,
-we may find $\langle\vec{m}\rangle$ by first multiplying
-Eq.~\ref{eq:generic_cme_ss} by $m$ and then summing over $m$.
-Carrying out this procedure we have
+By analogy to the above procedure for finding $\langle\vec{m}^0\rangle$, we may
+find $\langle\vec{m}\rangle$ by first multiplying Eq.~\ref{eq:generic_cme_ss} by
+$m$ and then summing over $m$. Carrying out this procedure we have
 \begin{equation}
@@
 The summations in the last two terms can be reindexed just as we did for
-$\langle\vec{m}^0\rangle$, freely adding or removing terms from
-the sum which are zero. For the first,
+$\langle\vec{m}^0\rangle$, freely adding or removing terms from the sum which
+are zero. For the first,
 \begin{equation}
@@
 \end{equation}
-So like $\langle\vec{m}^0\rangle$, $\langle\vec{m}\rangle$ is also found
-by simply solving a set of linear equations after first solving for
+So like $\langle\vec{m}^0\rangle$, $\langle\vec{m}\rangle$ is also found by
+simply solving a set of linear equations after first solving for
 $\langle\vec{m}^0\rangle$ from Eq.~\ref{eq:generic_cme_vecm0}.
@@
 occupying the $i-$th promoter state, so the dot product with $\vec{1}^\dagger$
-amounts to simply summing the elements of $\langle\vec{m}\rangle$.
-Rather than solving Eq.~\ref{eq:generic_cme_deriv_040} for $\langle\vec{m}\rangle$
-and then computing $\vec{1}^\dagger\cdot\langle\vec{m}\rangle$, we may
-take a shortcut by muliplying Eq.~\ref{eq:generic_cme_deriv_040} by
-$\vec{1}^\dagger$ first. The key observation that makes this useful is that
+amounts to simply summing the elements of $\langle\vec{m}\rangle$. Rather than
+solving Eq.~\ref{eq:generic_cme_deriv_040} for $\langle\vec{m}\rangle$ and then
+computing $\vec{1}^\dagger\cdot\langle\vec{m}\rangle$, we may take a shortcut by
+multiplying Eq.~\ref{eq:generic_cme_deriv_040} by $\vec{1}^\dagger$ first. The
+key observation that makes this useful is that
 \begin{equation}
@@
 We note that the in equilibrium models of transcription such as in
-Figure~\ref{fig1:means_cartoons}, it is usually \textit{assumed} that the
-mean mRNA level is given by the ratio of initiation rate $r$ to degradation
-rate $\gamma$ multiplied by the probability of finding the system in the
-RNAP-bound state. Reassuringly, we have recovered exactly this result
-from the master equation picture: the product
+Figure~\ref{fig1:means_cartoons}, it is usually \textit{assumed} that the mean
+mRNA level is given by the ratio of initiation rate $r$ to degradation rate
+$\gamma$ multiplied by the probability of finding the system in the RNAP-bound
+state. Reassuringly, we have recovered exactly this result from the master
+equation picture: the product
 $\vec{1}^\dagger\cdot\mathbf{R_A}\langle\vec{m}^0\rangle$ picks out the
-probability of the active promoter state from $\langle\vec{m}^0\rangle$
-and multiplies it by the initiation rate $r$.
+probability of the active promoter state from $\langle\vec{m}^0\rangle$ and
+multiplies it by the initiation rate $r$.
 
@@
 \end{equation}
-Reindexing the sums of the last two terms proceeds just as it did for the
-zeroth and first moments. Explicitly, we have
+Reindexing the sums of the last two terms proceeds just as it did for the zeroth
+and first moments. Explicitly, we have
 \begin{equation}
@@
 \end{equation}
-As we found when computing $\langle{m}\rangle$ from $\langle\vec{m}\rangle$,
-we can spare ourselves some algebra by multiplying
+As we found when computing $\langle{m}\rangle$ from $\langle\vec{m}\rangle$, we
+can spare ourselves some algebra by multiplying
 Eq.~\ref{eq:generic_cme_deriv_060} by $\vect{1}^\dagger$, which then reduces to
@@
 
-Finally we have all the preliminary results needed to write a general expression for the Fano factor $\nu$. The Fano factor is defined as the ratio of variance to mean, which can be written as
+Finally we have all the preliminary results needed to write a general expression
+for the Fano factor $\nu$. The Fano factor is defined as the ratio of variance
+to mean, which can be written as
 \begin{equation}
@@
 \end{equation}
-Note a subtle notational trap here: $\langle{m}\rangle
-= \frac{1}{\gamma} \vec{1}^\dagger\cdot\mathbf{R_A}\langle\vec{m}^0\rangle$
-rather than the by-eye similar but wrong expression $\langle{m}\rangle
-\ne \frac{1}{\gamma} \vec{1}^\dagger\cdot\mathbf{R_A}\langle\vec{m}\rangle$,
-so the last term in Eq.~\ref{eq:generic_fano_appdx} is in general quite nontrivial.
-For a generic promoter, Eq.~\ref{eq:generic_fano_appdx} may be greater than, less than, or equal to one, as asserted in Section~\ref{sec:beyond_means}.
-We have not found the general form Eq.~\ref{eq:generic_fano_appdx} terribly intuitive and instead defer discussion to specific examples.
+Note a subtle notational trap here: $\langle{m}\rangle = \frac{1}{\gamma}
+\vec{1}^\dagger\cdot\mathbf{R_A}\langle\vec{m}^0\rangle$ rather than the by-eye
+similar but wrong expression $\langle{m}\rangle \ne \frac{1}{\gamma}
+\vec{1}^\dagger\cdot\mathbf{R_A}\langle\vec{m}\rangle$, so the last term in
+Eq.~\ref{eq:generic_fano_appdx} is in general quite nontrivial. For a generic
+promoter, Eq.~\ref{eq:generic_fano_appdx} may be greater than, less than, or
+equal to one, as asserted in Section~\ref{sec:beyond_means}. We have not found
+the general form Eq.~\ref{eq:generic_fano_appdx} terribly intuitive and instead
+defer discussion to specific examples.
 
 \subsubsection{Summary of general results}
-For ease of reference, we collect and reprint here the key results derived
-in this section that are used in the main text and subsequent subsections.
-Mean mRNA copy number and Fano factor are given by
-Eqs.~\ref{eq:generic_mean_m_appdx} and \ref{eq:generic_fano_appdx}, which are
+For ease of reference, we collect and reprint here the key results derived in
+this section that are used in the main text and subsequent subsections. Mean
+mRNA copy number and Fano factor are given by Eqs.~\ref{eq:generic_mean_m_appdx}
+and \ref{eq:generic_fano_appdx}, which are
 \begin{equation}
@@
 \end{equation}
-respectively.
-To compute these two quantities, we need the expressions for
+respectively. To compute these two quantities, we need the expressions for
 $\langle\vec{m}^0\rangle$ and $\langle\vec{m}\rangle$ given by solving
@@
 \end{equation}
-Some comments are in order before we consider particular models. First,
-note that to obtain $\langle\vec{m}\rangle$ and $\nu$, we need not bother
-solving for all components of the vectors $\langle\vec{m}^0\rangle$ and
-$\langle\vec{m}\rangle$, but only the components which are multiplied by
-nonzero elements of $\mathbf{R_A}$. The only component of
-$\langle\vec{m}^0\rangle$ that ever survives is the transciptionally active
-state, and for the models we consider here, there is only ever one such state.
-This will save us some amount of algebra below.
+Some comments are in order before we consider particular models. First, note
+that to obtain $\langle\vec{m}\rangle$ and $\nu$, we need not bother solving for
+all components of the vectors $\langle\vec{m}^0\rangle$ and
+$\langle\vec{m}\rangle$, but only the components which are multiplied by nonzero
+elements of $\mathbf{R_A}$. The only component of $\langle\vec{m}^0\rangle$ that
+ever survives is the transciptionally active state, and for the models we
+consider here, there is only ever one such state. This will save us some amount
+of algebra below.
 
 Also note that we are computing Fano factors to verify the results of
-Section~\ref{sec:beyond_means}, concerning the constitutive promoter models
-in Figure~\ref{fig2:constit_cartoons} which are analogs of the simple
-repression models in Figure~\ref{fig1:means_cartoons}. We can translate
-the matrices from the simple repression models to the constitutive case
-by simply all occurences of repressor rates and removing the row and column
-corresponding to the repressor bound state. The results for
-$\langle{m}\rangle$ computed in the repressed case can be easily translated
-to the constitutive case, rather than recalculating from scratch, by taking
-the limit $k_R^+\rightarrow 0$, since this amounts to sending repressor
-copy number to zero.
+Section~\ref{sec:beyond_means}, concerning the constitutive promoter models in
+Figure~\ref{fig2:constit_cartoons} which are analogs of the simple repression
+models in Figure~\ref{fig1:means_cartoons}. We can translate the matrices from
+the simple repression models to the constitutive case by simply substituting all
+occurrences of repressor rates by zero and removing the row and column
+corresponding to the repressor bound state. The results for $\langle{m}\rangle$
+computed in the repressed case can be easily translated to the constitutive
+case, rather than recalculating from scratch, by taking the limit
+$k_R^+\rightarrow 0$, since this amounts to sending repressor copy number to
+zero.
 
 Finally, we point out that it would be possible to compute
-$\langle\vec{m}^0\rangle$ more simply using the diagram methods from King
-and Altman~\cite{King1956} (also independently discovered by
-Hill~\cite{Hill1966}). But to our knowledge this method cannot be
-applied to compute $\langle\vec{m}\rangle$ or $\nu$, so since we would need
-to resort to solving the matrix equations anyways for $\langle\vec{m}\rangle$,
-we choose not to introduce the extra conceptual burden of the diagram
-methods simply for computing $\langle\vec{m}^0\rangle$.
+$\langle\vec{m}^0\rangle$ more simply using the diagram methods from King and
+Altman~\cite{King1956} (also independently discovered by Hill~\cite{Hill1966}).
+But to our knowledge this method cannot be applied to compute
+$\langle\vec{m}\rangle$ or $\nu$, so since we would need to resort to solving
+the matrix equations anyways for $\langle\vec{m}\rangle$, we choose not to
+introduce the extra conceptual burden of the diagram methods simply for
+computing $\langle\vec{m}^0\rangle$.
 
@@
 \subsubsection{Mean mRNA}
-For nonequilibrium model one in Figure~\ref{fig1:means_cartoons},
-we have already shown the full master equation in
-Eq.~\ref{eq:poisson_promoter_cme} and Eq.~\ref{eq:poisson_promoter_cme_appdx},
-but for completeness we reprint it again as
+For nonequilibrium model one in Figure~\ref{fig1:means_cartoons}, we have
+already shown the full master equation in Eq.~\ref{eq:poisson_promoter_cme} and
+Eq.~\ref{eq:poisson_promoter_cme_appdx}, but for completeness we reprint it
+again as
 \begin{align}
@@
 This is a direct transcription of the states and rates in
-Figure~\ref{fig1:means_cartoons}. This may be converted to the matrix form
-of the master equation shown in Eq.~\ref{eq:generic_cme_appdx} with matrices
+Figure~\ref{fig1:means_cartoons}. This may be converted to the matrix form of
+the master equation shown in Eq.~\ref{eq:generic_cme_appdx} with matrices
 \begin{equation}
@@
 \end{equation}
-where $\mathbf{R_A}$ and $\mathbf{R_D}$ are equal, so we drop the subscript
-and denote both simply by $\mathbf{R}$. Since our interest is only in
-steady-state we dropped the time dependence as well.
+where $\mathbf{R_A}$ and $\mathbf{R_D}$ are equal, so we drop the subscript and
+denote both simply by $\mathbf{R}$. Since our interest is only in steady-state
+we dropped the time dependence as well.
 
 First we need $\langle\vec{m}^0\rangle$. Label its components as $p_R$ and
-$p_U$, the probabilities of finding the system in either promoter state,
-and note that only $p_U$ survives multiplication by $\mathbf{R}$, since
+$p_U$, the probabilities of finding the system in either promoter state, and
+note that only $p_U$ survives multiplication by $\mathbf{R}$, since
 \begin{equation}
@@
 As mentioned earlier in Section~\ref{sec:m0_appdx}, the two rows are linearly
-dependent, so taking only the first row and using normalization to set
-$p_R = 1-p_U$ gives
+dependent, so taking only the first row and using normalization to set $p_R =
+1-p_U$ gives
 \begin{equation}
@@
 To verify that the Fano factor for model one in
-Figure~\ref{fig2:constit_cartoons} is in fact 1 as claimed in the main text,
+Figure~\ref{fig2:constit_cartoons}(A) is in fact 1 as claimed in the main text,
 note that in this limit $p_U = 1$ and $\langle{m}\rangle = r/\gamma$. All
-elements of $\mathbf{K}$ are zero, and $\mathbf{R_A}-\mathbf{R_D} = 0$,
-so Eq.~\ref{eq:vecm_appdx_ref} reduces to
+elements of $\mathbf{K}$ are zero, and $\mathbf{R_A}-\mathbf{R_D} = 0$, so
+Eq.~\ref{eq:vecm_appdx_ref} reduces to
 \begin{equation}
@@
 \end{align}
-which can be condensed to the matrix form of Eq.~\ref{eq:generic_cme_appdx}
-with matrices given by
+which can be condensed to the matrix form of Eq.~\ref{eq:generic_cme_appdx} with
+matrices given by
 \begin{equation}
@@
 As for model one, we must first find $\mathbf{R_A} \langle\vect{m}^0\rangle$.
-Denote its components as $p_R$, $p_E$, $p_P$, the probabilities of being
-found in repressor bound, empty, or RNAP-bound states, respectively.
-Only $p_P$ is necessary to find since
+Denote its components as $p_R$, $p_E$, $p_P$, the probabilities of being found
+in repressor bound, empty, or RNAP-bound states, respectively. Only $p_P$ is
+necessary to find since
 \begin{equation}
@@
 \end{equation}
-Substituting this in Eq.~\ref{eq:model2_deriv_appdx_020} and multiplying by $\mathbf{R_A}$ produces
+Substituting this in Eq.~\ref{eq:model2_deriv_appdx_020} and multiplying by
+$\mathbf{R_A}$ produces
 \begin{equation}
@@
 \subsubsection{Fano factor}
-To compute the Fano factor, we first remove the repressor bound state from
-the matrices describing the model, which reduce to
+To compute the Fano factor, we first remove the repressor bound state from the
+matrices describing the model, which reduce to
 \begin{equation}
@@
 \end{equation}
-Similarly we remove the repressor bound state from
-$\mathbf{R_A} \langle\vec{m}^0\rangle$ and take the $k_R^+\rightarrow 0$
-limit, which simplifies to 
+Similarly we remove the repressor bound state from $\mathbf{R_A}
+\langle\vec{m}^0\rangle$ and take the $k_R^+\rightarrow 0$ limit, which
+simplifies to 
 \begin{equation}
@@
 where we labeled the components of $\langle\vec{m}\rangle$ as $m_E$ and $m_P$,
-since they are the mean mRNA counts conditional upon the system residing in
-the empty or polymerase bound states, respectively. Unlike for
-$\langle\vec{m}^0\rangle$, the rows of this matrix are linearly independent
-so we simply solve this matrix equation as is. We can immediately eliminate
-$m_E$ since $m_E = m_P (k_P^- + r + \gamma)/k_P^+$ from the second row, and
+since they are the mean mRNA counts conditional upon the system residing in the
+empty or polymerase bound states, respectively. Unlike for
+$\langle\vec{m}^0\rangle$, the rows of this matrix are linearly independent so
+we simply solve this matrix equation as is. We can immediately eliminate $m_E$
+since $m_E = m_P (k_P^- + r + \gamma)/k_P^+$ from the second row, and
 substituting into the first row gives an equation for $m_P$ alone, which is
@@
 
-\subsection{Nonequilibrium Model Three
-    - Multistep Transcription Initiation and Escape}
+\subsection{Nonequilibrium Model Three - Multistep Transcription Initiation and
+Escape}
+
 \subsubsection{Mean mRNA}
 In close analogy to model two above, nonequilibrium model three from
-Figure~\ref{fig1:means_cartoons} can be described by our generic master equation
-Eq.~\ref{eq:generic_cme_appdx} with promoter transition matrix given by
+Figure~\ref{fig1:means_cartoons}(C) can be described by our generic master
+equation Eq.~\ref{eq:generic_cme_appdx} with promoter transition matrix given by
 \begin{equation}
@@
 \end{equation}
-where the four components of $\langle\vec{m}^0\rangle$ correspond to the
-four promoter states repressor bound, empty, RNAP-bound closed complex, and
-RNAP-bound open complex. As explained in Section~\ref{sec:m0_appdx}, we
-are free to discard one linearly dependent row from this matrix and replace
-it with the normalization condition $p_R + p_E + p_C + p_O = 1$.
-Using normalization to eliminate $p_R$ from the first row gives
+where the four components of $\langle\vec{m}^0\rangle$ correspond to the four
+promoter states repressor bound, empty, RNAP-bound closed complex, and
+RNAP-bound open complex. As explained in Section~\ref{sec:m0_appdx}, we are free
+to discard one linearly dependent row from this matrix and replace it with the
+normalization condition $p_R + p_E + p_C + p_O = 1$. Using normalization to
+eliminate $p_R$ from the first row gives
 \begin{equation}
@@
 \end{equation}
-If we substitute this in the third row, then the last two rows constitute
-two equations in $p_C$ and $p_O$ given by
+If we substitute this in the third row, then the last two rows constitute two
+equations in $p_C$ and $p_O$ given by
 \begin{align}
@@
 \end{equation}
-Once again $p_O$, the transcriptionally active state, is the only component
-of $\langle\vec{m}^0\rangle$ that survives multiplication by $\mathbf{R_A}$,
-and $\mathbf{R_A}\langle\vec{m}^0\rangle = r p_O$. So
+Once again $p_O$, the transcriptionally active state, is the only component of
+$\langle\vec{m}^0\rangle$ that survives multiplication by $\mathbf{R_A}$, and
+$\mathbf{R_A}\langle\vec{m}^0\rangle = r p_O$. So
 \begin{equation}
@@
 excise the repressor states and rates from the problem. More precisely, we
-construct the matrix
-$(\mathbf{K} - \mathbf{R_D} + \mathbf{R_A} - \gamma\mathbf{I})$ and
-substitute it into Eq.~\ref{eq:vecm_appdx_ref} which is now
+construct the matrix $(\mathbf{K} - \mathbf{R_D} + \mathbf{R_A} -
+\gamma\mathbf{I})$ and substitute it into Eq.~\ref{eq:vecm_appdx_ref} which is
+now
 \begin{equation}
@@
 \end{equation}
-where we labeled the unbound, closed complex, and open complex componenets of
-$\langle\vec{m}\rangle$ as $m_E$, $m_C$, and $m_O$, respectively. $p_O$
-is given by the limit of Eq.~\ref{eq:model3_pO} as $k_R^+\rightarrow 0$,
-which is
+where we labeled the unbound, closed complex, and open complex components of
+$\langle\vec{m}\rangle$ as $m_E$, $m_C$, and $m_O$, respectively. $p_O$ is given
+by the limit of Eq.~\ref{eq:model3_pO} as $k_R^+\rightarrow 0$, which is
 \begin{equation}
@@
 \end{equation}
-where we define $\mathcal{Z}$ for upcoming convenience as this sum of terms
-will appear repeatedly. We can use the second equation to eliminate $m_E$,
-finding $m_E = m_C(k_P^- + k_O + \gamma)/k_P^+$, and the third to eliminate
-$m_C$, which is simply $m_C = m_O(r+\gamma)/k_O$. Substituting these both into
-the first equation gives a single equation for the variable of interest, $m_O$,
+where we define $\mathcal{Z}$ for upcoming convenience as this sum of terms will
+appear repeatedly. We can use the second equation to eliminate $m_E$, finding
+$m_E = m_C(k_P^- + k_O + \gamma)/k_P^+$, and the third to eliminate $m_C$, which
+is simply $m_C = m_O(r+\gamma)/k_O$. Substituting these both into the first
+equation gives a single equation for the variable of interest, $m_O$,
 \begin{equation}
@@
 Now $\vec{1}^\dagger\cdot\mathbf{R_A}\langle\vec{m}\rangle = r m_O$, and
-$\langle{m}\rangle = rp_O/\gamma$, so if we substitute these two quantities
-into Eq.~\ref{eq:fano_appdx_ref}, we will readily obtain the Fano factor as
+$\langle{m}\rangle = rp_O/\gamma$, so if we substitute these two quantities into
+Eq.~\ref{eq:fano_appdx_ref}, we will readily obtain the Fano factor as
 \begin{equation}
@@
 
-\subsection{Nonequilibrium Model Four
-    - ``Active'' and ``Inactive'' States}
+\subsection{Nonequilibrium Model Four - ``Active'' and ``Inactive'' States}
 \subsubsection{Mean mRNA}
 The mathematical specification of this model is almost identical to model two.
-The matrix $\mathbf{K}$ is identical, as is $\mathbf{R_D}$. The only
-difference is that now $\mathbf{R_A}=\mathbf{R_D}$, i.e., both are diagonal,
-in contrast to model two where $\mathbf{R_A}$ has an off-diagonal element, as in
-Eq.~\ref{eq:model2_matrices_appdx}.
-Then the analog of Eq.~\ref{eq:model2_K-R+R_for_m0} for finding
-$\langle{m}^0\rangle$ is
+The matrix $\mathbf{K}$ is identical, as is $\mathbf{R_D}$. The only difference
+is that now $\mathbf{R_A}=\mathbf{R_D}$, i.e., both are diagonal, in contrast to
+model two where $\mathbf{R_A}$ has an off-diagonal element, as in
+Eq.~\ref{eq:model2_matrices_appdx}. Then the analog of
+Eq.~\ref{eq:model2_K-R+R_for_m0} for finding $\langle{m}^0\rangle$ is
 \begin{equation}
@@
 \end{equation}
-In fact we need not do this calculation explicitly and can instead recycle
-the calculation of mean mRNA $\langle{m}\rangle$ from model two. The
-matrices are identical except for the relabeling
-$k_P^- \longleftrightarrow (k_P^- + r)$, and a careful look through the
-derivation of $\langle{m}\rangle$ for model two shows that the parameters
-$k_P^-$ and $r$ only ever appear as the sum $k_P^- + r$. So taking
-$\langle{m}\rangle$ from model two, Eq.~\ref{eq:model2_meanm_appdx}, and
-relabeling $(k_P^- + r) \rightarrow k_P^-$ gives us our answer for model
-four, simply
+In fact we need not do this calculation explicitly and can instead recycle the
+calculation of mean mRNA $\langle{m}\rangle$ from model two. The matrices are
+identical except for the relabeling $k_P^- \longleftrightarrow (k_P^- + r)$, and
+a careful look through the derivation of $\langle{m}\rangle$ for model two shows
+that the parameters $k_P^-$ and $r$ only ever appear as the sum $k_P^- + r$. So
+taking $\langle{m}\rangle$ from model two, Eq.~\ref{eq:model2_meanm_appdx}, and
+relabeling $(k_P^- + r) \rightarrow k_P^-$ gives us our answer for model four,
+simply
 \begin{equation}
@@
 Likewise, for computing the Fano factor of this model we may take a shortcut.
-Consider the constitutive model four from Figure~\ref{fig2:constit_cartoons}
-for which we want to compute the Fano factor and compare it to nonequilibrium
-model one of simple repression in Figure~\ref{fig1:means_cartoons}.
-Mathematically these are exactly the same model, just with rates labeled
-differently and the meaning of the promoter states interpreted differently.
-Furthermore, nonequilibrium model one from Figure~\ref{fig1:means_cartoons}
-was the model considered by Jones et.\ al.~\cite{Jones2014}, where they
-derived the Fano factor for that model to be
+Consider the constitutive model four from Figure~\ref{fig2:constit_cartoons} for
+which we want to compute the Fano factor and compare it to nonequilibrium model
+one of simple repression in Figure~\ref{fig1:means_cartoons}. Mathematically
+these are exactly the same model, just with rates labeled differently and the
+meaning of the promoter states interpreted differently. Furthermore,
+nonequilibrium model one from Figure~\ref{fig1:means_cartoons} was the model
+considered by Jones et.\ al.~\cite{Jones2014}, where they derived the Fano
+factor for that model to be
 \begin{equation}

### 2020-06-04 doc/appendix_01_nonbursty.tex
MRM writes short intro to chemical master eq. with new figure

@@
 \label{sec:non_bursty}
-
-\mrm{I personally think that the paper so far has done a great job at being very 
-pedagogical. That is why I am of the opinion that before jumping into the 
-derivation of moments and so on there needs to be a $p(m, t + \Delta t)$ kind of 
-derivation.}
 
@@
 cartoons is as a chemical master equation such as Eq.~\ref{eq:2state_rep_cme}
-for model one, which for completeness we reproduce here as
+for model one. Before jumping into the derivations of the general computation of
+the mean mRNA level and the Fano factor we will work through the derivation of an example master equation. In particular we will focus on model 1 from 
+Figure~\ref{fig1:means_cartoons}(C). The general steps are applicable to all
+other chemical master equations in this work.
+
+\subsection{Derivation of chemical master equation}
+
+\begin{figure}[h!]
+\centering
+\includegraphics{../figures/si/figS0X_two_state.pdf}
+\caption{
+\textbf{Two-state promoter chemical master equation.}
+(A) Schematic of the two state promoter simple repression model. Rates $k_R^+$
+and $k_R^-$ are the association and dissociation rates of the transcriptional
+repressor, respectively, $r$ is the transcription initiation rate, and $\gamma$
+is the mRNA degradation rate. (B) Schematic depiction of the mRNA count state
+transitions. The model in (A) only allows for jumps in mRNA of size 1. The
+production of mRNA can only happen when the promoter is in the transcriptionally
+active state.}
+\label{fig:two_state}
+\end{figure}
+
+The chemical master equation describes the continuous time evolution of a
+continuous or discrete probability distribution function. In our specific case
+we want to describe the time evolution of the discrete mRNA distribution. What
+this means is that we want to compute the probability of having $m$ mRNA
+molecules at time $t + \Delta t$, where $\Delta t$ is a sufficiently small time
+interval such that only one of the possible reactions take place. For the
+example that we will workout here in detail we chose the two-state stochastic
+simple repression model schematized in Figure~\ref{fig:two_state}(A). To derive
+the master equation we will focus more on the representation shown in
+Figure~\ref{fig:two_state}(B), where the transitions between different mRNA
+counts and promoter states is more explicitly depicted. Given that the DNA
+promoter can exist in one of two states -- transcriptionally active state, and
+with repressor bound -- we will keep track not only of the mRNA count, but on
+which state the promoter is. For this we will keep track of two probability
+distributions: The probability of having $m$ mRNAs at time $t$ when the promoter
+is in the transcriptionally active state $A$, $p_A(m, t)$, and the equivalent
+probability but when the promoter is in the repressor bound state $R$, $p_R(m,
+t)$.
+
+Since mRNA production can only take place on the transcriptionally active state
+we will focus on this function for our derivation. The repressor bound state
+will have an equivalent equation without terms involving the production of
+mRNAs. We begin by listing the possible state transition that can happen for a
+particular mRNA count with the promoter in the active state. For state changes
+in a small time window $\Delta t$ that ``jump into'' state $m$ in the
+transcriptionally active state we have
+\begin{itemize}
+    \item Produce an mRNA, jumping from $m-1$ to $m$.
+    \item Degrade an mRNA, jumping from $m+1$ to $m$.
+    \item Transition from the repressor bound state $R$ to the active state $A$.
+\end{itemize}
+Likewise, for state transitions that ``jump out'' of state $m$ in the
+transcriptionally inactive state we have
+\begin{itemize}
+    \item Produce an mRNA, jumping from $m$ to $m+1$.
+    \item Degrade an mRNA, jumping from $m$ to $m-1$.
+    \item Transition from the active state $A$ to the repressor bound state $R$.
+\end{itemize}
+The mRNA production does not depend on the current number of mRNAs, therefore
+these state transitions happen with probability $r\Delta t$. The same is true
+for the promoter state transitions, each happen with probability $k_R^\pm \Delta
+t$. As for the mRNA degradation events does must depend on the current number of
+mRNA molecules. Each molecule has a constant probability $\gamma \Delta t$ of 
+being degraded, so the total probability for an mRNA degradation event to happen
+is computed by multiplying this probability by the current number of mRNAs.
+
+To see these terms in action let us compute the probability of having $m$ mRNA
+at time $t + \Delta t$ in the transcriptionally active state. This takes the 
+form
+\begin{equation}
+\begin{aligned}
+p_A(m, t + \Delta t) &= p_A(m, t) \\
+&+ \overbrace{(r\Delta t) p_A(m-1, t)}^{m-1 \rightarrow m}
+- \overbrace{(r\Delta t) p_A(m, t)}^{m \rightarrow m+1}\\
+% \;\;\;\text{(mRNA production)} \\
+&+ \overbrace{(m+1)(\gamma \Delta t) p_A(m+1, t)}^{m+1 \rightarrow m}
+- \overbrace{m(\gamma \Delta t) p_A(m, t)}^{m \rightarrow m-1}\\
+% \;\;\;\text{(mRNA degradation)} \\
+&+\overbrace{(k_R^- \Delta t) p_R(m, t)}^{R \rightarrow A}
+-\overbrace{(k_R^+ \Delta t) p_A(m, t)}^{A \rightarrow R},
+% \;\;\;\text{(promoter state transition)}
+\end{aligned}
+\end{equation}
+where the overbrace indicates the corresponding state transitions. Notice that
+the second to last term on the right-hand side is multiplied by $p_R(m, t)$ 
+since the transition from state $R$ to state $A$ depends on the probability of
+being in state $R$ to begin with. It is through this term that both probability
+distribution functions are coupled. An equivalent equation can be written for
+the probability of having $m$ mRNA at time $t + \Delta t$ while in the repressor
+bound state, the only difference being that the mRNA production rates are 
+removed, and the sign for the promoter state transitions are inverted. This is
+\begin{equation}
+\begin{aligned}
+p_R(m, t + \Delta t) &= p_R(m, t) \\
+&+ \overbrace{(m+1)(\gamma \Delta t) p_R(m+1, t)}^{m+1 \rightarrow m}
+- \overbrace{m(\gamma \Delta t) p_R(m, t)}^{m \rightarrow m-1}\\
+% \;\;\;\text{(mRNA degradation)} \\
+&-\overbrace{(k_R^- \Delta t) p_R(m, t)}^{R \rightarrow A}
++\overbrace{(k_R^+ \Delta t) p_A(m, t)}^{A \rightarrow R}.
+\end{aligned}
+\end{equation}
+
+All we have to do now are simple algebraic steps in order to simplify the 
+equations. Let's focus on the transcriptionally active state $A$. First we will
+send the term $p_A(m, t)$ to the right-hand side, and then we will divide both
+sides of the equation by $\Delta t$. This results in
+\begin{equation}
+\begin{aligned}
+\frac{p_A(m, t + \Delta t) - p_A(m, t)}{\Delta t} &=
+r p_A(m-1, t) - r p_A(m, t)\\
+&+ (m+1)\gamma p_A(m+1, t)
+- m \gamma p_A(m, t)\\
+&+k_R^- p_R(m, t)
+-k_R^+ p_A(m, t).
+\end{aligned}
+\end{equation}
+Upon taking the limit when $\Delta t \rightarrow 0$ we can transform the 
+left-hand side into a derivative, obtaining the chemical master equation
+\begin{equation}
+\begin{aligned}
+\frac{d p_A(m, t)}{dt} &=
+r p_A(m-1, t) - r p_A(m, t)\\
+&+ (m+1)\gamma p_A(m+1, t)
+- m \gamma p_A(m, t)\\
+&+k_R^- p_R(m, t)
+-k_R^+ p_A(m, t).
+\end{aligned}
+\end{equation}
+Doing equivalent manipulations for the repressor bound state gives an ODE of the
+form
+\begin{equation}
+\begin{aligned}
+\frac{d p_R(m, t)}{dt} &=
+(m+1)\gamma p_R(m+1, t)
+- m \gamma p_R(m, t)\\
+&-k_R^- p_R(m, t)
++k_R^+ p_A(m, t).
+\end{aligned}
+\end{equation}
+In the next section we will write these couple ODEs in a more compat form using
+matrix notation.
+
+\subsection{Matrix form of the multi-state chemical master equation}
+
+Having derived an example chemical master equation we now focus on writing a
+general matrix form for the kinetic models 1-4 shown in
+Figure~\ref{fig1:means_cartoons}(C) in the main text. In each of these four
+models, the natural mathematization of their cartoons is as a chemical master
+equation such as Eq.~\ref{eq:2state_rep_cme} for model one, which for
+completeness we reproduce here as
 \begin{align}

### 2020-06-04 doc/appendix_01_nonbursty.tex
RP finished appendix

@@
 not derived in Sections~\ref{section_02_means} and \ref{sec:beyond_means} of the
-main text. In each of these four models, the natural mathematization of their
+main text. In each of these four models, the natural mathematicization of their
 cartoons is as a chemical master equation such as Eq.~\ref{eq:2state_rep_cme}
@@
 transitions. The model in (A) only allows for jumps in mRNA of size 1. The
-production of mRNA can only happen when the promoter is in the transcriptionally
+production of mRNA can only occur when the promoter is in the transcriptionally
 active state.}
@@
 molecules at time $t + \Delta t$, where $\Delta t$ is a sufficiently small time
-interval such that only one of the possible reactions take place. For the
-example that we will workout here in detail we chose the two-state stochastic
+interval such that only one of the possible reactions take place during that time interval. For the
+example that we will work out here in detail we chose the two-state stochastic
 simple repression model schematized in Figure~\ref{fig:two_state}(A). To derive
@@
 
-Since mRNA production can only take place on the transcriptionally active state
+Since mRNA production can only take place in the transcriptionally active state
 we will focus on this function for our derivation. The repressor bound state
@@
     \item Degrade an mRNA, jumping from $m+1$ to $m$.
-    \item Transition from the repressor bound state $R$ to the active state $A$.
+    \item Transition from the repressor bound state $R$ with m mRNAs  to the active state $A$
+    with m mRNAs.
 \end{itemize}
@@
     \item Degrade an mRNA, jumping from $m$ to $m-1$.
-    \item Transition from the active state $A$ to the repressor bound state $R$.
+    \item Transition from the active state $A$ with m mRNAs to the repressor bound state $R$ with
+    m mRNAs.
 \end{itemize}
 The mRNA production does not depend on the current number of mRNAs, therefore
-these state transitions happen with probability $r\Delta t$. The same is true
-for the promoter state transitions, each happen with probability $k_R^\pm \Delta
-t$. As for the mRNA degradation events does must depend on the current number of
-mRNA molecules. Each molecule has a constant probability $\gamma \Delta t$ of 
+these state transitions occur with probability $r\Delta t$. The same is true
+for the promoter state transitions, each occurs with probability $k_R^\pm \Delta
+t$. As for the mRNA degradation events, these transitions depend on the current number of
+mRNA molecules since the more molecules of mRNA there are, the more will decay during
+a given time interval. Each molecule has a constant probability $\gamma \Delta t$ of 
 being degraded, so the total probability for an mRNA degradation event to happen
@@
 since the transition from state $R$ to state $A$ depends on the probability of
-being in state $R$ to begin with. It is through this term that both probability
-distribution functions are coupled. An equivalent equation can be written for
+being in state $R$ to begin with. It is through this term that the dynamics of the two probability
+distribution functions ($p_R(m,t)$ and $p_A(m,t)$ are coupled. An equivalent equation can be written for
 the probability of having $m$ mRNA at time $t + \Delta t$ while in the repressor
@@
 \end{equation}
-In the next section we will write these couple ODEs in a more compat form using
+In the next section we will write these coupled ODEs in a more compact form using
 matrix notation.
@@
 Figure~\ref{fig1:means_cartoons}(C) in the main text. In each of these four
-models, the natural mathematization of their cartoons is as a chemical master
+models, the natural mathematicization of their cartoons is as a chemical master
 equation such as Eq.~\ref{eq:2state_rep_cme} for model one, which for
@@
 $p_P$, and $p_E$ for the repressor bound, empty, and polymerase bound states,
-respectively. By analogy to~\ref{eq:poisson_promoter_cme_appdx}, the master
+respectively. By analogy to eqn.~\ref{eq:poisson_promoter_cme_appdx}, the master
 equation for model two can be written
@@
 steady-state mRNA distribution from Eq.~\ref{eq:generic_cme_appdx}. Our
-treatment of this is analogous to that given in~\cite{Sanchez2011}
+treatment of this is analogous to that given in Refs.~\cite{Sanchez2011}
 and~\cite{Razo-Mejia2020}. We first obtain the steady-state limit of
@@
 where $\vec{1}^\dagger$ is a row vector whose elements are all 1.
-Eq.~\ref{eq:m_from_vecm} holds since the $i$-th element of the column vector
+Eq.~\ref{eq:m_from_vecm} holds since the $i^{th}$ element of the column vector
 $\langle\vec{m}\rangle$ is the mean mRNA value conditional on the system
-occupying the $i-$th promoter state, so the dot product with $\vec{1}^\dagger$
+occupying the $i^{th}$ promoter state, so the dot product with $\vec{1}^\dagger$
 amounts to simply summing the elements of $\langle\vec{m}\rangle$. Rather than

### 2020-06-04 doc/appendix_01_nonbursty.tex
MRM pass through appendix 1 to fix spacing. Section is basically ready

@@
 for model one. Before jumping into the derivations of the general computation of
-the mean mRNA level and the Fano factor we will work through the derivation of an example master equation. In particular we will focus on model 1 from 
+the mean mRNA level and the Fano factor we will work through the derivation of
+an example master equation. In particular we will focus on model 1 from
 Figure~\ref{fig1:means_cartoons}(C). The general steps are applicable to all
@@
 molecules at time $t + \Delta t$, where $\Delta t$ is a sufficiently small time
-interval such that only one of the possible reactions take place during that time interval. For the
-example that we will work out here in detail we chose the two-state stochastic
-simple repression model schematized in Figure~\ref{fig:two_state}(A). To derive
-the master equation we will focus more on the representation shown in
-Figure~\ref{fig:two_state}(B), where the transitions between different mRNA
-counts and promoter states is more explicitly depicted. Given that the DNA
-promoter can exist in one of two states -- transcriptionally active state, and
-with repressor bound -- we will keep track not only of the mRNA count, but on
-which state the promoter is. For this we will keep track of two probability
-distributions: The probability of having $m$ mRNAs at time $t$ when the promoter
-is in the transcriptionally active state $A$, $p_A(m, t)$, and the equivalent
-probability but when the promoter is in the repressor bound state $R$, $p_R(m,
-t)$.
+interval such that only one of the possible reactions take place during that
+time interval. For the example that we will work out here in detail we chose the
+two-state stochastic simple repression model schematized in
+Figure~\ref{fig:two_state}(A). To derive the master equation we will focus more
+on the representation shown in Figure~\ref{fig:two_state}(B), where the
+transitions between different mRNA counts and promoter states is more explicitly
+depicted. Given that the DNA promoter can exist in one of two states --
+transcriptionally active state, and with repressor bound -- we will keep track
+not only of the mRNA count, but on which state the promoter is. For this we will
+keep track of two probability distributions: The probability of having $m$ mRNAs
+at time $t$ when the promoter is in the transcriptionally active state $A$,
+$p_A(m, t)$, and the equivalent probability but when the promoter is in the
+repressor bound state $R$, $p_R(m, t)$.
 
@@
     \item Degrade an mRNA, jumping from $m+1$ to $m$.
-    \item Transition from the repressor bound state $R$ with m mRNAs  to the active state $A$
-    with m mRNAs.
+    \item Transition from the repressor bound state $R$ with $m$ mRNAs  to the
+    active state $A$ with $m$ mRNAs.
 \end{itemize}
@@
     \item Degrade an mRNA, jumping from $m$ to $m-1$.
-    \item Transition from the active state $A$ with m mRNAs to the repressor bound state $R$ with
-    m mRNAs.
+    \item Transition from the active state $A$ with $m$ mRNAs to the repressor
+    bound state $R$ with $m$ mRNAs.
 \end{itemize}
 The mRNA production does not depend on the current number of mRNAs, therefore
-these state transitions occur with probability $r\Delta t$. The same is true
-for the promoter state transitions, each occurs with probability $k_R^\pm \Delta
-t$. As for the mRNA degradation events, these transitions depend on the current number of
-mRNA molecules since the more molecules of mRNA there are, the more will decay during
-a given time interval. Each molecule has a constant probability $\gamma \Delta t$ of 
-being degraded, so the total probability for an mRNA degradation event to happen
-is computed by multiplying this probability by the current number of mRNAs.
+these state transitions occur with probability $r\Delta t$. The same is true for
+the promoter state transitions, each occurs with probability $k_R^\pm \Delta t$.
+As for the mRNA degradation events, these transitions depend on the current
+number of mRNA molecules since the more molecules of mRNA there are, the more
+will decay during a given time interval. Each molecule has a constant
+probability $\gamma \Delta t$ of being degraded, so the total probability for an
+mRNA degradation event to happen is computed by multiplying this probability by
+the current number of mRNAs.
 
@@
 since the transition from state $R$ to state $A$ depends on the probability of
-being in state $R$ to begin with. It is through this term that the dynamics of the two probability
-distribution functions ($p_R(m,t)$ and $p_A(m,t)$ are coupled. An equivalent equation can be written for
-the probability of having $m$ mRNA at time $t + \Delta t$ while in the repressor
-bound state, the only difference being that the mRNA production rates are 
-removed, and the sign for the promoter state transitions are inverted. This is
+being in state $R$ to begin with. It is through this term that the dynamics of
+the two probability distribution functions ($p_R(m,t)$ and $p_A(m,t)$) are
+coupled. An equivalent equation can be written for the probability of having $m$
+mRNA at time $t + \Delta t$ while in the repressor bound state, the only
+difference being that the mRNA production rates are removed, and the sign for
+the promoter state transitions are inverted. This is
 \begin{equation}

### 2020-06-04 code/figures/si/figS0X_ppc.py
script for PPC of unreg promoters in SI

@@ -1,0 +1,133 @@
+#%%
+import pickle
+from git import Repo #for directory convenience
+
+import numpy as np
+import scipy.stats as st
+import pandas as pd
+import arviz as az
+
+import matplotlib.pyplot as plt
+import matplotlib
+import matplotlib.patheffects as path_effects
+import seaborn as sns
+import bebi103.viz
+
+import srep.viz
+from srep.utils import load_FISH_by_promoter
+
+#%%
+repo = Repo("./", search_parent_directories=True)
+# repo_rootdir holds the absolute path to the top-level of our repo
+repo_rootdir = repo.working_tree_dir
+
+colors = srep.viz.color_selector('constit')
+srep.viz.plotting_style()
+#%%
+
+# Import necessary data
+
+repo = Repo("./", search_parent_directories=True)
+# repo_rootdir holds the absolute path to the top-level of our repo
+repo_rootdir = repo.working_tree_dir
+
+colors = srep.viz.color_selector('constit')
+srep.viz.plotting_style()
+
+pklfile = open(f"{repo_rootdir}/data/mcmc_samples/constit_post_inf.pkl", 'rb')
+all_samples = pickle.load(pklfile)
+pklfile.close()
+
+df_unreg, df_energies, = load_FISH_by_promoter(("unreg", "energies"))
+df_energies.sort_values('Energy (kT)', inplace=True)
+
+# %%
+
+# Initialize figure
+fig = plt.figure(constrained_layout=False, figsize=(6, 12))
+
+# Set gridspec
+gs = fig.add_gridspec(
+    nrows=18, 
+    ncols=3, 
+    # left=0.05, 
+    # right=0.45,
+    # top=1, 
+    # bottom=0.55,
+    wspace=0.15,
+    hspace=0.15,
+)
+
+# Initialize list to save ax objects
+ax = list()
+# Loop through rows and columns
+for i in range(6):
+    for j in range(3):
+        # Generate axis for ECDF
+        ax_ecdf = fig.add_subplot(gs[(i * 3):(i * 3 + 2), j])
+        ax_diff = fig.add_subplot(gs[(i * 3 + 2), j])
+        # Join axis
+        ax_diff.get_shared_x_axes().join(ax_ecdf, ax_diff)
+        # Turn off axis labels
+        if j != 0:
+            ax_ecdf.get_yaxis().set_ticklabels([])
[…]

### 2020-06-04 srep/viz.py

@@ -343,20 +343,6 @@ def ppc_ecdf_pair(posterior_samples, ppc_var, df, percentiles=(95, 75, 50, 25), 
     
     return ax
 
-# %%
-fig, ax = plt.subplots(1, 2, figsize= (6, 2.5))
-srep.viz.ppc_ecdf_pair(
-    all_samples['5DL10'],
-    'mRNA_counts_ppc',
-    df_unreg[df_unreg['experiment'] == '5DL10'],
-    ax=ax,
-    data_color='black',
-    color='green',
-    data_label='5DL10'
-    )
-plt.legend()
-# %%
-
 def ecdf(data):
     """
     data[0] should be a 1D array of observed values, and

### 2020-06-04 code/figures/si/figS0X_ppc.py
update figure dimensions

@@ -44,12 +44,12 @@ df_energies.sort_values('Energy (kT)', inplace=True)
 # %%
 
 # Initialize figure
-fig = plt.figure(constrained_layout=False, figsize=(6, 12))
+fig = plt.figure(constrained_layout=False, figsize=(7.5, 8.75))
 
 # Set gridspec
 gs = fig.add_gridspec(
-    nrows=18, 
-    ncols=3, 
+    nrows=15, 
+    ncols=4, 
     # left=0.05, 
     # right=0.45,
     # top=1, 
@@ -60,36 +60,41 @@ gs = fig.add_gridspec(
 
 # Initialize list to save ax objects
 ax = list()
+# Initialize plot counter
+counter = 0
 # Loop through rows and columns
-for i in range(6):
-    for j in range(3):
+for i in range(5):
+    for j in range(4):
         # Generate axis for ECDF
-        ax_ecdf = fig.add_subplot(gs[(i * 3):(i * 3 + 2), j])
-        ax_diff = fig.add_subplot(gs[(i * 3 + 2), j])
-        # Join axis
-        ax_diff.get_shared_x_axes().join(ax_ecdf, ax_diff)
-        # Turn off axis labels
-        if j != 0:
-            ax_ecdf.get_yaxis().set_ticklabels([])
-            ax_diff.get_yaxis().set_ticklabels([])
+        if counter < len(df_energies):
+            #update counter
+            counter += 1
+            ax_ecdf = fig.add_subplot(gs[(i * 3):(i * 3 + 2), j])
+            ax_diff = fig.add_subplot(gs[(i * 3 + 2), j])
+            # Join axis
+            ax_diff.get_shared_x_axes().join(ax_ecdf, ax_diff)
+            # Turn off axis labels
+            if j != 0:
+                ax_ecdf.get_yaxis().set_ticklabels([])
+                ax_diff.get_yaxis().set_ticklabels([])
 
-        ax_ecdf.get_xaxis().set_ticklabels([])
-        if i != 5:
-            ax_diff.get_xaxis().set_ticklabels([])
-        # Set axis label
-        if j == 0:
-            ax_ecdf.set_ylabel("ECDF")
-            ax_diff.set_ylabel("ECDF diff.")
-        if i == 5:
-            ax_diff.set_xlabel("mRNA/cell")
-        # Join axis to first plot
-        if (i != 0) | (j != 0):
-            ax_ecdf.get_shared_x_axes().join(ax_ecdf, ax[0][0])
[…]

### 2020-06-04 code/figures/si/figS0X_ppc.py
sort promoters by mean mRNA count

@@ -99,8 +99,12 @@ for i in range(5):
 # Align y axis label
 fig.align_ylabels(ax)
 
+# Compute the mean of each promoter mRNA count
+df_mean = df_unreg.groupby("experiment")["mRNA_cell"].apply(np.mean)
+# Sort promoters by mean
+df_mean = df_mean.sort_values(ascending=False)
 # Find unique promoters
-promoters = df_energies["Name"].unique()
+promoters = df_mean.index.values
 
 # Loop through unique promoters and plot ECDF
 for i, p in enumerate(promoters):

### 2020-06-04 code/figures/si/figS0X_ppc.py
update figure to label by number and energy

@@ -50,10 +50,6 @@ fig = plt.figure(constrained_layout=False, figsize=(7.5, 8.75))
 gs = fig.add_gridspec(
     nrows=15, 
     ncols=4, 
-    # left=0.05, 
-    # right=0.45,
-    # top=1, 
-    # bottom=0.55,
     wspace=0.15,
     hspace=0.15,
 )
@@ -67,8 +63,7 @@ for i in range(5):
     for j in range(4):
         # Generate axis for ECDF
         if counter < len(df_energies):
-            #update counter
-            counter += 1
+            
             ax_ecdf = fig.add_subplot(gs[(i * 3):(i * 3 + 2), j])
             ax_diff = fig.add_subplot(gs[(i * 3 + 2), j])
             # Join axis
@@ -79,13 +74,13 @@ for i in range(5):
                 ax_diff.get_yaxis().set_ticklabels([])
 
             ax_ecdf.get_xaxis().set_ticklabels([])
-            if i != 5:
+            if counter not in range(14, 18):
                 ax_diff.get_xaxis().set_ticklabels([])
             # Set axis label
             if j == 0:
                 ax_ecdf.set_ylabel("ECDF")
                 ax_diff.set_ylabel("ECDF diff.")
-            if i == 5:
+            if counter in range(14, 18):
                 ax_diff.set_xlabel("mRNA/cell")
             # Join axis to first plot
             if (i != 0) | (j != 0):
@@ -95,16 +90,17 @@ for i in range(5):
                 ax_diff.get_shared_y_axes().join(ax_diff, ax[0][1])
             # Add to list
             ax.append([ax_ecdf, ax_diff])
+            #update counter
+            counter += 1
 
 # Align y axis label
 fig.align_ylabels(ax)
 
-# Compute the mean of each promoter mRNA count
-df_mean = df_unreg.groupby("experiment")["mRNA_cell"].apply(np.mean)
-# Sort promoters by mean
-df_mean = df_mean.sort_values(ascending=False)
+# Set xlimit
+ax_ecdf.set_xlim([-2, 50])
+
 # Find unique promoters
-promoters = df_mean.index.values
+promoters = df_energies["Name"].values
 
 # Loop through unique promoters and plot ECDF
 for i, p in enumerate(promoters):
@@ -120,14 +116,37 @@ for i, p in enumerate(promoters):
         data_size=1,
     )
 
+    # Add cross-reference number
+    text = ax[i][0].text(
[…]

### 2020-06-04 doc/appendix_03_bayesian_inference.tex
add PPC figure to SI for unregulated promoters

@@
 Section~\ref{section_04_bayesian_inference} in the main text details the
-results of applying this inference to the single-cell mRNA counts data.
+results of applying this inference to the single-cell mRNA counts data. There
+we show the posterior distribution for the two parameters for different 
+promoters. Figure~\ref{figS:ppc_unreg} shows the so-called posterior predictive
+checks (see main text for explanation) for all 18 unregulated promoters shown
+in the main text.
+
+\begin{figure}[p]
+\centering
+\includegraphics{../figures/si/figS0X_ppc.pdf}
+\caption{\textbf{Theory-data comparison of inference on unregulated promoters.}
+Comparison of the inference (red shaded area) vs the experimental measurements
+(black lines) for 18 different unregulated promoters with different mean mRNA
+expression levels from Ref.~\cite{Jones2014}. Upper panels show the empirical
+cumulative distribution function (ECDF), while the lower panels show the
+differences with respect to the median of the posterior samples. White numbers
+are the same as in Figure~\ref{fig1:means_cartoons} for cross comparison. The
+predicted binding energies $\beta\Delta\varepsilon_p$ were obtained from the
+energy matrix model in Ref.~\cite{Brewster2012}}
+\label{figS:ppc_unreg}
+\end{figure}
 

### 2020-06-05 doc/abstract.tex
MRM pass through abstract. Looks good
Not quite final pass on everything, Appdx S3 still needs more

--- 2020-05-21 doc/abstract.tex
@@
 of this so-called simple-repression motif, based both on equilibrium and kinetic
-thinking, can be used to derive the predicted levels of gene expression. These
-different models are invoked to confront a variety of different data on mean,
-variance and full gene expression distribution, illustrating the extent to which
-such models can be distinguished.
+thinking, can be used to derive the predicted levels of gene expression and
+shed light on the often surprising past success of the equilbrium models. These
+different models are then invoked to confront a variety of different data on mean,
+variance and full gene expression distributiona, illustrating the extent to which
+such models can and cannot be distinguished, and suggesting a two-state model with a
+distribution of burst sizes as the most potent of the seven for describing the
+simple-repression motif.
 \end{abstract}

### 2020-06-05 doc/appendix_01_nonbursty.tex
Not quite final pass on everything, Appdx S3 still needs more

@@
 \subsection{Derivation of chemical master equation}
+\label{sec:cme_from_cartoon}
 
@@
 will have an equivalent equation without terms involving the production of
-mRNAs. We begin by listing the possible state transition that can happen for a
+mRNAs. We begin by listing the possible state transitions that can occur for a
 particular mRNA count with the promoter in the active state. For state changes
@@
 these state transitions occur with probability $r\Delta t$. The same is true for
-the promoter state transitions, each occurs with probability $k_R^\pm \Delta t$.
+the promoter state transitions; each occurs with probability $k_R^\pm \Delta t$.
 As for the mRNA degradation events, these transitions depend on the current
@@
 probability $\gamma \Delta t$ of being degraded, so the total probability for an
-mRNA degradation event to happen is computed by multiplying this probability by
+mRNA degradation event to occur is computed by multiplying this probability by
 the current number of mRNAs.
@@
 All we have to do now are simple algebraic steps in order to simplify the 
-equations. Let's focus on the transcriptionally active state $A$. First we will
+equations. Let us focus on the transcriptionally active state $A$. First we will
 send the term $p_A(m, t)$ to the right-hand side, and then we will divide both
@@
 whose elements are simply the probabilities of finding the system in each of the
-possible promoter states. To reduce notational clutter in what follows, sums
-over $m$ will be assumed to run from 0 to $\infty$ unless otherwise specified,
-i.e., $\sum_{m=0}^\infty$ will be denoted simply as $\sum_m$. It will also be
+possible promoter states. It will be
 convenient to denote by $\vec{1}^\dagger$ a row vector of the same length as
@@
 \begin{equation}
-0 = \sum_m \left[
+0 = \sum_{m=0}^\infty \left[
     \left( \mathbf{K} - \mathbf{R_D} - \gamma m \mathbf{I} \right) \vec{p}(m)
@@
 0 = (\mathbf{K} - \mathbf{R_D}) \langle\vec{m}^0\rangle
-    - \gamma \langle\vec{m}\rangle + \mathbf{R_A} \sum_m \vec{p}(m-1)
-    + \gamma \sum_m (m+1)\vec{p}(m+1).
+    - \gamma \langle\vec{m}\rangle + \mathbf{R_A} \sum_{m=0}^\infty \vec{p}(m-1)
+    + \gamma \sum_{m=0}^\infty (m+1)\vec{p}(m+1).
 \label{eq:generic_cme_deriv_020}
@@
 instead.
+The reader may find it illuminating to skip ahead and see
+Eq.~\ref{eq:generic_cme_vecm0} in use with concrete examples, e.g.,
+Eq.~\ref{eq:model1_m0_giver_appdx} and what follows it, before
+continuing on through the general formulas for moments.
 
@@
 \begin{equation}
-0 = \sum_m m \left[
+0 = \sum_{m=0}^\infty m \left[
 \left( \mathbf{K} - \mathbf{R_D} - \gamma m \mathbf{I} \right) \vec{p}(m)
@@
 0 = (\mathbf{K} - \mathbf{R_D}) \langle\vec{m}\rangle
-    - \gamma \langle\vec{m}^2\rangle + \mathbf{R_A} \sum_m m\vec{p}(m-1)
-    + \gamma \sum_m m(m+1)\vec{p}(m+1).
+    - \gamma \langle\vec{m}^2\rangle + \mathbf{R_A} \sum_{m=0}^\infty m\vec{p}(m-1)
+    + \gamma \sum_{m=0}^\infty m(m+1)\vec{p}(m+1).
 \label{eq:generic_cme_deriv_030}
@@
 $\langle\vec{m}^0\rangle$, freely adding or removing terms from the sum which
-are zero. For the first,
+are zero. For the first term we find
 \begin{equation}
@@
 \end{equation}
-and for the second,
+and similarly for the second,
 \begin{equation}
@@
 \begin{equation}
-0 = \sum_m m^2 \left[
+0 = \sum_{m=0}^\infty m^2 \left[
 \left( \mathbf{K} - \mathbf{R_D} - \gamma m \mathbf{I} \right) \vec{p}(m)
@@
 0 = (\mathbf{K} - \mathbf{R_D}) \langle\vec{m}^2\rangle
-    - \gamma \langle\vec{m}^3\rangle + \mathbf{R_A} \sum_m m^2\vec{p}(m-1)
-    + \gamma \sum_m m^2(m+1)\vec{p}(m+1).
+    - \gamma \langle\vec{m}^3\rangle + \mathbf{R_A} \sum_{m=0}^\infty m^2\vec{p}(m-1)
+    + \gamma \sum_{m=0}^\infty m^2(m+1)\vec{p}(m+1).
 \label{eq:generic_cme_deriv_050}
@@
     \begin{pmatrix} p_R \\ p_U \end{pmatrix} = 0.
+\label{eq:model1_m0_giver_appdx}
 \end{equation}
@@
 
+\subsubsection{Generalizing $\nu<1$ to more fine-grained models}
+In the main text we argued that the convolution of multiple
+exponential distributions should be a distribution with a smaller
+fractional width than the corresponding exponential distribution
+with the same mean.
+This can be made more precise with a result from~\cite{Stewart2007},
+who showed that the convolution of multiple gamma distributions (of which the
+exponential distribution is a special case) is, to a very good approximation,
+also gamma distributed. Using their Eq.~2 for the distribution of the
+convolution, with shape parameters set to 1 to give exponential distributions,
+the total waiting time distribution has a ratio of variance to squared mean
+$\sigma^2/\mu^2 = \sum_i k_i^2/\left(\sum_i k_i\right)^2$, where the $k_i$ are
+the rates of the individual steps. Clearly this is less than 1 and therefore the
+total waiting time distribution is narrower than the corresponding exponential.
+
+We also claimed in the main text that for a process whose waiting time
+distribution is narrower than exponential, i.e., has $\sigma^2/\mu^2<1$,
+the distribution of counts should be less variable than a Poisson
+distribution, leading to a Fano factor $\nu<1$.
+This we argue by analogy to photon statistics where it is known that
+``antibunched'' arrivals, in other words more uniformly distributed in
+time relative to uncorrelated arrivals, generally gives rise to
+sub-Poissonian noise~\cite{Paul1982, Zou1990}. Although loopholes to this
+result are known to exist, to our knowledge they appear to arise from
+uniquely quantum effects so we do not expect they apply for our problem.
+Nevertheless we refrain from elevating this equivalence of
+kinetic cycles with sub-Poissonian noise to a ``theorem.''
+
 \subsection{Nonequilibrium Model Four - ``Active'' and ``Inactive'' States}
@@
 \begin{pmatrix} -k_R^- & k_R^+ & 0 \\
-        k_R^- & -k_R^+ -k_P^+ & k_P^-\\
-        0 & k_P^+ & -k_P^-
+        k_R^- & -k_R^+ -k^+ & k^-\\
+        0 & k^+ & -k^-
 \end{pmatrix}
 \begin{pmatrix}
-    p_R \\ p_E \\ p_P
+    p_R \\ p_I \\ p_A
 \end{pmatrix}
@@
 calculation of mean mRNA $\langle{m}\rangle$ from model two. The matrices are
-identical except for the relabeling $k_P^- \longleftrightarrow (k_P^- + r)$, and
+identical except for the relabeling $k^- \longleftrightarrow (k_P^- + r)$, and
 a careful look through the derivation of $\langle{m}\rangle$ for model two shows
@@
 taking $\langle{m}\rangle$ from model two, Eq.~\ref{eq:model2_meanm_appdx}, and
-relabeling $(k_P^- + r) \rightarrow k_P^-$ gives us our answer for model four,
+relabeling $(k_P^- + r) \rightarrow k^-$ gives us our answer for model four,
 simply
@@
 \langle{m}\rangle = \frac{r}{\gamma}
-        \frac{k_P^+ k_R^-} {k_P^+ k_R^- + k_P^- (k_R^+ + k_R^-)}.
+        \frac{k^+ k_R^-} {k^+ k_R^- + k^- (k_R^+ + k_R^-)}.
 \end{equation}

### 2020-06-05 doc/appendix_02_bursty.tex
Not quite final pass on everything, Appdx S3 still needs more

@@
 In order to solve Eq.~\ref{eq:si_master_unreg} we will use the generating
-function method \mrm{reference?}. The probability generating function is defined
-as
+function method~\cite{vanKampen2007}. The probability generating function
+is defined as
 \begin{align}
@@
 {d F(z, \tau) \over d\tau} =
-- \sum_m z^m m p(m, \tau) +
+- \sum_m z^m m p(m, \tau)
 + \sum_m z^m (m + 1) p(m + 1, \tau)
@@
 probability function $p(m, \tau)$. Upon finding the function for $F(z, \tau)$,
-we will recover this probability distribution by evaluating $F(z=1, \tau)$ in
-other cases we can evaluate $F(z=0, \tau)$ to recover instead the moments of the
-distribution. The point here is that we know that when all the dust settles we
-will evaluate $z$ to be less than one \mrm{someone needs to fact check me}.
+we will recover this probability distribution by evaluating derivatives
+of $F(z, \tau)$ at $z=0$, whereas we can evaluate derivatives of
+$F(z, \tau)$ at $z=1$ to instead recover the moments of the
+distribution. The point here is that when the dust settles we
+will evaluate $z$ to be less than or equal to one.
 Furthermore, we know that the parameter of the geometric distribution $\theta$
-must be between zero and one. With these two facts we can safely state that $| z
-(1 - \theta) | < 1$. Defining $n \equiv m - m'$ we rewrite the last sum in
+must be strictly between zero and one. With these two facts we can safely state that
+$| z (1 - \theta) | < 1$. Defining $n \equiv m - m'$ we rewrite the last sum in
 Eq.~\ref{eq:si_generating_02} as
@@
 where we use the geometric series since, as stated before, $| z (1 - \theta) | <
-1$. Putting these results together, the PDE for the generation function is
+1$. Putting these results together, the PDE for the generating function is
 \begin{equation}
@@
 key fact follows from the definition
-$F(z,t) = \sum_m z^m p(m,t)$,
-normalization requires that
+$F(z,t) = \sum_m z^m p(m,t)$.
+Clearly normalization of the distribution requires that
 $F(z=1, t) = \sum_m p(m,t) = 1$.
-\footnote{
-Sometimes the generating function may be undefined \textit{at} $z=1$ but the
-limit still holds. Also many authors change variables from $z$ to other
-things, so one must keep track of how the normalization condition transforms.
-}
-Doing the integrals (and producing constant $c$) gives
+A subtlety is that sometimes the generating function may be undefined
+\textit{at} $z=1$, in which case the limit as $z$ approaches $1$ from
+below suffices to define the normalization condition.
+We also warn the reader that, while it is frequently convenient to change
+variables from $z$ to a different independent variable, one must
+carefully track how the normalization condition transforms.
+
+Continuing on, we evaluate the integrals (producing a constant $c$) which gives
 \begin{align}
@@
 mean $b=(1-\theta)/\theta$.
-\footnote{
-$\theta$ can be viewed as a Bernoulli trial probability: after each transcript
-in a burst is produced, $\theta$ is the probability that the burst terminates
-and $(1-\theta)$ is the probability that the burst continues.
-}
 
@@
 linearly independent solutions. More precisely, 
-\mrm{the use of variable $\gamma$ here is confusing. Is it okay to change it to
-$\delta$ or something else in order to avoid associating it with the mRNA
-degradation rate?}
-\begin{align}
-f_R(w) = C_1 {_2F_1}(\alpha, \beta, \gamma; w)
-+ C_2 w^{1-\gamma}{_2F_1}(1+\alpha-\gamma, 1+\beta-\gamma, 2-\gamma; w)
+\begin{align}
+f_R(w) = C^{(1)} {_2F_1}(\alpha, \beta, \delta; w)
++ C^{(2)} w^{1-\delta}{_2F_1}(1+\alpha-\delta, 1+\beta-\delta, 2-\delta; w)
 \end{align}
@@
 \\
-\gamma &= 1 + k_R^- + k_R^+
+\delta &= 1 + k_R^- + k_R^+
 \end{split}
 \end{align}
-and constants $C_1$ and $C_2$ to be set by boundary conditions. Solving for
+and constants $C^{(1)}$ and $C^{(2)}$ to be set by boundary conditions. Solving for
 $\alpha$ and $\beta$, we find
-\footnote{
-Note that $\alpha$ and $\beta$ are interchangeable in the definition of
-${_2F_1}$ and differ only in the sign preceeding the radical.
-}
 \begin{align}
@@
 \\
-\gamma &= 1 + k_R^- + k_R^+.
+\delta &= 1 + k_R^- + k_R^+.
 \end{split}
 \end{align}
+Note that $\alpha$ and $\beta$ are interchangeable in the definition of
+${_2F_1}$ and differ only in the sign preceeding the radical.
 Since the normalization condition requires that $f_R$ be finite at $w=0$,
-we can immediately set $C_2=0$ to discard the second solution.
+we can immediately set $C^{(2)}=0$ to discard the second solution.
 This is because all the rate constants are strictly positive,
-so $\gamma>1$ and therefore $w^{1-\gamma}$ blows up as $w\rightarrow0$.
+so $\delta>1$ and therefore $w^{1-\delta}$ blows up as $w\rightarrow0$.
 Now that we have $f_R$, we would like to find the generating function
@@
 $f_R$ and $f_A$ to our new, more convenient, variable $w$.
-Plugging our solution for $f_R(w) = C_1{_2F_1}(\alpha, \beta, \gamma; w)$
+Plugging our solution for $f_R(w) = C^{(1)}{_2F_1}(\alpha, \beta, \delta; w)$
 into $f_A$, we will require the differentiation rule for ${_2F_1}$,
@@
 \begin{align}
-\pderiv[f_R]{w} = C_1\frac{\alpha\beta}{\gamma}
-                {_2F_1}(\alpha+1, \beta+1, \gamma+1; w),
+\pderiv[f_R]{w} = C^{(1)}\frac{\alpha\beta}{\delta}
+                {_2F_1}(\alpha+1, \beta+1, \delta+1; w),
 \end{align}
@@
 \begin{align}
-f_A(w) = \frac{C_1}{k_R^+}
+f_A(w) = \frac{C^{(1)}}{k_R^+}
 \left(
-k_R^- {_2F_1}(\alpha, \beta, \gamma; w)
-+ w\frac{\alpha\beta}{\gamma} {_2F_1}(\alpha+1, \beta+1, \gamma+1; w)
+k_R^- {_2F_1}(\alpha, \beta, \delta; w)
++ w\frac{\alpha\beta}{\delta} {_2F_1}(\alpha+1, \beta+1, \delta+1; w)
 \right)
@@
 \begin{align}
-f(w) = C_1\left(1 + \frac{k_R^-}{k_R^+}\right)
-        {_2F_1}(\alpha, \beta, \gamma; w)
-+ w \frac{C_1}{k_R^+} \frac{\alpha\beta}{\gamma}
-        {_2F_1}(\alpha+1, \beta+1, \gamma+1; w).
+f(w) = C^{(1)}\left(1 + \frac{k_R^-}{k_R^+}\right)
+        {_2F_1}(\alpha, \beta, \delta; w)
++ w \frac{C^{(1)}}{k_R^+} \frac{\alpha\beta}{\delta}
+        {_2F_1}(\alpha+1, \beta+1, \delta+1; w).
 \end{align}
@@
 \begin{align}
-w\frac{\alpha\beta}{\gamma} {_2F_1}(\alpha+1, \beta+1, \gamma+1; w)
+w\frac{\alpha\beta}{\delta} {_2F_1}(\alpha+1, \beta+1, \delta+1; w)
 =
-(\gamma-1)\left(
-{_2F_1}(\alpha, \beta, \gamma-1; w) - {_2F_1}(\alpha, \beta, \gamma; w)
+(\delta-1)\left(
+{_2F_1}(\alpha, \beta, \delta-1; w) - {_2F_1}(\alpha, \beta, \delta; w)
 \right).
@@
 \begin{align}
-f(w) = \frac{C_1}{k_R^+}
+f(w) = \frac{C^{(1)}}{k_R^+}
 \left[
         \left(k_R^+ + k_R^-\right)
-        {_2F_1}(\alpha, \beta, \gamma; w)
-+ (\gamma-1)\left(
-        {_2F_1}(\alpha, \beta, \gamma-1; w) - {_2F_1}(\alpha, \beta, \gamma; w)
+        {_2F_1}(\alpha, \beta, \delta; w)
++ (\delta-1)\left(
+        {_2F_1}(\alpha, \beta, \delta-1; w) - {_2F_1}(\alpha, \beta, \delta; w)
         \right)
@@
 \end{align}
-and since $\gamma-1 = k_R^+ + k_R^-$, the first and third terms cancel,
+and since $\delta-1 = k_R^+ + k_R^-$, the first and third terms cancel,
 leaving only
 \begin{align}
-f(w) = C_1\frac{k_R^+ + k_R^-}{k_R^+} {_2F_1}(\alpha, \beta, \gamma-1; w).
+f(w) = C^{(1)}\frac{k_R^+ + k_R^-}{k_R^+} {_2F_1}(\alpha, \beta, \delta-1; w).
 \end{align}
 Now we enforce normalization, demanding $f(w=0) = f(z=1) = 1$.
-${_2F_1}(\alpha, \beta, \gamma-1; 0) = 1$, so we must have
-$C_1 = k_R^+ / (k_R^+ + k_R^-)$ and consequently
+${_2F_1}(\alpha, \beta, \delta-1; 0) = 1$, so we must have
+$C^{(1)} = k_R^+ / (k_R^+ + k_R^-)$ and consequently
 \begin{align}
@@
 the form $\alpha(\alpha+1)(\alpha+2)\cdots(\alpha+m-1)$ in the expression for
-$p(m)$ and similarly for $\beta$ and $\gamma$. These resemble ratios of
-factorials, but since $\alpha$, $\beta$, and $\gamma$ are not necessarily
+$p(m)$ and similarly for $\beta$ and $\delta$. These resemble ratios of
+factorials, but since $\alpha$, $\beta$, and $\delta$ are not necessarily
 integer, we should express the ratios using gamma functions instead. More
@@
 which is finally the probability distribution we sought to derive.
-
-\subsubsection{Moments of two-state bursty promoter mRNA distribution}
-
-We can verify the correctness of~\eq{eq:p_m_bursty+rep_appdx} by comparison with
-Gillespie simulation. We can also easily compute moments of the distribution
-from the generating function, and check these against raw moments computed
-directly from the master equation.
-
-The $j$-th raw moment of the distribution can be computed from the generating
-function from the formula
-\begin{align}
-\langle m^j \rangle
-= \left. \left(z \frac{\partial}{\partial z}\right)^j f(z)\right|_{z=1}.
-\end{align}
-Note that \textit{probabilities} are generated by differentiating and evaluating
-at $z=0$, while \textit{moments} are generated by differentiating and evaluating
-at $z=1$. This also aligns with the normalization constraint, which can be
-thought of as the zeroth raw moment.
-
-The mean mRNA is given by
-\begin{align}
-\langle m \rangle = \left.z \pderiv[f]{z} \right|_{z=1}
-= \frac{\alpha\beta b}{k_R^+ + k_R^-}
-= k_i b\frac{k_R^-}{k_R^+ + k_R^-},
-\end{align}
-which is intuitive and appealingly simple: the mean mRNA is just the mean mRNA
-in the absence of repression ($k_i b$, the burst rate time mean burst size)
-times the fraction of time the repressor is bound, $k_R^-/(k_R^+ + k_R^-)$.
-
-To compute the variance, let us first compute
-\begin{align}
-\begin{split}
-\langle m^2\rangle
-&= \left .z\pderiv{z}\left(z\pderiv[f]{z}\right) \right|_{z=1}
-= \left. z\left(\pderiv[f]{z} + z\psecderiv[f]{z}\right)\right|_{z=1}
-\\
-&= \frac{\alpha\beta b}{k_R^+ + k_R^-}
-        + \frac{\alpha(\alpha+1)\beta(\beta+1) b^2}
-                {(k_R^+ + k_R^-)(k_R^+ + k_R^- + 1)}
-\\
-&= \frac{k_i b k_R^-}{k_R^+ + k_R^-}
-        + \frac{k_i k_R^- b^2 (1 + k_i + k_R^+ + k_R^- + k_i k_R^-)}
-                {(k_R^+ + k_R^-)(k_R^+ + k_R^- + 1)},
-\end{split}
-\end{align}
-from which the variance is
-\begin{align}
-\begin{split}
-var(m) = \langle m^2\rangle - \langle m\rangle^2
-&= \frac{k_i b k_R^-}{k_R^+ + k_R^-}
-\left(
-1 + \frac{b(1 + k_i + k_R^+ + k_R^- + k_i k_R^-)} {k_R^+ + k_R^- + 1}
-- \frac{k_i b k_R^-}{k_R^+ + k_R^-}
-\right)
-\\
-&= k_i b\frac{k_R^-}{k_R^+ + k_R^-}
-\left(
-1 + b + k_i b\frac{1 + k_R^-} {k_R^+ + k_R^- + 1}
-- \frac{k_i b k_R^-}{k_R^+ + k_R^-}
-\right)
-\\
-&= k_i b\frac{k_R^-}{k_R^+ + k_R^-}
-\left(
-1 + b + k_i b\frac{k_R^+}{(k_R^+ + k_R^-)(k_R^+ + k_R^- + 1)}
-\right).
-\end{split}
-\end{align}
-The Fano factor is the quantity in parentheses. This can be interpreted as
-the variability arising from the negative binomial distribution of the
-unregulated promoter $(1+b)$ plus additional noise arising from repression.
-
-\mmnote{Still to writeup: moment derivation direct from CME to compare}
-\mrm{stopped reading here}
 
@@
 from $\mathcal{O}(1)$ to $\mathcal{O}(10^2)$,
-even if $\alpha$, $\beta$, and $\gamma$
+even if $\alpha$, $\beta$, and $\delta$
 in~\eq{eq:p_m_bursty+rep_appdx} are small,
@@
 ${_2F_1}$ for the other 50-100 values of $m$ using the recurrence
-relation.\footnote{
-There are even more sophisticated ways of utilizing the recurrence
+relation.
+In fact there are even more sophisticated ways of utilizing the recurrence
 relations that might have netted another factor of 2 speed-up, and
@@
 not seem worth the time to pursue.
-}
-
-However, there two further wrinkles.
+
+However, there are two further wrinkles.
 The first is that
@@
 model,~\eq{eq:p_m_bursty+rep_appdx}, we need to evaluate hypergeometric functions
-of the form ${_2F_1}(\alpha+m, \beta+m, \gamma+m; -b)$
+of the form ${_2F_1}(\alpha+m, \beta+m, \delta+m; -b)$
 for values of $m$ ranging from $0$ to $\mathcal{O}(100)$.
@@
 in~\cite{Gil2007} and search for a transformation we could apply to
-${_2F_1}(\alpha+m, \beta+m, \gamma+m; -b)$ that would place the $m$'s
+${_2F_1}(\alpha+m, \beta+m, \delta+m; -b)$ that would place the $m$'s
 (the variable being incremented by the recursion)
@@
 \begin{align}
-{_2F_1}(\alpha+m, \beta+m, \gamma+m; -b)
+{_2F_1}(\alpha+m, \beta+m, \delta+m; -b)
 =
 (1+b)^{-\alpha-m}
-        {_2F_1}\left(\alpha+m, \gamma-\beta, \gamma+m; \frac{b}{1+b}\right),
+        {_2F_1}\left(\alpha+m, \delta-\beta, \delta+m; \frac{b}{1+b}\right),
 \label{eq:rec_euler_pretransform}
@@
 &\propto
-{_2F_1}\left(-\alpha^\prime + \gamma^\prime - m,
-                -\beta^\prime + \gamma^\prime,
-                1-\alpha^\prime-\beta^\prime+\gamma^\prime-m;
+{_2F_1}\left(-\alpha^\prime + \delta^\prime - m,
+                -\beta^\prime + \delta^\prime,
+                1-\alpha^\prime-\beta^\prime+\delta^\prime-m;
                 1-z\right)
@@
                 \beta^\prime,
-                1+\alpha^\prime+\beta^\prime-\gamma^\prime+m;
+                1+\alpha^\prime+\beta^\prime-\delta^\prime+m;
                 1-z\right),
@@
 \begin{align}
-(\gamma^\prime - (\alpha^\prime + m)) f_{m-1}
+(\delta^\prime - (\alpha^\prime + m)) f_{m-1}
 +
-(2(\alpha^\prime+m) - \gamma^\prime + (\beta^\prime - \alpha^\prime)z)f_m
+(2(\alpha^\prime+m) - \delta^\prime + (\beta^\prime - \alpha^\prime)z)f_m
 + \alpha^\prime(z-1) f_{m+1} = 0.
@@
 \\
-\beta^\prime &= \gamma - \beta
-\\
-1 + \alpha^\prime + \beta^\prime - \gamma^\prime &= \gamma
+\beta^\prime &= \delta - \beta
+\\
+1 + \alpha^\prime + \beta^\prime - \delta^\prime &= \delta
 \\
@@
 \\
-\beta^\prime &= \gamma - \beta
-\\
-\gamma^\prime &= 1 + \alpha - \beta
+\beta^\prime &= \delta - \beta
+\\
+\delta^\prime &= 1 + \alpha - \beta
 \\
@@
 =
-\frac{\Gamma(1+\alpha^\prime-\gamma^\prime+m)}
-        {\Gamma(1+\alpha^\prime+\beta^\prime-\gamma^\prime+m)}
+\frac{\Gamma(1+\alpha^\prime-\delta^\prime+m)}
+        {\Gamma(1+\alpha^\prime+\beta^\prime-\delta^\prime+m)}
 \times
@@
 from large $m$ to small $m$.
-So the procedure is to compute the value of ${_2F_1}$ for the two
-largest $m$ values of interest using arbitrary precision arithmetic,
-compute the prefactors to construct
-$y_{4,\text{max}(m)}$ and $y_{4,\text{max}(m)-1}$,
-recursively compute $y_{4,m}$ for all $m$ less than $\text{max}(m)$ down
-to $m=0$, then cancel off the prefactors of the resulting values of
+
+Let us finally outline the complete procedure as an algorithm to be implemented:
+\begin{enumerate}
+\item Compute the value of ${_2F_1}$ for the two
+largest $m$ values of interest using arbitrary precision arithmetic.
+\item Compute the prefactors to construct
+$y_{4,\text{max}(m)}$ and $y_{4,\text{max}(m)-1}$.
+\item Recursively compute $y_{4,m}$ for all $m$ less than $\text{max}(m)$ down
+to $m=0$.
+\item Cancel off the prefactors of the resulting values of
 $y_{4,m}$ for all $m$ to produce ${_2F_1}$ for all desired $m$ values.
-\mmnote{Maybe I should reformat this paragraph into
-a pseudocode version of the algorithm?
-Instead of, or in addition to?}
+\end{enumerate}
 

### 2020-06-05 doc/appendix_03_bayesian_inference.tex
Not quite final pass on everything, Appdx S3 still needs more

@@
 \subsection{The problem of parameter inference}
-
-\mmnote{Bayes theorem. Elaborate on posterior as plausibility over space of
-model parameters.}
 
@@
 numerical values to these parameters we can compute concrete predictions that
-can be contrasted with experimental data. But for these predictions to match the
+can be contrasted with experimental data. For these predictions to match the
 data the parameter values have to carefully be chosen from the whole parameter
-space. But how do we go about asserting the effectiveness of different regions
+space. But how do we go about assessing the effectiveness of different regions
 of parameter space to speak to the ability of our model to reproduce the
@@
 depend on each other Bayes theorem gives us a recipe for how to update our 
-beliefs about one, let's say $B$, given that we get to observe something about
+beliefs about one, let us say $B$, given some state of knowledge, or lack thereof, about
 $A$. In its most classic form Bayes theorem is written as
@@
 $P(\vec{\theta} \mid D)$, where $D$ is the experimental data, quantifies the
-plausibility of a set of parameter values given that we get to observe some
+plausibility of a set of parameter values given our observation of some
 particular dataset. In other words, through the application of Bayes formula we
@@
 that the formulation of our model is not going to capture the phenomena we claim
-to want to understand. This prior information is caputerd in the \textit{prior
+to want to understand. This prior information is captured in the \textit{prior
 probability} $P(\vec{\theta})$. The relationship between how parameter values
@@
 
+We also emphasize that the dichotomy we have presented between prior and
+likelihood is more subtle. Although it is often stated that our prior
+knowledge is entirely encapsulated by the obviously named prior
+probability $P(\vec{\theta})$, this is usually too simplistic.
+The form(s) we choose for our likelihood function
+$P(D \mid \vec{\theta})$ also draw heavily on our prior domain expertise
+and the assumptions, implicit and explicit, that these choices encode are
+at least as important, and often inseparable from,
+the prior probability, as persuasively argued in~\cite{Gelman2017}.
+
 \subsubsection{The likelihood function}
-
-\mmnote{Choosing likelihoods. Easy here, just the steady-state distributions of
-our master equations, assuming iid single-cell measurements.}
 
@@
         \item Deterministic models: Models for which a concrete selection of
-        parameter values give a single output. Say differently, rigid models 
+        parameter values give a single output. Said differently, models 
         with a one-to-one mapping between inputs and outputs.
-        \item Probabilistic models: As the name says it, models that rather than
-        having a one-to-one mapping, describe the full probability distribution
-        of possible outputs.
+        \item Probabilistic models: As the name suggests, models that, rather than
+        having a one-to-one input-output mapping, describe the full
+        probability distribution of possible outputs.
 \end{itemize}
-In this paper we focus on inference done with probabilistic models. After all
+In this paper we focus on inference done with probabilistic models. After all,
 the chemical master equations we wrote down describe the time evolutions of the
@@
 statistical model on top of our model to describe deviations from the
-predictions; our models themselves focus on predicting such variations from cell
-count to cell count.
+predictions. Instead our models themselves focus on predicting such variation
+in cell count.
 
 \subsubsection{Prior selection}
-
-\mmnote{Choosing priors. Not critical here because we are data-rich and
-nonhierarchical, as long as we don't exclude ground truth. In other
-applications, prior can matter a lot. This is a feature, not a bug: it's
-revealing hidden structure that was implied by your modeling assumptions and you
-just didn't realize it. Predictive checks to verify you're not excluding any
-plausible values.}
-
 The different models explored in this work embraced different levels of
@@
 
+\subsubsection{Expectations and marginalizations}
+One of the great powers of Bayesian analysis is \textit{marginalization}.
+For models with more than one or two parameters, it is generally unwise
+to attempt to draw inferences directly from the full joint posterior
+distribution $P(\vec{\theta} \mid D)$...
+
 \mmnote{
@@
 \item MCMC.
-Lot's of fun here, and deep analogies to stat mech and
-Hamiltonian mechanics worth exploring. In a perfect world, you'd
+Lot's of fun and deep analogies to stat mech and
+Hamiltonian mechanics we'll allude to but not explore here.
+The idea of MCMC: in a perfect world, you'd
 wave a magic wand and generate independent draws from target
@@
 So the posterior is fine, but is the model actually not wrong?
-PPC is essential to verify that your model is a plausible
-generative process for the data.
-\item Discuss identifiability, degeneracy, and failed PPCs in main text. Better thru example anyway.
+PPC is essential to verify that your model is
+not blatantly contradicted by the data.
 \end{itemize}
@@
 obtain a single value for each of the association and dissociation rates. What
-this means, as described in~\ref{section_04_bayesian_inference} of the main text
+this means, as described in Section~\ref{section_04_bayesian_inference} of the main text
 is that we have a seven dimensional parameter space with four possible
@@
 
-\mrm{This paragraph needs its own subsection in my opinion if it is to be 
-included in the SI.}
-This likelihood is rather inscrutable. We did not find any of the known
-analytical approximations for ${_2F_1}$ terribly useful in gaining intuition, so
-we instead resorted to numerics. One insight we found
-\mmnote{a plot would probably help explain, maybe SI, need to explain better}
-was that for very strong or very weak repression, the distribution in
-Eq.~\ref{eq:p_m_bursty+rep} is well approximated by a negative binomial with
-burst size $b$ and burst rate $k_i$ equal to their constitutive \textit{lacUV5}
-values, except with $k_i$ multiplied by the fold-change
-$\left(1+k_R^+/k_R^-\right)^{-1}$. In other words, once again only the
-\textit{ratio} $k_R^+/k_R^-$ was detectable. But for intermediate repression,
-the distribution was visibly broadened with Fano factor greater than $1+b$, the
-value for the corresponding constitutive case. This indicates that the repressor
-rates had left an imprint on the distribution, and perhaps intuitively, this
-intermediate regime occurs for values of $k_R^\pm$ comparable to the burst rate
-$k_i$. Put another way, if the repressor rates are much faster or much slower
-than $k_i$, then there is a timescale separation and effectively only one
-timescale remains, $k_i\left(1+k_R^+/k_R^-\right)^{-1}$. Only when all three
-rates in the problem are comparable does the mRNA distribution retain detectable
-information about them.
-
 Next we specify priors. As for the constitutive model, weakly informative
@@
 \mmnote{Outline:
-\begin{itemize}
-\item Hey look it mostly works. With full model, 9D model is well identified,
-though most of the individual experiments unsurprisingly are not:
-only 3 of the 9 datasets were identifiable if fit in isolation
-and for the rest only the ratio of repressor rates is identifiable.
-Only by making assumptions that certain rates
-are equal across datasets was identifiability achieved.
-PPC is not perfect, it kinda misses some
-rep/op pairs but overall it's surprisingly predictive considering how strong
-were the assumptions that went into it.
-\item Do we chalk this up to experimental imperfections or is there interesting
-biology to be found in the disagreements? I'm inclined towards the former. Even
-just catching cells in truly apples-to-apples growth phases, or matching up
-parameters of imaging sessions weeks or months apart, is hardly trivial.
-\item The alternative is to refine our model and/or add on complexity, and it's
-not at all clear what additions to make to the model, especially how to produce
-the long tail of repressed strains over UV5. I see no way to get that without
-getting into complicated time-correlation/history effects, where
-transcription...rebounds?? After being repressed?? Sounds odd.
-\item Comparison between my inferred rates, equilibrium binding E, and
-single-molecule measurements. Not quite as slam-dunk as I'd hoped because of
-$\gamma$ dependence, but rates are w/in factor of 2 or better, seems quite good
-really! With exception of one Oid data data point, free energies are w/in about
-0.5 kT also. That's not massively larger than the repeatability threshold of
-fold-change measurements, at maybe 0.1 to 0.3 kT.
-\item Surprising that so many model parameters can be inferred from just a few
+Perhaps urprising that so many model parameters can be inferred from just a few
 distributions that, by eye, don't have strong features! Does require some
@@
 quite reasonable.
-\end{itemize}
 }

### 2020-06-05 doc/section_01_introduction.tex
Not quite final pass on everything, Appdx S3 still needs more

@@
 Gertz2009, Sherman2012, Saiz2013, Ko1991, Peccoud1995, Record1996, Kepler2001,
-Sanchez2008, Shahrezaei2008, Sanchez2011, Michel2010}. One root to such
+Sanchez2008, Shahrezaei2008, Sanchez2011, Michel2010}. One route to such
 predictive understanding is to focus on the simplest regulatory architecture and
@@
 dissection of transcription \cite{Oehler1994, Oehler1990}. All of the proposed
-models models coarse-grain away some of the important microscopic features of
+models coarse-grain away some of the important microscopic features of
 this architecture that have been elucidated by generations of geneticists,
@@
 principles of such coarse-graining in order to increase the predictive power of
-these models. In doing so we hope future models of regulatory response will be
+these models. In doing so, we hope future models of regulatory response will be
 able to serve the powerful predictive role needed to take synthetic biology from
@@
 meaning. For example, we are not satisfied with the strictly phenomenological
-approach offered by the commonly used Hill functions~\cite{Frank2013}. Studies
+approach offered by the commonly used Hill functions. As argued
+in~\cite{Frank2013}, Hill functions are ubiquitous precisely because they
+coarse-grain away all biophysical details into inscrutable parameters. Studies
 like~\cite{Razo-Mejia2018} have demonstrated that Hill functions are clearly
@@
 then those processes are irrelevant for our purposes. Forward modeling and
-inverse (statistical inferential) modeling are necessary to tackles such
+inverse (statistical inferential) modeling are necessary to tackle such
 questions.

### 2020-06-05 doc/section_02_means.tex
Finish appendix on mean and Fanos for nonbursty models
Not quite final pass on everything, Appdx S3 still needs more

@@
 \mbox{average rate of transcription} = \sum_i r_i p_i,
+\label{eq:transcrip_prop_pbound}
 \end{equation}
@@
 and without a transcription factor present, any biological noise common to both
-conditions can be made to cancel away. In addition, the presence of different
-operator sites, i.e. binding sites for the repressor, is known to change the
-basal expression level in the absence of repressor~\mmnote{look up
-cite}\mrm{does reg-seq mention anything like this in their SI? I definitely saw
-that on my own experiments, but didn't mention it in the paper}. Computing
-fold-changes neatly cancels this effect as well.
+conditions can be made to cancel away.
+% \mmnote{I'd love to include the factoid below but I can't find
+% a cite to back it up...}
+% In addition, the presence of different
+% operator sites, i.e. binding sites for the repressor, is known to change the
+% basal expression level in the absence of repressor~\mmnote{look up
+% cite}\mrm{does reg-seq mention anything like this in their SI? I definitely saw
+% that on my own experiments, but didn't mention it in the paper}. Computing
+% fold-changes neatly cancels this effect as well.
 
@@
 the promoter is idealized as existing in one of two states, either repressor
-bound or repressor unbound. The rate of transcription, and therefore the level
-of gene expression, is assumed to be proportional to the fraction of time spent
-in the repressor unbound state. Mathematically, this amounts to the statement
-that the mean mRNA level is given by
+bound or repressor unbound. The rate of transcription
+is assumed to be proportional to the fraction of time spent
+in the repressor unbound state.
+From the relative statistical weights listed in
+Figure~\ref{fig1:means_cartoons}, the probability $p_U$ of being in the
+unbound state is
+\begin{equation}
+p_U = \left(1 + \frac{R}{N_{NS}} e^{-\beta\Delta\varepsilon_R}\right)^{-1}.
+\end{equation}
+The mean rate of transcription is then given by $r p_U$, as assumed by
+Eq.~\ref{eq:transcrip_prop_pbound}. The mean number of mRNA is
+set by the balance of average mRNA transcription and degradation rates,
+so it follows that the mean mRNA level is given by
 \begin{equation}
@@
 The fold-change is  found as the ratio of mean mRNA with and without repressor
-as introduced in eqn.~\ref{eq:fc_def}. Invoking that definition results in
+as introduced in Eq.~\ref{eq:fc_def}. Invoking that definition results in
 \begin{equation}
@@
 which clearly matches the form of the master curve in
-Figure~\ref{fig1:means_cartoons}(C) with $\rho=1$ and $\Delta F_R = 
+Figure~\ref{fig1:means_cartoons}(D) with $\rho=1$ and $\Delta F_R = 
 \beta\Delta\varepsilon_r - \log (R / N_{NS})$.
@@
 as the coarse-graining of any equilibrium promoter model in which no
-transcriptionally active states have transcription factor bound. Put
-differently, there is no overlap between TF bound states and transcriptionally
+transcriptionally active states have transcription factor bound, or put
+differently, when there is no overlap between
+transcription factor bound states and transcriptionally
 active states. We will see this explicitly in the 3-state equilibrium model
@@
 
-Let $p_{m,R}(t)$ denote the joint probability of finding the system in the
+Let $p_R(m,t)$ denote the joint probability of finding the system in the
 repressor bound state $R$ with $m$ mRNA molecules present at time $t$. Similarly
-define $p_{m,U}(t)$ for the repressor unbound state $U$. This model is governed
-by coupled master equations giving the time evolution of $p_{m,R}(t)$ and
-$p_{m,U}(t)$~\cite{Sanchez2008, Sanchez2011, Phillips2019} which we can write as
+define $p_U(m,t)$ for the repressor unbound state $U$. This model is governed
+by coupled master equations giving the time evolution of $p_R(m,t)$ and
+$p_U(m,t)$~\cite{Sanchez2008, Sanchez2011, Phillips2019} which we can write as
 \begin{align}
 \begin{split}
-\deriv{t}p_{m,R}(t) =& 
-- \overbrace{k_R^- p_{m,R}(t)}^{R \rightarrow U}
-+ \overbrace{k_R^+ p_{m,U}(t)}^{U \rightarrow R}
-+ \overbrace{(m+1)\gamma p_{m+1,R}(t)}^{m + 1 \rightarrow m}
-- \overbrace{\gamma p_{m,R}(t)}^{m \rightarrow m - 1}
+\deriv{t}p_R(m,t) =& 
+- \overbrace{k_R^- p_R(m,t)}^{R \rightarrow U}
++ \overbrace{k_R^+ p_U(m,t)}^{U \rightarrow R}
++ \overbrace{(m+1)\gamma p_R(m+1,t)}^{m + 1 \rightarrow m}
+- \overbrace{\gamma p_R(m,t)}^{m \rightarrow m - 1}
 \\
-\deriv{t}p_{m,U}(t) =&\; 
-\overbrace{k_R^- p_{m,R}(t)}^{R \rightarrow U}
-- \overbrace{k_R^+ p_{m,U}(t)}^{U \rightarrow R}
-+ \overbrace{rp_{m-1,U}(t)}^{m-1 \rightarrow m}
-- \overbrace{rp_{m,U}(t)}^{m \rightarrow m + 1}
+\deriv{t}p_U(m,t) =&\; 
+\overbrace{k_R^- p_R(m,t)}^{R \rightarrow U}
+- \overbrace{k_R^+ p_U(m,t)}^{U \rightarrow R}
++ \overbrace{rp_U(m-1,t)}^{m-1 \rightarrow m}
+- \overbrace{rp_U(m,t)}^{m \rightarrow m + 1}
 \\
-&+ \overbrace{(m+1)\gamma p_{m+1,U}(t)}^{m + 1 \rightarrow m}
-- \overbrace{\gamma p_{m,U}(t)}^{m \rightarrow m - 1},
+&+ \overbrace{(m+1)\gamma p_U(m+1,t)}^{m + 1 \rightarrow m}
+- \overbrace{\gamma p_U(m,t)}^{m \rightarrow m - 1},
 \label{eq:poisson_promoter_cme}
@@
 describe transcription initiation increasing the mRNA copy number by one.
+We direct the reader to Appendix~\ref{sec:cme_from_cartoon} for a careful
+treatment showing how the form of this master equation follows from the
+corresponding cartoon in Figure~\ref{fig1:means_cartoons}.
 
@@
 \begin{equation}
-\vect{p}_m = \begin{pmatrix} p_{m,R} \\ p_{m,U} \end{pmatrix},\
-\matr{K} = \begin{pmatrix} -k_R^- & k_R^+ \\ k_R^- & -k_R^+ \end{pmatrix},\
-\matr{R} = \begin{pmatrix} 0 & 0 \\ 0 & r \end{pmatrix},\
+\vec{p}(m) = \begin{pmatrix} p_R(m) \\ p_U(m) \end{pmatrix},\
+\mathbf{K} = \begin{pmatrix} -k_R^- & k_R^+ \\ k_R^- & -k_R^+ \end{pmatrix},\
+\mathbf{R} = \begin{pmatrix} 0 & 0 \\ 0 & r \end{pmatrix},\
 \label{eq:2state_cme_matrices}
@@
 \begin{equation}
-\deriv{t}\vect{p}_m(t) =
-\left( \matr{K} - \matr{R} - \gamma m \matr{I} \right) \vect{p}_m(t)
-                + \matr{R} \vect{p}_{m-1}(t)
-                + \gamma (m+1) \matr{I} \vect{p}_{m+1}(t),
+\deriv{t}\vec{p}(m,t) =
+\left( \mathbf{K} - \mathbf{R} - \gamma m \mathbf{I} \right) \vec{p}(m,t)
+                + \mathbf{R} \vec{p}(m-1,t)
+                + \gamma (m+1) \mathbf{I} \vec{p}(m+1,t),
 \label{eq:2state_rep_cme}
 \end{equation}
-where $\matr{I}$ is the identity matrix. Taking steady state by setting time
+where $\mathbf{I}$ is the identity matrix. Taking steady state by setting time
 derivatives to zero, the mean mRNA can be found to be
@@
         \left(1 + \frac{k_R^+}{k_R^-}\right)^{-1},
+\label{eq:mean_m_model1}
 \end{equation}
@@
 The master equation governing this model is analogous
-to~\eqrange{eq:2state_cme_matrices}{eq:2state_rep_cme} for model 1 above. The
+to Eqs.~\ref{eq:2state_cme_matrices}-\ref{eq:2state_rep_cme} for model 1 above. The
 main subtlety arises since transcription initiation accompanies a promoter state
-change. This can be understood by analogy to $\matr{K}$. The off-diagonal and
-diagonal elements of $\matr{K}$ correspond to transitions arriving at or
+change. This can be understood by analogy to $\mathbf{K}$. The off-diagonal and
+diagonal elements of $\mathbf{K}$ correspond to transitions arriving at or
 departing from, respectively, the promoter state of interest. If transcription
@@
 departing transitions have different initial copy numbers of mRNA, unlike for
-$\matr{K}$ where they are the same (see Appendix~\cite{sec:non_bursty}). The
+$\mathbf{K}$ where they are the same (see Appendix~\ref{sec:non_bursty}). The
 master equation for this model is
 \begin{equation}
-\deriv{t}\vect{p}_m(t) =
-\left( \matr{K} - \matr{R_D} - \gamma m \matr{I} \right) \vect{p}_m(t)
-                + \matr{R_A} \vect{p}_{m-1}(t) +
-                \gamma (m+1) \matr{I} \vect{p}_{m+1}(t),
+\deriv{t}\vec{p}(m,t) =
+\left( \mathbf{K} - \mathbf{R_D} - \gamma m \mathbf{I} \right) \vec{p}(m,t)
+                + \mathbf{R_A} \vec{p}(m-1,t) +
+                \gamma (m+1) \mathbf{I} \vec{p}(m+1,t),
 \label{eq:3state_rep_cme}
@@
 \begin{equation}
-\vect{p}_m = \begin{pmatrix} p_{m,R} \\ p_{m,E} \\ p_{m,P} \end{pmatrix},\
-\matr{K} = \begin{pmatrix} -k_R^- & k_R^+ & 0 \\
+\vec{p}(m) = \begin{pmatrix} p_R(m) \\ p_E(m) \\ p_P(m) \end{pmatrix},\
+\mathbf{K} = \begin{pmatrix} -k_R^- & k_R^+ & 0 \\
                         k_R^- & -k_R^+ -k_P^+ & k_P^- \\
@@
 \begin{equation}
-\matr{R_A} = \begin{pmatrix}
+\mathbf{R_A} = \begin{pmatrix}
                 0 & 0 & 0 \\ 
@@
                 \end{pmatrix},\
-\matr{R_D} = \begin{pmatrix}
+\mathbf{R_D} = \begin{pmatrix}
                 0 & 0 & 0 \\ 
@@
 \end{equation}
-The elements of $\vect{p}_m$ encode the probabilities of having $m$ mRNA present
+The elements of $\vec{p}(m)$ encode the probabilities of having $m$ mRNA present
 along with the promoter having repressor bound ($R$), being empty ($E$), or
-having RNAP bound ($P$), respectively. $\matr{R_A}$ describes probability flux
-\textit{arriving} at the state $\vect{p}_m$ from a state with one fewer mRNA,
-namely $\vect{p}_{m-1}$, and $\matr{R_D}$ describes probability flux
-\textit{departing} from the state $\vect{p}_m$ for a state with one more mRNA,
-namely $\vect{p}_{m+1}$. $\matr{K}$ is closely analogous to model 1.
+having RNAP bound ($P$), respectively. $\mathbf{R_A}$ describes probability flux
+\textit{arriving} at the state $\vec{p}(m)$ from a state with one fewer mRNA,
+namely $\vec{p}(m-1)$, and $\mathbf{R_D}$ describes probability flux
+\textit{departing} from the state $\vec{p}(m)$ for a state with one more mRNA,
+namely $\vec{p}(m+1)$. $\mathbf{K}$ is closely analogous to model 1.
 
@@
         \frac{k_R^- k_P^+}
-        {k_R^- k_P^+ + k_R^- (k_P^- + r) + k_R^+ (k_P^- + r)}.
+        {k_R^- k_P^+ + k_R^- (k_P^- + r) + k_R^+ (k_P^- + r)},
+\label{eq:model2_meanm}
 \end{equation}
-Fold-change is again found from the ratio prescribed by~\eq{eq:fc_def}, from
+and with details again deferred to Appendix~\ref{sec:non_bursty}.
+Fold-change is again found from the ratio prescribed by Eq.~\ref{eq:fc_def}, from
 which we have
@@
 \begin{equation}
-\deriv{t}\vect{p}_m(t) =
-\left( \matr{K} - \matr{R_D} - \gamma m \matr{I} \right) \vect{p}_m(t)
-                + \matr{R_A} \vect{p}_{m-1}(t) +
-                \gamma (m+1) \matr{I} \vect{p}_{m+1}(t),
+\deriv{t}\vec{p}(m,t) =
+\left( \mathbf{K} - \mathbf{R_D} - \gamma m \mathbf{I} \right) \vec{p}(m,t)
+                + \mathbf{R_A} \vec{p}(m-1,t) +
+                \gamma (m+1) \mathbf{I} \vec{p}(m+1,t),
 \end{equation}
@@
 \begin{equation}
-\vect{p}_m = \begin{pmatrix} p_{m,R} \\ p_{m,E} \\
-                             p_{m,C} \\ p_{m,O}\end{pmatrix},\
-\matr{K} = \begin{pmatrix} -k_R^- & k_R^+ & 0 & 0\\
+\vec{p}(m) = \begin{pmatrix} p_R(m) \\ p_E(m) \\
+                             p_C(m) \\ p_O(m)\end{pmatrix},\
+\mathbf{K} = \begin{pmatrix} -k_R^- & k_R^+ & 0 & 0\\
                         k_R^- & -k_R^+ -k_P^+ & k_P^- & 0 \\
@@
 ($E$), RNAP closed complex ($C$), and RNAP open complex ($O$). Besides
-increasing dimension by one, the only new feature in $\matr{K}$ is the
+increasing dimension by one, the only new feature in $\mathbf{K}$ is the
 rate $k_O$, representing the (irreversible) rate of open complex formation from
@@
 \begin{equation}
-\matr{R_A} = \begin{pmatrix}
+\mathbf{R_A} = \begin{pmatrix}
         0 & 0 & 0 & 0 \\ 
@@
                 \end{pmatrix},\
-\matr{R_D} = \begin{pmatrix}
+\mathbf{R_D} = \begin{pmatrix}
         0 & 0 & 0 & 0 \\ 
@@
         {k_R^- [(k_P^+ (k_O + r) + r(k_P^- + k_O)] + k_R^+ r(k_P^- + k_O)},
+\label{eq:model3_mean_m}
 \end{equation}
@@
 transcription steps may still be collapsed to the form of the master curve for
-the means giving in~\fig{fig1:means_cartoons}(D) so long as the repressor
+the means given in Figure~\ref{fig1:means_cartoons}(D), so long as the repressor
 binding is exclusive with transcriptionally active states. We offer this as a
@@
 \begin{equation}
-\matr{R} = \begin{pmatrix}
+\mathbf{R} = \begin{pmatrix}
                 0 & 0 & 0 \\ 
@@
 \end{equation}
-The master equation takes verbatim the same form as it did for model
-1,~\eq{eq:2state_rep_cme}. Meanwhile the promoter transition matrix $K$ is the
-same as~\eq{eq:3state_cme_matrices_pt1} from model 2, although we relabel the
+The master equation takes verbatim the same form as it did for model 1,
+Eq.~\ref{eq:2state_rep_cme}. Meanwhile the promoter transition
+matrix $\mathbf{K}$ is the same as Eq.~\ref{eq:3state_cme_matrices_pt1}
+from model 2, although we relabel the
 rate constants from $k_P^\pm$ to $k^\pm$ to reiterate that these are not simply
@@
 with some characteristic rate $r$. The current model by contrast produces, at
-some Poisson rate $k_i$, \textit{bursts} of mRNA transcripts with a
-geometrically distributed burst size and a mean burst size $b$.
+some Poisson rate $k_i$, \textit{bursts} of mRNA transcripts.
+The burst sizes are assumed to be geometrically distributed with
+a mean burst size $b$.
 
@@
 Why does the master curve of~\fig{fig1:means_cartoons}(D) exist at all?
-\textcolor{blue}{This brings to mind the deep questions posed in,
-e.g.,~\cite{Frank2013} and~\cite{Frank2014a}, asking us to search for general
-``symmetries'' of the problem instead of specifying generative models of the
-known microscopics of a system} \mrm{I don't understand this last sentence. Can
-we make it more transparent?}. In our case, the key feature seems to be the
+This brings to mind the deep questions posed in,
+e.g.,~\cite{Frank2013} and~\cite{Frank2014a}, suggesting we consider multiple
+plausible models of a system and search for their common patterns to tease
+out which broad features are and are not important.
+In our case, the key feature seems to be the
 exclusive nature of repressor and RNAP binding, which allows the parameter
@@
 cell's physiology is to manipulate RNAP binding sites. Work such
-as~\cite{Kinney2010} has shown that models of sequence-dependent RNAP can be
+as~\cite{Kinney2010} has shown that models of sequence-dependent RNAP affinity can be
 inferred from data, and the authors of~\cite{Brewster2012} showed that the model
@@
 nonequilibrium model 4 in~\fig{fig1:means_cartoons}(C), with zero repressor
-(looking ahead, this is equivalent to model 4 in~\fig{fig2:constit_cartoons}),
+(looking ahead, this is equivalent to model 4
+in Figure~\ref{fig2:constit_cartoons}(A)),
 to single-cell mRNA counts data from~\cite{Brewster2014}. This provided a
@@
 $\Delta(\Delta F_R) \approx \Delta\rho$. Suppose we are lucky and $\Delta\rho$
-happens to be $\sim0.1$, on the high end of our range estimated above. A
+is $\sim0.1$, on the high end of our range estimated above. A
 determination of $\Delta\varepsilon_R/k_BT$ with an uncertainty of barely 0.1
@@
 promising path to pursue experimentally. It is doubtful that inference of
-repressor kinetic rates would be any easier.\footnote{\mmnote{In fact the
-uncertainties quoted in Hammar et.\ al.~\cite{Hammar2014} from Elf's group are
-small enough to do this calculation, but HJ \& I think their error bars are
-insanely optimistic. Not sure I want to blatantly call them out on it though??}}
+repressor kinetic rates would be any easier.%\footnote{\mmnote{In fact the
+% uncertainties quoted in Hammar et.\ al.~\cite{Hammar2014} from Elf's group are
+% small enough to do this calculation, but HJ \& I think their error bars are
+% insanely optimistic. Not sure I want to blatantly call them out on it though??}}
 
@@
 distributions.
-\marginpar{\it MM: Add a notation aside somewhere that all association rates are written as zeroth-order rates.
-In other words, the concentration of the molecule is hidden inside, so for instance, $k_R^+$ is proportional to repressor copy number.}

### 2020-06-05 doc/section_03_beyond_means.tex
Finish appendix on mean and Fanos for nonbursty models
Not quite final pass on everything, Appdx S3 still needs more

@@
 we can choose, and we show several potential kinetic models for constitutive
-promoters in~\fig{fig2:constit_cartoons}. Let us consider the suitability of
-each model for our purposes in turn.
+promoters in Figure~\ref{fig2:constit_cartoons}(A). Let us consider
+the suitability of each model for our purposes in turn.
 
@@
 The simplest model of constitutive expression that we can imagine is shown as
-model 1 in Figure~\ref{fig2:constit_cartoons} and assumes that transcripts are
+model 1 in Figure~\ref{fig2:constit_cartoons}(A) and assumes that transcripts are
 produced as a Poisson process from a single promoter state. This is the picture
@@
 the corrections to Poisson behavior are derived as additive corrections to the
-Fano factor. For disproving the ``universal noise curve'' from So et
-al~\cite{So2011}, this picture was excellent. It is appealing in its simplicity,
+Fano factor. For disproving the ``universal noise curve'' from So et.\
+al.~\cite{So2011}, this picture was excellent. It is appealing in its simplicity,
 with only two parameters, the initiation rate $r$ and degradation rate $\gamma$.
@@
 the promoter is either empty or bound by RNAP ~\cite{Phillips2015a,
-Phillips2019}. Here,  as shown in model 2 of Figure~\ref{fig2:constit_cartoons},
+Phillips2019}. Here,  as shown in model 2 of Figure~\ref{fig2:constit_cartoons}(A),
 transcription initiation results in a state transition from the bound to the
@@
              \left(\gamma + \rate{k}{P}{+} + \rate{k}{P}{-} + r\right)}.
+\label{eq:model2_fano}
 \end{align}
@@
 Perhaps adding some of this mechanistic detail as shown in model 3 of
-Figure~\ref{fig2:constit_cartoons} might rescue the previous model. The next
+Figure~\ref{fig2:constit_cartoons}(A) might rescue the previous model. The next
 simplest refinement of that model could consider open complex formation and
@@
 Unfortunately, as hinted at in earlier work~\cite{Mitarai2015}, this model too
-has Fano factor $\nu<1$. This can be seen by starting with their Eq.~A1 for the
-Fano factor of the analogous model with repressor and taking the limit as
-repressor concentration goes to zero. After substantial algebra, the result is
+has Fano factor $\nu<1$. We again leave the algebraic details for
+Appendix~\ref{sec:non_bursty} and merely state the result that
 \begin{align}
-\nu = 1 - \frac{r k_O k_P^+}{\mathcal{Z}}
-\frac{k_P^+ + k_P^- + k_O + \gamma}
-        {\mathcal{Z} + \gamma(k_P^+ + k_P^- + k_O + \gamma) + \gamma^2},
+\nu = 1 - \frac{r k_P^+ k_O}{\mathcal{Z}}
+\frac{k_P^+ + k_P^- + k_O + r + \gamma}
+    {\mathcal{Z} + \gamma(k_P^+ + k_P^- + k_O + r) + \gamma^2},
+\label{eq:model3_fano}
 \end{align}
@@
 
-\marginpar{\it RP: very fun next paragraph - I like the argument, need to think more}
 In fact, we suspect \textit{any} model in which transcription proceeds through a
@@
 exponential with the same mean.
-\footnote{This can be made more precise with a result from~\cite{Stewart2007},
-who showed that the convolution of multiple gamma distributions (of which the
-exponential distribution is a special case) is, to a very good approximation,
-also gamma distributed. Using their Eq.~2 for the distribution of the
-convolution, with shape parameters set to 1 to give exponential distributions,
-the total waiting time distribution has a ratio of variance to squared mean
-$\sigma^2/\mu^2 = \sum_i k_i^2/\left(\sum_i k_i\right)^2$, where the $k_i$ are
-the rates of the individual steps. Clearly this is less than 1 and therefore the
-total waiting time distribution is narrower than the corresponding exponential.}
 A less disperse waiting time distribution means transcription initations are
@@
 to Poisson, giving $\nu<1$.
-\footnote{This last step, while intuitive, can be argued by analogy to photon
-statistics where antibunching gives rise to sub-Poissonian noise~\cite{Paul1982,
-Zou1990}. Although loopholes exist, we do not expect they apply for our problem.
-Nevertheless we refrain from elevating this cycle/sub-Poissonian equivalence to
-a ``theorem.'' \mmnote{Antibunching is the ``obvious'' analogy for me but it'd
-be totally out of left field for most readers, any suggestions on a more
-comprehensible reference??}}
+(In Appendix~\ref{sec:non_bursty} we present a more precise version
+of the intuitive arguments in this paragraph.)
 Regardless of the merits of this model in describing the noise properties of
@@
 
-Inspired by~\cite{Razo-Mejia2020}, we revisit an analog of model 2, but as with
-the models considered in \ref{section_02_means}, the interpretation of the two
+Inspired by~\cite{Razo-Mejia2020}, we next revisit an analog of model 2 in
+Figure~\ref{fig2:constit_cartoons}(A), but as with
+the analogous models considered in Section~\ref{section_02_means},
+the interpretation of the two
 states is changed. Rather than explicitly viewing them as RNAP bound and
@@
 process only when the promoter is in one of the two states in this model,
-transcription initiations should now be ``bunched,'' in contrast to the
+transcription initiations should now be ``bunched'' in time, in contrast to the
 ``anti-bunching'' of models 2 and 3 above. One might further guess that this
@@
 \nu &= 1 + \frac{r k^-}{(k^+ + k^- + \gamma)(k^+ + k^-)},
+\label{eq:model4_fano}
 \end{align}
@@
 
-\marginpar{\it RP: best description I have ever seen of this.  Good job}
+% \marginpar{\it RP: best description I have ever seen of this.  Good job}
 To develop some intuition, consider model 4 in the limit $k^+ \ll k^- \lesssim
@@
 transcriptional bursting, as reported in,
-e.g.,~\cite{Golding2005,Chong2014,Sevier2016}\mmnote{should probably add some
-more cites here}. Let us make this correspondence more precise. The mean dwell
+e.g.,~\cite{Golding2005,Chong2014,Sevier2016}.
+%\mmnote{should probably add some more cites here}.
+Let us make this correspondence more precise. The mean dwell
 time in the active state is $1/k^-$. While in this state, transcripts are
@@
 timescale left in the problem, $1/k^+$.
-\footnote{If instead it were the case that $k^+ \sim k^-$, then the waiting time
-$1/k^+ + 1/k^-$ would have a peaked distribution, but this does not appear to be
-the case for any of datasets from~\cite{Jones2014}.}
+% \footnote{If instead it were the case that $k^+ \sim k^-$, then the waiting time
+% $1/k^+ + 1/k^-$ would have a peaked distribution, but this does not appear to be
+% the case for any of datasets from~\cite{Jones2014}.}
 
@@
 are probably irrelevant for setting the population distribution of mRNA. The
-timescales are simply too fast, and as we will see later
-in~\fig{fig:constit_post}, the noise seems to be governed by slower timescales.
+timescales are simply too fast, and as we will see later in
+Figures~\ref{fig:constit_post_full} and~\ref{fig4:repressed_post_full},
+the noise seems to be governed by slower timescales.
 Perhaps in hindsight this is not surprising: intuitively, the degradation rate

### 2020-06-05 doc/section_04_bayesian_inference.tex
Not quite final pass on everything, Appdx S3 still needs more

@@
 by allowing both Fano factor $\nu>1$, but also by remedying, by design, the
-problems of parameter degeneracy than model 4 in
+problems of parameter degeneracy that model 4 in
 Figure~\ref{fig2:constit_cartoons} suffered~\cite{Razo-Mejia2020}. Does this
@@
 \begin{equation}
-{d\over dt}p_{m,U}(t) = 
-        rp_{m-1,U}(t) 
-        - rp_{m,U}(t)
-        + (m+1)\gamma p_{m+1,U}(t) 
-        - \gamma p_{m,U}(t),
+{d\over dt}p_U(m)(t) = 
+        rp_U(m-1)(t) 
+        - rp_U(m)(t)
+        + (m+1)\gamma p_U(m+1)(t) 
+        - \gamma p_U(m)(t),
 \end{equation}
@@
 where $D = \{m_1, m_2, \ldots, m_N \}$ are the single-cell mRNA experimental
-counts. As is standard we will neglected the denominator $p(D)$ on the right
+counts. As is standard we will neglect the denominator $p(D)$ on the right
 hand side since it is independent of $\lambda$ and serves only as a
@@
 likelihood of the full inference problem $p(D\mid\lambda)$ is simply a product
-of the single cell likelihoods given by \eq{eq:poisson_inference010} above, so
+of the single cell likelihoods given by Eq.~\ref{eq:poisson_inference010} above, so
 \begin{equation}
@@
 As detailed in Appendix~\ref{sec:bayesian} this particular choice of prior is
-known as the \textit{conjugate} prior for a Poisson likelihood. This type of
-priors have nice properties to work with, in particular a conjugate prior allows
-us to get a closed form for the posterior distribution $p(\lambda \mid D)$. For
-our case it can be shown that for a Poisson distribution likelihood with its
+known as the \textit{conjugate} prior for a Poisson likelihood.
+Conjugate priors have the convenient properties that a closed form exists for the posterior distribution $p(\lambda \mid D)$ - unusual in Bayesian inference problems - and the closed form posterior takes the same form as the prior.
+For
+our case of a Poisson distribution likelihood with its
 Gamma distribution conjugate prior, the posterior distribution is also a Gamma
-distribution ~\cite{Gelman2013}. Specifically the two parameters $\alpha'$ and
+distribution~\cite{Gelman2013}. Specifically the two parameters $\alpha'$ and
 $\beta'$ for this posterior distribution take the form $\alpha' = \alpha +
@@
 was ``wrong'' since it could not reproduce features of the data such as a Fano
-factor $> 1$. The fact that we got such a narrow posterior distribution for our
+factor $> 1$. The fact that we obtain such a narrow posterior distribution for our
 parameter $P(\lambda \mid D)$ does not equate to the model being adequate to
-describe the data. What this means is that given the data $D$, the parameter 
-$\lambda$ for our model is highly constrained to take a narrow range of values,
-but a narrow posterior distribution does not speak to the goodness of fit of the
-model. As we will see later in Figure~\ref{fig:constit_post_full} after
-exploring the bursty promoter model, indeed there is not good correspondence
-when contrasting the Poisson model with the experimental data.
+describe the data. What this means is that given the data $D$, only values
+in a narrow range are remotely plausible for the parameter $\lambda$,
+but a narrow posterior distribution does not necessarily
+mean the model accurately depicts reality.
+As we will see later in Figure~\ref{fig:constit_post_full} after
+exploring the bursty promoter model, indeed the correspondence
+when contrasting the Poisson model with the experimental data is quite poor.
 
@@
 mRNA degradation rate $\gamma$. As sketched earlier, to think of the negative
-binomial distribution in terms of an intuitive story, we imagine the arrival of
+binomial distribution in terms of an intuitive ``story,'' in the precise
+meaning of~\cite{Blitzstein2015}, we imagine the arrival of
 bursts as a Poisson process with rate $k_i$, where each burst has a
@@
 constitutive model considered in~\cite{Razo-Mejia2020} (nonequilibrium model 4
-in~\ref{fig2:constit_cartoons}), this model is well-identified with both
+in Figure~\ref{fig2:constit_cartoons}(A)), this model is well-identified with both
 parameters determined to a fractional uncertainty of 5-10\%. The strong
@@
 encloses approximately 95\% of the samples, and thus 95\% of the probability
-mass, of its posterior distribution. Theory-experiment comparisons (not shown)
+mass, of its posterior distribution. Theory-experiment comparisons,
+shown in Figure~\ref{figS:ppc_unreg} in Appendix~\ref{sec:bayesian},
 display a similar level of agreement between data and predictive samples as for
 the bursty model with \textit{lacUV5} in \fig{fig:constit_post_full}(B).
-\mmnote{Might be worth showing some/all of these PPCs in supplement for
-completeness? We could do postage stamp sized diff plots and probably fit all 18
-promoters on 1-2 pages.}\mrm{Definitely need to show them in SI.}
 
@@
 
-\mmnote{Outline, still to cover:
-\begin{itemize}
-\item There is a puzzle in comparing w/ Chong2014 supercoiling
-model: if supercoiling is the thing, why are my burst sizes all
-the same but burst rates vary? And why is the duty cycle of my
-promoter so low, ie., bursts so short? COmpare their fig 7E;
-their $\beta/\alpha$ is my $k^+/k^-$. They have one or two genes
-with very small $\beta/\alpha$, does the \textit{galK} locus just
-happen to be that, or is there a deeper disagreement? Hard to say
-w/o more data. Are we
-missing important things?? Unclear, we leave as open question. Good enough for
-our purposes: posterior is identifiable, PPC is great, and of the models we've
-thought of it is unique in satisfying both.
-\end{itemize}
-}
+We mention one further comparison, between our inferred parameters and the
+work of Chong et.\ al.~\cite{Chong2014}, which is interesting and puzzling.
+Beautiful experiments in~\cite{Chong2014} convincingly argue that
+supercoiling accumulated from the production of mRNA transcripts is key
+in setting the burstiness of mRNA production.
+In their model, this supercoiling occurs on the scale of $\sim100$~kb
+domains of DNA. This suggests that all genes on a given domain should burst
+in synchrony, and that the difference between highly and lowly expressed
+genes is the \textit{size} of transcriptional bursts, not the
+\textit{time between} bursts.
+But here, all burst sizes we infer in Figure~\ref{fig:constit_post_full}(C)
+are comparable and burst rates vary wildly. It is not immediately clear
+how to square this circle. Furthermore, Figure~7E in~\cite{Chong2014}
+reports values of the quantity they label $\beta/\alpha$ and we label
+$k^+/k^-$ in model 4 from Figure~\ref{fig2:constit_cartoons}. In contrast
+to the findings of~\cite{Razo-Mejia2020}, Chong et.\ al.\ do not find
+$k^+/k^-\ll1$ for most of the genes they consider. This begs the question:
+is the \textit{galK} chromosomal locus used for the reporter constructs
+in~\cite{Razo-Mejia2020} and~\cite{Jones2014} merely an outlier,
+or is there a deeper puzzle here waiting to be resolved? Without more
+apples-to-apples data we can only speculate, and we leave it as an
+intriguing open question for the field.
+
+Despite such puzzles, our goal here is not to unravel the mysterious origins
+of burstiness in transcription. Our remaining task in this work is a
+determination of the physical reality of equilibrium binding energies in
+Figure~\ref{fig1:means_cartoons}, as codified by the equilibrium-nonequilibrium
+equivalence of Eq.~\ref{eq:deltaFR_eq_noneq_equiv}.
+For our phenomenological needs here model 5 in
+Figure~\ref{fig2:constit_cartoons} is more than adequate: the posterior
+distributions in Figure~\ref{fig:constit_post_full}(C) are cleanly
+identifiable and the predictive checks in Figure~\ref{figS:ppc_unreg}
+indicate no discrepancies between the model and the mRNA single-moleucle
+count data of~\cite{Jones2014}. Of the models we have considered it
+is unique in satisfying both these requirements. So we will happily use
+it as a foundation to build upon in the next section when we add regulation.
 
@@
 factor does not affect the mean burst size nor the burst initiation rate. The
-regulation happens as the promoter is taken away from the transcriptionally
+regulation occurs as the promoter is taken away from the transcriptionally
 active state when the promoter is bound by repressor. We assume that there is
@@
 a subset of the datasets from Jones et.\ al.~\cite{Jones2014}, as discussed more
-in Appendix~\ref{sec:bayes}.
+in Appendix~\ref{sec:bayesian}.
 
@@
 concentration of the small-molecule that controlled the expression of the LacI
-repressor (see Appendix~\ref{sec:bayes}). This is because for this particular
+repressor (see Appendix~\ref{sec:bayesian}). This is because for this particular
 dataset the repressor copy numbers were not measured directly, but it is safe to
@@
 \end{equation}
-where $_2F_1$ is the confluent hypergeometric function of the second kind,
+where $_2F_1$ is the confluent hypergeometric function of the second kind and
 $\alpha$ and $\beta$, defined for notational convenience, are
@@
 we instead resorted to numerics. One insight we found
-\mmnote{a plot would probably help explain, maybe SI, need to explain better}
 was that for very strong or very weak repression, the distribution in
@@
 promoters above. All details for our prior distributions are listed in 
-Appendix~\ref{sec:bayes}.
+Appendix~\ref{sec:bayesian}.
 
@@
 also a small offset between experiment and simulation for O2 at the higher
-repressor copy numbers, especially at 2~ng/mL. We suspect this may be caused by
-the very large number of zero mRNA counts for Oid, 2~ng/mL skewing it's partner
-datasets through the shared association rate, and so we are not too concerned by
-this discrepancy either. None of these fairly minor quibbles cause us to
+repressor copy numbers, especially at 2 and 10~ng/mL.
+From the estimate of repressor copy numbers from~\cite{Jones2014}, it is
+possible that the repressor copy numbers here are becoming large enough to
+partially invalidate our assumption of a separation of timescales between
+burst duration and repressor association rate. Another possibility is that
+the very large number of zero mRNA counts for Oid, 2~ng/mL is skewing it's partner
+datasets through the shared association rate.
+None of these fairly minor quibbles cause us to
 seriously doubt the overall correctness of our model, which further validates

### 2020-06-05 doc/section_05_discussion.tex
Not quite final pass on everything, Appdx S3 still needs more

@@
 sequenced. But there is a troubling Achilles heel buried in all of that genomic
-data and that is our inability to find and interpret regulatory sequence. In
+data, which is our inability to find and interpret regulatory sequence. In
 many cases, this is not possible even qualitatively, let alone the possibility
@@
 of gene expression found in a population of cells, the intrinsic cell-to-cell
-variability, and the full gene expression distribution found in such population
+variability, and the full gene expression distribution found in such a population
 as would be reported in a single molecule mRNA Fluorescence in situ
@@
 Fano factors (variance/mean) that are less than one, at odds with any and all
-experimental data that we are aware of\cite{So2011, Jones2014}. This theoretical
+experimental data that we are aware of~\cite{So2011, Jones2014}. This theoretical
 result allows us to directly discard a subset of the models (models 1-3 in
@@
 to be able to reproduce the data. A previous exploration of model 4 demonstrated
-the sloppy nature of the model in which data on single-cell mRNA counts alone
+the ``sloppy''~\cite{Transtrum2015} nature of the model
+in which data on single-cell mRNA counts alone
 cannot constrain the value of all parameters simultaneously
 \cite{Razo-Mejia2020}. Here we demonstrate that the proposed one-state bursty
-promoter model (model 5 in Figure~\ref{fig2:constit_cartoons}) emerges as a
+promoter model (model 5 in Figure~\ref{fig2:constit_cartoons}(A)) emerges as a
 limit of the commonly used two-state promoter model \cite{Peccoud1995,
@@
 It is interesting to speculate what microscopic details are being coarse-grained
-by our burst rate and burst size in Figure~\ref{fig2:constit_cartoons} model 5.
+by our burst rate and burst size in Figure~\ref{fig2:constit_cartoons}, model 5.
 Chromosomal locus is one possible influence we have not addressed in this work,
@@
 them can now be formulated as a rigorous quantitative question. In particular,
-many aspects of the simple repression motif seem to merit the claim that they
-are sufficiently well understood to use that understanding to design new
-versions of that architecture based upon predictions about how repressor copy
+we believe understanding of the simple repression motif has 
+advanced sufficiently that the design of new
+versions of the architecture is now possible,
+based upon predictions about how repressor copy
 number and DNA binding site strength control expression. In our view, the next
 step in the progression is to first perform similar rigorous analyses of the
-fundamental ``basis set'' of regulatory architectures. Natural followups to this
-work is the exploration of motifs such as simple activation that is
+fundamental ``basis set'' of regulatory architectures. Natural follow-ups to this
+work are explorations of motifs such as simple activation that is
 regulated by a single activator binding site, and the repressor-activator
 architecture, mediated by the binding of both a single activator and a single
-repressor and beyond. With the individual input-output functions in hand,
+repressor, and beyond. With the individual input-output functions in hand,
 similar quantitative dissections including the rigorous analysis of their tuning

### 2020-06-06 code/figures/main/fig02B.py
fix figures according to Griffin's comments

@@ -77,7 +77,7 @@ ax.text(
 # super-Poissonian
 ax.axhspan(1, 10, facecolor='#E8B19D', alpha=0.5, zorder=-100)
 ax.text(
-    5
+    5,
     2.5,
     "super-Poissonian",
     fontsize=10,
@@ -86,14 +86,6 @@ ax.text(
 )
 # Poissonian
 ax.axhline(1, color="black", linestyle=":", linewidth=3)
-ax.text(
-    5,
-    1,
-    "Poissonian",
-    fontsize=10,
-    horizontalalignment="center",
-    verticalalignment="bottom",
-)
 
 # Define colormap
 cmap = "magma"

### 2020-06-06 code/figures/main/fig03.py

@@ -74,8 +74,9 @@ ax_a1.get_xaxis().set_visible(False)
 ax_a1.get_yaxis().set_visible(False)
 ax_a3.get_xaxis().set_visible(False)
 ax_a3.get_yaxis().set_visible(False)
-
-
+# Set facecolor white
+ax_a1.set_facecolor("white")
+ax_a3.set_facecolor("white")
 
 # Set girdspec for UV5 posterior predictive checks
 gs_b = fig.add_gridspec(
@@ -119,20 +120,19 @@ plt.gcf().text(0.48, 0.48, "(D)", fontsize=14)
 # Set UV5 parameters samples
 alpha_samples = all_samples['UV5'].posterior.alpha.values.flatten()
 b_samples = all_samples['UV5'].posterior.b.values.flatten()
+log_prob = all_samples['UV5'].sample_stats.lp.values.flatten()
 
-ax_a2.plot(
+ax_a2.scatter(
     b_samples[::3],
     alpha_samples[::3],
-    'k.',
-    markersize=3,
-    alpha=0.2,
+    c=log_prob[::3],
+    cmap="gray_r",
+    marker="o",
+    s=5,
+    alpha=0.5,
     label='UV5 posterior samples')
 ax_a2.set_ylabel(r'$k_i$ (bursts per mRNA lifetime)')
 ax_a2.set_xlabel(r'$b$ (transcripts per burst)')
-
-# Add grid lines
-[ax_a1.axvline(x, color="white", linewidth=0.5) 
-for x in [3.25, 3.5, 3.75]]
 
 # Plot marginal distributions
 ax_a1.hist(b_samples,
@@ -190,7 +190,7 @@ srep.viz.predictive_ecdf(
     discrete=True,
     ax=ax_b,
     pred_label='Model 5 (N. Binom)',
-    data_label='UV5 data, Jones et. al.',
+    data_label='UV5 data,\n Jones et. al. 2014',
     data_color='black',
     data_size=1, #linewidth
     )
@@ -207,12 +207,15 @@ uv5 = ax_b.plot([], [], color="black")
 
 ax_b.legend(
     [(poisson_2[0], poisson_1[0]), (neg_binom_2[0], neg_binom_1[0]), (uv5[0],),], 
-    ["Model 1 (Poisson)", "Model 5 (N. Binom)", "UV5 data, Jones et. al."],
+    [
+        "Model 1 (Poisson)", 
[…]

### 2020-06-06 code/figures/main/fig04.py

@@ -331,7 +331,6 @@ for k, ax in enumerate([ax_3, ax_4, ax_5]):
     ax.set_xlabel('mRNA counts per cell')
     ax.set_xlim(-2.5, 55)
     srep.viz.titlebox(ax, title, bgcolor="#FFEDC0")
-    # ax.legend(loc='lower right', fontsize='small')
 # Set yaxis label for first plot
 ax_3.set_ylabel('ECDF')
 
@@ -350,8 +349,7 @@ for key, item in aTc_col_dict.items():
     p1 = ax_5.plot([], [], color=colors[-1])
     p2= ax_5.fill([], [], color=colors[0])
     # 6. Fix label name
-    label = expt.split("_")[1]
-    label = label.replace("ngmL", " ng/mL")
+    label = key.replace("ngmL", " ng/mL")
     label = label.replace("0p5", "0.5")
     # 7. Append to lists
     plot_list.append((p2[0], p1[0],))

### 2020-06-06 doc/abstract.tex
Implemented most of GC's edits plus more of my own

@@
 different models are then invoked to confront a variety of different data on mean,
-variance and full gene expression distributiona, illustrating the extent to which
+variance and full gene expression distributions, illustrating the extent to which
 such models can and cannot be distinguished, and suggesting a two-state model with a

### 2020-06-06 doc/appendix_01_nonbursty.tex
Implemented most of GC's edits plus more of my own

@@
 In this section we detail the calculation of mean mRNA levels, fold-changes in
-expression, and Fano factors for nonequilibrium promoter models one through four
+expression, and Fano factors for nonequilibrium promoter models 1 through 4
 in Figure~\ref{fig1:means_cartoons}. These are the results that were quoted but
@@
 cartoons is as a chemical master equation such as Eq.~\ref{eq:2state_rep_cme}
-for model one. Before jumping into the derivations of the general computation of
+for model 1. Before jumping into the derivations of the general computation of
 the mean mRNA level and the Fano factor we will work through the derivation of
@@
 models, the natural mathematicization of their cartoons is as a chemical master
-equation such as Eq.~\ref{eq:2state_rep_cme} for model one, which for
+equation such as Eq.~\ref{eq:2state_rep_cme} for model 1, which for
 completeness we reproduce here as
@@
 
-Next consider the cartoon for nonequilibrium model two in
+Next consider the cartoon for nonequilibrium model 2 in
 Figure~\ref{fig1:means_cartoons}(C). Now we must track probabilities $p_R$,
 $p_P$, and $p_E$ for the repressor bound, empty, and polymerase bound states,
-respectively. By analogy to eqn.~\ref{eq:poisson_promoter_cme_appdx}, the master
-equation for model two can be written
+respectively. By analogy to Eq.~\ref{eq:poisson_promoter_cme_appdx}, the master
+equation for model 2 can be written
 \begin{align}
@@
 \end{equation}
-which is just Eq.~\ref{eq:3state_rep_cme} in the main text. Straightforward if
-tedious algebra verifies that Eqs.~\ref{eq:model2_cme_appdx}
+which is just Eq.~\ref{eq:3state_rep_cme} in the main text. Straightforward
+albeit tedious algebra verifies that Eqs.~\ref{eq:model2_cme_appdx}
 and~\ref{eq:generic_cme_appdx} are in fact equivalent.
@@
 Although we derived Eq.~\ref{eq:generic_cme_appdx} for the particular case of
-nonequilibrium model two in Figure~\ref{fig1:means_cartoons}, in fact the
+nonequilibrium model 2 in Figure~\ref{fig1:means_cartoons}, in fact the
 chemical master equations for all of the nonequilibrium models in
-Figure~\ref{fig1:means_cartoons} except for model five can be cast in this form.
-(We treat model five separately in Appendix~\ref{sec:gen_fcn_appdx}.) Model
-three introduces no new subtleties beyond model two and
+Figure~\ref{fig1:means_cartoons} except for model 5 can be cast in this form.
+(We treat model 5 separately in Appendix~\ref{sec:gen_fcn_appdx}.) Model
+3 introduces no new subtleties beyond model 2 and
 Eq.~\ref{eq:generic_cme_appdx} applies equally well to it, simply with different
-matrices of dimension four instead of three. Models one and four are both
+matrices of dimension four instead of three. Models 1 and 4 are both
 handled by Eq.~\ref{eq:2state_rep_cme} in the main text, which is just
@@
 Recalling that our goal in this section is to derive expressions for mean mRNA
-and Fano factor for nonequilibrium models one through four in
+and Fano factor for nonequilibrium models 1 through four in
 Figure~\ref{fig1:means_cartoons}, and since all four of these models are
@@
 averaging is over all possible mRNA copy numbers and promoter states $S$. For
-example, for model one in Figure~\ref{fig1:means_cartoons}(C), the sum on $S$
+example, for model 1 in Figure~\ref{fig1:means_cartoons}(C), the sum on $S$
 would cover repressor bound and unbound states ($R$ and $U$ respectively), for
-model two, the sum would cover repressor bound, polymerase bound, and empty
+model 2, the sum would cover repressor bound, polymerase bound, and empty
 states ($R$, $P$, and $E$), and so on for the other models.
@@
 squared, conditional on the promoter being in a certain state. For example, for
-model one in Figure~\ref{fig1:means_cartoons} which has repressor bound and
+model 1 in Figure~\ref{fig1:means_cartoons} which has repressor bound and
 unbound states labeled $R$ and $U$, $\langle\vec{m}^2\rangle$ would be
@@
 \subsubsection{Mean mRNA}
-For nonequilibrium model one in Figure~\ref{fig1:means_cartoons}, we have
+For nonequilibrium model 1 in Figure~\ref{fig1:means_cartoons}, we have
 already shown the full master equation in Eq.~\ref{eq:poisson_promoter_cme} and
@@
 \subsubsection{Fano factor}
-To verify that the Fano factor for model one in
+To verify that the Fano factor for model 1 in
 Figure~\ref{fig2:constit_cartoons}(A) is in fact 1 as claimed in the main text,
@@
 \subsubsection{Mean mRNA}
-As shown earlier, the full master equation for model two in
+As shown earlier, the full master equation for model 2 in
 Figure~\ref{fig1:means_cartoons} is
@@
 \end{equation}
-As for model one, we must first find $\mathbf{R_A} \langle\vect{m}^0\rangle$.
+As for model 1, we must first find $\mathbf{R_A} \langle\vect{m}^0\rangle$.
 Denote its components as $p_R$, $p_E$, $p_P$, the probabilities of being found
@@
 \subsubsection{Mean mRNA}
-In close analogy to model two above, nonequilibrium model three from
+In close analogy to model 2 above, nonequilibrium model 3 from
 Figure~\ref{fig1:means_cartoons}(C) can be described by our generic master
@@
 \subsubsection{Mean mRNA}
-The mathematical specification of this model is almost identical to model two.
+The mathematical specification of this model is almost identical to model 2.
 The matrix $\mathbf{K}$ is identical, as is $\mathbf{R_D}$. The only difference
 is that now $\mathbf{R_A}=\mathbf{R_D}$, i.e., both are diagonal, in contrast to
-model two where $\mathbf{R_A}$ has an off-diagonal element, as in
+model 2 where $\mathbf{R_A}$ has an off-diagonal element, as in
 Eq.~\ref{eq:model2_matrices_appdx}. Then the analog of
@@
 In fact we need not do this calculation explicitly and can instead recycle the
-calculation of mean mRNA $\langle{m}\rangle$ from model two. The matrices are
+calculation of mean mRNA $\langle{m}\rangle$ from model 2. The matrices are
 identical except for the relabeling $k^- \longleftrightarrow (k_P^- + r)$, and
-a careful look through the derivation of $\langle{m}\rangle$ for model two shows
+a careful look through the derivation of $\langle{m}\rangle$ for model 2 shows
 that the parameters $k_P^-$ and $r$ only ever appear as the sum $k_P^- + r$. So
-taking $\langle{m}\rangle$ from model two, Eq.~\ref{eq:model2_meanm_appdx}, and
+taking $\langle{m}\rangle$ from model 2, Eq.~\ref{eq:model2_meanm_appdx}, and
 relabeling $(k_P^- + r) \rightarrow k^-$ gives us our answer for model four,
@@
 meaning of the promoter states interpreted differently. Furthermore,
-nonequilibrium model one from Figure~\ref{fig1:means_cartoons} was the model
+nonequilibrium model 1 from Figure~\ref{fig1:means_cartoons} was the model
 considered by Jones et.\ al.~\cite{Jones2014}, where they derived the Fano

### 2020-06-06 doc/appendix_02_bursty.tex
Implemented most of GC's edits plus more of my own

@@
 that this model is rich enough that it can capture the expression pattern of
-bacteria constitutive promoters. Figure~\ref{figS1:bursty_one_state} shows two
+bacterial constitutive promoters. Figure~\ref{figS1:bursty_one_state} shows two
 different schematic representations of the model.
@@
 \lambda \sum_{m'=m + 1}^{\infty} G_{m'-m} p(m, \tau) = \lambda p(m, t) \left[
-\underbrace{\sum_{m'=\textcolor{red}{m}}^{\infty}G_{m'-m}}
+\underbrace{\sum_{m'={m}}^{\infty}G_{m'-m}}
 _{\text{re-index sum to include burst size zero}} -
@@
 \centering
-\includegraphics{../figures/si/figS0X_reindex_sum.pdf}
-\caption{\textbf{Reindexing double sum.} Schematic for how to proceed with the
-reindexing of the sum $\sum_{m=0}^\infty \sum_{m'=0}^m$. The trick consists on
-noticing that this double sum adds up all of the $(m, m')$ pairs in the lower
-triangle part of the $m, m'$ plane.}
+\includegraphics[width=5cm]{../figures/si/figS0X_reindex_sum.pdf}
+\caption{\textbf{Reindexing double sum.} Schematic for reindexing the sum
+$\sum_{m=0}^\infty \sum_{m'=0}^m$. Blue circles depict the 2D grid of
+nonnegative integers restricted to the lower triangular part of the $m, m'$
+plane. The trick is that this double sum runs over all $(m, m')$ pairs with
+$m'\le m$. Summing $m$ first instead of $m'$ requires determining the
+boundary: the upper boundary of the $m'$-first double sum becomes the
+lower boundary of the $m$-first double sum.}
 \label{figS2:sum_reindex}

### 2020-06-06 doc/appendix_03_bayesian_inference.tex
Implemented most of GC's edits plus more of my own

@@
 
-\subsubsection{Bayes theorem}
+\subsubsection{Bayes' theorem}
 
-Bayes theorem is a simple mathematical statement that can apply to \textit{any}
+Bayes' theorem is a simple mathematical statement that can apply to \textit{any}
 logical conjecture. For two particular events $A$ and $B$ that potentially 
-depend on each other Bayes theorem gives us a recipe for how to update our 
+depend on each other Bayes' theorem gives us a recipe for how to update our 
 beliefs about one, let us say $B$, given some state of knowledge, or lack thereof, about
-$A$. In its most classic form Bayes theorem is written as
+$A$. In its most classic form Bayes' theorem is written as
 \begin{equation}
@@
 plausibility of a set of parameter values given our observation of some
-particular dataset. In other words, through the application of Bayes formula we
+particular dataset. In other words, through the application of Bayes' formula we
 update our beliefs on the possible values that parameters can take upon learning
@@
 capable of generating equivalent datasets to the one we got to observe in an 
-experiment. In this way Bayesian inference consists of applying Bayes formula 
+experiment. In this way Bayesian inference consists of applying Bayes' formula 
 as 
@@
 \end{equation}
-Notice than rather than writing the full form of Bayes theorem, we limit 
+Notice than rather than writing the full form of Bayes' theorem, we limit 
 ourselves to the terms that depend on our quantity of interest -- that is the 
@@
 coarse-graining that resulted in a diverse number of parameters for different
-models. For each of these model configurations Bayes theorem demands from us to
+models. For each of these model configurations Bayes' theorem demands from us to
 represent our preconceptions on the possible parameter values in the form of the
@@
 \end{equation}
-Although it is common practice to use non-informative, or maximally
+Although it is not uncommon practice to use non-informative, or maximally
 uninformative priors, we are of the mindset that this is a disservice to the
-philosophical and practical implications of Bayes theorem. It sounds almost
+philosophical and practical implications of Bayes' theorem. It sounds almost
 contradictory to claim that can we represent our thinking about a natural
-phenomena in the form of a mathematical model -- which itself is a bold claim of
-deeply understanding the features that matter in the object of study -- but we
-have absolutely no idea what the parameter values ought to be. We therefore make
+phenomenon in the form of a mathematical model -- in the context of Bayesian
+inference this means choosing a form for the likehihoods, and even making
+this choice presupposes prior understanding or assumptions as to the
+relevant features in the system under study -- but that we have absolutely
+no idea what the parameter values could or could not be. We therefore make
 use of our own expertise, many times in the form of order-of-magnitude
-estimates, to write down prior distributions for our parameters. In summary then
-we don’t want to be maximally uninformative because we aren’t completely
-uninformed.
+estimates, to write down weakly-informative prior distributions
+for our parameters.
 
@@
 as they are broad enough. A way to see why this is the case is to simply look at
-Bayes theorem. For $N ~ 1000-3000$ datum all of the independent of each other
-and $n \ll 10^3$ parameters Bayes theorem reads as
+Bayes' theorem. For $N ~ 1000-3000$ datum all of the independent of each other
+and $n \ll 10^3$ parameters Bayes' theorem reads as
 \begin{equation}
@@
 \subsubsection{Expectations and marginalizations}
-One of the great powers of Bayesian analysis is \textit{marginalization}.
-For models with more than one or two parameters, it is generally unwise
-to attempt to draw inferences directly from the full joint posterior
-distribution $P(\vec{\theta} \mid D)$...
+For models with more than one or two parameters, it is generally difficult
+to visualize or reason about the full joint posterior distribution
+$P(\vec{\theta} \mid D)$ directly.
+One of the great powers of Bayesian analysis is \textit{marginalization},
+allowing us to reduce the dimensionality to only the parameters of
+immediate interest by averaging over the other dimensions.
+Formally, for a three dimensional model with parameters
+$\theta_1$, $\theta_2$, and $\theta_3$, we can for instance
+marginalize away $\theta_3$ to produce a 2D posterior as
+\begin{equation}
+P(\theta_1, \theta_2 \mid D) \propto
+        \int_{\theta_3} d\theta_3 \,P(\theta_1, \theta_2, \theta_3 \mid D),
+\end{equation}
+or we can marginalize away $\theta_1$ and $\theta_3$ to produce the
+1D marginal posterior of $\theta_2$ alone, which would be
+\begin{equation}
+P(\theta_2 \mid D) \propto
+        \int_{\theta_1} d\theta_1 \int_{\theta_3} d\theta_3
+        \,P(\theta_1, \theta_2, \theta_3 \mid D).
+\end{equation}
+Conceptually, this is what we did in generating the 2D slices of the
+full 9D model in Figure~\ref{fig4:repressed_post_full}(A).
+In practice, this marginalization is even easier with Markov Chain Monte Carlo
+samples in hand. Since each point is simply a list of parameter values,
+we simply ignore the parameters which we want to marginalize
+away~\cite{Gelman2013}.
 
-\mmnote{
-\begin{itemize}
-% \item Bayes theorem.
-% Elaborate on posterior as plausibility over
-% space of model parameters.
-% \item Choosing likelihoods.
-% Easy here, just the steady-state distributions of our master
-% equations, assuming iid single-cell measurements.
-% \item Choosing priors.
-% Not critical here because we are data-rich and nonhierarchical,
-% as long as we don't exclude ground truth. In other applications,
-% prior can matter a lot. This is a feature, not a bug: it's
-% revealing hidden structure that was implied by your modeling
-% assumptions and you just didn't realize it. Predictive checks to
-% verify you're not excluding any plausible values.
-\item Marginalization/expectations.
-Analytically hard, trivial with sampling.
-\item MCMC.
-Lot's of fun and deep analogies to stat mech and
-Hamiltonian mechanics we'll allude to but not explore here.
-The idea of MCMC: in a perfect world, you'd
-wave a magic wand and generate independent draws from target
-posterior. But we don't know how. Next best: generate a Markov
-chain of $N$ \textit{correlated} samples, estimate an
-autocorrelation time $\tau$, then you've got roughly $N/\tau$
-effective samples. Obviously with some assumptions.
-\item Posterior predictive checks.
-So the posterior is fine, but is the model actually not wrong?
-PPC is essential to verify that your model is
-not blatantly contradicted by the data.
-\end{itemize}
-}
+\subsubsection{Markov Chain Monte Carlo}
+The theory and practice of Bayesian inference with Markov Chain Monte Carlo
+(MCMC) is a rich subject with fascinating and deep analogies to statistical
+mechanics, even drawing on classical Hamiltonian mechanics and general
+relativity in its modern incarnations. We refer the interested reader
+to~\cite{Gelman2013} and~\cite{Betancourt2018} for excellent introductions.
+Here we merely give a brief summary of
+the MCMC computations carried out in this work.
+
+We used the Python package \texttt{emcee} for most of the MCMC sampling
+in this work. For the constitutive promoter inference, we also ran
+sampling with the excellent Stan modeling language as a check. We did not
+use Stan for the inference of the simple repression model because
+implementing the gradients of the hypergeometric function ${_2F_1}$
+appearing in Eq.~\ref{eq:p_m_bursty+rep_appdx}, the probability
+distribution for our bursty model with repression, would have been
+an immensely challenging task. \texttt{emcee} was more than adequate for
+our purposes, and we were perhaps lucky that the 9-D posterior model
+for the model of simple repression with bursty promoter was quite well
+behaved and did not require the extra power of the Hamiltonian Monte
+Carlo algorithm provided by Stan~\cite{Carpenter2017}.
+Source code for all statistical inference will be made available at
+\url{https://github.com/RPGroup-PBoC/bursty_transcription}.
 
@@
 As for the Poisson promoter model, this expression for the steady-state mRNA
-distribution is exactly the likelihood we want to use in Bayes theorem. Again
+distribution is exactly the likelihood we want to use in Bayes' theorem. Again
 denoting the single-cell mRNA count data as $D=\{m_1, m_2,\dots, m_N\}$, here
-Bayes theorem takes the form
+Bayes' theorem takes the form
 \begin{equation}
@@
 \end{split}
-\label{eq:p_m_bursty+rep}
+\label{eq:p_m_bursty+rep_infreprint}
 \end{equation}
@@
 \end{equation}
-where the likelihood is specified by Eq.~\ref{eq:p_m_bursty+rep}. We ran MCMC
-sampling on the full nine dimensional posterior specified by this generative
-model.
+where the likelihood is specified by Eq.~\ref{eq:p_m_bursty+rep_infreprint}.
+We ran MCMC sampling on the full nine dimensional posterior specified
+by this generative model.
 
@@
 so no inference was possible.
-
-\mmnote{Outline:
-Perhaps urprising that so many model parameters can be inferred from just a few
-distributions that, by eye, don't have strong features! Does require some
-assumptions, e.g., about equivalence of rates across experiments, but that seems
-quite reasonable.
-}

### 2020-06-06 doc/section_01_introduction.tex
Implemented most of GC's edits plus more of my own

@@
 transcription factors and RNA polymerases on the DNA. This challenge stems first
-and foremost from our ignorance about what those architectures even are with
+and foremost from our ignorance about what those architectures even are, with
 more than 60\% of the genes even in an ostensibly well understood organism such
-as {\it E. coli} with no regulatory insights at all~\cite{Rydenfelt2014-2,
+as {\it E. coli} having no regulatory insights at all~\cite{Rydenfelt2014-2,
 Belliveau2018, Ghatak2019, Santos_Zavaleta2019}. But even once we have
@@
 pushed~\cite{Garcia2011, Phillips2019}. If we demonstrate that we can pass that
-test by successfully predicting both the means and variance in gene expression,
+test by successfully predicting both the means and variance in gene expression
+at the mRNA level,
 then that provides a more solid foundation upon which to launch into more
-complex problems.
+complex problems - for instance, some of the previously unknown architectures
+uncovered in~\cite{Belliveau2018} and~\cite{Ireland2020}.
 
@@
 this architecture that have been elucidated by generations of geneticists,
-molecular biologists and biochemists. Our aim is to understand the underlying
-principles of such coarse-graining in order to increase the predictive power of
-these models. In doing so, we hope future models of regulatory response will be
+molecular biologists and biochemists.
+One goal in exploring such coarse-grainings is to build towards the
+future models of regulatory response that will be
 able to serve the powerful predictive role needed to take synthetic biology from
@@
 at the molecular level propagate to input-output functions at the genetic
-circuit level. We want concepts, not mere facts. In particular a key question
+circuit level. In particular a key question
 is: at this level of coarse-graining, what microscopic details do we need to

### 2020-06-06 doc/section_02_means.tex
Implemented most of GC's edits plus more of my own

@@
 repressor ($R > 0$) on the gene expression level compared to an unregulated
-promoter ($R = 0$). The second equality in~\ref{eq:fc_def} follows from assuming
+promoter ($R = 0$). The second equality in Eq.~\ref{eq:fc_def} follows from assuming
 that the translation efficiency, i.e., the number of proteins translated per
@@
 \end{equation}
-which clearly matches the form of the master curve in
+which matches the form of the master curve in
 Figure~\ref{fig1:means_cartoons}(D) with $\rho=1$ and $\Delta F_R = 
@@
 generalized to incorporate small-molecule inducers that bind the repressor. 
-% Interestingly, the effect of this generalization is
-% simply to renormalize $R$ from the total number of repressors to a smaller
-% effective number, roughly (though not exactly) the number which are unbound by
-% inducers.
+The effect of this generalization is, roughly speaking,
+simply to rescale $R$ from the total number of repressors to a smaller
+effective number of available repressors which are unbound by inducers.
+We point out that the same generalization can be incorporated quite easily
+into any of our models in Figure~\ref{fig1:means_cartoons} by simply
+rescaling the repressor copy number $R$ in the equilibrium models, or
+equivalently $k_R^+$ in the nonequilibrium models.
 
-The mean mRNA, as derived in the appendix from a similar enumeration of states
-and weights as the previous model, is
+The mean mRNA copy number, as derived in Appendix~\ref{sec:non_bursty}
+from a similar enumeration of states and weights as the previous model, is
 \begin{equation}
@@
 \end{equation}
-but obviously with a higher-dimensional state space and different matrices. The
+but with a higher-dimensional state space and different matrices. The
 state vector and promoter transition matrix are now
@@
 increasing dimension by one, the only new feature in $\mathbf{K}$ is the
-rate $k_O$, representing the (irreversible) rate of open complex formation from
-the closed complex. The initiation matrices are given by
+rate $k_O$, representing the rate of open complex formation from
+the closed complex, which we assume for simplicity to be irreversible
+in keeping with some~\cite{Mitarai2015} but not all~\cite{DeHaseth1998}
+past literature. The initiation matrices are given by
 \begin{equation}
@@
 promoter and modeled initiation of a transcript as triggering a promoter state
-change, making the hopefully uncontroversial assumption that an RNAP can only
+change, making the assumption that an RNAP can only
 make one transcript at a time. In contrast, in the present model the promoter
@@
 The burst sizes are assumed to be geometrically distributed with
-a mean burst size $b$.
+a mean burst size $b$, which we will motivate in Section~\ref{sec:beyond_means}
+when we derive this model as a certain limiting case of model 4.
 
@@
 
-\subsubsection{Summary of Results across models for Fold-Changes in Mean
+\subsection{Discussion of Results Across Models for Fold-Changes in Mean
 Expression}
-What have we learned from our exploration of the models
-in~\fig{fig1:means_cartoons} and their predictions of fold-changes in mean
-expression levels? The key outcome of our analysis is the existence of a master
-curve shown in \fig{fig1:means_cartoons}(D) to which the fold-change predictions
+The key outcome of our analysis of the models in
+Figure~\ref{fig1:means_cartoons} is the existence of a master curve shown
+in Figure~\ref{fig1:means_cartoons}(D) to which the fold-change predictions
 of all the models collapse. This master curve is parametrized by only two
@@
 
-Why does the master curve of~\fig{fig1:means_cartoons}(D) exist at all?
+Why does the master curve of Figure~\ref{fig1:means_cartoons}(D) exist at all?
 This brings to mind the deep questions posed in,
@@
 describing the repressor, $\Delta F_R$, to cleanly separate from all other
-details of the promoter architecture, which are encapsulated in $\rho$. In the
-graph-theoretic language which was applied to similar problems
-in~\cite{Gunawardena2012} and~\cite{Ahsendorf2014}, the repressor-bound state is
-a leaf of the graph describing transitions between promoter states. Arbitrary
-nonequilibrium behavior can occur on the rest of the promoter state space, but
-it may all be swept up in the effective parameter $\rho$, to which the repressor
-makes no contribution.
+details of the promoter architecture, which are encapsulated in $\rho$.
+Arbitrary nonequilibrium behavior can occur on the rest of the
+promoter state space, but it may all be swept up in the effective
+parameter $\rho$, to which the repressor makes no contribution.
+We point the interested reader to~\cite{Gunawardena2012}
+and~\cite{Ahsendorf2014} for an interesting analysis of similar problems
+using a graph-theoretic language.
 
@@
 the resulting collection the label $\Delta F_R$, it is unclear if the physical
-interpretation as an effective free energy makes sense. As is well known, free
+interpretation as an effective free energy makes sense. The problem is that free
 energies cannot be assigned unambiguously to states out of equilibrium because
@@
 and/or supercoiling accumulation and release~\cite{Chong2014, Sevier2016},
-and/or, RNAP clusters analogous to those obvserved in
+and/or, RNAP clusters analogous to those observed in
 eukaryotes~\cite{Cisse2013, Cho2016} and recently also observed in
@@
 One might instead imagine a bottom-up modeling approach, where we mathematicize
-a model of what we hypothesize the important steps are and are not, use in vitro
+a model of what we hypothesize the important steps are and are not, use \textit{in vitro}
 data for the steps deemed important, and \textit{predict} what $\rho$ should be.
@@
 different models will likely make indistinguishable predictions, and without any
-way to experimentally perturb in vivo, there is no clear way to test whether the
-modeling assumptions are correct.
+way to experimentally perturb \textit{in vivo}, there is no clear way
+to test whether the modeling assumptions are correct.
 
@@
 
+Another possible route to dissect transcription details without a global
+perturbation would be to use phage polymerase with phage-specific promoters.
+While such results would carry some caveats, e.g., whether the repression of
+the phage polymerase is a good analog to the repression of the native RNAP,
+it could nevertheless be worthy of consideration.
+
 We have already pointed out that the master curve of~\fig{fig1:means_cartoons}
@@
 meaning therefore $\rho\approx1$.  This approximation can be well justified on
-the basis of the number of RNAP per cell and the strength of binding of RNAP to
+the basis of the number of RNAP and $\sigma$ factors per cell
+and the strength of binding of RNAP to
 DNA at weak promoters. This is suggestive, but how can we be sure that $\rho$ is
@@
 We would also like to highlight the relevance of~\cite{Landman2019} here.
-Landman et.\ al.\ reanalyzed and compared in vivo and in vitro data on the lacI
-repressor's binding affinity to its various operator sequences. (The in vivo
+Landman et.\ al.\ reanalyzed and compared \textit{in vivo} and
+\textit{in vitro} data on the lacI repressor's binding affinity to its
+various operator sequences. (The \textit{in vivo}
 data was from, essentially, fitting our master curve to expression
-measurements.) They find broad agreement between the in vitro and in vivo
+measurements.) They find broad agreement between the
+\textit{in vitro} and \textit{in vivo}
 values. This reinforces the suspicion that the equilibrium $\Delta\varepsilon_R$

### 2020-06-06 doc/section_03_beyond_means.tex
Implemented most of GC's edits plus more of my own

@@
 In~\cite{Jones2014}, the authors carefully attribute measured deviations from
-Fano = 1 to intensity variability in fluorescent spots, gene copy number
+Fano = 1 to intensity variability in fluorescence measurements, gene copy number
 variation, and copy number fluctuations of the transcription machinery, e.g.,
@@
 exponential distribution. The former is a convolution of multiple exponentials,
-and intuitively should be more peaked with a smaller fractional width than an
+and intuitively the waiting time distribution for a multistep process
+should be more peaked with a smaller fractional width than a single
 exponential with the same mean.
@@
 which is exactly the geometric distribution with standard parameter
-$\theta\equiv k^-/(k^- + r)$ and domain $n \in \{0, 1, 2, \dots\}$ (this is one
-of two common conventions for the geometric distribution). The mean of the
-geometric distribution, with this convention, is
+$\theta\equiv k^-/(k^- + r)$ and domain $n \in \{0, 1, 2, \dots\}$.
+The mean of the geometric distribution, with this convention, is
 \begin{align}

### 2020-06-06 doc/section_04_bayesian_inference.tex
Implemented most of GC's edits plus more of my own

@@
 problem is then to find the probability distribution for the parameter value
-$\lambda$ given the experimental data. By Bayes theorem this can be written as
+$\lambda$ given the experimental data. By Bayes' theorem this can be written as
 \begin{equation}
@@
 theorem. Again denoting the single-cell mRNA count data as $D=\{m_1, m_2,\dots,
-m_N\}$, here Bayes theorem takes the form
+m_N\}$, here Bayes' theorem takes the form
 \begin{equation}
@@
 commonly the case for parameters for which our previous knowledge based on our
-domain expertise spans several orders of magnitude. For this we chose log-Normal
+domain expertise spans several orders of magnitude. For this we chose log-normal
 distributions for both $k_i$ and $b$. Details on the mean and variance of these
@@
 predictive ECDF, indicating this model is actually able to generate the observed
-data and increasing our confidence that this model is at least not wrong.
+data and increasing our confidence that this model
+is sufficient to parametrize the physical reality of the system.
 
-\textit{lacUV5} is our primary target here, since it forms the core of all the
+The commonly used promoter sequence \textit{lacUV5} is our primary
+target here, since it forms the core of all the
 simple repression constructs of~\cite{Jones2014} that we consider in
@@
 a range of $\lesssim10^1$ (and with the exception of promoter 6, just a span of
-3-5x). This suggests that $k_i$, not $b$, is the key dynamic variable that
+3 to 5-fold). This suggests that $k_i$, not $b$, is the key dynamic variable that
 promoter sequence tunes.
@@
 While the associations represented by Eq.~\ref{eq:bursty_equil_corresp1} and
-Eq.~\ref{eq:bursty_equil_corresp2} appear to be born out by the data in
+Eq.~\ref{eq:bursty_equil_corresp2} appear to be borne out by the data in
 Figure~\ref{fig:constit_post_full}, we do not find the association of parameters
@@
 for each aTc induction concentration (corresponding to an unknown mean repressor
-copy number). (B, upper panel) Ratios of our inferred unbinding rates are
+copy number).
+(B, upper) Ratios of our inferred unbinding rates are
 compared with operator binding energy differences measured by Garcia and
@@
 al.~\cite{Razo-Mejia2018} (circles). Blue glyphs compare O2-O1, while orange
-compare O1-Oid. (B, lower panel) Unbinding rates for O1 (cyan) and Oid (red)
-inferred in this work are compared with single-molecule measurements of the same
-from Hammar et.\ at.\ 2014. The comparison requires an assumption about mRNA
-lifetime $\gamma^{-1}$, which was not measured, and we plot representative
-choices assuming $\gamma^{-1}=3$~min (triangles) or $\gamma^{-1}=5$~min
-(squares). (C) Theory-experiment comparison are shown for each of the datasets
+compare O1-Oid. Points with perfect agreement would lie on the dotted line.
+(B, lower) Unbinding rates for O1 (cyan) and Oid (red)
+inferred in this work are compared with single-molecule measurements
+from Hammar et.\ al.~\cite{Hammar2014}. We plot the comparison assuming illustrative mRNA
+lifetimes of $\gamma^{-1}=3$~min (triangles) or $\gamma^{-1}=5$~min
+(squares). Dotted line is as in upper panel.
+(C) Theory-experiment comparison are shown for each of the datasets
 used in the inference of the model in (A). Observed single-molecule mRNA counts
-data from~\cite{Jones2014} is plotted as black lines. The median of the randomly
+data from~\cite{Jones2014} are plotted as black lines. The median of the randomly
 generated samples for each condition is plotted as a dark colored line. Lighter
 colored bands enclose 95\% of all samples for a given operator/repressor copy
-number pair. Samples are sorted by operator and plotted separately for visual
-clarity. The unregulated promoter, UV5, is shown with each as a reference
-point.}
+number pair.
+%Samples are sorted by operator and plotted separately for visual clarity.
+The unregulated promoter, \textit{lacUV5}, is shown with each as a reference.}
 \label{fig4:repressed_post_full}
@@
 assume that a given concentration of the inducer resulted in a specific mean
-repressor copy number\cite{Chure2019a}. Also note that the authors
+repressor copy number~\cite{Chure2019a}. Also note that the authors
 of~\cite{Jones2014} report estimates of LacI copy number per cell rather than
@@
 
-Having stated the problem, Bayes theorem reads simply
+Having stated the problem, Bayes' theorem reads
 \begin{equation}
@@
 Next we specify priors. As for the constitutive model, weakly informative
-lognormal priors are a natural choice for all our rates. We found that if the
+log-normal priors are a natural choice for all our rates. We found that if the
 priors were too weak, our MCMC sampler would often become stuck in regions of
@@
 simultaneously verifying that the marginal posteriors for each parameter were
-not artificially constrained or distorted by the presence of the prior. The only
-exception to this is the highly informative priors we placed on $k_i$ and $b$,
-since we have strong knowledge of them from our inference of constitutive
-promoters above. All details for our prior distributions are listed in 
+not artificially constrained or distorted by the presence of the prior.
+% The only
+% exception to this is the highly informative priors we placed on $k_i$ and $b$,
+% since we have strong knowledge of them from our inference of constitutive
+% promoters above.
+All details for our prior distributions are listed in 
 Appendix~\ref{sec:bayesian}.
@@
 uncertainties as well, as detailed in Appendix~\ref{sec:bayesian}.
-\mmnote{Explain in appendix how marginalization with sampling makes this
-``trivial'' to compute from samples of the joint posterior, b/c sampling makes
-marginalization trivial.} We then compare these log ratios of rates to the
+% \mmnote{Explain in appendix how marginalization with sampling makes this
+% ``trivial'' to compute from samples of the joint posterior, b/c sampling makes
+% marginalization trivial.}
+We then compare these log ratios of rates to the
 binding energy differences $\Delta\varepsilon_{O1} - \Delta\varepsilon_{Oid}$
@@
 values from both~\cite{Garcia2011a} and~\cite{Razo-Mejia2018}. Three of the four
-values are within $\sim0.5 k_BT$ of the diagonal representing perfect agreement,
-which is comparable to the $\sim\text{few}\times0.1 k_BT$ variability between
+values are within $\sim0.5~k_BT$ of the diagonal representing perfect agreement,
+which is comparable to the $\sim\text{few}\times0.1~k_BT$ variability between
 the independent determinations of the same quantities between~\cite{Garcia2011a}
@@
 burst duration and repressor association rate. Another possibility is that
-the very large number of zero mRNA counts for Oid, 2~ng/mL is skewing it's partner
+the very large number of zero mRNA counts for Oid, 2~ng/mL is skewing its partner
 datasets through the shared association rate.

### 2020-06-06 doc/section_05_discussion.tex
Implemented most of GC's edits plus more of my own

@@
 variability, and the full gene expression distribution found in such a population
-as would be reported in a single molecule mRNA Fluorescence in situ
+as would be reported in a single molecule mRNA Fluorescence \textit{in situ}
 Hybridization experiment, for example.
@@
 operons?
-
-It seems clear therefore that chromosomal locus plays a major role in mean
-expression levels as well as noise. What features of different loci are and are
-not important? Sigma factors could certainly play a part, but it seems many
-other mechanisms are at play: supercoiling, and more generally 3D organization
-of the chromosome, read-through of neighboring genes, and surely some
-fascinating unknown unknowns. Could our preferred coarse-grained model capture
-the variability across different loci? If so, and we were to repeat the
-parameter inference as done in this work, is there a simple theoretical model we
-could build to understand the resulting parameters?
+What features of different loci are and are not important?
+Could our preferred coarse-grained model capture the variability across
+different loci? If so, and we were to repeat the parameter inference
+as done in this work, is there a simple theoretical model we could build
+to understand the resulting parameters?
 
@@
 dissection of regulatory response in bacteria, with similar beautiful work
-emerging in examples such as the arabinose and gal operons as
+emerging in examples such as the \textit{ara} and \textit{gal} operons as
 well~\cite{Dunn1984b, Oehler1990, Weickert1993, Oehler1994, Schleif2000,

### 2020-06-14 doc/section_07_acknowledgments.tex
add missing sections

% !TEX root = ./busty_transcription.tex
\section*{Acknowledgments}

We thank Rob Brewster for providing the raw single-molecule mRNA FISH data. We
thank Justin Bois for his key support with the Bayesian inference section. We
would also like to thank Griffin Chure for invaluable feedback on the
manuscript. This material is based upon work supported by the National Science
Foundation Graduate Research Fellowship under Grant No. DGE‐1745301. This work
was also supported by La Fondation Pierre-Gilles de Gennes, the Rosen Center at
Caltech, and the NIH 1R35 GM118043 (MIRA). M.R.M. was supported by the Caldwell
CEMI fellowship.

### 2020-09-08 doc/resubmission/abstract.tex
Update to the introduction

--- 2020-06-06 doc/abstract.tex
@@
 and inhibiting transcription. In this paper, we show how seven distinct models
-of this so-called simple-repression motif, based both on equilibrium and kinetic
-thinking, can be used to derive the predicted levels of gene expression and
-shed light on the often surprising past success of the equilbrium models. These
-different models are then invoked to confront a variety of different data on mean,
-variance and full gene expression distributions, illustrating the extent to which
-such models can and cannot be distinguished, and suggesting a two-state model with a
-distribution of burst sizes as the most potent of the seven for describing the
-simple-repression motif.
+of this so-called simple-repression motif, based both on thermodynamic and
+kinetic thinking, can be used to derive the predicted levels of gene expression
+and shed light on the often surprising past success of the thermodynamic models.
+These different models are then invoked to confront a variety of different data
+on mean, variance and full gene expression distributions, illustrating the
+extent to which such models can and cannot be distinguished, and suggesting a
+two-state model with a distribution of burst sizes as the most potent of the
+seven for describing the simple-repression motif.
 \end{abstract}

### 2020-09-08 doc/resubmission/section_01_introduction.tex
Update to the introduction

--- 2020-06-06 doc/section_01_introduction.tex
@@
 a result of spatiotemporal cues that integrate chemical, mechanical and other
-types of signals. The original repressor-operator model conceived by Jacob and
-Monod in the context of bacterial metabolism has now been transformed into the
-much broader subject of gene regulatory networks in living organisms of all
-kinds~\cite{Jacob1961, Britten1969, Ben-TabouDe-Leon2007}. One of the remaining
+types of signals. As our ability to experimentally observe and measure the
+dynamical processes that constitute the central dogma improves, there is an
+opportunity to undertake a theory-experiment dialogue in order to sharpen our
+understanding of such a fundamental biological process. One of the remaining
 outstanding challenges to have emerged in the genomic era is our continued
@@
 Gertz2009, Sherman2012, Saiz2013, Ko1991, Peccoud1995, Record1996, Kepler2001,
-Sanchez2008, Shahrezaei2008, Sanchez2011, Michel2010}. One route to such
-predictive understanding is to focus on the simplest regulatory architecture and
-to push the theory-experiment dialogue as far and as hard as it can be
-pushed~\cite{Garcia2011, Phillips2019}. If we demonstrate that we can pass that
-test by successfully predicting both the means and variance in gene expression
-at the mRNA level,
-then that provides a more solid foundation upon which to launch into more
-complex problems - for instance, some of the previously unknown architectures
-uncovered in~\cite{Belliveau2018} and~\cite{Ireland2020}.
+Sanchez2008, Shahrezaei2008, Sanchez2011, Michel2010, Iyer-Biswas2009}. One
+route to such predictive understanding is to focus on the simplest regulatory
+architecture and to push the theory-experiment dialogue to increase the
+predictive power of our theoretical models~\cite{Garcia2011, Phillips2019}. If
+we demonstrate that we can pass that test by successfully predicting both the
+means and variance in gene expression at the mRNA level, then that provides a
+more solid foundation upon which to launch into more complex problems - for
+instance, some of the previously unknown architectures uncovered
+in~\cite{Belliveau2018} and~\cite{Ireland2020}.
 
@@
 binding site as developed in pioneering early work on the quantitative
-dissection of transcription \cite{Oehler1994, Oehler1990}. All of the proposed
-models coarse-grain away some of the important microscopic features of
-this architecture that have been elucidated by generations of geneticists,
-molecular biologists and biochemists.
-One goal in exploring such coarse-grainings is to build towards the
-future models of regulatory response that will be
-able to serve the powerful predictive role needed to take synthetic biology from
-a brilliant exercise in enlightened empiricism to a rational design framework as
-in any other branch of engineering. More precisely, we want phenomenology in the
-sense of coarse-graining away atomistic detail, but still retaining biophysical
-meaning. For example, we are not satisfied with the strictly phenomenological
-approach offered by the commonly used Hill functions. As argued
-in~\cite{Frank2013}, Hill functions are ubiquitous precisely because they
-coarse-grain away all biophysical details into inscrutable parameters. Studies
-like~\cite{Razo-Mejia2018} have demonstrated that Hill functions are clearly
-insufficient since each new situation requires a completely new set of
-parameters. Such work requires a quantitative theory of how biophysical changes
-at the molecular level propagate to input-output functions at the genetic
-circuit level. In particular a key question
-is: at this level of coarse-graining, what microscopic details do we need to
-explicitly model, and how do we figure that out? For example, do we need to
-worry about all or even any of the steps that individual RNA polymerases go
-through each time they make a transcript? Turning the question around, can we
-see any imprint of those processes in the available data? If the answer is no,
-then those processes are irrelevant for our purposes. Forward modeling and
-inverse (statistical inferential) modeling are necessary to tackle such
-questions.
+dissection of transcription \cite{Oehler1994, Oehler1990}. One of the main
+features of of the models we explore is that, by construction, all features
+related to the microstates in which the repressor is bound to the promoter can
+be separated from the microstates in which the RNA polymerase (RNAP) is bound.
+From a modeling perspective, this means that some of the models can be written
+as effective two-state models for which there is a rich
+literature~\cite{Peccoud1995,Bintu2005a,Shahrezaei2008,Iyer-Biswas2009,Sanchez2013,Jones2014}.
+Here, we systematically compare the predictions of several models with different
+levels of coarse graining written in terms of thermodynamic and kinetic
+parameters. One goal in exploring such coarse-grainings is to build towards the
+future models of regulatory response that will be able to serve the powerful
+predictive role needed to take synthetic biology from a brilliant exercise in
+enlightened empiricism to a rational design framework as in any other branch of
+engineering. More precisely, we want phenomenology in the sense of
+coarse-graining away atomistic detail, but still retaining biophysical meaning.
+In particular a key question is: at this level of coarse-graining, what
+microscopic details do we need to explicitly model, and how do we figure that
+out? For example, do we need to worry about all or even any of the steps that
+individual RNA polymerases go through each time they make a transcript? Turning
+the question around, can we see any imprint of those processes in the available
+data? If the answer is no, then those processes are irrelevant for our purposes.
+Forward modeling and inverse (statistical inferential) modeling are necessary to
+tackle such questions. We combine both approaches in order to discard models 
+that cannot empirically satisfy the main features of experimental data. First we
+apply forward modeling to demonstrate that none of the models are 
+distinguishable at the level of mean gene expression. We then extend the 
+modeling to look at higher moments of the distribution, eliminating models that
+do not empirically satisfy the observed cell-to-cell variability. Finally we 
+arrive to a minimal model on which we can apply inverse modeling in order to 
+infer the parameters that explain the data.
 
 Figure~\ref{fig1:means_cartoons}(A) shows the qualitative picture of simple
-repression that is implicit in the repressor-operator model. An operator, the
-binding site on the DNA for a repressor protein, may be found occupied by a
+repression that is implicit in the repressor-operator model. An operator, i.e.,
+the binding site on the DNA for a repressor protein, may be found occupied by a
 repressor, in which case transcription is blocked from occurring. Alternatively,
@@
 \centering
-\includegraphics[width=\textwidth]{../figures/main/fig01.pdf}
+\includegraphics[width=\textwidth]{../../figures/main/fig01.pdf}
 \caption{\textbf{An overview of the simple repression motif at the level of
@@
 The logic of the remainder of the paper is as follows. In
-section~\ref{section_02_means}, we show how both thermodynamic models and
-kinetic models based upon the chemical master equation all culminate in the same
+section~\ref{section_02_means}, we show how both thermodynamic models
+(Figure~\ref{fig1:means_cartoons}(B)) and kinetic models based upon the chemical
+master equation (Figure~\ref{fig1:means_cartoons}(C)) all culminate in the same
 underlying functional form for the fold-change in the average level of gene
@@
 the question of how to discriminate between the different models.
-
-%\mmnote{Key ideas, no particular order, that I haven't written down before:
-%\begin{itemize}
-%\item Our goal is to build phenomenological models of input-output functions of
-%genetic circuits. More precisely, we want phenomenology in the sense of
-%coarse-graining away atomistic detail, but still retaining biophysical meaning,
-%e.g., we don't want to coarse-grain as far as Hill functions. Why not? We are
-%motivated by studies like~\cite{Razo-Mejia2020}, for which Hill functions are
-%insufficient. Such work requires a quantitative theory of how biophysical
-%changes at the molecular level propagate to input-output functions at the
-%genetic circuit level. We want concepts, not mere facts. In particular a key
-%question is: at this level of coarse graining, what microscopic details do I
-%need to explicitly model, and how do we figure that out? For example, do I need
-%to worry about all or even any of the steps that individual RNAPs go through
-%each time they make a transcript? Turning the question around, can we see any
-%imprint of those processes in the available data? If the answer is no, then
-%those processes are irrelevant for our purposes. Forward modeling and inverse
-%(statistical inferential) modeling can complement each other beautifully here.
-%\end{itemize}}
-
-%Biochemistry tells us of a progression of steps, but theory tells us
-%that coarse graining can be exact

### 2020-09-08 doc/resubmission/appendix_01_nonbursty.tex
first attempt to shrink section 02. Sending to appendix 01 in the mean time'

--- 2020-06-06 doc/appendix_01_nonbursty.tex
@@
 
+\mrm{THIS NEEDS TO BE FIXED}
+\subsubsection{The Two-State Equilibrium Model}
+In this simplest model, depicted as (1) in Figure~\ref{fig1:means_cartoons}(B),
+the promoter is idealized as existing in one of two states, either repressor
+bound or repressor unbound. The rate of transcription
+is assumed to be proportional to the fraction of time spent
+in the repressor unbound state.
+From the relative statistical weights listed in
+Figure~\ref{fig1:means_cartoons}, the probability $p_U$ of being in the
+unbound state is
+\begin{equation}
+p_U = \left(1 + \frac{R}{N_{NS}} e^{-\beta\Delta\varepsilon_R}\right)^{-1}.
+\end{equation}
+The mean rate of transcription is then given by $r p_U$, as assumed by
+Eq.~\ref{eq:transcrip_prop_pbound}. The mean number of mRNA is
+set by the balance of average mRNA transcription and degradation rates,
+so it follows that the mean mRNA level is given by
+\begin{equation}
+\langle m \rangle = \frac{r}{\gamma}
+        \left(1 + \frac{R}{N_{NS}} e^{-\beta\Delta\varepsilon_R}\right)^{-1},
+\end{equation}
+where $r$ is the transcription rate from the repressor unbound state, $\gamma$
+is the mRNA degradation rate, $R$ is repressor copy number, $N_{NS}$ is the
+number of nonspecific binding sites in the genome where repressors spend most
+of their time when not bound to the operator, $\beta \equiv 1/k_BT$, and
+$\Delta\varepsilon_R$ is the binding energy of a repressor to its operator site.
+The derivation of this result is deferred to Appendix~\ref{sec:non_bursty}.
+
+The fold-change is  found as the ratio of mean mRNA with and without repressor
+as introduced in Eq.~\ref{eq:fc_def}. Invoking that definition results in
+\begin{equation}
+\text{fold-change}
+= \left(1 + \frac{R}{N_{NS}} e^{-\beta\Delta\varepsilon_R}\right)^{-1},
+\end{equation}
+which matches the form of the master curve in
+Figure~\ref{fig1:means_cartoons}(D) with $\rho=1$ and $\Delta F_R = 
+\beta\Delta\varepsilon_r - \log (R / N_{NS})$.
+
+In fact it was noted in~\cite{Chure2019} that this two-state model can be viewed
+as the coarse-graining of any equilibrium promoter model in which no
+transcriptionally active states have transcription factor bound, or put
+differently, when there is no overlap between
+transcription factor bound states and transcriptionally
+active states. We will see this explicitly in the 3-state equilibrium model
+below, but perhaps surprising is that an analogous result carries over even to
+the nonequilibrium models we consider later.
+
+\subsubsection{The Three-State Equilibrium Model}
+Compared to the previous model, here we fine-grain the repressor unbound state
+into two separate states: empty, and RNAP bound as shown in (2) in
+Figure~\ref{fig1:means_cartoons}(B). This picture was used in~\cite{Garcia2011a}
+as we use it here, and in~\cite{Razo-Mejia2018} and~\cite{Chure2019} it was
+generalized to incorporate small-molecule inducers that bind the repressor. 
+The effect of this generalization is, roughly speaking,
+simply to rescale $R$ from the total number of repressors to a smaller
+effective number of available repressors which are unbound by inducers.
+We point out that the same generalization can be incorporated quite easily
+into any of our models in Figure~\ref{fig1:means_cartoons} by simply
+rescaling the repressor copy number $R$ in the equilibrium models, or
+equivalently $k_R^+$ in the nonequilibrium models.
+
+The mean mRNA copy number, as derived in Appendix~\ref{sec:non_bursty}
+from a similar enumeration of states and weights as the previous model, is
+\begin{equation}
+\langle m \rangle = \frac{r}{\gamma}
+\frac{\frac{P}{N_{NS}} e^{-\beta\Delta\varepsilon_P}}
+        {
+        1 + \frac{R}{N_{NS}} e^{-\beta\Delta\varepsilon_R}
+        + \frac{P}{N_{NS}} e^{-\beta\Delta\varepsilon_P}
+        },
+\end{equation}
+where the new variables are $\Delta\varepsilon_P$, the difference in RNAP
+binding energy to its specific site (the promoter) relative to an average
+nonspecific background site, and the RNAP copy number, $P$. The fold-change
+again follows immediately as
+\begin{align}
+\text{fold-change}
+&= \frac{\frac{P}{N_{NS}} e^{-\beta\Delta\varepsilon_P}}
+        {
+        1 + \frac{R}{N_{NS}} e^{-\beta\Delta\varepsilon_R}
+        + \frac{P}{N_{NS}} e^{-\beta\Delta\varepsilon_P}
+        }
+\frac{1 + \frac{P}{N_{NS}} e^{-\beta\Delta\varepsilon_P}}
+        {\frac{P}{N_{NS}} e^{-\beta\Delta\varepsilon_P}}
+\\
+&= \left(
+1 + \frac{\frac{R}{N_{NS}} e^{-\beta\Delta\varepsilon_R}}
+        {1 + \frac{P}{N_{NS}} e^{-\beta\Delta\varepsilon_P}}
+\right)^{-1}
+\\
+&= (1 + \exp(-\Delta F_R - \log\rho))^{-1},
+\end{align}
+with $\Delta F_R = \beta\Delta\varepsilon_R - \log(R/N_{NS})$ and $\rho = 1 +
+\frac{P}{N_{NS}}\mathrm{e}^{-\beta\Delta\varepsilon_P}$ as shown
+in~\fig{fig1:means_cartoons}(B). Thus far, we see that the two thermodynamic
+models, despite making different coarse-graining commitments, result in the same
+functional form for the fold-change in mean gene expression.  We now explore how
+kinetic models fare when faced with computing the same observable.
+\mrm{ENDS HERE}
+
+\mrm{THIS NEEDS TO BE FIXED}
+\subsubsection{The Poisson Promoter Nonequilibrium Model}
+For our first kinetic model, we  imitate the states considered in the Two-State
+Equilibrium Model and consider the simplest possible picture with only two
+states, repressor bound and unbound. This is exactly the model used for the main
+results of~\cite{Jones2014}. In this picture, repressor association and
+dissociation rates from its operator site, $k_R^+$ and $k_R^-$, respectively,
+govern transitions between the two states. When the system is in the unbound
+state, transcription initiates at rate $r$, which represents a coarse-graining
+of all the downstream processes into a single effective rate. mRNA is degraded
+at rate $\gamma$ as already exploited in the previous models.
+
+Let $p_R(m,t)$ denote the joint probability of finding the system in the
+repressor bound state $R$ with $m$ mRNA molecules present at time $t$. Similarly
+define $p_U(m,t)$ for the repressor unbound state $U$. This model is governed
+by coupled master equations giving the time evolution of $p_R(m,t)$ and
+$p_U(m,t)$~\cite{Sanchez2008, Sanchez2011, Phillips2019} which we can write as
+\begin{align}
+\begin{split}
+\deriv{t}p_R(m,t) =& 
+- \overbrace{k_R^- p_R(m,t)}^{R \rightarrow U}
++ \overbrace{k_R^+ p_U(m,t)}^{U \rightarrow R}
++ \overbrace{(m+1)\gamma p_R(m+1,t)}^{m + 1 \rightarrow m}
+- \overbrace{\gamma p_R(m,t)}^{m \rightarrow m - 1}
+\\
+\deriv{t}p_U(m,t) =&\; 
+\overbrace{k_R^- p_R(m,t)}^{R \rightarrow U}
+- \overbrace{k_R^+ p_U(m,t)}^{U \rightarrow R}
++ \overbrace{rp_U(m-1,t)}^{m-1 \rightarrow m}
+- \overbrace{rp_U(m,t)}^{m \rightarrow m + 1}
+\\
+&+ \overbrace{(m+1)\gamma p_U(m+1,t)}^{m + 1 \rightarrow m}
+- \overbrace{\gamma p_U(m,t)}^{m \rightarrow m - 1},
+\label{eq:poisson_promoter_cme}
+\end{split}
+\end{align}
+where each term on the right corresponds to a transition between two states of
+the promoter as indicated by the overbrace label. In each equation, the first
+two terms describe transitions between promoter states due to repressors
+unbinding and binding, respectively. The final two terms describe degradation of
+mRNA, decreasing the copy number by one, and the terms with coefficient $r$
+describe transcription initiation increasing the mRNA copy number by one.
+We direct the reader to Appendix~\ref{sec:cme_from_cartoon} for a careful
+treatment showing how the form of this master equation follows from the
+corresponding cartoon in Figure~\ref{fig1:means_cartoons}.
+
+We can greatly simplify the notation, which will be especially useful for the
+more complicated models yet to come, by re-expressing the master equation in
+vector form~\cite{Phillips2012}. The promoter states are collected into a vector
+and the rate constants are collected into matrices as
+\begin{equation}
+\vec{p}(m) = \begin{pmatrix} p_R(m) \\ p_U(m) \end{pmatrix},\
+\mathbf{K} = \begin{pmatrix} -k_R^- & k_R^+ \\ k_R^- & -k_R^+ \end{pmatrix},\
+\mathbf{R} = \begin{pmatrix} 0 & 0 \\ 0 & r \end{pmatrix},\
+\label{eq:2state_cme_matrices}
+\end{equation}
+so that the master equation may be condensed as
+\begin{equation}
+\deriv{t}\vec{p}(m,t) =
+\left( \mathbf{K} - \mathbf{R} - \gamma m \mathbf{I} \right) \vec{p}(m,t)
+                + \mathbf{R} \vec{p}(m-1,t)
+                + \gamma (m+1) \mathbf{I} \vec{p}(m+1,t),
+\label{eq:2state_rep_cme}
+\end{equation}
+where $\mathbf{I}$ is the identity matrix. Taking steady state by setting time
+derivatives to zero, the mean mRNA can be found to be
+\begin{equation}
+\langle m \rangle = \frac{r}{\gamma}
+        \left(1 + \frac{k_R^+}{k_R^-}\right)^{-1},
+\label{eq:mean_m_model1}
+\end{equation}
+with the algebra details again deferred to Appendix~\ref{sec:non_bursty}. Recall
+$k_R^+$ is proportional to the repressor copy number, so in computing
+fold-change, absence of repressor corresponds to $k_R^+\rightarrow0$. Therefore
+fold-change in this model is simply
+\begin{equation}
+\text{fold-change} = \left(1 + \frac{k_R^+}{k_R^-}\right)^{-1},
+\end{equation}
+again matching the master curve of~\fig{fig1:means_cartoons}(D) with $\rho=1$.
+
+\subsubsection{Nonequilibrium Model Two - RNAP Bound and Unbound States}
+Our second kinetic model depicted in Figure~\ref{fig1:means_cartoons}(C) mirrors
+the second equilibrium model of Figure~\ref{fig1:means_cartoons}(B) by
+fine-graining  the repressor unbound state of nonequilibrium model 1, resolving
+it into an empty promoter state and an RNAP-bound state. Note in this picture,
+in contrast with model 4 below, transcription initiation is accompanied by a
+promoter state change, in keeping with the interpretation as RNAP-bound and
+empty states: if an RNAP successfully escapes the promoter and proceeds to
+elongation of a transcript, clearly it is no longer bound at the promoter.
+Therefore another RNAP must bind before another transcript can be initiated.
+
+The master equation governing this model is analogous
+to Eqs.~\ref{eq:2state_cme_matrices}-\ref{eq:2state_rep_cme} for model 1 above. The
+main subtlety arises since transcription initiation accompanies a promoter state
+change. This can be understood by analogy to $\mathbf{K}$. The off-diagonal and
+diagonal elements of $\mathbf{K}$ correspond to transitions arriving at or
+departing from, respectively, the promoter state of interest. If transcription
+initiation is accompanied by promoter state changes, we must have separate
+matrices for arriving and departing transcription events since the arriving and
+departing transitions have different initial copy numbers of mRNA, unlike for
+$\mathbf{K}$ where they are the same (see Appendix~\ref{sec:non_bursty}). The
+master equation for this model is
+\begin{equation}
+\deriv{t}\vec{p}(m,t) =
+\left( \mathbf{K} - \mathbf{R_D} - \gamma m \mathbf{I} \right) \vec{p}(m,t)
+                + \mathbf{R_A} \vec{p}(m-1,t) +
+                \gamma (m+1) \mathbf{I} \vec{p}(m+1,t),
+\label{eq:3state_rep_cme}
+\end{equation}
+with the state vector and promoter transition matrix defined as
+\begin{equation}
+\vec{p}(m) = \begin{pmatrix} p_R(m) \\ p_E(m) \\ p_P(m) \end{pmatrix},\
+\mathbf{K} = \begin{pmatrix} -k_R^- & k_R^+ & 0 \\
+                        k_R^- & -k_R^+ -k_P^+ & k_P^- \\
+                        0 & k_P^+ & -k_P^- 
+                \end{pmatrix},
+\label{eq:3state_cme_matrices_pt1}
+\end{equation}
+and the initiation matrices given by
+\begin{equation}
+\mathbf{R_A} = \begin{pmatrix}
+                0 & 0 & 0 \\ 
+                0 & 0 & r \\ 
+                0 & 0 & 0
+                \end{pmatrix},\
+\mathbf{R_D} = \begin{pmatrix}
+                0 & 0 & 0 \\ 
+                0 & 0 & 0 \\ 
+                0 & 0 & r
+                \end{pmatrix}.
+\label{eq:3state_cme_matrices_pt2}
+\end{equation}
+The elements of $\vec{p}(m)$ encode the probabilities of having $m$ mRNA present
+along with the promoter having repressor bound ($R$), being empty ($E$), or
+having RNAP bound ($P$), respectively. $\mathbf{R_A}$ describes probability flux
+\textit{arriving} at the state $\vec{p}(m)$ from a state with one fewer mRNA,
+namely $\vec{p}(m-1)$, and $\mathbf{R_D}$ describes probability flux
+\textit{departing} from the state $\vec{p}(m)$ for a state with one more mRNA,
+namely $\vec{p}(m+1)$. $\mathbf{K}$ is closely analogous to model 1.
+
+Mean mRNA at steady state is found analogously to model 1, with the result
+\begin{equation}
+\langle m\rangle = \frac{r}{\gamma}
+        \frac{k_R^- k_P^+}
+        {k_R^- k_P^+ + k_R^- (k_P^- + r) + k_R^+ (k_P^- + r)},
+\label{eq:model2_meanm}
+\end{equation}
+and with details again deferred to Appendix~\ref{sec:non_bursty}.
+Fold-change is again found from the ratio prescribed by Eq.~\ref{eq:fc_def}, from
+which we have
+\begin{align}
+\text{fold-change}
+&=      \frac{k_R^- k_P^+}
+        {k_R^- k_P^+ + k_R^- (k_P^- + r) + k_R^+ (k_P^- + r)}
+        \frac{k_P^+ + k_P^- + r}{k_P^+}
+\\
+&=      \left(1 + \frac{k_R^+}{k_R^-}
+                \frac{k_P^- + r}{k_P^+ + k_P^- + r}
+        \right)^{-1}
+\\
+&=      \left(1 + \frac{k_R^+}{k_R^-}
+        \left(1 + \frac{k_P^+}{k_P^- + r}\right)^{-1}
+        \right)^{-1},
+\end{align}
+which follows the master curve of~\fig{fig1:means_cartoons}(D) with $\rho = 1 +
+k_P^+/(k_P^- + r)$ as claimed.
+
+\subsubsection{Nonequilibrium Model Three - Multistep Transcription Initiation
+and Escape}
+One might reasonably complain that the first two ``nonequilibrium'' models we
+have considered are straw men. Their steady states necessarily satisfy detailed
+balance which is equivalent to thermodynamic equilibrium. Why is this the case?
+At steady state there is by definition no net probability flux in or out of each
+promoter state, but since the promoter states form a linear chain, there is only
+one way in or out of the repressor bound and RNAP bound states, implying each
+edge must actually have a net zero probability flux, which is the definition of
+detailed balance (usually phrased as equality of forward and reverse transition
+fluxes).
+
+Now we consider model 3 in Figure~\ref{fig1:means_cartoons}(C) which allows the
+possibility of true nonequilibrium steady-state fluxes through the promoter
+states. We point out that this model was considered previously
+in~\cite{Mitarai2015} where a comparison was made with model 1 as used
+in~\cite{Jones2014}. The authors of~\cite{Mitarai2015} argued that the
+additional complexity is essential to properly account for the noise in the mRNA
+distribution. We will weigh in on both models later when we consider observables
+beyond fold-change.
+
+The master equation governing this model is identical in form to model 2 above,
+namely
+\begin{equation}
+\deriv{t}\vec{p}(m,t) =
+\left( \mathbf{K} - \mathbf{R_D} - \gamma m \mathbf{I} \right) \vec{p}(m,t)
+                + \mathbf{R_A} \vec{p}(m-1,t) +
+                \gamma (m+1) \mathbf{I} \vec{p}(m+1,t),
+\end{equation}
+but with a higher-dimensional state space and different matrices. The
+state vector and promoter transition matrix are now
+\begin{equation}
+\vec{p}(m) = \begin{pmatrix} p_R(m) \\ p_E(m) \\
+                             p_C(m) \\ p_O(m)\end{pmatrix},\
+\mathbf{K} = \begin{pmatrix} -k_R^- & k_R^+ & 0 & 0\\
+                        k_R^- & -k_R^+ -k_P^+ & k_P^- & 0 \\
+                        0 & k_P^+ & -k_P^- - k_O & 0 \\
+                        0 & 0 & k_O & 0
+                \end{pmatrix},
+\end{equation}
+with the four promoter states, in order, being repressor bound ($R$), empty
+($E$), RNAP closed complex ($C$), and RNAP open complex ($O$). Besides
+increasing dimension by one, the only new feature in $\mathbf{K}$ is the
+rate $k_O$, representing the rate of open complex formation from
+the closed complex, which we assume for simplicity to be irreversible
+in keeping with some~\cite{Mitarai2015} but not all~\cite{DeHaseth1998}
+past literature. The initiation matrices are given by
+\begin{equation}
+\mathbf{R_A} = \begin{pmatrix}
+        0 & 0 & 0 & 0 \\ 
+        0 & 0 & 0 & r \\ 
+        0 & 0 & 0 & 0 \\ 
+        0 & 0 & 0 & 0
+                \end{pmatrix},\
+\mathbf{R_D} = \begin{pmatrix}
+        0 & 0 & 0 & 0 \\ 
+        0 & 0 & 0 & 0 \\ 
+        0 & 0 & 0 & 0 \\ 
+        0 & 0 & 0 & r
+                \end{pmatrix},
+\end{equation}
+again closely analogous to nonequilibrium model 2.
+
+The expression for mean mRNA is substantially more complicated now, as worked
+out in Appendix~\ref{sec:non_bursty} where we find
+\begin{equation}
+\langle m\rangle = \frac{r}{\gamma}
+        \frac{k_R^- k_P^+ k_O}
+        {k_R^- [(k_P^+ (k_O + r) + r(k_P^- + k_O)] + k_R^+ r(k_P^- + k_O)},
+\label{eq:model3_mean_m}
+\end{equation}
+which can be simplified to
+\begin{equation}
+\langle m\rangle
+= \frac{r}{\gamma}
+\frac{\frac{k_P^+ k_O}{r(k_O + k_P^-)}}
+        {1 + \frac{k_P^+ (k_O + r)}{r(k_O + k_P^-)} + \frac{k_R^+}{k_R^-}}.
+\end{equation}
+The strategy is to isolate the terms involving the repressor, so that now the
+fold-change is seen to be simply
+\begin{align}
+\text{fold-change}
+&= \frac{\frac{k_P^+ k_O}{r(k_O + k_P^-)}}
+        {1 + \frac{k_P^+ (k_O + r)}{r(k_O + k_P^-)} + \frac{k_R^+}{k_R^-}}
+        \frac{1 + \frac{k_P^+ (k_O + r)}{r(k_O + k_P^-)}}
+                {\frac{k_P^+ k_O}{r(k_O + k_P^-)}}
+\\
+&= \left(
+        1 + \frac{k_R^+}{k_R^-}
+        \left(1 + \frac{k_P^+ (k_O + r)}{r(k_O + k_P^-)}\right)^{-1}
+\right)^{-1},
+\end{align}
+surprisingly reducing to the master curve of~\fig{fig1:means_cartoons}(D) once
+again, with $\rho = 1 + \frac{k_P^+ (k_O + r)}{r(k_O + k_P^-)}$.
+
+This example hints that an arbitrarily fine-grained model of downstream
+transcription steps may still be collapsed to the form of the master curve for
+the means given in Figure~\ref{fig1:means_cartoons}(D), so long as the repressor
+binding is exclusive with transcriptionally active states. We offer this as a
+conjecture, and we suspect that a careful argument using the King-Altman diagram
+method~\cite{King1956, Hill1966} might furnish a ``proof.'' Our focus here is
+not on full generality but rather to survey an assortment of plausible models
+for simple repression  that have been proposed in the literature.
+
+\subsubsection{Nonequilibrium Model Four - ``Active'' and ``Inactive'' States}
+Model 4 in Figure~\ref{fig1:means_cartoons}(C) is at the core of the theory
+in~\cite{Razo-Mejia2020}. At a glance the cartoon for this model may appear very
+similar to model 2, and mathematically it is, but the interpretation is rather
+different. In model 2, we interpreted the third state literally as an RNAP-bound
+promoter and modeled initiation of a transcript as triggering a promoter state
+change, making the assumption that an RNAP can only
+make one transcript at a time. In contrast, in the present model the promoter
+state does \textit{not} change when a transcript is initiated. So we no longer
+interpret these states as literally RNAP bound and unbound but instead as
+coarse-grained ``active'' and ``inactive'' states, the details of which we leave
+unspecified for now. We will comment more on this model below when we discuss
+Fano factors of models.
+
+Mathematically this model is very similar to models 1 and 2. Like model 1, the
+matrix $R$ describing transcription initiation is diagonal, namely
+\begin{equation}
+\mathbf{R} = \begin{pmatrix}
+                0 & 0 & 0 \\ 
+                0 & 0 & 0 \\ 
+                0 & 0 & r
+        \end{pmatrix}.
+\end{equation}
+The master equation takes verbatim the same form as it did for model 1,
+Eq.~\ref{eq:2state_rep_cme}. Meanwhile the promoter transition
+matrix $\mathbf{K}$ is the same as Eq.~\ref{eq:3state_cme_matrices_pt1}
+from model 2, although we relabel the
+rate constants from $k_P^\pm$ to $k^\pm$ to reiterate that these are not simply
+RNAP binding and unbinding rates.
+
+Carrying out the algebra, the mean mRNA can be found to be
+\begin{equation}
+\langle m\rangle = \frac{r}{\gamma}
+\frac{k_R^- k^+}
+{k_R^- k^+ + k_R^- k^- + k_R^+ k^-},
+\end{equation}
+and the fold-change readily follows,
+\begin{align}
+\text{fold-change}
+&=      \frac{k_R^- k^+}{k_R^- k^+ + k_R^- k^- + k_R^+ k^-}
+        \frac{k_R^- k^+ + k_R^- k^-}{k_R^- k^+}
+\\
+&=      \left(1 + \frac{k_R^+}{k_R^-}
+                \left(1 + \frac{k^+}{k^-}\right)^{-1}
+        \right)^{-1},
+\end{align}
+from which we see $\rho = 1 + k^+/k^-$ as shown in~\fig{fig1:means_cartoons}(C).
+
+\subsubsection{Nonequilibrium Model Five - Bursty Promoter}
+The final model we consider shown in Figure~\ref{fig1:means_cartoons}(C) is an
+intuitive analog to model 1, with just two states, repressor bound or unbound,
+and transition rates between them of $k_R^+$ and $k_R^-$. In model 1, when in
+the unbound state, single mRNA transcripts are produced as a Poisson process
+with some characteristic rate $r$. The current model by contrast produces, at
+some Poisson rate $k_i$, \textit{bursts} of mRNA transcripts.
+The burst sizes are assumed to be geometrically distributed with
+a mean burst size $b$, which we will motivate in Section~\ref{sec:beyond_means}
+when we derive this model as a certain limiting case of model 4.
+
+From this intuitive picture and by analogy to model 1, then, it should be
+plausible that the mean mRNA level is
+\begin{equation}
+\langle m\rangle = \frac{k_i b}{\gamma}
+        \left(1 + \frac{k_R^+}{k_R^-}\right)^{-1},
+\end{equation}
+which will turn out to be correct from a careful calculation. For now, we simply
+note that just like model 1, the fold-change becomes
+\begin{equation}
+\text{fold-change} = \left(1 + \frac{k_R^+}{k_R^-}\right)^{-1}
+\end{equation}
+with $\rho=1$ also like model 1.
+We will also see later how this model emerges as a natural limit of model 4.
+\mrm{ENDS HERE}
+
 \subsection{Derivation of chemical master equation}
@@
 \centering
-\includegraphics{../figures/si/figS0X_two_state.pdf}
+\includegraphics{../../figures/si/figS0X_two_state.pdf}
 \caption{

### 2020-09-08 doc/resubmission/section_02_means.tex
first attempt to shrink section 02. Sending to appendix 01 in the mean time'

% !TEX root = ./busty_transcription.tex
\section{Mean Gene Expression}\label{section_02_means}

As noted in the previous section, there are two broad classes of models in play
for computing the input-output functions of regulatory architectures as shown in
Figure~\ref{fig1:means_cartoons}. In both classes of model, the promoter is
imagined to exist in a discrete set of states of occupancy, with each such state
of occupancy accorded its own rate of transcription -- including no 
transcription for many of these states. The models are probabilistic with each
state assigned some probability and the overall rate of transcription given by 
\begin{equation}
\mbox{average rate of transcription} = \sum_i r_i p_i,
\label{eq:transcrip_prop_pbound}
\end{equation}
where $i$ labels the distinct states, $p_i$ is the probability of the
$i^{\text{th}}$ state, and $r_i$ is the rate of transcription of that state.
Ultimately, the different models differ along several key axes: what states to
consider and how to compute the probabilities of those states.

The first class of models that are the focus of the present section on
predicting the mean level of gene expression, sometimes known as thermodynamic
models, invoke the tools of equilibrium statistical mechanics to compute the
probabilities of the promoter microstates~\cite{Ackers1982, Shea1985,
Buchler2003, Vilar2003a, Vilar2003b, Bintu2005a, Bintu2005c, Gertz2009,
Sherman2012, Saiz2013}. As seen in Figure~\ref{fig1:means_cartoons}(B), even
within the class of thermodynamic models, we can make different commitments
about the underlying microscopic states of the promoter.  Indeed, the list of
options considered here does not at all exhaust the suite of different
microscopic states we can assign to the promoter.

The second class of models that allow us to access the mean gene expression use
chemical master equations to compute the probabilities of the different
microscopic states ~\cite{Ko1991, Peccoud1995, Record1996, Kepler2001,
Sanchez2008, Shahrezaei2008, Sanchez2011, Michel2010}. As seen in
Figure~\ref{fig1:means_cartoons}(C), we consider a host of different
nonequilibrium models, each of which will have its own result for both the mean
(this section) and noise (next section) in gene expression.

\subsection{Fold-changes are indistinguishable across models}
As a first stop on our search for the ``right'' model of simple repression, let
us consider what we can learn from theory and experimental measurements on the
average level of gene expression in a population of cells. One experimental
strategy that has been particularly useful (if incomplete since it misses out on
gene expression dynamics) is to measure the fold-change in mean
expression~\cite{Garcia2011}. The fold-change is defined as
\begin{equation}
\text{fold-change}
= \frac{\langle \text{gene expression with repressor present} \rangle}
        {\langle \text{gene expression with repressor absent} \rangle}
= \frac{\langle m (R > 0) \rangle}{\langle m (R = 0) \rangle}
\label{eq:fc_def}
\end{equation}
where angle brackets $\left\langle \cdot \right\rangle$ denote the average over
a population of cells and mean mRNA $\langle m\rangle$ is viewed as a function
of repressor copy number $R$. What this means is that the fold-change in gene
expression is a relative measurement of the effect of the transcriptional
repressor ($R > 0$) on the gene expression level compared to an unregulated
promoter ($R = 0$). The second equality in Eq.~\ref{eq:fc_def} follows from
assuming that the translation efficiency, i.e., the number of proteins
translated per mRNA, is the same in both conditions. In other words, we assume
that mean protein level is proportional to mean mRNA level, and that the
proportionality constant is the same in both conditions and therefore cancels
out in the ratio. This is reasonable since the cells in the two conditions are
identical except for the presence of the transcription factor, and the model
assumes that the transcription factor has no direct effect on translation.

Fold-change has proven a very convenient observable in past
work~\cite{Garcia2011a, Brewster2014, Razo-Mejia2018, Chure2019}. Part of its
utility in dissecting transcriptional regulation is its ratiometric nature,
which removes many secondary effects that are present when making an absolute
gene expression measurement. Also, by measuring otherwise identical cells with
and without a transcription factor present, any biological noise common to both
conditions can be made to cancel away.

\fig{fig1:means_cartoons} depicts a smorgasbord of mathematicized cartoons for
simple repression using both thermodynamic and kinetic models that have appeared
in previous literature. For each cartoon, we calculate the fold-change in mean
gene expression as predicted by that model, deferring some algebraic details to
Appendix~\ref{sec:non_bursty}. What we will find is that all cartoons collapse
to a single master curve, shown in \fig{fig1:means_cartoons}(D), which contains
just two parameters. We label the parameters $\Delta F_R$, an effective free
energy parametrizing the repressor-DNA interaction, and $\rho$, which subsumes
all details of transcription in the absence of repressors. We will offer some
intuition for why this master curve exists and discuss why at the level of the
mean expression, we are unable to discriminate ``right'' from ``wrong'' cartoons
given only measurements of fold-changes in expression.

\subsubsection{Two- and Three-state Thermodynamic Models}
We begin our analysis with models (1) and (2) in
Figure~\ref{fig1:means_cartoons}(B). In each of these models the promoter
is idealized as existing in a set of discrete states; the difference being
whether or not the RNAP bound state is included or not. Gene expression is then
assumed to be proportional to the probability of the promoter being in either
the empty state (model (1)) or the RNAP-bound state (model (2)). We direct the
reader to Appendix~\ref{sec:non_bursty} for details on the derivation of the
fold-change. For our purposes here, it suffices to state that the functional
form of the fold-change for model (1) is
\begin{equation}
\text{fold-change}
= \left(1 + \frac{R}{N_{NS}} e^{-\beta\Delta\varepsilon_R}\right)^{-1},
\end{equation}
where $R$ is the number of repressors per cell, $N_{NS}$ is the number of
non-specific binding sites where the repressor can bind, $\Delta\varepsilon_R$
is the repressor-operator binding energy, and $\beta \equiv (k_BT)^{-1}$. This
equation matches the form of the master curve in
Figure~\ref{fig1:means_cartoons}(D) with $\rho=1$ and $\Delta F_R =
\beta\Delta\varepsilon_r - \log (R / N_{NS})$. For model (2) we have a similar
situation. The fold-change takes the form
\begin{align}
\text{fold-change}
&= \left(
1 + \frac{\frac{R}{N_{NS}} e^{-\beta\Delta\varepsilon_R}}
        {1 + \frac{P}{N_{NS}} e^{-\beta\Delta\varepsilon_P}}
\right)^{-1}
\\
&= (1 + \exp(-\Delta F_R + \log\rho))^{-1},
\end{align}
where $P$ is the number of RNAP per cell, and $\Delta\varepsilon_P$ is the
RNAP-promoter binding energy. For this model we have $\Delta F_R =
\beta\Delta\varepsilon_R - \log(R/N_{NS})$ and $\rho = 1 +
\frac{P}{N_{NS}}\mathrm{e}^{-\beta\Delta\varepsilon_P}$. Thus far, we see that
the two thermodynamic models, despite making different coarse-graining
commitments, result in the same functional form for the fold-change in mean gene
expression.  We now explore how kinetic models fare when faced with computing
the same observable.

\subsubsection{Kinetic models}
One of the main difference between models shown in
Figure~\ref{fig1:means_cartoons}(C), casted in the language of chemical master
equations, compared with the thermodynamic models discussed in the previous 
section is the probability space over which they are built. Rather than keeping
track only of the microstate of the promoter, and assuming that gene expression
is proportional to the probability of the promoter being in a certain
microstate, chemical master equation models are built on the entire probability
state of both the promoter microstate, and the current mRNA count. Therefore, in
order to compute the fold-change, we must compute the mean mRNA count on each of
the promoter microstates, and add them all together~\cite{Sanchez2013}.

Again, we spare all details of the derivation to Appendix~\ref{sec:non_bursty}.
Here we just highlight the general findings for all five kinetic models. As
already shown in Figure~\ref{fig1:means_cartoons}(C) and (D), all the kinetic
models explored can be collapsed into the master curve. Given that the
repressor-bound state only connects to the rest of the promoter dynamics via its
binding and unbinding rates, $k_R^+$ and $k_R^-$ respectively, all models can
effectively be separated into two categories: a single repressor-bound state,
and all other promoter states with different levels of coarse graining. This
structure then guarantees that, at steady-state, detailed balance between these
two groups is satisfied. What this implies is that the steady-state distribution
of each of the non-repressor states has the same functional form with or without
the repressor, allowing us to write the fold-change as a product of the ratio of
the binding and unbinding rates of the promoter, and the promoter details. This
is 
\mrm{This is based in one of the comments from reviewer \# 2. It is empirically
true for all models we explored, but I don't know how to prove that it is a 
general statement for any unimaginable complex model.}
\begin{align}
\text{fold-change} &= \left( 1 + \frac{k_R^+}{k_R^-} \rho \right)^{-1},\\
&= (1 + \exp(-\Delta F_R + \log(\rho) ))^{-1},
\end{align}
where $\Delta F_R \equiv -\log(k_R^+/k_R^-)$, and the functional forms of $\rho$
for each model change as shown in Figure~\ref{fig1:means_cartoons}(C) (See
Appendix~\ref{sec:non_bursty} for further details).

The key outcome of our analysis of the models in
Figure~\ref{fig1:means_cartoons} is the existence of a master curve shown
in Figure~\ref{fig1:means_cartoons}(D) to which the fold-change predictions
of all the models collapse. This master curve is parametrized by only two
effective parameters: $\Delta F_R$, which characterizes the number of repressors
and their binding strength to the DNA, and $\rho$, which characterizes all other
features of the promoter architecture. The key assumption underpinning this
result is that no transcription occurs when a repressor is bound to its
operator. Given this outcome, a mean-based metric such as the fold-change that
can be readily measured experimentally is insufficient to discern between these
different levels of coarse-graining. In the next section we extend the analysis
of the models to higher moments of the mRNA distribution as we continue to look
for the ``right'' level of coarse graining.

% \subsection{Discussion of Results Across Models for Fold-Changes in Mean
% Expression}
% The key outcome of our analysis of the models in
% Figure~\ref{fig1:means_cartoons} is the existence of a master curve shown
% in Figure~\ref{fig1:means_cartoons}(D) to which the fold-change predictions
% of all the models collapse. This master curve is parametrized by only two
% effective parameters: $\Delta F_R$, which characterizes the number of repressors
% and their binding strength to the DNA, and $\rho$, which characterizes all other
% features of the promoter architecture. The key assumption underpinning this
% result is that no transcription occurs when a repressor is bound to its
% operator. Note, however, that we are agnostic about the molecular mechanism
% which achieves this; steric effects are one plausible mechanism, but, for
% instance, ``action-at-a-distance'' mediated by kinked DNA due to repressors
% bound tens or hundreds of nucleotides upstream of a promoter is plausible as
% well.

% Why does the master curve of Figure~\ref{fig1:means_cartoons}(D) exist at all?
% This brings to mind the deep questions posed in,
% e.g.,~\cite{Frank2013} and~\cite{Frank2014a}, suggesting we consider multiple
% plausible models of a system and search for their common patterns to tease
% out which broad features are and are not important.
% In our case, the key feature seems to be the
% exclusive nature of repressor and RNAP binding, which allows the parameter
% describing the repressor, $\Delta F_R$, to cleanly separate from all other
% details of the promoter architecture, which are encapsulated in $\rho$.
% Arbitrary nonequilibrium behavior can occur on the rest of the
% promoter state space, but it may all be swept up in the effective
% parameter $\rho$, to which the repressor makes no contribution.
% We point the interested reader to~\cite{Gunawardena2012}
% and~\cite{Ahsendorf2014} for an interesting analysis of similar problems
% using a graph-theoretic language.

% As suggested in~\cite{Chure2019}, we believe this master curve should generalize
% to architectures with multiple repressor binding sites, as long as the
% exclusivity of transcription factor binding and transcription initiation is
% maintained. The interpretation of $\Delta F_R$ is then of an effective free
% energy of all repressor bound states. In an equilibrium picture this is simply
% given by the log of the sum of Boltzmann weights of all repressor bound states,
% which looks like the log of a partition function of a subsystem. In a
% nonequilibrium picture, while we can still mathematically gather terms and give
% the resulting collection the label $\Delta F_R$, it is unclear if the physical
% interpretation as an effective free energy makes sense. The problem is that free
% energies cannot be assigned unambiguously to states out of equilibrium because
% the free energy change along a generic path traversing the state space is path
% dependent, unlike at equilibrium. A consequence of this is that, out of
% equilibrium, $\Delta F_R$ is no longer a simple sum of Boltzmann weights.
% Instead it resembles a restricted sum of King-Altman diagrams~\cite{King1956,
% Hill1966}. Following the work of Hill~\cite{Hill1989}, it may yet be possible to
% interpret this expression as an effective free energy, but this remains unclear
% to us. We leave this an open problem for future work.

% If we relax the requirement of exclusive repressor-RNAP binding, one could
% imagine models in which repressor and RNAP doubly-bound states are allowed,
% where the repressor's effect is to \text{reduce} the transcription rate rather
% than setting it to zero. Our results do not strictly apply to such a model,
% although we note that if the repressor's reduction of the transcription rate is
% substantial, such a model might still be well-approximated by one of the models
% in~\fig{fig1:means_cartoons}.

% One may worry that our ``one curve to rule them all'' is a mathematical
% tautology. In fact we \textit{agree} with this criticism if $\Delta F_R$ is
% ``just a fitting parameter'' and cannot be meaningfully interpreted as a real,
% physical free energy. An analogy to Hill functions is appropriate here. One of
% their great strengths and weaknesses, depending on the use they are put to, is
% that their parameters coarse-grain many details and are generally not
% interpretable in terms of microscopic models, for deep reasons discussed at
% length in~\cite{Frank2013}. By contrast, our master curve claims to have the
% best of both worlds: a coarse-graining of all details besides the repressor into
% a single effective parameter $\rho$, while simultaneously retaining an
% interpretation of $\Delta F_R$ as a physically meaningful and interpretable free
% energy. Our task, then, is to prove or disprove this claim.

% How do we test this and probe the theory with fold-change measurements? There is
% a fundamental limitation in that the master curve is essentially a one-parameter
% function of $\Delta F_R + \log\rho$. Worse, there are many \textit{a priori}
% plausible microscopic mechanisms that could contribute to the value of $\rho$,
% such as RNAP binding and escape kinetics~\cite{DeHaseth1998, Mitarai2015},
% and/or supercoiling accumulation and release~\cite{Chong2014, Sevier2016},
% and/or, RNAP clusters analogous to those observed in
% eukaryotes~\cite{Cisse2013, Cho2016} and recently also observed in
% bacteria~\cite{Ladouceur2020}. Even if $\Delta F_R$ is measured to high
% precision, inferring the potential microscopic contributions to $\rho$, buried
% inside a log no less, from fold-change measurements seems beyond reach. As a
% statistical inference problem it is entirely nonidentifiable, in the language
% of~\cite{Gelman2013}, Section 4.3.

% If we cannot simply infer values of $\rho$ from measurements of fold-change, can
% we perturb some of the parameters that make up $\rho$ and measure the change?
% Unfortunately we suspect this is off-limits experimentally: most of the
% potential contributors to $\rho$ are global processes that affect many or all
% genes. For instance, changing RNAP association rates by changing RNAP copy
% numbers, or changing supercoiling kinetics by changing topoisomerase copy
% numbers, would massively perturb the entire cell's physiology and confound any
% determination of $\rho$.

% One might instead imagine a bottom-up modeling approach, where we mathematicize
% a model of what we hypothesize the important steps are and are not, use \textit{in vitro}
% data for the steps deemed important, and \textit{predict} what $\rho$ should be.
% But again, because of the one-parameter nature of the master curve, many
% different models will likely make indistinguishable predictions, and without any
% way to experimentally perturb \textit{in vivo}, there is no clear way
% to test whether the modeling assumptions are correct.

% In light of this, we prefer the view that parameters and rates are not directly
% comparable between cartoons in~\fig{fig1:means_cartoons}. Rather, parameters in
% the simpler cartoons represent coarse-grained combinations of parameters in the
% finer-grained models. For instance, by equating $\rho$ between any two models,
% one can derive various possible correspondences between the two models'
% parameters. Note that these correspondences are clearly not unique, since many
% possible associations could be made. It then is a choice as to what microscopic
% interpretations the model-builder prefers for the parameters in a particular
% cartoon, and as to which coarse-grainings lend intuition and which seem
% nonsensical. Indeed, since it remains an open question what microscopic features
% dominate $\rho$ (as suggested above, perhaps RNAP binding and escape
% kinetics~\cite{DeHaseth1998, Mitarai2015}, or supercoiling accumulation and
% release~\cite{Chong2014, Sevier2016}, or, something more exotic like RNAP
% clusters~\cite{Cisse2013, Cho2016, Ladouceur2020}), we are hesitant to put too
% much weight on any one microscopic interpretation of model parameters that make
% up $\rho$.

% One possible tuning knob to probe $\rho$ that would not globally perturb the
% cell's physiology is to manipulate RNAP binding sites. Work such
% as~\cite{Kinney2010} has shown that models of sequence-dependent RNAP affinity can be
% inferred from data, and the authors of~\cite{Brewster2012} showed that the model
% of~\cite{Kinney2010} has predictive power by using the model to \textit{design}
% binding sites of a desired affinity. But for our purposes, this begs the
% question: the authors of~\cite{Kinney2010} \textit{assumed} a particular model
% (essentially our 3-state equilibrium model but without the repressor), so it is
% unclear how or if such methods can be turned around to \textit{compare}
% different models of promoter function.

% Another possible route to dissect transcription details without a global
% perturbation would be to use phage polymerase with phage-specific promoters.
% While such results would carry some caveats, e.g., whether the repression of
% the phage polymerase is a good analog to the repression of the native RNAP,
% it could nevertheless be worthy of consideration.

% We have already pointed out that the master curve of~\fig{fig1:means_cartoons}
% is essentially a one-parameter model, the one parameter being $\Delta F_R +
% \log\rho$. By now the reader may be alarmed as to how can we even determine
% $\Delta F_R$ and $\rho$ independently of each other, never mind shedding a lens
% on the internal structure of $\rho$ itself. A hint is provided by the weak
% promoter approximation, invoked repeatedly in prior studies~\cite{Bintu2005c,
% Garcia2011a, Razo-Mejia2018} of simple repression using the 3-state equilibrium
% model in~\fig{fig1:means_cartoons}(B). In that picture, the weak promoter
% approximation means $\frac{P}{N_{NS}}\exp(-\beta\Delta\varepsilon_P) \ll 1$,
% meaning therefore $\rho\approx1$.  This approximation can be well justified on
% the basis of the number of RNAP and $\sigma$ factors per cell
% and the strength of binding of RNAP to
% DNA at weak promoters. This is suggestive, but how can we be sure that $\rho$ is
% not, for instance, actually $10^2$ and that $\Delta F_R$ hides a compensatory
% factor? A resolution is offered by an independent inference of $\rho$ in the
% absence of repressors. This was done in~\cite{Razo-Mejia2020} by fitting
% nonequilibrium model 4 in~\fig{fig1:means_cartoons}(C), with zero repressor
% (looking ahead, this is equivalent to model 4
% in Figure~\ref{fig2:constit_cartoons}(A)),
% to single-cell mRNA counts data from~\cite{Brewster2014}. This provided a
% determination of $k^+$ and $k^-$, from which their ratio is estimated to be no
% more than a few $10^{-1}$ and possibly as small as $10^{-2}$.

% The realization that $\rho\approx1$ to an excellent approximation,
% \textit{independent} of which model in~\fig{fig1:means_cartoons} one prefers,
% goes a long way towards explaining the surprising success of equilibrium models
% of simple repression. Even though our 2- and 3-state models get so many details
% of transcription wrong, it does not matter because fold-change is a cleverly
% designed ratio. Since $\rho$ subsumes all details except the repressor, and
% $\log\rho\approx0$, fitting these simple models to fold-change measurements can
% still give a surprisingly good estimate of repressor binding energies. So the
% ratiometric construction of fold-change fulfills its intended purpose of
% canceling out all features of the promoter architecture except the repressor
% itself. Nevertheless it is perhaps surprising how effectively it does so:
% \textit{a priori}, one might not have expected $\rho$ to be quite so close to 1.

% We would also like to highlight the relevance of~\cite{Landman2019} here.
% Landman et.\ al.\ reanalyzed and compared \textit{in vivo} and
% \textit{in vitro} data on the lacI repressor's binding affinity to its
% various operator sequences. (The \textit{in vivo}
% data was from, essentially, fitting our master curve to expression
% measurements.) They find broad agreement between the
% \textit{in vitro} and \textit{in vivo}
% values. This reinforces the suspicion that the equilibrium $\Delta\varepsilon_R$
% repressor binding energies do in fact represent real physical free energies.
% Again, \textit{a priori} this did not have to be the case, even knowing that
% $\rho\approx1$.

% In principle, if $\Delta F_R$ can be measured to sufficient precision, then
% deviations from $\rho=1$ become a testable matter of experiment. In practice, it
% is probably unrealistic to measure repressor rates $k_R^+$ or $k_R^-$ or
% fold-changes in expression levels (and hence $\Delta\varepsilon_R$) precisely
% enough to detect the expected tiny deviations from $\rho=1$. We can estimate the
% requisite precision in $\Delta F_R$ to resolve a given $\Delta\rho$ by noting,
% since $\rho\approx1$, that $\log(1+\Delta\rho)\approx \Delta\rho$, so
% $\Delta(\Delta F_R) \approx \Delta\rho$. Suppose we are lucky and $\Delta\rho$
% is $\sim0.1$, on the high end of our range estimated above. A
% determination of $\Delta\varepsilon_R/k_BT$ with an uncertainty of barely 0.1
% was achieved in the excellent measurements of~\cite{Razo-Mejia2018}, so this
% requires a very difficult determination of $\Delta F_R$ for a very crude
% determination of $\rho$, which suggests, to put it lightly, this is not a
% promising path to pursue experimentally. It is doubtful that inference of
% repressor kinetic rates would be any easier.%\footnote{\mmnote{In fact the
% % uncertainties quoted in Hammar et.\ al.~\cite{Hammar2014} from Elf's group are
% % small enough to do this calculation, but HJ \& I think their error bars are
% % insanely optimistic. Not sure I want to blatantly call them out on it though??}}

% Moving forward, we have weak evidence supporting the interpretation of $\Delta
% F_R$ as a physically real free energy~\cite{Landman2019} and other work casting
% doubt~\cite{Hammar2014}. How might we resolve the confusion? If there is no
% discriminatory power to test the theory and distinguish the various models with
% measurements of fold-changes in means, how do we probe the theory? Clearly to
% discriminate between the nonequilibrium models in~\fig{fig1:means_cartoons}, we
% need to go beyond means to ask questions about kinetics, noise and even full
% distributions of mRNA copy numbers over a population of cells. If the
% ``one-curve-to-rule-them-all'' is more than a mathematical tautology, then the
% free energy of repressor binding inferred from fold-change measurements should
% agree with repressor binding and unbinding rates. In other words, the
% equilibrium and nonequilibrium definitions of $\Delta F_R$ must agree, meaning
% \begin{equation}
% \Delta F_R = \beta\Delta\varepsilon_R - \log(R/N_{NS})
%         = - \log(k_R^+/k_R^-),
% \label{eq:deltaFR_eq_noneq_equiv}
% \end{equation}
% must hold, where $\beta\Delta\varepsilon_R$ is inferred from the master curve
% fit to fold-change measurements, and $k_R^+$ and $k_R^-$ are inferred in some
% orthogonal manner. Single molecule measurements such as from~\cite{Hammar2014}
% have directly observed these rates, and in the remainder of this work we explore
% a complementary approach: inferring repressor binding and unbinding rates
% $k_R^+$ and $k_R^-$ from single-molecule measurements of mRNA population
% distributions.

### 2020-09-10 doc/resubmission/section_02_means.tex
implement Rob's comments

@@
 Appendix~\ref{sec:non_bursty}. What we will find is that all cartoons collapse
-to a single master curve, shown in \fig{fig1:means_cartoons}(D), which contains
-just two parameters. We label the parameters $\Delta F_R$, an effective free
-energy parametrizing the repressor-DNA interaction, and $\rho$, which subsumes
-all details of transcription in the absence of repressors. We will offer some
-intuition for why this master curve exists and discuss why at the level of the
-mean expression, we are unable to discriminate ``right'' from ``wrong'' cartoons
-given only measurements of fold-changes in expression.
+to a single master curve as shown in \fig{fig1:means_cartoons}(D), which
+contains just two parameters. We label the parameters $\Delta F_R$, an effective
+free energy parametrizing the repressor-DNA interaction, and $\rho$, which
+subsumes all details of transcription in the absence of repressors. We will
+offer some intuition for why this master curve exists and discuss why at the
+level of the mean expression, we are unable to discriminate ``right'' from
+``wrong'' cartoons given only measurements of fold-changes in expression.
 
@@
 
-Again, we spare all details of the derivation to Appendix~\ref{sec:non_bursty}.
+Again, we consign all details of the derivation to Appendix~\ref{sec:non_bursty}.
 Here we just highlight the general findings for all five kinetic models. As
 already shown in Figure~\ref{fig1:means_cartoons}(C) and (D), all the kinetic
-models explored can be collapsed into the master curve. Given that the
+models explored can be collapsed onto the master curve. Given that the
 repressor-bound state only connects to the rest of the promoter dynamics via its
@@
 result is that no transcription occurs when a repressor is bound to its
-operator. Given this outcome, a mean-based metric such as the fold-change that
-can be readily measured experimentally is insufficient to discern between these
+operator. Given this outcome, i.e., the degeneracy of the different models at
+the level of fold-change, a mean-based metric such as the fold-change that can
+be readily measured experimentally is insufficient to discern between these
 different levels of coarse-graining. In the next section we extend the analysis
-of the models to higher moments of the mRNA distribution as we continue to look
-for the ``right'' level of coarse graining.
+of the models to higher moments of the mRNA distribution as we continue to
+examine the discriminatory power of these different models.
 

### 2020-09-11 doc/resubmission/appendix_01_nonbursty.tex
proposed changes to beyond means section

@@
 We will also see later how this model emerges as a natural limit of model 4.
+\mrm{ENDS HERE}
+
+\mrm{THIS NEEDS TO BE FIXED}
+\subsubsection{Noise in the Two-State Promoter, RNAP Bound or Unbound.}
+
+Our second model of constitutive transcription posits an architecture in which
+the promoter is either empty or bound by RNAP ~\cite{Phillips2015a,
+Phillips2019}. Here,  as shown in model 2 of Figure~\ref{fig2:constit_cartoons}(A),
+transcription initiation results in a state transition from the bound to the
+unbound state, reflecting the microscopic reality that an RNAP that has begun to
+elongate a transcript is no longer available at the start site to begin another.
+As shown in Appendix~\ref{sec:non_bursty}, the Fano factor in this model is
+given by
+\begin{align}
+    \nu = 1 -
+        \frac{r\rate{k}{P}{+}}
+            {\left(\rate{k}{P}{+} + \rate{k}{P}{-} + r\right)
+             \left(\gamma + \rate{k}{P}{+} + \rate{k}{P}{-} + r\right)}.
+\label{eq:model2_fano}
+\end{align}
+The problem with this picture is that the Fano factor is always $<1$. To make
+ contact with the experimental reality of $\nu>1$ as shown in
+ Figure~\ref{fig2:constit_cartoons}(B), clearly some corrections will be needed.
+ While this model adds an appealing element of microscopic reality, we are
+ forced to reject it as the additional complexity is unable to capture the
+ phenomenology of interest. Obviously the promoter state does in fact proceed
+ through cycles of RNAP binding, initiating, and elongating, but it seems that
+ the super-Poissonian noise in mRNA copy number we want to model must be
+ governed by other features of the system.
+
+\subsubsection{Noise in the Three-State Promoter, Multistep Transcription
+Initiation and Escape.}
+
+How might we remedy the deficits of model 2? It is known~\cite{DeHaseth1998}
+that once RNAP initially binds the promoter region, a multitude of distinct
+steps occur sequentially before RNAP finally escapes into the elongation phase.
+Perhaps adding some of this mechanistic detail as shown in model 3 of
+Figure~\ref{fig2:constit_cartoons}(A) might rescue the previous model. The next
+simplest refinement of that model could consider open complex formation and
+promoter escape as separate steps rather than as a single effective step. In
+other words, we construct model 3 by adding a single extra state to model 2, and
+we will label the two RNAP-bound states as the closed and open complexes,
+despite the true biochemical details certainly being more complex. For example,
+earlier work extended this model by adding an additional repressor bound state
+and did not explicitly consider the limit with no repressor that we analyze
+here~\cite{Mitarai2015}. Again, our goal here is not a complete accounting of
+all the relevant biochemical detail; this is an exploratory search for the
+important features that a model needs to include to square with the known
+experimental reality of constitutive expression.
+
+Unfortunately, as hinted at in earlier work~\cite{Mitarai2015}, this model too
+has Fano factor $\nu<1$. We again leave the algebraic details for
+Appendix~\ref{sec:non_bursty} and merely state the result that
+\begin{align}
+\nu = 1 - \frac{r k_P^+ k_O}{\mathcal{Z}}
+\frac{k_P^+ + k_P^- + k_O + r + \gamma}
+    {\mathcal{Z} + \gamma(k_P^+ + k_P^- + k_O + r) + \gamma^2},
+\label{eq:model3_fano}
+\end{align}
+where we defined $\mathcal{Z} = r(k_O + k_P^-) + k_P^+(k_O + r)$ for notational
+tidiness. This is necessarily less than 1 for arbitrary rate constants.
+
+In fact, we suspect \textit{any} model in which transcription proceeds through a
+multistep cycle must necessarily have $\nu<1$. The intuitive argument compares
+the waiting time distribution to traverse the cycle with the waiting time for a
+Poisson promoter (model 1) with the same mean time. The latter is simply an
+exponential distribution. The former is a convolution of multiple exponentials,
+and intuitively the waiting time distribution for a multistep process
+should be more peaked with a smaller fractional width than a single
+exponential with the same mean.
+A less disperse waiting time distribution means transcription initations are
+more uniformly distributed in time relative to a Poisson process. Hence the
+distribution of mRNA over a population of cells should be less variable compared
+to Poisson, giving $\nu<1$.
+(In Appendix~\ref{sec:non_bursty} we present a more precise version
+of the intuitive arguments in this paragraph.)
+Regardless of the merits of this model in describing the noise properties of
+constitutive transcription initiation, it ultimately fails the simplest
+quantitative feature of the data, namely that the Fano factor $> 1$ and hence
+we must discard this mechanistic picture and search elsewhere.
 \mrm{ENDS HERE}

### 2020-09-11 doc/resubmission/section_02_means.tex
proposed changes to beyond means section

@@
 \subsubsection{Two- and Three-state Thermodynamic Models}
-We begin our analysis with models (1) and (2) in
+We begin our analysis with models 1 and 2 in
 Figure~\ref{fig1:means_cartoons}(B). In each of these models the promoter
@@
 assumed to be proportional to the probability of the promoter being in either
-the empty state (model (1)) or the RNAP-bound state (model (2)). We direct the
+the empty state (model 1) or the RNAP-bound state (model (2)). We direct the
 reader to Appendix~\ref{sec:non_bursty} for details on the derivation of the
 fold-change. For our purposes here, it suffices to state that the functional
-form of the fold-change for model (1) is
+form of the fold-change for model 1 is
 \begin{equation}
@@
 Figure~\ref{fig1:means_cartoons}(D) with $\rho=1$ and $\Delta F_R =
-\beta\Delta\varepsilon_r - \log (R / N_{NS})$. For model (2) we have a similar
+\beta\Delta\varepsilon_r - \log (R / N_{NS})$. For model 2 we have a similar
 situation. The fold-change takes the form

### 2020-09-11 doc/resubmission/section_03_beyond_means.tex
implement Rob's comments
proposed changes to beyond means section

--- 2020-06-06 doc/section_03_beyond_means.tex
@@
 \centering
-\includegraphics[width=\textwidth]{../figures/main/fig02.pdf}
+\includegraphics[width=\textwidth]{../../figures/main/fig02.pdf}
 \caption{\textbf{Comparison of different models for noise in the constitutive promoter.}
@@
 
-\subsubsection{Noise in the Poisson Promoter Model}
+\subsubsection{Poisson Noise promoter}
 
 The simplest model of constitutive expression that we can imagine is shown as
-model 1 in Figure~\ref{fig2:constit_cartoons}(A) and assumes that transcripts are
-produced as a Poisson process from a single promoter state. This is the picture
-from Jones et.\ al.~\cite{Jones2014} that was used to interpret a systematic
-study of gene expression noise over a series of promoters designed to have
-different strengths. This model insists that the ``true'' steady-state mRNA
+model 1 in Figure~\ref{fig2:constit_cartoons}(A) and assumes that transcripts
+are produced as a Poisson process from a single promoter state. This is the
+picture from Jones et.\ al.~\cite{Jones2014} that was used to interpret a
+systematic study of gene expression noise over a series of promoters designed to
+have different strengths. This model insists that the ``true'' steady-state mRNA
 distribution is Poisson, implying the Fano factor $\nu$ must be 1.
@@
 the corrections to Poisson behavior are derived as additive corrections to the
-Fano factor. For disproving the ``universal noise curve'' from So et.\
-al.~\cite{So2011}, this picture was excellent. It is appealing in its simplicity,
-with only two parameters, the initiation rate $r$ and degradation rate $\gamma$.
-Since $\gamma$ is independently known from other experiments, and the mean mRNA
-copy number is $r/\gamma$, $r$ is easily inferred from data. In other words, the
-model is not excessively complex for the data at hand. But for many interesting
-questions, for instance in the recent work~\cite{Razo-Mejia2020}, knowledge of
-means and variances alone is insufficient, and a description of the full
-distribution of molecular counts is necessary. For this we need a (slightly)
-more complex model than model 1 that would allow us to incorporate the
-non-Poissonian features of constitutive promoters directly into a master
-equation formulation.
+Fano factor. For \mrm{challenging} the ``universal noise curve'' from So et.\
+al.~\cite{So2011}, this picture was excellent. It is appealing in its
+simplicity, with only two parameters, the initiation rate $r$ and degradation
+rate $\gamma$. Since $\gamma$ is independently known from other experiments, and
+the mean mRNA copy number is $r/\gamma$, $r$ is easily inferred from data. In
+other words, the model is not excessively complex for the data at hand. But for
+many interesting questions, for instance in the recent
+work~\cite{Razo-Mejia2020}, knowledge of means and variances alone is
+insufficient, and a description of the full distribution of molecular counts is
+necessary. For this we need a (slightly) more complex model than model 1 that
+would allow us to incorporate the non-Poissonian features of constitutive
+promoters directly into a master equation formulation.
 
-\subsubsection{Noise in the Two-State Promoter, RNAP Bound or Unbound.}
+\mrm{I added the word challenging instead of disproving because of the 
+following comment from reviewer 1. But maybe more needs to be discussed about 
+this?}
 
-Our second model of constitutive transcription posits an architecture in which
-the promoter is either empty or bound by RNAP ~\cite{Phillips2015a,
-Phillips2019}. Here,  as shown in model 2 of Figure~\ref{fig2:constit_cartoons}(A),
-transcription initiation results in a state transition from the bound to the
-unbound state, reflecting the microscopic reality that an RNAP that has begun to
-elongate a transcript is no longer available at the start site to begin another.
-As shown in Appendix~\ref{sec:non_bursty}, the Fano factor in this model is
-given by
-\begin{align}
-    \nu = 1 -
-        \frac{r\rate{k}{P}{+}}
-            {\left(\rate{k}{P}{+} + \rate{k}{P}{-} + r\right)
-             \left(\gamma + \rate{k}{P}{+} + \rate{k}{P}{-} + r\right)}.
-\label{eq:model2_fano}
-\end{align}
-The problem with this picture is that the Fano factor is always $<1$. To make
- contact with the experimental reality of $\nu>1$ as shown in
- Figure~\ref{fig2:constit_cartoons}(B), clearly some corrections will be needed.
- While this model adds an appealing element of microscopic reality, we are
- forced to reject it as the additional complexity is unable to capture the
- phenomenology of interest. Obviously the promoter state does in fact proceed
- through cycles of RNAP binding, initiating, and elongating, but it seems that
- the super-Poissonian noise in mRNA copy number we want to model must be
- governed by other features of the system.
+\mrm{
+    “disproving the universal noise curve from So et.
+al.
+[59]” (page 18).
+How was the universal noise curve disproved?
+Universal noise/mean relations, similar to those reported in ref \#59, were repeatedly observed, in both bacteria and eukaryotes (see, e.g., PMID: 24311680, PMID: 25858977).
+Moreover, later reports showed that the trend is observed even when accounting for gene dosage effects (PMID: 26965629).
+Thus, the statement above requires clarification.
+}
 
-\subsubsection{Noise in the Three-State Promoter, Multistep Transcription
-Initiation and Escape.}
+\subsubsection{Sub-Poissoninan noise promoters with RNAP escape}
+A natural extension of the one-state promoter studied in the previous section is
+to explicitly include an empty promoter state. This state allows for single RNAP
+to bind and unbind from the promoter with rates $k_P^+$ and $k_P^-$,
+respectively, before engaging in a transcriptional event. Once the RNAP
+irreversibly escapes from the promoter to transcribe the mRNA, the promoter goes
+back to this empty state. Models 2 and 3 from
+Figure~\ref{fig2:constit_cartoons}(A) show two different versions of this
+process. In model 2 the polymerase can directly engage in a transcription event
+with rate $r$ after binding to the promoter, while model 3 includes an
+irreversible step with rate $k_O$ that coarse-grains the known multi-step
+process involved in transcriptional initiation~\cite{DeHaseth1998}.
 
-How might we remedy the deficits of model 2? It is known~\cite{DeHaseth1998}
-that once RNAP initially binds the promoter region, a multitude of distinct
-steps occur sequentially before RNAP finally escapes into the elongation phase.
-Perhaps adding some of this mechanistic detail as shown in model 3 of
-Figure~\ref{fig2:constit_cartoons}(A) might rescue the previous model. The next
-simplest refinement of that model could consider open complex formation and
-promoter escape as separate steps rather than as a single effective step. In
-other words, we construct model 3 by adding a single extra state to model 2, and
-we will label the two RNAP-bound states as the closed and open complexes,
-despite the true biochemical details certainly being more complex. For example,
-earlier work extended this model by adding an additional repressor bound state
-and did not explicitly consider the limit with no repressor that we analyze
-here~\cite{Mitarai2015}. Again, our goal here is not a complete accounting of
-all the relevant biochemical detail; this is an exploratory search for the
-important features that a model needs to include to square with the known
-experimental reality of constitutive expression.
-
-Unfortunately, as hinted at in earlier work~\cite{Mitarai2015}, this model too
-has Fano factor $\nu<1$. We again leave the algebraic details for
-Appendix~\ref{sec:non_bursty} and merely state the result that
-\begin{align}
-\nu = 1 - \frac{r k_P^+ k_O}{\mathcal{Z}}
-\frac{k_P^+ + k_P^- + k_O + r + \gamma}
-    {\mathcal{Z} + \gamma(k_P^+ + k_P^- + k_O + r) + \gamma^2},
-\label{eq:model3_fano}
-\end{align}
-where we defined $\mathcal{Z} = r(k_O + k_P^-) + k_P^+(k_O + r)$ for notational
-tidiness. This is necessarily less than 1 for arbitrary rate constants.
-
-In fact, we suspect \textit{any} model in which transcription proceeds through a
-multistep cycle must necessarily have $\nu<1$. The intuitive argument compares
-the waiting time distribution to traverse the cycle with the waiting time for a
-Poisson promoter (model 1) with the same mean time. The latter is simply an
-exponential distribution. The former is a convolution of multiple exponentials,
-and intuitively the waiting time distribution for a multistep process
-should be more peaked with a smaller fractional width than a single
-exponential with the same mean.
-A less disperse waiting time distribution means transcription initations are
-more uniformly distributed in time relative to a Poisson process. Hence the
+Although physically intuitive given our understanding of how a single transcript
+is produced, we are forced to discard both of these models on the basis that the
+Fano factor for both of them is strictly $< 1$ as shown in
+Figure~\ref{fig2:constit_cartoons}(A) (See Appendix~\ref{sec:non_bursty} for
+detailed derivations). This is in disagreement with the experimental data
+from~\cite{Jones2014}, reproduced in Figure~\ref{fig2:constit_cartoons}(B). All
+of these unregulated promoters --which only differ on the promoter sequence--
+show strictly super-Poissonian noise (Fano factor $> 1$). In fact, we suspect
+\textit{any} model in which transcription proceeds through a multistep cycle
+must necessarily have $\nu<1$. The intuitive argument compares the waiting time
+distribution to traverse the cycle with the waiting time for a Poisson promoter
+(model 1) with the same mean time. The latter is simply an exponential
+distribution. The former is a convolution of multiple exponentials, and
+intuitively the waiting time distribution for a multistep process should be more
+peaked with a smaller fractional width than a single exponential with the same
+mean. A less disperse waiting time distribution means transcription initations
+are more uniformly distributed in time relative to a Poisson process. Hence the
 distribution of mRNA over a population of cells should be less variable compared
-to Poisson, giving $\nu<1$.
-(In Appendix~\ref{sec:non_bursty} we present a more precise version
-of the intuitive arguments in this paragraph.)
-Regardless of the merits of this model in describing the noise properties of
-constitutive transcription initiation, it ultimately fails the simplest
-quantitative feature of the data, namely that the Fano factor $> 1$ and hence
-we must discard this mechanistic picture and search elsewhere.
+to Poisson, giving $\nu<1$. (In Appendix~\ref{sec:non_bursty} we present a more
+precise version of the intuitive arguments in this paragraph).
 
@@
 physical meaning to the two states in this model.
+\mrm{I like this paragraph's philosphy, but I don't know if it should be cut out 
+for the sake of shrinking the paper further.}
 
@@
 
-The final model we consider is 
-inspired by the failure mode of model 4. The key observation above
-was that, as found in~\cite{Razo-Mejia2020}, only two parameters, $k^+$ and the
-ratio $r/k^-$, could be directly inferred from the single-cell mRNA data
-from~\cite{Jones2014}. So let us take this seriously and imagine a model where
-these are the only two model parameters. What would this model look like?
+The final model we consider is inspired by the failure mode of model 4. The key
+observation above was that, as found in~\cite{Razo-Mejia2020}, only two
+parameters, $k^+$ and the ratio $r/k^-$, could be directly inferred from the
+single-cell mRNA data from~\cite{Jones2014}. So let us take this seriously and
+imagine a model where these are the only two model parameters. What would this
+model look like?
 
-% \marginpar{\it RP: best description I have ever seen of this.  Good job}
 To develop some intuition, consider model 4 in the limit $k^+ \ll k^- \lesssim
@@
 transcripts. This should call to mind the well-known phenomenon of
-transcriptional bursting, as reported in,
-e.g.,~\cite{Golding2005,Chong2014,Sevier2016}.
-%\mmnote{should probably add some more cites here}.
-Let us make this correspondence more precise. The mean dwell
-time in the active state is $1/k^-$. While in this state, transcripts are
-produced at a rate $r$ per unit time. So on average, $r/k^-$ transcripts are
-produced before the system switches to the inactive state. Once in the inactive
-state, the system dwells there for an average time $1/k^+$ before returning to
-the active state and repeating the process. $r/k^-$ resembles an average burst
-size, and $1/k^+$ resembles the time interval between burst events. More
-precisely, $1/k^+$ is the mean time between the end of one burst and the start
-of the next, whereas $1/k^+ + 1/k^-$ would be the mean interval between the
-start of two successive burst events, but in the limit $k^+ \ll k^-$, $1/k^+ +
-1/k^- \approx 1/k^+$. Note that this limit ensures that the waiting time between
-bursts is approximately exponentially distributed, with mean set by the only
-timescale left in the problem, $1/k^+$.
-% \footnote{If instead it were the case that $k^+ \sim k^-$, then the waiting time
-% $1/k^+ + 1/k^-$ would have a peaked distribution, but this does not appear to be
-% the case for any of datasets from~\cite{Jones2014}.}
+transcriptional bursting, as observed in,
+e.g.,~\cite{Golding2005,Chong2014,Sevier2016,Sepulveda2016}. Let us make this
+correspondence more precise. The mean dwell time in the active state is $1/k^-$.
+While in this state, transcripts are produced at a rate $r$ per unit time. So on
+average, $r/k^-$ transcripts are produced before the system switches to the
+inactive state. Once in the inactive state, the system dwells there for an
+average time $1/k^+$ before returning to the active state and repeating the
+process. $r/k^-$ resembles an average burst size, and $1/k^+$ resembles the time
+interval between burst events. More precisely, $1/k^+$ is the mean time between
+the end of one burst and the start of the next, whereas $1/k^+ + 1/k^-$ would be
+the mean interval between the start of two successive burst events, but in the
+limit $k^+ \ll k^-$, $1/k^+ + 1/k^- \approx 1/k^+$. Note that this limit ensures
+that the waiting time between bursts is approximately exponentially distributed,
+with mean set by the only timescale left in the problem, $1/k^+$. If instead it
+were the case that $k^+ \sim k^-$, then the waiting time $1/k^+ + 1/k^-$ would
+have a peaked distribution, but this does not appear to be the case for any of
+datasets from~\cite{Jones2014}.
 
@@
 \end{equation}
-Since the dwell time $T$ is unobservable, we actually want $P(n)$, the dwell
-time distribution with no conditioning on $T$. Basic rules of probability theory
-tell us we can write $P(n)$ in terms of $P(n\mid T)$ as
+Since the dwell time $T$ is unobservable from single molecule mRNA counts, we
+actually want $P(n)$, the dwell time distribution with no conditioning on $T$.
+Basic rules of probability theory tell us we can write $P(n)$ in terms of
+$P(n\mid T)$ as
 \begin{equation}
@@
 \end{align}
-where we use $k_i$ to denote the burst initiation rate, $\text{Geom}(n;b)$ is the
-geometric distribution with mean~$b$, i.e., $\text{Geom}(n;b) =
+where we use $k_i$ to denote the burst initiation rate, $\text{Geom}(n;b)$ is
+the geometric distribution with mean~$b$, i.e., $\text{Geom}(n;b) =
 \frac{1}{1+b}\left(\frac{b}{1+b}\right)^n$ (with domain over nonnegative
@@
 later~\cite{Jones2014}.
+\mrm{Another great paragraph that could potentially be removed.}
 

### 2020-09-14 doc/resubmission/section_02_means.tex
implement Robs 20200913 comments

@@
 of occupancy accorded its own rate of transcription -- including no 
-transcription for many of these states. The models are probabilistic with each
+transcription for many of these states. This discretization of a potentially
+continuous number of promoter states (due to effects such as supercoiling of 
+DNA~\cite{Chong2014, Sevier2016}) is analogous to how the Monod-Wyman-Changeux
+model of allostery coarse-grains continuous molecule conformations into a finite
+number of states~\cite{Martins2011}. The models are probabilistic with each
 state assigned some probability and the overall rate of transcription given by 

### 2020-09-14 doc/resubmission/section_03_beyond_means.tex
implement Robs 20200913 comments

@@
 we will focus on the Fano factor, defined as the variance/mean. This metric
-serves as a powerful discriminatory tool from the null model that the
-steady-state mRNA distribution must be Poisson, giving a Fano factor of one.
+serves as a powerful discriminatory tool to compare our different models to the
+null model that the steady-state mRNA distribution must be Poisson, resulting a
+Fano factor of one.
 
@@
 constitutive promoters. In model (1), transcripts are produced in a Poisson
-process~\cite{Sanchez2013, Jones2014}. Model (2) features explicit modeling of
+process~\cite{Sanchez2013, Jones2014}. Model (2) features explicit treatment of
 RNAP binding/unbinding kinetics~\cite{Phillips2015a}. Model (3) is a more
@@
 systematic study of gene expression noise over a series of promoters designed to
-have different strengths. This model insists that the ``true'' steady-state mRNA
-distribution is Poisson, implying the Fano factor $\nu$ must be 1.
-In~\cite{Jones2014}, the authors carefully attribute measured deviations from
-Fano = 1 to intensity variability in fluorescence measurements, gene copy number
-variation, and copy number fluctuations of the transcription machinery, e.g.,
-RNAP itself. In this picture, the master equation makes no appearance, and all
-the corrections to Poisson behavior are derived as additive corrections to the
-Fano factor. For \mrm{challenging} the ``universal noise curve'' from So et.\
-al.~\cite{So2011}, this picture was excellent. It is appealing in its
-simplicity, with only two parameters, the initiation rate $r$ and degradation
-rate $\gamma$. Since $\gamma$ is independently known from other experiments, and
-the mean mRNA copy number is $r/\gamma$, $r$ is easily inferred from data. In
-other words, the model is not excessively complex for the data at hand. But for
-many interesting questions, for instance in the recent
-work~\cite{Razo-Mejia2020}, knowledge of means and variances alone is
-insufficient, and a description of the full distribution of molecular counts is
-necessary. For this we need a (slightly) more complex model than model 1 that
-would allow us to incorporate the non-Poissonian features of constitutive
-promoters directly into a master equation formulation.
-
-\mrm{I added the word challenging instead of disproving because of the 
-following comment from reviewer 1. But maybe more needs to be discussed about 
-this?}
-
-\mrm{
-    “disproving the universal noise curve from So et.
-al.
-[59]” (page 18).
-How was the universal noise curve disproved?
-Universal noise/mean relations, similar to those reported in ref \#59, were repeatedly observed, in both bacteria and eukaryotes (see, e.g., PMID: 24311680, PMID: 25858977).
-Moreover, later reports showed that the trend is observed even when accounting for gene dosage effects (PMID: 26965629).
-Thus, the statement above requires clarification.
-}
+have different strengths, but no regulation. This model insists that the
+``true'' steady-state mRNA distribution is Poisson, implying the Fano factor
+$\nu$ must be 1. In~\cite{Jones2014}, the authors carefully attribute measured
+deviations from Fano = 1 to intensity variability in fluorescence measurements,
+gene copy number variation, and copy number fluctuations of the transcription
+machinery, e.g., RNAP itself. In this picture, all the corrections to Poisson
+behavior are derived as additive corrections to the Fano factor. For challenging
+the ``universal noise curve'' from So et.\ al.~\cite{So2011}, this picture was
+excellent. It is appealing in its simplicity, with only two parameters, the
+initiation rate $r$ and degradation rate $\gamma$. In other words, the model is
+not excessively complex for the data at hand. But for many interesting
+questions, for instance in the recent work~\cite{Razo-Mejia2020}, attributing
+all deviations from the model to extrinsic noise sources, limits the kinds of
+predictions that can be done. To make progress then we need a (slightly) more
+complex model than model 1 that would allow us to incorporate the non-Poissonian
+features of constitutive promoters directly into a master equation formulation.
 
@@
 from~\cite{Jones2014}, reproduced in Figure~\ref{fig2:constit_cartoons}(B). All
-of these unregulated promoters --which only differ on the promoter sequence--
+of these unregulated promoters --which only differ in their promoter sequence--
 show strictly super-Poissonian noise (Fano factor $> 1$). In fact, we suspect
@@
 physical meaning to the two states in this model.
-\mrm{I like this paragraph's philosphy, but I don't know if it should be cut out 
-for the sake of shrinking the paper further.}
 
@@
 later~\cite{Jones2014}.
-\mrm{Another great paragraph that could potentially be removed.}
 

### 2020-09-22 code/figures/si/figS0X_histograms.py
add histogram version of fig 4C

@@ -1,0 +1,175 @@
+# %%
+import enum
+import re
+import dill
+from git import Repo #for directory convenience
+
+import numpy as np
+import pandas as pd
+
+import emcee
+import arviz as az
+
+import matplotlib.pyplot as plt
+import seaborn as sns
+import bebi103.viz
+
+import srep
+
+srep.viz.plotting_style()
+pboc_colors = srep.viz.color_selector('pboc')
+
+# %%
+fig, ax = plt.subplots(4, 3, figsize=(7.5, 10), sharex=True, sharey=False)
+
+# # Modify tick font size
+# for a in ax:
+#     a.tick_params(axis="both", which="major", labelsize=8)
+
+repo = Repo("./", search_parent_directories=True)
+# repo_rootdir holds the absolute path to the top-level of our repo
+repo_rootdir = repo.working_tree_dir
+
+# Select PBoC color palette
+colors = srep.viz.color_selector('pboc')
+# Set PBoC plotting style
+srep.viz.plotting_style()
+
+
+# load in the pickled samples
+pklfile = open(
+    f"{repo_rootdir}/data/mcmc_samples/repression_pooled_expts.pkl", 'rb'
+    )
+model, sampler, ppc_uv5, ppc_rep = dill.load(pklfile)
+pklfile.close()
+inf_dat = az.convert_to_inference_data(
+    sampler, var_names=model.var_labels
+    )
+
+data_uv5, data_rep = srep.utils.condense_data(model.expts)
+
+n_dim = np.shape(model.var_labels)
+
+# Define operators
+op_array = ["Oid", "O1", "O2"]
+
+# Define aTc concentrations
+aTc_array = ["0p5ngmL", "1ngmL", "2ngmL", "10ngmL"]
+
+# Set global colors for aTc concentrations
+aTc_colors = ('blue', 'betancourt', 'green', 'orange')
+aTc_col_dict = dict(zip(aTc_array , aTc_colors))
+
+# organize all the options upfront
+all_expts = (
+    ("Oid_2ngmL", "Oid_1ngmL"),
+    ("O1_1ngmL", "O1_2ngmL", "O1_10ngmL"),
+    ("O2_0p5ngmL", "O2_1ngmL", "O2_2ngmL", "O2_10ngmL")
+)
+
+# Loop through operators concentrations
+for op_idx, op in enumerate(op_array):
+    # List experiments available for operator
+    op_exp = all_expts[op_idx]
+
[…]

### 2020-09-22 code/figures/si/figS0X_histograms.py
update histogram plot

@@ -20,7 +20,7 @@ srep.viz.plotting_style()
 pboc_colors = srep.viz.color_selector('pboc')
 
 # %%
-fig, ax = plt.subplots(4, 3, figsize=(7.5, 10), sharex=True, sharey=False)
+fig, ax = plt.subplots(4, 3, figsize=(8.5, 10), sharex=False, sharey=False)
 
 # # Modify tick font size
 # for a in ax:
@@ -103,16 +103,16 @@ for op_idx, op in enumerate(op_array):
             )
             # Remove residual ticks from the original left axis
             ax[aTc_idx, op_idx].tick_params(color="w", width=0)
-        # Add xlabel to bottom plots
-        if aTc_idx == 3:
-            ax[aTc_idx, op_idx].set_xlabel("mRNA / cell")
+        
         # Add ylabel to left plots
-        if op_idx == 0:
-            ax[aTc_idx, op_idx].set_ylabel("probability")
+        # if op_idx == 0:
+        #     ax[aTc_idx, op_idx].set_ylabel("probability")
 
         # Check if experiment exists, if not, skip experiment
         if expt not in op_exp:
             ax[aTc_idx, op_idx].set_facecolor("#D3D3D3")
+            ax[aTc_idx, op_idx].tick_params(axis='x', colors='white')
+            ax[aTc_idx, op_idx].tick_params(axis='y', colors='white')
             continue
 
         # Find experiment index
@@ -160,13 +160,18 @@ for op_idx, op in enumerate(op_array):
             hist_data,
             where="post",
             color="black",
-            linewidth=1
+            linewidth=1.25
         )
+        # Set x-label
+        ax[aTc_idx, op_idx].set_xlabel("mRNA / cell")
+        ax[aTc_idx, op_idx].set_ylabel("probability")
 
-        ax[aTc_idx, op_idx].set_xlim(-2.5, 55)
+        # Set axis limit
+        upper_limit = np.where(hist_data > 5E-3)[0][-1]
+        ax[aTc_idx, op_idx].set_xlim(0, upper_limit)
         
 # Adjust spacing between plots
-plt.subplots_adjust(hspace=0.08, wspace=0.3)
+plt.subplots_adjust(hspace=0.3, wspace=0.4)
 
 
 plt.savefig(

### 2020-09-22 code/figures/si/figS0X_histograms.py
fix figure name

@@ -175,6 +175,6 @@ plt.subplots_adjust(hspace=0.3, wspace=0.4)
 
 
 plt.savefig(
-    f"{repo_rootdir}/figures/si/fig0X_histograms.pdf", bbox_inches='tight'
+    f"{repo_rootdir}/figures/si/figS0X_histograms.pdf", bbox_inches='tight'
 )
 # %%

### 2020-09-24 doc/resubmission/appendix_02_bursty.tex
updates before bringing back Dropbox documents

--- 2020-06-06 doc/appendix_02_bursty.tex
@@
 \centering
-\includegraphics{../figures/si/figS0X_bursty_states.pdf}
+\includegraphics{../../figures/si/figS0X_bursty_states.pdf}
 \caption{
@@
 \centering
-\includegraphics[width=5cm]{../figures/si/figS0X_reindex_sum.pdf}
+\includegraphics[width=5cm]{../../figures/si/figS0X_reindex_sum.pdf}
 \caption{\textbf{Reindexing double sum.} Schematic for reindexing the sum
@@
 \centering
-\includegraphics{../figures/si/figS0X_Shahrezaei_promoter.pdf}
+\includegraphics{../../figures/si/figS0X_Shahrezaei_promoter.pdf}
 \caption{\textbf{Schematic of three-stage promoter from~\cite{Shahrezaei2008}.}

### 2020-09-24 doc/resubmission/section_02_means.tex
updates before bringing back Dropbox documents

@@
 microscopic states ~\cite{Ko1991, Peccoud1995, Record1996, Kepler2001,
-Sanchez2008, Shahrezaei2008, Sanchez2011, Michel2010}. As seen in
-Figure~\ref{fig1:means_cartoons}(C), we consider a host of different
-nonequilibrium models, each of which will have its own result for both the mean
-(this section) and noise (next section) in gene expression.
+Sanchez2008, Shahrezaei2008, Sanchez2011, Michel2010}. The main differences
+between both modeling approaches can be summarized as: 1) Although for both
+classes of models the steps involving transcriptional events are assumed to be
+strictly irreversible, thermodynamic models force the regulation, i.e., the
+control over the expression exerted by the repressor, to be in equilibrium. This
+does not need to be the case for kinetic models. 2) Thermodynamic models ignore
+the mRNA count from the state of the Markov process, while kinetic models keep
+track of both the promoter state and the mRNA count. 3) Finally, thermodynamic
+and kinetic models coarse-grain to different degrees the molecular mechanisms
+through which RNAP enters the transcriptional event. As seen in
+Figure~\ref{fig1:means_cartoons}(C), we consider a host of different kinetic
+models, each of which will have its own result for both the mean (this section)
+and noise (next section) in gene expression.
 

### 2020-09-24 doc/resubmission/appendix_01_nonbursty.tex
update resubmission version of paper

@@
 In this section we detail the calculation of mean mRNA levels, fold-changes in
-expression, and Fano factors for nonequilibrium promoter models 1 through 4
+expression, and Fano factors for both thermodynamic and kinetic promoter models
 in Figure~\ref{fig1:means_cartoons}. These are the results that were quoted but
 not derived in Sections~\ref{section_02_means} and \ref{sec:beyond_means} of the
-main text. In each of these four models, the natural mathematicization of their
-cartoons is as a chemical master equation such as Eq.~\ref{eq:2state_rep_cme}
-for model 1. Before jumping into the derivations of the general computation of
-the mean mRNA level and the Fano factor we will work through the derivation of
-an example master equation. In particular we will focus on model 1 from
-Figure~\ref{fig1:means_cartoons}(C). The general steps are applicable to all
-other chemical master equations in this work.
+main text. In each of these models, the natural mathematicization of their
+cartoons is either an equilibrium model based on statisticla mechanics, or a
+chemical master equation. These derivations will go through the specifics of the
+models in Figure~\ref{fig1:means_cartoons}. We point the readers to some other
+excellent reviews on the general
+frameworks~\cite{Bintu2005a,Sanchez2013,Phillips2015a}.
 
 \mrm{THIS NEEDS TO BE FIXED}
+\subsection{Thremodynamic models of gene regulation}
+The first class of models we will explore are the so-called thermodynamic models
+of gene regulation~\cite{Ackers1982}. The premise for these models is that we
+imagine the steps leading to transcription --binding and unbinding of
+transcription factors and of RNAP-- to occur in a timescale much faster than the
+downstream processes --open complex formation and RNAP
+escape~\cite{Browning2004}. What this allows us to do is to assume that the
+steps leading to transcription can be modeled as being in quasi-equilibrium.
+This is enormously advantageous since we can use the theoretical tools of 
+equilibrium statisticla mechanics to model such process.
+
+This means that we can enlist the possible microstates in which we can find the
+promoter in order to calculate their probabilities based on quantities such as
+molecular counts and interaction energies. We then assume that the gene
+expression level is proportional to the probability of finding the promoter in
+the transcriptionally active microstate. As mentioned in the main text, these
+models can only deal with the mean gene expression level. This is because the
+probability space we consider is that of the state of the repressor, rather than
+both the state of the repressor and the mRNA counts. Let's now work through
+models 1 and 2 from Figure~\ref{fig1:means_cartoons}(B).
+
 \subsubsection{The Two-State Equilibrium Model}
@@
 \mrm{THIS NEEDS TO BE FIXED}
+
+Before jumping into the derivations of the general
+computation of the mean mRNA level and the Fano factor we will work through the
+derivation of an example master equation. In particular we will focus on model 1
+from Figure~\ref{fig1:means_cartoons}(C). The general steps are applicable to
+all other chemical master equations in this work.
+
 \subsubsection{The Poisson Promoter Nonequilibrium Model}

### 2020-09-24 doc/resubmission/appendix_03_bayesian_inference.tex
updates before bringing back Dropbox documents
update resubmission version of paper

--- 2020-06-06 doc/appendix_03_bayesian_inference.tex
@@
 \centering
-\includegraphics{../figures/si/figS0X_ppc.pdf}
+\includegraphics{../../figures/si/figS0X_ppc.pdf}
 \caption{\textbf{Theory-data comparison of inference on unregulated promoters.}
@@
 \centering
-\includegraphics{../figures/si/figS0X_aTc_circuit.pdf}
+\includegraphics{../../figures/si/figS0X_aTc_circuit.pdf}
 \caption{\textbf{aTc controlled expression of LacI repressor.} Schematic of the
@@
 so no inference was possible.
+
+As suggested by one of our reviewers, in order for readers to judge the
+agreement between our predictions and the experimental data for the regulated
+case, we include Figure~\ref{figS:eg_histograms} that plot the same data as
+Figure~\ref{fig4:repressed_post_full}(C), but rather than showing ECDF, we show
+histograms of the individual distributions.
+
+\begin{figure}[p]
+\centering
+\includegraphics[width=\textwidth]{../../figures/si/figS0X_histograms.pdf}
+\caption{\textbf{Theory-experiment comparision of mRNA distributions for
+regulated promoters.} Comparison of the inference (color shaded area
+encloses the 95\% of all posterior predictive check samples while the color
+solid line represents the median of all samples) vs the experimental
+measurements (black lines) for different regulated promoters with different
+operators (columns) and aTc concentrations (rows) from
+Ref.~\cite{Jones2014}.}
+\label{figS:reg_histograms}
+\end{figure}

### 2020-09-24 doc/resubmission/section_01_introduction.tex
implement Rob's comments
update resubmission version of paper

@@
 Shea1985, Buchler2003, Vilar2003a, Vilar2003b, Bintu2005a, Bintu2005c,
-Gertz2009, Sherman2012, Saiz2013, Ko1991, Peccoud1995, Record1996, Kepler2001,
-Sanchez2008, Shahrezaei2008, Sanchez2011, Michel2010, Iyer-Biswas2009}. One
-route to such predictive understanding is to focus on the simplest regulatory
-architecture and to push the theory-experiment dialogue to increase the
-predictive power of our theoretical models~\cite{Garcia2011, Phillips2019}. If
-we demonstrate that we can pass that test by successfully predicting both the
-means and variance in gene expression at the mRNA level, then that provides a
-more solid foundation upon which to launch into more complex problems - for
-instance, some of the previously unknown architectures uncovered
-in~\cite{Belliveau2018} and~\cite{Ireland2020}.
+Kuhlman2007, Gertz2009, Sherman2012, Saiz2013, Ko1991, Peccoud1995, Record1996,
+Kepler2001, Sanchez2008, Shahrezaei2008, Sanchez2011, Michel2010,
+Iyer-Biswas2009}. One route to such predictive understanding is to focus on the
+simplest regulatory architecture and to push the theory-experiment dialogue to
+increase the predictive power of our theoretical models~\cite{Garcia2011,
+Phillips2019}. If we demonstrate that we can pass that test by successfully
+predicting both the means and variance in gene expression at the mRNA level,
+then that provides a more solid foundation upon which to launch into more
+complex problems - for instance, some of the previously unknown architectures
+uncovered in~\cite{Belliveau2018} and~\cite{Ireland2020}.
 
@@
 as effective two-state models for which there is a rich
-literature~\cite{Peccoud1995,Bintu2005a,Shahrezaei2008,Iyer-Biswas2009,Sanchez2013,Jones2014}.
-Here, we systematically compare the predictions of several models with different
-levels of coarse graining written in terms of thermodynamic and kinetic
-parameters. One goal in exploring such coarse-grainings is to build towards the
-future models of regulatory response that will be able to serve the powerful
-predictive role needed to take synthetic biology from a brilliant exercise in
-enlightened empiricism to a rational design framework as in any other branch of
-engineering. More precisely, we want phenomenology in the sense of
-coarse-graining away atomistic detail, but still retaining biophysical meaning.
-In particular a key question is: at this level of coarse-graining, what
-microscopic details do we need to explicitly model, and how do we figure that
-out? For example, do we need to worry about all or even any of the steps that
-individual RNA polymerases go through each time they make a transcript? Turning
-the question around, can we see any imprint of those processes in the available
-data? If the answer is no, then those processes are irrelevant for our purposes.
-Forward modeling and inverse (statistical inferential) modeling are necessary to
-tackle such questions. We combine both approaches in order to discard models 
-that cannot empirically satisfy the main features of experimental data. First we
-apply forward modeling to demonstrate that none of the models are 
-distinguishable at the level of mean gene expression. We then extend the 
-modeling to look at higher moments of the distribution, eliminating models that
-do not empirically satisfy the observed cell-to-cell variability. Finally we 
-arrive to a minimal model on which we can apply inverse modeling in order to 
-infer the parameters that explain the data.
+literature~\cite{Peccoud1995, Shahrezaei2008, Iyer-Biswas2009, Tkacik2009,
+Sanchez2013, Jones2014, So2011, Munsky2012}. Here, we systematically compare the
+predictions of several models with different levels of coarse graining written
+in terms of thermodynamic and kinetic parameters. One goal in exploring such
+coarse-grainings is to build towards the future models of regulatory response
+that will be able to serve the powerful predictive role needed to take synthetic
+biology from a brilliant exercise in enlightened empiricism to a rational design
+framework as in any other branch of engineering. More precisely, we want
+phenomenology in the sense of coarse-graining away atomistic detail, but still
+retaining biophysical meaning. In particular a key question is: at this level of
+coarse-graining, what microscopic details do we need to explicitly model, and
+how do we figure that out? For example, do we need to worry about all or even
+any of the steps that individual RNA polymerases go through each time they make
+a transcript? Turning the question around, can we see any imprint of those
+processes in the available data? If the answer is no, then those processes are
+irrelevant for our purposes. Forward modeling and inverse (statistical
+inferential) modeling are necessary to tackle such questions. We combine both
+approaches in order to discard models that cannot empirically satisfy the main
+features of experimental data. First we apply forward modeling to demonstrate
+that none of the models are distinguishable at the level of mean gene
+expression. We then extend the modeling to look at higher moments of the
+distribution, eliminating models that do not empirically satisfy the observed
+cell-to-cell variability. Finally we arrive at a minimal model on which we can
+apply inverse modeling in order to infer the parameters that explain the data.
 
@@
 underlying functional form for the fold-change in the average level of gene
-expression as shown in Figure~\ref{fig1:means_cartoons}(D).
-Section~\ref{sec:beyond_means} goes beyond an analysis of the mean gene
-expression by asking how the same models presented in
-Figure~\ref{fig1:means_cartoons}(C) can be used to explore noise in gene
+expression with an effective free energy $\Delta F_R$ capturing the regulation
+given by the transcription factor, and a term $\rho$ describing the level of
+coarse-graining of the transcriptional events as shown in
+Figure~\ref{fig1:means_cartoons}(D). Section~\ref{sec:beyond_means} goes beyond
+an analysis of the mean gene expression by asking how the same models presented
+in Figure~\ref{fig1:means_cartoons}(C) can be used to explore noise in gene
 expression. To make contact with experiment, all of these models must make a

### 2020-09-24 doc/resubmission/section_03_beyond_means.tex
update resubmission version of paper

@@
 transcriptional bursting, as observed in,
-e.g.,~\cite{Golding2005,Chong2014,Sevier2016,Sepulveda2016}. Let us make this
-correspondence more precise. The mean dwell time in the active state is $1/k^-$.
-While in this state, transcripts are produced at a rate $r$ per unit time. So on
-average, $r/k^-$ transcripts are produced before the system switches to the
-inactive state. Once in the inactive state, the system dwells there for an
-average time $1/k^+$ before returning to the active state and repeating the
+e.g.,~\cite{Golding2005,Chong2014,Senecal2014,Sevier2016,Sepulveda2016}. Let us
+make this correspondence more precise. The mean dwell time in the active state
+is $1/k^-$. While in this state, transcripts are produced at a rate $r$ per unit
+time. So on average, $r/k^-$ transcripts are produced before the system switches
+to the inactive state. Once in the inactive state, the system dwells there for
+an average time $1/k^+$ before returning to the active state and repeating the
 process. $r/k^-$ resembles an average burst size, and $1/k^+$ resembles the time

### 2020-09-24 doc/resubmission/section_05_discussion.tex
update resubmission version of paper

--- 2020-06-06 doc/section_05_discussion.tex
@@
 hopeful approximations or biological naivete \cite{Razo-Mejia2018, Chure2019}.
+The fact that thermodynamic and kinetic models for the simple repression motif
+can not only be reconciled, but are indistinguishable from each other at the
+level of mean mRNA count is a consequence of the possible transitions between
+promoter microstates. The promoter microstates in all models shown in
+Figure~\ref{figure1:means_cartoons} can explicitly be separated into two groups:
+1) promoters with repressor bound, and 2) all other promoters states. The single
+path between these two groups of states guarantees detailed-balance between the
+groups. We can therefore separate \textit{expression} related transitions, which
+by definition must be out of equilibrium, with \textit{regulation} transitions
+that can in principle be in equilibrium. This implies that the probability
+distribution of the promoter microstates not involving the repressor take the
+same functional form, allowing us to write the fold-change as an effective free
+energy involving a regulation term, $\Delta F_R$, and a promoter details term,
+$\rho$. We offer this argument as a conjecture, and we suspect that a careful
+argument using the graph-theoretic framework based on the Matrix-Tree
+Theorem~\cite{Gunawardena2012}, might furnish a ``proof'' not with some 
+challenges as this theoretical framework applies, at the present, to finite 
+graphs.
+
 To further dissect the relative merits of the different models, we must appeal
@@
 Fano factors (variance/mean) that are less than one, at odds with any and all
-experimental data that we are aware of~\cite{So2011, Jones2014}. This theoretical
-result allows us to directly discard a subset of the models (models 1-3 in
-Figure~\ref{fig2:constit_cartoons}(A)) since they cannot be reconciled with
-experimental observations. The two remaining models (models 4 and 5 in
+experimental data that we are aware of~\cite{So2011, Jones2014}. This
+theoretical result allows us to directly discard a subset of the models (models
+1-3 in Figure~\ref{fig2:constit_cartoons}(A)) since they cannot be reconciled
+with experimental observations. The two remaining models (models 4 and 5 in
 Figure~\ref{fig2:constit_cartoons}) appear to contain enough microscopic realism
 to be able to reproduce the data. A previous exploration of model 4 demonstrated
-the ``sloppy''~\cite{Transtrum2015} nature of the model
-in which data on single-cell mRNA counts alone
-cannot constrain the value of all parameters simultaneously
-\cite{Razo-Mejia2020}. Here we demonstrate that the proposed one-state bursty
-promoter model (model 5 in Figure~\ref{fig2:constit_cartoons}(A)) emerges as a
-limit of the commonly used two-state promoter model \cite{Peccoud1995,
-Shahrezaei2008, So2011, Sanchez2013, Jones2014}. We put the idea to the test
-that this level of coarse-graining is rich enough to reproduce previous
-experimental observations. In particular we perform Bayesian inference to
-determine the two parameters describing the full steady-state mRNA distribution,
-finding that the model is able to provide a quantitative description of a
-plethora of promoter sequences with different mean levels of expression and
-noise.
+the ``sloppy''~\cite{Transtrum2015} nature of the model in which data on
+single-cell mRNA counts alone cannot constrain the value of all parameters
+simultaneously \cite{Razo-Mejia2020}. Here we demonstrate that the proposed
+one-state bursty promoter model (model 5 in
+Figure~\ref{fig2:constit_cartoons}(A)) emerges as a limit of the commonly used
+two-state promoter model \cite{Peccoud1995, Shahrezaei2008, So2011, Sanchez2013,
+Jones2014}. We put the idea to the test that this level of coarse-graining is
+rich enough to reproduce previous experimental observations. In particular we
+perform Bayesian inference to determine the two parameters describing the full
+steady-state mRNA distribution, finding that the model is able to provide a
+quantitative description of a plethora of promoter sequences with different mean
+levels of expression and noise.
 
@@
 of self-consistent parameters, demonstrating not only a predictive framework,
-but also reconciling the equilibrium and non-equilibrium views of the same
-simple repression constructs. More specifically, we obtained values for the
+but also reconciling the thermodynamic and kinetic views of the same simple
+repression constructs. More specifically, we obtained values for the
 transcription factor association and dissociation rates by performing Bayesian
@@
 such transcription factors for the binding site. The free energy value obtained
-from these kinetic rates -- computed as the log ratio of the rates -- agrees
-with previous inferences performed only from mean gene expression measurements,
-that assumed an equilibrium rather than a kinetic
-framework~\cite{Garcia2011a, Razo-Mejia2018}.
+from these kinetic rates --computed as the log ratio of the rates-- agrees with
+previous inferences performed only from mean gene expression measurements, that
+assumed a thermodynamic rather than a kinetic
+framework~\cite{Garcia2011a,Razo-Mejia2018}. Interestingly, to constraint the
+binding and unbinding rates inferred for Figure~\ref{fig4:repressed_post_full},
+all of the rates need to be simultaneously fit. This is because a single
+combination of operator (repressor binding site) and aTc concentration
+(repressor copy number) cannot constraint both rates independently, but only
+their ratio (See Appendix~\ref{sec:bayesian} for further details).
 
-It is interesting to speculate what microscopic details are being coarse-grained
-by our burst rate and burst size in Figure~\ref{fig2:constit_cartoons}, model 5.
+It is important to mention the potential problems present in all models,
+including model 5 from Figure~\ref{fig2:constit_cartoons}(A). Irreversible
+reactions as the ones used in all transcription events lead to physical
+paradoxes such as infinite entropy production rates. These are mathematical
+conveniences rather than physical reality, leaving the inferred rates for the
+transcription burst rate and burst size as an effective rate with lots of
+(potentially) reversible processes. Also, the state of the art modeling imagines
+the promoter existing in a series of discrete states. This assumption might not
+be entirely true as effects such as DNA supercoiling~\cite{Chong2014,Sevier2016}
+or DNA looping~\cite{Boedicker2013a} can be considered continuous
+variables; although there are formal ways of integrating out such degrees of
+freedom. Furthermore, when computing the fold-change in gene expression
+(Section~\ref{section_02_means}) we assume that the translation efficiency of
+mRNA transcript is equal in cells with and without regulation. There might be
+some known biological effects that could limit the applicability of this
+assumption such as cooperation between the transcription and translation 
+machinery~\cite{Proshkin2010}, or coupling of mRNA transcription and 
+degradation~\cite{Chen2015} among other potential effects. Another implicit 
+assumption in all of the models discussed here is that the effects due to 
+changes in gene copy number during the cell cycle have no effects on the 
+expression profile. For moderate growth rates as the ones used 
+in~\cite{Jones2014} from where we obtained the data, there are at any point at
+most two copies of the gene. On a recent publication, we explored the effects
+that this change in gene dosage can have in gene expression, finding that the
+effects are much more drastic at the protein level~\cite{Razo-Mejia2020}.
+
+More specifically, for model 5 from Figure~\ref{fig2:constit_cartoons}(A), it is
+interesting to speculate what microscopic details are being coarse-grained by
+our burst rate and burst size in Figure~\ref{fig2:constit_cartoons}(A), model 5.
 Chromosomal locus is one possible influence we have not addressed in this work,
@@
 the difference also be due to the different chromosomal locations of the two
-operons?
-What features of different loci are and are not important?
-Could our preferred coarse-grained model capture the variability across
-different loci? If so, and we were to repeat the parameter inference
-as done in this work, is there a simple theoretical model we could build
-to understand the resulting parameters?
+operons? What features of different loci are and are not important? Could our
+preferred coarse-grained model capture the variability across different loci? If
+so, and we were to repeat the parameter inference as done in this work, is there
+a simple theoretical model we could build to understand the resulting
+parameters?
 
@@
 established the simple-repression motif as an arena for the quantitative
-dissection of regulatory response in bacteria, with similar beautiful work
-emerging in examples such as the \textit{ara} and \textit{gal} operons as
+dissection of regulatory response in bacteria, with similar work emerging in
+examples such as the \textit{ara} and \textit{gal} operons as
 well~\cite{Dunn1984b, Oehler1990, Weickert1993, Oehler1994, Schleif2000,
@@
 them can now be formulated as a rigorous quantitative question. In particular,
-we believe understanding of the simple repression motif has 
-advanced sufficiently that the design of new
-versions of the architecture is now possible,
-based upon predictions about how repressor copy
-number and DNA binding site strength control expression. In our view, the next
-step in the progression is to first perform similar rigorous analyses of the
-fundamental ``basis set'' of regulatory architectures. Natural follow-ups to this
-work are explorations of motifs such as simple activation that is
-regulated by a single activator binding site, and the repressor-activator
-architecture, mediated by the binding of both a single activator and a single
-repressor, and beyond. With the individual input-output functions in hand,
-similar quantitative dissections including the rigorous analysis of their tuning
-parameters can be undertaken for the ``basis set'' of full gene-regulatory
-networks such as switches, feed-forward architectures and oscillators for
-example, building upon the recent impressive bonanza of efforts from systems
-biologists and synthetic biologists~\cite{Milo2002, Alon2007}.
+we believe understanding of the simple repression motif has advanced
+sufficiently that the design of new versions of the architecture is now
+possible, based upon predictions about how repressor copy number and DNA binding
+site strength control expression. In our view, the next step in the progression
+is to first perform similar rigorous analyses of the fundamental ``basis set''
+of regulatory architectures. Natural follow-ups to this work are explorations of
+motifs such as simple activation that is regulated by a single activator binding
+site, and the repressor-activator architecture, mediated by the binding of both
+a single activator and a single repressor, and beyond. With the individual
+input-output functions in hand, similar quantitative dissections including the
+rigorous analysis of their tuning parameters can be undertaken for the ``basis
+set'' of full gene-regulatory networks such as switches, feed-forward
+architectures and oscillators for example, building upon the recent impressive
+efforts from systems biologists and synthetic biologists~\cite{Milo2002,
+Alon2007}.

### 2020-09-24 doc/resubmission/appendix_01_nonbursty.tex
add thermodynamic models section to SI

@@
 
-\mrm{THIS NEEDS TO BE FIXED}
 \subsection{Thremodynamic models of gene regulation}
@@
 expression level is proportional to the probability of finding the promoter in
-the transcriptionally active microstate. As mentioned in the main text, these
-models can only deal with the mean gene expression level. This is because the
-probability space we consider is that of the state of the repressor, rather than
-both the state of the repressor and the mRNA counts. Let's now work through
-models 1 and 2 from Figure~\ref{fig1:means_cartoons}(B).
+the transcriptionally active microstate, i.e.,
+\begin{equation}
+\text{average rate of transcription} = \sum_i r_i p_i,
+\label{eq:rate_transcription_si}
+\end{equation}
+where $i$ labels the distinct states, $p_i$ is the probability of the ith state,
+and $r_i$ is the rate of transcription of that state. As mentioned in the main
+text, these models can only deal with the mean gene expression level. This is
+because the probability space we consider is that of the state of the repressor,
+rather than both the state of the repressor and the mRNA counts. Let's now work
+through models 1 and 2 from Figure~\ref{fig1:means_cartoons}(B).
 
@@
 From the relative statistical weights listed in
-Figure~\ref{fig1:means_cartoons}, the probability $p_U$ of being in the
+Figure~\ref{fig1:means_cartoons}(B), the probability $p_U$ of being in the
 unbound state is
@@
 The mean rate of transcription is then given by $r p_U$, as assumed by
-Eq.~\ref{eq:transcrip_prop_pbound}. The mean number of mRNA is
+Eq.~\ref{eq:rate_transcription_si}. The mean number of mRNA is
 set by the balance of average mRNA transcription and degradation rates,
@@
 is the mRNA degradation rate, $R$ is repressor copy number, $N_{NS}$ is the
-number of nonspecific binding sites in the genome where repressors spend most
-of their time when not bound to the operator, $\beta \equiv 1/k_BT$, and
-$\Delta\varepsilon_R$ is the binding energy of a repressor to its operator site.
-The derivation of this result is deferred to Appendix~\ref{sec:non_bursty}.
+number of nonspecific binding sites in the genome where repressors spend most of
+their time when not bound to the operator (binding site), $\beta \equiv 1/k_BT$,
+and $\Delta\varepsilon_R$ is the binding energy of a repressor to its operator
+site. The derivation of this result can be found in~\cite{Phillips2015a}.
 
@@
 transcriptionally active states have transcription factor bound, or put
-differently, when there is no overlap between
-transcription factor bound states and transcriptionally
-active states. We will see this explicitly in the 3-state equilibrium model
-below, but perhaps surprising is that an analogous result carries over even to
-the nonequilibrium models we consider later.
+differently, when there is no overlap between transcription factor bound states
+and transcriptionally active states. We will see this explicitly in the 3-state
+equilibrium model below, but perhaps surprising is that an analogous result
+carries over even to the nonequilibrium models we consider later.
 
@@
 as we use it here, and in~\cite{Razo-Mejia2018} and~\cite{Chure2019} it was
-generalized to incorporate small-molecule inducers that bind the repressor. 
-The effect of this generalization is, roughly speaking,
-simply to rescale $R$ from the total number of repressors to a smaller
-effective number of available repressors which are unbound by inducers.
-We point out that the same generalization can be incorporated quite easily
-into any of our models in Figure~\ref{fig1:means_cartoons} by simply
-rescaling the repressor copy number $R$ in the equilibrium models, or
-equivalently $k_R^+$ in the nonequilibrium models.
+generalized to incorporate small-molecule inducers that bind the repressor. The
+effect of this generalization is, roughly speaking, simply to rescale $R$ from
+the total number of repressors to a smaller effective number of available
+repressors which are unbound by inducers. We point out that the same
+generalization can be incorporated quite easily into any of our models in
+Figure~\ref{fig1:means_cartoons} by simply rescaling the repressor copy number
+$R$ in the thermodynamic models, or equivalently $k_R^+$ in the kinetic models.
 
@@
 kinetic models fare when faced with computing the same observable.
-\mrm{ENDS HERE}
 

### 2020-09-24 doc/resubmission/appendix_01_nonbursty.tex
finished transfering main text to SI for appendix 1

@@
 
-\mrm{THIS NEEDS TO BE FIXED}
-
-Before jumping into the derivations of the general
-computation of the mean mRNA level and the Fano factor we will work through the
-derivation of an example master equation. In particular we will focus on model 1
-from Figure~\ref{fig1:means_cartoons}(C). The general steps are applicable to
-all other chemical master equations in this work.
-
-\subsubsection{The Poisson Promoter Nonequilibrium Model}
-For our first kinetic model, we  imitate the states considered in the Two-State
-Equilibrium Model and consider the simplest possible picture with only two
-states, repressor bound and unbound. This is exactly the model used for the main
-results of~\cite{Jones2014}. In this picture, repressor association and
-dissociation rates from its operator site, $k_R^+$ and $k_R^-$, respectively,
-govern transitions between the two states. When the system is in the unbound
-state, transcription initiates at rate $r$, which represents a coarse-graining
-of all the downstream processes into a single effective rate. mRNA is degraded
-at rate $\gamma$ as already exploited in the previous models.
-
-Let $p_R(m,t)$ denote the joint probability of finding the system in the
-repressor bound state $R$ with $m$ mRNA molecules present at time $t$. Similarly
-define $p_U(m,t)$ for the repressor unbound state $U$. This model is governed
-by coupled master equations giving the time evolution of $p_R(m,t)$ and
-$p_U(m,t)$~\cite{Sanchez2008, Sanchez2011, Phillips2019} which we can write as
-\begin{align}
-\begin{split}
-\deriv{t}p_R(m,t) =& 
-- \overbrace{k_R^- p_R(m,t)}^{R \rightarrow U}
-+ \overbrace{k_R^+ p_U(m,t)}^{U \rightarrow R}
-+ \overbrace{(m+1)\gamma p_R(m+1,t)}^{m + 1 \rightarrow m}
-- \overbrace{\gamma p_R(m,t)}^{m \rightarrow m - 1}
-\\
-\deriv{t}p_U(m,t) =&\; 
-\overbrace{k_R^- p_R(m,t)}^{R \rightarrow U}
-- \overbrace{k_R^+ p_U(m,t)}^{U \rightarrow R}
-+ \overbrace{rp_U(m-1,t)}^{m-1 \rightarrow m}
-- \overbrace{rp_U(m,t)}^{m \rightarrow m + 1}
-\\
-&+ \overbrace{(m+1)\gamma p_U(m+1,t)}^{m + 1 \rightarrow m}
-- \overbrace{\gamma p_U(m,t)}^{m \rightarrow m - 1},
-\label{eq:poisson_promoter_cme}
-\end{split}
-\end{align}
-where each term on the right corresponds to a transition between two states of
-the promoter as indicated by the overbrace label. In each equation, the first
-two terms describe transitions between promoter states due to repressors
-unbinding and binding, respectively. The final two terms describe degradation of
-mRNA, decreasing the copy number by one, and the terms with coefficient $r$
-describe transcription initiation increasing the mRNA copy number by one.
-We direct the reader to Appendix~\ref{sec:cme_from_cartoon} for a careful
-treatment showing how the form of this master equation follows from the
-corresponding cartoon in Figure~\ref{fig1:means_cartoons}.
-
-We can greatly simplify the notation, which will be especially useful for the
-more complicated models yet to come, by re-expressing the master equation in
-vector form~\cite{Phillips2012}. The promoter states are collected into a vector
-and the rate constants are collected into matrices as
-\begin{equation}
-\vec{p}(m) = \begin{pmatrix} p_R(m) \\ p_U(m) \end{pmatrix},\
-\mathbf{K} = \begin{pmatrix} -k_R^- & k_R^+ \\ k_R^- & -k_R^+ \end{pmatrix},\
-\mathbf{R} = \begin{pmatrix} 0 & 0 \\ 0 & r \end{pmatrix},\
-\label{eq:2state_cme_matrices}
-\end{equation}
-so that the master equation may be condensed as
-\begin{equation}
-\deriv{t}\vec{p}(m,t) =
-\left( \mathbf{K} - \mathbf{R} - \gamma m \mathbf{I} \right) \vec{p}(m,t)
-                + \mathbf{R} \vec{p}(m-1,t)
-                + \gamma (m+1) \mathbf{I} \vec{p}(m+1,t),
-\label{eq:2state_rep_cme}
-\end{equation}
-where $\mathbf{I}$ is the identity matrix. Taking steady state by setting time
-derivatives to zero, the mean mRNA can be found to be
-\begin{equation}
-\langle m \rangle = \frac{r}{\gamma}
-        \left(1 + \frac{k_R^+}{k_R^-}\right)^{-1},
-\label{eq:mean_m_model1}
-\end{equation}
-with the algebra details again deferred to Appendix~\ref{sec:non_bursty}. Recall
-$k_R^+$ is proportional to the repressor copy number, so in computing
-fold-change, absence of repressor corresponds to $k_R^+\rightarrow0$. Therefore
-fold-change in this model is simply
-\begin{equation}
-\text{fold-change} = \left(1 + \frac{k_R^+}{k_R^-}\right)^{-1},
-\end{equation}
-again matching the master curve of~\fig{fig1:means_cartoons}(D) with $\rho=1$.
-
-\subsubsection{Nonequilibrium Model Two - RNAP Bound and Unbound States}
-Our second kinetic model depicted in Figure~\ref{fig1:means_cartoons}(C) mirrors
-the second equilibrium model of Figure~\ref{fig1:means_cartoons}(B) by
-fine-graining  the repressor unbound state of nonequilibrium model 1, resolving
-it into an empty promoter state and an RNAP-bound state. Note in this picture,
-in contrast with model 4 below, transcription initiation is accompanied by a
-promoter state change, in keeping with the interpretation as RNAP-bound and
-empty states: if an RNAP successfully escapes the promoter and proceeds to
-elongation of a transcript, clearly it is no longer bound at the promoter.
-Therefore another RNAP must bind before another transcript can be initiated.
-
-The master equation governing this model is analogous
-to Eqs.~\ref{eq:2state_cme_matrices}-\ref{eq:2state_rep_cme} for model 1 above. The
-main subtlety arises since transcription initiation accompanies a promoter state
-change. This can be understood by analogy to $\mathbf{K}$. The off-diagonal and
-diagonal elements of $\mathbf{K}$ correspond to transitions arriving at or
-departing from, respectively, the promoter state of interest. If transcription
-initiation is accompanied by promoter state changes, we must have separate
-matrices for arriving and departing transcription events since the arriving and
-departing transitions have different initial copy numbers of mRNA, unlike for
-$\mathbf{K}$ where they are the same (see Appendix~\ref{sec:non_bursty}). The
-master equation for this model is
-\begin{equation}
-\deriv{t}\vec{p}(m,t) =
-\left( \mathbf{K} - \mathbf{R_D} - \gamma m \mathbf{I} \right) \vec{p}(m,t)
-                + \mathbf{R_A} \vec{p}(m-1,t) +
-                \gamma (m+1) \mathbf{I} \vec{p}(m+1,t),
-\label{eq:3state_rep_cme}
-\end{equation}
-with the state vector and promoter transition matrix defined as
-\begin{equation}
-\vec{p}(m) = \begin{pmatrix} p_R(m) \\ p_E(m) \\ p_P(m) \end{pmatrix},\
-\mathbf{K} = \begin{pmatrix} -k_R^- & k_R^+ & 0 \\
-                        k_R^- & -k_R^+ -k_P^+ & k_P^- \\
-                        0 & k_P^+ & -k_P^- 
-                \end{pmatrix},
-\label{eq:3state_cme_matrices_pt1}
-\end{equation}
-and the initiation matrices given by
-\begin{equation}
-\mathbf{R_A} = \begin{pmatrix}
-                0 & 0 & 0 \\ 
-                0 & 0 & r \\ 
-                0 & 0 & 0
-                \end{pmatrix},\
-\mathbf{R_D} = \begin{pmatrix}
-                0 & 0 & 0 \\ 
-                0 & 0 & 0 \\ 
-                0 & 0 & r
-                \end{pmatrix}.
-\label{eq:3state_cme_matrices_pt2}
-\end{equation}
-The elements of $\vec{p}(m)$ encode the probabilities of having $m$ mRNA present
-along with the promoter having repressor bound ($R$), being empty ($E$), or
-having RNAP bound ($P$), respectively. $\mathbf{R_A}$ describes probability flux
-\textit{arriving} at the state $\vec{p}(m)$ from a state with one fewer mRNA,
-namely $\vec{p}(m-1)$, and $\mathbf{R_D}$ describes probability flux
-\textit{departing} from the state $\vec{p}(m)$ for a state with one more mRNA,
-namely $\vec{p}(m+1)$. $\mathbf{K}$ is closely analogous to model 1.
-
-Mean mRNA at steady state is found analogously to model 1, with the result
-\begin{equation}
-\langle m\rangle = \frac{r}{\gamma}
-        \frac{k_R^- k_P^+}
-        {k_R^- k_P^+ + k_R^- (k_P^- + r) + k_R^+ (k_P^- + r)},
-\label{eq:model2_meanm}
-\end{equation}
-and with details again deferred to Appendix~\ref{sec:non_bursty}.
-Fold-change is again found from the ratio prescribed by Eq.~\ref{eq:fc_def}, from
-which we have
-\begin{align}
-\text{fold-change}
-&=      \frac{k_R^- k_P^+}
-        {k_R^- k_P^+ + k_R^- (k_P^- + r) + k_R^+ (k_P^- + r)}
-        \frac{k_P^+ + k_P^- + r}{k_P^+}
-\\
-&=      \left(1 + \frac{k_R^+}{k_R^-}
-                \frac{k_P^- + r}{k_P^+ + k_P^- + r}
-        \right)^{-1}
-\\
-&=      \left(1 + \frac{k_R^+}{k_R^-}
-        \left(1 + \frac{k_P^+}{k_P^- + r}\right)^{-1}
-        \right)^{-1},
-\end{align}
-which follows the master curve of~\fig{fig1:means_cartoons}(D) with $\rho = 1 +
-k_P^+/(k_P^- + r)$ as claimed.
-
-\subsubsection{Nonequilibrium Model Three - Multistep Transcription Initiation
-and Escape}
-One might reasonably complain that the first two ``nonequilibrium'' models we
-have considered are straw men. Their steady states necessarily satisfy detailed
-balance which is equivalent to thermodynamic equilibrium. Why is this the case?
-At steady state there is by definition no net probability flux in or out of each
-promoter state, but since the promoter states form a linear chain, there is only
-one way in or out of the repressor bound and RNAP bound states, implying each
-edge must actually have a net zero probability flux, which is the definition of
-detailed balance (usually phrased as equality of forward and reverse transition
-fluxes).
-
-Now we consider model 3 in Figure~\ref{fig1:means_cartoons}(C) which allows the
-possibility of true nonequilibrium steady-state fluxes through the promoter
-states. We point out that this model was considered previously
-in~\cite{Mitarai2015} where a comparison was made with model 1 as used
-in~\cite{Jones2014}. The authors of~\cite{Mitarai2015} argued that the
-additional complexity is essential to properly account for the noise in the mRNA
-distribution. We will weigh in on both models later when we consider observables
-beyond fold-change.
-
-The master equation governing this model is identical in form to model 2 above,
-namely
-\begin{equation}
-\deriv{t}\vec{p}(m,t) =
-\left( \mathbf{K} - \mathbf{R_D} - \gamma m \mathbf{I} \right) \vec{p}(m,t)
-                + \mathbf{R_A} \vec{p}(m-1,t) +
-                \gamma (m+1) \mathbf{I} \vec{p}(m+1,t),
-\end{equation}
-but with a higher-dimensional state space and different matrices. The
-state vector and promoter transition matrix are now
-\begin{equation}
-\vec{p}(m) = \begin{pmatrix} p_R(m) \\ p_E(m) \\
-                             p_C(m) \\ p_O(m)\end{pmatrix},\
-\mathbf{K} = \begin{pmatrix} -k_R^- & k_R^+ & 0 & 0\\
-                        k_R^- & -k_R^+ -k_P^+ & k_P^- & 0 \\
-                        0 & k_P^+ & -k_P^- - k_O & 0 \\
-                        0 & 0 & k_O & 0
-                \end{pmatrix},
-\end{equation}
-with the four promoter states, in order, being repressor bound ($R$), empty
-($E$), RNAP closed complex ($C$), and RNAP open complex ($O$). Besides
-increasing dimension by one, the only new feature in $\mathbf{K}$ is the
-rate $k_O$, representing the rate of open complex formation from
-the closed complex, which we assume for simplicity to be irreversible
-in keeping with some~\cite{Mitarai2015} but not all~\cite{DeHaseth1998}
-past literature. The initiation matrices are given by
-\begin{equation}
-\mathbf{R_A} = \begin{pmatrix}
-        0 & 0 & 0 & 0 \\ 
-        0 & 0 & 0 & r \\ 
-        0 & 0 & 0 & 0 \\ 
-        0 & 0 & 0 & 0
-                \end{pmatrix},\
-\mathbf{R_D} = \begin{pmatrix}
-        0 & 0 & 0 & 0 \\ 
-        0 & 0 & 0 & 0 \\ 
-        0 & 0 & 0 & 0 \\ 
-        0 & 0 & 0 & r
-                \end{pmatrix},
-\end{equation}
-again closely analogous to nonequilibrium model 2.
-
-The expression for mean mRNA is substantially more complicated now, as worked
-out in Appendix~\ref{sec:non_bursty} where we find
-\begin{equation}
-\langle m\rangle = \frac{r}{\gamma}
-        \frac{k_R^- k_P^+ k_O}
-        {k_R^- [(k_P^+ (k_O + r) + r(k_P^- + k_O)] + k_R^+ r(k_P^- + k_O)},
-\label{eq:model3_mean_m}
-\end{equation}
-which can be simplified to
-\begin{equation}
-\langle m\rangle
-= \frac{r}{\gamma}
-\frac{\frac{k_P^+ k_O}{r(k_O + k_P^-)}}
-        {1 + \frac{k_P^+ (k_O + r)}{r(k_O + k_P^-)} + \frac{k_R^+}{k_R^-}}.
-\end{equation}
-The strategy is to isolate the terms involving the repressor, so that now the
-fold-change is seen to be simply
-\begin{align}
-\text{fold-change}
-&= \frac{\frac{k_P^+ k_O}{r(k_O + k_P^-)}}
-        {1 + \frac{k_P^+ (k_O + r)}{r(k_O + k_P^-)} + \frac{k_R^+}{k_R^-}}
-        \frac{1 + \frac{k_P^+ (k_O + r)}{r(k_O + k_P^-)}}
-                {\frac{k_P^+ k_O}{r(k_O + k_P^-)}}
-\\
-&= \left(
-        1 + \frac{k_R^+}{k_R^-}
-        \left(1 + \frac{k_P^+ (k_O + r)}{r(k_O + k_P^-)}\right)^{-1}
-\right)^{-1},
-\end{align}
-surprisingly reducing to the master curve of~\fig{fig1:means_cartoons}(D) once
-again, with $\rho = 1 + \frac{k_P^+ (k_O + r)}{r(k_O + k_P^-)}$.
-
-This example hints that an arbitrarily fine-grained model of downstream
-transcription steps may still be collapsed to the form of the master curve for
-the means given in Figure~\ref{fig1:means_cartoons}(D), so long as the repressor
-binding is exclusive with transcriptionally active states. We offer this as a
-conjecture, and we suspect that a careful argument using the King-Altman diagram
-method~\cite{King1956, Hill1966} might furnish a ``proof.'' Our focus here is
-not on full generality but rather to survey an assortment of plausible models
-for simple repression  that have been proposed in the literature.
-
-\subsubsection{Nonequilibrium Model Four - ``Active'' and ``Inactive'' States}
-Model 4 in Figure~\ref{fig1:means_cartoons}(C) is at the core of the theory
-in~\cite{Razo-Mejia2020}. At a glance the cartoon for this model may appear very
-similar to model 2, and mathematically it is, but the interpretation is rather
-different. In model 2, we interpreted the third state literally as an RNAP-bound
-promoter and modeled initiation of a transcript as triggering a promoter state
-change, making the assumption that an RNAP can only
-make one transcript at a time. In contrast, in the present model the promoter
-state does \textit{not} change when a transcript is initiated. So we no longer
-interpret these states as literally RNAP bound and unbound but instead as
-coarse-grained ``active'' and ``inactive'' states, the details of which we leave
-unspecified for now. We will comment more on this model below when we discuss
-Fano factors of models.
-
-Mathematically this model is very similar to models 1 and 2. Like model 1, the
-matrix $R$ describing transcription initiation is diagonal, namely
-\begin{equation}
-\mathbf{R} = \begin{pmatrix}
-                0 & 0 & 0 \\ 
-                0 & 0 & 0 \\ 
-                0 & 0 & r
-        \end{pmatrix}.
-\end{equation}
-The master equation takes verbatim the same form as it did for model 1,
-Eq.~\ref{eq:2state_rep_cme}. Meanwhile the promoter transition
-matrix $\mathbf{K}$ is the same as Eq.~\ref{eq:3state_cme_matrices_pt1}
-from model 2, although we relabel the
-rate constants from $k_P^\pm$ to $k^\pm$ to reiterate that these are not simply
-RNAP binding and unbinding rates.
-
-Carrying out the algebra, the mean mRNA can be found to be
-\begin{equation}
-\langle m\rangle = \frac{r}{\gamma}
-\frac{k_R^- k^+}
-{k_R^- k^+ + k_R^- k^- + k_R^+ k^-},
-\end{equation}
-and the fold-change readily follows,
-\begin{align}
-\text{fold-change}
-&=      \frac{k_R^- k^+}{k_R^- k^+ + k_R^- k^- + k_R^+ k^-}
-        \frac{k_R^- k^+ + k_R^- k^-}{k_R^- k^+}
-\\
-&=      \left(1 + \frac{k_R^+}{k_R^-}
-                \left(1 + \frac{k^+}{k^-}\right)^{-1}
-        \right)^{-1},
-\end{align}
-from which we see $\rho = 1 + k^+/k^-$ as shown in~\fig{fig1:means_cartoons}(C).
-
-\subsubsection{Nonequilibrium Model Five - Bursty Promoter}
-The final model we consider shown in Figure~\ref{fig1:means_cartoons}(C) is an
-intuitive analog to model 1, with just two states, repressor bound or unbound,
-and transition rates between them of $k_R^+$ and $k_R^-$. In model 1, when in
-the unbound state, single mRNA transcripts are produced as a Poisson process
-with some characteristic rate $r$. The current model by contrast produces, at
-some Poisson rate $k_i$, \textit{bursts} of mRNA transcripts.
-The burst sizes are assumed to be geometrically distributed with
-a mean burst size $b$, which we will motivate in Section~\ref{sec:beyond_means}
-when we derive this model as a certain limiting case of model 4.
-
-From this intuitive picture and by analogy to model 1, then, it should be
-plausible that the mean mRNA level is
-\begin{equation}
-\langle m\rangle = \frac{k_i b}{\gamma}
-        \left(1 + \frac{k_R^+}{k_R^-}\right)^{-1},
-\end{equation}
-which will turn out to be correct from a careful calculation. For now, we simply
-note that just like model 1, the fold-change becomes
-\begin{equation}
-\text{fold-change} = \left(1 + \frac{k_R^+}{k_R^-}\right)^{-1}
-\end{equation}
-with $\rho=1$ also like model 1.
-We will also see later how this model emerges as a natural limit of model 4.
-\mrm{ENDS HERE}
-
-\mrm{THIS NEEDS TO BE FIXED}
-\subsubsection{Noise in the Two-State Promoter, RNAP Bound or Unbound.}
-
-Our second model of constitutive transcription posits an architecture in which
-the promoter is either empty or bound by RNAP ~\cite{Phillips2015a,
-Phillips2019}. Here,  as shown in model 2 of Figure~\ref{fig2:constit_cartoons}(A),
-transcription initiation results in a state transition from the bound to the
-unbound state, reflecting the microscopic reality that an RNAP that has begun to
-elongate a transcript is no longer available at the start site to begin another.
-As shown in Appendix~\ref{sec:non_bursty}, the Fano factor in this model is
-given by
-\begin{align}
-    \nu = 1 -
-        \frac{r\rate{k}{P}{+}}
-            {\left(\rate{k}{P}{+} + \rate{k}{P}{-} + r\right)
-             \left(\gamma + \rate{k}{P}{+} + \rate{k}{P}{-} + r\right)}.
-\label{eq:model2_fano}
-\end{align}
-The problem with this picture is that the Fano factor is always $<1$. To make
- contact with the experimental reality of $\nu>1$ as shown in
- Figure~\ref{fig2:constit_cartoons}(B), clearly some corrections will be needed.
- While this model adds an appealing element of microscopic reality, we are
- forced to reject it as the additional complexity is unable to capture the
- phenomenology of interest. Obviously the promoter state does in fact proceed
- through cycles of RNAP binding, initiating, and elongating, but it seems that
- the super-Poissonian noise in mRNA copy number we want to model must be
- governed by other features of the system.
-
-\subsubsection{Noise in the Three-State Promoter, Multistep Transcription
-Initiation and Escape.}
-
-How might we remedy the deficits of model 2? It is known~\cite{DeHaseth1998}
-that once RNAP initially binds the promoter region, a multitude of distinct
-steps occur sequentially before RNAP finally escapes into the elongation phase.
-Perhaps adding some of this mechanistic detail as shown in model 3 of
-Figure~\ref{fig2:constit_cartoons}(A) might rescue the previous model. The next
-simplest refinement of that model could consider open complex formation and
-promoter escape as separate steps rather than as a single effective step. In
-other words, we construct model 3 by adding a single extra state to model 2, and
-we will label the two RNAP-bound states as the closed and open complexes,
-despite the true biochemical details certainly being more complex. For example,
-earlier work extended this model by adding an additional repressor bound state
-and did not explicitly consider the limit with no repressor that we analyze
-here~\cite{Mitarai2015}. Again, our goal here is not a complete accounting of
-all the relevant biochemical detail; this is an exploratory search for the
-important features that a model needs to include to square with the known
-experimental reality of constitutive expression.
-
-Unfortunately, as hinted at in earlier work~\cite{Mitarai2015}, this model too
-has Fano factor $\nu<1$. We again leave the algebraic details for
-Appendix~\ref{sec:non_bursty} and merely state the result that
-\begin{align}
-\nu = 1 - \frac{r k_P^+ k_O}{\mathcal{Z}}
-\frac{k_P^+ + k_P^- + k_O + r + \gamma}
-    {\mathcal{Z} + \gamma(k_P^+ + k_P^- + k_O + r) + \gamma^2},
-\label{eq:model3_fano}
-\end{align}
-where we defined $\mathcal{Z} = r(k_O + k_P^-) + k_P^+(k_O + r)$ for notational
-tidiness. This is necessarily less than 1 for arbitrary rate constants.
-
-In fact, we suspect \textit{any} model in which transcription proceeds through a
-multistep cycle must necessarily have $\nu<1$. The intuitive argument compares
-the waiting time distribution to traverse the cycle with the waiting time for a
-Poisson promoter (model 1) with the same mean time. The latter is simply an
-exponential distribution. The former is a convolution of multiple exponentials,
-and intuitively the waiting time distribution for a multistep process
-should be more peaked with a smaller fractional width than a single
-exponential with the same mean.
-A less disperse waiting time distribution means transcription initations are
-more uniformly distributed in time relative to a Poisson process. Hence the
-distribution of mRNA over a population of cells should be less variable compared
-to Poisson, giving $\nu<1$.
-(In Appendix~\ref{sec:non_bursty} we present a more precise version
-of the intuitive arguments in this paragraph.)
-Regardless of the merits of this model in describing the noise properties of
-constitutive transcription initiation, it ultimately fails the simplest
-quantitative feature of the data, namely that the Fano factor $> 1$ and hence
-we must discard this mechanistic picture and search elsewhere.
-\mrm{ENDS HERE}
-
 \subsection{Derivation of chemical master equation}
 \label{sec:cme_from_cartoon}
+
+Before jumping into the derivations of the general computation of the mean mRNA
+level and the Fano factor we will work through the derivation of an example
+master equation. In particular we will focus on model 1 from
+Figure~\ref{fig1:means_cartoons}(C). The general steps are applicable to all
+other chemical master equations in this work.
 
@@
 
+% \subsubsection{Nonequilibrium Model Four - ``Active'' and ``Inactive'' States}
+% Model 4 in Figure~\ref{fig1:means_cartoons}(C) is at the core of the theory
+% in~\cite{Razo-Mejia2020}. At a glance the cartoon for this model may appear very
+% similar to model 2, and mathematically it is, but the interpretation is rather
+% different. In model 2, we interpreted the third state literally as an RNAP-bound
+% promoter and modeled initiation of a transcript as triggering a promoter state
+% change, making the assumption that an RNAP can only
+% make one transcript at a time. In contrast, in the present model the promoter
+% state does \textit{not} change when a transcript is initiated. So we no longer
+% interpret these states as literally RNAP bound and unbound but instead as
+% coarse-grained ``active'' and ``inactive'' states, the details of which we leave
+% unspecified for now. We will comment more on this model below when we discuss
+% Fano factors of models.
+
+% Mathematically this model is very similar to models 1 and 2. Like model 1, the
+% matrix $R$ describing transcription initiation is diagonal, namely
+% \begin{equation}
+% \mathbf{R} = \begin{pmatrix}
+%                 0 & 0 & 0 \\ 
+%                 0 & 0 & 0 \\ 
+%                 0 & 0 & r
+%         \end{pmatrix}.
+% \end{equation}
+% The master equation takes verbatim the same form as it did for model 1,
+% Eq.~\ref{eq:2state_rep_cme}. Meanwhile the promoter transition
+% matrix $\mathbf{K}$ is the same as Eq.~\ref{eq:3state_cme_matrices_pt1}
+% from model 2, although we relabel the
+% rate constants from $k_P^\pm$ to $k^\pm$ to reiterate that these are not simply
+% RNAP binding and unbinding rates.
+
+% Carrying out the algebra, the mean mRNA can be found to be
+% \begin{equation}
+% \langle m\rangle = \frac{r}{\gamma}
+% \frac{k_R^- k^+}
+% {k_R^- k^+ + k_R^- k^- + k_R^+ k^-},
+% \end{equation}
+% and the fold-change readily follows,
+% \begin{align}
+% \text{fold-change}
+% &=      \frac{k_R^- k^+}{k_R^- k^+ + k_R^- k^- + k_R^+ k^-}
+%         \frac{k_R^- k^+ + k_R^- k^-}{k_R^- k^+}
+% \\
+% &=      \left(1 + \frac{k_R^+}{k_R^-}
+%                 \left(1 + \frac{k^+}{k^-}\right)^{-1}
+%         \right)^{-1},
+% \end{align}
+% from which we see $\rho = 1 + k^+/k^-$ as shown in~\fig{fig1:means_cartoons}(C).
+
+% \subsubsection{Nonequilibrium Model Five - Bursty Promoter}
+% The final model we consider shown in Figure~\ref{fig1:means_cartoons}(C) is an
+% intuitive analog to model 1, with just two states, repressor bound or unbound,
+% and transition rates between them of $k_R^+$ and $k_R^-$. In model 1, when in
+% the unbound state, single mRNA transcripts are produced as a Poisson process
+% with some characteristic rate $r$. The current model by contrast produces, at
+% some Poisson rate $k_i$, \textit{bursts} of mRNA transcripts.
+% The burst sizes are assumed to be geometrically distributed with
+% a mean burst size $b$, which we will motivate in Section~\ref{sec:beyond_means}
+% when we derive this model as a certain limiting case of model 4.
+
+% From this intuitive picture and by analogy to model 1, then, it should be
+% plausible that the mean mRNA level is
+% \begin{equation}
+% \langle m\rangle = \frac{k_i b}{\gamma}
+%         \left(1 + \frac{k_R^+}{k_R^-}\right)^{-1},
+% \end{equation}
+% which will turn out to be correct from a careful calculation. For now, we simply
+% note that just like model 1, the fold-change becomes
+% \begin{equation}
+% \text{fold-change} = \left(1 + \frac{k_R^+}{k_R^-}\right)^{-1}
+% \end{equation}
+% with $\rho=1$ also like model 1.
+% We will also see later how this model emerges as a natural limit of model 4.
+% \mrm{ENDS HERE}
+
+% \mrm{THIS NEEDS TO BE FIXED}
+% \subsubsection{Noise in the Two-State Promoter, RNAP Bound or Unbound.}
+
+% Our second model of constitutive transcription posits an architecture in which
+% the promoter is either empty or bound by RNAP ~\cite{Phillips2015a,
+% Phillips2019}. Here,  as shown in model 2 of Figure~\ref{fig2:constit_cartoons}(A),
+% transcription initiation results in a state transition from the bound to the
+% unbound state, reflecting the microscopic reality that an RNAP that has begun to
+% elongate a transcript is no longer available at the start site to begin another.
+% As shown in Appendix~\ref{sec:non_bursty}, the Fano factor in this model is
+% given by
+% \begin{align}
+%     \nu = 1 -
+%         \frac{r\rate{k}{P}{+}}
+%             {\left(\rate{k}{P}{+} + \rate{k}{P}{-} + r\right)
+%              \left(\gamma + \rate{k}{P}{+} + \rate{k}{P}{-} + r\right)}.
+% \label{eq:model2_fano}
+% \end{align}
+% The problem with this picture is that the Fano factor is always $<1$. To make
+%  contact with the experimental reality of $\nu>1$ as shown in
+%  Figure~\ref{fig2:constit_cartoons}(B), clearly some corrections will be needed.
+%  While this model adds an appealing element of microscopic reality, we are
+%  forced to reject it as the additional complexity is unable to capture the
+%  phenomenology of interest. Obviously the promoter state does in fact proceed
+%  through cycles of RNAP binding, initiating, and elongating, but it seems that
+%  the super-Poissonian noise in mRNA copy number we want to model must be
+%  governed by other features of the system.
+
+% \subsubsection{Noise in the Three-State Promoter, Multistep Transcription
+% Initiation and Escape.}
+
+% How might we remedy the deficits of model 2? It is known~\cite{DeHaseth1998}
+% that once RNAP initially binds the promoter region, a multitude of distinct
+% steps occur sequentially before RNAP finally escapes into the elongation phase.
+% Perhaps adding some of this mechanistic detail as shown in model 3 of
+% Figure~\ref{fig2:constit_cartoons}(A) might rescue the previous model. The next
+% simplest refinement of that model could consider open complex formation and
+% promoter escape as separate steps rather than as a single effective step. In
+% other words, we construct model 3 by adding a single extra state to model 2, and
+% we will label the two RNAP-bound states as the closed and open complexes,
+% despite the true biochemical details certainly being more complex. For example,
+% earlier work extended this model by adding an additional repressor bound state
+% and did not explicitly consider the limit with no repressor that we analyze
+% here~\cite{Mitarai2015}. Again, our goal here is not a complete accounting of
+% all the relevant biochemical detail; this is an exploratory search for the
+% important features that a model needs to include to square with the known
+% experimental reality of constitutive expression.
+
+% Unfortunately, as hinted at in earlier work~\cite{Mitarai2015}, this model too
+% has Fano factor $\nu<1$. We again leave the algebraic details for
+% Appendix~\ref{sec:non_bursty} and merely state the result that
+% \begin{align}
+% \nu = 1 - \frac{r k_P^+ k_O}{\mathcal{Z}}
+% \frac{k_P^+ + k_P^- + k_O + r + \gamma}
+%     {\mathcal{Z} + \gamma(k_P^+ + k_P^- + k_O + r) + \gamma^2},
+% \label{eq:model3_fano}
+% \end{align}
+% where we defined $\mathcal{Z} = r(k_O + k_P^-) + k_P^+(k_O + r)$ for notational
+% tidiness. This is necessarily less than 1 for arbitrary rate constants.
+
+% In fact, we suspect \textit{any} model in which transcription proceeds through a
+% multistep cycle must necessarily have $\nu<1$. The intuitive argument compares
+% the waiting time distribution to traverse the cycle with the waiting time for a
+% Poisson promoter (model 1) with the same mean time. The latter is simply an
+% exponential distribution. The former is a convolution of multiple exponentials,
+% and intuitively the waiting time distribution for a multistep process
+% should be more peaked with a smaller fractional width than a single
+% exponential with the same mean.
+% A less disperse waiting time distribution means transcription initations are
+% more uniformly distributed in time relative to a Poisson process. Hence the
+% distribution of mRNA over a population of cells should be less variable compared
+% to Poisson, giving $\nu<1$.
+% (In Appendix~\ref{sec:non_bursty} we present a more precise version
+% of the intuitive arguments in this paragraph.)
+% Regardless of the merits of this model in describing the noise properties of
+% constitutive transcription initiation, it ultimately fails the simplest
+% quantitative feature of the data, namely that the Fano factor $> 1$ and hence
+% we must discard this mechanistic picture and search elsewhere.
+% \mrm{ENDS HERE}
+
 \subsection{Matrix form of the multi-state chemical master equation}
@@
 models, the natural mathematicization of their cartoons is as a chemical master
-equation such as Eq.~\ref{eq:2state_rep_cme} for model 1, which for
-completeness we reproduce here as
+equation such as the coupled chemical master equations we derived for model 1,
+which for completeness we reproduce here as
 \begin{align}
@@
 \end{equation}
-which is just Eq.~\ref{eq:3state_rep_cme} in the main text. Straightforward
-albeit tedious algebra verifies that Eqs.~\ref{eq:model2_cme_appdx}
-and~\ref{eq:generic_cme_appdx} are in fact equivalent.
+Straightforward albeit tedious algebra verifies that
+Eqs.~\ref{eq:model2_cme_appdx} and~\ref{eq:generic_cme_appdx} are in fact
+equivalent.
 
 Although we derived Eq.~\ref{eq:generic_cme_appdx} for the particular case of
-nonequilibrium model 2 in Figure~\ref{fig1:means_cartoons}, in fact the
-chemical master equations for all of the nonequilibrium models in
-Figure~\ref{fig1:means_cartoons} except for model 5 can be cast in this form.
-(We treat model 5 separately in Appendix~\ref{sec:gen_fcn_appdx}.) Model
-3 introduces no new subtleties beyond model 2 and
-Eq.~\ref{eq:generic_cme_appdx} applies equally well to it, simply with different
-matrices of dimension four instead of three. Models 1 and 4 are both
-handled by Eq.~\ref{eq:2state_rep_cme} in the main text, which is just
-Eq.~\ref{eq:generic_cme_appdx} except in the special case of $\mathbf{R_D} =
-\mathbf{R_A} \equiv \mathbf{R}$, since in these two models transcription
-initiation events do not change promoter state.
+kinetic model 2 in Figure~\ref{fig1:means_cartoons}, in fact the chemical master
+equations for all of the kinetic models in Figure~\ref{fig1:means_cartoons}
+except for model 5 can be cast in this form. (We treat model 5 separately in
+Appendix~\ref{sec:gen_fcn_appdx}.) Model 3 introduces no new subtleties beyond
+model 2 and Eq.~\ref{eq:generic_cme_appdx} applies equally well to it, simply
+with different matrices of dimension four instead of three. Models 1 and 4 are
+both handled by a special case of Eq.~\ref{eq:generic_cme_appdx} where
+$\mathbf{R_D} = \mathbf{R_A} \equiv \mathbf{R}$, since in these two models
+transcription initiation events do not change promoter state.
 
@@
 
-\subsection{Nonequilibrium Model One - Poisson Promoter}
+\subsection{Kinetic Model One - Poisson Promoter}
 \subsubsection{Mean mRNA}
-For nonequilibrium model 1 in Figure~\ref{fig1:means_cartoons}, we have
-already shown the full master equation in Eq.~\ref{eq:poisson_promoter_cme} and
-Eq.~\ref{eq:poisson_promoter_cme_appdx}, but for completeness we reprint it
-again as
+For kinetic model 1 in Figure~\ref{fig1:means_cartoons}, we have already shown
+the full master equation in Eq.~\ref{eq:poisson_promoter_cme_appdx}, but for
+completeness we reprint it again as
 \begin{align}
@@
 \end{split}
+\label{eq:2state_rep_cme}
 \end{align}
 This is a direct transcription of the states and rates in
-Figure~\ref{fig1:means_cartoons}. This may be converted to the matrix form of
+Figure~\ref{fig1:means_cartoons}(C). This may be converted to the matrix form of
 the master equation shown in Eq.~\ref{eq:generic_cme_appdx} with matrices
@@
 \begin{equation}
-\langle{m}\rangle = \frac{r}{\gamma} \frac{k_R^-}{k_R^- + k_R^+}
-\end{equation}
-as asserted in Eq.~\ref{eq:mean_m_model1} of the main text.
+\langle{m}\rangle = \frac{r}{\gamma} \frac{k_R^-}{k_R^- + k_R^+}.
+\end{equation}
+
+\subsubsection{Fold-Change}
+$k_R^+$ is proportional to the repressor copy number, so in computing
+fold-change, absence of repressor corresponds to $k_R^+\rightarrow0$. Therefore
+fold-change in this model is simply
+\begin{equation}
+\text{fold-change} = \left(1 + \frac{k_R^+}{k_R^-}\right)^{-1},
+\end{equation}
+matching the master curve of~\fig{fig1:means_cartoons}(D) with $\rho=1$.
 
@@
 
-\subsection{Nonequilibrium Model Two - RNAP Bound and Unbound States}
-\subsubsection{Mean mRNA}
-As shown earlier, the full master equation for model 2 in
-Figure~\ref{fig1:means_cartoons} is
+\subsection{Kinetic Model Two - RNAP Bound and Unbound States}
+
+Our second kinetic model depicted in Figure~\ref{fig1:means_cartoons}(C) mirrors
+the second thermodynamic model of Figure~\ref{fig1:means_cartoons}(B) by
+fine-graining  the repressor unbound state of nonequilibrium model 1, resolving
+it into an empty promoter state and an RNAP-bound state. Note in this picture,
+in contrast with model 4 below, transcription initiation is accompanied by a
+promoter state change, in keeping with the interpretation as RNAP-bound and
+empty states: if an RNAP successfully escapes the promoter and proceeds to
+elongation of a transcript, clearly it is no longer bound at the promoter.
+Therefore another RNAP must bind before another transcript can be initiated.
+
+The master equation governing this model is analogous to
+Eq.~\ref{eq:2state_rep_cme} for model 1 above. The main subtlety arises since
+transcription initiation accompanies a promoter state change. This can be
+understood by analogy to $\mathbf{K}$. The off-diagonal and diagonal elements of
+$\mathbf{K}$ correspond to transitions arriving at or departing from,
+respectively, the promoter state of interest. If transcription initiation is
+accompanied by promoter state changes, we must have separate matrices for
+arriving and departing transcription events since the arriving and departing
+transitions have different initial copy numbers of mRNA, unlike for $\mathbf{K}$
+where they are the same. The master equation for this model is given by three
+coupled differential equations of the form
 \begin{align}
@@
 \\
-&- \overbrace{rp_P(m,t)}^{m \rightarrow m + 1,\;A \rightarrow U},
+&- \overbrace{rp_P(m,t)}^{m \rightarrow m + 1,\;A \rightarrow U}.
 \end{split}
 \end{align}
-which can be condensed to the matrix form of Eq.~\ref{eq:generic_cme_appdx} with
-matrices given by
-\begin{equation}
+We can rewrite this equation using the matrix format as
+\begin{equation}
+\deriv{t}\vec{p}(m,t) =
+\left( \mathbf{K} - \mathbf{R_D} - \gamma m \mathbf{I} \right) \vec{p}(m,t)
+                + \mathbf{R_A} \vec{p}(m-1,t) +
+                \gamma (m+1) \mathbf{I} \vec{p}(m+1,t),
+\label{eq:3state_rep_cme}
+\end{equation}
+with the state vector and promoter transition matrix defined as
+\begin{equation}
+\vec{p}(m) = \begin{pmatrix} p_R(m) \\ p_E(m) \\ p_P(m) \end{pmatrix},\
 \mathbf{K} = \begin{pmatrix} -k_R^- & k_R^+ & 0 \\
@@
                         0 & k_P^+ & -k_P^- 
-                \end{pmatrix},\
+                \end{pmatrix},
+\label{eq:3state_cme_matrices_pt1}
+\end{equation}
+and the initiation matrices given by
+\begin{equation}
 \mathbf{R_A} = \begin{pmatrix}
@@
                 \end{pmatrix}.
-\label{eq:model2_matrices_appdx}
-\end{equation}
+\label{eq:3state_cme_matrices_pt2}
+\end{equation}
+The elements of $\vec{p}(m)$ encode the probabilities of having $m$ mRNA present
+along with the promoter having repressor bound ($R$), being empty ($E$), or
+having RNAP bound ($P$), respectively. $\mathbf{R_A}$ describes probability flux
+\textit{arriving} at the state $\vec{p}(m)$ from a state with one fewer mRNA,
+namely $\vec{p}(m-1)$, and $\mathbf{R_D}$ describes probability flux
+\textit{departing} from the state $\vec{p}(m)$ for a state with one more mRNA,
+namely $\vec{p}(m+1)$. $\mathbf{K}$ is closely analogous to model 1.
+
+\subsubsection{Mean mRNA}
 As for model 1, we must first find $\mathbf{R_A} \langle\vect{m}^0\rangle$.
@@
 \langle{m}\rangle = \frac{r}{\gamma}
-        \frac{k_P^+ k_R^-} {k_P^+ k_R^- + (k_P^- + r)(k_R^+ + k_R^-)},
+        \frac{k_P^+ k_R^-} {k_P^+ k_R^- + (k_P^- + r)(k_R^+ + k_R^-)}.
 \label{eq:model2_meanm_appdx}
 \end{equation}
-as claimed in Eq.~\ref{eq:model2_meanm} in the main text.
+
+\subsubsection{Fold-Change}
+Fold-change is again found from the ratio prescribed by Eq.~\ref{eq:fc_def},
+from which we have
+\begin{align}
+\text{fold-change}
+&=      \frac{k_R^- k_P^+}
+        {k_R^- k_P^+ + k_R^- (k_P^- + r) + k_R^+ (k_P^- + r)}
+        \frac{k_P^+ + k_P^- + r}{k_P^+}
+\\
+&=      \left(1 + \frac{k_R^+}{k_R^-}
+                \frac{k_P^- + r}{k_P^+ + k_P^- + r}
+        \right)^{-1}
+\\
+&=      \left(1 + \frac{k_R^+}{k_R^-}
+        \left(1 + \frac{k_P^+}{k_P^- + r}\right)^{-1}
+        \right)^{-1},
+\end{align}
+which follows the master curve of~\fig{fig1:means_cartoons}(D) with $\rho = 1 +
+k_P^+/(k_P^- + r)$ as claimed.
 
@@
 \begin{equation}
-\nu = 1 - \frac{r k_P^+ } {(k_P^+ + k_P^- + r)(k_P^+ + k_P^- + r + \gamma)},
-\end{equation}
-exactly Eq.~\ref{eq:model2_fano} in the main text.
-
-\subsection{Nonequilibrium Model Three - Multistep Transcription Initiation and
+\nu = 1 - \frac{r k_P^+ } {(k_P^+ + k_P^- + r)(k_P^+ + k_P^- + r + \gamma)}.
+\end{equation}
+
+\subsection{Kinetic Model Three - Multistep Transcription Initiation and
 Escape}
 
-\subsubsection{Mean mRNA}
-In close analogy to model 2 above, nonequilibrium model 3 from
-Figure~\ref{fig1:means_cartoons}(C) can be described by our generic master
-equation Eq.~\ref{eq:generic_cme_appdx} with promoter transition matrix given by
-\begin{equation}
-\mathbf{K} =
-\begin{pmatrix} -k_R^- & k_R^+ & 0 & 0\\
+One might reasonably complain that the first two ``nonequilibrium'' models we
+have considered are straw men. Their steady states necessarily satisfy detailed
+balance which is equivalent to thermodynamic equilibrium. Why is this the case?
+At steady state there is by definition no net probability flux in or out of each
+promoter state, but since the promoter states form a linear chain, there is only
+one way in or out of the repressor bound and RNAP bound states, implying each
+edge must actually have a net zero probability flux, which is the definition of
+detailed balance (usually phrased as equality of forward and reverse transition
+fluxes).
+
+Now we consider model 3 in Figure~\ref{fig1:means_cartoons}(C) which allows the
+possibility of true nonequilibrium steady-state fluxes through the promoter
+states. We point out that this model was considered previously
+in~\cite{Mitarai2015} where a comparison was made with model 1 as used
+in~\cite{Jones2014}. The authors of~\cite{Mitarai2015} argued that the
+additional complexity is essential to properly account for the noise in the mRNA
+distribution. As mentioned in the main text, we weigh on this conclusion based
+on the fact that this model cannot reproduce the observable cell-to-cell 
+variability in mRNA copy number.
+
+The master equation governing this model is identical in form to model 2 above,
+namely
+\begin{equation}
+\deriv{t}\vec{p}(m,t) =
+\left( \mathbf{K} - \mathbf{R_D} - \gamma m \mathbf{I} \right) \vec{p}(m,t)
+                + \mathbf{R_A} \vec{p}(m-1,t) +
+                \gamma (m+1) \mathbf{I} \vec{p}(m+1,t),
+\end{equation}
+but with a higher-dimensional state space and different matrices. The
+state vector and promoter transition matrix are now
+\begin{equation}
+\vec{p}(m) = \begin{pmatrix} p_R(m) \\ p_E(m) \\
+                             p_C(m) \\ p_O(m)\end{pmatrix},\
+\mathbf{K} = \begin{pmatrix} -k_R^- & k_R^+ & 0 & 0\\
                         k_R^- & -k_R^+ -k_P^+ & k_P^- & 0 \\
@@
                         0 & 0 & k_O & 0
-\end{pmatrix}
-\end{equation}
-and transcription matrices given by
-\begin{equation}
-\mathbf{R_A} =
-\begin{pmatrix}
+                \end{pmatrix},
+\end{equation}
+with the four promoter states, in order, being repressor bound ($R$), empty
+($E$), RNAP closed complex ($C$), and RNAP open complex ($O$). Besides
+increasing dimension by one, the only new feature in $\mathbf{K}$ is the
+rate $k_O$, representing the rate of open complex formation from
+the closed complex, which we assume for simplicity to be irreversible
+in keeping with some~\cite{Mitarai2015} but not all~\cite{DeHaseth1998}
+past literature. The initiation matrices are given by
+\begin{equation}
+\mathbf{R_A} = \begin{pmatrix}
         0 & 0 & 0 & 0 \\ 
@@
                 \end{pmatrix},\
-\mathbf{R_D} =
-\begin{pmatrix}
+\mathbf{R_D} = \begin{pmatrix}
         0 & 0 & 0 & 0 \\ 
@@
         0 & 0 & 0 & r
-\end{pmatrix}.
-\end{equation}
-$\langle\vec{m}^0\rangle$ is again given by Eq.~\ref{eq:vecm0_appdx_ref},
-which in this case takes the form
+                \end{pmatrix},
+\end{equation}
+again closely analogous to nonequilibrium model 2.
+
+\subsubsection{Mean mRNA}
+In close analogy to model 2 above, $\langle\vec{m}^0\rangle$ is again given by
+Eq.~\ref{eq:vecm0_appdx_ref}, which in this case takes the form
 \begin{equation}
@@
 which equals Eq.~\ref{eq:model3_mean_m} in the main text.
+
+\subsubsection{Fold-Change}
+The mean mRNA is much messier than for the other models. The strategy to compute
+the fold-change is to isolate the terms involving the repressor, so that now the
+fold-change is seen to be simply
+\begin{align}
+\text{fold-change}
+&= \frac{\frac{k_P^+ k_O}{r(k_O + k_P^-)}}
+        {1 + \frac{k_P^+ (k_O + r)}{r(k_O + k_P^-)} + \frac{k_R^+}{k_R^-}}
+        \frac{1 + \frac{k_P^+ (k_O + r)}{r(k_O + k_P^-)}}
+                {\frac{k_P^+ k_O}{r(k_O + k_P^-)}}
+\\
+&= \left(
+        1 + \frac{k_R^+}{k_R^-}
+        \left(1 + \frac{k_P^+ (k_O + r)}{r(k_O + k_P^-)}\right)^{-1}
+\right)^{-1},
+\end{align}
+surprisingly reducing to the master curve of~\fig{fig1:means_cartoons}(D) once
+again, with $\rho = 1 + \frac{k_P^+ (k_O + r)}{r(k_O + k_P^-)}$.
+
+This example hints that an arbitrarily fine-grained model of downstream
+transcription steps may still be collapsed to the form of the master curve for
+the means given in Figure~\ref{fig1:means_cartoons}(D), so long as the repressor
+binding is exclusive with transcriptionally active states. We offer this
+argument as a conjecture, and we suspect that a careful argument using the
+graph-theoretic framework based on the Matrix-Tree
+Theorem~\cite{Gunawardena2012}, might furnish a ``proof'' not with some
+challenges as this theoretical framework applies, at the present, to finite
+graphs.
 
@@
         \frac{k_P^+ + k_P^- + k_O + r + \gamma}
-            {\mathcal{Z} + \gamma(k_P^+ + k_P^- + k_O + r) + \gamma^2},
-\end{equation}
-as stated in Eq.~\ref{eq:model3_fano} in the main text.
+            {\mathcal{Z} + \gamma(k_P^+ + k_P^- + k_O + r) + \gamma^2}.
+\end{equation}
 
@@
 
-\subsection{Nonequilibrium Model Four - ``Active'' and ``Inactive'' States}
+\subsection{Kinetic Model Four - ``Active'' and ``Inactive'' States}
+Model 4 in Figure~\ref{fig1:means_cartoons}(C) is at the core of the theory
+in~\cite{Razo-Mejia2020}. At a glance the cartoon for this model may appear very
+similar to model 2, and mathematically it is, but the interpretation is rather
+different. In model 2, we interpreted the third state literally as an RNAP-bound
+promoter and modeled initiation of a transcript as triggering a promoter state
+change, making the assumption that an RNAP can only
+make one transcript at a time. In contrast, in the present model the promoter
+state does \textit{not} change when a transcript is initiated. So we no longer
+interpret these states as literally RNAP bound and unbound but instead as
+coarse-grained ``active'' and ``inactive'' states, the details of which we leave
+unspecified for now. We will comment more on this model below when we discuss
+Fano factors of models.
+
+Mathematically this model is very similar to models 1 and 2. Like model 1, the
+matrix $R$ describing transcription initiation is diagonal, namely
+\begin{equation}
+\mathbf{R} = \begin{pmatrix}
+                0 & 0 & 0 \\ 
+                0 & 0 & 0 \\ 
+                0 & 0 & r
+        \end{pmatrix}.
+\end{equation}
+The master equation takes verbatim the same form as it did for model 1,
+Eq.~\ref{eq:2state_rep_cme}. Meanwhile the promoter transition
+matrix $\mathbf{K}$ is the same as Eq.~\ref{eq:3state_cme_matrices_pt1}
+from model 2, although we relabel the
+rate constants from $k_P^\pm$ to $k^\pm$ to reiterate that these are not simply
+RNAP binding and unbinding rates.
+
 \subsubsection{Mean mRNA}
@@
 \end{equation}
+
+\subsubsection{Fold-Change}
+
+The fold-change readily follows,
+\begin{align}
+\text{fold-change}
+&=      \frac{k_R^- k^+}{k_R^- k^+ + k_R^- k^- + k_R^+ k^-}
+        \frac{k_R^- k^+ + k_R^- k^-}{k_R^- k^+}
+\\
+&=      \left(1 + \frac{k_R^+}{k_R^-}
+                \left(1 + \frac{k^+}{k^-}\right)^{-1}
+        \right)^{-1},
+\end{align}
+from which we see $\rho = 1 + k^+/k^-$ as shown in~\fig{fig1:means_cartoons}(C).
 
@@
 \begin{equation}
-\nu = 1 + \frac{r k^-}{(k^- + k^+)(k^- + k^+ + \gamma)},
-\end{equation}
-as quoted in Eq.~\ref{eq:model4_fano} and in Figure~\ref{fig2:constit_cartoons}.
+\nu = 1 + \frac{r k^-}{(k^- + k^+)(k^- + k^+ + \gamma)}.
+\end{equation}

### 2020-09-26 doc/resubmission/section_01_introduction.tex
implement Griffin's comments from 20200925

@@
 a given promoter architecture, there remains the predictive challenge of
-understanding its input-output properties, an objective that can be met by a
-myriad of approaches using the tools of statistical physics~\cite{Ackers1982,
+understanding its input-output properties, an objective that can be met by
+myriad approaches using the tools of statistical physics~\cite{Ackers1982,
 Shea1985, Buchler2003, Vilar2003a, Vilar2003b, Bintu2005a, Bintu2005c,
@@
 
-To that end, in this paper we examine a wide variety of distinct models for the
+To that end, in this work we examine a wide variety of distinct models for the
 simple repression regulatory architecture. This genetic architecture consists of
@@
 a transcript? Turning the question around, can we see any imprint of those
-processes in the available data? If the answer is no, then those processes are
-irrelevant for our purposes. Forward modeling and inverse (statistical
-inferential) modeling are necessary to tackle such questions. We combine both
-approaches in order to discard models that cannot empirically satisfy the main
-features of experimental data. First we apply forward modeling to demonstrate
-that none of the models are distinguishable at the level of mean gene
-expression. We then extend the modeling to look at higher moments of the
-distribution, eliminating models that do not empirically satisfy the observed
-cell-to-cell variability. Finally we arrive at a minimal model on which we can
-apply inverse modeling in order to infer the parameters that explain the data.
+processes in the available data? If the answer is no, then those processes,
+though fascinating in their own right, are irrelevant for our purposes. Forward
+modeling and inverse (statistical inferential) modeling are necessary to tackle
+such questions. We combine both approaches in order to discard models that
+cannot empirically satisfy the main features of experimental data. First we
+apply forward modeling to demonstrate that none of the models are
+distinguishable at the level of mean gene expression. We then extend the
+modeling to look at higher moments of the distribution, eliminating models that
+do not empirically satisfy the observed cell-to-cell variability. Finally we
+arrive at a minimal model on which we can apply inverse modeling in order to
+infer the parameters that explain the data.
 
@@
 sterically blocked from binding to its promoter sequence. Current evidence
-suggests this is sometimes, but not always the case, and it remains an
+suggests this is sometimes (but not always) the case and it remains an
 interesting open question precisely how a repressor bound far upstream is able
@@
 Figure~\ref{fig1:means_cartoons}(D). Section~\ref{sec:beyond_means} goes beyond
-an analysis of the mean gene expression by asking how the same models presented
-in Figure~\ref{fig1:means_cartoons}(C) can be used to explore noise in gene
-expression. To make contact with experiment, all of these models must make a
-commitment to some numerical values for the key parameters found in each such
+an analysis of the mean level of gene expression by asking how the same models
+presented in Figure~\ref{fig1:means_cartoons}(C) can be used to explore noise in
+gene expression. To make contact with experiments, all of these models must make
+a commitment to some numerical values for the key parameters found in each such
 model. Therefore in Section~\ref{section_04_bayesian_inference} we explore the

### 2020-09-26 doc/resubmission/section_02_means.tex
implement Griffin's comments from 20200925

@@
 continuous number of promoter states (due to effects such as supercoiling of
-DNA~\cite{Chong2014, Sevier2016}) is analogous to how the Monod-Wyman-Changeux
-model of allostery coarse-grains continuous molecule conformations into a finite
-number of states~\cite{Martins2011}. The models are probabilistic with each
-state assigned some probability and the overall rate of transcription given by 
+DNA~\cite{Chong2014, Sevier2016}  or DNA looping~\cite{Boedicker2013a}) is
+analogous to how the Monod-Wyman-Changeux model of allostery coarse-grains
+continuous molecule conformations into a finite number of
+states~\cite{Martins2011}. The models are probabilistic with each state assigned
+some probability and the overall rate of transcription given by 
 \begin{equation}
@@
 $i^{\text{th}}$ state, and $r_i$ is the rate of transcription of that state.
-Ultimately, the different models differ along several key axes: what states to
-consider and how to compute the probabilities of those states.
+Ultimately, the different models differ along several key aspects: what states
+to consider and how to compute the probabilities of those states.
 
@@
 within the class of thermodynamic models, we can make different commitments
-about the underlying microscopic states of the promoter.  Indeed, the list of
-options considered here does not at all exhaust the suite of different
-microscopic states we can assign to the promoter.
+about the underlying microscopic states of the promoter. Model 1 considers only
+two states: a state in which a repressor (with copy number $R$) binds to an
+operator and a transcriptionally active state. The free energy difference
+between the repressor binding the operator, i.e. a specific binding site, and
+one of the $N_{NS}$ non-specific sites is given by $\Delta\varepsilon_R$ (given
+in $k_BT$ units as $\beta\equiv (k_BT)^{-1}$). Model 2 expands this model to
+include an empty promoter where no transcription happens, as well as a state on
+which one of the $P$ RNAPs binds to the promoter with binding energy
+$\Delta\varepsilon_P$. Indeed, the list of options considered here does not at
+all exhaust the suite of different microscopic states we can assign to the
+promoter.
 
@@
 = \frac{\langle m (R > 0) \rangle}{\langle m (R = 0) \rangle}
+= \frac{\langle p (R > 0) \rangle}{\langle p (R = 0) \rangle}
 \label{eq:fc_def}
@@
 where angle brackets $\left\langle \cdot \right\rangle$ denote the average over
-a population of cells and mean mRNA $\langle m\rangle$ is viewed as a function
-of repressor copy number $R$. What this means is that the fold-change in gene
-expression is a relative measurement of the effect of the transcriptional
-repressor ($R > 0$) on the gene expression level compared to an unregulated
-promoter ($R = 0$). The second equality in Eq.~\ref{eq:fc_def} follows from
-assuming that the translation efficiency, i.e., the number of proteins
-translated per mRNA, is the same in both conditions. In other words, we assume
-that mean protein level is proportional to mean mRNA level, and that the
-proportionality constant is the same in both conditions and therefore cancels
-out in the ratio. This is reasonable since the cells in the two conditions are
-identical except for the presence of the transcription factor, and the model
-assumes that the transcription factor has no direct effect on translation.
+a population of cells and mean mRNA $\langle m\rangle$ and mean protein $\langle
+p\rangle$ are viewed as a function of repressor copy number $R$. What this means
+is that the fold-change in gene expression is a relative measurement of the
+effect of the transcriptional repressor ($R > 0$) on the gene expression level
+compared to an unregulated promoter ($R = 0$). The third equality in
+Eq.~\ref{eq:fc_def} follows from assuming that the translation efficiency, i.e.,
+the number of proteins translated per mRNA, is the same in both conditions. In
+other words, we assume that mean protein level is proportional to mean mRNA
+level, and that the proportionality constant is the same in both conditions and
+therefore cancels out in the ratio. This is reasonable since the cells in the
+two conditions are identical except for the presence of the transcription
+factor, and the model assumes that the transcription factor has no direct effect
+on translation.
 
@@
 
-\fig{fig1:means_cartoons} depicts a smorgasbord of mathematicized cartoons for
-simple repression using both thermodynamic and kinetic models that have appeared
-in previous literature. For each cartoon, we calculate the fold-change in mean
-gene expression as predicted by that model, deferring some algebraic details to
-Appendix~\ref{sec:non_bursty}. What we will find is that all cartoons collapse
-to a single master curve as shown in \fig{fig1:means_cartoons}(D), which
-contains just two parameters. We label the parameters $\Delta F_R$, an effective
-free energy parametrizing the repressor-DNA interaction, and $\rho$, which
-subsumes all details of transcription in the absence of repressors. We will
-offer some intuition for why this master curve exists and discuss why at the
-level of the mean expression, we are unable to discriminate ``right'' from
-``wrong'' cartoons given only measurements of fold-changes in expression.
+Figure~\ref{fig1:means_cartoons}(B) and (C) depicts an array of mathematicized
+cartoons for simple repression using both thermodynamic and kinetic models,
+respectively, that have appeared in previous literature. For each cartoon, we
+calculate the fold-change in mean gene expression as predicted by that model,
+deferring most algebraic details to Appendix~\ref{sec:non_bursty}. What we find
+is that for all cartoons the fold-change can be written as a Fermi function
+
+\begin{equation}
+ \text{fold-change} = \left( 1 + \exp(-\Delta F_R + \log(\rho))  \right)^{-1},
+ \label{eq:deltaFR_eq_noneq_equiv}
+\end{equation}
+where the effective free energy contains two terms: the parameters $\Delta F_R$,
+an effective free energy parametrizing the repressor-DNA interaction, and
+$\rho$, a term derived from the level of coarse-graining used to model all
+repressor-free states. In other words, the effective free energy of the Fermi
+function can be written as the additive effect of the regulation given by the
+repressor via $\Delta F_R$, and the kinetic scheme used to describe the steps
+that lead to a transcriptional event via $\log(\rho)$ (See
+Figure~\ref{fig1:means_cartoons}(D), left panel). This implies that all models
+collapse to a single master curve as shown in
+Figure~\ref{fig1:means_cartoons}(D). We will offer some intuition for why this
+master curve exists and discuss why at the level of the mean expression, we are
+unable to discriminate ``right'' from ``wrong'' cartoons given only measurements
+of fold-changes in expression.
 
 \subsubsection{Two- and Three-state Thermodynamic Models}
+
 We begin our analysis with models 1 and 2 in
-Figure~\ref{fig1:means_cartoons}(B). In each of these models the promoter
-is idealized as existing in a set of discrete states; the difference being
-whether or not the RNAP bound state is included or not. Gene expression is then
-assumed to be proportional to the probability of the promoter being in either
-the empty state (model 1) or the RNAP-bound state (model (2)). We direct the
-reader to Appendix~\ref{sec:non_bursty} for details on the derivation of the
-fold-change. For our purposes here, it suffices to state that the functional
-form of the fold-change for model 1 is
+Figure~\ref{fig1:means_cartoons}(B). In each of these models the promoter is
+idealized as existing in a set of discrete states; the difference being whether
+or not the RNAP bound state is included or not. Gene expression is then assumed
+to be proportional to the probability of the promoter being in either the empty
+state (model 1) or the RNAP-bound state (model (2)). We direct the reader to
+Appendix~\ref{sec:non_bursty} for details on the derivation of the fold-change.
+For our purposes here, it suffices to state that the functional form of the
+fold-change for model 1 is
 \begin{equation}
@@
 situation. The fold-change takes the form
-\begin{align}
+\begin{eqnarray}
 \text{fold-change}
@@
 &= (1 + \exp(-\Delta F_R + \log\rho))^{-1},
-\end{align}
+\end{eqnarray}
 where $P$ is the number of RNAP per cell, and $\Delta\varepsilon_P$ is the
@@
 
-Again, we consign all details of the derivation to Appendix~\ref{sec:non_bursty}.
-Here we just highlight the general findings for all five kinetic models. As
-already shown in Figure~\ref{fig1:means_cartoons}(C) and (D), all the kinetic
-models explored can be collapsed onto the master curve. Given that the
-repressor-bound state only connects to the rest of the promoter dynamics via its
-binding and unbinding rates, $k_R^+$ and $k_R^-$ respectively, all models can
-effectively be separated into two categories: a single repressor-bound state,
-and all other promoter states with different levels of coarse graining. This
-structure then guarantees that, at steady-state, detailed balance between these
-two groups is satisfied. What this implies is that the steady-state distribution
-of each of the non-repressor states has the same functional form with or without
-the repressor, allowing us to write the fold-change as a product of the ratio of
-the binding and unbinding rates of the promoter, and the promoter details. This
-is 
+We relegate all details of the derivation to Appendix~\ref{sec:non_bursty}, and
+highlight here the general findings for all five kinetic models. As already
+shown in Figure~\ref{fig1:means_cartoons}(C) and (D), all the kinetic models
+explored can be collapsed onto the master curve. Given that the repressor-bound
+state only connects to the rest of the promoter dynamics via its binding and
+unbinding rates, $k_R^+$ and $k_R^-$ respectively, all models can effectively be
+separated into two categories: a single repressor-bound state, and all other
+promoter states with different levels of coarse graining. This structure then
+guarantees that, at steady-state, detailed balance between these two groups is
+satisfied. What this implies is that the steady-state distribution of each of
+the non-repressor states has the same functional form with or without the
+repressor, allowing us to write the fold-change as a product of the ratio of the
+binding and unbinding rates of the promoter, and the promoter details. This is
 \mrm{This is based in one of the comments from reviewer \# 2. It is empirically
@@
 general statement for any unimaginable complex model.}
-\begin{align}
+\begin{eqnarray}
 \text{fold-change} &= \left( 1 + \frac{k_R^+}{k_R^-} \rho \right)^{-1},\\
 &= (1 + \exp(-\Delta F_R + \log(\rho) ))^{-1},
-\end{align}
+\end{eqnarray}
 where $\Delta F_R \equiv -\log(k_R^+/k_R^-)$, and the functional forms of $\rho$
-for each model change as shown in Figure~\ref{fig1:means_cartoons}(C) (See
-Appendix~\ref{sec:non_bursty} for further details).
+for each model change as shown in Figure~\ref{fig1:means_cartoons}(C). Another
+intuitive way two think about these to terms is as follows: in all kinetic
+models shown in Figure~\ref{fig1:means_cartoons}(C) the repressor-bound state
+can only be reached from a single repressor-free state. The ratio of these two
+states --repressor-bound and adjacent repressor-free state-- must remain the
+same for all models, regardless of the details included in other promoter states
+if $\Delta F_R$ represents an effective free energy of the repressor binding the
+DNA operator. The presence of other states then draws probability density from
+the promoter being in either of these two states, making the ratio between the
+repressor-bound state and \textit{all} repressor-free states different. The log
+difference in this ratio is given by $\log(\rho)$. Since model 1 and model 5 of
+Figure~\ref{fig1:means_cartoons}(C) consist of a single repressor-free state,
+$\rho$ is then necessarily 1 (See Appendix~\ref{sec:non_bursty} for further
+details).
 
@@
 be readily measured experimentally is insufficient to discern between these
-different levels of coarse-graining. In the next section we extend the analysis
-of the models to higher moments of the mRNA distribution as we continue to
-examine the discriminatory power of these different models.
-
-% \subsection{Discussion of Results Across Models for Fold-Changes in Mean
-% Expression}
-% The key outcome of our analysis of the models in
-% Figure~\ref{fig1:means_cartoons} is the existence of a master curve shown
-% in Figure~\ref{fig1:means_cartoons}(D) to which the fold-change predictions
-% of all the models collapse. This master curve is parametrized by only two
-% effective parameters: $\Delta F_R$, which characterizes the number of repressors
-% and their binding strength to the DNA, and $\rho$, which characterizes all other
-% features of the promoter architecture. The key assumption underpinning this
-% result is that no transcription occurs when a repressor is bound to its
-% operator. Note, however, that we are agnostic about the molecular mechanism
-% which achieves this; steric effects are one plausible mechanism, but, for
-% instance, ``action-at-a-distance'' mediated by kinked DNA due to repressors
-% bound tens or hundreds of nucleotides upstream of a promoter is plausible as
-% well.
-
-% Why does the master curve of Figure~\ref{fig1:means_cartoons}(D) exist at all?
-% This brings to mind the deep questions posed in,
-% e.g.,~\cite{Frank2013} and~\cite{Frank2014a}, suggesting we consider multiple
-% plausible models of a system and search for their common patterns to tease
-% out which broad features are and are not important.
-% In our case, the key feature seems to be the
-% exclusive nature of repressor and RNAP binding, which allows the parameter
-% describing the repressor, $\Delta F_R$, to cleanly separate from all other
-% details of the promoter architecture, which are encapsulated in $\rho$.
-% Arbitrary nonequilibrium behavior can occur on the rest of the
-% promoter state space, but it may all be swept up in the effective
-% parameter $\rho$, to which the repressor makes no contribution.
-% We point the interested reader to~\cite{Gunawardena2012}
-% and~\cite{Ahsendorf2014} for an interesting analysis of similar problems
-% using a graph-theoretic language.
-
-% As suggested in~\cite{Chure2019}, we believe this master curve should generalize
-% to architectures with multiple repressor binding sites, as long as the
-% exclusivity of transcription factor binding and transcription initiation is
-% maintained. The interpretation of $\Delta F_R$ is then of an effective free
-% energy of all repressor bound states. In an equilibrium picture this is simply
-% given by the log of the sum of Boltzmann weights of all repressor bound states,
-% which looks like the log of a partition function of a subsystem. In a
-% nonequilibrium picture, while we can still mathematically gather terms and give
-% the resulting collection the label $\Delta F_R$, it is unclear if the physical
-% interpretation as an effective free energy makes sense. The problem is that free
-% energies cannot be assigned unambiguously to states out of equilibrium because
-% the free energy change along a generic path traversing the state space is path
-% dependent, unlike at equilibrium. A consequence of this is that, out of
-% equilibrium, $\Delta F_R$ is no longer a simple sum of Boltzmann weights.
-% Instead it resembles a restricted sum of King-Altman diagrams~\cite{King1956,
-% Hill1966}. Following the work of Hill~\cite{Hill1989}, it may yet be possible to
-% interpret this expression as an effective free energy, but this remains unclear
-% to us. We leave this an open problem for future work.
-
-% If we relax the requirement of exclusive repressor-RNAP binding, one could
-% imagine models in which repressor and RNAP doubly-bound states are allowed,
-% where the repressor's effect is to \text{reduce} the transcription rate rather
-% than setting it to zero. Our results do not strictly apply to such a model,
-% although we note that if the repressor's reduction of the transcription rate is
-% substantial, such a model might still be well-approximated by one of the models
-% in~\fig{fig1:means_cartoons}.
-
-% One may worry that our ``one curve to rule them all'' is a mathematical
-% tautology. In fact we \textit{agree} with this criticism if $\Delta F_R$ is
-% ``just a fitting parameter'' and cannot be meaningfully interpreted as a real,
-% physical free energy. An analogy to Hill functions is appropriate here. One of
-% their great strengths and weaknesses, depending on the use they are put to, is
-% that their parameters coarse-grain many details and are generally not
-% interpretable in terms of microscopic models, for deep reasons discussed at
-% length in~\cite{Frank2013}. By contrast, our master curve claims to have the
-% best of both worlds: a coarse-graining of all details besides the repressor into
-% a single effective parameter $\rho$, while simultaneously retaining an
-% interpretation of $\Delta F_R$ as a physically meaningful and interpretable free
-% energy. Our task, then, is to prove or disprove this claim.
-
-% How do we test this and probe the theory with fold-change measurements? There is
-% a fundamental limitation in that the master curve is essentially a one-parameter
-% function of $\Delta F_R + \log\rho$. Worse, there are many \textit{a priori}
-% plausible microscopic mechanisms that could contribute to the value of $\rho$,
-% such as RNAP binding and escape kinetics~\cite{DeHaseth1998, Mitarai2015},
-% and/or supercoiling accumulation and release~\cite{Chong2014, Sevier2016},
-% and/or, RNAP clusters analogous to those observed in
-% eukaryotes~\cite{Cisse2013, Cho2016} and recently also observed in
-% bacteria~\cite{Ladouceur2020}. Even if $\Delta F_R$ is measured to high
-% precision, inferring the potential microscopic contributions to $\rho$, buried
-% inside a log no less, from fold-change measurements seems beyond reach. As a
-% statistical inference problem it is entirely nonidentifiable, in the language
-% of~\cite{Gelman2013}, Section 4.3.
-
-% If we cannot simply infer values of $\rho$ from measurements of fold-change, can
-% we perturb some of the parameters that make up $\rho$ and measure the change?
-% Unfortunately we suspect this is off-limits experimentally: most of the
-% potential contributors to $\rho$ are global processes that affect many or all
-% genes. For instance, changing RNAP association rates by changing RNAP copy
-% numbers, or changing supercoiling kinetics by changing topoisomerase copy
-% numbers, would massively perturb the entire cell's physiology and confound any
-% determination of $\rho$.
-
-% One might instead imagine a bottom-up modeling approach, where we mathematicize
-% a model of what we hypothesize the important steps are and are not, use \textit{in vitro}
-% data for the steps deemed important, and \textit{predict} what $\rho$ should be.
-% But again, because of the one-parameter nature of the master curve, many
-% different models will likely make indistinguishable predictions, and without any
-% way to experimentally perturb \textit{in vivo}, there is no clear way
-% to test whether the modeling assumptions are correct.
-
-% In light of this, we prefer the view that parameters and rates are not directly
-% comparable between cartoons in~\fig{fig1:means_cartoons}. Rather, parameters in
-% the simpler cartoons represent coarse-grained combinations of parameters in the
-% finer-grained models. For instance, by equating $\rho$ between any two models,
-% one can derive various possible correspondences between the two models'
-% parameters. Note that these correspondences are clearly not unique, since many
-% possible associations could be made. It then is a choice as to what microscopic
-% interpretations the model-builder prefers for the parameters in a particular
-% cartoon, and as to which coarse-grainings lend intuition and which seem
-% nonsensical. Indeed, since it remains an open question what microscopic features
-% dominate $\rho$ (as suggested above, perhaps RNAP binding and escape
-% kinetics~\cite{DeHaseth1998, Mitarai2015}, or supercoiling accumulation and
-% release~\cite{Chong2014, Sevier2016}, or, something more exotic like RNAP
-% clusters~\cite{Cisse2013, Cho2016, Ladouceur2020}), we are hesitant to put too
-% much weight on any one microscopic interpretation of model parameters that make
-% up $\rho$.
-
-% One possible tuning knob to probe $\rho$ that would not globally perturb the
-% cell's physiology is to manipulate RNAP binding sites. Work such
-% as~\cite{Kinney2010} has shown that models of sequence-dependent RNAP affinity can be
-% inferred from data, and the authors of~\cite{Brewster2012} showed that the model
-% of~\cite{Kinney2010} has predictive power by using the model to \textit{design}
-% binding sites of a desired affinity. But for our purposes, this begs the
-% question: the authors of~\cite{Kinney2010} \textit{assumed} a particular model
-% (essentially our 3-state equilibrium model but without the repressor), so it is
-% unclear how or if such methods can be turned around to \textit{compare}
-% different models of promoter function.
-
-% Another possible route to dissect transcription details without a global
-% perturbation would be to use phage polymerase with phage-specific promoters.
-% While such results would carry some caveats, e.g., whether the repression of
-% the phage polymerase is a good analog to the repression of the native RNAP,
-% it could nevertheless be worthy of consideration.
-
-% We have already pointed out that the master curve of~\fig{fig1:means_cartoons}
-% is essentially a one-parameter model, the one parameter being $\Delta F_R +
-% \log\rho$. By now the reader may be alarmed as to how can we even determine
-% $\Delta F_R$ and $\rho$ independently of each other, never mind shedding a lens
-% on the internal structure of $\rho$ itself. A hint is provided by the weak
-% promoter approximation, invoked repeatedly in prior studies~\cite{Bintu2005c,
-% Garcia2011a, Razo-Mejia2018} of simple repression using the 3-state equilibrium
-% model in~\fig{fig1:means_cartoons}(B). In that picture, the weak promoter
-% approximation means $\frac{P}{N_{NS}}\exp(-\beta\Delta\varepsilon_P) \ll 1$,
-% meaning therefore $\rho\approx1$.  This approximation can be well justified on
-% the basis of the number of RNAP and $\sigma$ factors per cell
-% and the strength of binding of RNAP to
-% DNA at weak promoters. This is suggestive, but how can we be sure that $\rho$ is
-% not, for instance, actually $10^2$ and that $\Delta F_R$ hides a compensatory
-% factor? A resolution is offered by an independent inference of $\rho$ in the
-% absence of repressors. This was done in~\cite{Razo-Mejia2020} by fitting
-% nonequilibrium model 4 in~\fig{fig1:means_cartoons}(C), with zero repressor
-% (looking ahead, this is equivalent to model 4
-% in Figure~\ref{fig2:constit_cartoons}(A)),
-% to single-cell mRNA counts data from~\cite{Brewster2014}. This provided a
-% determination of $k^+$ and $k^-$, from which their ratio is estimated to be no
-% more than a few $10^{-1}$ and possibly as small as $10^{-2}$.
-
-% The realization that $\rho\approx1$ to an excellent approximation,
-% \textit{independent} of which model in~\fig{fig1:means_cartoons} one prefers,
-% goes a long way towards explaining the surprising success of equilibrium models
-% of simple repression. Even though our 2- and 3-state models get so many details
-% of transcription wrong, it does not matter because fold-change is a cleverly
-% designed ratio. Since $\rho$ subsumes all details except the repressor, and
-% $\log\rho\approx0$, fitting these simple models to fold-change measurements can
-% still give a surprisingly good estimate of repressor binding energies. So the
-% ratiometric construction of fold-change fulfills its intended purpose of
-% canceling out all features of the promoter architecture except the repressor
-% itself. Nevertheless it is perhaps surprising how effectively it does so:
-% \textit{a priori}, one might not have expected $\rho$ to be quite so close to 1.
-
-% We would also like to highlight the relevance of~\cite{Landman2019} here.
-% Landman et.\ al.\ reanalyzed and compared \textit{in vivo} and
-% \textit{in vitro} data on the lacI repressor's binding affinity to its
-% various operator sequences. (The \textit{in vivo}
-% data was from, essentially, fitting our master curve to expression
-% measurements.) They find broad agreement between the
-% \textit{in vitro} and \textit{in vivo}
-% values. This reinforces the suspicion that the equilibrium $\Delta\varepsilon_R$
-% repressor binding energies do in fact represent real physical free energies.
-% Again, \textit{a priori} this did not have to be the case, even knowing that
-% $\rho\approx1$.
-
-% In principle, if $\Delta F_R$ can be measured to sufficient precision, then
-% deviations from $\rho=1$ become a testable matter of experiment. In practice, it
-% is probably unrealistic to measure repressor rates $k_R^+$ or $k_R^-$ or
-% fold-changes in expression levels (and hence $\Delta\varepsilon_R$) precisely
-% enough to detect the expected tiny deviations from $\rho=1$. We can estimate the
-% requisite precision in $\Delta F_R$ to resolve a given $\Delta\rho$ by noting,
-% since $\rho\approx1$, that $\log(1+\Delta\rho)\approx \Delta\rho$, so
-% $\Delta(\Delta F_R) \approx \Delta\rho$. Suppose we are lucky and $\Delta\rho$
-% is $\sim0.1$, on the high end of our range estimated above. A
-% determination of $\Delta\varepsilon_R/k_BT$ with an uncertainty of barely 0.1
-% was achieved in the excellent measurements of~\cite{Razo-Mejia2018}, so this
-% requires a very difficult determination of $\Delta F_R$ for a very crude
-% determination of $\rho$, which suggests, to put it lightly, this is not a
-% promising path to pursue experimentally. It is doubtful that inference of
-% repressor kinetic rates would be any easier.%\footnote{\mmnote{In fact the
-% % uncertainties quoted in Hammar et.\ al.~\cite{Hammar2014} from Elf's group are
-% % small enough to do this calculation, but HJ \& I think their error bars are
-% % insanely optimistic. Not sure I want to blatantly call them out on it though??}}
-
-% Moving forward, we have weak evidence supporting the interpretation of $\Delta
-% F_R$ as a physically real free energy~\cite{Landman2019} and other work casting
-% doubt~\cite{Hammar2014}. How might we resolve the confusion? If there is no
-% discriminatory power to test the theory and distinguish the various models with
-% measurements of fold-changes in means, how do we probe the theory? Clearly to
-% discriminate between the nonequilibrium models in~\fig{fig1:means_cartoons}, we
-% need to go beyond means to ask questions about kinetics, noise and even full
-% distributions of mRNA copy numbers over a population of cells. If the
-% ``one-curve-to-rule-them-all'' is more than a mathematical tautology, then the
-% free energy of repressor binding inferred from fold-change measurements should
-% agree with repressor binding and unbinding rates. In other words, the
-% equilibrium and nonequilibrium definitions of $\Delta F_R$ must agree, meaning
-% \begin{equation}
-% \Delta F_R = \beta\Delta\varepsilon_R - \log(R/N_{NS})
-%         = - \log(k_R^+/k_R^-),
-% \label{eq:deltaFR_eq_noneq_equiv}
-% \end{equation}
-% must hold, where $\beta\Delta\varepsilon_R$ is inferred from the master curve
-% fit to fold-change measurements, and $k_R^+$ and $k_R^-$ are inferred in some
-% orthogonal manner. Single molecule measurements such as from~\cite{Hammar2014}
-% have directly observed these rates, and in the remainder of this work we explore
-% a complementary approach: inferring repressor binding and unbinding rates
-% $k_R^+$ and $k_R^-$ from single-molecule measurements of mRNA population
-% distributions.
+different levels of coarse-graining.

### 2020-09-26 doc/resubmission/section_03_beyond_means.tex
implement Griffin's comments from 20200925

@@
 calculation of the Fano factor produces
-\begin{align}
-\nu &= 1 + \frac{r k^-}{(k^+ + k^- + \gamma)(k^+ + k^-)},
+\begin{equation}
+\nu = 1 + \frac{r k^-}{(k^+ + k^- + \gamma)(k^+ + k^-)},
 \label{eq:model4_fano}
-\end{align}
+\end{equation}
 which is strictly greater than 1, verifying the above intuition. Note we have
@@
 The mean of the geometric distribution, with this convention, is
-\begin{align}
+\begin{equation}
 \langle n\rangle = \frac{1 - \theta}{\theta}
@@
         = \frac{r}{k^-},
-\end{align}
+\end{equation}
 exactly as we guessed intuitively above.
@@
 Appendix~\ref{sec:non_bursty} takes the form
-\begin{align}
+\begin{eqnarray}
 \begin{split}
@@
 \end{split}
-\end{align}
+\end{eqnarray}
 where we use $k_i$ to denote the burst initiation rate, $\text{Geom}(n;b)$ is
@@
 
+\mrm{Griffin suggests removing this paragraph}
 In reflecting on these 5 models, the reader may feel that exploring a multitude

### 2020-09-26 doc/resubmission/section_04_bayesian_inference.tex
implement Rob's comments
implement Griffin's comments from 20200925

--- 2020-06-06 doc/section_04_bayesian_inference.tex
@@
 
-In this section of the paper, we continue our program of providing one complete
-description of the entire broad sweep of studies that have been made in the
-context of the repressor-operator model, dating all the way back to the original
-work of Jacob and Monod and including the visionary quantitative work of
-M\"{u}ller-Hill and collaborators~\cite{Oehler1990} and up to more recent
-studies \cite{Garcia2011a}. In addition, the aim is to reconcile the equilibrium
-and non-equilibrium perspectives that have been brought to bear on this problem.
-From Section~\ref{section_02_means}, this reconciliation depends on a key
+In this section we aim to reconcile the thermodynamic and kinetic perspectives
+that have been brought to bear on the repressor-operator model. From
+Section~\ref{section_02_means}, this reconciliation depends on a key
 quantitative question as codified by Eq.~\ref{eq:deltaFR_eq_noneq_equiv}: does
-the free energy of repressor binding, as described in the equilibrium models and
-indirectly inferred from gene expression measurements, agree with the
-corresponding values of repressor binding and unbinding rates in the
-non-equilibrium picture, measured or inferred more directly? In this section we
-tackle the statistical inference problem of inferring these repressor rates from
+the free energy of repressor binding, as described in the thermodynamic models
+and indirectly inferred from gene expression measurements, agree with the
+corresponding values of repressor binding and unbinding rates in the kinetic
+picture, measured or inferred more directly? In this section we tackle the
+statistical inference problem of inferring these repressor rates from
 single-cell mRNA counts data. But before we can turn to the full case of simple
@@
 problems of parameter degeneracy that model 4 in
-Figure~\ref{fig2:constit_cartoons} suffered~\cite{Razo-Mejia2020}. Does this
-stand up to closer scrutiny, namely, comparison to full mRNA distributions
-rather than simply their moments? We will test this thoroughly on single-cell
-mRNA counts for different unregulated promoters from Jones et.\
-al.~\cite{Jones2014}.
-
-It will be instructive, however, to first consider the Poisson promoter, model 1
-in Figure~\ref{fig2:constit_cartoons}. As we alluded to earlier, since the
-Poisson distribution has a Fano factor $\nu$ strictly equal to 1, and all of the
+Figure~\ref{fig2:constit_cartoons}(A) suffered~\cite{Razo-Mejia2020}. We will
+test this thoroughly on single-cell mRNA counts for different unregulated
+promoters from Jones et.\ al.~\cite{Jones2014}, but first we consider It will be
+instructive, however, to first consider the Poisson promoter, model 1 in
+Figure~\ref{fig2:constit_cartoons}. As we alluded to earlier, since the Poisson
+distribution has a Fano factor $\nu$ strictly equal to 1, and all of the
 observed data in Figure~\ref{fig2:constit_cartoons}(B) has Fano factor $\nu>1$,
@@
 
+\mrm{Griffin also feels this paragraph should go. Too many times reminding the 
+reader what we are up to. I think he is right.}
 Our objective for this section will then be to assess whether or not model 5 is
@@
 
-For this model the master equation of interest is
-Eq.~\ref{eq:poisson_promoter_cme} with repressor set to zero, i.e.,
-\begin{equation}
-{d\over dt}p_U(m)(t) = 
-        rp_U(m-1)(t) 
-        - rp_U(m)(t)
-        + (m+1)\gamma p_U(m+1)(t) 
-        - \gamma p_U(m)(t),
-\end{equation}
+Model 1 in Figure~\ref{fig2:constit_cartoons}(A) predicts a  mRNA distribution
 whose steady-state solution is given by a Poisson distribution with parameter
-$\lambda \equiv r / \gamma$~\cite{Sanchez2013}. The goal of our inference 
+$\lambda \equiv r / \gamma$, where $r$ is the mRNA production rate, and $\gamma$
+is the mRNA degradation rate~\cite{Sanchez2013}. The goal of our inference
 problem is then to find the probability distribution for the parameter value
@@
 \centering
-\includegraphics[width=\textwidth]{../figures/main/fig03.pdf}
+\includegraphics[width=\textwidth]{../../figures/main/fig03.pdf}
 \caption{\textbf{Constitutive promoter posterior inference and model
@@
 of~\cite{Brewster2012}, where these same 18 promoters were considered through
-the lens of the three-state equilibrium model (model 2 in
+the lens of the three-state thermodynamic model (model 2 in
 Figure~\ref{fig1:means_cartoons}(B)) and binding energies $\Delta\varepsilon_P$
@@
 By taking the weak promoter approximation for the equilibrium model ($P/N_{NS} 
-\exp(-\beta\Delta\varepsilon_r) \ll 1$) results in
+\exp(-\beta\Delta\varepsilon_r) \ll 1$) results in~\cite{Brewster2012}
 \begin{equation}
@@
 
-We mention one further comparison, between our inferred parameters and the
-work of Chong et.\ al.~\cite{Chong2014}, which is interesting and puzzling.
-Beautiful experiments in~\cite{Chong2014} convincingly argue that
-supercoiling accumulated from the production of mRNA transcripts is key
-in setting the burstiness of mRNA production.
-In their model, this supercoiling occurs on the scale of $\sim100$~kb
-domains of DNA. This suggests that all genes on a given domain should burst
-in synchrony, and that the difference between highly and lowly expressed
-genes is the \textit{size} of transcriptional bursts, not the
-\textit{time between} bursts.
-But here, all burst sizes we infer in Figure~\ref{fig:constit_post_full}(C)
-are comparable and burst rates vary wildly. It is not immediately clear
-how to square this circle. Furthermore, Figure~7E in~\cite{Chong2014}
-reports values of the quantity they label $\beta/\alpha$ and we label
-$k^+/k^-$ in model 4 from Figure~\ref{fig2:constit_cartoons}. In contrast
-to the findings of~\cite{Razo-Mejia2020}, Chong et.\ al.\ do not find
-$k^+/k^-\ll1$ for most of the genes they consider. This begs the question:
-is the \textit{galK} chromosomal locus used for the reporter constructs
-in~\cite{Razo-Mejia2020} and~\cite{Jones2014} merely an outlier,
-or is there a deeper puzzle here waiting to be resolved? Without more
-apples-to-apples data we can only speculate, and we leave it as an
-intriguing open question for the field.
+\mrm{ I would vote to remove this paragraph completely.}
+We mention one further comparison, between our inferred parameters and the work
+of Chong et.\ al.~\cite{Chong2014}, which is interesting and puzzling.
+Experiments in~\cite{Chong2014} convincingly argue that supercoiling accumulated
+from the production of mRNA transcripts is key in setting the burstiness of mRNA
+production. In their model, this supercoiling occurs on the scale of
+$\sim100$~kb domains of DNA. This suggests that all genes on a given domain
+should burst in synchrony, and that the difference between highly and lowly
+expressed genes is the \textit{size} of transcriptional bursts, not the
+\textit{time between} bursts. But here, all burst sizes we infer in
+Figure~\ref{fig:constit_post_full}(C) are comparable and burst rates vary
+wildly. It is not immediately clear how to square this circle. Furthermore,
+Figure~7E in~\cite{Chong2014} reports values of the quantity they label
+$\beta/\alpha$ and we label $k^+/k^-$ in model 4 from
+Figure~\ref{fig2:constit_cartoons}. In contrast to the findings
+of~\cite{Razo-Mejia2020}, Chong et.\ al.\ do not find $k^+/k^-\ll1$ for most of
+the genes they consider. This begs the question: is the \textit{galK}
+chromosomal locus used for the reporter constructs in~\cite{Razo-Mejia2020}
+and~\cite{Jones2014} merely an outlier, or is there a deeper puzzle here waiting
+to be resolved? Without more apples-to-apples data we can only speculate, and we
+leave it as an intriguing open question for the field.
 
-Despite such puzzles, our goal here is not to unravel the mysterious origins
-of burstiness in transcription. Our remaining task in this work is a
-determination of the physical reality of equilibrium binding energies in
-Figure~\ref{fig1:means_cartoons}, as codified by the equilibrium-nonequilibrium
-equivalence of Eq.~\ref{eq:deltaFR_eq_noneq_equiv}.
-For our phenomenological needs here model 5 in
-Figure~\ref{fig2:constit_cartoons} is more than adequate: the posterior
-distributions in Figure~\ref{fig:constit_post_full}(C) are cleanly
-identifiable and the predictive checks in Figure~\ref{figS:ppc_unreg}
-indicate no discrepancies between the model and the mRNA single-moleucle
-count data of~\cite{Jones2014}. Of the models we have considered it
-is unique in satisfying both these requirements. So we will happily use
-it as a foundation to build upon in the next section when we add regulation.
+Despite such puzzles, our goal here is not to unravel the mysterious origins of
+burstiness in transcription. Our remaining task in this work is a determination
+of the physical reality of thermodynamic binding energies in
+Figure~\ref{fig1:means_cartoons}, as codified by the thermodynamic-kinetic
+equivalence of Eq.~\ref{eq:deltaFR_eq_noneq_equiv}. For our phenomenological
+needs here model 5 in Figure~\ref{fig2:constit_cartoons}(A) is more than
+adequate: the posterior distributions in Figure~\ref{fig:constit_post_full}(C)
+are cleanly identifiable and the predictive checks in
+Figure~\ref{figS:ppc_unreg} indicate no discrepancies between the model and the
+mRNA single-molecule count data of~\cite{Jones2014}. Of the models we have
+considered it is unique in satisfying both these requirements. So we will
+happily use it as a foundation to build upon in the next section when we add
+regulation.
 
@@
 Now that we have a satisfactory model in hand for constitutive promoters, we
-would like to return to the main thread: can we reconcile the equilibrium and
-nonequilibrium models by putting to the test
-Eq.~\ref{eq:deltaFR_eq_noneq_equiv}, the correspondence between indirectly
-inferred equilibrium binding energies and nonequilibrium kinetic rates? To make
-this comparison, is it possible to infer repressor binding and unbinding rates
-from mRNA distributions over a population of cells as measured by
-single-molecule Fluorescence \textit{in situ} Hybridization in~\cite{Jones2014}?
-If so, how do these inferred rates compare to direct single-molecule
-measurements such as from~\cite{Hammar2014} and to binding energies such as
-from~\cite{Garcia2011a} and~\cite{Razo-Mejia2018}, which were inferred under the
-assumptions of the equilibrium models in Figure~\ref{fig1:means_cartoons}(B)?
-And can this comparison shed light on the unreasonable effectiveness of the
-equilibrium models, for instance, in their application in~\cite{Chure2019,
-Chure2019a}?
+would like to return to the main thread: can we reconcile the thermodynamic and
+kinetic models by putting to the test Eq.~\ref{eq:deltaFR_eq_noneq_equiv}, the
+correspondence between indirectly inferred thermodynamic binding energies and
+kinetic rates? To make this comparison, is it possible to infer repressor
+binding and unbinding rates from mRNA distributions over a population of cells
+as measured by single-molecule Fluorescence \textit{in situ} Hybridization
+in~\cite{Jones2014}? If so, how do these inferred rates compare to direct
+single-molecule measurements such as from~\cite{Hammar2014} and to binding
+energies such as from~\cite{Garcia2011a} and~\cite{Razo-Mejia2018}, which were
+inferred under the assumptions of the thermodynamic models in
+Figure~\ref{fig1:means_cartoons}(B)? And can this comparison shed light on the
+unreasonable effectiveness of the equilibrium models, for instance, in their
+application in~\cite{Chure2019, Chure2019a}?
 
@@
 \centering
-\includegraphics[width=\textwidth]{../figures/main/fig04.pdf}
+\includegraphics[width=\textwidth]{../../figures/main/fig04.pdf}
 \caption{\textbf{Simple repression parameter inference and comparison.}
@@
 nine different conditions, spanning several combinations of three unique
-repressor binding sites and four unique repressor copy numbers. We assume that
-the values of $k_i$ and $b$ are known, since we have already cleanly inferred
-them from constitutive promoter data, and further we assume that these values
-are the same across datasets with different repressor binding sites and copy
-numbers. In other words, we assume that the regulation of the transcription
-factor does not affect the mean burst size nor the burst initiation rate. The
-regulation occurs as the promoter is taken away from the transcriptionally
-active state when the promoter is bound by repressor. We assume that there is
-one unbinding rate parameter for each repressor binding site, and likewise one
-binding rate for each unique repressor copy number. This makes our model seven
-dimensional, or nine if one counts $k_i$ and $b$ as well. Note that we use only
-a subset of the datasets from Jones et.\ al.~\cite{Jones2014}, as discussed more
-in Appendix~\ref{sec:bayesian}.
+repressor binding sites (the so-called $Oid$, $O1$, and $O2$ operators) and four
+unique repressor copy numbers. We assume that the values of $k_i$ and $b$ are
+known, since we have already cleanly inferred them from constitutive promoter
+data, and further we assume that these values are the same across datasets with
+different repressor binding sites and copy numbers. In other words, we assume
+that the regulation of the transcription factor does not affect the mean burst
+size nor the burst initiation rate. The regulation occurs as the promoter is
+taken away from the transcriptionally active state when the promoter is bound by
+repressor. We assume that there is one unbinding rate parameter for each
+repressor binding site, and likewise one binding rate for each unique repressor
+copy number. This makes our model seven dimensional, or nine if one counts $k_i$
+and $b$ as well. Note that we use only a subset of the datasets from Jones et.\
+al.~\cite{Jones2014}, as discussed more in Appendix~\ref{sec:bayesian}.
 
@@
 direct measurements. However, these estimates were made assuming the validity of
-the equilibrium models in Figure~\ref{fig1:means_cartoons}, and since testing
+the thermodynamic models in Figure~\ref{fig1:means_cartoons}, and since testing
 these models is our present goal, it would be circular logic if we were to make
 the same assumption. Therefore we will make no assumptions about the LacI copy
-number for a given inducer concentrations.
+number for a given inducer concentration.
 
@@
 \subsubsection{Comparison with prior measurements of repressor binding energies}
-Our primary goal in this work is to reconcile the kinetic and equilibrium
+Our primary goal in this work is to reconcile the kinetic and thermodynamic
 pictures of simple repression. Towards this end we would like to compare the
@@
 and~\cite{Razo-Mejia2018}. If the agreement is close, then it suggests that the
-equilibrium models are not wrong and the repressor binding energies they contain
-correspond to physically real free energies, not mere fit parameters.
+thermodynamic models are not wrong and the repressor binding energies they
+contain correspond to physically real free energies, not mere fit parameters.
 
 Figure~\ref{fig4:repressed_post_full}(B) shows both comparisons, with the top
-panel comparing to equilibrium binding energies and the bottom panel comparing
+panel comparing to thermodynamic binding energies and the bottom panel comparing
 to single-molecule measurements. First consider the top panel and its comparison
@@
 also a small offset between experiment and simulation for O2 at the higher
-repressor copy numbers, especially at 2 and 10~ng/mL.
-From the estimate of repressor copy numbers from~\cite{Jones2014}, it is
-possible that the repressor copy numbers here are becoming large enough to
-partially invalidate our assumption of a separation of timescales between
-burst duration and repressor association rate. Another possibility is that
-the very large number of zero mRNA counts for Oid, 2~ng/mL is skewing its partner
-datasets through the shared association rate.
-None of these fairly minor quibbles cause us to
+repressor copy numbers, especially at 2 and 10~ng/mL. From the estimate of
+repressor copy numbers from~\cite{Jones2014}, it is possible that the repressor
+copy numbers here are becoming large enough to partially invalidate our
+assumption of a separation of timescales between burst duration and repressor
+association rate. Another possibility is that the very large number of zero mRNA
+counts for Oid, 2~ng/mL is skewing its partner datasets through the shared
+association rate. None of these fairly minor potential caveats cause us to
 seriously doubt the overall correctness of our model, which further validates

### 2020-09-26 doc/resubmission/abstract.tex
more changes

--- 2020-06-06 doc/abstract.tex
@@
 and inhibiting transcription. In this paper, we show how seven distinct models
-of this so-called simple-repression motif, based both on equilibrium and kinetic
-thinking, can be used to derive the predicted levels of gene expression and
-shed light on the often surprising past success of the equilbrium models. These
-different models are then invoked to confront a variety of different data on mean,
-variance and full gene expression distributions, illustrating the extent to which
-such models can and cannot be distinguished, and suggesting a two-state model with a
-distribution of burst sizes as the most potent of the seven for describing the
-simple-repression motif.
+of this so-called simple-repression motif, based both on equilibrium
+thermodynamic and chemical kinetic, can be used to derive the predicted levels
+of gene expression and shed light on the often surprising past success of the
+thermodynamic models. These different models are then invoked to confront a
+variety of different data on mean, variance and full gene expression
+distributions, illustrating the extent to which such models can and cannot be
+distinguished, and suggesting a two-state model with a distribution of burst
+sizes as the most potent of the seven for describing the simple-repression
+motif.
 \end{abstract}

### 2020-09-26 code/figures/main/fig01D.py
code to plot mastercurve

@@ -1,0 +1,15 @@
+# -*- coding: utf-8 -*-
+# %%
+import numpy as np
+import pandas as pd
+import matplotlib.pyplot as plt
+import srep.viz
+colors = srep.viz.color_selector('pboc')
+srep.viz.plotting_style(grid=False)
+
+fig, ax = plt.subplots(1, 1, figsize=(5, 3), dpi=150)
+x = np.linspace(-6,6)
+ax.plot(x, 1/(1+np.exp(-x)))
+ax.set_ylabel('fold-change')
+ax.set_xlabel(r"$\Delta F_R + \log(\rho)$")
+plt.savefig('../../../figures/main/fig01D.pdf', bbox_inches='tight')

### 2020-10-14 doc/resubmission/section_02_means.tex
update to latest version of main text

--- 2020-09-24 doc/resubmission/section_02_means.tex
@@
 
-\fig{fig1:means_cartoons} depicts a smorgasbord of mathematicized cartoons for
-simple repression using both thermodynamic and kinetic models that have appeared
-in previous literature. For each cartoon, we calculate the fold-change in mean
-gene expression as predicted by that model, deferring some algebraic details to
-Appendix~\ref{sec:non_bursty}. What we will find is that all cartoons collapse
-to a single master curve as shown in \fig{fig1:means_cartoons}(D), which
-contains just two parameters. We label the parameters $\Delta F_R$, an effective
-free energy parametrizing the repressor-DNA interaction, and $\rho$, which
-subsumes all details of transcription in the absence of repressors. We will
-offer some intuition for why this master curve exists and discuss why at the
-level of the mean expression, we are unable to discriminate ``right'' from
-``wrong'' cartoons given only measurements of fold-changes in expression.
+Figure~\ref{fig1:means_cartoons}(B) and (C) depicts a smorgasbord of
+mathematicized cartoons for simple repression using both thermodynamic and
+kinetic models, respectively, that have appeared in previous literature. For
+each cartoon, we calculate the fold-change in mean gene expression as predicted
+by that model, deferring most algebraic details to
+Appendix~\ref{sec:non_bursty}. What we will find is that for all cartoons the
+fold-change can be written as a Fermi function 
+\begin{equation}
+ \text{fold-change} = \left( 1 + \exp(-\Delta F_R + \log(\rho))  \right)^{-1},
+ \label{eq:deltaFR_eq_noneq_equiv}
+\end{equation}
+where the effective free energy contains two terms: the parameters $\Delta F_R$,
+an effective free energy parametrizing the repressor-DNA interaction, and
+$\rho$, a term derived from the level of coarse-graining used to model all
+repressor-free states. In other words, the effective free energy of the Fermi
+function can be written as the additive effect of the regulation given by the
+repressor via $\Delta F_R$, and the kinetic scheme used to describe the steps
+that lead to a transcriptional event via $\log(\rho)$ (See
+Figure~\ref{fig1:means_cartoons}(D), left panel). This implies that all models
+collapse to a single master curve as shown in
+Figure~\ref{fig1:means_cartoons}(D). We will offer some intuition for why this
+master curve exists and discuss why at the level of the mean expression, we are
+unable to discriminate ``right'' from ``wrong'' cartoons given only measurements
+of fold-changes in expression.
 
@@
 
-Again, we consign all details of the derivation to Appendix~\ref{sec:non_bursty}.
-Here we just highlight the general findings for all five kinetic models. As
-already shown in Figure~\ref{fig1:means_cartoons}(C) and (D), all the kinetic
-models explored can be collapsed onto the master curve. Given that the
-repressor-bound state only connects to the rest of the promoter dynamics via its
-binding and unbinding rates, $k_R^+$ and $k_R^-$ respectively, all models can
-effectively be separated into two categories: a single repressor-bound state,
-and all other promoter states with different levels of coarse graining. This
-structure then guarantees that, at steady-state, detailed balance between these
-two groups is satisfied. What this implies is that the steady-state distribution
-of each of the non-repressor states has the same functional form with or without
-the repressor, allowing us to write the fold-change as a product of the ratio of
-the binding and unbinding rates of the promoter, and the promoter details. This
-is 
-\mrm{This is based in one of the comments from reviewer \# 2. It is empirically
-true for all models we explored, but I don't know how to prove that it is a 
-general statement for any unimaginable complex model.}
+Again, we consign all details of the derivation to
+Appendix~\ref{sec:non_bursty}. Here we just highlight the general findings for
+all five kinetic models. As already shown in Figure~\ref{fig1:means_cartoons}(C)
+and (D), all the kinetic models explored can be collapsed onto the master curve.
+Given that the repressor-bound state only connects to the rest of the promoter
+dynamics via its binding and unbinding rates, $k_R^+$ and $k_R^-$ respectively,
+all models can effectively be separated into two categories: a single
+repressor-bound state, and all other promoter states with different levels of
+coarse graining. This structure then guarantees that, at steady-state, detailed
+balance between these two groups is satisfied. What this implies is that the
+steady-state distribution of each of the non-repressor states has the same
+functional form with or without the repressor, allowing us to write the
+fold-change as a product of the ratio of the binding and unbinding rates of the
+promoter, and the promoter details. This is \mrm{This is based in one of the
+comments from reviewer \# 2. It is empirically true for all models we explored,
+but I don't know how to prove that it is a general statement for any
+unimaginable complex model.}
 \begin{align}
@@
 where $\Delta F_R \equiv -\log(k_R^+/k_R^-)$, and the functional forms of $\rho$
-for each model change as shown in Figure~\ref{fig1:means_cartoons}(C) (See
-Appendix~\ref{sec:non_bursty} for further details).
+for each model change as shown in Figure~\ref{fig1:means_cartoons}(C). Another
+intuitive way two think about these to terms is as follows: in all kinetic
+models shown in Figure~\ref{fig1:means_cartoons}(C) the repressor-bound state
+can only be reached from a single repressor-free state. The ratio of these two
+states --repressor-bound and adjacent repressor-free state-- must remain the
+same for all models, regardless of the details included in other promoter states
+if $\Delta F_R$ represents an effective free energy of the repressor binding the
+DNA operator. The presence of other states then draws probability density from
+the promoter being in either of these two states, making the ratio between the
+repressor-bound state and \textit{all} repressor-free states different. The log
+difference in this ratio is given by $\log(\rho)$. Since model 1 and model 5 of
+Figure~\ref{fig1:means_cartoons}(C) consist of a single repressor-free state,
+$\rho$ is then necessarily 1 (See Appendix~\ref{sec:non_bursty} for further
+details).
 

### 2020-10-14 doc/resubmission/section_04_bayesian_inference.tex
update to latest version of main text

@@
 
-In this section we aim to reconcile the thermodynamic and kinetic perspectives
-that have been brought to bear on the repressor-operator model. From
-Section~\ref{section_02_means}, this reconciliation depends on a key
-quantitative question as codified by Eq.~\ref{eq:deltaFR_eq_noneq_equiv}: does
-the free energy of repressor binding, as described in the thermodynamic models
-and indirectly inferred from gene expression measurements, agree with the
+In this section of the paper, we continue our program of providing one complete
+description of the entire broad sweep of studies that have been made in the
+context of the repressor-operator model, dating all the way back to the original
+work of Jacob and Monod and including the visionary quantitative work of
+M\"{u}ller-Hill and collaborators~\cite{Oehler1990} and up to more recent
+studies \cite{Garcia2011a}. In addition, the aim is to reconcile the
+thermodynamic and kinetic perspectives that have been brought to bear on this
+problem. From Section~\ref{section_02_means}, this reconciliation depends on a
+key quantitative question as codified by Eq.~\ref{eq:deltaFR_eq_noneq_equiv}:
+does the free energy of repressor binding, as described in the thermodynamic
+models and indirectly inferred from gene expression measurements, agree with the
 corresponding values of repressor binding and unbinding rates in the kinetic
@@
 problems of parameter degeneracy that model 4 in
-Figure~\ref{fig2:constit_cartoons}(A) suffered~\cite{Razo-Mejia2020}. We will
-test this thoroughly on single-cell mRNA counts for different unregulated
-promoters from Jones et.\ al.~\cite{Jones2014}, but first we consider It will be
-instructive, however, to first consider the Poisson promoter, model 1 in
-Figure~\ref{fig2:constit_cartoons}. As we alluded to earlier, since the Poisson
-distribution has a Fano factor $\nu$ strictly equal to 1, and all of the
+Figure~\ref{fig2:constit_cartoons}(A) suffered~\cite{Razo-Mejia2020}. Does this
+stand up to closer scrutiny, namely, comparison to full mRNA distributions
+rather than simply their moments? We will test this thoroughly on single-cell
+mRNA counts for different unregulated promoters from Jones et.\
+al.~\cite{Jones2014}.
+
+It will be instructive, however, to first consider the Poisson promoter, model 1
+in Figure~\ref{fig2:constit_cartoons}. As we alluded to earlier, since the
+Poisson distribution has a Fano factor $\nu$ strictly equal to 1, and all of the
 observed data in Figure~\ref{fig2:constit_cartoons}(B) has Fano factor $\nu>1$,
@@
 
-\mrm{Griffin also feels this paragraph should go. Too many times reminding the 
-reader what we are up to. I think he is right.}
 Our objective for this section will then be to assess whether or not model 5 is

### 2020-10-14 doc/resubmission/appendix_01_nonbursty.tex
more changes
fix appendix 1...again

@@
 This is enormously advantageous since we can use the theoretical tools of 
-equilibrium statisticla mechanics to model such process.
+equilibrium statistical mechanics to model such process.
 
@@
 expression level is proportional to the probability of finding the promoter in
-the transcriptionally active microstate, i.e.,
-\begin{equation}
-\text{average rate of transcription} = \sum_i r_i p_i,
-\label{eq:rate_transcription_si}
-\end{equation}
-where $i$ labels the distinct states, $p_i$ is the probability of the ith state,
-and $r_i$ is the rate of transcription of that state. As mentioned in the main
-text, these models can only deal with the mean gene expression level. This is
-because the probability space we consider is that of the state of the repressor,
-rather than both the state of the repressor and the mRNA counts. Let's now work
-through models 1 and 2 from Figure~\ref{fig1:means_cartoons}(B).
+the transcriptionally active microstate. As mentioned in the main text, these
+models can only deal with the mean gene expression level. This is because the
+probability space we consider is that of the state of the repressor, rather than
+both the state of the repressor and the mRNA counts. Let's now work through
+models 1 and 2 from Figure~\ref{fig1:means_cartoons}(B).
 
@@
 From the relative statistical weights listed in
-Figure~\ref{fig1:means_cartoons}(B), the probability $p_U$ of being in the
+Figure~\ref{fig1:means_cartoons}, the probability $p_U$ of being in the
 unbound state is
@@
 The mean rate of transcription is then given by $r p_U$, as assumed by
-Eq.~\ref{eq:rate_transcription_si}. The mean number of mRNA is
+Eq.~\ref{eq:transcrip_prop_pbound}. The mean number of mRNA is
 set by the balance of average mRNA transcription and degradation rates,
@@
 is the mRNA degradation rate, $R$ is repressor copy number, $N_{NS}$ is the
-number of nonspecific binding sites in the genome where repressors spend most of
-their time when not bound to the operator (binding site), $\beta \equiv 1/k_BT$,
-and $\Delta\varepsilon_R$ is the binding energy of a repressor to its operator
-site. The derivation of this result can be found in~\cite{Phillips2015a}.
+number of nonspecific binding sites in the genome where repressors spend most
+of their time when not bound to the operator, $\beta \equiv 1/k_BT$, and
+$\Delta\varepsilon_R$ is the binding energy of a repressor to its operator site.
+
+\paragraph{Fold-change}
 
@@
 transcriptionally active states have transcription factor bound, or put
-differently, when there is no overlap between transcription factor bound states
-and transcriptionally active states. We will see this explicitly in the 3-state
-equilibrium model below, but perhaps surprising is that an analogous result
-carries over even to the nonequilibrium models we consider later.
+differently, when there is no overlap between
+transcription factor bound states and transcriptionally
+active states. We will see this explicitly in the 3-state equilibrium model
+below, but perhaps surprising is that an analogous result carries over even to
+the kinetic models we consider later.
 
@@
 as we use it here, and in~\cite{Razo-Mejia2018} and~\cite{Chure2019} it was
-generalized to incorporate small-molecule inducers that bind the repressor. The
-effect of this generalization is, roughly speaking, simply to rescale $R$ from
-the total number of repressors to a smaller effective number of available
-repressors which are unbound by inducers. We point out that the same
-generalization can be incorporated quite easily into any of our models in
-Figure~\ref{fig1:means_cartoons} by simply rescaling the repressor copy number
-$R$ in the thermodynamic models, or equivalently $k_R^+$ in the kinetic models.
+generalized to incorporate small-molecule inducers that bind the repressor. 
+The effect of this generalization is, roughly speaking,
+simply to rescale $R$ from the total number of repressors to a smaller
+effective number of available repressors which are unbound by inducers.
+We point out that the same generalization can be incorporated quite easily
+into any of our models in Figure~\ref{fig1:means_cartoons} by simply
+rescaling the repressor copy number $R$ in the equilibrium models, or
+equivalently $k_R^+$ in the kinetic models.
 
@@
 binding energy to its specific site (the promoter) relative to an average
-nonspecific background site, and the RNAP copy number, $P$. The fold-change
-again follows immediately as
+nonspecific background site, and the RNAP copy number, $P$.
+
+\paragraph{Fold-change}
+The fold-change again follows immediately as
 \begin{align}
@@
 
-\subsection{Derivation of chemical master equation}
-\label{sec:cme_from_cartoon}
-
 Before jumping into the derivations of the general computation of the mean mRNA
@@
 other chemical master equations in this work.
+
+% \subsubsection{Nonequilibrium Model Five - Bursty Promoter}
+% The final model we consider shown in Figure~\ref{fig1:means_cartoons}(C) is an
+% intuitive analog to model 1, with just two states, repressor bound or unbound,
+% and transition rates between them of $k_R^+$ and $k_R^-$. In model 1, when in
+% the unbound state, single mRNA transcripts are produced as a Poisson process
+% with some characteristic rate $r$. The current model by contrast produces, at
+% some Poisson rate $k_i$, \textit{bursts} of mRNA transcripts.
+% The burst sizes are assumed to be geometrically distributed with
+% a mean burst size $b$, which we will motivate in Section~\ref{sec:beyond_means}
+% when we derive this model as a certain limiting case of model 4.
+
+% From this intuitive picture and by analogy to model 1, then, it should be
+% plausible that the mean mRNA level is
+% \begin{equation}
+% \langle m\rangle = \frac{k_i b}{\gamma}
+%         \left(1 + \frac{k_R^+}{k_R^-}\right)^{-1},
+% \end{equation}
+% which will turn out to be correct from a careful calculation. For now, we simply
+% note that just like model 1, the fold-change becomes
+% \begin{equation}
+% \text{fold-change} = \left(1 + \frac{k_R^+}{k_R^-}\right)^{-1}
+% \end{equation}
+% with $\rho=1$ also like model 1.
+% We will also see later how this model emerges as a natural limit of model 4.
+% \mrm{ENDS HERE}
+
+% \mrm{THIS NEEDS TO BE FIXED}
+% \subsubsection{Noise in the Two-State Promoter, RNAP Bound or Unbound.}
+
+% Our second model of constitutive transcription posits an architecture in which
+% the promoter is either empty or bound by RNAP ~\cite{Phillips2015a,
+% Phillips2019}. Here,  as shown in model 2 of Figure~\ref{fig2:constit_cartoons}(A),
+% transcription initiation results in a state transition from the bound to the
+% unbound state, reflecting the microscopic reality that an RNAP that has begun to
+% elongate a transcript is no longer available at the start site to begin another.
+% As shown in Appendix~\ref{sec:non_bursty}, the Fano factor in this model is
+% given by
+% \begin{align}
+%     \nu = 1 -
+%         \frac{r\rate{k}{P}{+}}
+%             {\left(\rate{k}{P}{+} + \rate{k}{P}{-} + r\right)
+%              \left(\gamma + \rate{k}{P}{+} + \rate{k}{P}{-} + r\right)}.
+% \label{eq:model2_fano}
+% \end{align}
+% The problem with this picture is that the Fano factor is always $<1$. To make
+%  contact with the experimental reality of $\nu>1$ as shown in
+%  Figure~\ref{fig2:constit_cartoons}(B), clearly some corrections will be needed.
+%  While this model adds an appealing element of microscopic reality, we are
+%  forced to reject it as the additional complexity is unable to capture the
+%  phenomenology of interest. Obviously the promoter state does in fact proceed
+%  through cycles of RNAP binding, initiating, and elongating, but it seems that
+%  the super-Poissonian noise in mRNA copy number we want to model must be
+%  governed by other features of the system.
+
+% \subsubsection{Noise in the Three-State Promoter, Multistep Transcription
+% Initiation and Escape.}
+
+% How might we remedy the deficits of model 2? It is known~\cite{DeHaseth1998}
+% that once RNAP initially binds the promoter region, a multitude of distinct
+% steps occur sequentially before RNAP finally escapes into the elongation phase.
+% Perhaps adding some of this mechanistic detail as shown in model 3 of
+% Figure~\ref{fig2:constit_cartoons}(A) might rescue the previous model. The next
+% simplest refinement of that model could consider open complex formation and
+% promoter escape as separate steps rather than as a single effective step. In
+% other words, we construct model 3 by adding a single extra state to model 2, and
+% we will label the two RNAP-bound states as the closed and open complexes,
+% despite the true biochemical details certainly being more complex. For example,
+% earlier work extended this model by adding an additional repressor bound state
+% and did not explicitly consider the limit with no repressor that we analyze
+% here~\cite{Mitarai2015}. Again, our goal here is not a complete accounting of
+% all the relevant biochemical detail; this is an exploratory search for the
+% important features that a model needs to include to square with the known
+% experimental reality of constitutive expression.
+
+% Unfortunately, as hinted at in earlier work~\cite{Mitarai2015}, this model too
+% has Fano factor $\nu<1$. We again leave the algebraic details for
+% Appendix~\ref{sec:non_bursty} and merely state the result that
+% \begin{align}
+% \nu = 1 - \frac{r k_P^+ k_O}{\mathcal{Z}}
+% \frac{k_P^+ + k_P^- + k_O + r + \gamma}
+%     {\mathcal{Z} + \gamma(k_P^+ + k_P^- + k_O + r) + \gamma^2},
+% \label{eq:model3_fano}
+% \end{align}
+% where we defined $\mathcal{Z} = r(k_O + k_P^-) + k_P^+(k_O + r)$ for notational
+% tidiness. This is necessarily less than 1 for arbitrary rate constants.
+
+% In fact, we suspect \textit{any} model in which transcription proceeds through a
+% multistep cycle must necessarily have $\nu<1$. The intuitive argument compares
+% the waiting time distribution to traverse the cycle with the waiting time for a
+% Poisson promoter (model 1) with the same mean time. The latter is simply an
+% exponential distribution. The former is a convolution of multiple exponentials,
+% and intuitively the waiting time distribution for a multistep process
+% should be more peaked with a smaller fractional width than a single
+% exponential with the same mean.
+% A less disperse waiting time distribution means transcription initations are
+% more uniformly distributed in time relative to a Poisson process. Hence the
+% distribution of mRNA over a population of cells should be less variable compared
+% to Poisson, giving $\nu<1$.
+% (In Appendix~\ref{sec:non_bursty} we present a more precise version
+% of the intuitive arguments in this paragraph.)
+% Regardless of the merits of this model in describing the noise properties of
+% constitutive transcription initiation, it ultimately fails the simplest
+% quantitative feature of the data, namely that the Fano factor $> 1$ and hence
+% we must discard this mechanistic picture and search elsewhere.
+% \mrm{ENDS HERE}
+
+\subsection{Derivation of chemical master equation}
+\label{sec:cme_from_cartoon}
 
@@
 
-% \subsubsection{Nonequilibrium Model Four - ``Active'' and ``Inactive'' States}
-% Model 4 in Figure~\ref{fig1:means_cartoons}(C) is at the core of the theory
-% in~\cite{Razo-Mejia2020}. At a glance the cartoon for this model may appear very
-% similar to model 2, and mathematically it is, but the interpretation is rather
-% different. In model 2, we interpreted the third state literally as an RNAP-bound
-% promoter and modeled initiation of a transcript as triggering a promoter state
-% change, making the assumption that an RNAP can only
-% make one transcript at a time. In contrast, in the present model the promoter
-% state does \textit{not} change when a transcript is initiated. So we no longer
-% interpret these states as literally RNAP bound and unbound but instead as
-% coarse-grained ``active'' and ``inactive'' states, the details of which we leave
-% unspecified for now. We will comment more on this model below when we discuss
-% Fano factors of models.
-
-% Mathematically this model is very similar to models 1 and 2. Like model 1, the
-% matrix $R$ describing transcription initiation is diagonal, namely
-% \begin{equation}
-% \mathbf{R} = \begin{pmatrix}
-%                 0 & 0 & 0 \\ 
-%                 0 & 0 & 0 \\ 
-%                 0 & 0 & r
-%         \end{pmatrix}.
-% \end{equation}
-% The master equation takes verbatim the same form as it did for model 1,
-% Eq.~\ref{eq:2state_rep_cme}. Meanwhile the promoter transition
-% matrix $\mathbf{K}$ is the same as Eq.~\ref{eq:3state_cme_matrices_pt1}
-% from model 2, although we relabel the
-% rate constants from $k_P^\pm$ to $k^\pm$ to reiterate that these are not simply
-% RNAP binding and unbinding rates.
-
-% Carrying out the algebra, the mean mRNA can be found to be
-% \begin{equation}
-% \langle m\rangle = \frac{r}{\gamma}
-% \frac{k_R^- k^+}
-% {k_R^- k^+ + k_R^- k^- + k_R^+ k^-},
-% \end{equation}
-% and the fold-change readily follows,
-% \begin{align}
-% \text{fold-change}
-% &=      \frac{k_R^- k^+}{k_R^- k^+ + k_R^- k^- + k_R^+ k^-}
-%         \frac{k_R^- k^+ + k_R^- k^-}{k_R^- k^+}
-% \\
-% &=      \left(1 + \frac{k_R^+}{k_R^-}
-%                 \left(1 + \frac{k^+}{k^-}\right)^{-1}
-%         \right)^{-1},
-% \end{align}
-% from which we see $\rho = 1 + k^+/k^-$ as shown in~\fig{fig1:means_cartoons}(C).
-
-% \subsubsection{Nonequilibrium Model Five - Bursty Promoter}
-% The final model we consider shown in Figure~\ref{fig1:means_cartoons}(C) is an
-% intuitive analog to model 1, with just two states, repressor bound or unbound,
-% and transition rates between them of $k_R^+$ and $k_R^-$. In model 1, when in
-% the unbound state, single mRNA transcripts are produced as a Poisson process
-% with some characteristic rate $r$. The current model by contrast produces, at
-% some Poisson rate $k_i$, \textit{bursts} of mRNA transcripts.
-% The burst sizes are assumed to be geometrically distributed with
-% a mean burst size $b$, which we will motivate in Section~\ref{sec:beyond_means}
-% when we derive this model as a certain limiting case of model 4.
-
-% From this intuitive picture and by analogy to model 1, then, it should be
-% plausible that the mean mRNA level is
-% \begin{equation}
-% \langle m\rangle = \frac{k_i b}{\gamma}
-%         \left(1 + \frac{k_R^+}{k_R^-}\right)^{-1},
-% \end{equation}
-% which will turn out to be correct from a careful calculation. For now, we simply
-% note that just like model 1, the fold-change becomes
-% \begin{equation}
-% \text{fold-change} = \left(1 + \frac{k_R^+}{k_R^-}\right)^{-1}
-% \end{equation}
-% with $\rho=1$ also like model 1.
-% We will also see later how this model emerges as a natural limit of model 4.
-% \mrm{ENDS HERE}
-
-% \mrm{THIS NEEDS TO BE FIXED}
-% \subsubsection{Noise in the Two-State Promoter, RNAP Bound or Unbound.}
-
-% Our second model of constitutive transcription posits an architecture in which
-% the promoter is either empty or bound by RNAP ~\cite{Phillips2015a,
-% Phillips2019}. Here,  as shown in model 2 of Figure~\ref{fig2:constit_cartoons}(A),
-% transcription initiation results in a state transition from the bound to the
-% unbound state, reflecting the microscopic reality that an RNAP that has begun to
-% elongate a transcript is no longer available at the start site to begin another.
-% As shown in Appendix~\ref{sec:non_bursty}, the Fano factor in this model is
-% given by
-% \begin{align}
-%     \nu = 1 -
-%         \frac{r\rate{k}{P}{+}}
-%             {\left(\rate{k}{P}{+} + \rate{k}{P}{-} + r\right)
-%              \left(\gamma + \rate{k}{P}{+} + \rate{k}{P}{-} + r\right)}.
-% \label{eq:model2_fano}
-% \end{align}
-% The problem with this picture is that the Fano factor is always $<1$. To make
-%  contact with the experimental reality of $\nu>1$ as shown in
-%  Figure~\ref{fig2:constit_cartoons}(B), clearly some corrections will be needed.
-%  While this model adds an appealing element of microscopic reality, we are
-%  forced to reject it as the additional complexity is unable to capture the
-%  phenomenology of interest. Obviously the promoter state does in fact proceed
-%  through cycles of RNAP binding, initiating, and elongating, but it seems that
-%  the super-Poissonian noise in mRNA copy number we want to model must be
-%  governed by other features of the system.
-
-% \subsubsection{Noise in the Three-State Promoter, Multistep Transcription
-% Initiation and Escape.}
-
-% How might we remedy the deficits of model 2? It is known~\cite{DeHaseth1998}
-% that once RNAP initially binds the promoter region, a multitude of distinct
-% steps occur sequentially before RNAP finally escapes into the elongation phase.
-% Perhaps adding some of this mechanistic detail as shown in model 3 of
-% Figure~\ref{fig2:constit_cartoons}(A) might rescue the previous model. The next
-% simplest refinement of that model could consider open complex formation and
-% promoter escape as separate steps rather than as a single effective step. In
-% other words, we construct model 3 by adding a single extra state to model 2, and
-% we will label the two RNAP-bound states as the closed and open complexes,
-% despite the true biochemical details certainly being more complex. For example,
-% earlier work extended this model by adding an additional repressor bound state
-% and did not explicitly consider the limit with no repressor that we analyze
-% here~\cite{Mitarai2015}. Again, our goal here is not a complete accounting of
-% all the relevant biochemical detail; this is an exploratory search for the
-% important features that a model needs to include to square with the known
-% experimental reality of constitutive expression.
-
-% Unfortunately, as hinted at in earlier work~\cite{Mitarai2015}, this model too
-% has Fano factor $\nu<1$. We again leave the algebraic details for
-% Appendix~\ref{sec:non_bursty} and merely state the result that
-% \begin{align}
-% \nu = 1 - \frac{r k_P^+ k_O}{\mathcal{Z}}
-% \frac{k_P^+ + k_P^- + k_O + r + \gamma}
-%     {\mathcal{Z} + \gamma(k_P^+ + k_P^- + k_O + r) + \gamma^2},
-% \label{eq:model3_fano}
-% \end{align}
-% where we defined $\mathcal{Z} = r(k_O + k_P^-) + k_P^+(k_O + r)$ for notational
-% tidiness. This is necessarily less than 1 for arbitrary rate constants.
-
-% In fact, we suspect \textit{any} model in which transcription proceeds through a
-% multistep cycle must necessarily have $\nu<1$. The intuitive argument compares
-% the waiting time distribution to traverse the cycle with the waiting time for a
-% Poisson promoter (model 1) with the same mean time. The latter is simply an
-% exponential distribution. The former is a convolution of multiple exponentials,
-% and intuitively the waiting time distribution for a multistep process
-% should be more peaked with a smaller fractional width than a single
-% exponential with the same mean.
-% A less disperse waiting time distribution means transcription initations are
-% more uniformly distributed in time relative to a Poisson process. Hence the
-% distribution of mRNA over a population of cells should be less variable compared
-% to Poisson, giving $\nu<1$.
-% (In Appendix~\ref{sec:non_bursty} we present a more precise version
-% of the intuitive arguments in this paragraph.)
-% Regardless of the merits of this model in describing the noise properties of
-% constitutive transcription initiation, it ultimately fails the simplest
-% quantitative feature of the data, namely that the Fano factor $> 1$ and hence
-% we must discard this mechanistic picture and search elsewhere.
-% \mrm{ENDS HERE}
-
 \subsection{Matrix form of the multi-state chemical master equation}
@@
 models, the natural mathematicization of their cartoons is as a chemical master
-equation such as the coupled chemical master equations we derived for model 1,
-which for completeness we reproduce here as
+equation such as Eq.~\ref{eq:2state_rep_cme} for model 1, which for
+completeness we reproduce here as
 \begin{align}
@@
 
-Next consider the cartoon for nonequilibrium model 2 in
+Next consider the cartoon for kinetic model 2 in
 Figure~\ref{fig1:means_cartoons}(C). Now we must track probabilities $p_R$,
@@
 Although we derived Eq.~\ref{eq:generic_cme_appdx} for the particular case of
-kinetic model 2 in Figure~\ref{fig1:means_cartoons}, in fact the chemical master
-equations for all of the kinetic models in Figure~\ref{fig1:means_cartoons}
-except for model 5 can be cast in this form. (We treat model 5 separately in
-Appendix~\ref{sec:gen_fcn_appdx}.) Model 3 introduces no new subtleties beyond
-model 2 and Eq.~\ref{eq:generic_cme_appdx} applies equally well to it, simply
-with different matrices of dimension four instead of three. Models 1 and 4 are
-both handled by a special case of Eq.~\ref{eq:generic_cme_appdx} where
-$\mathbf{R_D} = \mathbf{R_A} \equiv \mathbf{R}$, since in these two models
-transcription initiation events do not change promoter state.
+kinetic model 2 in Figure~\ref{fig1:means_cartoons}, in fact the
+chemical master equations for all of the kinetic models in
+Figure~\ref{fig1:means_cartoons} except for model 5 can be cast in this form.
+(We treat model 5 separately in Appendix~\ref{sec:gen_fcn_appdx}.) Model
+3 introduces no new subtleties beyond model 2 and
+Eq.~\ref{eq:generic_cme_appdx} applies equally well to it, simply with different
+matrices of dimension four instead of three. Models 1 and 4 are both
+handled by Eq.~\ref{eq:2state_rep_cme} in the main text, which is just
+Eq.~\ref{eq:generic_cme_appdx} except in the special case of $\mathbf{R_D} =
+\mathbf{R_A} \equiv \mathbf{R}$, since in these two models transcription
+initiation events do not change promoter state.
 
 Recalling that our goal in this section is to derive expressions for mean mRNA
-and Fano factor for nonequilibrium models 1 through four in
+and Fano factor for kinetic models 1 through four in
 Figure~\ref{fig1:means_cartoons}, and since all four of these models are
@@
 \subsection{Kinetic Model One - Poisson Promoter}
-\subsubsection{Mean mRNA}
-For kinetic model 1 in Figure~\ref{fig1:means_cartoons}, we have already shown
-the full master equation in Eq.~\ref{eq:poisson_promoter_cme_appdx}, but for
-completeness we reprint it again as
+
+This first model (model 1 in Figure~\ref{fig1:means_cartoons}) is the one that
+we used in the previous section to motivate the derivation of the chemical
+master equation. Eq.~\ref{eq:poisson_promoter_cme_appdx} shows the two coupled
+master equations, but for completeness we reprint it again as
 \begin{align}
@@
 \end{split}
-\label{eq:2state_rep_cme}
+\label{eq:model1_kinetic_cme}
 \end{align}
 This is a direct transcription of the states and rates in
-Figure~\ref{fig1:means_cartoons}(C). This may be converted to the matrix form of
+Figure~\ref{fig1:means_cartoons}. This may be converted to the matrix form of
 the master equation shown in Eq.~\ref{eq:generic_cme_appdx} with matrices
@@
 
-First we need $\langle\vec{m}^0\rangle$. Label its components as $p_R$ and
-$p_U$, the probabilities of finding the system in either promoter state, and
-note that only $p_U$ survives multiplication by $\mathbf{R}$, since
+\subsubsection{Mean mRNA}
+To compute the mean mRNA we need $\langle\vec{m}^0\rangle$. Label its components
+as $p_R$ and $p_U$, the probabilities of finding the system in either promoter
+state, and note that only $p_U$ survives multiplication by $\mathbf{R}$, since
 \begin{equation}
@@
 
-\subsubsection{Fold-Change}
-$k_R^+$ is proportional to the repressor copy number, so in computing
-fold-change, absence of repressor corresponds to $k_R^+\rightarrow0$. Therefore
-fold-change in this model is simply
-\begin{equation}
-\text{fold-change} = \left(1 + \frac{k_R^+}{k_R^-}\right)^{-1},
-\end{equation}
-matching the master curve of~\fig{fig1:means_cartoons}(D) with $\rho=1$.
+\subsubsection{Fold-change}
+The fold-change for this model is easily computed since the unregulated promoter
+has a mean mRNA $\langle m \rangle = r / \gamma$, resulting in a fold-change of
+the form
+\begin{equation}
+\text{fold-change} = \frac{k_R^-}{k_R^- + k_R^+}.
+\end{equation}
+This can be rewritten as a Fermi function of the form
+\begin{equation}
+\text{fold-change} = \frac{1}{1 + \exp[\log(k_R^+ / k_R^-)]},
+\end{equation}
+giving $\rho = 1$ as indicated in Figure~\ref{fig1:means_cartoons}.
 
@@
 
-\subsection{Kinetic Model Two - RNAP Bound and Unbound States}
+\subsection{Kinetic Model Two - RNAP Bound and Unbound States with RNAP escape}
 
 Our second kinetic model depicted in Figure~\ref{fig1:means_cartoons}(C) mirrors
-the second thermodynamic model of Figure~\ref{fig1:means_cartoons}(B) by
-fine-graining  the repressor unbound state of nonequilibrium model 1, resolving
-it into an empty promoter state and an RNAP-bound state. Note in this picture,
-in contrast with model 4 below, transcription initiation is accompanied by a
+the second equilibrium model of Figure~\ref{fig1:means_cartoons}(B) by
+fine-graining  the repressor unbound state of kinetic model 1, resolving it into
+an empty promoter state and an RNAP-bound state. Note in this picture, in
+contrast with model 4, transcription initiation is accompanied by a
 promoter state change, in keeping with the interpretation as RNAP-bound and
@@
 The master equation governing this model is analogous to
-Eq.~\ref{eq:2state_rep_cme} for model 1 above. The main subtlety arises since
-transcription initiation accompanies a promoter state change. This can be
-understood by analogy to $\mathbf{K}$. The off-diagonal and diagonal elements of
-$\mathbf{K}$ correspond to transitions arriving at or departing from,
-respectively, the promoter state of interest. If transcription initiation is
-accompanied by promoter state changes, we must have separate matrices for
-arriving and departing transcription events since the arriving and departing
-transitions have different initial copy numbers of mRNA, unlike for $\mathbf{K}$
-where they are the same. The master equation for this model is given by three
-coupled differential equations of the form
+Eq.~\ref{eq:model1_kinetic_cme} for model 1 above. The main subtlety arises
+since transcription initiation accompanies a promoter state change. The full
+master equation for model 2 in Figure~\ref{fig1:means_cartoons} is then of the
+form
 \begin{align}
@@
 + \overbrace{(m+1)\gamma p_E(m+1,t)}^{m + 1 \rightarrow m}
-- \overbrace{\gamma mp_E(m,t)}^{m \rightarrow m - 1}.
+- \overbrace{\gamma mp_E(m,t)}^{m \rightarrow m - 1}
 \\
@@
 - \overbrace{k_P^+ p_E(m,t)}^{U \rightarrow A}
-+ \overbrace{rp_P(m-1,t)}^{m-1 \rightarrow m,\;A \rightarrow U}
++ \overbrace{rp_P(m-1,t)}^{m-1 \rightarrow m,\;A \rightarrow U},
 \\
@@
 + \overbrace{(m+1)\gamma p_P(m+1,t)}^{m + 1 \rightarrow m}
-- \overbrace{\gamma mp_P(m,t)}^{m \rightarrow m - 1}.
+- \overbrace{\gamma mp_P(m,t)}^{m \rightarrow m - 1}
 \\
@@
 \end{align}
-We can rewrite this equation using the matrix format as
+
+To write down this in matrix notation one needs to be careful on how we define
+the matrices involving a transcriptional event. The off-diagonal and diagonal
+elements of $\mathbf{K}$ correspond to transitions arriving at or departing
+from, respectively, the promoter state of interest. If transcription initiation
+is accompanied by promoter state changes, we must have separate matrices for
+arriving and departing transcription events since the arriving and departing
+transitions have different initial copy numbers of mRNA, unlike for $\mathbf{K}$
+where they are the same. The master equation for this model is
 \begin{equation}
@@
 
-\subsubsection{Fold-Change}
-Fold-change is again found from the ratio prescribed by Eq.~\ref{eq:fc_def},
-from which we have
+\subsubsection{Fold-change}
+Fold-change is again found from the ratio prescribed by Eq.~\ref{eq:fc_def} in
+the main text, from which we have
 \begin{align}
@@
 \begin{equation}
-\nu = 1 - \frac{r k_P^+ } {(k_P^+ + k_P^- + r)(k_P^+ + k_P^- + r + \gamma)}.
-\end{equation}
+\nu = 1 - \frac{r k_P^+ } {(k_P^+ + k_P^- + r)(k_P^+ + k_P^- + r + \gamma)},
+\end{equation}
+exactly Eq.~\ref{eq:model2_fano} in the main text.
 
@@
 Escape}
-
-One might reasonably complain that the first two ``nonequilibrium'' models we
+One might reasonably complain that the first two ``kinetic'' models we
 have considered are straw men. Their steady states necessarily satisfy detailed
@@
 additional complexity is essential to properly account for the noise in the mRNA
-distribution. As mentioned in the main text, we weigh on this conclusion based
-on the fact that this model cannot reproduce the observable cell-to-cell 
-variability in mRNA copy number.
+distribution. We will weigh in on both models later when we consider observables
+beyond fold-change.
 
@@
 \end{equation}
-again closely analogous to nonequilibrium model 2.
+again closely analogous to kinetic model 2.
 
 \subsubsection{Mean mRNA}
-In close analogy to model 2 above, $\langle\vec{m}^0\rangle$ is again given by
-Eq.~\ref{eq:vecm0_appdx_ref}, which in this case takes the form
+The expression for mean mRNA is substantially more complicated now.
+$\langle\vec{m}^0\rangle$ is again given by Eq.~\ref{eq:vecm0_appdx_ref}, which
+in this case takes the form
 \begin{equation}
@@
     \frac{k_P^+ k_R^- k_O}{k_P^+ k_R^- k_O + r k_P^+ k_R^- +
-                            r (k_P^- + k_O) (k_R^+ + k_R^-)},
-\end{equation}
-which equals Eq.~\ref{eq:model3_mean_m} in the main text.
-
-\subsubsection{Fold-Change}
-The mean mRNA is much messier than for the other models. The strategy to compute
-the fold-change is to isolate the terms involving the repressor, so that now the
+                            r (k_P^- + k_O) (k_R^+ + k_R^-)}.
+\end{equation}
+which can be simplified to
+\begin{equation}
+\langle m\rangle
+= \frac{r}{\gamma}
+\frac{\frac{k_P^+ k_O}{r(k_O + k_P^-)}}
+        {1 + \frac{k_P^+ (k_O + r)}{r(k_O + k_P^-)} + \frac{k_R^+}{k_R^-}}.
+\end{equation}
+
+\subsubsection{Fold-change}
+
+The strategy is to isolate the terms involving the repressor, so that now the
 fold-change is seen to be simply
@@
 again, with $\rho = 1 + \frac{k_P^+ (k_O + r)}{r(k_O + k_P^-)}$.
-
-This example hints that an arbitrarily fine-grained model of downstream
-transcription steps may still be collapsed to the form of the master curve for
-the means given in Figure~\ref{fig1:means_cartoons}(D), so long as the repressor
-binding is exclusive with transcriptionally active states. We offer this
-argument as a conjecture, and we suspect that a careful argument using the
-graph-theoretic framework based on the Matrix-Tree
-Theorem~\cite{Gunawardena2012}, might furnish a ``proof'' not with some
-challenges as this theoretical framework applies, at the present, to finite
-graphs.
 
@@
 \subsubsection{Generalizing $\nu<1$ to more fine-grained models}
-In the main text we argued that the convolution of multiple
-exponential distributions should be a distribution with a smaller
-fractional width than the corresponding exponential distribution
-with the same mean.
-This can be made more precise with a result from~\cite{Stewart2007},
-who showed that the convolution of multiple gamma distributions (of which the
-exponential distribution is a special case) is, to a very good approximation,
-also gamma distributed. Using their Eq.~2 for the distribution of the
-convolution, with shape parameters set to 1 to give exponential distributions,
-the total waiting time distribution has a ratio of variance to squared mean
-$\sigma^2/\mu^2 = \sum_i k_i^2/\left(\sum_i k_i\right)^2$, where the $k_i$ are
-the rates of the individual steps. Clearly this is less than 1 and therefore the
-total waiting time distribution is narrower than the corresponding exponential.
+In the main text we argued that the convolution of multiple exponential
+distributions should be a distribution with a smaller fractional width than the
+corresponding exponential distribution with the same mean. This can be made more
+precise with a result from~\cite{Stewart2007}, who showed that the convolution
+of multiple gamma distributions (of which the exponential distribution is a
+special case) is, to a very good approximation, also gamma distributed. Using
+their Eq.~2 for the distribution of the convolution, with shape parameters set
+to 1 to give exponential distributions, the total waiting time distribution has
+a ratio of variance to squared mean $\sigma^2/\mu^2 = \sum_i k_i^2/\left(\sum_i
+k_i\right)^2$, where the $k_i$ are the rates of the individual steps. Clearly
+this is less than 1 and therefore the total waiting time distribution is
+narrower than the corresponding exponential.
 
 We also claimed in the main text that for a process whose waiting time
-distribution is narrower than exponential, i.e., has $\sigma^2/\mu^2<1$,
-the distribution of counts should be less variable than a Poisson
-distribution, leading to a Fano factor $\nu<1$.
-This we argue by analogy to photon statistics where it is known that
-``antibunched'' arrivals, in other words more uniformly distributed in
-time relative to uncorrelated arrivals, generally gives rise to
-sub-Poissonian noise~\cite{Paul1982, Zou1990}. Although loopholes to this
-result are known to exist, to our knowledge they appear to arise from
-uniquely quantum effects so we do not expect they apply for our problem.
-Nevertheless we refrain from elevating this equivalence of
-kinetic cycles with sub-Poissonian noise to a ``theorem.''
+distribution is narrower than exponential, i.e., has $\sigma^2/\mu^2<1$, the
+distribution of counts should be less variable than a Poisson distribution,
+leading to a Fano factor $\nu<1$. This we argue by analogy to photon statistics
+where it is known that ``antibunched'' arrivals, in other words more uniformly
+distributed in time relative to uncorrelated arrivals, generally gives rise to
+sub-Poissonian noise~\cite{Paul1982, Zou1990}. Although loopholes to this result
+are known to exist, to our knowledge they appear to arise from uniquely quantum
+effects so we do not expect they apply for our problem. Nevertheless we refrain
+from elevating this equivalence of kinetic cycles with sub-Poissonian noise to a
+``theorem.''
 
 \subsection{Kinetic Model Four - ``Active'' and ``Inactive'' States}
+
 Model 4 in Figure~\ref{fig1:means_cartoons}(C) is at the core of the theory
@@
 The master equation takes verbatim the same form as it did for model 1,
-Eq.~\ref{eq:2state_rep_cme}. Meanwhile the promoter transition
-matrix $\mathbf{K}$ is the same as Eq.~\ref{eq:3state_cme_matrices_pt1}
-from model 2, although we relabel the
-rate constants from $k_P^\pm$ to $k^\pm$ to reiterate that these are not simply
-RNAP binding and unbinding rates.
+Eq.~\ref{eq:2state_rep_cme}. Meanwhile the promoter transition matrix
+$\mathbf{K}$ is the same as Eq.~\ref{eq:3state_cme_matrices_pt1} from model 2,
+although we relabel the rate constants from $k_P^\pm$ to $k^\pm$ to reiterate
+that these are not simply RNAP binding and unbinding rates.
+
+% Carrying out the algebra, the mean mRNA can be found to be
+% \begin{equation}
+% \langle m\rangle = \frac{r}{\gamma}
+% \frac{k_R^- k^+}
+% {k_R^- k^+ + k_R^- k^- + k_R^+ k^-},
+% \end{equation}
+% and
 
@@
 is that now $\mathbf{R_A}=\mathbf{R_D}$, i.e., both are diagonal, in contrast to
-model 2 where $\mathbf{R_A}$ has an off-diagonal element, as in
-Eq.~\ref{eq:model2_matrices_appdx}. Then the analog of
+model 2 where $\mathbf{R_A}$ has an off-diagonal element. Then the analog of
 Eq.~\ref{eq:model2_K-R+R_for_m0} for finding $\langle{m}^0\rangle$ is
@@
 
-\subsubsection{Fold-Change}
-
+\subsubsection{Fold-change}
 The fold-change readily follows,
@@
 Consider the constitutive model four from Figure~\ref{fig2:constit_cartoons} for
-which we want to compute the Fano factor and compare it to nonequilibrium model
+which we want to compute the Fano factor and compare it to kinetic model
 one of simple repression in Figure~\ref{fig1:means_cartoons}. Mathematically
@@
 meaning of the promoter states interpreted differently. Furthermore,
-nonequilibrium model 1 from Figure~\ref{fig1:means_cartoons} was the model
+kinetic model 1 from Figure~\ref{fig1:means_cartoons} was the model
 considered by Jones et.\ al.~\cite{Jones2014}, where they derived the Fano
@@
 \begin{equation}
-\nu = 1 + \frac{r k^-}{(k^- + k^+)(k^- + k^+ + \gamma)}.
-\end{equation}
+\nu = 1 + \frac{r k^-}{(k^- + k^+)(k^- + k^+ + \gamma)},
+\end{equation}
+as quoted in Eq.~\ref{eq:model4_fano} and in Figure~\ref{fig2:constit_cartoons}.

### 2020-10-14 doc/resubmission/section_02_means.tex
change noneq for kinetic
implement Rob's changes to main text

@@
 continuous number of promoter states (due to effects such as supercoiling of
-DNA~\cite{Chong2014, Sevier2016}  or DNA looping~\cite{Boedicker2013a}) is
+DNA~\cite{Chong2014, Sevier2016} or DNA looping \cite{Boedicker2013a}) is
 analogous to how the Monod-Wyman-Changeux model of allostery coarse-grains
@@
 by that model, deferring most algebraic details to
-Appendix~\ref{sec:non_bursty}. What we will find is that for all cartoons the
-fold-change can be written as a Fermi function 
+Appendix~\ref{sec:non_bursty}. What we will find is that
+for all cartoons the fold-change can be written as a Fermi function 
 \begin{equation}
@@
 that lead to a transcriptional event via $\log(\rho)$ (See
-Figure~\ref{fig1:means_cartoons}(D), left panel). This implies that all models
+Figure~~\ref{fig1:means_cartoons}(D), left panel). This implies all models
 collapse to a single master curve as shown in
@@
 for each model change as shown in Figure~\ref{fig1:means_cartoons}(C). Another
-intuitive way two think about these to terms is as follows: in all kinetic
-models shown in Figure~\ref{fig1:means_cartoons}(C) the repressor-bound state
-can only be reached from a single repressor-free state. The ratio of these two
-states --repressor-bound and adjacent repressor-free state-- must remain the
-same for all models, regardless of the details included in other promoter states
-if $\Delta F_R$ represents an effective free energy of the repressor binding the
-DNA operator. The presence of other states then draws probability density from
-the promoter being in either of these two states, making the ratio between the
-repressor-bound state and \textit{all} repressor-free states different. The log
-difference in this ratio is given by $\log(\rho)$. Since model 1 and model 5 of
-Figure~\ref{fig1:means_cartoons}(C) consist of a single repressor-free state,
-$\rho$ is then necessarily 1 (See Appendix~\ref{sec:non_bursty} for further
-details).
+intuitive way to think about these two terms is as follows: in all kinetic
+models shown in Figure~1(C) the repressor-bound state can only be reached from a
+single repressor-free state. The ratio of these two states --repressor-bound and
+adjacent repressor-free state-- must remain the same for all models, regardless
+of the details included in other promoter states if $\Delta F_R$ represents an
+effective free energy of the repressor binding the DNA operator. The presence of
+other states then draws probability density from the promoter being in either of
+these two states, making the ratio between the repressor-bound state and
+\textit{all} repressor-free states different. The log difference in this ratio
+is given by $\log(\rho)$. Since model 1 and model 5 of Figure~1(C) consist of a
+single repressor-free state, $\rho$ is then necessarily 1 (See
+Appendix~\ref{sec:non_bursty} for further details).
 
@@
 % absence of repressors. This was done in~\cite{Razo-Mejia2020} by fitting
-% nonequilibrium model 4 in~\fig{fig1:means_cartoons}(C), with zero repressor
+% kinetic model 4 in~\fig{fig1:means_cartoons}(C), with zero repressor
 % (looking ahead, this is equivalent to model 4

### 2020-10-14 doc/resubmission/section_05_discussion.tex
implement Rob's changes to main text

@@
 level of mean mRNA count is a consequence of the possible transitions between
-promoter microstates. The promoter microstates in all models shown in
-Figure~\ref{figure1:means_cartoons} can explicitly be separated into two groups:
-1) promoters with repressor bound, and 2) all other promoters states. The single
-path between these two groups of states guarantees detailed-balance between the
-groups. We can therefore separate \textit{expression} related transitions, which
-by definition must be out of equilibrium, with \textit{regulation} transitions
-that can in principle be in equilibrium. This implies that the probability
-distribution of the promoter microstates not involving the repressor take the
-same functional form, allowing us to write the fold-change as an effective free
-energy involving a regulation term, $\Delta F_R$, and a promoter details term,
-$\rho$. We offer this argument as a conjecture, and we suspect that a careful
-argument using the graph-theoretic framework based on the Matrix-Tree
+promoter microstates. The promoter microstates in all models shown in Figure~1
+can explicitly be separated into two groups: 1) promoters with repressor bound,
+and 2) all other promoters states. The single path between these two groups of
+states guarantees detailed-balance between the groups. We can therefore separate
+\textit{expression} related transitions, which by definition must be out of
+equilibrium, with \textit{regulation} transitions that can in principle be in
+equilibrium. This implies that the probability distribution of the promoter
+microstates not involving the repressor take the same functional form, allowing
+us to write the fold-change as an effective free energy involving a regulation
+term, $\Delta F_R$, and a promoter details term, $\rho$. We offer this argument
+as a conjecture, and we suspect that a careful argument using the
+graph-theoretic framework based on the Matrix-Tree
 Theorem~\cite{Gunawardena2012}, might furnish a ``proof'' not with some
@@
 
-It is important to mention the potential problems present in all models,
-including model 5 from Figure~\ref{fig2:constit_cartoons}(A). Irreversible
-reactions as the ones used in all transcription events lead to physical
-paradoxes such as infinite entropy production rates. These are mathematical
-conveniences rather than physical reality, leaving the inferred rates for the
-transcription burst rate and burst size as an effective rate with lots of
-(potentially) reversible processes. Also, the state of the art modeling imagines
-the promoter existing in a series of discrete states. This assumption might not
-be entirely true as effects such as DNA supercoiling~\cite{Chong2014,Sevier2016}
-or DNA looping~\cite{Boedicker2013a} can be considered continuous
-variables; although there are formal ways of integrating out such degrees of
-freedom. Furthermore, when computing the fold-change in gene expression
-(Section~\ref{section_02_means}) we assume that the translation efficiency of
-mRNA transcript is equal in cells with and without regulation. There might be
-some known biological effects that could limit the applicability of this
-assumption such as cooperation between the transcription and translation 
+It is important to be clear on the nature of the assumptions  present in all
+models, including model 5 from Figure~\ref{fig2:constit_cartoons}(A). First, as
+with many of the models used routinely in the analysis of transcription, the
+models considered here imagine the promoter as existing in a series of discrete
+states. This assumption is clearly an oversimplification due to effects such as
+DNA supercoiling~\cite{Chong2014,Sevier2016} or DNA
+looping~\cite{Boedicker2013a} which are both parameterized by  continuous
+variables describing DNA shape; although there are formal and very interesting
+ways of integrating out such degrees of freedom. Another facet of the models
+used here is that they can feature irreversible processes characterized by only
+a single rate process with no corresponding return pathway at odds with
+thermodynamics. These irreversible reactions lead to physical paradoxes such as
+infinite entropy production rates. Such models are mathematically convenient and
+serve only as an approximation to the more realistic situation in which the
+forward rates are much larger than the backward rates. One consequence is that
+the inferred rates for the transcription burst rate and burst size  can only be
+thought of as effective rate constants. Another idealization of the models
+considered here is that when computing the fold-change in gene expression
+(Section 2) we assume that the translation efficiency of an mRNA transcript is
+the same in cells with and without regulation. There might be some known
+biological effects that could limit the applicability of this assumption such as
+cooperation between the transcription and translation
 machinery~\cite{Proshkin2010}, or coupling of mRNA transcription and
@@
 changes in gene copy number during the cell cycle have no effects on the
-expression profile. For moderate growth rates as the ones used 
-in~\cite{Jones2014} from where we obtained the data, there are at any point at
-most two copies of the gene. On a recent publication, we explored the effects
-that this change in gene dosage can have in gene expression, finding that the
-effects are much more drastic at the protein level~\cite{Razo-Mejia2020}.
+expression profile. For moderate growth rates such as those used
+in~\cite{Jones2014} from which we obtained the data, at all points during the
+cell cycle there are at most two copies of the gene of interest. In a recent
+publication, we explored the effects that this change in gene dosage can have in
+gene expression, finding that the effects are much more drastic at the protein
+level~\cite{Razo-Mejia2020}.
 

### 2020-10-15 doc/resubmission/section_01_introduction.tex
implement most of RP comments from 20201015

--- 2020-09-24 doc/resubmission/section_01_introduction.tex
@@
 Belliveau2018, Ghatak2019, Santos_Zavaleta2019}. But even once we have
-established the identity of key transcription factors and their binding sites of
-a given promoter architecture, there remains the predictive challenge of
+established the identity of key transcription factors and their binding sites
+for a given promoter architecture, there remains the predictive challenge of
 understanding its input-output properties, an objective that can be met by a
@@
 dissection of transcription \cite{Oehler1994, Oehler1990}. One of the main
-features of of the models we explore is that, by construction, all features
-related to the microstates in which the repressor is bound to the promoter can
-be separated from the microstates in which the RNA polymerase (RNAP) is bound.
-From a modeling perspective, this means that some of the models can be written
-as effective two-state models for which there is a rich
+features of the models we explore is that, by construction, all features related
+to the microstates in which the repressor is bound to the promoter can be
+separated from the microstates in which the RNA polymerase (RNAP) is bound. From
+a modeling perspective, this means that some of the models can be written as
+effective two-state models for which there is a rich
 literature~\cite{Peccoud1995, Shahrezaei2008, Iyer-Biswas2009, Tkacik2009,

### 2020-10-15 doc/resubmission/section_02_means.tex
implement most of RP comments from 20201015

@@
 
-The first class of models that are the focus of the present section on
-predicting the mean level of gene expression, sometimes known as thermodynamic
-models, invoke the tools of equilibrium statistical mechanics to compute the
-probabilities of the promoter microstates~\cite{Ackers1982, Shea1985,
-Buchler2003, Vilar2003a, Vilar2003b, Bintu2005a, Bintu2005c, Gertz2009,
-Sherman2012, Saiz2013}. As seen in Figure~\ref{fig1:means_cartoons}(B), even
-within the class of thermodynamic models, we can make different commitments
-about the underlying microscopic states of the promoter. Model 1 considers only
-two states: a state in which a repressor (with copy number $R$) binds to an
-operator and a transcriptionally active state. The free energy difference
-between the repressor binding the operator, i.e. a specific binding site, and
-one of the $N_{NS}$ non-specific sites is given by $\Delta\varepsilon_R$ (given
-in $k_BT$ units as $\beta\equiv (k_BT)^{-1}$). Model 2 expands this model to
-include an empty promoter where no transcription happens, as well as a state on
-which one of the $P$ RNAPs binds to the promoter with binding energy
-$\Delta\varepsilon_P$. Indeed, the list of options considered here does not at
-all exhaust the suite of different microscopic states we can assign to the
-promoter.
+The first class of models that are the subject of the present section focus on
+predicting the mean level of gene expression. These models, sometimes known as
+thermodynamic models, invoke the tools of equilibrium statistical mechanics to
+compute the probabilities of the promoter microstates~\cite{Ackers1982,
+Shea1985, Buchler2003, Vilar2003a, Vilar2003b, Bintu2005a, Bintu2005c,
+Gertz2009, Sherman2012, Saiz2013}. As seen in
+Figure~\ref{fig1:means_cartoons}(B), even within the class of thermodynamic
+models, we can make different commitments about the underlying microscopic
+states of the promoter. Model 1 considers only two states: a state in which a
+repressor (with copy number $R$) binds to an operator and a transcriptionally
+active state. The free energy difference between the repressor binding the
+operator, i.e. a specific binding site, and one of the $N_{NS}$ non-specific
+sites is given by $\Delta\varepsilon_R$ (given in $k_BT$ units with $\beta\equiv
+(k_BT)^{-1}$). Model 2 expands this model to include an empty promoter where no
+transcription occurs, as well as a state in which one of the $P$ RNAPs binds to
+the promoter with binding energy $\Delta\varepsilon_P$. Indeed, the list of
+options considered here does not at all exhaust the suite of different
+microscopic states we can assign to the promoter. The essence of thermodynamic 
+models is to assign a discrete set of states and to use equilibrium statistical 
+mechanics to compute the probabilities of occupancy of those states.
 
@@
 and without a transcription factor present, any biological noise common to both
-conditions can be made to cancel away.
+conditions can be made to cancel out.
 
@@
 by that model, deferring most algebraic details to
-Appendix~\ref{sec:non_bursty}. What we will find is that
-for all cartoons the fold-change can be written as a Fermi function 
+Appendix~\ref{sec:non_bursty}. What we will find is that for all cartoons the
+fold-change can be written as a Fermi function of the form
 \begin{equation}
@@
 One of the main difference between models shown in
-Figure~\ref{fig1:means_cartoons}(C), casted in the language of chemical master
+Figure~\ref{fig1:means_cartoons}(C), cast in the language of chemical master
 equations, compared with the thermodynamic models discussed in the previous
@@
 fold-change as a product of the ratio of the binding and unbinding rates of the
-promoter, and the promoter details. This is \mrm{This is based in one of the
-comments from reviewer \# 2. It is empirically true for all models we explored,
-but I don't know how to prove that it is a general statement for any
-unimaginable complex model.}
+promoter, and the promoter details. This results in a fold-change of the form
 \begin{align}

### 2020-10-15 doc/resubmission/section_03_beyond_means.tex
implement most of RP comments from 20201015

--- 2020-09-24 doc/resubmission/section_03_beyond_means.tex
@@
 In this section, our objective is to explore the same models considered in the
-previous section, but now with reference to the the question of how well they
+previous section, but now with reference to the question of how well they
 describe the distribution of gene expression levels, with special reference to
 the variance in these distributions. To that end, we repeat the same pattern as
-in the previous section by examining the models one by one. In particular
-we will focus on the Fano factor, defined as the variance/mean. This metric
-serves as a powerful discriminatory tool to compare our different models to the
-null model that the steady-state mRNA distribution must be Poisson, resulting a
-Fano factor of one.
+in the previous section by examining the models one by one. In particular we
+will focus on the Fano factor, defined as the variance/mean. This metric serves
+as a powerful discriminatory tool to compare our different models to the null
+model that the steady-state mRNA distribution must be Poisson, resulting a Fano
+factor of one.
 
@@
 however, there are some lingering puzzles with interpreting this supercoiling
-hypothesis, so we leave it as a speculation and refrain from assigning definite
-physical meaning to the two states in this model.
+hypothesis, so we leave it as a speculative hypothesis and refrain from
+assigning definite physical meaning to the two states in this model.
 

### 2020-10-15 doc/resubmission/section_04_bayesian_inference.tex
implement most of RP comments from 20201015

@@
 
-\mrm{ I would vote to remove this paragraph completely.}
-We mention one further comparison, between our inferred parameters and the work
-of Chong et.\ al.~\cite{Chong2014}, which is interesting and puzzling.
-Experiments in~\cite{Chong2014} convincingly argue that supercoiling accumulated
-from the production of mRNA transcripts is key in setting the burstiness of mRNA
-production. In their model, this supercoiling occurs on the scale of
-$\sim100$~kb domains of DNA. This suggests that all genes on a given domain
-should burst in synchrony, and that the difference between highly and lowly
-expressed genes is the \textit{size} of transcriptional bursts, not the
-\textit{time between} bursts. But here, all burst sizes we infer in
-Figure~\ref{fig:constit_post_full}(C) are comparable and burst rates vary
-wildly. It is not immediately clear how to square this circle. Furthermore,
-Figure~7E in~\cite{Chong2014} reports values of the quantity they label
-$\beta/\alpha$ and we label $k^+/k^-$ in model 4 from
-Figure~\ref{fig2:constit_cartoons}. In contrast to the findings
-of~\cite{Razo-Mejia2020}, Chong et.\ al.\ do not find $k^+/k^-\ll1$ for most of
-the genes they consider. This begs the question: is the \textit{galK}
-chromosomal locus used for the reporter constructs in~\cite{Razo-Mejia2020}
-and~\cite{Jones2014} merely an outlier, or is there a deeper puzzle here waiting
-to be resolved? Without more apples-to-apples data we can only speculate, and we
-leave it as an intriguing open question for the field.
-
-Despite such puzzles, our goal here is not to unravel the mysterious origins of
-burstiness in transcription. Our remaining task in this work is a determination
-of the physical reality of thermodynamic binding energies in
-Figure~\ref{fig1:means_cartoons}, as codified by the thermodynamic-kinetic
-equivalence of Eq.~\ref{eq:deltaFR_eq_noneq_equiv}. For our phenomenological
-needs here model 5 in Figure~\ref{fig2:constit_cartoons}(A) is more than
-adequate: the posterior distributions in Figure~\ref{fig:constit_post_full}(C)
-are cleanly identifiable and the predictive checks in
-Figure~\ref{figS:ppc_unreg} indicate no discrepancies between the model and the
-mRNA single-molecule count data of~\cite{Jones2014}. Of the models we have
-considered it is unique in satisfying both these requirements. So we will
-happily use it as a foundation to build upon in the next section when we add
-regulation.
+Our remaining task in this work is a determination of the physical reality of
+thermodynamic binding energies in Figure~\ref{fig1:means_cartoons}, as codified
+by the thermodynamic-kinetic equivalence of Eq.~\ref{eq:deltaFR_eq_noneq_equiv}.
+For our phenomenological needs here model 5 in
+Figure~\ref{fig2:constit_cartoons}(A) is more than adequate: the posterior
+distributions in Figure~\ref{fig:constit_post_full}(C) are cleanly identifiable
+and the predictive checks in Figure~\ref{figS:ppc_unreg} indicate no
+discrepancies between the model and the mRNA single-molecule count data
+of~\cite{Jones2014}. Of the models we have considered it is unique in satisfying
+both these requirements. As a result we will use it as a foundation to build
+upon in the next section when we add regulation.
 
@@
 This likelihood is rather inscrutable. We did not find any of the known
-analytical approximations for ${_2F_1}$ terribly useful in gaining intuition, so
-we instead resorted to numerics. One insight we found
-was that for very strong or very weak repression, the distribution in
-Eq.~\ref{eq:p_m_bursty+rep} is well approximated by a negative binomial with
-burst size $b$ and burst rate $k_i$ equal to their constitutive \textit{lacUV5}
-values, except with $k_i$ multiplied by the fold-change
-$\left(1+k_R^+/k_R^-\right)^{-1}$. In other words, once again only the ratio
-$k_R^+/k_R^-$ was detectable. But for intermediate repression, the distribution
-was visibly broadened with Fano factor greater than $1+b$, the value for the
-corresponding constitutive case. This indicates that the repressor rates had
-left an imprint on the distribution, and perhaps intuitively, this intermediate
-regime occurs for values of $k_R^\pm$ comparable to the burst rate $k_i$. Put
-another way, if the repressor rates are much faster or much slower than $k_i$,
-then there is a timescale separation and effectively only one timescale remains,
-$k_i\left(1+k_R^+/k_R^-\right)^{-1}$. Only when all three rates in the problem
-are comparable does the mRNA distribution retain detectable information about
-them.
+analytical approximations for ${_2F_1}$ useful in gaining intuition, so we
+instead resorted to numerics. One insight we found was that for very strong or
+very weak repression, the distribution in Eq.~\ref{eq:p_m_bursty+rep} is well
+approximated by a negative binomial with burst size $b$ and burst rate $k_i$
+equal to their constitutive \textit{lacUV5} values, except with $k_i$ multiplied
+by the fold-change $\left(1+k_R^+/k_R^-\right)^{-1}$. In other words, once again
+only the ratio $k_R^+/k_R^-$ was detectable. But for intermediate repression,
+the distribution was visibly broadened with Fano factor greater than $1+b$, the
+value for the corresponding constitutive case. This indicates that the repressor
+rates had left an imprint on the distribution, and perhaps intuitively, this
+intermediate regime occurs for values of $k_R^\pm$ comparable to the burst rate
+$k_i$. Put another way, if the repressor rates are much faster or much slower
+than $k_i$, then there is a timescale separation and effectively only one
+timescale remains, $k_i\left(1+k_R^+/k_R^-\right)^{-1}$. Only when all three
+rates in the problem are comparable does the mRNA distribution retain detectable
+information about them.
 

### 2020-10-15 doc/resubmission/section_05_discussion.tex
implement most of RP comments from 20201015

@@
 assumed a thermodynamic rather than a kinetic
-framework~\cite{Garcia2011a,Razo-Mejia2018}. Interestingly, to constraint the
+framework~\cite{Garcia2011a,Razo-Mejia2018}. Interestingly, to constrain the
 binding and unbinding rates inferred for Figure~\ref{fig4:repressed_post_full},

### 2020-10-16 doc/resubmission/section_02_means.tex
minimal changes to fix some errors

@@
 situation. The fold-change takes the form
-\begin{align}
+\begin{eqnarray}
 \text{fold-change}
-&= \left(
+&=& \left(
 1 + \frac{\frac{R}{N_{NS}} e^{-\beta\Delta\varepsilon_R}}
@@
 \\
-&= (1 + \exp(-\Delta F_R + \log\rho))^{-1},
-\end{align}
+&=& (1 + \exp(-\Delta F_R + \log\rho))^{-1},
+\end{eqnarray}
 where $P$ is the number of RNAP per cell, and $\Delta\varepsilon_P$ is the
@@
 promoter, and the promoter details. This results in a fold-change of the form
-\begin{align}
-\text{fold-change} &= \left( 1 + \frac{k_R^+}{k_R^-} \rho \right)^{-1},\\
-&= (1 + \exp(-\Delta F_R + \log(\rho) ))^{-1},
-\end{align}
+\begin{eqnarray}
+ \text{fold-change} &=& \left( 1 + \frac{k_R^+}{k_R^-} \rho \right)^{-1},\\
+&=& (1 + \exp(-\Delta F_R + \log(\rho) ))^{-1},
+\end{eqnarray}
 where $\Delta F_R \equiv -\log(k_R^+/k_R^-)$, and the functional forms of $\rho$

### 2020-10-16 doc/resubmission/section_03_beyond_means.tex
minimal changes to fix some errors

@@
 
-One interpretation~\cite{Chong2014, Sevier2016} for the active and inactive
-states is that they represent the promoter's supercoiling state: transitions to
-the inactive state are caused by accumulation of positive supercoiling, which
-inhibits transcription, and transitions back to ``active'' are caused by gyrase
-or other topoisomerases relieving the supercoiling. This is an interesting
-possibility because it would mean the timescale for promoter state transitions
-is driven by topoisomerase kinetics, not by RNAP kinetics. From in vitro
-measurements, the former are suggested to be of order minutes~\cite{Chong2014}.
-Contrast this with model 2, where the state transitions are assumed to be
-governed by RNAP, which, assuming a copy number per cell of order $10^3$, has a
-diffusion-limited association rate $k_{on} \sim 10^2~\text{s}^{-1}$ to a target
-promoter. Combined with known $K_d$'s of order $\mu$M, this gives an RNAP
-dissociation rate $k_{off}$ of order $10^2$~s$^{-1}$. As we will show below,
-however, there are some lingering puzzles with interpreting this supercoiling
-hypothesis, so we leave it as a speculative hypothesis and refrain from
-assigning definite physical meaning to the two states in this model.
+One interpretation~\cite{Chong2014, Sevier2016, Sevier2018} for the active and
+inactive states is that they represent the promoter's supercoiling state:
+transitions to the inactive state are caused by accumulation of positive
+supercoiling, which inhibits transcription, and transitions back to ``active''
+are caused by gyrase or other topoisomerases relieving the supercoiling. This is
+an interesting possibility because it would mean the timescale for promoter
+state transitions is driven by topoisomerase kinetics, not by RNAP kinetics.
+From in vitro measurements, the former are suggested to be of order
+minutes~\cite{Chong2014}. Contrast this with model 2, where the state
+transitions are assumed to be governed by RNAP, which, assuming a copy number
+per cell of order $10^3$, has a diffusion-limited association rate $k_{on} \sim
+10^2~\text{s}^{-1}$ to a target promoter. Combined with known $K_d$'s of order
+$\mu$M, this gives an RNAP dissociation rate $k_{off}$ of order $10^2$~s$^{-1}$.
+As we will show below, however, there are some lingering puzzles with
+interpreting this supercoiling hypothesis, so we leave it as a speculative
+hypothesis and refrain from assigning definite physical meaning to the two
+states in this model.
 
@@
 Appendix~\ref{sec:non_bursty} takes the form
-\begin{align}
+\begin{equation}
 \begin{split}
@@
 \end{split}
-\end{align}
+\end{equation}
 where we use $k_i$ to denote the burst initiation rate, $\text{Geom}(n;b)$ is

### 2020-10-16 doc/resubmission/section_03_beyond_means.tex
remove challenging so et al phrase

@@
 machinery, e.g., RNAP itself. In this picture, all the corrections to Poisson
-behavior are derived as additive corrections to the Fano factor. For challenging
-the ``universal noise curve'' from So et.\ al.~\cite{So2011}, this picture was
-excellent. It is appealing in its simplicity, with only two parameters, the
-initiation rate $r$ and degradation rate $\gamma$. In other words, the model is
-not excessively complex for the data at hand. But for many interesting
-questions, for instance in the recent work~\cite{Razo-Mejia2020}, attributing
-all deviations from the model to extrinsic noise sources, limits the kinds of
-predictions that can be done. To make progress then we need a (slightly) more
-complex model than model 1 that would allow us to incorporate the non-Poissonian
-features of constitutive promoters directly into a master equation formulation.
+behavior are derived as additive corrections to the Fano factor. This picture is
+appealing in its simplicity, with only two parameters, the initiation rate $r$
+and degradation rate $\gamma$. In other words, the model is not excessively
+complex for the data at hand. But for many interesting questions, for instance
+in the recent work~\cite{Razo-Mejia2020}, attributing all deviations from the
+model to extrinsic noise sources, limits the kinds of predictions that can be
+done. To make progress then we need a (slightly) more complex model than model 1
+that would allow us to incorporate the non-Poissonian features of constitutive
+promoters directly into a master equation formulation.
 

### 2020-10-16 doc/resubmission/appendix_01_nonbursty.tex
fix ?? issues

@@
 models, the natural mathematicization of their cartoons is as a chemical master
-equation such as Eq.~\ref{eq:2state_rep_cme} for model 1, which for
-completeness we reproduce here as
+equation. For model 1 we have the master equation
 \begin{align}
@@
 Although we derived Eq.~\ref{eq:generic_cme_appdx} for the particular case of
-kinetic model 2 in Figure~\ref{fig1:means_cartoons}, in fact the
-chemical master equations for all of the kinetic models in
-Figure~\ref{fig1:means_cartoons} except for model 5 can be cast in this form.
-(We treat model 5 separately in Appendix~\ref{sec:gen_fcn_appdx}.) Model
-3 introduces no new subtleties beyond model 2 and
-Eq.~\ref{eq:generic_cme_appdx} applies equally well to it, simply with different
-matrices of dimension four instead of three. Models 1 and 4 are both
-handled by Eq.~\ref{eq:2state_rep_cme} in the main text, which is just
-Eq.~\ref{eq:generic_cme_appdx} except in the special case of $\mathbf{R_D} =
-\mathbf{R_A} \equiv \mathbf{R}$, since in these two models transcription
-initiation events do not change promoter state.
+kinetic model 2 in Figure~\ref{fig1:means_cartoons}, in fact the chemical master
+equations for all of the kinetic models in Figure~\ref{fig1:means_cartoons}
+except for model 5 can be cast in this form. (We treat model 5 separately in
+Appendix~\ref{sec:gen_fcn_appdx}.) Model 3 introduces no new subtleties beyond
+model 2 and Eq.~\ref{eq:generic_cme_appdx} applies equally well to it, simply
+with different matrices of dimension four instead of three. Models 1 and 4 are
+both Eq.~\ref{eq:generic_cme_appdx} except that $\mathbf{R_D} = \mathbf{R_A}
+\equiv \mathbf{R}$, since in these two models transcription initiation events do
+not change promoter state.
 
@@
 \begin{equation}
-\nu = 1 - \frac{r k_P^+ } {(k_P^+ + k_P^- + r)(k_P^+ + k_P^- + r + \gamma)},
-\end{equation}
-exactly Eq.~\ref{eq:model2_fano} in the main text.
+\nu = 1 - \frac{r k_P^+ } {(k_P^+ + k_P^- + r)(k_P^+ + k_P^- + r + \gamma)}.
+\end{equation}
 
@@
 \end{equation}
-The master equation takes verbatim the same form as it did for model 1,
-Eq.~\ref{eq:2state_rep_cme}. Meanwhile the promoter transition matrix
-$\mathbf{K}$ is the same as Eq.~\ref{eq:3state_cme_matrices_pt1} from model 2,
-although we relabel the rate constants from $k_P^\pm$ to $k^\pm$ to reiterate
-that these are not simply RNAP binding and unbinding rates.
+The master equation takes verbatim the same form as it did for model 1.
+Meanwhile the promoter transition matrix $\mathbf{K}$ is the same as
+Eq.~\ref{eq:3state_cme_matrices_pt1} from model 2, although we relabel the rate
+constants from $k_P^\pm$ to $k^\pm$ to reiterate that these are not simply RNAP
+binding and unbinding rates.
 

### 2020-10-16 doc/resubmission/appendix_03_bayesian_inference.tex
fix ?? issues

@@
 agreement between our predictions and the experimental data for the regulated
-case, we include Figure~\ref{figS:eg_histograms} that plot the same data as
+case, we include Figure~\ref{figS:reg_histograms} that plot the same data as
 Figure~\ref{fig4:repressed_post_full}(C), but rather than showing ECDF, we show

### 2020-10-17 doc/resubmission/appendix_03_bayesian_inference.tex
send Poisson promoter inference to appendix

@@
 us build practical intuition into the implementation of the Bayesian inference
-pipeline as we noted in Section~\ref{sec:beyond_means} of the main text that
-this model cannot be reconciled with experimental data from observables such as
-the Fano factor. In other words, we acknowledge that this model is ``wrong,''
-but we still see value in going through the analysis since the simple nature of
-the model translates into a neat statistical analysis.
+pipeline. As we noted in Section~\ref{sec:beyond_means} of the main text, this
+model cannot be reconciled with experimental data from observables such as the
+Fano factor. In other words, we acknowledge that this model is ``wrong,'' but we
+still see value in going through the analysis since the simple nature of the
+model translates into a neat statistical analysis.
 
@@
 
-As specified in the main test, the mRNA steady-state distribution for model 1 in
-Figure~\ref{fig2:constit_cartoons}(A) is Poisson with parameter $\lambda$.
+Model 1 in Figure~\ref{fig2:constit_cartoons}(A) predicts a  mRNA distribution
+whose steady-state solution is given by a Poisson distribution with parameter
+$\lambda \equiv r / \gamma$, where $r$ is the mRNA production rate, and $\gamma$
+is the mRNA degradation rate~\cite{Sanchez2013}. The goal of our inference
+problem is then to find the probability distribution for the parameter value
+$\lambda$ given the experimental data. By Bayes' theorem this can be written as
+\begin{equation}
+p(\lambda \mid D) = {p(D \mid \lambda) p(\lambda) \over p(D)},
+\end{equation}
+where $D = \{m_1, m_2, \ldots, m_N \}$ are the single-cell mRNA experimental
+counts. As is standard we will neglect the denominator $p(D)$ on the right
+hand side since it is independent of $\lambda$ and serves only as a
+normalization factor.
+
+The steady-state solution for the master equation defines the likelihood term
+for a single cell $p(m \mid \lambda)$. What this means is that for a given
+choice of parameter $\lambda$, under model 1 of
+Figure~\ref{fig2:constit_cartoons}(A), we expect to observe $m$ mRNAs in a
+single cell with probability
+\begin{equation}
+p(m\mid\lambda) = \frac{\lambda^m e^{-\lambda}}{m!}.
+\label{eq:poisson_inference010}
+\end{equation}
+Assuming each cell's mRNA count in our dataset is independent of others, the
+likelihood of the full inference problem $p(D\mid\lambda)$ is simply a product
+of the single cell likelihoods given by Eq.~\ref{eq:poisson_inference010} above, so
+\begin{equation}
+p(D\mid\lambda) = \prod_{k=1}^N \frac{\lambda^{m_k}e^{-\lambda}}{m_k!}.
+\end{equation}
+
 Throughout this Appendix we will appeal to the convenient notation for

### 2020-10-17 doc/resubmission/section_04_bayesian_inference.tex
minimal changes to fix some errors
send Poisson promoter inference to appendix

@@
 problem is that of Bayesian probability. We will then build a Bayesian inference
-pipeline to fit the model parameters to data. To gain intuition on how this
-analysis is done we will begin with the ``wrong'' model 1 in
-Figure~\ref{fig2:constit_cartoons}(A). We will use the full dataset of
+pipeline to fit the model parameters to data. We will use the full dataset of
 single-cell mRNA counts from~\cite{Jones2014} used in
-Figure~\ref{fig2:constit_cartoons}(B).
-
-\subsubsection{Model 1: Poisson promoter}
-
-Model 1 in Figure~\ref{fig2:constit_cartoons}(A) predicts a  mRNA distribution
-whose steady-state solution is given by a Poisson distribution with parameter
-$\lambda \equiv r / \gamma$, where $r$ is the mRNA production rate, and $\gamma$
-is the mRNA degradation rate~\cite{Sanchez2013}. The goal of our inference
-problem is then to find the probability distribution for the parameter value
-$\lambda$ given the experimental data. By Bayes' theorem this can be written as
-\begin{equation}
-p(\lambda \mid D) = {p(D \mid \lambda) p(\lambda) \over p(D)},
-\end{equation}
-where $D = \{m_1, m_2, \ldots, m_N \}$ are the single-cell mRNA experimental
-counts. As is standard we will neglect the denominator $p(D)$ on the right
-hand side since it is independent of $\lambda$ and serves only as a
-normalization factor.
-
-The steady-state solution for the master equation defines the likelihood term
-for a single cell $p(m \mid \lambda)$. What this means is that for a given
-choice of parameter $\lambda$, under model 1 of
-Figure~\ref{fig2:constit_cartoons}(A), we expect to observe $m$ mRNAs in a
-single cell with probability
-\begin{equation}
-p(m\mid\lambda) = \frac{\lambda^m e^{-\lambda}}{m!}.
-\label{eq:poisson_inference010}
-\end{equation}
-Assuming each cell's mRNA count in our dataset is independent of others, the
-likelihood of the full inference problem $p(D\mid\lambda)$ is simply a product
-of the single cell likelihoods given by Eq.~\ref{eq:poisson_inference010} above, so
-\begin{equation}
-p(D\mid\lambda) = \prod_{k=1}^N \frac{\lambda^{m_k}e^{-\lambda}}{m_k!}.
-\end{equation}
-
-To proceed we need to specify a prior distribution $p(\lambda)$. In this case we
-are extremely data-rich, as the dataset from Jones et.\ al~\cite{Jones2014} has
-of order 1000-3000 single-cell measurements for each promoter, so our choice of
-prior matters little here, as long as it is sufficiently broad. For details on
-the prior selection we refer the reader to Appendix~\ref{sec:bayesian}. For our
-purpose here it suffices to specify that we use as prior a Gamma distribution.
-This particular choice of prior introduces two new parameters, $\alpha$ and
-$\beta$, which parametrize the gamma distribution itself, which we use to encode
-the range of $\lambda$ values we view as reasonable. Recall $\lambda$ is the
-mean steady-state mRNA count per cell, which \textit{a priori} could plausibly
-be anywhere from 0 to a few hundred. $\alpha=1$ and $\beta=1/50$ achieve this,
-since the gamma distribution is strictly positive with mean $\alpha/\beta$ and
-standard deviation $\sqrt{\alpha}/\beta$.
-
-As detailed in Appendix~\ref{sec:bayesian} this particular choice of prior is
-known as the \textit{conjugate} prior for a Poisson likelihood.
-Conjugate priors have the convenient properties that a closed form exists for the posterior distribution $p(\lambda \mid D)$ - unusual in Bayesian inference problems - and the closed form posterior takes the same form as the prior.
-For
-our case of a Poisson distribution likelihood with its
-Gamma distribution conjugate prior, the posterior distribution is also a Gamma
-distribution~\cite{Gelman2013}. Specifically the two parameters $\alpha'$ and
-$\beta'$ for this posterior distribution take the form $\alpha' = \alpha +
-\bar{m} N$ and $\beta' = \beta + N$, where we defined the sample mean $\bar{m} =
-\frac{1}{N}\sum_{k=1}^N m_k$ for notational convenience, and $N$ is the number
-of cells in our dataset. Furthermore, given that $N$ is $\mathcal{O}(10^3)$ and
-$\langle m\rangle \gtrsim 0.1$ for all promoters measured in~\cite{Jones2014}
-our data easily overwhelms the choice of prior, and allows us to approximate the
-Gamma distribution with a Gaussian distribution with mean $\bar{m}$ and variance
-$\bar{m} / N$ with marginal errors. As an example with real numbers, for the
-\textit{lacUV5} promoter, Jones et.\ al~\cite{Jones2014} measured 2648 cells
-with an average mRNA count per cell of $\bar{m} \approx 18.7$. For this case our
-posterior distribution $P(\lambda \mid D)$ would be a Gaussian distribution with
-mean $\mu = 18.7$, and a standard deviation $\sigma \approx 0.08$. This suggests
-we have inferred our model's one parameter to a precision of order 1\%.
-
-We remind the reader that we began this section claiming that the Poisson model
-was ``wrong'' since it could not reproduce features of the data such as a Fano
-factor $> 1$. The fact that we obtain such a narrow posterior distribution for our
-parameter $P(\lambda \mid D)$ does not equate to the model being adequate to
-describe the data. What this means is that given the data $D$, only values
-in a narrow range are remotely plausible for the parameter $\lambda$,
-but a narrow posterior distribution does not necessarily
-mean the model accurately depicts reality.
-As we will see later in Figure~\ref{fig:constit_post_full} after
-exploring the bursty promoter model, indeed the correspondence
-when contrasting the Poisson model with the experimental data is quite poor.
+Figure~\ref{fig2:constit_cartoons}(B). To gain intuition on how this analysis is
+done we analyzed the ``wrong'' model 1 in Figure~\ref{fig2:constit_cartoons}(A).
+For this analysis, we direct the reader to Appendix~\ref{sec:bayesian} as an
+intuitive introduction to the Bayesian framework with a simpler model.
 
@@
 Figure~\ref{fig:constit_post_full}(A). In contrast to the active/inactive
-constitutive model considered in~\cite{Razo-Mejia2020} (nonequilibrium model 4
-in Figure~\ref{fig2:constit_cartoons}(A)), this model is well-identified with both
+constitutive model considered in~\cite{Razo-Mejia2020} (kinetic model 4 in
+Figure~\ref{fig2:constit_cartoons}(A)), this model is well-identified with both
 parameters determined to a fractional uncertainty of 5-10\%. The strong
@@
 Having found the model's posterior to be well-identified as with the Poisson
-promoter, the next step is to compare both models with experimental data. To do
-this for the case of the bursty promoter, for each of the parameter samples
-shown in Figure~\ref{fig:constit_post_full}(A) we generated negative
-bionomial-distributed mRNA counts. As MCMC samples parameter space
-proportionally to the posterior distribution, this set of random samples span
-the range of possible values that we would expect given the correspondence
+promoter (Appendix~\ref{sec:bayesian}), the next step is to compare both models
+with experimental data. To do this for the case of the bursty promoter, for each
+of the parameter samples shown in Figure~\ref{fig:constit_post_full}(A) we
+generated negative bionomial-distributed mRNA counts. As MCMC samples parameter
+space proportionally to the posterior distribution, this set of random samples
+span the range of possible values that we would expect given the correspondence
 between our theoretical model and the experimental data. A similar procedure can
@@
 predictive ECDF, indicating this model is actually able to generate the observed
-data and increasing our confidence that this model
-is sufficient to parametrize the physical reality of the system.
+data and increasing our confidence that this model is sufficient to parametrize
+the physical reality of the system.
 
@@
 $\alpha$ and $\beta$, defined for notational convenience, are
-\begin{align}
+\begin{equation}
 \begin{split}
@@
 \end{split}
-\end{align}
+\end{equation}
 

### 2020-10-17 code/figures/main/fig02B.py
fix typo in colorbar

@@ -103,7 +103,7 @@ im = ax.scatter(
 # Add colorbar
 cbar = fig.colorbar(im, ax=ax, pad=0.01)
 # Set colorbar legend
-cbar.set_label(r"$\Delta\epsilon_r \; (k_BT)$")
+cbar.set_label(r"$\Delta\epsilon_P \; (k_BT)$")
 
 # Annotate points
 for i, (row, data) in enumerate(df_summary.iterrows()):

### 2020-10-17 code/figures/main/fig03.py
fix typo in colorbar

@@ -312,7 +312,7 @@ im = ax_d.scatter(
 
 im.set_visible(False)
 cbar = fig.colorbar(im, ax=ax_d, pad=0.01)
-cbar.set_label(r"$\Delta\epsilon_r \; (k_BT)$")
+cbar.set_label(r"$\Delta\epsilon_P \; (k_BT)$")
 
 # add a guideline for the eye for the predicted log(k_i) ~ - binding E
 guide_x = np.linspace(-7,-3.5)

### 2020-10-17 doc/resubmission/appendix_01_nonbursty.tex
minor fixes to typos

@@
 transitions between different mRNA counts and promoter states is more explicitly
-depicted. Given that the DNA promoter can exist in one of two states --
-transcriptionally active state, and with repressor bound -- we will keep track
+depicted. Given that the DNA promoter can exist in one of two states--
+transcriptionally active state, and with repressor bound-- we will keep track
 not only of the mRNA count, but on which state the promoter is. For this we will

### 2020-10-17 doc/resubmission/appendix_02_bursty.tex
minor fixes to typos

@@
 Given the normalization constraint of the geometric distribution, adding the
-probability of all possible burst sizes -- including size zero since we
-re-indexed the sum -- allows us to write
+probability of all possible burst sizes --including size zero since we
+re-indexed the sum-- allows us to write
 \begin{equation}

### 2020-10-17 doc/resubmission/appendix_03_bayesian_inference.tex
minor fixes to typos

@@
 experimental observations? The language of probability, and more specifically of
-Bayesian statistics is -- we think -- the natural language to tackle this
+Bayesian statistics is --we think-- the natural language to tackle this
 question.
@@
 Notice than rather than writing the full form of Bayes' theorem, we limit 
-ourselves to the terms that depend on our quantity of interest -- that is the 
-parameter values themselves $\vec{\theta}$ -- as the denominator $P(D)$ only
+ourselves to the terms that depend on our quantity of interest --that is the 
+parameter values themselves $\vec{\theta}$-- as the denominator $P(D)$ only
 serves as a normalization constant.
@@
 contradictory to claim that can we represent our thinking about a natural
-phenomenon in the form of a mathematical model -- in the context of Bayesian
+phenomenon in the form of a mathematical model --in the context of Bayesian
 inference this means choosing a form for the likehihoods, and even making
 this choice presupposes prior understanding or assumptions as to the
-relevant features in the system under study -- but that we have absolutely
+relevant features in the system under study-- but that we have absolutely
 no idea what the parameter values could or could not be. We therefore make

### 2020-10-17 doc/resubmission/section_02_means.tex
minor fixes to typos

@@
 imagined to exist in a discrete set of states of occupancy, with each such state
-of occupancy accorded its own rate of transcription -- including no
+of occupancy accorded its own rate of transcription --including no
 transcription for many of these states. This discretization of a potentially
@@
 and without a transcription factor present, any biological noise common to both
-conditions can be made to cancel out.
-
-Figure~\ref{fig1:means_cartoons}(B) and (C) depicts a smorgasbord of
-mathematicized cartoons for simple repression using both thermodynamic and
-kinetic models, respectively, that have appeared in previous literature. For
-each cartoon, we calculate the fold-change in mean gene expression as predicted
-by that model, deferring most algebraic details to
+conditions can be made to cancel out. Figure~\ref{fig1:means_cartoons}(B) and
+(C) depicts a smorgasbord of mathematicized cartoons for simple repression using
+both thermodynamic and kinetic models, respectively, that have appeared in
+previous literature. For each cartoon, we calculate the fold-change in mean gene
+expression as predicted by that model, deferring most algebraic details to
 Appendix~\ref{sec:non_bursty}. What we will find is that for all cartoons the

### 2020-10-17 doc/resubmission/section_04_bayesian_inference.tex
minor fixes to typos

@@
 distribution is exactly the likelihood we want to use when stating Bayes
-theorem. Again denoting the single-cell mRNA count data as $D=\{m_1, m_2,\dots,
-m_N\}$, here Bayes' theorem takes the form
+theorem. Denoting the single-cell mRNA count data as $D=\{m_1, m_2,\dots,
+m_N\}$, Bayes' theorem takes the form
 \begin{equation}
@@
 \end{equation}
-We already have our likelihood -- the product of $N$ negative binomials as
-Eq.~\ref{eq:neg_bionom} -- so we only need to choose priors on $k_i$ and $b$.
-For the datasets from~\cite{Jones2014} that we are analyzing, as for the Poisson
+We already have our likelihood --the product of $N$ negative binomials as
+Eq.~\ref{eq:neg_bionom}-- so we only need to choose priors on $k_i$ and $b$. For
+the datasets from~\cite{Jones2014} that we are analyzing, as for the Poisson
 promoter model above we are still data-rich so the prior's influence remains
@@
 was fit on \textit{lacUV5} data from~\cite{Jones2014}. (B) The empirical
-distribution function (ECDF) of the observed population distribution of mRNA
-transcripts under the control of a constitutive \textit{lacUV5} promoter is
-shown in black. The median posterior predictive ECDFs for models (1), Poisson,
-and (5), negative binomial, are plotted in dark green and dark blue,
+cumulative distribution function (ECDF) of the observed population distribution
+of mRNA transcripts under the control of a constitutive \textit{lacUV5} promoter
+is shown in black. The median posterior predictive ECDFs for models (1),
+Poisson, and (5), negative binomial, are plotted in dark green and dark blue,
 respectively. Lighter green and blue regions enclose 95\% of all posterior
@@
 \end{equation}
-By taking the weak promoter approximation for the equilibrium model ($P/N_{NS} 
+By taking the weak promoter approximation for the thermodynamic model ($P/N_{NS}
 \exp(-\beta\Delta\varepsilon_r) \ll 1$) results in~\cite{Brewster2012}
@@
 Figure~\ref{fig1:means_cartoons}(B)? And can this comparison shed light on the
-unreasonable effectiveness of the equilibrium models, for instance, in their
+unreasonable effectiveness of the thermodynamic models, for instance, in their
 application in~\cite{Chure2019, Chure2019a}?
@@
 for each aTc induction concentration (corresponding to an unknown mean repressor
-copy number).
-(B, upper) Ratios of our inferred unbinding rates are
-compared with operator binding energy differences measured by Garcia and
+copy number). (B, upper) Ratios of our inferred unbinding rates are compared
+with operator binding energy differences measured by Garcia and
 Phillips~\cite{Garcia2011a} (triangles) and Razo-Mejia et.\
 al.~\cite{Razo-Mejia2018} (circles). Blue glyphs compare O2-O1, while orange
-compare O1-Oid. Points with perfect agreement would lie on the dotted line.
-(B, lower) Unbinding rates for O1 (cyan) and Oid (red)
-inferred in this work are compared with single-molecule measurements
-from Hammar et.\ al.~\cite{Hammar2014}. We plot the comparison assuming illustrative mRNA
-lifetimes of $\gamma^{-1}=3$~min (triangles) or $\gamma^{-1}=5$~min
-(squares). Dotted line is as in upper panel.
-(C) Theory-experiment comparison are shown for each of the datasets
-used in the inference of the model in (A). Observed single-molecule mRNA counts
-data from~\cite{Jones2014} are plotted as black lines. The median of the randomly
-generated samples for each condition is plotted as a dark colored line. Lighter
-colored bands enclose 95\% of all samples for a given operator/repressor copy
-number pair.
-%Samples are sorted by operator and plotted separately for visual clarity.
-The unregulated promoter, \textit{lacUV5}, is shown with each as a reference.}
+compare O1-Oid. Points with perfect agreement would lie on the dotted line. (B,
+lower) Unbinding rates for O1 (green) and Oid (red) inferred in this work are
+compared with single-molecule measurements from Hammar et.\
+al.~\cite{Hammar2014}. We plot the comparison assuming illustrative mRNA
+lifetimes of $\gamma^{-1}=3$~min (triangles) or $\gamma^{-1}=5$~min (squares).
+Dotted line is as in upper panel. (C) Theory-experiment comparison are shown for
+each of the datasets used in the inference of the model in (A). Observed
+single-molecule mRNA counts data from~\cite{Jones2014} are plotted as black
+lines. The median of the randomly generated samples for each condition is
+plotted as a dark colored line. Lighter colored bands enclose 95\% of all
+samples for a given operator/repressor copy number pair. The unregulated
+promoter, \textit{lacUV5}, is shown with each as a reference.}
 \label{fig4:repressed_post_full}
@@
 nine different conditions, spanning several combinations of three unique
-repressor binding sites (the so-called $Oid$, $O1$, and $O2$ operators) and four
+repressor binding sites (the so-called Oid, O1, and O2 operators) and four
 unique repressor copy numbers. We assume that the values of $k_i$ and $b$ are
@@
 On the whole then, we regard this as striking confirmation of the validity of
-the equilibrium models. Their lynchpin parameter is a phenomenological free
+the thermodynamic models. Their lynchpin parameter is a phenomenological free
 energy of repressor binding that has previously only been inferred indirectly.
@@
 uncertainties that remain in the entire theory-experiment dialogue.
-%\mmnote{Does this deserve qualifiers, like ``to a good approximation,'' or ``as a first order model'' or something?
-I like the simple finality without, but I worry it's too brazen or subtlely misleading.}
 
@@
 seriously doubt the overall correctness of our model, which further validates
-its use to compare the equilibrium models' binding energies to the
-nonequilibrium models' repressor kinetics, as we originally set out to do.
+its use to compare the thermodynamic models' binding energies to the kinetic
+models' repressor binding and unbinding rates, as we originally set out to do.

### 2020-10-22 doc/resubmission/appendix_01_nonbursty.tex
final pass before resubmission

@@
 main text. In each of these models, the natural mathematicization of their
-cartoons is either an equilibrium model based on statisticla mechanics, or a
+cartoons is either an equilibrium model based on statistical mechanics, or a
 chemical master equation. These derivations will go through the specifics of the
@@
 
-\subsection{Thremodynamic models of gene regulation}
+\subsection{Thermodynamic models of gene regulation}
 The first class of models we will explore are the so-called thermodynamic models
@@
 other chemical master equations in this work.
-
-% \subsubsection{Nonequilibrium Model Five - Bursty Promoter}
-% The final model we consider shown in Figure~\ref{fig1:means_cartoons}(C) is an
-% intuitive analog to model 1, with just two states, repressor bound or unbound,
-% and transition rates between them of $k_R^+$ and $k_R^-$. In model 1, when in
-% the unbound state, single mRNA transcripts are produced as a Poisson process
-% with some characteristic rate $r$. The current model by contrast produces, at
-% some Poisson rate $k_i$, \textit{bursts} of mRNA transcripts.
-% The burst sizes are assumed to be geometrically distributed with
-% a mean burst size $b$, which we will motivate in Section~\ref{sec:beyond_means}
-% when we derive this model as a certain limiting case of model 4.
-
-% From this intuitive picture and by analogy to model 1, then, it should be
-% plausible that the mean mRNA level is
-% \begin{equation}
-% \langle m\rangle = \frac{k_i b}{\gamma}
-%         \left(1 + \frac{k_R^+}{k_R^-}\right)^{-1},
-% \end{equation}
-% which will turn out to be correct from a careful calculation. For now, we simply
-% note that just like model 1, the fold-change becomes
-% \begin{equation}
-% \text{fold-change} = \left(1 + \frac{k_R^+}{k_R^-}\right)^{-1}
-% \end{equation}
-% with $\rho=1$ also like model 1.
-% We will also see later how this model emerges as a natural limit of model 4.
-% \mrm{ENDS HERE}
-
-% \mrm{THIS NEEDS TO BE FIXED}
-% \subsubsection{Noise in the Two-State Promoter, RNAP Bound or Unbound.}
-
-% Our second model of constitutive transcription posits an architecture in which
-% the promoter is either empty or bound by RNAP ~\cite{Phillips2015a,
-% Phillips2019}. Here,  as shown in model 2 of Figure~\ref{fig2:constit_cartoons}(A),
-% transcription initiation results in a state transition from the bound to the
-% unbound state, reflecting the microscopic reality that an RNAP that has begun to
-% elongate a transcript is no longer available at the start site to begin another.
-% As shown in Appendix~\ref{sec:non_bursty}, the Fano factor in this model is
-% given by
-% \begin{align}
-%     \nu = 1 -
-%         \frac{r\rate{k}{P}{+}}
-%             {\left(\rate{k}{P}{+} + \rate{k}{P}{-} + r\right)
-%              \left(\gamma + \rate{k}{P}{+} + \rate{k}{P}{-} + r\right)}.
-% \label{eq:model2_fano}
-% \end{align}
-% The problem with this picture is that the Fano factor is always $<1$. To make
-%  contact with the experimental reality of $\nu>1$ as shown in
-%  Figure~\ref{fig2:constit_cartoons}(B), clearly some corrections will be needed.
-%  While this model adds an appealing element of microscopic reality, we are
-%  forced to reject it as the additional complexity is unable to capture the
-%  phenomenology of interest. Obviously the promoter state does in fact proceed
-%  through cycles of RNAP binding, initiating, and elongating, but it seems that
-%  the super-Poissonian noise in mRNA copy number we want to model must be
-%  governed by other features of the system.
-
-% \subsubsection{Noise in the Three-State Promoter, Multistep Transcription
-% Initiation and Escape.}
-
-% How might we remedy the deficits of model 2? It is known~\cite{DeHaseth1998}
-% that once RNAP initially binds the promoter region, a multitude of distinct
-% steps occur sequentially before RNAP finally escapes into the elongation phase.
-% Perhaps adding some of this mechanistic detail as shown in model 3 of
-% Figure~\ref{fig2:constit_cartoons}(A) might rescue the previous model. The next
-% simplest refinement of that model could consider open complex formation and
-% promoter escape as separate steps rather than as a single effective step. In
-% other words, we construct model 3 by adding a single extra state to model 2, and
-% we will label the two RNAP-bound states as the closed and open complexes,
-% despite the true biochemical details certainly being more complex. For example,
-% earlier work extended this model by adding an additional repressor bound state
-% and did not explicitly consider the limit with no repressor that we analyze
-% here~\cite{Mitarai2015}. Again, our goal here is not a complete accounting of
-% all the relevant biochemical detail; this is an exploratory search for the
-% important features that a model needs to include to square with the known
-% experimental reality of constitutive expression.
-
-% Unfortunately, as hinted at in earlier work~\cite{Mitarai2015}, this model too
-% has Fano factor $\nu<1$. We again leave the algebraic details for
-% Appendix~\ref{sec:non_bursty} and merely state the result that
-% \begin{align}
-% \nu = 1 - \frac{r k_P^+ k_O}{\mathcal{Z}}
-% \frac{k_P^+ + k_P^- + k_O + r + \gamma}
-%     {\mathcal{Z} + \gamma(k_P^+ + k_P^- + k_O + r) + \gamma^2},
-% \label{eq:model3_fano}
-% \end{align}
-% where we defined $\mathcal{Z} = r(k_O + k_P^-) + k_P^+(k_O + r)$ for notational
-% tidiness. This is necessarily less than 1 for arbitrary rate constants.
-
-% In fact, we suspect \textit{any} model in which transcription proceeds through a
-% multistep cycle must necessarily have $\nu<1$. The intuitive argument compares
-% the waiting time distribution to traverse the cycle with the waiting time for a
-% Poisson promoter (model 1) with the same mean time. The latter is simply an
-% exponential distribution. The former is a convolution of multiple exponentials,
-% and intuitively the waiting time distribution for a multistep process
-% should be more peaked with a smaller fractional width than a single
-% exponential with the same mean.
-% A less disperse waiting time distribution means transcription initations are
-% more uniformly distributed in time relative to a Poisson process. Hence the
-% distribution of mRNA over a population of cells should be less variable compared
-% to Poisson, giving $\nu<1$.
-% (In Appendix~\ref{sec:non_bursty} we present a more precise version
-% of the intuitive arguments in this paragraph.)
-% Regardless of the merits of this model in describing the noise properties of
-% constitutive transcription initiation, it ultimately fails the simplest
-% quantitative feature of the data, namely that the Fano factor $> 1$ and hence
-% we must discard this mechanistic picture and search elsewhere.
-% \mrm{ENDS HERE}
 
@@
 equation. For model 1 we have the master equation
-\begin{align}
+\begin{equation}
 \begin{split}
@@
 \end{split}
-\end{align}
+\end{equation}
 Here $p_R(m,t)$ and $p_U(m,t)$ are the probabilities of finding the system with
@@
 equation for model 2 can be written
-\begin{align}
+\begin{equation}
 \begin{split}
@@
 \end{split}
-\end{align}
+\end{equation}
 $k_P^+$ and $k_P^-$ are defined in close analogy to $k_R^+$ and $k_R^-$, except
@@
 elements of $\mathbf{R_A}$. The only component of $\langle\vec{m}^0\rangle$ that
-ever survives is the transciptionally active state, and for the models we
+ever survives is the transcriptionally active state, and for the models we
 consider here, there is only ever one such state. This will save us some amount
@@
 master equations, but for completeness we reprint it again as
-\begin{align}
+\begin{equation}
 \begin{split}
@@
 \label{eq:model1_kinetic_cme}
-\end{align}
+\end{equation}
 This is a direct transcription of the states and rates in
@@
 form
-\begin{align}
+\begin{equation}
 \begin{split}
@@
 condition leads to a set of three linearly independent equations, namely
-\begin{align}
+\begin{eqnarray}
 -k_R^- p_R + k_R^+ p_E &= 0 \\
@@
 p_R + p_E + p_P &= 1.
-\end{align}
+\end{eqnarray}
 Using $p_R = 1 - p_E - p_P$ to eliminate $p_R$ in the first and solving the
@@
 the main text, from which we have
-\begin{align}
+\begin{eqnarray}
 \text{fold-change}
-&=      \frac{k_R^- k_P^+}
+&=&      \frac{k_R^- k_P^+}
         {k_R^- k_P^+ + k_R^- (k_P^- + r) + k_R^+ (k_P^- + r)}
@@
 \\
-&=      \left(1 + \frac{k_R^+}{k_R^-}
+&=&      \left(1 + \frac{k_R^+}{k_R^-}
                 \frac{k_P^- + r}{k_P^+ + k_P^- + r}
@@
 \\
-&=      \left(1 + \frac{k_R^+}{k_R^-}
+&=&      \left(1 + \frac{k_R^+}{k_R^-}
         \left(1 + \frac{k_P^+}{k_P^- + r}\right)^{-1}
         \right)^{-1},
-\end{align}
+\end{eqnarray}
 which follows the master curve of~\fig{fig1:means_cartoons}(D) with $\rho = 1 +
@@
 equations in $p_C$ and $p_O$ given by
-\begin{align}
-k_P^+k_R^-(1-p_C-p_O) - (k_P^- + k_O)(k_R^+ + k_R^-) p_C &= 0
-\\
-k_O p_C - r p_O &= 0.
-\end{align}
+\begin{eqnarray}
+k_P^+k_R^-(1-p_C-p_O) - (k_P^- + k_O)(k_R^+ + k_R^-) p_C &=& 0
+\\
+k_O p_C - r p_O &=& 0.
+\end{eqnarray}
 Solving for $p_C = p_O r/k_O$ in the second and substituting into the first
@@
 fold-change is seen to be simply
-\begin{align}
+\begin{eqnarray}
 \text{fold-change}
-&= \frac{\frac{k_P^+ k_O}{r(k_O + k_P^-)}}
+&=& \frac{\frac{k_P^+ k_O}{r(k_O + k_P^-)}}
         {1 + \frac{k_P^+ (k_O + r)}{r(k_O + k_P^-)} + \frac{k_R^+}{k_R^-}}
@@
 \\
-&= \left(
+&=& \left(
         1 + \frac{k_R^+}{k_R^-}
@@
 \right)^{-1},
-\end{align}
+\end{eqnarray}
 surprisingly reducing to the master curve of~\fig{fig1:means_cartoons}(D) once
@@
 The fold-change readily follows,
-\begin{align}
+\begin{eqnarray}
 \text{fold-change}
-&=      \frac{k_R^- k^+}{k_R^- k^+ + k_R^- k^- + k_R^+ k^-}
+&=&      \frac{k_R^- k^+}{k_R^- k^+ + k_R^- k^- + k_R^+ k^-}
         \frac{k_R^- k^+ + k_R^- k^-}{k_R^- k^+}
 \\
-&=      \left(1 + \frac{k_R^+}{k_R^-}
+&=&      \left(1 + \frac{k_R^+}{k_R^-}
                 \left(1 + \frac{k^+}{k^-}\right)^{-1}
         \right)^{-1},
-\end{align}
+\end{eqnarray}
 from which we see $\rho = 1 + k^+/k^-$ as shown in~\fig{fig1:means_cartoons}(C).

### 2020-10-22 doc/resubmission/appendix_02_bursty.tex
final pass before resubmission

@@
 master equation
-\begin{align}
+\begin{equation}
 {d \over d\tau}p(m, \tau) = 
@@
 \label{eq:si_master_unreg}
-\end{align}
+\end{equation}
 
@@
 is defined as
-\begin{align}
+\begin{equation}
 F(z,t) = \sum_{m=0}^\infty z^m p(m,t),
-\end{align}
+\end{equation}
 where $z$ is just a dummy variable that will help us later on to obtain the
@@
 Eq.~\ref{eq:si_generating_01} can be rewritten as
-\begin{align}
-\sum_{m} z^{m} \cdot m \cdot p(m, \tau) &=
+\begin{eqnarray}
+\sum_{m} z^{m} \cdot m \cdot p(m, \tau) &=&
 \sum_{m} z \frac{\partial z^{m}}{\partial z} p(m, \tau), \\
-&=\sum_{m} z \frac{\partial}{\partial z}\left(z^{m} p(m, \tau)\right), \\
-&=z \frac{\partial}{\partial z}\left(\sum_{m} z^{m} p(m, \tau)\right), \\
-&= z {\partial F(z, \tau) \over \partial z}.
-\end{align}
+&=&\sum_{m} z \frac{\partial}{\partial z}\left(z^{m} p(m, \tau)\right), \\
+&=&z \frac{\partial}{\partial z}\left(\sum_{m} z^{m} p(m, \tau)\right), \\
+&=& z {\partial F(z, \tau) \over \partial z}.
+\end{eqnarray}
 For the second term on the right hand side of Eq.~\ref{eq:si_generating_01} we
 define $k \equiv m + 1$. This allows us to write
-\begin{align}
-\sum_{m=0}^{\infty} z^{m} \cdot(m+1) \cdot p(m+1, \tau) &=
+\begin{eqnarray}
+\sum_{m=0}^{\infty} z^{m} \cdot(m+1) \cdot p(m+1, \tau) &=&
 \sum_{k=1}^{\infty} z^{k-1} \cdot k \cdot p(k, \tau), \\
-&=z^{-1} \sum_{k=1}^{\infty} z^{k} \cdot k \cdot p(k, \tau), \\
-&=z^{-1} \sum_{k=0}^{\infty} z^{k} \cdot k \cdot p(k, \tau), \\
-&=z^{-1} \left(z \frac{\partial F(z)}{\partial z}\right), \\
-&=\frac{\partial F(z)}{\partial z}.
-\end{align}
+&=&z^{-1} \sum_{k=1}^{\infty} z^{k} \cdot k \cdot p(k, \tau), \\
+&=&z^{-1} \sum_{k=0}^{\infty} z^{k} \cdot k \cdot p(k, \tau), \\
+&=&z^{-1} \left(z \frac{\partial F(z)}{\partial z}\right), \\
+&=&\frac{\partial F(z)}{\partial z}.
+\end{eqnarray}
 
@@
 results in
-\begin{align}
+\begin{equation}
 \lambda \sum_m z^m \sum_{m'=0}^m G_{m-m'} p(m', \tau) =
@@
 \theta (1 - \theta)^{m-m'} p(m', \tau),
-\end{align}
+\end{equation}
 where we also substituted the definition of the geometric distribution $G_{k} =
 \theta (1 - \theta)^k$. Redistributing the sums we can write
-\begin{align}
+\begin{equation}
 \lambda \sum_{m'=0}^{\infty} \sum_{m=m'}^{\infty} z^m 
@@
 \label{eq:si_generating_02}
-\end{align}
+\end{equation}
 
@@
 Eq.~\ref{eq:si_generating_02} as
-\begin{align}
-\sum_{m=m'}^{\infty} \left[z (1-\theta)\right]^{m} &= 
+\begin{eqnarray}
+\sum_{m=m'}^{\infty} \left[z (1-\theta)\right]^{m} &=&
 \sum_{n=0}^{\infty} \left[z (1-\theta)\right]^{n + m'} \\
-&= \left[ z (1 - \theta) \right]^{m'} 
+&=& \left[ z (1 - \theta) \right]^{m'} 
 \sum_{n=0}^{\infty} \left[ z (1 - \theta) \right]^{n} \\
-&= \left[ z (1 - \theta) \right]^{m'} 
+&=& \left[ z (1 - \theta) \right]^{m'} 
 \left( {1 \over 1 - z (1 - \theta)} \right),
-\end{align}
+\end{eqnarray}
 where we use the geometric series since, as stated before, $| z (1 - \theta) | <
@@
 steady-state, the PDE reduces to the ODE
-\begin{align}
+\begin{equation}
 \deriv[F]{z} = \frac{\xi}{1-z\xi}\lambda F,
-\end{align}
+\end{equation}
 which we can integrate as
-\begin{align}
+\begin{equation}
 \int \frac{dF}{F} = \int \frac{\lambda\xi dz}{1-\xi z}.
-\end{align}
+\end{equation}
 The initial conditions for generating functions can be subtle and confusing. The
@@
 Continuing on, we evaluate the integrals (producing a constant $c$) which gives
-\begin{align}
-\ln F &= -\lambda \ln(1-\xi z) + c
+\begin{eqnarray}
+\ln F &=& -\lambda \ln(1-\xi z) + c
 \\
-F &= \frac{c}{(1-\xi z)^\lambda}.
-\end{align}
+F &=& \frac{c}{(1-\xi z)^\lambda}.
+\end{eqnarray}
 Only one choice for $c$ can satisfy initial conditions, producing
-\begin{align}
+\begin{equation}
 F(z) = \left(\frac{1-\xi}{1-\xi z}\right)^\lambda
@@
 \label{eq:gen_fn}
-\end{align}
+\end{equation}
 
@@
 states, namely
-\begin{align}
-{d\over dt}p_R(m,t) =& 
+\begin{equation}
+{d\over dt}p_R(m,t) =
 k_R^+ p_A(m,t) - k_R^- p_R(m,t)
-        + (m+1)\gamma p_R(m+1,t) - m\gamma p_R(m,t)
-\\
+        + (m+1)\gamma p_R(m+1,t) - m\gamma p_R(m,t),
+\end{equation}
+and
+\begin{equation}
 \begin{split}
-{d\over dt}p_A(m,t) =& - k_R^+ p_A(m,t) + k_R^- p_R(m,t)
+{d\over dt}p_A(m,t) = - k_R^+ p_A(m,t) + k_R^- p_R(m,t)
         + (m+1)\gamma p_A(m+1,t) - m\gamma p_A(m,t) 
 \\
-&- k_i p_A(m,t) + k_i \sum_{m^\prime=0}^m \theta(1-\theta)^{m-m^\prime} p_A(m^\prime,t),
+- k_i p_A(m,t) + k_i \sum_{m^\prime=0}^m \theta(1-\theta)^{m-m^\prime} p_A(m^\prime,t),
 \end{split}
-\end{align}
+\end{equation}
 where $p_R(m,t)$ is the probability of the system having $m$ mRNA copies and
@@
 the time derivatives to zero, giving
-\begin{align}
-0 =& k_R^+ p_A(m) - k_R^- p_R(m) + (m+1) p_R(m+1) - m p_R(m)
+\begin{equation}
+0 = k_R^+ p_A(m) - k_R^- p_R(m) + (m+1) p_R(m+1) - m p_R(m),
+\end{equation}
+and
+\begin{equation}
+\begin{split}
+0 = - k_R^+ p_A(m) + k_R^- p_R(m) + (m+1) p_A(m+1) - m p_A(m) 
 \\
-\begin{split}
-0 =& - k_R^+ p_A(m) + k_R^- p_R(m) + (m+1) p_A(m+1) - m p_A(m) 
-\\
-&- k_i p_A(m) + k_i \sum_{m^\prime=0}^m \theta(1-\theta)^{m-m^\prime} p_A(m^\prime),
+- k_i p_A(m) + k_i \sum_{m^\prime=0}^m \theta(1-\theta)^{m-m^\prime} p_A(m^\prime),
 \end{split}
-\end{align}
+\end{equation}
 where for convenience we kept the same notation for all rates, but these are
@@
 state,
-\begin{align}
+\begin{equation}
 f_A(z) = \sum_{m=0}^\infty z^m p_A(m),
@@
 f_R(z) = \sum_{m=0}^\infty z^m p_R(m).
-\end{align}
+\end{equation}
 Our real objective is the generating function $f(z)$ that generates the mRNA
@@
 case, so the coupled ODEs for the generating functions read
-\begin{align}
-0 =& k_R^+ f_A(z) - k_R^- f_R(z) + \pderiv{z} f_R(z) - z \pderiv{z} f_R(z)
-\\
+\begin{equation}
+0 = k_R^+ f_A(z) - k_R^- f_R(z) + \pderiv{z} f_R(z) - z \pderiv{z} f_R(z),
+\end{equation}
+and
+\begin{equation}
 \begin{split}
@@
 \end{split}
-\end{align}
+\end{equation}
 and after changing variables $\xi = 1 - \theta$ as before and rearranging, we
 have
-\begin{align}
-0 &= k_R^+ f_A(z) - k_R^- f_R(z) + (1-z) \pderiv{z} f_R(z)
+\begin{eqnarray}
+0 &=& k_R^+ f_A(z) - k_R^- f_R(z) + (1-z) \pderiv{z} f_R(z)
 \\
-0 &=  - k_R^+ f_A(z) + k_R^- f_R(z) + (1 - z) \pderiv{z} f_A(z)
+0 &=&  - k_R^+ f_A(z) + k_R^- f_R(z) + (1 - z) \pderiv{z} f_A(z)
 + k_i \frac{(z-1)\xi}{1-z\xi} f_A(z),
-\end{align}
+\end{eqnarray}
 We can transform this problem from two coupled first-order ODEs to a single
@@
 giving
-\begin{align}
+\begin{equation}
 \begin{split}
@@
 \end{split}
-\end{align}
+\end{equation}
 where, to reduce notational clutter, we have dropped the explicit $z$ dependence
 of $f_A$ and $f_R$. Simplifying we have
-\begin{align}
+\begin{equation}
 0 = \psecderiv[f_R]{z}
@@
         + \frac{k_i k_R^- \xi}{(1-z\xi)(1-z)}f_R.
-\end{align}
+\end{equation}
 This can be recognized as the hypergeometric differential equation, with
@@
 First map the $z=1$ singularity to 0 by the change of variables $v=z-1$, giving
-\begin{align}
+\begin{equation}
 0 = \psecderiv[f_R]{v}
@@
         + \frac{k_i k_R^- \xi}{((1+v)\xi - 1)v}f_R.
-\end{align}
+\end{equation}
 Now two singularities are at $v=0$ and $v=\infty$. The third is determined by
@@
 bill. In other words, the combined change of variables
-\begin{align}
+\begin{equation}
 w = \frac{\xi}{1-\xi} (z-1)
-\end{align}
+\end{equation}
 maps $z = \{1, \xi^{-1}, \infty\}$ to $w =\{0, 1, \infty\}$ as desired. Plugging
@@
 $(1 + v)\xi - 1 = (1 - \xi)(w - 1)$ gives
-\begin{align}
+\begin{equation}
 0 = \left(\frac{\xi}{1-\xi}\right)^2 \psecderiv[f_R]{w}
@@
 + \frac{k_i k_R^- \xi^2}{(1-\xi)^2 w(w-1)}f_R.
-\end{align}
+\end{equation}
 This is close to the standard form of the hypergeometric differential equation,
 and some cancellation and rearrangement gives
-\begin{align}
+\begin{equation}
 0 = w(w-1)\psecderiv[f_R]{w}
@@
 + k_i k_R^- f_R.
-\end{align}
+\end{equation}
 and a little more algebra produces
-\begin{align}
+\begin{equation}
 0 = w(1-w)\psecderiv[f_R]{w}
@@
 - k_i k_R^- f_R,
-\end{align}
+\end{equation}
 which is the standard form. From this we can read off the solution in terms of
@@
 linearly independent solutions. More precisely, 
-\begin{align}
+\begin{equation}
 f_R(w) = C^{(1)} {_2F_1}(\alpha, \beta, \delta; w)
 + C^{(2)} w^{1-\delta}{_2F_1}(1+\alpha-\delta, 1+\beta-\delta, 2-\delta; w)
-\end{align}
+\end{equation}
 with parameters determined by
-\begin{align}
+\begin{equation}
 \begin{split}
@@
 \end{split}
-\end{align}
+\end{equation}
 and constants $C^{(1)}$ and $C^{(2)}$ to be set by boundary conditions. Solving for
 $\alpha$ and $\beta$, we find
-\begin{align}
+\begin{equation}
 \begin{split}
@@
 \end{split}
-\end{align}
+\end{equation}
 Note that $\alpha$ and $\beta$ are interchangeable in the definition of
-${_2F_1}$ and differ only in the sign preceeding the radical.
+${_2F_1}$ and differ only in the sign preceding the radical.
 Since the normalization condition requires that $f_R$ be finite at $w=0$,
@@
 We can recover $f_A$ from our solution for $f_R$, namely
-\begin{align}
+\begin{equation}
 f_A(z) = \frac{1}{k_R^+}\left(k_R^- f_R(z) + (z-1) \pderiv[f_R]{z}\right)
-\end{align}
+\end{equation}
 or
-\begin{align}
+\begin{equation}
 f_A(w) = \frac{1}{k_R^+}\left(k_R^- f_R(w) + w \pderiv[f_R]{w}\right),
-\end{align}
+\end{equation}
 where in the second line we transformed our original relation between
@@
 which tells us
-\begin{align}
+\begin{equation}
 \pderiv[f_R]{w} = C^{(1)}\frac{\alpha\beta}{\delta}
                 {_2F_1}(\alpha+1, \beta+1, \delta+1; w),
-\end{align}
+\end{equation}
 from which it follows that
-\begin{align}
+\begin{equation}
 f_A(w) = \frac{C^{(1)}}{k_R^+}
@@
 \right)
-\end{align}
+\end{equation}
 and therefore
-\begin{align}
+\begin{equation}
 f(w) = C^{(1)}\left(1 + \frac{k_R^-}{k_R^+}\right)
@@
         {_2F_1}(\alpha+1, \beta+1, \delta+1; w).
-\end{align}
+\end{equation}
 To proceed, we need one of the (many) useful identities known for
 hypergeometric functions, in particular
-\begin{align}
+\begin{equation}
 w\frac{\alpha\beta}{\delta} {_2F_1}(\alpha+1, \beta+1, \delta+1; w)
@@
 \right).
-\end{align}
+\end{equation}
 Substituting this for the second term in $f(w)$, we find
-\begin{align}
+\begin{equation}
 f(w) = \frac{C^{(1)}}{k_R^+}
@@
 \right],
-\end{align}
+\end{equation}
 and since $\delta-1 = k_R^+ + k_R^-$, the first and third terms cancel,
 leaving only
-\begin{align}
+\begin{equation}
 f(w) = C^{(1)}\frac{k_R^+ + k_R^-}{k_R^+} {_2F_1}(\alpha, \beta, \delta-1; w).
-\end{align}
+\end{equation}
 Now we enforce normalization, demanding $f(w=0) = f(z=1) = 1$.
@@
 $C^{(1)} = k_R^+ / (k_R^+ + k_R^-)$ and consequently
-\begin{align}
+\begin{equation}
 f(w) =  {_2F_1}(\alpha, \beta, k_R^+ + k_R^-; w).
-\end{align}
+\end{equation}
 Recalling that the mean burst size $b = (1-\theta)/\theta = \xi/(1-\xi)$
@@
 we can transform back to the original variable $z$ to find the tidy result
-\begin{align}
+\begin{equation}
 f(z) =  {_2F_1}(\alpha, \beta, k_R^+ + k_R^-; b(z-1)),
-\end{align}
+\end{equation}
 with $\alpha$ and $\beta$ given above by
-\begin{align}
+\begin{equation}
 \begin{split}
@@
 \end{split}
-\end{align}
+\end{equation}
 Finally we are in sight of the original goal. We can generate the steady-state
 probability distribution of interest by differentiating the generating function,
-\begin{align}
+\begin{equation}
 p(m) = m! \left.\frac{\partial^m}{\partial z^m} f(z) \right|_{z=0},
-\end{align}
+\end{equation}
 which follows easily from its definition. Some contemplation reveals that
@@
 precisely, one finds
-\begin{align}
+\begin{equation}
 p(m) = \frac{
@@
 \label{eq:p_m_bursty+rep_appdx}
-\end{align}
+\end{equation}
 which is finally the probability distribution we sought to derive.
@@
 indexed by an integer $n$,
-\begin{align}
+\begin{equation}
 f_n = {_2F_1}(a+\epsilon_i n, b+\epsilon_j n, c+\epsilon_k n; z),
-\end{align}
+\end{equation}
 for a fixed choice of $\epsilon_i, \epsilon_j, \epsilon_k \in \{0,\pm 1\}$
@@
 Then there exist known recurrence relations of the form
-\begin{align}
+\begin{equation}
 A_n f_{n-1} + B_n f_{n} + C_n f_{n+1} = 0,
-\end{align}
+\end{equation}
 where $A_n, B_n$, and $C_n$ are some functions of $a,b,c$, and $z$.
@@
 After some ``guess and check,'' we found that the transformation
-\begin{align}
+\begin{equation}
 {_2F_1}(\alpha+m, \beta+m, \delta+m; -b)
@@
 \label{eq:rec_euler_pretransform}
-\end{align}
+\end{equation}
 produces a ${_2F_1}$ on the right hand side that closely resembles
@@
 Explicitly, these solutions are
-\begin{align}
+\begin{eqnarray}
 y_{3,m}
-&\propto
+&\propto&
 {_2F_1}\left(-\alpha^\prime + \delta^\prime - m,
@@
 y_{4,m}
-&\propto
+&\propto&
 {_2F_1}\left(\alpha^\prime + m,
@@
 \label{eq:minimal_soln_sans_prefac}
-\end{align}
+\end{eqnarray}
 where we have omitted prefactors which are unimportant for now.
@@
 in~\cite{Abramowitz1964}, which is
-\begin{align}
+\begin{equation}
 (\delta^\prime - (\alpha^\prime + m)) f_{m-1}
@@
 \label{eq:chosen_rec_rel}
-\end{align}
+\end{equation}
 Now we must solve for the parameters appearing in the recurrence relation
 in terms of our parameters, namely by setting
-\begin{align}
+\begin{equation}
 \begin{split}
@@
 \end{split}
-\end{align}
+\end{equation}
 and solving to find
-\begin{align}
+\begin{equation}
 \begin{split}
@@
 \end{split}
-\end{align}
+\end{equation}
 Finally we have everything we need. The minimal solution
-\begin{align}
+\begin{equation}
 y_{4,m}
@@
                 1-z\right),
-\end{align}
+\end{equation}
 where we have now included the necessary prefactors,

### 2020-10-22 doc/resubmission/appendix_03_bayesian_inference.tex
final pass before resubmission

@@
 phenomenon in the form of a mathematical model --in the context of Bayesian
-inference this means choosing a form for the likehihoods, and even making
+inference this means choosing a form for the likelihoods, and even making
 this choice presupposes prior understanding or assumptions as to the
@@
 \end{equation}
-where the simbol ``$\sim$'' can be read as \textit{is distributed according to}.
+where the symbol ``$\sim$'' can be read as \textit{is distributed according to}.
 So the previous equation can be read as: the mRNA copy number $m$ is distributed
@@
 repressor was under the control of an inducible promoter regulated by the TetR
-repressor as schematized in Figre~\ref{figS:aTc_circuit}. When TetR binds to the
+repressor as schematized in Figure~\ref{figS:aTc_circuit}. When TetR binds to the
 small molecule anhydrotetracycline (aTc), it shifts to an inactive conformation
@@
 where $\alpha$ and $\beta$, defined for notational convenience, are
-\begin{align}
+\begin{equation}
 \begin{split}
@@
 \includegraphics[width=\textwidth]{../../figures/si/figS0X_histograms.pdf}
-\caption{\textbf{Theory-experiment comparision of mRNA distributions for
+\caption{\textbf{Theory-experiment comparison of mRNA distributions for
 regulated promoters.} Comparison of the inference (color shaded area

### 2020-12-06 doc/resubmission/section_01_introduction.tex
minor fixes to typos
minor changes after 2nd round of reviews. Paper is ready

@@
 effective two-state models for which there is a rich
-literature~\cite{Peccoud1995, Shahrezaei2008, Iyer-Biswas2009, Tkacik2009,
-Sanchez2013, Jones2014, So2011, Munsky2012}. Here, we systematically compare the
-predictions of several models with different levels of coarse graining written
-in terms of thermodynamic and kinetic parameters. One goal in exploring such
-coarse-grainings is to build towards the future models of regulatory response
-that will be able to serve the powerful predictive role needed to take synthetic
-biology from a brilliant exercise in enlightened empiricism to a rational design
-framework as in any other branch of engineering. More precisely, we want
-phenomenology in the sense of coarse-graining away atomistic detail, but still
-retaining biophysical meaning. In particular a key question is: at this level of
-coarse-graining, what microscopic details do we need to explicitly model, and
-how do we figure that out? For example, do we need to worry about all or even
-any of the steps that individual RNA polymerases go through each time they make
-a transcript? Turning the question around, can we see any imprint of those
-processes in the available data? If the answer is no, then those processes are
-irrelevant for our purposes. Forward modeling and inverse (statistical
-inferential) modeling are necessary to tackle such questions. We combine both
-approaches in order to discard models that cannot empirically satisfy the main
-features of experimental data. First we apply forward modeling to demonstrate
-that none of the models are distinguishable at the level of mean gene
-expression. We then extend the modeling to look at higher moments of the
-distribution, eliminating models that do not empirically satisfy the observed
-cell-to-cell variability. Finally we arrive at a minimal model on which we can
-apply inverse modeling in order to infer the parameters that explain the data.
+literature~\cite{Peccoud1995, Raj2006, Shahrezaei2008, Iyer-Biswas2009,
+Tkacik2009, Sanchez2013, Jones2014, So2011, Munsky2012, Kumar2015}. Here, we
+systematically compare the predictions of several models with different levels
+of coarse graining written in terms of thermodynamic and kinetic parameters. One
+goal in exploring such coarse-grainings is to build towards the future models of
+regulatory response that will be able to serve the powerful predictive role
+needed to take synthetic biology from a brilliant exercise in enlightened
+empiricism to a rational design framework as in any other branch of engineering.
+More precisely, we want phenomenology in the sense of coarse-graining away
+atomistic detail, but still retaining biophysical meaning. In particular a key
+question is: at this level of coarse-graining, what microscopic details do we
+need to explicitly model, and how do we figure that out? For example, do we need
+to worry about all or even any of the steps that individual RNA polymerases go
+through each time they make a transcript? Turning the question around, can we
+see any imprint of those processes in the available data? If the answer is no,
+then those processes are irrelevant for our purposes. Forward modeling and
+inverse (statistical inferential) modeling are necessary to tackle such
+questions. We combine both approaches in order to discard models that cannot
+empirically satisfy the main features of experimental data. First we apply
+forward modeling to demonstrate that none of the models are distinguishable at
+the level of mean gene expression. We then extend the modeling to look at higher
+moments of the distribution, eliminating models that do not empirically satisfy
+the observed cell-to-cell variability. Finally we arrive at a minimal model on
+which we can apply inverse modeling in order to infer the parameters that
+explain the data.
 
@@
 $\rho$ which subsumes all regulatory details of the architecture that do not
-directly involve the repressor. (B) Simple repression models from an equilibrium
-perspective. (C) Equivalent models cast in chemical kinetics language. (D) The
-``master curve'' to which all cartoons in (B) and (C) collapse.}
+directly involve the repressor. (B) Simple repression models from a
+thermodynamic perspective. (C) Equivalent models cast in chemical kinetics
+language. (D) The ``master curve'' to which all cartoons in (B) and (C)
+collapse.}
 \label{fig1:means_cartoons}

### 2020-12-06 doc/resubmission/section_02_means.tex
final pass before resubmission
minor changes after 2nd round of reviews. Paper is ready

@@
 gene expression dynamics) is to measure the fold-change in mean
-expression~\cite{Garcia2011}. The fold-change is defined as
+expression~\cite{Garcia2011}. The fold-change $FC$ is defined as
 \begin{equation}
-\text{fold-change}
-= \frac{\langle \text{gene expression with repressor present} \rangle}
-        {\langle \text{gene expression with repressor absent} \rangle}
-= \frac{\langle m (R > 0) \rangle}{\langle m (R = 0) \rangle}
-= \frac{\langle p (R > 0) \rangle}{\langle p (R = 0) \rangle}
+FC(R)
+= \frac{\langle \text{gene expression with }R > 0 \rangle}
+        {\langle \text{gene expression with }R = 0 \rangle}
+= \frac{\langle m (R) \rangle}{\langle m (0) \rangle}
+= \frac{\langle p (R) \rangle}{\langle p (0) \rangle},
 \label{eq:fc_def}
@@
 \begin{equation}
- \text{fold-change} = \left( 1 + \exp(-\Delta F_R + \log(\rho))  \right)^{-1},
+FC(R) = \left( 1 + \exp(-\Delta F_R(R) + \log(\rho))  \right)^{-1},
  \label{eq:deltaFR_eq_noneq_equiv}
@@
 \begin{equation}
-\text{fold-change}
+FC(R)
 = \left(1 + \frac{R}{N_{NS}} e^{-\beta\Delta\varepsilon_R}\right)^{-1},
@@
 \begin{eqnarray}
-\text{fold-change}
+FC(R)
 &=& \left(
@@
 \begin{eqnarray}
- \text{fold-change} &=& \left( 1 + \frac{k_R^+}{k_R^-} \rho \right)^{-1},\\
+FC &=& \left( 1 + \frac{k_R^+}{k_R^-} \rho \right)^{-1},\\
 &=& (1 + \exp(-\Delta F_R + \log(\rho) ))^{-1},
@@
 The key outcome of our analysis of the models in
-Figure~\ref{fig1:means_cartoons} is the existence of a master curve shown
-in Figure~\ref{fig1:means_cartoons}(D) to which the fold-change predictions
-of all the models collapse. This master curve is parametrized by only two
-effective parameters: $\Delta F_R$, which characterizes the number of repressors
-and their binding strength to the DNA, and $\rho$, which characterizes all other
-features of the promoter architecture. The key assumption underpinning this
-result is that no transcription occurs when a repressor is bound to its
-operator. Given this outcome, i.e., the degeneracy of the different models at
-the level of fold-change, a mean-based metric such as the fold-change that can
-be readily measured experimentally is insufficient to discern between these
-different levels of coarse-graining. In the next section we extend the analysis
-of the models to higher moments of the mRNA distribution as we continue to
-examine the discriminatory power of these different models.
-
-% \subsection{Discussion of Results Across Models for Fold-Changes in Mean
-% Expression}
-% The key outcome of our analysis of the models in
-% Figure~\ref{fig1:means_cartoons} is the existence of a master curve shown
-% in Figure~\ref{fig1:means_cartoons}(D) to which the fold-change predictions
-% of all the models collapse. This master curve is parametrized by only two
-% effective parameters: $\Delta F_R$, which characterizes the number of repressors
-% and their binding strength to the DNA, and $\rho$, which characterizes all other
-% features of the promoter architecture. The key assumption underpinning this
-% result is that no transcription occurs when a repressor is bound to its
-% operator. Note, however, that we are agnostic about the molecular mechanism
-% which achieves this; steric effects are one plausible mechanism, but, for
-% instance, ``action-at-a-distance'' mediated by kinked DNA due to repressors
-% bound tens or hundreds of nucleotides upstream of a promoter is plausible as
-% well.
-
-% Why does the master curve of Figure~\ref{fig1:means_cartoons}(D) exist at all?
-% This brings to mind the deep questions posed in,
-% e.g.,~\cite{Frank2013} and~\cite{Frank2014a}, suggesting we consider multiple
-% plausible models of a system and search for their common patterns to tease
-% out which broad features are and are not important.
-% In our case, the key feature seems to be the
-% exclusive nature of repressor and RNAP binding, which allows the parameter
-% describing the repressor, $\Delta F_R$, to cleanly separate from all other
-% details of the promoter architecture, which are encapsulated in $\rho$.
-% Arbitrary nonequilibrium behavior can occur on the rest of the
-% promoter state space, but it may all be swept up in the effective
-% parameter $\rho$, to which the repressor makes no contribution.
-% We point the interested reader to~\cite{Gunawardena2012}
-% and~\cite{Ahsendorf2014} for an interesting analysis of similar problems
-% using a graph-theoretic language.
-
-% As suggested in~\cite{Chure2019}, we believe this master curve should generalize
-% to architectures with multiple repressor binding sites, as long as the
-% exclusivity of transcription factor binding and transcription initiation is
-% maintained. The interpretation of $\Delta F_R$ is then of an effective free
-% energy of all repressor bound states. In an equilibrium picture this is simply
-% given by the log of the sum of Boltzmann weights of all repressor bound states,
-% which looks like the log of a partition function of a subsystem. In a
-% nonequilibrium picture, while we can still mathematically gather terms and give
-% the resulting collection the label $\Delta F_R$, it is unclear if the physical
-% interpretation as an effective free energy makes sense. The problem is that free
-% energies cannot be assigned unambiguously to states out of equilibrium because
-% the free energy change along a generic path traversing the state space is path
-% dependent, unlike at equilibrium. A consequence of this is that, out of
-% equilibrium, $\Delta F_R$ is no longer a simple sum of Boltzmann weights.
-% Instead it resembles a restricted sum of King-Altman diagrams~\cite{King1956,
-% Hill1966}. Following the work of Hill~\cite{Hill1989}, it may yet be possible to
-% interpret this expression as an effective free energy, but this remains unclear
-% to us. We leave this an open problem for future work.
-
-% If we relax the requirement of exclusive repressor-RNAP binding, one could
-% imagine models in which repressor and RNAP doubly-bound states are allowed,
-% where the repressor's effect is to \text{reduce} the transcription rate rather
-% than setting it to zero. Our results do not strictly apply to such a model,
-% although we note that if the repressor's reduction of the transcription rate is
-% substantial, such a model might still be well-approximated by one of the models
-% in~\fig{fig1:means_cartoons}.
-
-% One may worry that our ``one curve to rule them all'' is a mathematical
-% tautology. In fact we \textit{agree} with this criticism if $\Delta F_R$ is
-% ``just a fitting parameter'' and cannot be meaningfully interpreted as a real,
-% physical free energy. An analogy to Hill functions is appropriate here. One of
-% their great strengths and weaknesses, depending on the use they are put to, is
-% that their parameters coarse-grain many details and are generally not
-% interpretable in terms of microscopic models, for deep reasons discussed at
-% length in~\cite{Frank2013}. By contrast, our master curve claims to have the
-% best of both worlds: a coarse-graining of all details besides the repressor into
-% a single effective parameter $\rho$, while simultaneously retaining an
-% interpretation of $\Delta F_R$ as a physically meaningful and interpretable free
-% energy. Our task, then, is to prove or disprove this claim.
-
-% How do we test this and probe the theory with fold-change measurements? There is
-% a fundamental limitation in that the master curve is essentially a one-parameter
-% function of $\Delta F_R + \log\rho$. Worse, there are many \textit{a priori}
-% plausible microscopic mechanisms that could contribute to the value of $\rho$,
-% such as RNAP binding and escape kinetics~\cite{DeHaseth1998, Mitarai2015},
-% and/or supercoiling accumulation and release~\cite{Chong2014, Sevier2016},
-% and/or, RNAP clusters analogous to those observed in
-% eukaryotes~\cite{Cisse2013, Cho2016} and recently also observed in
-% bacteria~\cite{Ladouceur2020}. Even if $\Delta F_R$ is measured to high
-% precision, inferring the potential microscopic contributions to $\rho$, buried
-% inside a log no less, from fold-change measurements seems beyond reach. As a
-% statistical inference problem it is entirely nonidentifiable, in the language
-% of~\cite{Gelman2013}, Section 4.3.
-
-% If we cannot simply infer values of $\rho$ from measurements of fold-change, can
-% we perturb some of the parameters that make up $\rho$ and measure the change?
-% Unfortunately we suspect this is off-limits experimentally: most of the
-% potential contributors to $\rho$ are global processes that affect many or all
-% genes. For instance, changing RNAP association rates by changing RNAP copy
-% numbers, or changing supercoiling kinetics by changing topoisomerase copy
-% numbers, would massively perturb the entire cell's physiology and confound any
-% determination of $\rho$.
-
-% One might instead imagine a bottom-up modeling approach, where we mathematicize
-% a model of what we hypothesize the important steps are and are not, use \textit{in vitro}
-% data for the steps deemed important, and \textit{predict} what $\rho$ should be.
-% But again, because of the one-parameter nature of the master curve, many
-% different models will likely make indistinguishable predictions, and without any
-% way to experimentally perturb \textit{in vivo}, there is no clear way
-% to test whether the modeling assumptions are correct.
-
-% In light of this, we prefer the view that parameters and rates are not directly
-% comparable between cartoons in~\fig{fig1:means_cartoons}. Rather, parameters in
-% the simpler cartoons represent coarse-grained combinations of parameters in the
-% finer-grained models. For instance, by equating $\rho$ between any two models,
-% one can derive various possible correspondences between the two models'
-% parameters. Note that these correspondences are clearly not unique, since many
-% possible associations could be made. It then is a choice as to what microscopic
-% interpretations the model-builder prefers for the parameters in a particular
-% cartoon, and as to which coarse-grainings lend intuition and which seem
-% nonsensical. Indeed, since it remains an open question what microscopic features
-% dominate $\rho$ (as suggested above, perhaps RNAP binding and escape
-% kinetics~\cite{DeHaseth1998, Mitarai2015}, or supercoiling accumulation and
-% release~\cite{Chong2014, Sevier2016}, or, something more exotic like RNAP
-% clusters~\cite{Cisse2013, Cho2016, Ladouceur2020}), we are hesitant to put too
-% much weight on any one microscopic interpretation of model parameters that make
-% up $\rho$.
-
-% One possible tuning knob to probe $\rho$ that would not globally perturb the
-% cell's physiology is to manipulate RNAP binding sites. Work such
-% as~\cite{Kinney2010} has shown that models of sequence-dependent RNAP affinity can be
-% inferred from data, and the authors of~\cite{Brewster2012} showed that the model
-% of~\cite{Kinney2010} has predictive power by using the model to \textit{design}
-% binding sites of a desired affinity. But for our purposes, this begs the
-% question: the authors of~\cite{Kinney2010} \textit{assumed} a particular model
-% (essentially our 3-state equilibrium model but without the repressor), so it is
-% unclear how or if such methods can be turned around to \textit{compare}
-% different models of promoter function.
-
-% Another possible route to dissect transcription details without a global
-% perturbation would be to use phage polymerase with phage-specific promoters.
-% While such results would carry some caveats, e.g., whether the repression of
-% the phage polymerase is a good analog to the repression of the native RNAP,
-% it could nevertheless be worthy of consideration.
-
-% We have already pointed out that the master curve of~\fig{fig1:means_cartoons}
-% is essentially a one-parameter model, the one parameter being $\Delta F_R +
-% \log\rho$. By now the reader may be alarmed as to how can we even determine
-% $\Delta F_R$ and $\rho$ independently of each other, never mind shedding a lens
-% on the internal structure of $\rho$ itself. A hint is provided by the weak
-% promoter approximation, invoked repeatedly in prior studies~\cite{Bintu2005c,
-% Garcia2011a, Razo-Mejia2018} of simple repression using the 3-state equilibrium
-% model in~\fig{fig1:means_cartoons}(B). In that picture, the weak promoter
-% approximation means $\frac{P}{N_{NS}}\exp(-\beta\Delta\varepsilon_P) \ll 1$,
-% meaning therefore $\rho\approx1$.  This approximation can be well justified on
-% the basis of the number of RNAP and $\sigma$ factors per cell
-% and the strength of binding of RNAP to
-% DNA at weak promoters. This is suggestive, but how can we be sure that $\rho$ is
-% not, for instance, actually $10^2$ and that $\Delta F_R$ hides a compensatory
-% factor? A resolution is offered by an independent inference of $\rho$ in the
-% absence of repressors. This was done in~\cite{Razo-Mejia2020} by fitting
-% kinetic model 4 in~\fig{fig1:means_cartoons}(C), with zero repressor
-% (looking ahead, this is equivalent to model 4
-% in Figure~\ref{fig2:constit_cartoons}(A)),
-% to single-cell mRNA counts data from~\cite{Brewster2014}. This provided a
-% determination of $k^+$ and $k^-$, from which their ratio is estimated to be no
-% more than a few $10^{-1}$ and possibly as small as $10^{-2}$.
-
-% The realization that $\rho\approx1$ to an excellent approximation,
-% \textit{independent} of which model in~\fig{fig1:means_cartoons} one prefers,
-% goes a long way towards explaining the surprising success of equilibrium models
-% of simple repression. Even though our 2- and 3-state models get so many details
-% of transcription wrong, it does not matter because fold-change is a cleverly
-% designed ratio. Since $\rho$ subsumes all details except the repressor, and
-% $\log\rho\approx0$, fitting these simple models to fold-change measurements can
-% still give a surprisingly good estimate of repressor binding energies. So the
-% ratiometric construction of fold-change fulfills its intended purpose of
-% canceling out all features of the promoter architecture except the repressor
-% itself. Nevertheless it is perhaps surprising how effectively it does so:
-% \textit{a priori}, one might not have expected $\rho$ to be quite so close to 1.
-
-% We would also like to highlight the relevance of~\cite{Landman2019} here.
-% Landman et.\ al.\ reanalyzed and compared \textit{in vivo} and
-% \textit{in vitro} data on the lacI repressor's binding affinity to its
-% various operator sequences. (The \textit{in vivo}
-% data was from, essentially, fitting our master curve to expression
-% measurements.) They find broad agreement between the
-% \textit{in vitro} and \textit{in vivo}
-% values. This reinforces the suspicion that the equilibrium $\Delta\varepsilon_R$
-% repressor binding energies do in fact represent real physical free energies.
-% Again, \textit{a priori} this did not have to be the case, even knowing that
-% $\rho\approx1$.
-
-% In principle, if $\Delta F_R$ can be measured to sufficient precision, then
-% deviations from $\rho=1$ become a testable matter of experiment. In practice, it
-% is probably unrealistic to measure repressor rates $k_R^+$ or $k_R^-$ or
-% fold-changes in expression levels (and hence $\Delta\varepsilon_R$) precisely
-% enough to detect the expected tiny deviations from $\rho=1$. We can estimate the
-% requisite precision in $\Delta F_R$ to resolve a given $\Delta\rho$ by noting,
-% since $\rho\approx1$, that $\log(1+\Delta\rho)\approx \Delta\rho$, so
-% $\Delta(\Delta F_R) \approx \Delta\rho$. Suppose we are lucky and $\Delta\rho$
-% is $\sim0.1$, on the high end of our range estimated above. A
-% determination of $\Delta\varepsilon_R/k_BT$ with an uncertainty of barely 0.1
-% was achieved in the excellent measurements of~\cite{Razo-Mejia2018}, so this
-% requires a very difficult determination of $\Delta F_R$ for a very crude
-% determination of $\rho$, which suggests, to put it lightly, this is not a
-% promising path to pursue experimentally. It is doubtful that inference of
-% repressor kinetic rates would be any easier.%\footnote{\mmnote{In fact the
-% % uncertainties quoted in Hammar et.\ al.~\cite{Hammar2014} from Elf's group are
-% % small enough to do this calculation, but HJ \& I think their error bars are
-% % insanely optimistic. Not sure I want to blatantly call them out on it though??}}
-
-% Moving forward, we have weak evidence supporting the interpretation of $\Delta
-% F_R$ as a physically real free energy~\cite{Landman2019} and other work casting
-% doubt~\cite{Hammar2014}. How might we resolve the confusion? If there is no
-% discriminatory power to test the theory and distinguish the various models with
-% measurements of fold-changes in means, how do we probe the theory? Clearly to
-% discriminate between the nonequilibrium models in~\fig{fig1:means_cartoons}, we
-% need to go beyond means to ask questions about kinetics, noise and even full
-% distributions of mRNA copy numbers over a population of cells. If the
-% ``one-curve-to-rule-them-all'' is more than a mathematical tautology, then the
-% free energy of repressor binding inferred from fold-change measurements should
-% agree with repressor binding and unbinding rates. In other words, the
-% equilibrium and nonequilibrium definitions of $\Delta F_R$ must agree, meaning
-% \begin{equation}
-% \Delta F_R = \beta\Delta\varepsilon_R - \log(R/N_{NS})
-%         = - \log(k_R^+/k_R^-),
-% \label{eq:deltaFR_eq_noneq_equiv}
-% \end{equation}
-% must hold, where $\beta\Delta\varepsilon_R$ is inferred from the master curve
-% fit to fold-change measurements, and $k_R^+$ and $k_R^-$ are inferred in some
-% orthogonal manner. Single molecule measurements such as from~\cite{Hammar2014}
-% have directly observed these rates, and in the remainder of this work we explore
-% a complementary approach: inferring repressor binding and unbinding rates
-% $k_R^+$ and $k_R^-$ from single-molecule measurements of mRNA population
-% distributions.
+Figure~\ref{fig1:means_cartoons} is the existence of a master curve shown in
+Figure~\ref{fig1:means_cartoons}(D) to which the fold-change predictions of all
+the models collapse. This master curve is parametrized by only two effective
+parameters: $\Delta F_R$, which characterizes the number of repressors and their
+binding strength to the DNA, and $\rho$, which characterizes all other features
+of the promoter architecture. The key assumption underpinning this result is
+that no transcription occurs when a repressor is bound to its operator. Given
+this outcome, i.e., the degeneracy of the different models at the level of
+fold-change, a mean-based metric such as the fold-change that can be readily
+measured experimentally is insufficient to discern between these different
+levels of coarse-graining. The natural extension that the field has followed for
+the most part is to explore higher moments of the gene expression distribution
+in order to establish if those contain the key insights into the mechanistic
+nature of the gene transcription process~\cite{Iyer-Biswas2009,Munsky2012}.
+Following a similar trend, in the next section we extend the analysis of the
+models to higher moments of the mRNA distribution as we continue to examine the
+discriminatory power of these different models.

### 2020-12-06 doc/resubmission/section_03_beyond_means.tex
remove comment in latex file
final pass before resubmission
minor changes after 2nd round of reviews. Paper is ready

@@
 can be viewed as a certain limit of model (4) in which transcripts are produced
-in bursts, and initiation of bursts is a Poisson process. 
-% \mmnote{Haven't found model 5 formulated quite like this in the literature,
-% usually burstiness is handled like model 4, but I should probably search a bit
-% more.} 
-The right column shows the Fano factor $\nu$ (variance/mean) for each model.
-Note especially the crucial diagnostic: (2) and (3) have $\nu$ strictly below 1,
-while only for (4) and (5) can $\nu$ exceed 1. Models with Fano factors $\le 1$
-cannot produce the single-cell data observed in part (B) without introducing
-additional assumptions and model complexity. (B) Data from \cite{Jones2014}.
-Mean mRNA count vs. Fano factor (variance/mean) for different promoters as
-determined with single-molecule mRNA Fluorescence \textit{in situ}
-Hybridization. The colorbar indicates the predicted binding affinity of RNAP to
-the promoter sequence as determined in \cite{Brewster2012}. Numbers serve for
-cross comparison with data presented in Figure 3.}
+in bursts, and initiation of bursts is a Poisson process. The right column shows
+the Fano factor $\nu$ (variance/mean) for each model. Note especially the
+crucial diagnostic: (2) and (3) have $\nu$ strictly below 1, while only for (4)
+and (5) can $\nu$ exceed 1. Models with Fano factors $\le 1$ cannot produce the
+single-cell data observed in part (B) without introducing additional assumptions
+and model complexity. (B) Data from \cite{Jones2014}. Mean mRNA count vs. Fano
+factor (variance/mean) for different promoters as determined with
+single-molecule mRNA Fluorescence \textit{in situ} Hybridization. The colorbar
+indicates the predicted binding affinity of RNAP to the promoter sequence as
+determined in \cite{Brewster2012}. Numbers serve for cross comparison with data
+presented in Figure 3.}
 \label{fig2:constit_cartoons}
@@
 peaked with a smaller fractional width than a single exponential with the same
-mean. A less disperse waiting time distribution means transcription initations
+mean. A less disperse waiting time distribution means transcription initiations
 are more uniformly distributed in time relative to a Poisson process. Hence the
@@
 many binding sites over the genome, which should serve to pull them out of
-circulation and keep their effective association rates from rising too large.
-Therefore, there is \textit{perhaps} a timescale separation possible between
-transcription factor association rates and burst durations, but this assumption
-could very well break down, so we will have to keep it in mind when we infer
-repressor rates from the Jones et.\ al.\ single-cell mRNA counts data
-later~\cite{Jones2014}.
+circulation and keep their effective association rates from rising too
+large~\cite{Schmidt2016}. Therefore, there is \textit{perhaps} a timescale
+separation possible between transcription factor association rates and burst
+durations, but this assumption could very well break down, so we will have to
+keep it in mind when we infer repressor rates from the Jones et.\ al.\
+single-cell mRNA counts data later~\cite{Jones2014}.
 

### 2020-12-06 doc/resubmission/section_04_bayesian_inference.tex
minor changes after 2nd round of reviews. Paper is ready

@@
 Appendix~\ref{sec:gen_fcn_appdx}, the steady-state mRNA distribution in this
-model is a negative binomial distribution, given by
+model is a negative binomial distribution~\cite{Kumar2015}, given by
 \begin{equation}
@@
 
-On the whole then, we regard this as striking confirmation of the validity of
-the thermodynamic models. Their lynchpin parameter is a phenomenological free
-energy of repressor binding that has previously only been inferred indirectly.
-Our result shows that the microscopic interpretation of this free energy, as the
-log of a ratio of transition rates, does indeed hold true to within the inherent
-uncertainties that remain in the entire theory-experiment dialogue.
+On the whole then, we find these to be results very suggestive. Although not
+conclusively, our results point at the possible interpretation of the
+phenomenological free energy of transcription factor binding as the log of a
+ratio of transition rates. This did not need to be the case; the free energy was
+inferred from bulk measurements of mean gene expression relative to an
+unregulated strain --the fold-change-- while the rates were extracted from
+fitting single-molecule counts to the full mRNA steady state distribution.
+Further single molecule experiments along the lines of~\cite{Hammar2014} could
+shed further light into this idea.
+% On the whole then, we regard this as striking confirmation of the validity of
+% the thermodynamic models. Their lynchpin parameter is a phenomenological free
+% energy of repressor binding that has previously only been inferred indirectly.
+% Our result shows that the microscopic interpretation of this free energy, as the
+% log of a ratio of transition rates, does indeed hold true to within the inherent
+% uncertainties that remain in the entire theory-experiment dialogue.
 

### 2020-12-06 doc/resubmission/section_05_discussion.tex
final pass before resubmission
minor changes after 2nd round of reviews. Paper is ready

@@
 Hybridization experiment, for example.
+We do so by contrasting two theoretical frameworks to think about the problem.
+On the one hand, thermodynamic models are convenient ways to describe the
+problem of gene regulation because of the relatively small number of parameters
+compared with kinetic models, and the agnostic nature with respect to the full
+reaction topology that leads to a transcription event. In addition, either
+implicitly or explicitly, many approaches to gene regulation focus strictly on
+occupancy and ideas such as $K_d$s, and thus is of great interest to test these
+models. This comes with the limitation that thermodynamic models are only able
+to make statements about mean gene expression of a quasi-steady state process.
+On the other hand, kinetic models break those limitations, being able to make
+explicit predictions about higher moments of the gene expression distribution,
+and the transient states towards steady state, at the cost of propagating the
+number of parameters, and demanding the full set of reactions
 
@@
 a single rate process with no corresponding return pathway at odds with
-thermodynamics. These irreversible reactions lead to physical paradoxes such as
-infinite entropy production rates. Such models are mathematically convenient and
-serve only as an approximation to the more realistic situation in which the
-forward rates are much larger than the backward rates. One consequence is that
-the inferred rates for the transcription burst rate and burst size  can only be
-thought of as effective rate constants. Another idealization of the models
-considered here is that when computing the fold-change in gene expression
-(Section 2) we assume that the translation efficiency of an mRNA transcript is
-the same in cells with and without regulation. There might be some known
-biological effects that could limit the applicability of this assumption such as
-cooperation between the transcription and translation
-machinery~\cite{Proshkin2010}, or coupling of mRNA transcription and
-degradation~\cite{Chen2015} among other potential effects. Another implicit
-assumption in all of the models discussed here is that the effects due to
-changes in gene copy number during the cell cycle have no effects on the
-expression profile. For moderate growth rates such as those used
+thermodynamics. Such models are mathematically convenient and serve only as an
+approximation to the more realistic situation in which the forward rates are
+much larger than the backward rates. One consequence is that the inferred rates
+for the transcription burst rate and burst size  can only be thought of as
+effective rate constants. Another idealization of the models considered here is
+that when computing the fold-change in gene expression (Section 2) we assume
+that the translation efficiency of an mRNA transcript is the same in cells with
+and without regulation. There might be some known biological effects that could
+limit the applicability of this assumption such as cooperation between the
+transcription and translation machinery~\cite{Proshkin2010}, or coupling of mRNA
+transcription and degradation~\cite{Chen2015} among other potential effects.
+Another implicit assumption in all of the models discussed here is that the
+effects due to changes in gene copy number during the cell cycle have no effects
+on the expression profile. For moderate growth rates such as those used
 in~\cite{Jones2014} from which we obtained the data, at all points during the
